### 2008-02-05 trunk/Papers/dl-gre-08/bisim.tex
added structure and some references

\section{Computing bisimulation classes} \label{sec:bisim}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "dl-gre-08"
%%% End:

### 2008-02-05 trunk/Papers/dl-gre-08/introduction.tex
added structure and some references

\section{Introduction} \label{sec:introduction}

\cite{Dale1995}
\cite{dale91:_gener_refer_expres_invol_relat}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "dl-gre-08"
%%% End:

### 2008-02-13 trunk/Papers/dl-gre-08/related.tex
added structure and some references
added some more references

--- 2008-02-05 trunk/Papers/dl-gre-08/introduction.tex
@@
-\section{Introduction} \label{sec:introduction}
+\section{Related Work} \label{sec:related}
 
 \cite{Dale1995}
-\cite{dale91:_gener_refer_expres_invol_relat}
+
+\cite{deemter01:_gener_refer_expres}
+
+\cite{dale91:_gener_refer_expres_invol_relat} -- avoid infinite
+regress by requiring that each semantic atom can be used only once in
+the description, i.e.\ no \emph{edge} may be used twice.
+
+\cite{gardent02:_gener_minim_defin_descr}
+
+\cite{Krahmer2003} -- avoid infinite regress automatically because
+it doesn't count as progress to add an edge to the subgraph that was
+already in it, i.e.\ cleaner version of Dale and Haddock heuristic.
+Subgraphs can be seen as DL concepts made up of conjunction and
+existential quantifiers (no negation).
+
+\cite{kelleher06:_increm_gener_of_spatial_refer} -- generate GREs with
+spatial relations. Polynomial algorithm (check this). Generates REs
+for landmarks recursively, i.e.\ suffer from Dale and Haddock infinite
+regress in principle. Avoid this by using each individual only once in
+the entire description, i.e.\ no \emph{node} may be used
+twice. Advantage over ours: take salience of landmarks into account.
 

### 2008-03-10 trunk/Papers/dl-gre-08/discussion.tex
added structure and some references
some more references
added some more references
update on evaluation section

--- 2008-02-05 trunk/Papers/dl-gre-08/bisim.tex
@@
-\section{Computing bisimulation classes} \label{sec:bisim}
+\section{Discussion} \label{sec:discussion}
+
+%evaluate on TUNA \cite{deemter06:_build_seman_trans_corpus_for} or a
+%richer version of it
+
+%or on the cabinets corpus
+%\cite{viethen06:_algor_for_gener_refer_expres}
+
+We will now describe two experiments evaluating different aspects of
+the performance of our algorithm. First, we will look at its running
+time on randomly generated domain graphs of different sizes. Second,
+we will compare the output of our algorithm to human-generated
+descriptions.
+
+\subsection{Efficiency}
+
+\subsection{Quality of Output}
+
+Now, we will compare the descriptions generated by our algorithm to
+those humans produce. For this purpose, we use a corpus of
+human-generated referring expressions collected and made available by
+Jette Viethen and Robert
+Dale\footnote{http://www.ics.mq.edu.au/~jviethen/drawers}.  The data
+was collected in an experiment where participants were asked to
+describe one of 16 filing cabinet drawers. The drawers had different
+colors and were arranged in a four-by-four grid, as shown in Figure
+\ref{fig:drawers}. The human-generated descriptions use four
+non-relational (the drawer's \textsf{color}, its \textsf{column} and
+\textsf{row} number, and whether it is in a \textsf{corner}) and five
+relational properties (\textsf{above, below, next to, left of, right
+of}). Of the 118 referring expressions, only 15 use relational properties.
+\cite{viethen06:_algor_for_gener_refer_expres}
+describe the data in more detail and present results of evaluating the
+Full Brevity algorithm by \cite{}, the Incremental Algorithm by
+\cite{} and the relational algorithm by \cite{} on this corpus.
+
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{drawers.eps}
+\caption{A schematic view of the filing cabinets.\todo{needs nicer pic}}\label{fig:drawers}
+\end{figure}
+
+As described above \todo{needs to be described}, Dale and Reiter's
+Incremental Algorithm is a special case of our algorithm, where no
+roles are used and a specific order is imposed on the predicates. We
+replicated Viethen and Dale's experiment, which tested the Incremental
+Algorithm on the drawer domain with all possible preference ordering
+given the four predicates
+\textsf{color, column, row}, and \textsf{corner}
+\cite{viethen06:_algor_for_gener_refer_expres}.  Unsurprisingly, we
+get the same results as they did: 98 of the 103 non-relational
+descriptions produced by humans can be generated using one of the 24
+possible preference orderings.
+
+\newcite{viethen06:_algor_for_gener_refer_expres} also tested Dale and
+Haddock's \shortcite{dale91:_gener_refer_expres_invol_relat}
+Relational Algorithm on the drawer domain. Surprisingly, it could not
+generate \textit{any} of the descriptions produced by humans. To
+evaluate our algorithm's use of relations, we tested it on a
+representation of the domain using the predicates \textsf{color} and
+\textsf{corner} and the roles \textsf{above, below, next to, left of},
+and \textsf{right of}. We did not include the \textsf{column} and
+\textsf{row} predicates because every drawer can be uniquely
+identified through those predicates. Since our algorithm has to
+consider predicates before relations, this would mean that no relation
+ever gets used. Like for the Incremental Algorithm, we tested all
+possible orderings of predicates and roles (with the restriction that
+all predicates come before all roles). Our algorithm generates 10 of
+the 15 relational descriptions. Of the five that it could not
+generate, two mentioned column information, which we excluded from our
+domain, and three (which interestingly were all produced by the same
+person) are of the format \textit{the blue/pink/yellow/orange drawer
+above/below/next to/right of/left of the two blue/pink/yellow/orange
+drawers}. This would correspond to the following output from our
+algorithm: $C_1 \& \exists R . (C_2 \& \exists R . C_2)$ (where $C_i$
+are color predicates and $R$ a role). While our algorithm does not
+generate descriptions that contain the same relation on two levels of
+embedding for these drawers, it does generate some other, simpler
+descriptions for them; for example, $\textsf{orange} \& \exists
+\textsf{left\_of} . (\textsf{pink}
+\& \textsf{corner})$ which could be rendered as
+\textit{the orange drawer left of the pink one in the corner}, or
+$\textsf{orange} \& \exists \textsf{below} . \textsf{yellow}$,
+which could be rendered as \textit{the orange drawer below the yellow
+drawer}.
 

### 2008-03-11 trunk/Papers/dl-gre-08/abstract.tex
wrote first versions of intro, related work, efficiency

\begin{abstract}
  In this paper, we propose to reinterpret the problem of generating
  referring expressions as the problem of computing a concept in a
  description logic that is only satisfied by the referent.  This view
  offers a new unifying perspective under which existing GRE
  algorithms can be compared.  We also show that by applying existing
  algorithms for computing bisimulation classes in description logic,
  we can obtain extremely efficient algorithms for relational
  referring expressions without any danger of running into infinite
  regress.
\end{abstract}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "dl-gre-08"
%%% End:

### 2008-03-11 trunk/Papers/dl-gre-08/discussion.tex
wrote first versions of intro, related work, efficiency

@@
 descriptions.
-
-\subsection{Efficiency}
 
@@
 \begin{figure}
-\includegraphics[width=0.4\textwidth]{drawers.eps}
+\includegraphics[width=0.4\textwidth]{drawers}
 \caption{A schematic view of the filing cabinets.\todo{needs nicer pic}}\label{fig:drawers}
@@
 
+\todo{should we show some more examples? - ak}
+
+\subsection{Efficiency}
+
+Both the \el\ and the \alc\ algorithms took about 15 milliseconds to
+compute distinguishing concepts for all 16 individuals in the Viethen
+\& Dale dataset.\footnote{We measured all runtimes on a MacBook Pro
+  (Intel Core 2 Duo, 2.16 GHz) running Java 1.6 beta, and allowed the
+  Java VM to warm up, i.e.\ just-in-time compile all bytecode, before
+  taking the measurements.}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\columnwidth]{runtimes}
+  \caption{Average runtimes (in ms) of the two algorithms on random graphs of
+    different sizes.} 
+  \label{fig:runtimes}
+\end{figure}
+
+In order to get a more comprehensive picture of our system's
+efficiency, we ran the two algorithms on random graphs with increasing
+numbers of nodes.  \todo{harmonize this with what Carlos calls them in
+  Section~\ref{sec:bisim}} Each graph had ten different atoms and four
+different roles, a 10\% chance for a node to have each atom, and a
+10\% chance for each pair of nodes to be connected by each role.  The
+results (averaged over 10 runs for each graph size) are shown in
+Fig.~\ref{fig:runtimes}.  As the figure shows, the \el\ algorithm
+takes about 400 ms on average to generate relational REs for all
+individuals in the graph with 100 nodes, i.e.\ 4 ms on average for
+each individual.  The \alc\ algorithm is slower by a constant factor
+of about 2-3 which probably reflects implementation differences.  In
+both cases, we think we can conclude that the systems are fast enough
+for practical use.
+
+\subsection{Interface to realization}
+
+\begin{itemize}
+\item there's a risk to generate a concept that can't be realized,
+  especially for the \alc\ algorithm
+\item everybody except perhaps SPUD has this problem
+\item problem is worse in our case because it is harder to control the
+  order in which atoms and relations are explored
+\item would be interesting for future work to think about how
+  search in the bisim algorithms can be controlled by the realizer
+\end{itemize}
+
 %%% Local Variables: 

### 2008-03-11 trunk/Papers/dl-gre-08/introduction.tex
added some more references
wrote first versions of intro, related work, efficiency

@@
 
-\cite{Dale1995}
-\cite{dale91:_gener_refer_expres_invol_relat}
+The generation of referring expressions is one of the most active and
+successful research areas in NLG.  Starting with Dale and Reiter's
+incremental algorithm \cite{Dale1995}, various researchers have added
+extensions such as reference to sets \cite{Stone2000}, more expressive
+logical connectives \cite{deemter01:_gener_refer_expres}, and
+relational expressions \cite{dale91:_gener_refer_expres_invol_relat}.
+
+Referring expressions involving relations, in particular, have
+received increasing attention recently, particularly in the context of
+spatial referring expressions in situated generation (e.g.\
+\cite{kelleher06:_increm_gener_of_spatial_refer}), where it seems
+particularly natural to use expressions such as ``the book on the
+table''.  However, the classical algorithm by
+\newcite{dale91:_gener_refer_expres_invol_relat} was recently shown to
+be unable to generate satisfying REs in practice
+\cite{viethen06:_algor_for_gener_refer_expres}.  Furthermore, the Dale
+\& Haddock algorithm and most of its successors
+\cite{Krahmer2003,kelleher06:_increm_gener_of_spatial_refer} are
+vulnerable to the problem of ``infinite regress'', where the algorithm
+jumps back and forth between generating descriptions for two related
+individuals infinitely, as in ``the book on the table which supports a
+book on the table \ldots''.
+
+In this paper, we propose a fundamentally new way of looking at the
+problem of generating referring expressions: We take a referring
+expression describing an individual $a$ to be a concept of some
+description logic (DL) which is satisfied by $a$ and no other
+individual in the universe.  This allows us to organize existing
+approaches to GRE in an expressiveness hierarchy.  For instance, the
+classical Dale \& Reiter algorithms compute concepts in a description
+logic that consists only of conjunctions of atoms, whereas Dale \&
+Haddock and other relational algorithms compute concepts in the logic
+\el, which permits conjunction and existential quantification.
+
+Furthermore, we show how to apply algorithms for computing the
+\emph{bisimulation classes} of a DL model to GRE.  Roughly speaking,
+bisimulation classes are sets of individuals that satisfy exactly the
+same concepts; if an individual is alone in its bisimulation class,
+then the concept corresponding to this class represents a uniquely
+referring expression.  We present bisimulation classes algorithms for
+the description logics \el\ (only conjunction and existential
+quantification) and \alc\ (full modal logic).  By essentially
+computing referring expressions for all individuals in the domain at
+once, these algorithms systematically avoid the infinite regress
+problem.  The \el\ algorithm, which generates the same class of
+descriptions as Dale \& Haddock, is capable of generating 67\% of the
+relational REs in the
+\newcite{viethen06:_algor_for_gener_refer_expres} dataset correctly,
+and takes about 400 ms to compute relational REs for all 100
+individuals in a random scene.
+
+The paper is structured as follows.  In Section~\ref{sec:bisim}, we
+will first define description logic and fundamental concepts such as
+bisimulation.  We will then show how to compute bisimulation classes
+in Section~\ref{sec:gre} and apply these algorithms to GRE.  In
+Section~\ref{sec:discussion}, we evaluate our algorithms and discuss
+these results.  Section~\ref{sec:related} compares our approach to
+related research; in particular, it shows how various prominent GRE
+algorithms can be seen as special cases of ours.
+Section~\ref{sec:conclusion} concludes and points to future work.
 

### 2008-03-11 trunk/Papers/dl-gre-08/related.tex
wrote first versions of intro, related work, efficiency

--- 2008-02-05 trunk/Papers/dl-gre-08/introduction.tex
@@
-\section{Introduction} \label{sec:introduction}
+\section{Related Work} \label{sec:related}
 
-\cite{Dale1995}
-\cite{dale91:_gener_refer_expres_invol_relat}
+\begin{figure}
+  \centering
+  \begin{tabular}{l|l}
+    GRE algorithm & DL variant \\ \hline
+    \newcite{Dale1995} & conjunctions of atoms \\
+    \newcite{deemter01:_gener_refer_expres} & propositional logic \\
+    \newcite{dale91:_gener_refer_expres_invol_relat} & \el \\
+    \newcite{Krahmer2003} & \el \\
+    \newcite{kelleher06:_increm_gener_of_spatial_refer} & \el \\
+    \newcite{gardent02:_gener_minim_defin_descr} & \alc
+  \end{tabular}
+  \caption{DL variants used by different GRE algorithms.}
+  \label{fig:related}
+\end{figure}
+
+The view of GRE as a problem of computing DL concepts allows us to
+organize existing approaches to GRE with respect to the logical
+connectives of DL which they can use in the referring expressions they
+generate.  This is summarized for some approaches in
+Fig.~\ref{fig:related}.  At the most unexpressive end of the spectrum,
+\newcite{Dale1995} only generate referring expressions that correspond
+to conjunctions of atomic concepts.
+\newcite{deemter01:_gener_refer_expres} adds the other propositional
+connectives (negation and disjunction) to this, whereas
+\newcite{dale91:_gener_refer_expres_invol_relat} and its successors
+\cite{Krahmer2003,kelleher06:_increm_gener_of_spatial_refer} add
+existential quantifiers.  \newcite{gardent02:_gener_minim_defin_descr}
+combines these strands into an algorithm which computes arbitrary
+concepts in \alc. \todo{Is this really what Claire does?}
+
+In this paper, we have presented two algorithms for concepts in \el\
+and \alc.  These algorithms compute referring expressions for all
+individuals in the domain at once, and thus there is no danger of
+infinite regress.  Other approaches for \el\ and higher logics
+typically have to invest some effort into avoiding infinite regress --
+e.g.\ by requiring that no edge may be used twice
+\cite{dale91:_gener_refer_expres_invol_relat} or that no node may be
+used twice \cite{kelleher06:_increm_gener_of_spatial_refer}.  The
+graph-based algorithm by \newcite{Krahmer2003} avoids infinite regress
+with similar ease as our algorithms -- adding an edge to a subgraph
+twice doesn't change the subgraph --, but this comes at the price of
+NP-complete complexity.
+
+Because they compute referring expressions for all individuals in the
+domain together, our algorithms will be strongest in static settings,
+such as the generation of descriptions for museum exhibits, in which
+the individuals and their properties don't change much.  Nevertheless,
+they remain fast enough for real-time use in dynamic settings.
+Another advantage that recent algorithms for relational REs
+\cite{Krahmer2003,kelleher06:_increm_gener_of_spatial_refer} have over
+ours is that they can take weights and salience into account.  How
+this could be incorporated into our algorithms is an interesting
+question for future research.
 

### 2008-03-11 trunk/Papers/dl-gre-08/discussion.tex
Whee! Even faster!

@@
 Fig.~\ref{fig:runtimes}.  As the figure shows, the \el\ algorithm
-takes about 400 ms on average to generate relational REs for all
+takes about 350 ms on average to generate relational REs for all
 individuals in the graph with 100 nodes, i.e.\ 4 ms on average for
-each individual.  The \alc\ algorithm is slower by a constant factor
-of about 2-3 which probably reflects implementation differences.  In
-both cases, we think we can conclude that the systems are fast enough
-for practical use.
+each individual.  The \alc\ algorithm is even faster, at about 200 ms
+for the graph with 100 nodes.  In both cases, we conclude that the
+systems are fast enough for practical use.
 

### 2008-03-13 trunk/Papers/dl-gre-08/bisim.tex
some more references
Carlos: First ideas on the intro to bisim.tex.  Added some packages in dl-gre-08.

@@
-\section{Computing bisimulation classes} \label{sec:bisim}
+\newcommand{\gM}{\mathcal{M}}
+\newcommand{\gL}{\mathcal{L}}
+
+\section{Description Languages and Bisimulation} \label{sec:bisim}
+
+In computational linguistics, the genereting referring expression problem (the GRE
+problem) can be intuitively presented as follows: given some information about
+a group of objects (e.g., grounded facts about individuals in an ongoing dialogue),
+and a grammar, found a grammatically correct expression (usually a noun phrase) that uniquely identify a given object.
+
+By taking a logic perspective, we can recast the GRE problem as an inference problem.
+We can think that the information about the group of objects is provided in the form
+of a model $\gM$ of a given logical language $\gL$.  Now, the inference problem
+associated with the GRE problem will be to find a formula in $\gL$ such that it
+exacly identify a given object $o$.  Formally, if $|\cdot|^\gM$ is the interpretation
+function for $\gL$ which for each formula in $\gL$ and for each model $\gM$ returns
+the set of elements in $\gM$ for which the formula $\varphi$ holds, then we can define
+\medskip
+
+\noindent
+{\small
+\begin{tabular}{l} \hline
+\textsc{$\gL$-GRE Problem}\\ \hline
+\ \ \ Input: A model $\gM$, an object $o$ in the domain of $\gM$\\
+\ \ \ Output: a formula $\varphi \in \gL$ such that $|\varphi|^\gM = \{o\}$\\
+\hspace*{1.4cm} (if such a formula exists).\\ \hline
+\end{tabular}}
+\medskip
+
+As it can be seen from the definition, the problem depends on which particular
+logical language $\gL$ we are interested in, and also (as the original GRE problem)
+it might terminate with failure.  Depending on the model $\gM$, the chosen object $o$
+and the expressive power of $\gL$ it can be the case that there is no formula of
+$\gL$ that uniquely identifies $o$.
+
+Consider the following example.  We have three objects related as indicated in
+Figure~\ref{fig-car1}
+\begin{figure}
+\begin{center}
+\includegraphics{pic.pdf}
+\end{center}
+\caption{Distinguishable/indistinguishable points.}\label{fig-car1}
+\end{figure}
+
+The points $o$ and $o'$ can easily be distinguished in first-order logic:
+the formula $R(x,x)$ is true of $o$ and false in $o'$. The corresponding
+referring expression for $o$ would be ``a point relating to itself''.
+But if we can only use propositional and relational information (equivalently, our grammar does not contain the word ``itself'') then $o$ and $o'$ are indistinguishable. Both
+are ``points which are related to points which are related to points\ldots''
+
+\cite{dovier04:_effic_algor_for_comput_bisim_equiv}
 

### 2008-03-13 trunk/Papers/dl-gre-08/bisim.tex
Added definitions of alc and el. Working on bisimulations.

@@
 \ \ \ Input: A model $\gM$, an object $o$ in the domain of $\gM$\\
-\ \ \ Output: a formula $\varphi \in \gL$ such that $|\varphi|^\gM = \{o\}$\\
+\ \ \ Output: A formula $\varphi \in \gL$ such that $|\varphi|^\gM = \{o\}$\\
 \hspace*{1.4cm} (if such a formula exists).\\ \hline
@@
 Consider the following example.  We have three objects related as indicated in
-Figure~\ref{fig-car1}
+Figure~\ref{fig-car1}.
 \begin{figure}
 \begin{center}
-\includegraphics{pic.pdf}
+\includegraphics[scale=.8]{pic.pdf}
 \end{center}
@@
 \end{figure}
-
+%
 The points $o$ and $o'$ can easily be distinguished in first-order logic:
@@
 are ``points which are related to points which are related to points\ldots''
+
+In what follows we will introduce two logical languages (called \el and \alc, see~\cite{handbook}) which can naturally be mapped to standard grammatically correct expressions.  \el allows
+for conjunctions of atomic and existentially quantified relational expressions (e.g.,
+``a red book on a table'') while \alc is the extension of \el with full negation
+(e.g., ``a book which is not red and is not on the floor'').
+
+\begin{definition}
+Formulas of $\alc$ are generated by the following grammar:
+$$
+\form ::= p \mid \neg \varphi \mid \varphi \wedge \varphi' \mid \exists R. \varphi
+$$
+where $p$ is in the set of propositional symbols \prop, $R$ is in the set of relational symbols \rel, $\varphi$ and $\varphi'$ are in \form. $\el$ is the
+negation-free fragment of $\alc$.
+
+Formulas of both $\alc$ and $\el$ are interpreted in relational first-order models
+$\gM = (\Delta^\gM,|\cdot|^\gM)$ where $W$ is an arbitrary non-empy set, and $|\cdot|^\gM$ is an interpretation
+function such that:
+$$
+\begin{array}{ccl}
+|p|^\gM & \subseteq & \Delta^\gM  \mbox{ for $p \in \prop$}\\
+|R|^\gM & \subseteq & \Delta\times W  \mbox{ for $R \in \rel$}\\
+|\neg \varphi|^\gM & = & W \backslash |\varphi|^\gM\\
+|\varphi \wedge \varphi'|^\gM & = & |\varphi|^\gM \cap |\varphi'|^\gM\\
+|\exists R.\varphi|^\gM & = & \{o \mid \mbox{for some } o' ((o,o') \in |R|^\gM)\\
+& & \mbox{ and } o' \in |\varphi|^\gM \}.\\
+\end{array}
+$$
+\end{definition}
+
+We can now exactly characterize the problem of when a formula distinguishing two
+elements in a given model exists. We will do so using the notion of \emph{bisimulaiton}.
+
+\begin{definition}
+Given a model $\gM = (\Delta^\gM,)$
+\end{definition}
 

### 2008-03-13 trunk/Papers/dl-gre-08/bisim.tex
More text and added picture.

@@
 $$
+Given a model $\gM$ and an object $o$ in the domain of $\gM$, let
+$\propm(o) = \{p \in \prop \mid o \in |p|^\gM\}$.
 \end{definition}
@@
 We can now exactly characterize the problem of when a formula distinguishing two
-elements in a given model exists. We will do so using the notion of \emph{bisimulaiton}.
+elements in a given model exists. We will do so using the notion of \emph{bisimulation}.
 
 \begin{definition}
-Given a model $\gM = (\Delta^\gM,)$
+Given a model $\gM = (\Delta^\gM,|\cdot|^\gM)$, two elements $o$ and $o'$ of $\Delta^\gM$ are $\alc$-bisimilar if it is possible to find a relation $Z \subseteq \Delta^\gM \times \Delta^\gM$ such that
+\begin{enumerate}
+\item $(o,o') \in Z$.
+\item if $(e_1, e_2) \in Z$ then $\propm(e_1) = \propm(e_2)$.
+\item if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_1'$ such that
+$(e_1,e_1') \in |R|^\gM$ then there is $e_2'$ such that $(e_2,e_2') \in |R|^\gM$ and
+$(e_1',e_2') \in Z$.
+\item if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_2'$ such that
+$(e_2,e_2') \in |R|^\gM$ then there is $e_1'$ such that $(e_1,e_1') \in |R|^\gM$ and
+$(e_1',e_2') \in Z$.
+\end{enumerate}
 \end{definition}
+
+\begin{theorem}[\cite{modallogic}]
+Given a model $\gM$, if  $o$ and $o'$ are two $\alc$-bisimilar elements in $\gM$
+then there is no $\alc$-formula $\varphi$ such that $o \in |\varphi|^\gM$ and
+$o' \not \in |\varphi|^\gM$.
+\end{theorem}
+
+In other words, $\alc$-bisimilar elements are indistinguishable in the $\alc$ langauge.
+By defining a suitable notion of $\el$-bisimulation, the same result can be proved for
+the $\el$ language.
 

### 2008-03-14 trunk/Papers/dl-gre-08/bisim.tex
Changed 'object' to individual. Changed '\wedge' to `\sqcap'. Added convers of bisimulation preservation. Resto to write the Paige & Tarjan algorithm.

--- 2008-03-13 trunk/Papers/dl-gre-08/bisim.tex
@@
 
-In computational linguistics, the genereting referring expression problem (the GRE
+In computational linguistics, the generating referring expression problem (the GRE
 problem) can be intuitively presented as follows: given some information about
-a group of objects (e.g., grounded facts about individuals in an ongoing dialogue),
-and a grammar, found a grammatically correct expression (usually a noun phrase) that uniquely identify a given object.
+a group of individuals (e.g., grounded facts about individuals in an ongoing dialogue),
+and a grammar, found a grammatically correct expression (usually a noun phrase) that uniquely identify a given individual.
 
 By taking a logic perspective, we can recast the GRE problem as an inference problem.
-We can think that the information about the group of objects is provided in the form
+We can think that the information about the group of individuals is provided in the form
 of a model $\gM$ of a given logical language $\gL$.  Now, the inference problem
 associated with the GRE problem will be to find a formula in $\gL$ such that it
-exacly identify a given object $o$.  Formally, if $|\cdot|^\gM$ is the interpretation
+exacly identify a given individual $i$.  Formally, if $|\cdot|^\gM$ is the interpretation
 function for $\gL$ which for each formula in $\gL$ and for each model $\gM$ returns
@@
 \textsc{$\gL$-GRE Problem}\\ \hline
-\ \ \ Input: A model $\gM$, an object $o$ in the domain of $\gM$\\
-\ \ \ Output: a formula $\varphi \in \gL$ such that $|\varphi|^\gM = \{o\}$\\
+\ \ \ Input: A model $\gM$, an individual $i$ in the domain of $\gM$\\
+\ \ \ Output: A formula $\varphi \in \gL$ such that $|\varphi|^\gM = \{i\}$\\
 \hspace*{1.4cm} (if such a formula exists).\\ \hline
@@
 As it can be seen from the definition, the problem depends on which particular
-logical language $\gL$ we are interested in, and also (as the original GRE problem)
-it might terminate with failure.  Depending on the model $\gM$, the chosen object $o$
+logical language $\gL$ we are interested in, and (as the original GRE problem)
+it might terminate with failure.  Depending on the model $\gM$, the chosen individual $i$
 and the expressive power of $\gL$ it can be the case that there is no formula of
-$\gL$ that uniquely identifies $o$.
+$\gL$ that uniquely identifies $i$.
 
-Consider the following example.  We have three objects related as indicated in
-Figure~\ref{fig-car1}
+Consider the following example.  We have three individuals related as indicated in
+Figure~\ref{fig-car1}.
 \begin{figure}
 \begin{center}
-\includegraphics{pic.pdf}
+\includegraphics[scale=.8]{pic.pdf}
 \end{center}
@@
 \end{figure}
+%
+The individuals $i$ and $i'$ can easily be distinguished in first-order logic:
+the formula $R(x,x)$ is true of $i$ and false in $i'$. The corresponding
+referring expression for $i$ would be ``an individual relating to itself''.
+But if we can only use propositional and relational information (equivalently, our grammar does not contain the word ``itself'') then $i$ and $i'$ are indistinguishable. Both
+are ``individuals which are related to individuals which are related to individuals\ldots''
 
-The points $o$ and $o'$ can easily be distinguished in first-order logic:
-the formula $R(x,x)$ is true of $o$ and false in $o'$. The corresponding
-referring expression for $o$ would be ``a point relating to itself''.
-But if we can only use propositional and relational information (equivalently, our grammar does not contain the word ``itself'') then $o$ and $o'$ are indistinguishable. Both
-are ``points which are related to points which are related to points\ldots''
+In what follows we will introduce two logical languages (called \el and \alc, see~\cite{baad:desc03}) which can naturally be mapped to standard grammatically correct expressions.  \el allows
+for conjunctions of atomic and existentially quantified relational expressions (e.g.,
+``a red book on a table'') while \alc is the extension of \el with full negation
+(e.g., ``a book which is not red and is not on the floor'').
+
+\begin{definition}
+Formulas of $\alc$ are generated by the following grammar:
+$$
+\form ::= p \mid \neg \varphi \mid \varphi \sqcap \varphi' \mid \exists R. \varphi
+$$
+where $p$ is in the set of propositional symbols \prop, $R$ is in the set of relational symbols \rel, $\varphi$ and $\varphi'$ are in \form. $\el$ is the
+negation-free fragment of $\alc$.
+
+Formulas of both $\alc$ and $\el$ are interpreted in relational first-order models
+$\gM = (\Delta^\gM,|\cdot|^\gM)$ where $W$ is an arbitrary non-empy set, and $|\cdot|^\gM$ is an interpretation
+function such that:
+$$
+\begin{array}{ccl}
+|p|^\gM & \subseteq & \Delta^\gM  \mbox{ for $p \in \prop$}\\
+|R|^\gM & \subseteq & \Delta^\gM \times \Delta^\gM  \mbox{ for $R \in \rel$}\\
+|\neg \varphi|^\gM & = & W \backslash |\varphi|^\gM\\
+|\varphi \wedge \varphi'|^\gM & = & |\varphi|^\gM \cap |\varphi'|^\gM\\
+|\exists R.\varphi|^\gM & = & \{i \mid \mbox{for some } i' ((i,i') \in |R|^\gM)\\
+& & \mbox{ and } i' \in |\varphi|^\gM \}.\\
+\end{array}
+$$
+Given a model $\gM$ and an individual $i$ in the domain of $\gM$, let
+$\propm(i) = \{p \in \prop \mid i \in |p|^\gM\}$.
+\end{definition}
+
+We can now exactly characterize the problem of when a formula distinguishing two
+individuals in a given model exists. We will do so using the notion of \emph{bisimulation}.
+
+\begin{definition}
+Given a model $\gM = (\Delta^\gM,|\cdot|^\gM)$, two elements $i$ and $i'$ of $\Delta^\gM$ are $\alc$-bisimilar if it is possible to find a relation $Z \subseteq \Delta^\gM \times \Delta^\gM$ such that
+\begin{enumerate}
+\item $(i,i') \in Z$.
+\item if $(e_1, e_2) \in Z$ then $\propm(e_1) = \propm(e_2)$.
+\item if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_1'$ such that
+$(e_1,e_1') \in |R|^\gM$ then there is $e_2'$ such that $(e_2,e_2') \in |R|^\gM$ and
+$(e_1',e_2') \in Z$.
+\item if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_2'$ such that
+$(e_2,e_2') \in |R|^\gM$ then there is $e_1'$ such that $(e_1,e_1') \in |R|^\gM$ and
+$(e_1',e_2') \in Z$.
+\end{enumerate}
+\end{definition}
+
+\begin{theorem}[\cite{blac:moda01}]\label{bisim}
+Given a model $\gM$, if  $i$ and $i'$ are two $\alc$-bisimilar elements in $\gM$
+then there is no $\alc$-formula $\varphi$ such that $i \in |\varphi|^\gM$ and
+$i' \not \in |\varphi|^\gM$.
+\end{theorem}
+
+In other words, $\alc$-bisimilar elements are indistinguishable in the $\alc$ langauge.
+If we only consider finite models (actually, it is enought if we consider models where each element has only a finite number of immediate successors) then the converse of Theorem~\ref{bisim} also holds.
+
+\begin{theorem}[\cite{blac:moda01}]\label{bisim2}
+Given a model $\gM$, $i$ and $i'$ are two $\alc$-bisimilar elements in $\gM$
+if and only if there is no $\alc$-formula $\varphi$ such that $i \in |\varphi|^\gM$ and
+$i' \not \in |\varphi|^\gM$.
+\end{theorem}
+
+By defining a suitable notion of $\el$-bisimulation, the same results can be proved for
+the $\el$ language.
+
+\cite{paig:thre87}
 

### 2008-03-16 trunk/Papers/dl-gre-08/bisim.tex
explained my algorithm

@@
 |\neg \varphi|^\gM & = & W \backslash |\varphi|^\gM\\
-|\varphi \wedge \varphi'|^\gM & = & |\varphi|^\gM \cap |\varphi'|^\gM\\
+|\varphi \sqcap \varphi'|^\gM & = & |\varphi|^\gM \cap |\varphi'|^\gM\\
 |\exists R.\varphi|^\gM & = & \{i \mid \mbox{for some } i' ((i,i') \in |R|^\gM)\\

### 2008-03-16 trunk/Papers/dl-gre-08/gre.tex
added structure and some references
wrote first versions of intro, related work, efficiency
explained my algorithm

--- 2008-02-05 trunk/Papers/dl-gre-08/bisim.tex
@@
-\section{Computing bisimulation classes} \label{sec:bisim}
+\section{Generating referring expressions} \label{sec:gre}
+
+We can connect description logics to the problem of generating
+referring expressions via the concept of \emph{bisimulation classes}.
+A bisimulation class is a maximal subset $C$ of the domain such that
+all members of $C$ are pairwise bisimilar.  For the finite models we
+deal with in GRE, this amounts to saying that there are certain
+concepts that are satisfied by all individuals in $C$ and no
+individuals that are not in $C$ (by Theorems~\ref{bisim} and
+\ref{bisim2}).  If we represent referring expressions as DL concepts,
+this means that there exists a unique referring expression for some
+individual $a$ iff $a$ is alone in its bisimulation class; and if it
+is, we can use one of the concepts that are characteristic for $C$ as
+the referring expression.
+
+Representing referring expressions as DL concepts is a very natural
+perspective.  The purely conjunctive bag of propositional symbols that
+e.g.\ \newcite{Dale1995} compute is clearly a conjunction in
+description logic.  Relational expressions as in
+\newcite{dale91:_gener_refer_expres_invol_relat} or
+\newcite{Krahmer2003} are simply concepts of \el; for instance, ``the
+book on the table'' corresponds to the concept $\mathsf{book} \sqcap
+\exists \mathsf{on}. \mathsf{table}$.  And other propositional
+connectives, such as disjunction and negation
+\cite{deemter01:_gener_refer_expres}, can clearly be represented in
+\alc\ as well.
+
+All this means that we can reduce the problem of computing a unique RE
+to the problem of computing the bisimulation classes of a model.  We
+will now present two algorithms -- one for \alc\ and one for \el\ --
+for doing this.  In fact, each of these algorithms computes all
+bisimulation classes of a model at once, effectively computing a RE
+for every single individual in the domain simultaneously.
+
+\subsection{Bisimulation classes for \alc}
+
+\subsection{Bisimulation classes for \el}
+
+The algorithm we just presented is driven by case distinctions between
+concepts $\varphi$ and $\neg \varphi$, and thus adds large amounts of
+negations to the characteristic concepts.  This can be inconvenient
+in NLG, because such concepts may be expressible only by rather clumsy
+natural-language expressions. \todo{example?}  We will therefore
+introduce a new algorithm which is limited to computing
+(negation-free) concepts of \el.
+
+As before, the algorithm maintains a set $\C = \{C_1,\ldots,C_n\}$ of
+concepts (this time of \el) such that $\interp{C_1} \cup \ldots \cup
+\interp{C_n} = \Delta$, and which it refines iteratively.  However,
+where the \alc\ algorithm maintains the invariant that
+$\interp{C_1},\ldots,\interp{C_n}$ is a partition of $\Delta$, we
+weaken this invariant to the requirement that there are no $m \geq 2$
+pairwise different indices $1 \leq i_1,\ldots,i_m \leq n$ such that
+$\interp{C_{i_1}} = \interp{C_{i_2}} \cup \ldots \cup
+\interp{C_{i_m}}$.  We call the concept $C_{i_1}$ \emph{subsumed} if
+such a decomposition exists.
+
+\begin{algorithm}[t]
+\dontprintsemicolon
+\caption{\el\ bisimulation classes}
+\label{algo:bisim-el}
+$\C \leftarrow \{\top\}$ \;
+\For{$p \in \prop$}{
+  add$_\el$(p, \C) \;
+}
+\While{not all concepts in $\C$ singleton}{
+  \For{$C \in \C$, $R \in \rel$}{
+    add$_\el$($\exists R.C$, $\C$) \;
+  }
+  \If{made no changes to $\C$}{
+    exit\;
+  }
+}
+\end{algorithm}
+
+\begin{algorithm}[t]
+\dontprintsemicolon
+\caption{add$_\el$($C$, $\C$)}
+\label{algo:bisim-add-el}
+\For{$C' \in \C$ non-singleton}{
+  \If{$C' \sqcap C$ is not subsumed in $\C$}{
+    add $C' \sqcap C$ to $\C$ \;
+    remove subsumed concepts from $\C$\;
+  }
+}
+\end{algorithm}
+
+Algorithms~\ref{algo:bisim-el} and \ref{algo:bisim-add-el} compute
+the \el\ bisimulation classes for a given model.
+Algorithm~\ref{algo:bisim-el} iterates over all propositional and
+relational symbols of the signature to construct new concepts until
+either all concepts in $\C$ are singleton (i.e., there is only one
+individual that satisfies them), or no progress has been made in the
+previous iteration.  In each iteration, it calls the procedure
+add$_\el$($C$, $\C$), which intersects $C$ with all current concepts
+that are not already singleton, adds the result to $\C$ and then
+removes all concepts in $\C$ that have become subsumed.  Removal of
+subsumed concepts can be implemented efficiently by arranging the
+concepts in $\C$ in a a graph that connects two concepts with an edge
+if one is a maximal proper subset of the other.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=5cm]{pic-dale-haddock}
+  \caption{The Dale and Haddock scenario.}
+  \label{fig:dale-haddock}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\columnwidth]{pic-el-vs-alc}
+  \caption{Illustrating the difference between \el\ and \alc.}
+  \label{fig:el-vs-alc}
+\end{figure}
+
+Let's see what this algorithm does on the example shown in
+Fig.~\ref{fig:dale-haddock}, which is taken from
+\newcite{dale91:_gener_refer_expres_invol_relat}.  In the first loop,
+the algorithm will add new concepts $\mathsf{floor}$, $\mathsf{bowl}$,
+$\mathsf{cup}$, and $\mathsf{table}$; because adding these concepts
+makes $\top$ subsumed, $\C$ will consist of these four concepts after
+the loop.  Not all of these concepts are singleton; for instance,
+$\interp{\mathsf{cup}}$ contains two individuals.  So we iterate over
+the relations to make the concepts more precise.  After the first
+iteration over the relations, we have $\C = \{ \mathsf{floor},
+\mathsf{bowl} \sqcap \exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl}
+\sqcap \exists \mathsf{on}.\mathsf{table}, \mathsf{cup},
+\mathsf{table} \}$. Notice that $\mathsf{bowl}$ has become subsumed,
+but we haven't distinguished the cups and tables further, but we can
+use the distinction between the bowls to distinguish the cups in the
+second iteration.  The result of this is $\C = \{ \mathsf{floor},
+\mathsf{bowl} \sqcap \exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl}
+\sqcap \exists \mathsf{on}.\mathsf{table}, \mathsf{cup} \sqcap \exists
+\mathsf{in}. (\mathsf{bowl} \sqcap \exists
+\mathsf{on}.\mathsf{floor}), \mathsf{cup} \sqcap \exists
+\mathsf{in}. (\mathsf{bowl} \sqcap \exists
+\mathsf{on}.\mathsf{table}), \mathsf{table} \}$.  At this point, all
+concepts except for $\mathsf{table}$ are singleton, and further
+iterations don't allow us to refine $\mathsf{table}$; so the algorithm
+terminates.  Each of the singleton bisimulation classes is represented
+by a characteristic concept; for instance, $\C$ tells us that
+$\mathsf{cup} \sqcap \exists \mathsf{in}. (\mathsf{bowl} \sqcap
+\exists \mathsf{on}.\mathsf{table})$ is only satisfied by $c_2$, so we
+may refer to $c_2$ as ``the cup in the bowl on the table''.
+
+Compared to the \alc\ algorithm, the \el\ algorithm will generally
+compute larger bisimulation classes, and compute singleton
+bisimulation classes less frequently, because the concepts it can
+build can ``see'' fewer differences between individuals.  For example,
+in the model shown in Fig.~\ref{fig:el-vs-alc}, the \alc\ algorithm
+will compute the concepts $\{\neg \exists R. (\neg P \sqcap \neg Q),
+\exists R. (\neg P \sqcap \neg Q), Q, \neg P \sqcap \neg Q\}$ (each of
+which is singleton), but $a$ and $b$ are \el-bisimilar in this model,
+and the only \el-concept that is true in $d$ is $\top$, so the \el\
+algorithm can only compute the concepts $\{\top, P, P \sqcap Q\}$.
+Furthermore, because the \el\ algorithm maintains a weaker invariant
+for the classes, it may consider more of these classes simultaneously.
+Prima facie, this means that it has worst-case exponential runtime,
+given that the whole domain has an exponential number of subsets.
+However, it is possible that a more careful analysis of the
+combinatorics of the considered subsets reveal that the algorithm is
+actually polynomial; it certainly behaves polynomially in our
+evaluations (see below).
 

### 2008-03-16 trunk/Papers/dl-gre-08/gre.tex
explained my algorithm

@@
 \For{$C' \in \C$ non-singleton}{
-  \If{$C' \sqcap C$ is not subsumed in $\C$}{
+  \If{$C' \sqcap C$ is not subsumed in $\C$ {\bf and} 
+    $\interp{C' \sqcap C} \neq \emptyset$ {\bf and}
+    $\interp{C' \sqcap C} \neq \interp{C'}$}{
     add $C' \sqcap C$ to $\C$ \;

### 2008-03-18 trunk/Papers/dl-gre-08/discussion.tex
example outputs

@@
 
-Now, we will compare the descriptions generated by our algorithm to
-those humans produce. For this purpose, we use a corpus of
+To compare the descriptions generated by our algorithm to
+those humans produce, we use a corpus of
 human-generated referring expressions collected and made available by
@@
 \ref{fig:drawers}. The human-generated descriptions use four
-non-relational (the drawer's \textsf{color}, its \textsf{column} and
+non-relational properties (the drawer's \textsf{color}, its \textsf{column} and
 \textsf{row} number, and whether it is in a \textsf{corner}) and five
@@
 of}). Of the 118 referring expressions, only 15 use relational properties.
-\cite{viethen06:_algor_for_gener_refer_expres}
+Viethen and Dale \shortcite{viethen06:_algor_for_gener_refer_expres}
 describe the data in more detail and present results of evaluating the
@@
 descriptions produced by humans can be generated using one of the 24
-possible preference orderings.
+possible preference orderings. Unfortunately, there isn't one ordering
+that covers all those 98 descriptions - a set of four different
+orderings is necessary.
 
@@
 possible orderings of predicates and roles (with the restriction that
-all predicates come before all roles). Our algorithm generates 10 of
-the 15 relational descriptions. Of the five that it could not
-generate, two mentioned column information, which we excluded from our
-domain, and three (which interestingly were all produced by the same
-person) are of the format \textit{the blue/pink/yellow/orange drawer
-above/below/next to/right of/left of the two blue/pink/yellow/orange
-drawers}. This would correspond to the following output from our
-algorithm: $C_1 \& \exists R . (C_2 \& \exists R . C_2)$ (where $C_i$
-are color predicates and $R$ a role). While our algorithm does not
-generate descriptions that contain the same relation on two levels of
-embedding for these drawers, it does generate some other, simpler
-descriptions for them; for example, $\textsf{orange} \& \exists
+all predicates come before all roles). For 10 of the 15 human-produced
+relational descriptions there is an ordering such that using this
+ordering our algorithm can generate this description. Figure
+\ref{fig:example_outputs} shows the output of our \el\ algorithm for
+one preference ordering of predicates. To cover 10 of the 15
+human-produced descriptions three different orderings are
+necessary. Of the five human-produced escriptions that the \el\
+algorithm can not generate, two mention column information, which we
+excluded from our domain, and three (which interestingly were all
+produced by the same person) are of the format \textit{the
+blue/pink/yellow/orange drawer above/below/next to/right of/left of
+the two blue/pink/yellow/orange drawers}. This would correspond to the
+following output from our algorithm: $C_1 \& \exists R . (C_2 \&
+\exists R . C_2)$ (where $C_i$ are color predicates and $R$ a
+role). While our algorithm does not generate descriptions that contain
+the same relation on two levels of embedding for these drawers, it
+does generate some other, simpler descriptions for them; for example,
+$\textsf{orange} \& \exists
 \textsf{left\_of} . (\textsf{pink}
@@
 \todo{should we show some more examples? - ak}
+
+\begin{figure*}
+
+\begin{small}
+\begin{tabular}{|lp{0.5\textwidth}p{0.4\textwidth}|}\hline
+drawer & human-produce description & output of the \el\ algorithm\\\hline
+2 & \textit{the orange drawer above the blue drawer} & $\exists \textsf{above}.(\textsf{blue}) \& \textsf{orange}$\\
+4 & \textit{the yellow drawer on the top of the pink one} & $\exists \textsf{above}.(\textsf{pink}) \& \textsf{yellow}$\\
+5 & \textit{the pink drawer in the fourth column below the yellow one} & $\exists \textsf{above}.(\textsf{orange}) \& \textsf{pink}$\\
+6 & \textit{the yellow drawer on top of the yellow drawer} & $\exists \textsf{above}.(\textsf{yellow}) \& \textsf{yellow}$\\
+6 & \textit{the yellow drawer that's above another yellow drawer} & \\
+6 & \textit{the drawer after the two blue ones in horizontal sequence} & \\
+7 & \textit{the blue drawer below the orange drawer in the second column} & $\exists \textsf{next}.(\exists \textsf{above}.(\textsf{orange}) \& \textsf{blue}) \& \exists \textsf{above}.(\textsf{blue}) \& \textsf{blue}$\\
+7 & \textit{the blue drawer below the orange one} & \\
+10 & \textit{the blue drawer above the pink drawer} & $\exists \textsf{above}.(\textsf{pink}) \& \textsf{blue}$\\
+10 & \textit{the blue drawer above the pink drawer} & \\
+11 & \textit{the yellow drawer next to the orange drawer} & $\exists \textsf{above}.(\textsf{orange}) \& \textsf{yellow}$\\
+11 & \textit{the yellow one next to the orange one} & \\
+12 & \textit{the orange drawer below the pink drawer} & $\exists \textsf{above}.(\textsf{pink} \& \textsf{corner}) \& \textsf{orange}$\\
+14 & \textit{the orange drawer below the two yellow drawers} & $\exists \textsf{next}.(\textsf{pink} \& \textsf{corner}) \& \exists \textsf{next}.(\textsf{pink}) \& \textsf{orange}$\\
+14 & \textit{the orange drawer below the two yellow drawers} & \\\hline
+\end{tabular}
+\end{small}
+
+\caption{Human produced descriptions and output of our \el\  algorithm using the ordering \textsf{corner}, \textsf{color}, \textsf{above, next, below, right, left}}
+\label{fig:example_outputs}
+\end{figure*}
+
+% ordering no. 235
+%List(corner, green, pink, orange, blue)
+%List(above, next, below, right, left)
 
@@
 
-\begin{itemize}
-\item there's a risk to generate a concept that can't be realized,
-  especially for the \alc\ algorithm
-\item everybody except perhaps SPUD has this problem
-\item problem is worse in our case because it is harder to control the
-  order in which atoms and relations are explored
-\item would be interesting for future work to think about how
-  search in the bisim algorithms can be controlled by the realizer
-\end{itemize}
+%\begin{itemize}
+%\item there's a risk to generate a concept that can't be realized,
+%  especially for the \alc\ algorithm
+%\item everybody except perhaps SPUD has this problem
+%\item problem is worse in our case because it is harder to control the
+%  order in which atoms and relations are explored
+%\item would be interesting for future work to think about how
+%  search in the bisim algorithms can be controlled by the realizer
+%\end{itemize}
+
+Like other algorithms that do content determination without taking
+into account linguistic contraints, we have no guarantee that the
+descriptions generated by our algorithm can actually be realized. For
+example, none of the descriptions out algorithms generate in the
+drawer domain contain a property which would commonly be realized as a
+noun (the property \textsf{drawer} is not included because it applies
+to all entities in this domain). If we fed this output directly into a
+realizer, the realizer would thus be faced with the task of having to
+produce a noun phrase without being given anything that could serve as
+the head noun. Of course, this particular problem could easily be
+dealt with through a post-processing step that adds a type property to
+the description. But at least for the \alc\ algorithm it is not clear
+that all non-realizable descriptions could be fixed through such
+simple post-processing solutions.
+
+Partly in response to this problem, Horacek \shortcite{} proposes to
+interleave content determination with realization and build the
+linguistic representation of a natural language description in
+parallel with building the representation of its semantic content.
+
+SPUD \cite{} is another system that avoids this problem by
+interleaving content determination with realization. SPUD goes one
+step further and instead of generating referring expressions in
+isolation, deals with them in the context of the sentence they appear
+in taking into account semantic information contributed by other parts
+of that sentence. This allows them to generate utterances like
+\textit{``remove the rabbit from the hat''} which can be used in contexts
+containing several rabbits and several hats as long as there is only
+one rabbit-hat pair where the rabbit is in the hat. While our
+algorithm does generate referring expressions for all entities in the
+context at once (and thereby avoids the problem of infinite regress
+\todo{discussed earlier}), it is currently not designed to take into
+account additional information provided by the context. When used to
+generate the expressions referring to the relevant rabbit and hat in
+the situation described above, our algorithms would produce
+\textit{``the rabbit in the hat''} and \textit{``the hat with the
+rabbit''}, respectively, yielding \textit{``remove the rabbit in the
+hat from the hat with the rabbit''}.
+
+To improve the interface between our algorithm and a realization
+component, we therefore need to investigate ways of how linguistic
+contraints on realization can be used to guide our content
+determination algorithms and how it can be integrated into the process
+of generating a whole utterance in such a way that relevant
+information contributed by other parts of the utterance can be taken
+into account.
 

### 2008-03-18 trunk/Papers/dl-gre-08/conclusion.tex
added structure and some references
wrote first versions of intro, related work, efficiency
iterated 4.3 and 5; updated runtimes picture

--- 2008-02-05 trunk/Papers/dl-gre-08/bisim.tex
@@
-\section{Computing bisimulation classes} \label{sec:bisim}
+\section{Conclusion} \label{sec:conclusion}
+
+referring to sets (or is this trivial?)
+
+where do we talk about mutually constraining descriptions like ``the
+rabbit in the hat''?
+
+will make implementation available
+
+first time to actually show efficient runtimes for generation of
+relational REs
+
+Because they compute referring expressions for all individuals in the
+domain together, our algorithms will be strongest in static settings,
+such as the generation of descriptions for museum exhibits, in which
+the individuals and their properties don't change much.  Nevertheless,
+they remain fast enough for real-time use in dynamic settings.
+Another advantage that recent algorithms for relational REs
+\cite{Krahmer2003,kelleher06:_increm_gener_of_spatial_refer} have over
+ours is that they can take weights and salience into account.  How
+this could be incorporated into our algorithms is an interesting
+question for future research.
+
+relational descriptions in data are comparatively rare and
+inconsistent; not sufficiently annotated in TUNA
+\cite{deemter06:_build_seman_trans_corpus_for}; need more.
+
+Things that we care about but that are never said explicitly so far:
+
+\begin{itemize}
+\item in a static scenario, we only have to compute REs once -- and in
+  any application that uses many REs before the context changes (e.g.\
+  museums), this is exactly what we want to do
+\item it is difficult to control the search for concepts by linguistic
+  constrtaints
+\item our GRE algorithms can take orderings over the propositional and
+  relational symbols into account
+\end{itemize}
 

### 2008-03-18 trunk/Papers/dl-gre-08/discussion.tex
iterated 4.3 and 5; updated runtimes picture

@@
 \section{Discussion} \label{sec:discussion}
-
-%evaluate on TUNA \cite{deemter06:_build_seman_trans_corpus_for} or a
-%richer version of it
-
-%or on the cabinets corpus
-%\cite{viethen06:_algor_for_gener_refer_expres}
 
@@
   \includegraphics[width=\columnwidth]{runtimes}
-  \caption{Average runtimes (in ms) of the two algorithms on random graphs of
-    different sizes.} 
+  \caption{Average runtimes (in ms) of the two algorithms on random
+    models of     different sizes.} 
   \label{fig:runtimes}
@@
 
-In order to get a more comprehensive picture of our system's
-efficiency, we ran the two algorithms on random graphs with increasing
-numbers of nodes.  \todo{harmonize this with what Carlos calls them in
-  Section~\ref{sec:bisim}} Each graph had ten different atoms and four
-different roles, a 10\% chance for a node to have each atom, and a
-10\% chance for each pair of nodes to be connected by each role.  The
-results (averaged over 10 runs for each graph size) are shown in
-Fig.~\ref{fig:runtimes}.  As the figure shows, the \el\ algorithm
-takes about 350 ms on average to generate relational REs for all
-individuals in the graph with 100 nodes, i.e.\ 4 ms on average for
-each individual.  The \alc\ algorithm is even faster, at about 200 ms
-for the graph with 100 nodes.  In both cases, we conclude that the
-systems are fast enough for practical use.
+In order to get a more comprehensive picture of the algorithms'
+efficiency, we ran them on random models with increasing numbers of
+individuals.  Each model had random interpretations for ten different
+propositional and four relational symbols; each individual had a 10\%
+chance to be in the extension of each propositional symbol, and each
+pair of individuals had a 10\% chance to be related by a relational
+symbol.  The results (averaged over 10 runs for each model size) are
+shown in Fig.~\ref{fig:runtimes}.  As the figure shows, the \el\
+algorithm takes about 350 ms on average to generate relational REs for
+all individuals in the model of size 100, i.e.\ less than 4 ms on
+average for each individual.  The \alc\ algorithm is even faster, at
+about 140 ms for the model of size 100.  As far as we know, these are
+by far the fastest published runtimes for any relational GRE algorithm
+in the literature.
+
+Another encouraging aspect of both runtime graphs is that they seem to
+be approximated well by quadratic functions.  So although the
+worst-case runtime of the \el\ algorithm might be exponential, both
+algorithms empirically exhibit (low) polynomial runtime on the inputs
+we considered.
 
@@
 
-Like other algorithms that do content determination without taking
-into account linguistic contraints, we have no guarantee that the
-descriptions generated by our algorithm can actually be realized. For
-example, none of the descriptions out algorithms generate in the
-drawer domain contain a property which would commonly be realized as a
-noun (the property \textsf{drawer} is not included because it applies
-to all entities in this domain). If we fed this output directly into a
-realizer, the realizer would thus be faced with the task of having to
-produce a noun phrase without being given anything that could serve as
-the head noun. Of course, this particular problem could easily be
-dealt with through a post-processing step that adds a type property to
-the description. But at least for the \alc\ algorithm it is not clear
-that all non-realizable descriptions could be fixed through such
-simple post-processing solutions.
+Our GRE algorithms do not guarantee that the concepts they compute can
+actually be realized in language.  For example, none of the concepts
+our algorithms computed in the Viethen \& Dale domain contained an
+atom that would commonly be realized as a noun; the property
+$\mathsf{drawer}$ is never used because it applies to all individuals
+in the domain.  This particular problem could easily be worked around
+in a post-processing step.  However, another problem arises from the
+massive use of negation in the \alc\ algorithm; it will be hard for
+any realizer to find a reasonable way of expressing the concept $\neg
+\exists R.(\neg P \sqcap \neg Q)$ from the example in
+Fig.~\ref{fig:el-vs-alc} as a smooth noun phrase.  Although we agree
+with \newcite{deemter01:_gener_refer_expres} and others that the
+careful use of negation and disjunction can improve REs, these
+connectives must not be overused.  Thus we consider the concepts
+computed by the \el\ algorithm ``safer'' with respect to realization.
 
-Partly in response to this problem, Horacek \shortcite{} proposes to
-interleave content determination with realization and build the
-linguistic representation of a natural language description in
-parallel with building the representation of its semantic content.
+Of course, we share the problem of interfacing GRE and realization
+with every other approach that separates these two modules, i.e.\
+almost the entire GRE literature (notable exceptions are e.g.\ SPUD
+\cite{Stone2003a} and \newcite{Horacek1997}).  In principle, we
+believe that it is a good idea to handle sentence planning and
+realization in a single module; for instance, SPUD can use its
+awareness of the syntactic context to generate succinct REs as in
+``take the rabbit from the hat''.  We hope that the ideas we have
+explored here for efficient and expressive RE generation can
+eventually be combined with recent efficient algorithms for integrated
+sentence planning and realization, such as in \newcite{KolSto07}.
 
-SPUD \cite{} is another system that avoids this problem by
-interleaving content determination with realization. SPUD goes one
-step further and instead of generating referring expressions in
-isolation, deals with them in the context of the sentence they appear
-in taking into account semantic information contributed by other parts
-of that sentence. This allows them to generate utterances like
-\textit{``remove the rabbit from the hat''} which can be used in contexts
-containing several rabbits and several hats as long as there is only
-one rabbit-hat pair where the rabbit is in the hat. While our
-algorithm does generate referring expressions for all entities in the
-context at once (and thereby avoids the problem of infinite regress
-\todo{discussed earlier}), it is currently not designed to take into
-account additional information provided by the context. When used to
-generate the expressions referring to the relevant rabbit and hat in
-the situation described above, our algorithms would produce
-\textit{``the rabbit in the hat''} and \textit{``the hat with the
-rabbit''}, respectively, yielding \textit{``remove the rabbit in the
-hat from the hat with the rabbit''}.
-
-To improve the interface between our algorithm and a realization
-component, we therefore need to investigate ways of how linguistic
-contraints on realization can be used to guide our content
-determination algorithms and how it can be integrated into the process
-of generating a whole utterance in such a way that relevant
-information contributed by other parts of the utterance can be taken
-into account.
+One problem that arises in our approach specifically is that the
+bisimulation classes algorithms derive some measure of efficiency from
+their freedom to build concepts without having to respect any
+linguistic constraints.  It seems straightforward to extend Krahmer et
+al.'s \shortcite{Krahmer2003} approach such that it only considers
+subgraphs that can actually be realized, because their algorithm
+proceeds by a genuine search for uniquely identifying subgraphs, and
+will simply take a different branch of the search if some subgraph is
+useless.  Our algorithms don't search in the same way; any constraints
+that disallow certain refinements of a partition jeopardize the
+algorithms' ability to find a partition into singletons in polynomial
+time.  A particularly vivid example is the fact that both algorithms
+crucially split over all propositional symbols before they start
+iterating over relational symbols; changing this may make them
+incomplete.  Investigating this interplay between efficiency and
+linguistic constraints is an interesting avenue for future research.
 

### 2008-03-18 trunk/Papers/dl-gre-08/related.tex
explained my algorithm
iterated 4.3 and 5; updated runtimes picture

--- 2008-02-05 trunk/Papers/dl-gre-08/introduction.tex
@@
-\section{Introduction} \label{sec:introduction}
+\section{A unified perspective on GRE} \label{sec:related}
 
-\cite{Dale1995}
-\cite{dale91:_gener_refer_expres_invol_relat}
+\begin{figure}
+  \centering
+  \begin{tabular}{l|l}
+    GRE algorithm & DL variant \\ \hline
+    \newcite{Dale1995} & only conjunction \\
+    \newcite{deemter01:_gener_refer_expres} & propositional connectives \\
+    \newcite{dale91:_gener_refer_expres_invol_relat} & \el \\
+    \newcite{Krahmer2003} & \el \\
+    \newcite{kelleher06:_increm_gener_of_spatial_refer} & \el \\
+    \newcite{gardent02:_gener_minim_defin_descr} & \alc
+  \end{tabular}
+  \caption{DL variants used by different GRE algorithms.}
+  \label{fig:related}
+\end{figure}
+
+The view of GRE as a problem of computing DL concepts allows us to
+organize existing approaches to GRE with respect to the logical
+connectives of DL which they can use in the referring expressions they
+generate.  We have summarized this for some approaches in
+Fig.~\ref{fig:related}.  We already mentioned some of these in
+Section~\ref{sec:gre} already.  In addition, the graphs
+\newcite{Krahmer2003} use to represent referring expressions can be
+seen as concepts of \el\ that are only satisfied at those points at
+which they can be embedded, and the descriptions generated by
+\newcite{gardent02:_gener_minim_defin_descr} are all concepts of \alc,
+without requiring that the bisimulation class must be singleton (to be
+able to refer to sets).
+
+Next to the inventory of DL connectives, we can also compare the
+algorithms that different approaches use to compute the concepts.  In
+a purely propositional setting, where there are no relations, our \el\
+algorithm performs essentially the same computation as the Incremental
+Algorithm \cite{Dale1995}, i.e.\ split the domain according to the
+positive atoms.  However, in the relational case, our algorithms
+operate very differently than any of the ones listed in
+Fig.~\ref{fig:related}.  Rather than attempting to generate a RE for a
+single individual and then doing the same for their neighbors if
+necessary, with all the associated risks of infinite regress that this
+entails (and that then have to be avoided by additional heuristics),
+we successively refine a decomposition of the domain into smaller
+classes, effectively computing REs for all individuals at the same
+time.  That is, our algorithm is not a search algorithm in the sense
+of \newcite{DBLP:conf/ijcai/BohnetD05}, and at least the \alc\
+algorithm can therefore compute concepts of minimal relational depth
+in worst-case polynomial time.  Perhaps closest in spirit is the
+Krahmer et al.\ graph algorithm, which also computes \el\ concepts by
+extending them successively; but this algorithm, too, is focused on
+reference to a single individual, and it has NP-complete worst-case
+complexity.
 

### 2008-03-19 trunk/Papers/dl-gre-08/discussion.tex

@@
 \ref{fig:drawers}. The human-generated descriptions use four
-non-relational properties (the drawer's \textsf{color}, its \textsf{column} and
-\textsf{row} number, and whether it is in a \textsf{corner}) and five
-relational properties (\textsf{above, below, next to, left of, right
-of}). Of the 118 referring expressions, only 15 use relational properties.
+non-relational properties (the drawer's $\mathsf{color}$, its $\mathsf{column}$ and
+$\mathsf{row}$ number, and whether it is in a $\mathsf{corner}$) and five
+relational properties ($\mathsf{above, below, next\ to, left\ of,
+right\ of}$). Of the 118 referring expressions, only 15 use relational properties.
 Viethen and Dale \shortcite{viethen06:_algor_for_gener_refer_expres}
 describe the data in more detail and present results of evaluating the
-Full Brevity algorithm by \cite{}, the Incremental Algorithm by
-\cite{} and the relational algorithm by \cite{} on this corpus.
+Full Brevity algorithm, the Incremental Algorithm (both by
+\cite{Dale1995}) and the relational algorithm by
+\cite{dale91:_gener_refer_expres_invol_relat} on this corpus.
 
 \begin{figure}
-\includegraphics[width=0.4\textwidth]{drawers}
+\includegraphics[width=0.3\textwidth]{drawers.jpg}
 \caption{A schematic view of the filing cabinets.\todo{needs nicer pic}}\label{fig:drawers}
@@
 As described above \todo{needs to be described}, Dale and Reiter's
-Incremental Algorithm is a special case of our algorithm, where no
-roles are used and a specific order is imposed on the predicates. We
-replicated Viethen and Dale's experiment, which tested the Incremental
-Algorithm on the drawer domain with all possible preference ordering
-given the four predicates
-\textsf{color, column, row}, and \textsf{corner}
+Incremental Algorithm is a special case of our \el\ algorithm, where
+no roles are used and a specific order is imposed on the
+predicates. We replicated Viethen and Dale's experiment, which tested
+the Incremental Algorithm on the drawer domain with all possible
+preference orderings given the four predicates $\mathsf{color, column,
+row}$, and $\mathsf{corner}$
 \cite{viethen06:_algor_for_gener_refer_expres}.  Unsurprisingly, we
@@
 
-\newcite{viethen06:_algor_for_gener_refer_expres} also tested Dale and
-Haddock's \shortcite{dale91:_gener_refer_expres_invol_relat}
-Relational Algorithm on the drawer domain. Surprisingly, it could not
-generate \textit{any} of the descriptions produced by humans. To
-evaluate our algorithm's use of relations, we tested it on a
-representation of the domain using the predicates \textsf{color} and
-\textsf{corner} and the roles \textsf{above, below, next to, left of},
-and \textsf{right of}. We did not include the \textsf{column} and
-\textsf{row} predicates because every drawer can be uniquely
-identified through those predicates. Since our algorithm has to
-consider predicates before relations, this would mean that no relation
-ever gets used. Like for the Incremental Algorithm, we tested all
-possible orderings of predicates and roles (with the restriction that
-all predicates come before all roles). For 10 of the 15 human-produced
-relational descriptions there is an ordering such that using this
-ordering our algorithm can generate this description. Figure
-\ref{fig:example_outputs} shows the output of our \el\ algorithm for
-one preference ordering of predicates. To cover 10 of the 15
-human-produced descriptions three different orderings are
-necessary. Of the five human-produced escriptions that the \el\
-algorithm can not generate, two mention column information, which we
-excluded from our domain, and three (which interestingly were all
-produced by the same person) are of the format \textit{the
-blue/pink/yellow/orange drawer above/below/next to/right of/left of
-the two blue/pink/yellow/orange drawers}. This would correspond to the
-following output from our algorithm: $C_1 \& \exists R . (C_2 \&
-\exists R . C_2)$ (where $C_i$ are color predicates and $R$ a
-role). While our algorithm does not generate descriptions that contain
-the same relation on two levels of embedding for these drawers, it
-does generate some other, simpler descriptions for them; for example,
-$\textsf{orange} \& \exists
-\textsf{left\_of} . (\textsf{pink}
-\& \textsf{corner})$ which could be rendered as
-\textit{the orange drawer left of the pink one in the corner}, or
-$\textsf{orange} \& \exists \textsf{below} . \textsf{yellow}$,
-which could be rendered as \textit{the orange drawer below the yellow
-drawer}.
-
-\todo{should we show some more examples? - ak}
-
-\begin{figure*}
+\begin{figure*}[thb]
 
 \begin{small}
-\begin{tabular}{|lp{0.5\textwidth}p{0.4\textwidth}|}\hline
-drawer & human-produce description & output of the \el\ algorithm\\\hline
-2 & \textit{the orange drawer above the blue drawer} & $\exists \textsf{above}.(\textsf{blue}) \& \textsf{orange}$\\
-4 & \textit{the yellow drawer on the top of the pink one} & $\exists \textsf{above}.(\textsf{pink}) \& \textsf{yellow}$\\
-5 & \textit{the pink drawer in the fourth column below the yellow one} & $\exists \textsf{above}.(\textsf{orange}) \& \textsf{pink}$\\
-6 & \textit{the yellow drawer on top of the yellow drawer} & $\exists \textsf{above}.(\textsf{yellow}) \& \textsf{yellow}$\\
-6 & \textit{the yellow drawer that's above another yellow drawer} & \\
-6 & \textit{the drawer after the two blue ones in horizontal sequence} & \\
-7 & \textit{the blue drawer below the orange drawer in the second column} & $\exists \textsf{next}.(\exists \textsf{above}.(\textsf{orange}) \& \textsf{blue}) \& \exists \textsf{above}.(\textsf{blue}) \& \textsf{blue}$\\
-7 & \textit{the blue drawer below the orange one} & \\
-10 & \textit{the blue drawer above the pink drawer} & $\exists \textsf{above}.(\textsf{pink}) \& \textsf{blue}$\\
-10 & \textit{the blue drawer above the pink drawer} & \\
-11 & \textit{the yellow drawer next to the orange drawer} & $\exists \textsf{above}.(\textsf{orange}) \& \textsf{yellow}$\\
-11 & \textit{the yellow one next to the orange one} & \\
-12 & \textit{the orange drawer below the pink drawer} & $\exists \textsf{above}.(\textsf{pink} \& \textsf{corner}) \& \textsf{orange}$\\
-14 & \textit{the orange drawer below the two yellow drawers} & $\exists \textsf{next}.(\textsf{pink} \& \textsf{corner}) \& \exists \textsf{next}.(\textsf{pink}) \& \textsf{orange}$\\
-14 & \textit{the orange drawer below the two yellow drawers} & \\\hline
+\begin{tabular}{|lp{0.9\textwidth}|}\hline
+\multirow{2}{*}{drawer} & \textit{human-produce description} \\
+& $\mathsf{output\ of\ the\ \el\ algorithm}$\\\hline
+2 & \textit{the orange drawer above the blue drawer} \\
+& (a) $\exists \mathsf{above}.(\mathsf{blue}) \sqcap \mathsf{orange}$ (b) $\exists \mathsf{above}.(\exists \mathsf{below}.(\mathsf{orange}) \sqcap \mathsf{blue}) \sqcap \mathsf{orange}$ (c) $\exists \mathsf{next}.(\mathsf{blue}) \sqcap \exists \mathsf{next}.(\mathsf{pink}) \sqcap \mathsf{orange}$\\
+4 & \textit{the yellow drawer on the top of the pink one} \\ 
+& (a) $\exists \mathsf{above}.(\mathsf{pink}) \sqcap \mathsf{yellow}$  (b) $\exists \mathsf{above}.(\mathsf{pink}) \sqcap \mathsf{corner} \sqcap \mathsf{green}$  (c) $\exists \mathsf{above}.(\exists \mathsf{next}.(\mathsf{green}) \sqcap \mathsf{pink}) \sqcap \mathsf{green} \sqcap \mathsf{corner}$\\
+5 & $\ast$ \textit{the pink drawer in the fourth column below the yellow one}\\
+& (a) $\exists \mathsf{above}.(\mathsf{orange}) \sqcap \mathsf{pink}$  (b) $\exists \mathsf{below}.(\mathsf{green}) \sqcap \mathsf{pink}$  (c) $\exists \mathsf{above}.(\exists \mathsf{next}.(\mathsf{green}) \sqcap \mathsf{orange}) \sqcap \exists \mathsf{next}.(\mathsf{green}) \sqcap \mathsf{pink}$\\
+%6 & \textit{the yellow drawer on top of the yellow drawer} / \textit{the yellow drawer that's above another yellow drawer} / $\ast$ \textit{the drawer after the two blue ones in horizontal sequence}\\
+6 & \textit{the yellow drawer on top of the yellow drawer} (2$\times$) / $\ast$ \textit{the drawer after the two blue ones in horizontal sequence}\\
+& (a) $\exists \mathsf{above}.(\mathsf{yellow}) \sqcap \mathsf{yellow}$  (b) $\exists \mathsf{below}.(\mathsf{pink}) \sqcap \mathsf{green}$  (c) $\exists \mathsf{next}.(\mathsf{blue}) \sqcap \exists \mathsf{next}.(\mathsf{pink}) \sqcap \mathsf{green}$\\
+7 & \textit{the blue drawer below the orange one} / $\ast$ \textit{the blue drawer below the orange drawer in the second column}\\ 
+& (a) $\exists \mathsf{next}.(\exists \mathsf{above}.(\mathsf{orange}) \sqcap \mathsf{blue}) \sqcap \exists \mathsf{above}.(\mathsf{blue}) \sqcap \mathsf{blue}$  (b) $\exists \mathsf{below}.(\mathsf{orange}) \sqcap \mathsf{blue}$  (c) $\exists \mathsf{next}.(\mathsf{blue}) \sqcap \exists \mathsf{next}.(\mathsf{green}) \sqcap \mathsf{blue}$\\
+10 & \textit{the blue drawer above the pink drawer} (2$\times$)\\
+& (a) $\exists \mathsf{above}.(\mathsf{pink}) \sqcap \mathsf{blue}$  (b) $\exists \mathsf{above}.(\mathsf{pink}) \sqcap \exists \mathsf{below}.(\mathsf{blue}) \sqcap \mathsf{blue}$  (c) $\exists \mathsf{next}.(\mathsf{orange}) \sqcap \exists \mathsf{next}.(\mathsf{green}) \sqcap \mathsf{blue}$\\
+%11 & \textit{the yellow drawer next to the orange drawer} / \textit{the yellow one next to the orange one}\\
+11 & \textit{the yellow drawer next to the orange drawer} (2$\times$)\\
+& (a) $\exists \mathsf{above}.(\mathsf{orange}) \sqcap \mathsf{yellow}$  (b) $\exists \mathsf{below}.(\mathsf{green}) \sqcap \mathsf{green}$  (c) $\exists \mathsf{next}.(\mathsf{orange}) \sqcap \mathsf{green}$\\
+12 & \textit{the orange drawer below the pink drawer}\\ 
+& (a) $\exists \mathsf{above}.(\mathsf{pink} \sqcap \mathsf{corner}) \sqcap \mathsf{orange}$  (b) $\exists \mathsf{below}.(\mathsf{pink}) \sqcap \mathsf{orange}$  (c) $\exists \mathsf{next}.(\mathsf{green}) \sqcap \mathsf{orange}$\\
+14 & $\ast$ \textit{the orange drawer below the two yellow drawers} (2$\times$)\\
+& (a) $\exists \mathsf{next}.(\mathsf{pink} \sqcap \mathsf{corner}) \sqcap \exists \mathsf{next}.(\mathsf{pink}) \sqcap \mathsf{orange}$  (b) $\exists \mathsf{below}.(\mathsf{green}) \sqcap \mathsf{orange}$  (c) $\exists \mathsf{next}.(\mathsf{pink} \sqcap \mathsf{corner}) \sqcap \mathsf{orange}$\\\hline
 \end{tabular}
@@
 
-\caption{Human produced descriptions and output of our \el\  algorithm using the ordering \textsf{corner}, \textsf{color}, \textsf{above, next, below, right, left}}
+\caption{Human produced descriptions and output of our \el\  algorithm
+using the orderings (a) $\mathsf{corner,}$ $\mathsf{color, above,
+next, below, right, left}$, (b) $\mathsf{color, corner, below, above,
+left, right, next}$ and (c) $\mathsf{corner, color, below}$
+$\mathsf{above, left, right, next}$. Descriptions that the \el\
+algorithm cannot generate with any ordering are marked by $\ast$.}
 \label{fig:example_outputs}
@@
 %List(above, next, below, right, left)
+
+% ordering no. 19
+%List(green, pink, orange, blue, corner)
+%List(below, above, left, right, next)
+
+% ordering no. 174
+% List(corner, green, pink, orange, blue)
+% List(next, above, below, left, right)
+
+\newcite{viethen06:_algor_for_gener_refer_expres} also tested Dale and
+Haddock's \shortcite{dale91:_gener_refer_expres_invol_relat}
+Relational Algorithm on the drawer domain. Surprisingly, it could not
+generate \textit{any} of the descriptions produced by humans. To
+evaluate our algorithm's use of relations, we tested it on a
+representation of the domain using the predicates $\mathsf{color}$ and
+$\mathsf{corner}$ and the roles $\mathsf{above, below, next\ to, left\ of}$,
+and $\mathsf{right\ of}$. We did not include the $\mathsf{column}$ and
+$\mathsf{row}$ predicates because every drawer can be uniquely
+identified through those predicates. Since our algorithm has to
+consider predicates before relations, this would mean that no relation
+ever gets used. Like for the Incremental Algorithm, we tested all
+possible orderings of predicates and roles (with the restriction that
+all predicates come before all roles). 10 of the 15 human-produced
+descriptions could thus be generated using one of the 240 different
+orderings. Figure \ref{fig:example_outputs} shows the output of our
+\el\ algorithm for three preference orderings. Together these three
+orderings cover 10 of the human-produced descriptions. 
+Of the five human-produced escriptions that the \el\
+algorithm can not generate, two mention column information, which we
+excluded from our domain, and three involve references to sets
+(\textit{the two blues ones in horizontal sequence}/\textit{the two
+yellow drawers}). While our algorithm cannot reproduce those
+descriptions, it does  generate other, simpler descriptions for them.
+
+If we were to use this algorithm in an application, we would of course
+have to make a decision on which ordering to use. Dale and Reiter
+\shortcite{Dale1995} seem to assume that there is one ordering for a
+given domain. The data from the drawer domain do not support this
+hypothesis: there is no single ordering that covers all human-produced
+descriptions. It is not even the case that one speaker uses only one
+ordering in this domain; even though some subjects show a clear trend
+towards certain orderings. An interesting open research question is,
+therefore, what factors determine which ordering is used.
+
+To further investigate this question both in the absence and in the
+presence of relational properties, we need more human-produced
+descriptions using relational properties. The grid structure of the
+drawer domain lent itself well to non-relational descriptions using
+row and column information and 103 out of 118 human-produced
+descriptions are indeed non-relational. The TUNA corpus
+\cite{deemter06:_build_seman_trans_corpus_for} used a similar
+grid-based mode of presentation and the results are similar in terms
+of the number of relational descriptions used. However, many domains
+do not have such an internal structure which makes the use of
+relations unnecessary. We are currently preparing an experiment to
+collect referring expression data similar to the drawer corpus but
+using a blocks world domain that will hopefully
 
@@
 
-\begin{figure}
+\begin{figure}[thb]
   \centering

### 2008-03-19 trunk/Papers/dl-gre-08/gre.tex
simplified the picture
Written the algorithm for ALC and changed the one for EL.

--- 2008-03-16 trunk/Papers/dl-gre-08/gre.tex
@@
 \label{algo:bisim-el}
-$\C \leftarrow \{\top\}$ \;
+\KwIn{A model $\gM = (\Delta^\gM, |\cdot|^\gM)$}
+\KwOut{A set \RE of pairs $(\varphi,|\varphi|^\gM)$ such that
+$\{S \mid (\varphi,S) \in \RE\}$ is the set of $\el$-bisimulation
+classes of $\gM$.}
+\RE $\leftarrow \{(\top,\Delta^\gM)\}$ \;
 \For{$p \in \prop$}{
-  add$_\el$(p, \C) \;
+  add$_\el$(p, \RE) \;
 }
-\While{not all concepts in $\C$ singleton}{
-  \For{$C \in \C$, $R \in \rel$}{
-    add$_\el$($\exists R.C$, $\C$) \;
+\While{for some $(\varphi,S) \in \RE$, $|S| > 1$}{
+  \For{$(\varphi,S) \in$ \RE, $R \in \rel$}{
+    $\RE \leftarrow$ add$_\el$($\exists R.\varphi$, \RE) \;
   }
-  \If{made no changes to $\C$}{
+  \If{made no changes to $\RE$}{
     exit\;
@@
 \dontprintsemicolon
-\caption{add$_\el$($C$, $\C$)}
+\caption{add$_\el$($\varphi$, $\RE$)}
 \label{algo:bisim-add-el}
-\For{$C' \in \C$ non-singleton}{
-  \If{$C' \sqcap C$ is not subsumed in $\C$}{
-    add $C' \sqcap C$ to $\C$ \;
-    remove subsumed concepts from $\C$\;
+\For{$(\psi,S) \in \RE$ with $|S| > 1$}{
+  \If{$\psi \sqcap \varphi$ is not subsumed in $\RE$ {\bf and}
+    $|\psi \sqcap \varphi|^\gM \neq \emptyset$ {\bf and}
+    $|\psi \sqcap \varphi|^\gM \neq |\psi|^\gM$}{
+    add $(\psi \sqcap \varphi,|\psi \sqcap \varphi|^\gM)$ to $\RE$ \;
+    remove subsumed concepts from $\RE$\;
   }
+}
+\end{algorithm}
+
+\begin{algorithm}[t]
+\dontprintsemicolon
+\caption{}
+\label{algo:bisim-alc}
+\KwIn{A model $\gM = (\Delta^\gM, |\cdot|^\gM)$}
+\KwOut{A set \RE of pairs $(\varphi,|\varphi|^\gM)$ such that
+$\{S \mid (\varphi,S) \in \RE\}$ is the set of $\alc$-bisimulation
+classes of $\gM$.}
+$\RE \leftarrow \{(\top,\Delta^\gM)\}$
+
+\For{$p \in \prop$}{
+   \For{$(\varphi,S) \in \RE$}{
+      split$((\varphi,S),p,\RE)$
+   }
+}
+
+\While{for some $(\varphi,S) \in \RE, |S|>1$}{
+   \For{$(\varphi,S) \in \RE, R \in \rel$}{
+      \For{any set $\{(\psi_1,S_1),\ldots,(\psi_n,S_n)\} \in \RE$}{
+         split$((\varphi,S),\exists R.\bigsqcup_i \psi_i,\RE)$
+         }
+   }
+   \If{made no changes to \RE}{
+      exit
+      }
+}
+
+\end{algorithm}
+
+\begin{algorithm}
+\caption{split$((\varphi,S),\psi, \RE)$}
+   \If{$S \cap |\psi|^\gM \not = \emptyset$ and
+    $S \backslash |\psi|^\gM \not = \emptyset$}{
+         add  $(\varphi \sqcap \psi     , S \cap |\psi|^\gM)$ and
+               $(\varphi \sqcap \neg \psi, S \backslash |\psi|^\gM)$ to \RE \;
+         delete $(\varphi,S)$ from \RE \;
       }
@@
 will compute the concepts $\{\neg \exists R. (\neg P \sqcap \neg Q),
-\exists R. (\neg P \sqcap \neg Q), Q, \neg P \sqcap \neg Q\}$ (each of
-which is singleton), but $a$ and $b$ are \el-bisimilar in this model,
+Q, \exists R. (\neg P \sqcap \neg Q), \neg P \sqcap \neg Q\}$ (each of
+which is singleton), but $a$ and $c$ are \el-bisimilar in this model,
 and the only \el-concept that is true in $d$ is $\top$, so the \el\

### 2008-03-19 trunk/Papers/dl-gre-08/discussion.tex
drawer pic

@@
 
-We will now describe two experiments evaluating different aspects of
-the performance of our algorithm. First, we will look at its running
-time on randomly generated domain graphs of different sizes. Second,
-we will compare the output of our algorithm to human-generated
-descriptions.
+We will now describe two experiments evaluating the quality of the \el\
+algorithm's output and the efficiency of both of our algorithms, and
+we discuss the interface between our algorithms and realization.
+
+%different aspects of
+%the performance of our algorithm. First, we will look at its running
+%time on randomly generated domain graphs of different sizes. Second,
+%we will compare the output of our algorithm to human-generated
+%descriptions.
 
@@
 \begin{figure}
-\includegraphics[width=0.3\textwidth]{drawers.jpg}
-\caption{A schematic view of the filing cabinets.\todo{needs nicer pic}}\label{fig:drawers}
+\begin{center}
+\includegraphics[width=0.25\textwidth]{drawers.png}
+\end{center}\vspace*{-3ex}
+\caption{A schematic view of the filing cabinets.}\label{fig:drawers}\vspace*{-3ex}
 \end{figure}
@@
 
-\begin{small}
-\begin{tabular}{|lp{0.9\textwidth}|}\hline
-\multirow{2}{*}{drawer} & \textit{human-produce description} \\
+\begin{footnotesize}
+\begin{tabular}{|l|p{0.95\textwidth}|}\hline
+\multirow{2}{*}{id} & \textit{human-produce description} \\
 & $\mathsf{output\ of\ the\ \el\ algorithm}$\\\hline
-2 & \textit{the orange drawer above the blue drawer} \\
-& (a) $\exists \mathsf{above}.(\mathsf{blue}) \sqcap \mathsf{orange}$ (b) $\exists \mathsf{above}.(\exists \mathsf{below}.(\mathsf{orange}) \sqcap \mathsf{blue}) \sqcap \mathsf{orange}$ (c) $\exists \mathsf{next}.(\mathsf{blue}) \sqcap \exists \mathsf{next}.(\mathsf{pink}) \sqcap \mathsf{orange}$\\
-4 & \textit{the yellow drawer on the top of the pink one} \\ 
-& (a) $\exists \mathsf{above}.(\mathsf{pink}) \sqcap \mathsf{yellow}$  (b) $\exists \mathsf{above}.(\mathsf{pink}) \sqcap \mathsf{corner} \sqcap \mathsf{green}$  (c) $\exists \mathsf{above}.(\exists \mathsf{next}.(\mathsf{green}) \sqcap \mathsf{pink}) \sqcap \mathsf{green} \sqcap \mathsf{corner}$\\
-5 & $\ast$ \textit{the pink drawer in the fourth column below the yellow one}\\
-& (a) $\exists \mathsf{above}.(\mathsf{orange}) \sqcap \mathsf{pink}$  (b) $\exists \mathsf{below}.(\mathsf{green}) \sqcap \mathsf{pink}$  (c) $\exists \mathsf{above}.(\exists \mathsf{next}.(\mathsf{green}) \sqcap \mathsf{orange}) \sqcap \exists \mathsf{next}.(\mathsf{green}) \sqcap \mathsf{pink}$\\
+\multirow{2}{*}{2} & \textit{the orange drawer above the blue drawer} \\
+&  $\bm{\mathsf{orange} \sqcap \exists \mathsf{above}.\mathsf{blue}}$ / $\mathsf{orange} \sqcap \exists \mathsf{above}.(\exists \mathsf{below}.(\mathsf{orange}) \sqcap \mathsf{blue})$ / $ \mathsf{orange} \sqcap \exists \mathsf{next}.(\mathsf{blue}) \sqcap \exists \mathsf{next}.(\mathsf{pink})$\\\hline
+\multirow{2}{*}{4} & \textit{the yellow drawer on the top of the pink one} \\ 
+&  $\bm{\mathsf{yellow} \sqcap \exists \mathsf{above}.\mathsf{pink}}$  / $\mathsf{yellow}  \sqcap \mathsf{corner} \sqcap \exists \mathsf{above}.\mathsf{pink}$  / $\mathsf{yellow} \sqcap \mathsf{corner} \sqcap \exists \mathsf{above}.(\exists \mathsf{next}.(\mathsf{yellow}) \sqcap \mathsf{pink})$\\\hline
+\multirow{2}{*}{5} & $\ast$ \textit{the pink drawer in the fourth column below the yellow one}\\
+&  $\mathsf{pink} \sqcap \exists \mathsf{above}.\mathsf{orange}$  / $\mathsf{pink} \sqcap \exists \mathsf{below}.\mathsf{yellow}$  / $ \mathsf{pink} \sqcap \exists \mathsf{next}.(\mathsf{yellow}) \sqcap \exists \mathsf{above}.(\exists \mathsf{next}.(\mathsf{yellow}) \sqcap \mathsf{orange})$\\\hline
 %6 & \textit{the yellow drawer on top of the yellow drawer} / \textit{the yellow drawer that's above another yellow drawer} / $\ast$ \textit{the drawer after the two blue ones in horizontal sequence}\\
-6 & \textit{the yellow drawer on top of the yellow drawer} (2$\times$) / $\ast$ \textit{the drawer after the two blue ones in horizontal sequence}\\
-& (a) $\exists \mathsf{above}.(\mathsf{yellow}) \sqcap \mathsf{yellow}$  (b) $\exists \mathsf{below}.(\mathsf{pink}) \sqcap \mathsf{green}$  (c) $\exists \mathsf{next}.(\mathsf{blue}) \sqcap \exists \mathsf{next}.(\mathsf{pink}) \sqcap \mathsf{green}$\\
-7 & \textit{the blue drawer below the orange one} / $\ast$ \textit{the blue drawer below the orange drawer in the second column}\\ 
-& (a) $\exists \mathsf{next}.(\exists \mathsf{above}.(\mathsf{orange}) \sqcap \mathsf{blue}) \sqcap \exists \mathsf{above}.(\mathsf{blue}) \sqcap \mathsf{blue}$  (b) $\exists \mathsf{below}.(\mathsf{orange}) \sqcap \mathsf{blue}$  (c) $\exists \mathsf{next}.(\mathsf{blue}) \sqcap \exists \mathsf{next}.(\mathsf{green}) \sqcap \mathsf{blue}$\\
-10 & \textit{the blue drawer above the pink drawer} (2$\times$)\\
-& (a) $\exists \mathsf{above}.(\mathsf{pink}) \sqcap \mathsf{blue}$  (b) $\exists \mathsf{above}.(\mathsf{pink}) \sqcap \exists \mathsf{below}.(\mathsf{blue}) \sqcap \mathsf{blue}$  (c) $\exists \mathsf{next}.(\mathsf{orange}) \sqcap \exists \mathsf{next}.(\mathsf{green}) \sqcap \mathsf{blue}$\\
+\multirow{2}{*}{6} & \textit{the yellow drawer on top of the yellow drawer} (2$\times$) / $\ast$ \textit{the drawer after the two blue ones in horizontal sequence}\\
+&  $\bm{\mathsf{yellow} \sqcap \exists \mathsf{above}.\mathsf{yellow}}$  / $\mathsf{yellow} \sqcap \exists \mathsf{below}.\mathsf{pink}$  / $\mathsf{yellow} \sqcap \exists \mathsf{next}.(\mathsf{blue}) \sqcap \exists \mathsf{next}.(\mathsf{pink})$\\\hline
+\multirow{2}{*}{7} & \textit{the blue drawer below the orange one} / $\ast$ \textit{the blue drawer below the orange drawer in the second column}\\ 
+& $\mathsf{blue} \sqcap \exists \mathsf{above}.(\mathsf{blue}) \sqcap \exists \mathsf{next}.(\exists \mathsf{above}.(\mathsf{orange}) \sqcap \mathsf{blue})$ / $\bm{\mathsf{blue} \sqcap \exists \mathsf{below}.(\mathsf{orange})}$ \textbf{/} $\mathsf{blue} \sqcap \exists \mathsf{next}.(\mathsf{blue}) \sqcap \exists \mathsf{next}.(\mathsf{yellow})$\\\hline
+\multirow{2}{*}{10} & \textit{the blue drawer above the pink drawer} (2$\times$)\\
+&  $\bm{\mathsf{blue} \sqcap \exists \mathsf{above}.(\mathsf{pink})}$  / $\mathsf{blue} \sqcap \exists \mathsf{above}.(\mathsf{pink}) \sqcap \exists \mathsf{below}.(\mathsf{blue})$  / $\mathsf{blue} \sqcap \exists \mathsf{next}.(\mathsf{orange}) \sqcap \exists \mathsf{next}.(\mathsf{yellow})$\\\hline
 %11 & \textit{the yellow drawer next to the orange drawer} / \textit{the yellow one next to the orange one}\\
-11 & \textit{the yellow drawer next to the orange drawer} (2$\times$)\\
-& (a) $\exists \mathsf{above}.(\mathsf{orange}) \sqcap \mathsf{yellow}$  (b) $\exists \mathsf{below}.(\mathsf{green}) \sqcap \mathsf{green}$  (c) $\exists \mathsf{next}.(\mathsf{orange}) \sqcap \mathsf{green}$\\
-12 & \textit{the orange drawer below the pink drawer}\\ 
-& (a) $\exists \mathsf{above}.(\mathsf{pink} \sqcap \mathsf{corner}) \sqcap \mathsf{orange}$  (b) $\exists \mathsf{below}.(\mathsf{pink}) \sqcap \mathsf{orange}$  (c) $\exists \mathsf{next}.(\mathsf{green}) \sqcap \mathsf{orange}$\\
-14 & $\ast$ \textit{the orange drawer below the two yellow drawers} (2$\times$)\\
-& (a) $\exists \mathsf{next}.(\mathsf{pink} \sqcap \mathsf{corner}) \sqcap \exists \mathsf{next}.(\mathsf{pink}) \sqcap \mathsf{orange}$  (b) $\exists \mathsf{below}.(\mathsf{green}) \sqcap \mathsf{orange}$  (c) $\exists \mathsf{next}.(\mathsf{pink} \sqcap \mathsf{corner}) \sqcap \mathsf{orange}$\\\hline
+\multirow{2}{*}{11} & \textit{the yellow drawer next to the orange drawer} (2$\times$)\\
+&  $\mathsf{yellow} \sqcap \exists \mathsf{above}.\mathsf{orange}$  / $\mathsf{yellow} \sqcap \exists \mathsf{below}.\mathsf{yellow}$  / $\bm{\mathsf{yellow} \sqcap \exists \mathsf{next}.\mathsf{orange}}$\\\hline
+\multirow{2}{*}{12} & \textit{the orange drawer below the pink drawer}\\ 
+&  $\mathsf{orange} \sqcap \exists \mathsf{above}.(\mathsf{pink} \sqcap \mathsf{corner})$  / $\bm{\mathsf{orange} \sqcap \exists \mathsf{below}.\mathsf{pink}}$  / $\mathsf{orange} \sqcap \exists \mathsf{next}.\mathsf{yellow}$\\\hline
+\multirow{2}{*}{14} & $\ast$ \textit{the orange drawer below the two yellow drawers} (2$\times$)\\
+&  $ \mathsf{orange} \sqcap \exists \mathsf{next}.(\mathsf{pink} \sqcap \mathsf{corner}) \sqcap \exists \mathsf{next}.(\mathsf{pink})$  / $\mathsf{orange} \sqcap \exists \mathsf{below}.\mathsf{yellow}$  / $\mathsf{orange} \sqcap \exists \mathsf{next}.(\mathsf{pink} \sqcap \mathsf{corner})$\\\hline
 \end{tabular}
-\end{small}
+\end{footnotesize}
 
-\caption{Human produced descriptions and output of our \el\  algorithm
-using the orderings (a) $\mathsf{corner,}$ $\mathsf{color, above,
-next, below, right, left}$, (b) $\mathsf{color, corner, below, above,
-left, right, next}$ and (c) $\mathsf{corner, color, below}$
-$\mathsf{above, left, right, next}$. Descriptions that the \el\
-algorithm cannot generate with any ordering are marked by $\ast$.}
+\caption{Human produced descriptions from
+\cite{viethen06:_algor_for_gener_refer_expres} for nine different drawers (id)
+ and output of our \el\ algorithm using three different orderings.
+ Descriptions that the \el\ algorithm cannot generate with any
+ ordering are marked by $\ast$. Generated descriptions that match one
+ produced by humans are in bold face. 2$\times$ indicates that the
+ same description was produced twice.}
+
 \label{fig:example_outputs}
 \end{figure*}
+
+%(a) $\mathsf{corner,}$ $\mathsf{color, above,
+%next, below, right, left}$, (b) $\mathsf{color, corner, below, above,
+%left, right, next}$ and (c) $\mathsf{corner, color, below}$
+%$\mathsf{above, left, right, next}$. Descriptions that the \el\
+%algorithm cannot generate with any ordering are marked by $\ast$.}
 
@@
 $\mathsf{row}$ predicates because every drawer can be uniquely
-identified through those predicates. Since our algorithm has to
-consider predicates before relations, this would mean that no relation
+identified through those predicates. Since our algorithm (for
+efficiency reasons)
+considers predicates before relations, this would mean that no relation
 ever gets used. Like for the Incremental Algorithm, we tested all

### 2008-03-20 trunk/Papers/dl-gre-08/bisim.tex
Written the algorithm for ALC and changed the one for EL.
Main changes in Section 2 and in the algorithms. Intro in section 2 completed.  Still working in the end of Section 2.  Section 3 only changed by addition of text that was in Section 2 before. I still have to work here.
In addition, added some references to the bib.

@@
 
-In computational linguistics, the generating referring expression problem (the GRE
-problem) can be intuitively presented as follows: given some information about
-a group of individuals (e.g., grounded facts about individuals in an ongoing dialogue),
-and a grammar, found a grammatically correct expression (usually a noun phrase) that uniquely identify a given individual.
+One of the main ideas we want to investigate in this paper is
+how we can use \emph{formulas} from description logics (see~\cite{baad:desc03}) to represent
+referring expressions (RE). And how we can use the notion of \emph{bisimulation}
+(see~\cite{blac:moda01}) to obtain very efficient algorithms for computing referring expressions.
 
-By taking a logic perspective, we can recast the GRE problem as an inference problem.
-We can think that the information about the group of individuals is provided in the form
-of a model $\gM$ of a given logical language $\gL$.  Now, the inference problem
-associated with the GRE problem will be to find a formula in $\gL$ such that it
-exacly identify a given individual $i$.  Formally, if $|\cdot|^\gM$ is the interpretation
-function for $\gL$ which for each formula in $\gL$ and for each model $\gM$ returns
-the set of elements in $\gM$ for which the formula $\varphi$ holds, then we can define
-\medskip
+In what follows we will introduce in detail two description logics called \el and \alc.  Intuitively, \el allows
+for conjunctions of atomic and existentially quantified relational expressions (e.g.,
+we can write formulas in \el that could be realized as ``a red book on a table''). \alc is the extension of \el with full negation
+(e.g., we can write formulas in \alc which could be realized as ``a book which is not red and is not on the floor'').
+Computing bisimulations in \alc is very efficient, but \alc can contain expressions which are difficult to realize (because of the presence of negation).
+On the other hand,
+\el expressions are easily realizable, but the bisimulation algorithm is more complex.
 
-\noindent
-{\small
-\begin{tabular}{l} \hline
-\textsc{$\gL$-GRE Problem}\\ \hline
-\ \ \ Input: A model $\gM$, an individual $i$ in the domain of $\gM$\\
-\ \ \ Output: A formula $\varphi \in \gL$ such that $|\varphi|^\gM = \{i\}$\\
-\hspace*{1.4cm} (if such a formula exists).\\ \hline
-\end{tabular}}
-\medskip
-
-As it can be seen from the definition, the problem depends on which particular
-logical language $\gL$ we are interested in, and (as the original GRE problem)
-it might terminate with failure.  Depending on the model $\gM$, the chosen individual $i$
-and the expressive power of $\gL$ it can be the case that there is no formula of
-$\gL$ that uniquely identifies $i$.
-
-Consider the following example.  We have three individuals related as indicated in
-Figure~\ref{fig-car1}.
-\begin{figure}
-\begin{center}
-\includegraphics[scale=.8]{pic.pdf}
-\end{center}
-\caption{Distinguishable/indistinguishable points.}\label{fig-car1}
-\end{figure}
-%
-The individuals $i$ and $i'$ can easily be distinguished in first-order logic:
-the formula $R(x,x)$ is true of $i$ and false in $i'$. The corresponding
-referring expression for $i$ would be ``an individual relating to itself''.
-But if we can only use propositional and relational information (equivalently, our grammar does not contain the word ``itself'') then $i$ and $i'$ are indistinguishable. Both
-are ``individuals which are related to individuals which are related to individuals\ldots''
-
-In what follows we will introduce two logical languages (called \el and \alc, see~\cite{baad:desc03}) which can naturally be mapped to standard grammatically correct expressions.  \el allows
-for conjunctions of atomic and existentially quantified relational expressions (e.g.,
-``a red book on a table'') while \alc is the extension of \el with full negation
-(e.g., ``a book which is not red and is not on the floor'').
+But let start by formally defining the description languages we are going to use, and the notion of bisimulation.
 
@@
 
-We can now exactly characterize the problem of when a formula distinguishing two
-individuals in a given model exists. We will do so using the notion of \emph{bisimulation}.
+The previous definition formally introduces the syntax and semantics of the
+logics \alc and \el.  In the paper we will sometime refer to some other description logics (mainly in Section~\ref{xxx}).  For completeness we mention them briefly here,
+and we refer the reader to~\cite{baad:desc03} for further details:
+
+\begin{center}
+\begin{tabular}{lp{6cm}}
+$\mathcal{CL}$ & is obtained from $\el$ by dropping existential quantification\\
+$\mathcal{PL}$ & is obtained form $\alc$ by dropping existential quantification\\
+$\mathcal{ELU}_{(\neg)}$ & is obtained from $\el$ by adding disjunction and atomic negation
+\end{tabular}
+\end{center}
+
+\noindent
+It should be clear from the definitions above that $\mathcal{CL}$ is conjunctive logic, while $\mathcal{PL}$ is propositional logic.
+
+We now introduce the notion of \emph{bisimulation}.  We will discuss in detail
+the case for the $\alc$ language.  Further details can be found
+in~\cite{blac:moda01,kurt:expr99}.
 
@@
 
-\begin{theorem}[\cite{blac:moda01}]\label{bisim}
+\begin{theorem}[]\label{bisim}
 Given a model $\gM$, if  $i$ and $i'$ are two $\alc$-bisimilar elements in $\gM$

### 2008-03-20 trunk/Papers/dl-gre-08/gre.tex
Main changes in Section 2 and in the algorithms. Intro in section 2 completed.  Still working in the end of Section 2.  Section 3 only changed by addition of text that was in Section 2 before. I still have to work here.
In addition, added some references to the bib.

--- 2008-03-16 trunk/Papers/dl-gre-08/gre.tex
@@
 \section{Generating referring expressions} \label{sec:gre}
+
+In computational linguistics, the generating referring expression problem (the GRE
+problem) can be intuitively presented as follows: given some information about
+a group of individuals (e.g., grounded facts about individuals in an ongoing dialogue),
+and a grammar, found a grammatically correct expression (usually a noun phrase) that uniquely identify a given individual.
+
+By taking a logic perspective, we can recast the GRE problem as an inference problem.
+We can think that the information about the group of individuals is provided in the form
+of a model $\gM$ of a given logical language $\gL$.  Now, the inference problem
+associated with the GRE problem will be to find a formula in $\gL$ such that it
+exaclty identifies a given individual $i$.  Formally, if $|\cdot|^\gM$ is the interpretation
+function for $\gL$ which for each formula in $\gL$ and for each model $\gM$ returns
+the set of elements in $\gM$ for which the formula $\varphi$ holds, then we can define:
+\medskip
+
+\noindent
+{\small
+\begin{center}
+\begin{tabular}{ll} \hline
+\multicolumn{2}{l}{
+\textsc{$\gL$-GRE Problem}}\\ \hline
+\ \ Input: & A model $\gM$ and an individual $i$ in the\\
+& \hspace*{0.5cm} domain of $\gM$.\\
+\ \ Output: & A formula $\varphi \in \gL$ such that $|\varphi|^\gM = \{i\}$\\
+& \hspace*{0.5cm} (if such a formula exists).\\ \hline
+\end{tabular}
+\end{center}}
+\medskip
+
+As it can be seen from the definition, the problem depends on which particular
+logical language $\gL$ we are interested in, and (as the original GRE problem)
+it might terminate with failure.  Depending on \emph{a)} the model $\gM$,
+\emph{b)} the chosen individual $i$
+and \emph{c} the expressive power of $\gL$, it can be the case that there is no formula of
+$\gL$ that uniquely identifies $i$ in $\gM$.
+
+Consider the following example.  We have three individuals related as indicated in
+Figure~\ref{fig-car1}.
+\begin{figure}
+\begin{center}
+\includegraphics[scale=.8]{pic.pdf}
+\end{center}
+\caption{Distinguishable/indistinguishable points.}\label{fig-car1}
+\end{figure}
+%
+The individuals $i$ and $i'$ can easily be distinguished in first-order logic:
+the formula $R(x,x)$ is true of $i$ and false in $i'$. The corresponding
+referring expression for $i$ would be ``an individual relating to itself''.
+But if we can only use propositional and relational information (equivalently, our grammar does not contain the word ``itself'') then $i$ and $i'$ are indistinguishable. Both
+are ``individuals which are related to individuals which are related to individuals\ldots''
 
@@
 
+% \begin{algorithm}[t]
+% \dontprintsemicolon
+% \caption{\el\ bisimulation classes}
+% \label{algo:bisim-el}
+% \KwIn{A model $\gM = (\Delta^\gM, |\cdot|^\gM)$}
+% \KwOut{A set \RE of pairs $(\varphi,|\varphi|^\gM)$ such that
+% $\{S \mid (\varphi,S) \in \RE\}$ is the set of $\el$-bisimulation
+% classes of $\gM$.}
+% \RE $\leftarrow \{(\top,\Delta^\gM)\}$ \;
+% \For{$p \in \prop$}{
+%   add$_\el$(p, \RE) \;
+% }
+% \While{for some $(\varphi,S) \in \RE$, $|S| > 1$}{
+%   \For{$(\varphi,S) \in$ \RE, $R \in \rel$}{
+%     add$_\el$($\exists R.\varphi$, \RE) \;
+%   }
+%   \If{made no changes to $\RE$}{
+%     exit\;
+%   }
+% }
+% \end{algorithm}
+
 \begin{algorithm}[t]
 \dontprintsemicolon
-\caption{\el\ bisimulation classes}
-\label{algo:bisim-el}
-$\C \leftarrow \{\top\}$ \;
+\caption{}
+\label{algo:bisim-alc}
+\KwIn{A model $\gM = (\Delta^\gM, |\cdot|^\gM)$}
+\KwOut{A set \RE of formulas  such that
+$\{|\varphi|^\gM \in \RE\}$ is the set of $\mathcal{L}$-bisimulation
+classes of $\gM$.}
+
+$\RE \leftarrow \{\top\}$
+
 \For{$p \in \prop$}{
-  add$_\el$(p, \C) \;
+      add$_\mathcal{L}(p,\RE)$
    }
-\While{not all concepts in $\C$ singleton}{
-  \For{$C \in \C$, $R \in \rel$}{
-    add$_\el$($\exists R.C$, $\C$) \;
+
+\While{for some $\varphi \in \RE, |\varphi|^\gM>1$}{
+   \For{$\varphi \in \RE, R \in \rel$}{
+         add$_\mathcal{L}(\exists R.\varphi,\RE)$
    }
-  \If{made no changes to $\C$}{
-    exit\;
+   \If{made no changes to \RE}{
+      exit
+      }
+}
+
+\end{algorithm}
+
+\begin{algorithm}
+\caption{add$_\alc(\varphi,\RE)$}
+\For{$\psi \in \RE$ with $|\psi|^\gM > 1$}{
+   \If{$|\psi \sqcap \varphi|^\gM \not = \emptyset$ and
+       $|\psi \sqcap \neg \varphi|^\gM \not = \emptyset$}{
+         add $\psi \sqcap \varphi$ and
+               $\psi \sqcap \neg \varphi$ to \RE \;
+         remove $\psi$ from \RE \;
       }
@@
 \dontprintsemicolon
-\caption{add$_\el$($C$, $\C$)}
+\caption{add$_\el$($\varphi$, $\RE$)}
 \label{algo:bisim-add-el}
-\For{$C' \in \C$ non-singleton}{
-  \If{$C' \sqcap C$ is not subsumed in $\C$}{
-    add $C' \sqcap C$ to $\C$ \;
-    remove subsumed concepts from $\C$\;
+\For{$\psi \in \RE$ with $|\psi|^\gM > 1$}{
+  \If{$\psi \sqcap \varphi$ is not subsumed in $\RE$ {\bf and}
+    $|\psi \sqcap \varphi|^\gM \neq \emptyset$ {\bf and}
+    $|\psi \sqcap \varphi|^\gM \neq |\psi|^\gM$}{
+    add $\psi \sqcap \varphi$ to $\RE$ \;
+    remove subsumed concepts from $\RE$\;
   }

### 2008-03-20 trunk/Papers/dl-gre-08/conclusion.tex
iterated discussion; moved some of it into conclusion

@@
 
+One finding in our experiments with the Viethen and Dale dataset is
+that there is no single ordering that covers all human-produced
+description, which seems to be in contrast to Dale and Reiter's
+\shortcite{Dale1995} assumption that there is only one ordering for
+each given domain.  In fact, it is not even the case that each speaker
+consistently uses just one ordering.  An interesting open research
+question is thus what factors determine which ordering is used.
+Unfortunately, both in the Viethen and Dale dataset and in the TUNA
+corpus \cite{deemter06:_build_seman_trans_corpus_for}, only a minority
+of referring expressions is relational, maybe because these domains
+lend themselves very well to row/column style propositional REs.  We
+are currently preparing an experiment to collect REs in a domain in
+which relational REs will be more frequent.
+
 %%% Local Variables: 

### 2008-03-20 trunk/Papers/dl-gre-08/discussion.tex
iterated discussion; moved some of it into conclusion

--- 2008-03-18 trunk/Papers/dl-gre-08/discussion.tex
@@
 
-We will now describe two experiments evaluating different aspects of
-the performance of our algorithm. First, we will look at its running
-time on randomly generated domain graphs of different sizes. Second,
-we will compare the output of our algorithm to human-generated
-descriptions.
+We will now describe two experiments evaluating the quality of the \el\
+algorithm's output and the efficiency of both of our algorithms, and
+we discuss the interface between our algorithms and realization.
 
-\subsection{Quality of Output}
+%different aspects of
+%the performance of our algorithm. First, we will look at its running
+%time on randomly generated domain graphs of different sizes. Second,
+%we will compare the output of our algorithm to human-generated
+%descriptions.
 
-To compare the descriptions generated by our algorithm to
-those humans produce, we use a corpus of
-human-generated referring expressions collected and made available by
-Jette Viethen and Robert
-Dale\footnote{http://www.ics.mq.edu.au/~jviethen/drawers}.  The data
-was collected in an experiment where participants were asked to
-describe one of 16 filing cabinet drawers. The drawers had different
-colors and were arranged in a four-by-four grid, as shown in Figure
-\ref{fig:drawers}. The human-generated descriptions use four
-non-relational properties (the drawer's \textsf{color}, its \textsf{column} and
-\textsf{row} number, and whether it is in a \textsf{corner}) and five
-relational properties (\textsf{above, below, next to, left of, right
-of}). Of the 118 referring expressions, only 15 use relational properties.
-Viethen and Dale \shortcite{viethen06:_algor_for_gener_refer_expres}
-describe the data in more detail and present results of evaluating the
-Full Brevity algorithm by \cite{}, the Incremental Algorithm by
-\cite{} and the relational algorithm by \cite{} on this corpus.
+\subsection{Evaluation: Output Quality}
+
+To compare the descriptions generated by our algorithm to those humans
+produce, we use a corpus of human-generated referring expressions
+collected and made available by Jette Viethen and Robert
+Dale.\footnote{http://www.ics.mq.edu.au/\~{}jviethen/drawers}  They
+asked human subjects to describe one of 16 filing cabinet drawers. The
+drawers had different colors and were arranged in a four-by-four grid
+(see Fig.~\ref{fig:drawers}). The human subjects used seven
+non-relational properties (the drawer's color, its $\mathsf{column}$
+and $\mathsf{row}$ number, and whether it is in a $\mathsf{corner}$)
+and five relational properties ($\mathsf{above, below, next\ to, left\
+  of, right\ of}$). Of the 118 referring expressions, only 15 use
+relations.
 
 \begin{figure}
-\includegraphics[width=0.4\textwidth]{drawers}
-\caption{A schematic view of the filing cabinets.\todo{needs nicer pic}}\label{fig:drawers}
+\begin{center}
+\includegraphics[width=0.25\textwidth]{drawers.png}
+\end{center}\vspace*{-3ex}
+\caption{A schematic view of the filing cabinets.}\label{fig:drawers}\vspace*{-3ex}
 \end{figure}
 
-As described above \todo{needs to be described}, Dale and Reiter's
-Incremental Algorithm is a special case of our algorithm, where no
-roles are used and a specific order is imposed on the predicates. We
-replicated Viethen and Dale's experiment, which tested the Incremental
-Algorithm on the drawer domain with all possible preference ordering
-given the four predicates
-\textsf{color, column, row}, and \textsf{corner}
-\cite{viethen06:_algor_for_gener_refer_expres}.  Unsurprisingly, we
-get the same results as they did: 98 of the 103 non-relational
-descriptions produced by humans can be generated using one of the 24
-possible preference orderings. Unfortunately, there isn't one ordering
-that covers all those 98 descriptions - a set of four different
-orderings is necessary.
+\newcite{viethen06:_algor_for_gener_refer_expres} describe the data in
+more detail and present results of evaluating the Full Brevity
+algorithm, the Incremental Algorithm (both by \newcite{Dale1995}), and
+the relational algorithm \cite{dale91:_gener_refer_expres_invol_relat}
+on this corpus.  Because all these algorithms are sensitive to a
+predefined ordering in which predicates and relations are added to the
+descriptions, their evaluation methodology is to try all possible
+orderings and evaluate what percentage of descriptions an algorithm
+can generate with any ordering.  It turned out that the Incremental
+Algorithm is capable of generating 98 of the 103 non-relational
+descriptions (using four different orderings to cover all the data).
+However, the Dale and Haddock algorithm was unable to generate even a
+single one of the human-generated relational descriptions.
 
-\newcite{viethen06:_algor_for_gener_refer_expres} also tested Dale and
-Haddock's \shortcite{dale91:_gener_refer_expres_invol_relat}
-Relational Algorithm on the drawer domain. Surprisingly, it could not
-generate \textit{any} of the descriptions produced by humans. To
-evaluate our algorithm's use of relations, we tested it on a
-representation of the domain using the predicates \textsf{color} and
-\textsf{corner} and the roles \textsf{above, below, next to, left of},
-and \textsf{right of}. We did not include the \textsf{column} and
-\textsf{row} predicates because every drawer can be uniquely
-identified through those predicates. Since our algorithm has to
-consider predicates before relations, this would mean that no relation
-ever gets used. Like for the Incremental Algorithm, we tested all
-possible orderings of predicates and roles (with the restriction that
-all predicates come before all roles). For 10 of the 15 human-produced
-relational descriptions there is an ordering such that using this
-ordering our algorithm can generate this description. Figure
-\ref{fig:example_outputs} shows the output of our \el\ algorithm for
-one preference ordering of predicates. To cover 10 of the 15
-human-produced descriptions three different orderings are
-necessary. Of the five human-produced escriptions that the \el\
-algorithm can not generate, two mention column information, which we
-excluded from our domain, and three (which interestingly were all
-produced by the same person) are of the format \textit{the
-blue/pink/yellow/orange drawer above/below/next to/right of/left of
-the two blue/pink/yellow/orange drawers}. This would correspond to the
-following output from our algorithm: $C_1 \& \exists R . (C_2 \&
-\exists R . C_2)$ (where $C_i$ are color predicates and $R$ a
-role). While our algorithm does not generate descriptions that contain
-the same relation on two levels of embedding for these drawers, it
-does generate some other, simpler descriptions for them; for example,
-$\textsf{orange} \& \exists
-\textsf{left\_of} . (\textsf{pink}
-\& \textsf{corner})$ which could be rendered as
-\textit{the orange drawer left of the pink one in the corner}, or
-$\textsf{orange} \& \exists \textsf{below} . \textsf{yellow}$,
-which could be rendered as \textit{the orange drawer below the yellow
-drawer}.
+\input example-outputs
 
-\todo{should we show some more examples? - ak}
+We replicated Viethen and Dale's experiment for the \el\ algorithm
+presented above.  In the non-relational case, our algorithm generates
+exactly the same 98 correct descriptions as the Incremental Algorithm;
+this is because it performs essentially the same computation if there
+are no relations.  To evaluate the relational case, we first removed
+the $\mathsf{column}$ and $\mathsf{row}$ predicates from the domain,
+because these predicates make it possible to refer to each drawer with
+non-relational descriptions, and our algorithm would always prefer
+these.  Using three (of 240) different orderings, our algorithm was
+able to generate 10 of the 15 human-produced relational descriptions
+correctly.  This is shown in more detail in
+Fig.~\ref{fig:example_outputs}; notice that four of the 15
+descriptions occurred twice in the corpus.  Of the five human-produced
+escriptions that the \el\ algorithm cannot generate, two mention
+column information, which we excluded from our domain, and three
+involve references to sets (\textit{the two blues ones in horizontal
+  sequence}/\textit{the two yellow drawers}). While our algorithm
+cannot reproduce those descriptions, it does generate other, simpler
+descriptions for them.
 
-\begin{figure*}
-
-\begin{small}
-\begin{tabular}{|lp{0.5\textwidth}p{0.4\textwidth}|}\hline
-drawer & human-produce description & output of the \el\ algorithm\\\hline
-2 & \textit{the orange drawer above the blue drawer} & $\exists \textsf{above}.(\textsf{blue}) \& \textsf{orange}$\\
-4 & \textit{the yellow drawer on the top of the pink one} & $\exists \textsf{above}.(\textsf{pink}) \& \textsf{yellow}$\\
-5 & \textit{the pink drawer in the fourth column below the yellow one} & $\exists \textsf{above}.(\textsf{orange}) \& \textsf{pink}$\\
-6 & \textit{the yellow drawer on top of the yellow drawer} & $\exists \textsf{above}.(\textsf{yellow}) \& \textsf{yellow}$\\
-6 & \textit{the yellow drawer that's above another yellow drawer} & \\
-6 & \textit{the drawer after the two blue ones in horizontal sequence} & \\
-7 & \textit{the blue drawer below the orange drawer in the second column} & $\exists \textsf{next}.(\exists \textsf{above}.(\textsf{orange}) \& \textsf{blue}) \& \exists \textsf{above}.(\textsf{blue}) \& \textsf{blue}$\\
-7 & \textit{the blue drawer below the orange one} & \\
-10 & \textit{the blue drawer above the pink drawer} & $\exists \textsf{above}.(\textsf{pink}) \& \textsf{blue}$\\
-10 & \textit{the blue drawer above the pink drawer} & \\
-11 & \textit{the yellow drawer next to the orange drawer} & $\exists \textsf{above}.(\textsf{orange}) \& \textsf{yellow}$\\
-11 & \textit{the yellow one next to the orange one} & \\
-12 & \textit{the orange drawer below the pink drawer} & $\exists \textsf{above}.(\textsf{pink} \& \textsf{corner}) \& \textsf{orange}$\\
-14 & \textit{the orange drawer below the two yellow drawers} & $\exists \textsf{next}.(\textsf{pink} \& \textsf{corner}) \& \exists \textsf{next}.(\textsf{pink}) \& \textsf{orange}$\\
-14 & \textit{the orange drawer below the two yellow drawers} & \\\hline
-\end{tabular}
-\end{small}
-
-\caption{Human produced descriptions and output of our \el\  algorithm using the ordering \textsf{corner}, \textsf{color}, \textsf{above, next, below, right, left}}
-\label{fig:example_outputs}
-\end{figure*}
-
-% ordering no. 235
-%List(corner, green, pink, orange, blue)
-%List(above, next, below, right, left)
-
-\subsection{Efficiency}
+\subsection{Evaluation: Efficiency}
 
@@
 
-\begin{figure}
+\begin{figure}[t]
   \centering
@@
 their freedom to build concepts without having to respect any
-linguistic constraints.  It seems straightforward to extend Krahmer et
-al.'s \shortcite{Krahmer2003} approach such that it only considers
-subgraphs that can actually be realized, because their algorithm
-proceeds by a genuine search for uniquely identifying subgraphs, and
-will simply take a different branch of the search if some subgraph is
-useless.  Our algorithms don't search in the same way; any constraints
-that disallow certain refinements of a partition jeopardize the
-algorithms' ability to find a partition into singletons in polynomial
-time.  A particularly vivid example is the fact that both algorithms
-crucially split over all propositional symbols before they start
-iterating over relational symbols; changing this may make them
-incomplete.  Investigating this interplay between efficiency and
-linguistic constraints is an interesting avenue for future research.
+linguistic constraints.  It seems straightforward, for instance, to
+extend Krahmer et al.'s \shortcite{Krahmer2003} approach such that it
+only considers subgraphs that can actually be realized, because their
+algorithm proceeds by a genuine search for uniquely identifying
+subgraphs, and will simply take a different branch of the search if
+some subgraph is useless.  This would be harder in our case.  Our
+algorithms don't search in the same way; if we disallow certain
+refinements of a partition, we have to allow the algorithms to
+backtrack and thus jeopardize the worst-case polynomial runtime.
+Investigating this interplay between efficiency and linguistic
+constraints is an interesting avenue for future research.
 

### 2008-03-20 trunk/Papers/dl-gre-08/conclusion.tex
iterated 5-6

--- 2008-02-05 trunk/Papers/dl-gre-08/bisim.tex
@@
-\section{Computing bisimulation classes} \label{sec:bisim}
+\section{Conclusion} \label{sec:conclusion}
+
+In this paper, we have presented a new perspective on GRE as the
+problem of computing concepts of description logic with given
+extensions.  We have also shown how such concepts can be computed
+efficiently by applying existing algorithms for computing bisimulation
+classes.  Our algorithms are able to generate 95\% of the
+non-relational and 67\% of the relational REs from
+\newcite{viethen06:_algor_for_gener_refer_expres}.  They are also
+extremely efficient (140 ms to generate REs for all individuals in a
+random model of size 100); to our knowledge, these are by far the
+fastest runtimes for relational GRE reported in the literature.  We
+will make our implementations available online.
+
+Because they compute referring expressions for all individuals in the
+domain at once, our algorithms will perform especially strongly in
+static settings, such as the generation of descriptions for museum
+exhibits, in which the individuals and their properties don't change
+much.  In such a setting, REs could be computed once when the
+application starts, and then reused throughout.  Nevertheless, they
+remain fast enough for real-time use in dynamic settings.  One
+interesting question for future research is how bisimulation
+algorithms that can incrementally update the bisimulation classes when
+the model changes, such as \todo{cite}, can be adapted to GRE.
+
+One finding in our experiments with the Viethen and Dale dataset is
+that there is no single ordering that covers all human-produced
+descriptions, which seems to be in contrast to Dale and Reiter's
+\shortcite{Dale1995} assumption that there is only one ordering for
+each given domain.  In fact, it is not even the case that each speaker
+consistently uses just one ordering.  An interesting open research
+question is thus what factors determine which ordering is used.
+Unfortunately, both in the Viethen and Dale dataset and in the TUNA
+corpus \cite{deemter06:_build_seman_trans_corpus_for}, only a minority
+of referring expressions is relational, maybe because these domains
+lend themselves very well to row/column style propositional REs.  We
+are currently preparing an experiment to collect REs in a domain in
+which relational REs will be more frequent.
 

### 2008-03-20 trunk/Papers/dl-gre-08/related.tex
iterated 5-6

--- 2008-02-13 trunk/Papers/dl-gre-08/related.tex
@@
-\section{Related Work} \label{sec:related}
+\section{A unified perspective on GRE} \label{sec:related}
 
-\cite{Dale1995}
+\begin{figure}
+  \centering
+  \begin{tabular}{l|l}
+    GRE algorithm & DL variant \\ \hline
+    \newcite{Dale1995} & $\mathcal{CL}$ \\
+    \newcite{deemter01:_gener_refer_expres} & $\mathcal{PL}$ \\
+    \newcite{dale91:_gener_refer_expres_invol_relat} & \el \\
+    \newcite{Krahmer2003} & \el \\
+    \newcite{kelleher06:_increm_gener_of_spatial_refer} & \el \\
+    \newcite{gardent02:_gener_minim_defin_descr} & $\mathcal{ELU}_{(\neg)}$
+  \end{tabular}
+  \caption{DL variants used by different GRE algorithms.}
+  \label{fig:related}
+\end{figure}
 
-\cite{deemter01:_gener_refer_expres}
+Viewing GRE as a problem of generating DL concepts offers a new,
+unified perspective of the GRE problem: It is the problem of computing
+a DL concept with a certain (typically singleton) extension.  Many
+existing approaches can be subsumed under this view; we have
+summarized this for some of them in Fig.~\ref{fig:related}, along with
+the DL fragment they use.  We already discussed some of these
+approaches in Section~\ref{sec:gre}.  Furthermore, the graphs
+\newcite{Krahmer2003} use to represent referring expressions can be
+seen as concepts of \el\ that are only satisfied at those points at
+which they can be embedded; the non-relational but negative and
+disjunctive descriptions generated by
+\newcite{deemter01:_gener_refer_expres} are simply concepts of
+$\mathcal{PL}$; and \newcite{gardent02:_gener_minim_defin_descr} then
+generalizes this into generating concepts of $\mathcal{ELU}_{(\neg)}$,
+i.e.\ \el\ plus disjunction and atomic negation.
 
-\cite{dale91:_gener_refer_expres_invol_relat} -- avoid infinite
-regress by requiring that each semantic atom can be used only once in
-the description, i.e.\ no \emph{edge} may be used twice.
+The approach presented here fits very well into this landscape, and in
+a sense completes the picture by showing how to generate concepts for
+arbitrary bisimulation classes in \alc, which combines all connectives
+used in any of these previous approaches.  Where our approach breaks
+new ground is in the way these concepts are computed: It successively
+refines a decomposition of the domain into subsets.  In this way, it
+is reminiscent of the Incremental Algorithm, and in fact the IA can be
+seen as the special case of the \el\ algorithm for the non-relational
+case.
 
-\cite{gardent02:_gener_minim_defin_descr}
-
-\cite{Krahmer2003} -- avoid infinite regress automatically because
-it doesn't count as progress to add an edge to the subgraph that was
-already in it, i.e.\ cleaner version of Dale and Haddock heuristic.
-Subgraphs can be seen as DL concepts made up of conjunction and
-existential quantifiers (no negation).
-
-\cite{kelleher06:_increm_gener_of_spatial_refer} -- generate GREs with
-spatial relations. Polynomial algorithm (check this). Generates REs
-for landmarks recursively, i.e.\ suffer from Dale and Haddock infinite
-regress in principle. Avoid this by using each individual only once in
-the entire description, i.e.\ no \emph{node} may be used
-twice. Advantage over ours: take salience of landmarks into account.
+However, our approach is in contrast with
+\newcite{dale91:_gener_refer_expres_invol_relat} and its successors,
+such as \newcite{kelleher06:_increm_gener_of_spatial_refer}, which
+attempt to generate a RE for a single individual for successive
+individuals in the model and therefore run a risk of infinite regress.
+Our algorithms are guaranteed to terminate (and with concepts of the
+smallest relational depth that cover all the data), and infinite
+regress is automatically excluded.  Perhaps closest in spirit is the
+Krahmer et al.\ graph algorithm, which also computes \el\ concepts by
+extending them successively; but this algorithm, too, is focused on
+reference to a single individual, and it has NP-complete worst-case
+complexity.
 

### 2008-03-20 trunk/Papers/dl-gre-08/gre.tex
improved pictures

@@
 Consider the following example.  We have three individuals related as indicated in
-Figure~\ref{fig-car1}.
-\begin{figure}
-\begin{center}
-\includegraphics[scale=.8]{pic.pdf}
-\end{center}
-\caption{Distinguishable/indistinguishable points.}\label{fig-car1}
-\end{figure}
+Figure~\ref{fig:dale-haddock}d.
+
+%\begin{figure}
+%\begin{center}
+%\includegraphics[scale=.8]{pic.pdf}
+%\end{center}
+%\caption{Distinguishable/indistinguishable points.}\label{fig-car1}
+%\end{figure}
 %
@@
 
-\begin{figure}
+\begin{figure*}
   \centering
-  \includegraphics[width=5cm]{pic-dale-haddock}
-  \caption{The Dale and Haddock scenario.}
+  \includegraphics[width=\textwidth]{pic-dale-haddock}
+  \caption{(a) The \newcite{dale91:_gener_refer_expres_invol_relat}
+    scenario; (b) the \newcite{Stone1998a} scenario; (c) illustrating
+    the difference between \el\ and \alc; (d)
+    distinguishable/indistinguishable points.}
   \label{fig:dale-haddock}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\columnwidth]{pic-el-vs-alc}
-  \caption{Illustrating the difference between \el\ and \alc.}
-  \label{fig:el-vs-alc}
-\end{figure}
+% \begin{figure}
+%   \centering
+%   \includegraphics[width=\columnwidth]{pic-el-vs-alc}
+%   \caption{Illustrating the difference between \el\ and \alc.}
+%   \label{fig:el-vs-alc}
+% \end{figure}
 

### 2008-03-20 trunk/Papers/dl-gre-08/discussion.tex
removed other SPUD reference

@@
 with every other approach that separates these two modules, i.e.\
-almost the entire GRE literature (notable exceptions are e.g.\ SPUD
-\cite{Stone2003a} and \newcite{Horacek1997}).  In principle, we
+almost the entire GRE literature (notable exceptions are e.g.\
+\newcite{Horacek1997} and SPUD \cite{Stone1998a}).  In principle, we
 believe that it is a good idea to handle sentence planning and

### 2008-03-20 trunk/Papers/dl-gre-08/bisim.tex
First version of sections 2 and 3 completed.  I'm now working on the examples (section 3.2).  Perhaps sections 2 and 3 could be shortened (we are already on page 9).

@@
 
-\begin{theorem}[]\label{bisim}
-Given a model $\gM$, if  $i$ and $i'$ are two $\alc$-bisimilar elements in $\gM$
-then there is no $\alc$-formula $\varphi$ such that $i \in |\varphi|^\gM$ and
-$i' \not \in |\varphi|^\gM$.
-\end{theorem}
+Why are bisimulations relevant for computing referring expressions?
+The following result provide us with a crucial characterization:
 
-In other words, $\alc$-bisimilar elements are indistinguishable in the $\alc$ langauge.
-If we only consider finite models (actually, it is enought if we consider models where each element has only a finite number of immediate successors) then the converse of Theorem~\ref{bisim} also holds.
-
-\begin{theorem}[\cite{blac:moda01}]\label{bisim2}
-Given a model $\gM$, $i$ and $i'$ are two $\alc$-bisimilar elements in $\gM$
+\begin{theorem}[\cite{blac:moda01}]\label{bisim}
+Given a finite model $\gM$, $i$ and $i'$ are two $\alc$-bisimilar elements in $\gM$
 if and only if there is no $\alc$-formula $\varphi$ such that $i \in |\varphi|^\gM$ and
@@
 
-By defining a suitable notion of $\el$-bisimulation, the same results can be proved for
-the $\el$ language.
-
-\cite{paig:thre87}
-
-\cite{dovier04:_effic_algor_for_comput_bisim_equiv}
+In other words, $\alc$-bisimilar elements are indistinguishable in the $\alc$ language. Hence, bisimulation gives us a test for non-referenceability: if a
+model $\gM$ contains two different individuals $i$ and $i'$ which are
+$\alc$-bisimilar, then there is no formula of $\alc$ that can be used to uniquely
+refer to $i$ or $i'$.  We will discuss which is the exact relation between
+bisimulation and generation of referring expressions in the next section.
+For the moment we just remark that a suitable notion of $\el$-bisimulation can also be defined~\cite{kurt:expr99}.
 

### 2008-03-20 trunk/Papers/dl-gre-08/gre.tex
First version of sections 2 and 3 completed.  I'm now working on the examples (section 3.2).  Perhaps sections 2 and 3 could be shortened (we are already on page 9).

@@
 concepts that are satisfied by all individuals in $C$ and no
-individuals that are not in $C$ (by Theorems~\ref{bisim} and
-\ref{bisim2}).  If we represent referring expressions as DL concepts,
+individuals that are not in $C$ (by Theorems~\ref{bisim}).  If we represent referring expressions as description logic formulas,
 this means that there exists a unique referring expression for some
-individual $a$ iff $a$ is alone in its bisimulation class; and if it
-is, we can use one of the concepts that are characteristic for $C$ as
+individual $i$ iff $i$ is alone in its bisimulation class; and if it
+is, we can use one of the formulas that are characteristic for $C$ as
 the referring expression.
@@
 Representing referring expressions as DL concepts is a very natural
-perspective.  The purely conjunctive bag of propositional symbols that
-e.g.\ \newcite{Dale1995} compute is clearly a conjunction in
+perspective.  The purely conjunctive set of propositional symbols that,
+e.g.,~\newcite{Dale1995} compute is clearly a conjunction in
 description logic.  Relational expressions as in
-\newcite{dale91:_gener_refer_expres_invol_relat} or
-\newcite{Krahmer2003} are simply concepts of \el; for instance, ``the
-book on the table'' corresponds to the concept $\mathsf{book} \sqcap
-\exists \mathsf{on}. \mathsf{table}$.  And other propositional
-connectives, such as disjunction and negation
-\cite{deemter01:_gener_refer_expres}, can clearly be represented in
-\alc\ as well.
+\newcite{dale91:_gener_refer_expres_invol_relat} are concepts of \el.
+We will discuss this idea further in Section~\ref{xxx}.
 
-All this means that we can reduce the problem of computing a unique RE
+Summing up, we can reduce the problem of computing a unique RE
 to the problem of computing the bisimulation classes of a model.  We
@@
 
-\subsection{Bisimulation classes for \alc}
+\subsection{Bisimulation classes for \alc and \el}
 
-\subsection{Bisimulation classes for \el}
+There exists very efficient algorithms
+for computing the bisimulation classes of a model~\cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
+These algorithms were originally used for automata minimization: given an
+automata $A$ they would return a smaller automata $A'$ which recognized the
+same language.  We can adapt these algorithm to compute a formula representing each bisimulation class.  Again, let's first
+discuss the $\alc$ case.
 
-The algorithm we just presented is driven by case distinctions between
-concepts $\varphi$ and $\neg \varphi$, and thus adds large amounts of
-negations to the characteristic concepts.  This can be inconvenient
-in NLG, because such concepts may be expressible only by rather clumsy
-natural-language expressions. \todo{example?}  We will therefore
-introduce a new algorithm which is limited to computing
-(negation-free) concepts of \el.
+We assume given a model $\gM=(\Delta^\gM, |\cdot|^\gM)$ containing all the information about the current
+situation.  We will compute a set $\RE$ of $\alc$ formulas such that the
+set $\{|\varphi|^\gM \mid \varphi \in \RE\}$ is a partition of $\Delta^\gM$
+that contains all the bisimulation classes of $\gM$.  The pseudocode is
+shown in Algorithm~\ref{algo:bisim-l} and Algorithm~\ref{algo:bisim-add-alc}.
 
-As before, the algorithm maintains a set $\C = \{C_1,\ldots,C_n\}$ of
-concepts (this time of \el) such that $\interp{C_1} \cup \ldots \cup
-\interp{C_n} = \Delta$, and which it refines iteratively.  However,
-where the \alc\ algorithm maintains the invariant that
-$\interp{C_1},\ldots,\interp{C_n}$ is a partition of $\Delta$, we
-weaken this invariant to the requirement that there are no $m \geq 2$
-pairwise different indices $1 \leq i_1,\ldots,i_m \leq n$ such that
-$\interp{C_{i_1}} = \interp{C_{i_2}} \cup \ldots \cup
-\interp{C_{i_m}}$.  We call the concept $C_{i_1}$ \emph{subsumed} if
-such a decomposition exists.
-
-% \begin{algorithm}[t]
-% \dontprintsemicolon
-% \caption{\el\ bisimulation classes}
-% \label{algo:bisim-el}
-% \KwIn{A model $\gM = (\Delta^\gM, |\cdot|^\gM)$}
-% \KwOut{A set \RE of pairs $(\varphi,|\varphi|^\gM)$ such that
-% $\{S \mid (\varphi,S) \in \RE\}$ is the set of $\el$-bisimulation
-% classes of $\gM$.}
-% \RE $\leftarrow \{(\top,\Delta^\gM)\}$ \;
-% \For{$p \in \prop$}{
-%   add$_\el$(p, \RE) \;
-% }
-% \While{for some $(\varphi,S) \in \RE$, $|S| > 1$}{
-%   \For{$(\varphi,S) \in$ \RE, $R \in \rel$}{
-%     add$_\el$($\exists R.\varphi$, \RE) \;
-%   }
-%   \If{made no changes to $\RE$}{
-%     exit\;
-%   }
-% }
-% \end{algorithm}
+\begin{theorem}\label{theo:bisimulation-classes}
+Given a model $\gM$, the set $\mathcal{C}$ of all $\alc$-bisimulation classes
+of $\gM$ is computed by Algorithms~\ref{algo:bisim-l} and~\ref{algo:bisim-add-alc} in polynomial time.
+Moreover, for each class $C \in \mathcal{C}$ the algorithm returns an $\alc$ formula $\varphi$ such that
+$C = |\varphi|^\gM$.
+\end{theorem}
 
@@
 \dontprintsemicolon
-\caption{}
-\label{algo:bisim-alc}
+\caption{Computes $\mathcal{L}$-bisimulation classes}
+\label{algo:bisim-l}
 \KwIn{A model $\gM = (\Delta^\gM, |\cdot|^\gM)$}
@@
 }
-
 \end{algorithm}
 
-\begin{algorithm}
+Algorithms~\ref{algo:bisim-l} and \ref{algo:bisim-add-alc} compute
+the \alc-bi\-sim\-u\-la\-tion classes for a given model.
+Algorithm~\ref{algo:bisim-l} iterates over all propositional and
+relational symbols of the signature to construct new formulas until
+either all formulas in $\RE$ denote singletons (i.e., there is only one
+individual that satisfies them), or no progress has been made in the
+previous iteration.  In each iteration, it calls the procedure
+add$_\alc$($\varphi$, $\RE$), which intersects $\varphi$ with any formula $\psi
+\in \RE$ which does not denote a singleton and which is
+not equivalent to $\varphi$ and to $\neg \varphi$. In this case it
+replace $\psi$ in $\RE$ by $\psi \sqcap \varphi$ and $\psi \sqcap \neg \varphi$ to
+$\RE$.
+
+The $\alc$ algorithm is driven by case distinctions between
+concepts $\varphi$ and $\neg \varphi$, and thus freely adds 
+negations to the characteristic formulas in $\RE$.  This can be inconvenient
+in natural language generation, because such concepts may be expressible only by rather clumsy natural-language expressions. We will see examples in Section~\ref{xxx}.
+
+An algorithm which is limited to computing
+(negation-free) formulas in \el can be obtained by combining
+Algorithms~\ref{algo:bisim-l} and~\ref{algo:bisim-add-el}.
+
+As before, the algorithm maintains a set $\RE = \{\varphi_1,\ldots,\varphi_n\}$ of
+formulas (this time of \el) such that $\interp{\varphi_1}^\gM \cup \ldots \cup
+\interp{\varphi_n}^\gM = \Delta^\gM$, and which it refines iteratively.  However,
+where the \alc\ algorithm maintains the invariant that
+$\interp{\varphi_1},\ldots,\interp{\varphi_n}$ is a partition of $\Delta$, we
+weaken this invariant to the requirement that there are no $m \geq 2$
+pairwise different indices $1 \leq i_1,\ldots,i_m \leq n$ such that
+$\interp{\varphi_{i_1}} = \interp{\varphi_{i_2}} \cup \ldots \cup
+\interp{\varphi_{i_m}}$.  We call the formula $\varphi_{i_1}$ \emph{subsumed} if
+such a decomposition exists.
+
+\begin{algorithm}[t]
 \caption{add$_\alc(\varphi,\RE)$}
+\label{algo:bisim-add-alc}
 \For{$\psi \in \RE$ with $|\psi|^\gM > 1$}{
@@
 \end{algorithm}
-
+%
 \begin{algorithm}[t]
@@
 
-Algorithms~\ref{algo:bisim-el} and \ref{algo:bisim-add-el} compute
-the \el\ bisimulation classes for a given model.
-Algorithm~\ref{algo:bisim-el} iterates over all propositional and
-relational symbols of the signature to construct new concepts until
-either all concepts in $\C$ are singleton (i.e., there is only one
-individual that satisfies them), or no progress has been made in the
-previous iteration.  In each iteration, it calls the procedure
-add$_\el$($C$, $\C$), which intersects $C$ with all current concepts
-that are not already singleton, adds the result to $\C$ and then
-removes all concepts in $\C$ that have become subsumed.  Removal of
-subsumed concepts can be implemented efficiently by arranging the
-concepts in $\C$ in a a graph that connects two concepts with an edge
-if one is a maximal proper subset of the other.
+The procedure add$_\el$($\varphi$, $\RE$) shown in Algorithms~\ref{algo:bisim-add-el} intersects $\varphi$ with all current formulas in $\RE$
+that do not denote a singleton, adds the result to $\RE$ and then
+removes all formulas in $\RE$ that have become subsumed.  Removal of
+subsumed formulas from $\RE$ can be implemented efficiently by arranging
+them by inclusion in terms of their denotation.
 
@@
 % \end{figure}
+
+\subsection{Some Examples}
 
@@
 %%% End:
+
+% \begin{algorithm}[t]
+% \dontprintsemicolon
+% \caption{\el\ bisimulation classes}
+% \label{algo:bisim-el}
+% \KwIn{A model $\gM = (\Delta^\gM, |\cdot|^\gM)$}
+% \KwOut{A set \RE of pairs $(\varphi,|\varphi|^\gM)$ such that
+% $\{S \mid (\varphi,S) \in \RE\}$ is the set of $\el$-bisimulation
+% classes of $\gM$.}
+% \RE $\leftarrow \{(\top,\Delta^\gM)\}$ \;
+% \For{$p \in \prop$}{
+%   add$_\el$(p, \RE) \;
+% }
+% \While{for some $(\varphi,S) \in \RE$, $|S| > 1$}{
+%   \For{$(\varphi,S) \in$ \RE, $R \in \rel$}{
+%     add$_\el$($\exists R.\varphi$, \RE) \;
+%   }
+%   \If{made no changes to $\RE$}{
+%     exit\;
+%   }
+% }
+% \end{algorithm}

### 2008-03-20 trunk/Papers/dl-gre-08/abstract.tex

@@
   In this paper, we propose to reinterpret the problem of generating
-  referring expressions as the problem of computing a concept in a
+  referring expressions (GRE) as the problem of computing a formula in a
   description logic that is only satisfied by the referent.  This view

### 2008-03-20 trunk/Papers/dl-gre-08/introduction.tex

@@
 The generation of referring expressions is one of the most active and
-successful research areas in NLG.  Starting with Dale and Reiter's
+successful research areas in natural language generation.  Starting with Dale and Reiter's
 incremental algorithm \cite{Dale1995}, various researchers have added
@@
 problem of generating referring expressions: We take a referring
-expression describing an individual $a$ to be a concept of some
-description logic (DL) which is satisfied by $a$ and no other
+expression describing an individual $i$ to be a formula of some
+description logic (DL) which is satisfied by $i$ and no other
 individual in the universe.  This allows us to organize existing
@@
 
-Furthermore, we show how to apply algorithms for computing the
-\emph{bisimulation classes} of a DL model to GRE.  Roughly speaking,
+Furthermore, we show how to addapt algorithms for computing the
+\emph{bisimulation classes} of a DL model to the GRE problem.  Intuitively,
 bisimulation classes are sets of individuals that satisfy exactly the
@@
 once, these algorithms systematically avoid the infinite regress
-problem.  The \el\ algorithm, which generates the same class of
+problem.  The \el\ algorithm, which generates the same kind of
 descriptions as Dale \& Haddock, is capable of generating 67\% of the
@@
 bisimulation.  We will then show how to compute bisimulation classes
-in Section~\ref{sec:gre} and apply these algorithms to GRE.  In
+ and how to apply these algorithms to GRE in Section~\ref{sec:gre}.  In
 Section~\ref{sec:discussion}, we evaluate our algorithms and discuss
-these results.  Section~\ref{sec:related} compares our approach to
+our results.  Section~\ref{sec:related} compares our approach to
 related research; in particular, it shows how various prominent GRE

### 2008-03-20 trunk/Papers/dl-gre-08/bisim.tex

@@
 (e.g., we can write formulas in \alc which could be realized as ``a book which is not red and is not on the floor'').
-Computing bisimulations in \alc is very efficient, but \alc can contain expressions which are difficult to realize (because of the presence of negation).
-On the other hand,
-\el expressions are easily realizable, but the bisimulation algorithm is more complex.
+%Computing bisimulations in \alc is very efficient, but \alc can contain expressions which are difficult to realize (because of the presence of negation).  On the other hand,
+%\el expressions are easily realizable, but the bisimulation algorithm is more complex.
 
-But let start by formally defining the description languages we are going to use, and the notion of bisimulation.
+Let start by formally defining the description languages we are going to use, and the notion of bisimulation.
 
@@
 The previous definition formally introduces the syntax and semantics of the
-logics \alc and \el.  In the paper we will sometime refer to some other description logics (mainly in Section~\ref{xxx}).  For completeness we mention them briefly here,
+logics \alc and \el.  In the paper we will sometime refer to some other description logics (mainly in Section~\ref{sec:related}).  For completeness we mention them briefly here,
 and we refer the reader to~\cite{baad:desc03} for further details:
@@
 \begin{center}
-\begin{tabular}{lp{6cm}}
-$\mathcal{CL}$ & is obtained from $\el$ by dropping existential quantification\\
-$\mathcal{PL}$ & is obtained form $\alc$ by dropping existential quantification\\
-$\mathcal{ELU}_{(\neg)}$ & is obtained from $\el$ by adding disjunction and atomic negation
+\begin{tabular}{lp{5.9cm}}
+$\mathcal{CL}$ & is obtained from $\el$ by dropping existential quantification.\\
+$\mathcal{PL}$ & is obtained form $\alc$ by dropping existential quantification.\\
+$\mathcal{ELU}_{(\neg)}$ & is obtained from $\el$ by adding disjunction and atomic negation.
 \end{tabular}
@@
 We now introduce the notion of \emph{bisimulation}.  We will discuss in detail
-the case for the $\alc$ language.  Further details can be found
+the case for the $\alc$ language,further details can be found
 in~\cite{blac:moda01,kurt:expr99}.
@@
 \begin{definition}
-Given a model $\gM = (\Delta^\gM,|\cdot|^\gM)$, two elements $i$ and $i'$ of $\Delta^\gM$ are $\alc$-bisimilar if it is possible to find a relation $Z \subseteq \Delta^\gM \times \Delta^\gM$ such that
-\begin{enumerate}
-\item $(i,i') \in Z$.
-\item if $(e_1, e_2) \in Z$ then $\propm(e_1) = \propm(e_2)$.
-\item if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_1'$ such that
+Given a model $\gM = (\Delta^\gM,|\cdot|^\gM)$, two elements $i$ and $i'$ of $\Delta^\gM$ are $\alc$-bisimilar if there is a relation $Z \subseteq \Delta^\gM \times \Delta^\gM$ such that
+\begin{itemize}
+\item[i)] $(i,i') \in Z$.\\[-1.5em]
+\item[ii)] if $(e_1, e_2) \in Z$ then $\propm(e_1) = \propm(e_2)$.\\[-1.5em]
+\item[iii)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_1'$ such that
 $(e_1,e_1') \in |R|^\gM$ then there is $e_2'$ such that $(e_2,e_2') \in |R|^\gM$ and
-$(e_1',e_2') \in Z$.
-\item if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_2'$ such that
+$(e_1',e_2') \in Z$.\\[-1.5em]
+\item[iv)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_2'$ such that
 $(e_2,e_2') \in |R|^\gM$ then there is $e_1'$ such that $(e_1,e_1') \in |R|^\gM$ and
 $(e_1',e_2') \in Z$.
-\end{enumerate}
+\end{itemize}
 \end{definition}
@@
 Why are bisimulations relevant for computing referring expressions?
-The following result provide us with a crucial characterization:
+The following result provide a crucial characterization~\cite{blac:moda01}:
 
-\begin{theorem}[\cite{blac:moda01}]\label{bisim}
-Given a finite model $\gM$, $i$ and $i'$ are two $\alc$-bisimilar elements in $\gM$
+\begin{theorem}\label{bisim}
+Given a finite model $\gM$, $i$ and $i'$ are $\alc$-bisimilar in $\gM$
 if and only if there is no $\alc$-formula $\varphi$ such that $i \in |\varphi|^\gM$ and

### 2008-03-20 trunk/Papers/dl-gre-08/gre.tex

@@
 
-In computational linguistics, the generating referring expression problem (the GRE
-problem) can be intuitively presented as follows: given some information about
+In computational linguistics, the GRE
+problem can be intuitively presented as follows: given some information about
 a group of individuals (e.g., grounded facts about individuals in an ongoing dialogue),
-and a grammar, found a grammatically correct expression (usually a noun phrase) that uniquely identify a given individual.
+and a grammar, find a grammatically correct expression (usually a noun phrase) that uniquely identify a given individual.
 
 By taking a logic perspective, we can recast the GRE problem as an inference problem.
-We can think that the information about the group of individuals is provided in the form
+We can assume that the information about the group of individuals is provided in the form
 of a model $\gM$ of a given logical language $\gL$.  Now, the inference problem
 associated with the GRE problem will be to find a formula in $\gL$ such that it
-exaclty identifies a given individual $i$.  Formally, if $|\cdot|^\gM$ is the interpretation
-function for $\gL$ which for each formula in $\gL$ and for each model $\gM$ returns
-the set of elements in $\gM$ for which the formula $\varphi$ holds, then we can define:
+exaclty identifies a given individual $i$.
 \medskip
@@
 Consider the following example.  We have three individuals related as indicated in
-Figure~\ref{fig:dale-haddock}d.
+Fig.~\ref{fig:dale-haddock}d.
 
@@
 all members of $C$ are pairwise bisimilar.  For the finite models we
-deal with in GRE, this amounts to saying that there are certain
-concepts that are satisfied by all individuals in $C$ and no
-individuals that are not in $C$ (by Theorems~\ref{bisim}).  If we represent referring expressions as description logic formulas,
+deal with in GRE, this amounts to saying that all individuals in $C$ and only those satisfy the same formulas (Theorems~\ref{bisim}).
+If we represent referring expressions as description logic formulas,
 this means that there exists a unique referring expression for some
@@
 \newcite{dale91:_gener_refer_expres_invol_relat} are concepts of \el.
-We will discuss this idea further in Section~\ref{xxx}.
+We will discuss this idea further in Section~\ref{sec:related}.
 
@@
 for computing the bisimulation classes of a model~\cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
-These algorithms were originally used for automata minimization: given an
+These algorithms were originally used for automata minimization (given an
 automata $A$ they would return a smaller automata $A'$ which recognized the
-same language.  We can adapt these algorithm to compute a formula representing each bisimulation class.  Again, let's first
+same language), but we can adapt these algorithm to compute a formula representing each bisimulation class.  Again, let's first
 discuss the $\alc$ case.
@@
 \dontprintsemicolon
-\caption{Computes $\mathcal{L}$-bisimulation classes}
+\caption{Compute $\mathcal{L}$-bisimulation classes}
 \label{algo:bisim-l}
@@
 \in \RE$ which does not denote a singleton and which is
-not equivalent to $\varphi$ and to $\neg \varphi$. In this case it
-replace $\psi$ in $\RE$ by $\psi \sqcap \varphi$ and $\psi \sqcap \neg \varphi$ to
+not equivalent to $\varphi$ and to $\neg \varphi$. In this case, it
+replaces $\psi$ in $\RE$ by $\psi \sqcap \varphi$ and $\psi \sqcap \neg \varphi$ to
 $\RE$.
@@
 negations to the characteristic formulas in $\RE$.  This can be inconvenient
-in natural language generation, because such concepts may be expressible only by rather clumsy natural-language expressions. We will see examples in Section~\ref{xxx}.
+in natural language generation, because such concepts may be expressible only by rather clumsy natural-language expressions. We will see examples in Section~\ref{sec:examples}.
 
@@
 
+Because the \el\ algorithm maintains a weaker invariant
+for the classes, it may consider more of these classes simultaneously.
+Prima facie, this means that it has worst-case exponential runtime,
+given that the whole domain has an exponential number of subsets.
+However, it is possible that a more careful analysis of the
+combinatorics of the considered subsets reveal that the algorithm is
+actually polynomial; it certainly behaves polynomially in our
+evaluations (see below).
+
 \begin{figure*}
@@
 
-\subsection{Some Examples}
+\subsection{Some Examples}\label{sec:examples}
 
-Let's see what this algorithm does on the example shown in
-Fig.~\ref{fig:dale-haddock}, which is taken from
-\newcite{dale91:_gener_refer_expres_invol_relat}.  In the first loop,
-the algorithm will add new concepts $\mathsf{floor}$, $\mathsf{bowl}$,
-$\mathsf{cup}$, and $\mathsf{table}$; because adding these concepts
-makes $\top$ subsumed, $\C$ will consist of these four concepts after
-the loop.  Not all of these concepts are singleton; for instance,
-$\interp{\mathsf{cup}}$ contains two individuals.  So we iterate over
-the relations to make the concepts more precise.  After the first
-iteration over the relations, we have $\C = \{ \mathsf{floor},
+We will now see what our algorithms do in some interesting
+examples. Consider the model shown in
+Fig.~\ref{fig:dale-haddock}a, which is taken from
+\newcite{dale91:_gener_refer_expres_invol_relat}, and
+let's run the $\el$ algorithm.
+
+In the first loop,
+the formulas $\mathsf{floor}$, $\mathsf{bowl}$,
+$\mathsf{cup}$, and $\mathsf{table}$ will be added to $\RE$ because they subsume $\top$.  Not all of these formulas denote singletons; for instance,
+$|\mathsf{cup}|\gM$ contains two individuals.  So we iterate over
+the relations to refine our formulas.  After the first
+iteration over the relations, we have $\RE = \{ \mathsf{floor},
 \mathsf{bowl} \sqcap \exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl}
@@
 \mathsf{table} \}$. Notice that $\mathsf{bowl}$ has become subsumed,
-but we haven't distinguished the cups and tables further, but we can
+but we haven't distinguished the cups and tables further.
+
+Now we can
 use the distinction between the bowls to distinguish the cups in the
-second iteration.  The result of this is $\C = \{ \mathsf{floor},
+second iteration.  The result of this is $\RE = \{ \mathsf{floor},
 \mathsf{bowl} \sqcap \exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl}
@@
 \mathsf{in}. (\mathsf{bowl} \sqcap \exists
-\mathsf{on}.\mathsf{table}), \mathsf{table} \}$.  At this point, all
-concepts except for $\mathsf{table}$ are singleton, and further
+\mathsf{on}.\mathsf{table}), \mathsf{table} \}$.
+
+At this point, all
+formulas except $\mathsf{table}$ denote singletons, and further
 iterations don't allow us to refine $\mathsf{table}$; so the algorithm
-terminates.  Each of the singleton bisimulation classes is represented
-by a characteristic concept; for instance, $\C$ tells us that
+terminates.
+
+Each of the singleton bisimulation classes is represented
+by a characteristic formula; for instance, $\RE$ tells us that
 $\mathsf{cup} \sqcap \exists \mathsf{in}. (\mathsf{bowl} \sqcap
@@
 build can ``see'' fewer differences between individuals.  For example,
-in the model shown in Fig.~\ref{fig:el-vs-alc}, the \alc\ algorithm
-will compute the concepts $\{\neg \exists R. (\neg P \sqcap \neg Q),
-Q, \exists R. (\neg P \sqcap \neg Q), \neg P \sqcap \neg Q\}$ (each of
-which is singleton), but $a$ and $c$ are \el-bisimilar in this model,
+in the model shown in Fig.~\ref{fig:dale-haddock}c, the \alc algorithm
+will compute the set $\RE = \{\neg \exists R. (\neg P \sqcap \neg Q),
+Q, \exists R. (\neg P \sqcap \neg Q), \neg P \sqcap \neg Q\}$ (each formula denotes a singleton), but $a$ and $c$ are \el-bisimilar in this model,
 and the only \el-concept that is true in $d$ is $\top$, so the \el\
-algorithm can only compute the concepts $\{\top, P, P \sqcap Q\}$.
-Furthermore, because the \el\ algorithm maintains a weaker invariant
-for the classes, it may consider more of these classes simultaneously.
-Prima facie, this means that it has worst-case exponential runtime,
-given that the whole domain has an exponential number of subsets.
-However, it is possible that a more careful analysis of the
-combinatorics of the considered subsets reveal that the algorithm is
-actually polynomial; it certainly behaves polynomially in our
-evaluations (see below).
+algorithm can only compute the formulas $\{\top, P, P \sqcap Q\}$.
 

### 2008-03-20 trunk/Papers/dl-gre-08/bisim.tex
iterated over 2-3; paper is down to 8 pages + 3 refs

@@
-\newcommand{\gM}{\mathcal{M}}
-\newcommand{\gL}{\mathcal{L}}
-
 \section{Description Languages and Bisimulation} \label{sec:bisim}
 
-One of the main ideas we want to investigate in this paper is
-how we can use \emph{formulas} from description logics (see~\cite{baad:desc03}) to represent
-referring expressions (RE). And how we can use the notion of \emph{bisimulation}
-(see~\cite{blac:moda01}) to obtain very efficient algorithms for computing referring expressions.
-
-In what follows we will introduce in detail two description logics called \el and \alc.  Intuitively, \el allows
-for conjunctions of atomic and existentially quantified relational expressions (e.g.,
-we can write formulas in \el that could be realized as ``a red book on a table''). \alc is the extension of \el with full negation
-(e.g., we can write formulas in \alc which could be realized as ``a book which is not red and is not on the floor'').
-%Computing bisimulations in \alc is very efficient, but \alc can contain expressions which are difficult to realize (because of the presence of negation).  On the other hand,
-%\el expressions are easily realizable, but the bisimulation algorithm is more complex.
-
-Let start by formally defining the description languages we are going to use, and the notion of bisimulation.
+The main point of this paper is to represent referring expressions
+(REs) as formulas of description logic \cite{baad:desc03}.  In order
+to make this point, we will now first define the two description
+logics we will be working with -- \alc\ and \el.
 
 \begin{definition}
-Formulas of $\alc$ are generated by the following grammar:
+Formulas $\varphi$ of $\alc$ are generated by the following grammar:
 $$
-\form ::= p \mid \neg \varphi \mid \varphi \sqcap \varphi' \mid \exists R. \varphi
+\varphi,\varphi' ::= p \mid \neg \varphi \mid \varphi \sqcap \varphi'
+\mid \exists R. \varphi
 $$
-where $p$ is in the set of propositional symbols \prop, $R$ is in the set of relational symbols \rel, $\varphi$ and $\varphi'$ are in \form. $\el$ is the
-negation-free fragment of $\alc$.
+where $p$ is in the set of propositional symbols \prop, $R$ is in the
+set of relational symbols \rel.
 
-Formulas of both $\alc$ and $\el$ are interpreted in relational first-order models
-$\gM = (\Delta^\gM,|\cdot|^\gM)$ where $W$ is an arbitrary non-empy set, and $|\cdot|^\gM$ is an interpretation
-function such that:
+$\el$ is the negation-free fragment of $\alc$.
+
+Formulas of both $\alc$ and $\el$ are interpreted in ordinary
+relational first-order models $\gM = (\Delta^\gM,|\cdot|^\gM)$ where
+$\Delta^\gM$ is some non-empy set, and $|\cdot|^\gM$ is an
+interpretation function such that:
 $$
@@
 |R|^\gM & \subseteq & \Delta^\gM \times \Delta^\gM  \mbox{ for $R \in \rel$}\\
-|\neg \varphi|^\gM & = & W \backslash |\varphi|^\gM\\
+|\neg \varphi|^\gM & = & \Delta^\gM \backslash |\varphi|^\gM\\
 |\varphi \sqcap \varphi'|^\gM & = & |\varphi|^\gM \cap |\varphi'|^\gM\\
-|\exists R.\varphi|^\gM & = & \{i \mid \mbox{for some } i' ((i,i') \in |R|^\gM)\\
+|\exists R.\varphi|^\gM & = & \{i \mid \mbox{for some } i', (i,i') \in |R|^\gM\\
 & & \mbox{ and } i' \in |\varphi|^\gM \}.\\
@@
 
-The previous definition formally introduces the syntax and semantics of the
-logics \alc and \el.  In the paper we will sometime refer to some other description logics (mainly in Section~\ref{sec:related}).  For completeness we mention them briefly here,
-and we refer the reader to~\cite{baad:desc03} for further details:
+Every formula of a description logic denotes a set of individuals in
+the domain; thus we can use such formulas to describe sets.  For
+instance, in the model in Fig.~\ref{fig:dale-haddock}b, the formula
+$\mathsf{flower}$ denotes the set $\{f_1,f_2\}$; the formula
+$\mathsf{flower} \sqcap \exists \mathsf{in}.\mathsf{hat}$ denotes
+$\{f_2\}$; and the formula $\mathsf{flower} \sqcap \neg
+\exists.\mathsf{hat}$ denotes $\{f_1\}$.
 
-\begin{center}
-\begin{tabular}{lp{5.9cm}}
-$\mathcal{CL}$ & is obtained from $\el$ by dropping existential quantification.\\
-$\mathcal{PL}$ & is obtained form $\alc$ by dropping existential quantification.\\
-$\mathcal{ELU}_{(\neg)}$ & is obtained from $\el$ by adding disjunction and atomic negation.
-\end{tabular}
-\end{center}
+Different description logics differ in the inventory of logical
+connectives they permit.  The difference between \alc\ and \el\ is
+that \alc\ permits negation and \el\ doesn't.  There are many other
+description logics in the literature; some that we will get back to in
+Section~\ref{sec:related} are $\mathcal{CL}$ (\el\ without existential
+quantification, i.e.\ only conjunctions of atoms); $\mathcal{PL}$
+(\alc\ without existential quantification, i.e.\ propositional logic);
+and $\mathcal{ELU}_{(\neg)}$ (\el\ plus disjunction and atomic
+negation).
 
-\noindent
-It should be clear from the definitions above that $\mathcal{CL}$ is conjunctive logic, while $\mathcal{PL}$ is propositional logic.
-
-We now introduce the notion of \emph{bisimulation}.  We will discuss in detail
-the case for the $\alc$ language,further details can be found
-in~\cite{blac:moda01,kurt:expr99}.
+A key notion in studying description logics is \emph{bisimulation}.
+We will discuss in detail the case for $\alc$; further details can be
+found in~\cite{blac:moda01,kurt:expr99}.
 
 \begin{definition}
-Given a model $\gM = (\Delta^\gM,|\cdot|^\gM)$, two elements $i$ and $i'$ of $\Delta^\gM$ are $\alc$-bisimilar if there is a relation $Z \subseteq \Delta^\gM \times \Delta^\gM$ such that
+Given a model $\gM = (\Delta^\gM,|\cdot|^\gM)$, two elements $i$ and
+$i'$ of $\Delta^\gM$ are \emph{$\alc$-bisimilar} if there is a relation $Z
+\subseteq \Delta^\gM \times \Delta^\gM$ such that 
 \begin{itemize}
@@
 \end{itemize}
+A maximal subset of $\Delta^\gM$ all of whose elements are pairwise
+bisimilar is called an \emph{\alc-bisimulation class} of $\gM$.
 \end{definition}
 
-Why are bisimulations relevant for computing referring expressions?
-The following result provide a crucial characterization~\cite{blac:moda01}:
+The crucial property that makes bisimulation relevant for GRE is the
+following theorem \cite{blac:moda01}.
 
 \begin{theorem}\label{bisim}
-Given a finite model $\gM$, $i$ and $i'$ are $\alc$-bisimilar in $\gM$
-if and only if there is no $\alc$-formula $\varphi$ such that $i \in |\varphi|^\gM$ and
-$i' \not \in |\varphi|^\gM$.
+  Given a finite model $\gM$, $i$ and $i'$ are $\alc$-bisimilar in
+  $\gM$ if and only if there is no $\alc$-formula $\varphi$ such that
+  $i \in |\varphi|^\gM$ and $i' \not \in |\varphi|^\gM$.
 \end{theorem}
 
-In other words, $\alc$-bisimilar elements are indistinguishable in the $\alc$ language. Hence, bisimulation gives us a test for non-referenceability: if a
-model $\gM$ contains two different individuals $i$ and $i'$ which are
-$\alc$-bisimilar, then there is no formula of $\alc$ that can be used to uniquely
-refer to $i$ or $i'$.  We will discuss which is the exact relation between
-bisimulation and generation of referring expressions in the next section.
-For the moment we just remark that a suitable notion of $\el$-bisimulation can also be defined~\cite{kurt:expr99}.
+In other words, $\alc$-bisimilar individuals are indistinguishable in
+the $\alc$ language.  In the example in Fig.~\ref{fig:dale-haddock}a,
+the two individuals $t_1$ and $t_2$ are \alc-bisimilar; and indeed,
+they satisfy exactly the same \alc-formulas.  Although we can't go
+into details here, it is possible to define a suitable notion of
+$\el$-bisimulation for which the analogous theorem holds
+\cite{kurt:expr99}.  In general, \el-bisimulation is a weaker notion
+than \alc-bisimulation; for instance, in Fig.~\ref{fig:dale-haddock},
+$f_1$ and $f_2$ are \el-bisimilar \todo{oops! here's the asymmetry,
+  there is a formula for $f_2$ but not for $f_1$} (no \el-formula will
+distinguish them), but we have already seen that they satisfy
+different \alc-formulas.
 

### 2008-03-20 trunk/Papers/dl-gre-08/discussion.tex
iterated over 2-3; paper is down to 8 pages + 3 refs

@@
 
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=\columnwidth]{runtimes}
-  \caption{Average runtimes (in ms) of the two algorithms on random
-    models of     different sizes.} 
-  \label{fig:runtimes}
-\end{figure}
-
 In order to get a more comprehensive picture of the algorithms'
@@
 \subsection{Interface to realization}
+\label{sec:discussion-realization}
 
@@
 massive use of negation in the \alc\ algorithm; it will be hard for
-any realizer to find a reasonable way of expressing the concept $\neg
-\exists R.(\neg P \sqcap \neg Q)$ from the example in
-Fig.~\ref{fig:el-vs-alc} as a smooth noun phrase.  Although we agree
-with \newcite{deemter01:_gener_refer_expres} and others that the
-careful use of negation and disjunction can improve REs, these
-connectives must not be overused.  Thus we consider the concepts
-computed by the \el\ algorithm ``safer'' with respect to realization.
+any realizer to find a reasonable way of expressing a concept like
+$\neg \exists R.(\neg P \sqcap \neg Q)$ as a smooth noun phrase.
+Although we agree with \newcite{deemter01:_gener_refer_expres} and
+others that the careful use of negation and disjunction can improve
+REs, these connectives must not be overused.  Thus we consider the
+concepts computed by the \el\ algorithm ``safer'' with respect to
+realization.
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\columnwidth]{runtimes}
+  \caption{Average runtimes (in ms) of the two algorithms on random
+    models of     different sizes.} 
+  \label{fig:runtimes}
+\end{figure}
 

### 2008-03-20 trunk/Papers/dl-gre-08/gre.tex
iterated over 2-3; paper is down to 8 pages + 3 refs

@@
 
-In computational linguistics, the GRE
-problem can be intuitively presented as follows: given some information about
-a group of individuals (e.g., grounded facts about individuals in an ongoing dialogue),
-and a grammar, find a grammatically correct expression (usually a noun phrase) that uniquely identify a given individual.
+Now we apply description logic to GRE.  The core claim of this paper
+is that the GRE problem can be viewed as the problem of computing a
+formula of description logic whose extension is exactly a given set of
+individuals.
 
-By taking a logic perspective, we can recast the GRE problem as an inference problem.
-We can assume that the information about the group of individuals is provided in the form
-of a model $\gM$ of a given logical language $\gL$.  Now, the inference problem
-associated with the GRE problem will be to find a formula in $\gL$ such that it
-exaclty identifies a given individual $i$.
 \medskip
-
 \noindent
@@
 \textsc{$\gL$-GRE Problem}}\\ \hline
-\ \ Input: & A model $\gM$ and an individual $i$ in the\\
-& \hspace*{0.5cm} domain of $\gM$.\\
-\ \ Output: & A formula $\varphi \in \gL$ such that $|\varphi|^\gM = \{i\}$\\
+\ \ Input: & A model $\gM$ and a set $A \subseteq \Delta^\gM$.\\
+\ \ Output: & A formula $\varphi \in \gL$ such that $|\varphi|^\gM = A$\\
 & \hspace*{0.5cm} (if such a formula exists).\\ \hline
@@
 
-As it can be seen from the definition, the problem depends on which particular
-logical language $\gL$ we are interested in, and (as the original GRE problem)
-it might terminate with failure.  Depending on \emph{a)} the model $\gM$,
-\emph{b)} the chosen individual $i$
-and \emph{c} the expressive power of $\gL$, it can be the case that there is no formula of
-$\gL$ that uniquely identifies $i$ in $\gM$.
+In the examples above, it is because $\mathsf{flower} \sqcap \exists
+\mathsf{in}. \mathsf{hat}$ denotes exactly $\{f_2\}$ that we can say
+``the flower in the hat'' to refer to $f_2$.  This generalizes the
+common view of GRE as attribute selection \cite{Dale1995}: From our
+perspective, a bag of attributes is simply a conjunction of atoms.
+However, more expressive approaches to GRE fit into our framework as
+well; for example, relational expressions as in
+\newcite{dale91:_gener_refer_expres_invol_relat} are formulas of \el.
+Different approaches will deliver different REs, and indeed will be
+able to build REs at all in different circumstances; this reflects the
+differences between the DL languages they use to represent their REs
+in.  This idea provides a unified perspective for comparing different
+GRE approaches, and we will discuss how to see several further
+approaches in Section~\ref{sec:related}.
 
-Consider the following example.  We have three individuals related as indicated in
-Fig.~\ref{fig:dale-haddock}d.
+For the rest of this paper, we assume that we are generating a
+singular RE, i.e.\ $A$ will generally be a singleton set.  In this
+case, we will only be able to generate a formula that denotes exactly
+$A = \{a\}$ (i.e., a RE that uniquely refers to $a$) iff there is no
+other individual $b$ that is bisimilar to $a$; otherwise, any formula
+that is satisfied by $a$ is also satisfied by $b$.  This is why there
+is no RE that uniquely refers to $t_2$ in Fig.~\ref{fig:dale-haddock}a
+and only uses the relations ``on'' and ``in'' (and not their
+inverses).  In other words, $A$ must be a bisimulation class.
+Conversely, if we know that $a$ is alone in its bisimulation class, we
+know that there is a concept that describes only $a$ and nothing
+else.  This means that we can recast the problem of generating
+referring expressions as the problem of computing the bisimulation
+classes of a model.
 
-%\begin{figure}
-%\begin{center}
-%\includegraphics[scale=.8]{pic.pdf}
-%\end{center}
-%\caption{Distinguishable/indistinguishable points.}\label{fig-car1}
-%\end{figure}
-%
-The individuals $i$ and $i'$ can easily be distinguished in first-order logic:
-the formula $R(x,x)$ is true of $i$ and false in $i'$. The corresponding
-referring expression for $i$ would be ``an individual relating to itself''.
-But if we can only use propositional and relational information (equivalently, our grammar does not contain the word ``itself'') then $i$ and $i'$ are indistinguishable. Both
-are ``individuals which are related to individuals which are related to individuals\ldots''
-
-We can connect description logics to the problem of generating
-referring expressions via the concept of \emph{bisimulation classes}.
-A bisimulation class is a maximal subset $C$ of the domain such that
-all members of $C$ are pairwise bisimilar.  For the finite models we
-deal with in GRE, this amounts to saying that all individuals in $C$ and only those satisfy the same formulas (Theorems~\ref{bisim}).
-If we represent referring expressions as description logic formulas,
-this means that there exists a unique referring expression for some
-individual $i$ iff $i$ is alone in its bisimulation class; and if it
-is, we can use one of the formulas that are characteristic for $C$ as
-the referring expression.
-
-Representing referring expressions as DL concepts is a very natural
-perspective.  The purely conjunctive set of propositional symbols that,
-e.g.,~\newcite{Dale1995} compute is clearly a conjunction in
-description logic.  Relational expressions as in
-\newcite{dale91:_gener_refer_expres_invol_relat} are concepts of \el.
-We will discuss this idea further in Section~\ref{sec:related}.
-
-Summing up, we can reduce the problem of computing a unique RE
-to the problem of computing the bisimulation classes of a model.  We
-will now present two algorithms -- one for \alc\ and one for \el\ --
-for doing this.  In fact, each of these algorithms computes all
-bisimulation classes of a model at once, effectively computing a RE
-for every single individual in the domain simultaneously.
+In the rest of this section, we will present algorithms that compute
+the bisimulation classes of a given model for \alc\ and \el, along
+with a characteristic formula for each class.  These algorithms
+efficiently compute \emph{all} bisimulation classes of a model, and
+not just the one for some target referent.  In effect, they compute
+REs for all individuals in some model at the same time.  The REs
+generated by the \el\ algorithm correspond to the relational REs
+generated e.g.\ by \newcite{dale91:_gener_refer_expres_invol_relat}
+(but without any danger of infinite regress), whereas the REs
+generated by the \alc\ algorithm can freely use negation as well.
 
@@
 
-There exists very efficient algorithms
-for computing the bisimulation classes of a model~\cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
-These algorithms were originally used for automata minimization (given an
-automata $A$ they would return a smaller automata $A'$ which recognized the
-same language), but we can adapt these algorithm to compute a formula representing each bisimulation class.  Again, let's first
-discuss the $\alc$ case.
+There is substantial literature on the topic of computing the
+\alc-bisimulation classes of a given model efficiently
+\cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
+However, these algorithms will only compute the bisimulation classes
+themselves. We extend the \newcite{hopc:algo71} algorithm in such a
+way that it computes, for each bisimulation class, also a formula that
+denotes exactly this class.  We can then use these formulas as
+representations of the REs.
 
-We assume given a model $\gM=(\Delta^\gM, |\cdot|^\gM)$ containing all the information about the current
-situation.  We will compute a set $\RE$ of $\alc$ formulas such that the
-set $\{|\varphi|^\gM \mid \varphi \in \RE\}$ is a partition of $\Delta^\gM$
-that contains all the bisimulation classes of $\gM$.  The pseudocode is
-shown in Algorithm~\ref{algo:bisim-l} and Algorithm~\ref{algo:bisim-add-alc}.
-
-\begin{theorem}\label{theo:bisimulation-classes}
-Given a model $\gM$, the set $\mathcal{C}$ of all $\alc$-bisimulation classes
-of $\gM$ is computed by Algorithms~\ref{algo:bisim-l} and~\ref{algo:bisim-add-alc} in polynomial time.
-Moreover, for each class $C \in \mathcal{C}$ the algorithm returns an $\alc$ formula $\varphi$ such that
-$C = |\varphi|^\gM$.
-\end{theorem}
+The pseudocode for our \alc\ algorithm is shown as
+Algorithm~\ref{algo:bisim-l} and Algorithm~\ref{algo:bisim-add-alc}.
+Given a model $\gM=(\Delta^\gM, |\cdot|^\gM)$, the algorithm computes
+a set $\RE$ of $\alc$ formulas such that the set $\{|\varphi|^\gM \mid
+\varphi \in \RE\}$ is the set of \alc-bisimulation classes of $\gM$.
+The algorithm starts with $\RE = \{\top\}$ (where $\interp{\top} =
+\Delta^\gM$), and successively refines $\RE$ by making its elements
+denote smaller and smaller sets.  It maintains the invariant that at
+the start and end of every iteration, $\{|\varphi|^\gM \mid \varphi
+\in \RE\}$ is always a partition of $\Delta^\gM$.  The algorithm
+iterates over all propositional and relational symbols of the
+signature to construct new formulas until either all formulas in $\RE$
+denote singletons (i.e., there is only one individual that satisfies
+them), or no progress has been made in the previous iteration.  In
+each iteration, it calls the procedure add$_\alc$($\varphi$, $\RE$),
+which intersects $\varphi$ with any formula $\psi \in \RE$ which does
+not denote a singleton and which is not equivalent to $\varphi$ and to
+$\neg \varphi$. In this case, it replaces $\psi$ in $\RE$ by $\psi
+\sqcap \varphi$ and $\psi \sqcap \neg \varphi$ to $\RE$.
 
@@
 
-Algorithms~\ref{algo:bisim-l} and \ref{algo:bisim-add-alc} compute
-the \alc-bi\-sim\-u\-la\-tion classes for a given model.
-Algorithm~\ref{algo:bisim-l} iterates over all propositional and
-relational symbols of the signature to construct new formulas until
-either all formulas in $\RE$ denote singletons (i.e., there is only one
-individual that satisfies them), or no progress has been made in the
-previous iteration.  In each iteration, it calls the procedure
-add$_\alc$($\varphi$, $\RE$), which intersects $\varphi$ with any formula $\psi
-\in \RE$ which does not denote a singleton and which is
-not equivalent to $\varphi$ and to $\neg \varphi$. In this case, it
-replaces $\psi$ in $\RE$ by $\psi \sqcap \varphi$ and $\psi \sqcap \neg \varphi$ to
-$\RE$.
+The \alc\ algorithm is capable of computing the \alc-bisimulation
+class of the model very efficiently, in time $O(nm)$ where $n$ is the
+number of individuals in the domain, and $m$ is the number of pairs of
+individuals that are related by some relation.  However, it will
+freely introduce negations in the case distinctions, which can make
+the resulting formula hard to realize (see also
+Section~\ref{sec:discussion-realization}).  This is why we also
+present an algorithm for the \el-bisimulation classes; \el\
+corresponds to positive relational REs, which are generally easy to
+realize.
 
-The $\alc$ algorithm is driven by case distinctions between
-concepts $\varphi$ and $\neg \varphi$, and thus freely adds 
-negations to the characteristic formulas in $\RE$.  This can be inconvenient
-in natural language generation, because such concepts may be expressible only by rather clumsy natural-language expressions. We will see examples in Section~\ref{sec:examples}.
+We obtain the \el\ algorithm by replacing the call to add$_{\alc}$ in
+Algorithm~\ref{algo:bisim-l} by a call to add$_{\el}$, which is defined
+in Algorithm~\ref{algo:bisim-add-el}.  As before, the algorithm
+maintains a set $\RE = \{\varphi_1,\ldots,\varphi_n\}$ of formulas
+(this time of \el) such that $\interp{\varphi_1}^\gM \cup \ldots \cup
+\interp{\varphi_n}^\gM = \Delta^\gM$, and which it refines
+iteratively.  However, where the \alc\ algorithm maintains the
+invariant that $\interp{\varphi_1},\ldots,\interp{\varphi_n}$ is a
+partition of $\Delta$, we weaken this invariant to the requirement
+that there are no $m \geq 2$ pairwise different indices $1 \leq
+i_1,\ldots,i_m \leq n$ such that $\interp{\varphi_{i_1}} =
+\interp{\varphi_{i_2}} \cup \ldots \cup \interp{\varphi_{i_m}}$.  We
+call the formula $\varphi_{i_1}$ \emph{subsumed} if such a
+decomposition exists.
 
-An algorithm which is limited to computing
-(negation-free) formulas in \el can be obtained by combining
-Algorithms~\ref{algo:bisim-l} and~\ref{algo:bisim-add-el}.
+Because the \el\ algorithm maintains a weaker invariant for the
+classes, it may consider more of these classes simultaneously.  Prima
+facie, this means that it has worst-case exponential runtime, given
+that the whole domain has an exponential number of subsets.  However,
+it is possible that a more careful analysis of the combinatorics of
+the considered subsets will reveal that the algorithm is actually
+polynomial.
 
-As before, the algorithm maintains a set $\RE = \{\varphi_1,\ldots,\varphi_n\}$ of
-formulas (this time of \el) such that $\interp{\varphi_1}^\gM \cup \ldots \cup
-\interp{\varphi_n}^\gM = \Delta^\gM$, and which it refines iteratively.  However,
-where the \alc\ algorithm maintains the invariant that
-$\interp{\varphi_1},\ldots,\interp{\varphi_n}$ is a partition of $\Delta$, we
-weaken this invariant to the requirement that there are no $m \geq 2$
-pairwise different indices $1 \leq i_1,\ldots,i_m \leq n$ such that
-$\interp{\varphi_{i_1}} = \interp{\varphi_{i_2}} \cup \ldots \cup
-\interp{\varphi_{i_m}}$.  We call the formula $\varphi_{i_1}$ \emph{subsumed} if
-such a decomposition exists.
+% \begin{figure}
+%   \centering
+%   \includegraphics[width=\columnwidth]{pic-el-vs-alc}
+%   \caption{Illustrating the difference between \el\ and \alc.}
+%   \label{fig:el-vs-alc}
+% \end{figure}
+
+\subsection{Some Examples}\label{sec:examples}
+
+We will now see what our algorithms do in some interesting
+examples. Consider the model shown in
+Fig.~\ref{fig:dale-haddock}a, which is taken from
+\newcite{dale91:_gener_refer_expres_invol_relat}, and
+let's run the $\el$ algorithm.
 
@@
 
-The procedure add$_\el$($\varphi$, $\RE$) shown in Algorithms~\ref{algo:bisim-add-el} intersects $\varphi$ with all current formulas in $\RE$
-that do not denote a singleton, adds the result to $\RE$ and then
-removes all formulas in $\RE$ that have become subsumed.  Removal of
-subsumed formulas from $\RE$ can be implemented efficiently by arranging
-them by inclusion in terms of their denotation.
+The algorithm starts with $\RE = \{\top\}$.  In the first loop, it
+adds the formulas $\mathsf{floor}$, $\mathsf{bowl}$, $\mathsf{cup}$,
+and $\mathsf{table}$, and then removes $\top$ because it is now
+subsumed.  Not all of these formulas denote singletons; for instance,
+$|\mathsf{cup}|^\gM$ contains two individuals.  So we iterate over the
+relations to refine our formulas.  After the first iteration over the
+relations, we have $\RE = \{ \mathsf{floor}, \mathsf{bowl} \sqcap
+\exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl} \sqcap \exists
+\mathsf{on}.\mathsf{table}, \mathsf{cup}, \mathsf{table} \}$. Notice
+that $\mathsf{bowl}$ has become subsumed, but we haven't distinguished
+the cups and tables further.
 
-Because the \el\ algorithm maintains a weaker invariant
-for the classes, it may consider more of these classes simultaneously.
-Prima facie, this means that it has worst-case exponential runtime,
-given that the whole domain has an exponential number of subsets.
-However, it is possible that a more careful analysis of the
-combinatorics of the considered subsets reveal that the algorithm is
-actually polynomial; it certainly behaves polynomially in our
-evaluations (see below).
-
-\begin{figure*}
-  \centering
-  \includegraphics[width=\textwidth]{pic-dale-haddock}
-  \caption{(a) The \newcite{dale91:_gener_refer_expres_invol_relat}
-    scenario; (b) the \newcite{Stone1998a} scenario; (c) illustrating
-    the difference between \el\ and \alc; (d)
-    distinguishable/indistinguishable points.}
-  \label{fig:dale-haddock}
-\end{figure*}
-
-% \begin{figure}
-%   \centering
-%   \includegraphics[width=\columnwidth]{pic-el-vs-alc}
-%   \caption{Illustrating the difference between \el\ and \alc.}
-%   \label{fig:el-vs-alc}
-% \end{figure}
-
-\subsection{Some Examples}\label{sec:examples}
-
-We will now see what our algorithms do in some interesting
-examples. Consider the model shown in
-Fig.~\ref{fig:dale-haddock}a, which is taken from
-\newcite{dale91:_gener_refer_expres_invol_relat}, and
-let's run the $\el$ algorithm.
-
-In the first loop,
-the formulas $\mathsf{floor}$, $\mathsf{bowl}$,
-$\mathsf{cup}$, and $\mathsf{table}$ will be added to $\RE$ because they subsume $\top$.  Not all of these formulas denote singletons; for instance,
-$|\mathsf{cup}|\gM$ contains two individuals.  So we iterate over
-the relations to refine our formulas.  After the first
-iteration over the relations, we have $\RE = \{ \mathsf{floor},
-\mathsf{bowl} \sqcap \exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl}
-\sqcap \exists \mathsf{on}.\mathsf{table}, \mathsf{cup},
-\mathsf{table} \}$. Notice that $\mathsf{bowl}$ has become subsumed,
-but we haven't distinguished the cups and tables further.
-
-Now we can
-use the distinction between the bowls to distinguish the cups in the
-second iteration.  The result of this is $\RE = \{ \mathsf{floor},
-\mathsf{bowl} \sqcap \exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl}
-\sqcap \exists \mathsf{on}.\mathsf{table}, \mathsf{cup} \sqcap \exists
+Now we can use the distinction between the bowls to distinguish the
+cups in the second iteration.  The result of this is $\RE = \{
+\mathsf{floor}, \mathsf{bowl} \sqcap \exists
+\mathsf{on}.\mathsf{floor}, \mathsf{bowl} \sqcap \exists
+\mathsf{on}.\mathsf{table}, \mathsf{cup} \sqcap \exists
 \mathsf{in}. (\mathsf{bowl} \sqcap \exists
@@
 
-At this point, all
-formulas except $\mathsf{table}$ denote singletons, and further
-iterations don't allow us to refine $\mathsf{table}$; so the algorithm
-terminates.
+At this point, all formulas except $\mathsf{table}$ denote singletons,
+and further iterations don't allow us to refine $\mathsf{table}$; so
+the algorithm terminates.  Each of the singleton bisimulation classes
+is represented by a characteristic formula; for instance, $\RE$ tells
+us that $\mathsf{cup} \sqcap \exists \mathsf{in}. (\mathsf{bowl}
+\sqcap \exists \mathsf{on}.\mathsf{table})$ is only satisfied by
+$c_2$, so we may refer to $c_2$ as ``the cup in the bowl on the
+table''.  Notice that the algorithm didn't focus on any particular
+individual; it simultaneously generated REs for all individuals except
+for the two tables.
 
-Each of the singleton bisimulation classes is represented
-by a characteristic formula; for instance, $\RE$ tells us that
-$\mathsf{cup} \sqcap \exists \mathsf{in}. (\mathsf{bowl} \sqcap
-\exists \mathsf{on}.\mathsf{table})$ is only satisfied by $c_2$, so we
-may refer to $c_2$ as ``the cup in the bowl on the table''.
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\columnwidth]{pic-dale-haddock}
+  \caption{(a) The \newcite{dale91:_gener_refer_expres_invol_relat}
+    scenario; (b) the \newcite{Stone1998a} scenario.}
+  \label{fig:dale-haddock}
+\end{figure}
 
-Compared to the \alc\ algorithm, the \el\ algorithm will generally
-compute larger bisimulation classes, and compute singleton
-bisimulation classes less frequently, because the concepts it can
-build can ``see'' fewer differences between individuals.  For example,
-in the model shown in Fig.~\ref{fig:dale-haddock}c, the \alc algorithm
-will compute the set $\RE = \{\neg \exists R. (\neg P \sqcap \neg Q),
-Q, \exists R. (\neg P \sqcap \neg Q), \neg P \sqcap \neg Q\}$ (each formula denotes a singleton), but $a$ and $c$ are \el-bisimilar in this model,
-and the only \el-concept that is true in $d$ is $\top$, so the \el\
-algorithm can only compute the formulas $\{\top, P, P \sqcap Q\}$.
+The \el\ algorithm has a harder time with the example in
+Fig.~\ref{fig:dale-haddock}b \cite{Stone1998a}.  While it will
+correctly identify $r_1$ as ``the rabbit in the hat'' and $f_2$ as
+``the flower in the bathtub'', it will not be able to compute a
+RE for $f_1$; indeed, the algorithm terminates with $\RE$ containing
+both $\mathsf{flower}$ and $\mathsf{flower} \sqcap \exists
+\mathsf{in}.\mathsf{hat}$.  By contrast, the \alc\ algorithm will
+split $\mathsf{flower}$ into the two new formulas $\mathsf{flower}
+\sqcap \exists \mathsf{in}.\mathsf{hat}$ and $\mathsf{flower} \sqcap
+\neg \exists \mathsf{in}.\mathsf{hat}$, generating a unique RE for
+$f_1$ as well.
 

### 2008-03-20 trunk/Papers/dl-gre-08/introduction.tex
iterated over 2-3; paper is down to 8 pages + 3 refs

@@
 
-The generation of referring expressions is one of the most active and
-successful research areas in natural language generation.  Starting with Dale and Reiter's
-incremental algorithm \cite{Dale1995}, various researchers have added
-extensions such as reference to sets \cite{Stone2000}, more expressive
-logical connectives \cite{deemter01:_gener_refer_expres}, and
-relational expressions \cite{dale91:_gener_refer_expres_invol_relat}.
+The generation of referring expressions (GRE) is one of the most
+active and successful research areas in natural language generation.
+Starting with Dale and Reiter's incremental algorithm \cite{Dale1995},
+various researchers have added extensions such as reference to sets
+\cite{Stone2000}, more expressive logical connectives
+\cite{deemter01:_gener_refer_expres}, and relational expressions
+\cite{dale91:_gener_refer_expres_invol_relat}.
 

### 2008-03-21 trunk/Papers/dl-gre-08/discussion.tex

@@
 
-\subsection{Evaluation: Output Quality}
+\subsection{Evaluation: output quality}
 
@@
 the relational algorithm \cite{dale91:_gener_refer_expres_invol_relat}
-on this corpus.  Because all these algorithms are sensitive to a
-predefined ordering in which predicates and relations are added to the
-descriptions, their evaluation methodology is to try all possible
-orderings and evaluate what percentage of descriptions an algorithm
-can generate with any ordering.  It turned out that the Incremental
-Algorithm is capable of generating 98 of the 103 non-relational
-descriptions (using four different orderings to cover all the data).
-However, the Dale and Haddock algorithm was unable to generate even a
-single one of the human-generated relational descriptions.
+on this corpus. The Incremental Algorithm is dependent on a predefined
+ordering in which properties are added to the description. Viethen and
+Dale, therefore, try all possible orderings and evaluate what
+percentage of descriptions an algorithm can generate with any of
+them. The Full Brevity and the Relational Algorithms choose properties
+based on their disciminatory power and only use the orderings as tie
+breakers. Viethen and Dale found that the Incremental Algorithm is
+capable of generating 98 of the 103 non-relational
+descriptions. However, the Dale and Haddock algorithm was unable to
+generate even a single one of the human-generated relational
+descriptions.
+
+%(using four different orderings to cover all the data).
 
@@
 We replicated Viethen and Dale's experiment for the \el\ algorithm
-presented above.  In the non-relational case, our algorithm generates
-exactly the same 98 correct descriptions as the Incremental Algorithm;
-this is because it performs essentially the same computation if there
-are no relations.  To evaluate the relational case, we first removed
-the $\mathsf{column}$ and $\mathsf{row}$ predicates from the domain,
-because these predicates make it possible to refer to each drawer with
-non-relational descriptions, and our algorithm would always prefer
-these.  Using three (of 240) different orderings, our algorithm was
-able to generate 10 of the 15 human-produced relational descriptions
-correctly.  This is shown in more detail in
+presented above. In the non-relational case, our results are the same
+as their's for the Incremental Algorithm: the \el\ algorithm generates
+98 of the 103 non-relational description; this is because the two
+algorithms perform essentially the same computations if there are no
+relations.
+
+To evaluate the relational case, we first removed the
+$\mathsf{column}$ and $\mathsf{row}$ predicates from the domain,
+because these predicates make it possible to uniquely identify each
+drawer with non-relational descriptions, and our algorithm would
+always prefer these.  Using three (of 240) different orderings, our
+algorithm was able to generate 10 of the 15 human-produced relational
+descriptions correctly.  This is shown in more detail in
 Fig.~\ref{fig:example_outputs}; notice that four of the 15
 descriptions occurred twice in the corpus.  Of the five human-produced
-escriptions that the \el\ algorithm cannot generate, two mention
+descriptions that the \el\ algorithm cannot generate, two mention
 column information, which we excluded from our domain, and three
 involve references to sets (\textit{the two blues ones in horizontal
-  sequence}/\textit{the two yellow drawers}). While our algorithm
-cannot reproduce those descriptions, it does generate other, simpler
+sequence}/\textit{the two yellow drawers}). While our algorithm cannot
+reproduce those descriptions, it does generate other, simpler
 descriptions for them.
 
-\subsection{Evaluation: Efficiency}
+\subsection{Evaluation: efficiency}
 
@@
 compute distinguishing concepts for all 16 individuals in the Viethen
-\& Dale dataset.\footnote{We measured all runtimes on a MacBook Pro
+and Dale dataset.\footnote{We measured all runtimes on a MacBook Pro
   (Intel Core 2 Duo, 2.16 GHz) running Java 1.6 beta, and allowed the
@@
 actually be realized in language.  For example, none of the concepts
-our algorithms computed in the Viethen \& Dale domain contained an
+our algorithms computed in the Viethen and Dale domain contained an
 atom that would commonly be realized as a noun; the property
@@
   \caption{Average runtimes (in ms) of the two algorithms on random
-    models of     different sizes.} 
+    models with different numbers of individuals.}%%of different sizes.} 
   \label{fig:runtimes}

### 2008-03-21 trunk/Papers/dl-gre-08/bisim.tex

@@
-\section{Description Languages and Bisimulation} \label{sec:bisim}
+\section{Description Logics and Simulation} \label{sec:bisim}
 
-The main point of this paper is to represent referring expressions
-(REs) as formulas of description logic \cite{baad:desc03}.  In order
+As we mentioned, in this paper we will represent referring expressions
+(REs) as formulas of description logic~\cite{baad:desc03}.  In order
 to make this point, we will now first define the two description
@@
 
-A key notion in studying description logics is \emph{bisimulation}.
+A key notion in studying description logics is \emph{simulation}.
 We will discuss in detail the case for $\alc$; further details can be
@@
 Given a model $\gM = (\Delta^\gM,|\cdot|^\gM)$, two elements $i$ and
-$i'$ of $\Delta^\gM$ are \emph{$\alc$-bisimilar} if there is a relation $Z
+$i'$ of $\Delta^\gM$ are \emph{$\alc$-similar} if there is a relation $Z
 \subseteq \Delta^\gM \times \Delta^\gM$ such that 

### 2008-03-21 trunk/Papers/dl-gre-08/discussion.tex

@@
 compute distinguishing concepts for all 16 individuals in the Viethen
-and Dale dataset.\footnote{We measured all runtimes on a MacBook Pro
-  (Intel Core 2 Duo, 2.16 GHz) running Java 1.6 beta, and allowed the
-  Java VM to warm up, i.e.\ just-in-time compile all bytecode, before
+and Dale dataset.\footnote{Runtimes are meassured on a MacBook Pro
+  (Intel Core 2 Duo, 2.16 GHz) running Java 1.6 beta, and we allowed the
+  Java VM to warm up, i.e., just-in-time compile all bytecode, before
   taking the measurements.}

### 2008-03-21 trunk/Papers/dl-gre-08/related.tex

@@
 
-The approach presented here fits very well into this landscape, and in
-a sense completes the picture by showing how to generate concepts for
+The approach presented here fits well into this landscape, and it completes the picture by showing how to generate concepts for
 arbitrary bisimulation classes in \alc, which combines all connectives

### 2008-03-21 trunk/Papers/dl-gre-08/related.tex

@@
   \centering
-  \begin{tabular}{l|l}
+  \begin{small}
+  \begin{tabular}{l|p{0.2\textwidth}}
     GRE algorithm & DL variant \\ \hline
@@
     \newcite{dale91:_gener_refer_expres_invol_relat} & \el \\
-    \newcite{Krahmer2003} & \el \\
     \newcite{kelleher06:_increm_gener_of_spatial_refer} & \el \\
-    \newcite{gardent02:_gener_minim_defin_descr} & $\mathcal{ELU}_{(\neg)}$
+    \newcite{gardent02:_gener_minim_defin_descr} & $\mathcal{ELU}_{(\neg)}$\\
+    \newcite{Krahmer2003} & \el\ + equality between individuals
   \end{tabular}
+  \end{small}
   \caption{DL variants used by different GRE algorithms.}
@@
 the DL fragment they use.  We already discussed some of these
-approaches in Section~\ref{sec:gre}.  Furthermore, the graphs
-\newcite{Krahmer2003} use to represent referring expressions can be
-seen as concepts of \el\ that are only satisfied at those points at
-which they can be embedded; the non-relational but negative and
-disjunctive descriptions generated by
+approaches in Section~\ref{sec:gre}.  Furthermore, the non-relational
+but negative and disjunctive descriptions generated by
 \newcite{deemter01:_gener_refer_expres} are simply concepts of
-$\mathcal{PL}$; and \newcite{gardent02:_gener_minim_defin_descr} then
+$\mathcal{PL}$; and \newcite{gardent02:_gener_minim_defin_descr}
 generalizes this into generating concepts of $\mathcal{ELU}_{(\neg)}$,
 i.e.\ \el\ plus disjunction and atomic negation.
+The graphs \newcite{Krahmer2003} use to represent referring
+expressions go beyond DL in that they can express equality between
+individuals. For example, Krahmer et al. could distinguish a dog that
+bites itself from dogs that bite other dogs; a distinction which
+cannot be made using description logic formulas. Their referring
+expressions can be seen as formulas of a hybrid logic (another related
+and well-studied family of logics \todo{citation?? or do we want to
+mention hybrid logic at all?}), more specifically the logic \el\ +
+equality between individuals.
 
-The approach presented here fits well into this landscape, and it completes the picture by showing how to generate concepts for
+The approach presented here fits well into this landscape, and it
+completes the picture by showing how to generate concepts for
 arbitrary bisimulation classes in \alc, which combines all connectives
@@
 refines a decomposition of the domain into subsets.  In this way, it
-is reminiscent of the Incremental Algorithm, and in fact the IA can be
-seen as the special case of the \el\ algorithm for the non-relational
-case.
+is reminiscent of the Incremental Algorithm, which in fact can be seen
+as the special case of the \el\ algorithm for the non-relational case.
 
-However, our approach is in contrast with
+However, unlike
 \newcite{dale91:_gener_refer_expres_invol_relat} and its successors,
-such as \newcite{kelleher06:_increm_gener_of_spatial_refer}, which
-attempt to generate a RE for a single individual for successive
-individuals in the model and therefore run a risk of infinite regress.
-Our algorithms are guaranteed to terminate (and with concepts of the
-smallest relational depth that cover all the data), and infinite
-regress is automatically excluded.  Perhaps closest in spirit is the
+such as \newcite{kelleher06:_increm_gener_of_spatial_refer}, we do not
+have to take special precautions to avoid infinite regress. While Dale
+and Haddock's algorithm attempts to generate a RE for a single
+individual for successive individuals in the model, our algorithms
+consider all individuals in parallel and are guaranteed to terminate
+with concepts of the smallest relational depth that cover all the data
+\todo{is this true?}. Perhaps closest in spirit is the
 Krahmer et al.\ graph algorithm, which also computes \el\ concepts by

### 2008-03-21 trunk/Papers/dl-gre-08/bisim.tex
Changed talk in Section 2 from bisimulation to simulation.

@@
 where $p$ is in the set of propositional symbols \prop, $R$ is in the
-set of relational symbols \rel.
-
-$\el$ is the negation-free fragment of $\alc$.
+set of relational symbols \rel. $\el$ is the negation-free fragment of $\alc$.
 
@@
 $$
-Given a model $\gM$ and an individual $i$ in the domain of $\gM$, let
+
+Given a model $\gM$ and an individual $i$ in $\gM$, let
 $\propm(i) = \{p \in \prop \mid i \in |p|^\gM\}$.
@@
 \begin{definition}
-Given a model $\gM = (\Delta^\gM,|\cdot|^\gM)$, two elements $i$ and
-$i'$ of $\Delta^\gM$ are \emph{$\alc$-similar} if there is a relation $Z
+Given a model $\gM = (\Delta^\gM,|\cdot|^\gM)$, and two individuals $i$ and $i'$ in $\gM$, an \emph{$\alc$-simulation} linking $i$ to $i'$  is a relation $Z
 \subseteq \Delta^\gM \times \Delta^\gM$ such that 
@@
 \end{itemize}
-A maximal subset of $\Delta^\gM$ all of whose elements are pairwise
-bisimilar is called an \emph{\alc-bisimulation class} of $\gM$.
+%A maximal subset of $\Delta^\gM$ all of whose elements are pairwise
+%bisimilar is called an \emph{\alc-bisimulation class} of $\gM$.
 \end{definition}
 
-The crucial property that makes bisimulation relevant for GRE is the
+We will say that an individual $i$ is \emph{\alc-similar} to $i'$
+in a given model $\gM$ if for any formula $\varphi \in \alc$ $i \in \interp{\varphi}$ then
+$i' \in \interp{\varphi}$.  Notice that, given that the language includes negation,
+the notion of $\alc$-similarity is symmetric: if $i$ is \alc-similar to $i'$ then
+$i'$ is \alc-similar to $i$.  This is not the case for all description
+logic.  In particular, the notion of $\el$-similarity is not symmetric.
+Given a model $\gM$ and an individual $i$ in $\gM$, we will write \simm$_\alc(i)$, the similarity set, to all the elements of $\gM$ which are \alc-similar to $i$.x
+
+The crucial property that makes simulation relevant for GRE is the
 following theorem \cite{blac:moda01}.
@@
 \begin{theorem}\label{bisim}
-  Given a finite model $\gM$, $i$ and $i'$ are $\alc$-bisimilar in
-  $\gM$ if and only if there is no $\alc$-formula $\varphi$ such that
-  $i \in |\varphi|^\gM$ and $i' \not \in |\varphi|^\gM$.
+  Given a finite model $\gM$ and two individuals $i$ and $i'$ in $\gM$, there is an $\alc$-simulation linking $i$ to $i'$ if and only if $i$ is $\alc$-similar to $i'$.
 \end{theorem}
 
-In other words, $\alc$-bisimilar individuals are indistinguishable in
-the $\alc$ language.  In the example in Fig.~\ref{fig:dale-haddock}a,
-the two individuals $t_1$ and $t_2$ are \alc-bisimilar; and indeed,
-they satisfy exactly the same \alc-formulas.  Although we can't go
+Notice that, because for $\alc$ the notion of similarity si symmetric, $\alc$-similar individuals are indistinguishable in the $\alc$ language.  In the example in Fig.~\ref{fig:dale-haddock}a,
+the two individuals $t_1$ and $t_2$ are \alc-similar; and indeed,
+they satisfy exactly the same \alc-formulas.
+
+Although we can't go
 into details here, it is possible to define a suitable notion of
-$\el$-bisimulation for which the analogous theorem holds
-\cite{kurt:expr99}.  In general, \el-bisimulation is a weaker notion
-than \alc-bisimulation; for instance, in Fig.~\ref{fig:dale-haddock},
-$f_1$ and $f_2$ are \el-bisimilar \todo{oops! here's the asymmetry,
-  there is a formula for $f_2$ but not for $f_1$} (no \el-formula will
-distinguish them), but we have already seen that they satisfy
-different \alc-formulas.
+$\el$-simulation for which the analogous theorem holds
+\cite{kurt:expr99}.  In general, \el-simulation is a weaker notion
+than \alc-simulation; for instance, in Fig.~\ref{fig:dale-haddock},
+$f_1$ is \el-similar to $f_2$, but not vice-versa.
 

### 2008-03-21 trunk/Papers/dl-gre-08/bisim.tex

@@
 relational first-order models $\gM = (\Delta^\gM,|\cdot|^\gM)$ where
-$\Delta^\gM$ is some non-empy set, and $|\cdot|^\gM$ is an
+$\Delta^\gM$ is a non-empy set and $|\cdot|^\gM$ is an
 interpretation function such that:
@@
 $\{f_2\}$; and the formula $\mathsf{flower} \sqcap \neg
-\exists.\mathsf{hat}$ denotes $\{f_1\}$.
+\exists \mathsf{in}.\mathsf{hat}$ denotes $\{f_1\}$.
 
@@
 Section~\ref{sec:related} are $\mathcal{CL}$ (\el\ without existential
-quantification, i.e.\ only conjunctions of atoms); $\mathcal{PL}$
-(\alc\ without existential quantification, i.e.\ propositional logic);
+quantification, i.e., only conjunctions of atoms); $\mathcal{PL}$
+(\alc\ without existential quantification, i.e., propositional logic);
 and $\mathcal{ELU}_{(\neg)}$ (\el\ plus disjunction and atomic

### 2008-03-21 trunk/Papers/dl-gre-08/gre.tex
gre with similarity sets

@@
 Now we apply description logic to GRE.  The core claim of this paper
-is that the GRE problem can be viewed as the problem of computing a
-formula of description logic whose extension is exactly a given set of
-individuals.
+is that it is natural and useful to view the GRE problem as the
+problem of computing a formula of description logic whose extension is
+a given target set of individuals.
 
@@
 \mathsf{in}. \mathsf{hat}$ denotes exactly $\{f_2\}$ that we can say
-``the flower in the hat'' to refer to $f_2$.  This generalizes the
-common view of GRE as attribute selection \cite{Dale1995}: From our
-perspective, a bag of attributes is simply a conjunction of atoms.
-However, more expressive approaches to GRE fit into our framework as
-well; for example, relational expressions as in
-\newcite{dale91:_gener_refer_expres_invol_relat} are formulas of \el.
-Different approaches will deliver different REs, and indeed will be
-able to build REs at all in different circumstances; this reflects the
-differences between the DL languages they use to represent their REs
-in.  This idea provides a unified perspective for comparing different
-GRE approaches, and we will discuss how to see several further
-approaches in Section~\ref{sec:related}.
+``the flower in the hat'' to refer to $f_2$.  This perspective
+provides a general framework into which many existing GRE approaches
+fit: Traditional attribute selection \cite{Dale1995} corresponds to
+building purely DL formulas that are conjunctions of atoms; relational
+REs as in \newcite{dale91:_gener_refer_expres_invol_relat} are
+formulas of \el; and so on.  We will pursue the idea of organizing GRE
+approaches with respect to the variant of DL they use further in
+Section~\ref{sec:related}.
 
 For the rest of this paper, we assume that we are generating a
-singular RE, i.e.\ $A$ will generally be a singleton set.  In this
-case, we will only be able to generate a formula that denotes exactly
-$A = \{a\}$ (i.e., a RE that uniquely refers to $a$) iff there is no
-other individual $b$ that is bisimilar to $a$; otherwise, any formula
-that is satisfied by $a$ is also satisfied by $b$.  This is why there
-is no RE that uniquely refers to $t_2$ in Fig.~\ref{fig:dale-haddock}a
-and only uses the relations ``on'' and ``in'' (and not their
-inverses).  In other words, $A$ must be a bisimulation class.
-Conversely, if we know that $a$ is alone in its bisimulation class, we
-know that there is a concept that describes only $a$ and nothing
-else.  This means that we can recast the problem of generating
-referring expressions as the problem of computing the bisimulation
-classes of a model.
+singular RE, i.e.\ the target referent $A$ will generally be a
+singleton set.  In this case, we will only be able to generate a
+formula that denotes exactly $A = \{a\}$ (i.e., a RE that uniquely
+refers to $a$) if there is no other individual $b$ to which $a$ is
+similar; otherwise, any formula that is satisfied by $a$ is also
+satisfied by $b$.  Conversely, if we know that $a$ is not similar to
+any other individual, then there is a concept that is satisfied by $a$
+and not by anything else; this concept can serve as a unique singular
+RE.  In other words, we can reduce the $\gL$-GRE problem for a given
+model to the problem of computing the $\gL$-similarity sets of this
+model.
 
 In the rest of this section, we will present algorithms that compute
-the bisimulation classes of a given model for \alc\ and \el, along
-with a characteristic formula for each class.  These algorithms
-efficiently compute \emph{all} bisimulation classes of a model, and
-not just the one for some target referent.  In effect, they compute
-REs for all individuals in some model at the same time.  The REs
-generated by the \el\ algorithm correspond to the relational REs
-generated e.g.\ by \newcite{dale91:_gener_refer_expres_invol_relat}
-(but without any danger of infinite regress), whereas the REs
-generated by the \alc\ algorithm can freely use negation as well.
+the similarity sets of a given model for \alc\ and \el, and for each
+similarity set, a characteristic formula that denotes exactly this
+set.  In the \alc\ case, we adapt a standard algorithm from the
+literature for computing \emph{bisimulation classes}; we will then
+further adapt this algorithm for \el.  In effect, both algorithms
+compute REs for all individuals in some model at the same time -- very
+efficiently and without any danger of infinite regress.
 
-\subsection{Bisimulation classes for \alc and \el}
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\columnwidth]{pic-dale-haddock}
+  \caption{(a) The \newcite{dale91:_gener_refer_expres_invol_relat}
+    scenario; (b) the \newcite{Stone1998a} scenario.}
+  \label{fig:dale-haddock}
+\end{figure}
 
-There is substantial literature on the topic of computing the
-\alc-bisimulation classes of a given model efficiently
+\subsection{Computing similarity sets}
+
+Theorem~\ref{bisim} states that in \alc, similarity sets correspond
+exactly to \emph{bisimulation classes}, i.e.\ maximal sets of
+individuals that can be linked by simulations.  Bisimulation classes
+have been studied extensively in the literature, and there are several
+efficient algorithms for computing \alc-bisimulation classes
 \cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
-However, these algorithms will only compute the bisimulation classes
-themselves. We extend the \newcite{hopc:algo71} algorithm in such a
-way that it computes, for each bisimulation class, also a formula that
-denotes exactly this class.  We can then use these formulas as
-representations of the REs.
+However, these algorithms will only compute the similarity sets
+themselves. Here we extend the \newcite{hopc:algo71} algorithm such
+that it computes, for each set, also a formula that denotes exactly
+this set.  We can then use these formulas as representations of the
+REs.
 
@@
 Algorithm~\ref{algo:bisim-l} and Algorithm~\ref{algo:bisim-add-alc}.
-Given a model $\gM=(\Delta^\gM, |\cdot|^\gM)$, the algorithm computes
-a set $\RE$ of $\alc$ formulas such that the set $\{|\varphi|^\gM \mid
-\varphi \in \RE\}$ is the set of \alc-bisimulation classes of $\gM$.
+Given a model $\gM=(\Delta, \interp{\cdot})$, the algorithm computes a
+set $\RE$ of $\alc$ formulas such that the set $\{\interp{\varphi}
+\mid \varphi \in \RE\}$ is the set of \alc-similarity sets of $\gM$.
 The algorithm starts with $\RE = \{\top\}$ (where $\interp{\top} =
-\Delta^\gM$), and successively refines $\RE$ by making its elements
-denote smaller and smaller sets.  It maintains the invariant that at
-the start and end of every iteration, $\{|\varphi|^\gM \mid \varphi
-\in \RE\}$ is always a partition of $\Delta^\gM$.  The algorithm
-iterates over all propositional and relational symbols of the
-signature to construct new formulas until either all formulas in $\RE$
-denote singletons (i.e., there is only one individual that satisfies
-them), or no progress has been made in the previous iteration.  In
-each iteration, it calls the procedure add$_\alc$($\varphi$, $\RE$),
-which intersects $\varphi$ with any formula $\psi \in \RE$ which does
-not denote a singleton and which is not equivalent to $\varphi$ and to
-$\neg \varphi$. In this case, it replaces $\psi$ in $\RE$ by $\psi
-\sqcap \varphi$ and $\psi \sqcap \neg \varphi$ to $\RE$.
+\Delta$), and successively refines $\RE$ by making its elements denote
+smaller and smaller sets.  It maintains the invariant that at the
+start and end of every iteration, $\{\interp{\varphi} \mid \varphi \in
+\RE\}$ is always a partition of $\Delta$.  The algorithm iterates over
+all propositional and relational symbols of the signature to construct
+new formulas until either all formulas in $\RE$ denote singletons
+(i.e., there is only one individual that satisfies them), or no
+progress has been made in the previous iteration.  In each iteration,
+it calls the procedure add$_\alc$($\varphi$, $\RE$), which intersects
+$\varphi$ with any formula $\psi \in \RE$ which does not denote a
+singleton and which is not equivalent to $\varphi$ and to $\neg
+\varphi$. In this case, it replaces $\psi$ in $\RE$ by $\psi \sqcap
+\varphi$ and $\psi \sqcap \neg \varphi$.
+
+The \alc\ algorithm is capable of computing the \alc-similarity sets
+of the model very efficiently, in time $O(nm)$ where $n$ is the number
+of individuals in the domain, and $m$ is the number of pairs of
+individuals that are related by some relation.  However, it will
+freely introduce negations in the case distinctions, which can make
+the resulting formula hard to realize (see also
+Section~\ref{sec:discussion-realization}).  This is why we also
+present an algorithm for the \el-similarity sets; \el\ corresponds to
+positive relational REs, which are generally much easier to realize.
+
+We obtain the \el\ algorithm by replacing the call to add$_{\alc}$ in
+Algorithm~\ref{algo:bisim-l} by a call to add$_{\el}$, which is
+defined in Algorithm~\ref{algo:bisim-add-el}.  As before, the
+algorithm maintains a set $\RE = \{\varphi_1,\ldots,\varphi_n\}$ of
+formulas (this time of \el) such that $\interp{\varphi_1} \cup \ldots
+\cup \interp{\varphi_n} = \Delta$, and which it refines iteratively.
+However, where the \alc\ algorithm maintains the invariant that
+$\interp{\varphi_1},\ldots,\interp{\varphi_n}$ is a partition of
+$\Delta$, we weaken this invariant to the requirement that there are
+no $m \geq 2$ pairwise different indices $1 \leq i_1,\ldots,i_m \leq
+n$ such that $\interp{\varphi_{i_1}} = \interp{\varphi_{i_2}} \cup
+\ldots \cup \interp{\varphi_{i_m}}$.  We call the formula
+$\varphi_{i_1}$ \emph{subsumed} if such a decomposition exists.
+
+Because it maintains a weaker invariant, the set $\RE$ may contain
+more formulas at the same time in the \el\ algorithm than in the \alc\
+algorithm.  Prima facie, this means that it has worst-case exponential
+runtime, given that the whole domain has an exponential number of
+subsets.  However, it is possible that a more careful analysis of the
+combinatorics of the considered subsets will reveal that the algorithm
+is actually polynomial.
 
@@
 \dontprintsemicolon
-\caption{Compute $\mathcal{L}$-bisimulation classes}
+\caption{Computing the $\mathcal{L}$-similarity sets}
 \label{algo:bisim-l}
-\KwIn{A model $\gM = (\Delta^\gM, |\cdot|^\gM)$}
+\KwIn{A model $\gM = (\Delta, \interp{\cdot})$}
 \KwOut{A set \RE of formulas  such that
-$\{|\varphi|^\gM \in \RE\}$ is the set of $\mathcal{L}$-bisimulation
-classes of $\gM$.}
+$\{\interp{\varphi} \mid \varphi \in \RE\}$ is the set of
+$\mathcal{L}$-similarity 
+sets of $\gM$.}
 
@@
 
-The \alc\ algorithm is capable of computing the \alc-bisimulation
-class of the model very efficiently, in time $O(nm)$ where $n$ is the
-number of individuals in the domain, and $m$ is the number of pairs of
-individuals that are related by some relation.  However, it will
-freely introduce negations in the case distinctions, which can make
-the resulting formula hard to realize (see also
-Section~\ref{sec:discussion-realization}).  This is why we also
-present an algorithm for the \el-bisimulation classes; \el\
-corresponds to positive relational REs, which are generally easy to
-realize.
+Notice that both algorithms try to refine $\RE$ first according to the
+propositional symbols, and then by relational expressions of
+increasing depth.  The algorithms would remain correct if we
+interleaved the addition of propositional and relational symbols;
+however, the algorithms would become slower because the propositional
+refinements would have to be percolated through the relational chains
+in subsequent iterations.
 
-We obtain the \el\ algorithm by replacing the call to add$_{\alc}$ in
-Algorithm~\ref{algo:bisim-l} by a call to add$_{\el}$, which is defined
-in Algorithm~\ref{algo:bisim-add-el}.  As before, the algorithm
-maintains a set $\RE = \{\varphi_1,\ldots,\varphi_n\}$ of formulas
-(this time of \el) such that $\interp{\varphi_1}^\gM \cup \ldots \cup
-\interp{\varphi_n}^\gM = \Delta^\gM$, and which it refines
-iteratively.  However, where the \alc\ algorithm maintains the
-invariant that $\interp{\varphi_1},\ldots,\interp{\varphi_n}$ is a
-partition of $\Delta$, we weaken this invariant to the requirement
-that there are no $m \geq 2$ pairwise different indices $1 \leq
-i_1,\ldots,i_m \leq n$ such that $\interp{\varphi_{i_1}} =
-\interp{\varphi_{i_2}} \cup \ldots \cup \interp{\varphi_{i_m}}$.  We
-call the formula $\varphi_{i_1}$ \emph{subsumed} if such a
-decomposition exists.
+\subsection{Some examples}\label{sec:examples}
 
-Because the \el\ algorithm maintains a weaker invariant for the
-classes, it may consider more of these classes simultaneously.  Prima
-facie, this means that it has worst-case exponential runtime, given
-that the whole domain has an exponential number of subsets.  However,
-it is possible that a more careful analysis of the combinatorics of
-the considered subsets will reveal that the algorithm is actually
-polynomial.
-
-% \begin{figure}
-%   \centering
-%   \includegraphics[width=\columnwidth]{pic-el-vs-alc}
-%   \caption{Illustrating the difference between \el\ and \alc.}
-%   \label{fig:el-vs-alc}
-% \end{figure}
-
-\subsection{Some Examples}\label{sec:examples}
-
-We will now see what our algorithms do in some interesting
-examples. Consider the model shown in
-Fig.~\ref{fig:dale-haddock}a, which is taken from
-\newcite{dale91:_gener_refer_expres_invol_relat}, and
-let's run the $\el$ algorithm.
+We will now see what our algorithms do for some interesting
+examples. Consider the model shown in Fig.~\ref{fig:dale-haddock}a,
+which is taken from \newcite{dale91:_gener_refer_expres_invol_relat},
+and let's run the $\el$ algorithm.
 
@@
 \label{algo:bisim-add-alc}
-\For{$\psi \in \RE$ with $|\psi|^\gM > 1$}{
-   \If{$|\psi \sqcap \varphi|^\gM \not = \emptyset$ and
-       $|\psi \sqcap \neg \varphi|^\gM \not = \emptyset$}{
+\For{$\psi \in \RE$ with $|\interp{\psi}| > 1$}{
+   \If{$\interp{\psi \sqcap \varphi} \not = \emptyset$ and
+       $\interp{\psi \sqcap \neg \varphi} \not = \emptyset$}{
          add $\psi \sqcap \varphi$ and
@@
 \label{algo:bisim-add-el}
-\For{$\psi \in \RE$ with $|\psi|^\gM > 1$}{
+\For{$\psi \in \RE$ with $|\interp{\psi}| > 1$}{
   \If{$\psi \sqcap \varphi$ is not subsumed in $\RE$ {\bf and}
-    $|\psi \sqcap \varphi|^\gM \neq \emptyset$ {\bf and}
-    $|\psi \sqcap \varphi|^\gM \neq |\psi|^\gM$}{
+    $\interp{\psi \sqcap \varphi} \neq \emptyset$ {\bf and}
+    $\interp{\psi \sqcap \varphi} \neq \interp{\psi}$}{
     add $\psi \sqcap \varphi$ to $\RE$ \;
-    remove subsumed concepts from $\RE$\;
+    remove subsumed formulas from $\RE$\;
   }
@@
 subsumed.  Not all of these formulas denote singletons; for instance,
-$|\mathsf{cup}|^\gM$ contains two individuals.  So we iterate over the
-relations to refine our formulas.  After the first iteration over the
-relations, we have $\RE = \{ \mathsf{floor}, \mathsf{bowl} \sqcap
+$\interp{\mathsf{cup}}$ contains two individuals.  So we iterate over
+the relations to refine our formulas.  After the first iteration over
+the relations, we have $\RE = \{ \mathsf{floor}, \mathsf{bowl} \sqcap
 \exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl} \sqcap \exists
@@
 
-Now we can use the distinction between the bowls to distinguish the
-cups in the second iteration.  The result of this is $\RE = \{
-\mathsf{floor}, \mathsf{bowl} \sqcap \exists
-\mathsf{on}.\mathsf{floor}, \mathsf{bowl} \sqcap \exists
-\mathsf{on}.\mathsf{table}, \mathsf{cup} \sqcap \exists
+Now we can use the split between the bowls to distinguish the cups in
+the second iteration.  The result of this is $\RE = \{ \mathsf{floor},
+\mathsf{bowl} \sqcap \exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl}
+\sqcap \exists \mathsf{on}.\mathsf{table}, \mathsf{cup} \sqcap \exists
 \mathsf{in}. (\mathsf{bowl} \sqcap \exists
@@
 and further iterations don't allow us to refine $\mathsf{table}$; so
-the algorithm terminates.  Each of the singleton bisimulation classes
-is represented by a characteristic formula; for instance, $\RE$ tells
-us that $\mathsf{cup} \sqcap \exists \mathsf{in}. (\mathsf{bowl}
-\sqcap \exists \mathsf{on}.\mathsf{table})$ is only satisfied by
-$c_2$, so we may refer to $c_2$ as ``the cup in the bowl on the
-table''.  Notice that the algorithm didn't focus on any particular
-individual; it simultaneously generated REs for all individuals except
-for the two tables.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=\columnwidth]{pic-dale-haddock}
-  \caption{(a) The \newcite{dale91:_gener_refer_expres_invol_relat}
-    scenario; (b) the \newcite{Stone1998a} scenario.}
-  \label{fig:dale-haddock}
-\end{figure}
+the algorithm terminates.  Each formula with a singleton extension
+$\{a\}$ is a unique description of $a$; for instance, $\mathsf{cup}
+\sqcap \exists \mathsf{in}. (\mathsf{bowl} \sqcap \exists
+\mathsf{on}.\mathsf{table})$ is only satisfied by $c_2$, so we may
+refer to $c_2$ as ``the cup in the bowl on the table''.  Notice that
+the algorithm didn't focus on any particular individual; it
+simultaneously generated REs for all individuals except for the two
+tables.
 
@@
 correctly identify $r_1$ as ``the rabbit in the hat'' and $f_2$ as
-``the flower in the bathtub'', it will not be able to compute a
-RE for $f_1$; indeed, the algorithm terminates with $\RE$ containing
-both $\mathsf{flower}$ and $\mathsf{flower} \sqcap \exists
-\mathsf{in}.\mathsf{hat}$.  By contrast, the \alc\ algorithm will
-split $\mathsf{flower}$ into the two new formulas $\mathsf{flower}
-\sqcap \exists \mathsf{in}.\mathsf{hat}$ and $\mathsf{flower} \sqcap
-\neg \exists \mathsf{in}.\mathsf{hat}$, generating a unique RE for
-$f_1$ as well.
+``the flower in the bathtub'', it will not be able to compute a RE for
+$f_1$ because $f_2$ is \el-similar to $f_1$.  Indeed, the algorithm
+terminates with $\RE$ containing both $\mathsf{flower}$ and
+$\mathsf{flower} \sqcap \exists \mathsf{in}.\mathsf{hat}$.  This is a
+typical pattern for asymmetrical cases of similarity in \el: If there
+are two formulas $\varphi_1$ and $\varphi_2$ in the output set with
+$\interp{\varphi_1} \subseteq \interp{\varphi_2}$, then there is
+generally some individual $b \in \interp{\varphi_2} -
+\interp{\varphi_1}$ such that all individuals in $\interp{\varphi_1}$
+are similar to $b$, but not vice versa.  By contrast, the \alc\
+algorithm can exploit the greater expressivity of \alc\ to split
+$\mathsf{flower}$ into the two new formulas $\mathsf{flower} \sqcap
+\exists \mathsf{in}.\mathsf{hat}$ and $\mathsf{flower} \sqcap \neg
+\exists \mathsf{in}.\mathsf{hat}$, generating a unique RE for $f_1$ as
+well.
 
@@
 %%% End:
-
-% \begin{algorithm}[t]
-% \dontprintsemicolon
-% \caption{\el\ bisimulation classes}
-% \label{algo:bisim-el}
-% \KwIn{A model $\gM = (\Delta^\gM, |\cdot|^\gM)$}
-% \KwOut{A set \RE of pairs $(\varphi,|\varphi|^\gM)$ such that
-% $\{S \mid (\varphi,S) \in \RE\}$ is the set of $\el$-bisimulation
-% classes of $\gM$.}
-% \RE $\leftarrow \{(\top,\Delta^\gM)\}$ \;
-% \For{$p \in \prop$}{
-%   add$_\el$(p, \RE) \;
-% }
-% \While{for some $(\varphi,S) \in \RE$, $|S| > 1$}{
-%   \For{$(\varphi,S) \in$ \RE, $R \in \rel$}{
-%     add$_\el$($\exists R.\varphi$, \RE) \;
-%   }
-%   \If{made no changes to $\RE$}{
-%     exit\;
-%   }
-% }
-% \end{algorithm}

### 2008-03-21 trunk/Papers/dl-gre-08/introduction.tex
intro

@@
 active and successful research areas in natural language generation.
-Starting with Dale and Reiter's incremental algorithm \cite{Dale1995},
+Building upon Dale and Reiter's work \cite{dale89cooking,Dale1995},
 various researchers have added extensions such as reference to sets
@@
 
-In this paper, we propose a fundamentally new way of looking at the
-problem of generating referring expressions: We take a referring
-expression describing an individual $i$ to be a formula of some
-description logic (DL) which is satisfied by $i$ and no other
-individual in the universe.  This allows us to organize existing
-approaches to GRE in an expressiveness hierarchy.  For instance, the
-classical Dale \& Reiter algorithms compute concepts in a description
-logic that consists only of conjunctions of atoms, whereas Dale \&
-Haddock and other relational algorithms compute concepts in the logic
-\el, which permits conjunction and existential quantification.
+In this paper, we propose to view the problem of generating referring
+expressions (REs) as the problem of computing a formula of description
+logic (DL) that denotes exactly the set of individuals that we want to
+refer to.  This very natural idea has been mentioned in passing before
+\cite{Krahmer2003,gardent07:_gener_bridg_defin_descr}; however, we
+take it one step further by proposing DL as an interlingua for
+comparing the REs produced by different approaches to GRE.  In this
+way, we can organize existing GRE approaches in an expressiveness
+hierarchy.  For instance, the classical Dale and Reiter algorithms
+compute purely conjunctive formulas;
+\newcite{deemter01:_gener_refer_expres} extends this language by
+adding the other propositional connectives, whereas
+\newcite{dale91:_gener_refer_expres_invol_relat} extends it by
+allowing existential quantification.
 
-Furthermore, we show how to addapt algorithms for computing the
-\emph{bisimulation classes} of a DL model to the GRE problem.  Intuitively,
-bisimulation classes are sets of individuals that satisfy exactly the
-same concepts; if an individual is alone in its bisimulation class,
-then the concept corresponding to this class represents a uniquely
-referring expression.  We present bisimulation classes algorithms for
-the description logics \el\ (only conjunction and existential
-quantification) and \alc\ (full modal logic).  By essentially
-computing referring expressions for all individuals in the domain at
-once, these algorithms systematically avoid the infinite regress
-problem.  The \el\ algorithm, which generates the same kind of
-descriptions as Dale \& Haddock, is capable of generating 67\% of the
-relational REs in the
-\newcite{viethen06:_algor_for_gener_refer_expres} dataset correctly,
-and takes about 400 ms to compute relational REs for all 100
-individuals in a random scene.
+Furthermore, the view of GRE as a problem of computing DL formulas
+with a given extension allows us to apply existing algorithms for the
+latter problem to obtain efficient algorithms for GRE.  We present
+algorithms that compute such formulas for the description logics \el\
+(which allows only conjunction and existential quantification) and
+\alc\ (which also allows negation).  These algorithms effectively
+compute REs for all individuals in the domain at the same time, which
+allows them to systematically avoid the infinite regress problem.  The
+\el\ algorithm is capable of generating 67\% of the relational REs in
+the \newcite{viethen06:_algor_for_gener_refer_expres} dataset, in about
+15 milliseconds.  The \alc\ algorithm is even faster; it computes
+relational REs for all 100 individuals in a random model in 140
+milliseconds.
 
 The paper is structured as follows.  In Section~\ref{sec:bisim}, we
-will first define description logic and fundamental concepts such as
-bisimulation.  We will then show how to compute bisimulation classes
- and how to apply these algorithms to GRE in Section~\ref{sec:gre}.  In
-Section~\ref{sec:discussion}, we evaluate our algorithms and discuss
-our results.  Section~\ref{sec:related} compares our approach to
-related research; in particular, it shows how various prominent GRE
-algorithms can be seen as special cases of ours.
+will first define description logics.  We will then show how to
+generate REs by computing DL similarity sets for \alc\ and \el\ in
+Section~\ref{sec:gre}.  In Section~\ref{sec:discussion}, we evaluate
+our algorithms and discuss our results.  Section~\ref{sec:related}
+compares our approach to related research; in particular, it shows how
+various prominent GRE algorithms fit into the DL framework.
 Section~\ref{sec:conclusion} concludes and points to future work.

### 2008-03-21 trunk/Papers/dl-gre-08/conclusion.tex
removed other SPUD reference
conclusion

@@
 
-In this paper, we have presented a new perspective on GRE as the
-problem of computing concepts of description logic with given
-extensions.  We have also shown how such concepts can be computed
-efficiently by applying existing algorithms for computing bisimulation
-classes.  Our algorithms are able to generate 95\% of the
-non-relational and 67\% of the relational REs from
-\newcite{viethen06:_algor_for_gener_refer_expres}.  They are also
-extremely efficient (140 ms to generate REs for all individuals in a
-random model of size 100); to our knowledge, these are by far the
-fastest runtimes for relational GRE reported in the literature.  We
-will make our implementations available online.
+In this paper, we have explored the idea of viewing the generation of
+singular REs as the problem of computing a DL formula with a given
+extension.  We have shown how such formulas can be computed
+efficiently (for \alc\ and \el) by adapting existing algorithms for
+computing similarity sets.  The \el\ algorithm is able to generate
+95\% of the non-relational and 67\% of the relational REs from
+\newcite{viethen06:_algor_for_gener_refer_expres}.  Both algorithms
+are extremely efficient (350 ms and 140 ms respectively to generate
+relational REs for all individuals in a random model with 100
+individuals); to our knowledge, these are by far the fastest runtimes
+for relational GRE reported in the literature.  We will make our
+implementation available online.
 
@@
 exhibits, in which the individuals and their properties don't change
-much.  In such a setting, REs could be computed once when the
-application starts, and then reused throughout.  Nevertheless, they
-remain fast enough for real-time use in dynamic settings.  One
-interesting question for future research is how bisimulation
-algorithms that can incrementally update the bisimulation classes when
-the model changes, such as \todo{cite}, can be adapted to GRE.
+much.  However, even in more dynamic settings, our algorithms have a
+chance to outperform search algorithms like Dale and Haddock's in the
+average case because they can't get stuck in unproductive branches of
+the search space.  Nevertheless, one interesting question for future
+research is how to adapt bisimulation algorithms that can
+incrementally update the bisimulation classes when the model changes.
+Similarly, it would be interesting to explore how linguistic
+constraints and different attribute orderings can be taken into
+account without slowing the algorithms down, and how the algorithms
+can be adapted to compute REs for sets.
 
-One finding in our experiments with the Viethen and Dale dataset is
-that there is no single ordering that covers all human-produced
+In experimenting with the Viethen and Dale data, we found that
+there is no single ordering that covers all human-produced
 descriptions, which seems to be in contrast to Dale and Reiter's

### 2008-03-21 trunk/Papers/dl-gre-08/bisim.tex
removed simulations

@@
-\section{Description Logics and Simulation} \label{sec:bisim}
+\section{Description logics and similarity}
+\label{sec:bisim}
 
-As we mentioned, in this paper we will represent referring expressions
-(REs) as formulas of description logic~\cite{baad:desc03}.  In order
-to make this point, we will now first define the two description
-logics we will be working with -- \alc\ and \el.
+In this paper, we will represent referring expressions (REs) as
+formulas of description logic~\cite{baad:desc03}.  In order to make
+this point, we will now first define the two description logics we
+will be working with: \alc\ and \el.
 
 \begin{definition}
-Formulas $\varphi$ of $\alc$ are generated by the following grammar:
+  Formulas (or concepts) $\varphi$ of $\alc$ are generated by the
+  following grammar:
 $$
@@
 Formulas of both $\alc$ and $\el$ are interpreted in ordinary
-relational first-order models $\gM = (\Delta^\gM,|\cdot|^\gM)$ where
-$\Delta^\gM$ is a non-empy set and $|\cdot|^\gM$ is an
+relational first-order models $\gM = (\Delta^\gM,\interp{\cdot}^\gM)$ where
+$\Delta^\gM$ is a non-empy set and $\interp{\cdot}^\gM$ is an
 interpretation function such that:
@@
 \begin{array}{ccl}
-|p|^\gM & \subseteq & \Delta^\gM  \mbox{ for $p \in \prop$}\\
-|R|^\gM & \subseteq & \Delta^\gM \times \Delta^\gM  \mbox{ for $R \in \rel$}\\
-|\neg \varphi|^\gM & = & \Delta^\gM \backslash |\varphi|^\gM\\
-|\varphi \sqcap \varphi'|^\gM & = & |\varphi|^\gM \cap |\varphi'|^\gM\\
-|\exists R.\varphi|^\gM & = & \{i \mid \mbox{for some } i', (i,i') \in |R|^\gM\\
-& & \mbox{ and } i' \in |\varphi|^\gM \}.\\
+\interp{p}^\gM & \subseteq & \Delta^\gM  \mbox{ for $p \in \prop$}\\
+\interp{R}^\gM & \subseteq & \Delta^\gM \times \Delta^\gM  \mbox{ for $R \in \rel$}\\
+\interp{\neg \varphi}^\gM & = & \Delta^\gM \backslash \interp{\varphi}^\gM\\
+\interp{\varphi \sqcap \varphi'}^\gM & = & \interp{\varphi}^\gM \cap \interp{\varphi'}^\gM\\
+\interp{\exists R.\varphi}^\gM & = & \{i \mid \mbox{for some } i', (i,i') \in \interp{R}^\gM\\
+& & \mbox{ and } i' \in \interp{\varphi}^\gM \}.\\
 \end{array}
@@
 Given a model $\gM$ and an individual $i$ in $\gM$, let
-$\propm(i) = \{p \in \prop \mid i \in |p|^\gM\}$.
+$\propm(i) = \{p \in \prop \mid i \in \interp{p}^\gM\}$.
 \end{definition}
@@
 
+A key notion in description logics that we will use here is
+\emph{similarity} \todo{citation?}.  For any description logic $\gL$,
+we will say that an individual $i$ is \emph{\gL-similar} to $i'$ in a
+given model $\gM$ if for any formula $\varphi \in \gL$ such that $i
+\in \interp{\varphi}$, we also have $i' \in \interp{\varphi}$.
+Intuitively, this means that there is no $\gL$-formula that would
+distinguish $i$ from $i'$.  We say that the \emph{\gL-similarity set}
+of some individual $i$ is the set of all individuals to which $i$ is
+\gL-similar.
+
+Notice that similarity is not necessarily a symmetrical relation: For
+instance, $f_1$ is \el-similar to $f_2$ in
+Fig.~\ref{fig:dale-haddock}b, but $f_2$ is not \el-similar to $f_1$
+(it satisfies the formula $\exists \mathsf{in}.\mathsf{hat}$ and $f_1$
+doesn't).  However, \alc-similarity is a symmetrical relation because
+\alc\ contains negation; and indeed, $f_1$ is not \alc-similar to
+$f_2$ either because it satisfied $\neg \exists
+\mathsf{in}.\mathsf{hat}$.  Because \alc\ is more expressive than \el,
+it is generally possible for some individual $a$ to be \el-similar but
+not \alc-similar to some individual $b$, but not vice versa.
+
+\todo{need to cite \cite{blac:moda01,kurt:expr99} in the appropriate
+  places}
+
+\ignore{
+
+Notice
+that, given that the language includes negation, the notion of
+$\alc$-similarity is symmetric: if $i$ is \alc-similar to $i'$ then
+$i'$ is \alc-similar to $i$.  This is not the case for all description
+logic.  In particular, the notion of $\el$-similarity is not
+symmetric.  Given a model $\gM$ and an individual $i$ in $\gM$, we
+will write \simm$_\alc(i)$, the similarity set, to all the elements of
+$\gM$ which are \alc-similar to $i$.x
+
 A key notion in studying description logics is \emph{simulation}.
@@
 \begin{definition}
-Given a model $\gM = (\Delta^\gM,|\cdot|^\gM)$, and two individuals $i$ and $i'$ in $\gM$, an \emph{$\alc$-simulation} linking $i$ to $i'$  is a relation $Z
+Given a model $\gM = (\Delta^\gM,\interp{\cdot}^\gM)$, and two individuals $i$ and $i'$ in $\gM$, an \emph{$\alc$-simulation} linking $i$ to $i'$  is a relation $Z
 \subseteq \Delta^\gM \times \Delta^\gM$ such that 
@@
 \item[iii)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_1'$ such that
-$(e_1,e_1') \in |R|^\gM$ then there is $e_2'$ such that $(e_2,e_2') \in |R|^\gM$ and
+$(e_1,e_1') \in \interp{R}^\gM$ then there is $e_2'$ such that $(e_2,e_2') \in \interp{R}^\gM$ and
 $(e_1',e_2') \in Z$.\\[-1.5em]
 \item[iv)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_2'$ such that
-$(e_2,e_2') \in |R|^\gM$ then there is $e_1'$ such that $(e_1,e_1') \in |R|^\gM$ and
+$(e_2,e_2') \in \interp{R}^\gM$ then there is $e_1'$ such that $(e_1,e_1') \in \interp{R}^\gM$ and
 $(e_1',e_2') \in Z$.
@@
 \end{definition}
-
-We will say that an individual $i$ is \emph{\alc-similar} to $i'$
-in a given model $\gM$ if for any formula $\varphi \in \alc$ $i \in \interp{\varphi}$ then
-$i' \in \interp{\varphi}$.  Notice that, given that the language includes negation,
-the notion of $\alc$-similarity is symmetric: if $i$ is \alc-similar to $i'$ then
-$i'$ is \alc-similar to $i$.  This is not the case for all description
-logic.  In particular, the notion of $\el$-similarity is not symmetric.
-Given a model $\gM$ and an individual $i$ in $\gM$, we will write \simm$_\alc(i)$, the similarity set, to all the elements of $\gM$ which are \alc-similar to $i$.x
 
@@
 
+}
+
 %%% Local Variables:

### 2008-03-21 trunk/Papers/dl-gre-08/gre.tex
removed simulations

@@
 \ \ Input: & A model $\gM$ and a set $A \subseteq \Delta^\gM$.\\
-\ \ Output: & A formula $\varphi \in \gL$ such that $|\varphi|^\gM = A$\\
+\ \ Output: & A formula $\varphi \in \gL$ such that
+$\interp{\varphi}^\gM = A$\\ 
 & \hspace*{0.5cm} (if such a formula exists).\\ \hline
@@
 set.  In the \alc\ case, we adapt a standard algorithm from the
-literature for computing \emph{bisimulation classes}; we will then
+literature for computing \emph{simulation classes}; we will then
 further adapt this algorithm for \el.  In effect, both algorithms
@@
 
-Theorem~\ref{bisim} states that in \alc, similarity sets correspond
-exactly to \emph{bisimulation classes}, i.e.\ maximal sets of
-individuals that can be linked by simulations.  Bisimulation classes
-have been studied extensively in the literature, and there are several
-efficient algorithms for computing \alc-bisimulation classes
+It can be shown that for \alc, the similarity sets of a finite model
+coincide exactly with the \emph{simulation classes} of this model
+\cite{blac:moda01}.  We don't have the space here to define simulation
+classes, but they have been studied extensively in the literature, and
+there are several efficient algorithms for computing \alc-simulation
+classes
 \cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
-However, these algorithms will only compute the similarity sets
+However, these algorithms will only compute the simulation classes
 themselves. Here we extend the \newcite{hopc:algo71} algorithm such
-that it computes, for each set, also a formula that denotes exactly
-this set.  We can then use these formulas as representations of the
-REs.
+that it computes, along with each set, also a formula that denotes
+exactly this set.  We can then use these formulas as representations
+of the REs.
 
@@
 
-\While{for some $\varphi \in \RE, |\varphi|^\gM>1$}{
+\While{exists some $\varphi \in \RE, |\interp{\varphi}|^\gM>1$}{
    \For{$\varphi \in \RE, R \in \rel$}{
@@
 \mathsf{in}. (\mathsf{bowl} \sqcap \exists
-\mathsf{on}.\mathsf{table}), \mathsf{table} \}$.
-
-At this point, all formulas except $\mathsf{table}$ denote singletons,
-and further iterations don't allow us to refine $\mathsf{table}$; so
-the algorithm terminates.  Each formula with a singleton extension
-$\{a\}$ is a unique description of $a$; for instance, $\mathsf{cup}
-\sqcap \exists \mathsf{in}. (\mathsf{bowl} \sqcap \exists
+\mathsf{on}.\mathsf{table}), \mathsf{table} \}$.  At this point, all
+formulas except $\mathsf{table}$ denote singletons, and further
+iterations don't allow us to refine $\mathsf{table}$; so the algorithm
+terminates.  Each formula with a singleton extension $\{a\}$ is a
+unique description of $a$; for instance, $\mathsf{cup} \sqcap \exists
+\mathsf{in}. (\mathsf{bowl} \sqcap \exists
 \mathsf{on}.\mathsf{table})$ is only satisfied by $c_2$, so we may
@@
 simultaneously generated REs for all individuals except for the two
-tables.
+tables (which are similar to each other).
 

### 2008-03-21 trunk/Papers/dl-gre-08/gre.tex
small stuff

@@
 Given a model $\gM=(\Delta, \interp{\cdot})$, the algorithm computes a
-set $\RE$ of $\alc$ formulas such that the set $\{\interp{\varphi}
-\mid \varphi \in \RE\}$ is the set of \alc-similarity sets of $\gM$.
-The algorithm starts with $\RE = \{\top\}$ (where $\interp{\top} =
+set $\RE$ of $\alc$ formulas such that $\{\interp{\varphi} \mid
+\varphi \in \RE\}$ is the set of \alc-similarity sets of $\gM$.  The
+algorithm starts with $\RE = \{\top\}$ (where $\interp{\top} =
 \Delta$), and successively refines $\RE$ by making its elements denote
@@
 \RE\}$ is always a partition of $\Delta$.  The algorithm iterates over
-all propositional and relational symbols of the signature to construct
-new formulas until either all formulas in $\RE$ denote singletons
-(i.e., there is only one individual that satisfies them), or no
-progress has been made in the previous iteration.  In each iteration,
-it calls the procedure add$_\alc$($\varphi$, $\RE$), which intersects
-$\varphi$ with any formula $\psi \in \RE$ which does not denote a
-singleton and which is not equivalent to $\varphi$ and to $\neg
-\varphi$. In this case, it replaces $\psi$ in $\RE$ by $\psi \sqcap
-\varphi$ and $\psi \sqcap \neg \varphi$.
+all propositional and relational symbols in $\prop$ and $\rel$ to
+construct new formulas until either all formulas in $\RE$ denote
+singletons (i.e., there is only one individual that satisfies them),
+or no progress has been made in the previous iteration.  In each
+iteration, it calls the procedure add$_\alc$($\varphi$, $\RE$), which
+intersects $\varphi$ with any formula $\psi \in \RE$ which does not
+denote a singleton and which is not equivalent to $\varphi$ and to
+$\neg \varphi$. In this case, it replaces $\psi$ in $\RE$ by $\psi
+\sqcap \varphi$ and $\psi \sqcap \neg \varphi$.
 
 The \alc\ algorithm is capable of computing the \alc-similarity sets
-of the model very efficiently, in time $O(nm)$ where $n$ is the number
-of individuals in the domain, and $m$ is the number of pairs of
-individuals that are related by some relation.  However, it will
-freely introduce negations in the case distinctions, which can make
-the resulting formula hard to realize (see also
-Section~\ref{sec:discussion-realization}).  This is why we also
-present an algorithm for the \el-similarity sets; \el\ corresponds to
-positive relational REs, which are generally much easier to realize.
+of the model in time $O(nm)$, where $n$ is the number of individuals
+in the domain, and $m$ is the number of pairs of individuals that are
+related by some relation.  However, it will freely introduce negations
+in the case distinctions, which can make the resulting formula hard to
+realize (see also Section~\ref{sec:discussion-realization}).  This is
+why we also present an algorithm for the \el-similarity sets; \el\
+corresponds to positive relational REs, which are generally much
+easier to realize.
 

### 2008-03-21 trunk/Papers/dl-gre-08/conclusion.tex

@@
 the search space. Nevertheless, one interesting question for future
-research is how to adapt bisimulation algorithms that can
-incrementally update the bisimulation classes when the model changes.
-Similarly, it would be interesting to explore how linguistic
-constraints and different attribute orderings can be taken into
-account without slowing the algorithms down, and how the algorithms
-can be adapted to compute REs for sets.
+research is how to incrementally update simulation classes when the
+model changes. Similarly, it would be interesting to explore how
+linguistic constraints and different attribute orderings can be taken
+into account without slowing the algorithms down, and how the
+algorithms can be adapted to compute REs for sets. In exploring these
+extensions we will be able to draw on a rich body of literature that
+considers many variants of simulation algorithms, including for
+example incremental simulation algorithms or algorithms for
+identifying sets of individuals.
+
+%Nevertheless, one interesting question for future
+%research is how to adapt simulation algorithms that can
+%incrementally update the simulation classes when the model changes.
+%Similarly, it would be interesting to explore how linguistic
+%constraints and different attribute orderings can be taken into
+%account without slowing the algorithms down, and how the algorithms
+%can be adapted to compute REs for sets.
 

### 2008-03-21 trunk/Papers/dl-gre-08/discussion.tex

--- 2008-03-21 trunk/Papers/dl-gre-08/discussion.tex
@@
 Both the \el\ and the \alc\ algorithms took about 15 milliseconds to
-compute distinguishing concepts for all 16 individuals in the Viethen
-and Dale dataset.\footnote{We measured all runtimes on a MacBook Pro
-  (Intel Core 2 Duo, 2.16 GHz) running Java 1.6 beta, and allowed the
-  Java VM to warm up, i.e.\ just-in-time compile all bytecode, before
+compute distinguishing formulas for all 16 individuals in the Viethen
+and Dale dataset.\footnote{Runtimes are meassured on a MacBook Pro
+  (Intel Core 2 Duo, 2.16 GHz) running Java 1.6 beta, and we allowed the
+  Java VM to warm up, i.e., just-in-time compile all bytecode, before
   taking the measurements.}
@@
 %\begin{itemize}
-%\item there's a risk to generate a concept that can't be realized,
+%\item there's a risk to generate a formula that can't be realized,
 %  especially for the \alc\ algorithm
@@
 
-Our GRE algorithms do not guarantee that the concepts they compute can
-actually be realized in language.  For example, none of the concepts
+Our GRE algorithms do not guarantee that the formula they compute can
+actually be realized in language.  For example, none of the formulas
 our algorithms computed in the Viethen and Dale domain contained an
@@
 massive use of negation in the \alc\ algorithm; it will be hard for
-any realizer to find a reasonable way of expressing a concept like
+any realizer to find a reasonable way of expressing a formula like
 $\neg \exists R.(\neg P \sqcap \neg Q)$ as a smooth noun phrase.
@@
 REs, these connectives must not be overused.  Thus we consider the
-concepts computed by the \el\ algorithm ``safer'' with respect to
+formulas computed by the \el\ algorithm ``safer'' with respect to
 realization.
@@
 bisimulation classes algorithms derive some measure of efficiency from
-their freedom to build concepts without having to respect any
+their freedom to build formulas without having to respect any
 linguistic constraints.  It seems straightforward, for instance, to

### 2008-03-21 trunk/Papers/dl-gre-08/gre.tex

@@
 satisfied by $b$.  Conversely, if we know that $a$ is not similar to
-any other individual, then there is a concept that is satisfied by $a$
-and not by anything else; this concept can serve as a unique singular
+any other individual, then there is a formula that is satisfied by $a$
+and not by anything else; this formula can serve as a unique singular
 RE.  In other words, we can reduce the $\gL$-GRE problem for a given

### 2008-03-21 trunk/Papers/dl-gre-08/introduction.tex

@@
 \cite{viethen06:_algor_for_gener_refer_expres}.  Furthermore, the Dale
-\& Haddock algorithm and most of its successors
+and Haddock algorithm and most of its successors
 \cite{Krahmer2003,kelleher06:_increm_gener_of_spatial_refer} are
@@
 
-In this paper, we propose to view the problem of generating referring
-expressions (REs) as the problem of computing a formula of description
-logic (DL) that denotes exactly the set of individuals that we want to
-refer to.  This very natural idea has been mentioned in passing before
+In this paper, we propose to view GRE as the problem of computing a
+formula of description logic (DL) that denotes exactly the set of
+individuals that we want to refer to.  This very natural idea has been
+mentioned in passing before
 \cite{Krahmer2003,gardent07:_gener_bridg_defin_descr}; however, we
 take it one step further by proposing DL as an interlingua for
-comparing the REs produced by different approaches to GRE.  In this
-way, we can organize existing GRE approaches in an expressiveness
-hierarchy.  For instance, the classical Dale and Reiter algorithms
-compute purely conjunctive formulas;
+comparing the referring expressions (REs) produced by different
+approaches to GRE.  In this way, we can organize existing GRE
+approaches in an expressiveness hierarchy.  For instance, the
+classical Dale and Reiter algorithms compute purely conjunctive
+formulas;
 \newcite{deemter01:_gener_refer_expres} extends this language by

### 2008-03-21 trunk/Papers/dl-gre-08/related.tex

@@
     \newcite{gardent02:_gener_minim_defin_descr} & $\mathcal{ELU}_{(\neg)}$\\
-    \newcite{Krahmer2003} & \el\ + equality between individuals
+%    \newcite{Krahmer2003} & \el\ + equality between individuals
   \end{tabular}
@@
 generalizes this into generating concepts of $\mathcal{ELU}_{(\neg)}$,
-i.e.\ \el\ plus disjunction and atomic negation.
-The graphs \newcite{Krahmer2003} use to represent referring
-expressions go beyond DL in that they can express equality between
-individuals. For example, Krahmer et al. could distinguish a dog that
-bites itself from dogs that bite other dogs; a distinction which
-cannot be made using description logic formulas. Their referring
-expressions can be seen as formulas of a hybrid logic (another related
-and well-studied family of logics \todo{citation?? or do we want to
-mention hybrid logic at all?}), more specifically the logic \el\ +
-equality between individuals.
-
+i.e., \el\ plus disjunction and atomic negation.
 The approach presented here fits well into this landscape, and it
@@
 arbitrary bisimulation classes in \alc, which combines all connectives
-used in any of these previous approaches.  Where our approach breaks
-new ground is in the way these concepts are computed: It successively
+used in any of these previous approaches.
+
+Perhaps closest in spirit to our approach is 
+Krahmer et al.\ graph algorithm~\shortcite{Krahmer2003}, which also computes concepts by
+extending them successively.  Interestingly, Krahmer et al.\ themselves
+discuss the possibility of a logic based approach by considering graph as
+models of a logical language and using logical tools.  They explicitly
+mentions hybrid logics which can be seen as very expressive description
+logics~\cite{arec:hybr05b}.
+Their algorithm, though, focuses on
+reference to a single individual and has NP-complete worst-case
+complexity.
+
+%The graphs  use to represent referring
+%expressions go beyond DL in that they can express equality between
+%individuals. For example, Krahmer et al. could distinguish a dog that
+%bites itself from dogs that bite other dogs; a distinction which
+%cannot be made using description logic formulas. Their referring
+%expressions can be seen as formulas of a hybrid logic (another related
+%and well-studied family of logics \todo{citation?? or do we want to
+%mention hybrid logic at all?}), more specifically the logic \el\ +
+%equality between individuals.
+
+Where our approach breaks new ground is in the way these concepts
+are computed: It successively
 refines a decomposition of the domain into subsets.  In this way, it
@@
 as the special case of the \el\ algorithm for the non-relational case.
-
 However, unlike
 \newcite{dale91:_gener_refer_expres_invol_relat} and its successors,
-such as \newcite{kelleher06:_increm_gener_of_spatial_refer}, we do not
-have to take special precautions to avoid infinite regress. While Dale
+we do not have to take special precautions to avoid infinite regress. While Dale
 and Haddock's algorithm attempts to generate a RE for a single
 individual for successive individuals in the model, our algorithms
-consider all individuals in parallel and are guaranteed to terminate
-with concepts of the smallest relational depth that cover all the data
-\todo{is this true?}. Perhaps closest in spirit is the
-Krahmer et al.\ graph algorithm, which also computes \el\ concepts by
-extending them successively; but this algorithm, too, is focused on
-reference to a single individual, and it has NP-complete worst-case
-complexity.
+consider all individuals in parallel and are guaranteed to always terminate.
 

### 2008-03-21 trunk/Papers/dl-gre-08/related.tex

@@
 
-Viewing GRE as a problem of generating DL concepts offers a new,
+Viewing GRE as a problem of generating DL formulas offers a new,
 unified perspective of the GRE problem: It is the problem of computing
-a DL concept with a certain (typically singleton) extension.  Many
+a DL formula with a certain (typically singleton) extension.  Many
 existing approaches can be subsumed under this view; we have
@@
 but negative and disjunctive descriptions generated by
-\newcite{deemter01:_gener_refer_expres} are simply concepts of
+\newcite{deemter01:_gener_refer_expres} are simply formulas of
 $\mathcal{PL}$; and \newcite{gardent02:_gener_minim_defin_descr}
-generalizes this into generating concepts of $\mathcal{ELU}_{(\neg)}$,
+generalizes this into generating formulas of $\mathcal{ELU}_{(\neg)}$,
 i.e., \el\ plus disjunction and atomic negation.
 The approach presented here fits well into this landscape, and it
-completes the picture by showing how to generate concepts for
+completes the picture by showing how to generate formulas for
 arbitrary bisimulation classes in \alc, which combines all connectives
@@
 Perhaps closest in spirit to our approach is 
-Krahmer et al.\ graph algorithm~\shortcite{Krahmer2003}, which also computes concepts by
+Krahmer et al.\ graph algorithm~\shortcite{Krahmer2003}, which also computes formulas by
 extending them successively.  Interestingly, Krahmer et al.\ themselves
@@
 
-Where our approach breaks new ground is in the way these concepts
+Where our approach breaks new ground is in the way these formulas
 are computed: It successively

### 2008-03-21 trunk/Papers/dl-gre-08/gre.tex
minor edits

@@
 fit: Traditional attribute selection \cite{Dale1995} corresponds to
-building purely DL formulas that are conjunctions of atoms; relational
-REs as in \newcite{dale91:_gener_refer_expres_invol_relat} are
-formulas of \el; and so on.  We will pursue the idea of organizing GRE
-approaches with respect to the variant of DL they use further in
+building DL formulas that are conjunctions of atoms; relational REs as
+in \newcite{dale91:_gener_refer_expres_invol_relat} are formulas of
+\el; and so on.  We will pursue the idea of organizing GRE approaches
+with respect to the variant of DL they use further in
 Section~\ref{sec:related}.
@@
 For the rest of this paper, we assume that we are generating a
-singular RE, i.e.\ the target referent $A$ will generally be a
-singleton set.  In this case, we will only be able to generate a
-formula that denotes exactly $A = \{a\}$ (i.e., a RE that uniquely
-refers to $a$) if there is no other individual $b$ to which $a$ is
-similar; otherwise, any formula that is satisfied by $a$ is also
-satisfied by $b$.  Conversely, if we know that $a$ is not similar to
-any other individual, then there is a formula that is satisfied by $a$
-and not by anything else; this formula can serve as a unique singular
-RE.  In other words, we can reduce the $\gL$-GRE problem for a given
-model to the problem of computing the $\gL$-similarity sets of this
-model.
+singular RE, i.e.\ the target referent $A$ will be a singleton set.
+In this case, we will only be able to generate a formula that denotes
+exactly $A = \{a\}$ (i.e., a RE that uniquely refers to $a$) if there
+is no other individual $b$ to which $a$ is similar; otherwise, any
+formula that is satisfied by $a$ is also satisfied by $b$.
+Conversely, if we know that $a$ is not similar to any other
+individual, then there is a formula that is satisfied by $a$ and not
+by anything else; this formula can serve as a unique singular RE.  In
+other words, we can reduce the $\gL$-GRE problem for a given model to
+the problem of computing the $\gL$-similarity sets of this model.
 
@@
 
-The \alc\ algorithm is capable of computing the \alc-similarity sets
-of the model in time $O(nm)$, where $n$ is the number of individuals
-in the domain, and $m$ is the number of pairs of individuals that are
-related by some relation.  However, it will freely introduce negations
-in the case distinctions, which can make the resulting formula hard to
-realize (see also Section~\ref{sec:discussion-realization}).  This is
-why we also present an algorithm for the \el-similarity sets; \el\
+The \alc\ algorithm computes the \alc-similarity sets of the model in
+time $O(nm)$, where $n$ is the number of individuals in the domain,
+and $m$ is the number of pairs of individuals that are related by some
+relation.  However, it will freely introduce negations in the case
+distinctions, which can make the resulting formula hard to realize
+(see also Section~\ref{sec:discussion-realization}).  This is why we
+also present an algorithm for the \el-similarity sets; \el\
 corresponds to positive relational REs, which are generally much
@@
 ``the flower in the bathtub'', it will not be able to compute a RE for
-$f_1$ because $f_2$ is \el-similar to $f_1$.  Indeed, the algorithm
+$f_1$ because $f_1$ is \el-similar to $f_2$.  Indeed, the algorithm
 terminates with $\RE$ containing both $\mathsf{flower}$ and

### 2008-03-21 trunk/Papers/dl-gre-08/discussion.tex
capitalization
minor edits\

@@
 
-\subsection{Evaluation: output quality}
+\subsection{Evaluation: Output quality}
 
@@
 them. The Full Brevity and the Relational Algorithms choose properties
-based on their disciminatory power and only use the orderings as tie
+based on their discriminatory power and only use the orderings as tie
 breakers. Viethen and Dale found that the Incremental Algorithm is
@@
 presented above. In the non-relational case, our results are the same
-as their's for the Incremental Algorithm: the \el\ algorithm generates
-98 of the 103 non-relational description; this is because the two
+as theirs for the Incremental Algorithm: the \el\ algorithm generates
+98 of the 103 non-relational description.  This is because the two
 algorithms perform essentially the same computations if there are no
@@
 
-\subsection{Evaluation: efficiency}
+\subsection{Evaluation: Efficiency}
 

### 2008-03-21 trunk/Papers/dl-gre-08/bisim.tex
edits on the phone

@@
 
-In this paper, we will represent referring expressions (REs) as
-formulas of description logic~\cite{baad:desc03}.  In order to make
-this point, we will now first define the two description logics we
-will be working with: \alc\ and \el.
+In this paper, we will represent referring expressions as formulas of
+description logic~\cite{baad:desc03}.  In order to make this point, we
+will now define the two description logics we will be working with:
+\alc\ and \el.
 
-\begin{definition}
-  Formulas (or concepts) $\varphi$ of $\alc$ are generated by the
-  following grammar:
+\emph{Formulas} (or \emph{concepts}) $\varphi$ of $\alc$ are generated
+by the following grammar:
 $$
@@
 $$
-where $p$ is in the set of propositional symbols \prop, $R$ is in the
-set of relational symbols \rel. $\el$ is the negation-free fragment of $\alc$.
+where $p$ is in the set of propositional symbols \prop, and $R$ is in
+the set of relational symbols \rel. $\el$ is the negation-free
+fragment of $\alc$.
 
 Formulas of both $\alc$ and $\el$ are interpreted in ordinary
-relational first-order models $\gM = (\Delta^\gM,\interp{\cdot}^\gM)$ where
-$\Delta^\gM$ is a non-empy set and $\interp{\cdot}^\gM$ is an
+relational first-order models $\gM = (\Delta,\interp{\cdot})$ where
+$\Delta$ is a non-empy set and $\interp{\cdot}$ is an
 interpretation function such that:
@@
 \begin{array}{ccl}
-\interp{p}^\gM & \subseteq & \Delta^\gM  \mbox{ for $p \in \prop$}\\
-\interp{R}^\gM & \subseteq & \Delta^\gM \times \Delta^\gM  \mbox{ for $R \in \rel$}\\
-\interp{\neg \varphi}^\gM & = & \Delta^\gM \backslash \interp{\varphi}^\gM\\
-\interp{\varphi \sqcap \varphi'}^\gM & = & \interp{\varphi}^\gM \cap \interp{\varphi'}^\gM\\
-\interp{\exists R.\varphi}^\gM & = & \{i \mid \mbox{for some } i', (i,i') \in \interp{R}^\gM\\
-& & \mbox{ and } i' \in \interp{\varphi}^\gM \}.\\
+\interp{p} & \subseteq & \Delta  \mbox{ for $p \in \prop$}\\
+\interp{R} & \subseteq & \Delta \times \Delta  \mbox{ for $R \in \rel$}\\
+\interp{\neg \varphi} & = & \Delta - \interp{\varphi}\\
+\interp{\varphi \sqcap \varphi'} & = & \interp{\varphi} \cap \interp{\varphi'}\\
+\interp{\exists R.\varphi} & = & \{i \mid \mbox{for some } i', (i,i') \in \interp{R}\\
+& & \mbox{ and } i' \in \interp{\varphi} \}.\\
 \end{array}
 $$
-
-Given a model $\gM$ and an individual $i$ in $\gM$, let
-$\propm(i) = \{p \in \prop \mid i \in \interp{p}^\gM\}$.
-\end{definition}
 
@@
 Different description logics differ in the inventory of logical
-connectives they permit.  The difference between \alc\ and \el\ is
-that \alc\ permits negation and \el\ doesn't.  There are many other
-description logics in the literature; some that we will get back to in
-Section~\ref{sec:related} are $\mathcal{CL}$ (\el\ without existential
-quantification, i.e., only conjunctions of atoms); $\mathcal{PL}$
-(\alc\ without existential quantification, i.e., propositional logic);
-and $\mathcal{ELU}_{(\neg)}$ (\el\ plus disjunction and atomic
-negation).
+connectives they allow: While \alc\ permits negation, \el\
+doesn't. There are many other description logics in the literature;
+some that we will get back to in Section~\ref{sec:related} are
+$\mathcal{CL}$ (\el\ without existential quantification, i.e., only
+conjunctions of atoms); $\mathcal{PL}$ (\alc\ without existential
+quantification, i.e., propositional logic); and
+$\mathcal{ELU}_{(\neg)}$ (\el\ plus disjunction and atomic negation).
 
-A key notion in description logics that we will use here is
-\emph{similarity} \todo{citation?}.  For any description logic $\gL$,
+\todo{formula preservation} A key notion in description logics that we
+will use here is \emph{similarity}.  For any description logic $\gL$,
 we will say that an individual $i$ is \emph{\gL-similar} to $i'$ in a
@@
 \in \interp{\varphi}$, we also have $i' \in \interp{\varphi}$.
-Intuitively, this means that there is no $\gL$-formula that would
-distinguish $i$ from $i'$.  We say that the \emph{\gL-similarity set}
-of some individual $i$ is the set of all individuals to which $i$ is
-\gL-similar.
+Equivalently, there is no $\gL$-formula that holds of $i$ but not of
+$i'$.  We say that the \emph{\gL-similarity set} of some individual
+$i$ is the set of all individuals to which $i$ is \gL-similar.
 
@@
 doesn't).  However, \alc-similarity is a symmetrical relation because
-\alc\ contains negation; and indeed, $f_1$ is not \alc-similar to
-$f_2$ either because it satisfied $\neg \exists
+the language contains negation; and indeed, $f_1$ is not \alc-similar
+to $f_2$ either because it satisfied $\neg \exists
 \mathsf{in}.\mathsf{hat}$.  Because \alc\ is more expressive than \el,
@@
 not \alc-similar to some individual $b$, but not vice versa.
-
-\todo{need to cite \cite{blac:moda01,kurt:expr99} in the appropriate
-  places}
 
@@
 \begin{definition}
-Given a model $\gM = (\Delta^\gM,\interp{\cdot}^\gM)$, and two individuals $i$ and $i'$ in $\gM$, an \emph{$\alc$-simulation} linking $i$ to $i'$  is a relation $Z
-\subseteq \Delta^\gM \times \Delta^\gM$ such that 
+Given a model $\gM = (\Delta,\interp{\cdot})$, and two individuals $i$ and $i'$ in $\gM$, an \emph{$\alc$-simulation} linking $i$ to $i'$  is a relation $Z
+\subseteq \Delta \times \Delta$ such that 
 \begin{itemize}
@@
 \item[iii)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_1'$ such that
-$(e_1,e_1') \in \interp{R}^\gM$ then there is $e_2'$ such that $(e_2,e_2') \in \interp{R}^\gM$ and
+$(e_1,e_1') \in \interp{R}$ then there is $e_2'$ such that $(e_2,e_2') \in \interp{R}$ and
 $(e_1',e_2') \in Z$.\\[-1.5em]
 \item[iv)] if $(e_1,e_2) \in Z$ and for some $R \in \rel$ there is $e_2'$ such that
-$(e_2,e_2') \in \interp{R}^\gM$ then there is $e_1'$ such that $(e_1,e_1') \in \interp{R}^\gM$ and
+$(e_2,e_2') \in \interp{R}$ then there is $e_1'$ such that $(e_1,e_1') \in \interp{R}$ and
 $(e_1',e_2') \in Z$.
 \end{itemize}
-%A maximal subset of $\Delta^\gM$ all of whose elements are pairwise
+%A maximal subset of $\Delta$ all of whose elements are pairwise
 %bisimilar is called an \emph{\alc-bisimulation class} of $\gM$.

### 2008-03-21 trunk/Papers/dl-gre-08/conclusion.tex
edits on the phone

@@
 extension.  We have shown how such formulas can be computed
-efficiently (for \alc\ and \el) by adapting existing algorithms for
-computing similarity sets.  The \el\ algorithm is able to generate
-95\% of the non-relational and 67\% of the relational REs from
+efficiently (for \alc\ and \el) by adapting existing algorithms from
+the literature.  The \el\ algorithm is able to generate 95\% of the
+non-relational and 67\% of the relational REs from
 \newcite{viethen06:_algor_for_gener_refer_expres}.  Both algorithms
@@
 extensions we will be able to draw on a rich body of literature that
-considers many variants of simulation algorithms, including for
-example incremental simulation algorithms or algorithms for
-identifying sets of individuals.
+considers many variants of simulation algorithms.
 

### 2008-03-21 trunk/Papers/dl-gre-08/discussion.tex
edits on the phone

@@
 drawers had different colors and were arranged in a four-by-four grid
-(see Fig.~\ref{fig:drawers}). The human subjects used seven
-non-relational properties (the drawer's color, its $\mathsf{column}$
-and $\mathsf{row}$ number, and whether it is in a $\mathsf{corner}$)
-and five relational properties ($\mathsf{above, below, next\ to, left\
-  of, right\ of}$). Of the 118 referring expressions, only 15 use
-relations.
+(see Fig.~\ref{fig:drawers}). The human subjects used four
+non-relational properties (the drawer's $\mathsf{color}$, its
+$\mathsf{column}$ and $\mathsf{row}$ number, and whether it is in a
+$\mathsf{corner}$) and five relational properties ($\mathsf{above,
+  below, next\ to, left\ of, right\ of}$). Of the 118 referring
+expressions obtained in the experiment, only 15 use relations.
 
@@
 as theirs for the Incremental Algorithm: the \el\ algorithm generates
-98 of the 103 non-relational description.  This is because the two
-algorithms perform essentially the same computations if there are no
-relations.
+98 of the 103 non-relational descriptions, using four (of the
+possible) orderings.  This is because the two algorithms perform
+essentially the same computations if there are no relations.
 
@@
 drawer with non-relational descriptions, and our algorithm would
-always prefer these.  Using three (of 240) different orderings, our
-algorithm was able to generate 10 of the 15 human-produced relational
-descriptions correctly.  This is shown in more detail in
-Fig.~\ref{fig:example_outputs}; notice that four of the 15
-descriptions occurred twice in the corpus.  Of the five human-produced
-descriptions that the \el\ algorithm cannot generate, two mention
-column information, which we excluded from our domain, and three
-involve references to sets (\textit{the two blues ones in horizontal
-sequence}/\textit{the two yellow drawers}). While our algorithm cannot
-reproduce those descriptions, it does generate other, simpler
-descriptions for them.
+always prefer these.  Using three (of the 240 possible) different
+orderings, our algorithm was able to generate 10 of the 15
+human-produced relational descriptions correctly.  This is shown in
+more detail in Fig.~\ref{fig:example_outputs}; notice that four of the
+15 descriptions occurred twice in the corpus.  Of the five
+human-produced descriptions that the \el\ algorithm cannot generate,
+two mention column information, which we excluded from our domain, and
+three involve references to sets (\textit{the two blues ones in
+  horizontal sequence}/\textit{the two yellow drawers}). While our
+algorithm cannot reproduce those descriptions, it does generate other,
+simpler descriptions for them.
 

### 2008-03-21 trunk/Papers/dl-gre-08/gre.tex
edits on the phone

@@
 \textsc{$\gL$-GRE Problem}}\\ \hline
-\ \ Input: & A model $\gM$ and a set $A \subseteq \Delta^\gM$.\\
+\ \ Input: & A model $\gM$ and a target set $A \subseteq \Delta$.\\
 \ \ Output: & A formula $\varphi \in \gL$ such that
-$\interp{\varphi}^\gM = A$\\ 
+$\interp{\varphi} = A$\\ 
 & \hspace*{0.5cm} (if such a formula exists).\\ \hline
@@
 in \newcite{dale91:_gener_refer_expres_invol_relat} are formulas of
-\el; and so on.  We will pursue the idea of organizing GRE approaches
-with respect to the variant of DL they use further in
+\el; and so on.  We will further pursue the idea of organizing GRE
+approaches with respect to the variant of DL they use in
 Section~\ref{sec:related}.
@@
 In the rest of this section, we will present algorithms that compute
-the similarity sets of a given model for \alc\ and \el, and for each
-similarity set, a characteristic formula that denotes exactly this
-set.  In the \alc\ case, we adapt a standard algorithm from the
-literature for computing \emph{simulation classes}; we will then
-further adapt this algorithm for \el.  In effect, both algorithms
-compute REs for all individuals in some model at the same time -- very
-efficiently and without any danger of infinite regress.
+the similarity sets of a given model for \alc\ and \el, together with
+characteristic formulas that denote them.  In the \alc\ case, we adapt
+a standard algorithm from the literature for computing
+\emph{simulation classes}; we will then further adapt this algorithm
+for \el.  In effect, both algorithms compute REs for all individuals
+in some model at the same time -- very efficiently and without any
+danger of infinite regress.
 
@@
 It can be shown that for \alc, the similarity sets of a finite model
-coincide exactly with the \emph{simulation classes} of this model
-\cite{blac:moda01}.  We don't have the space here to define simulation
-classes, but they have been studied extensively in the literature, and
-there are several efficient algorithms for computing \alc-simulation
-classes
+coincide exactly with the \emph{simulation classes} of this model.
+Simulation classes have been studied extensively in the literature
+\cite{blac:moda01,kurt:expr99}, and there are several efficient
+algorithms for computing \alc-simulation classes
 \cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
@@
 The \alc\ algorithm computes the \alc-similarity sets of the model in
-time $O(nm)$, where $n$ is the number of individuals in the domain,
-and $m$ is the number of pairs of individuals that are related by some
-relation.  However, it will freely introduce negations in the case
-distinctions, which can make the resulting formula hard to realize
-(see also Section~\ref{sec:discussion-realization}).  This is why we
-also present an algorithm for the \el-similarity sets; \el\
-corresponds to positive relational REs, which are generally much
-easier to realize.
+time $O(n^3)$, where $n$ is the number of individuals in the domain.
+However, it will freely introduce negations in the case distinctions,
+which can make the resulting formula hard to realize (see also
+Section~\ref{sec:discussion-realization}).  This is why we also
+present an algorithm for the \el-similarity sets; \el\ corresponds to
+positive relational REs, which are generally much easier to realize.
 
@@
 n$ such that $\interp{\varphi_{i_1}} = \interp{\varphi_{i_2}} \cup
-\ldots \cup \interp{\varphi_{i_m}}$.  We call the formula
-$\varphi_{i_1}$ \emph{subsumed} if such a decomposition exists.
+\ldots \cup \interp{\varphi_{i_m}}$.  We call $\varphi_{i_1}$
+\emph{subsumed} if such a decomposition exists.
 
@@
 correctly identify $r_1$ as ``the rabbit in the hat'' and $f_2$ as
-``the flower in the bathtub'', it will not be able to compute a RE for
-$f_1$ because $f_1$ is \el-similar to $f_2$.  Indeed, the algorithm
+``the flower in the hat'', it will not be able to compute a RE for $f_1$
+because $f_1$ is \el-similar to $f_2$.  Indeed, the algorithm
 terminates with $\RE$ containing both $\mathsf{flower}$ and

### 2008-03-21 trunk/Papers/dl-gre-08/introduction.tex
edits on the phone

@@
 
-Referring expressions involving relations, in particular, have
+Referring expressions (REs) involving relations, in particular, have
 received increasing attention recently, particularly in the context of
@@
 take it one step further by proposing DL as an interlingua for
-comparing the referring expressions (REs) produced by different
-approaches to GRE.  In this way, we can organize existing GRE
-approaches in an expressiveness hierarchy.  For instance, the
-classical Dale and Reiter algorithms compute purely conjunctive
-formulas;
+comparing the REs produced by different approaches to GRE.  In this
+way, we can organize existing GRE approaches in an expressiveness
+hierarchy.  For instance, the classical Dale and Reiter algorithms
+compute purely conjunctive formulas;
 \newcite{deemter01:_gener_refer_expres} extends this language by

### 2008-03-21 trunk/Papers/dl-gre-08/related.tex
edits on the phone

@@
 
-Viewing GRE as a problem of generating DL formulas offers a new,
-unified perspective of the GRE problem: It is the problem of computing
-a DL formula with a certain (typically singleton) extension.  Many
-existing approaches can be subsumed under this view; we have
-summarized this for some of them in Fig.~\ref{fig:related}, along with
-the DL fragment they use.  We already discussed some of these
-approaches in Section~\ref{sec:gre}.  Furthermore, the non-relational
-but negative and disjunctive descriptions generated by
-\newcite{deemter01:_gener_refer_expres} are simply formulas of
-$\mathcal{PL}$; and \newcite{gardent02:_gener_minim_defin_descr}
-generalizes this into generating formulas of $\mathcal{ELU}_{(\neg)}$,
-i.e., \el\ plus disjunction and atomic negation.
-The approach presented here fits well into this landscape, and it
-completes the picture by showing how to generate formulas for
-arbitrary bisimulation classes in \alc, which combines all connectives
-used in any of these previous approaches.
+Viewing GRE as a problem of generating DL formulas offers a unified
+perspective of the GRE problem: It is the problem of computing a DL
+formula with a given extension.  Many existing approaches can be
+subsumed under this view; we have summarized this for some of them in
+Fig.~\ref{fig:related}, along with the DL fragment they use.  We
+already discussed some of these approaches in Section~\ref{sec:gre}.
+Furthermore, the non-relational but negative and disjunctive
+descriptions generated by \newcite{deemter01:_gener_refer_expres} are
+simply formulas of $\mathcal{PL}$; and
+\newcite{gardent02:_gener_minim_defin_descr} generalizes this into
+generating formulas of $\mathcal{ELU}_{(\neg)}$, i.e., \el\ plus
+disjunction and atomic negation.  The approach presented here fits
+well into this landscape, and it completes the picture by showing how
+to generate REs in \alc, which combines all connectives used in any of
+these previous approaches.
 

### 2008-03-21 trunk/Papers/dl-gre-08/bisim.tex
the rest of the edits we discussed

@@
 
-\todo{formula preservation} A key notion in description logics that we
-will use here is \emph{similarity}.  For any description logic $\gL$,
-we will say that an individual $i$ is \emph{\gL-similar} to $i'$ in a
-given model $\gM$ if for any formula $\varphi \in \gL$ such that $i
-\in \interp{\varphi}$, we also have $i' \in \interp{\varphi}$.
+Below, we will use a key notion of formula preservation that we call
+\emph{similarity}.  For any description logic $\gL$, we will say that
+an individual $i$ is \emph{\gL-similar} to $i'$ in a given model $\gM$
+if for any formula $\varphi \in \gL$ such that $i \in
+\interp{\varphi}$, we also have $i' \in \interp{\varphi}$.
 Equivalently, there is no $\gL$-formula that holds of $i$ but not of

### 2008-03-21 trunk/Papers/dl-gre-08/related.tex
the rest of the edits we discussed

@@
 
-Perhaps closest in spirit to our approach is 
-Krahmer et al.\ graph algorithm~\shortcite{Krahmer2003}, which also computes formulas by
-extending them successively.  Interestingly, Krahmer et al.\ themselves
-discuss the possibility of a logic based approach by considering graph as
-models of a logical language and using logical tools.  They explicitly
-mentions hybrid logics which can be seen as very expressive description
-logics~\cite{arec:hybr05b}.
-Their algorithm, though, focuses on
-reference to a single individual and has NP-complete worst-case
-complexity.
+Where our approach breaks new ground is in the way these formulas are
+computed: It successively refines a decomposition of the domain into
+subsets.  In this way, it is reminiscent of the Incremental Algorithm,
+which in fact can be seen as the special case of the \el\ algorithm
+for the non-relational case.  However, unlike
+\newcite{dale91:_gener_refer_expres_invol_relat} and its successors,
+such as \newcite{kelleher06:_increm_gener_of_spatial_refer},
+we do not have to take special precautions to avoid infinite
+regress. While Dale and Haddock's algorithm attempts to generate a RE
+for a single individual for successive individuals in the model, our
+algorithms consider all individuals in parallel and are guaranteed to
+always terminate.
+
+Perhaps closest in spirit to our approach is Krahmer et al.'s graph
+algorithm~\shortcite{Krahmer2003}, which also computes REs by
+extending them successively.  However, their subgraphs go beyond the
+expressive power of \alc\ in that they can distinguish between ``the
+dog that bites a dog'' and ``the dog that bites itself''.  The price
+they pay for this increase in expressive power is an NP-complete
+worst-case complexity.  Interestingly, Krahmer et al.\ themselves
+discuss the possibility of seeing their subgraphs as formulas of
+hybrid logic which are satisfied at the points where the subgraph can
+be embedded; and hybrid logics can be seen as very expressive
+description logics~\cite{arec:hybr05b} as well.
 
@@
 
-Where our approach breaks new ground is in the way these formulas
-are computed: It successively
-refines a decomposition of the domain into subsets.  In this way, it
-is reminiscent of the Incremental Algorithm, which in fact can be seen
-as the special case of the \el\ algorithm for the non-relational case.
-However, unlike
-\newcite{dale91:_gener_refer_expres_invol_relat} and its successors,
-we do not have to take special precautions to avoid infinite regress. While Dale
-and Haddock's algorithm attempts to generate a RE for a single
-individual for successive individuals in the model, our algorithms
-consider all individuals in parallel and are guaranteed to always terminate.
-
 %%% Local Variables: 

### 2008-03-21 trunk/Papers/dl-gre-08/bisim.tex
moved figures around

@@
 Below, we will use a key notion of formula preservation that we call
-\emph{similarity}.  For any description logic $\gL$, we will say that
+\emph{similarity}.  For any DL $\gL$, we will say that
 an individual $i$ is \emph{\gL-similar} to $i'$ in a given model $\gM$

### 2008-03-21 trunk/Papers/dl-gre-08/gre.tex
moved figures around

@@
 Simulation classes have been studied extensively in the literature
-\cite{blac:moda01,kurt:expr99}, and there are several efficient
-algorithms for computing \alc-simulation classes
+(see e.g. \newcite{blac:moda01}; \newcite{kurt:expr99}), and there are several
+efficient algorithms for computing \alc-simulation classes
 \cite{hopc:algo71,paig:thre87,dovier04:_effic_algor_for_comput_bisim_equiv}.
@@
 exactly this set.  We can then use these formulas as representations
-of the REs.
+of the referring expressions.
 
@@
 
+The algorithm starts with $\RE = \{\top\}$.  In the first loop, it
+adds the formulas $\mathsf{floor}$, $\mathsf{bowl}$, $\mathsf{cup}$,
+and $\mathsf{table}$, and then removes $\top$ because it is now
+subsumed.  Not all of these formulas denote singletons; for instance,
+$\interp{\mathsf{cup}}$ contains two individuals.  So we iterate over
+the relations to refine our formulas.  After the first iteration over
+the relations, we have $\RE = \{ \mathsf{floor}, \mathsf{bowl} \sqcap
+\exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl} \sqcap \exists
+\mathsf{on}.\mathsf{table}, \mathsf{cup}, \mathsf{table} \}$. Notice
+that $\mathsf{bowl}$ has become subsumed, but we haven't distinguished
+the cups and tables further.
+
 \begin{algorithm}[t]
@@
 \end{algorithm}
-
-The algorithm starts with $\RE = \{\top\}$.  In the first loop, it
-adds the formulas $\mathsf{floor}$, $\mathsf{bowl}$, $\mathsf{cup}$,
-and $\mathsf{table}$, and then removes $\top$ because it is now
-subsumed.  Not all of these formulas denote singletons; for instance,
-$\interp{\mathsf{cup}}$ contains two individuals.  So we iterate over
-the relations to refine our formulas.  After the first iteration over
-the relations, we have $\RE = \{ \mathsf{floor}, \mathsf{bowl} \sqcap
-\exists \mathsf{on}.\mathsf{floor}, \mathsf{bowl} \sqcap \exists
-\mathsf{on}.\mathsf{table}, \mathsf{cup}, \mathsf{table} \}$. Notice
-that $\mathsf{bowl}$ has become subsumed, but we haven't distinguished
-the cups and tables further.
 

### 2008-03-21 trunk/Papers/dl-gre-08/discussion.tex
moved figures around

@@
 
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=\columnwidth]{runtimes}
-  \caption{Average runtimes (in ms) of the two algorithms on random
-    models with different numbers of individuals.}%%of different sizes.} 
-  \label{fig:runtimes}
-\end{figure}
-
 Of course, we share the problem of interfacing GRE and realization
@@
 
-One problem that arises in our approach specifically is that the
-bisimulation classes algorithms derive some measure of efficiency from
-their freedom to build formulas without having to respect any
-linguistic constraints.  It seems straightforward, for instance, to
-extend Krahmer et al.'s \shortcite{Krahmer2003} approach such that it
-only considers subgraphs that can actually be realized, because their
-algorithm proceeds by a genuine search for uniquely identifying
-subgraphs, and will simply take a different branch of the search if
-some subgraph is useless.  This would be harder in our case.  Our
-algorithms don't search in the same way; if we disallow certain
-refinements of a partition, we have to allow the algorithms to
-backtrack and thus jeopardize the worst-case polynomial runtime.
-Investigating this interplay between efficiency and linguistic
-constraints is an interesting avenue for future research.
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\columnwidth]{runtimes}
+  \caption{Average runtimes (in ms) of the two algorithms on random
+    models with different numbers of individuals.}%%of different sizes.} 
+  \label{fig:runtimes}
+\end{figure}
+
+One problem that arises in our approach specifically is that both
+algorithms derive some measure of efficiency from their freedom to
+build formulas without having to respect any linguistic constraints.
+It seems straightforward, for instance, to extend Krahmer et al.'s
+\shortcite{Krahmer2003} approach such that it only considers subgraphs
+that can actually be realized, because their algorithm proceeds by a
+genuine search for uniquely identifying subgraphs, and will simply
+take a different branch of the search if some subgraph is useless.
+This would be harder in our case.  Our algorithms don't search in the
+same way; if we disallow certain refinements of a partition, we have
+to allow the algorithms to backtrack and thus jeopardize the
+worst-case polynomial runtime.  Investigating this interplay between
+efficiency and linguistic constraints is an interesting avenue for
+future research.
 

### 2008-03-26 trunk/Papers/rabbit-from-hat/howto.tex
structure

\section{Generating succinct expressions}
\label{sec:howto}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "rabbit-from-hat"
%%% End:

### 2008-03-27 trunk/Papers/rabbit-from-hat/howto.tex
wrote some text in sect 4-5

@@
-\section{Generating succinct expressions}
+\section{Succinct expressions in planning}
 \label{sec:howto}
+
+The primary reason for FF's failure to compute a plan is due to the
+fact that it computes all instances of all actions and literals before
+it starts the search.  This heuristic allows it to avoid performing
+unifications during the search, and is thus a good idea on the
+planning competition benchmarks, whose domain sizes are not very
+large.  However, in our case FF has to deal with potentially huge
+numbers of instances.  If we assume a maximum plan size of 10, the
+action \textbf{S-takefrom-1} alone has $31 \cdot 15^3 \cdot 4 =
+418,500$ instances -- and there are ten copies of it!  For grammars of
+more realistic sizes, it is easy to imagine how the number of
+instances might run into the trillions.
+
+\begin{itemize}
+\item PKS: don't have to instantiate
+\item new PKS: can build names on the fly by function application
+\item external modules seem a comparatively small part, but if nothing
+  else, they save an argument slot.
+\end{itemize}
 

### 2008-03-27 trunk/Papers/rabbit-from-hat/howto.tex
finished the thought in sect 5

@@
 
-\begin{itemize}
-\item PKS: don't have to instantiate
-\item new PKS: can build names on the fly by function application
-\item external modules seem a comparatively small part, but if nothing
-  else, they save an argument slot.
-\end{itemize}
+A second problem is that FF, which is a forward planner, must choose
+an instance of \textbf{S-takefrom-1} before it generates the referring
+expressions that will be realized in the two noun phrases.  However,
+at this point the planner doesn't know yet what choices of distractor
+sets will be convenient to realize in the REs.  In the above example,
+the only choice that can be realized easily is $s_3$ -- in the three
+other cases, there is no single noun that will eliminate all
+distractors --, but the planner must still try all three
+instances. \todo{is this true? or can ff do something clever with the
+  heuristic?}  In a sense, we have replaced the problems that arise
+from a pipeline model which sequences sentence planning before
+realization with the converse problems of performing realization
+before the sentence planner had a chance to suggest REs.
+
+Finally, a significant amount of runtime is spent on actually
+computing the different splits before the planner itself is run.  In
+the worst case, a domain with $n$ individuals may require us to
+distinguish \todo{what?} different splits.  These not only blow up the
+size of the initial state, but they also take \todo{a long} time to
+compute.  In our implementation, computing the initial state for four
+rabbits, hats, flowers, and bathtubs each takes about 40 minutes.
+This problem is due to the fact that we don't know for what
+individuals and relations we will need to know the splits during the
+planning process, and so we must compute all of them to be on the safe
+side.
+
+We will now solve both of these problems by switching from FF to the
+PKS planner \cite{Petrick-Bacchus:02} and extending PKS in two
+different ways: allowing terms including functions in the effects, and
+attaching external reasoning modules to the planner.  PKS computes
+instances of actions and literals only by need, when they are actually
+used by the planner, and thus automatically avoids the first problem
+mentioned above.  Furthermore, by moving all the bookkeeping about
+distractors into an external module which maintains a constraint
+network, we avoid the second and third problem as well.  The search
+for a realization is still driven by the planner; the planner then
+queries the constraint network as an external module.
+
+\todo{now we have to talk about how to actually do it}
 

### 2008-05-15 trunk/Papers/dl-gre-08/discussion.tex
autoren rein

@@
 \begin{center}
-\includegraphics[width=0.25\textwidth]{drawers.png}
+\includegraphics[width=0.2\textwidth]{drawers.png}
 \end{center}\vspace*{-3ex}
@@
 
-To evaluate the relational case, we first removed the
-$\mathsf{column}$ and $\mathsf{row}$ predicates from the domain,
-because these predicates make it possible to uniquely identify each
-drawer with non-relational descriptions, and our algorithm would
-always prefer these.  Using three (of the 240 possible) different
-orderings, our algorithm was able to generate 10 of the 15
-human-produced relational descriptions correctly.  This is shown in
-more detail in Fig.~\ref{fig:example_outputs}; notice that four of the
-15 descriptions occurred twice in the corpus.  Of the five
+%To evaluate the relational case, we first removed the
+%$\mathsf{column}$ and $\mathsf{row}$ predicates from the domain,
+%because these predicates make it possible to uniquely identify each
+%drawer with non-relational descriptions, and our algorithm would
+%always prefer these.  Using three (of the 240 possible) different
+%orderings, our algorithm was able to generate 10 of the 15
+%human-produced relational descriptions correctly.  This is shown in
+%more detail in Fig.~\ref{fig:example_outputs}; notice that four of the
+%15 descriptions occurred twice in the corpus.
+
+When we add relations, our algorithm is able to generate 10 of the 15
+human-produced relational descriptions correctly (in addition to the
+98 non-relational descriptions). This is shown in more detail in
+Fig.~\ref{fig:example_outputs}, which gives example outputs of the
+\el\ algorithm for three different orderings. (Notice that four of the
+15 descriptions occurred twice in the corpus.) Of the five
 human-produced descriptions that the \el\ algorithm cannot generate,
-two mention column information, which we excluded from our domain, and
 three involve references to sets (\textit{the two blues ones in
-  horizontal sequence}/\textit{the two yellow drawers}). While our
+  horizontal sequence}/\textit{the two yellow drawers}), and two
+contain so much redundant information that our algorithm, which stops
+once it has found a distinguishing description, cannot
+reproduce it. While the
 algorithm cannot reproduce those descriptions, it does generate other,
 simpler descriptions for them.
+
+%two mention column information, which we excluded from our domain, and
 

### 2008-05-16 trunk/Papers/dl-gre-08/bisim.tex
good except for one extra line

@@
 relational first-order models $\gM = (\Delta,\interp{\cdot})$ where
-$\Delta$ is a non-empy set and $\interp{\cdot}$ is an
+$\Delta$ is a non-empty set and $\interp{\cdot}$ is an
 interpretation function such that:
@@
 the language contains negation; and indeed, $f_1$ is not \alc-similar
-to $f_2$ either because it satisfied $\neg \exists
+to $f_2$ either because it satisfies $\neg \exists
 \mathsf{in}.\mathsf{hat}$.  Because \alc\ is more expressive than \el,
-it is generally possible for some individual $a$ to be \el-similar but
+it is  possible for some individual $a$ to be \el-similar but
 not \alc-similar to some individual $b$, but not vice versa.

### 2008-05-16 trunk/Papers/dl-gre-08/conclusion.tex
autoren rein
good except for one extra line

@@
 individuals); to our knowledge, these are by far the fastest runtimes
-for relational GRE reported in the literature.  We will make our
-implementation available online.
+for relational GRE reported in the literature.  We have made our
+implementation available online at
+\url{http://code.google.com/p/crisp-nlg/wiki/DlGre}.
 
@@
 linguistic constraints and different attribute orderings can be taken
-into account without slowing the algorithms down, and how the
-algorithms can be adapted to compute REs for sets. In exploring these
-extensions we will be able to draw on a rich body of literature that
-considers many variants of simulation algorithms.
+into account without slowing the algorithms down, how our algorithms
+could be integrated with more standard DL T-Box inferences, and how the
+algorithms can be adapted to use inverse relations or to compute REs
+for sets. In exploring these extensions we will be able to draw on a
+rich body of literature that considers many variants of simulation
+algorithms.
 
-%Nevertheless, one interesting question for future
-%research is how to adapt simulation algorithms that can
-%incrementally update the simulation classes when the model changes.
-%Similarly, it would be interesting to explore how linguistic
-%constraints and different attribute orderings can be taken into
-%account without slowing the algorithms down, and how the algorithms
-%can be adapted to compute REs for sets.
-
-In experimenting with the Viethen and Dale data, we found that
-there is no single ordering that covers all human-produced
-descriptions, which seems to be in contrast to Dale and Reiter's
+In experimenting with the Viethen and Dale data, we found that there
+is no single ordering that covers all human-produced descriptions,
+which seems to be in contrast to Dale and Reiter's
 \shortcite{Dale1995} assumption that there is only one ordering for
@@
 lend themselves very well to row/column style propositional REs.  We
-are currently preparing an experiment to collect REs in a domain in
-which relational REs will be more frequent.
+are currently collecting REs in a domain in which propositional REs
+are less preferred.
+
+\begin{small}
+\paragraph{Acknowledgments.} We are grateful to Hector Geffner (who
+independently suggested to view GRE as computation of DL formulas),
+Kees van Deemter, and Emiel Krahmer for interesting discussions.  We
+also thank Jette Viethen and Robert Dale for making their corpus
+available, and the reviewers for their comments.
+\end{small}
 

### 2008-05-16 trunk/Papers/dl-gre-08/discussion.tex
good except for one extra line

@@
 compute distinguishing formulas for all 16 individuals in the Viethen
-and Dale dataset.\footnote{Runtimes are meassured on a MacBook Pro
-  (Intel Core 2 Duo, 2.16 GHz) running Java 1.6 beta, and we allowed the
+and Dale dataset.\footnote{Runtimes were measured on a MacBook Pro
+  (Intel Core 2 Duo, 2.16 GHz) running Java 1.6 beta.  We allowed the
   Java VM to warm up, i.e., just-in-time compile all bytecode, before
@@
 
-Another encouraging aspect of both runtime graphs is that they seem to
-be approximated well by quadratic functions.  So although the
-worst-case runtime of the \el\ algorithm might be exponential, both
-algorithms empirically exhibit (low) polynomial runtime on the inputs
-we considered.
+% Another encouraging aspect of both runtime graphs is that they seem to
+% be approximated well by quadratic functions.  So although the
+% worst-case runtime of the \el\ algorithm might be exponential, both
+% algorithms empirically exhibit (low) polynomial runtime on the inputs
+% we considered.
 
@@
 $\neg \exists R.(\neg P \sqcap \neg Q)$ as a smooth noun phrase.
-Although we agree with \newcite{deemter01:_gener_refer_expres} and
+Although we agree with \newcite{deemter02:_gener_refer_expres} and
 others that the careful use of negation and disjunction can improve

### 2008-05-16 trunk/Papers/dl-gre-08/gre.tex
good except for one extra line

@@
 is that it is natural and useful to view the GRE problem as the
-problem of computing a formula of description logic whose extension is
-a given target set of individuals.
+problem of computing a formula of some description logic $\gL$ whose
+extension is a given target set $A$ of individuals.
 
@@
 For the rest of this paper, we assume that we are generating a
-singular RE, i.e.\ the target referent $A$ will be a singleton set.
-In this case, we will only be able to generate a formula that denotes
-exactly $A = \{a\}$ (i.e., a RE that uniquely refers to $a$) if there
-is no other individual $b$ to which $a$ is similar; otherwise, any
-formula that is satisfied by $a$ is also satisfied by $b$.
-Conversely, if we know that $a$ is not similar to any other
-individual, then there is a formula that is satisfied by $a$ and not
-by anything else; this formula can serve as a unique singular RE.  In
-other words, we can reduce the $\gL$-GRE problem for a given model to
-the problem of computing the $\gL$-similarity sets of this model.
+singular RE, i.e.\ the target set $A$ will be singleton.  In this
+case, we will only be able to generate a formula that denotes exactly
+$A = \{a\}$ (i.e., a RE that uniquely refers to $a$) if there is no
+other individual $b$ to which $a$ is similar; otherwise, any formula
+that is satisfied by $a$ is also satisfied by $b$.  Conversely, if we
+know that $a$ is not similar to any other individual, then there is a
+formula that is satisfied by $a$ and not by anything else; this
+formula can serve as a unique singular RE.  In other words, we can
+reduce the $\gL$-GRE problem for a given model to the problem of
+computing the $\gL$-similarity sets of this model.  Notice that this
+use of similarity sets can be seen as a generalization of van
+Deemter's \shortcite{deemter02:_gener_refer_expres} ``satellite sets''
+to relational descriptions.
 
@@
 The pseudocode for our \alc\ algorithm is shown as
-Algorithm~\ref{algo:bisim-l} and Algorithm~\ref{algo:bisim-add-alc}.
-Given a model $\gM=(\Delta, \interp{\cdot})$, the algorithm computes a
-set $\RE$ of $\alc$ formulas such that $\{\interp{\varphi} \mid
-\varphi \in \RE\}$ is the set of \alc-similarity sets of $\gM$.  The
-algorithm starts with $\RE = \{\top\}$ (where $\interp{\top} =
-\Delta$), and successively refines $\RE$ by making its elements denote
-smaller and smaller sets.  It maintains the invariant that at the
-start and end of every iteration, $\{\interp{\varphi} \mid \varphi \in
-\RE\}$ is always a partition of $\Delta$.  The algorithm iterates over
-all propositional and relational symbols in $\prop$ and $\rel$ to
-construct new formulas until either all formulas in $\RE$ denote
-singletons (i.e., there is only one individual that satisfies them),
-or no progress has been made in the previous iteration.  In each
-iteration, it calls the procedure add$_\alc$($\varphi$, $\RE$), which
-intersects $\varphi$ with any formula $\psi \in \RE$ which does not
-denote a singleton and which is not equivalent to $\varphi$ and to
-$\neg \varphi$. In this case, it replaces $\psi$ in $\RE$ by $\psi
-\sqcap \varphi$ and $\psi \sqcap \neg \varphi$.
+Algorithm~\ref{algo:bisim-l} (with $\gL=\alc$) and
+Algorithm~\ref{algo:bisim-add-alc}.  Given a model $\gM=(\Delta,
+\interp{\cdot})$, the algorithm computes a set $\RE$ of $\alc$
+formulas such that $\{\interp{\varphi} \mid \varphi \in \RE\}$ is the
+set of \alc-similarity sets of $\gM$.  The algorithm starts with $\RE
+= \{\top\}$ (where $\interp{\top} = \Delta$), and successively refines
+$\RE$ by making its elements denote smaller and smaller sets.  It
+maintains the invariant that at the start and end of every iteration,
+$\{\interp{\varphi} \mid \varphi \in \RE\}$ is always a partition of
+$\Delta$.  The algorithm iterates over all propositional and
+relational symbols in $\prop$ and $\rel$ to construct new formulas
+until either all formulas in $\RE$ denote singletons (i.e., there is
+only one individual that satisfies them), or no progress has been made
+in the previous iteration.  In each iteration, it calls the procedure
+add$_\alc$($\varphi$, $\RE$), which intersects $\varphi$ with any
+formula $\psi \in \RE$ which does not denote a singleton and which is
+not equivalent to $\varphi$ and to $\neg \varphi$. In this case, it
+replaces $\psi$ in $\RE$ by $\psi \sqcap \varphi$ and $\psi \sqcap
+\neg \varphi$.
 
@@
 more formulas at the same time in the \el\ algorithm than in the \alc\
-algorithm.  Prima facie, this means that it has worst-case exponential
-runtime, given that the whole domain has an exponential number of
-subsets.  However, it is possible that a more careful analysis of the
-combinatorics of the considered subsets will reveal that the algorithm
-is actually polynomial.
+algorithm.  Given that $\Delta$ has an exponential number of subsets,
+there is a risk that the \el\ algorithm might have worst-case
+exponential runtime (although we are not aware of such worst-case
+examples).  We leave a more careful runtime analysis for future work.
 

### 2008-05-16 trunk/Papers/dl-gre-08/introduction.tex
good except for one extra line

@@
 \cite{Stone2000}, more expressive logical connectives
-\cite{deemter01:_gener_refer_expres}, and relational expressions
+\cite{deemter02:_gener_refer_expres}, and relational expressions
 \cite{dale91:_gener_refer_expres_invol_relat}.
@@
 \cite{viethen06:_algor_for_gener_refer_expres}.  Furthermore, the Dale
-and Haddock algorithm and most of its successors
-\cite{Krahmer2003,kelleher06:_increm_gener_of_spatial_refer} are
-vulnerable to the problem of ``infinite regress'', where the algorithm
-jumps back and forth between generating descriptions for two related
-individuals infinitely, as in ``the book on the table which supports a
-book on the table \ldots''.
+and Haddock algorithm and most of its successors (such as
+\cite{kelleher06:_increm_gener_of_spatial_refer}) are vulnerable to
+the problem of ``infinite regress'', where the algorithm jumps back
+and forth between generating descriptions for two related individuals
+infinitely, as in ``the book on the table which supports a book on the
+table \ldots''.
 
@@
 compute purely conjunctive formulas;
-\newcite{deemter01:_gener_refer_expres} extends this language by
+\newcite{deemter02:_gener_refer_expres} extends this language by
 adding the other propositional connectives, whereas

### 2008-05-16 trunk/Papers/dl-gre-08/related.tex
moved figures around
good except for one extra line

@@
 \section{A unified perspective on GRE} \label{sec:related}
-
-\begin{figure}
-  \centering
-  \begin{small}
-  \begin{tabular}{l|p{0.2\textwidth}}
-    GRE algorithm & DL variant \\ \hline
-    \newcite{Dale1995} & $\mathcal{CL}$ \\
-    \newcite{deemter01:_gener_refer_expres} & $\mathcal{PL}$ \\
-    \newcite{dale91:_gener_refer_expres_invol_relat} & \el \\
-    \newcite{kelleher06:_increm_gener_of_spatial_refer} & \el \\
-    \newcite{gardent02:_gener_minim_defin_descr} & $\mathcal{ELU}_{(\neg)}$\\
-%    \newcite{Krahmer2003} & \el\ + equality between individuals
-  \end{tabular}
-  \end{small}
-  \caption{DL variants used by different GRE algorithms.}
-  \label{fig:related}
-\end{figure}
 
@@
 Furthermore, the non-relational but negative and disjunctive
-descriptions generated by \newcite{deemter01:_gener_refer_expres} are
+descriptions generated by \newcite{deemter02:_gener_refer_expres} are
 simply formulas of $\mathcal{PL}$; and
@@
 for a single individual for successive individuals in the model, our
-algorithms consider all individuals in parallel and are guaranteed to
-always terminate.
+algorithms consider all individuals in parallel.  It monotonically
+refines a partition of the model and never needs to backtrack, and
+therefore is always guaranteed to terminate.
+
+\begin{figure}
+  \centering
+  \begin{small}
+  \begin{tabular}{l|p{0.2\textwidth}}
+    GRE algorithm & DL variant \\ \hline
+    \newcite{Dale1995} & $\mathcal{CL}$ \\
+    \newcite{deemter02:_gener_refer_expres} & $\mathcal{PL}$ \\
+    \newcite{dale91:_gener_refer_expres_invol_relat} & \el \\
+    \newcite{kelleher06:_increm_gener_of_spatial_refer} & \el \\
+    \newcite{gardent02:_gener_minim_defin_descr} & $\mathcal{ELU}_{(\neg)}$\\
+%    \newcite{Krahmer2003} & \el\ + equality between individuals
+  \end{tabular}
+  \end{small}
+  \caption{DL variants used by different GRE algorithms.}
+  \label{fig:related}
+\end{figure}
 
@@
 be embedded; and hybrid logics can be seen as very expressive
-description logics~\cite{arec:hybr05b} as well.
+description logics as well \cite{arec:hybr05b}.
 

### 2008-05-16 trunk/Papers/dl-gre-08/gre.tex
fits on 8 pages; finished as far as I am concerned

@@
 
-We will now see what our algorithms do for some interesting
-examples. Consider the model shown in Fig.~\ref{fig:dale-haddock}a,
-which is taken from \newcite{dale91:_gener_refer_expres_invol_relat},
-and let's run the $\el$ algorithm.
-
-The algorithm starts with $\RE = \{\top\}$.  In the first loop, it
-adds the formulas $\mathsf{floor}$, $\mathsf{bowl}$, $\mathsf{cup}$,
-and $\mathsf{table}$, and then removes $\top$ because it is now
-subsumed.  Not all of these formulas denote singletons; for instance,
+Let's try our algorithms on some examples.  We first run the \el\
+algorithm on the model shown in Fig.~\ref{fig:dale-haddock}a, which is
+taken from \newcite{dale91:_gener_refer_expres_invol_relat}.  The
+algorithm starts with $\RE = \{\top\}$.  In the first loop, it adds
+the formulas $\mathsf{floor}$, $\mathsf{bowl}$, $\mathsf{cup}$, and
+$\mathsf{table}$, and then removes $\top$ because it is now subsumed.
+Not all of these formulas denote singletons; for instance,
 $\interp{\mathsf{cup}}$ contains two individuals.  So we iterate over

### 2008-05-16 trunk/Papers/dl-gre-08/discussion.tex

@@
 \end{center}\vspace*{-3ex}
-\caption{A schematic view of the filing cabinets.}\label{fig:drawers}\vspace*{-3ex}
+\caption{A schematic view of the filing cabinets.}\label{fig:drawers}\vspace*{-1.5ex}
 \end{figure}
@@
 human-produced relational descriptions correctly (in addition to the
-98 non-relational descriptions). This is shown in more detail in
-Fig.~\ref{fig:example_outputs}, which gives example outputs of the
-\el\ algorithm for three different orderings. (Notice that four of the
-15 descriptions occurred twice in the corpus.) Of the five
-human-produced descriptions that the \el\ algorithm cannot generate,
-three involve references to sets (\textit{the two blues ones in
-  horizontal sequence}/\textit{the two yellow drawers}), and two
-contain so much redundant information that our algorithm, which stops
-once it has found a distinguishing description, cannot
-reproduce it. While the
-algorithm cannot reproduce those descriptions, it does generate other,
-simpler descriptions for them.
+98 non-relational descriptions).  Fig.~\ref{fig:example_outputs} gives
+example outputs of the \el\ algorithm for three different orderings,
+which together achieve this coverage.  Of the five human-produced
+descriptions that the \el\ algorithm cannot generate, three involve
+references to sets (\textit{the two blues ones in horizontal
+sequence}/\textit{the two yellow drawers}), and two contain so much
+redundant information that our algorithm, which similar to the
+Incremental Algorithm allows for some redundancy but stops once it has
+found a distinguishing description, cannot reproduce them. It does,
+however, generate other, simpler descriptions for all of the
+human-produced descriptions that cannot be reproduced exactly.
 
@@
 symbol.  The results (averaged over 10 runs for each model size) are
-shown in Fig.~\ref{fig:runtimes}.  As the figure shows, the \el\
+shown in Fig.~\ref{fig:runtimes}.  The \el\
 algorithm takes about 350 ms on average to generate relational REs for
@@
   \centering
-  \includegraphics[width=\columnwidth]{runtimes}
+  \includegraphics[width=\columnwidth]{runtimes}\vspace*{-2ex}
   \caption{Average runtimes (in ms) of the two algorithms on random
-    models with different numbers of individuals.}%%of different sizes.} 
-  \label{fig:runtimes}
+    models with different numbers of individuals.}
+  \label{fig:runtimes}\vspace*{-1.5ex}
 \end{figure}
 
-One problem that arises in our approach specifically is that both
+One problem that arises in our approach is that both
 algorithms derive some measure of efficiency from their freedom to

### 2008-05-16 trunk/Papers/dl-gre-08/discussion.tex
Included my small corrections, and rewrote end of section 3.1 to address the issue propositions vs. relational symbols.

@@
 algorithm, the Incremental Algorithm (both by \newcite{Dale1995}), and
-the relational algorithm \cite{dale91:_gener_refer_expres_invol_relat}
+the Relational Algorithm \cite{dale91:_gener_refer_expres_invol_relat}
 on this corpus. The Incremental Algorithm is dependent on a predefined
@@
 capable of generating 98 of the 103 non-relational
-descriptions. However, the Dale and Haddock algorithm was unable to
+descriptions. However, the Relational Algorithm was unable to
 generate even a single one of the human-generated relational
@@
 algorithm takes about 350 ms on average to generate relational REs for
-all individuals in the model of size 100, i.e.\ less than 4 ms on
+all individuals in the model of size 100, i.e., less than 4 ms on
 average for each individual.  The \alc\ algorithm is even faster, at
@@
 Of course, we share the problem of interfacing GRE and realization
-with every other approach that separates these two modules, i.e.\
-almost the entire GRE literature (notable exceptions are e.g.\
-\newcite{Horacek1997} and SPUD \cite{Stone1998a}).  In principle, we
+with every other approach that separates these two modules, i.e.,
+almost the entire GRE literature (notable exceptions are, e.g.,~\newcite{Horacek1997} and SPUD \cite{Stone1998a}).  In principle, we
 believe that it is a good idea to handle sentence planning and

### 2008-05-16 trunk/Papers/dl-gre-08/gre.tex
Included my small corrections, and rewrote end of section 3.1 to address the issue propositions vs. relational symbols.

@@
 For the rest of this paper, we assume that we are generating a
-singular RE, i.e.\ the target set $A$ will be singleton.  In this
+singular RE, i.e., the target set $A$ will be a singleton.  In this
 case, we will only be able to generate a formula that denotes exactly
@@
     scenario; (b) the \newcite{Stone1998a} scenario.}
-  \label{fig:dale-haddock}
+  \label{fig:dale-haddock}\vspace*{-1ex}
 \end{figure}
@@
 Simulation classes have been studied extensively in the literature
-(see e.g. \newcite{blac:moda01}; \newcite{kurt:expr99}), and there are several
+(see e.g.,~\newcite{blac:moda01}; \newcite{kurt:expr99}), and there are several
 efficient algorithms for computing \alc-simulation classes
@@
 exponential runtime (although we are not aware of such worst-case
-examples).  We leave a more careful runtime analysis for future work.
+examples).  We leave a more careful complexity analysis for future work.
 
@@
 
-Notice that both algorithms try to refine $\RE$ first according to the
+Notice that we presented both algorithms as first refining $\RE$  according to the
 propositional symbols, and then by relational expressions of
-increasing depth.  The algorithms would remain correct if we
-interleaved the addition of propositional and relational symbols;
-however, the algorithms would become slower because the propositional
-refinements would have to be percolated through the relational chains
-in subsequent iterations.
+increasing depth.
+But propositional symbols can easily be encoded by means of new relation symbols (e.g., we could represent that $f_1$ is a flower in Fig.~\ref{fig:dale-haddock} as a reflexive loop labeled  $\mathsf{flower}$).
+By means of this encoding, any arbitrary ordering of preference can be used when generating $\RE$, and we don't need to distinguish between propositions and relations.
 

### 2008-05-16 trunk/Papers/dl-gre-08/introduction.tex
Included my small corrections, and rewrote end of section 3.1 to address the issue propositions vs. relational symbols.

@@
 Referring expressions (REs) involving relations, in particular, have
-received increasing attention recently, particularly in the context of
+received increasing attention recently, specially in the context of
 spatial referring expressions in situated generation (e.g.\

### 2008-05-16 trunk/Papers/dl-gre-08/related.tex
Included my small corrections, and rewrote end of section 3.1 to address the issue propositions vs. relational symbols.

@@
 Viewing GRE as a problem of generating DL formulas offers a unified
-perspective of the GRE problem: It is the problem of computing a DL
+perspective: It is the problem of computing a DL
 formula with a given extension.  Many existing approaches can be
@@
 subsets.  In this way, it is reminiscent of the Incremental Algorithm,
-which in fact can be seen as the special case of the \el\ algorithm
-for the non-relational case.  However, unlike
+which in fact can be seen as the special case of the \el\ algorithm.
+However, unlike
 \newcite{dale91:_gener_refer_expres_invol_relat} and its successors,
@@
 regress. While Dale and Haddock's algorithm attempts to generate a RE
-for a single individual for successive individuals in the model, our
+for a single individual, for successive individuals in the model, our
 algorithms consider all individuals in parallel.  It monotonically
@@
   \caption{DL variants used by different GRE algorithms.}
-  \label{fig:related}
+  \label{fig:related}\vspace*{-1.5ex}
 \end{figure}
 
-Perhaps closest in spirit to our approach is Krahmer et al.'s graph
+Perhaps closest in spirit to our approach is Krahmer et al.\ graph
 algorithm~\shortcite{Krahmer2003}, which also computes REs by
@@
 be embedded; and hybrid logics can be seen as very expressive
-description logics as well \cite{arec:hybr05b}.
+description logics \cite{arec:hybr05b}.
 

### 2008-05-16 trunk/Papers/dl-gre-08/related.tex

@@
 
-Perhaps closest in spirit to our approach is Krahmer et al.\ graph
+Perhaps closest in spirit to our approach is Krahmer et al.'s graph
 algorithm~\shortcite{Krahmer2003}, which also computes REs by

### 2008-05-16 trunk/Papers/dl-gre-08/discussion.tex
minor changes

@@
   sequence}/\textit{the two yellow drawers}), and two contain so much
-redundant information that our algorithm, which similar to the
-Incremental Algorithm allows for some redundancy but stops once it has
-found a distinguishing description, cannot reproduce them. It does,
-however, generate other, simpler descriptions for all of the
-human-produced descriptions that cannot be reproduced exactly.
+redundant information that our algorithm cannot reproduce them:
+Similarly to the Incremental Algorithm, it allows for some redundancy,
+but stops once it has found a distinguishing description. It does,
+however, generate other, simpler descriptions for these three
+referents. 
+%human-produced descriptions that cannot be reproduced exactly.
 

### 2008-05-16 trunk/Papers/dl-gre-08/gre.tex
Changed a sentence in the conclusions to put it in line with the change at the end  of section 3.1

@@
 
-Notice that we presented both algorithms as first refining $\RE$  according to the
-propositional symbols, and then by relational expressions of
-increasing depth.
-But propositional symbols can easily be encoded by means of new relation symbols (e.g., we could represent that $f_1$ is a flower in Fig.~\ref{fig:dale-haddock} as a reflexive loop labeled  $\mathsf{flower}$).
-By means of this encoding, any arbitrary ordering of preference can be used when generating $\RE$, and we don't need to distinguish between propositions and relations.
+We presented both algorithms as first refining $\RE$
+according to propositional symbols, and then by relational
+expressions of increasing depth. But, actually, propositional symbols can be encoded using new relational symbols (e.g., we could represent
+that $f_1$ is a flower in Fig.~\ref{fig:dale-haddock} as a relation labeled $\mathsf{flower}$ from $f_1$ to an additional dummy element $d$).
+In this way, any
+arbitrary ordering of preference can be used when generating $\RE$,
+and we don't need to distinguish between propositions and relations.
 

### 2008-05-16 trunk/Papers/dl-gre-08/gre.tex
final version INLG

@@
 
-We presented both algorithms as first refining $\RE$
-according to propositional symbols, and then by relational
-expressions of increasing depth. But, actually, propositional symbols can be encoded using new relational symbols (e.g., we could represent
-that $f_1$ is a flower in Fig.~\ref{fig:dale-haddock} as a relation labeled $\mathsf{flower}$ from $f_1$ to an additional dummy element $d$).
-In this way, any
-arbitrary ordering of preference can be used when generating $\RE$,
-and we don't need to distinguish between propositions and relations.
+We presented both algorithms as first refining $\RE$ according to
+propositional symbols, and then by relational expressions of
+increasing depth. But actually, propositional symbols can be encoded
+using new relational symbols (e.g., we could represent that $f_1$ is a
+flower in Fig.~\ref{fig:dale-haddock} as a relation labeled
+$\mathsf{flower}$ from $f_1$ to an additional dummy element $d$).  In
+this way, we don't need to distinguish between propositions and
+relations, and any arbitrary preference ordering of properties can be
+used.
 

### 2008-05-20 trunk/Papers/rmrs-08/abstract.tex
first dummy version

--- 2008-03-11 trunk/Papers/dl-gre-08/abstract.tex
@@
 \begin{abstract}
-  In this paper, we propose to reinterpret the problem of generating
-  referring expressions as the problem of computing a concept in a
-  description logic that is only satisfied by the referent.  This view
-  offers a new unifying perspective under which existing GRE
-  algorithms can be compared.  We also show that by applying existing
-  algorithms for computing bisimulation classes in description logic,
-  we can obtain extremely efficient algorithms for relational
-  referring expressions without any danger of running into infinite
-  regress.
+  la la la
 \end{abstract}
@@
 %%% mode: latex
-%%% TeX-master: "dl-gre-08"
+%%% TeX-master: "rmrs-08"
 %%% End:

### 2008-05-20 trunk/Papers/rmrs-08/entailment.tex
first dummy version

\section{RMRS comparison as entailment}
\label{sec:entailment}

Now that we have the formal tools in place, let's go back to the
problem of checking compatibility of semantic outputs.  We can phrase
this as RMRS entailment.  This is progress already.

Then show the representation theorem of entailment as solved form
comparison:

\begin{thm}
  Let $\varphi, \varphi'$ be two RMRSs.  Then $\varphi \models
  \varphi'$ iff for every solved form $S$ of $\varphi$, there is a
  solved form $S'$ of $\varphi'$ such that $S$ is an extension of
  $S'$.
\end{thm}

(Of course, this doesn't give us an efficient algorithm for entailment
checking, because the number of solved forms may be exponential on
each side.  And we still don't know the complexity of deciding
entailment (even for normal dominance constraints).  Nevertheless,
this is nice, because it illustrates that our model theory is both
pretty and useful for proving stuff.)

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "rmrs-08"
%%% End:

### 2008-05-20 trunk/Papers/rmrs-08/introduction.tex
first dummy version

\section{Introduction} \label{sec:introduction}

What we'll do here.  Problem: integrate semantic representations
computed by deep and shallow parsing.  We represent both in RMRS,
which we define cleanly here for the first time.  We also introduce a
model theory for RMRS.  This allows us to define ``compatibility'' of
deep and shallow semantic representations as entailment of RMRS
representations; we define entailment and characterize it
syntactically as extension of solved forms.

just so we have a citation here: \cite{Copestake_etal:MRS}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "rmrs-08"
%%% End:

### 2008-05-20 trunk/Papers/rmrs-08/motivation.tex
first dummy version

\section{Deep and shallow semantic construction}
\label{sec:motivation}

Motivate what we're doing and provide background.  Example outputs of
deep and shallow parsers.  Illustrate and state the problem.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "rmrs-08"
%%% End:

### 2008-05-20 trunk/Papers/rmrs-08/rmrs.tex
first dummy version

\section{Robust Minimal Recursion Semantics}
\label{sec:rmrs}

Define syntax of RMRS; model theory of RMRS; solved forms.  Explain
the role of the RMRS models (= trees) in the semantic construction
pipeline, in particular with respect to decoding trees into semantic
representations of the base language formalism.  Present examples in
MRS syntax first, then in our RMRS language.  Finally, compare our
version of RMRS against MRS and dominance constraints.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "rmrs-08"
%%% End:

### 2008-05-30 trunk/Papers/rmrs-08/entailment.tex
very prelim version

@@
 
-Now that we have the formal tools in place, let's go back to the
-problem of checking compatibility of semantic outputs.  We can phrase
-this as RMRS entailment.  This is progress already.
+As discussed in Sections~\ref{sec:intro} and~\ref{sec:motivation}, 
+{\sc rmrs}'s design aims to make it possible to compare
+the semantic output of several language processors of varying depth,
+and even combine them when appropriate.  The following theorem 
+demonstrates that {\sc rmrs}-entailment---defined in terms of
+models and or solved forms---provides the logical 
+basis for such parser integration:
 
-Then show the representation theorem of entailment as solved form
-comparison:
-
-\begin{thm}
-  Let $\varphi, \varphi'$ be two RMRSs.  Then $\varphi \models
+% NOTE: I'm fudging that this applies only to normal dominance
+% constraints that are overlap-free.  We should put this back if there
+% is room!
+%
+\begin{thm}\label{thm:big-one}
+  Let $\varphi, \varphi'$ be two {\sc rmrs}s.  Then $\varphi \models
   \varphi'$ iff for every solved form $S$ of $\varphi$, there is a
@@
 
-(Of course, this doesn't give us an efficient algorithm for entailment
-checking, because the number of solved forms may be exponential on
-each side.  And we still don't know the complexity of deciding
-entailment (even for normal dominance constraints).  Nevertheless,
-this is nice, because it illustrates that our model theory is both
-pretty and useful for proving stuff.)
+This theorem doesn't provide 
+efficient
+algorithms for entailment checking, and of course this would be
+required to make parser comparison and integration with {\sc rmrs}
+practical.  Computing the solved forms of $\varphi$ and $\varphi'$ may
+be exponential, and the complexity of the {\sc rmrs} validity problem
+is still unknown.  However, the above theorem illustrates that our
+model theory for {\sc rmrs} functions in the way it should for parser
+comparison and integration, and it can thus be used to prove important
+logical properties of shallow parsing systems that adopt {\sc rmrs}.
+
+The outline proof of Theorem~\ref{thm:big-one} is as follows:
+\begin{proof}
+  ``$\Leftarrow$'': Let $M,\alpha$ an arbitrary model and variable
+  assignment that satisfy $\varphi$.  There is a solved form $s$ of
+  $\varphi$ such that $M,\alpha \models s$.  By assumption, this means
+  that there is a solved form $s'$ of $\varphi'$ such that $s$ is an
+  extension of $s'$.  Therefore, $M,\alpha \models s'$ and hence
+  $M,\alpha \models \varphi'$.
+
+  ``$\Rightarrow$'': Assume that $\varphi \models \varphi'$, and let
+  $s$ be an arbitrary solved form of $\varphi$.  We need to show that
+  we can remove atoms from $s$ such that $s$ remains an extension of
+  the resulting $s'$ and $s'$ is a solved form of $\varphi'$.
+
+  We construct $s'$ as the conjunction of all atoms in $s$ that only
+  contain variables occurring in $\varphi'$ and all atom $X \dom Y$
+  such that $X$ and $Y$ occur in $\varphi'$ and $(X,Y) \in D(s)$.
+
+  It is obvious that $s$ is an extension of $s'$.  Furthermore, $s'$
+  is in solved form, because $s$ was a tree already and we didn't
+  introduce new edges.  The tricky part is to show that $s'$ is a
+  solved form of $\varphi'$.  $s'$ contains all labeling atoms of
+  $\varphi'$ according to
+  Lemma~\ref{lem:entailment-preserves-labeling-atoms}; a similar
+  argument holds for inequality atoms.  Finally, $D(\varphi')
+  \subseteq D(s')$ because $M,\alpha \models s'$ and $s'$ is
+  tree-shaped and normal and therefore contains all statements about
+  dominance that are true in $M$.
+\end{proof}
+
+\begin{lemma} \label{lem:entailment-preserves-labeling-atoms}
+  Under the assumptions of
+  Theorem~\ref{thm:big-one}, if $\varphi \models
+  \varphi'$, then $\varphi'$ only contains labeling atoms that also
+  occur in $\varphi$.  
+\end{lemma}
+\begin{proof}
+  Let $\varphi \models \varphi'$, but let $\varphi'$ contain a
+  labeling atom $X \Neq f(X_1,\ldots,X_n)$ that doesn't occur in
+  $\varphi$.  Call this atom $A$.
+
+  Furthermore, let $M,\alpha \models \varphi$ arbitrary.  Because
+  $M,\alpha \models \varphi'$, we must have $M(\alpha(X)) = f$.
+
+  Now replace the label of $\alpha(X)$ in $M$ by some other symbol $g$
+  of arity $n$.  Call the resulting model $M'$.  We still have
+  $M',\alpha \models \varphi$ because $\varphi$ doesn't contain a
+  labeling atom for $X$.  But now $M',\alpha \not\models \varphi'$.
+  This contradicts the entailment assumption.
+\end{proof}
+
+It is really necessary to state
+Theorem~\ref{thm:big-one} in terms of extensions of
+solved forms.  It is \emph{not} true that $\varphi \models \varphi'$
+implies that $\varphi$ is an extension of $\varphi'$.  A
+counterexample is shown in Fig.~\ref{fig:counterexample}, where the
+right-hand constraint entails the left-hand constraint (they are
+equivalent), but is not an extension of it.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.6]{pic-extension-counterexample}
+  \caption{A counterexample to a simpler version of Theorem~\ref{thm:big-one}.}
+  \label{fig:counterexample}
+\end{figure}
 

### 2008-05-30 trunk/Papers/rmrs-08/introduction.tex
very prelim version

@@
-\section{Introduction} \label{sec:introduction}
+\section{Introduction} \label{sec:intro}
 
-What we'll do here.  Problem: integrate semantic representations
-computed by deep and shallow parsing.  We represent both in RMRS,
-which we define cleanly here for the first time.  We also introduce a
-model theory for RMRS.  This allows us to define ``compatibility'' of
-deep and shallow semantic representations as entailment of RMRS
-representations; we define entailment and characterize it
-syntactically as extension of solved forms.
+% What we'll do here.  Problem: RMRS is designed to support building
+% semantic components to shallow parsers, and is intended to offer a way
+% of integrating semantic representations
+% computed by deep and shallow parsing.  We demonstrate the feasibility
+% of this for the first time, by introducing a model theory for RMRS,
+% defining entailment among RMRS representations and characterising this
+% entailment relation 
+% syntactically as extension of solved forms (ref to solved forms).  
+% We show that the ``compatability'' of
+% deep and shallow semantic representations is definable in terms of the
+% entailment 
+% relation between the RMRS
+% representations.
 
-just so we have a citation here: \cite{Copestake_etal:MRS}
+% just so we have a citation here: \cite{Copestake_etal:MRS}
+
+Representing semantics as a logical form that supports automated
+inference and model construction is vital for deeper language
+engineering tasks, such as dialogue systems.
+These need access to the vast array of
+semantic information that is captured in conventional logical forms,
+including (but not limited to) information about semantic scope and
+predicate argument structure.  Hand-crafted grammars yield detailed
+logical forms (e.g., \cite{butt:etal:1999,copestake:flickinger:2000}), but deep parsing tends to lack
+robustness: hand-crafted grammars fail to cover all words and
+linguistic constructions; and unedited text and speech contains
+ill-formed phrases, which by design deep grammars do not handle.  
+\hidden{
+Deep
+parsers on their own also lack ambiguity resolution techniques,
+producing all possible analyses.  Integration of statistical parse
+selection with deep parsing is promising (e.g.,
+\cite{toutanova:etal:2004}), but is not a solved problem.
+}
+
+Robost language processors that produce a single conventional logical
+form for a given natural language string are beginning to emerge
+(e.g.,
+\cite{bos:etal:2004,rupp:etal:2000,wong:mooney:2006,zettlemoyer:collins:2005}).
+But the output of these systems don't relate to any gold standard deep
+parse as produced by expert grammar developers (for instance, while
+the training corpus used in \cite{zettlemoyer:collins:2005} features
+control phenomena in the language strings, their logical forms don't
+represent it).  This makes it hard to judge the logical forms that the
+models derive from a linguistic perspective; nor can one integrate
+their output with that of a hand-crafted grammar when desired.
+
+This paper focusses on a particular approach to producing semantic
+information from robust parsers, exemplified in
+\cite{copestake:2003,frank:2004}, among others.  Their strategy is to
+utilise semantic underspecification to semi-automatically build
+semantic components to shallow parsers, so that the output neither
+over-determines nor under-determines the semantic information that is
+revealed by the (shallow) syntactic analysis.  The semantic formalism
+used to express this is Robust Minimal Recursion Semantics ({\sc
+  rmrs}, \cite{copestake:2003}); this is an extension of {\sc mrs}
+\cite{copestake:etal:2005} that is designed to be maximally flexible
+in the type of semantic information that can be left underspecified:
+it can express partial information about semantic scope, the values of
+arguments to predicate symbols and/or their argument position, the
+arity of the predicate symbols and the sorts of arguments they take.
+We show in Section~\ref{sec:motivation} that all these features are
+needed when information about lexical subcategorisation or
+syntactic dependencies is missing---a characteristic feature of
+shallow parsers.  Several researchers have demonstrated that {\sc
+  rmrs} is a suitable framework on which to semi-automatically
+construct semantic components to shallow parsers, ranging in depth
+from {\sc pos} taggers (refs) to chunk parsers (frank ref) and
+intermediate parsers ({\sc rasp} ref).
+
+A major motivation for adopting {\sc rmrs} over other techniques for
+robustly deriving logical forms is the promise that it can form the
+basis for integrating the output of several parsers, and be compared
+in particular with the output of a hand-crafted grammar.  This paper
+demonstrates the feasibility of this integration for the first time,
+by introducing a model theory for {\sc rmrs}, that in turn defines
+entailment among {\sc rmrs} representations.  This entailment relation
+is also characterised syntactically as an extension of solved forms
+(ref to solved forms).  We show that the proof theory and model theory
+of {\sc rmrs} that results provides a formal basis for integrating the
+semantic output of several shallow parsers, for checking the
+satisfiability of a shallow parse, and for testing its compatibility
+with a deep parse.
 

### 2008-05-30 trunk/Papers/rmrs-08/motivation.tex
very prelim version

@@
 
-Motivate what we're doing and provide background.  Example outputs of
-deep and shallow parsers.  Illustrate and state the problem.
+% Motivate what we're doing and provide background.  Example outputs of
+% deep and shallow parsers.  Illustrate and state the problem.
+%
+% Try to find an eg where the comparison isn't easy to see (from Anna
+% Ritchie's deliverable?). 
+% NOTE: on Ritichie: Perhaps "Three dogs bark on Tuesday in
+% June" from page 11?  It's hard to follow what's going on, from the
+% sheer size of the RMRSs and the need to allow for variation on the
+% variable names.  But I don't know how to extract this figure
+% techinically; nor whether we have room.
+
+We motivate and describe the main features of {\sc rmrs} with a (toy)
+example.  Sentence 
+(\ref{eg:cat}a) features a semantic scope ambiguity that, according to
+most deep grammars, syntax
+doesn't resolve.    The {\sc mrs} (\ref{eg:cat}b) derived by the
+English Resource Grammar ({\sc erg}, \cite{copestake:flickinger:2000})
+reflects this by underspecifying socpe: (i) each predication is
+labelled ($l_1$, $l_4$, etc); (ii) scopal arguments are indicated by
+holes (e.g., $h_2$ and $h_3$ in the quantifier {\em every}); and (iii)
+constraints on scope are expressed by the qeq conditions (roughly, $h
+=_q l$ means that only zero or more quantifiers lie between the scopal
+argument $h$ and the predication labelled by $l$ in any fully-resolved
+logical form); and (iv) these scope
+constraints allow for more than one way of `plugging' the holes with
+labels.\footnote{The event variable in the predicate symbol
+  corresponding to the adjective is motivated by copulas---{\em The
+    cat is fat} arguably has the logical form 
+  $\mbox{\_the\_q\_1}(x,\mbox{\_cat\_n\_1}(x),\mbox{\_fat\_j\_1}(e,x))$.}
+\begin{examples}
+\item   \label{eg:cat}
+\begin{subexamples}
+\item
+Every fat cat chased some dog.
+% \item
+% $\mbox{\_every\_q\_1}(x, \mbox{\_fat\_j\_1}(e',x) \wedge
+% \mbox{\_cat\_n\_1}(x),$\\
+% \hspace*{0.5in}$\mbox{\_some\_q\_1}(y, \mbox{\_dog\_n\_1}(y),
+% \mbox{\_chase\_v\_1}(e_{spast},x) )$\\
+% $\mbox{\_some\_q\_1}(y, \mbox{\_dog\_n\_1}(y),$\\
+% \hspace*{0.5in}$\mbox{\_every\_q\_1}(x, \mbox{\_fat\_j\_1}(e',x)
+% \wedge 
+% \mbox{\_cat\_n\_1}(x), 
+% \mbox{\_chase\_v\_1}(e_{spast},x)))$
+\item   $l_1\handel\mbox{\_every\_q\_1}(x,h_2,h_3)$\\
+$l_4\handel\mbox{\_fat\_j\_1}(e',x)$\\
+$l_4\handel\mbox{\_cat\_n\_1}(x)$\\
+$l_5\handel\mbox{\_some\_n\_1}(y,h_6,h_7)$\\
+$l_8\handel\mbox{\_dog\_n\_1}(y)$\\
+$l_9\handel\mbox{\_chased\_n\_1}(e,x,y)$\\
+$h_0 =_q l_9, h_2 =_q l_4, h_6 =_q l_9$
+\end{subexamples}
+\end{examples}
+(\ref{eg:cat}b) also exhibits {\sc mrs}' compact description of
+conjunction through label sharing (see $l_4$ above), and the {\sc
+  erg}'s naming convention for predicate symbols: each predicate is
+constructed from the word lemma, its syntactic category and a sense
+number, and the leading underscore identifies lexical predicates.
+
+Conceptually, this {\sc mrs} is a partial description of all its
+possible (fully specific) interpretations in context.  
+These are expressed in a so-called {\em base language}, where each
+predicate symbol in the {\sc mrs} is a proxy for one or more
+base-language constructors of the same arity
+Assuming a very close
+correspondence between the {\sc mrs} predicate symbols and the
+base-language constructors they 
+stand for, the {\sc mrs}
+(\ref{eg:cat}b) (partially) describes both the base-language formulae in
+Figure~\ref{fig:1}, shown as trees to emphasise that an {\sc mrs} is a
+partial description of the formulae's {\em form}; not its
+(base-language) entailments. 
+In essence, the labels in the {\sc mrs} reify the
+nodes in the tree, and each of these trees encodes one way of equating
+holes with labels that respects qeq constraints and binding conditions
+on the variables.
+
+\begin{figure}[h]
+
+Put trees in here
+
+\caption{The Form of alternative base logical forms of
+  (\ref{eg:cat}a).}
+\label{fig:1}
+\end{figure}
+
+The {\sc mrs} (\ref{eg:cat}c) is a notational variant of the {\sc
+  rmrs} (\ref{eg:cat}d). {\sc rmrs}s offer a more factorised
+representation: its predicate symbols are all unary and the other
+arguments to the (base-language) constructor that it corresponds to
+are represented by separate binary relations on the unique {\em
+  anchor} of the predicate symbol ($a_1, a_{41},\ldots$) together with
+variable and label equalities (e.g., $x=x_1$, $l_{41}=l_{42}$).
+\begin{examples}
+\item   [\ref{eg:cat}]
+\begin{subexamples}
+\item   [c]
+$l_1:a_1\handel\mbox{\_every\_q\_1}(x_1),$\\
+\hspace*{0.1in}$\mbox{RSTR}(a_1,h_2),
+\mbox{BODY}(a_1,h_3)$\\ 
+$l_{41}:a_{41}\handel\mbox{\_fat\_j\_1}(e'), \mbox{ARG1}(a_{41},x_2)$\\
+$l_{42}:a_{42}\handel\mbox{\_cat\_n\_1}(x_3)$\\
+$l_5:a_5\handel\mbox{\_some\_q\_1}(x_4)$,\\
+\hspace*{0.1in}$\mbox{RSTR}(a_5,h_6),
+\mbox{BODY}(a_5,h_7)$\\ 
+$l_8:a_8\handel\mbox{\_dog\_n\_1}(x_5)$\\
+$l_9:a_9\handel\mbox{\_chase\_v\_1}(e_{spast})$,\\
+\hspace*{0.1in}$\mbox{ARG1}(a_9,x_6),
+\mbox{ARG2}(a_9,x_7)$\\ 
+$h_2=_q l_{42}, l_{41}=l_{42}, h_6 =_q l_8, h_0=_q l_9$\\
+$x_1=x_2,x_2=x_3, x_4=x_5,$\\
+$x_3=x_6, x_5=x_7$
+\end{subexamples}
+\end{examples}
+
+This factored representation allows one to build semantic components
+to shallow parsers where lexical or syntactic information is absent.
+An extreme example would be a {\sc 
+  pos} tagger: one can build its semantic component simply by deriving
+lexical predicate symbols from the word lemmas and their tags and ensuring
+that the predication for each word receives a unique label, anchor and
+argument, as given in (\ref{eg:cat-pos}):
+\begin{examples}
+\item   \label{eg:cat-pos}
+\begin{subexamples}
+\item   Every\_{\sc at1} fat\_{\sc jj} cat\_{\sc nn1} chased\_{\sc
+    vvd} some\_{\sc dd} dog\_{\sc nn1}
+\item   $l_1:a_1:\mbox{\_every\_q}(x)$, \\
+$l_{41}:a_{41}:\mbox{\_fat\_a}(e')$,\\
+$l_{42}:a_{42}:\mbox{\_cat\_n}(x_3)$\\
+$l_9:a_9:\mbox{\_chased\_v}(e_{past})$, \\
+$l_5:a_5:\mbox{\_some\_q}(x_4)$, \\
+$l_8:a_8:\mbox{\_dog\_n}(x_5)$
+\end{subexamples}
+\end{examples}
+Semantic relations, sense tags and the arity of the base-language
+constructors are
+missing from (\ref{eg:cat-pos}b) because the {\sc pos} tagger doesn't
+reveal information about syntactic constituency, word sense or lexical
+subcategorisation respectively.  
+But the {\sc rmrs}s (\ref{eg:cat}c) and
+(\ref{eg:cat-pos}b) are entirely compatible, the former being more
+specific than the latter.  At least, this holds so long as the
+base-language constructors that the {\sc rmrs} predicate symbol
+{\_name\_t\_s} corresponds to is a 
+subset of those corresponding to the predicate symbol {\em \_name\_t},
+for any word lemma {\em name} and {\sc pos} tag $t$ (written 
+$\mbox{{\em \_name\_t\_s}} \sqsubset \mbox{{\em name\_t}}$).
+
+The values of non-scopal arguments are underspecified through the
+absence of ARG relations and/or through the absence of
+equalities and inequalities among the variables $x_1, x_2,\ldots$.
+The need to
+underspecify such semantic dependencies is required when
+the shallow parser doesn't record syntactic dependencies.
+Indeed, there is no single {\sc mrs} which is satisfied by both the
+trees in Figure~\ref{fig:2}, assuming of course that disjunction is not part of
+the {\sc mrs} language.  In words, these trees collectively specify
+the (partial) semantic information that $x$ is either the second
+argument (i.e., the direct object) or the third argument (i.e., the
+indirect object) to {\em give}, but we don't know which.  
+A shallow
+processor that misses long distance dependencies in
+(\ref{eg:which-dog}) cannot
+discern that the bone is the second argument to {\em
+  give} in (\ref{eg:which-dog}a), while the dog is the third argument
+in (\ref{eg:which-dog}b).
+\begin{examples}
+\item   \label{eg:which-dog}
+\begin{subexamples}
+\item   To which dog did Kim give the bone?
+\item   Which bone did Kim give the dog?
+\end{subexamples}
+\end{examples}
+Thus to maximise the semantic contribution
+of the parser while not overdetermining it, one should express that
+the argument position of the noun phrases in {\em give the bone} vs.\
+{\em give the dog} are underspecified between the second and third
+position, as shown in Figure~\ref{fig:2}.   This will be expressed in
+{\sc rmrs} via the binary relation  $\mbox{{\em ARG}}_{\{2,3\}}(a,x)$
+between the anchor $a$ of the predicate symbol for {\em give} and the
+individual $x$.   More generally, $\mbox{{\em ARG}}_n(a,x)$ means that
+$x$ is an argument to the predicate anchored at $a$, but its argument
+position is unspecified.
+
+\begin{figure}
+
+\leaf{$e$}
+\leaf{$\ldots$}
+\leaf{$x$}
+\leaf{$\ldots$}
+\branch{4}{\_give\_v\_b1}
+\tree
+\hfill
+\leaf{$e$}
+\leaf{$\ldots$}
+\leaf{$\ldots$}
+\leaf{$x$}
+\branch{4}{\_give\_v\_b1}
+\tree
+
+\caption{$x$ is a participant in a
+  giving event, but $x$'s role is unknown.}
+\label{fig:2}
+\end{figure}
+
+The binary ARG relations also supports underspecifying the {\em arity}
+of the base-language constructors, as is required when lexical
+subcategorisation information is missing from the shallow parser.
+While a deep parser assumes complete information about lexical
+subcategorisation in any of its derivations, thereby fixing the arity
+of the base-language constructors in the fully-specific logical form
+to be the same as the {\sc mrs} predicate symbol $P$ that's introduced by
+the lexeme, the predicate symbols in an {\sc rmrs} yielded from a
+shallow parse should, in principle at least,
+correspond to base language constructors of mixed arity.  For
+instance, the English verb {\em bake} can be used causatively ({\em
+  Kim baked the potatoes}), in which case semantically it corresponds
+to a 3-place predicate symbol {\em \_bake\_v\_b1} in the base
+language, or it may be used 
+inchoatively ({\em the potatoes are baking}), that semantically
+corresponds to a 2-place predicate {\em \_bake\_v\_b2}.  So the 
+predicate symbol {\em 
+  \_bake\_v} in the {\sc rmrs} that's introduced by a {\sc pos} tagger
+must be able to denote either of these
+base-language predicates in the {\sc rmrs}'s interpretation.
+
+\hidden{
+Finally, {\sc rmrs} allows one to underspecify the {\em sort} of an
+argument by including in its language meta-variables $u_1,u_2,\ldots$
+over individual, event, and handle variables.  For instance, the predicate
+symbol {\em \_be\_v} produced by a {\sc pos} tagger is resolvable to several
+3-place base-language constructors whose argument-types 
+vary: the base-language constructor corresponding
+to its use in {\em John is a man} takes an event and two individuals
+as arguments; but in {\em It is John who talked} the constructor takes
+an event, an individual and a proposition as arguments.
+}
 

### 2008-05-30 trunk/Papers/rmrs-08/rmrs.tex
very prelim version

@@
 
-Define syntax of RMRS; model theory of RMRS; solved forms.  Explain
-the role of the RMRS models (= trees) in the semantic construction
-pipeline, in particular with respect to decoding trees into semantic
-representations of the base language formalism.  Present examples in
-MRS syntax first, then in our RMRS language.  Finally, compare our
-version of RMRS against MRS and dominance constraints.
+% Define syntax of RMRS; model theory of RMRS; solved forms.  Explain
+% the role of the RMRS models (= trees) in the semantic construction
+% pipeline, in particular with respect to decoding trees into semantic
+% representations of the base language formalism.  Present examples in
+% MRS syntax first, then in our RMRS language.  Finally, compare our
+% version of RMRS against MRS and dominance constraints.
+
+\subsection{RMRS Syntax}
+
+To formalise the basic ideas from Section~\ref{sec:motivation}, 
+we start by defining the syntax of the {\sc rmrs} language.
+We adopt a
+syntax in the style of {\sc clls} (ref), in that labels, anchors and
+holes are all represented as {\em node variables}
+$\nvar=\{X,Y,X_1,\ldots\}$---the distinction among them comes
+from their position in the formulae of an {\sc rmrs} description (see
+Definition~\ref{defn:rmrs-syntax}). 
+The {\sc rmrs} vocabulary also includes
+base variables $\ovar$, consisting of
+individual variables $\{x,x_1,y,\ldots\}$ and event variables
+$\{e_1,\ldots\}$ (one can also include supersorts to
+these, but we forego doing so here).  Finally, the vocabulary includes
+0-ary predicate symbols $\pred=\{P,Q,P_1,\ldots\}$.
+
+\begin{definition}\label{defn:rmrs-syntax}
+An {\sc rmrs} is an underspecified description $\varphi$ consisting of a
+finite set of atoms 
+of the forms in (\ref{eg:atomic-forms}), such that $\varphi$
+satisfies the
+following three constraints:
+\begin{enumerate}
+\item   If $\ep{X}{Y}{P}, \ep{Z}{Y}{Q}\in \varphi$, then $X=Z$ and $P=Q$
+  (i.e., each predication in $\varphi$ has a unique anchor);
+\item   If $\ARG_S(X,v)\in \varphi$, then $X:Y:P\not\in \varphi$ and
+  $ARG_{S'}(Y,X)\not\in \varphi$ for any $Y\in \nvar$ and
+  $P\in \pred$ (i.e., the first argument to an ARG-relation must be an
+  anchor).
+\item   If $\ep{X}{Y}{P}\in \varphi$, then $\ep{Y}{Z}{Q},
+  \ep{Z}{X}{Q}\not\in \varphi$ for all 
+  $Z\in \nvar$ or $Q\in \pred$ (i.e., an anchor cannot be a label, or
+  {\em vice versa}).
+\end{enumerate}
+\begin{examples}
+\item   \label{eg:atomic-forms}
+$\begin{array}{rrll}
+\varphi &::= &\ep{X}{Y}{P} \\ %& \mbox{labelled elementary predication} \\
+&| &\ARG_S(X,v) \\ % & \mbox{ARG relation for base variables} \\
+&| &\ARG_S(X,Y) \\ % & \mbox{ARG relation for node variables} \\
+&| &X \qeq Y \\ % & \mbox{QEQ} \\
+&| &v_1 = v_2 \;|\; v_1 \neq v_2 \\ % & \mbox{base variable
+                                % (in)equality} \\
+&| &X = Y \;|\; X \neq Y \\ % & \mbox{node variable (in)equality} \\
+&| &P \spec Q % & \mbox{SPEC relation between predicate symbols}
+\end{array}
+$
+\end{examples}
+\end{definition}
+
+The subscript $S$ in the ARG relations stands for a subset of $\N$ which
+is either finite or $\N$ 
+itself.  The original {\sc rmrs} syntax $\ARG_i$ is 
+an abbreviation of $\ARG_{\{i\}}$ and $\ARG_n$ is an abbreviation of
+$\ARG_\N$.  Furthermore, $l:a:P(v)$ in the {\sc rmrs}s from
+Section~\ref{sec:motivation} and earlier papers (e.g.,
+\cite{copestake:2003}) 
+is a notational variant of $\{X:Y:P, ARG_{\{0\}}(v)\}$ in
+Definition~\ref{defn:rmrs-syntax}.
+
+\subsection{Model Theory}
+
+The model theory formalises the relationship between an {\sc rmrs} and
+the fully specific, alternative logical forms that it describes, as
+expressed in the base language.
+As suggested in Section~\ref{sec:motivation},  each model $\tau$
+corresponds to a unique formula of the base language, conceived as a
+tree (see Figure~\ref{fig:1}).  {\em Satisfaction} must then give rise
+to a notion of {\em 
+  truth} and {\em entailment} that respectively correspond to an {\sc
+  rmrs} $\varphi$ being a partial description of $\tau$, and one {\sc
+  rmrs} $\varphi$ being more specific than another $\varphi'$ (in the sense that
+$\varphi$ imposes more
+constraints than $\varphi'$ on the possible fully-specific logical forms).
+
+Since each model corresponds to a (unique) base language formula, defining
+models depends on the base-language's vocabulary $\Sigma$.  More
+formally, $\Sigma$ is a sorted signature of
+base-language constructors $f$ of sorts $s_1\times\ldots
+s_n\rightarrow s$,
+where $s,s_1,\ldots,s_n\in {\cal S}$, ${\cal S}$ being the sort
+hierarchy for $\Sigma$.  We call $n\geq 0$ the {\em arity} of $f$, and
+we write $\Sigma_i$ and $\Sigma_{\geq i}$ for the constructors in $\Sigma$
+with arity exactly $i$ and at least $i$ respectively.
+
+The models of {\sc rmrs} are then defined to be finite constructor
+trees:
+\begin{definition}\label{defn:models}
+A {\em finite constructor tree} $\tau$ is
+a function $\tau:D
+\rightarrow \Sigma$ such that $D$ is a \emph{tree domain} (i.e., a subset of
+$\N^*$ which is closed under prefix and left sibling) and the number
+of children of each node $u \in D$ is equal to the arity of $\tau(u)$.
+We define the \emph{sort} $s(u)$ of each node $u$ in $\tau$ bottom-up
+by saying that if $u$ has children $u_1,\ldots,u_n$ and $s(\tau(u)) =
+s_1 \times \ldots \times s_n \rightarrow s$, then $s(u) = s$ iff
+$s(u_i) \leq s_i$ for all $i$.  $\tau$ is \emph{well-sorted} if
+all nodes can be assigned sorts in this way.  In what follows, we only consider
+well-sorted finite constructor trees.
+\end{definition}
+
+The same tree $\tau$ can
+represent formulas from different-style base languages (e.g.,
+extensional and modal languages, {\sc drt}, etc) by 
+leaving the encoding flexible.  But for reasons 
+of space, we don't
+explore this here; for simplicity, we
+assume a sort hierarchy ${\cal S}$ for $\Sigma$ that is
+closely related to the sorts of {\sc rmrs} descriptions
+themselves.  So ${\cal S}$ contains the sorts
+$h$ (for subformulas), $x$ (for individual variables) and $e$ (for event
+variables).
+We also
+assume that $\Sigma$ contains conjunction constructors
+$\wedge_i: h\times\ldots h\rightarrow h$ of arity $i$ for every
+$i\in \N$.  We write $D(\tau)$ for the tree domain of a
+constructor tree $\tau$, and further define the following relations
+between nodes in a finite constructor tree:
+
+\begin{definition}\label{defn:dominance}
+$u \qeqdom v$ iff $v = u \cdot 2^k$ for some $k \geq 0$ and all
+  symbols $\tau(u), \tau(u \cdot 2), \ldots, \tau(u \cdot 2^{k-1})$
+  are quantifiers.
+
+$u \wedgedom v$ iff $u \leq v$ (i.e.\ it is a prefix), and all
+  symbols on the path from $u$ to $v$ (not including $v$) are one of
+  the $\wedge_i \in \Sigma$.
+\end{definition}
+
+The {\em satisfaction} relation between an {\sc rmrs} description $\varphi$
+and a finite constructor tree $\tau$ is defined in terms of several
+assignment functions.  The need for a 
+node variable
+assignment function
+$\alpha:
+\nvar
+\rightarrow D(\tau)$ is clear; the
+constraints on scope defined by the {\sc rmrs} must be resolvable to
+$\tau$'s specific
+specific scope information, and for any given $\tau$ there may be more
+than one way to do this.
+Secondly, to ensure that, for
+instance, the predications corresponding to {\em every} and {\em cat}
+in the {\sc rmrs}
+(\ref{eg:cat-pos}b)  partially describe fully-specific
+logical forms that accurately record the semantic dependency between
+these lexemes in (\ref{eg:cat}a) as well as partially describing
+fully-specific logical forms where they are not semantically
+dependent (e.g., as required when the same predications are produced
+from the {\sc pos} tagged sentence {\em every dog chased some cat}),
+the relationship of satisfaction 
+should ensure that the variables from $\ovar$ in an
+{\sc rmrs} description can
+map in a many-to-one way to variables in $D(\tau)$.  So
+satisfaction must also be defined in
+terms of 
+a (0-ary) variable assignment function
+$g:\ovar \rightarrow \Sigma_0$ from {\sc rmrs} variables to 
+base-language variables that respects the sorts, i.e.\
+$s(g(v)) \leq s(v)$ for all $v \in \ovar$.  Finally, satisfaction must
+be defined in terms of a relation $\sigma \subseteq
+\pred \times \Sigma_{\geq 1}$ that maps each
+{\sc rmrs} predicate symbol to a set of constructors from $\Sigma$.
+
+\begin{definition}\label{defn:satisfaction}
+Satisfaction of atoms is defined as follows:
+$$
+\begin{array}{l}
+  \tau, \alpha, g, \sigma \models  \ep{X}{Y}{P}
+\mbox{ iff}\\
+\qquad \tau(\alpha(Y)) \in \sigma(P) \mbox{ and } \alpha(X) \wedgedom
+  \alpha(Y) \\
+  \tau, \alpha, g, \sigma \models \ARG_S(X,a)
+\mbox{ iff exists  $i \in S$ s.t.}\\
+\qquad  \alpha(X) \cdot i \in D(\tau) \mbox{ and }
+  \tau(\alpha(X) \cdot 
+  i) = g(a) \\
+  \tau, \alpha, g, \sigma \models \ARG_S(X,Y)
+\mbox{ iff  exists $i \in S$ s.t.}\\
+\qquad \alpha(X) \cdot i \in D(\tau), \alpha(X) \cdot
+  i = \alpha(Y) \mbox{ and }\\
+\qquad
+s(\alpha(X) \cdot i) = h \\
+  \tau, \alpha, g, \sigma \models X \qeq Y
+\mbox{ iff } \alpha(X) \qeqdom \alpha(Y) \\
+  \tau, \alpha, g, \sigma \models X ={/}\neq Y
+\mbox{ iff } \\
+\qquad \alpha(X) ={/}\neq \alpha(Y) \\
+  \tau, \alpha, g, \sigma \models v_1 ={/}\neq v_2
+\mbox{ iff}\\
+\qquad g(v_1) ={/}\neq g(v_2) \\
+  \tau, \alpha, g, \sigma \models P \spec Q
+\mbox{iff $\sigma(P) \subseteq \sigma(Q)$}
+\end{array}
+$$
+A 4-tuple $\tau,\alpha,g,\sigma$ satisfies an underspecified
+representation $\varphi$ (written $\tau,\alpha,g,\sigma \models \varphi$)
+iff it satisfies all of its
+elements.
+\end{definition}
+Satisfaction encapsulates information about
+semantic scope ambiguities, missing
+information about 
+semantic dependencies and/or missing information about lexical
+subcategorisation and lexical senses.   For $j=\{1,2\}$, suppose that
+$\tau_j,\alpha_j,g_j,\sigma\models \varphi$ and let $\vocab(\varphi)$
+be the vocabulary that features in $\varphi$.  Then 
+$\varphi$ exhibits a semantic scope
+ambiguity if there is a $Y,Y'\in \vocab(\varphi)$ such that $\alpha_1(Y) \qeqdom
+\alpha_1(Y')$ and $\alpha_2(Y')\qeqdom \alpha_2(Y)$.   It exhibits
+missing information about semantic dependencies if $\exists v,v'\in
+\ovar$ in $\vocab(\varphi)$ such that $g_1(v)=g_1(v')$ and
+$g_2(v)\neq g_2(v')$.  It 
+exhibits missing lexical subcategorisation information if $\exists
+Y\in \vocab(\varphi)$ such that $\tau_1(\alpha_1(Y))$ is a constructor of a
+different type from $\tau_2(\alpha_2(Y))$.  And it exhibits missing
+lexical sense information if $\tau_1(\alpha_1(Y))$ and
+$\tau_2(\alpha_2(Y))$ are different base-language constructors, but of
+the same type.
+
+Truth, validity and entailment can now be defined in terms of
+satisfiability 
+in the usual way:
+\begin{definition}\label{defn:entailment}
+Let $\tau$ be a model, $\alpha$ a node variable assignment function,
+$g$ a 0-ary variable assignment function, $\sigma$ a mapping from
+$\pred$ to the powerset of $\Sigma$, and $\varphi,\varphi'$ {\sc rmrs}
+descriptions.  Then:
+\begin{description}
+\item   [truth:] $\tau\models \varphi$ iff $\exists \alpha,g,\sigma$  such
+  that $\tau,\alpha,g,\sigma\models \varphi$
+\item   [validity:] $\models \varphi$ iff $\forall \tau$, $\tau\models \varphi$.
+\item   [entailment:] $\varphi\models \varphi'$ iff $\forall \tau$, if
+  $\tau\models \varphi$ then $\tau\models \varphi'$.
+\end{description}
+\end{definition}
+% NOTE: This definition of entailment isn't right, because it allows
+% the assignment $\sigma$ from predicates to constructors to vary
+% when testing the satisfiability of $\varphi$ vs. $\varphi'$ with
+% respect to $\tau$.  I think that to get the notion of entailment we
+% want, $\sigma$ should actually be part of the {\sc rmrs} *model*.
+% In other words, an {\sc rmrs} model $M$ is a tuple $\langle \tau,
+% \sigma\rangle$, where $\tau$ is a finte constructor tree and
+% $\sigma$ is a mapping from $\Pred$ to the power set of $\Sigma$.
+% Then truth and entailment are defined correctly.  I believe our
+% proofs of things like Theorem~\ref{thm:big-one} actually assumes
+% that $\sigma$ is constant when interpreting $\varphi$ vs.\
+% $\varphi'$, no?
+
+\subsection{Solved Forms}
+
+Following the framework of dominance constraints, one can define a fragment
+of {\sc rmrs} descriptions, known as {\em solved forms}, which are
+guaranteed to be satisfiable.  For dominance constraints, solved forms
+are sets of constraints whose node variables form a forrest under the
+dominance relation $\dom$.  Solved forms for {\sc rmrs} must impose
+additional constraints, so as to ensure that all the (underspecified)
+ARG relations are satisfiable.  Indeed, we help to ensure this by
+substituting assuming that {\sc rmrs} solved forms have any variable
+equalities substituted into the ARG relations themselves.
+
+\begin{definition}\label{defn:solved-forms}
+  An RMRS $\varphi$ is \emph{in solved form} iff:
+  \begin{enumerate}
+  \item every variable in $\varphi$ is either a hole, a label or an
+    anchor (but not two of 
+    these);
+%% NOTE: The more constrained definition of an {\sc rmrs} description
+%% that we have now makes the set of labels and anchors mutually
+%% exclusive.  It does not make labels and holes mutually exclusive;
+%% nor anchors and holes mutually exclusive
+%% however.  
+  \item $\varphi$ doesn't contain equality, inequality, and SPEC
+    ($\sqsubseteq$) atoms;
+%   \item no downward-looking label occurs on the left-hand side of two
+%    different EPs;
+  \item if $\ARG_S(Y,i)$ is in $\varphi$, then $|S| = 1$;
+  \item for any label $Y$ and index set $S$, there
+    are no two atoms $\ARG_S(Y,i)$ and $\ARG_S(Y,i')$ in $\varphi$;
+  \item no label occurs on the right-hand side of two
+    different $\qeqdom$ atoms.
+  \end{enumerate}
+\end{definition}
+
+\begin{claim}
+  Every RMRS in solved form is satisfiable.
+\end{claim}
+\begin{proof}
+by constructing a model $\tau$ that
+contains for each predicate $P\in \varphi$ a member of $\sigma(P)$ with
+maximum arity, and that satisfies the ARGs and dominances in $\varphi$.
+\end{proof}
+
+One can now define the solved forms of an {\sc rmrs}
+description:
+\begin{definition}  \label{defn:solved-form-of}
+  The \emph{syntactic dominance relation} in an {\sc  rmrs} $\varphi$ is the
+  reflexive, transitive closure of the binary relation $$
+\begin{array}{r c l}
+\{(X,Y) & \mid &  \mbox{$\varphi$ contains $X \qeqdom Y$ or}\\
+&& \mbox{$ARG_S(X,Y)$ for some
+    $S$}\}
+\end{array}
+$$  
+We write $D(\varphi)$ for the syntactic dominance
+  relation of $\varphi$.
+
+  An RMRS $\varphi'$ is \emph{a solved form of} the RMRS $\varphi$
+  iff $\varphi'$ is in solved form and there is a substitution $s$
+  that maps the  node and base variables of $\varphi$ to the node and
+  base variables of $\varphi'$ such that
+  \begin{enumerate}
+  \item $\varphi'$ contains the EP $\ep{X'}{Y'}{P}$ iff there are
+    variables $X,Y$ such that $\ep{X}{Y}{P}$ is in $\varphi$, $X' =
+    s(X)$, and $Y' = s(Y)$;
+  \item for every atom $\ARG_S(X,i)$ in $\varphi$, there is exactly
+    one atom $\ARG_{S'}(X',i')$ in $\varphi$ with $X' = s(X)$, $i' =
+    s(i)$, and $S' \subseteq S$;
+  \item $D(\varphi') \supseteq s(D(\varphi))$.
+  \end{enumerate}
+\end{definition}
+
+\begin{claim}
+  For every model $\tau$ of some {\sc rmrs} $\varphi$, there is a solved form
+  $\varphi'$ of $\varphi$ such that $\tau$ is also a model of $\varphi'$.
+\end{claim}
+\begin{proof}
+  will be by constructing a solved form that respects the equalities,
+  ARGs, and dominances in $\tau$.
+\end{proof}
+
+\begin{prop}
+  Every {\sc rmrs} $\varphi$ has only a finite number of solved forms, up to
+  renaming of variables.
+\end{prop}
+\begin{proof}
+  Up to renaming of variables, there is only a finite number of
+  substitutions on the node and base variables of $\varphi$.  Let $s$
+  be such a substitution.  This fixes the set of EPs of any solved
+  form of $\varphi$ that is based on $s$ uniquely.  There is only a
+  finite set of choices for the subsets $S'$ in condition 2 of
+  Def.~\ref{defn:solved-form-of}, and there is only a finite set of
+  choices of new $\qeqdom$ atoms that satisfy condition 3.  Therefore,
+  the set of solved forms of $\varphi$ is finite.
+\end{proof}
 

### 2008-06-05 trunk/Papers/experiences/conclusion.tex
populating icaps paper

--- 2008-03-21 trunk/Papers/dl-gre-08/conclusion.tex
@@
 \section{Conclusion} \label{sec:conclusion}
-
-In this paper, we have explored the idea of viewing the generation of
-singular REs as the problem of computing a DL formula with a given
-extension.  We have shown how such formulas can be computed
-efficiently (for \alc\ and \el) by adapting existing algorithms from
-the literature.  The \el\ algorithm is able to generate 95\% of the
-non-relational and 67\% of the relational REs from
-\newcite{viethen06:_algor_for_gener_refer_expres}.  Both algorithms
-are extremely efficient (350 ms and 140 ms respectively to generate
-relational REs for all individuals in a random model with 100
-individuals); to our knowledge, these are by far the fastest runtimes
-for relational GRE reported in the literature.  We will make our
-implementation available online.
-
-Because they compute referring expressions for all individuals in the
-domain at once, our algorithms will perform especially strongly in
-static settings, such as the generation of descriptions for museum
-exhibits, in which the individuals and their properties don't change
-much.  However, even in more dynamic settings, our algorithms have a
-chance to outperform search algorithms like Dale and Haddock's in the
-average case because they can't get stuck in unproductive branches of
-the search space. Nevertheless, one interesting question for future
-research is how to incrementally update simulation classes when the
-model changes. Similarly, it would be interesting to explore how
-linguistic constraints and different attribute orderings can be taken
-into account without slowing the algorithms down, and how the
-algorithms can be adapted to compute REs for sets. In exploring these
-extensions we will be able to draw on a rich body of literature that
-considers many variants of simulation algorithms.
-
-%Nevertheless, one interesting question for future
-%research is how to adapt simulation algorithms that can
-%incrementally update the simulation classes when the model changes.
-%Similarly, it would be interesting to explore how linguistic
-%constraints and different attribute orderings can be taken into
-%account without slowing the algorithms down, and how the algorithms
-%can be adapted to compute REs for sets.
-
-In experimenting with the Viethen and Dale data, we found that
-there is no single ordering that covers all human-produced
-descriptions, which seems to be in contrast to Dale and Reiter's
-\shortcite{Dale1995} assumption that there is only one ordering for
-each given domain.  In fact, it is not even the case that each speaker
-consistently uses just one ordering.  An interesting open research
-question is thus what factors determine which ordering is used.
-Unfortunately, both in the Viethen and Dale dataset and in the TUNA
-corpus \cite{deemter06:_build_seman_trans_corpus_for}, only a minority
-of referring expressions is relational, maybe because these domains
-lend themselves very well to row/column style propositional REs.  We
-are currently preparing an experiment to collect REs in a domain in
-which relational REs will be more frequent.
 
@@
 %%% mode: latex
-%%% TeX-master: "dl-gre-08"
+%%% TeX-master: "experiences"
 %%% End:

### 2008-06-05 trunk/Papers/experiences/discussion.tex
populating icaps paper

--- 2008-05-16 trunk/Papers/dl-gre-08/discussion.tex
@@
 
-We will now describe two experiments evaluating the quality of the \el\
-algorithm's output and the efficiency of both of our algorithms, and
-we discuss the interface between our algorithms and realization.
-
-%different aspects of
-%the performance of our algorithm. First, we will look at its running
-%time on randomly generated domain graphs of different sizes. Second,
-%we will compare the output of our algorithm to human-generated
-%descriptions.
-
-\subsection{Evaluation: Output quality}
-
-To compare the descriptions generated by our algorithm to those humans
-produce, we use a corpus of human-generated referring expressions
-collected and made available by Jette Viethen and Robert
-Dale.\footnote{http://www.ics.mq.edu.au/\~{}jviethen/drawers} They
-asked human subjects to describe one of 16 filing cabinet drawers. The
-drawers had different colors and were arranged in a four-by-four grid
-(see Fig.~\ref{fig:drawers}). The human subjects used four
-non-relational properties (the drawer's $\mathsf{color}$, its
-$\mathsf{column}$ and $\mathsf{row}$ number, and whether it is in a
-$\mathsf{corner}$) and five relational properties ($\mathsf{above,
-  below, next\ to, left\ of, right\ of}$). Of the 118 referring
-expressions obtained in the experiment, only 15 use relations.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.2\textwidth]{drawers.png}
-\end{center}\vspace*{-3ex}
-\caption{A schematic view of the filing cabinets.}\label{fig:drawers}\vspace*{-1.5ex}
-\end{figure}
-
-\newcite{viethen06:_algor_for_gener_refer_expres} describe the data in
-more detail and present results of evaluating the Full Brevity
-algorithm, the Incremental Algorithm (both by \newcite{Dale1995}), and
-the Relational Algorithm \cite{dale91:_gener_refer_expres_invol_relat}
-on this corpus. The Incremental Algorithm is dependent on a predefined
-ordering in which properties are added to the description. Viethen and
-Dale, therefore, try all possible orderings and evaluate what
-percentage of descriptions an algorithm can generate with any of
-them. The Full Brevity and the Relational Algorithms choose properties
-based on their discriminatory power and only use the orderings as tie
-breakers. Viethen and Dale found that the Incremental Algorithm is
-capable of generating 98 of the 103 non-relational
-descriptions. However, the Relational Algorithm was unable to
-generate even a single one of the human-generated relational
-descriptions.
-
-%(using four different orderings to cover all the data).
-
-\input example-outputs
-
-We replicated Viethen and Dale's experiment for the \el\ algorithm
-presented above. In the non-relational case, our results are the same
-as theirs for the Incremental Algorithm: the \el\ algorithm generates
-98 of the 103 non-relational descriptions, using four (of the
-possible) orderings.  This is because the two algorithms perform
-essentially the same computations if there are no relations.
-
-%To evaluate the relational case, we first removed the
-%$\mathsf{column}$ and $\mathsf{row}$ predicates from the domain,
-%because these predicates make it possible to uniquely identify each
-%drawer with non-relational descriptions, and our algorithm would
-%always prefer these.  Using three (of the 240 possible) different
-%orderings, our algorithm was able to generate 10 of the 15
-%human-produced relational descriptions correctly.  This is shown in
-%more detail in Fig.~\ref{fig:example_outputs}; notice that four of the
-%15 descriptions occurred twice in the corpus.
-
-When we add relations, our algorithm is able to generate 10 of the 15
-human-produced relational descriptions correctly (in addition to the
-98 non-relational descriptions).  Fig.~\ref{fig:example_outputs} gives
-example outputs of the \el\ algorithm for three different orderings,
-which together achieve this coverage.  Of the five human-produced
-descriptions that the \el\ algorithm cannot generate, three involve
-references to sets (\textit{the two blues ones in horizontal
-sequence}/\textit{the two yellow drawers}), and two contain so much
-redundant information that our algorithm, which similar to the
-Incremental Algorithm allows for some redundancy but stops once it has
-found a distinguishing description, cannot reproduce them. It does,
-however, generate other, simpler descriptions for all of the
-human-produced descriptions that cannot be reproduced exactly.
-
-%two mention column information, which we excluded from our domain, and
-
-\subsection{Evaluation: Efficiency}
-
-Both the \el\ and the \alc\ algorithms took about 15 milliseconds to
-compute distinguishing formulas for all 16 individuals in the Viethen
-and Dale dataset.\footnote{Runtimes were measured on a MacBook Pro
-  (Intel Core 2 Duo, 2.16 GHz) running Java 1.6 beta.  We allowed the
-  Java VM to warm up, i.e., just-in-time compile all bytecode, before
-  taking the measurements.}
-
-In order to get a more comprehensive picture of the algorithms'
-efficiency, we ran them on random models with increasing numbers of
-individuals.  Each model had random interpretations for ten different
-propositional and four relational symbols; each individual had a 10\%
-chance to be in the extension of each propositional symbol, and each
-pair of individuals had a 10\% chance to be related by a relational
-symbol.  The results (averaged over 10 runs for each model size) are
-shown in Fig.~\ref{fig:runtimes}.  The \el\
-algorithm takes about 350 ms on average to generate relational REs for
-all individuals in the model of size 100, i.e., less than 4 ms on
-average for each individual.  The \alc\ algorithm is even faster, at
-about 140 ms for the model of size 100.  As far as we know, these are
-by far the fastest published runtimes for any relational GRE algorithm
-in the literature.
-
-% Another encouraging aspect of both runtime graphs is that they seem to
-% be approximated well by quadratic functions.  So although the
-% worst-case runtime of the \el\ algorithm might be exponential, both
-% algorithms empirically exhibit (low) polynomial runtime on the inputs
-% we considered.
-
-\subsection{Interface to realization}
-\label{sec:discussion-realization}
-
-%\begin{itemize}
-%\item there's a risk to generate a formula that can't be realized,
-%  especially for the \alc\ algorithm
-%\item everybody except perhaps SPUD has this problem
-%\item problem is worse in our case because it is harder to control the
-%  order in which atoms and relations are explored
-%\item would be interesting for future work to think about how
-%  search in the bisim algorithms can be controlled by the realizer
-%\end{itemize}
-
-Our GRE algorithms do not guarantee that the formula they compute can
-actually be realized in language.  For example, none of the formulas
-our algorithms computed in the Viethen and Dale domain contained an
-atom that would commonly be realized as a noun; the property
-$\mathsf{drawer}$ is never used because it applies to all individuals
-in the domain.  This particular problem could easily be worked around
-in a post-processing step.  However, another problem arises from the
-massive use of negation in the \alc\ algorithm; it will be hard for
-any realizer to find a reasonable way of expressing a formula like
-$\neg \exists R.(\neg P \sqcap \neg Q)$ as a smooth noun phrase.
-Although we agree with \newcite{deemter02:_gener_refer_expres} and
-others that the careful use of negation and disjunction can improve
-REs, these connectives must not be overused.  Thus we consider the
-formulas computed by the \el\ algorithm ``safer'' with respect to
-realization.
-
-Of course, we share the problem of interfacing GRE and realization
-with every other approach that separates these two modules, i.e.,
-almost the entire GRE literature (notable exceptions are, e.g.,~\newcite{Horacek1997} and SPUD \cite{Stone1998a}).  In principle, we
-believe that it is a good idea to handle sentence planning and
-realization in a single module; for instance, SPUD can use its
-awareness of the syntactic context to generate succinct REs as in
-``take the rabbit from the hat''.  We hope that the ideas we have
-explored here for efficient and expressive RE generation can
-eventually be combined with recent efficient algorithms for integrated
-sentence planning and realization, such as in \newcite{KolSto07}.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=\columnwidth]{runtimes}\vspace*{-2ex}
-  \caption{Average runtimes (in ms) of the two algorithms on random
-    models with different numbers of individuals.}
-  \label{fig:runtimes}\vspace*{-1.5ex}
-\end{figure}
-
-One problem that arises in our approach is that both
-algorithms derive some measure of efficiency from their freedom to
-build formulas without having to respect any linguistic constraints.
-It seems straightforward, for instance, to extend Krahmer et al.'s
-\shortcite{Krahmer2003} approach such that it only considers subgraphs
-that can actually be realized, because their algorithm proceeds by a
-genuine search for uniquely identifying subgraphs, and will simply
-take a different branch of the search if some subgraph is useless.
-This would be harder in our case.  Our algorithms don't search in the
-same way; if we disallow certain refinements of a partition, we have
-to allow the algorithms to backtrack and thus jeopardize the
-worst-case polynomial runtime.  Investigating this interplay between
-efficiency and linguistic constraints is an interesting avenue for
-future research.
+planning competition makes for great starting point for users
 

### 2008-06-05 trunk/Papers/experiences/experiences.tex
populating icaps paper

\documentclass[letterpaper]{article}

\usepackage{aaai} 
\usepackage{times} 
\usepackage{helvet} 
\usepackage{courier} 
\title{Experiences with planning for natural language generation}
\author{Alexander Koller \and Ron Petrick \\
School of Informatics \\
University of Edinburgh}

\begin{document}

\maketitle

\begin{abstract}
  lkajsd
\end{abstract}

\input introduction
\input domains
\input experiments
\input discussion
\input conclusion

\bibliography{bibliography}
\bibliographystyle{aaai}

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2008-06-05 trunk/Papers/experiences/introduction.tex
populating icaps paper

\section{Introduction} \label{sec:introduction}

Natural language generation (NLG; \citeauthor{reiter00building} 2000)
is one of the major subfields of natural language processing.  It is
concerned with computing natural language sentences or texts that
convey a given piece of information to the user.  Problems in NLG are
often more complex than their counterparts in parsing \cite{KolStr02},
and indeed efficiency has not traditionally been the main focus of
NLG.  On the other hand, NLG has some obvious parallels with planning:
One tries to change the mental state of the hearer by applying actions
which correspond to uttering words or sentences.  This parallel has a
long tradition in NLG \cite{appelt:planning,young94dpocl}, and there
has recently been a resurgence of interest in applying modern planning
techniques to NLG \cite{Steedman-Petrick:07,KolSto07,benotti08b}.
These new approaches combine an interest in planning as a modelling
tool for natural language with a hope for improved efficiency by
exploiting modern planning algorithms to help with the search.

In this paper, we present two recent planning domains that arise in
the context of applying planning to natural language generation.  We
will first review the sentence generation domain, which was cast as a
planning problem by Koller and Stone \citeyear{KolSto07}.  In this
domain, the goal is to generate a single sentence that expresses a
given piece of meaning; the plan encodes this sentence, and the
actions correspond to uttering individual words.  We will then
introduce the GIVE domain (``Generating Instructions in Virtual
Environments'').  GIVE is a new shared task that was recently posed as
a challenge for the NLG community
\cite{alexander07:_shared_task_propos}.  It uses planning as one
module of a NLG system which generates natural-language instructions
that will guide the user towards performing a given task in a virtual
environment.

We will then discuss some experiences with using off-the-shelf
planners in these domains.  In particular, we explore the efficiency
of FF \cite{HoffmannNebel01} and SGPLAN
\cite{hsu06:_new_featur_in_sgplan_for} on instances of our planning
domains of varying sizes.  Our findings are mixed.  On the one hand,
it turns out that modern planners handle the search problems that
arise in NLG quite easily.  On the other hand, they spend tremendous
amounts of time on preprocessing; for instance, FF spends 90\% of its
runtime in the sentence generation domain on grounding out the
literals and actions, and ends up slower than an ad-hoc Java
implementation of GraphPlan that avoids grounding.  We argue that for
domains like ours, which are dominated by the number of actions and the
universe size rather than the combinatorics of the search problem, the
overall runtime of a planner could be improved by grounding more
selectively.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "experiences"
%%% End:

### 2008-06-05 trunk/Papers/experiences/domains.tex
populating icaps paper
a bit about GIVE

--- 2008-06-05 trunk/Papers/experiences/discussion.tex
@@
-\section{Discussion} \label{sec:discussion}
+\section{Planning in NLG} \label{sec:domains}
 
-planning competition makes for great starting point for users
+In this section, we present the two domains: first the sentence
+generation domain, then the instruction giving domain.
+
+\subsection{Sentence generation as planning}
+
+\subsection{Planning in instruction giving}
+
+The object of the GIVE Challenge (``Generating Instructions in Virtual
+Environments''; Koller et al.\
+\citeyear{alexander07:_shared_task_propos}) is to build an NLG system
+which is able to give natural-language instructions which will guide a
+human user in performing some task in a virtual environment.  A map of
+an example world is shown in Fig.~\ref{fig:give-development-world};
+instructions for the first steps could be ``turn left; press the
+button; turn right and walk through the door;'' and so forth.  From an
+NLG perspective, GIVE makes for an interesting challenge because it is
+a theory-neutral task that exercises all components of an NLG system,
+and emphasizes the study of communication in a (simulated) physical
+environment.  It also has the advantage that the user and the NLG
+system can be physically in different places, as long as the 3D client
+and the NLG system are connected over a network.  This makes it
+possible to evaluate GIVE NLG systems on a large scale over the
+Internet.  The first GIVE evaluation will take place in late 2008;
+currently seven research teams from five countries are working on
+developing systems to participate in the challenge.
+
+\begin{figure}
+\centering
+\ldots 
+\caption{Map of the GIVE development world.}
+  \label{fig:give-development-world}
+\end{figure}
+
+\todo{explain world? -- yes, definitely}
+
+One crucial component of a GIVE system is a planner which will compute
+a domain plan that consists of the actions the user has to execute to
+achieve the goal.  For these purposes, the virtual world is
+discretized into a planning problem in which each object has a
+symbolic position and orientation.  The first few action instances in
+the example domain could be $turn-left(north,west)$, $move(pos-3-2,
+pos-2-2, west)$, $manipulate-b_1-off-on(pos-2-2)$,
+$turn-right(west,north)$, \todo{complete this} and so on.  Under the
+current implementation of the planning problem, the shortest plan for
+getting the player from the initial state to the goal (in which they
+are holding the trophy) has \todo{107} steps, most of which are
+``move'' actions.  The bulk of the GIVE problem is therefore a variant
+of the Gridworld problem, which also involves finding a route through
+a world with discrete positions -- but with the need to press buttons
+in the right order and with somewhat more complicated room shapes.
+
+It is interesting to note that although the task of a GIVE NLG system
+could be seen as mapping a domain plan to a sequence of NL
+instructions, this mapping is not trivial.  It would be
+\emph{possible} to express the last part of the above example plan as
+``turn right; walk forward; walk forward; turn right; walk forward;
+walk forward; turn left; walk forward'', but this is a very clumsy way
+of phrasing the instruction ``turn around and walk through the door''
+that will frustrate users and lead to bad evaluation results.  In
+other words, it is important to aggregate multiple movement actions
+into a single NL instruction.  On the other hand, it may be necessary
+to generate instructions that don't correspond to domain actions.  For
+instance, it will be necessary to find a correct referring expression
+for $b_1$ in order to express the action
+$manipulate-b_1-off-on(pos-2-2)$.  The interaction between referring
+and instructing can go much deeper than that.  Imagine, for example,
+that the user just entered the top right room, and we want to instruct
+them to press $b_4$.  At that point in time, the user can see six
+different buttons, and generating a single instruction of the form
+``push X'' is difficult.  However, an instruction sequence like ``walk
+forward; turn left; now press the button right in front of you'' will
+be successful and natural.  Here the first two instructions do double
+duty by preparing the simple referring expression ``the button right
+in front of you'', which is only valid after the user's position and
+orientation changed.  A generation system that is supposed to generate
+such instruction sequences must be tightly integrated with the planner
+in order to exploit the simultaneous effect of the first two
+instructions on the world and on the communicative situation.
+
+\todo{equivalent plans; plan recognition}
 

### 2008-06-05 trunk/Papers/experiences/domains.tex
wrote something about the experiments

@@
 \centering
-\ldots 
+\includegraphics[width=1 \columnwidth]{give_world_no_expl}
 \caption{Map of the GIVE development world.}
@@
 getting the player from the initial state to the goal (in which they
-are holding the trophy) has \todo{107} steps, most of which are
-``move'' actions.  The bulk of the GIVE problem is therefore a variant
-of the Gridworld problem, which also involves finding a route through
-a world with discrete positions -- but with the need to press buttons
-in the right order and with somewhat more complicated room shapes.
+are holding the trophy) has 104 steps, most of which are ``move''
+actions.  The bulk of the GIVE problem is therefore a variant of the
+Gridworld problem, which also involves finding a route through a world
+with discrete positions -- but with the need to press buttons in the
+right order and with somewhat more complicated room shapes.
 
@@
 
-\todo{equivalent plans; plan recognition}
+\todo{equivalent plans; plan recognition; real-time requirements}
 

### 2008-06-05 trunk/Papers/experiences/experiments.tex
populating icaps paper
wrote something about the experiments

--- 2008-06-05 trunk/Papers/experiences/discussion.tex
@@
-\section{Discussion} \label{sec:discussion}
+\section{Experiments} \label{sec:experiments}
 
-planning competition makes for great starting point for users
+As a first impression, it is a good idea to use planning in both of
+these domains.  In the GIVE domain, SGPLAN 5.2.2 computes a domain
+plan from the initial state to the goal in 0.3 seconds, which is fast
+enough for the application.\footnote{All runtimes reported here were
+  measured on a Pentium 4 CPU at 3 GHz.  Java programs were allowed to
+  ``warm up'', i.e.\ the planner was run five times in a row and the
+  first four measurements discarded to ensure that the JVM had
+  just-in-time compiled all relevant bytecode.}  In the sentence
+generation domain, FF dramatically outperforms the best previously
+known algorithm for the same problem (a reimplementation of
+\cite{Stone2003a}), although the latter is a greedy search algorithm
+with a heuristic that is hand-tailored to the domain.
+
+However, we also observed that although FF manages the search for a
+plan very efficiently, it spends comparatively large amounts of time
+on computing instances, most of which are then never used during the
+search.  Below, we report on a series of experiments that are designed
+to improve our understanding of this situation.
+
+\subsection{Experiment 1: Sentence generation}
+\label{sec:exper-1:-sent}
+
+For the first experiment, we generated a series of sentence generation
+problems which required the planner to compute a plan representing the
+sentence ``Mary likes the Adj$_1$ \ldots Adj$_n$ rabbit.''  Each
+problem instance assumed a certain number $m$ of rabbits that were
+distinguished by $n \leq m$ different properties, such that all $n$
+properties were required to distinguish the target referent from all
+other rabbits.  The $n$ properties are realized as $n$ different
+adjectives, in any order.  This setup allows us to control the plan
+length (a plan using $n$ properties will have length $n+4$) and the
+universe size (the universe will contain $m+1$ individuals in addition
+to the differently-typed individuals for encoding the grammar).
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.4\textwidth]{pic-runtime-modifiers-with-sgplan}
+  \caption{Runtimes in the sentence generation domain.  The horizontal
+  axis represents parameters $(m,n)$ from $(1,1)$ to $(10,10)$ in
+  lexicographical order.}
+  \label{fig:runtimes-crisp}
+\end{figure}
+
+The results of this experiment are shown in
+Fig.~\ref{fig:runtimes-crisp}; we plot the input parameters $(m,n)$
+(in lexicographic order) on the horizontal axis and the runtime in
+milliseconds on the vertical axis.  The figure reveals a number of
+interesting insights.  First of all, FF significantly outperforms
+SGPLAN in this domain.\footnote{Experiments with SGPLAN were run with
+  a pre-release version of SGPLAN that was kindly provided by Chih-Wei
+  Hsu, as the release version of SGPLAN 5.2.2 has a bug which causes
+  it to crash on some instances.}  Second, FF's runtime is dominated
+by its initial grounding step, in which it computes the ground
+instances of all predicates and actions used in the planning problem
+to avoid unification during search: The ratio of grounding time to
+total runtime is generally above 85\%, and rises above 99\% at $m=11$,
+which is still a small universe in this application.\footnote{The
+  ``grounding'' time reported here is what FF reports as ``time spent:
+  instantiating action templates''.}  This is also reflected in the
+much higher sensitivity of FF to the domain size: If we fix $n=1$, FF
+takes 60 ms to compute a plan at $m=1$, but 2.4 seconds at $m=10$; by
+contrast, our GraphPlan implementation takes 30 ms at $m=1$ and still
+only 50 ms at $m=10$.
+
+As a consequence, even FF is outperformed quite clearly by an ad-hoc
+Java implementation of GraphPlan which only computes instances of
+predicates and actions as they are discovered while the planning graph
+gets built.  On the other hand, the runtime that FF actually spends on
+search is again considerably lower than that of our GraphPlan
+implementation, and for any given $m$, GraphPlan's runtime grows much
+faster with $n$ (i.e.\ the plan size) than FF's.  We conclude that on
+the sentence planning domain, FF does very well on search, but then
+spends about ten times as much runtime again on grounding.
+
+\subsection{Experiment 2: Minimal GIVE worlds}
+\label{sec:exper-2:-minim}
+
+To evaluate the performance of the planners on the planning problems
+that arise in the GIVE domain, we construct a sequence of test worlds
+as illustrated in Fig.~\ref{fig:give-minimal}.  These worlds consist
+of a $2n$ by $h$ grid of positions, such that there are buttons at
+positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$.  The player
+starts in position $(1,1)$ and must press all buttons in order to
+complete the game.  The world is generated as a GIVE world
+description, and then converted into a planning problem by the GIVE
+software. \todo{at what level of detail should we describe this?}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.4 \textwidth]{pic-buttons}
+  \caption{Minimal GIVE world.}
+  \label{fig:give-minimal}
+\end{figure}
+
+Some results from this experiment (for $h=20$, with $n$ ranging from
+$1$ to $40$) are shown in Fig.~\ref{fig:give-runtime-minimal}.  The
+most obvious result is that FF is unable to solve any problems beyond
+$n=13$ on the experimentation machine within the memory limit of 1 GB.
+SGPLAN, on the other hand, solves instances beyond $n=40$ without
+major problems.  The time spent on grounding is not a major factor in
+either planner, probably because the planners need more time to
+actually compute the plan -- at $n=40$, the optimal plan has a length
+of about 1600!
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.4\textwidth]{pic-runtime-buttons}
+  \caption{Runtimes of FF and SGPLAN on the minimal GIVE worlds for $h=20$. The horizontal axis is $n$, ranging from 1 to 40.}
+  \label{fig:give-runtime-minimal}
+\end{figure}
+
+\subsection{Experiment 3: GIVE worlds with junk positions}
+\label{sec:experiment-3:-give}
+
+Finally, we varied the GIVE world in order to judge the effect of the
+universe size in the GIVE domain (see Fig.~\ref{fig:give-junk}).  We
+added another $w$ by $h$ positions to the right of a minimal world as
+in Experiment 2.  Other than that, we left the initial state and goal
+untouched, i.e.\ the new positions were not needed in any plan; this
+approximates the situation in the actual GIVE domain, where most grid
+positions never need to be used.  As before, we generate a GIVE world
+description and then convert it into a planning problem.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.4\textwidth]{pic-empty-buttons}
+  \caption{GIVE world with junk positions.}
+  \label{fig:give-junk}
+\end{figure}
+
+Some results for $h=20$ and $n=5$ (with $w$ ranging from $1$ to $70$)
+are shown in Fig.~\ref{fig:give-runtime-junk}.  Again, FF runs out of
+memory, this time at $w=17$, whereas SGPLAN happily solved inputs
+beyond $w=70$.  However, unlike in Experiment 2, both planners now
+spend a substantial propertion of their time on grounding again.  In
+SGPLAN, this translates to a ``parsing time'' (which
+\todo{presumably} includes grounding) which grows from 180 ms to 21.7
+seconds as $w$ grows from $1$ to $75$; the rest of the runtime (which
+includes the search) only grows from 400 ms to 2.3 seconds.  This
+difference is particularly dramatic given that the actual optimal plan
+is an identical plan of about 100 steps for each $w$.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.4\textwidth]{pic-runtime-empty-world}
+  \caption{Runtimes of FF and SGPLAN on the GIVE worlds with junk
+    positions for $h=20$ and $n=5$. The horizontal axis is $w$,
+    ranging from 1 to 70.}
+  \label{fig:give-runtime-junk}
+\end{figure}
 

### 2008-06-05 trunk/Papers/experiences/conclusion.tex
now we have something in every section

@@
 \section{Conclusion} \label{sec:conclusion}
+
+In this paper, we have introduced two novel planning domains related
+to problems in natural language generation: the sentence generation
+problem and the GIVE navigation problem.  We have then reported on a
+number of experiments in which we applied off-the-shelf planners to
+these domains.  Our main results were that while modern planners do
+pretty well in controlling the complexity of search, they suffer from
+practical problems that limit their performance in unexpected places.
+In particular, in both domains the grounding step that both planners
+perform tends to take much more time than the search itself.
+
+\todo{say something clever about the future}
 

### 2008-06-05 trunk/Papers/experiences/discussion.tex
wrote something about the experiments
now we have something in every section

@@
 
-planning competition makes for great starting point for users
+The positive conclusion we can draw from the experiments reported
+above is that for the most part, modern planners are pretty good at
+controlling the search for the NLG-related inputs that we looked at.
+This is particularly true for FF in the sentence generation domain,
+which generates nontrivial 14-word sentences in about two seconds.
+There is still room for improvement (a special-purpose algorithm for
+referring expressions could generate ten of these 14 words in a few
+milliseconds), and the search becomes slower -- for obvious reasons --
+in the almost completely unconstrained world of Experiment 2, but
+search is not the main efficiency bottleneck.
+
+The most restrictive bottleneck in the two domains we have looked at
+in this paper is the initial grounding step that almost all modern
+planners perform.  Our conclusion from the experiments reported above
+is that this is a good strategy in domains where the typical plan
+length dominates the typical universe size, i.e.\ where the initial
+investment into grounding pays off in saved unifications during the
+search.  However, as we have shown, there are natural planning domains
+in which relatively short plans must be computed in large universes;
+it is not unrealistic for the universe in the sentence generation
+domain to consist of thousands of domain individuals and tens of
+thousands of actions, some of which take three domain individuals as
+parameters.  The common strategy of splitting the universe over
+several types of individuals will not be very effective here.
+
+We believe that if planning is to become a mature technology which
+real-world users will believe in, it must find a way to deal with the
+grounding problem more gracefully than current systems do.  The
+research focus in developing planning algorithms has quite rightfully
+been on dealing with the search, and for this it is convenient --
+depending on the particular algorithm, necessary -- to ground out all
+predicates and actions before starting the search itself.  However, we
+think that the domains presented here are not that unusual in their
+ratio of plan length against universe size.  At the end of the day,
+real-world users will care about the \emph{total} runtime of a
+planner, and this is more than just the search.
+
+By and large, it is a relatively pleasant experience to come as a
+``customer'' to the planning community: Thanks to the competitions, it
+is easy to identify and download a fast implementation, at least for
+Linux.  However, in the course of our experiments we found (and
+reported) bugs in both SGPLAN and FF.  And as our experiments have
+shown, there is no one silver bullet: Even planners as closely related
+as FF and SGPLAN can differ significantly in their performance on
+different domains.
 

### 2008-06-09 trunk/Papers/experiences/domains.tex
added some text for the crisp domain

@@
 \subsection{Sentence generation as planning}
+
+Sentence generation is the problem of computing, from a grammar and a
+semantic representation, a single sentence that expresses this piece
+of meaning.  This problem is traditionally split into several steps
+\cite{reiter00building}.  In a first step, called \emph{sentence
+  planning}, the semantic representation is first enriched with more
+information; for instance, \emph{referring expressions}, which refer
+to individuals that we want to talk about, are determined at this
+point.  In a second step, called \emph{surface realization}, this
+enriched representation is then translated into a natural-language
+sentence, using the grammar.
+
+However, determining referring expressions (REs) and realization
+interact, so it would be beneficial to perform both steps together.
+This was the goal of the SPUD system \cite{Stone2003a}, which
+performed both steps together using a top-down generation algorithm
+based on tree-adjoining grammars \cite{joshi;etal1997} whose lexical
+entries were equipped with semantic and pragmatic information.
+Unfortunately, SPUD suffered from having to explore a huge search
+space, and had to resort to a non-optimal greedy search strategy to
+retain reasonable efficiency.
+
+\cite{KolSto07} attempt to improve the efficiency by translating the
+SPUD sentence generation problem into a planning problem and using a
+planning algorithm for generation.  We will now illustrate how they do
+this by a (simplified) example.
+
+\begin{figure}
+  \centering
+  xxxx
+  \caption{The example grammar.}
+  \label{fig:white-rabbit-sleeps-grammar}
+\end{figure}
+
+\begin{figure}
+  \centering
+  xxxx
+  \caption{Derivation of ``The white rabbit sleeps.''}
+  \label{fig:white-rabbit-sleeps-deriv}
+\end{figure}
+
+Consider a knowledge base containing the individuals $r_1$ and $r_2$;
+$r_1$ and $r_2$ are rabbits, $r_1$ is white and $r_2$ is brown, and
+$r_1$ sleeps.  Let's say we want to express the information
+$\{\mathsf{sleep}(r_1)\}$ using the tree-adjoining grammar shown in
+Fig.~\ref{fig:white-rabbit-sleeps-grammar}.  This grammar consists of
+\emph{elementary trees}, each of which contributes certain
+\emph{semantic content}.  We can instantiate these trees by
+substituting individuals for \emph{semantic roles}, such as
+$\mathsf{self}$ and $\mathsf{subj}$, and then combine the tree
+instances as shown in Fig.~\ref{fig:white-rabbit-sleeps-deriv} to
+obtain the sentence ``The white rabbit sleeps''.  The SPUD algorithm
+computes this grammatical derivation by starting with the elementary
+tree for ``sleeps''; this satisfies the need to convey the semantic
+information, but introduces a need to generate a noun phrase (NP) for
+the subject, which also refers uniquely to the target referent $r_1$.
+In a second step, it substitutes the tree for ``the rabbit'' into the
+open NP leaf, which makes the derivation grammatically complete; but
+as there are two different individuals that could be described as
+``the rabbit'' -- technically, $r_2$ is still a \emph{distractor} for
+the referring expression --, we are still not finished.  This is why
+we add the tree for ``white'' to the derivation by the
+\emph{adjunction} operation.  This makes the derivation syntactically
+and semantically complete, and we are done.
+
+\begin{figure}
+  \centering
+  xxxx
+  \caption{Generating ``The white rabbit sleeps'' as a planning problem.}
+  \label{fig:white-rabbit-as-planning}
+\end{figure}
+
+This process has clear parallels to planning: We manipulate a state by
+applying actions in order to achieve a goal.  We can make this
+connection precise by translating the SPUD problem into a planning
+problem; the relevant excerpt of this problem is shown in
+Fig.~\ref{fig:white-rabbit-as-planning}.  Each action in this planning
+problem corresponds to the addition of a single elementary tree to the
+derivation; the first parameter of the action is a node name in the
+derivation tree, and the further parameters stand for the individuals
+to which the semantic roles will be instantiated.  The syntactic
+preconditions and effects are encoded using $\mathsf{open}$ and
+$\mathsf{canadjoin}$ literals; the status of each referring expression
+is tracked using $\mathsf{distractor}$ literals.  Notice that the
+action effects contain terms of the form $\mathsf{subj}(u)$, which
+construct new node names.  In order to represent the planning problem
+in PDDL, these terms can be eliminated by estimating an upper bound
+$n$ for the plan length, making $n$ copies of each action, ensuring
+that copy $i$ can only be applied in step $i$, and replacing the term
+$\mathsf{subj}(u)$ in an action copy by the constant
+$\mathsf{subj}_i$.
+
+In the example, the following plan solves the planning problem:
+\begin{enumerate}
+\item $\mathsf{sleeps}(\mathsf{root}, r_1)$
+\item $\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}), r_1)$
+\item $\mathsf{white}(\mathsf{subj}(\mathsf{root}), r_1)$
+\end{enumerate}
+
+The grammatical derivation in
+Fig.~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the generated
+sentence, can be systematically reconstructed from this plan.
+Therefore we can solve the sentence generation problem via the detour
+through planning and bring current search heuristics for planning to
+bear on generation.
 

### 2008-06-19 trunk/Papers/experiences/introduction.tex
changes to author block and introduction

@@
-\section{Introduction} \label{sec:introduction}
+\section{Introduction}
+\label{sec:introduction}
 
-Natural language generation (NLG; \citeauthor{reiter00building} 2000)
-is one of the major subfields of natural language processing.  It is
-concerned with computing natural language sentences or texts that
-convey a given piece of information to the user.  Problems in NLG are
-often more complex than their counterparts in parsing \cite{KolStr02},
-and indeed efficiency has not traditionally been the main focus of
-NLG.  On the other hand, NLG has some obvious parallels with planning:
-One tries to change the mental state of the hearer by applying actions
-which correspond to uttering words or sentences.  This parallel has a
-long tradition in NLG \cite{appelt:planning,young94dpocl}, and there
-has recently been a resurgence of interest in applying modern planning
-techniques to NLG \cite{Steedman-Petrick:07,KolSto07,benotti08b}.
-These new approaches combine an interest in planning as a modelling
-tool for natural language with a hope for improved efficiency by
-exploiting modern planning algorithms to help with the search.
+Natural language generation (NLG; \citeauthor{reiter00building}
+\citeyear{reiter00building}) is one of the major subfields of natural
+language processing, concerned with computing natural language sentences or
+texts that convey a given piece of information to the user. Intuitively,
+this task can be viewed as a problem involving actions, beliefs, and goals:
+an agent communicating with another agent tries to change the mental state
+of the hearer by applying actions which correspond to the utterance of
+words or sentences. This characterization of NLG suggests obvious parallels
+to automated planning---a view which has a long tradition in NLG
+\cite{appelt:planning,young94dpocl}.  More recently, there has been a
+resurgence of interest in applying modern planning techniques to NLG
+\cite{Steedman-Petrick:07,KolSto07,benotti08b}.  While efficiency has not
+been the main focus of NLG, problems in NLG are often more complex than
+their counterparts in other areas of natural language processing (such as
+parsing \cite{KolStr02}), giving rise to computationally challenging
+domains. These new approaches combine an interest in planning as a
+modelling tool for natural language, with a hope of improved efficiency by
+exploiting modern approaches and algorithms.
+
+%Natural language generation (NLG; \citeauthor{reiter00building} 2000)
+%is one of the major subfields of natural language processing.  It is
+%concerned with computing natural language sentences or texts that
+%convey a given piece of information to the user.  Problems in NLG are
+%often more complex than their counterparts in parsing \cite{KolStr02},
+%and indeed efficiency has not traditionally been the main focus of
+%NLG.  On the other hand, NLG has some obvious parallels with planning:
+%One tries to change the mental state of the hearer by applying actions
+%which correspond to uttering words or sentences.  This parallel has a
+%long tradition in NLG \cite{appelt:planning,young94dpocl}, and there
+%has recently been a resurgence of interest in applying modern planning
+%techniques to NLG \cite{Steedman-Petrick:07,KolSto07,benotti08b}.
+%These new approaches combine an interest in planning as a modelling
+%tool for natural language with a hope for improved efficiency by
+%exploiting modern planning algorithms to help with the search.
+
+%%%%%
+%The purpose of this paper is twofold.
 

### 2008-06-19 trunk/Papers/experiences/introduction.tex
more changes to the introduction

@@
 
+The focus of this paper is twofold. We begin by presenting two recent
+planning domains that arise in the context of natural language generation:
+the sentence generation domain and the GIVE domain. In the sentence
+generation task, the goal is to generate a single sentence that expresses a
+given meaning. \citeauthor{KolSto07}~\shortcite{KolSto07} cast this domain
+as a planning problem, where the plan encodes the necessary sentence and
+the actions correspond to uttering individual words. In the GIVE domain
+(``Generating Instructions in Virtual Environments''), we describe a new
+shared task that was recently posed as a challenge for the NLG community
+\cite{alexander07:_shared_task_propos}. GIVE uses planning as one module of
+an NLG system that generates natural-language instructions to guide a
+user performing a given task in a virtual environment.
+
+We will then discuss some of our experiences using off-the-shelf planners
+in these domains. In particular, we explore the efficiency of FF
+\cite{HoffmannNebel01} and SGPLAN \cite{hsu06:_new_featur_in_sgplan_for} on
+varying-sized instances of our planning domains. We also compare these
+results against an ad-hoc Java implementation of GraphPlan.  Our findings
+are mixed. On the one hand, it turns out that modern planners handle some
+of the search problems that arise in NLG quite easily. (Large problem
+instances remain a challenge.) On the other hand, these same planners spend
+tremendous amounts of time on preprocessing. For instance, FF spends 90\%
+of its runtime in the sentence generation domain on grounding out literals
+and actions, and ends up slower than our version of GraphPlan that avoids
+grounding. For domains like ours, which are dominated by the number of
+actions and the universe size rather than the combinatorics of the search
+problem, this observation suggests that the overall runtime of a planner
+could be improved by grounding more selectively.
+
 %Natural language generation (NLG; \citeauthor{reiter00building} 2000)
@@
 
-%%%%%
-%The purpose of this paper is twofold.
+%In this paper, we present two recent planning domains that arise in the
+%context of applying planning to natural language generation.  We will
+%first review the sentence generation domain, which was cast as a planning
+%problem by \citeauthor{KolSto07}~\shortcite{KolSto07}.  In this domain,
+%the goal is to generate a single sentence that expresses a given piece of
+%meaning; the plan encodes this sentence, and the actions correspond to
+%uttering individual words.  We will then introduce the GIVE domain
+%(``Generating Instructions in Virtual Environments'').  GIVE is a new
+%shared task that was recently posed as a challenge for the NLG community
+%\cite{alexander07:_shared_task_propos}.  It uses planning as one module of
+%a NLG system which generates natural-language instructions that will guide
+%the user towards performing a given task in a virtual environment.
 
-In this paper, we present two recent planning domains that arise in
-the context of applying planning to natural language generation.  We
-will first review the sentence generation domain, which was cast as a
-planning problem by Koller and Stone \citeyear{KolSto07}.  In this
-domain, the goal is to generate a single sentence that expresses a
-given piece of meaning; the plan encodes this sentence, and the
-actions correspond to uttering individual words.  We will then
-introduce the GIVE domain (``Generating Instructions in Virtual
-Environments'').  GIVE is a new shared task that was recently posed as
-a challenge for the NLG community
-\cite{alexander07:_shared_task_propos}.  It uses planning as one
-module of a NLG system which generates natural-language instructions
-that will guide the user towards performing a given task in a virtual
-environment.
-
-We will then discuss some experiences with using off-the-shelf
-planners in these domains.  In particular, we explore the efficiency
-of FF \cite{HoffmannNebel01} and SGPLAN
-\cite{hsu06:_new_featur_in_sgplan_for} on instances of our planning
-domains of varying sizes.  Our findings are mixed.  On the one hand,
-it turns out that modern planners handle the search problems that
-arise in NLG quite easily.  On the other hand, they spend tremendous
-amounts of time on preprocessing; for instance, FF spends 90\% of its
-runtime in the sentence generation domain on grounding out the
-literals and actions, and ends up slower than an ad-hoc Java
-implementation of GraphPlan that avoids grounding.  We argue that for
-domains like ours, which are dominated by the number of actions and the
-universe size rather than the combinatorics of the search problem, the
-overall runtime of a planner could be improved by grounding more
-selectively.
+%We will then discuss some experiences with using off-the-shelf planners in
+%these domains.  In particular, we explore the efficiency of FF
+%\cite{HoffmannNebel01} and SGPLAN \cite{hsu06:_new_featur_in_sgplan_for} on
+%instances of our planning domains of varying sizes.  Our findings are
+%mixed.  On the one hand, it turns out that modern planners handle the
+%search problems that arise in NLG quite easily.  On the other hand, they
+%spend tremendous amounts of time on preprocessing; for instance, FF spends
+%90\% of its runtime in the sentence generation domain on grounding out the
+%literals and actions, and ends up slower than an ad-hoc Java implementation
+%of GraphPlan that avoids grounding.  We argue that for domains like ours,
+%which are dominated by the number of actions and the universe size rather
+%than the combinatorics of the search problem, the overall runtime of a
+%planner could be improved by grounding more selectively.
 

### 2008-06-19 trunk/Papers/experiences/introduction.tex
more changes to the introduction

@@
 are mixed. On the one hand, it turns out that modern planners handle some
-of the search problems that arise in NLG quite easily. (Large problem
-instances remain a challenge.) On the other hand, these same planners spend
-tremendous amounts of time on preprocessing. For instance, FF spends 90\%
-of its runtime in the sentence generation domain on grounding out literals
-and actions, and ends up slower than our version of GraphPlan that avoids
-grounding. For domains like ours, which are dominated by the number of
-actions and the universe size rather than the combinatorics of the search
-problem, this observation suggests that the overall runtime of a planner
-could be improved by grounding more selectively.
+of the search problems that arise in NLG quite easily (although large
+problem instances remain a challenge). On the other hand, these same
+planners spend tremendous amounts of time on preprocessing. For instance,
+FF spends 90\% of its runtime in the sentence generation domain on
+grounding out literals and actions, and ends up slower than our version of
+GraphPlan that avoids grounding. For domains like ours, which are dominated
+by the number of actions and the universe size, rather than the
+combinatorics of the search problem, this observation suggests that the
+overall runtime of a planner could be improved by grounding more
+selectively. We therefore offer these domains as challenges for the
+planning community.
 

### 2008-06-20 trunk/Papers/experiences/discussion.tex
small change to introduction; changed to discussion section

@@
 
-The positive conclusion we can draw from the experiments reported
-above is that for the most part, modern planners are pretty good at
-controlling the search for the NLG-related inputs that we looked at.
-This is particularly true for FF in the sentence generation domain,
-which generates nontrivial 14-word sentences in about two seconds.
-There is still room for improvement (a special-purpose algorithm for
-referring expressions could generate ten of these 14 words in a few
-milliseconds), and the search becomes slower -- for obvious reasons --
-in the almost completely unconstrained world of Experiment 2, but
-search is not the main efficiency bottleneck.
+The positive conclusion we can draw from the experiments reported above is
+that modern planners are fairly good at controlling the search for many of
+the moderately-sized NLG problems we looked at. This is particularly true
+for FF in the sentence generation domain, which generates nontrivial
+14-word sentences in about two seconds. There is still room for
+improvement, however. For instance, by comparison a special-purpose
+algorithm for referring expressions could generate ten of these 14-word
+sentences in a few milliseconds. Although search efficiency is not the main
+performance concern in these domains, it nevertheless becomes a factor in
+scenarios like Experiment~2, where the world is almost completely
+unconstrained and (unsurprisingly) the search becomes much slower.
 
-The most restrictive bottleneck in the two domains we have looked at
-in this paper is the initial grounding step that almost all modern
-planners perform.  Our conclusion from the experiments reported above
-is that this is a good strategy in domains where the typical plan
-length dominates the typical universe size, i.e.\ where the initial
-investment into grounding pays off in saved unifications during the
-search.  However, as we have shown, there are natural planning domains
-in which relatively short plans must be computed in large universes;
-it is not unrealistic for the universe in the sentence generation
-domain to consist of thousands of domain individuals and tens of
-thousands of actions, some of which take three domain individuals as
-parameters.  The common strategy of splitting the universe over
-several types of individuals will not be very effective here.
+%The positive conclusion we can draw from the experiments reported
+%above is that for the most part, modern planners are pretty good at
+%controlling the search for the NLG-related inputs that we looked at.
+%This is particularly true for FF in the sentence generation domain,
+%which generates nontrivial 14-word sentences in about two seconds.
+%There is still room for improvement (a special-purpose algorithm for
+%referring expressions could generate ten of these 14 words in a few
+%milliseconds), and the search becomes slower -- for obvious reasons --
+%in the almost completely unconstrained world of Experiment 2, but
+%search is not the main efficiency bottleneck.
+
+The most restrictive bottleneck in the two domains we investigated is the
+initial grounding step that many modern planners perform. While this may be
+a good strategy for traditional planning benchmarks and IPC domains---where
+plan length often dominates universe size, and an initial investment into
+grounding pays off in saved instantiations during the search---this
+approach is less efective in the domains we've considered. As our
+experiements have shown, there are natural planning domains in which
+relatively short plans must be computed in large universes. For instance,
+in the case of the sentence generation domain, it is not unrealistic for
+the universe to consist of thousands of domain individuals and tens of
+thousands of actions, some of which require three domain individuals as
+parameters. The common strategy of splitting the universe over several
+types of individuals will not be very effective here.
+
+%The most restrictive bottleneck in the two domains we have looked at
+%in this paper is the initial grounding step that almost all modern
+%planners perform.  Our conclusion from the experiments reported above
+%is that this is a good strategy in domains where the typical plan
+%length dominates the typical universe size, i.e.\ where the initial
+%investment into grounding pays off in saved unifications during the
+%search.  However, as we have shown, there are natural planning domains
+%in which relatively short plans must be computed in large universes;
+%it is not unrealistic for the universe in the sentence generation
+%domain to consist of thousands of domain individuals and tens of
+%thousands of actions, some of which take three domain individuals as
+%parameters.  The common strategy of splitting the universe over
+%several types of individuals will not be very effective here.
 

### 2008-06-20 trunk/Papers/experiences/introduction.tex
small change to introduction; changed to discussion section

@@
 in these domains. In particular, we explore the efficiency of FF
-\cite{HoffmannNebel01} and SGPLAN \cite{hsu06:_new_featur_in_sgplan_for} on
-varying-sized instances of our planning domains. We also compare these
-results against an ad-hoc Java implementation of GraphPlan.  Our findings
-are mixed. On the one hand, it turns out that modern planners handle some
-of the search problems that arise in NLG quite easily (although large
-problem instances remain a challenge). On the other hand, these same
-planners spend tremendous amounts of time on preprocessing. For instance,
-FF spends 90\% of its runtime in the sentence generation domain on
-grounding out literals and actions, and ends up slower than our version of
-GraphPlan that avoids grounding. For domains like ours, which are dominated
-by the number of actions and the universe size, rather than the
-combinatorics of the search problem, this observation suggests that the
-overall runtime of a planner could be improved by grounding more
+\cite{HoffmannNebel01} and SGPLAN
+\cite{hsu06:_new_featur_in_sgplan_for}---planners that have been successful
+on traditional benchmarks, and in the International Planning
+Competition (IPC)---on varying-sized instances of our planning domains. We
+also compare these results against an ad-hoc Java implementation of
+GraphPlan.  Our findings are mixed. On the one hand, it turns out that
+modern planners handle some of the search problems that arise in NLG quite
+easily (although large problem instances remain a challenge). On the other
+hand, these same planners spend tremendous amounts of time on
+preprocessing. For instance, FF spends 90\% of its runtime in the sentence
+generation domain on grounding out literals and actions, and ends up slower
+than our version of GraphPlan that avoids grounding. For domains like ours,
+which are dominated by the number of actions and the universe size, rather
+than the combinatorics of the search problem, this observation suggests
+that the overall runtime of a planner could be improved by grounding more
 selectively. We therefore offer these domains as challenges for the

### 2008-06-20 trunk/Papers/experiences/domains.tex
small changes to the planning in NLG section
pictures
details on the give problem

@@
 
-In this section, we present the two domains: first the sentence
-generation domain, then the instruction giving domain.
+In this section, we present our planning domains: the sentence
+generation domain and the instruction giving domain.
 
@@
   \centering
-  xxxx
+  \includegraphics[width=\columnwidth]{pic-grammar}
   \caption{The example grammar.}
@@
   \centering
-  xxxx
+  \includegraphics[width=\columnwidth]{pic-derivation}
   \caption{Derivation of ``The white rabbit sleeps.''}
@@
 which is able to give natural-language instructions which will guide a
-human user in performing some task in a virtual environment.  A map of
-an example world is shown in Fig.~\ref{fig:give-development-world};
-instructions for the first steps could be ``turn left; press the
-button; turn right and walk through the door;'' and so forth.  From an
+human user in performing some task in a virtual environment.  From an
 NLG perspective, GIVE makes for an interesting challenge because it is
@@
 Internet.  The first GIVE evaluation will take place in late 2008;
-currently seven research teams from five countries are working on
+currently eight research teams from five countries are working on
 developing systems to participate in the challenge.
@@
 
-\todo{explain world? -- yes, definitely}
+A map of an example world is shown in
+Fig.~\ref{fig:give-development-world}.  In this world, the user's task
+is to pick up a trophy in the top left room.  The trophy is hidden in
+a safe behind a picture, and the user must therefore first move the
+picture away and open the safe by pushing a sequence of buttons (the
+small square boxes on the walls) in the correct order.  In order to
+get to all these buttons, they must also open the lower door and
+deactivate the alarm tile by pushing further buttons.  To simplify
+both the planning and the NLG task, the world is discretized into
+tiles of equal size; the user can turn by 90 degree steps in either
+direction, and can move from the center of one tile to the center of
+the next.  Some actions that the user can take are shown in
+Fig.~\ref{fig:give-planning}.  The first few steps of one shortest
+plan to achieve the goal are:
+\begin{enumerate}
+\item $\mathsf{turn-left}(\mathsf{north}, \mathsf{west})$
+\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$
+\item $\mathsf{manipulate-b1-off-on}(\mathsf{pos\_5\_2})$
+\item $\mathsf{turn-right}(\mathsf{west}, \mathsf{north})$
+\end{enumerate}
 
-One crucial component of a GIVE system is a planner which will compute
-a domain plan that consists of the actions the user has to execute to
-achieve the goal.  For these purposes, the virtual world is
-discretized into a planning problem in which each object has a
-symbolic position and orientation.  The first few action instances in
-the example domain could be $turn-left(north,west)$, $move(pos-3-2,
-pos-2-2, west)$, $manipulate-b_1-off-on(pos-2-2)$,
-$turn-right(west,north)$, \todo{complete this} and so on.  Under the
-current implementation of the planning problem, the shortest plan for
-getting the player from the initial state to the goal (in which they
-are holding the trophy) has 104 steps, most of which are ``move''
-actions.  The bulk of the GIVE problem is therefore a variant of the
-Gridworld problem, which also involves finding a route through a world
-with discrete positions -- but with the need to press buttons in the
-right order and with somewhat more complicated room shapes.
+\begin{figure}
+  \centering
+  xxxxx
+  \caption{Excerpt from the GIVE planning problem.}
+  \label{fig:give-planning}
+\end{figure}
 
-It is interesting to note that although the task of a GIVE NLG system
-could be seen as mapping a domain plan to a sequence of NL
-instructions, this mapping is not trivial.  It would be
-\emph{possible} to express the last part of the above example plan as
-``turn right; walk forward; walk forward; turn right; walk forward;
-walk forward; turn left; walk forward'', but this is a very clumsy way
-of phrasing the instruction ``turn around and walk through the door''
-that will frustrate users and lead to bad evaluation results.  In
-other words, it is important to aggregate multiple movement actions
-into a single NL instruction.  On the other hand, it may be necessary
-to generate instructions that don't correspond to domain actions.  For
-instance, it will be necessary to find a correct referring expression
-for $b_1$ in order to express the action
-$manipulate-b_1-off-on(pos-2-2)$.  The interaction between referring
-and instructing can go much deeper than that.  Imagine, for example,
-that the user just entered the top right room, and we want to instruct
-them to press $b_4$.  At that point in time, the user can see six
-different buttons, and generating a single instruction of the form
-``push X'' is difficult.  However, an instruction sequence like ``walk
-forward; turn left; now press the button right in front of you'' will
-be successful and natural.  Here the first two instructions do double
-duty by preparing the simple referring expression ``the button right
-in front of you'', which is only valid after the user's position and
-orientation changed.  A generation system that is supposed to generate
-such instruction sequences must be tightly integrated with the planner
-in order to exploit the simultaneous effect of the first two
-instructions on the world and on the communicative situation.
+This shortest plan consists of 108 action instances, most of which are
+``move'' actions.  The bulk of the GIVE problem is therefore a variant
+of the Gridworld problem, which also involves finding a route through
+a world with discrete positions -- but with the need to press buttons
+in the right order and with somewhat more complicated room shapes.  In
+this example, the NLG system could generate an instruction sequence
+starting with ``turn left; walk forward; press the button; turn right;
+walk forward; walk forward; turn right; walk forward; walk forward;
+turn left; walk forward''.  However, this would be clumsy and boring,
+and it is to be expected that frustrated users will quickly cancel an
+interaction with such a system, resulting in a bad evaluation score.
+Having just pressed the button, it would be much better to say ``turn
+around and walk through the door''.  This means that the mapping from
+action instances to natural language instructions is not trivial.
+Conversely, it may also be necessary to express a single planning step
+with several instructions.  Having just entered the top left room, it
+may be easier for the user to understand the instruction ``walk to the
+center of the room; turn right; now press the green button in front of
+you'' than the instruction ``press the green button on the wall to
+your right''.  To plan the referring expression ``the green button in
+front of you'' at a time when the button is not yet in front of the
+user, the NLG system must keep track of the hypothetical changes in
+what objects are visible to the user.  Thus the acts of referring and
+instructing, and the modules for planning and NLG, must be integrated
+tightly.
 

### 2008-06-21 trunk/Papers/experiences/conclusion.tex
small change to introduction; changes to discussion and conclusion

--- 2008-06-05 trunk/Papers/experiences/conclusion.tex
@@
 \section{Conclusion} \label{sec:conclusion}
+
+In this paper, we introduced two novel planning domains arising from
+problems in natural language generation: the sentence generation domain and
+the GIVE navigation domain. We also reported on the results of a number of
+experiments in which we applied off-the-shelf planners to these domains.
+Our main results were mixed. While modern planners do a pretty good job of
+controlling the complexity of search, they also suffer from practical
+problems that limit their performance in unexpected places. In particular,
+in both domains the grounding step performed by both planners dominates the
+time it takes to perform the search itself. The NLG community's recent
+interest in planning also presents a valuable opportunity for planning
+researchers: provided some of the challenges we have highlighted can be
+addressed, projects like GIVE offer a constructive platform for the
+planning community to showcase their techniques to a wider audience---and
+to improve the quality of their tools for real-world planning tasks.
+
+%promising, worthwhile, far-reaching, productive
+
+%In this paper, we have introduced two novel planning domains related
+%to problems in natural language generation: the sentence generation
+%problem and the GIVE navigation problem.  We have then reported on a
+%number of experiments in which we applied off-the-shelf planners to
+%these domains.  Our main results were that while modern planners do
+%pretty well in controlling the complexity of search, they suffer from
+%practical problems that limit their performance in unexpected places.
+%In particular, in both domains the grounding step that both planners
+%perform tends to take much more time than the search itself.
 

### 2008-06-21 trunk/Papers/experiences/discussion.tex
small change to introduction; changes to discussion and conclusion

@@
 sentences in a few milliseconds. Although search efficiency is not the main
-performance concern in these domains, it nevertheless becomes a factor in
+performance metric in these domains, it nevertheless becomes a factor in
 scenarios like Experiment~2, where the world is almost completely
@@
 
-We believe that if planning is to become a mature technology which
-real-world users will believe in, it must find a way to deal with the
-grounding problem more gracefully than current systems do.  The
-research focus in developing planning algorithms has quite rightfully
-been on dealing with the search, and for this it is convenient --
-depending on the particular algorithm, necessary -- to ground out all
-predicates and actions before starting the search itself.  However, we
-think that the domains presented here are not that unusual in their
-ratio of plan length against universe size.  At the end of the day,
-real-world users will care about the \emph{total} runtime of a
-planner, and this is more than just the search.
+The recent trend in planning research has (rightfully) focused on the
+development of alorithms that control search in sophisticated ways,
+resulting in a host of planners that are more powerful and more successful
+than their predecessors. Depending on the particular algorithm, however,
+these planners often ground out sets of predicates and actions before
+starting the search itself, which has a very pronounced effect on domains
+such as those described above. Since our domains are not that unusual in
+their structure and composition, we hope that the lessons learnt from our
+experiences can help improve the performance of current systems. We believe
+that such improvements are necessary if planning is to become a more mature
+technology that can offer tools to a wider community of users. At the end
+of the day, real-world users will care about the \emph{total} runtime of a
+planner, and this is more than just the search time.
+%We offer our larger problem instances as future challenges for the
+%planning community.
 
-By and large, it is a relatively pleasant experience to come as a
-``customer'' to the planning community: Thanks to the competitions, it
-is easy to identify and download a fast implementation, at least for
-Linux.  However, in the course of our experiments we found (and
-reported) bugs in both SGPLAN and FF.  And as our experiments have
-shown, there is no one silver bullet: Even planners as closely related
-as FF and SGPLAN can differ significantly in their performance on
-different domains.
+%We believe that if planning is to become a mature technology which
+%real-world users will believe in, it must find a way to deal with the
+%grounding problem more gracefully than current systems do.  The
+%research focus in developing planning algorithms has quite rightfully
+%been on dealing with the search, and for this it is convenient --
+%depending on the particular algorithm, necessary -- to ground out all
+%predicates and actions before starting the search itself.  However, we
+%think that the domains presented here are not that unusual in their
+%ratio of plan length against universe size.  At the end of the day,
+%real-world users will care about the \emph{total} runtime of a
+%planner, and this is more than just the search.
+
+By and large, our experiences with the planning community from the point of
+view of a ``customer'' (one of the authors is not a planning researcher)
+have been relatively pleasant. Thanks to the planning competitions, it is
+easy to identify and download a fast implementation, at least for Linux.
+However, in the course of our experiments we found (and reported) bugs in
+both SGPLAN and FF. We also discovered that deciding between the range of
+available planners is not always straightforward. As our experiments have
+shown, even planners as closely related as FF and SGPLAN can differ
+significantly in their performance on different domains.
+
+%By and large, it is a relatively pleasant experience to come as a
+%``customer'' to the planning community: Thanks to the competitions, it
+%is easy to identify and download a fast implementation, at least for
+%Linux.  However, in the course of our experiments we found (and
+%reported) bugs in both SGPLAN and FF.  And as our experiments have
+%shown, there is no one silver bullet: Even planners as closely related
+%as FF and SGPLAN can differ significantly in their performance on
+%different domains.
 

### 2008-06-21 trunk/Papers/experiences/introduction.tex
small change to introduction; changes to discussion and conclusion

--- 2008-06-19 trunk/Papers/experiences/introduction.tex
@@
 in these domains. In particular, we explore the efficiency of FF
-\cite{HoffmannNebel01} and SGPLAN \cite{hsu06:_new_featur_in_sgplan_for} on
-varying-sized instances of our planning domains. We also compare these
-results against an ad-hoc Java implementation of GraphPlan.  Our findings
-are mixed. On the one hand, it turns out that modern planners handle some
-of the search problems that arise in NLG quite easily (although large
-problem instances remain a challenge). On the other hand, these same
+\cite{HoffmannNebel01} and SGPLAN
+\cite{hsu06:_new_featur_in_sgplan_for}---planners that have been successful
+on traditional benchmarks, and in the International Planning Competition
+(IPC)---on a range of problem instances from our planning domains.  We also
+compare these results against an ad-hoc Java implementation of GraphPlan.
+Our findings are mixed. On the one hand, it turns out that modern planners
+handle some of the search problems that arise in NLG quite easily (although
+large problem instances remain a challenge). On the other hand, these same
 planners spend tremendous amounts of time on preprocessing. For instance,

### 2008-06-22 trunk/Papers/experiences/domains.tex
added PDDL actions; changes to Planning in NLG section (GIVE subsection)

@@
 
+\begin{figure}[t]
+\centering
+{\small%
+\begin{verbatim}
+(:action add-sleeps
+   :parameters (?u - node
+                ?xself - individual
+                ?xsubj - individual)
+   :precondition (and (subst S ?u)
+      (referent ?u ?xself)
+      (sleep ?xself ?xsubj))
+   :effect (and (not (subst S ?u))
+      (expressed sleep ?xself ?xsubj)
+      (subst NP (subj ?u))
+      (referent (subj ?u) ?xsubj)
+      (forall (?y - individual)
+         (when (not (= ?y ?xself))
+            (distractor (subj ?u) ?y)))))
+
+(:action add-rabbit
+   :parameters (?u - node
+                ?xself - individual)
+   :precondition (and (subst NP ?u)
+      (referent ?u ?xself)
+      (rabbit ?xself))
+   :effect (and (not (subst NP ?u))
+      (canadjoin N ?u)
+      (forall (?y - individual)
+         (when (not (rabbit ?y))
+           (not (distractor ?u ?y))))))
+
+(:action add-white
+   :parameters (?u - node
+                ?xself - individual)
+   :precondition (and (canadjoin N ?u)
+      (referent ?u ?xself)
+      (rabbit ?xself))
+   :effect (forall (?y - individual)
+      (when (not (white ?y))
+         (not (distractor ?u ?y)))))
+\end{verbatim}}%
+\caption{PDDL actions for generating the sentence ``The white rabbit
+sleeps''.}
+\label{fig:pddl-crisp}
+\end{figure}
+
 This process has clear parallels to planning: We manipulate a state by
@@
 The object of the GIVE Challenge (``Generating Instructions in Virtual
-Environments''; Koller et al.\
-\citeyear{alexander07:_shared_task_propos}) is to build an NLG system
-which is able to give natural-language instructions which will guide a
-human user in performing some task in a virtual environment.  From an
-NLG perspective, GIVE makes for an interesting challenge because it is
-a theory-neutral task that exercises all components of an NLG system,
-and emphasizes the study of communication in a (simulated) physical
-environment.  It also has the advantage that the user and the NLG
-system can be physically in different places, as long as the 3D client
-and the NLG system are connected over a network.  This makes it
-possible to evaluate GIVE NLG systems on a large scale over the
+Environments''; Koller et al.\ \citeyear{alexander07:_shared_task_propos})
+is to build an NLG system which is able to produce natural-language
+instructions which will guide a human user in performing some task in a
+virtual environment. From an NLG perspective, GIVE makes for an
+interesting challenge because it is a theory-neutral task that exercises
+all components of an NLG system, and emphasizes the study of communication
+in a (simulated) physical environment. It also has the advantage that the
+user and the NLG system can be physically in different places, as long as
+the 3D client and the NLG system are connected over a network. This makes
+it possible to evaluate GIVE NLG systems on a large scale over the
 Internet. The first GIVE evaluation will take place in late 2008;
 currently eight research teams from five countries are working on
-developing systems to participate in the challenge.
+developing systems to participate in the challenge.\footnote{See the GIVE
+ website for more details about this project:
+ \url{homepages.inf.ed.ac.uk/v1akolle/proj/give/}.}
 
@@
 \centering
-\includegraphics[width=1 \columnwidth]{give_world_no_expl}
+%\includegraphics[width=1 \columnwidth]{give_world_no_expl}
+\includegraphics[width=1 \columnwidth]{give_world_2}
 \caption{Map of the GIVE development world.}
@@
 
-A map of an example world is shown in
-Fig.~\ref{fig:give-development-world}.  In this world, the user's task
-is to pick up a trophy in the top left room.  The trophy is hidden in
-a safe behind a picture, and the user must therefore first move the
-picture away and open the safe by pushing a sequence of buttons (the
-small square boxes on the walls) in the correct order.  In order to
-get to all these buttons, they must also open the lower door and
-deactivate the alarm tile by pushing further buttons.  To simplify
-both the planning and the NLG task, the world is discretized into
-tiles of equal size; the user can turn by 90 degree steps in either
-direction, and can move from the center of one tile to the center of
-the next.  Some actions that the user can take are shown in
-Fig.~\ref{fig:give-planning}.  The first few steps of one shortest
-plan to achieve the goal are:
-\begin{enumerate}
-\item $\mathsf{turn-left}(\mathsf{north}, \mathsf{west})$
-\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$
-\item $\mathsf{manipulate-b1-off-on}(\mathsf{pos\_5\_2})$
-\item $\mathsf{turn-right}(\mathsf{west}, \mathsf{north})$
-\end{enumerate}
+A map of an example GIVE world is shown in
+Fig.~\ref{fig:give-development-world}.  In this world, the user's task is
+to pick up a trophy in the top left room.  The trophy is hidden in a safe
+behind a picture, so the user must first move the picture out of the way
+and open the safe by pushing a sequence of buttons (the small square boxes
+on the walls) in the correct order.  In order to get to all these buttons,
+the user must also open the lower door and deactivate the alarm tile by
+pushing further buttons.  To simplify both the planning and the NLG task,
+the world is discretized into tiles of equal size; the user can turn by 90
+degree steps in either direction, and can move from the centre of one tile
+to the centre of the next. Some of the actions the user can take are shown
+in Fig.~\ref{fig:give-planning}, in PDDL syntax.
 
-\begin{figure}
-  \centering
-  xxxxx
-  \caption{Excerpt from the GIVE planning problem.}
+\begin{figure}[t]
+{\small%
+\begin{verbatim}
+(:action move
+   :parameters (?from - position
+                ?to - position
+                ?ori - orientation)
+   :precondition (and (player-pos ?from) 
+      (adjacent ?from ?to ?ori) 
+      (player-orient ?ori)
+      (not-blocked ?from ?to)
+      (not-alarmed ?to))
+   :effect (and (not (player-pos ?from))
+      (player-pos ?to)))
+
+(:action turn-left
+   :parameters (?ori - orientation
+                ?newOri - orientation)
+   :precondition (and (player-orient ?ori)
+      (next-orient-left ?ori ?newOri))
+   :effect (and (not (player-orient ?ori))
+      (player-orient ?newOri)))
+
+(:action turn-right
+   :parameters (?ori - orientation
+                ?newOri - orientation)
+   :precondition (and (player-orient ?ori)
+      (next-orient-right ?ori ?newOri))
+   :effect (and (not (player-orient ?ori))
+      (player-orient ?newOri)))
+
+(:action manipulate-b1-off-on
+   :parameters (?pos - position)
+   :precondition (and (state b1 off)
+      (player-pos ?pos)
+      (position b1 ?pos))
+   :effect (and (not (state b1 off))
+      (state b1 on)
+      (not (state d1 closed))
+      (state d1 open) 
+      (not (blocked pos_6_5 pos_6_4))
+      (not (blocked pos_6_4 pos_6_5))))
+\end{verbatim}}%
+\caption{PDDL actions for the GIVE domain.}
 \label{fig:give-planning}
@@
 
-This shortest plan consists of 108 action instances, most of which are
-``move'' actions.  The bulk of the GIVE problem is therefore a variant
-of the Gridworld problem, which also involves finding a route through
-a world with discrete positions -- but with the need to press buttons
-in the right order and with somewhat more complicated room shapes.  In
-this example, the NLG system could generate an instruction sequence
-starting with ``turn left; walk forward; press the button; turn right;
-walk forward; walk forward; turn right; walk forward; walk forward;
-turn left; walk forward''.  However, this would be clumsy and boring,
-and it is to be expected that frustrated users will quickly cancel an
+The plans generated in the GIVE domain are often nontrivial. For instance,
+in the example above the shortest plan consists of 108 action steps; the
+first few steps are as follows:
+%
+\begin{enumerate}
+\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north}, \mathsf{west})$
+\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$
+\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$
+\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$
+\end{enumerate}
+%
+(Most of the actions in this plan are ``move'' actions.) The bulk of the
+GIVE problem can be considered a variant of the Gridworld problem, which
+also involves finding a route through a world with discrete positions---but
+with the additional need to press buttons in the right order, reason about
+many more objects in the world, and navigate somewhat more complicated room
+shapes.
+
+The task of generating natural language instructions in this domain is also
+more complicated than just a straightforward translation of the action
+steps in a plan. For instance, the NLG system could generate an instruction
+sequence starting with ``turn left; walk forward; press the button; turn
+right; walk forward; walk forward; turn right; walk forward; walk forward;
+turn left; walk forward''.  However, this would be clumsy and boring, and
+it is to be expected that frustrated users will quickly cancel an
 interaction with such a system, resulting in a bad evaluation score.
@@
 action instances to natural language instructions is not trivial.
-Conversely, it may also be necessary to express a single planning step
-with several instructions.  Having just entered the top left room, it
-may be easier for the user to understand the instruction ``walk to the
-center of the room; turn right; now press the green button in front of
-you'' than the instruction ``press the green button on the wall to
-your right''.  To plan the referring expression ``the green button in
-front of you'' at a time when the button is not yet in front of the
-user, the NLG system must keep track of the hypothetical changes in
-what objects are visible to the user.  Thus the acts of referring and
-instructing, and the modules for planning and NLG, must be integrated
-tightly.
+Conversely, it may also be necessary to express a single planning step with
+several instructions.  Having just entered the top left room, it may be
+easier for the user to understand the instruction ``walk to the centre of
+the room; turn right; now press the green button in front of you'' than the
+instruction ``press the green button on the wall to your right''.  To plan
+the referring expression ``the green button in front of you'' at a time
+when the button is not yet in front of the user, the NLG system must keep
+track of the hypothetical changes in what objects are visible to the user.
+Thus the acts of referring and instructing, and the modules for planning
+and NLG, must be tightly integrated.
 
-\todo{equivalent plans; plan recognition; real-time requirements}
+Plan execution monitoring also plays an important role in the GIVE problem.
+At a high level, the system needs to monitor a user's actions and compare
+them against the generated instruction set to determine if the user has
+correctly followed directions or not. In the case of the latter, new
+instructions may have to be generated. In practice, the situation
+can be quite complicated since the mental state of the user is not known
+and so the system must observe the user's actions in real time. For
+example, a user directed to ``turn around and walk through the door'' may
+not necessarily perform these actions to the letter, i.e., immediately
+turning 180 degrees and proceeding directly to the door. Instead, the user
+might take a roundabout route through the room, eventually exiting out the
+door. Although the user's actions do not match the generated instructions
+exactly, they meet the intended goal. The system must be able to identify
+such ``equivalent'' plans and not immediately generate new instructions as
+soon as the user's actions have gone off course. Furthermore, a user can
+communicate certain intentions to the system, both through action and
+inaction. For instance, the system may correctly infer that a user has
+failed to follow instructions if the user exits a room when given a
+directive to ``walk to the centre of the room''. However, the system should
+also be able to make a similar conclusion if a user simply does nothing
+when given the instruction. Thus, such real-time plan monitoring is
+essential in the GIVE domain, and also affects other modules in the system.
 

### 2008-06-22 trunk/Papers/experiences/experiences.tex
a bit about GIVE
changes to author block and introduction
added PDDL actions; changes to Planning in NLG section (GIVE subsection)

@@
 \usepackage{courier}
-\title{Experiences with planning for natural language generation}
-\author{Alexander Koller \and Ron Petrick \\
+
+\usepackage{graphicx}
+\usepackage{url}
+
+\newcommand{\todo}[1]{\textbf{(#1)}}
+
+\title{Experiences with Planning for Natural Language Generation}
+\author{Alexander Koller \and Ronald Petrick \\
 School of Informatics \\
-University of Edinburgh}
+University of Edinburgh \\
+Edinburgh \ EH8 9AB, Scotland, UK \\
+\texttt{\{a.koller,r.petrick\}@ed.ac.uk}}
+
+\nocopyright
 

### 2008-06-22 trunk/Papers/experiences/experiments.tex
minor edits
added PDDL actions; changes to Planning in NLG section (GIVE subsection)

@@
 plan from the initial state to the goal in 0.3 seconds, which is fast
-enough for the application.\footnote{All runtimes reported here were
-  measured on a Pentium 4 CPU at 3 GHz.  Java programs were allowed to
-  ``warm up'', i.e.\ the planner was run five times in a row and the
-  first four measurements discarded to ensure that the JVM had
-  just-in-time compiled all relevant bytecode.}  In the sentence
-generation domain, FF dramatically outperforms the best previously
-known algorithm for the same problem (a reimplementation of
-\cite{Stone2003a}), although the latter is a greedy search algorithm
-with a heuristic that is hand-tailored to the domain.
+enough for the application.\footnote{All reported runtimes were
+  measured on a Pentium 4 CPU running at 3 GHz. Java programs were allowed
+  to ``warm up'', i.e.\ the planner was run five times in a row and the
+  first four measurements discarded to ensure that the JVM had just-in-time
+  compiled all relevant bytecode.}  In the sentence
+generation domain, FF dramatically outperforms the best previously known
+algorithm for the same problem (a reimplementation of \cite{Stone2003a}),
+although the latter is a greedy search algorithm with a heuristic that is
+hand-tailored to the domain.
 
@@
   \centering
-  \includegraphics[width=0.4\textwidth]{pic-runtime-modifiers-with-sgplan}
+  \includegraphics[width=1\columnwidth]{pic-runtime-modifiers-with-sgplan}
   \caption{Runtimes in the sentence generation domain.  The horizontal
@@
   \centering
-  \includegraphics[width=0.4 \textwidth]{pic-buttons}
+  \includegraphics[width=1\columnwidth]{pic-buttons}
   \caption{Minimal GIVE world.}
@@
   \centering
-  \includegraphics[width=0.4\textwidth]{pic-runtime-buttons}
+  \includegraphics[width=1\columnwidth]{pic-runtime-buttons}
   \caption{Runtimes of FF and SGPLAN on the minimal GIVE worlds for $h=20$. The horizontal axis is $n$, ranging from 1 to 40.}
@@
   \centering
-  \includegraphics[width=0.4\textwidth]{pic-empty-buttons}
+  \includegraphics[width=1\columnwidth]{pic-empty-buttons}
   \caption{GIVE world with junk positions.}
@@
   \centering
-  \includegraphics[width=0.4\textwidth]{pic-runtime-empty-world}
+  \includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
   \caption{Runtimes of FF and SGPLAN on the GIVE worlds with junk

### 2008-06-22 trunk/Papers/experiences/domains.tex
changed to Planning in NLG section (sentence generation subsection)

@@
 Sentence generation is the problem of computing, from a grammar and a
-semantic representation, a single sentence that expresses this piece
-of meaning.  This problem is traditionally split into several steps
-\cite{reiter00building}.  In a first step, called \emph{sentence
-  planning}, the semantic representation is first enriched with more
-information; for instance, \emph{referring expressions}, which refer
-to individuals that we want to talk about, are determined at this
-point.  In a second step, called \emph{surface realization}, this
-enriched representation is then translated into a natural-language
-sentence, using the grammar.
+semantic representation, a single sentence that expresses this piece of
+meaning.  This problem is traditionally split into several steps
+\cite{reiter00building}.  In a first step, called \emph{sentence planning},
+the semantic representation is first enriched with more information; for
+instance, \emph{referring expressions}, which refer to individuals
+that we want to talk about, are determined at this point. In a second
+step, called \emph{surface realization}, this enriched representation is
+then translated into a natural-language sentence, using the grammar.
 
-However, determining referring expressions (REs) and realization
-interact, so it would be beneficial to perform both steps together.
-This was the goal of the SPUD system \cite{Stone2003a}, which
-performed both steps together using a top-down generation algorithm
-based on tree-adjoining grammars \cite{joshi;etal1997} whose lexical
-entries were equipped with semantic and pragmatic information.
-Unfortunately, SPUD suffered from having to explore a huge search
-space, and had to resort to a non-optimal greedy search strategy to
+In practice, the process of determining referring expressions typically
+interacts with the realization step and so it turns out to be beneficial to
+perform both of these steps together. This was the goal of the SPUD system
+\cite{Stone2003a}, which performed both steps together using a top-down
+generation algorithm based on tree-adjoining grammars \cite{joshi;etal1997}
+whose lexical entries were equipped with semantic and pragmatic
+information. Unfortunately, SPUD suffered from having to explore a huge
+search space, and had to resort to a non-optimal greedy search strategy to
 retain reasonable efficiency.
 
-\cite{KolSto07} attempt to improve the efficiency by translating the
-SPUD sentence generation problem into a planning problem and using a
-planning algorithm for generation.  We will now illustrate how they do
-this by a (simplified) example.
+%However, determining referring expressions and realization
+%interact, so it would be beneficial to perform both steps together.
+%This was the goal of the SPUD system \cite{Stone2003a}, which
+%performed both steps together using a top-down generation algorithm
+%based on tree-adjoining grammars \cite{joshi;etal1997} whose lexical
+%entries were equipped with semantic and pragmatic information.
+%Unfortunately, SPUD suffered from having to explore a huge search
+%space, and had to resort to a non-optimal greedy search strategy to
+%retain reasonable efficiency.
 
@@
 
-Consider a knowledge base containing the individuals $r_1$ and $r_2$;
-$r_1$ and $r_2$ are rabbits, $r_1$ is white and $r_2$ is brown, and
-$r_1$ sleeps.  Let's say we want to express the information
-$\{\mathsf{sleep}(r_1)\}$ using the tree-adjoining grammar shown in
-Fig.~\ref{fig:white-rabbit-sleeps-grammar}.  This grammar consists of
-\emph{elementary trees}, each of which contributes certain
-\emph{semantic content}.  We can instantiate these trees by
-substituting individuals for \emph{semantic roles}, such as
-$\mathsf{self}$ and $\mathsf{subj}$, and then combine the tree
-instances as shown in Fig.~\ref{fig:white-rabbit-sleeps-deriv} to
-obtain the sentence ``The white rabbit sleeps''.  The SPUD algorithm
-computes this grammatical derivation by starting with the elementary
-tree for ``sleeps''; this satisfies the need to convey the semantic
-information, but introduces a need to generate a noun phrase (NP) for
-the subject, which also refers uniquely to the target referent $r_1$.
-In a second step, it substitutes the tree for ``the rabbit'' into the
-open NP leaf, which makes the derivation grammatically complete; but
-as there are two different individuals that could be described as
-``the rabbit'' -- technically, $r_2$ is still a \emph{distractor} for
-the referring expression --, we are still not finished.  This is why
-we add the tree for ``white'' to the derivation by the
-\emph{adjunction} operation.  This makes the derivation syntactically
-and semantically complete, and we are done.
+%\citeauthor{KolSto07}~\shortcite{KolSto07} attempt to improve the
+%efficiency by translating the SPUD sentence generation problem into a
+%planning problem and using a planning algorithm for generation.  We will
+%now illustrate how they do this by a (simplified) example.
 
-\begin{figure}
-  \centering
-  xxxx
-  \caption{Generating ``The white rabbit sleeps'' as a planning problem.}
-  \label{fig:white-rabbit-as-planning}
-\end{figure}
+To improve on the efficiency of SPUD,
+\citeauthor{KolSto07}~\shortcite{KolSto07} translate the sentence
+generation problem into a planning problem and use a planning algorithm for
+generation. We illustrate this process on a simplified example.
+
+Consider a knowledge base containing the individuals $r_1$ and $r_2$, and a
+set of attributes encoding the fact that $r_1$ and $r_2$ are rabbits, $r_1$
+is white and $r_2$ is brown, and $r_1$ sleeps.  Say that we want to express
+the information $\{\mathsf{sleep}(r_1)\}$ using the tree-adjoining grammar
+shown in Fig.~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists
+of \emph{elementary trees} (i.e., the disjoint trees in the figure), each
+of which contributes certain \emph{semantic content}. We can instantiate
+these trees by substituting individuals for \emph{semantic roles}, such as
+$\mathsf{self}$ and $\mathsf{subj}$, and then combine the tree instances as
+shown in Fig.~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
+``The white rabbit sleeps''. The SPUD algorithm computes this grammatical
+derivation by starting with the elementary tree for ``sleeps''. This tree
+satisfies the need to convey the semantic information, but introduces a
+need to generate a noun phrase (NP) for the subject, which also refers
+uniquely to the target referent $r_1$. In a second step, SPUD substitutes
+the tree for ``the rabbit'' into the open NP leaf, which makes the
+derivation grammatically complete. Since there are two different
+individuals that could be described as ``the rabbit''---technically, $r_2$
+is still a \emph{distractor} (i.e., a \todo{one line definition here}) for
+the referring expression---we are still not finished. To complete the
+derivation, the tree for ``white'' is added to the existing structure by
+the \emph{adjunction} operation, making the derivation syntactically and
+semantically complete.
 
@@
 \caption{PDDL actions for generating the sentence ``The white rabbit
-sleeps''.}
-\label{fig:pddl-crisp}
+sleeps.''}
+%\caption{Generating ``The white rabbit sleeps'' as a planning problem.}
+\label{fig:white-rabbit-as-planning}
 \end{figure}
 
-This process has clear parallels to planning: We manipulate a state by
-applying actions in order to achieve a goal.  We can make this
-connection precise by translating the SPUD problem into a planning
-problem; the relevant excerpt of this problem is shown in
-Fig.~\ref{fig:white-rabbit-as-planning}.  Each action in this planning
-problem corresponds to the addition of a single elementary tree to the
-derivation; the first parameter of the action is a node name in the
-derivation tree, and the further parameters stand for the individuals
-to which the semantic roles will be instantiated.  The syntactic
-preconditions and effects are encoded using $\mathsf{open}$ and
-$\mathsf{canadjoin}$ literals; the status of each referring expression
-is tracked using $\mathsf{distractor}$ literals.  Notice that the
-action effects contain terms of the form $\mathsf{subj}(u)$, which
-construct new node names.  In order to represent the planning problem
-in PDDL, these terms can be eliminated by estimating an upper bound
-$n$ for the plan length, making $n$ copies of each action, ensuring
-that copy $i$ can only be applied in step $i$, and replacing the term
-$\mathsf{subj}(u)$ in an action copy by the constant
-$\mathsf{subj}_i$.
+The process described above has clear parallels to planning: we manipulate
+a state by applying actions in order to achieve a goal. We can make this
+connection even more precise by translating the SPUD problem into a
+planning problem. For instance, Fig.~\ref{fig:white-rabbit-as-planning}
+shows the corresponding PDDL actions for the above generation task, where
+each action corresponds to an operation that adds a single elementary tree
+to the derivation.  In each case, the first parameter of the action is a
+node name in the derivation tree, and the remaining parameters stand for
+the individuals to which the semantic roles will be instantiated.  The
+syntactic preconditions and effects are encoded using $\mathsf{open}$ and
+$\mathsf{canadjoin}$ literals; the status of each referring expression is
+tracked using $\mathsf{distractor}$ literals. Notice that the action
+effects contain terms of the form $\mathsf{subj}(u)$, which construct new
+node names. In order to meet the syntactic requirements of PDDL, these
+terms can be eliminated by estimating an upper bound $n$ for the plan
+length, making $n$ copies of each action, ensuring that copy $i$ can only
+be applied in step $i$, and replacing the term $\mathsf{subj}(u)$ in an
+action copy by the constant $\mathsf{subj}_i$.
 
-In the example, the following plan solves the planning problem:
+Once we've defined appropriate actions, we can solve the generation
+problem as an ordinary planning problem. For instance, the following plan
+solves our previous example:
+%
 \begin{enumerate}
@@
 \end{enumerate}
-
-The grammatical derivation in
+%
+Using this plan, the grammatical derivation in
 Fig.~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the generated
-sentence, can be systematically reconstructed from this plan.
-Therefore we can solve the sentence generation problem via the detour
-through planning and bring current search heuristics for planning to
-bear on generation.
+sentence, can be systematically reconstructed. Thus, we can solve the
+sentence generation problem via the detour through planning and bring
+current search heuristics for planning to bear on generation.
 

### 2008-06-23 trunk/Papers/experiences/experiments.tex
changes to experiments section (experiment 1)

--- 2008-06-05 trunk/Papers/experiences/discussion.tex
@@
-\section{Discussion} \label{sec:discussion}
+\section{Experiments} \label{sec:experiments}
 
-planning competition makes for great starting point for users
+We now describe the results of a set of experiments designed to evaluate
+the suitability of particular planners (FF, SGPLAN, and an ad-hoc
+implementation of GraphPlan) in our two NLG domains. As a first impression,
+our results indicate that planning is a promising tool for both domains. In
+the GIVE domain, SGPLAN 5.2.2 computes a domain plan from the initial state
+to the goal in 0.3 seconds, which is fast enough for moderately-sized
+problems instances in the application.\footnote{All runtimes were
+  measured on a Pentium 4 CPU running at 3 GHz. Java programs were allowed
+  to ``warm up'', i.e.\ the planner was run five times and the first four
+  measurements discarded to ensure that the JVM had just-in-time compiled
+  all relevant bytecode.}
+In the sentence generation domain, FF dramatically outperforms the best
+previously known algorithm for the same problem (a reimplementation of
+\cite{Stone2003a}), although the latter is a greedy search algorithm with a
+heuristic that is hand-tailored to the domain.
+
+We also observe that although FF manages the search for a plan very
+efficiently, it spends comparatively large amounts of time computing
+instantiations of predicates and actions, most of which are then never used
+during the search. As a result, FF's ``grounding time'' dominates its
+overall planning time, leading to some conflicting results. Our experiments
+below seek to improve our understanding of this situation.
+
+\subsection{Experiment 1: Sentence generation}
+\label{sec:exper-1:-sent}
+
+In the first experiment, we generate a series of sentence generation
+problems which require the planner to compute a plan representing the
+sentence ``Mary likes the Adj$_1$ \ldots Adj$_n$ rabbit.''  Each problem
+instance assumes a certain number $m$ of rabbits that were distinguished by
+$n \leq m$ different properties, such that all $n$ properties are required
+to distinguish the target referent from all other rabbits.  The $n$
+properties are realized as $n$ different adjectives, in any order.  This
+setup allows us to control the plan length (a plan with $n$ properties will
+have length $n+4$) and the universe size (the universe will contain $m+1$
+individuals in addition to the differently-typed individuals used to encode
+the grammar).
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1\columnwidth]{pic-runtime-modifiers-with-sgplan}
+  \caption{Runtimes in the sentence generation domain. The horizontal
+  axis represents parameters $(m,n)$ from $(1,1)$ to $(10,10)$ in
+  lexicographical order.}
+  \label{fig:runtimes-crisp}
+\end{figure}
+
+The results of this experiment are shown in
+Fig.~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
+(in lexicographic order) on the horizontal axis and the runtime is shown in
+milliseconds on the vertical axis. These results reveal a number of
+interesting insights. First, FF significantly outperforms
+SGPLAN in this domain.\footnote{Experiments with SGPLAN use a
+ pre-release version of SGPLAN, kindly provided by Chih-Wei Hsu. The
+ release version of SGPLAN 5.2.2 had a bug, causing it to crash on some
+ instances.}
+Second, FF's runtime is dominated by its initial grounding step, in which
+it computes the ground instances of all predicates and actions used in the
+planning problem to avoid unnecessary instantiations during search. In
+particular, the ratio of grounding time to total runtime is generally above
+85\%, and rises to above 99\% at $m=11$, which is still a small universe in
+this application.\footnote{The
+  ``grounding'' time reported here is what FF reports as ``time spent:
+  instantiating action templates''.}  This is also reflected in the
+much higher sensitivity of FF to the domain size: if we fix $n=1$, FF takes
+60 ms to compute a plan at $m=1$, but 2.4 seconds at $m=10$. By contrast,
+our naive GraphPlan implementation takes 30 ms at $m=1$ and still only 50
+ms at $m=10$.
+
+Our results illustrate that FF is outperformed quite clearly by an ad-hoc
+Java implementation of GraphPlan which only computes instances of
+predicates and actions as they are discovered, while the planning graph is
+being built. On the other hand, the runtime that FF actually spends on
+search is again considerably lower than that of our GraphPlan
+implementation, and for any given $m$, GraphPlan's runtime grows much
+faster with $n$ (i.e., the plan size) than FF's. As a result, for
+moderately-sized instances of the sentence planning domain FF does very
+well on search, but spends about ten times as much runtime again on
+grounding. Larger (but still realistically-sized) instances of the sentence
+generation problem are still problematic for the planners we tested.
+
+\subsection{Experiment 2: Minimal GIVE worlds}
+\label{sec:exper-2:-minim}
+
+To evaluate the performance of the planners on the planning problems
+that arise in the GIVE domain, we construct a sequence of test worlds
+as illustrated in Fig.~\ref{fig:give-minimal}.  These worlds consist
+of a $2n$ by $h$ grid of positions, such that there are buttons at
+positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$.  The player
+starts in position $(1,1)$ and must press all buttons in order to
+complete the game.  The world is generated as a GIVE world
+description, and then converted into a planning problem by the GIVE
+software. \todo{at what level of detail should we describe this?}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.8\columnwidth]{pic-buttons}
+  \caption{Minimal GIVE world.}
+  \label{fig:give-minimal}
+\end{figure}
+
+Some results from this experiment (for $h=20$, with $n$ ranging from
+$1$ to $40$) are shown in Fig.~\ref{fig:give-runtime-minimal}.  The
+most obvious result is that FF is unable to solve any problems beyond
+$n=13$ on the experimentation machine within the memory limit of 1 GB.
+SGPLAN, on the other hand, solves instances beyond $n=40$ without
+major problems.  The time spent on grounding is not a major factor in
+either planner, probably because the planners need more time to
+actually compute the plan -- at $n=40$, the optimal plan has a length
+of about 1600!
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1\columnwidth]{pic-runtime-buttons}
+  \caption{Runtimes of FF and SGPLAN on the minimal GIVE worlds for $h=20$. The horizontal axis is $n$, ranging from 1 to 40.}
+  \label{fig:give-runtime-minimal}
+\end{figure}
+
+\subsection{Experiment 3: GIVE worlds with extra positions}
+\label{sec:experiment-3:-give}
+
+Finally, we varied the GIVE world in order to judge the effect of the
+universe size in the GIVE domain (see Fig.~\ref{fig:give-junk}).  We
+added another $w$ by $h$ positions to the right of a minimal world as
+in Experiment 2.  Other than that, we left the initial state and goal
+untouched, i.e.\ the new positions were not needed in any plan; this
+approximates the situation in the actual GIVE domain, where most grid
+positions never need to be used.  As before, we generate a GIVE world
+description and then convert it into a planning problem.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1\columnwidth]{pic-empty-buttons}
+  \caption{GIVE world with extra positions.}
+  \label{fig:give-junk}
+\end{figure}
+
+Some results for $h=20$ and $n=5$ (with $w$ ranging from $1$ to $70$)
+are shown in Fig.~\ref{fig:give-runtime-junk}.  Again, FF runs out of
+memory, this time at $w=17$, whereas SGPLAN happily solved inputs
+beyond $w=70$.  However, unlike in Experiment 2, both planners now
+spend a substantial propertion of their time on grounding again.  In
+SGPLAN, this translates to a ``parsing time'' (which
+\todo{presumably} includes grounding) which grows from 180 ms to 21.7
+seconds as $w$ grows from $1$ to $75$; the rest of the runtime (which
+includes the search) only grows from 400 ms to 2.3 seconds.  This
+difference is particularly dramatic given that the actual optimal plan
+is an identical plan of about 100 steps for each $w$.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
+  \caption{Runtimes of FF and SGPLAN on the GIVE worlds with junk
+    positions for $h=20$ and $n=5$. The horizontal axis is $w$,
+    ranging from 1 to 70.}
+  \label{fig:give-runtime-junk}
+\end{figure}
 

### 2008-06-23 trunk/Papers/experiences/domains.tex
small textual changes

@@
 the information $\{\mathsf{sleep}(r_1)\}$ using the tree-adjoining grammar
-shown in Fig.~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists
+shown in Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists
 of \emph{elementary trees} (i.e., the disjoint trees in the figure), each
@@
 $\mathsf{self}$ and $\mathsf{subj}$, and then combine the tree instances as
-shown in Fig.~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
+shown in Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
 ``The white rabbit sleeps''. The SPUD algorithm computes this grammatical
@@
 connection even more precise by translating the SPUD problem into a
-planning problem. For instance, Fig.~\ref{fig:white-rabbit-as-planning}
+planning problem. For instance, Figure~\ref{fig:white-rabbit-as-planning}
 shows the corresponding PDDL actions for the above generation task, where
@@
 Using this plan, the grammatical derivation in
-Fig.~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the generated
+Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the generated
 sentence, can be systematically reconstructed. Thus, we can solve the
@@
 A map of an example GIVE world is shown in
-Fig.~\ref{fig:give-development-world}.  In this world, the user's task is
+Figure~\ref{fig:give-development-world}.  In this world, the user's task is
 to pick up a trophy in the top left room.  The trophy is hidden in a safe
@@
 degree steps in either direction, and can move from the centre of one tile
-to the centre of the next. Some of the actions the user can take are shown
-in Fig.~\ref{fig:give-planning}, in PDDL syntax.
+to the centre of the next. Figure~\ref{fig:give-planning} shows some of the
+available actions in PDDL syntax.
 

### 2008-06-23 trunk/Papers/experiences/experiments.tex
small textual changes

@@
 The results of this experiment are shown in
-Fig.~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
+Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
 (in lexicographic order) on the horizontal axis and the runtime is shown in
@@
 
-To evaluate the performance of the planners on the planning problems
-that arise in the GIVE domain, we construct a sequence of test worlds
-as illustrated in Fig.~\ref{fig:give-minimal}.  These worlds consist
-of a $2n$ by $h$ grid of positions, such that there are buttons at
-positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$.  The player
-starts in position $(1,1)$ and must press all buttons in order to
-complete the game.  The world is generated as a GIVE world
-description, and then converted into a planning problem by the GIVE
-software. \todo{at what level of detail should we describe this?}
+In the second experiment, we evaluate the performance of the planners on
+problems arising in the GIVE domain. We construct a series of test worlds,
+similar to the one illustrated in Figure~\ref{fig:give-minimal}. These worlds
+consist of a $2n$ by $h$ grid of positions, such that there are buttons at
+positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The player starts
+in position $(1,1)$ and must press all buttons in order to complete the
+game. The world is generated as a GIVE world description, and then
+automatically converted into a planning problem by the GIVE software.  (We
+refer the reader to \todo{cite?} for more details about this process.)
 
@@
 Some results from this experiment (for $h=20$, with $n$ ranging from
-$1$ to $40$) are shown in Fig.~\ref{fig:give-runtime-minimal}.  The
+$1$ to $40$) are shown in Figure~\ref{fig:give-runtime-minimal}.  The
 most obvious result is that FF is unable to solve any problems beyond
@@
 Finally, we varied the GIVE world in order to judge the effect of the
-universe size in the GIVE domain (see Fig.~\ref{fig:give-junk}).  We
+universe size in the GIVE domain (see Figure~\ref{fig:give-junk}).  We
 added another $w$ by $h$ positions to the right of a minimal world as
@@
 Some results for $h=20$ and $n=5$ (with $w$ ranging from $1$ to $70$)
-are shown in Fig.~\ref{fig:give-runtime-junk}.  Again, FF runs out of
+are shown in Figure~\ref{fig:give-runtime-junk}.  Again, FF runs out of
 memory, this time at $w=17$, whereas SGPLAN happily solved inputs

### 2008-06-23 trunk/Papers/experiences/discussion.tex
changes to experiments (GIVE subsection)

@@
 
-The positive conclusion we can draw from the experiments reported above is
-that modern planners are fairly good at controlling the search for many of
-the moderately-sized NLG problems we looked at. This is particularly true
-for FF in the sentence generation domain, which generates nontrivial
-14-word sentences in about two seconds. There is still room for
+The positive conclusion we can draw from the simple experiments reported
+above is that modern planners are fairly good at controlling the search for
+many of the moderately-sized NLG problems we looked at. This is
+particularly true for FF in the sentence generation domain, which generates
+nontrivial 14-word sentences in about two seconds. There is still room for
 improvement, however. For instance, by comparison a special-purpose

### 2008-06-23 trunk/Papers/experiences/experiments.tex
changes to experiments (GIVE subsection)

@@
 
-We now describe the results of a set of experiments designed to evaluate
-the suitability of particular planners (FF, SGPLAN, and an ad-hoc
+We now describe the results of three simple experiments designed to
+evaluate the suitability of particular planners (FF, SGPLAN, and an ad-hoc
 implementation of GraphPlan) in our two NLG domains. As a first impression,
@@
 problems arising in the GIVE domain. We construct a series of test worlds,
-similar to the one illustrated in Figure~\ref{fig:give-minimal}. These worlds
-consist of a $2n$ by $h$ grid of positions, such that there are buttons at
-positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The player starts
-in position $(1,1)$ and must press all buttons in order to complete the
-game. The world is generated as a GIVE world description, and then
-automatically converted into a planning problem by the GIVE software.  (We
-refer the reader to \todo{cite?} for more details about this process.)
+similar to the one illustrated in Figure~\ref{fig:give-minimal}. These
+worlds consist of a $2n$ by $h$ grid of positions, such that there are
+buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The
+player starts in position $(1,1)$ and must press all buttons in order to
+successfully complete the game. The world is generated as a GIVE world
+description, and then automatically converted into a planning problem by
+the GIVE software. (We refer the reader to \todo{add citation} for more
+details about this process.)
 
@@
 
-Some results from this experiment (for $h=20$, with $n$ ranging from
-$1$ to $40$) are shown in Figure~\ref{fig:give-runtime-minimal}.  The
-most obvious result is that FF is unable to solve any problems beyond
-$n=13$ on the experimentation machine within the memory limit of 1 GB.
-SGPLAN, on the other hand, solves instances beyond $n=40$ without
-major problems.  The time spent on grounding is not a major factor in
-either planner, probably because the planners need more time to
-actually compute the plan -- at $n=40$, the optimal plan has a length
-of about 1600!
+Results for the $h=20$ case, with $n$ ranging from $1$ to $40$, are shown
+in Figure~\ref{fig:give-runtime-minimal}.  The most obvious result is that
+FF is unable to solve any problems beyond $n=13$ on our experimentation
+machine within the memory limit of 1 GB.  SGPLAN, on the other hand, solves
+instances beyond $n=40$ without major problems.  The time spent on
+grounding is not a major factor in either planner, probably because the
+planners need more time to actually compute the plan---for instance, the
+optimal plan for the problem $n=40$ has a length of about 1600 steps.
 
@@
 
-Finally, we varied the GIVE world in order to judge the effect of the
-universe size in the GIVE domain (see Figure~\ref{fig:give-junk}).  We
-added another $w$ by $h$ positions to the right of a minimal world as
-in Experiment 2.  Other than that, we left the initial state and goal
-untouched, i.e.\ the new positions were not needed in any plan; this
-approximates the situation in the actual GIVE domain, where most grid
-positions never need to be used.  As before, we generate a GIVE world
+In the final experiment, we vary the structure of the GIVE world in order
+to judge the effect that universe size has on the planning problem in this
+domain.  Starting with the ordinary GIVE world described in Experiment~2,
+we add another $w$ by $h$ empty ``junk'' positions to the right of the
+minimal world (see Figure~\ref{fig:give-junk}). These new positions are not
+actually needed in any plan, but approximate the situation in the actual
+GIVE domain, where most grid positions never need to be used. We leave the
+initial state and goal untouched. As before, we generate a GIVE world
 description and then convert it into a planning problem.
@@
   \includegraphics[width=1\columnwidth]{pic-empty-buttons}
-  \caption{GIVE world with extra positions.}
+  \caption{GIVE world with extra ``junk'' positions.}
   \label{fig:give-junk}
@@
 
-Some results for $h=20$ and $n=5$ (with $w$ ranging from $1$ to $70$)
-are shown in Figure~\ref{fig:give-runtime-junk}.  Again, FF runs out of
-memory, this time at $w=17$, whereas SGPLAN happily solved inputs
-beyond $w=70$.  However, unlike in Experiment 2, both planners now
-spend a substantial propertion of their time on grounding again.  In
-SGPLAN, this translates to a ``parsing time'' (which
-\todo{presumably} includes grounding) which grows from 180 ms to 21.7
-seconds as $w$ grows from $1$ to $75$; the rest of the runtime (which
-includes the search) only grows from 400 ms to 2.3 seconds.  This
-difference is particularly dramatic given that the actual optimal plan
-is an identical plan of about 100 steps for each $w$.
+Results for the $h=20$, $n=5$ case (with $w$ ranging from $1$ to $70$) are
+shown in Figure~\ref{fig:give-runtime-junk}. As in Experiment~2, FF again
+runs out of memory, this time at $w=17$. SGPLAN happily solves inputs
+beyond $w=70$. However, unlike in Experiment 2, both planners now spend a
+substantial propertion of their time on grounding. In SGPLAN, this
+translates to a ``parsing time'' (which \todo{presumably} includes
+grounding) which grows from 180 ms to 21.7 seconds as $w$ grows from $1$ to
+$75$. The rest of the runtime (which also includes the search time) only
+grows from 400 ms to 2.3 seconds. This difference is particularly dramatic
+given that the actual optimal plan in each case is an identical plan of
+about 100 steps. The overall runtimes in these instances is also concerning
+since runtimes over a couple seconds negatively affect the response time of
+the GIVE system.
 

### 2008-06-23 trunk/Papers/experiences/experiments.tex
changes to experiments (GIVE subsection)

@@
 given that the actual optimal plan in each case is an identical plan of
-about 100 steps. The overall runtimes in these instances is also concerning
-since runtimes over a couple seconds negatively affect the response time of
-the GIVE system.
+about 100 steps. The planning times for these instances are also concerning
+since times over a couple seconds negatively affect the overall response
+time of the system, which must react in real time to user actions.
 

### 2008-06-23 trunk/Papers/experiences/conclusion.tex
small change to conclusions

@@
 researchers: provided some of the challenges we have highlighted can be
-addressed, projects like GIVE offer a constructive platform for the
-planning community to showcase their techniques to a wider audience---and
-to improve the quality of their tools for real-world planning tasks.
+addressed, projects like GIVE (and other NLG-inspired problem domains)
+offer a constructive platform for the planning community to showcase their
+techniques to a wider audience---and to improve the quality of their tools
+for real-world planning tasks.
 

### 2008-06-23 trunk/Papers/experiences/experiences.tex
small change

@@
 \begin{abstract}
-  lkajsd
+We investigate the application of modern planning techniques to domains
+arising from problems in natural language generation (NLG). In particular,
+we consider two novel NLG-inspired planning problems, the sentence
+generation domain and the GIVE (``Generating Instructions in Virtual
+Environment'') domain, and investigate the efficiency of FF and SGPLAN in
+these domains. We also compare our results against an ad-hoc implementation
+of GraphPlan in Java. Our results are mixed. While modern planners are able
+to quickly solve many moderately-sized instances of the NLG problems, the
+overall planning time is dominated by grounding, rather than search, which
+has a pronounced effect on our domains which require relatively short plans
+in large universes.
+
+We offer these domains as challenges for the planning community.
 \end{abstract}

### 2008-06-23 trunk/Papers/experiences/experiences.tex
added abstract

@@
 of GraphPlan in Java. Our results are mixed. While modern planners are able
-to quickly solve many moderately-sized instances of the NLG problems, the
-overall planning time is dominated by grounding, rather than search, which
-has a pronounced effect on our domains which require relatively short plans
-in large universes.
-
-We offer these domains as challenges for the planning community.
+to quickly solve many moderately-sized instances of our problems, the
+overall planning time is dominated by the grounding step that these
+planners perform (rather than search). This has a pronounced effect on our
+domains which require relatively short plans in large universes.  We share
+our experiences and offer these domains as challenges for the planning
+community.
 \end{abstract}

### 2008-06-23 trunk/Papers/experiences/discussion.tex
spellcheck

@@
 grounding pays off in saved instantiations during the search---this
-approach is less efective in the domains we've considered. As our
-experiements have shown, there are natural planning domains in which
+approach is less effective in the domains we've considered. As our
+experiments have shown, there are natural planning domains in which
 relatively short plans must be computed in large universes. For instance,
@@
 The recent trend in planning research has (rightfully) focused on the
-development of alorithms that control search in sophisticated ways,
+development of algorithms that control search in sophisticated ways,
 resulting in a host of planners that are more powerful and more successful

### 2008-06-23 trunk/Papers/experiences/experiments.tex
spellcheck
better graphs

@@
   \includegraphics[width=1\columnwidth]{pic-runtime-modifiers-with-sgplan}
-  \caption{Runtimes in the sentence generation domain. The horizontal
-  axis represents parameters $(m,n)$ from $(1,1)$ to $(10,10)$ in
-  lexicographical order.}
+  \caption{Runtimes (in ms) in the sentence generation domain.  The
+    horizontal axis represents parameters $(m,n)$ from $(1,1)$ to
+    $(10,10)$ in lexicographical order.}
   \label{fig:runtimes-crisp}
@@
   \includegraphics[width=1\columnwidth]{pic-runtime-buttons}
-  \caption{Runtimes of FF and SGPLAN on the minimal GIVE worlds for $h=20$. The horizontal axis is $n$, ranging from 1 to 40.}
+  \caption{Runtimes (in seconds) of FF and SGPLAN on the minimal GIVE
+    worlds for $h=20$. The horizontal axis is $n$.}
   \label{fig:give-runtime-minimal}
@@
 beyond $w=70$. However, unlike in Experiment 2, both planners now spend a
-substantial propertion of their time on grounding. In SGPLAN, this
+substantial proportion of their time on grounding. In SGPLAN, this
 translates to a ``parsing time'' (which \todo{presumably} includes
@@
   \includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
-  \caption{Runtimes of FF and SGPLAN on the GIVE worlds with junk
-    positions for $h=20$ and $n=5$. The horizontal axis is $w$,
-    ranging from 1 to 70.}
+  \caption{Runtimes (in seconds) of FF and SGPLAN on the GIVE worlds with junk
+    positions for $h=20$ and $n=5$. The horizontal axis is $w$.}
   \label{fig:give-runtime-junk}

### 2008-06-23 trunk/Papers/experiences/domains.tex
small changes

@@
 communicate certain intentions to the system, both through action and
-inaction. For instance, the system may correctly infer that a user has
-failed to follow instructions if the user exits a room when given a
-directive to ``walk to the centre of the room''. However, the system should
-also be able to make a similar conclusion if a user simply does nothing
-when given the instruction. Thus, such real-time plan monitoring is
-essential in the GIVE domain, and also affects other modules in the system.
+inaction. For instance, the system should infer that a user has failed to
+follow instructions if the user exits a room when given a directive to
+``walk to the centre of the room''. However, the system should also make a
+similar conclusion if a user simply does nothing when given the
+instruction. Such real-time plan execution monitoring is essential in this
+domain.
+
+%However, the system should also be able to make a similar conclusion if a
+%user simply does nothing when given the instruction. Thus, such real-time
+%plan monitoring is essential in the GIVE domain, and also affects other
+%modules in the system.
 

### 2008-06-23 trunk/Papers/experiences/domains.tex
I'm essentially happy now

@@
 
-In practice, the process of determining referring expressions typically
-interacts with the realization step and so it turns out to be beneficial to
-perform both of these steps together. This was the goal of the SPUD system
-\cite{Stone2003a}, which performed both steps together using a top-down
-generation algorithm based on tree-adjoining grammars \cite{joshi;etal1997}
-whose lexical entries were equipped with semantic and pragmatic
-information. Unfortunately, SPUD suffered from having to explore a huge
-search space, and had to resort to a non-optimal greedy search strategy to
-retain reasonable efficiency.
-
-%However, determining referring expressions and realization
-%interact, so it would be beneficial to perform both steps together.
-%This was the goal of the SPUD system \cite{Stone2003a}, which
-%performed both steps together using a top-down generation algorithm
-%based on tree-adjoining grammars \cite{joshi;etal1997} whose lexical
-%entries were equipped with semantic and pragmatic information.
-%Unfortunately, SPUD suffered from having to explore a huge search
-%space, and had to resort to a non-optimal greedy search strategy to
-%retain reasonable efficiency.
+In practice, the process of determining referring expressions
+typically interacts with the realization step and so it turns out to
+be beneficial to perform both of these steps together. This was the
+goal of the SPUD system \cite{Stone2003a}, which did this using a
+top-down generation algorithm based on tree-adjoining grammars
+\cite{joshi;etal1997} whose lexical entries were equipped with
+semantic and pragmatic information. Unfortunately, SPUD suffered from
+having to explore a huge search space, and had to resort to a
+non-optimal greedy search strategy to retain reasonable efficiency.
+To improve on the efficiency of SPUD,
+\citeauthor{KolSto07}~\shortcite{KolSto07} translate the sentence
+generation problem into a planning problem and use a planning
+algorithm for generation.
 
@@
 
-%\citeauthor{KolSto07}~\shortcite{KolSto07} attempt to improve the
-%efficiency by translating the SPUD sentence generation problem into a
-%planning problem and using a planning algorithm for generation.  We will
-%now illustrate how they do this by a (simplified) example.
+We illustrate this process on a simplified example.  Consider a
+knowledge base containing the individuals $e$, $r_1$ and $r_2$, and a
+set of attributes encoding the fact that $r_1$ and $r_2$ are rabbits,
+$r_1$ is white and $r_2$ is brown, and $e$ is an event in which $r_1$
+sleeps.  Say that we want to express the information
+$\{\mathsf{sleep}(e,r_1)\}$ using the tree-adjoining grammar shown in
+Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
+\emph{elementary trees} (i.e., the disjoint trees in the figure), each
+of which contributes certain \emph{semantic content}. We can
+instantiate these trees by substituting individuals for \emph{semantic
+  roles}, such as $\mathsf{self}$ and $\mathsf{subj}$, and then
+combine the tree instances as shown in
+Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
+``The white rabbit sleeps''.
 
-To improve on the efficiency of SPUD,
-\citeauthor{KolSto07}~\shortcite{KolSto07} translate the sentence
-generation problem into a planning problem and use a planning algorithm for
-generation. We illustrate this process on a simplified example.
-
-Consider a knowledge base containing the individuals $r_1$ and $r_2$, and a
-set of attributes encoding the fact that $r_1$ and $r_2$ are rabbits, $r_1$
-is white and $r_2$ is brown, and $r_1$ sleeps.  Say that we want to express
-the information $\{\mathsf{sleep}(r_1)\}$ using the tree-adjoining grammar
-shown in Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists
-of \emph{elementary trees} (i.e., the disjoint trees in the figure), each
-of which contributes certain \emph{semantic content}. We can instantiate
-these trees by substituting individuals for \emph{semantic roles}, such as
-$\mathsf{self}$ and $\mathsf{subj}$, and then combine the tree instances as
-shown in Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
-``The white rabbit sleeps''. The SPUD algorithm computes this grammatical
-derivation by starting with the elementary tree for ``sleeps''. This tree
-satisfies the need to convey the semantic information, but introduces a
-need to generate a noun phrase (NP) for the subject, which also refers
-uniquely to the target referent $r_1$. In a second step, SPUD substitutes
-the tree for ``the rabbit'' into the open NP leaf, which makes the
-derivation grammatically complete. Since there are two different
-individuals that could be described as ``the rabbit''---technically, $r_2$
-is still a \emph{distractor} (i.e., a \todo{one line definition here}) for
-the referring expression---we are still not finished. To complete the
-derivation, the tree for ``white'' is added to the existing structure by
-the \emph{adjunction} operation, making the derivation syntactically and
+We compute this grammatical derivation top-down by starting with the
+elementary tree for ``sleeps''. This tree satisfies the need to convey
+the semantic information, but introduces a need to generate a noun
+phrase (NP) for the subject; this NP must refer uniquely to the target
+referent $r_1$. In a second step, we substitute the tree for ``the
+rabbit'' into the open NP leaf, which makes the derivation
+grammatically complete. Since there are two different individuals that
+could be described as ``the rabbit''---technically, $r_2$ is still a
+\emph{distractor} (i.e., based on the description ``the rabbit'', the
+hearer might erroneously think that we're talking about $r_2$ and not
+$r_1$)---we are still not finished. To complete the derivation, the
+tree for ``white'' is added to the existing structure by the
+\emph{adjunction} operation, making the derivation syntactically and
 semantically complete.
@@
 A map of an example GIVE world is shown in
-Figure~\ref{fig:give-development-world}.  In this world, the user's task is
-to pick up a trophy in the top left room.  The trophy is hidden in a safe
-behind a picture, so the user must first move the picture out of the way
-and open the safe by pushing a sequence of buttons (the small square boxes
-on the walls) in the correct order.  In order to get to all these buttons,
-the user must also open the lower door and deactivate the alarm tile by
-pushing further buttons.  To simplify both the planning and the NLG task,
-the world is discretized into tiles of equal size; the user can turn by 90
-degree steps in either direction, and can move from the centre of one tile
-to the centre of the next. Figure~\ref{fig:give-planning} shows some of the
-available actions in PDDL syntax.
+Figure~\ref{fig:give-development-world}.  In this world, the user's
+task is to pick up a trophy in the top left room.  The trophy is
+hidden in a safe behind a picture, so the user must first move the
+picture out of the way and open the safe by pushing a sequence of
+buttons (the small square boxes on the walls) in the correct order.
+In order to get to all these buttons, the user must also open the door
+in the centre of the map and deactivate the alarm tile by pushing
+further buttons.  To simplify both the planning and the NLG task, the
+world is discretized into tiles of equal size; the user can turn by 90
+degree steps in either direction, and can move from the centre of one
+tile to the centre of the next. Figure~\ref{fig:give-planning} shows
+some of the available actions in PDDL syntax.
 
@@
 \end{enumerate}
-%
-(Most of the actions in this plan are ``move'' actions.) The bulk of the
-GIVE problem can be considered a variant of the Gridworld problem, which
-also involves finding a route through a world with discrete positions---but
-with the additional need to press buttons in the right order, reason about
-many more objects in the world, and navigate somewhat more complicated room
-shapes.
 
-The task of generating natural language instructions in this domain is also
-more complicated than just a straightforward translation of the action
-steps in a plan. For instance, the NLG system could generate an instruction
-sequence starting with ``turn left; walk forward; press the button; turn
-right; walk forward; walk forward; turn right; walk forward; walk forward;
-turn left; walk forward''.  However, this would be clumsy and boring, and
-it is to be expected that frustrated users will quickly cancel an
-interaction with such a system, resulting in a bad evaluation score.
-Having just pressed the button, it would be much better to say ``turn
-around and walk through the door''.  This means that the mapping from
-action instances to natural language instructions is not trivial.
-Conversely, it may also be necessary to express a single planning step with
-several instructions.  Having just entered the top left room, it may be
-easier for the user to understand the instruction ``walk to the centre of
-the room; turn right; now press the green button in front of you'' than the
-instruction ``press the green button on the wall to your right''.  To plan
-the referring expression ``the green button in front of you'' at a time
-when the button is not yet in front of the user, the NLG system must keep
-track of the hypothetical changes in what objects are visible to the user.
-Thus the acts of referring and instructing, and the modules for planning
-and NLG, must be tightly integrated.
+Most of the actions in this plan are ``move'' actions.  This makes the
+bulk of the GIVE problem very similar to the Gridworld problem, which
+also involves finding a route through a world with discrete
+positions---but with the additional need to press buttons in the right
+order, reason about many more objects in the world, and navigate
+somewhat more complicated room shapes.
 
-Plan execution monitoring also plays an important role in the GIVE problem.
-At a high level, the system needs to monitor a user's actions and compare
-them against the generated instruction set to determine if the user has
-correctly followed directions or not. In the case of the latter, new
-instructions may have to be generated. In practice, the situation
-can be quite complicated since the mental state of the user is not known
-and so the system must observe the user's actions in real time. For
-example, a user directed to ``turn around and walk through the door'' may
-not necessarily perform these actions to the letter, i.e., immediately
-turning 180 degrees and proceeding directly to the door. Instead, the user
-might take a roundabout route through the room, eventually exiting out the
-door. Although the user's actions do not match the generated instructions
-exactly, they meet the intended goal. The system must be able to identify
-such ``equivalent'' plans and not immediately generate new instructions as
-soon as the user's actions have gone off course. Furthermore, a user can
-communicate certain intentions to the system, both through action and
-inaction. For instance, the system should infer that a user has failed to
-follow instructions if the user exits a room when given a directive to
-``walk to the centre of the room''. However, the system should also make a
-similar conclusion if a user simply does nothing when given the
-instruction. Such real-time plan execution monitoring is essential in this
-domain.
+The GIVE task comes with a number of interesting research challenges
+both for the NLG and the planning communities.  On the NLG side, one
+major problem is that the natural language instruction sequences can't
+simply be obtained by verbalising each action instance in the plan in
+turn.  For instance, the NLG system could generate an instruction
+sequence starting with ``turn left; walk forward; press the button;
+turn right; walk forward; walk forward; turn right; walk forward; walk
+forward; turn left; walk forward''.  However, this would be clumsy and
+boring, and it is to be expected that frustrated users will quickly
+cancel an interaction with such a system, resulting in a bad
+evaluation score.  Having just pressed the button, it would be much
+better to say ``turn around and walk through the door''; i.e.,
+multiple action instances should be merged into a single
+instruction. Conversely, it may also be necessary to express a single
+planning step with several instructions.  Having just entered the top
+left room, it may be easier for the user to understand the instruction
+``walk to the centre of the room; turn right; now press the green
+button in front of you'' than the instruction ``press the green button
+on the wall to your right''.  To plan the referring expression ``the
+green button in front of you'' at a time when the button is not yet in
+front of the user, the NLG system must keep track of the hypothetical
+changes in what objects are visible to the user.  Thus the acts of
+referring and instructing, and the modules for planning and NLG, must
+be tightly integrated.
 
-%However, the system should also be able to make a similar conclusion if a
-%user simply does nothing when given the instruction. Thus, such real-time
-%plan monitoring is essential in the GIVE domain, and also affects other
-%modules in the system.
+From a planning perspective, the use of planning in GIVE imposes very
+strict runtime requirements on the planner: Planning must happen in
+real time, because if the system takes too much time between deciding
+to give an instruction and actually giving this instruction, the user
+may have walked or turned away, thus making the instruction invalid.
+Furthermore, plan execution monitoring also plays an important role in
+the GIVE problem.  At a high level, the system needs to monitor a
+user's actions and compare them against the generated instruction set
+to determine if the user has correctly followed directions or not. In
+the case of the latter, new instructions may have to be generated. In
+practice, the situation can be quite complicated since the mental
+state of the user is not known and so the system must observe the
+user's actions in real time. For example, a user directed to ``turn
+around and walk through the door'' may not necessarily perform these
+actions to the letter, i.e., immediately turning 180 degrees and
+proceeding directly to the door. Instead, the user might take a
+roundabout route through the room, eventually exiting out the
+door. Although the user's actions do not match the generated
+instructions exactly, they meet the intended goal. The system must be
+able to identify such ``equivalent'' plans and not immediately
+generate new instructions as soon as the user's actions have gone off
+course. Furthermore, a user can communicate certain intentions to the
+system, both through action and inaction. For instance, the system
+should infer that a user has failed to follow instructions if the user
+exits a room when given a directive to ``walk to the centre of the
+room''. However, the system should also make a similar conclusion if a
+user simply does nothing when given the instruction.
 

### 2008-06-23 trunk/Papers/experiences/experiments.tex
I'm essentially happy now

@@
   ``grounding'' time reported here is what FF reports as ``time spent:
-  instantiating action templates''.}  This is also reflected in the
-much higher sensitivity of FF to the domain size: if we fix $n=1$, FF takes
-60 ms to compute a plan at $m=1$, but 2.4 seconds at $m=10$. By contrast,
-our naive GraphPlan implementation takes 30 ms at $m=1$ and still only 50
-ms at $m=10$.
+  instantiating action templates''.}
 
-Our results illustrate that FF is outperformed quite clearly by an ad-hoc
-Java implementation of GraphPlan which only computes instances of
-predicates and actions as they are discovered, while the planning graph is
-being built. On the other hand, the runtime that FF actually spends on
-search is again considerably lower than that of our GraphPlan
-implementation, and for any given $m$, GraphPlan's runtime grows much
-faster with $n$ (i.e., the plan size) than FF's. As a result, for
-moderately-sized instances of the sentence planning domain FF does very
-well on search, but spends about ten times as much runtime again on
-grounding. Larger (but still realistically-sized) instances of the sentence
-generation problem are still problematic for the planners we tested.
+In fact, FF spends so much time on grounding that it is consistently
+outperformed by an ad-hoc Java implementation of GraphPlan which only
+computes instances of predicates and actions as they are discovered,
+while the planning graph is being built -- despite the fact that FF is
+consistently much faster as far as pure search time is concerned.
+Looking at the difference from a different angle, FF's performance is
+much more sensitive to the domain size: if we fix $n=1$, FF takes 60
+ms to compute a plan at $m=1$, but 2.4 seconds (for the same plan) at
+$m=10$; our GraphPlan implementation takes 30 ms at $m=1$ and still
+only 50 ms at $m=10$.  Conversely, GraphPlan's runtime grows much
+faster with the plan size (i.e., with growing values of $n$ for a
+fixed $m$).  Larger (but still realistically-sized) instances of the
+sentence generation problem are still problematic for the planners we
+tested.
 
@@
 
-In the second experiment, we evaluate the performance of the planners on
-problems arising in the GIVE domain. We construct a series of test worlds,
-similar to the one illustrated in Figure~\ref{fig:give-minimal}. These
-worlds consist of a $2n$ by $h$ grid of positions, such that there are
-buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The
-player starts in position $(1,1)$ and must press all buttons in order to
-successfully complete the game. The world is generated as a GIVE world
-description, and then automatically converted into a planning problem by
-the GIVE software. (We refer the reader to \todo{add citation} for more
-details about this process.)
+In the second experiment, we evaluate the performance of the planners
+on problems arising in the GIVE domain. We construct a series of test
+worlds, similar to the one illustrated in
+Figure~\ref{fig:give-minimal}. These worlds consist of a $2n$ by $h$
+grid of positions, such that there are buttons at positions $(2i-1,1)$
+and $(2i,h)$ for $1 \leq i \leq n$. The player starts in position
+$(1,1)$ and must press all buttons in order to successfully complete
+the game. The world is generated as a GIVE world description, and then
+automatically converted into a planning problem as in
+Fig.~\ref{fig:give-planning} by the GIVE software.
 
@@
 
-Results for the $h=20$, $n=5$ case (with $w$ ranging from $1$ to $70$) are
-shown in Figure~\ref{fig:give-runtime-junk}. As in Experiment~2, FF again
-runs out of memory, this time at $w=17$. SGPLAN happily solves inputs
-beyond $w=70$. However, unlike in Experiment 2, both planners now spend a
-substantial proportion of their time on grounding. In SGPLAN, this
-translates to a ``parsing time'' (which \todo{presumably} includes
-grounding) which grows from 180 ms to 21.7 seconds as $w$ grows from $1$ to
-$75$. The rest of the runtime (which also includes the search time) only
-grows from 400 ms to 2.3 seconds. This difference is particularly dramatic
-given that the actual optimal plan in each case is an identical plan of
-about 100 steps. The planning times for these instances are also concerning
-since times over a couple seconds negatively affect the overall response
-time of the system, which must react in real time to user actions.
+Results for the $h=20$, $n=5$ case (with $w$ ranging from $1$ to $70$)
+are shown in Figure~\ref{fig:give-runtime-junk}. As in Experiment~2,
+FF again runs out of memory, this time at $w=17$. SGPLAN happily
+solves inputs beyond $w=70$. However, unlike in Experiment 2, both
+planners now spend a substantial proportion of their time on
+grounding. In SGPLAN, this translates to a ``parsing time'' (which we
+assume includes grounding) which grows from 180 ms to 21.7 seconds as
+$w$ grows from $1$ to $75$. The rest of the runtime (which also
+includes the search time) only grows from 400 ms to 2.3 seconds. This
+difference is particularly dramatic given that the actual optimal plan
+in each case is an identical plan of about 100 steps. The planning
+times for these instances are also concerning since times over a
+couple seconds negatively affect the overall response time of the
+system, which must react in real time to user actions.
 

### 2008-06-23 trunk/Papers/experiences/introduction.tex
small change
I'm essentially happy now

@@
 \citeyear{reiter00building}) is one of the major subfields of natural
-language processing, concerned with computing natural language sentences or
-texts that convey a given piece of information to the user. Intuitively,
-this task can be viewed as a problem involving actions, beliefs, and goals:
-an agent communicating with another agent tries to change the mental state
-of the hearer by applying actions which correspond to the utterance of
-words or sentences. This characterization of NLG suggests obvious parallels
-to automated planning---a view which has a long tradition in NLG
+language processing, concerned with computing natural language
+sentences or texts that convey a given piece of information to the
+user. Intuitively, this task can be viewed as a problem involving
+actions, beliefs, and goals: an agent communicating with another agent
+tries to change the mental state of the hearer by applying actions
+which correspond to the utterance of words or sentences. This
+characterization of NLG suggests obvious parallels to automated
+planning---a view which has a long tradition in NLG
 \cite{appelt:planning,young94dpocl}.  More recently, there has been a
 resurgence of interest in applying modern planning techniques to NLG
-\cite{Steedman-Petrick:07,KolSto07,benotti08b}.  While efficiency has not
-been the main focus of NLG, problems in NLG are often more complex than
-their counterparts in other areas of natural language processing (such as
-parsing \cite{KolStr02}), giving rise to computationally challenging
-domains. These new approaches combine an interest in planning as a
-modelling tool for natural language, with a hope of improved efficiency by
-exploiting modern approaches and algorithms.
+\cite{Steedman-Petrick:07,KolSto07,benotti08b}.  While efficiency has
+not always been the main focus of NLG, problems in NLG are often more
+complex than their counterparts in other areas of natural language
+processing (such as parsing \cite{KolStr02}), giving rise to
+computationally challenging domains. These new approaches combine an
+interest in planning as a modelling tool for natural language, with a
+hope of improved efficiency by exploiting modern approaches and
+algorithms.
 
@@
 (IPC)---on a range of problem instances from our planning domains.  We also
-compare these results against an ad-hoc Java implementation of GraphPlan.
-Our findings are mixed. On the one hand, it turns out that modern planners
-handle some of the search problems that arise in NLG quite easily (although
-large problem instances remain a challenge). On the other hand, these same
-planners spend tremendous amounts of time on preprocessing. For instance,
-FF spends 90\% of its runtime in the sentence generation domain on
-grounding out literals and actions, and ends up slower than our version of
-GraphPlan that avoids grounding. For domains like ours, which are dominated
-by the number of actions and the universe size, rather than the
-combinatorics of the search problem, this observation suggests that the
-overall runtime of a planner could be improved by grounding more
+compare these results against an ad-hoc Java implementation of GraphPlan
+\cite{Blum1997}.  Our findings are mixed. On the one hand, it turns out
+that modern planners handle some of the search problems that arise in NLG
+quite easily (although large problem instances remain a challenge). On the
+other hand, these same planners spend tremendous amounts of time on
+preprocessing. For instance, FF spends 90\% of its runtime in the sentence
+generation domain on grounding out literals and actions, and ends up slower
+than our version of GraphPlan that avoids grounding. For domains like ours,
+which are dominated by the number of actions and the universe size, rather
+than the combinatorics of the search problem, this observation suggests
+that the overall runtime of a planner could be improved by grounding more
 selectively. We therefore offer these domains as challenges for the

### 2008-06-23 trunk/Papers/experiences/experiments.tex
small grammatical changes

@@
 Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
-(in lexicographic order) on the horizontal axis and the runtime is shown in
+in lexicographic order on the horizontal axis and the runtime is shown in
 milliseconds on the vertical axis. These results reveal a number of
@@
 
-In fact, FF spends so much time on grounding that it is consistently
+In fact, the time spent by FF on grounding is such that it is consistently
 outperformed by an ad-hoc Java implementation of GraphPlan which only
-computes instances of predicates and actions as they are discovered,
-while the planning graph is being built -- despite the fact that FF is
-consistently much faster as far as pure search time is concerned.
-Looking at the difference from a different angle, FF's performance is
-much more sensitive to the domain size: if we fix $n=1$, FF takes 60
-ms to compute a plan at $m=1$, but 2.4 seconds (for the same plan) at
-$m=10$; our GraphPlan implementation takes 30 ms at $m=1$ and still
-only 50 ms at $m=10$.  Conversely, GraphPlan's runtime grows much
-faster with the plan size (i.e., with growing values of $n$ for a
-fixed $m$).  Larger (but still realistically-sized) instances of the
-sentence generation problem are still problematic for the planners we
-tested.
+computes instances of predicates and actions as they are discovered (i.e.,
+while the planning graph is being built)---despite the fact that FF is
+consistently much faster as far as pure search time is concerned.  Looking
+at the difference from another angle, FF's performance is much more
+sensitive to the domain size: if we fix $n=1$, FF takes 60 ms to compute a
+plan at $m=1$, but 2.4 seconds (for the same plan) at $m=10$; our GraphPlan
+implementation takes 30 ms at $m=1$ and still only requires 50 ms at
+$m=10$.  Conversely, GraphPlan's runtime grows much faster with the plan
+size (i.e., with growing values of $n$ for a fixed $m$).  Larger (but still
+realistically-sized) instances of the sentence generation problem are still
+problematic for the planners we tested.
 
@@
 grid of positions, such that there are buttons at positions $(2i-1,1)$
-and $(2i,h)$ for $1 \leq i \leq n$. The player starts in position
-$(1,1)$ and must press all buttons in order to successfully complete
-the game. The world is generated as a GIVE world description, and then
-automatically converted into a planning problem as in
-Fig.~\ref{fig:give-planning} by the GIVE software.
+and $(2i,h)$ for $1 \leq i \leq n$. The player starts in position $(1,1)$
+and must press all the buttons in order to successfully complete the game.
+The world is generated as a GIVE world description, and then automatically
+converted into a planning problem (as in Figure~\ref{fig:give-planning}) by
+the GIVE software.
 
@@
 In the final experiment, we vary the structure of the GIVE world in order
-to judge the effect that universe size has on the planning problem in this
+to judge the effect of universe size on the planning problem in this
 domain.  Starting with the ordinary GIVE world described in Experiment~2,
@@
 actually needed in any plan, but approximate the situation in the actual
-GIVE domain, where most grid positions never need to be used. We leave the
-initial state and goal untouched. As before, we generate a GIVE world
-description and then convert it into a planning problem.
+GIVE domain, where most grid positions are never used. We leave the initial
+state and goal untouched. As before, we generate a GIVE world description
+and then convert it into a planning problem.
 

### 2008-06-23 tags/spark-08-submit/conclusion.tex
SPARK 2008 workshop submission

--- 2008-06-23 trunk/Papers/experiences/conclusion.tex
@@
 controlling the complexity of search, they also suffer from practical
-problems that limit their performance in unexpected places. In particular,
+problems that limit their performance in unexpected ways. In particular,
 in both domains the grounding step performed by both planners dominates the

### 2008-07-01 trunk/Papers/rmrs-08/introduction.tex
restructured intro

@@
 
-% What we'll do here.  Problem: RMRS is designed to support building
-% semantic components to shallow parsers, and is intended to offer a way
-% of integrating semantic representations
-% computed by deep and shallow parsing.  We demonstrate the feasibility
-% of this for the first time, by introducing a model theory for RMRS,
-% defining entailment among RMRS representations and characterising this
-% entailment relation 
-% syntactically as extension of solved forms (ref to solved forms).  
-% We show that the ``compatability'' of
-% deep and shallow semantic representations is definable in terms of the
-% entailment 
-% relation between the RMRS
-% representations.
+Representing semantics as a logical form that supports automated
+inference and model generation is vital for deeper language
+engineering tasks, such as dialogue systems.  Such detailed semantic
+representations can be obtained from hand-crafted deep grammars (e.g.,
+\cite{butt:etal:1999,copestake:flickinger:2000}), but deep parsing
+tends to lack robustness: hand-crafted grammars fail to cover all
+words and linguistic constructions; and unedited text and speech
+contains ill-formed phrases, which by design deep grammars do not
+handle.
 
-% just so we have a citation here: \cite{Copestake_etal:MRS}
+On the other hand, there has been a recent trend towards systems for
+robust wide-coverage semantic construction.  For instance,
+\cite{bos:etal:2004} build upon a statistical CCG parser, exploiting
+the elegant syntax-semantics interface that CCG provides;
+\cite{zettlemoyer:collins:2007} and others combine syntactic analyses
+with shallow features to learn string-to-semantics mappings; and
+\todo{say something about RASP and Anette's work}.
 
-Representing semantics as a logical form that supports automated
-inference and model construction is vital for deeper language
-engineering tasks, such as dialogue systems.
-These need access to the vast array of
-semantic information that is captured in conventional logical forms,
-including (but not limited to) information about semantic scope and
-predicate argument structure.  Hand-crafted grammars yield detailed
-logical forms (e.g., \cite{butt:etal:1999,copestake:flickinger:2000}), but deep parsing tends to lack
-robustness: hand-crafted grammars fail to cover all words and
-linguistic constructions; and unedited text and speech contains
-ill-formed phrases, which by design deep grammars do not handle.  
+However, there are certain semantic phenomena that such shallow
+approaches can't capture, including quantifier scope, optional
+arguments, and long-distance dependencies.  Forcing a shallow parser
+to make a decision about these phenomena is therefore a major source
+of errors.  It would be preferrable to give the parser the option to
+declare itself not sufficiently informed to make the decision, and
+leave the decision open -- i.e., to compute a partial semantic
+representation that neither over-determines nor under-determines the
+semantic information revealed by the (shallow) syntactic analysis, and
+can later be completed by further inferences.
+
+In this paper, we focus on one particular approach to representing
+partial semantic information and producing it from robust parsers:
+Robust Minimal Recursion Semantics ({\sc rmrs},
+\cite{copestake:2003}).  {\sc rmrs} is a generalisation of {\sc mrs}
+\cite{copestake:etal:2005} designed to support underspecification both
+of scope relations and of the predicate-argument structure, and has
+been demonstrated to support semantic constructions of shallow parsers
+ranging in depth from {\sc pos} taggers (refs) to chunk parsers (frank
+ref) and intermediate parsers ({\sc rasp} ref).  {\sc mrs} analyses
+derived from deep grammars, such as the English Resource Grammar
+\cite{copestake:flickinger:2000} can also be seen as special cases of
+{\sc rmrs} representations.
+
+The key contribution we make is to cast {\sc rmrs}, for the first
+time, as a logic with a clearly defined model theory.  This has a
+number of advantages.  First, it allows us to formally relate \rmrs\ to
+other framework such as \mrs\ and predicate logic in order to
+establish the compatibility of \rmrs\ with the approaches to robust
+semantic analysis mentioned above.  Second, it provides us with a
+notion of logical entailment, which we can use to verify the
+compatibility of \rmrs\ representations computed from different
+parsers, such as a deep and a shallow parser.  Finally, it is only
+through rigid formalisation that we can give users of \rmrs\ access to
+inference systems and to algorithms for efficiently computing the
+possible fully specific semantic representations described by a
+partial \rmrs.
+
+We will first illustrate the differences and challenges with deep and
+shallow semantic construction in an example in
+Section~\ref{sec:motivation}.  We will then define the syntax and
+model-theoretic semantics of \rmrs\ and introduce some important
+formal concepts in Section~\ref{sec:rmrs}.  Finally, we demonstrate
+the practical usefulness of our formalisation in
+Section~\ref{sec:entailment} by modelling compatibility of
+partial semantic representations as \rmrs\ entailment and providing a
+syntactic criterion for characterising entailment.
+Section~\ref{sec:conclusion} concludes and points to future work.
+
 \hidden{
-Deep
-parsers on their own also lack ambiguity resolution techniques,
-producing all possible analyses.  Integration of statistical parse
-selection with deep parsing is promising (e.g.,
-\cite{toutanova:etal:2004}), but is not a solved problem.
-}
+
+The language in which these partial representations are expressed
+
+Moreover, in a hybrid system that attempts to combine the outputs of
+deep and shallow parsers,
 
@@
 
-This paper focusses on a particular approach to producing semantic
-information from robust parsers, exemplified in
+This paper focuses on a particular approach to producing partial
+semantic information from robust parsers, exemplified in
 \cite{copestake:2003,frank:2004}, among others.  Their strategy is to
@@
 revealed by the (shallow) syntactic analysis.  The semantic formalism
-used to express this is Robust Minimal Recursion Semantics ({\sc
-  rmrs}, \cite{copestake:2003}); this is an extension of {\sc mrs}
-\cite{copestake:etal:2005} that is designed to be maximally flexible
+used to express this is Robust Minimal Recursion Semantics
+
+; this is a generalisation of
+
+ that is designed to be maximally flexible
 in the type of semantic information that can be left underspecified:
@@
 We show in Section~\ref{sec:motivation} that all these features are
-needed when information about lexical subcategorisation or
-syntactic dependencies is missing---a characteristic feature of
-shallow parsers.  Several researchers have demonstrated that {\sc
-  rmrs} is a suitable framework on which to semi-automatically
-construct semantic components to shallow parsers, ranging in depth
-from {\sc pos} taggers (refs) to chunk parsers (frank ref) and
-intermediate parsers ({\sc rasp} ref).
+needed when information about lexical subcategorisation or syntactic
+dependencies is missing---a characteristic feature of shallow parsers.
+Several researchers have demonstrated that {\sc rmrs} is a suitable
+framework on which to semi-automatically construct semantic components
+to shallow parsers, ranging in depth .
 
@@
 
+}
+
 %%% Local Variables: 

### 2008-07-03 trunk/Papers/rmrs-08/abstract.tex
3rd July

@@
 \begin{abstract}
-  la la la
+  One way to construct semantic
+  representations in a robust manner is to enhance shallow language
+  processors with semantic components (e.g., \cite{frank:2004,
+    briscoe:etal:2006}).  Here, we provide a model theory for a
+  semantic formalism 
+  that is designed to enable this,
+  namely Robust Minimal Recursion Semantics ({\sc rmrs},
+  \cite{copestake:2003}).  We show that {\sc rmrs} supports a notion
+  of entailment 
+  that has a number of desirable properties: specifically, it allows
+  {\sc rmrs} to form the basis for comparing the semantic output of
+  different parses of the same string, even when the parses vary in
+  the depth of their analysis, and in particular to test the
+  compatibility of the semantic output of a shallow parser with that
+  of a deep parser (where such deep parses exist).
 \end{abstract}

### 2008-07-03 trunk/Papers/rmrs-08/introduction.tex
3rd July

@@
 Representing semantics as a logical form that supports automated
-inference and model generation is vital for deeper language
+inference and model construction is vital for deeper language
 engineering tasks, such as dialogue systems.  Such detailed semantic
@@
 robust wide-coverage semantic construction.  For instance,
-\cite{bos:etal:2004} build upon a statistical CCG parser, exploiting
-the elegant syntax-semantics interface that CCG provides;
-\cite{zettlemoyer:collins:2007} and others combine syntactic analyses
-with shallow features to learn string-to-semantics mappings; and
-\todo{say something about RASP and Anette's work}.
+\cite{bos:etal:2004} exploit
+{\sc ccg}'s elegant syntax-semantics interface to produce a unique
+logical form for each derivation of
+a statistical {\sc ccg} parser;
+\cite{zettlemoyer:collins:2007} and others learn
+string-to-logical-form mappings, exploting informative features from
+the output of
+shallow language processors; and \cite{copestake:2003} and
+\cite{frank:2004}, among others,
+endow robust language processors with semantic components.
 
-However, there are certain semantic phenomena that such shallow
-approaches can't capture, including quantifier scope, optional
-arguments, and long-distance dependencies.  Forcing a shallow parser
-to make a decision about these phenomena is therefore a major source
+However, there are certain semantic
+phenomena that such shallow
+approaches can't capture reliably, including quantifier scope, optional
+arguments,
+% COMMENT: this issue about optional arguments is really too cryptic.
+% Basically, there is no way for assessing by *any* learned language
+% processor whether a single word form that appears in
+% a number of different syntactic valencies is an example of lexical
+% sense ambiguity, or they are all examples of the same sense being
+% used in syntactically different ways (i.e., we have optional
+% arguments).  Only hand-crafted grammars can tell you this.  So, for
+% example, Bos et al (2004) would get bet(x,y) for "I bet 5 pounds",
+% and bet(x,y,z) for "I bet you 5 pounds" and bet(x,p) for "I bet it's
+% going to rain": I.e., the vocab of the semantic formalism is
+% constructed automatically from NL word forms, and the predicate
+% symbols in that automatically constructed vocab vary in (a) valency,
+% and (b) the semantic dependencies represented by a given argument
+% position (e.g., the second argument of "bet" in "I bet 5 pounds" is
+% 5 pounds, and for "I bet it's raining" it's "it's raining".
+% 
+and long-distance dependencies.  
+% TODO: search Steedman and Clark EMNLP 2004 for the performance
+% figures on object relative clauses (which is constructions like
+% "The man that I met", "the man that I think likes me", but not "the
+% man that likes me"; they get about 50% of them exactly right; this
+% is impressive but shows it's error prone.
+Forcing a shallow parser
+to make a decision about these phenomena therefore can be a major source
 of errors.  It would be preferrable to give the parser the option to
 declare itself not sufficiently informed to make the decision, and
-leave the decision open -- i.e., to compute a partial semantic
+to leave the decision open -- i.e., to compute a partial semantic
 representation that neither over-determines nor under-determines the
 semantic information revealed by the (shallow) syntactic analysis, and
-can later be completed by further inferences.
+to complete the semantic information later, via inferences that
+exploit information extraneous to the shallow parser.
 
 In this paper, we focus on one particular approach to representing
-partial semantic information and producing it from robust parsers:
+partial semantic information:
 Robust Minimal Recursion Semantics ({\sc rmrs},
 \cite{copestake:2003}).  {\sc rmrs} is a generalisation of {\sc mrs}
-\cite{copestake:etal:2005} designed to support underspecification both
-of scope relations and of the predicate-argument structure, and has
-been demonstrated to support semantic constructions of shallow parsers
-ranging in depth from {\sc pos} taggers (refs) to chunk parsers (frank
-ref) and intermediate parsers ({\sc rasp} ref).  {\sc mrs} analyses
+\cite{copestake:etal:2005}, designed to support underspecification both
+of scope relations and of the predicate-argument structure.
+\cite{copestake:2003, frank:2004} use it to
+support semantic constructions by shallow parsers
+ranging in depth from {\sc pos} taggers to chunk parsers  
+and intermediate parsers such as {\sc rasp}
+\cite{briscoe:etal:2006}.  {\sc mrs} analyses
 derived from deep grammars, such as the English Resource Grammar

### 2008-07-03 trunk/Papers/rmrs-08/motivation.tex
3rd July

@@
 
-We motivate and describe the main features of {\sc rmrs} with a (toy)
+We motivate and describe the main features of {\sc rmrs} with a toy
 example.  Sentence 
@@
 English Resource Grammar ({\sc erg}, \cite{copestake:flickinger:2000})
-reflects this by underspecifying socpe: (i) each predication is
+reflects this: (i) each predication is
 labelled ($l_1$, $l_4$, etc); (ii) scopal arguments are indicated by
-holes (e.g., $h_2$ and $h_3$ in the quantifier {\em every}); and (iii)
+holes (e.g., $h_2$ and $h_3$ in the quantifier {\em every}); (iii)
 constraints on scope are expressed by the qeq conditions (roughly, $h
@@
   corresponding to the adjective is motivated by copulas---{\em The
-    cat is fat} arguably has the logical form 
+    cat is fat} has logical form 
   $\mbox{\_the\_q\_1}(x,\mbox{\_cat\_n\_1}(x),\mbox{\_fat\_j\_1}(e,x))$.}
@@
 $l_4\handel\mbox{\_cat\_n\_1}(x)$\\
-$l_5\handel\mbox{\_some\_n\_1}(y,h_6,h_7)$\\
-$l_8\handel\mbox{\_dog\_n\_1}(y)$\\
-$l_9\handel\mbox{\_chased\_n\_1}(e,x,y)$\\
-$h_0 =_q l_9, h_2 =_q l_4, h_6 =_q l_9$
+$l_5\handel\mbox{\_chased\_n\_1}(e,x,y)$\\
+$l_6\handel\mbox{\_some\_n\_1}(y,h_7,h_8)$\\
+$l_9\handel\mbox{\_dog\_n\_1}(y)$\\
+$h_2 =_q l_4, h_7 =_q l_9$
 \end{subexamples}
@@
 predicate symbol in the {\sc mrs} is a proxy for one or more
-base-language constructors of the same arity
+base-language constructors of the same arity.
 Assuming a very close
@@
 $l_{42}:a_{42}\handel\mbox{\_cat\_n\_1}(x_3)$\\
-$l_5:a_5\handel\mbox{\_some\_q\_1}(x_4)$,\\
-\hspace*{0.1in}$\mbox{RSTR}(a_5,h_6),
-\mbox{BODY}(a_5,h_7)$\\ 
-$l_8:a_8\handel\mbox{\_dog\_n\_1}(x_5)$\\
-$l_9:a_9\handel\mbox{\_chase\_v\_1}(e_{spast})$,\\
-\hspace*{0.1in}$\mbox{ARG1}(a_9,x_6),
-\mbox{ARG2}(a_9,x_7)$\\ 
-$h_2=_q l_{42}, l_{41}=l_{42}, h_6 =_q l_8, h_0=_q l_9$\\
-$x_1=x_2,x_2=x_3, x_4=x_5,$\\
-$x_3=x_6, x_5=x_7$
+$l_5:a_5\handel\mbox{\_chase\_v\_1}(e_{spast})$,\\
+\hspace*{0.1in}$\mbox{ARG1}(a_5,x_4),
+\mbox{ARG2}(a_5,x_5)$\\ 
+$l_6:a_6\handel\mbox{\_some\_q\_1}(x_6)$,\\
+\hspace*{0.1in}$\mbox{RSTR}(a_6,h_7),
+\mbox{BODY}(a_6,h_8)$\\ 
+$l_9:a_9\handel\mbox{\_dog\_n\_1}(x_7)$\\
+$h_2=_q l_{42}, l_{41}=l_{42}, h_7 =_q l_9$\\
+$x_1=x_2, x_2=x_3, x_3=x_4,$\\
+$x_5=x_6, x_5=x_7$
 \end{subexamples}
@@
 $l_{42}:a_{42}:\mbox{\_cat\_n}(x_3)$\\
-$l_9:a_9:\mbox{\_chased\_v}(e_{past})$, \\
-$l_5:a_5:\mbox{\_some\_q}(x_4)$, \\
-$l_8:a_8:\mbox{\_dog\_n}(x_5)$
+$l_5:a_5:\mbox{\_chased\_v}(e_{past})$, \\
+$l_6:a_6:\mbox{\_some\_q}(x_4)$, \\
+$l_9:a_9:\mbox{\_dog\_n}(x_5)$
 \end{subexamples}
@@
 subset of those corresponding to the predicate symbol {\em \_name\_t},
-for any word lemma {\em name} and {\sc pos} tag $t$ (written 
+for any word lemma {\em name}, {\sc pos} tag $t$ and sense $s$ (written 
 $\mbox{{\em \_name\_t\_s}} \sqsubset \mbox{{\em name\_t}}$).
@@
 Thus to maximise the semantic contribution
-of the parser while not overdetermining it, one should express that
-the argument position of the noun phrases in {\em give the bone} vs.\
-{\em give the dog} are underspecified between the second and third
-position, as shown in Figure~\ref{fig:2}.   This will be expressed in
+of the parser while not overdetermining it, the phrases
+{\em give the bone} vs.\
+{\em give the dog} should whether the semantic index of the noun
+phrase is the second or third argument to {\em give},
+as shown in Figure~\ref{fig:2}.   This will be expressed in
 {\sc rmrs} via the binary relation  $\mbox{{\em ARG}}_{\{2,3\}}(a,x)$

### 2008-07-03 trunk/Papers/rmrs-08/abstract.tex
3rd July

@@
   representations in a robust manner is to enhance shallow language
-  processors with semantic components (e.g., \cite{frank:2004,
-    briscoe:etal:2006}).  Here, we provide a model theory for a
+  processors with semantic components (e.g.,
+  \cite{frank:2004,briscoe:etal:2006}).  Here, we provide a model
+  theory for a 
   semantic formalism 

### 2008-07-03 trunk/Papers/rmrs-08/introduction.tex
3rd July

@@
 representations can be obtained from hand-crafted deep grammars (e.g.,
-\cite{butt:etal:1999,copestake:flickinger:2000}), but deep parsing
+\cite{butt:etal:1999,copestake:flickinger:2000}), but this
 tends to lack robustness: hand-crafted grammars fail to cover all
@@
 However, there are certain semantic
-phenomena that such shallow
-approaches can't capture reliably, including quantifier scope, optional
+phenomena that these robust
+approaches don't capture reliably, including quantifier scope, optional
 arguments,
@@
 % 
-and long-distance dependencies.  
-% TODO: search Steedman and Clark EMNLP 2004 for the performance
-% figures on object relative clauses (which is constructions like
-% "The man that I met", "the man that I think likes me", but not "the
-% man that likes me"; they get about 50% of them exactly right; this
-% is impressive but shows it's error prone.
+and long-distance dependencies (for instance, \cite{clark:etal:2004}
+report that the parser used in \cite{bos:etal:2004} 
+yields semantic dependencies from the unbounded
+dependencies borne from
+object extraction---e.g., {\em the man that I met\dots}---with 63\%
+accuracy).
 Forcing a shallow parser
 to make a decision about these phenomena therefore can be a major source
-of errors.  It would be preferrable to give the parser the option to
+of errors.  Depending on the application, it may be preferrable to
+give the parser the option to 
 declare itself not sufficiently informed to make the decision, and
-to leave the decision open -- i.e., to compute a partial semantic
+to leave the decision open---i.e., to compute a partial semantic
 representation that neither over-determines nor under-determines the
@@
 \cite{copestake:2003}).  {\sc rmrs} is a generalisation of {\sc mrs}
-\cite{copestake:etal:2005}, designed to support underspecification both
+\cite{copestake:etal:2005} that is designed to support underspecification both
 of scope relations and of the predicate-argument structure.
-\cite{copestake:2003, frank:2004} use it to
+\cite{copestake:2003,frank:2004} use it to
 support semantic constructions by shallow parsers
@@
 number of advantages.  First, it allows us to formally relate \rmrs\ to
-other framework such as \mrs\ and predicate logic in order to
+other frameworks such as \mrs\ and predicate logic in order to
 establish the compatibility of \rmrs\ with the approaches to robust
@@
 inference systems and to algorithms for efficiently computing the
-possible fully specific semantic representations described by a
-partial \rmrs.
+possible fully specific semantic representations described by the \rmrs
+output of a shallow parser.
 

### 2008-07-07 trunk/Papers/rmrs-08/introduction.tex
Just in case

@@
 and long-distance dependencies (for instance, \cite{clark:etal:2004}
-report that the parser used in \cite{bos:etal:2004} 
-yields semantic dependencies from the unbounded
-dependencies borne from
-object extraction---e.g., {\em the man that I met\dots}---with 63\%
-accuracy).
-Forcing a shallow parser
-to make a decision about these phenomena therefore can be a major source
-of errors.  Depending on the application, it may be preferrable to
-give the parser the option to 
-declare itself not sufficiently informed to make the decision, and
-to leave the decision open---i.e., to compute a partial semantic
-representation that neither over-determines nor under-determines the
-semantic information revealed by the (shallow) syntactic analysis, and
-to complete the semantic information later, via inferences that
-exploit information extraneous to the shallow parser.
+report that the parser used in \cite{bos:etal:2004} yields semantic
+dependencies from the unbounded dependencies borne from object
+extraction---e.g., {\em the man that I met\dots}---with 63\%
+accuracy).  Forcing a shallow parser to make a decision about these
+phenomena therefore can be a major source of errors.  Depending on the
+application, it may be preferrable to give the parser the option to to
+leave the decision open when it's not sufficiently informed---i.e., to
+compute a partial semantic representation that neither over-determines
+nor under-determines the semantic information revealed by the
+(shallow) syntactic analysis, and to complete the semantic information
+later, via inferences that exploit information extraneous to the
+parser.
 

### 2008-07-20 trunk/Papers/experiences/conclusion.tex
revised for reviewer comments

--- 2008-06-05 trunk/Papers/experiences/conclusion.tex
@@
 
-In this paper, we have introduced two novel planning domains related
-to problems in natural language generation: the sentence generation
-problem and the GIVE navigation problem.  We have then reported on a
-number of experiments in which we applied off-the-shelf planners to
-these domains.  Our main results were that while modern planners do
-pretty well in controlling the complexity of search, they suffer from
-practical problems that limit their performance in unexpected places.
-In particular, in both domains the grounding step that both planners
-perform tends to take much more time than the search itself.
+In this paper, we introduced two novel planning domains arising from
+problems in natural language generation: the sentence generation
+domain and the GIVE navigation domain. We also reported on the results
+of a number of experiments in which we applied off-the-shelf planners
+to these domains.
 
-\todo{say something clever about the future}
+Our results were mixed. While modern planners do a pretty good job of
+controlling the complexity of search, they also suffer from practical
+problems that limit their performance in unexpected ways. In
+particular, in both domains the grounding step performed by both
+planners dominates the time it takes to perform the search itself. In
+future work, it would be interesting to explore planners such as SHOP2
+\cite{DBLP:journals/jair/NauAIKMWY03}, which do not perform the
+initial grounding step.  We are also considering to propose some of
+our domains as IPC benchmarks.
+
+The NLG community's recent interest in planning presents a valuable
+opportunity for planning researchers: provided some of the challenges
+we have highlighted can be addressed, projects like GIVE (and other
+NLG-inspired problem domains) offer a constructive platform for the
+planning community to showcase their techniques to a wider
+audience---and to improve the quality of their tools for real-world
+planning tasks.  Furthermore, GIVE highlights planning-related
+challenges such as plan execution monitoring and plan presentation
+and summarisation and offers a platform on which such technologies can
+be evaluated in experiments with human users.
+
+\paragraph{Acknowledgments.} This work is a collaboration that arose
+in the context of the ``planning and language'' group at the
+University of Edinburgh.  We thank all members of this group,
+especially Hector Geffner and Mark Steedman, for interesting
+discussions.  This work was supported by the DFG Research Fellowship
+``CRISP: Efficient integrated realization and microplanning'' and the
+EU project XYZ PACO-PLUS.
 

### 2008-07-20 trunk/Papers/experiences/discussion.tex
revised for reviewer comments

@@
 
-The positive conclusion we can draw from the simple experiments reported
-above is that modern planners are fairly good at controlling the search for
-many of the moderately-sized NLG problems we looked at. This is
-particularly true for FF in the sentence generation domain, which generates
-nontrivial 14-word sentences in about two seconds. There is still room for
-improvement, however. For instance, by comparison a special-purpose
-algorithm for referring expressions could generate ten of these 14-word
-sentences in a few milliseconds. Although search efficiency is not the main
-performance metric in these domains, it nevertheless becomes a factor in
+The positive conclusion we can draw from the simple experiments
+reported above is that modern planners are fairly good at controlling
+the search for many of the moderately-sized NLG problems we looked
+at. This is particularly true for FF in the sentence generation
+domain, which generates nontrivial 14-word sentences in about two
+seconds. There is still room for improvement, however; special-purpose
+algorithms generate much larger referring expressions in milliseconds
+\cite{AreKolStr08}.  Although search efficiency is not the main
+concern in these domains, it nevertheless becomes a factor in
 scenarios like Experiment~2, where the world is almost completely
@@
 
-%The positive conclusion we can draw from the experiments reported
-%above is that for the most part, modern planners are pretty good at
-%controlling the search for the NLG-related inputs that we looked at.
-%This is particularly true for FF in the sentence generation domain,
-%which generates nontrivial 14-word sentences in about two seconds.
-%There is still room for improvement (a special-purpose algorithm for
-%referring expressions could generate ten of these 14 words in a few
-%milliseconds), and the search becomes slower -- for obvious reasons --
-%in the almost completely unconstrained world of Experiment 2, but
-%search is not the main efficiency bottleneck.
-
-The most restrictive bottleneck in the two domains we investigated is the
-initial grounding step that many modern planners perform. While this may be
-a good strategy for traditional planning benchmarks and IPC domains---where
-plan length often dominates universe size, and an initial investment into
-grounding pays off in saved instantiations during the search---this
-approach is less effective in the domains we've considered. As our
-experiments have shown, there are natural planning domains in which
-relatively short plans must be computed in large universes. For instance,
-in the case of the sentence generation domain, it is not unrealistic for
-the universe to consist of thousands of domain individuals and tens of
-thousands of actions, some of which require three domain individuals as
-parameters. The common strategy of splitting the universe over several
-types of individuals will not be very effective here.
-
-%The most restrictive bottleneck in the two domains we have looked at
-%in this paper is the initial grounding step that almost all modern
-%planners perform.  Our conclusion from the experiments reported above
-%is that this is a good strategy in domains where the typical plan
-%length dominates the typical universe size, i.e.\ where the initial
-%investment into grounding pays off in saved unifications during the
-%search.  However, as we have shown, there are natural planning domains
-%in which relatively short plans must be computed in large universes;
-%it is not unrealistic for the universe in the sentence generation
-%domain to consist of thousands of domain individuals and tens of
-%thousands of actions, some of which take three domain individuals as
-%parameters.  The common strategy of splitting the universe over
-%several types of individuals will not be very effective here.
+The most restrictive bottleneck in the two domains we investigated is
+the initial grounding step that many modern planners perform. While
+this may be a good strategy for traditional planning benchmarks and
+IPC domains---where plan length often dominates universe size, and an
+initial investment into grounding pays off in saved instantiations
+during the search---this approach is less effective in the domains
+we've considered. As our experiments have shown, there are natural
+planning domains in which relatively short plans must be computed in
+large universes. For instance, in the case of the sentence generation
+domain, it is not unrealistic to look for plans of length 20 over a
+universe consisting of thousands of domain individuals and tens of
+thousands of actions, some of which require three domain individuals
+as parameters. The common strategy of splitting the universe over
+several types of individuals will not be very effective here.
 
@@
 development of algorithms that control search in sophisticated ways,
-resulting in a host of planners that are more powerful and more successful
-than their predecessors. Depending on the particular algorithm, however,
-these planners often ground out sets of predicates and actions before
-starting the search itself, which has a very pronounced effect on domains
-such as those described above. Since our domains are not that unusual in
-their structure and composition, we hope that the lessons learnt from our
-experiences can help improve the performance of current systems. We believe
-that such improvements are necessary if planning is to become a more mature
-technology that can offer tools to a wider community of users. At the end
-of the day, real-world users will care about the \emph{total} runtime of a
-planner, and this is more than just the search time.
-%We offer our larger problem instances as future challenges for the
-%planning community.
-
-%We believe that if planning is to become a mature technology which
-%real-world users will believe in, it must find a way to deal with the
-%grounding problem more gracefully than current systems do.  The
-%research focus in developing planning algorithms has quite rightfully
-%been on dealing with the search, and for this it is convenient --
-%depending on the particular algorithm, necessary -- to ground out all
-%predicates and actions before starting the search itself.  However, we
-%think that the domains presented here are not that unusual in their
-%ratio of plan length against universe size.  At the end of the day,
-%real-world users will care about the \emph{total} runtime of a
-%planner, and this is more than just the search.
+resulting in a host of planners that are more powerful and more
+successful than their predecessors. However, the common strategy of
+grounding out sets of predicates and actions in implementations of
+these algorithms has a very pronounced effect on domains such as those
+described above. Reachability analysis has the potential to help---we
+could gain a lot by only computing reachable instances---, but this
+too is typically done on the grounded problem. \todo{check if this is
+  true} Since our domains are not that unusual in their structure and
+composition, we hope that the lessons learnt from our experiences can
+help improve the performance of current systems. We believe that such
+improvements are necessary if planning is to become a more mature
+technology that can offer tools to a wider community of users. At the
+end of the day, real-world users will care about the \emph{total}
+runtime of a planner, and this is more than just the search time.
 
@@
 
-%By and large, it is a relatively pleasant experience to come as a
-%``customer'' to the planning community: Thanks to the competitions, it
-%is easy to identify and download a fast implementation, at least for
-%Linux.  However, in the course of our experiments we found (and
-%reported) bugs in both SGPLAN and FF.  And as our experiments have
-%shown, there is no one silver bullet: Even planners as closely related
-%as FF and SGPLAN can differ significantly in their performance on
-%different domains.
-
 %%% Local Variables: 

### 2008-07-20 trunk/Papers/experiences/domains.tex
revised for reviewer comments

@@
 
-In this section, we present our planning domains: the sentence
-generation domain and the instruction giving domain.
+We begin by presenting our two planning domains: the sentence generation
+domain and the instruction giving domain.
 
@@
 
-In practice, the process of determining referring expressions
-typically interacts with the realization step and so it turns out to
-be beneficial to perform both of these steps together. This was the
-goal of the SPUD system \cite{Stone2003a}, which did this using a
-top-down generation algorithm based on tree-adjoining grammars
-\cite{joshi;etal1997} whose lexical entries were equipped with
-semantic and pragmatic information. Unfortunately, SPUD suffered from
-having to explore a huge search space, and had to resort to a
-non-optimal greedy search strategy to retain reasonable efficiency.
-To improve on the efficiency of SPUD,
+In practice, the process of determining referring expressions typically
+interacts with the realization step and so it turns out to be beneficial to
+perform both of these steps together. This was the goal of the SPUD system
+\cite{Stone2003a}, which performed this task using a top-down generation
+algorithm based on tree-adjoining grammars \cite{joshi;etal1997} whose
+lexical entries were equipped with semantic and pragmatic information.
+Unfortunately, SPUD suffered from having to explore a huge search space,
+and had to resort to a non-optimal greedy search strategy to retain
+reasonable efficiency.  To improve on the efficiency of SPUD,
 \citeauthor{KolSto07}~\shortcite{KolSto07} translate the sentence
-generation problem into a planning problem and use a planning
-algorithm for generation.
+generation problem into a planning problem and use a planning algorithm for
+generation.
 
@@
 
-We compute this grammatical derivation top-down by starting with the
-elementary tree for ``sleeps''. This tree satisfies the need to convey
-the semantic information, but introduces a need to generate a noun
-phrase (NP) for the subject; this NP must refer uniquely to the target
-referent $r_1$. In a second step, we substitute the tree for ``the
-rabbit'' into the open NP leaf, which makes the derivation
-grammatically complete. Since there are two different individuals that
-could be described as ``the rabbit''---technically, $r_2$ is still a
-\emph{distractor} (i.e., based on the description ``the rabbit'', the
-hearer might erroneously think that we're talking about $r_2$ and not
-$r_1$)---we are still not finished. To complete the derivation, the
-tree for ``white'' is added to the existing structure by the
-\emph{adjunction} operation, making the derivation syntactically and
-semantically complete.
+We compute this grammatical derivation in a top-down manner, starting with
+the elementary tree for ``sleeps''. This tree satisfies the need to convey
+the semantic information, but introduces a need to generate a noun phrase
+(NP) for the subject; this NP must refer uniquely to the target referent
+$r_1$. In a second step, we substitute the tree for ``the rabbit'' into the
+open NP leaf, which makes the derivation grammatically complete. Since
+there are two different individuals that could be described as ``the
+rabbit''---technically, $r_2$ is still a \emph{distractor} (i.e., based on
+the description ``the rabbit'', the hearer might erroneously think that
+we're talking about $r_2$ and not $r_1$)---we are still not finished. To
+complete the derivation, the tree for ``white'' is added to the existing
+structure by an \emph{adjunction} operation, making the derivation
+syntactically and semantically complete.
 
@@
 
-The process described above has clear parallels to planning: we manipulate
-a state by applying actions in order to achieve a goal. We can make this
-connection even more precise by translating the SPUD problem into a
-planning problem. For instance, Figure~\ref{fig:white-rabbit-as-planning}
-shows the corresponding PDDL actions for the above generation task, where
-each action corresponds to an operation that adds a single elementary tree
-to the derivation.  In each case, the first parameter of the action is a
-node name in the derivation tree, and the remaining parameters stand for
-the individuals to which the semantic roles will be instantiated.  The
-syntactic preconditions and effects are encoded using $\mathsf{open}$ and
-$\mathsf{canadjoin}$ literals; the status of each referring expression is
-tracked using $\mathsf{distractor}$ literals. Notice that the action
-effects contain terms of the form $\mathsf{subj}(u)$, which construct new
-node names. In order to meet the syntactic requirements of PDDL, these
-terms can be eliminated by estimating an upper bound $n$ for the plan
-length, making $n$ copies of each action, ensuring that copy $i$ can only
-be applied in step $i$, and replacing the term $\mathsf{subj}(u)$ in an
-action copy by the constant $\mathsf{subj}_i$.
+The process described above has clear parallels to planning: we
+manipulate a state by applying actions in order to achieve a goal. We
+can make this connection even more precise by translating the SPUD
+problem into a planning problem.\footnote{See
+  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
+  which this conversion is implemented.}  For instance,
+Figure~\ref{fig:white-rabbit-as-planning} shows the corresponding PDDL
+actions for the above generation task, where each action corresponds
+to an operation that adds a single elementary tree to the derivation.
+In each case, the first parameter of the action is a node name in the
+derivation tree, and the remaining parameters stand for the
+individuals to which the semantic roles will be instantiated.  The
+syntactic preconditions and effects are encoded using $\mathsf{open}$
+and $\mathsf{canadjoin}$ literals; the status of each referring
+expression is tracked using $\mathsf{distractor}$ literals. Notice
+that the action effects contain terms of the form $\mathsf{subj}(u)$,
+which construct new node names. In order to meet the syntactic
+requirements of PDDL, these terms can be eliminated by estimating an
+upper bound $n$ for the plan length, making $n$ copies of each action,
+ensuring that copy $i$ can only be applied in step $i$, and replacing
+the term $\mathsf{subj}(u)$ in an action copy by the constant
+$\mathsf{subj}_i$.
 
-Once we've defined appropriate actions, we can solve the generation
+Once the appropriate actions are defined, we can solve the generation
 problem as an ordinary planning problem. For instance, the following plan
@@
 Using this plan, the grammatical derivation in
-Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the generated
-sentence, can be systematically reconstructed. Thus, we can solve the
-sentence generation problem via the detour through planning and bring
-current search heuristics for planning to bear on generation.
+Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
+generated sentence ``the white rabbit sleeps'', can be systematically
+reconstructed. Thus, we can solve the sentence generation problem via
+the detour through planning and bring current search heuristics for
+planning to bear on generation.
 
@@
 The object of the GIVE Challenge (``Generating Instructions in Virtual
-Environments''; Koller et al.\ \citeyear{alexander07:_shared_task_propos})
-is to build an NLG system which is able to produce natural-language
-instructions which will guide a human user in performing some task in a
-virtual environment. From an NLG perspective, GIVE makes for an
-interesting challenge because it is a theory-neutral task that exercises
-all components of an NLG system, and emphasizes the study of communication
-in a (simulated) physical environment. It also has the advantage that the
-user and the NLG system can be physically in different places, as long as
-the 3D client and the NLG system are connected over a network. This makes
-it possible to evaluate GIVE NLG systems on a large scale over the
-Internet. The first GIVE evaluation will take place in late 2008;
-currently eight research teams from five countries are working on
-developing systems to participate in the challenge.\footnote{See the GIVE
- website for more details about this project:
- \url{homepages.inf.ed.ac.uk/v1akolle/proj/give/}.}
+Environments''; Koller et al.\
+\citeyear{alexander07:_shared_task_propos}) is to build an NLG system
+which is able to produce natural-language instructions which will
+guide a human user in performing some task in a virtual
+environment. From an NLG perspective, GIVE makes for an interesting
+challenge because it is a theory-neutral task that exercises all
+components of an NLG system, and emphasizes the study of communication
+in a (simulated) physical environment. It also has the advantage that
+the user and the NLG system can be in physically different places, as
+long as the 3D client and the NLG system are connected over a
+network. This makes it possible to evaluate GIVE NLG systems on a
+large scale over the Internet. The first GIVE evaluation will take
+place in late 2008; currently eight research teams from five countries
+are working on developing systems to participate in the
+challenge.\footnote{See the GIVE website for more details about this
+  project: \url{http://www.give-challenge.org/}.}
 
@@
 major problem is that the natural language instruction sequences can't
-simply be obtained by verbalising each action instance in the plan in
+simply be obtained by verbalising each action instance in a plan in
 turn.  For instance, the NLG system could generate an instruction
@@
 evaluation score.  Having just pressed the button, it would be much
-better to say ``turn around and walk through the door''; i.e.,
-multiple action instances should be merged into a single
-instruction. Conversely, it may also be necessary to express a single
+better to say ``turn around and walk through the door''. That is,
+multiple action instances should be merged into a single instruction;
+plans must be \emph{summarised} in order to present them to the
+user. Conversely, it may also be necessary to express a single
 planning step with several instructions.  Having just entered the top
@@
 
-From a planning perspective, the use of planning in GIVE imposes very
-strict runtime requirements on the planner: Planning must happen in
-real time, because if the system takes too much time between deciding
-to give an instruction and actually giving this instruction, the user
-may have walked or turned away, thus making the instruction invalid.
-Furthermore, plan execution monitoring also plays an important role in
-the GIVE problem.  At a high level, the system needs to monitor a
-user's actions and compare them against the generated instruction set
-to determine if the user has correctly followed directions or not. In
-the case of the latter, new instructions may have to be generated. In
-practice, the situation can be quite complicated since the mental
-state of the user is not known and so the system must observe the
-user's actions in real time. For example, a user directed to ``turn
-around and walk through the door'' may not necessarily perform these
-actions to the letter, i.e., immediately turning 180 degrees and
-proceeding directly to the door. Instead, the user might take a
-roundabout route through the room, eventually exiting out the
-door. Although the user's actions do not match the generated
-instructions exactly, they meet the intended goal. The system must be
-able to identify such ``equivalent'' plans and not immediately
-generate new instructions as soon as the user's actions have gone off
-course. Furthermore, a user can communicate certain intentions to the
-system, both through action and inaction. For instance, the system
-should infer that a user has failed to follow instructions if the user
-exits a room when given a directive to ``walk to the centre of the
-room''. However, the system should also make a similar conclusion if a
-user simply does nothing when given the instruction.
+From a planning perspective, GIVE imposes very strict runtime requirements
+on the planner and its associated modules: planning must happen in real
+time and the system must respond to a user in a timely fashion.  If the
+system takes too much time deliberating over an instruction to give, rather
+than actually giving this instruction, the user may have walked or turned
+away, thus making the instruction invalid.  Furthermore, plan execution
+monitoring also plays an important role in the GIVE problem.  At a high
+level, the system needs to monitor a user's actions and compare them
+against the generated instruction set to determine if the user has
+correctly followed directions or not. In the case of the latter, new
+instructions may have to be generated. In practice, the situation can be
+quite complicated since the mental state of the user is not known and so
+the system must observe the user's actions in real time. For example, a
+user directed to ``turn around and walk through the door'' may not
+necessarily perform these actions to the letter, i.e., immediately turning
+180 degrees and proceeding directly to the door. Instead, the user might
+take a roundabout route through the room, eventually exiting out the door.
+Although the user's actions do not match the generated instructions
+exactly, they meet the intended goal. The system must be able to identify
+such ``equivalent'' plans and not immediately generate new instructions as
+soon as the user's actions have gone off course. Furthermore, a user can
+communicate certain intentions to the system, both through action and
+inaction. For instance, the system should infer that a user has failed to
+follow instructions if the user exits a room when given a directive to
+``walk to the centre of the room''. The system should also make a
+similar conclusion if a user simply does nothing when given the
+instruction.
+
+%between deciding to give an instruction and
 

### 2008-07-20 trunk/Papers/experiences/experiments.tex
revised for reviewer comments

@@
 
-Results for the $h=20$ case, with $n$ ranging from $1$ to $40$, are shown
-in Figure~\ref{fig:give-runtime-minimal}.  The most obvious result is that
-FF is unable to solve any problems beyond $n=13$ on our experimentation
-machine within the memory limit of 1 GB.  SGPLAN, on the other hand, solves
-instances beyond $n=40$ without major problems.  The time spent on
-grounding is not a major factor in either planner, probably because the
-planners need more time to actually compute the plan---for instance, the
-optimal plan for the problem $n=40$ has a length of about 1600 steps.
+Results for the $h=20$ case, with $n$ ranging from $1$ to $40$, are
+shown in Figure~\ref{fig:give-runtime-minimal}.  The most obvious
+result is that FF is unable to solve any problems beyond $n=13$ on our
+experimentation machine within the memory limit of 1 GB.  SGPLAN, on
+the other hand, solves instances beyond $n=40$ without major
+problems.\footnote{Our ad-hoc implementation of GraphPlan is unable to
+  find any of these plans.}  The time spent on grounding is not a
+major factor in either planner, probably because the planners need
+more time to actually compute the plan---for instance, the optimal
+plan for the problem $n=40$ has a length of about 1600 steps.  As an
+example, the plan for $n=h=1$ looks as follows:
+
+\begin{enumerate}
+\item $\mathsf{move}(\mathsf{pos\_1\_1},\mathsf{pos\_1\_2}, \mathsf{north})$
+\item $\mathsf{manipulate}\textsf{-}\mathsf{u1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_1\_2})$
+\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{north}, \mathsf{east})$
+\item $\mathsf{move}(\mathsf{pos\_1\_2}, \mathsf{pos\_2\_2},
+  \mathsf{east})$
+\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{east}, \mathsf{south})$
+\item $\mathsf{move}(\mathsf{pos\_2\_2}, \mathsf{pos\_2\_1}, \mathsf{south})$
+\item $\mathsf{manipulate}\textsf{-}\mathsf{l1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_2\_1})$
+\end{enumerate}
 
@@
 difference is particularly dramatic given that the actual optimal plan
-in each case is an identical plan of about 100 steps. The planning
-times for these instances are also concerning since times over a
-couple seconds negatively affect the overall response time of the
-system, which must react in real time to user actions.
+in each case is an identical plan of about 100 steps (namely, the one
+we would have found in Experiment 2 for $h=20$ and $n=5$). The
+planning times for these instances are also concerning since times
+over a couple seconds negatively affect the overall response time of
+the system, which must react in real time to user actions.
 

### 2008-07-20 trunk/Papers/experiences/introduction.tex
revised for reviewer comments

@@
 which correspond to the utterance of words or sentences. This
-characterization of NLG suggests obvious parallels to automated
+characterisation of NLG suggests obvious parallels to automated
 planning---a view which has a long tradition in NLG
@@
 
-We will then discuss some of our experiences using off-the-shelf planners
-in these domains. In particular, we explore the efficiency of FF
+We then discuss some of our experiences using off-the-shelf planners
+in our two domains. In particular, we explore the efficiency of FF
 \cite{HoffmannNebel01} and SGPLAN
-\cite{hsu06:_new_featur_in_sgplan_for}---planners that have been successful
-on traditional benchmarks, and in the International Planning Competition
-(IPC)---on a range of problem instances from our planning domains.  We also
-compare these results against an ad-hoc Java implementation of GraphPlan
-\cite{Blum1997}.  Our findings are mixed. On the one hand, it turns out
-that modern planners handle some of the search problems that arise in NLG
-quite easily (although large problem instances remain a challenge). On the
-other hand, these same planners spend tremendous amounts of time on
-preprocessing. For instance, FF spends 90\% of its runtime in the sentence
-generation domain on grounding out literals and actions, and ends up slower
-than our version of GraphPlan that avoids grounding. For domains like ours,
-which are dominated by the number of actions and the universe size, rather
-than the combinatorics of the search problem, this observation suggests
-that the overall runtime of a planner could be improved by grounding more
-selectively. We therefore offer these domains as challenges for the
-planning community.
+\cite{hsu06:_new_featur_in_sgplan_for}---planners that are easily
+available, support expressive preconditions and effects out of the
+box, and have been successful on traditional benchmarks, and in the
+International Planning Competition (IPC)---on a range of problem
+instances from our planning domains.  We also compare these results
+against an ad-hoc Java implementation of GraphPlan \cite{Blum1997}.
+Our findings are mixed. On the one hand, it turns out that modern
+planners handle some of the search problems that arise in NLG quite
+easily (although large problem instances still remain a challenge).
+On the other hand, these same planners spend tremendous amounts of
+time on preprocessing. For instance, FF spends 90\% of its runtime in
+the sentence generation domain on grounding out literals and actions,
+and ends up slower than our version of GraphPlan that avoids
+grounding. For domains like ours, which are dominated by the number of
+actions and the universe size, rather than the combinatorics of the
+search problem, this observation suggests that the overall runtime of
+a planner could be improved by grounding more selectively. We
+therefore offer these domains as challenges for the planning
+community.
 

### 2008-07-24 trunk/Papers/experiences/conclusion.tex
new graphs; small changes to acknowledgements

@@
 
-\paragraph{Acknowledgments.} This work is a collaboration that arose
-in the context of the ``planning and language'' group at the
-University of Edinburgh.  We thank all members of this group,
-especially Hector Geffner and Mark Steedman, for interesting
-discussions.  This work was supported by the DFG Research Fellowship
-``CRISP: Efficient integrated realization and microplanning'' and the
-EU project XYZ PACO-PLUS.
+\section{Acknowledgements}
+This work arose in the context of the Planning and Language Interest Group
+at the University of Edinburgh. We thank all members of this group,
+especially Hector Geffner and Mark Steedman, for interesting discussions.
+This work was partially supported by the DFG Research Fellowship ``CRISP:
+Efficient integrated realization and microplanning'' and by the European
+Commission through the PACO-PLUS project (FP6-2004-IST-4-27657).
+
+%\paragraph{Acknowledgments.} This work is a collaboration that arose
+%in the context of the ``planning and language'' group at the
+%University of Edinburgh.  We thank all members of this group,
+%especially Hector Geffner and Mark Steedman, for interesting
+%discussions.  This work was supported by the DFG Research Fellowship
+%``CRISP: Efficient integrated realization and microplanning'' and the
+%EU project XYZ PACO-PLUS.
 

### 2008-07-25 trunk/Papers/experiences/conclusion.tex
small changes

@@
 of a number of experiments in which we applied off-the-shelf planners
-to these domains.
+to a set of problem instances in these domains.
 
@@
 controlling the complexity of search, they also suffer from practical
-problems that limit their performance in unexpected ways. In
-particular, in both domains the grounding step performed by both
-planners dominates the time it takes to perform the search itself. In
-future work, it would be interesting to explore planners such as SHOP2
-\cite{DBLP:journals/jair/NauAIKMWY03}, which do not perform the
-initial grounding step.  We are also considering to propose some of
-our domains as IPC benchmarks.
+problems that limit their performance in unexpected ways. In particular, in
+both domains the grounding step performed by FF and SGPLAN dominates the
+time it takes to perform the search itself. In future work, we hope to
+extend our analysis by exploring other planners, such as SHOP2
+\cite{DBLP:journals/jair/NauAIKMWY03}, which do not perform the initial
+grounding step. We also believe our domains would provide suitable
+challenges for planners entered in future editions of the IPC.
+%We are also considering to propose some of our domains as IPC benchmarks.
 
 The NLG community's recent interest in planning presents a valuable
-opportunity for planning researchers: provided some of the challenges
-we have highlighted can be addressed, projects like GIVE (and other
-NLG-inspired problem domains) offer a constructive platform for the
-planning community to showcase their techniques to a wider
-audience---and to improve the quality of their tools for real-world
-planning tasks.  Furthermore, GIVE highlights planning-related
-challenges such as plan execution monitoring and plan presentation
-and summarisation and offers a platform on which such technologies can
-be evaluated in experiments with human users.
+opportunity for planning researchers.  While GIVE highlights
+planning-related challenges such as plan execution monitoring and plan
+presentation (i.e., summarisation and elaboration), it also offers a
+platform on which such technologies can be evaluated in experiments with
+human users.  Furthermore, more mainstream research on reasoning under
+uncertainty, planning with incomplete information and sensing, and resource
+management can also be applied in this domain setting. Provided some of the
+challenges we have highlighted can be addressed, projects like GIVE (and
+other NLG-inspired problem domains) offer a constructive platform for the
+planning community to showcase their techniques to a wider audience---and
+to improve the quality of their tools for real-world planning tasks.
+
+%The NLG community's recent interest in planning presents a valuable
+%opportunity for planning researchers: provided some of the challenges
+%we have highlighted can be addressed, projects like GIVE (and other
+%NLG-inspired problem domains) offer a constructive platform for the
+%planning community to showcase their techniques to a wider
+%audience---and to improve the quality of their tools for real-world
+%planning tasks.  Furthermore, GIVE highlights planning-related
+%challenges such as plan execution monitoring and plan presentation
+%and summarisation and offers a platform on which such technologies can
+%be evaluated in experiments with human users.
 

### 2008-07-25 trunk/Papers/experiences/discussion.tex
small changes

@@
 at. This is particularly true for FF in the sentence generation
-domain, which generates nontrivial 14-word sentences in about two
-seconds. There is still room for improvement, however; special-purpose
-algorithms generate much larger referring expressions in milliseconds
-\cite{AreKolStr08}.  Although search efficiency is not the main
-concern in these domains, it nevertheless becomes a factor in
+domain, which generates nontrivial 14-word sentences in about two seconds.
+There is still room for improvement, however: special-purpose
+algorithms generate much larger referring expressions in
+milliseconds \cite{AreKolStr08}.  Although search efficiency is not the
+main concern in these domains, it nevertheless becomes a factor in
 scenarios like Experiment~2, where the world is almost completely
@@
 during the search---this approach is less effective in the domains
-we've considered. As our experiments have shown, there are natural
+we've considered. As our experiments show, there are natural
 planning domains in which relatively short plans must be computed in
@@
 successful than their predecessors. However, the common strategy of
-grounding out sets of predicates and actions in implementations of
-these algorithms has a very pronounced effect on domains such as those
-described above. Reachability analysis has the potential to help---we
-could gain a lot by only computing reachable instances---, but this
-too is typically done on the grounded problem. \todo{check if this is
-  true} Since our domains are not that unusual in their structure and
-composition, we hope that the lessons learnt from our experiences can
-help improve the performance of current systems. We believe that such
-improvements are necessary if planning is to become a more mature
-technology that can offer tools to a wider community of users. At the
-end of the day, real-world users will care about the \emph{total}
-runtime of a planner, and this is more than just the search time.
+grounding out sets of predicates and actions in implementations of these
+algorithms has a very pronounced effect on domains such as those described
+above. More sophisticated techniques based on reachability analysis may
+have the potential to help improve this situation, provided such
+approaches can avoid the drawbacks resulting from unnecessary grounding.
+Since our domains are not that unusual in their structure and composition,
+we hope that the lessons learnt from our experiences can help improve the
+performance of current systems.  We believe that such improvements are
+necessary if planning is to become a more mature technology that can offer
+tools to a wider community of users.  At the end of the day, real-world
+users will care about the \emph{total} runtime of a planner, and this is
+more than just the search time.
+%has the potential to help---we could gain a lot by only computing
+%reachable instances---, but this too is typically done on the grounded
+%problem.  \todo{check if this is true}
 

### 2008-07-25 trunk/Papers/experiences/domains.tex
small changes

@@
 \begin{enumerate}
-\item $\mathsf{sleeps}(\mathsf{root}, r_1)$
-\item $\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}), r_1)$
-\item $\mathsf{white}(\mathsf{subj}(\mathsf{root}), r_1)$
+\item $\mathsf{sleeps}(\mathsf{root}, r_1)$,
+\item $\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}), r_1)$,
+\item $\mathsf{white}(\mathsf{subj}(\mathsf{root}), r_1)$.
 \end{enumerate}
@@
 further buttons.  To simplify both the planning and the NLG task, the
-world is discretized into tiles of equal size; the user can turn by 90
+world is discretised into tiles of equal size; the user can turn by 90
 degree steps in either direction, and can move from the centre of one
@@
 The plans generated in the GIVE domain are often nontrivial. For instance,
-in the example above the shortest plan consists of 108 action steps; the
+in the above example the shortest plan consists of 108 action steps; the
 first few steps are as follows:
@@
 \begin{enumerate}
-\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north}, \mathsf{west})$
-\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$
-\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$
+\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north},
+\mathsf{west})$,
+\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
+\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$,
+\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
 \end{enumerate}
@@
 major problem is that the natural language instruction sequences can't
-simply be obtained by verbalising each action instance in a plan in
-turn.  For instance, the NLG system could generate an instruction
-sequence starting with ``turn left; walk forward; press the button;
-turn right; walk forward; walk forward; turn right; walk forward; walk
-forward; turn left; walk forward''.  However, this would be clumsy and
-boring, and it is to be expected that frustrated users will quickly
-cancel an interaction with such a system, resulting in a bad
-evaluation score.  Having just pressed the button, it would be much
-better to say ``turn around and walk through the door''. That is,
-multiple action instances should be merged into a single instruction;
-plans must be \emph{summarised} in order to present them to the
-user. Conversely, it may also be necessary to express a single
-planning step with several instructions.  Having just entered the top
-left room, it may be easier for the user to understand the instruction
-``walk to the centre of the room; turn right; now press the green
-button in front of you'' than the instruction ``press the green button
-on the wall to your right''.  To plan the referring expression ``the
-green button in front of you'' at a time when the button is not yet in
-front of the user, the NLG system must keep track of the hypothetical
-changes in what objects are visible to the user.  Thus the acts of
-referring and instructing, and the modules for planning and NLG, must
-be tightly integrated.
+simply be obtained by verbalising each action instance in a plan in turn.
+For instance, the NLG system could generate an instruction sequence
+starting with ``turn left; walk forward; press the button; turn right; walk
+forward; walk forward; turn right; walk forward; walk forward; turn left;
+walk forward''.  However, such directions would be clumsy and boring, and
+it is to be expected that frustrated users will quickly cancel an
+interaction with such a system, resulting in a bad evaluation score.
+Having just pressed the button, it would be much better to say ``turn
+around and walk through the door''. That is, plans must be
+\emph{summarised} by merging multiple action instances into a single
+instruction, in order to present them to the user. Conversely, it may also
+be necessary to \emph{elaborate} on a single planning step by expressing it
+with several instructions.  Having just entered the top left room, it may
+be easier for the user to understand the instruction ``walk to the centre
+of the room; turn right; now press the green button in front of you'',
+rather than the instruction ``press the green button on the wall to your
+right''.  To plan the referring expression ``the green button in front of
+you'' at a time when the button is not yet in front of the user, the NLG
+system must keep track of the hypothetical changes in what objects are
+visible to the user.  Thus the acts of referring and instructing, and the
+modules for planning and NLG, must be tightly integrated.
 

### 2008-07-25 trunk/Papers/experiences/experiences.tex
added abstract
spellcheck
small changes

@@
 overall planning time is dominated by the grounding step that these
-planners perform (rather than search). This has a pronounced effect on our
-domains which require relatively short plans in large universes.  We share
-our experiences and offer these domains as challenges for the planning
-community.
+planners perform, rather than search. This has a pronounced effect on our
+domains which require relatively short plans but have large universes. We
+share our experiences and offer these domains as challenges for the
+planning community.
 \end{abstract}

### 2008-07-25 trunk/Papers/experiences/introduction.tex
small changes

@@
 
-We then discuss some of our experiences using off-the-shelf planners
-in our two domains. In particular, we explore the efficiency of FF
+We then discuss some of our experiences using off-the-shelf planners in our
+two domains. In particular, we explore the efficiency of FF
 \cite{HoffmannNebel01} and SGPLAN
 \cite{hsu06:_new_featur_in_sgplan_for}---planners that are easily
-available, support expressive preconditions and effects out of the
-box, and have been successful on traditional benchmarks, and in the
-International Planning Competition (IPC)---on a range of problem
-instances from our planning domains.  We also compare these results
-against an ad-hoc Java implementation of GraphPlan \cite{Blum1997}.
-Our findings are mixed. On the one hand, it turns out that modern
-planners handle some of the search problems that arise in NLG quite
-easily (although large problem instances still remain a challenge).
-On the other hand, these same planners spend tremendous amounts of
-time on preprocessing. For instance, FF spends 90\% of its runtime in
-the sentence generation domain on grounding out literals and actions,
-and ends up slower than our version of GraphPlan that avoids
-grounding. For domains like ours, which are dominated by the number of
-actions and the universe size, rather than the combinatorics of the
-search problem, this observation suggests that the overall runtime of
-a planner could be improved by grounding more selectively. We
-therefore offer these domains as challenges for the planning
-community.
+available, support an expressive subset of PDDL for encoding preconditions
+and effects, and have been successful on traditional benchmarks and in the
+International Planning Competition (IPC)---on a range of problem instances
+from our planning domains.  We also compare these results against an ad-hoc
+Java implementation of GraphPlan \cite{Blum1997}.  Our findings are mixed.
+On the one hand, it turns out that modern planners handle some of the
+search problems that arise in NLG quite easily (although large problem
+instances still remain a challenge).  On the other hand, these same
+planners spend tremendous amounts of time on preprocessing. For instance,
+FF spends 90\% of its runtime in the sentence generation domain on
+grounding out literals and actions, and ends up slower than our version of
+GraphPlan that avoids grounding. For domains like ours, which are dominated
+by the number of actions and the universe size, rather than the
+combinatorics of the search problem, this observation suggests that the
+overall runtime of a planner could be improved by grounding more
+selectively. We therefore offer these domains as challenges for the
+planning community.
+%available, support expressive preconditions and effects out of the box,
 

### 2008-07-25 trunk/Papers/experiences/domains.tex
changes to conclusions; added new footnote in domains section

@@
 \emph{summarised} by merging multiple action instances into a single
-instruction, in order to present them to the user. Conversely, it may also
-be necessary to \emph{elaborate} on a single planning step by expressing it
-with several instructions.  Having just entered the top left room, it may
-be easier for the user to understand the instruction ``walk to the centre
-of the room; turn right; now press the green button in front of you'',
-rather than the instruction ``press the green button on the wall to your
-right''.  To plan the referring expression ``the green button in front of
-you'' at a time when the button is not yet in front of the user, the NLG
-system must keep track of the hypothetical changes in what objects are
-visible to the user.  Thus the acts of referring and instructing, and the
-modules for planning and NLG, must be tightly integrated.
+instruction, in order to present them to the
+user.\footnote{This task is similar to the construction of ``macro''
+actions, but is primarily intended for plan presentation.}
+Conversely, it may also be necessary to \emph{elaborate} on a single
+planning step by expressing it with several instructions.  Having just
+entered the top left room, it may be easier for the user to understand the
+instruction ``walk to the centre of the room; turn right; now press the
+green button in front of you'', rather than the instruction ``press the
+green button on the wall to your right''.  To plan the referring expression
+``the green button in front of you'' at a time when the button is not yet
+in front of the user, the NLG system must keep track of the hypothetical
+changes in what objects are visible to the user.  Thus the acts of
+referring and instructing, and the modules for planning and NLG, must be
+tightly integrated.
 

### 2008-09-04 trunk/Papers/rmrs-08/motivation.tex
went over section 3

@@
 
+\includegraphics[width=6cm]{pic-cat-chased-dog}
+
 Put trees in here
@@
 \end{examples}
+
+\todo{The syntax used in this RMRS is incompatible with the later
+  definition.}
+
 Semantic relations, sense tags and the arity of the base-language

### 2008-09-04 trunk/Papers/rmrs-08/rmrs.tex
went over section 3

--- 2008-05-20 trunk/Papers/rmrs-08/rmrs.tex
@@
 
-Define syntax of RMRS; model theory of RMRS; solved forms.  Explain
-the role of the RMRS models (= trees) in the semantic construction
-pipeline, in particular with respect to decoding trees into semantic
-representations of the base language formalism.  Present examples in
-MRS syntax first, then in our RMRS language.  Finally, compare our
-version of RMRS against MRS and dominance constraints.
+We will now make the basic ideas from Section \ref{sec:motivation}
+more precise.  We will first define the syntax of the \rmrs\ language;
+this is a notational variant of earlier definitions in the
+literature.  We will then define a model theory for our version of
+\rmrs, and conclude this section by carrying over the notion of
+\emph{solved forms} from CLLS \cite{egg:etal:2001}.
+
+\subsection{RMRS Syntax}
+
+We define the syntax of \rmrs\ in the style of CLLS
+\cite{egg:etal:2001}.  We assume an infinite set of {\em node
+  variables} $\nvar=\{X,Y,X_1,\ldots\}$, which we will use as labels,
+anchors, and holes; the distinction between these uses will come from
+their position in the formula.  Furthermore, we assume an infinite set
+of \emph{base variables} $\ovar$, consisting of individual variables
+$\{x,x_1,y,\ldots\}$ and event variables $\{e_1,\ldots\}$.  Finally,
+we assume a vocabulary of \emph{predicate symbols}
+$\pred=\{P,Q,P_1,\ldots\}$, each of which has an \emph{arity} $\ar(P)
+\geq 0$.
+
+We can now define formulas of \rmrs\ as follows.
+
+\begin{definition}\label{defn:rmrs-syntax}
+  An \emph{\rmrs} is a finite set $\varphi$ of atoms of one of the
+  following forms; $S \subseteq \N$ is a set of numbers that is either
+  finite or $\N$ itself.
+
+$\begin{array}{rrll}
+A &::= &\ep{X}{Y}{P} \\ %& \mbox{labelled elementary predication} \\
+&| &\ARG_S(X,v) \\ % & \mbox{ARG relation for base variables} \\
+&| &\ARG_S(X,Y) \\ % & \mbox{ARG relation for node variables} \\
+&| &X \qeq Y \\ % & \mbox{QEQ} \\
+&| &v_1 = v_2 \;|\; v_1 \neq v_2 \\ % & \mbox{base variable
+                                % (in)equality} \\
+&| &X = Y \;|\; X \neq Y \\ % & \mbox{node variable (in)equality} \\
+&| &P \spec Q % & \mbox{SPEC relation between predicate symbols}
+\end{array}
+$
+
+A node variable $X$ is called a \emph{label} in the \rmrs\
+$\varphi$ iff $\varphi$ contains an atom of the form $\ARG_S(X,i)$ or
+an elementary predication of the form $\ep{Y}{X}{P}$; $X$ is called an
+\emph{anchor} iff $\varphi$ contains an atom of the form
+$\ep{X}{Y}{P}$ or $Y \qeq X$; and $X$ is called a \emph{hole} iff
+$\varphi$ contains an atom of the form $\ARG_S(Y,X)$ or $X \qeq Y$.
+
+An \rmrs\ $\varphi$ is called \emph{normal} iff it satisfies the
+following conditions:
+\begin{enumerate}
+\item Every node variable in $\varphi$ is either a label, an anchor,
+  or a hole, but not two of these.
+\item If $\ep{X}{Y}{P}, \ep{Z}{Y}{Q}\in \varphi$, then $X=Z$ and $P=Q$
+  (i.e., each predication has a unique anchor).
+\end{enumerate}
+\end{definition}
+
+\rmrs\ as presented here combines similarities to earlier definitions
+of \rmrs\ (e.g., \cite{copestake:2003}) and to dominance constraints
+\cite{egg:etal:2001}.  Compared to earlier versions of \rmrs, we write
+$\ARG_{\{i\}}$ (with a singleton set $S$) instead of $\ARG_i$, and
+$\ARG_\N$ instead of $\ARG_n$.  Furthermore, we can express the
+elementary predication $\ep{l}{a}{P(v)}$ with the atoms
+$\{\ep{a}{l}{P}, \ARG_{\{0\}}(l,v)\}$; note that the positions of the
+anchor and label have changed.  In other words, the syntax defined
+here differs in some details, but is more general. The primary
+differences to dominance constraints are the distinction of labels and
+anchors to support EP conjunctions directly, the separation of
+atoms for specifying the labelling and the parent-child relations, and
+the use of $\qeq$ instead of dominance.  One could thus see our
+version of \rmrs\ as a redefinition of classical \rmrs\ in the
+dominance constraints style, which will simplify the formal work
+below.
+
+\todo{don't split stuff up into tuple parts}
+
+\subsection{Model Theory}
+
+The model theory formalises the relationship between an {\sc rmrs} and
+the fully specific, alternative logical forms that it describes,
+expressed in the base language.  We represent such a logical form as a
+tree $\tau$.  The models of the \rmrs\ will then be such trees, and we
+can define notions such as satisfaction and entailment over these
+models in the usual way.  Satisfaction corresponds to the relation
+between an \rmrs\ and a fully specific semantic representation that it
+describes, whereas entailment corresponds to the relation between
+a more and a less specific \rmrs.
+
+In this paper, we assume for simplicity that the base language is
+predicate logic; in this case, $\tau$ is simply the structure tree of
+the predicate logic formula.  Other base languages may require
+different mappings between formulas and trees.  We assume that
+$\Sigma$ is a ranked signature consisting of the symbols of the base
+language with appropriate arities.  To encode predicate logic, we
+assume the unary constructor $\neg$ and binary constructors $\wedge$,
+$\rightarrow$, etc.; the binary constructors $\forall x$ and $\exists
+x$ for all base-language variables $x$; all the constants and
+base-language variables as nullary constructors; and, for each
+predicate symbol $P$ of arity $n$, a constructor $P$ of arity $n$.
+For other base languages, a different signature could be chosen; we
+only assume for convenience that the signature contains a binary
+constructor $\wedge$ to represent conjunction.  We write $\Sigma_i$
+for the set of all constructors in $\Sigma$ that have arity $i$.
+
+The models of {\sc rmrs} are then defined to be finite constructor
+trees:
+\begin{definition}\label{defn:models}
+  A {\em finite constructor tree} $\tau$ is a function $\tau:D
+  \rightarrow \Sigma$ such that $D$ is a \emph{tree domain} (i.e., a
+  subset of $\N^*$ which is closed under prefix and left sibling) and
+  the number of children of each node $u \in D$ is equal to the arity
+  of $\tau(u)$.  
+\end{definition}
+
+We write $D(\tau)$ for the tree domain of a constructor tree $\tau$,
+and further define the following relations between nodes in a finite
+constructor tree:
+
+\begin{definition}\label{defn:dominance}
+$u \qeqdom v$ iff $v = u \cdot 2^k$ for some $k \geq 0$ and all
+  symbols $\tau(u), \tau(u \cdot 2), \ldots, \tau(u \cdot 2^{k-1})$
+  are quantifiers.  (This is the ``qeq'' relation familiar from the
+  \mrs\ and \rmrs\ literature.)
+
+  $u \wedgedom v$ iff $u \leq v$ (i.e.\ it is a prefix), and all
+  symbols on the path from $u$ to $v$ (not including $v$) are
+  $\wedge$.
+\end{definition}
+
+The {\em satisfaction} relation between an {\sc rmrs} $\varphi$ and a
+finite constructor tree $\tau$ is defined in terms of several
+assignment functions.  First, we assume a node variable assignment
+function $\alpha: \nvar \rightarrow D(\tau)$ that maps the node
+variables in an \rmrs\ to the nodes of $\tau$.  Second, we assume a
+base language assignment function $g:\ovar \rightarrow \Sigma_0$ that
+maps the base variables in the \rmrs\ to nullary constructors
+representing variables in the base language.  This function enables us
+to map two variables that seem different in the \rmrs\ to the same
+base-language variable, introducing e.g.\ semantic dependencies that a
+POS tagger didn't catch.  For instance, $g$ might map $x_4$ and $x_5$
+in (\ref{eg:cat-pos}b) to $y$ to specify the restrictor of the
+quantifier.  Finally, we assume a relation $\sigma \subseteq \pred
+\times \Sigma_{\geq 1}$ that maps each {\sc rmrs} predicate symbol to
+a set of constructors from $\Sigma$.  This relation allows us to
+underspecify lexical ambiguities; for instance,
+$\sigma(\mathrm{\_cat\_n})$ might contain the symbol $\mathrm{\_cat\_n\_1}$
+in (\ref{eg:cat}c) along with symbols for other word senses.
+
+\begin{definition}\label{defn:satisfaction}
+Satisfaction of atoms is defined as follows:
+$$
+\begin{array}{l}
+  \tau, \alpha, g, \sigma \models  \ep{X}{Y}{P}
+\mbox{ iff}\\
+\qquad \tau(\alpha(Y)) \in \sigma(P) \mbox{ and } \alpha(X) \wedgedom
+  \alpha(Y) \\
+  \tau, \alpha, g, \sigma \models \ARG_S(X,a)
+\mbox{ iff exists  $i \in S$ s.t.}\\
+\qquad  \alpha(X) \cdot i \in D(\tau) \mbox{ and }
+  \tau(\alpha(X) \cdot 
+  i) = g(a) \\
+  \tau, \alpha, g, \sigma \models \ARG_S(X,Y)
+\mbox{ iff  exists $i \in S$ s.t.}\\
+\qquad \alpha(X) \cdot i \in D(\tau), \alpha(X) \cdot
+  i = \alpha(Y) \mbox{ and }\\
+\qquad
+s(\alpha(X) \cdot i) = h \\
+  \tau, \alpha, g, \sigma \models X \qeq Y
+\mbox{ iff } \alpha(X) \qeqdom \alpha(Y) \\
+  \tau, \alpha, g, \sigma \models X ={/}\neq Y
+\mbox{ iff } \\
+\qquad \alpha(X) ={/}\neq \alpha(Y) \\
+  \tau, \alpha, g, \sigma \models v_1 ={/}\neq v_2
+\mbox{ iff}\\
+\qquad g(v_1) ={/}\neq g(v_2) \\
+  \tau, \alpha, g, \sigma \models P \spec Q
+\mbox{iff $\sigma(P) \subseteq \sigma(Q)$}
+\end{array}
+$$
+A 4-tuple $\tau,\alpha,g,\sigma$ satisfies an \rmrs\ $\varphi$
+(written $\tau,\alpha,g,\sigma \models \varphi$) iff it satisfies all
+of its elements.
+\end{definition}
+
+Notice that one \rmrs\ may be satisfied by multiple trees; we can take
+the \rmrs\ to be a partial description of each of these trees.  In
+particular, \rmrs s may represent semantic scope ambiguities, missing
+information about semantic dependencies, and missing information about
+lexical subcategorisation and lexical senses.  For $j=\{1,2\}$,
+suppose that $\tau_j,\alpha_j,g_j,\sigma\models \varphi$ and let
+$\nvar$ be the node variables from which $\varphi$ is built.  Then
+$\varphi$ exhibits a semantic scope ambiguity if there are variables
+$Y,Y'\in \nvar$ such that $\alpha_1(Y) \qeqdom \alpha_1(Y')$ and
+$\alpha_2(Y')\qeqdom \alpha_2(Y)$.  It exhibits missing information
+about semantic dependencies if there are base-language variables $
+v,v'\in \ovar$ such that $g_1(v)=g_1(v')$ and $g_2(v)\neq g_2(v')$.
+It exhibits missing lexical subcategorisation information if there is
+a variable $ Y \in \nvar$ such that $\tau_1(\alpha_1(Y))$ is a
+constructor of a different type from $\tau_2(\alpha_2(Y))$ \todo{what
+  does this mean?}.  And it exhibits missing lexical sense information
+if $\tau_1(\alpha_1(Y))$ and $\tau_2(\alpha_2(Y))$ are different
+base-language constructors, but of the same type.
+
+Let's look at the \rmrs\ in (\ref{eg:cat-pos}b) by way of example.
+This \rmrs\ is normal, and it is of a particularly simple form because
+it consists only of elementary predications.  The \rmrs\ is satisfied
+(among others) by the tree in Fig.~\ref{fig:1} together with some
+particular $\alpha$, $g$, and $\sigma$.  The assignment $\alpha$ will
+map the variables $l_1$ and $a_1$ to the root node; $l_{41}$ and
+$l_{42}$ to its second child (labeled with ``$\wedge$''); $a_{41}$ to
+the first child of \emph{that} node (i.e.\ the node 21, labelled with
+``fat'') and $a_{42}$ to the node 22; and so forth.  $g$ will map $x$
+and $x_3$ to $x$, $e'$ to $e$, and so on.  And $\sigma$ will map each
+\rmrs\ predicate symbol (which represents a word) to the set of its
+word senses, e.g.\ \_cat\_n to a set containing \_cat\_n\_1 and
+possibly others.  It is then easy to verify that every single atom in
+the \rmrs\ is satisfied -- most interestingly, the EPs
+$\ep{l_{41}}{a_{41}}{\_fat\_a(e')}$ and
+$\ep{l_{42}}{a_{42}}{\_cat\_n(x_3)}$ are satisfied because
+$\alpha(l_{41}) \wedgedom \alpha(a_{41})$ and $\alpha(l_{42})
+\wedgedom \alpha(a_{42})$.
+
+Truth, validity and entailment can now be defined in terms of
+satisfiability in the usual way:
+\begin{definition}\label{defn:entailment}
+\begin{description}
+\item   [truth:] $\tau\models \varphi$ iff $\exists \alpha,g,\sigma$  such
+  that $\tau,\alpha,g,\sigma\models \varphi$
+\item   [validity:] $\models \varphi$ iff $\forall \tau$, $\tau\models \varphi$.
+\item   [entailment:] $\varphi\models \varphi'$ iff $\forall \tau$, if
+  $\tau\models \varphi$ then $\tau\models \varphi'$.
+\end{description}
+\end{definition}
+% NOTE: This definition of entailment isn't right, because it allows
+% the assignment $\sigma$ from predicates to constructors to vary
+% when testing the satisfiability of $\varphi$ vs. $\varphi'$ with
+% respect to $\tau$.  I think that to get the notion of entailment we
+% want, $\sigma$ should actually be part of the {\sc rmrs} *model*.
+% In other words, an {\sc rmrs} model $M$ is a tuple $\langle \tau,
+% \sigma\rangle$, where $\tau$ is a finte constructor tree and
+% $\sigma$ is a mapping from $\Pred$ to the power set of $\Sigma$.
+% Then truth and entailment are defined correctly.  I believe our
+% proofs of things like Theorem~\ref{thm:big-one} actually assumes
+% that $\sigma$ is constant when interpreting $\varphi$ vs.\
+% $\varphi'$, no?
+
+\subsection{Solved Forms}
+
+One feature which the above definition of \rmrs\ shares with dominance
+constraints is that any satisfiable \rmrs\ has an infinite number of
+models which only differ in the areas that the \rmrs\ didn't ``talk
+about''.  An example is shown in Fig.~\ref{fig:fat-black-cat}.  This
+tree, which represents the meaning of ``the fat black cat chased the
+dog'', differs from the model in Fig.~\ref{fig:1} in the subtree below
+the restrictor of the universal quantifier.  It is still a model of
+(\ref{eg:cat-pos}b) -- all the EPs are satisfied.  However, there is
+more ``semantic material'' (the ``black'' subformula) in the space
+between the labels $l_{41}$ and $l_{42}$ and the anchors $a_{41}$ and
+$a_{42}$.  The \rmrs\ didn't specify that this material should be
+there; but it also didn't specify that it can't be there, and indeed
+there is no way in which the \rmrs\ can enforce that no extra material
+can be added.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=6cm]{pic-more-stuff}
+  \caption{Another tree which satisfies (\ref{eg:cat-pos}b).}
+  \label{fig:fat-black-cat}
+\end{figure}
+
+In the context of robust semantic processing, this is a desirable
+feature of \rmrs, because it means that when we enrich an \rmrs\
+obtained from a shallow processor with more semantic information, we
+don't change the set of models; we only \emph{restrict} the set of
+models further and further towards the semantic representation we are
+trying to reconstruct.  However, the fact that all satisfiable \rmrs s
+have an infinite number of models is a problem for algorithms that try
+to decide satisfiability or compute the fully specific semantic
+representations -- how can this be done in finite time?
+
+In this paper, we follow the dominance constraint tradition in
+defining a fragment of \rmrs s called \emph{solved forms}.  All \rmrs
+s in solved form are guaranteed to be satisfiable, and thus each
+solved form represents an infinite class of models.  However, we will
+set it up in such a way that each satisfiable \rmrs\ has only a finite
+number of solved forms which partition the space of possible models
+into classes with only ``irrelevant'' differences as above.  A solver
+can then enumerate the solved forms rather than all models.
+
+Intuitively, we expect that an \rmrs\ in solved form is fully
+specified with respect to scope ambiguities and the predicate-argument
+structure, all variable equalities and inequalities have been
+resolved, and only contains lexical ambiguities that are ``harmless'',
+inthat any choice of a constructor for the predicate symbol will
+satisfy the \rmrs.  This is made precise below.
+
+\todo{Note: We don't really talk about SPEC anywhere else.}
+
+\begin{definition}\label{defn:solved-forms}
+  An RMRS $\varphi$ is \emph{in solved form} iff:
+  \begin{enumerate}
+  \item every variable in $\varphi$ is either a hole, a label or an
+    anchor (but not two of these);
+  \item $\varphi$ doesn't contain equality, inequality, and SPEC
+    ($\sqsubseteq$) atoms;
+  \item if $\ARG_S(Y,i)$ is in $\varphi$, then $|S| = 1$;
+  \item for any label $Y$ and index set $S$, there
+    are no two atoms $\ARG_S(Y,i)$ and $\ARG_S(Y,i')$ in $\varphi$;
+  \item if $Y$ is an anchor in some EP $\ep{X}{Y}{P}$ and $k$ is the
+    maximum number such that $\ARG_{\{k\}}(X,i)$ is in $\varphi$ for
+    any $i$, then there is a constructor $p \in \sigma(P)$ whose arity
+    is at least $k$;
+  \item no label occurs on the right-hand side of two
+    different $\qeqdom$ atoms.
+  \end{enumerate}
+\end{definition}
+
+All these conditions together ensure that it is possible to ``read
+off'' a model from the solved form.
+
+\begin{prop}
+  Every RMRS in solved form is satisfiable.
+\end{prop}
+\begin{proof}[Proof (sketch; see also
+  \cite{Duchier00dominanceconstraints})] 
+  For each EP, we choose to label the anchor with the constructor $p$
+  of sufficiently high arity whose existence we assumed; we determine
+  the edges between an anchor and its children from the uniquely
+  determined $\ARG$ atoms; plugging labels into holes is
+  straightforward because no label is qeq-dominated by more than one
+  label; and then we can fill the spaces between the labels and
+  anchors with conjunctions.
+\end{proof}
+
+\begin{definition}  \label{defn:solved-form-of}
+  The \emph{syntactic dominance relation} in an {\sc  rmrs} $\varphi$ is the
+  reflexive, transitive closure of the binary relation $$
+\begin{array}{r c l}
+\{(X,Y) & \mid &  \mbox{$\varphi$ contains $X \qeqdom Y$ or}\\
+&& \mbox{$ARG_S(X,Y)$ for some
+    $S$}\}
+\end{array}
+$$  
+We write $D(\varphi)$ for the syntactic dominance
+  relation of $\varphi$.
+
+  An RMRS $\varphi'$ is \emph{a solved form of} the RMRS $\varphi$
+  iff $\varphi'$ is in solved form and there is a substitution $s$
+  that maps the  node and base variables of $\varphi$ to the node and
+  base variables of $\varphi'$ such that
+  \begin{enumerate}
+  \item $\varphi'$ contains the EP $\ep{X'}{Y'}{P}$ iff there are
+    variables $X,Y$ such that $\ep{X}{Y}{P}$ is in $\varphi$, $X' =
+    s(X)$, and $Y' = s(Y)$;
+  \item for every atom $\ARG_S(X,i)$ in $\varphi$, there is exactly
+    one atom $\ARG_{S'}(X',i')$ in $\varphi$ with $X' = s(X)$, $i' =
+    s(i)$, and $S' \subseteq S$;
+  \item $D(\varphi') \supseteq s(D(\varphi))$.
+  \end{enumerate}
+\end{definition}
+
+\begin{prop}
+  For every tuple $(\tau,\alpha,g,\sigma)$ that satisfies some \rmrs\
+  $\varphi$, there is a solved form $\varphi'$ of $\varphi$ such that
+  $(\tau,\alpha,g,\sigma)$ also satisfies $\varphi'$.
+\end{prop}
+\begin{proof}
+  We construct the substitution $s$ from $\alpha$ and $g$.  Then we
+  add all qeqs that are satisfied by $\alpha$ and restrict the $\ARG$
+  atoms to those child indices that are actually used in $\tau$.  The
+  result is in solved form because $\tau$ is a tree; it is a solved
+  form of $\varphi$ by construction.
+\end{proof}
+
+\todo{At this point we should have some examples of solved forms.  The
+  running example is a bit lame here because it already is in solved
+  form, and therefore only has one solved form.}
+
+\begin{prop}
+  Every {\sc rmrs} $\varphi$ has only a finite number of solved forms, up to
+  renaming of variables.
+\end{prop}
+\begin{proof}
+  Up to renaming of variables, there is only a finite number of
+  substitutions on the node and base variables of $\varphi$.  Let $s$
+  be such a substitution.  This fixes the set of EPs of any solved
+  form of $\varphi$ that is based on $s$ uniquely.  There is only a
+  finite set of choices for the subsets $S'$ in condition 2 of
+  Def.~\ref{defn:solved-form-of}, and there is only a finite set of
+  choices of new $\qeqdom$ atoms that satisfy condition 3.  Therefore,
+  the set of solved forms of $\varphi$ is finite.
+\end{proof}
+
+\subsection{Extensions}
+
+In the earlier \rmrs\ literature, each term of an \rmrs\ is equipped
+with a \emph{sort}.  For instance, individual variables are typically
+assigned the sort $x$, event variables the sort $e$, and holes the
+4sort $h$; sometimes subsorts of these are considered (e.g.\ $e_{PAST}$
+for events in the past tense), and all the sorts arranged in a sort
+hierarchy ${\cal S}$.
+
+While we have defined \rmrs\ without sorts above to simplify the
+presentation, it is straightforward to add sorts to the definition.
+For this, one assumes that the signature $\Sigma$ is sorted, i.e.\
+assigns a sort $s_1\times\ldots s_n\rightarrow s$ to each constructor,
+where $n$ is the constructor's arity (possibly zero) and
+$s,s_1,\ldots,s_n\in {\cal S}$ are atomic sorts.  We can then
+restrict the models of \rmrs\ to trees that are \emph{well-sorted} in
+the usual sense, i.e.\ those in which we can infer a sort for each
+subtree, and require that the variable assignment functions likewise
+respect the sorts.  If we then modify Def.~\ref{defn:solved-forms}
+such that the constructor $p$ of sufficiently high arity is also
+consistent with the sorts of the known arguments -- i.e., if $p$ has
+sort $s_1 \times \ldots \times s_n \rightarrow s$ and the \rmrs\
+contains an atom $\ARG_{\{k\}}(Y,i)$ and $i$ is of sort $s'$, then
+$s'$ is a subsort of $s_k$ --, all the above propositions about solved
+forms remain true.
+
+Another extension that is worth discussing is the idea of replacing
+the qeq atom, $X \qeqdom Y$, by ordinary dominance $X \dom Y$ in the
+sense of dominance constraints, which is satisfied iff there is a path
+from $\alpha(X)$ to $\alpha(Y)$ in the tree.  Dominance is clearly a
+less restrictive condition than qeq, but it has been shown that the
+distinction between the two relations is irrelevant in \mrs s that were
+derived from a corpus by a deep grammar \cite{FucKolNieTha04}.  In our
+above definition of \rmrs, it is possible to replace qeq by dominance;
+this changes \rmrs\ to a different logic (let's call it
+Dominance-\rmrs), for which however all the above propositions about
+solved forms still hold if ``qeq'' is systematically replaced by
+``dominance''.
 

### 2008-09-05 trunk/Papers/rmrs-08/abstract.tex
language more concise and partial conclusion added

--- 2008-05-20 trunk/Papers/rmrs-08/abstract.tex
@@
 \begin{abstract}
-  la la la
+  One way to construct semantic
+  representations in a robust manner is to enhance shallow language
+  processors with semantic components.  Here, we provide a model
+  theory for a 
+  semantic formalism 
+  that is designed for this,
+  namely Robust Minimal Recursion Semantics ({\sc rmrs},
+  \cite{copestake:2003}).  We show that {\sc rmrs} supports a notion
+  of entailment 
+  that allows it
+  to form the basis for comparing the semantic output of
+  different parses of varying depth.
 \end{abstract}

### 2008-09-05 trunk/Papers/rmrs-08/conclusion.tex
first dummy version
language more concise and partial conclusion added

--- 2008-05-20 trunk/Papers/rmrs-08/abstract.tex
@@
-\begin{abstract}
-  la la la
-\end{abstract}
+\section{Conclusion}
+\label{sec:conclusion}
+
+In this paper we have motivated and described \rmrs---a semantic
+framework that can be used to specify
+semantic components to robust shallow parsers.  Accordingly, \rmrs\
+is designed to be
+maximally flexible on the type of semantic information
+that can be left underspecified, so that the semantic output of a
+shallow parser needn't over-determine or under-determine the semantic
+information that can be extracted from the shallow syntactic analysis.
+
+We defined its syntax and model-theoretic semantics, and following the
+tradition from dominance constraints, we defined a syntactic criterion
+for characterising entailment.  The syntactic notion of {\em solved
+  forms} is useful because it means that even when an underspecified
+\rmrs\ corresonds to an infinite number of possible fully-specific
+interpretations in context, there is a finite way of identifying all
+the `relevant' differences among those fully-specific interpretations.
+We showed that the model theory supports theorems that allow one to
+use \rmrs\ representations as a way of comparing and even integrating
+the semantic output of different parses of the same natural language
+string, even when they vary in depth.
 

### 2008-09-05 trunk/Papers/rmrs-08/introduction.tex
language more concise and partial conclusion added

@@
 inference and model construction is vital for deeper language
-engineering tasks, such as dialogue systems.  Such detailed semantic
-representations can be obtained from hand-crafted deep grammars (e.g.,
-\cite{butt:etal:1999,copestake:flickinger:2000}), but this
+engineering tasks, such as dialogue systems.  Logical forms
+can be obtained from hand-crafted deep grammars
+\cite{butt:etal:1999,copestake:flickinger:2000}, but this
 tends to lack robustness: hand-crafted grammars fail to cover all
-words and linguistic constructions; and unedited text and speech
-contains ill-formed phrases, which by design deep grammars do not
-handle.
+words and linguistic constructions and by design they do not handle
+ill-formed phrases.
 
@@
 string-to-logical-form mappings, exploting informative features from
-the output of
 shallow language processors; and \cite{copestake:2003} and
-\cite{frank:2004}, among others,
+\cite{frank:2004}
 endow robust language processors with semantic components.
@@
 and long-distance dependencies (for instance, \cite{clark:etal:2004}
-report that the parser used in \cite{bos:etal:2004} yields semantic
-dependencies from the unbounded dependencies borne from object
-extraction---e.g., {\em the man that I met\dots}---with 63\%
-accuracy).  Forcing a shallow parser to make a decision about these
-phenomena therefore can be a major source of errors.  Depending on the
+report that the parser used in \cite{bos:etal:2004} yields 63\%
+accuracy on object
+extraction; e.g., {\em the man that I met\dots}).
+Forcing a shallow parser to make a decision about these
+phenomena can therefore be error-prone.  Depending on the
 application, it may be preferrable to give the parser the option to to
@@
 (shallow) syntactic analysis, and to complete the semantic information
-later, via inferences that exploit information extraneous to the
+later, using information extraneous to the
 parser.
@@
 \cite{copestake:2003}).  {\sc rmrs} is a generalisation of {\sc mrs}
-\cite{copestake:etal:2005} that is designed to support underspecification both
-of scope relations and of the predicate-argument structure.
+\cite{copestake:etal:2005} that is designed to support underspecification
+of scope and predicate-argument structure.
 \cite{copestake:2003,frank:2004} use it to
-support semantic constructions by shallow parsers
+specify semantic components to shallow parsers
 ranging in depth from {\sc pos} taggers to chunk parsers  
@@
 derived from deep grammars, such as the English Resource Grammar
-\cite{copestake:flickinger:2000} can also be seen as special cases of
-{\sc rmrs} representations.
+({\sc erg}, \cite{copestake:flickinger:2000}) are
+special cases of 
+{\sc rmrs}.
 
@@
 inference systems and to algorithms for efficiently computing the
-possible fully specific semantic representations described by the \rmrs
+possible fully specific semantic representations described by the \rmrs\
 output of a shallow parser.
 
-We will first illustrate the differences and challenges with deep and
-shallow semantic construction in an example in
-Section~\ref{sec:motivation}.  We will then define the syntax and
-model-theoretic semantics of \rmrs\ and introduce some important
-formal concepts in Section~\ref{sec:rmrs}.  Finally, we demonstrate
-the practical usefulness of our formalisation in
-Section~\ref{sec:entailment} by modelling compatibility of
-partial semantic representations as \rmrs\ entailment and providing a
+Section~\ref{sec:motivation} illustrates via example the differences
+and challenges with deep and 
+shallow semantic construction.  
+Section~\ref{sec:rmrs} defines the syntax and
+model-theory of \rmrs\.  In Section~\ref{sec:entailment} we
+demonstrate 
+the practical usefulness of our formalisation 
+by modelling compatibility of
+partial semantic representations as \rmrs\ entailment and by providing a
 syntactic criterion for characterising entailment.
@@
 \hidden{
-
 The language in which these partial representations are expressed

### 2008-09-05 trunk/Papers/rmrs-08/motivation.tex
language more concise and partial conclusion added

@@
 
-We motivate and describe the main features of {\sc rmrs} with a toy
+We motivate and describe {\sc rmrs} with a toy
 example.  Sentence 
@@
 doesn't resolve.    The {\sc mrs} (\ref{eg:cat}b) derived by the
-English Resource Grammar ({\sc erg}, \cite{copestake:flickinger:2000})
+{\sc erg} \cite{copestake:flickinger:2000}
 reflects this: (i) each predication is
 labelled ($l_1$, $l_4$, etc); (ii) scopal arguments are indicated by
-holes (e.g., $h_2$ and $h_3$ in the quantifier {\em every}); (iii)
+holes (e.g., $h_2$, $h_3$); (iii)
 constraints on scope are expressed by the qeq conditions (roughly, $h
@@
 constraints allow for more than one way of `plugging' the holes with
-labels.\footnote{The event variable in the predicate symbol
-  corresponding to the adjective is motivated by copulas---{\em The
+labels.\footnote{The event variable in $\_\mbox{{\em fat}}\_j\_1$ is 
+  for copulas. {\em The
     cat is fat} has logical form 
@@
 (\ref{eg:cat}b) also exhibits {\sc mrs}' compact description of
-conjunction through label sharing (see $l_4$ above), and the {\sc
-  erg}'s naming convention for predicate symbols: each predicate is
+conjunction through label sharing (see $l_4$), and the {\sc
+  erg}'s naming convention for predicate symbols: each lexical predicate is
 constructed from the word lemma, its syntactic category and a sense
-number, and the leading underscore identifies lexical predicates.
+number, and the leading underscore distinguishes lexical predicates
+from non-lexical ones.
 
@@
 
-The {\sc mrs} (\ref{eg:cat}c) is a notational variant of the {\sc
-  rmrs} (\ref{eg:cat}d). {\sc rmrs}s offer a more factorised
-representation: its predicate symbols are all unary and the other
-arguments to the (base-language) constructor that it corresponds to
+The {\sc mrs} (\ref{eg:cat}b) is a notational variant of the {\sc
+  rmrs} (\ref{eg:cat}c). {\sc rmrs}s offer a more factorised
+representation: each predicate symbol is unary and the other
+arguments to its corresonding base-language constructor
 are represented by separate binary relations on the unique {\em
   anchor} of the predicate symbol ($a_1, a_{41},\ldots$) together with
-variable and label equalities (e.g., $x=x_1$, $l_{41}=l_{42}$).
+variable equalities (e.g., $x=x_1$).
 \begin{examples}
@@
   pos} tagger: one can build its semantic component simply by deriving
-lexical predicate symbols from the word lemmas and their tags and ensuring
-that the predication for each word receives a unique label, anchor and
-argument, as given in (\ref{eg:cat-pos}):
+lexical predicate symbols from the word lemmas and tags and assigning each
+predication a unique label, anchor and
+argument, as in (\ref{eg:cat-pos}):
 \begin{examples}
@@
 \end{examples}
-
-\todo{The syntax used in this RMRS is incompatible with the later
-  definition.}
 

### 2008-09-05 trunk/Papers/rmrs-08/rmrs.tex
language more concise and partial conclusion added

@@
 \rmrs\ as presented here combines similarities to earlier definitions
-of \rmrs\ (e.g., \cite{copestake:2003}) and to dominance constraints
-\cite{egg:etal:2001}.  Compared to earlier versions of \rmrs, we write
+of \rmrs\ (e.g., \cite{copestake:2003}) as illustrated in
+(\ref{eg:cat}c) and
+to dominance constraints
+\cite{egg:etal:2001}.  We
 $\ARG_{\{i\}}$ (with a singleton set $S$) instead of $\ARG_i$, and
-$\ARG_\N$ instead of $\ARG_n$.  Furthermore, we can express the
-elementary predication $\ep{l}{a}{P(v)}$ with the atoms
-$\{\ep{a}{l}{P}, \ARG_{\{0\}}(l,v)\}$; note that the positions of the
-anchor and label have changed.  In other words, the syntax defined
+$\ARG_\N$ instead of $\ARG_n$.  Furthermore, the
+elementary predication $\ep{l}{a}{P(v)}$ is a notational variant of
+$\{\ep{l}{a}{P}, \ARG_{\{0\}}(l,v)\}$.
+In other words, the syntax defined
 here differs in some details, but is more general. The primary
 differences to dominance constraints are the distinction of labels and
-anchors to support EP conjunctions directly, the separation of
-atoms for specifying the labelling and the parent-child relations, and
-the use of $\qeq$ instead of dominance.  One could thus see our
-version of \rmrs\ as a redefinition of classical \rmrs\ in the
-dominance constraints style, which will simplify the formal work
+anchors that will separate the
+labelling and the parent-child relations, and
+the use of $\qeq$ instead of dominance.  Redefining
+\rmrs\ in the
+style of dominance constraints will simplify the formal work
 below.
@@
 expressed in the base language.  We represent such a logical form as a
-tree $\tau$.  The models of the \rmrs\ will then be such trees, and we
-can define notions such as satisfaction and entailment over these
-models in the usual way.  Satisfaction corresponds to the relation
-between an \rmrs\ and a fully specific semantic representation that it
-describes, whereas entailment corresponds to the relation between
+tree $\tau$ (see Figure~\ref{fig:1}).  So models of the \rmrs\ are
+such trees, and we 
+can define satisfaction and entailment over these
+models in the usual way.  Satisfaction corresponds to 
+an \rmrs\ being a partial description of the (unique) fully specific
+semantic representation that the model corresponds to, whereas
+entailment corresponds to the relation between
 a more and a less specific \rmrs.
@@
 predicate logic; in this case, $\tau$ is simply the structure tree of
-the predicate logic formula.  Other base languages may require
-different mappings between formulas and trees.  We assume that
-$\Sigma$ is a ranked signature consisting of the symbols of the base
-language with appropriate arities.  To encode predicate logic, we
-assume the unary constructor $\neg$ and binary constructors $\wedge$,
+the predicate logic formula.  We assume that
+$\Sigma$ is a ranked signature consisting of predicate logic's
+symbols:
+a unary constructor $\neg$ and binary constructors $\wedge$,
 $\rightarrow$, etc.; the binary constructors $\forall x$ and $\exists
@@
 predicate symbol $P$ of arity $n$, a constructor $P$ of arity $n$.
-For other base languages, a different signature could be chosen; we
+Other base languages may require a different signature $\Sigma$ and/or
+a different mapping between formulas and trees.  We
 only assume for convenience that the signature contains a binary
@@
 POS tagger didn't catch.  For instance, $g$ might map $x_4$ and $x_5$
-in (\ref{eg:cat-pos}b) to $y$ to specify the restrictor of the
-quantifier.  Finally, we assume a relation $\sigma \subseteq \pred
+in (\ref{eg:cat-pos}b) to $y$ to specify a variable binding.  Finally,
+we assume a relation $\sigma \subseteq \pred 
 \times \Sigma_{\geq 1}$ that maps each {\sc rmrs} predicate symbol to
@@
 Let's look at the \rmrs\ in (\ref{eg:cat-pos}b) by way of example.
-This \rmrs\ is normal, and it is of a particularly simple form because
-it consists only of elementary predications.  The \rmrs\ is satisfied
-(among others) by the tree in Fig.~\ref{fig:1} together with some
+This \rmrs\ is satisfied
+by the tree in Fig.~\ref{fig:1} (among others) together with some
 particular $\alpha$, $g$, and $\sigma$.  The assignment $\alpha$ will
@@
 about''.  An example is shown in Fig.~\ref{fig:fat-black-cat}.  This
-tree, which represents the meaning of ``the fat black cat chased the
+tree, which represents the meaning of ``every fat black cat chased some
 dog'', differs from the model in Fig.~\ref{fig:1} in the subtree below
 the restrictor of the universal quantifier.  It is still a model of
-(\ref{eg:cat-pos}b) -- all the EPs are satisfied.  However, there is
-more ``semantic material'' (the ``black'' subformula) in the space
-between the labels $l_{41}$ and $l_{42}$ and the anchors $a_{41}$ and
-$a_{42}$.  The \rmrs\ didn't specify that this material should be
+(\ref{eg:cat-pos}b)---all the EPs are satisfied.  However, there is
+more ``semantic material'' (the ``black'' subformula).
+The \rmrs\ didn't specify that this material should be
 there; but it also didn't specify that it can't be there, and indeed
@@
 to decide satisfiability or compute the fully specific semantic
-representations -- how can this be done in finite time?
+representations---how can this be done in finite time?
 
-In this paper, we follow the dominance constraint tradition in
-defining a fragment of \rmrs s called \emph{solved forms}.  All \rmrs
+In this paper, we follow the dominance constraint strategy for
+resolving this dilemma; we
+define a fragment of \rmrs s called \emph{solved forms}.  All \rmrs
 s in solved form are guaranteed to be satisfiable, and thus each
@@
 structure, all variable equalities and inequalities have been
-resolved, and only contains lexical ambiguities that are ``harmless'',
-inthat any choice of a constructor for the predicate symbol will
-satisfy the \rmrs.  This is made precise below.
+resolved, and only lexical sense ambiguities remain.  This is made
+precise below.
 
@@
 assigned the sort $x$, event variables the sort $e$, and holes the
-4sort $h$; sometimes subsorts of these are considered (e.g.\ $e_{PAST}$
+sort $h$; sometimes subsorts of these are considered (e.g.\ $e_{PAST}$
 for events in the past tense), and all the sorts arranged in a sort
@@
 
-While we have defined \rmrs\ without sorts above to simplify the
+While we have defined \rmrs\ without sorts to simplify the
 presentation, it is straightforward to add sorts to the definition.
@@
 such that the constructor $p$ of sufficiently high arity is also
-consistent with the sorts of the known arguments -- i.e., if $p$ has
+consistent with the sorts of the known arguments---i.e., if $p$ has
 sort $s_1 \times \ldots \times s_n \rightarrow s$ and the \rmrs\
 contains an atom $\ARG_{\{k\}}(Y,i)$ and $i$ is of sort $s'$, then
-$s'$ is a subsort of $s_k$ --, all the above propositions about solved
+$s'$ is a subsort of $s_k$---all the above propositions about solved
 forms remain true.
 
-Another extension that is worth discussing is the idea of replacing
-the qeq atom, $X \qeqdom Y$, by ordinary dominance $X \dom Y$ in the
-sense of dominance constraints, which is satisfied iff there is a path
+Another extension is to replace
+the qeq atom $X \qeqdom Y$ with dominance $X \dom Y$, 
+which is satisfied iff there is a path
 from $\alpha(X)$ to $\alpha(Y)$ in the tree.  Dominance is clearly a

### 2008-09-05 trunk/Papers/rmrs-08/entailment.tex
entailment bits

@@
 efficient
-algorithms for entailment checking, and of course this would be
-required to make parser comparison and integration with {\sc rmrs}
+algorithms for entailment checking, and this is needed
+for parser comparison and integration to be
 practical.  Computing the solved forms of $\varphi$ and $\varphi'$ may
 be exponential, and the complexity of the {\sc rmrs} validity problem
-is still unknown.  However, the above theorem illustrates that our
-model theory for {\sc rmrs} functions in the way it should for parser
+is unknown.  However, the above theorem illustrates that our
+model theory functions in the way it should for parser
 comparison and integration, and it can thus be used to prove important
@@
   ``$\Rightarrow$'': Assume that $\varphi \models \varphi'$, and let
-  $s$ be an arbitrary solved form of $\varphi$.  We need to show that
+  $s$ be a solved form of $\varphi$.  We need to show that
   we can remove atoms from $s$ such that $s$ remains an extension of
@@
   It is obvious that $s$ is an extension of $s'$.  Furthermore, $s'$
-  is in solved form, because $s$ was a tree already and we didn't
+  is in solved form, because $s$ was a tree and we didn't
   introduce new edges.  The tricky part is to show that $s'$ is a
@@
   \subseteq D(s')$ because $M,\alpha \models s'$ and $s'$ is
-  tree-shaped and normal and therefore contains all statements about
-  dominance that are true in $M$.
+  tree-shaped and normal and therefore contains all dominance statements
+  that are true in $M$.
 \end{proof}

### 2008-09-17 trunk/Papers/rmrs-08/conclusion.tex
iterated over whole paper

--- 2008-05-20 trunk/Papers/rmrs-08/abstract.tex
@@
-\begin{abstract}
-  la la la
-\end{abstract}
+\section{Conclusion}
+\label{sec:conclusion}
+
+In this paper, we have motivated and defined \rmrs---a semantic
+framework that has been used to represent, compare, and combine
+semantic information computed from deep and shallow parsers.  \rmrs\
+is designed to be maximally flexible on the type of semantic
+information that can be left underspecified -- it can underspecify
+lexical ambiguity, predicate-argument relations, and scope --, so that
+the semantic output of a shallow parser needn't over-determine or
+under-determine the semantic information that can be extracted from
+the shallow syntactic analysis.  This lays the formal foundations for
+a semantic representation formalism that is emerging as a standard in
+robust semantic processing.
+
+We then defined the concept of \emph{solved forms} of an \rmrs, in
+analogy to dominance graphs.  Solved forms abstract away from
+irrelevant differences between models, and have been a crucial concept
+in the development of efficient solvers for dominance graphs.  In the
+future, we hope to draw upon that body of literature to develop
+efficient algorithms for processing \rmrs s, including solvers which
+will compute fully specific semantic representations from an \rmrs,
+efficient inference algorithms, and algorithms for entailment
+checking.  We have argued that entailment is a central notion in
+comparing the outputs of different NLP tools, and provided a syntactic
+characterization for \rmrs\ entailment, which we hope will allow us to
+develop efficient entailment algorithms in the future.
 

### 2008-09-17 trunk/Papers/rmrs-08/entailment.tex
iterated over whole paper

--- 2008-05-20 trunk/Papers/rmrs-08/entailment.tex
@@
 
-Now that we have the formal tools in place, let's go back to the
-problem of checking compatibility of semantic outputs.  We can phrase
-this as RMRS entailment.  This is progress already.
+The definitions in the previous section serve an important theoretical
+purpose in that they provide a formal underpinning for \rmrs, which so
+far has been used for partial semantic representations in implemented
+systems despite the lack of a formal definition.  Next to the peace of
+mind that comes with the use of a formally understood formalism, a
+clean definition can also provide a foundation for the development of
+efficient algorithms; this has been shown in the past in the dominance
+graph literature, most recently \newcite{KolRegTha08}.  We hope that
+this paper can likewise provide a starting point for the development
+of efficient solvers and inference mechanisms for \rmrs.
 
-Then show the representation theorem of entailment as solved form
-comparison:
+On the other hand, the above definition of \rmrs\ provides new
+concepts which can help us phrase questions of practical grammar
+engineering in well-defined formal terms.  For instance, one crucial
+issue in developing a hybrid system that combines the outputs of deep
+and shallow processors is to determine whether the \rmrs s produced by
+the two systems are compatible.  Because we assume that the deep
+grammar is more informative than that of the shallow processor, this
+amounts to checking whether every piece of information in the shallow
+representation is consistent with what the deep grammar said.  In
+general, it isn't even obvious what ``consistency'' is supposed to
+mean exactly, given that variables might be renamed, $\ARG$ roles
+assigned in different ways, and structural relations underspecified.
+But given the formal definition of \rmrs, we can now propose that an
+\rmrs\ $\varphi'$ that was computed by a shallow processor is
+consistent with an \rmrs\ $\varphi$ that was computed by a deep
+grammar amounts to checking \emph{entailment} between the two \rmrs s,
+i.e.\ whether $\varphi \models \varphi'$.  If we could establish
+efficiently that $\varphi'$ is entailed by $\varphi$, this would tell
+us that all models that satisfy $\varphi$ are also models of
+$\varphi'$, i.e.\ the deep representation is formally more specific.
 
-\begin{thm}
-  Let $\varphi, \varphi'$ be two RMRSs.  Then $\varphi \models
+We cannot provide an efficient algorithm for testing entailment in
+this paper.  However, we propose the following novel syntactic
+characterization of entailment.
+
+\begin{definition}
+  We call an \rmrs\ $\varphi'$ an \emph{extension} of an \rmrs\
+  $\varphi$ iff \todo{fill this in -- don't have the def with me right
+    now}.
+\end{definition}
+
+\begin{thm}\label{thm:big-one}
+  Let $\varphi, \varphi'$ be two {\sc rmrs}s.  Then $\varphi \models
   \varphi'$ iff for every solved form $S$ of $\varphi$, there is a
   solved form $S'$ of $\varphi'$ such that $S$ is an extension of
-  $S'$.
+  $S'$. \todo{We don't know if this is true for qeq; the proof is
+    written for $\dom$.}  \todo{Also note that this is about a version
+    of entailment in which $\varphi$ and $\varphi'$ contain the same
+    variables, which makes the claim a bit weaker.}
 \end{thm}
+\begin{proof}[Proof (sketch)]
+``$\Leftarrow$'' follows from
+Props.~\ref{prop:solved-forms-are-satisfiable} and
+\ref{prop:models-satisfy-solved-forms}.
 
-(Of course, this doesn't give us an efficient algorithm for entailment
-checking, because the number of solved forms may be exponential on
-each side.  And we still don't know the complexity of deciding
-entailment (even for normal dominance constraints).  Nevertheless,
-this is nice, because it illustrates that our model theory is both
-pretty and useful for proving stuff.)
+``$\Rightarrow$'': The basic idea is to construct a solved form for
+$\varphi'$ by choosing a solved form for $\varphi$ and removing all
+atoms using variables that don't occur in $\varphi'$.  The hard part
+is the proof that the result is a solved form of $\varphi'$; this step
+involves proving that if $\varphi \models \varphi'$, then all EPs in
+$\varphi'$ also occur in $\varphi$.  
+\end{proof}
+
+This characterization does not by itself provide an efficient test for
+entailment because the number of solved forms of $\varphi$ and
+$\varphi'$ may be exponential, and thus it is infeasible to compute
+them all and then test for extensions.  Nevertheless, we hope that the
+theorem provides a first step towards an improved understanding of
+entailment that will lead to such an efficient algorithm in the
+future.
 

### 2008-09-17 trunk/Papers/rmrs-08/introduction.tex
iterated over whole paper

@@
 robust wide-coverage semantic construction.  For instance,
-\cite{bos:etal:2004} exploit
-{\sc ccg}'s elegant syntax-semantics interface to produce a unique
-logical form for each derivation of
-a statistical {\sc ccg} parser;
-\cite{zettlemoyer:collins:2007} and others learn
-string-to-logical-form mappings, exploting informative features from
-shallow language processors; and \cite{copestake:2003} and
-\cite{frank:2004}
-endow robust language processors with semantic components.
+\newcite{bos:etal:2004} exploit {\sc ccg}'s elegant syntax-semantics
+interface to produce a unique logical form for each derivation of a
+statistical {\sc ccg} parser; \newcite{zettlemoyer:collins:2007} and
+others learn string-to-logical-form mappings, exploting informative
+features from shallow language processors; and
+\newcite{copestake:2003} and \newcite{frank:2004} endow robust
+language processors with semantic components.
 
-However, there are certain semantic
-phenomena that these robust
-approaches don't capture reliably, including quantifier scope, optional
-arguments,
 % COMMENT: this issue about optional arguments is really too cryptic.
@@
 %
-and long-distance dependencies (for instance, \cite{clark:etal:2004}
-report that the parser used in \cite{bos:etal:2004} yields 63\%
-accuracy on object
-extraction; e.g., {\em the man that I met\dots}).
-Forcing a shallow parser to make a decision about these
-phenomena can therefore be error-prone.  Depending on the
-application, it may be preferrable to give the parser the option to to
-leave the decision open when it's not sufficiently informed---i.e., to
-compute a partial semantic representation that neither over-determines
-nor under-determines the semantic information revealed by the
-(shallow) syntactic analysis, and to complete the semantic information
-later, using information extraneous to the
+
+However, there are certain semantic phenomena that these robust
+approaches don't capture reliably, including quantifier scope,
+optional arguments, and long-distance dependencies (for instance,
+\newcite{clark:etal:2004} report that the parser used by
+\newcite{bos:etal:2004} yields only 63\% accuracy on object
+extraction; e.g., {\em the man that I met\dots}).  Forcing a shallow
+parser to make a decision about these phenomena can therefore be
+error-prone.  Depending on the application, it may be preferrable to
+give the parser the option to to leave the decision open when it's not
+sufficiently informed---i.e., to compute a partial semantic
+representation that neither over-determines nor under-determines the
+semantic information revealed by the (shallow) syntactic analysis, and
+to complete the semantic information later, using information
+extraneous to the parser.
+
+In this paper, we focus on one particular approach to representing
+partial semantic information: Robust Minimal Recursion Semantics
+(\rmrs, \cite{copestake:2003}).  \rmrs\ is a generalisation of \mrs\
+\cite{copestake:etal:2005} that is designed to support
+underspecification of scope and predicate-argument structure.  It is
+an emerging standard representation language for partial semantic
+information, and has been applied in several implemented systems.  For
+instance, \newcite{copestake:2003} and \newcite{frank:2004} use it to
+specify semantic components to shallow parsers ranging in depth from
+{\sc pos} taggers to chunk parsers and intermediate parsers such as
+{\sc rasp} \cite{briscoe:etal:2006}.  {\sc mrs} analyses derived from
+deep grammars, such as the English Resource Grammar ({\sc erg},
+\cite{copestake:flickinger:2000}) can be seen as special cases of {\sc
+  rmrs}.
+
+However, from a formal perspective, \rmrs\ is somewhat ad-hoc in that
+it has no clearly defined model theory; i.e., it is not defined what,
+exactly, any given \rmrs\ representation actually \emph{means}.  The
+key contribution we make is to cast {\sc rmrs}, for the first time, as
+a logic with a well-defined model theory.  This has a number of
+advantages.  First, it allows us to formally relate \rmrs\ to other
+frameworks such as \mrs\ and predicate logic in order to establish the
+compatibility of \rmrs\ with the approaches to robust semantic
+analysis mentioned above.  \todo{except we don't do this here --
+  perhaps we should} Second, it provides us with a notion of logical
+entailment, which we can use to verify the compatibility of \rmrs\
+representations computed from different parsers, such as a deep and a
+shallow parser.  Finally, it is only through rigid formalisation that
+we can give users of \rmrs\ access to inference systems and to
+algorithms for efficiently computing the possible fully specific
+semantic representations described by the \rmrs\ output of a shallow
 parser.
 
-In this paper, we focus on one particular approach to representing
-partial semantic information:
-Robust Minimal Recursion Semantics ({\sc rmrs},
-\cite{copestake:2003}).  {\sc rmrs} is a generalisation of {\sc mrs}
-\cite{copestake:etal:2005} that is designed to support underspecification
-of scope and predicate-argument structure.
-\cite{copestake:2003,frank:2004} use it to
-specify semantic components to shallow parsers
-ranging in depth from {\sc pos} taggers to chunk parsers  
-and intermediate parsers such as {\sc rasp}
-\cite{briscoe:etal:2006}.  {\sc mrs} analyses
-derived from deep grammars, such as the English Resource Grammar
-({\sc erg}, \cite{copestake:flickinger:2000}) are
-special cases of 
-{\sc rmrs}.
-
-The key contribution we make is to cast {\sc rmrs}, for the first
-time, as a logic with a clearly defined model theory.  This has a
-number of advantages.  First, it allows us to formally relate \rmrs\ to
-other frameworks such as \mrs\ and predicate logic in order to
-establish the compatibility of \rmrs\ with the approaches to robust
-semantic analysis mentioned above.  Second, it provides us with a
-notion of logical entailment, which we can use to verify the
-compatibility of \rmrs\ representations computed from different
-parsers, such as a deep and a shallow parser.  Finally, it is only
-through rigid formalisation that we can give users of \rmrs\ access to
-inference systems and to algorithms for efficiently computing the
-possible fully specific semantic representations described by the \rmrs\
-output of a shallow parser.
-
 Section~\ref{sec:motivation} illustrates via example the differences
-and challenges with deep and 
-shallow semantic construction.  
-Section~\ref{sec:rmrs} defines the syntax and
-model-theory of \rmrs\.  In Section~\ref{sec:entailment} we
-demonstrate 
-the practical usefulness of our formalisation 
-by modelling compatibility of
-partial semantic representations as \rmrs\ entailment and by providing a
+and challenges with deep and shallow semantic construction.
+Section~\ref{sec:rmrs} defines the syntax and model-theory of \rmrs\.
+In Section~\ref{sec:entailment} we demonstrate the practical
+usefulness of our formalisation by modelling compatibility of partial
+semantic representations as \rmrs\ entailment and by providing a
 syntactic criterion for characterising entailment.

### 2008-09-17 trunk/Papers/rmrs-08/motivation.tex
iterated over whole paper

@@
 
-% Motivate what we're doing and provide background.  Example outputs of
-% deep and shallow parsers.  Illustrate and state the problem.
-%
-% Try to find an eg where the comparison isn't easy to see (from Anna
-% Ritchie's deliverable?). 
-% NOTE: on Ritichie: Perhaps "Three dogs bark on Tuesday in
-% June" from page 11?  It's hard to follow what's going on, from the
-% sheer size of the RMRSs and the need to allow for variation on the
-% variable names.  But I don't know how to extract this figure
-% techinically; nor whether we have room.
+We start with an informal discussion of some toy \rmrs s, which are
+meant to illustrate how \rmrs\ can be used to represent partial
+semantic information, and why people find it useful for this purpose.
 
-We motivate and describe {\sc rmrs} with a toy
-example.  Sentence 
-(\ref{eg:cat}a) features a semantic scope ambiguity that, according to
-most deep grammars, syntax
-doesn't resolve.    The {\sc mrs} (\ref{eg:cat}b) derived by the
-{\sc erg} \cite{copestake:flickinger:2000}
-reflects this: (i) each predication is
-labelled ($l_1$, $l_4$, etc); (ii) scopal arguments are indicated by
-holes (e.g., $h_2$, $h_3$); (iii)
-constraints on scope are expressed by the qeq conditions (roughly, $h
-=_q l$ means that only zero or more quantifiers lie between the scopal
-argument $h$ and the predication labelled by $l$ in any fully-resolved
-logical form); and (iv) these scope
-constraints allow for more than one way of `plugging' the holes with
-labels.\footnote{The event variable in $\_\mbox{{\em fat}}\_j\_1$ is 
-  for copulas. {\em The
-    cat is fat} has logical form 
-  $\mbox{\_the\_q\_1}(x,\mbox{\_cat\_n\_1}(x),\mbox{\_fat\_j\_1}(e,x))$.}
+\newcommand{\sem}[1]{\mathsf{#1}}
+\newcommand{\sempred}[1]{\mathrm{#1}}
+
+Consider the following sentence:
 \begin{examples}
-\item   \label{eg:cat}
-\begin{subexamples}
+  \item Every fat cat chased some dog.
+\end{examples}
+
+This sentence exhibits several kinds of ambiguity, including a scope
+ambiguity between ``every fat cat'' and ``some dog'' and lexical
+ambiguities of the words ``cat'' and ``dog'' (which have 8 and 7 noun
+senses according to Wordnet, respectively).  Two of these readings are
+shown as logical forms below; they can be represented as trees as
+shown in Fig.~\ref{fig:1}.
+
+\begin{examples}
+\item $\sem{\_every\_q\_1}(x, \sem{\_fat\_j\_1}(e',x) \wedge
+    \sem{\_cat\_n\_1}(x), \sem{\_some\_q\_1}(y, \sem{\_dog\_n\_1}(y),
+    \sem{\_chased\_v\_1}(e,x,y)))$
+\label{ex:fat-cat-1}
+\item $\sem{\_some\_q\_1}(y, \sem{\_dog\_n\_2}(y),
+\sem{\_every\_q\_1}(x, \sem{\_fat\_j\_1}(e',x) \wedge
+    \sem{\_cat\_n\_2}(x), 
+    \sem{\_chased\_v\_1}(e,x,y)))$
+\label{ex:fat-cat-2}
+\end{examples}
+
+\begin{figure*}[t]
+\includegraphics[width=6cm]{pic-cat-chased-dog}
+\hspace{1cm}
+\includegraphics[width=6cm]{pic-cat-chased-dog-2}
+\caption{Structure trees of the semantic representations (\ref{ex:fat-cat-1}) and
+  (\ref{ex:fat-cat-2}). \label{fig:1}}
+\end{figure*}
+
+Now imagine that we are trying to extract semantic information from
+the output of a part-of-speech tagger by using the word lemmas as
+lexical predicate symbols.  Such a semantic representation is highly
+partial, as it will not say anything about predicate-argument
+relations, or resolve lexical ambiguity -- it will use predicate
+symbols such as $\sempred{\_cat\_n}$, which might resolve to the
+predicate symbols $\sem{\_cat\_n\_1}$ or $\sem{\_cat\_n\_2}$ in the
+complete semantic representation.  (Notice that we are using different
+fonts for the ambiguous and unambiguous predicate symbols.)  We can
+express this information in \rmrs\ as follows.
+
+\begin{examples}
+\item \label{ex:cat-pos}
+$l_1:a_1:\sempred{\_every\_q}(x)$, \\
+$l_{41}:a_{41}:\sempred{\_fat\_a}(e')$,\\
+$l_{42}:a_{42}:\sempred{\_cat\_n}(x_3)$\\
+$l_5:a_5:\sempred{\_chased\_v}(e_{past})$, \\
+$l_6:a_6:\sempred{\_some\_q}(x_4)$, \\
+$l_9:a_9:\sempred{\_dog\_n}(x_5)$
+\end{examples}
+
+This \rmrs\ consists of six \emph{elementary predications} (EPs), one
+for each word lemma in the sentence; each of them is prefixed by a
+\emph{label} and an \emph{anchor}, which are essentially variables
+that refer to nodes in the trees in Fig.~\ref{fig:1}.  We can say that
+the two trees \emph{satisfy} this \rmrs\ because it is possible to map
+the labels and anchors into each tree in such a way that the label of
+the node they are mapped to is consistent with the label specified in
+the elementary predication.  For instance, the first tree satisfies
+the \rmrs\ because we can map $l_1$ and $a_1$ to the root of that
+tree, and the root label $\sem{\_every\_q\_1}$ is consistent with the
+EP predicat $\sempred{\_every\_q}$.
+
+There are of course many other trees (and thus, fully specific
+semantic representations such as (\ref{ex:fat-cat-1})) that are
+described equally well by the same \rmrs; this is not surprising,
+given that the semantic information we got from the POS tagger is so
+incomplete.  If we have more information, say information about the
+predicate-argument structure and semantic dependencies from a partial
+parser such as \todo{what? CASS?}, we can represent these in a more
+detailed \rmrs, as follows:
+
+\begin{examples}
 \item 
-Every fat cat chased some dog.
-% \item
-% $\mbox{\_every\_q\_1}(x, \mbox{\_fat\_j\_1}(e',x) \wedge
-% \mbox{\_cat\_n\_1}(x),$\\
-% \hspace*{0.5in}$\mbox{\_some\_q\_1}(y, \mbox{\_dog\_n\_1}(y),
-% \mbox{\_chase\_v\_1}(e_{spast},x) )$\\
-% $\mbox{\_some\_q\_1}(y, \mbox{\_dog\_n\_1}(y),$\\
-% \hspace*{0.5in}$\mbox{\_every\_q\_1}(x, \mbox{\_fat\_j\_1}(e',x)
-% \wedge 
-% \mbox{\_cat\_n\_1}(x), 
-% \mbox{\_chase\_v\_1}(e_{spast},x)))$
-\item   $l_1\handel\mbox{\_every\_q\_1}(x,h_2,h_3)$\\
-$l_4\handel\mbox{\_fat\_j\_1}(e',x)$\\
-$l_4\handel\mbox{\_cat\_n\_1}(x)$\\
-$l_5\handel\mbox{\_chased\_n\_1}(e,x,y)$\\
-$l_6\handel\mbox{\_some\_n\_1}(y,h_7,h_8)$\\
-$l_9\handel\mbox{\_dog\_n\_1}(y)$\\
-$h_2 =_q l_4, h_7 =_q l_9$
-\end{subexamples}
+$l_1:a_1:\sempred{\_every\_q}(x)$, \\
+$l_{41}:a_{41}:\sempred{\_fat\_a}(e')$,\\
+$l_{42}:a_{42}:\sempred{\_cat\_n}(x_3)$\\
+$l_5:a_5:\sempred{\_chased\_v}(e_{past})$, \\
+$l_6:a_6:\sempred{\_some\_q}(x_4)$, \\
+$l_9:a_9:\sempred{\_dog\_n}(x_5)$
+\todo{add more stuff with ARG and equalities here}
+\label{ex:cat-partial-parser}
 \end{examples}
-(\ref{eg:cat}b) also exhibits {\sc mrs}' compact description of
-conjunction through label sharing (see $l_4$), and the {\sc
-  erg}'s naming convention for predicate symbols: each lexical predicate is
-constructed from the word lemma, its syntactic category and a sense
-number, and the leading underscore distinguishes lexical predicates
-from non-lexical ones.
 
-Conceptually, this {\sc mrs} is a partial description of all its
-possible (fully specific) interpretations in context.  
-These are expressed in a so-called {\em base language}, where each
-predicate symbol in the {\sc mrs} is a proxy for one or more
-base-language constructors of the same arity.
-Assuming a very close
-correspondence between the {\sc mrs} predicate symbols and the
-base-language constructors they 
-stand for, the {\sc mrs}
-(\ref{eg:cat}b) (partially) describes both the base-language formulae in
-Figure~\ref{fig:1}, shown as trees to emphasise that an {\sc mrs} is a
-partial description of the formulae's {\em form}; not its
-(base-language) entailments. 
-In essence, the labels in the {\sc mrs} reify the
-nodes in the tree, and each of these trees encodes one way of equating
-holes with labels that respects qeq constraints and binding conditions
-on the variables.
+This \rmrs\ uses two new types of atoms.  Atoms of the form $x=y$
+express that the two variables $x$ and $y$ should really be the same;
+in the example, the \rmrs\ uses two variable names $x$ and $x'$, which
+are mapped to the same variable $x$ in Fig.~\ref{fig:1}.  \todo{I'm
+  making these names up -- adjust to RMRS.}  Atoms of the form
+$\ARG_i(a,x)$ and $\ARG_i(a,h)$ express that a node representing the
+variable $x$ or a node denoted by the \emph{hole} $h$ is the $i$-th
+child of the node to which the anchor $a$ refers.  The separation of
+EPs and $\ARG$ atoms is perhaps the most important difference between
+\rmrs\ and earlier underspecification formalisms such as \mrs\ and
+dominance graphs, which required that a predicate and its arguments
+had to be specified together.  \rmrs\ goes beyond these formalisms in
+expressive power, especially if we allow atoms of the form
+$\ARG_{\{2,3\}}(a,x)$ to express uncertainty as to whether $x$ is the
+second or third child of the anchor $a$.
 
-\begin{figure}[h]
+Notice also that the labels of the EPs for ``fat'' and ``cat'' are
+stipulated to be equal in the \rmrs, whereas the anchors are not.
+In the tree, it is the anchors that are mapped to the nodes with the
+labels $\sem{\_fat\_a\_1}$ and $\sem{\_cat\_n\_1}$; the label is
+mapped to the conjunction node just above them.  In other words, the
+role of the anchor in an EP is to carry the predicate and connect to
+the arguments; the role of the label is to connect the EP to the
+surrounding formula.  The use of equal labels for two EPs stems from
+\mrs, where intersective modification was also modeled by using two
+EPs with the same label.
 
-\includegraphics[width=6cm]{pic-cat-chased-dog}
+Finally, here is an \rmrs\ that could be derived from a deep grammar
+such as the ERG:
 
-Put trees in here
-
-\caption{The Form of alternative base logical forms of
-  (\ref{eg:cat}a).}
-\label{fig:1}
-\end{figure}
-
-The {\sc mrs} (\ref{eg:cat}b) is a notational variant of the {\sc
-  rmrs} (\ref{eg:cat}c). {\sc rmrs}s offer a more factorised
-representation: each predicate symbol is unary and the other
-arguments to its corresonding base-language constructor
-are represented by separate binary relations on the unique {\em
-  anchor} of the predicate symbol ($a_1, a_{41},\ldots$) together with
-variable equalities (e.g., $x=x_1$).
 \begin{examples}
-\item   [\ref{eg:cat}]
-\begin{subexamples}
-\item   [c]
-$l_1:a_1\handel\mbox{\_every\_q\_1}(x_1),$\\
+\item $l_1:a_1\handel\mbox{\_every\_q\_1}(x_1),$\\
 \hspace*{0.1in}$\mbox{RSTR}(a_1,h_2),
@@
 $x_5=x_6, x_5=x_7$
-\end{subexamples}
+\todo{fix RSTR and BODY to ARG1 and ARG2? or talk about it in the text?}
+\label{ex:cat-erg}
 \end{examples}
 
-This factored representation allows one to build semantic components
-to shallow parsers where lexical or syntactic information is absent.
-An extreme example would be a {\sc 
-  pos} tagger: one can build its semantic component simply by deriving
-lexical predicate symbols from the word lemmas and tags and assigning each
-predication a unique label, anchor and
-argument, as in (\ref{eg:cat-pos}):
-\begin{examples}
-\item   \label{eg:cat-pos}
-\begin{subexamples}
-\item   Every\_{\sc at1} fat\_{\sc jj} cat\_{\sc nn1} chased\_{\sc
-    vvd} some\_{\sc dd} dog\_{\sc nn1}
-\item   $l_1:a_1:\mbox{\_every\_q}(x)$, \\
-$l_{41}:a_{41}:\mbox{\_fat\_a}(e')$,\\
-$l_{42}:a_{42}:\mbox{\_cat\_n}(x_3)$\\
-$l_5:a_5:\mbox{\_chased\_v}(e_{past})$, \\
-$l_6:a_6:\mbox{\_some\_q}(x_4)$, \\
-$l_9:a_9:\mbox{\_dog\_n}(x_5)$
-\end{subexamples}
-\end{examples}
+This is now an extremely detailed underspecified representation of the
+different semantic representations of the sentence, which is
+essentially a notational variant of the \mrs\ that the ERG would
+derive.  Notice the use of atoms like $h_2 \qeq l_{42}$, which specify
+a kind of ``outscopes'' relationship between the hole and the label,
+and can be used to underspecify the scope of the two quantifiers.
 
-Semantic relations, sense tags and the arity of the base-language
-constructors are
-missing from (\ref{eg:cat-pos}b) because the {\sc pos} tagger doesn't
-reveal information about syntactic constituency, word sense or lexical
-subcategorisation respectively.  
-But the {\sc rmrs}s (\ref{eg:cat}c) and
-(\ref{eg:cat-pos}b) are entirely compatible, the former being more
-specific than the latter.  At least, this holds so long as the
-base-language constructors that the {\sc rmrs} predicate symbol
-{\_name\_t\_s} corresponds to is a 
-subset of those corresponding to the predicate symbol {\em \_name\_t},
-for any word lemma {\em name}, {\sc pos} tag $t$ and sense $s$ (written 
-$\mbox{{\em \_name\_t\_s}} \sqsubset \mbox{{\em name\_t}}$).
+In summary, \rmrs\ is a formalism for representing partial information
+about semantic representations.  It generalizes \mrs\ in that it can
+represent the output of a deep grammar in a way that is a notational
+variant of \mrs; but in cases in which a deep grammar is not
+available, it can be used to capture whatever partial semantic
+information a shallower NLP tool was able to provide.  \rmrs\ is
+designed to serve as a generic formalism for partial semantic
+representations which abstracts over the details of different NLP
+tools and allows us to represent, process, and compare information
+from different tools in the same formalism.  In the long run, it could
+also be a platform for incrementally enriching partial information
+such as (\ref{ex:cat-pos}) in order to eventually approximate more
+detailed representations such as (\ref{ex:cat-partial-parser}).
 
-The values of non-scopal arguments are underspecified through the
-absence of ARG relations and/or through the absence of
-equalities and inequalities among the variables $x_1, x_2,\ldots$.
-The need to
-underspecify such semantic dependencies is required when
-the shallow parser doesn't record syntactic dependencies.
-Indeed, there is no single {\sc mrs} which is satisfied by both the
-trees in Figure~\ref{fig:2}, assuming of course that disjunction is not part of
-the {\sc mrs} language.  In words, these trees collectively specify
-the (partial) semantic information that $x$ is either the second
-argument (i.e., the direct object) or the third argument (i.e., the
-indirect object) to {\em give}, but we don't know which.  
-A shallow
-processor that misses long distance dependencies in
-(\ref{eg:which-dog}) cannot
-discern that the bone is the second argument to {\em
-  give} in (\ref{eg:which-dog}a), while the dog is the third argument
-in (\ref{eg:which-dog}b).
-\begin{examples}
-\item   \label{eg:which-dog}
-\begin{subexamples}
-\item   To which dog did Kim give the bone?
-\item   Which bone did Kim give the dog?
-\end{subexamples}
-\end{examples}
-Thus to maximise the semantic contribution
-of the parser while not overdetermining it, the phrases
-{\em give the bone} vs.\
-{\em give the dog} should whether the semantic index of the noun
-phrase is the second or third argument to {\em give},
-as shown in Figure~\ref{fig:2}.   This will be expressed in
-{\sc rmrs} via the binary relation  $\mbox{{\em ARG}}_{\{2,3\}}(a,x)$
-between the anchor $a$ of the predicate symbol for {\em give} and the
-individual $x$.   More generally, $\mbox{{\em ARG}}_n(a,x)$ means that
-$x$ is an argument to the predicate anchored at $a$, but its argument
-position is unspecified.
-
-\begin{figure}
-
-\leaf{$e$}
-\leaf{$\ldots$}
-\leaf{$x$}
-\leaf{$\ldots$}
-\branch{4}{\_give\_v\_b1}
-\tree
-\hfill
-\leaf{$e$}
-\leaf{$\ldots$}
-\leaf{$\ldots$}
-\leaf{$x$}
-\branch{4}{\_give\_v\_b1}
-\tree
-
-\caption{$x$ is a participant in a
-  giving event, but $x$'s role is unknown.}
-\label{fig:2}
-\end{figure}
-
-The binary ARG relations also supports underspecifying the {\em arity}
-of the base-language constructors, as is required when lexical
-subcategorisation information is missing from the shallow parser.
-While a deep parser assumes complete information about lexical
-subcategorisation in any of its derivations, thereby fixing the arity
-of the base-language constructors in the fully-specific logical form
-to be the same as the {\sc mrs} predicate symbol $P$ that's introduced by
-the lexeme, the predicate symbols in an {\sc rmrs} yielded from a
-shallow parse should, in principle at least,
-correspond to base language constructors of mixed arity.  For
-instance, the English verb {\em bake} can be used causatively ({\em
-  Kim baked the potatoes}), in which case semantically it corresponds
-to a 3-place predicate symbol {\em \_bake\_v\_b1} in the base
-language, or it may be used 
-inchoatively ({\em the potatoes are baking}), that semantically
-corresponds to a 2-place predicate {\em \_bake\_v\_b2}.  So the 
-predicate symbol {\em 
-  \_bake\_v} in the {\sc rmrs} that's introduced by a {\sc pos} tagger
-must be able to denote either of these
-base-language predicates in the {\sc rmrs}'s interpretation.
-
-\hidden{
-Finally, {\sc rmrs} allows one to underspecify the {\em sort} of an
-argument by including in its language meta-variables $u_1,u_2,\ldots$
-over individual, event, and handle variables.  For instance, the predicate
-symbol {\em \_be\_v} produced by a {\sc pos} tagger is resolvable to several
-3-place base-language constructors whose argument-types 
-vary: the base-language constructor corresponding
-to its use in {\em John is a man} takes an event and two individuals
-as arguments; but in {\em It is John who talked} the constructor takes
-an event, an individual and a proposition as arguments.
-}
+\todo{Leidensdruck!}
 

### 2008-09-17 trunk/Papers/rmrs-08/old/entailment.tex
iterated over whole paper

This theorem doesn't provide 
efficient
algorithms for entailment checking, and this is needed
for parser comparison and integration to be
practical.  Computing the solved forms of $\varphi$ and $\varphi'$ may
be exponential, and the complexity of the {\sc rmrs} validity problem
is unknown.  However, the above theorem illustrates that our
model theory functions in the way it should for parser
comparison and integration, and it can thus be used to prove important
logical properties of shallow parsing systems that adopt {\sc rmrs}.

The outline proof of Theorem~\ref{thm:big-one} is as follows:
\begin{proof}
  ``$\Leftarrow$'': Let $M,\alpha$ an arbitrary model and variable
  assignment that satisfy $\varphi$.  There is a solved form $s$ of
  $\varphi$ such that $M,\alpha \models s$.  By assumption, this means
  that there is a solved form $s'$ of $\varphi'$ such that $s$ is an
  extension of $s'$.  Therefore, $M,\alpha \models s'$ and hence
  $M,\alpha \models \varphi'$.

  ``$\Rightarrow$'': Assume that $\varphi \models \varphi'$, and let
  $s$ be a solved form of $\varphi$.  We need to show that
  we can remove atoms from $s$ such that $s$ remains an extension of
  the resulting $s'$ and $s'$ is a solved form of $\varphi'$.

  We construct $s'$ as the conjunction of all atoms in $s$ that only
  contain variables occurring in $\varphi'$ and all atom $X \dom Y$
  such that $X$ and $Y$ occur in $\varphi'$ and $(X,Y) \in D(s)$.

  It is obvious that $s$ is an extension of $s'$.  Furthermore, $s'$
  is in solved form, because $s$ was a tree and we didn't
  introduce new edges.  The tricky part is to show that $s'$ is a
  solved form of $\varphi'$.  $s'$ contains all labeling atoms of
  $\varphi'$ according to
  Lemma~\ref{lem:entailment-preserves-labeling-atoms}; a similar
  argument holds for inequality atoms.  Finally, $D(\varphi')
  \subseteq D(s')$ because $M,\alpha \models s'$ and $s'$ is
  tree-shaped and normal and therefore contains all dominance statements
  that are true in $M$.
\end{proof}

\begin{lemma} \label{lem:entailment-preserves-labeling-atoms}
  Under the assumptions of
  Theorem~\ref{thm:big-one}, if $\varphi \models
  \varphi'$, then $\varphi'$ only contains labeling atoms that also
  occur in $\varphi$.  
\end{lemma}
\begin{proof}
  Let $\varphi \models \varphi'$, but let $\varphi'$ contain a
  labeling atom $X \Neq f(X_1,\ldots,X_n)$ that doesn't occur in
  $\varphi$.  Call this atom $A$.

  Furthermore, let $M,\alpha \models \varphi$ arbitrary.  Because
  $M,\alpha \models \varphi'$, we must have $M(\alpha(X)) = f$.

  Now replace the label of $\alpha(X)$ in $M$ by some other symbol $g$
  of arity $n$.  Call the resulting model $M'$.  We still have
  $M',\alpha \models \varphi$ because $\varphi$ doesn't contain a
  labeling atom for $X$.  But now $M',\alpha \not\models \varphi'$.
  This contradicts the entailment assumption.
\end{proof}

It is really necessary to state
Theorem~\ref{thm:big-one} in terms of extensions of
solved forms.  It is \emph{not} true that $\varphi \models \varphi'$
implies that $\varphi$ is an extension of $\varphi'$.  A
counterexample is shown in Fig.~\ref{fig:counterexample}, where the
right-hand constraint entails the left-hand constraint (they are
equivalent), but is not an extension of it.

\begin{figure}
  \centering
  \includegraphics[scale=0.6]{pic-extension-counterexample}
  \caption{A counterexample to a simpler version of Theorem~\ref{thm:big-one}.}
  \label{fig:counterexample}
\end{figure}

### 2008-09-17 trunk/Papers/rmrs-08/rmrs.tex
iterated over whole paper

@@
 
-\rmrs\ as presented here combines similarities to earlier definitions
-of \rmrs\ (e.g., \cite{copestake:2003}) as illustrated in
-(\ref{eg:cat}c) and
-to dominance constraints
-\cite{egg:etal:2001}.  We
-$\ARG_{\{i\}}$ (with a singleton set $S$) instead of $\ARG_i$, and
-$\ARG_\N$ instead of $\ARG_n$.  Furthermore, the
-elementary predication $\ep{l}{a}{P(v)}$ is a notational variant of
-$\{\ep{l}{a}{P}, \ARG_{\{0\}}(l,v)\}$.
-In other words, the syntax defined
-here differs in some details, but is more general. The primary
-differences to dominance constraints are the distinction of labels and
-anchors that will separate the
-labelling and the parent-child relations, and
-the use of $\qeq$ instead of dominance.  Redefining
-\rmrs\ in the
-style of dominance constraints will simplify the formal work
-below.
+\rmrs\ as presented here combines similarities to earlier
+presentations of \rmrs\ (e.g., \cite{copestake:2003}) and to dominance
+constraints \cite{egg:etal:2001}.  We $\ARG_{\{i\}}$ (with a singleton
+set $S$) instead of $\ARG_i$, and $\ARG_\N$ instead of $\ARG_n$.
+Furthermore, the elementary predication $\ep{l}{a}{P(v)}$ is a
+notational variant of $\{\ep{l}{a}{P}, \ARG_{\{0\}}(l,v)\}$.  In other
+words, the syntax defined here differs in some details, but is more
+general. The primary differences to dominance constraints are the
+distinction of labels and anchors that will separate the labelling and
+the parent-child relations, and the use of $\qeq$ instead of
+dominance.  Redefining \rmrs\ in the style of dominance constraints
+will simplify the formal work below.
 
-\todo{don't split stuff up into tuple parts}
+\todo{we don't split stuff up into tuple parts, which earlier rmrs
+  did}
 
@@
 expressed in the base language.  We represent such a logical form as a
-tree $\tau$ (see Figure~\ref{fig:1}).  So models of the \rmrs\ are
-such trees, and we 
-can define satisfaction and entailment over these
-models in the usual way.  Satisfaction corresponds to 
-an \rmrs\ being a partial description of the (unique) fully specific
+tree $\tau$, such as the one in Fig.~\ref{fig:1}.  So models of the
+\rmrs\ are such trees, and we can define satisfaction and entailment
+over these models in the usual way.  Satisfaction corresponds to an
+\rmrs\ being a partial description of the (unique) fully specific
 semantic representation that the model corresponds to, whereas
-entailment corresponds to the relation between
-a more and a less specific \rmrs.
+entailment corresponds to the relation between a more and a less
+specific \rmrs.
 
@@
 predicate logic; in this case, $\tau$ is simply the structure tree of
-the predicate logic formula.  We assume that
-$\Sigma$ is a ranked signature consisting of predicate logic's
-symbols:
-a unary constructor $\neg$ and binary constructors $\wedge$,
-$\rightarrow$, etc.; the binary constructors $\forall x$ and $\exists
-x$ for all base-language variables $x$; all the constants and
-base-language variables as nullary constructors; and, for each
-predicate symbol $P$ of arity $n$, a constructor $P$ of arity $n$.
-Other base languages may require a different signature $\Sigma$ and/or
-a different mapping between formulas and trees.  We
-only assume for convenience that the signature contains a binary
-constructor $\wedge$ to represent conjunction.  We write $\Sigma_i$
-for the set of all constructors in $\Sigma$ that have arity $i$.
+the predicate logic formula.  We assume that $\Sigma$ is a ranked
+signature consisting of predicate logic's symbols: a unary constructor
+$\neg$ and binary constructors $\wedge$, $\rightarrow$, etc.; the
+binary constructors $\forall x$ and $\exists x$ for all base-language
+variables $x$; all the constants and base-language variables as
+nullary constructors; and, for each predicate symbol $P$ of arity $n$,
+a constructor $P$ of arity $n$.  Other base languages may require a
+different signature $\Sigma$ and/or a different mapping between
+formulas and trees.  We only assume for convenience that the signature
+contains a binary constructor $\wedge$ to represent conjunction.  We
+write $\Sigma_i$ for the set of all constructors in $\Sigma$ that have
+arity $i$.  We will follow the typographical convention that
+non-logical symbols in $\Sigma$ are written in
+$\mathsf{sans{-}serif}$, as opposed to the \rmrs\ predicate symbols
+like ``$\sempred{\_cat\_n}$''.
 
@@
 
-The {\em satisfaction} relation between an {\sc rmrs} $\varphi$ and a
+The {\em satisfaction} relation between an \rmrs\ $\varphi$ and a
 finite constructor tree $\tau$ is defined in terms of several
@@
 POS tagger didn't catch.  For instance, $g$ might map $x_4$ and $x_5$
-in (\ref{eg:cat-pos}b) to $y$ to specify a variable binding.  Finally,
-we assume a relation $\sigma \subseteq \pred 
-\times \Sigma_{\geq 1}$ that maps each {\sc rmrs} predicate symbol to
-a set of constructors from $\Sigma$.  This relation allows us to
-underspecify lexical ambiguities; for instance,
-$\sigma(\mathrm{\_cat\_n})$ might contain the symbol $\mathrm{\_cat\_n\_1}$
-in (\ref{eg:cat}c) along with symbols for other word senses.
+in (\ref{ex:cat-pos}) to $y$ to specify a variable binding.  Finally,
+we assume a relation $\sigma \subseteq \pred \times \Sigma_{\geq 1}$
+that maps each \rmrs\ predicate symbol to a set of constructors from
+$\Sigma$.  This relation allows us to underspecify lexical
+ambiguities; for instance, $\sigma(\sempred{\_cat\_n})$ might contain
+the symbol $\sem{\_cat\_n\_1}$ along with symbols for other word
+senses.
 
@@
 
-Let's look at the \rmrs\ in (\ref{eg:cat-pos}b) by way of example.
-This \rmrs\ is satisfied
-by the tree in Fig.~\ref{fig:1} (among others) together with some
-particular $\alpha$, $g$, and $\sigma$.  The assignment $\alpha$ will
-map the variables $l_1$ and $a_1$ to the root node; $l_{41}$ and
-$l_{42}$ to its second child (labeled with ``$\wedge$''); $a_{41}$ to
-the first child of \emph{that} node (i.e.\ the node 21, labelled with
-``fat'') and $a_{42}$ to the node 22; and so forth.  $g$ will map $x$
-and $x_3$ to $x$, $e'$ to $e$, and so on.  And $\sigma$ will map each
-\rmrs\ predicate symbol (which represents a word) to the set of its
-word senses, e.g.\ \_cat\_n to a set containing \_cat\_n\_1 and
-possibly others.  It is then easy to verify that every single atom in
-the \rmrs\ is satisfied -- most interestingly, the EPs
-$\ep{l_{41}}{a_{41}}{\_fat\_a(e')}$ and
+Let's look at the \rmrs\ in (\ref{ex:cat-pos}) by way of example.
+This \rmrs\ is satisfied by the tree in Fig.~\ref{fig:1} (among
+others) together with some particular $\alpha$, $g$, and $\sigma$.
+The assignment $\alpha$ will map the variables $l_1$ and $a_1$ to the
+root node; $l_{41}$ and $l_{42}$ to its second child (labeled with
+``$\wedge$''); $a_{41}$ to the first child of \emph{that} node (i.e.\
+the node 21, labelled with ``fat'') and $a_{42}$ to the node 22; and
+so forth.  $g$ will map $x$ and $x_3$ to $x$, $e'$ to $e$, and so on.
+And $\sigma$ will map each \rmrs\ predicate symbol (which represents a
+word) to the set of its word senses, e.g.\ \_cat\_n to a set
+containing \_cat\_n\_1 and possibly others.  It is then easy to verify
+that every single atom in the \rmrs\ is satisfied -- most
+interestingly, the EPs $\ep{l_{41}}{a_{41}}{\_fat\_a(e')}$ and
 $\ep{l_{42}}{a_{42}}{\_cat\_n(x_3)}$ are satisfied because
@@
 about''.  An example is shown in Fig.~\ref{fig:fat-black-cat}.  This
-tree, which represents the meaning of ``every fat black cat chased some
-dog'', differs from the model in Fig.~\ref{fig:1} in the subtree below
-the restrictor of the universal quantifier.  It is still a model of
-(\ref{eg:cat-pos}b)---all the EPs are satisfied.  However, there is
-more ``semantic material'' (the ``black'' subformula).
-The \rmrs\ didn't specify that this material should be
-there; but it also didn't specify that it can't be there, and indeed
-there is no way in which the \rmrs\ can enforce that no extra material
-can be added.
+tree, which represents the meaning of ``every fat black cat chased
+some dog'', differs from the model in Fig.~\ref{fig:1} in the subtree
+below the restrictor of the universal quantifier.  It is still a model
+of (\ref{ex:cat-pos})---all the EPs are satisfied.  However, there is
+more ``semantic material'' (the ``black'' subformula).  The \rmrs\
+didn't specify that this material should be there; but it also didn't
+specify that it can't be there, and indeed there is no way in which
+the \rmrs\ can enforce that no extra material can be added.
 
@@
   \includegraphics[width=6cm]{pic-more-stuff}
-  \caption{Another tree which satisfies (\ref{eg:cat-pos}b).}
+  \caption{Another tree which satisfies (\ref{ex:cat-pos}).}
   \label{fig:fat-black-cat}
@@
 
-\begin{prop}
+\begin{prop} \label{prop:solved-forms-are-satisfiable}
   Every RMRS in solved form is satisfiable.
@@
 
-\begin{prop}
+\begin{prop} \label{prop:models-satisfy-solved-forms}
   For every tuple $(\tau,\alpha,g,\sigma)$ that satisfies some \rmrs\
@@
 
-\begin{prop}
+\begin{prop}  \label{prop:finite-solved-forms}
   Every {\sc rmrs} $\varphi$ has only a finite number of solved forms, up to

### 2008-09-18 trunk/Papers/rmrs-08/introduction.tex
18/09/2008

@@
 robust wide-coverage semantic construction.  For instance,
-\newcite{bos:etal:2004} exploit {\sc ccg}'s elegant syntax-semantics
+\newcite{bos:etal:2004} exploit {\sc ccg}'s syntax-semantics
 interface to produce a unique logical form for each derivation of a
@@
 
-However, there are certain semantic phenomena that these robust
+There are certain semantic phenomena that these robust
 approaches don't capture reliably, including quantifier scope,
@@
 error-prone.  Depending on the application, it may be preferrable to
-give the parser the option to to leave the decision open when it's not
+give the parser the option to leave the decision open when it's not
 sufficiently informed---i.e., to compute a partial semantic
-representation that neither over-determines nor under-determines the
-semantic information revealed by the (shallow) syntactic analysis, and
-to complete the semantic information later, using information
+representation and to complete it later, using information
 extraneous to the parser.
@@
 deep grammars, such as the English Resource Grammar ({\sc erg},
-\cite{copestake:flickinger:2000}) can be seen as special cases of {\sc
+\cite{copestake:flickinger:2000}) are special cases of {\sc
   rmrs}.
@@
 However, from a formal perspective, \rmrs\ is somewhat ad-hoc in that
-it has no clearly defined model theory; i.e., it is not defined what,
-exactly, any given \rmrs\ representation actually \emph{means}.  The
-key contribution we make is to cast {\sc rmrs}, for the first time, as
-a logic with a well-defined model theory.  This has a number of
+it has no clearly defined model theory; i.e., what exactly, any given
+\rmrs\ representation actually \emph{means} is not defined.  The key
+contribution we make is to cast {\sc rmrs}, for the first time, as a
+logic with a well-defined model theory.  This has a number of
 advantages.  First, it allows us to formally relate \rmrs\ to other
@@
 compatibility of \rmrs\ with the approaches to robust semantic
-analysis mentioned above.  \todo{except we don't do this here --
-  perhaps we should} Second, it provides us with a notion of logical
+analysis mentioned above.  
+Second, it provides us with a notion of logical
 entailment, which we can use to verify the compatibility of \rmrs\ 

### 2008-09-18 trunk/Papers/rmrs-08/motivation.tex
18/09/2008

@@
+\newcommand{\sem}[1]{\mathsf{#1}}
+\newcommand{\sempred}[1]{\mathrm{#1}}
+
 \section{Deep and shallow semantic construction}
@@
 
-We start with an informal discussion of some toy \rmrs s, which are
-meant to illustrate how \rmrs\ can be used to represent partial
+We start with an informal discussion to 
+illustrate how \rmrs\ can be used to represent partial
 semantic information, and why people find it useful for this purpose.
-
-\newcommand{\sem}[1]{\mathsf{#1}}
-\newcommand{\sempred}[1]{\mathrm{#1}}
-
-Consider the following sentence:
+Consider the following (toy) sentence:
 \begin{examples}
@@
 \item $\sem{\_every\_q\_1}(x, \sem{\_fat\_j\_1}(e',x) \wedge
-    \sem{\_cat\_n\_1}(x), \sem{\_some\_q\_1}(y, \sem{\_dog\_n\_1}(y),
-    \sem{\_chased\_v\_1}(e,x,y)))$
+    \sem{\_cat\_n\_1}(x),$\\
+\hspace*{0.1in} $\sem{\_some\_q\_1}(y, \sem{\_dog\_n\_1}(y),$\\
+\hspace*{0.2in}$\sem{\_chased\_v\_1}(e,x,y)))$
 \label{ex:fat-cat-1}
-\item $\sem{\_some\_q\_1}(y, \sem{\_dog\_n\_2}(y),
-\sem{\_every\_q\_1}(x, \sem{\_fat\_j\_1}(e',x) \wedge
-    \sem{\_cat\_n\_2}(x), 
-    \sem{\_chased\_v\_1}(e,x,y)))$
+\item $\sem{\_some\_q\_1}(y, \sem{\_dog\_n\_2}(y),$\\
+\hspace*{0.1in}$\sem{\_every\_q\_1}(x, \sem{\_fat\_j\_1}(e',x) \wedge
+    \sem{\_cat\_n\_2}(x), $\\
+\hspace*{0.2in}$\sem{\_chased\_v\_1}(e,x,y)))$
 \label{ex:fat-cat-2}
@@
 partial, as it will not say anything about predicate-argument
-relations, or resolve lexical ambiguity -- it will use predicate
+relations or resolve lexical ambiguity -- it will use predicate
 symbols such as $\sempred{\_cat\_n}$, which might resolve to the
@@
 complete semantic representation.  (Notice that we are using different
-fonts for the ambiguous and unambiguous predicate symbols.)  We can
+fonts for the ambiguous and unambiguous predicate symbols.)  
+We can
 express this information in \rmrs\ as follows.
@@
 \item \label{ex:cat-pos}
-$l_1:a_1:\sempred{\_every\_q}(x)$, \\
+$l_1:a_1:\sempred{\_every\_q}(x_1)$, \\
 $l_{41}:a_{41}:\sempred{\_fat\_a}(e')$,\\
@@
 $l_5:a_5:\sempred{\_chased\_v}(e_{past})$, \\
-$l_6:a_6:\sempred{\_some\_q}(x_4)$, \\
-$l_9:a_9:\sempred{\_dog\_n}(x_5)$
+$l_6:a_6:\sempred{\_some\_q}(x_6)$, \\
+$l_9:a_9:\sempred{\_dog\_n}(x_7)$
 \end{examples}
 
-This \rmrs\ consists of six \emph{elementary predications} (EPs), one
+This \rmrs\ consists of six \emph{elementary predications} ({\sc ep}s), one
 for each word lemma in the sentence; each of them is prefixed by a
 \emph{label} and an \emph{anchor}, which are essentially variables
-that refer to nodes in the trees in Fig.~\ref{fig:1}.  We can say that
+that refer to nodes in the trees in Fig.~\ref{fig:1}.  
+The lack of semantic dependencies rendered by a {\sc pos} tagger
+also means we have unique arguments to each
+predicate symbol (e.g., see $x_1$ and $x_3$ in (\ref{ex:cat-pos}),
+compared with the {\em same} variable $x$ in the semantic
+representations (\ref{ex:fat-cat-1}) and (\ref{ex:fat-cat-2})).
+We can say that
 the two trees \emph{satisfy} this \rmrs\ because it is possible to map
-the labels and anchors into each tree in such a way that the label of
-the node they are mapped to is consistent with the label specified in
-the elementary predication.  For instance, the first tree satisfies
+the labels and anchors into nodes in each tree and variable names like
+$x_1$ and $x_3$ into variable names in the tree in such a way that the
+predication
+of
+the node that labels and anchors are mapped to is consistent with the
+that given in the {\sc ep}s of the {\sc rmrs}.
+For instance, the first tree satisfies
 the \rmrs\ because we can map $l_1$ and $a_1$ to the root of that
-tree, and the root label $\sem{\_every\_q\_1}$ is consistent with the
-EP predicat $\sempred{\_every\_q}$.
+tree, $x_1$ (and $x_3$) are both mapped to $x$, and the root label
+$\sem{\_every\_q\_1}$ is consistent with the 
+{\sc ep} predicate $\sempred{\_every\_q}$.
 
@@
 given that the semantic information we got from the POS tagger is so
-incomplete.  If we have more information, say information about the
-predicate-argument structure and semantic dependencies from a partial
-parser such as \todo{what? CASS?}, we can represent these in a more
+incomplete.  If we have more information, say information about
+subject and object relations from a chunk parser like Cass
+\cite{abney:1996}, we can represent these in a more
 detailed \rmrs, as follows:
@@
 \item 
-$l_1:a_1:\sempred{\_every\_q}(x)$, \\
+$l_1:a_1:\sempred{\_every\_q}(x_1)$, \\
 $l_{41}:a_{41}:\sempred{\_fat\_a}(e')$,\\
 $l_{42}:a_{42}:\sempred{\_cat\_n}(x_3)$\\
-$l_5:a_5:\sempred{\_chased\_v}(e_{past})$, \\
-$l_6:a_6:\sempred{\_some\_q}(x_4)$, \\
-$l_9:a_9:\sempred{\_dog\_n}(x_5)$
-\todo{add more stuff with ARG and equalities here}
+$l_5:a_5:\sempred{\_chased\_v}(e_{past})$, 
+\hspace*{0.1in} $\ARG1(a_5,x_4),
+\ARG2(a_5,x_5)$\\ 
+$l_6:a_6:\sempred{\_some\_q}(x_6)$, \\
+$l_9:a_9:\sempred{\_dog\_n}(x_7)$\\
+$x_3=x_4$, $x_5=x_7$
 \label{ex:cat-partial-parser}
@@
 
-This \rmrs\ uses two new types of atoms.  Atoms of the form $x=y$
-express that the two variables $x$ and $y$ should really be the same;
-in the example, the \rmrs\ uses two variable names $x$ and $x'$, which
-are mapped to the same variable $x$ in Fig.~\ref{fig:1}.  \todo{I'm
-  making these names up -- adjust to RMRS.}  Atoms of the form
-$\ARG_i(a,x)$ and $\ARG_i(a,h)$ express that a node representing the
-variable $x$ or a node denoted by the \emph{hole} $h$ is the $i$-th
-child of the node to which the anchor $a$ refers.  The separation of
-EPs and $\ARG$ atoms is perhaps the most important difference between
+This \rmrs\ uses two new types of atoms.  Atoms of the form $x_3=x_4$
+express that the two variables $x_3$ and $x_4$ must map to the same
+variable in any fully specific logical form; e.g., to the variable $x$
+in Fig.~\ref{fig:1}.  Atoms of the form
+$\ARG_i(a,x)$ and $\ARG_i(a,h)$ express that the $i$-th child of the
+node to which the anchor $a$ refers is the variable name that $x$ maps
+to (or the node that the {\em hole} $h$ denotes).  The separation of
+{\sc ep}s and $\ARG$ atoms is perhaps the most important difference between
 \rmrs\ and earlier underspecification formalisms such as \mrs\ and
@@
 had to be specified together.  \rmrs\ goes beyond these formalisms in
-expressive power, especially if we allow atoms of the form
+expressive power, because it allows one to state separately
+information about how many arguments a predicate takes, and the values of
+those (non-scopal and scopal) arguments.   If we also allow atoms of the form
 $\ARG_{\{2,3\}}(a,x)$ to express uncertainty as to whether $x$ is the
-second or third child of the anchor $a$.
-
-Notice also that the labels of the EPs for ``fat'' and ``cat'' are
-stipulated to be equal in the \rmrs, whereas the anchors are not.
-In the tree, it is the anchors that are mapped to the nodes with the
-labels $\sem{\_fat\_a\_1}$ and $\sem{\_cat\_n\_1}$; the label is
-mapped to the conjunction node just above them.  In other words, the
-role of the anchor in an EP is to carry the predicate and connect to
-the arguments; the role of the label is to connect the EP to the
-surrounding formula.  The use of equal labels for two EPs stems from
-\mrs, where intersective modification was also modeled by using two
-EPs with the same label.
+second or third child of the anchor $a$, then \rmrs\ can specify the
+arguments to a predicate while underspecifying their position.
 
@@
 \mbox{BODY}(a_1,h_3)$\\ 
-$l_{41}:a_{41}\handel\mbox{\_fat\_j\_1}(e'), \mbox{ARG1}(a_{41},x_2)$\\
+$l_{41}:a_{41}\handel\mbox{\_fat\_j\_1}(e'), \ARG1(a_{41},x_2)$\\
 $l_{42}:a_{42}\handel\mbox{\_cat\_n\_1}(x_3)$\\
 $l_5:a_5\handel\mbox{\_chase\_v\_1}(e_{spast})$,\\
-\hspace*{0.1in}$\mbox{ARG1}(a_5,x_4),
-\mbox{ARG2}(a_5,x_5)$\\ 
+\hspace*{0.1in}$\ARG1(a_5,x_4),
+\ARG2(a_5,x_5)$\\ 
 $l_6:a_6\handel\mbox{\_some\_q\_1}(x_6)$,\\
@@
 $x_5=x_6, x_5=x_7$
-\todo{fix RSTR and BODY to ARG1 and ARG2? or talk about it in the text?}
 \label{ex:cat-erg}
 \end{examples}
-
+RSTR and BODY are conventional names for the scopal arguments to
+predicate symbols corresponding to quantifiers.  
 This is now an extremely detailed underspecified representation of the
 different semantic representations of the sentence, which is
-essentially a notational variant of the \mrs\ that the ERG would
+essentially a notational variant of the \mrs\ that the {\sc erg} would
 derive.  Notice the use of atoms like $h_2 \qeq l_{42}$, which specify
@@
 and can be used to underspecify the scope of the two quantifiers.
+
+Notice also that the labels of the EPs for ``fat'' and ``cat'' are
+stipulated to be equal in the \rmrs\ (\ref{ex:cat-erg}), whereas the
+anchors are not. 
+In the tree, it is the anchors that are mapped to the nodes with the
+labels $\sem{\_fat\_a\_1}$ and $\sem{\_cat\_n\_1}$; the label is
+mapped to the conjunction node just above them.  In other words, the
+role of the anchor in an {\sc ep} is to connect a predicate to its
+arguments, while the role of the label is to connect the {\sc ep} to the
+surrounding formula.  Representing conjunction with label sharing
+stems from \mrs\ and provides compact semantic representations.
 

### 2008-09-19 trunk/Papers/rmrs-08/motivation.tex
leidensdruck

@@
 tools and allows us to represent, process, and compare information
-from different tools in the same formalism.  In the long run, it could
-also be a platform for incrementally enriching partial information
+from different tools in the same formalism, and in the long run
+provide a platform for incrementally enriching partial information
 such as (\ref{ex:cat-pos}) in order to eventually approximate more
@@
 
-\todo{Leidensdruck!}
+However, although people have started using \rmrs\ in their
+implemented systems, it is still relatively ad-hoc in that its formal
+syntax is not consistent across different papers, and its formal
+semantics is not defined; we don't know, formally, what an \rmrs\
+\emph{means} in terms of semantic representations like
+(\ref{ex:fat-cat-1}) and (\ref{ex:fat-cat-2}), and this hinders our
+ability to come up with efficient algorithms for processing \rmrs. The
+purpose of this paper is to lay the groundwork for fixing this problem.
 

### 2008-09-22 trunk/Papers/rmrs-08/conclusion.tex
version for Ann

@@
 is designed to be maximally flexible on the type of semantic
-information that can be left underspecified -- it can underspecify
-lexical ambiguity, predicate-argument relations, and scope --, so that
+information that can be left underspecified, so that
 the semantic output of a shallow parser needn't over-determine or
@@
 checking.  We have argued that entailment is a central notion in
-comparing the outputs of different NLP tools, and provided a syntactic
+comparing the outputs of different {\sc nlp} tools, and provided a syntactic
 characterization for \rmrs\ entailment, which we hope will allow us to

### 2008-09-22 trunk/Papers/rmrs-08/entailment.tex
version for Ann

@@
   solved form $S'$ of $\varphi'$ such that $S$ is an extension of
-  $S'$. \todo{We don't know if this is true for qeq; the proof is
-    written for $\dom$.}  \todo{Also note that this is about a version
-    of entailment in which $\varphi$ and $\varphi'$ contain the same
-    variables, which makes the claim a bit weaker.}
+  $S'$. \todo{Check proof holds for this version.}
+% Do we have a proof where $\varphi$ and $\varphi'$ can vary on
+% alphabetic variance on nodes and base variables?
 \end{thm}
@@
 is the proof that the result is a solved form of $\varphi'$; this step
-involves proving that if $\varphi \models \varphi'$, then all EPs in
+involves proving that if $\varphi \models \varphi'$, then all {\sc ep}s in
 $\varphi'$ also occur in $\varphi$.  

### 2008-09-22 trunk/Papers/rmrs-08/introduction.tex
version for Ann

@@
 error-prone.  Depending on the application, it may be preferrable to
-give the parser the option to leave the decision open when it's not
+give the parser the option to leave a semantic decision open when it's not
 sufficiently informed---i.e., to compute a partial semantic
@@
 
-In this paper, we focus on one particular approach to representing
-partial semantic information: Robust Minimal Recursion Semantics
-(\rmrs, \cite{copestake:2003}).  \rmrs\ is a generalisation of \mrs\
-\cite{copestake:etal:2005} that is designed to support
-underspecification of scope and predicate-argument structure.  It is
-an emerging standard representation language for partial semantic
-information, and has been applied in several implemented systems.  For
-instance, \newcite{copestake:2003} and \newcite{frank:2004} use it to
-specify semantic components to shallow parsers ranging in depth from
-{\sc pos} taggers to chunk parsers and intermediate parsers such as
-{\sc rasp} \cite{briscoe:etal:2006}.  {\sc mrs} analyses derived from
-deep grammars, such as the English Resource Grammar ({\sc erg},
-\cite{copestake:flickinger:2000}) are special cases of {\sc
-  rmrs}.
+In this paper, we focus on an approach to representing partial
+semantic information that supports this strategy: Robust Minimal
+Recursion Semantics (\rmrs, \cite{copestake:2003}).  \rmrs\ is a
+generalisation of \mrs\ \cite{copestake:etal:2005} that is designed to
+support underspecification of scope and predicate-argument structure.
+It is an emerging standard for representing partial
+semantic information, and has been applied in several implemented
+systems.  For instance, \newcite{copestake:2003} and
+\newcite{frank:2004} use it to specify semantic components to shallow
+parsers ranging in depth from {\sc pos} taggers to chunk parsers and
+intermediate parsers such as {\sc rasp} \cite{briscoe:etal:2006}.
+{\sc mrs} analyses derived from deep grammars, such as the English
+Resource Grammar ({\sc erg}, \cite{copestake:flickinger:2000}) are
+special cases of {\sc rmrs}.
 
 However, from a formal perspective, \rmrs\ is somewhat ad-hoc in that
-it has no clearly defined model theory; i.e., what exactly, any given
+it has no clearly defined model theory; i.e., what exactly any given
 \rmrs\ representation actually \emph{means} is not defined.  The key
@@
 and challenges with deep and shallow semantic construction.
-Section~\ref{sec:rmrs} defines the syntax and model-theory of \rmrs\.
+Section~\ref{sec:rmrs} defines the syntax and model-theory of \rmrs.
 In Section~\ref{sec:entailment} we demonstrate the practical

### 2008-09-22 trunk/Papers/rmrs-08/motivation.tex
version for Ann

@@
 ambiguities of the words ``cat'' and ``dog'' (which have 8 and 7 noun
-senses according to Wordnet, respectively).  Two of these readings are
+senses according to Wordnet, respectively).  Simplifying slightly by
+ignoring tense information (but see
+Section~\ref{sec:extensions}), two of these readings are
 shown as logical forms below; they can be represented as trees as
@@
 \hspace*{0.1in} $\sem{\_some\_q\_1}(y, \sem{\_dog\_n\_1}(y),$\\
-\hspace*{0.2in}$\sem{\_chased\_v\_1}(e,x,y)))$
+\hspace*{0.2in}$\sem{\_chase\_v\_1}(e,x,y)))$
 \label{ex:fat-cat-1}
@@
     \sem{\_cat\_n\_2}(x), $\\
-\hspace*{0.2in}$\sem{\_chased\_v\_1}(e,x,y)))$
+\hspace*{0.2in}$\sem{\_chase\_v\_1}(e,x,y)))$
 \label{ex:fat-cat-2}
@@
 partial, as it will not say anything about predicate-argument
-relations or resolve lexical ambiguity -- it will use predicate
+relations or resolve lexical ambiguity---it will use predicate
 symbols such as $\sempred{\_cat\_n}$, which might resolve to the
@@
 $l_{42}:a_{42}:\sempred{\_cat\_n}(x_3)$\\
-$l_5:a_5:\sempred{\_chased\_v}(e_{past})$, \\
+$l_5:a_5:\sempred{\_chase\_v}(e)$, \\
 $l_6:a_6:\sempred{\_some\_q}(x_6)$, \\
@@
 semantic representations such as (\ref{ex:fat-cat-1})) that are
-described equally well by the same \rmrs; this is not surprising,
-given that the semantic information we got from the POS tagger is so
+described equally well by the \rmrs\ (\ref{ex:cat-pos}); this is
+not surprising, 
+given that the semantic information we got from the {\sc pos} tagger is so
 incomplete.  If we have more information, say information about
@@
 $l_{42}:a_{42}:\sempred{\_cat\_n}(x_3)$\\
-$l_5:a_5:\sempred{\_chased\_v}(e_{past})$, 
+$l_5:a_5:\sempred{\_chase\_v}(e)$, \\
 \hspace*{0.1in} $\ARG1(a_5,x_4),
@@
 in Fig.~\ref{fig:1}.  Atoms of the form
-$\ARG_i(a,x)$ and $\ARG_i(a,h)$ express that the $i$-th child of the
-node to which the anchor $a$ refers is the variable name that $x$ maps
-to (or the node that the {\em hole} $h$ denotes).  The separation of
+$\ARG_i(a,z)$ and $\ARG_i(a,h)$ express that the $i$-th child
+(counting from 0) of the
+node to which the anchor $a$ refers is the variable name that $z$ denotes
+(or the node that the {\em hole} $h$ denotes).  The separation of
 {\sc ep}s and $\ARG$ atoms is perhaps the most important difference between
@@
 expressive power, because it allows one to state separately
-information about how many arguments a predicate takes, and the values of
+information about how many arguments a predicate takes (which is
+necessary for supporting parsers where information about lexical
+subcategorisation is absent), and the values of
 those (non-scopal and scopal) arguments.   If we also allow atoms of the form
@@
 
-Finally, here is an \rmrs\ that could be derived from a deep grammar
-such as the ERG:
-
+Finally, \rmrs\ (\ref{ex:cat-erg}) is a notational variant of the {\sc
+  mrs} derived by the {\sc erg}, a wide-coverage deep grammar:
 \begin{examples}
@@
 $l_{42}:a_{42}\handel\mbox{\_cat\_n\_1}(x_3)$\\
-$l_5:a_5\handel\mbox{\_chase\_v\_1}(e_{spast})$,\\
+$l_5:a_5\handel\mbox{\_chase\_v\_1}(e)$,\\
 \hspace*{0.1in}$\ARG1(a_5,x_4),
@@
 predicate symbols corresponding to quantifiers.   
-This is now an extremely detailed underspecified representation of the
-different semantic representations of the sentence, which is
-essentially a notational variant of the \mrs\ that the {\sc erg} would
-derive.  Notice the use of atoms like $h_2 \qeq l_{42}$, which specify
+Atoms like $h_2 \qeq l_{42}$ specify
 a kind of ``outscopes'' relationship between the hole and the label,
 and can be used to underspecify the scope of the two quantifiers.
-
-Notice also that the labels of the EPs for ``fat'' and ``cat'' are
+Notice also that the labels of the {\sc ep}s for ``fat'' and ``cat'' are
 stipulated to be equal in the \rmrs\ (\ref{ex:cat-erg}), whereas the
@@
 stems from \mrs\ and provides compact semantic representations.
+Finally, (\ref{ex:cat-erg}) intuitively represents more specific
+semantic information than (\ref{ex:cat-partial-parser}), so long as
+the base-language constructors that \_cat\_n\_1 can correspond to is a
+subset of those that \_cat\_n can correspond to (written
+$\mbox{\_cat\_n\_1}\sqsubseteq \mbox{\_cat\_n}$), and similalry for
+\_fat\_j\_1 etc.  This is because with these so-called SPEC relations
+on the predicate symbols in place, (\ref{ex:cat-erg}) simply adds
+ARGs, qeqs and variable equalities to (\ref{ex:cat-partial-parser}).
 
@@
 available, it can be used to capture whatever partial semantic
-information a shallower NLP tool was able to provide.  \rmrs\ is
+information a shallower {\sc nlp} tool was able to provide.  \rmrs\ is
 designed to serve as a generic formalism for partial semantic
-representations which abstracts over the details of different NLP
+representations which abstracts over the details of different {\sc nlp}
 tools and allows us to represent, process, and compare information
@@
 such as (\ref{ex:cat-pos}) in order to eventually approximate more
-detailed representations such as (\ref{ex:cat-partial-parser}).
+detailed representations such as (\ref{ex:cat-partial-parser}) or
+(\ref{ex:cat-erg}).
 

### 2008-09-22 trunk/Papers/rmrs-08/rmrs.tex
version for Ann

@@
 \rmrs, and conclude this section by carrying over the notion of
-\emph{solved forms} from CLLS \cite{egg:etal:2001}.
+\emph{solved forms} from {\sc clls} \cite{egg:etal:2001}.
 
@@
 
-We define the syntax of \rmrs\ in the style of CLLS
+We define the syntax of \rmrs\ in the style of {\sc clls}
 \cite{egg:etal:2001}.  We assume an infinite set of {\em node
@@
 
-A node variable $X$ is called a \emph{label} in the \rmrs\
-$\varphi$ iff $\varphi$ contains an atom of the form $\ARG_S(X,i)$ or
-an elementary predication of the form $\ep{Y}{X}{P}$; $X$ is called an
-\emph{anchor} iff $\varphi$ contains an atom of the form
-$\ep{X}{Y}{P}$ or $Y \qeq X$; and $X$ is called a \emph{hole} iff
+A node variable $X$ is called a \emph{label} iff $\varphi$ contains an
+atom of the form 
+$\ep{X}{Y}{P}$ or $Y \qeq X$; $X$ is called an
+ a \emph{anchor} in the \rmrs\
+$\varphi$ iff $\varphi$ contains an at atom of the form
+$\ep{Y}{X}{P}$ or $\ARG_S(X,i)$; and $X$ is
+called a \emph{hole} iff 
 $\varphi$ contains an atom of the form $\ARG_S(Y,X)$ or $X \qeq Y$.
@@
 presentations of \rmrs\ (e.g., \cite{copestake:2003}) and to dominance
-constraints \cite{egg:etal:2001}.  We $\ARG_{\{i\}}$ (with a singleton
+constraints \cite{egg:etal:2001}.  We use $\ARG_{\{i\}}$ (with a singleton
 set $S$) instead of $\ARG_i$, and $\ARG_\N$ instead of $\ARG_n$.
@@
 
-\todo{we don't split stuff up into tuple parts, which earlier rmrs
-  did}
-
 \subsection{Model Theory}
@@
 expressed in the base language.  We represent such a logical form as a
-tree $\tau$, such as the one in Fig.~\ref{fig:1}.  So models of the
+tree $\tau$, such as the ones in Fig.~\ref{fig:1}.  So models of the
 \rmrs\ are such trees, and we can define satisfaction and entailment
@@
 In this paper, we assume for simplicity that the base language is
-predicate logic; in this case, $\tau$ is simply the structure tree of
+predicate logic (e.g., see (\ref{ex:fat-cat-1}) and
+(\ref{ex:fat-cat-2})); in this case, $\tau$ is simply the structure
+tree of 
 the predicate logic formula.  We assume that $\Sigma$ is a ranked
@@
 binary constructors $\forall x$ and $\exists x$ for all base-language
-variables $x$; all the constants and base-language variables as
+variables $x$; all the constants and base-language variables are
 nullary constructors; and, for each predicate symbol $P$ of arity $n$,
@@
 $\mathsf{sans{-}serif}$, as opposed to the \rmrs\ predicate symbols
-like ``$\sempred{\_cat\_n}$''.
+like ``$\sempred{\_cat\_n}$'' and ``$\sempred{\_cat\_n\_1}$''.
 
@@
 suppose that $\tau_j,\alpha_j,g_j,\sigma\models \varphi$ and let
-$\nvar$ be the node variables from which $\varphi$ is built.  Then
+$\nvar$ and $\ovar$ be the node and base variables from which
+$\varphi$ is built.  Then 
 $\varphi$ exhibits a semantic scope ambiguity if there are variables
@@
 a variable $ Y \in \nvar$ such that $\tau_1(\alpha_1(Y))$ is a
-constructor of a different type from $\tau_2(\alpha_2(Y))$ \todo{what
-  does this mean?}.  And it exhibits missing lexical sense information
+constructor of a different type from $\tau_2(\alpha_2(Y))$  (i.e., the
+constructors are of a different arity or and/or their arguments are).
+And it exhibits missing lexical sense information
 if $\tau_1(\alpha_1(Y))$ and $\tau_2(\alpha_2(Y))$ are different
@@
 Let's look at the \rmrs\ in (\ref{ex:cat-pos}) by way of example.
-This \rmrs\ is satisfied by the tree in Fig.~\ref{fig:1} (among
+This \rmrs\ is satisfied by the trees in Fig.~\ref{fig:1} (among
 others) together with some particular $\alpha$, $g$, and $\sigma$.
-The assignment $\alpha$ will map the variables $l_1$ and $a_1$ to the
-root node; $l_{41}$ and $l_{42}$ to its second child (labeled with
-``$\wedge$''); $a_{41}$ to the first child of \emph{that} node (i.e.\
-the node 21, labelled with ``fat'') and $a_{42}$ to the node 22; and
-so forth.  $g$ will map $x$ and $x_3$ to $x$, $e'$ to $e$, and so on.
+For instance, consider the left-hand side tree in Fig.~\ref{fig:1}.
+The \rmrs\ (\ref{ex:cat-pos}) satisfies this tree with an assignment
+function $\alpha$ that maps the variables $l_1$ and $a_1$ to the
+root node, $l_{41}$ and $l_{42}$ to its second child (labeled with
+``$\wedge$''), $a_{41}$ to the first child of \emph{that} node (i.e.\
+the node 21, labelled with ``fat'') and $a_{42}$ to the node 22, and
+so forth.  $g$ will map $x_1$ and $x_3$ to $x$, and $x_6$ and $x_7$ to
+$y$, and so on.  
 And $\sigma$ will map each \rmrs\ predicate symbol (which represents a
 word) to the set of its word senses, e.g.\ \_cat\_n to a set
-containing \_cat\_n\_1 and possibly others.  It is then easy to verify
-that every single atom in the \rmrs\ is satisfied -- most
-interestingly, the EPs $\ep{l_{41}}{a_{41}}{\_fat\_a(e')}$ and
+containing $\sem{\_cat\_n\_1}$ and possibly others.  It is then easy to verify
+that every single atom in the \rmrs\ is satisfied---most
+interestingly, the {\sc ep}s $\ep{l_{41}}{a_{41}}{\_fat\_a(e')}$ and
 $\ep{l_{42}}{a_{42}}{\_cat\_n(x_3)}$ are satisfied because
@@
 below the restrictor of the universal quantifier.  It is still a model
-of (\ref{ex:cat-pos})---all the EPs are satisfied.  However, there is
+of (\ref{ex:cat-pos})---all the {\sc ep}s are satisfied.  However, there is
 more ``semantic material'' (the ``black'' subformula).  The \rmrs\
@@
 Intuitively, we expect that an \rmrs\ in solved form is fully
-specified with respect to scope ambiguities and the predicate-argument
+specified with respect to the predicate-argument
 structure, all variable equalities and inequalities have been
-resolved, and only lexical sense ambiguities remain.  This is made
+resolved, and only lexical sense ambiguities and (resolvable)
+scope ambiguities remain.  This is made
 precise below.
-
-\todo{Note: We don't really talk about SPEC anywhere else.}
 
@@
     are no two atoms $\ARG_S(Y,i)$ and $\ARG_S(Y,i')$ in $\varphi$;
-  \item if $Y$ is an anchor in some EP $\ep{X}{Y}{P}$ and $k$ is the
+  \item if $Y$ is an anchor in some {\sc ep} $\ep{X}{Y}{P}$ and $k$ is the
     maximum number such that $\ARG_{\{k\}}(X,i)$ is in $\varphi$ for
@@
   \cite{Duchier00dominanceconstraints})] 
-  For each EP, we choose to label the anchor with the constructor $p$
+  For each {\sc ep}, we choose to label the anchor with the constructor $p$
   of sufficiently high arity whose existence we assumed; we determine
@@
   \begin{enumerate}
-  \item $\varphi'$ contains the EP $\ep{X'}{Y'}{P}$ iff there are
+  \item $\varphi'$ contains the {\sc ep} $\ep{X'}{Y'}{P}$ iff there are
     variables $X,Y$ such that $\ep{X}{Y}{P}$ is in $\varphi$, $X' =
@@
   substitutions on the node and base variables of $\varphi$.  Let $s$
-  be such a substitution.  This fixes the set of EPs of any solved
+  be such a substitution.  This fixes the set of {\sc ep}s of any solved
   form of $\varphi$ that is based on $s$ uniquely.  There is only a
@@
 \subsection{Extensions}
+\label{sec:extensions}
 

### 2008-10-06 trunk/Papers/rmrs-08/conclusion.tex
Monday update

@@
 the semantic output of a shallow parser needn't over-determine or
-under-determine the semantic information that can be extracted from
+under-determine the semantics that can be extracted from
 the shallow syntactic analysis.  This lays the formal foundations for
-a semantic representation formalism that is emerging as a standard in
+a formalism that is emerging as a standard in
 robust semantic processing.
@@
 comparing the outputs of different {\sc nlp} tools, and provided a syntactic
-characterization for \rmrs\ entailment, which we hope will allow us to
+characterisation for \rmrs\ entailment, which we hope will allow us to
 develop efficient entailment algorithms in the future.

### 2008-10-06 trunk/Papers/rmrs-08/entailment.tex
Monday update

@@
-\section{RMRS comparison as entailment}
+\section{RMRS Comparison as Entailment}
 \label{sec:entailment}
 
-The definitions in the previous section serve an important theoretical
-purpose in that they provide a formal underpinning for \rmrs, which so
-far has been used for partial semantic representations in implemented
-systems despite the lack of a formal definition.  Next to the peace of
+The definitions from Section~\ref{sec:rmrs} serve an important theoretical
+purpose in that they provide a formal underpinning \rmrs's existing use
+for partial semantic representations in implemented
+systems.  Next to the peace of
 mind that comes with the use of a formally understood formalism, a
@@
 consistent with an \rmrs\ $\varphi$ that was computed by a deep
-grammar amounts to checking \emph{entailment} between the two \rmrs s,
-i.e.\ whether $\varphi \models \varphi'$.  If we could establish
+grammar if $\varphi$ entails $\varphi'$ (i.e., $\varphi \models
+\varphi'$).  If we could establish 
 efficiently that $\varphi'$ is entailed by $\varphi$, this would tell
@@
 this paper.  However, we propose the following novel syntactic
-characterization of entailment.
+characterisation of entailment.
 
@@
   solved form $S'$ of $\varphi'$ such that $S$ is an extension of
-  $S'$. \todo{Check proof holds for this version.}
-% Do we have a proof where $\varphi$ and $\varphi'$ can vary on
-% alphabetic variance on nodes and base variables?
+  $S'$. 
 \end{thm}
@@
 
-This characterization does not by itself provide an efficient test for
+This characterisation does not by itself provide an efficient test for
 entailment because the number of solved forms of $\varphi$ and

### 2008-10-06 trunk/Papers/rmrs-08/introduction.tex
Monday update

@@
 ill-formed phrases. 
-
-On the other hand, there has been a recent trend towards systems for
+There has thus been a trend recently towards systems for
 robust wide-coverage semantic construction.  For instance,
@@
 statistical {\sc ccg} parser; \newcite{zettlemoyer:collins:2007} and
-others learn string-to-logical-form mappings, exploting informative
+others learn string-to-logical-form mappings, exploiting informative
 features from shallow language processors; and
-\newcite{copestake:2003} and \newcite{frank:2004} endow robust
+\newcite{copestake:2003} and \newcite{frank:2004} endow shallow
 language processors with semantic components.
@@
 \newcite{clark:etal:2004} report that the parser used by
-\newcite{bos:etal:2004} yields only 63\% accuracy on object
-extraction; e.g., {\em the man that I met\dots}).  Forcing a shallow
+\newcite{bos:etal:2004} yields 63\% accuracy on object
+extraction; e.g., {\em the man that I met\dots}).  Forcing a robust
 parser to make a decision about these phenomena can therefore be
-error-prone.  Depending on the application, it may be preferrable to
+error-prone.  Depending on the application, it may be preferable to
 give the parser the option to leave a semantic decision open when it's not
@@
 semantic information that supports this strategy: Robust Minimal
-Recursion Semantics (\rmrs, \cite{copestake:2003}).  \rmrs\ is a
-generalisation of \mrs\ \cite{copestake:etal:2005} that is designed to
+Recursion Semantics (\rmrs, \cite{copestake:2003}).  \rmrs\ is
+designed to
 support underspecification of scope and predicate-argument structure.
@@
 intermediate parsers such as {\sc rasp} \cite{briscoe:etal:2006}.
-{\sc mrs} analyses derived from deep grammars, such as the English
+{\sc mrs} analyses \cite{copestake:etal:2005}
+derived from deep grammars, such as the English
 Resource Grammar ({\sc erg}, \cite{copestake:flickinger:2000}) are
@@
 
-Robost language processors that produce a single conventional logical
+Robust language processors that produce a single conventional logical
 form for a given natural language string are beginning to emerge

### 2008-10-06 trunk/Papers/rmrs-08/motivation.tex
Monday update

@@
 ambiguity between ``every fat cat'' and ``some dog'' and lexical
-ambiguities of the words ``cat'' and ``dog'' (which have 8 and 7 noun
-senses according to Wordnet, respectively).  Simplifying slightly by
+ambiguities of the nouns ``cat'' and ``dog'' (which have 8 and 7 
+WordNet senses respectively).  Simplifying slightly by
 ignoring tense information (but see
@@
 
-Now imagine that we are trying to extract semantic information from
+Now imagine trying to extract semantic information from
 the output of a part-of-speech tagger by using the word lemmas as
@@
 predicate symbols $\sem{\_cat\_n\_1}$ or $\sem{\_cat\_n\_2}$ in the
-complete semantic representation.  (Notice that we are using different
+complete semantic representation.  (Notice the different
 fonts for the ambiguous and unambiguous predicate symbols.)  
@@
 $l_1:a_1:\sempred{\_every\_q}(x_1)$, \\
-$l_{41}:a_{41}:\sempred{\_fat\_a}(e')$,\\
+$l_{41}:a_{41}:\sempred{\_fat\_j}(e')$,\\
 $l_{42}:a_{42}:\sempred{\_cat\_n}(x_3)$\\
@@
 The lack of semantic dependencies rendered by a {\sc pos} tagger
-also means we have unique arguments to each
+means we have unique arguments to each
 predicate symbol (e.g., see $x_1$ and $x_3$ in (\ref{ex:cat-pos}),
@@
 of
-the node that labels and anchors are mapped to is consistent with the
+the node that labels and anchors are mapped to is consistent with
 that given in the {\sc ep}s of the {\sc rmrs}.
 For instance, the first tree satisfies
-the \rmrs\ because we can map $l_1$ and $a_1$ to the root of that
-tree, $x_1$ (and $x_3$) are both mapped to $x$, and the root label
+the \rmrs\ because $l_1$ and $a_1$ can map to the root of that
+tree and $x_1$ to $x$, and the root label
 $\sem{\_every\_q\_1}$ is consistent with the 
@@
 In the tree, it is the anchors that are mapped to the nodes with the
-labels $\sem{\_fat\_a\_1}$ and $\sem{\_cat\_n\_1}$; the label is
+labels $\sem{\_fat\_j\_1}$ and $\sem{\_cat\_n\_1}$; the label is
 mapped to the conjunction node just above them.  In other words, the
@@
 Finally, (\ref{ex:cat-erg}) intuitively represents more specific
-semantic information than (\ref{ex:cat-partial-parser}), so long as
-the base-language constructors that \_cat\_n\_1 can correspond to is a
+semantic information than (\ref{ex:cat-pos}), so long as
+the constructors that \_cat\_n\_1 can correspond to is a
 subset of those that \_cat\_n can correspond to (written
-$\mbox{\_cat\_n\_1}\sqsubseteq \mbox{\_cat\_n}$), and similalry for
+$\mbox{\_cat\_n\_1}\sqsubseteq \mbox{\_cat\_n}$), and similarly for
 \_fat\_j\_1 etc.  This is because with these so-called SPEC relations
 on the predicate symbols in place, (\ref{ex:cat-erg}) simply adds
-ARGs, qeqs and variable equalities to (\ref{ex:cat-partial-parser}).
+ARGs, qeqs and variable equalities to (\ref{ex:cat-pos}).
 
 In summary, \rmrs\ is a formalism for representing partial information
-about semantic representations.  It generalizes \mrs\ in that it can
+about semantic representations.  It can
 represent the output of a deep grammar in a way that is a notational
@@
 available, it can be used to capture whatever partial semantic
-information a shallower {\sc nlp} tool was able to provide.  \rmrs\ is
-designed to serve as a generic formalism for partial semantic
+information a shallower {\sc nlp} tool can provide.  \rmrs\ is
+designed to be a generic formalism for partial semantic
 representations which abstracts over the details of different {\sc nlp}
@@
 (\ref{ex:fat-cat-1}) and (\ref{ex:fat-cat-2}), and this hinders our
-ability to come up with efficient algorithms for processing \rmrs. The
+ability to design efficient algorithms for processing \rmrs. The
 purpose of this paper is to lay the groundwork for fixing this problem.

### 2008-10-06 trunk/Papers/rmrs-08/rmrs.tex
Monday update

@@
 \cite{egg:etal:2001}.  We assume an infinite set of {\em node
-  variables} $\nvar=\{X,Y,X_1,\ldots\}$, which we will use as labels,
+  variables} $\nvar=\{X,Y,X_1,\ldots\}$, used as labels,
 anchors, and holes; the distinction between these uses will come from
@@
 we assume a vocabulary of \emph{predicate symbols}
-$\pred=\{P,Q,P_1,\ldots\}$, each of which has an \emph{arity} $\ar(P)
-\geq 0$.
+$\pred=\{P,Q,P_1,\ldots\}$.
 
@@
 $\ep{X}{Y}{P}$ or $Y \qeq X$; $X$ is called an
- a \emph{anchor} in the \rmrs\
-$\varphi$ iff $\varphi$ contains an at atom of the form
+\emph{anchor} in the \rmrs\
+$\varphi$ iff $\varphi$ contains an atom of the form
 $\ep{Y}{X}{P}$ or $\ARG_S(X,i)$; and $X$ is
@@
 set $S$) instead of $\ARG_i$, and $\ARG_\N$ instead of $\ARG_n$.
-Furthermore, the elementary predication $\ep{l}{a}{P(v)}$ is a
-notational variant of $\{\ep{l}{a}{P}, \ARG_{\{0\}}(l,v)\}$.  In other
+The elementary predication $\ep{l}{a}{P(v)}$ (see
+(\ref{ex:cat-pos})--(\ref{ex:cat-erg})) is a
+notational variant of $\{\ep{l}{a}{P}, \ARG_{\{0\}}(a,v)\}$.  In other
 words, the syntax defined here differs in some details, but is more
@@
 dominance.  Redefining \rmrs\ in the style of dominance constraints
-will simplify the formal work below.
+simplifies the formal work below.
 
@@
   symbols $\tau(u), \tau(u \cdot 2), \ldots, \tau(u \cdot 2^{k-1})$
-  are quantifiers.  (This is the ``qeq'' relation familiar from the
-  \mrs\ and \rmrs\ literature.)
+  are quantifiers.  (This is the ``qeq'' relation familiar from
+  \newcite{copestake:etal:2005}).
 
@@
 \qquad \alpha(X) \cdot i \in D(\tau), \alpha(X) \cdot
-  i = \alpha(Y) \mbox{ and }\\
-\qquad
-s(\alpha(X) \cdot i) = h \\
+  i = \alpha(Y)\\
   \tau, \alpha, g, \sigma \models X \qeq Y
@@
   \tau, \alpha, g, \sigma \models P \spec Q
-\mbox{iff $\sigma(P) \subseteq \sigma(Q)$}
+\mbox{ iff $\sigma(P) \subseteq \sigma(Q)$}
 \end{array}
@@
 the \rmrs\ to be a partial description of each of these trees.  In
-particular, \rmrs s may represent semantic scope ambiguities, missing
-information about semantic dependencies, and missing information about
+particular, \rmrs s may represent semantic scope ambiguities, and/or missing
+information about semantic dependencies, 
 lexical subcategorisation and lexical senses.  For $j=\{1,2\}$,
@@
 $\alpha_2(Y')\qeqdom \alpha_2(Y)$.  It exhibits missing information
-about semantic dependencies if there are base-language variables $
-v,v'\in \ovar$ such that $g_1(v)=g_1(v')$ and $g_2(v)\neq g_2(v')$.
+about semantic dependencies if there are base-language variables
+$v,v'\in \ovar$ such that $g_1(v)=g_1(v')$ and $g_2(v)\neq g_2(v')$. 
 It exhibits missing lexical subcategorisation information if there is
-a variable $ Y \in \nvar$ such that $\tau_1(\alpha_1(Y))$ is a
+a $ Y \in \nvar$ such that $\tau_1(\alpha_1(Y))$ is a
 constructor of a different type from $\tau_2(\alpha_2(Y))$  (i.e., the
-constructors are of a different arity or and/or their arguments are).
+constructors are of a different arity and/or their arguments are).
 And it exhibits missing lexical sense information
@@
 tree, which represents the meaning of ``every fat black cat chased
-some dog'', differs from the model in Fig.~\ref{fig:1} in the subtree
-below the restrictor of the universal quantifier.  It is still a model
+some dog'', is still a model
 of (\ref{ex:cat-pos})---all the {\sc ep}s are satisfied.  However, there is
@@
 specify that it can't be there, and indeed there is no way in which
-the \rmrs\ can enforce that no extra material can be added.
+the \rmrs\ can enforce that no extra material can be added (e.g.,
+consider predicate symbols that are introduced by construction rules
+in a deep grammar, which by design are absent from the semantic output
+of a {\sc pos} tagger).
 
@@
 
-\todo{At this point we should have some examples of solved forms.  The
-  running example is a bit lame here because it already is in solved
-  form, and therefore only has one solved form.}
+\todo{Draw a figure for (\ref{ex:cat-erg}), and explain that the trees
+  in Fig 1 are its solved forms.}
 

### 2008-10-06 trunk/Papers/rmrs-08/motivation.tex
picture fixes and font change

@@
 
-\begin{figure*}[t]
-\includegraphics[width=6cm]{pic-cat-chased-dog}
-\hspace{1cm}
-\includegraphics[width=6cm]{pic-cat-chased-dog-2}
+\begin{figure}[t]
+\centering
+\includegraphics[scale=0.8]{pic-cat-chased-dog} \\
+\includegraphics[scale=0.8]{pic-cat-chased-dog-2}
 \caption{Structure trees of the semantic representations (\ref{ex:fat-cat-1}) and
   (\ref{ex:fat-cat-2}). \label{fig:1}}
-\end{figure*}
+\end{figure}
 
@@
 $l_5:a_5:\sempred{\_chase\_v}(e)$, \\
-\hspace*{0.1in} $\ARG1(a_5,x_4),
-\ARG2(a_5,x_5)$\\ 
+\hspace*{0.1in} $\ARG_1(a_5,x_4),
+\ARG_2(a_5,x_5)$\\ 
 $l_6:a_6:\sempred{\_some\_q}(x_6)$, \\
@@
 \begin{examples}
-\item $l_1:a_1\handel\mbox{\_every\_q\_1}(x_1),$\\
-\hspace*{0.1in}$\mbox{RSTR}(a_1,h_2),
-\mbox{BODY}(a_1,h_3)$\\ 
-$l_{41}:a_{41}\handel\mbox{\_fat\_j\_1}(e'), \ARG1(a_{41},x_2)$\\
-$l_{42}:a_{42}\handel\mbox{\_cat\_n\_1}(x_3)$\\
-$l_5:a_5\handel\mbox{\_chase\_v\_1}(e)$,\\
-\hspace*{0.1in}$\ARG1(a_5,x_4),
-\ARG2(a_5,x_5)$\\ 
-$l_6:a_6\handel\mbox{\_some\_q\_1}(x_6)$,\\
-\hspace*{0.1in}$\mbox{RSTR}(a_6,h_7),
-\mbox{BODY}(a_6,h_8)$\\ 
-$l_9:a_9\handel\mbox{\_dog\_n\_1}(x_7)$\\
+\item $l_1:a_1\handel\sempred{\_every\_q\_1}(x_1),$\\
+\hspace*{0.1in}$\mathsf{RSTR}(a_1,h_2),
+\mathsf{BODY}(a_1,h_3)$\\ 
+$l_{41}:a_{41}\handel\sempred{\_fat\_j\_1}(e'), \ARG_1(a_{41},x_2)$\\
+$l_{42}:a_{42}\handel\sempred{\_cat\_n\_1}(x_3)$\\
+$l_5:a_5\handel\sempred{\_chase\_v\_1}(e)$,\\
+\hspace*{0.1in}$\ARG_1(a_5,x_4),
+\ARG_2(a_5,x_5)$\\ 
+$l_6:a_6\handel\sempred{\_some\_q\_1}(x_6)$,\\
+\hspace*{0.1in}$\mathsf{RSTR}(a_6,h_7),
+\mathsf{BODY}(a_6,h_8)$\\ 
+$l_9:a_9\handel\sempred{\_dog\_n\_1}(x_7)$\\
 $h_2=_q l_{42}, l_{41}=l_{42}, h_7 =_q l_9$\\
@@
 \end{examples}
-RSTR and BODY are conventional names for the scopal arguments to
-predicate symbols corresponding to quantifiers.   
-Atoms like $h_2 \qeq l_{42}$ specify
-a kind of ``outscopes'' relationship between the hole and the label,
-and can be used to underspecify the scope of the two quantifiers.
-Notice also that the labels of the {\sc ep}s for ``fat'' and ``cat'' are
-stipulated to be equal in the \rmrs\ (\ref{ex:cat-erg}), whereas the
-anchors are not. 
+$\mathsf{RSTR}$ and $\mathsf{BODY}$ are conventional names for the
+scopal arguments to predicate symbols corresponding to quantifiers.
+Atoms like $h_2 \qeq l_{42}$ specify a kind of ``outscopes''
+relationship between the hole and the label, and can be used to
+underspecify the scope of the two quantifiers.  Notice also that the
+labels of the {\sc ep}s for ``fat'' and ``cat'' are stipulated to be
+equal in the \rmrs\ (\ref{ex:cat-erg}), whereas the anchors are not.
 In the tree, it is the anchors that are mapped to the nodes with the
@@
 role of the anchor in an {\sc ep} is to connect a predicate to its
-arguments, while the role of the label is to connect the {\sc ep} to the
-surrounding formula.  Representing conjunction with label sharing
+arguments, while the role of the label is to connect the {\sc ep} to
+the surrounding formula.  Representing conjunction with label sharing
 stems from \mrs\ and provides compact semantic representations.
 Finally, (\ref{ex:cat-erg}) intuitively represents more specific
-semantic information than (\ref{ex:cat-pos}), so long as
-the constructors that \_cat\_n\_1 can correspond to is a
-subset of those that \_cat\_n can correspond to (written
-$\mbox{\_cat\_n\_1}\sqsubseteq \mbox{\_cat\_n}$), and similarly for
-\_fat\_j\_1 etc.  This is because with these so-called SPEC relations
-on the predicate symbols in place, (\ref{ex:cat-erg}) simply adds
-ARGs, qeqs and variable equalities to (\ref{ex:cat-pos}).
+semantic information than (\ref{ex:cat-pos}), so long as the
+constructors that \_cat\_n\_1 can correspond to is a subset of those
+that \_cat\_n can correspond to (written
+$\sempred{\_cat\_n\_1}\sqsubseteq \sempred{\_cat\_n}$), and similarly
+for \_fat\_j\_1 etc.  This is because with these so-called SPEC
+relations on the predicate symbols in place, (\ref{ex:cat-erg}) simply
+adds ARGs, qeqs and variable equalities to (\ref{ex:cat-pos}).
 

### 2008-10-08 trunk/Papers/rmrs-08/rmrs.tex
picture fixes and font change
quick changes to Extensions

@@
   \centering
-  \includegraphics[width=6cm]{pic-more-stuff}
+  \includegraphics[width=\columnwidth]{pic-more-stuff}
   \caption{Another tree which satisfies (\ref{ex:cat-pos}).}
@@
 
-In the earlier \rmrs\ literature, each term of an \rmrs\ is equipped
+So far we have included the dominance relation from
+\cite{egg:etal:2001} in the syntax and semantics of the formalism and
+not the qeq relation from \cite{copestake:etal:2005}.  This is partly
+because dominance is more expressive than qeq; for instance if a
+dependency parser links a determiner to a noun which in turn is
+linked to a verb, then the dominance relation is required to express
+the fact that
+that the argument to the predicate that's introduced by the verb is bound by
+the quantifier introduced by 
+the determiner (see earlier discussion).  However, it is very
+straightforward to add qeq to the syntax of the language and to extend
+the model theory accordingly.  This extension adds a new atom $X=_q Y$
+to Def.~\ref{defn:rmrs-syntax}, and $\tau,\alpha,g,\sigma$ will
+satisfy $X=_q Y$ iff $\alpha(X)$ and $\alpha(Y)$ are separated by
+zero or more nodes in $\tau$, each of which has a third child that is a
+quantifier.  It is possible to replace dominance by qeq in the above
+definition of \rmrs\ as well; and all the above propositions about
+solved forms still hold if ``dominance'' is systematically replaced
+with ``qeq''.
+
+Furthermore, grammar developers such as those in the {\sc delphin}
+community typically adopt conventions that restrict them to a
+fragment of the language from Def.~\ref{defn:rmrs-syntax} (once
+qeq is added to it), or they restrict attention to only a subset of
+the models (e.g., any variable that equals the variable
+that's bound by a quantifier is also bound by that quantifier). 
+Our formalism provides a general framework
+into which all these various fragments fit, and it's a matter of
+future work to investigate how these fragments afford efficient
+algorithms for performing inference over the \rmrs s.
+
+Another feature of the existing \rmrs\ literature is that each term of
+an \rmrs\ is equipped 
 with a \emph{sort}.  For instance, individual variables are typically
@@
 hierarchy ${\cal S}$.
-
 While we have defined \rmrs\ without sorts to simplify the
@@
 
-Another extension is to replace
-the qeq atom $X \qeqdom Y$ with dominance $X \dom Y$, 
-which is satisfied iff there is a path
-from $\alpha(X)$ to $\alpha(Y)$ in the tree.  Dominance is clearly a
-less restrictive condition than qeq, but it has been shown that the
-distinction between the two relations is irrelevant in \mrs s that were
-derived from a corpus by a deep grammar \cite{FucKolNieTha04}.  In our
-above definition of \rmrs, it is possible to replace qeq by dominance;
-this changes \rmrs\ to a different logic (let's call it
-Dominance-\rmrs), for which however all the above propositions about
-solved forms still hold if ``qeq'' is systematically replaced by
-``dominance''.
-
 %%% Local Variables: 

### 2008-10-08 trunk/Papers/rmrs-08/motivation.tex
changes to 2-3

@@
-\newcommand{\sem}[1]{\mathsf{#1}}
-\newcommand{\sempred}[1]{\mathrm{#1}}
-
 \section{Deep and shallow semantic construction}
@@
 
-We start with an informal discussion to 
-illustrate how \rmrs\ can be used to represent partial
-semantic information, and why people find it useful for this purpose.
-Consider the following (toy) sentence:
+We start with an informal illustration of why \rmrs\ is necessary and
+useful to represent partial semantic information.  Consider the
+following (toy) sentence:
+
 \begin{examples}
@@
 ambiguities of the nouns ``cat'' and ``dog'' (which have 8 and 7
-WordNet senses respectively).  Simplifying slightly by
-ignoring tense information (but see
-Section~\ref{sec:extensions}), two of these readings are
-shown as logical forms below; they can be represented as trees as
-shown in Fig.~\ref{fig:1}.
+WordNet senses respectively).  Simplifying slightly by ignoring tense
+information, two of these readings are shown as logical forms below;
+these can be represented as trees as shown in Fig.~\ref{fig:1}.
+%  (but see Section~\ref{sec:extensions})
 
@@
 
-Now imagine trying to extract semantic information from
-the output of a part-of-speech tagger by using the word lemmas as
-lexical predicate symbols.  Such a semantic representation is highly
-partial, as it will not say anything about predicate-argument
-relations or resolve lexical ambiguity---it will use predicate
-symbols such as $\sempred{\_cat\_n}$, which might resolve to the
-predicate symbols $\sem{\_cat\_n\_1}$ or $\sem{\_cat\_n\_2}$ in the
-complete semantic representation.  (Notice the different
-fonts for the ambiguous and unambiguous predicate symbols.)  
-We can
-express this information in \rmrs\ as follows.
+Now imagine we are trying to extract semantic information from the
+output of a part-of-speech tagger by using the word lemmas as lexical
+predicate symbols.  Such a semantic representation is highly partial,
+as it will not say anything about predicate-argument relations or
+resolve lexical ambiguity---it will use predicate symbols such as
+$\sempred{\_cat\_n}$, which might resolve to the predicate symbols
+$\sem{\_cat\_n\_1}$ or $\sem{\_cat\_n\_2}$ in the complete semantic
+representation.  (Notice the different fonts for the ambiguous and
+unambiguous predicate symbols.)
+
+Partial semantic information, especially with respect to ambiguity, is
+typically represented using an underspecification formalism such as
+\mrs\ \cite{copestake:etal:2005} or CLLS \cite{egg:etal:inpress}.  But
+most underspecification formalisms are unable to represent semantic
+information that is as partial as what we get from a POS tagger: They
+assume that the complete predicate-argument structure is known and
+there is no lexical ambiguity.  \rmrs\ \cite{copestake:2007a} was
+designed to address this problem.  In \rmrs, we can express the
+information we get from the POS tagger as follows:
 
@@
 
-This \rmrs\ consists of six \emph{elementary predications} ({\sc ep}s), one
-for each word lemma in the sentence; each of them is prefixed by a
+This \rmrs\ expresses only that certain predications must be made in
+the semantic representation -- but it doesn't say anything about the
+structure of the semantic representation, about most arguments of the
+predicates (note that $\sempred{\_chase\_v}(e)$ doesn't say who chases
+whom), or about the coindexation of variables ($\sempred{\_every\_q}$
+binds the variable $x_1$, whereas $\sempred{\_cat\_n}$ speaks about
+$x_3$), and maintains the lexical ambiguities.  Technically, it
+consists of six \emph{elementary predications} ({\sc ep}s), one for
+each word lemma in the sentence; each of them is prefixed by a
 \emph{label} and an \emph{anchor}, which are essentially variables
-that refer to nodes in the trees in Fig.~\ref{fig:1}.  
-The lack of semantic dependencies rendered by a {\sc pos} tagger
-means we have unique arguments to each
-predicate symbol (e.g., see $x_1$ and $x_3$ in (\ref{ex:cat-pos}),
-compared with the {\em same} variable $x$ in the semantic
-representations (\ref{ex:fat-cat-1}) and (\ref{ex:fat-cat-2})).
-We can say that
+that refer to nodes in the trees in Fig.~\ref{fig:1}.  We can say that
 the two trees \emph{satisfy} this \rmrs\ because it is possible to map
@@
 $x_1$ and $x_3$ into variable names in the tree in such a way that the
-predication
-of
-the node that labels and anchors are mapped to is consistent with
-that given in the {\sc ep}s of the {\sc rmrs}.
-For instance, the first tree satisfies
-the \rmrs\ because $l_1$ and $a_1$ can map to the root of that
-tree and $x_1$ to $x$, and the root label
-$\sem{\_every\_q\_1}$ is consistent with the 
-{\sc ep} predicate $\sempred{\_every\_q}$.
+predication of the node that labels and anchors are mapped to is
+consistent with that given in the {\sc ep}s of the {\sc rmrs}.  For
+instance, the first tree satisfies the \rmrs\ because $l_1$ and $a_1$
+can map to the root of that tree and $x_1$ to $x$, and the root label
+$\sem{\_every\_q\_1}$ is consistent with the {\sc ep} predicate
+$\sempred{\_every\_q}$.
+
+% The lack of semantic dependencies rendered by a {\sc pos} tagger means
+% we have unique arguments to each predicate symbol (e.g., see $x_1$ and
+% $x_3$ in (\ref{ex:cat-pos}), compared with the {\em same} variable $x$
+% in the semantic representations (\ref{ex:fat-cat-1}) and
+% (\ref{ex:fat-cat-2})).
 
@@
 semantic representations such as (\ref{ex:fat-cat-1})) that are
-described equally well by the \rmrs\ (\ref{ex:cat-pos}); this is
-not surprising, 
-given that the semantic information we got from the {\sc pos} tagger is so
-incomplete.  If we have more information, say information about
-subject and object relations from a chunk parser like Cass
-\cite{abney:1996}, we can represent these in a more
-detailed \rmrs, as follows:
+described equally well by the \rmrs\ (\ref{ex:cat-pos}); this is not
+surprising, given that the semantic information we got from the {\sc
+  pos} tagger is so incomplete.  If we have more information, say
+information about subject and object relations from a chunk parser
+like Cass \cite{abney:1996}, we can represent these in a more detailed
+\rmrs, as follows:
 
@@
 express that the two variables $x_3$ and $x_4$ must map to the same
-variable in any fully specific logical form; e.g., to the variable $x$
-in Fig.~\ref{fig:1}.  Atoms of the form
-$\ARG_i(a,z)$ and $\ARG_i(a,h)$ express that the $i$-th child
-(counting from 0) of the
-node to which the anchor $a$ refers is the variable name that $z$ denotes
-(or the node that the {\em hole} $h$ denotes).  The separation of
-{\sc ep}s and $\ARG$ atoms is perhaps the most important difference between
-\rmrs\ and earlier underspecification formalisms such as \mrs\ and
-dominance graphs, which required that a predicate and its arguments
-had to be specified together.  \rmrs\ goes beyond these formalisms in
-expressive power, because it allows one to state separately
-information about how many arguments a predicate takes (which is
-necessary for supporting parsers where information about lexical
-subcategorisation is absent), and the values of
-those (non-scopal and scopal) arguments.   If we also allow atoms of the form
-$\ARG_{\{2,3\}}(a,x)$ to express uncertainty as to whether $x$ is the
-second or third child of the anchor $a$, then \rmrs\ can specify the
-arguments to a predicate while underspecifying their position.
+variable in any fully specific logical form; e.g., both to the
+variable $x$ in Fig.~\ref{fig:1}.  Atoms of the form $\ARG_i(a,z)$ and
+$\ARG_i(a,h)$ express that the $i$-th child (counting from 0) of the
+node to which the anchor $a$ refers is the variable name that $z$
+denotes (or the node that the {\em hole} $h$ denotes).  So unlike
+earlier underspecification formalisms, \rmrs\ can specify the
+predicate of an atom separately from its (scopal and non-scopal)
+arguments; this is necessary for supporting parsers where information
+about lexical subcategorisation is absent). If we also allow atoms of
+the form $\ARG_{\{2,3\}}(a,x)$ to express uncertainty as to whether
+$x$ is the second or third child of the anchor $a$, then \rmrs\ can
+even specify the arguments to a predicate while underspecifying their
+position.  This is useful for \todo{what?}.
 
-Finally, \rmrs\ (\ref{ex:cat-erg}) is a notational variant of the {\sc
-  mrs} derived by the {\sc erg}, a wide-coverage deep grammar:
+Finally, the \rmrs\ (\ref{ex:cat-erg}) is a notational variant of the
+{\sc mrs} derived by the {\sc erg}, a wide-coverage deep grammar:
 \begin{examples}
@@
 \end{examples}
-$\mathsf{RSTR}$ and $\mathsf{BODY}$ are conventional names for the
-scopal arguments to predicate symbols corresponding to quantifiers.
-Atoms like $h_2 \qeq l_{42}$ specify a kind of ``outscopes''
-relationship between the hole and the label, and can be used to
-underspecify the scope of the two quantifiers.  Notice also that the
-labels of the {\sc ep}s for ``fat'' and ``cat'' are stipulated to be
-equal in the \rmrs\ (\ref{ex:cat-erg}), whereas the anchors are not.
-In the tree, it is the anchors that are mapped to the nodes with the
+
+Here we have full information about the predicate-argument structure
+and the variable binding; we only underspecify scope and some lexical
+ambiguity.  $\mathsf{RSTR}$ and $\mathsf{BODY}$ are conventional names
+for the $\ARG_1$ and $\ARG_2$ of a quantifier predicate symbol.  Atoms
+like $h_2 \qeq l_{42}$ (``qeq'') specify a certain kind of
+``outscopes'' relationship between the hole and the label, and are
+used here to underspecify the scope of the two quantifiers.
+
+Notice that the labels of the {\sc ep}s for ``fat'' and ``cat'' are
+stipulated to be equal in this \rmrs, whereas the anchors are not.  In
+the tree, it is the anchors that are mapped to the nodes with the
 labels $\sem{\_fat\_j\_1}$ and $\sem{\_cat\_n\_1}$; the label is
@@
 stems from \mrs\ and provides compact semantic representations.
-Finally, (\ref{ex:cat-erg}) intuitively represents more specific
-semantic information than (\ref{ex:cat-pos}), so long as the
-constructors that \_cat\_n\_1 can correspond to is a subset of those
-that \_cat\_n can correspond to (written
-$\sempred{\_cat\_n\_1}\sqsubseteq \sempred{\_cat\_n}$), and similarly
-for \_fat\_j\_1 etc.  This is because with these so-called SPEC
-relations on the predicate symbols in place, (\ref{ex:cat-erg}) simply
-adds ARGs, qeqs and variable equalities to (\ref{ex:cat-pos}).
 
-In summary, \rmrs\ is a formalism for representing partial information
-about semantic representations.  It can
-represent the output of a deep grammar in a way that is a notational
-variant of \mrs; but in cases in which a deep grammar is not
-available, it can be used to capture whatever partial semantic
-information a shallower {\sc nlp} tool can provide.  \rmrs\ is
-designed to be a generic formalism for partial semantic
-representations which abstracts over the details of different {\sc nlp}
-tools and allows us to represent, process, and compare information
-from different tools in the same formalism, and in the long run
-provide a platform for incrementally enriching partial information
-such as (\ref{ex:cat-pos}) in order to eventually approximate more
-detailed representations such as (\ref{ex:cat-partial-parser}) or
-(\ref{ex:cat-erg}).
+Finally, (\ref{ex:cat-erg}) uses predicate symbols like
+$\sempred{\_dog\_n\_1}$ that are meant to be more specific than the
+symbols like $\sempred{\_dog\_n}$ which the earlier \rmrs s used.
+This reflects the fact that the deep grammar will perform some lexical
+disambiguation, whereas the chunker and POS tagger didn't.  The fact
+that the former symbol should be more specific than the latter can be
+represented using SPEC atoms like $\sempred{\_dog\_n\_1}\sqsubseteq
+\sempred{\_dog\_n}$), which we have omitted in (\ref{ex:cat-erg}) for
+brevity.  Note that even a deep grammar will not fully disambiguate to
+semantic predicate symbols, such as WordNet senses, and so
+$\sempred{\_dog\_n\_1}$ can still be consistent with multiple symbols
+like $\sem{\_dog\_n\_1}$ and $\sem{\_dog\_n\_2}$ in the semantic
+representation.
 
-However, although people have started using \rmrs\ in their
-implemented systems, it is still relatively ad-hoc in that its formal
-syntax is not consistent across different papers, and its formal
-semantics is not defined; we don't know, formally, what an \rmrs\
-\emph{means} in terms of semantic representations like
-(\ref{ex:fat-cat-1}) and (\ref{ex:fat-cat-2}), and this hinders our
-ability to design efficient algorithms for processing \rmrs. The
-purpose of this paper is to lay the groundwork for fixing this problem.
+In summary, \rmrs\ is a formalism for partial semantic
+representations.  It allows us to represent the semantic information
+that can be extracted from a wide range of NLP tools in a uniform way.
+This is useful for hybrid systems which fall back to shallower
+analyses if deeper parsing is unsuccessful, or which try to match
+deeply parsed queries against shallow parses of large text corpora;
+and in fact, \rmrs\ is gaining popularity as a practical interchange
+format for exactly these purposes \cite{copestake:2003,frank:2004}.
+However, \rmrs\ is still relatively ad-hoc in that its formal syntax
+is not consistent across different papers, and its formal semantics is
+not defined; we don't know, formally, what an \rmrs\ \emph{means} in
+terms of semantic representations like (\ref{ex:fat-cat-1}) and
+(\ref{ex:fat-cat-2}), and this hinders our ability to design efficient
+algorithms for processing \rmrs. The purpose of this paper is to lay
+the groundwork for fixing this problem.
 

### 2008-10-08 trunk/Papers/rmrs-08/rmrs.tex
changes to 2-3

@@
 We will now make the basic ideas from Section \ref{sec:motivation}
-more precise.  We will first define the syntax of the \rmrs\ language;
-this is a notational variant of earlier definitions in the
-literature.  We will then define a model theory for our version of
-\rmrs, and conclude this section by carrying over the notion of
-\emph{solved forms} from {\sc clls} \cite{egg:etal:2001}.
+precise.  We will first define the syntax of the \rmrs\ language; this
+is a notational variant of earlier definitions in the literature.  We
+will then define a model theory for our version of \rmrs, and conclude
+this section by carrying over the notion of \emph{solved forms} from
+{\sc clls} \cite{egg:etal:2001}.
 
@@
 \cite{egg:etal:2001}.  We assume an infinite set of {\em node
-  variables} $\nvar=\{X,Y,X_1,\ldots\}$, used as labels,
-anchors, and holes; the distinction between these uses will come from
-their position in the formula.  Furthermore, we assume an infinite set
-of \emph{base variables} $\ovar$, consisting of individual variables
+  variables} $\nvar=\{X,Y,X_1,\ldots\}$, used as labels, anchors, and
+holes; the distinction between these uses will come from their
+position in the formula.  Furthermore, we assume an infinite set of
+\emph{base variables} $\ovar$, consisting of individual variables
 $\{x,x_1,y,\ldots\}$ and event variables $\{e_1,\ldots\}$.  Finally,
@@
   following forms; $S \subseteq \N$ is a set of numbers that is either
-  finite or $\N$ itself.
+  finite or $\N$ itself.\footnote{Throughout the paper, we assume that
+    $0 \in \N$.}
 
@@
 &| &\ARG_S(X,Y) \\ % & \mbox{ARG relation for node variables} \\
-&| &X \qeq Y \\ % & \mbox{QEQ} \\
+&| &X \dom Y \\ % & \mbox{dominance} \\
 &| &v_1 = v_2 \;|\; v_1 \neq v_2 \\ % & \mbox{base variable
@@
 A node variable $X$ is called a \emph{label} iff $\varphi$ contains an
-atom of the form 
-$\ep{X}{Y}{P}$ or $Y \qeq X$; $X$ is called an
-\emph{anchor} in the \rmrs\
-$\varphi$ iff $\varphi$ contains an atom of the form
-$\ep{Y}{X}{P}$ or $\ARG_S(X,i)$; and $X$ is
-called a \emph{hole} iff 
-$\varphi$ contains an atom of the form $\ARG_S(Y,X)$ or $X \qeq Y$.
-
-An \rmrs\ $\varphi$ is called \emph{normal} iff it satisfies the
-following conditions:
-\begin{enumerate}
-\item Every node variable in $\varphi$ is either a label, an anchor,
-  or a hole, but not two of these.
-\item If $\ep{X}{Y}{P}, \ep{Z}{Y}{Q}\in \varphi$, then $X=Z$ and $P=Q$
-  (i.e., each predication has a unique anchor).
-\end{enumerate}
+atom of the form $\ep{X}{Y}{P}$ or $Y \qeq X$; $X$ is called an
+\emph{anchor} in the \rmrs\ $\varphi$ iff $\varphi$ contains an atom
+of the form $\ep{Y}{X}{P}$ or $\ARG_S(X,i)$; and $X$ is called a
+\emph{hole} iff $\varphi$ contains an atom of the form $\ARG_S(Y,X)$
+or $X \qeq Y$.
 \end{definition}
@@
 \rmrs\ as presented here combines similarities to earlier
-presentations of \rmrs\ (e.g., \cite{copestake:2003}) and to dominance
-constraints \cite{egg:etal:2001}.  We use $\ARG_{\{i\}}$ (with a singleton
-set $S$) instead of $\ARG_i$, and $\ARG_\N$ instead of $\ARG_n$.
-The elementary predication $\ep{l}{a}{P(v)}$ (see
-(\ref{ex:cat-pos})--(\ref{ex:cat-erg})) is a
-notational variant of $\{\ep{l}{a}{P}, \ARG_{\{0\}}(a,v)\}$.  In other
-words, the syntax defined here differs in some details, but is more
-general. The primary differences to dominance constraints are the
-distinction of labels and anchors that will separate the labelling and
-the parent-child relations, and the use of $\qeq$ instead of
-dominance.  Redefining \rmrs\ in the style of dominance constraints
-simplifies the formal work below.
+presentations of \rmrs\ \cite{copestake:2003,copestake:2007b} and to
+CLLS/dominance constraints \cite{egg:etal:2001}.  For the most part,
+our syntax generalizes that of older versions of \rmrs: We use
+$\ARG_{\{i\}}$ (with a singleton set $S$) instead of $\ARG_i$ and
+$\ARG_\N$ instead of $\ARG_n$, and we take the EP $\ep{l}{a}{P(v)}$
+(as in Section~\ref{sec:motivation}) as an abbreviation of
+$\{\ep{l}{a}{P}, \ARG_{\{0\}}(a,v)\}$.  Similarly, we don't assume
+that labels, anchors, and holes are syntactically different objects;
+they receive their function from their positions in the formula.  One
+major difference is that we use dominance rather than qeq; see
+Section~\ref{sec:extensions} for a discussion.  Compared to dominance
+constraints, the primary difference is that we now have a mechanism
+for representing lexical ambiguity, and we can specify a predicate and
+its arguments separately.
 
-\subsection{Model Theory}
+\subsection{Model Theory} \label{sec:model-theory}
 
@@
 expressed in the base language.  We represent such a logical form as a
-tree $\tau$, such as the ones in Fig.~\ref{fig:1}.  So models of the
-\rmrs\ are such trees, and we can define satisfaction and entailment
-over these models in the usual way.  Satisfaction corresponds to an
-\rmrs\ being a partial description of the (unique) fully specific
-semantic representation that the model corresponds to, whereas
-entailment corresponds to the relation between a more and a less
-specific \rmrs.
+tree $\tau$, such as the ones in Fig.~\ref{fig:1}, and we can then
+define satisfaction of formulas in the usual way, by taking the tree
+as a model structure that interprets all predicate symbols specified
+above.
 
-In this paper, we assume for simplicity that the base language is
-predicate logic (e.g., see (\ref{ex:fat-cat-1}) and
-(\ref{ex:fat-cat-2})); in this case, $\tau$ is simply the structure
-tree of 
-the predicate logic formula.  We assume that $\Sigma$ is a ranked
-signature consisting of predicate logic's symbols: a unary constructor
-$\neg$ and binary constructors $\wedge$, $\rightarrow$, etc.; the
-binary constructors $\forall x$ and $\exists x$ for all base-language
-variables $x$; all the constants and base-language variables are
-nullary constructors; and, for each predicate symbol $P$ of arity $n$,
-a constructor $P$ of arity $n$.  Other base languages may require a
-different signature $\Sigma$ and/or a different mapping between
-formulas and trees.  We only assume for convenience that the signature
-contains a binary constructor $\wedge$ to represent conjunction.  We
-write $\Sigma_i$ for the set of all constructors in $\Sigma$ that have
-arity $i$.  We will follow the typographical convention that
-non-logical symbols in $\Sigma$ are written in
-$\mathsf{sans{-}serif}$, as opposed to the \rmrs\ predicate symbols
-like ``$\sempred{\_cat\_n}$'' and ``$\sempred{\_cat\_n\_1}$''.
+In this paper, we assume for simplicity that the base language is as
+in \mrs; essentially, $\tau$ becomes the structure tree of a formula
+of predicate logic.  We assume that $\Sigma$ is a ranked signature
+consisting of the symbols of predicate logic: a unary constructor
+$\neg$ and binary constructors $\wedge$, $\rightarrow$, etc.; a set of
+3-place quantifier symbols such as $\sem{\_forall\_q\_1}$ and
+$\sem{\_some\_q\_1}$ (with the children being the bound variable, the
+restrictor, and the scope); and, for each predicate symbol $P$ of
+arity $n$, a constructor $P$ of arity $n$ \todo{$n$ or $n+1$?}.  Other
+base languages may require a different signature $\Sigma$ and/or a
+different mapping between formulas and trees; the only strict
+requirement we make is that the signature contains a binary
+constructor $\wedge$ to represent conjunction.  We write $\Sigma_i$
+for the set of all constructors in $\Sigma$ that have arity $i$.  We
+will follow the typographical convention that non-logical symbols in
+$\Sigma$ are written in $\mathsf{sans{-}serif}$, as opposed to the
+\rmrs\ predicate symbols like ``$\sempred{\_cat\_n}$'' and
+``$\sempred{\_cat\_n\_1}$''.
 
@@
 \begin{definition}\label{defn:dominance}
-$u \qeqdom v$ iff $v = u \cdot 2^k$ for some $k \geq 0$ and all
-  symbols $\tau(u), \tau(u \cdot 2), \ldots, \tau(u \cdot 2^{k-1})$
-  are quantifiers.  (This is the ``qeq'' relation familiar from
-  \newcite{copestake:etal:2005}).
+  $u \dom v$ iff $u$ is a prefix of $v$, i.e.\ the node $u$ is above
+  the node $v$ in the tree.  (This is the dominance relation familiar
+  from \newcite{egg:etal:2001}.)
 
-  $u \wedgedom v$ iff $u \leq v$ (i.e.\ it is a prefix), and all
-  symbols on the path from $u$ to $v$ (not including $v$) are
-  $\wedge$.
+  $u \wedgedom v$ iff $u \dom v$, and all symbols on the path from $u$
+  to $v$ (including $u$ but not $v$) are $\wedge$.
 \end{definition}
@@
 maps the base variables in the \rmrs\ to nullary constructors
-representing variables in the base language.  This function enables us
-to map two variables that seem different in the \rmrs\ to the same
-base-language variable, introducing e.g.\ semantic dependencies that a
-POS tagger didn't catch.  For instance, $g$ might map $x_4$ and $x_5$
-in (\ref{ex:cat-pos}) to $y$ to specify a variable binding.  Finally,
-we assume a relation $\sigma \subseteq \pred \times \Sigma_{\geq 1}$
-that maps each \rmrs\ predicate symbol to a set of constructors from
-$\Sigma$.  This relation allows us to underspecify lexical
-ambiguities; for instance, $\sigma(\sempred{\_cat\_n})$ might contain
-the symbol $\sem{\_cat\_n\_1}$ along with symbols for other word
-senses.
+representing variables in the base language. Finally, we assume a
+relation $\sigma \subseteq \pred \times \Sigma_{\geq 1}$ that maps
+each \rmrs\ predicate symbol to a set of constructors from $\Sigma$.
+This relation allows us to underspecify lexical ambiguities; for
+instance, $\sigma(\sempred{\_cat\_n})$ might contain the symbol
+$\sem{\_cat\_n\_1}$ along with symbols for other word senses.
 
@@
   i = \alpha(Y)\\
-  \tau, \alpha, g, \sigma \models X \qeq Y
-\mbox{ iff } \alpha(X) \qeqdom \alpha(Y) \\
-  \tau, \alpha, g, \sigma \models X ={/}\neq Y
-\mbox{ iff } \\
-\qquad \alpha(X) ={/}\neq \alpha(Y) \\
-  \tau, \alpha, g, \sigma \models v_1 ={/}\neq v_2
-\mbox{ iff}\\
-\qquad g(v_1) ={/}\neq g(v_2) \\
+  \tau, \alpha, g, \sigma \models X \dom Y
+\mbox{ iff } \alpha(X) \dom \alpha(Y) \\
+  \tau, \alpha, g, \sigma \models X =  \!\!/\!\! \neq Y
+\mbox{ iff }  \alpha(X) =  \!\!/\!\! \neq \alpha(Y) \\
+  \tau, \alpha, g, \sigma \models v_1 =  \!\!/\!\! \neq v_2
+\mbox{ iff }  g(v_1) = \!\!/\!\! \neq g(v_2) \\
   \tau, \alpha, g, \sigma \models P \spec Q
@@
 the \rmrs\ to be a partial description of each of these trees.  In
-particular, \rmrs s may represent semantic scope ambiguities, and/or missing
-information about semantic dependencies, 
-lexical subcategorisation and lexical senses.  For $j=\{1,2\}$,
-suppose that $\tau_j,\alpha_j,g_j,\sigma\models \varphi$ and let
-$\nvar$ and $\ovar$ be the node and base variables from which
-$\varphi$ is built.  Then 
-$\varphi$ exhibits a semantic scope ambiguity if there are variables
-$Y,Y'\in \nvar$ such that $\alpha_1(Y) \qeqdom \alpha_1(Y')$ and
-$\alpha_2(Y')\qeqdom \alpha_2(Y)$.  It exhibits missing information
-about semantic dependencies if there are base-language variables
-$v,v'\in \ovar$ such that $g_1(v)=g_1(v')$ and $g_2(v)\neq g_2(v')$. 
-It exhibits missing lexical subcategorisation information if there is
-a $ Y \in \nvar$ such that $\tau_1(\alpha_1(Y))$ is a
-constructor of a different type from $\tau_2(\alpha_2(Y))$  (i.e., the
-constructors are of a different arity and/or their arguments are).
-And it exhibits missing lexical sense information
-if $\tau_1(\alpha_1(Y))$ and $\tau_2(\alpha_2(Y))$ are different
-base-language constructors, but of the same type.
+particular, \rmrs s may represent semantic scope ambiguities and/or
+missing information about semantic dependencies, lexical
+subcategorisation and lexical senses.  For $j=\{1,2\}$, suppose that
+$\tau_j,\alpha_j,g_j,\sigma\models \varphi$.  Then $\varphi$ exhibits
+a semantic scope ambiguity if there are variables $Y,Y'\in \nvar$ such
+that $\alpha_1(Y) \dom \alpha_1(Y')$ and $\alpha_2(Y')\dom
+\alpha_2(Y)$.  It exhibits missing information about semantic
+dependencies if there are base-language variables $v,v'\in \ovar$ such
+that $g_1(v)=g_1(v')$ and $g_2(v)\neq g_2(v')$.  It exhibits missing
+lexical subcategorisation information if there is a $ Y \in \nvar$
+such that $\tau_1(\alpha_1(Y))$ is a constructor of a different type
+from $\tau_2(\alpha_2(Y))$ (i.e., the constructors are of a different
+arity and/or their arguments are).  And it exhibits missing lexical
+sense information if $\tau_1(\alpha_1(Y))$ and $\tau_2(\alpha_2(Y))$
+are different base-language constructors, but of the same type.
 
@@
 The \rmrs\ (\ref{ex:cat-pos}) satisfies this tree with an assignment
-function $\alpha$ that maps the variables $l_1$ and $a_1$ to the
-root node, $l_{41}$ and $l_{42}$ to its second child (labeled with
+function $\alpha$ that maps the variables $l_1$ and $a_1$ to the root
+node, $l_{41}$ and $l_{42}$ to its second child (labeled with
 ``$\wedge$''), $a_{41}$ to the first child of \emph{that} node (i.e.\
@@
 so forth.  $g$ will map $x_1$ and $x_3$ to $x$, and $x_6$ and $x_7$ to
-$y$, and so on.  
-And $\sigma$ will map each \rmrs\ predicate symbol (which represents a
-word) to the set of its word senses, e.g.\ \_cat\_n to a set
-containing $\sem{\_cat\_n\_1}$ and possibly others.  It is then easy to verify
-that every single atom in the \rmrs\ is satisfied---most
-interestingly, the {\sc ep}s $\ep{l_{41}}{a_{41}}{\_fat\_a(e')}$ and
-$\ep{l_{42}}{a_{42}}{\_cat\_n(x_3)}$ are satisfied because
+$y$, and so on.  And $\sigma$ will map each \rmrs\ predicate symbol
+(which represents a word) to the set of its word senses, e.g.\
+\_cat\_n to a set containing $\sem{\_cat\_n\_1}$ and possibly others.
+It is then easy to verify that every single atom in the \rmrs\ is
+satisfied---most interestingly, the {\sc ep}s
+$\ep{l_{41}}{a_{41}}{\sempred{\_fat\_j}(e')}$ and
+$\ep{l_{42}}{a_{42}}{\sempred{\_cat\_n}(x_3)}$ are satisfied because
 $\alpha(l_{41}) \wedgedom \alpha(a_{41})$ and $\alpha(l_{42})
@@
 
-One feature which the above definition of \rmrs\ shares with dominance
-constraints is that any satisfiable \rmrs\ has an infinite number of
-models which only differ in the areas that the \rmrs\ didn't ``talk
-about''.  An example is shown in Fig.~\ref{fig:fat-black-cat}.  This
-tree, which represents the meaning of ``every fat black cat chased
-some dog'', is still a model
-of (\ref{ex:cat-pos})---all the {\sc ep}s are satisfied.  However, there is
-more ``semantic material'' (the ``black'' subformula).  The \rmrs\
-didn't specify that this material should be there; but it also didn't
-specify that it can't be there, and indeed there is no way in which
-the \rmrs\ can enforce that no extra material can be added (e.g.,
-consider predicate symbols that are introduced by construction rules
-in a deep grammar, which by design are absent from the semantic output
-of a {\sc pos} tagger).
+One aspect in which our definition of \rmrs\ is like dominance
+constraints and unlike \mrs\ is that any satisfiable \rmrs\ has an
+infinite number of models which only differ in the areas that the
+\rmrs\ didn't ``talk about''.  Consider e.g.\ (\ref{ex:cat-erg}).
+Read as an \mrs\ or as an \rmrs\ of the previous literature, this
+formula is an instruction to build a semantic representation out of
+the pieces for ``every fat cat'', ``some dog'', and ``chased''; a
+semantic representation as in Fig.~\ref{fig:fat-black-cat} would
+\emph{not} be taken as described by this \rmrs\.  However, under the
+semantics we proposed above, this tree is a correct model of
+(\ref{ex:cat-erg}) because all atoms are still satisfied; the \rmrs\
+didn't say anything about ``sleep'' or ``run'', but it couldn't
+enforce that the tree shouldn't contain those subformulas either.
 
@@
   \includegraphics[width=\columnwidth]{pic-more-stuff}
-  \caption{Another tree which satisfies (\ref{ex:cat-pos}).}
+  \caption{Another tree which satisfies (\ref{ex:cat-erg}).}
   \label{fig:fat-black-cat}
@@
 models further and further towards the semantic representation we are
-trying to reconstruct.  However, the fact that all satisfiable \rmrs s
-have an infinite number of models is a problem for algorithms that try
-to decide satisfiability or compute the fully specific semantic
-representations---how can this be done in finite time?
+trying to reconstruct.  Furthermore, it has been shown in the
+literature that a dominance-constraint style semantics for
+underspecified representations gives us more room to maneuver when
+developing efficient solvers than an \mrs-style semantics
+\cite{Althaus_etal:JoA}.
 
-In this paper, we follow the dominance constraint strategy for
-resolving this dilemma; we
-define a fragment of \rmrs s called \emph{solved forms}.  All \rmrs
-s in solved form are guaranteed to be satisfiable, and thus each
-solved form represents an infinite class of models.  However, we will
-set it up in such a way that each satisfiable \rmrs\ has only a finite
-number of solved forms which partition the space of possible models
-into classes with only ``irrelevant'' differences as above.  A solver
-can then enumerate the solved forms rather than all models.
+However, enumerating an infinite number of models is of course
+infeasible.  For this reason, we will now transfer the concept of
+\emph{solved forms} from dominance constraints to \rmrs.  An \rmrs\ in
+solved form is guaranteed to be satisfiable, and thus each solved form
+represents an infinite class of models.  However, each satisfiable
+\rmrs\ has only a finite number of solved forms which partition the
+space of possible models into classes with only ``irrelevant''
+differences as above.  A solver can then enumerate the solved forms
+rather than all models.
 
 Intuitively, we expect that an \rmrs\ in solved form is fully
-specified with respect to the predicate-argument
-structure, all variable equalities and inequalities have been
-resolved, and only lexical sense ambiguities and (resolvable)
-scope ambiguities remain.  This is made
+specified with respect to the predicate-argument structure, all
+variable equalities and inequalities and scope ambiguities have been
+resolved, and only lexical sense ambiguities remain.  This is made
 precise below.
@@
   \item no label occurs on the right-hand side of two
-    different $\qeqdom$ atoms.
+    different $\dom$ atoms.
   \end{enumerate}
@@
 
-All these conditions together ensure that it is possible to ``read
-off'' a model from the solved form.
+Because solved forms are so restricted, it is now possible ``read
+off'' at least one model from each solved form.  We don't have the
+space to present any proofs here; they proceed mostly along the same
+lines as the proofs in \newcite{Duchier00dominanceconstraints}.
 
@@
 \end{prop}
-\begin{proof}[Proof (sketch; see also
-  \cite{Duchier00dominanceconstraints})] 
-  For each {\sc ep}, we choose to label the anchor with the constructor $p$
-  of sufficiently high arity whose existence we assumed; we determine
-  the edges between an anchor and its children from the uniquely
-  determined $\ARG$ atoms; plugging labels into holes is
-  straightforward because no label is qeq-dominated by more than one
-  label; and then we can fill the spaces between the labels and
-  anchors with conjunctions.
-\end{proof}
+
+We can now define the solved forms \emph{of} an \rmrs\ $\varphi$;
+these finitely many \rmrs s in solved form partition the space of
+models of $\varphi$ into classes of models with trivial differences.
 
@@
 \begin{array}{r c l}
-\{(X,Y) & \mid &  \mbox{$\varphi$ contains $X \qeqdom Y$ or}\\
+\{(X,Y) & \mid &  \mbox{$\varphi$ contains $X \dom Y$ or}\\
 && \mbox{$ARG_S(X,Y)$ for some
@@
 \end{prop}
-\begin{proof}
-  We construct the substitution $s$ from $\alpha$ and $g$.  Then we
-  add all qeqs that are satisfied by $\alpha$ and restrict the $\ARG$
-  atoms to those child indices that are actually used in $\tau$.  The
-  result is in solved form because $\tau$ is a tree; it is a solved
-  form of $\varphi$ by construction.
-\end{proof}
-
-\todo{Draw a figure for (\ref{ex:cat-erg}), and explain that the trees
-  in Fig 1 are its solved forms.}
 
@@
 \end{prop}
-\begin{proof}
-  Up to renaming of variables, there is only a finite number of
-  substitutions on the node and base variables of $\varphi$.  Let $s$
-  be such a substitution.  This fixes the set of {\sc ep}s of any solved
-  form of $\varphi$ that is based on $s$ uniquely.  There is only a
-  finite set of choices for the subsets $S'$ in condition 2 of
-  Def.~\ref{defn:solved-form-of}, and there is only a finite set of
-  choices of new $\qeqdom$ atoms that satisfy condition 3.  Therefore,
-  the set of solved forms of $\varphi$ is finite.
-\end{proof}
+
+Let's look at an example for all these definitions.  All the \rmrs s
+presented in Section~\ref{sec:motivation}, where we replace $\qeq$ by
+$\dom$, are in solved form; this is least obvious for
+(\ref{ex:cat-erg}), but becomes clear once we notice that no label is
+on the right-hand side of two dominance atoms.  However, this is
+because the model that we can read off of (\ref{ex:cat-erg}) is
+similar to the one shown in Fig.~\ref{fig:fat-black-cat}; we simply
+take the three subtrees that are specified by ``every fat cat'',
+``some dog'', and ``chased'' and combine them as sisters.
 

### 2008-10-08 trunk/Papers/rmrs-08/introduction.tex
changes to 2-3
quick updates

@@
 \cite{butt:etal:1999,copestake:flickinger:2000}, but this
-tends to lack robustness: hand-crafted grammars fail to cover all
-words and linguistic constructions and by design they do not handle
+tends to lack robustness: they fail to cover all
+words and linguistic constructions and by design do not handle
 ill-formed phrases. 
 There has thus been a trend recently towards systems for
-robust wide-coverage semantic construction.  For instance,
-\newcite{bos:etal:2004} exploit {\sc ccg}'s syntax-semantics
-interface to produce a unique logical form for each derivation of a
-statistical {\sc ccg} parser; \newcite{zettlemoyer:collins:2007} and
-others learn string-to-logical-form mappings, exploiting informative
-features from shallow language processors; and
-\newcite{copestake:2003} and \newcite{frank:2004} endow shallow
-language processors with semantic components.
-
+robust wide-coverage semantic construction (e.g., 
+\cite{bos:etal:2004,zettlemoyer:collins:2007}).     
 % COMMENT: this issue about optional arguments is really too cryptic.
@@
 % 
-
-There are certain semantic phenomena that these robust
+But there are certain semantic phenomena that these robust
 approaches don't capture reliably, including quantifier scope,
@@
 semantic information that supports this strategy: Robust Minimal
-Recursion Semantics (\rmrs, \cite{copestake:2003}).  \rmrs\ is
+Recursion Semantics (\rmrs, \cite{copestake:2007a,copestake:2007b}).
+\rmrs\ is 
 designed to

### 2008-10-08 trunk/Papers/rmrs-08/motivation.tex
quick updates

@@
 
-Partial semantic information, especially with respect to ambiguity, is
+Partial semantic information is
 typically represented using an underspecification formalism such as
-\mrs\ \cite{copestake:etal:2005} or CLLS \cite{egg:etal:inpress}.  But
+\mrs\ \cite{copestake:etal:2005} or {\sc clls} \cite{egg:etal:inpress}.  But
 most underspecification formalisms are unable to represent semantic
-information that is as partial as what we get from a POS tagger: They
-assume that the complete predicate-argument structure is known and
-there is no lexical ambiguity.  \rmrs\ \cite{copestake:2007a} was
+information that is as partial as what we get from a {\sc pos} tagger: They
+assume that the complete predicate-argument structure is known.
+\rmrs\ \cite{copestake:2007a} was
 designed to address this problem.  In \rmrs, we can express the
-information we get from the POS tagger as follows:
+information we get from the {\sc pos} tagger as follows:
 
@@
 arguments; this is necessary for supporting parsers where information
-about lexical subcategorisation is absent). If we also allow atoms of
+about lexical subcategorisation is absent. If we also allow atoms of
 the form $\ARG_{\{2,3\}}(a,x)$ to express uncertainty as to whether
@@
 even specify the arguments to a predicate while underspecifying their
-position.  This is useful for \todo{what?}.
+position.  This is useful for specifying arguments to \_give\_v when a
+parser that is unableto process unbounded dependencies is faced with
+{\em Which bone did you give the dog?} vs.\ {\em To which dog did you
+  give the bone?}
 
@@
 This reflects the fact that the deep grammar will perform some lexical
-disambiguation, whereas the chunker and POS tagger didn't.  The fact
+disambiguation, whereas the chunker and {\sc pos} tagger didn't.  The fact
 that the former symbol should be more specific than the latter can be
@@
 representations.  It allows us to represent the semantic information
-that can be extracted from a wide range of NLP tools in a uniform way.
+that can be extracted from a wide range of {\sc nlp} tools in a uniform way.
 This is useful for hybrid systems which fall back to shallower

### 2008-10-08 trunk/Papers/rmrs-08/rmrs.tex
done with sect 3 for now

@@
 feature of \rmrs, because it means that when we enrich an \rmrs\
-obtained from a shallow processor with more semantic information, we
-don't change the set of models; we only \emph{restrict} the set of
-models further and further towards the semantic representation we are
-trying to reconstruct.  Furthermore, it has been shown in the
-literature that a dominance-constraint style semantics for
-underspecified representations gives us more room to maneuver when
-developing efficient solvers than an \mrs-style semantics
+obtained from a shallow processor with more semantic information --
+such as the relational information introduced by syntactic
+constructions such as appositives, noun-noun compounds, and
+nonrestrictive relative clauses --, we don't change the set of models;
+we only \emph{restrict} the set of models further and further towards
+the semantic representation we are trying to reconstruct.
+Furthermore, it has been shown in the literature that a
+dominance-constraint style semantics for underspecified
+representations gives us more room to maneuver when developing
+efficient solvers than an \mrs-style semantics
 \cite{Althaus_etal:JoA}.
@@
 Let's look at an example for all these definitions.  All the \rmrs s
-presented in Section~\ref{sec:motivation}, where we replace $\qeq$ by
-$\dom$, are in solved form; this is least obvious for
+presented in Section~\ref{sec:motivation} (replacing $\qeq$ by
+$\dom$), are in solved form; this is least obvious for
 (\ref{ex:cat-erg}), but becomes clear once we notice that no label is
@@
 because the model that we can read off of (\ref{ex:cat-erg}) is
-similar to the one shown in Fig.~\ref{fig:fat-black-cat}; we simply
-take the three subtrees that are specified by ``every fat cat'',
-``some dog'', and ``chased'' and combine them as sisters.
+similar to the one shown in Fig.~\ref{fig:fat-black-cat}; this model
+is problematic in several ways and in particular contains unbound
+variables.  If we restrict the class of models we consider to those in
+which variables are bound, we can add the information that the
+quantifiers outscope their bound variables to the \rmrs\ without
+changing its models further -- i.e., we add the atoms $h_3 \dom
+l_5$ and $h_8 \dom l_5$.  The extended \rmrs\ is no longer in solved
+form because the label $l_5$ is on the right-hand side of two
+dominance edges.  Instead, it has the following two solved forms:
+
+\begin{examples}
+\item \label{ex:sf-1}
+  $l_1\handel a_1\handel\sempred{\_every\_q\_1}(x_1)$,
+  $\mathsf{RSTR}(a_1,h_2)$, $\mathsf{BODY}(a_1,h_3)$,
+  $l_{41}\handel a_{41}\handel\sempred{\_fat\_j\_1}(e'), \ARG_1(a_{41},x_1)$,
+  $l_{41}\handel a_{42}\handel\sempred{\_cat\_n\_1}(x_1)$,
+  %
+  $l_6\handel a_6\handel\sempred{\_some\_q\_1}(x_6)$,
+  $\mathsf{RSTR}(a_6,h_7)$, $\mathsf{BODY}(a_6,h_8)$,
+  $l_9\handel a_9\handel\sempred{\_dog\_n\_1}(x_6)$,
+  %
+  $l_5\handel a_5\handel\sempred{\_chase\_v\_1}(e)$,
+  $\ARG_1(a_5,x_1)$, $\ARG_2(a_5,x_6)$,
+  $h_2 \dom l_{41}$, $h_3 \dom l_6$,
+  $h_7 \dom l_9$, $h_8 \dom l_5$
+\item \label{ex:sf-2}
+  $l_1\handel a_1\handel\sempred{\_every\_q\_1}(x_1)$,
+  $\mathsf{RSTR}(a_1,h_2)$, $\mathsf{BODY}(a_1,h_3)$,
+  $l_{41}\handel a_{41}\handel\sempred{\_fat\_j\_1}(e'), \ARG_1(a_{41},x_1)$,
+  $l_{41}\handel a_{42}\handel\sempred{\_cat\_n\_1}(x_1)$,
+  %
+  $l_6\handel a_6\handel\sempred{\_some\_q\_1}(x_6)$,
+  $\mathsf{RSTR}(a_6,h_7)$, $\mathsf{BODY}(a_6,h_8)$,
+  $l_9\handel a_9\handel\sempred{\_dog\_n\_1}(x_6)$,
+  %
+  $l_5\handel a_5\handel\sempred{\_chase\_v\_1}(e)$,
+  $\ARG_1(a_5,x_1)$, $\ARG_2(a_5,x_6)$,
+  $h_2 \dom l_{41}$, $h_3 \dom l_5$,
+  $h_7 \dom l_9$, $h_8 \dom l_1$
+\end{examples}
+
+Notice that we have eliminated all equalities by unifying the variable
+names, and we have fixed the relative scope of the two quantifiers.
+Each of these solved forms now stands for a separate class of models;
+for instance, the first model in Fig.~\ref{fig:1} is a model of
+(\ref{ex:sf-1}), whereas the second is a model of (\ref{ex:sf-2}).
 
@@
 
-So far we have included the dominance relation from
-\cite{egg:etal:2001} in the syntax and semantics of the formalism and
-not the qeq relation from \cite{copestake:etal:2005}.  This is partly
-because dominance is more expressive than qeq; for instance if a
-dependency parser links a determiner to a noun which in turn is
-linked to a verb, then the dominance relation is required to express
-the fact that
-that the argument to the predicate that's introduced by the verb is bound by
-the quantifier introduced by 
-the determiner (see earlier discussion).  However, it is very
-straightforward to add qeq to the syntax of the language and to extend
-the model theory accordingly.  This extension adds a new atom $X=_q Y$
-to Def.~\ref{defn:rmrs-syntax}, and $\tau,\alpha,g,\sigma$ will
-satisfy $X=_q Y$ iff $\alpha(X)$ and $\alpha(Y)$ are separated by
-zero or more nodes in $\tau$, each of which has a third child that is a
-quantifier.  It is possible to replace dominance by qeq in the above
-definition of \rmrs\ as well; and all the above propositions about
-solved forms still hold if ``dominance'' is systematically replaced
-with ``qeq''.
+So far we have based the syntax and semantics of \rmrs\ on the
+dominance relation from \newcite{egg:etal:2001} rather than the qeq
+relation from \newcite{copestake:etal:2005}.  This is partly because
+dominance is a weaker relation than qeq: If a dependency parser links
+a determiner to a noun and this noun to a verb, then we can use
+dominance but not qeq to represent that the predicate introduced by
+the verb is outscoped by the quantifier introduced by the determiner
+(see earlier discussion).  However, it is very straightforward to add
+qeq to the syntax of the language and to extend the model theory
+accordingly.  This extension adds a new atom $X=_q Y$ to
+Def.~\ref{defn:rmrs-syntax}, and $\tau,\alpha,g,\sigma$ will satisfy
+$X=_q Y$ iff $\alpha(X)$ and $\alpha(Y)$ are separated by zero or more
+nodes in $\tau$, each of which has a third child that is a
+quantifier. All the above propositions about solved forms still hold
+if ``dominance'' is systematically replaced with ``qeq''.
 
-Furthermore, grammar developers such as those in the {\sc delphin}
-community typically adopt conventions that restrict them to a
-fragment of the language from Def.~\ref{defn:rmrs-syntax} (once
-qeq is added to it), or they restrict attention to only a subset of
-the models (e.g., any variable that equals the variable
-that's bound by a quantifier is also bound by that quantifier). 
-Our formalism provides a general framework
-into which all these various fragments fit, and it's a matter of
-future work to investigate how these fragments afford efficient
-algorithms for performing inference over the \rmrs s.
+Furthermore, grammar developers such as those in the {\sc delph-in}
+community typically adopt conventions that restrict them to a fragment
+of the language from Def.~\ref{defn:rmrs-syntax} (once qeq is added to
+it), or they restrict attention to only a subset of the models (e.g.,
+ones in which all variables are bound). Our formalism provides a
+general framework into which all these various fragments fit, and it's
+a matter of future work to explore these fragments further.
 
 Another feature of the existing \rmrs\ literature is that each term of
-an \rmrs\ is equipped 
-with a \emph{sort}.  For instance, individual variables are typically
-assigned the sort $x$, event variables the sort $e$, and holes the
-sort $h$; sometimes subsorts of these are considered (e.g.\ $e_{PAST}$
-for events in the past tense), and all the sorts arranged in a sort
-hierarchy ${\cal S}$.
-While we have defined \rmrs\ without sorts to simplify the
-presentation, it is straightforward to add sorts to the definition.
-For this, one assumes that the signature $\Sigma$ is sorted, i.e.\
-assigns a sort $s_1\times\ldots s_n\rightarrow s$ to each constructor,
-where $n$ is the constructor's arity (possibly zero) and
-$s,s_1,\ldots,s_n\in {\cal S}$ are atomic sorts.  We can then
-restrict the models of \rmrs\ to trees that are \emph{well-sorted} in
-the usual sense, i.e.\ those in which we can infer a sort for each
-subtree, and require that the variable assignment functions likewise
-respect the sorts.  If we then modify Def.~\ref{defn:solved-forms}
-such that the constructor $p$ of sufficiently high arity is also
-consistent with the sorts of the known arguments---i.e., if $p$ has
-sort $s_1 \times \ldots \times s_n \rightarrow s$ and the \rmrs\
-contains an atom $\ARG_{\{k\}}(Y,i)$ and $i$ is of sort $s'$, then
-$s'$ is a subsort of $s_k$---all the above propositions about solved
-forms remain true.
+an \rmrs\ is equipped with a \emph{sort}.  For instance, individual
+variables are typically assigned the sort $x$, event variables the
+sort $e$, and holes the sort $h$; these are arranged together with
+their subsorts in a sort hierarchy ${\cal S}$.  While we have defined
+\rmrs\ without sorts to simplify the presentation, it is
+straightforward to add them.  For this, one assumes that the signature
+$\Sigma$ is sorted, i.e.\ assigns a sort $s_1\times\ldots
+s_n\rightarrow s$ to each constructor, where $n$ is the constructor's
+arity (possibly zero) and $s,s_1,\ldots,s_n\in {\cal S}$ are atomic
+sorts.  We restrict the models of \rmrs\ to trees that are
+\emph{well-sorted} in the usual sense, i.e.\ those in which we can
+infer a sort for each subtree, and require that the variable
+assignment functions likewise respect the sorts.  If we then modify
+Def.~\ref{defn:solved-forms} such that the constructor $p$ of
+sufficiently high arity is also consistent with the sorts of the known
+arguments---i.e., if $p$ has sort $s_1 \times \ldots \times s_n
+\rightarrow s$ and the \rmrs\ contains an atom $\ARG_{\{k\}}(Y,i)$ and
+$i$ is of sort $s'$, then $s'$ is a subsort of $s_k$---all the above
+propositions about solved forms remain true.
 

### 2008-10-08 trunk/Papers/rmrs-08/conclusion.tex

@@
 is designed to be maximally flexible on the type of semantic
-information that can be left underspecified, so that
-the semantic output of a shallow parser needn't over-determine or
-under-determine the semantics that can be extracted from
-the shallow syntactic analysis.  This lays the formal foundations for
-a formalism that is emerging as a standard in
-robust semantic processing.
+information that can be left underspecified, so that the semantic
+output of a shallow parser needn't over-determine or under-determine
+the semantics that can be extracted from the shallow syntactic
+analysis.  The key contribution of this paper was to lay the formal
+foundations for a formalism that is emerging as a standard in robust
+semantic processing.
 
-We then defined the concept of \emph{solved forms} of an \rmrs, in
-analogy to dominance graphs.  Solved forms abstract away from
-irrelevant differences between models, and have been a crucial concept
-in the development of efficient solvers for dominance graphs.  In the
-future, we hope to draw upon that body of literature to develop
-efficient algorithms for processing \rmrs s, including solvers which
-will compute fully specific semantic representations from an \rmrs,
-efficient inference algorithms, and algorithms for entailment
-checking.  We have argued that entailment is a central notion in
-comparing the outputs of different {\sc nlp} tools, and provided a syntactic
-characterisation for \rmrs\ entailment, which we hope will allow us to
-develop efficient entailment algorithms in the future.
+Although we have not directly provided new tools for modelling or
+processing language, we believe that a cleanly defined model theory
+for \rmrs\ is a crucial prerequisite for the future development of
+such tools.  We hope that future research will build upon this paper
+to develop efficient algorithms and implementations for solving \rmrs
+s, performing inferences that enrich \rmrs s from shallow analyses
+with deeper information, and checking consistency of \rmrs s that were
+obtained from different parsers.
 

### 2008-10-08 trunk/Papers/rmrs-08/entailment.tex

@@
-\section{RMRS Comparison as Entailment}
+\section{Future work}
 \label{sec:entailment}
 
-The definitions from Section~\ref{sec:rmrs} serve an important theoretical
-purpose in that they provide a formal underpinning \rmrs's existing use
-for partial semantic representations in implemented
-systems.  Next to the peace of
-mind that comes with the use of a formally understood formalism, a
-clean definition can also provide a foundation for the development of
-efficient algorithms; this has been shown in the past in the dominance
-graph literature, most recently \newcite{KolRegTha08}.  We hope that
-this paper can likewise provide a starting point for the development
-of efficient solvers and inference mechanisms for \rmrs.
+The definitions from Section~\ref{sec:rmrs} serve an important
+theoretical purpose in that they provide a formal underpinning \rmrs's
+existing use for partial semantic representations in implemented
+systems.  Next to the peace of mind that comes with the use of a
+well-understood formalism, we hope that the work reported here will
+serve as a starting point for future work on efficient solvers and
+inference mechanisms.
 
-On the other hand, the above definition of \rmrs\ provides new
-concepts which can help us phrase questions of practical grammar
-engineering in well-defined formal terms.  For instance, one crucial
-issue in developing a hybrid system that combines the outputs of deep
-and shallow processors is to determine whether the \rmrs s produced by
-the two systems are compatible.  Because we assume that the deep
-grammar is more informative than that of the shallow processor, this
-amounts to checking whether every piece of information in the shallow
-representation is consistent with what the deep grammar said.  In
-general, it isn't even obvious what ``consistency'' is supposed to
-mean exactly, given that variables might be renamed, $\ARG$ roles
-assigned in different ways, and structural relations underspecified.
-But given the formal definition of \rmrs, we can now propose that an
-\rmrs\ $\varphi'$ that was computed by a shallow processor is
-consistent with an \rmrs\ $\varphi$ that was computed by a deep
-grammar if $\varphi$ entails $\varphi'$ (i.e., $\varphi \models
-\varphi'$).  If we could establish 
-efficiently that $\varphi'$ is entailed by $\varphi$, this would tell
-us that all models that satisfy $\varphi$ are also models of
-$\varphi'$, i.e.\ the deep representation is formally more specific.
+One direction which we believe may develop from this paper is the
+development of efficient solvers for \rmrs.  As a first step, it would
+be interesting to define a practically useful fragment of \rmrs\ with
+polynomial-time satisfiability.  Our definition is sufficiently close
+to that of dominance constraints that we expect that it should be
+feasible to carry over the definition of \emph{normal dominance
+  constraints} \cite{Althaus_etal:JoA} to \rmrs; neither the lexical
+ambiguity of the node labels nor the separate specification of
+predicates and arguments should make satisfiability harder.
 
-We cannot provide an efficient algorithm for testing entailment in
-this paper.  However, we propose the following novel syntactic
-characterisation of entailment.
+Furthermore, the above definition of \rmrs\ provides new concepts
+which can help us phrase questions of practical grammar engineering in
+well-defined formal terms.  For instance, one crucial issue in
+developing a hybrid system that combines or compares the outputs of
+deep and shallow processors is to determine whether the \rmrs s
+produced by the two systems are compatible.  In the new formal terms
+we can use now, we can characterise compatibility of a more detailed
+\rmrs\ $\varphi$ (perhaps from a deep grammar) and a less detailed
+\rmrs\ $\varphi'$ simply as \emph{entailment} $\varphi \models
+\varphi'$.  If entailment holds, this tells us that all claims that
+$\varphi'$ makes about the semantic content of a sentence is
+consistent with the claims that $\varphi$ makes.
 
-\begin{definition}
-  We call an \rmrs\ $\varphi'$ an \emph{extension} of an \rmrs\
-  $\varphi$ iff \todo{fill this in -- don't have the def with me right
-    now}.
-\end{definition}
+At this point, we cannot provide an efficient algorithm for testing
+entailment of \rmrs.  However, we propose the following novel
+syntactic characterisation which may again serve as a starting point
+for research along those lines:
 
-\begin{thm}\label{thm:big-one}
+\todo{define ``extension''!!}
+
+\begin{prop}\label{thm:big-one}
   Let $\varphi, \varphi'$ be two {\sc rmrs}s.  Then $\varphi \models
@@
   $S'$. 
-\end{thm}
-\begin{proof}[Proof (sketch)]
-``$\Leftarrow$'' follows from
-Props.~\ref{prop:solved-forms-are-satisfiable} and
-\ref{prop:models-satisfy-solved-forms}.
-
-``$\Rightarrow$'': The basic idea is to construct a solved form for
-$\varphi'$ by choosing a solved form for $\varphi$ and removing all
-atoms using variables that don't occur in $\varphi'$.  The hard part
-is the proof that the result is a solved form of $\varphi'$; this step
-involves proving that if $\varphi \models \varphi'$, then all {\sc ep}s in
-$\varphi'$ also occur in $\varphi$.  
-\end{proof}
-
-This characterisation does not by itself provide an efficient test for
-entailment because the number of solved forms of $\varphi$ and
-$\varphi'$ may be exponential, and thus it is infeasible to compute
-them all and then test for extensions.  Nevertheless, we hope that the
-theorem provides a first step towards an improved understanding of
-entailment that will lead to such an efficient algorithm in the
-future.
+\end{prop}
 

### 2008-10-08 trunk/Papers/rmrs-08/introduction.tex

@@
 %
+
 But there are certain semantic phenomena that these robust
@@
 Recursion Semantics (\rmrs, \cite{copestake:2007a,copestake:2007b}).
-\rmrs\ is 
-designed to
-support underspecification of scope and predicate-argument structure.
-It is an emerging standard for representing partial
-semantic information, and has been applied in several implemented
-systems.  For instance, \newcite{copestake:2003} and
-\newcite{frank:2004} use it to specify semantic components to shallow
-parsers ranging in depth from {\sc pos} taggers to chunk parsers and
-intermediate parsers such as {\sc rasp} \cite{briscoe:etal:2006}.
-{\sc mrs} analyses \cite{copestake:etal:2005}
-derived from deep grammars, such as the English
-Resource Grammar ({\sc erg}, \cite{copestake:flickinger:2000}) are
-special cases of {\sc rmrs}.
+\rmrs\ is designed to support underspecification of lexical ambiguity,
+scope, and predicate-argument structure.  It is an emerging standard
+for representing partial semantic information, and has been applied in
+several implemented systems.  For instance, \newcite{copestake:2003}
+and \newcite{frank:2004} use it to specify semantic components to
+shallow parsers ranging in depth from {\sc pos} taggers to chunk
+parsers and intermediate parsers such as {\sc rasp}
+\cite{briscoe:etal:2006}.  {\sc mrs} analyses
+\cite{copestake:etal:2005} derived from deep grammars, such as the
+English Resource Grammar ({\sc erg}, \cite{copestake:flickinger:2000})
+are special cases of {\sc rmrs}.
 
-However, from a formal perspective, \rmrs\ is somewhat ad-hoc in that
-it has no clearly defined model theory; i.e., what exactly any given
-\rmrs\ representation actually \emph{means} is not defined.  The key
-contribution we make is to cast {\sc rmrs}, for the first time, as a
-logic with a well-defined model theory.  This has a number of
-advantages.  First, it allows us to formally relate \rmrs\ to other
-frameworks such as \mrs\ and predicate logic in order to establish the
-compatibility of \rmrs\ with the approaches to robust semantic
-analysis mentioned above.  
-Second, it provides us with a notion of logical
-entailment, which we can use to verify the compatibility of \rmrs\ 
-representations computed from different parsers, such as a deep and a
-shallow parser.  Finally, it is only through rigid formalisation that
-we can give users of \rmrs\ access to inference systems and to
-algorithms for efficiently computing the possible fully specific
-semantic representations described by the \rmrs\ output of a shallow
-parser.
+The key contribution we make is to cast {\sc rmrs}, for the first
+time, as a logic with a well-defined model theory.  Previously, no
+such model theory existed, and so \rmrs\ had to be used in a somewhat
+ad-hoc manner which left it open what exactly any given \rmrs\
+representation actually \emph{means}.  This has hindered quite
+practical progress in terms of understanding the relationship of
+\rmrs\ to other frameworks such as \mrs\ and predicate logic and in
+terms of the development of efficient algorithms.  As one application
+of the formal concepts that come with our definitions, we propose a
+novel way of characterising consistency of \rmrs\ analyses across
+different parsers as entailment.
 
-Section~\ref{sec:motivation} illustrates via example the differences
-and challenges with deep and shallow semantic construction.
+Section~\ref{sec:motivation} introduces \rmrs\ informally and
+illustrates why it is necessary and useful for representing semantic
+information across deep and shallow language processors.
 Section~\ref{sec:rmrs} defines the syntax and model-theory of \rmrs.
-In Section~\ref{sec:entailment} we demonstrate the practical
-usefulness of our formalisation by modelling compatibility of partial
-semantic representations as \rmrs\ entailment and by providing a
-syntactic criterion for characterising entailment.
-Section~\ref{sec:conclusion} concludes and points to future work.
+We finish in Section~\ref{sec:entailment} by pointing out some avenues
+for future research, and conclude in Section~\ref{sec:conclusion}.
 

### 2008-10-08 trunk/Papers/rmrs-08/rmrs.tex

@@
 semantic representation as in Fig.~\ref{fig:fat-black-cat} would
-\emph{not} be taken as described by this \rmrs\.  However, under the
+\emph{not} be taken as described by this \rmrs.  However, under the
 semantics we proposed above, this tree is a correct model of
@@
 $X=_q Y$ iff $\alpha(X)$ and $\alpha(Y)$ are separated by zero or more
-nodes in $\tau$, each of which has a third child that is a
+nodes in $\tau$, each of which has a rightmost child that is a
 quantifier. All the above propositions about solved forms still hold

### 2008-10-09 trunk/Papers/rmrs-08/entailment.tex
shortened and polished

@@
 The definitions from Section~\ref{sec:rmrs} serve an important
-theoretical purpose in that they provide a formal underpinning \rmrs's
+theoretical purpose in that they provide a formal underpinning for \rmrs's
 existing use for partial semantic representations in implemented

### 2008-10-09 trunk/Papers/rmrs-08/introduction.tex
shortened and polished

@@
 such model theory existed, and so \rmrs\ had to be used in a somewhat
-ad-hoc manner which left it open what exactly any given \rmrs\
+ad-hoc manner which left open what exactly any given \rmrs\
 representation actually \emph{means}.  This has hindered quite
-practical progress in terms of understanding the relationship of
+practical progress, both in terms of understanding the relationship of
 \rmrs\ to other frameworks such as \mrs\ and predicate logic and in
 terms of the development of efficient algorithms.  As one application
-of the formal concepts that come with our definitions, we propose a
+of the formal concepts that arise from our definitions, we use
+entailment to propose a
 novel way of characterising consistency of \rmrs\ analyses across
-different parsers as entailment.
+different parsers.
 

### 2008-10-09 trunk/Papers/rmrs-08/motivation.tex
shortened and polished

@@
 
-We start with an informal illustration of why \rmrs\ is necessary and
-useful to represent partial semantic information.  Consider the
+Consider the
 following (toy) sentence:
@@
 
-This sentence exhibits several kinds of ambiguity, including a scope
-ambiguity between ``every fat cat'' and ``some dog'' and lexical
-ambiguities of the nouns ``cat'' and ``dog'' (which have 8 and 7
-WordNet senses respectively).  Simplifying slightly by ignoring tense
+It exhibits several kinds of ambiguity, including a quantifier scope
+ambiguity and lexical
+ambiguities---e.g., the nouns ``cat'' and ``dog'' have 8 and 7
+WordNet senses respectively.  Simplifying slightly by ignoring tense
 information, two of these readings are shown as logical forms below;
@@
 \includegraphics[scale=0.8]{pic-cat-chased-dog-2}
-\caption{Structure trees of the semantic representations (\ref{ex:fat-cat-1}) and
-  (\ref{ex:fat-cat-2}). \label{fig:1}}
+\caption{Semantic representations (\ref{ex:fat-cat-1}) and
+  (\ref{ex:fat-cat-2}) as trees. \label{fig:1}}
 \end{figure}
@@
 output of a part-of-speech tagger by using the word lemmas as lexical
-predicate symbols.  Such a semantic representation is highly partial,
-as it will not say anything about predicate-argument relations or
-resolve lexical ambiguity---it will use predicate symbols such as
+predicate symbols.  Such a semantic representation is highly partial.
+It will use predicate symbols such as
 $\sempred{\_cat\_n}$, which might resolve to the predicate symbols
@@
 unambiguous predicate symbols.)
-
-Partial semantic information is
-typically represented using an underspecification formalism such as
-\mrs\ \cite{copestake:etal:2005} or {\sc clls} \cite{egg:etal:inpress}.  But
-most underspecification formalisms are unable to represent semantic
-information that is as partial as what we get from a {\sc pos} tagger: They
-assume that the complete predicate-argument structure is known.
+But
+most underspecification formalisms (e.g., {\sc mrs}
+\cite{copestake:etal:2005} and {\sc clls} \cite{egg:etal:2001}) are
+unable to represent semantic 
+information that is as partial as what we get from a {\sc pos} tagger
+because they cannot underspecify
+predicate-argument structure.
 \rmrs\ \cite{copestake:2007a} was
@@
 This \rmrs\ expresses only that certain predications must be made in
-the semantic representation -- but it doesn't say anything about the
-structure of the semantic representation, about most arguments of the
+the semantic representation -- but it doesn't say anything about
+semantic scope, about most arguments of the
 predicates (note that $\sempred{\_chase\_v}(e)$ doesn't say who chases
@@
 $x_1$ and $x_3$ into variable names in the tree in such a way that the
-predication of the node that labels and anchors are mapped to is
-consistent with that given in the {\sc ep}s of the {\sc rmrs}.  For
+predications of the nodes that labels and anchors are mapped to are 
+consistent with those in the {\sc ep}s of the {\sc rmrs}.  For
 instance, the first tree satisfies the \rmrs\ because $l_1$ and $a_1$
@@
 surprising, given that the semantic information we got from the {\sc
-  pos} tagger is so incomplete.  If we have more information, say
-information about subject and object relations from a chunk parser
-like Cass \cite{abney:1996}, we can represent these in a more detailed
-\rmrs, as follows:
+  pos} tagger is so incomplete.  If we have
+information about subjects and objects from a chunk parser
+like Cass \cite{abney:1996}, we can represent it in a more detailed
+\rmrs:
 
@@
 
-This \rmrs\ uses two new types of atoms.  Atoms of the form $x_3=x_4$
-express that the two variables $x_3$ and $x_4$ must map to the same
+This introduces two new types of atoms.   $x_3=x_4$
+expresses that the two variables $x_3$ and $x_4$ must map to the same
 variable in any fully specific logical form; e.g., both to the
@@
 even specify the arguments to a predicate while underspecifying their
-position.  This is useful for specifying arguments to \_give\_v when a
-parser that is unableto process unbounded dependencies is faced with
+position.  This can useful for specifying arguments to \_give\_v when a
+parser doesn't handle unbounded dependencies and is faced with
 {\em Which bone did you give the dog?} vs.\ {\em To which dog did you
@@
 
-Here we have full information about the predicate-argument structure
-and the variable binding; we only underspecify scope and some lexical
-ambiguity.  $\mathsf{RSTR}$ and $\mathsf{BODY}$ are conventional names
+$\mathsf{RSTR}$ and $\mathsf{BODY}$ are conventional names
 for the $\ARG_1$ and $\ARG_2$ of a quantifier predicate symbol.  Atoms
@@
 the surrounding formula.  Representing conjunction with label sharing
-stems from \mrs\ and provides compact semantic representations.
+stems from \mrs\ and provides compact representations.
 
@@
 represented using SPEC atoms like $\sempred{\_dog\_n\_1}\sqsubseteq
-\sempred{\_dog\_n}$), which we have omitted in (\ref{ex:cat-erg}) for
-brevity.  Note that even a deep grammar will not fully disambiguate to
+\sempred{\_dog\_n}$.
+Note that even a deep grammar will not fully disambiguate to
 semantic predicate symbols, such as WordNet senses, and so
@@
 like $\sem{\_dog\_n\_1}$ and $\sem{\_dog\_n\_2}$ in the semantic
-representation.
+representation.  However, unlike the symbols in the \rmrs\ output from a
+{\sc pos} tagger, those from a deep grammar will always be consistent
+with symbols of the {\em same} arity, because a deep grammar
+fully determines lexical subcategorisation.
 
@@
 format for exactly these purposes \cite{copestake:2003,frank:2004}.
-However, \rmrs\ is still relatively ad-hoc in that its formal syntax
-is not consistent across different papers, and its formal semantics is
+However, \rmrs\ is still relatively ad-hoc in that its
+formal semantics is
 not defined; we don't know, formally, what an \rmrs\ \emph{means} in

### 2008-10-09 trunk/Papers/rmrs-08/rmrs.tex
shortened and polished

@@
   following forms; $S \subseteq \N$ is a set of numbers that is either
-  finite or $\N$ itself.\footnote{Throughout the paper, we assume that
-    $0 \in \N$.}
+  finite or $\N$ itself (throughout the paper, we assume 
+    $0 \in \N$).
 
@@
 presentations of \rmrs\ \cite{copestake:2003,copestake:2007b} and to
-CLLS/dominance constraints \cite{egg:etal:2001}.  For the most part,
-our syntax generalizes that of older versions of \rmrs: We use
+{\sc clls}/dominance constraints \cite{egg:etal:2001}.  For the most part,
+our syntax generalises that of older versions of \rmrs: We use
 $\ARG_{\{i\}}$ (with a singleton set $S$) instead of $\ARG_i$ and
-$\ARG_\N$ instead of $\ARG_n$, and we take the EP $\ep{l}{a}{P(v)}$
-(as in Section~\ref{sec:motivation}) as an abbreviation of
+$\ARG_\N$ instead of $\ARG_n$, and the{\sc ep} $\ep{l}{a}{P(v)}$
+(as in Section~\ref{sec:motivation}) is an abbreviation of
 $\{\ep{l}{a}{P}, \ARG_{\{0\}}(a,v)\}$.  Similarly, we don't assume
@@
 $\neg$ and binary constructors $\wedge$, $\rightarrow$, etc.; a set of
-3-place quantifier symbols such as $\sem{\_forall\_q\_1}$ and
+3-place quantifier symbols such as $\sem{\_every\_q\_1}$ and
 $\sem{\_some\_q\_1}$ (with the children being the bound variable, the
-restrictor, and the scope); and, for each predicate symbol $P$ of
-arity $n$, a constructor $P$ of arity $n$ \todo{$n$ or $n+1$?}.  Other
+restrictor, and the scope); and, constructors of various arities for
+the predicate symbols; e.g., $\sem{\_chase\_v\_1}$ is of arity 3.  Other
 base languages may require a different signature $\Sigma$ and/or a
@@
 constructor $\wedge$ to represent conjunction.  We write $\Sigma_i$
-for the set of all constructors in $\Sigma$ that have arity $i$.  We
+for the set of all constructors in $\Sigma$ with arity $i$.  We
 will follow the typographical convention that non-logical symbols in
 $\Sigma$ are written in $\mathsf{sans{-}serif}$, as opposed to the
-\rmrs\ predicate symbols like ``$\sempred{\_cat\_n}$'' and
-``$\sempred{\_cat\_n\_1}$''.
+\rmrs\ predicate symbols like $\sempred{\_cat\_n}$ and
+$\sempred{\_cat\_n\_1}$.
 
@@
 base language assignment function $g:\ovar \rightarrow \Sigma_0$ that
-maps the base variables in the \rmrs\ to nullary constructors
+maps the base variables to nullary constructors
 representing variables in the base language. Finally, we assume a
@@
 each \rmrs\ predicate symbol to a set of constructors from $\Sigma$.
-This relation allows us to underspecify lexical ambiguities; for
-instance, $\sigma(\sempred{\_cat\_n})$ might contain the symbol
-$\sem{\_cat\_n\_1}$ along with symbols for other word senses.
+As we'll see shortly, this relation allows us to underspecify lexical
+ambiguities.
 
@@
 from $\tau_2(\alpha_2(Y))$ (i.e., the constructors are of a different
-arity and/or their arguments are).  And it exhibits missing lexical
+arity or they differ in whether their arguments are scopal vs.\
+non-scopal).  And it exhibits missing lexical
 sense information if $\tau_1(\alpha_1(Y))$ and $\tau_2(\alpha_2(Y))$
@@
 
-Let's look at the \rmrs\ in (\ref{ex:cat-pos}) by way of example.
-This \rmrs\ is satisfied by the trees in Fig.~\ref{fig:1} (among
+Let's look again at the \rmrs\ (\ref{ex:cat-pos}).
+This is satisfied by the trees in Fig.~\ref{fig:1} (among
 others) together with some particular $\alpha$, $g$, and $\sigma$.
@@
 $y$, and so on.  And $\sigma$ will map each \rmrs\ predicate symbol
-(which represents a word) to the set of its word senses, e.g.\
+(which represents a word) to the set of its fully resolved meanings, e.g.\
 \_cat\_n to a set containing $\sem{\_cat\_n\_1}$ and possibly others.
@@
 infinite number of models which only differ in the areas that the
-\rmrs\ didn't ``talk about''.  Consider e.g.\ (\ref{ex:cat-erg}).
-Read as an \mrs\ or as an \rmrs\ of the previous literature, this
+\rmrs\ didn't ``talk about''.  Reading (\ref{ex:cat-erg})
+as an \mrs\ or as an \rmrs\ of the previous literature, this
 formula is an instruction to build a semantic representation out of
@@
 feature of \rmrs, because it means that when we enrich an \rmrs\
-obtained from a shallow processor with more semantic information --
-such as the relational information introduced by syntactic
-constructions such as appositives, noun-noun compounds, and
-nonrestrictive relative clauses --, we don't change the set of models;
+obtained from a shallow processor with more semantic
+information---such as the relation symbols introduced by
+syntactic 
+constructions such as appositives, noun-noun compounds and free
+adjuncts---we don't change the set of models;
 we only \emph{restrict} the set of models further and further towards
@@
 \rmrs\ has only a finite number of solved forms which partition the
-space of possible models into classes with only ``irrelevant''
-differences as above.  A solver can then enumerate the solved forms
+space of possible models into classes with only `relevant'
+differences as in Fig.~\ref{fig:1}.  A solver can then enumerate the
+solved forms 
 rather than all models.
@@
 
-Because solved forms are so restricted, it is now possible ``read
-off'' at least one model from each solved form.  We don't have the
+Because solved forms are so restricted, it is now possible to `read
+off' at least one model from each solved form.  We don't have the
 space to present any proofs here; they proceed mostly along the same
@@
 \begin{prop} \label{prop:solved-forms-are-satisfiable}
-  Every RMRS in solved form is satisfiable.
+  Every {\sc rmrs} in solved form is satisfiable.
 \end{prop}
@@
 on the right-hand side of two dominance atoms.  However, this is
-because the model that we can read off of (\ref{ex:cat-erg}) is
-similar to the one shown in Fig.~\ref{fig:fat-black-cat}; this model
-is problematic in several ways and in particular contains unbound
-variables.  If we restrict the class of models we consider to those in
-which variables are bound, we can add the information that the
+because (at least) one model of (\ref{ex:cat-erg}) is
+shown in Fig.~\ref{fig:fat-black-cat}; this model
+is problematic in several ways and in particular contains an unbound
+variable $y$ even though it also contains a quantifier that binds $y$.
+If we restrict the class of models to those in 
+which such variables are bound (as \newcite{copestake:etal:2005} do),
+we can add the information that the 
 quantifiers outscope their bound variables to the \rmrs\ without
-changing its models further -- i.e., we add the atoms $h_3 \dom
-l_5$ and $h_8 \dom l_5$.  The extended \rmrs\ is no longer in solved
+changing its models further---i.e., we add the atoms $h_3 \dom
+l_5$ and $h_8 \dom l_5$.  Fig.~\ref{fig:fat-black-cat} is no longer a
+model for the extended \rmrs, whic in turn is no longer in solved
 form because the label $l_5$ is on the right-hand side of two
@@
 \item \label{ex:sf-1}
-  $l_1\handel a_1\handel\sempred{\_every\_q\_1}(x_1)$,
-  $\mathsf{RSTR}(a_1,h_2)$, $\mathsf{BODY}(a_1,h_3)$,
-  $l_{41}\handel a_{41}\handel\sempred{\_fat\_j\_1}(e'), \ARG_1(a_{41},x_1)$,
-  $l_{41}\handel a_{42}\handel\sempred{\_cat\_n\_1}(x_1)$,
+  $l_1\handel a_1\handel\sempred{\_every\_q\_1}(x_1)$,\\
+\hspace*{0.1in}$\mathsf{RSTR}(a_1,h_2)$, $\mathsf{BODY}(a_1,h_3)$,\\
+  $l_{41}\handel a_{41}\handel\sempred{\_fat\_j\_1}(e'), \ARG_1(a_{41},x_1)$,\\
+  $l_{41}\handel a_{42}\handel\sempred{\_cat\_n\_1}(x_1)$,\\
   %
-  $l_6\handel a_6\handel\sempred{\_some\_q\_1}(x_6)$,
-  $\mathsf{RSTR}(a_6,h_7)$, $\mathsf{BODY}(a_6,h_8)$,
-  $l_9\handel a_9\handel\sempred{\_dog\_n\_1}(x_6)$,
+  $l_6\handel a_6\handel\sempred{\_some\_q\_1}(x_6)$,\\
+\hspace*{0.1in}$\mathsf{RSTR}(a_6,h_7)$, $\mathsf{BODY}(a_6,h_8)$,\\
+  $l_9\handel a_9\handel\sempred{\_dog\_n\_1}(x_6)$,\\
   %
-  $l_5\handel a_5\handel\sempred{\_chase\_v\_1}(e)$,
-  $\ARG_1(a_5,x_1)$, $\ARG_2(a_5,x_6)$,
+  $l_5\handel a_5\handel\sempred{\_chase\_v\_1}(e)$,\\
+\hspace*{0.1in}  $\ARG_1(a_5,x_1)$, $\ARG_2(a_5,x_6)$,\\
   $h_2 \dom l_{41}$, $h_3 \dom l_6$,
@@
 \item \label{ex:sf-2}
-  $l_1\handel a_1\handel\sempred{\_every\_q\_1}(x_1)$,
-  $\mathsf{RSTR}(a_1,h_2)$, $\mathsf{BODY}(a_1,h_3)$,
-  $l_{41}\handel a_{41}\handel\sempred{\_fat\_j\_1}(e'), \ARG_1(a_{41},x_1)$,
-  $l_{41}\handel a_{42}\handel\sempred{\_cat\_n\_1}(x_1)$,
+  $l_1\handel a_1\handel\sempred{\_every\_q\_1}(x_1)$,\\
+\hspace*{0.1in}$\mathsf{RSTR}(a_1,h_2)$, $\mathsf{BODY}(a_1,h_3)$,\\
+  $l_{41}\handel a_{41}\handel\sempred{\_fat\_j\_1}(e'), \ARG_1(a_{41},x_1)$,\\
+  $l_{41}\handel a_{42}\handel\sempred{\_cat\_n\_1}(x_1)$,\\
   %
-  $l_6\handel a_6\handel\sempred{\_some\_q\_1}(x_6)$,
-  $\mathsf{RSTR}(a_6,h_7)$, $\mathsf{BODY}(a_6,h_8)$,
-  $l_9\handel a_9\handel\sempred{\_dog\_n\_1}(x_6)$,
+  $l_6\handel a_6\handel\sempred{\_some\_q\_1}(x_6)$,\\
+\hspace*{0.1in}$\mathsf{RSTR}(a_6,h_7)$, $\mathsf{BODY}(a_6,h_8)$,\\
+  $l_9\handel a_9\handel\sempred{\_dog\_n\_1}(x_6)$,\\
   %
-  $l_5\handel a_5\handel\sempred{\_chase\_v\_1}(e)$,
-  $\ARG_1(a_5,x_1)$, $\ARG_2(a_5,x_6)$,
+  $l_5\handel a_5\handel\sempred{\_chase\_v\_1}(e)$,\\
+\hspace*{0.1in}  $\ARG_1(a_5,x_1)$, $\ARG_2(a_5,x_6)$,\\
   $h_2 \dom l_{41}$, $h_3 \dom l_5$,
@@
 quantifier. All the above propositions about solved forms still hold
-if ``dominance'' is systematically replaced with ``qeq''.
+if ``dominance'' is replaced with ``qeq''.
 
@@
 of the language from Def.~\ref{defn:rmrs-syntax} (once qeq is added to
-it), or they restrict attention to only a subset of the models (e.g.,
-ones in which all variables are bound). Our formalism provides a
+it), or they restrict attention to only a subset of the models.
+Our formalism provides a
 general framework into which all these various fragments fit, and it's

### 2008-10-09 trunk/Papers/rmrs-08/introduction.tex
some new stuff

@@
 tends to lack robustness: they fail to cover all
-words and linguistic constructions and by design do not handle
+words and constructions and by design don't handle
 ill-formed phrases.
-There has thus been a trend recently towards systems for
+
+There has thus been a trend recently towards
 robust wide-coverage semantic construction (e.g., 
@@
 % 
-
 But there are certain semantic phenomena that these robust
@@
 Recursion Semantics (\rmrs, \cite{copestake:2007a,copestake:2007b}).
-\rmrs\ is designed to support underspecification of lexical ambiguity,
+\rmrs\ is designed to support underspecification of lexical information,
 scope, and predicate-argument structure.  It is an emerging standard

### 2008-10-09 trunk/Papers/rmrs-08/motivation.tex
some new stuff

@@
 WordNet senses respectively.  Simplifying slightly by ignoring tense
-information, two of these readings are shown as logical forms below;
+information, two of its readings are shown as logical forms below;
 these can be represented as trees as shown in Fig.~\ref{fig:1}.
@@
 This \rmrs\ expresses only that certain predications must be made in
-the semantic representation -- but it doesn't say anything about
+the semantic representation---but it doesn't say anything about
 semantic scope, about most arguments of the
@@
 binds the variable $x_1$, whereas $\sempred{\_cat\_n}$ speaks about
-$x_3$), and maintains the lexical ambiguities.  Technically, it
+$x_3$), and it maintains the lexical ambiguities.  Technically, it
 consists of six \emph{elementary predications} ({\sc ep}s), one for
@@
 This introduces two new types of atoms.   $x_3=x_4$
-expresses that the two variables $x_3$ and $x_4$ must map to the same
+means that $x_3$ and $x_4$ map to the same
 variable in any fully specific logical form; e.g., both to the
-variable $x$ in Fig.~\ref{fig:1}.  Atoms of the form $\ARG_i(a,z)$ and
+variable $x$ in Fig.~\ref{fig:1}.   $\ARG_i(a,z)$ and
 $\ARG_i(a,h)$ express that the $i$-th child (counting from 0) of the
@@
 earlier underspecification formalisms, \rmrs\ can specify the
-predicate of an atom separately from its (scopal and non-scopal)
+predicate of an atom separately from its
 arguments; this is necessary for supporting parsers where information
@@
 This reflects the fact that the deep grammar will perform some lexical
-disambiguation, whereas the chunker and {\sc pos} tagger didn't.  The fact
+disambiguation that the chunker and {\sc pos} tagger didn't.  The fact
 that the former symbol should be more specific than the latter can be
@@
 like $\sem{\_dog\_n\_1}$ and $\sem{\_dog\_n\_2}$ in the semantic
-representation.  However, unlike the symbols in the \rmrs\ output from a
-{\sc pos} tagger, those from a deep grammar will always be consistent
-with symbols of the {\em same} arity, because a deep grammar
+representation.  However, unlike the output of a
+{\sc pos} tagger, an \rmrs\ symbol that's output by a deep grammar is
+consistent 
+with symbols that all have the {\em same} arity, because a deep grammar
 fully determines lexical subcategorisation.
 
-In summary, \rmrs\ is a formalism for partial semantic
-representations.  It allows us to represent the semantic information
+In summary, \rmrs\ allows us to represent the (partial) semantic information
 that can be extracted from a wide range of {\sc nlp} tools in a uniform way.

### 2008-10-09 trunk/Papers/rmrs-08/rmrs.tex
some new stuff

@@
 
-\rmrs\ as presented here combines similarities to earlier
+Def.\ref{defn:rmrs-syntax}
+combines similarities to earlier
 presentations of \rmrs\ \cite{copestake:2003,copestake:2007b} and to
@@
 $\ARG_{\{i\}}$ (with a singleton set $S$) instead of $\ARG_i$ and
-$\ARG_\N$ instead of $\ARG_n$, and the{\sc ep} $\ep{l}{a}{P(v)}$
+$\ARG_\N$ instead of $\ARG_n$, and the {\sc ep} $\ep{l}{a}{P(v)}$
 (as in Section~\ref{sec:motivation}) is an abbreviation of
@@
 \begin{definition}\label{defn:dominance}
-  $u \dom v$ iff $u$ is a prefix of $v$, i.e.\ the node $u$ is above
+  $u \dom v$ iff $u$ is a prefix of $v$, i.e.\ the node $u$ is equal
+  to or above
   the node $v$ in the tree.  (This is the dominance relation familiar
@@
   $u \wedgedom v$ iff $u \dom v$, and all symbols on the path from $u$
-  to $v$ (including $u$ but not $v$) are $\wedge$.
+  to $v$ (not including $v$) are $\wedge$.
 \end{definition}
@@
 finite constructor tree $\tau$ is defined in terms of several
-assignment functions.  First, we assume a node variable assignment
-function $\alpha: \nvar \rightarrow D(\tau)$ that maps the node
-variables in an \rmrs\ to the nodes of $\tau$.  Second, we assume a
-base language assignment function $g:\ovar \rightarrow \Sigma_0$ that
+assignment functions.  First, a node variable assignment
+function $\alpha: \nvar \rightarrow D(\tau)$ maps the node
+variables in an \rmrs\ to the nodes of $\tau$.  Second, a
+base language assignment function $g:\ovar \rightarrow \Sigma_0$ 
 maps the base variables to nullary constructors
-representing variables in the base language. Finally, we assume a
-relation $\sigma \subseteq \pred \times \Sigma_{\geq 1}$ that maps
+representing variables in the base language. Finally, a
+relation $\sigma \subseteq \pred \times \Sigma_{\geq 1}$ maps
 each \rmrs\ predicate symbol to a set of constructors from $\Sigma$.
-As we'll see shortly, this relation allows us to underspecify lexical
+As we'll see shortly, this relation allows an \rmrs\ to underspecify lexical
 ambiguities.
@@
 dominance-constraint style semantics for underspecified
-representations gives us more room to maneuver when developing
+representations gives us more room to manoeuvre when developing
 efficient solvers than an \mrs-style semantics
@@
 \rmrs\ has only a finite number of solved forms which partition the
-space of possible models into classes with only `relevant'
-differences as in Fig.~\ref{fig:1}.  A solver can then enumerate the
+space of possible models into classes such that models within a class
+differ only in `irrelevant' details.  A solver can then enumerate the
 solved forms 
@@
 l_5$ and $h_8 \dom l_5$.  Fig.~\ref{fig:fat-black-cat} is no longer a
-model for the extended \rmrs, whic in turn is no longer in solved
+model for the extended \rmrs, which in turn is no longer in solved
 form because the label $l_5$ is on the right-hand side of two

### 2008-10-09 trunk/Papers/rmrs-08/entailment.tex
put proofs back in; played around with baselinestretch

@@
 syntactic characterisation which may again serve as a starting point
-for research along those lines:
-
-\todo{define ``extension''!!}
+for research along those lines. We call an \rmrs\ $\varphi'$ an
+\emph{extension} of the \rmrs\ $\varphi$ if $\varphi'$ contains all
+the EPs of $\varphi$ and $D(\varphi') \supseteq D(\varphi)$.
 
@@
 \end{prop}
+\begin{proof}[Proof (sketch)]
+``$\Leftarrow$'' follows from
+Props.~\ref{prop:solved-forms-are-satisfiable} and
+\ref{prop:models-satisfy-solved-forms}.
+
+``$\Rightarrow$'': We construct a solved form for
+$\varphi'$ by choosing a solved form for $\varphi$ and appropriate
+substitutions for mapping the variables of $\varphi$ and $\varphi'$
+onto each other, and removing all atoms using variables that don't
+occur in $\varphi'$ .  The hard part is the proof that the result is a
+solved form of $\varphi'$; this step involves proving that if $\varphi
+\models \varphi'$ with the same variable assignments, then all {\sc
+  ep}s in $\varphi'$ also occur in $\varphi$.
+\end{proof}
 

### 2008-10-09 trunk/Papers/rmrs-08/rmrs.tex
put proofs back in; played around with baselinestretch

@@
 
-Because solved forms are so restricted, it is now possible to `read
-off' at least one model from each solved form.  We don't have the
-space to present any proofs here; they proceed mostly along the same
-lines as the proofs in \newcite{Duchier00dominanceconstraints}.
+Because solved forms are so restricted, we can `read off' at least one
+model from each solved form.
 
@@
 \end{prop}
+\begin{proof}[Proof (sketch; see also
+  \cite{Duchier00dominanceconstraints})] 
+  For each {\sc ep}, we choose to label the anchor with the
+  constructor $p$ of sufficiently high arity whose existence we
+  assumed; we determine the edges between an anchor and its children
+  from the uniquely determined $\ARG$ atoms; plugging labels into
+  holes is straightforward because no label is dominated by more than
+  one label; and then we can fill the spaces between the labels and
+  anchors with conjunctions.
+\end{proof}
 
@@
 \end{prop}
+\begin{proof}
+  We construct the substitution $s$ from $\alpha$ and $g$.  Then we
+  add all dominance atoms that are satisfied by $\alpha$ and restrict
+  the $\ARG$ atoms to those child indices that are actually used in
+  $\tau$.  The result is in solved form because $\tau$ is a tree; it
+  is a solved form of $\varphi$ by construction.
+\end{proof}
 
@@
 \end{prop}
+\begin{proof}
+  Up to renaming of variables, there is only a finite number of
+  substitutions on the node and base variables of $\varphi$.  Let $s$
+  be such a substitution.  This fixes the set of {\sc ep}s of any
+  solved form of $\varphi$ that is based on $s$ uniquely.  There is
+  only a finite set of choices for the subsets $S'$ in condition 2 of
+  Def.~\ref{defn:solved-form-of}, and there is only a finite set of
+  choices of new dominance atoms that satisfy condition 3.  Therefore,
+  the set of solved forms of $\varphi$ is finite.
+\end{proof}
 

### 2008-10-10 trunk/Papers/rmrs-08/introduction.tex
titlebox back to normal

@@
 can be obtained from hand-crafted deep grammars
-\cite{butt:etal:1999,copestake:flickinger:2000}, but this
-tends to lack robustness: they fail to cover all
-words and constructions and by design don't handle
-ill-formed phrases.
-
+\cite{butt:etal:1999,copestake:flickinger:2000} but 
+this lacks robustness: not all
+words and constructions are covered and by design
+ill-formed phrases fail to parse. 
 There has thus been a trend recently towards
@@
 
-In this paper, we focus on an approach to representing partial
-semantic information that supports this strategy: Robust Minimal
-Recursion Semantics (\rmrs, \cite{copestake:2007a,copestake:2007b}).
+In this paper, we focus on an approach to
+semantic representation that supports this strategy: Robust Minimal
+Recursion Semantics (\rmrs, \newcite{copestake:2007a}).
 \rmrs\ is designed to support underspecification of lexical information,
 scope, and predicate-argument structure.  It is an emerging standard
-for representing partial semantic information, and has been applied in
+for representing partial semantics, and has been applied in
 several implemented systems.  For instance, \newcite{copestake:2003}
@@
 such model theory existed, and so \rmrs\ had to be used in a somewhat
-ad-hoc manner which left open what exactly any given \rmrs\
-representation actually \emph{means}.  This has hindered quite
+ad-hoc manner that left open exactly what any given \rmrs\
+representation actually \emph{means}.  This has hindered 
 practical progress, both in terms of understanding the relationship of
@@
 terms of the development of efficient algorithms.  As one application
-of the formal concepts that arise from our definitions, we use
+of our formalisation, we use
 entailment to propose a

### 2008-10-10 trunk/Papers/rmrs-08/motivation.tex
titlebox back to normal

@@
 following (toy) sentence:
-
 \begin{examples}
+\label{ex:cat}
   \item Every fat cat chased some dog.
 \end{examples}
-
 It exhibits several kinds of ambiguity, including a quantifier scope
@@
 
-Now imagine we are trying to extract semantic information from the
-output of a part-of-speech tagger by using the word lemmas as lexical
-predicate symbols.  Such a semantic representation is highly partial.
-It will use predicate symbols such as
-$\sempred{\_cat\_n}$, which might resolve to the predicate symbols
-$\sem{\_cat\_n\_1}$ or $\sem{\_cat\_n\_2}$ in the complete semantic
-representation.  (Notice the different fonts for the ambiguous and
-unambiguous predicate symbols.)
-But
-most underspecification formalisms (e.g., {\sc mrs}
+Now imagine trying to extract semantic information from the output of
+a part-of-speech ({\sc pos}) tagger by using the word lemmas as
+lexical predicate symbols.  Such a semantic representation is highly
+partial.  It will use predicate symbols such as $\sempred{\_cat\_n}$,
+which might resolve to the predicate symbols $\sem{\_cat\_n\_1}$ or
+$\sem{\_cat\_n\_2}$ in the complete semantic representation.  (Notice
+the different fonts for the ambiguous and unambiguous predicate
+symbols.)  But most underspecification formalisms (e.g., {\sc mrs}
 \cite{copestake:etal:2005} and {\sc clls} \cite{egg:etal:2001}) are
-unable to represent semantic 
-information that is as partial as what we get from a {\sc pos} tagger
-because they cannot underspecify
-predicate-argument structure.
-\rmrs\ \cite{copestake:2007a} was
-designed to address this problem.  In \rmrs, we can express the
-information we get from the {\sc pos} tagger as follows:
+unable to represent semantic information that is as partial as what we
+get from a {\sc pos} tagger because they cannot underspecify
+predicate-argument structure.  \rmrs\ \cite{copestake:2007a} is 
+designed to address this problem.  In \rmrs, the
+information we get from the {\sc pos} tagger is as follows:
 
@@
 
-This \rmrs\ expresses only that certain predications must be made in
-the semantic representation---but it doesn't say anything about
+This \rmrs\ expresses only that certain predications are present in
+the semantic representation---it doesn't say anything about
 semantic scope, about most arguments of the
-predicates (note that $\sempred{\_chase\_v}(e)$ doesn't say who chases
+predicates (e.g., $\sempred{\_chase\_v}(e)$ doesn't say who chases
 whom), or about the coindexation of variables ($\sempred{\_every\_q}$
@@
 the two trees \emph{satisfy} this \rmrs\ because it is possible to map
-the labels and anchors into nodes in each tree and variable names like
+the labels and anchors in (\ref{ex:cat-pos}) into nodes in each tree
+and variable names like 
 $x_1$ and $x_3$ into variable names in the tree in such a way that the
-predications of the nodes that labels and anchors are mapped to are 
-consistent with those in the {\sc ep}s of the {\sc rmrs}.  For
-instance, the first tree satisfies the \rmrs\ because $l_1$ and $a_1$
-can map to the root of that tree and $x_1$ to $x$, and the root label
+predications of the nodes that labels and anchors denote are 
+consistent with those in the {\sc ep}s of (\ref{ex:cat-pos}).  For
+$l_1$ and $a_1$
+can map to the root of the first tree in Fig.~\ref{fig:1} and $x_1$ to
+$x$, and the root label 
 $\sem{\_every\_q\_1}$ is consistent with the {\sc ep} predicate
@@
 described equally well by the \rmrs\ (\ref{ex:cat-pos}); this is not
-surprising, given that the semantic information we got from the {\sc
+surprising, given that the semantic output from the {\sc
   pos} tagger is so incomplete.  If we have
@@
 variable in any fully specific logical form; e.g., both to the
-variable $x$ in Fig.~\ref{fig:1}.   $\ARG_i(a,z)$ and
-$\ARG_i(a,h)$ express that the $i$-th child (counting from 0) of the
+variable $x$ in Fig.~\ref{fig:1}.   $\ARG_i(a,z)$ (and
+$\ARG_i(a,h)$) express that the $i$-th child (counting from 0) of the
 node to which the anchor $a$ refers is the variable name that $z$
@@
 used here to underspecify the scope of the two quantifiers.
-
 Notice that the labels of the {\sc ep}s for ``fat'' and ``cat'' are
-stipulated to be equal in this \rmrs, whereas the anchors are not.  In
+stipulated to be equal in (\ref{ex:cat-erg}), whereas the anchors are not.  In
 the tree, it is the anchors that are mapped to the nodes with the
@@
 Finally, (\ref{ex:cat-erg}) uses predicate symbols like
-$\sempred{\_dog\_n\_1}$ that are meant to be more specific than the
+$\sempred{\_dog\_n\_1}$ that are meant to be more specific than 
 symbols like $\sempred{\_dog\_n}$ which the earlier \rmrs s used.
-This reflects the fact that the deep grammar will perform some lexical
-disambiguation that the chunker and {\sc pos} tagger didn't.  The fact
+This reflects the fact that the deep grammar performs some lexical
+disambiguation that the chunker and {\sc pos} tagger don't.  The fact
 that the former symbol should be more specific than the latter can be
@@
 
-In summary, \rmrs\ allows us to represent the (partial) semantic information
-that can be extracted from a wide range of {\sc nlp} tools in a uniform way.
-This is useful for hybrid systems which fall back to shallower
-analyses if deeper parsing is unsuccessful, or which try to match
-deeply parsed queries against shallow parses of large text corpora;
+In summary, \rmrs\ allows us to represent in a uniform way
+the (partial) semantics
+that can be extracted from a wide range of {\sc nlp} tools.
+This is useful for hybrid systems which exploit shallower
+analyses when deeper parsing fails, or which try to match
+deeply parsed queries against shallow parses of large corpora;
 and in fact, \rmrs\ is gaining popularity as a practical interchange
-format for exactly these purposes \cite{copestake:2003,frank:2004}.
+format for exactly these purposes \cite{copestake:2003}.
 However, \rmrs\ is still relatively ad-hoc in that its

### 2008-10-10 trunk/Papers/rmrs-08/rmrs.tex
titlebox back to normal

@@
 
-We define the syntax of \rmrs\ in the style of {\sc clls}
+We define \rmrs\ syntax in the style of {\sc clls}
 \cite{egg:etal:2001}.  We assume an infinite set of {\em node
   variables} $\nvar=\{X,Y,X_1,\ldots\}$, used as labels, anchors, and
-holes; the distinction between these uses will come from their
-position in the formula.  Furthermore, we assume an infinite set of
+holes; the distinction between these will come from their
+position in the formulas.  We also assume an infinite set of
 \emph{base variables} $\ovar$, consisting of individual variables
-$\{x,x_1,y,\ldots\}$ and event variables $\{e_1,\ldots\}$.  Finally,
-we assume a vocabulary of \emph{predicate symbols}
+$\{x,x_1,y,\ldots\}$ and event variables $\{e_1,\ldots\}$, and a
+vocabulary of \emph{predicate symbols}
 $\pred=\{P,Q,P_1,\ldots\}$.
-
-We can now define formulas of \rmrs\ as follows.
+\rmrs\ formulas are defined as follows.
 
@@
 A node variable $X$ is called a \emph{label} iff $\varphi$ contains an
-atom of the form $\ep{X}{Y}{P}$ or $Y \qeq X$; $X$ is called an
-\emph{anchor} in the \rmrs\ $\varphi$ iff $\varphi$ contains an atom
-of the form $\ep{Y}{X}{P}$ or $\ARG_S(X,i)$; and $X$ is called a
+atom of the form $\ep{X}{Y}{P}$ or $Y \qeq X$; it is an 
+\emph{anchor} iff $\varphi$ contains an atom
+of the form $\ep{Y}{X}{P}$ or $\ARG_S(X,i)$; and it is a
 \emph{hole} iff $\varphi$ contains an atom of the form $\ARG_S(Y,X)$
@@
 
-Intuitively, we expect that an \rmrs\ in solved form is fully
+Intuitively, an \rmrs\ in solved form is fully
 specified with respect to the predicate-argument structure, all
@@
 Because solved forms are so restricted, we can `read off' at least one
-model from each solved form.
-
+model from each solved form:
 \begin{prop} \label{prop:solved-forms-are-satisfiable}
@@
   holes is straightforward because no label is dominated by more than
-  one label; and then we can fill the spaces between the labels and
-  anchors with conjunctions.
+  one label; and  spaces between the labels and
+  anchors are filled with conjunctions.
 \end{proof}
@@
 Another feature of the existing \rmrs\ literature is that each term of
-an \rmrs\ is equipped with a \emph{sort}.  For instance, individual
-variables are typically assigned the sort $x$, event variables the
-sort $e$, and holes the sort $h$; these are arranged together with
-their subsorts in a sort hierarchy ${\cal S}$.  While we have defined
-\rmrs\ without sorts to simplify the presentation, it is
-straightforward to add them.  For this, one assumes that the signature
-$\Sigma$ is sorted, i.e.\ assigns a sort $s_1\times\ldots
-s_n\rightarrow s$ to each constructor, where $n$ is the constructor's
-arity (possibly zero) and $s,s_1,\ldots,s_n\in {\cal S}$ are atomic
-sorts.  We restrict the models of \rmrs\ to trees that are
-\emph{well-sorted} in the usual sense, i.e.\ those in which we can
-infer a sort for each subtree, and require that the variable
+an \rmrs\ is equipped with a \emph{sort}.  In particular, individual
+variables $x$, event variables $e$ and holes $h$ are arranged together
+with their subsorts (e.g., $e_{\mbox{past}}$) and supersorts (e.g.,
+sort $i$ abstracts over $x$ and $e$) into a
+sort hierarchy ${\cal S}$. 
+For simplicity we defined \rmrs\ without sorts, but it is
+straightforward to add them.  For this, one
+assumes that the signature $\Sigma$ is sorted, i.e.\ assigns a sort
+$s_1\times\ldots s_n\rightarrow s$ to each constructor, where $n$ is
+the constructor's arity (possibly zero) and $s,s_1,\ldots,s_n\in {\cal
+  S}$ are atomic sorts.  We restrict the models of \rmrs\ to trees
+that are \emph{well-sorted} in the usual sense, i.e.\ those in which
+we can infer a sort for each subtree, and require that the variable
 assignment functions likewise respect the sorts.  If we then modify

### 2008-10-10 trunk/Papers/rmrs-08/conclusion.tex
final?

@@
 
-In this paper, we have motivated and defined \rmrs---a semantic
-framework that has been used to represent, compare, and combine
-semantic information computed from deep and shallow parsers.  \rmrs\
-is designed to be maximally flexible on the type of semantic
-information that can be left underspecified, so that the semantic
-output of a shallow parser needn't over-determine or under-determine
-the semantics that can be extracted from the shallow syntactic
-analysis.  The key contribution of this paper was to lay the formal
-foundations for a formalism that is emerging as a standard in robust
-semantic processing.
+In this paper, we motivated and defined \rmrs---a semantic framework
+that has been used to represent, compare, and combine semantic
+information computed from deep and shallow parsers.  \rmrs\ is
+designed to be maximally flexible on the type of semantic information
+that can be left underspecified, so that the semantic output of a
+shallow parser needn't over-determine or under-determine the semantics
+that can be extracted from the shallow syntactic analysis.  Our key
+contribution was to lay the formal foundations for a formalism that is
+emerging as a standard in robust semantic processing.
 
@@
 for \rmrs\ is a crucial prerequisite for the future development of
-such tools.  We hope that future research will build upon this paper
-to develop efficient algorithms and implementations for solving \rmrs
-s, performing inferences that enrich \rmrs s from shallow analyses
-with deeper information, and checking consistency of \rmrs s that were
-obtained from different parsers.
+such tools; this strategy was highly successful for dominance
+constraints \cite{Althaus_etal:JoA}. We hope that future research will
+build upon this paper to develop efficient algorithms and
+implementations for solving \rmrs s, performing inferences that enrich
+\rmrs s from shallow analyses with deeper information, and checking
+consistency of \rmrs s that were obtained from different parsers.
 

### 2008-10-10 trunk/Papers/rmrs-08/entailment.tex
final?

@@
 
-The definitions from Section~\ref{sec:rmrs} serve an important
-theoretical purpose in that they provide a formal underpinning for \rmrs's
-existing use for partial semantic representations in implemented
+The above definitions serve an important theoretical purpose in that
+they provide a formal underpinning for the use of \rmrs\ in practical
 systems.  Next to the peace of mind that comes with the use of a
 well-understood formalism, we hope that the work reported here will
-serve as a starting point for future work on efficient solvers and
-inference mechanisms.
+serve as a starting point for future research.
 
@@
 deep and shallow processors is to determine whether the \rmrs s
-produced by the two systems are compatible.  In the new formal terms
-we can use now, we can characterise compatibility of a more detailed
-\rmrs\ $\varphi$ (perhaps from a deep grammar) and a less detailed
-\rmrs\ $\varphi'$ simply as \emph{entailment} $\varphi \models
-\varphi'$.  If entailment holds, this tells us that all claims that
-$\varphi'$ makes about the semantic content of a sentence is
-consistent with the claims that $\varphi$ makes.
+produced by the two systems are compatible.  In the new formal terms,
+we can characterise compatibility of a more detailed \rmrs\ $\varphi$
+(perhaps from a deep grammar) and a less detailed \rmrs\ $\varphi'$
+simply as \emph{entailment} $\varphi \models \varphi'$.  If entailment
+holds, this tells us that all claims that $\varphi'$ makes about the
+semantic content of a sentence are consistent with the claims that
+$\varphi$ makes.
 
@@
 entailment of \rmrs.  However, we propose the following novel
-syntactic characterisation which may again serve as a starting point
-for research along those lines. We call an \rmrs\ $\varphi'$ an
-\emph{extension} of the \rmrs\ $\varphi$ if $\varphi'$ contains all
-the EPs of $\varphi$ and $D(\varphi') \supseteq D(\varphi)$.
+syntactic characterisation as a starting point for research along
+those lines. We call an \rmrs\ $\varphi'$ an \emph{extension} of the
+\rmrs\ $\varphi$ if $\varphi'$ contains all the EPs of $\varphi$ and
+$D(\varphi') \supseteq D(\varphi)$.
 

### 2008-10-10 trunk/Papers/rmrs-08/rmrs.tex
final?

@@
 
-Def.\ref{defn:rmrs-syntax}
-combines similarities to earlier
+Def.~\ref{defn:rmrs-syntax} combines similarities to earlier
 presentations of \rmrs\ \cite{copestake:2003,copestake:2007b} and to
-{\sc clls}/dominance constraints \cite{egg:etal:2001}.  For the most part,
-our syntax generalises that of older versions of \rmrs: We use
+{\sc clls}/dominance constraints \cite{egg:etal:2001}.  For the most
+part, our syntax generalises that of older versions of \rmrs: We use
 $\ARG_{\{i\}}$ (with a singleton set $S$) instead of $\ARG_i$ and
-$\ARG_\N$ instead of $\ARG_n$, and the {\sc ep} $\ep{l}{a}{P(v)}$
-(as in Section~\ref{sec:motivation}) is an abbreviation of
+$\ARG_\N$ instead of $\ARG_n$, and the {\sc ep} $\ep{l}{a}{P(v)}$ (as
+in Section~\ref{sec:motivation}) is an abbreviation of
 $\{\ep{l}{a}{P}, \ARG_{\{0\}}(a,v)\}$.  Similarly, we don't assume
@@
 $\sem{\_some\_q\_1}$ (with the children being the bound variable, the
-restrictor, and the scope); and, constructors of various arities for
+restrictor, and the scope); and constructors of various arities for
 the predicate symbols; e.g., $\sem{\_chase\_v\_1}$ is of arity 3.  Other
@@
   holes is straightforward because no label is dominated by more than
-  one label; and  spaces between the labels and
-  anchors are filled with conjunctions.
+  one hole; and spaces between the labels and anchors are filled with
+  conjunctions.
 \end{proof}
@@
 (\ref{ex:cat-erg}), but becomes clear once we notice that no label is
-on the right-hand side of two dominance atoms.  However, this is
-because (at least) one model of (\ref{ex:cat-erg}) is
-shown in Fig.~\ref{fig:fat-black-cat}; this model
-is problematic in several ways and in particular contains an unbound
-variable $y$ even though it also contains a quantifier that binds $y$.
-If we restrict the class of models to those in 
-which such variables are bound (as \newcite{copestake:etal:2005} do),
-we can add the information that the 
-quantifiers outscope their bound variables to the \rmrs\ without
-changing its models further---i.e., we add the atoms $h_3 \dom
-l_5$ and $h_8 \dom l_5$.  Fig.~\ref{fig:fat-black-cat} is no longer a
-model for the extended \rmrs, which in turn is no longer in solved
-form because the label $l_5$ is on the right-hand side of two
-dominance edges.  Instead, it has the following two solved forms:
+on the right-hand side of two dominance atoms.  However, the model
+constructed in the proof of
+Prop.~\ref{prop:solved-forms-are-satisfiable} looks a bit like
+Fig.~\ref{fig:fat-black-cat}; both models are problematic in several
+ways and in particular contain an unbound variable $y$ even though
+they also contains a quantifier that binds $y$.  If we restrict the
+class of models to those in which such variables are bound (as
+\newcite{copestake:etal:2005} do), we can enforce that the quantifiers
+outscope their bound variables without changing models of the \rmrs\
+further---i.e., we add the atoms $h_3 \dom l_5$ and $h_8 \dom l_5$.
+Fig.~\ref{fig:fat-black-cat} is no longer a model for the extended
+\rmrs, which in turn is no longer in solved form because the label
+$l_5$ is on the right-hand side of two dominance atoms.  Instead, it
+has the following two solved forms:
 
@@
 of the language from Def.~\ref{defn:rmrs-syntax} (once qeq is added to
-it), or they restrict attention to only a subset of the models.
-Our formalism provides a
+it), or they restrict attention to only a subset of the models (e.g.,
+ones with correctly bound variables, or ones which don't contain extra
+material like Fig.~\ref{fig:fat-black-cat}).  Our formalism provides a
 general framework into which all these various fragments fit, and it's

### 2009-01-16 trunk/Papers/rmrs-08/motivation.tex
typos done

@@
 predications of the nodes that labels and anchors denote are 
-consistent with those in the {\sc ep}s of (\ref{ex:cat-pos}).  For
+consistent with those in the {\sc ep}s of (\ref{ex:cat-pos})---e.g.,
 $l_1$ and $a_1$
-can map to the root of the first tree in Fig.~\ref{fig:1} and $x_1$ to
+can map to the root of the first tree in Fig.~\ref{fig:1}, $x_1$ to
 $x$, and the root label 
@@
 even specify the arguments to a predicate while underspecifying their
-position.  This can useful for specifying arguments to \_give\_v when a
+position.  This is useful for specifying arguments to \_give\_v when a
 parser doesn't handle unbounded dependencies and is faced with

### 2009-01-16 trunk/Papers/rmrs-08/rmrs.tex
typos done

@@
 they receive their function from their positions in the formula.  One
-major difference is that we use dominance rather than qeq; see
+major difference is that we use dominance ($\dom$) rather than qeq; see
 Section~\ref{sec:extensions} for a discussion.  Compared to dominance
@@
   \item for every atom $\ARG_S(X,i)$ in $\varphi$, there is exactly
-    one atom $\ARG_{S'}(X',i')$ in $\varphi$ with $X' = s(X)$, $i' =
+    one atom $\ARG_{S'}(X',i')$ in $\varphi'$ with $X' = s(X)$, $i' =
     s(i)$, and $S' \subseteq S$;
@@
 qeq to the syntax of the language and to extend the model theory
-accordingly.  This extension adds a new atom $X=_q Y$ to
+accordingly.  This extension adds a new atom $X \dom Y$ to
 Def.~\ref{defn:rmrs-syntax}, and $\tau,\alpha,g,\sigma$ will satisfy
-$X=_q Y$ iff $\alpha(X)$ and $\alpha(Y)$ are separated by zero or more
+$X \dom Y$ iff $\alpha(X)$ and $\alpha(Y)$ are separated by zero or more
 nodes in $\tau$, each of which has a rightmost child that is a

### 2009-01-20 trunk/Papers/rmrs-08/rmrs.tex
typos; fixed figure; acknowledgments

@@
 A node variable $X$ is called a \emph{label} iff $\varphi$ contains an
-atom of the form $\ep{X}{Y}{P}$ or $Y \qeq X$; it is an 
+atom of the form $\ep{X}{Y}{P}$ or $Y \dom X$; it is an 
 \emph{anchor} iff $\varphi$ contains an atom
@@
 \emph{hole} iff $\varphi$ contains an atom of the form $\ARG_S(Y,X)$
-or $X \qeq Y$.
+or $X \dom Y$.
 \end{definition}

### 2009-01-21 trunk/Papers/dl-gre-jair/discussion.tex
jair article: only reformatted inlg article

--- 2008-05-16 trunk/Papers/dl-gre-08/discussion.tex
@@
 \begin{center}
-\includegraphics[width=0.2\textwidth]{drawers.png}
+\todo{need to reformat pic}
+%%%\includegraphics[width=0.2\textwidth]{drawers.png}
 \end{center}\vspace*{-3ex}
@@
 
-\newcite{viethen06:_algor_for_gener_refer_expres} describe the data in
+\citeA{viethen06:_algor_for_gener_refer_expres} describe the data in
 more detail and present results of evaluating the Full Brevity
-algorithm, the Incremental Algorithm (both by \newcite{Dale1995}), and
+algorithm, the Incremental Algorithm (both by \citeA{Dale1995}), and
 the Relational Algorithm \cite{dale91:_gener_refer_expres_invol_relat}
@@
 redundant information that our algorithm cannot reproduce them:
-Similarly to the Incremental Algorithm, it allows for some redundancy,
-but stops once it has found a distinguishing description. It does,
-however, generate other, simpler descriptions for these three
-referents. 
+Similarly to the Incremental Algorithm, our algorithm allows for some
+redundancy, but stops once it has found a distinguishing
+description. It does, however, generate other, simpler descriptions
+for these referents.
+
+%these three referents.
 %human-produced descriptions that cannot be reproduced exactly.
-
 %two mention column information, which we excluded from our domain, and
@@
 $\neg \exists R.(\neg P \sqcap \neg Q)$ as a smooth noun phrase.
-Although we agree with \newcite{deemter02:_gener_refer_expres} and
+Although we agree with \citeA{deemter02:_gener_refer_expres} and
 others that the careful use of negation and disjunction can improve
@@
 with every other approach that separates these two modules, i.e.,
-almost the entire GRE literature (notable exceptions are, e.g.,~\newcite{Horacek1997} and SPUD \cite{Stone1998a}).  In principle, we
+almost the entire GRE literature (notable exceptions are, e.g.,~\citeA{Horacek1997} and SPUD \cite{Stone1998a}).  In principle, we
 believe that it is a good idea to handle sentence planning and
@@
 eventually be combined with recent efficient algorithms for integrated
-sentence planning and realization, such as in \newcite{KolSto07}.
+sentence planning and realization, such as in \citeA{KolSto07}.
 
@@
   \centering
-  \includegraphics[width=\columnwidth]{runtimes}\vspace*{-2ex}
+	\todo{need to reformat pic}
+  %%%\includegraphics[width=\columnwidth]{runtimes}\vspace*{-2ex}
   \caption{Average runtimes (in ms) of the two algorithms on random

### 2009-01-21 trunk/Papers/dl-gre-jair/gre.tex
jair article: only reformatted inlg article

--- 2008-05-16 trunk/Papers/dl-gre-08/gre.tex
@@
 building DL formulas that are conjunctions of atoms; relational REs as
-in \newcite{dale91:_gener_refer_expres_invol_relat} are formulas of
+in \citeA{dale91:_gener_refer_expres_invol_relat} are formulas of
 \el; and so on.  We will further pursue the idea of organizing GRE
@@
   \centering
-  \includegraphics[width=\columnwidth]{pic-dale-haddock}
-  \caption{(a) The \newcite{dale91:_gener_refer_expres_invol_relat}
-    scenario; (b) the \newcite{Stone1998a} scenario.}
+	\todo{need to reformat pic}
+  %%%%%%%%%%\includegraphics[width=\columnwidth]{pic-dale-haddock}
+  \caption{(a) The \citeA{dale91:_gener_refer_expres_invol_relat}
+    scenario; (b) the \citeA{Stone1998a} scenario.}
   \label{fig:dale-haddock}\vspace*{-1ex}
@@
 Simulation classes have been studied extensively in the literature
-(see e.g.,~\newcite{blac:moda01}; \newcite{kurt:expr99}), and there are several
+(see e.g.,~\citeA{blac:moda01}; \citeA{kurt:expr99}), and there are several
 efficient algorithms for computing \alc-simulation classes
@@
 However, these algorithms will only compute the simulation classes
-themselves. Here we extend the \newcite{hopc:algo71} algorithm such
+themselves. Here we extend the \citeA{hopc:algo71} algorithm such
 that it computes, along with each set, also a formula that denotes
@@
 algorithm on the model shown in Fig.~\ref{fig:dale-haddock}a, which is
-taken from \newcite{dale91:_gener_refer_expres_invol_relat}.  The
+taken from \citeA{dale91:_gener_refer_expres_invol_relat}.  The
 algorithm starts with $\RE = \{\top\}$.  In the first loop, it adds

### 2009-01-21 trunk/Papers/dl-gre-jair/main.tex
jair article: only reformatted inlg article

\documentclass[11pt]{article}

\usepackage{jair, theapa}

\usepackage{times}
\usepackage{url}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{multirow}

\input epsf %% at top of file

\usepackage[algoruled, linesnumbered,noend]{algorithm2e}

\newcommand{\gM}{\ensuremath{\mathcal{M}}}
\newcommand{\gL}{\ensuremath{\mathcal{L}}}

\newcommand{\C}{\ensuremath{\mathcal{C}}\xspace}
\newcommand{\M}{\ensuremath{\mathcal{M}}\xspace}
\newcommand{\RE}{\ensuremath{\mathit{RE}}\xspace}

\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]

\newcommand{\todo}[1]{\textbf{(#1)}}
\newcommand{\ignore}[1]{}

\newcommand{\el}{\ensuremath{\mathcal{EL}}\xspace}
\newcommand{\alc}{\ensuremath{\mathcal{ALC}}\xspace}
\newcommand{\form}{\mathsf{form}\xspace}
\newcommand{\prop}{\ensuremath{\mathsf{prop}}\xspace}
\newcommand{\simm}{\textsf{sim}}
\newcommand{\rel}{\ensuremath{\mathsf{rel}}\xspace}
\newcommand{\propm}{\ensuremath{\mathsf{prop}^\gM\xspace}}

\newcommand{\interp}[1]{|\!|#1|\!|}

\jairheading{0}{0000}{0-00}{0/0}{0/0}
\ShortHeadings{Referring Expressions as Formulas of Description Logic}
{Areces, Koller \& Striegnitz}
\firstpageno{1}

\begin{document}

\title{Referring Expressions as Formulas of Description Logic}

\author{\name Carlos Areces \email areces@loria.fr \\
       \addr INRIA Nancy Grand Est\\
       Nancy, France
       \AND
       \name Alexander Koller \email koller@coli.uni-sb.de \\
       \addr Saarland University\\
       Saarbr\"{u}cken, Germany
       \AND
       \name Kristina Striegnitz \email striegnk@union.edu \\
       \addr Union College\\
       Schenectady, NY, USA}

% For research notes, remove the comment character in the line below.
% \researchnote

\maketitle

\begin{abstract}
In this paper, we propose to reinterpret the problem of generating
referring expressions (GRE) as the problem of computing a formula in a
description logic that is only satisfied by the referent. This view
offers a new unifying perspective under which existing GRE algorithms
can be compared. We also show that by applying existing algorithms for
computing simulation classes in description logic, we can obtain
extremely efficient algorithms for relational referring expressions
without any danger of running into infinite regress.
\end{abstract}

\input introduction
\input bisim
\input gre
\input discussion
\input related
\input conclusion

\acks{We are grateful to Hector Geffner (who
independently suggested to view GRE as computation of DL formulas),
Kees van Deemter, and Emiel Krahmer for interesting discussions.  We
also thank Jette Viethen and Robert Dale for making their corpus
available, and the reviewers for their comments.
}

\appendix
\section*{In case we have an appendix.}

[section ommitted]

\vskip 0.2in
\bibliography{bibliography}
\bibliographystyle{theapa}

\end{document}

### 2009-01-28 trunk/Papers/rmrs-08/rmrs.tex
in my view, beta

@@
 restrictor, and the scope); and constructors of various arities for
-the predicate symbols; e.g., $\sem{\_chase\_v\_1}$ is of arity 3.  Other
-base languages may require a different signature $\Sigma$ and/or a
-different mapping between formulas and trees; the only strict
+the predicate symbols; e.g., $\sem{\_chase\_v\_1}$ is of arity 3.
+Other base languages may require a different signature $\Sigma$ and/or
+a different mapping between formulas and trees; the only strict
 requirement we make is that the signature contains a binary
 constructor $\wedge$ to represent conjunction.  We write $\Sigma_i$
-for the set of all constructors in $\Sigma$ with arity $i$.  We
-will follow the typographical convention that non-logical symbols in
-$\Sigma$ are written in $\mathsf{sans{-}serif}$, as opposed to the
-\rmrs\ predicate symbols like $\sempred{\_cat\_n}$ and
-$\sempred{\_cat\_n\_1}$.
+and $\Sigma_{\geq i}$ for the set of all constructors in $\Sigma$ with
+arity $i$ and at least $i$, respectively.  We will follow the
+typographical convention that non-logical symbols in $\Sigma$ are
+written in \textsf{sans-serif}, as opposed to the \rmrs\ predicate
+symbols like $\sempred{\_cat\_n}$ and $\sempred{\_cat\_n\_1}$.
 
 The models of {\sc rmrs} are then defined to be finite constructor
-trees:
+trees (see also \cite{egg:etal:2001}):
 \begin{definition}\label{defn:models}
@@
 \begin{definition}\label{defn:dominance}
-  $u \dom v$ iff $u$ is a prefix of $v$, i.e.\ the node $u$ is equal
-  to or above
-  the node $v$ in the tree.  (This is the dominance relation familiar
-  from \newcite{egg:etal:2001}.)
-
+  $u \dom v$ \emph{(dominance)} iff $u$ is a prefix of $v$, i.e.\ the node
+  $u$ is equal to or above the node $v$ in the tree.
   $u \wedgedom v$ iff $u \dom v$, and all symbols on the path from $u$
@@
 In the context of robust semantic processing, this is a desirable
-feature of \rmrs, because it means that when we enrich an \rmrs\
-obtained from a shallow processor with more semantic
-information---such as the relation symbols introduced by
-syntactic 
-constructions such as appositives, noun-noun compounds and free
-adjuncts---we don't change the set of models;
-we only \emph{restrict} the set of models further and further towards
-the semantic representation we are trying to reconstruct.
-Furthermore, it has been shown in the literature that a
+feature, because it means that when we enrich an \rmrs\ obtained from
+a shallow processor with more semantic information---such as the
+relation symbols introduced by syntactic constructions such as
+appositives, noun-noun compounds and free adjuncts---we don't change
+the set of models; we only \emph{restrict} the set of models further
+and further towards the semantic representation we are trying to
+reconstruct.  Furthermore, it has been shown in the literature that a
 dominance-constraint style semantics for underspecified
@@
 
-\begin{definition}  \label{defn:solved-form-of}
-  The \emph{syntactic dominance relation} in an {\sc  rmrs} $\varphi$ is the
+\begin{definition} \label{defn:solved-form-of} The \emph{syntactic
+    dominance relation} $D(\varphi)$ in an {\sc rmrs} $\varphi$ is the
   reflexive, transitive closure of the binary relation $$
@@
 $$
-We write $D(\varphi)$ for the syntactic dominance
-  relation of $\varphi$.
 
@@
 Let's look at an example for all these definitions.  All the \rmrs s
-presented in Section~\ref{sec:motivation} (replacing $\qeq$ by
-$\dom$), are in solved form; this is least obvious for
-(\ref{ex:cat-erg}), but becomes clear once we notice that no label is
-on the right-hand side of two dominance atoms.  However, the model
-constructed in the proof of
-Prop.~\ref{prop:solved-forms-are-satisfiable} looks a bit like
+presented in Section~\ref{sec:motivation} (replacing $\qeq$ by $\dom$)
+are in solved form; this is least obvious for (\ref{ex:cat-erg}), but
+becomes clear once we notice that no label is on the right-hand side
+of two dominance atoms.  However, the model constructed in the proof
+of Prop.~\ref{prop:solved-forms-are-satisfiable} looks a bit like
 Fig.~\ref{fig:fat-black-cat}; both models are problematic in several
@@
 qeq to the syntax of the language and to extend the model theory
-accordingly.  This extension adds a new atom $X \dom Y$ to
+accordingly.  This extension adds a new atom $X \qeq Y$ to
 Def.~\ref{defn:rmrs-syntax}, and $\tau,\alpha,g,\sigma$ will satisfy
-$X \dom Y$ iff $\alpha(X)$ and $\alpha(Y)$ are separated by zero or more
-nodes in $\tau$, each of which has a rightmost child that is a
-quantifier. All the above propositions about solved forms still hold
-if ``dominance'' is replaced with ``qeq''.
+$X \qeq Y$ iff $\alpha(X) \dom \alpha(Y)$, each node on the path is a
+quantifier, and each step in the path goes to the rightmost child. All
+the above propositions about solved forms still hold if ``dominance''
+is replaced with ``qeq''.
 

### 2009-01-28 trunk/Papers/rmrs-08/entailment.tex
EP fixed
final version?

@@
 
-The above definitions serve an important theoretical purpose in that
-they provide a formal underpinning for the use of \rmrs\ in practical
+The above definitions serve an important theoretical purpose: they
+formally underpin the use of \rmrs\ in practical
 systems.  Next to the peace of mind that comes with the use of a
@@
 
-One direction which we believe may develop from this paper is the
+One direction to pursue from this paper is the
 development of efficient solvers for \rmrs.  As a first step, it would
@@
 those lines. We call an \rmrs\ $\varphi'$ an \emph{extension} of the
-\rmrs\ $\varphi$ if $\varphi'$ contains all the EPs of $\varphi$ and
+\rmrs\ $\varphi$ if $\varphi'$ contains all the {\sc ep}s of $\varphi$ and
 $D(\varphi') \supseteq D(\varphi)$.

### 2009-01-28 trunk/Papers/rmrs-08/rmrs.tex
final version?

@@
 representing variables in the base language. Finally, a
-relation $\sigma \subseteq \pred \times \Sigma_{\geq 1}$ maps
+function $\sigma$ from $\pred$ to the power set of $\Sigma_{\geq 1}$ maps
 each \rmrs\ predicate symbol to a set of constructors from $\Sigma$.
@@
 relation from \newcite{copestake:etal:2005}.  This is partly because
-dominance is a weaker relation than qeq: If a dependency parser links
+dominance is the weaker relation: If a dependency parser links
 a determiner to a noun and this noun to a verb, then we can use
@@
 the verb is outscoped by the quantifier introduced by the determiner
-(see earlier discussion).  However, it is very straightforward to add
-qeq to the syntax of the language and to extend the model theory
-accordingly.  This extension adds a new atom $X \qeq Y$ to
+(see earlier discussion).  However, it is very straightforward to
+extend the syntax and semantics
+of the language to include the qeq relation.
+This extension adds a new atom $X \qeq Y$ to
 Def.~\ref{defn:rmrs-syntax}, and $\tau,\alpha,g,\sigma$ will satisfy

### 2009-04-27 trunk/Papers/ci-spark/manuscript.tex
added Ron's CI submission

%%%
%%% Computational Intelligence manuscript based on SPARK'08 paper
%%%
%%% Alexander Koller and Ronald Petrick
%%% 
%%% Updated: 2008/12/22
%%%
\documentclass[letterpaper]{article}
\usepackage{times}
\usepackage{graphicx}
\usepackage{url}
\usepackage[round]{natbib}
\usepackage{setspace}
\usepackage[paper=letterpaper,margin=1.0in]{geometry}

\title{Experiences with Planning for Natural Language Generation}
\author{\textbf{Alexander Koller} \\
Exzellenzcluster / FR 4.7 Computerlinguistik \\
Saarland University \\
Postfach 151150 \\
66041 Saarbrücken \\
Germany \\
Phone: +49 681 302 70040 \\
Fax: +49 681 302 4351 \\
E-mail: \url{koller@mmci.uni-saarland.de}
\and
\textbf{Ronald P. A. Petrick} \\
School of Informatics \\
University of Edinburgh \\
Informatics Forum, 10 Crichton Street \\
Edinburgh \ EH8 9AB \\
Scotland, United Kingdom \\
Phone: +44 131 650 4426 \\
Fax: +44 131 650 6899 \\
E-mail: \url{rpetrick@inf.ed.ac.uk}}

%\date{22 December 2008}
\date{}

\pagestyle{plain}
\pagenumbering{arabic}
\doublespacing

\bibpunct{(}{)}{;}{a}{}{,}

\begin{document}

\maketitle

%%%
%%% Abstract
%%%
\pagebreak
\begin{abstract}
We investigate the application of modern planning techniques to domains
arising from problems in natural language generation (NLG). In particular,
we consider two novel NLG-inspired planning problems, the sentence
generation domain and the GIVE (``Generating Instructions in Virtual
Environment'') domain, and investigate the efficiency of the FF and SGPLAN
planners in these domains. We also compare our results against an ad-hoc
implementation of GraphPlan in Java. Our results are mixed. While modern
planners are able to quickly solve many moderately-sized instances of our
problems, the overall planning time is dominated by the grounding step that
these planners perform, rather than search. This has a pronounced effect on
our domains which require relatively short plans but have large universes.
We share our experiences and offer these domains as challenges for the
planning community.
\end{abstract}

\bigskip\noindent
\textbf{Keywords:} natural language generation, planning

%%%
%%% Main text
%%%
\pagebreak
\section{Introduction}
\label{sec:introduction}

Natural language generation (NLG; \citealp{reiter00building}) is one of the
major subfields of natural language processing, concerned with computing
natural language sentences or texts that convey a given piece of
information to the user. Intuitively, this task can be viewed as a problem
involving actions, beliefs, and goals: an agent communicating with another
agent tries to change the mental state of the hearer by applying actions
which correspond to the utterance of words or sentences.
This characterization of NLG suggests obvious parallels to automated
planning---a view which has a long tradition in NLG.

Early approaches to generation as planning
\citep{perrault80,appelt:planning,hovy88,young94dpocl} focused primarily on
planning high-level structures, such as speech acts and discourse
relations.  While these approaches suffered from the inefficiency of the
planners that were available at the time, there has recently been a
resurgence of interest in applying modern planning techniques to NLG.
These newer papers revisit speech act planning based on the more expressive
planning formalisms available today \citep{Steedman-Petrick:07,benotti08b}
and exploit the improved efficiency of modern planners for an integrated
model of sentence planning and surface realization \citep{KolSto07}.
Problems in NLG are often computationally challenging (see, e.g.,
\citep{KolStr02}), and these new approaches combine an interest in planning
as a modelling tool for natural language, with a hope of improved
efficiency by exploiting modern approaches and algorithms.

The focus of this paper is twofold. First, we present two novel planning
domains that arise in the context of natural language generation: the
sentence generation domain and the GIVE domain. In the sentence generation
task, the goal is to generate a single sentence that expresses a given
meaning. \citet{KolSto07} cast this domain as a planning problem, where the
plan encodes the necessary sentence and the actions correspond to uttering
individual words.  In the GIVE domain (``Generating Instructions in Virtual
Environments''), we describe a new shared task that was recently posed as a
challenge for the NLG community \citep{alexander07:_shared_task_propos}.
GIVE uses planning as part of an NLG system that generates natural-language
instructions to guide a user in performing a given task in a virtual
environment.

Second, we discuss some of our experiences using off-the-shelf
planners in our two NLG-inspired domains. In particular, we explore the
efficiency of FF \citep{HoffmannNebel01} and SGPLAN
\citep{hsu06:_new_featur_in_sgplan_for}: two planners that are readily
available, support an expressive subset of the Planning Domain Definition
Language (PDDL; McDermott et al.~\citeyear{PDDL}) for encoding domains, and
have been successful on traditional benchmarks and problems from the
International Planning Competition (IPC)\footnote{See
 \url{http://ipc.icaps-conference.org/} for information about past
 editions of the IPC.}.
We test these planners on a range of problem instances in our planning
domains, and compare our results against an ad-hoc Java implementation of
GraphPlan \citep{Blum1997}. Overall, our findings are mixed. On the one
hand, we demonstrate that modern planners can easily handle some of the
moderately-sized search problems that arise in NLG. In certain cases, the
planners we tested are more efficient than specialized NLG algorithms
applied to the same problem. On the other hand, these same planners spend
tremendous amounts of time on preprocessing. For instance, in the sentence
generation domain, FF spends 90\% of its runtime grounding out literals and
actions, and ends up being slower than our version of GraphPlan that avoids
grounding. For domains like ours, which are dominated by the number of
actions and the universe size, rather than the combinatorics of the search
problem, this observation suggests that the overall runtime of a planner
could be improved by grounding more selectively. This is particularly true
for large problem instances, which still remain a challenge for the
planners we tested, and where the effects of the grounding approach become
quite pronounced. We therefore offer our NLG domains as challenges for the
planning community.

\section{Planning in NLG}
\label{sec:domains}

We begin by presenting our two planning domains: the sentence generation
domain and the instruction giving domain.

\subsection{Sentence generation as planning}

Sentence generation is the problem of computing from a grammar and a
semantic representation, a single sentence that expresses this piece of
meaning. Traditionally, this problem is split into several steps
\citep{reiter00building}. In a first step, called \emph{sentence planning},
the semantic representation is enriched with additional information. For
instance, \emph{referring expressions}, which refer to individuals (e.g.,
objects, people, etc.) in the domain of discourse, are determined at this
point. In a second step, called \emph{surface realization}, this enriched
representation is translated into a natural-language sentence, using the
grammar.

In practice, the process of determining referring expressions typically
interacts with the realization step and so it often turns out to be
beneficial to perform these two steps together. For instance, the SPUD
system (``Sentence Planning Using Description''; \citealt{Stone2003a})
performed this task using a top-down generation algorithm based on
tree-adjoining grammars \citep{joshi;etal1997} whose lexical entries were
equipped with semantic and pragmatic information. Unfortunately, SPUD had
to resort to a non-optimal greedy search strategy to cope with its large
search space and retain reasonable efficiency. To improve the efficiency
of SPUD, \citet{KolSto07} translate the sentence generation problem into a
planning problem and use a planning algorithm to perform the generation
task.

\begin{figure}[t]
  \centering
  \includegraphics[width=0.75\columnwidth]{pic-grammar}
  \caption{An example grammar in the sentence generation domain.}
  \label{fig:white-rabbit-sleeps-grammar}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.75\columnwidth]{pic-derivation-2}
  \caption{Derivation of ``The white rabbit sleeps.''}
  \label{fig:white-rabbit-sleeps-deriv}
\end{figure}

We illustrate this process on a simplified example.  Consider a
knowledge base containing the individuals $e$, $r_1$ and $r_2$, and a
set of attributes encoding the fact that $r_1$ and $r_2$ are rabbits,
$r_1$ is white and $r_2$ is brown, and $e$ is an event in which $r_1$
sleeps.  Say that we want to express the information
$\{\mathsf{sleep}(e,r_1)\}$ using the tree-adjoining grammar shown in
Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
\emph{elementary trees} (i.e., the disjoint trees in the figure), each
of which contributes certain \emph{semantic content}. We can
instantiate these trees by substituting individuals for \emph{semantic
  roles}, such as $\mathsf{self}$ and $\mathsf{subj}$, and then
combine the tree instances as shown in
Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
``The white rabbit sleeps''.

We compute this grammatical derivation in a top-down manner, starting with
the elementary tree for ``sleeps''. This tree satisfies the need to convey
the semantic information, but introduces a need to generate a noun phrase
(NP) for the subject; this NP must refer uniquely to the target referent
$r_1$. In a second step, we substitute the tree for ``the rabbit'' into the
open NP leaf, which makes the derivation grammatically complete. Since
there are two different individuals that could be described as ``the
rabbit''---technically, $r_2$ is still a \emph{distractor} (i.e., based on
the description ``the rabbit'', the hearer might erroneously think that
we're talking about $r_2$ and not $r_1$)---we are still not finished. To
complete the derivation, the tree for ``white'' is added to the existing
structure by an \emph{adjunction} operation, making the derivation
both syntactically and semantically complete.

\begin{figure}[p]
\centering
\begin{minipage}{0.5\textwidth}
{\small%
\begin{verbatim}
(:action add-sleeps
   :parameters (?u - node
                ?xself - individual
                ?xsubj - individual)
   :precondition
       (and (subst S ?u)
            (referent ?u ?xself)
            (sleep ?xself ?xsubj))
   :effect 
       (and (not (subst S ?u))
            (expressed sleep ?xself ?xsubj)
            (subst NP (subj ?u))
            (referent (subj ?u) ?xsubj)
            (forall (?y - individual)
                (when (not (= ?y ?xself))
                    (distractor (subj ?u) ?y)))))

(:action add-rabbit
   :parameters (?u - node
                ?xself - individual)
   :precondition 
       (and (subst NP ?u)
            (referent ?u ?xself)
            (rabbit ?xself))
   :effect 
       (and (not (subst NP ?u))
            (canadjoin N ?u)
            (forall (?y - individual)
                (when (not (rabbit ?y))
                    (not (distractor ?u ?y))))))

(:action add-white
   :parameters (?u - node
                ?xself - individual)
   :precondition 
       (and (canadjoin N ?u)
            (referent ?u ?xself)
            (rabbit ?xself))
   :effect 
       (forall (?y - individual)
           (when (not (white ?y))
               (not (distractor ?u ?y)))))
\end{verbatim}}%
\end{minipage}
\caption{PDDL actions for generating the sentence ``The white rabbit
sleeps.''}
\label{fig:white-rabbit-as-planning}
\end{figure}

The process described above has clear parallels to planning: we manipulate
states by applying sequences of actions to achieve a goal. We can make this
connection even more precise by translating the SPUD problem into a
planning problem.\footnote{See
  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
  which this conversion is implemented.}  For instance,
Figure~\ref{fig:white-rabbit-as-planning} shows the corresponding PDDL
actions for the above generation task, where each action corresponds
to an operation that adds a single elementary tree to the derivation.
In each case, the first parameter of the action is a node name in the
derivation tree, and the remaining parameters stand for the
individuals to which the semantic roles will be instantiated.  The
syntactic preconditions and effects are encoded using $\mathsf{open}$
and $\mathsf{canadjoin}$ literals; the status of each referring expression
is tracked using $\mathsf{distractor}$ literals. The resulting action
effects contain terms of the form $\mathsf{subj}(u)$, which construct new
node names, but aren't strictly permitted by PDDL. In order to meet the
syntactic requirements of PDDL, these terms can be eliminated by estimating
an upper bound $n$ for the plan length, making $n$ copies of each action,
ensuring that copy $i$ can only be applied in step $i$, and replacing the
term $\mathsf{subj}(u)$ in an action copy by the constant
$\mathsf{subj}_i$.

Once the appropriate actions are defined, we can solve the generation
problem as an ordinary planning problem. For instance, the following plan
solves our previous example:
%
\begin{enumerate}
\item $\mathsf{add}\textsf{-}\mathsf{sleeps}(\mathsf{root}, r_1)$,
\item $\mathsf{add}\textsf{-}\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}),r_1)$,
\item $\mathsf{add}\textsf{-}\mathsf{white}(\mathsf{subj}(\mathsf{root}),r_1)$.
\end{enumerate}
%
Using this plan, the grammatical derivation in
Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
generated sentence ``the white rabbit sleeps'', can be systematically
reconstructed. Thus, we can solve the sentence generation problem via
the detour through planning and bring current search heuristics for
planning to bear on generation.

\subsection{Planning in instruction giving}

The object of the GIVE Challenge (``Generating Instructions in Virtual
Environments''; \citealt{alexander07:_shared_task_propos}) is to build an
NLG system which is able to produce natural-language instructions which
will guide a human user in performing some task in a virtual environment.
From an NLG perspective, GIVE makes for an interesting challenge because it
is a theory-neutral task that exercises all components of an NLG system,
and emphasizes the study of communication in a (simulated) physical
environment. It also has the advantage that the user and the NLG system can
be in physically different places, as long as the 3D client and the NLG
system are connected over a network. This makes it possible to evaluate
GIVE NLG systems on a large scale over the Internet.
%
The first GIVE evaluation is currently underway. Five NLG systems from
three countries are in the process of being evaluated. At the time of
writing, 1028 different evaluation runs---each consisting of one user
interacting with one instance of the virtual environment and being guided
by an NLG system---have been collected, making GIVE the largest-scale
NLG evaluation effort ever.\footnote{See the GIVE website at
 \url{http://www.give-challenge.org/} for more details about this
 challenge.}

\begin{figure}[t]
\centering
%\includegraphics[width=1 \columnwidth]{give_world_no_expl}
\includegraphics[width=0.75\columnwidth]{give_world_2}
\caption{Map of the GIVE development world.}
  \label{fig:give-development-world}
\end{figure}

An example GIVE world map is shown in
Figure~\ref{fig:give-development-world}. To simplify both the planning and
the NLG task, the world is discretised into a set of tiles of equal size.
The user can turn by 90 degree steps in either direction, and can move from
the centre of one tile to the centre of the next tile, provided the path
between two tiles is not blocked. The world also consists of a set of
objects which can be manipulated by the user in various ways. For
instance, in the example world the user's task is to pick up a trophy in
the top left room. The trophy is hidden in a safe behind a picture, so the
user must first move the picture out of the way and open the safe by
pushing a sequence of buttons (the small square boxes on the walls) in the
correct order. In order to get to all these buttons, the user must also
open the door in the centre of the map and deactivate the alarm tile by
pushing further buttons. Figure~\ref{fig:give-planning} shows the encoding
of some of the available GIVE domain actions, in PDDL syntax.

\begin{figure}[p]
\centering
\begin{minipage}{0.5\textwidth}
{\small%
\begin{verbatim}
(:action move
   :parameters (?from - position
                ?to - position
                ?ori - orientation)
   :precondition 
       (and (player-pos ?from) 
            (adjacent ?from ?to ?ori) 
            (player-orient ?ori)
            (not-blocked ?from ?to)
            (not-alarmed ?to))
   :effect 
       (and (not (player-pos ?from))
            (player-pos ?to)))

(:action turn-left
   :parameters (?ori - orientation
                ?newOri - orientation)
   :precondition 
       (and (player-orient ?ori)
            (next-orient-left ?ori ?newOri))
   :effect 
       (and (not (player-orient ?ori))
            (player-orient ?newOri)))

(:action turn-right
   :parameters (?ori - orientation
                ?newOri - orientation)
   :precondition 
       (and (player-orient ?ori)
            (next-orient-right ?ori ?newOri))
   :effect 
       (and (not (player-orient ?ori))
            (player-orient ?newOri)))

(:action manipulate-b1-off-on
   :parameters (?pos - position)
   :precondition 
       (and (state b1 off)
            (player-pos ?pos)
            (position b1 ?pos))
   :effect 
       (and (not (state b1 off))
            (state b1 on)
            (not (state d1 closed))
            (state d1 open) 
            (not (blocked pos_6_5 pos_6_4))
            (not (blocked pos_6_4 pos_6_5))))
\end{verbatim}}%
\end{minipage}
\caption{PDDL actions for the GIVE domain.}
\label{fig:give-planning}
\end{figure}

The plans generated in the GIVE domain are often nontrivial. For instance,
in the above example the shortest plan consists of 108 action steps, with
the first few steps as follows:
%
\begin{enumerate}
\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north},
\mathsf{west})$,
\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$,
\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
\end{enumerate}

Most of the actions in the above plan are ``move'' actions. This makes the
bulk of the GIVE problem very similar to that of a Gridworld problem (see,
e.g., \citep{Tovey-Koenig:2000}), which also involves finding a route
through a two-dimensional world map with discrete positions. GIVE tasks are
usually much more complex, however, as illustrated by the above example
map: a successful plan must include steps for pressing buttons in the right
order, reasoning about large numbers of world objects, and navigating
through complicated room shapes.

The GIVE task comes with a number of interesting research challenges for
both the NLG and planning communities. On the NLG side, one major problem
is that verbalising each action instance in a plan makes for very
poor natural-language instructions. For instance, the NLG system could
generate an instruction sequence starting with ``turn left; walk forward;
press the button; turn right; walk forward; walk forward; turn right; walk
forward; walk forward; turn left; walk forward''. However, directions of
this form are linguistically clumsy and uninteresting, and it is to be
expected that frustrated users will quickly cancel an interaction with such
a system, resulting in a bad evaluation score. Having just pressed the
button, it would be much better to say ``turn around and walk through the
door''. That is, the system should \emph{summarise} plans by merging
multiple action instances into single instructions before presenting them
to the user.\footnote{This task is similar to the construction of
  ``macro'' actions, but is primarily intended for plan presentation
  purposes. See, e.g., \citep{Botea-etal:05} which contains a survey of
  recent approaches for generating macro actions in planning.}
Conversely, it may also be necessary to \emph{elaborate} on a single
planning step by expressing it with several instructions. Having just
entered the top left room, it may be easier for the user to understand the
instruction ``walk to the centre of the room; turn right; now press the
green button in front of you'', rather than the instruction ``press the
green button on the wall to your right''. Furthermore, in order to plan the
referring expression ``the green button in front of you'' at a time when
the button is not yet in front of the user, the NLG system must keep track
of the hypothetical changes in what objects are visible to the user. Thus,
the NLG acts of referring and instructing must be tightly integrated with
the modules for plan generation.

From a planning perspective, GIVE imposes very strict runtime
requirements on the plan generation task: planning must
happen in real time and the system must respond to a user in a timely
fashion.  If the system takes too much time deliberating over an
instruction to give, rather than actually giving this instruction, the
user may have walked or turned away, thus making the instruction
invalid.  Furthermore, plan execution monitoring also plays an
important role in the GIVE problem.  At a high level, the system needs
to monitor a user's actions and compare them against the generated
instruction set to determine if the user has correctly followed
directions or not. In the case of the latter, new instructions may
have to be generated. In practice, the situation can be quite
complicated since the mental state of the user is not known and so the
system must be inferred from observing the user's actions in real
time. For example, a user directed to ``turn around and walk through
the door'' may not necessarily perform these actions to the letter,
i.e., immediately turning 180 degrees and proceeding directly to the
door. Instead, the user might take a roundabout route through the
room, eventually exiting out the door.  Although the user's actions do
not match the generated instructions exactly, they meet the intended
goal. The system must be able to identify such ``equivalent'' plans
and not immediately generate new instructions as soon as the user's
actions have gone off course. Furthermore, a user can communicate
certain intentions to the system, both through action and
inaction. For instance, the system should infer that a user has failed
to follow instructions if the user exits a room when given a directive
to ``walk to the centre of the room''. The system should also make a
similar conclusion if a user simply does nothing when given the
instruction.

\section{Experiments}
\label{sec:experiments}

We now describe the results of four simple experiments designed to evaluate
the performance of three planners---FF \citep{HoffmannNebel01}, SGPLAN
5.2.2 \citep{hsu06:_new_featur_in_sgplan_for}, and an ad-hoc implementation
of GraphPlan \citep{Blum1997} written in Java---in our NLG domains. The FF
and SGPLAN planners were chosen due to their success in the International
Planning Competitions. Our implementation of GraphPlan is included as a
counterpoint to the other two, more complex, planners. Most notably, our
version of GraphPlan does not perform a preprocessing ``grounding'' step
like FF and SGPLAN but is much more selective in its predicate and action
instantiations.

On first impressions, our results indicate that planning is a promising
tool for both domains.  In the sentence generation domain, FF dramatically
outperforms the best previously known algorithm for the same problem (a
reimplementation of \citep{Stone2003a}), although the latter approach uses
a greedy search algorithm with a heuristic that is hand-tailored to the
domain. In the GIVE domain, SGPLAN computes a domain plan from the initial
state to the goal in 0.3 seconds, which is fast enough for moderately-sized
problems instances in the application.\footnote{All
  runtimes were measured on a Pentium 4 CPU running at 3 GHz. Java programs
  were allowed to ``warm up'', i.e., the planner was run five times and the
  first four measurements discarded to ensure that the JVM had just-in-time
  compiled all relevant bytecode.}

We also observe that although FF and SGPLAN manage the search for plans
very efficiently, they both spend comparatively large amounts of time
computing predicate and action instantiations in their preprocessing
stages, many of which are then never used during the search. FF's
``grounding time'' in particular dominates its overall planning time,
leading to some conflicting results. In some cases, our Java implementation
of GraphPlan significantly outperforms both state-of-the-art planners. The
experiments described below seek to improve our understanding of this
situation.

\subsection{Experiment 1: Sentence generation}
\label{sec:exper-1:-sent}

In the first experiment, we construct a series of sentence generation
problems which require the planner to compute a plan representing the
sentence ``Mary likes the Adj$_1$ \ldots Adj$_n$ rabbit.''  Each problem
instance assumes a certain number $m$ of rabbits that were distinguished by
$n \leq m$ different properties, such that all $n$ properties are required
to distinguish the target referent from all other rabbits.  The $n$
properties are realized as $n$ different adjectives, in any order.  This
setup allows us to control the plan length (a plan with $n$ properties will
have length $n+4$) and the universe size (the universe will contain $m+1$
individuals in addition to the differently-typed individuals used to encode
the grammar).

\begin{figure}
  \centering
  \includegraphics[width=0.75\columnwidth]{graph-exp1}
  %\includegraphics[width=1\columnwidth]{pic-runtime-modifiers-with-sgplan}
  \caption{Results for the sentence generation domain. The
    horizontal axis represents parameters $(m,n)$ from $(1,1)$ to
    $(10,10)$ in lexicographical order. The vertical axis is the
    runtime in milliseconds.}
  \label{fig:runtimes-crisp}
\end{figure}

The results of this experiment are shown in the graph in
Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
in lexicographic order on the horizontal axis and the runtime is shown in
milliseconds on the vertical axis. These results reveal a number of
interesting insights. First, FF significantly outperforms SGPLAN in this
domain, sometimes by a factor of 10 or more.\footnote{Experiments with
 SGPLAN use a pre-release version of SGPLAN, kindly provided to us by
 Chih-Wei Hsu.The release version of SGPLAN 5.2.2 had a bug which caused it
 to crash on some of our problem instances.}
Second, FF's runtime is dominated by its initial grounding step, in which
it computes the ground instances of all predicates and actions used in the
planning problem in order to avoid unnecessary instantiation during search.
In particular, the ratio of grounding time to total runtime is generally
above 85\%, and rises to above 99\% at $m=11$, which is still a relatively
small universe for this application.\footnote{The
  ``grounding'' time reported here is what FF reports as ``time spent:
  instantiating action templates''.}

In our testing examples, the time spent by FF on grounding is such that it
is consistently outperformed by our Java implementation of GraphPlan, which
only computes instances of predicates and actions as they are discovered
during the construction of the planning graph. Although FF is consistently
much faster as far as pure search time is concerned, our results indicate
that FF's performance is much more sensitive to the domain size: if we fix
$n=1$, FF takes 60 ms to compute a plan at $m=1$, but 2.4 seconds to
compute the same plan at $m=10$. By comparison, our GraphPlan
implementation takes 30 ms at $m=1$ and still only requires 50 ms at
$m=10$. Conversely, GraphPlan's runtime grows much faster with the plan
size (i.e., with growing values of $n$ for a fixed $m$). Larger, but still
realistically-sized, instances of the sentence generation problem are still
problematic for all the planners we tested.

\subsection{Experiment 2: Minimal GIVE worlds}
\label{sec:exper-2:-minim}

In the second experiment, we evaluate the performance of the planners on
problems arising in the GIVE domain. We construct a series of test worlds,
similar to the one illustrated in Figure~\ref{fig:give-minimal}. These
worlds consist of a $2n$ by $h$ grid of positions, such that there are
buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The
player starts in position $(1,1)$ and must press all the buttons
to successfully complete the game. The world is generated as a GIVE world
description, and then automatically converted into a planning problem by
the GIVE software. For instance, Figure~\ref{fig:give-planning} shows some
of the actions available in the GIVE domain in PDDL syntax.

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\columnwidth]{pic-buttons}
  \caption{Minimal GIVE world.}
  \label{fig:give-minimal}
\end{figure}

Results for the $h=20$ case, with $n$ ranging from $1$ to $40$, are shown
in Figure~\ref{fig:give-runtime-minimal}. The most obvious result is that
FF is unable to solve any problems beyond $n=13$ on our experimentation
machine within the memory limit of 1 GB. SGPLAN, on the other hand, solves
instances beyond $n=40$ without major problems. In this case, our
implementation of GraphPlan is unable to find any of these plans. The time
spent on grounding is not a major factor in either planner, probably
because the planners need more time to actually compute the plan---for
instance, the optimal plan for the problem $n=40$ has a length of about
1600 steps. As a concrete example, the following is a minimal plan for the
case of a $2$ by $2$ grid with $2$ buttons for the player to press (i.e.,
$n=1$, $h=2$):

\begin{enumerate}
\item $\mathsf{move}(\mathsf{pos\_1\_1},\mathsf{pos\_1\_2}, \mathsf{north})$,
\item $\mathsf{manipulate}\textsf{-}\mathsf{u1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_1\_2})$,
\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{north}, \mathsf{east})$,
\item $\mathsf{move}(\mathsf{pos\_1\_2}, \mathsf{pos\_2\_2},
  \mathsf{east})$,
\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{east}, \mathsf{south})$,
\item $\mathsf{move}(\mathsf{pos\_2\_2}, \mathsf{pos\_2\_1}, \mathsf{south})$,
\item $\mathsf{manipulate}\textsf{-}\mathsf{l1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_2\_1})$.
\end{enumerate}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.75\columnwidth]{graph-exp2}
  %\includegraphics[width=1\columnwidth]{pic-runtime-buttons}
  \caption{Results on the minimal GIVE
    domain for $h=20$. The horizontal axis is $n$. The vertical axis
    is the runtime in seconds.}
  \label{fig:give-runtime-minimal}
\end{figure}

\subsection{Experiment 3: GIVE worlds with extra positions}
\label{sec:experiment-3:-give}

In the third experiment, we vary the structure of the GIVE world in order
to judge the effect that universe size has on the planning problem in this
domain. Starting with the ordinary GIVE world described in Experiment~2, we
extend the world map by adding another $w$ by $h$ empty ``junk'' positions
to the right of the minimal world, as shown in Figure~\ref{fig:give-junk}.
These new positions are not actually required in any plan, but extend the
size of the state space and approximate the situation in the actual GIVE
domain where most grid positions are never used. We leave the initial state
and goal untouched. As before, we generate a GIVE world description and
then convert it into a planning problem in PDDL.

\begin{figure}[t]
  \centering
  \includegraphics[width=0.80\columnwidth]{pic-empty-buttons}
  \caption{GIVE world with extra ``junk'' positions.}
  \label{fig:give-junk}
\end{figure}

Results for the $h=20$, $n=5$ case with $w$ ranging from $1$ to $70$ are
shown in Figure~\ref{fig:give-runtime-junk}. As in Experiment~2, FF again
runs out of memory, this time at $w=17$, while SGPLAN easily solves inputs
beyond $w=70$. As in Experiment~2, our implementation of GraphPlan is
unable to generate any of these plans. Unlike Experiment 2, however, both
FF and SGPLAN now spend a substantial proportion of their time on
grounding. In SGPLAN, this translates to a ``parsing time'' (which we
assume includes grounding) which grows from 180 ms to 21.7 seconds as $w$
grows from $1$ to $75$. The rest of the runtime, which also includes the
search time, only grows from 400 ms to 2.3 seconds. This difference is
particularly dramatic given that the actual optimal plan in each case is an
identical plan of about 100 steps (i.e., the one we would have found in
Experiment 2 for the $h=20$, $n=5$ case). The planning times for these
instances are also concerning since times over a couple seconds will
negatively affect the overall response time of the system, which must react
in real time to user actions.

\begin{figure}[t]
  \centering
  \includegraphics[width=0.75\columnwidth]{graph-exp3}
  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
  \caption{Results on the GIVE domain with junk
    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
    The vertical axis is the runtime in seconds.}
  \label{fig:give-runtime-junk}
\end{figure}

\subsection{Experiment 4: GIVE worlds with inaccessible positions}
\label{sec:experiment-4:-give}

In the final experiment, we consider a variation on the GIVE worlds from
Experiment~3. As before, we begin with a base grid of $2n$ by $h$
positions, with buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i
\leq n$. We then generate a second grid of $w$ by $h$ positions and place
an additional button in this grid. Unlike Experiment~3, the second grid
does not extend the first grid but is ``disconnected'' so its positions are
inaccessible from the first grid, as shown in
Figure~\ref{fig:give-junk-nosoln}. Since the geometry of the grid makes it
impossible to construct a plan for pressing all the buttons in the world,
we instead investigate the time it takes a planner to arrive at the
conclusion that the problem cannot be solved.

Results for the $h=20$, $n=5$ case with $w$ ranging from 1 to 50 are shown
in Figure~\ref{fig:give-runtime-nosoln}.\footnote{FF and SGPLAN do
 not display runtimes when they fail to construct a plan so an external timing
 program was used to generate these results.}
In this case, the runtimes for FF rise sharply around $w=28$, producing
times that are unacceptable for an NLG system in practice. FF is unable
to complete problem instances above $w=35$. SGPLAN, by comparison, performs
exceptionally well and is able to complete problems instances up to $w=50$
in about 0.3 seconds and $w=80$ in about 0.6 seconds. These results
demonstrate that SGPLAN has a significant advantage over FF in this
experiment. 
%Our implementation of GraphPlan is unable to complete any of
%these problem instances.

\begin{figure}[t]
  \centering
  \includegraphics[width=0.85\columnwidth]{pic-empty-inaccessible}
  \caption{GIVE world with inaccessible positions.}
  \label{fig:give-junk-nosoln}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.85\columnwidth]{graph-exp4}
  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
  \caption{Results on the GIVE domain with inaccessible
    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
    The vertical axis is the runtime in seconds.}
  \label{fig:give-runtime-nosoln}
\end{figure}

\section{Discussion}
\label{sec:discussion}

The positive conclusion we draw from our simple set of experiments is that
modern planners are fairly good at controlling the search for many of the
moderately-sized NLG problems we looked at. This is particularly true for
FF in the sentence generation domain, which generates nontrivial 14-word
sentences in about two seconds. Our results also indicate there is still
room for improvement, however: special-purpose algorithms are able to
generate much larger referring expressions in milliseconds
\citep{AreKolStr08}. Although search efficiency is not the main concern in
these domains, it nevertheless becomes a factor in scenarios like
Experiment~2, where the world is almost completely unconstrained and
(unsurprisingly) the search becomes much slower.

The most restrictive bottleneck in the two NLG domains we investigated is
the initial grounding step that many modern planners perform. While this
may be a good strategy for traditional planning benchmarks and IPC
domains---where plan length often dominates universe size and an initial
investment into grounding pays off in saved instantiations during
search---this approach has a very pronounced effect in domains like ours.
As our experiments show, there are natural planning domains in which
relatively short plans must be computed in large universes. For instance,
in the sentence generation domain, it is not unrealistic to look for plans
of length 20 over a universe consisting of thousands of domain individuals
and tens of thousands of possible action instantiations, some of which
require three domain individuals as parameters. Planners like FF and SGPLAN
are less effective in such domains, to the point that grounding
often dominates the total running time of these planners. (In the case of
Experiment~1, our ad-hoc implementation of GraphPlan outperforms both
planners in the sentence generation task.)

Since our domains are not that unusual in their structure and composition,
we hope that the lessons learned from our experiences can help improve the
performance of current planning systems. For instance, the common strategy
of splitting the universe into several types of individuals will not be a
very effective solution in our NLG domains given the large number of
individuals that could exist for a given type. On the other hand,
sophisticated techniques based on reachability analysis may have the
potential to improve this situation, provided such approaches can avoid the
drawbacks of unnecessary grounding.\footnote{See
  \citep{Bryce:07} for a good survey on reachability heuristics for planning
  graph-based approaches to plan generation.}
While recent trends in planning research have (rightfully) focused on the
development of algorithms that control search in sophisticated ways,
resulting in state-of-the-art planners that are more powerful and
successful than their predecessors, we nevertheless believe that
improvements are necessary if planning is to become a more mature
technology that can offer tools to a wider community of users and
researchers. At the end of the day, real-world users will care about the
\emph{total} runtime of a planner, which is more than just the search time.

By and large, our experiences with the planning community from a user's
point of view have been favourable. Thanks to the planning competitions, it
is easy to identify and download a fast state-of-the-art planner, at least
for Linux. However, in the course of our experiments we found and reported
bugs in both FF and SGPLAN. We also discovered that deciding as to which
planner is appropriate for a particular task is not always straightforward,
as the best performing planner in the IPC is not necessarily suitable for
every planning task. As our experiments have shown, even planners as
closely related as FF and SGPLAN can differ significantly in their
performance on different domains (for instance, our results in
Experiment~4).

\section{Conclusion}
\label{sec:conclusion}

In this paper, we introduced two novel planning domains arising from
problems in natural language generation: the sentence generation domain and
the Generating Instructions in Virtual Environments (GIVE) challenge
domain. We also reported the results of a set of experiments designed to
evaluate the suitability of off-the-shelf planners (particularly FF and
SGPLAN) in these domains.

Our results were mixed. While the planners we tested do an impressive job
of controlling the complexity of search in many domains, they also suffer
from practical problems that limit their performance in unexpected ways: in
both of our NLG domains the grounding step performed by FF and SGPLAN
dominates the time it takes to perform the search itself. As future work,
we hope to extend our analysis to other planners that do not perform the
initial grounding step, planners such as SHOP2
\citep{DBLP:journals/jair/NauAIKMWY03} that have traditionally performed
well on a wide range of domains, and new planners introduced in the latest
edition of the IPC.\footnote{See
 \url{http://ipc.informatik.uni-freiburg.de/} for details of the 2008
 International Planning Competition (deterministic track).}

The NLG community's recent interest in planning presents a valuable
opportunity for planning researchers. Domains like GIVE highlight certain
challenges, such as plan execution monitoring and plan presentation (i.e.,
summarisation and elaboration), but also offer a platform on which such
technologies can be evaluated in experiments with human users. Although we
have focused on classical planning problems in this work, research related
to reasoning under uncertainty, resource management, and planning with
knowledge and sensing, can also be investigated in these settings. As such,
we believe our domains would provide suitable challenges for planners
entered in future editions of the IPC. Provided some of the challenges we
have identified can be addressed, NLG-inspired problem domains offer a
constructive platform for the planning community to showcase their
techniques to a wider audience---and to improve the quality of their tools
for real-world planning tasks.

\section*{Acknowledgements}

This work arose in the context of the Planning and Language Interest Group
at the University of Edinburgh. The authors would like to thank all members
of this group, especially Hector Geffner and Mark Steedman, for interesting
discussions. This work was partially supported by the DFG Research
Fellowship ``CRISP: Efficient integrated realization and microplanning''
and by the European Commission through the PACO-PLUS project
(FP6-2004-IST-4-27657).

\bibliography{references}
\bibliographystyle{ci}

\pagebreak
\listoffigures

\end{document}

### 2009-05-04 trunk/Papers/ci-spark/manuscript.tex
strengthened intro; updated graffle picture for PDF

@@
 Postfach 151150 \\
-66041 Saarbrücken \\
+66041 Saarbr\"ucken \\
 Germany \\
@@
 \bibpunct{(}{)}{;}{a}{}{,}
+
+\newcommand{\todo}[1]{\textbf{(#1)}}
 
@@
 
-Natural language generation (NLG; \citealp{reiter00building}) is one of the
-major subfields of natural language processing, concerned with computing
-natural language sentences or texts that convey a given piece of
-information to the user. Intuitively, this task can be viewed as a problem
-involving actions, beliefs, and goals: an agent communicating with another
-agent tries to change the mental state of the hearer by applying actions
-which correspond to the utterance of words or sentences.
-This characterization of NLG suggests obvious parallels to automated
-planning---a view which has a long tradition in NLG.
+Natural language generation (NLG; \citealp{reiter00building}) is one
+of the major subfields of natural language processing, concerned with
+computing natural language sentences or texts that convey a given
+piece of information to the user. This task can be viewed as a problem
+of achieving a (communicative) goal by successively applying
+(communicative) actions, and thus has clear intuitive parallels to
+automated planning.
 
-Early approaches to generation as planning
-\citep{perrault80,appelt:planning,hovy88,young94dpocl} focused primarily on
-planning high-level structures, such as speech acts and discourse
-relations.  While these approaches suffered from the inefficiency of the
-planners that were available at the time, there has recently been a
-resurgence of interest in applying modern planning techniques to NLG.
-These newer papers revisit speech act planning based on the more expressive
-planning formalisms available today \citep{Steedman-Petrick:07,benotti08b}
-and exploit the improved efficiency of modern planners for an integrated
-model of sentence planning and surface realization \citep{KolSto07}.
-Problems in NLG are often computationally challenging (see, e.g.,
-\citep{KolStr02}), and these new approaches combine an interest in planning
-as a modelling tool for natural language, with a hope of improved
-efficiency by exploiting modern approaches and algorithms.
+This view of generation as planning has a long tradition in NLG
+\citep{perrault80,appelt:planning,hovy88,young94dpocl}. Despite their
+theoretical attractiveness, these early approaches never gained
+mainstream use in applied systems, in part because of the inefficiency
+of the planners that were available at the time. However, the recent
+improvements in planner efficiency and expressiveness has sparked a
+renewed interest in applying modern planning techniques to NLG---both
+on the level of speech act planning
+\citep{Steedman-Petrick:07,benotti08b} and and on the level of
+sentence generation \citep{KolSto07}. The goal of this paper is to
+determine, by looking at some representative generation problems,
+whether planning has advanced to the point that this enterprise can be
+successful this time.
 
-The focus of this paper is twofold. First, we present two novel planning
-domains that arise in the context of natural language generation: the
-sentence generation domain and the GIVE domain. In the sentence generation
+The focus of this paper is twofold. First, we present two generation
+problems that have recently been cast as planning problems: the
+sentence generation task and the GIVE task. In the sentence generation
 task, the goal is to generate a single sentence that expresses a given
-meaning. \citet{KolSto07} cast this domain as a planning problem, where the
-plan encodes the necessary sentence and the actions correspond to uttering
-individual words.  In the GIVE domain (``Generating Instructions in Virtual
-Environments''), we describe a new shared task that was recently posed as a
-challenge for the NLG community \citep{alexander07:_shared_task_propos}.
-GIVE uses planning as part of an NLG system that generates natural-language
-instructions to guide a user in performing a given task in a virtual
-environment.
+meaning. \citet{KolSto07} cast this task as a planning problem, where
+the plan encodes the necessary sentence and the actions correspond to
+uttering individual words.  In the GIVE domain (``Generating
+Instructions in Virtual Environments''), we describe a new shared task
+that was recently posed as a challenge for the NLG community
+\citep{ByrKolStrCasDalMooObe09}.  GIVE uses planning as part of an NLG
+system that generates natural-language instructions to guide a user in
+performing a given task in a virtual environment.
 
 Second, we discuss some of our experiences using off-the-shelf
-planners in our two NLG-inspired domains. In particular, we explore the
-efficiency of FF \citep{HoffmannNebel01} and SGPLAN
+planners in our two NLG-inspired domains. In particular, we explore
+the efficiency of FF \citep{HoffmannNebel01} and SGPLAN
 \citep{hsu06:_new_featur_in_sgplan_for}: two planners that are readily
-available, support an expressive subset of the Planning Domain Definition
-Language (PDDL; McDermott et al.~\citeyear{PDDL}) for encoding domains, and
-have been successful on traditional benchmarks and problems from the
-International Planning Competition (IPC)\footnote{See
- \url{http://ipc.icaps-conference.org/} for information about past
- editions of the IPC.}.
-We test these planners on a range of problem instances in our planning
-domains, and compare our results against an ad-hoc Java implementation of
-GraphPlan \citep{Blum1997}. Overall, our findings are mixed. On the one
-hand, we demonstrate that modern planners can easily handle some of the
-moderately-sized search problems that arise in NLG. In certain cases, the
-planners we tested are more efficient than specialized NLG algorithms
-applied to the same problem. On the other hand, these same planners spend
-tremendous amounts of time on preprocessing. For instance, in the sentence
-generation domain, FF spends 90\% of its runtime grounding out literals and
-actions, and ends up being slower than our version of GraphPlan that avoids
-grounding. For domains like ours, which are dominated by the number of
-actions and the universe size, rather than the combinatorics of the search
-problem, this observation suggests that the overall runtime of a planner
-could be improved by grounding more selectively. This is particularly true
-for large problem instances, which still remain a challenge for the
-planners we tested, and where the effects of the grounding approach become
-quite pronounced. We therefore offer our NLG domains as challenges for the
-planning community.
+available, support an expressive subset of the Planning Domain
+Definition Language (PDDL; McDermott et al.~\citeyear{PDDL}) for
+encoding domains, and have been successful on traditional benchmarks
+and problems from the International Planning Competition
+(IPC)\footnote{See \url{http://ipc.icaps-conference.org/} for
+  information about past editions of the IPC.}.  We test these
+planners on a range of problem instances in our planning domains, and
+compare our results against an ad-hoc Java implementation of GraphPlan
+\citep{Blum1997}. Overall, our findings are mixed. On the one hand, we
+demonstrate that modern planners can easily handle the \emph{search}
+problems that arise in NLG on realistic inputs, which is promising
+given that the sentence generation task, at least, is NP-complete
+\citep{KolStr02}. On the other hand, these same planners spend
+tremendous amounts of time on preprocessing. For instance, in the
+sentence generation domain, FF spends 90\% of its runtime grounding
+out literals and actions, and ends up being slower than our version of
+GraphPlan that avoids grounding. For domains like ours, which are
+dominated by the number of actions and the universe size, rather than
+the combinatorics of the search problem, this observation suggests
+that the overall runtime of a planner could be improved by grounding
+more selectively. This is particularly true for large problem
+instances, which still remain a challenge for the planners we tested,
+and where the effects of the grounding approach become quite
+pronounced. We conclude that planning \emph{algorithms} have advanced
+to the point that generation as planning is now a realistic endeavour,
+but the concrete \emph{implementations} of these algorithms must be
+improved to be useful in this application, and offer our NLG domains
+as challenges for the planning community.
 
-\section{Planning in NLG}
+This paper is structured as follows. We begin by briefly reviewing
+some of the literature on NLG as planning
+(Section~\ref{sec:nlg-as-planning}. Then we define introduce the
+planning problems arising in two particular NLG tasks in more detail
+in Section~\ref{sec:domains}. We report on some experiments with these
+planning problems in Section~\ref{sec:experiments}, and discuss the
+results in Section~\ref{sec:discussion}. Section~\ref{sec:conclusion}
+concludes.
+
+\section{NLG as Planning} \label{sec:nlg-as-planning}
+
+The task of generating natural language from semantic representations
+is typically split up into two parts: the \emph{domain planning} task,
+which selects the information that is to be conveyed and structures it
+into sentence-sized chunks, and the \emph{sentence generation} tasks,
+which then translates each of these chunks into natural language
+sentences. The sentence generation task is often split up into two
+parts of its own---the \emph{sentence planning} task, which enriches
+the input by e.g.\ determining how to refer to objects and selects
+some lexical material, and the \emph{surface realization} task, which
+maps the enriched meaning representation into a sentence using a
+grammar. The chain of domain planning, sentence planning, and surface
+realization is sometimes called the ``NLG pipeline''
+\cite{reiter00building}.
+
+Viewing generation as a planning problem has a long tradition in the
+NLG literature, especially in discourse planning. \todo{Sketch
+  Perrault et al here. Say something very brief about Appelt, Hovy,
+  Moore. Say why all this didn't really work.}
+
+Most recently, interest in generation as planning has been revived for
+NLG tasks on various levels of the NLG pipeline. Below, we will look
+in some detail at \cite{KolSto07}'s recasting of sentence generation
+as planning; furthermore, \cite{Steedman-Petrick:07} and
+\cite{benotti08b} have applied planning to \todo{something}. Notice
+that although all of these approaches encode ``communicative actions''
+as planning operators, they still solve fundamentally different
+problems. For instance, Perrault et al.\ tackled discourse planning
+with planning operators that encode speech acts, each of which
+represents the meaning of a whole sentence and models the contribution
+of this sentence to the intentional structure of the entire
+discourse. By contrast, Koller and Stone solve a sentence generation
+problem using planning operators that encode utterances of single
+words and model the contribution of each word to the grammatical
+structure of the sentence. Whether these different levels of ``NLG as
+planning'' could be combined into a single planning problem, and
+whether this would make computational sense, is an open question.
+
+Problems in NLG at all levels of description can become
+computationally complex very easily. For instance, \citet{KolStr02}b
+showed that even surface realization is NP-complete under reasonable
+assumptions. This makes it necessary to investigate methods for
+guiding the search, which makes modern planning techniques a promising
+approach to NLG. The point of this paper is to investigate whether
+existing planners fulfill this promise of efficiency for the
+application to NLG.
+
+\section{Two NLG Tasks}
 \label{sec:domains}
 
-We begin by presenting our two planning domains: the sentence generation
-domain and the instruction giving domain.
+We will now focus the discussion towards two specific NLG problems:
+sentence generation in the sense of \cite{KolSto07}, and the
+generation of instructions in virtual environments
+\cite{ByrKolStrCasDalMooObe09}.
 
@@
   \centering
-  \includegraphics[width=0.75\columnwidth]{pic-derivation-2}
+  \includegraphics[width=0.75\columnwidth]{pic-derivation}
   \caption{Derivation of ``The white rabbit sleeps.''}

### 2009-05-04 trunk/Papers/ci-spark/manuscript.tex
gone over domains chapter

@@
 whether planning has advanced to the point that this enterprise can be
-successful this time.
+successful this time---and thus, to determine the usefulness of
+current planning technology to an application area that has a long
+tradition, but is not currently in the focus of the planning
+competitions.
 
@@
 Problems in NLG at all levels of description can become
-computationally complex very easily. For instance, \citet{KolStr02}b
+computationally complex very easily. For instance, \citet{KolStr02}
 showed that even surface realization is NP-complete under reasonable
@@
 
-Sentence generation is the problem of computing from a grammar and a
-semantic representation, a single sentence that expresses this piece of
-meaning. Traditionally, this problem is split into several steps
-\citep{reiter00building}. In a first step, called \emph{sentence planning},
-the semantic representation is enriched with additional information. For
-instance, \emph{referring expressions}, which refer to individuals (e.g.,
-objects, people, etc.) in the domain of discourse, are determined at this
-point. In a second step, called \emph{surface realization}, this enriched
-representation is translated into a natural-language sentence, using the
-grammar.
+Although the sentence generation task is often split up into two
+separate steps for sentence planning and surface realization, the
+decisions that an NLG system must make in these two modules often
+interact. For instance, in order to generate \emph{referring
+  expressions (REs)}, i.e.\ unique descriptions of individuals in the
+universe of discourse, it is necessary to know what properties of
+individuals can be expressed as nouns and adjectives. Grammatical
+information like this should belong exclusively to the surface
+realization module; but on the other hand, RE generation enriches the
+input meaning representation and is therefore usually seen as part of
+sentence planning.
 
-In practice, the process of determining referring expressions typically
-interacts with the realization step and so it often turns out to be
-beneficial to perform these two steps together. For instance, the SPUD
-system (``Sentence Planning Using Description''; \citealt{Stone2003a})
-performed this task using a top-down generation algorithm based on
-tree-adjoining grammars \citep{joshi;etal1997} whose lexical entries were
-equipped with semantic and pragmatic information. Unfortunately, SPUD had
-to resort to a non-optimal greedy search strategy to cope with its large
-search space and retain reasonable efficiency. To improve the efficiency
-of SPUD, \citet{KolSto07} translate the sentence generation problem into a
-planning problem and use a planning algorithm to perform the generation
-task.
+One seminal approach to solving the sentence planning and surface
+realization problems together is offered by the SPUD generation system
+(``Sentence Planning Using Description''; \citealt{Stone2003a}). This
+approach assumes a tree-adjoining grammar (TAG) \citep{joshi;etal1997}
+whose lexical entries are equipped with semantic and pragmatic
+information, as well as two sets of ground atoms: the
+\emph{communicative goal} that specifies what semantic information we
+must express, and a \emph{knowledge base}. Let's consider an example
+to illustrate the problem that SPUD solves. Consider a knowledge base
+containing the individuals $e$, $r_1$ and $r_2$, and a set of
+attributes encoding the fact that $r_1$ and $r_2$ are rabbits, $r_1$
+is white and $r_2$ is brown, and $e$ is an event in which $r_1$
+sleeps.  Say that we want to express the information
+$\{\mathsf{sleep}(e,r_1)\}$ using the tree-adjoining grammar shown in
+Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
+\emph{elementary trees} (i.e., the disjoint trees in the figure), each
+of which contributes certain \emph{semantic content}. We can
+instantiate these trees by substituting individuals for \emph{semantic
+  roles}, such as $\mathsf{self}$ and $\mathsf{subj}$, and then
+combine the tree instances as shown in
+Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
+``The white rabbit sleeps''.
 
@@
 
-We illustrate this process on a simplified example.  Consider a
-knowledge base containing the individuals $e$, $r_1$ and $r_2$, and a
-set of attributes encoding the fact that $r_1$ and $r_2$ are rabbits,
-$r_1$ is white and $r_2$ is brown, and $e$ is an event in which $r_1$
-sleeps.  Say that we want to express the information
-$\{\mathsf{sleep}(e,r_1)\}$ using the tree-adjoining grammar shown in
-Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
-\emph{elementary trees} (i.e., the disjoint trees in the figure), each
-of which contributes certain \emph{semantic content}. We can
-instantiate these trees by substituting individuals for \emph{semantic
-  roles}, such as $\mathsf{self}$ and $\mathsf{subj}$, and then
-combine the tree instances as shown in
-Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
-``The white rabbit sleeps''.
-
-We compute this grammatical derivation in a top-down manner, starting with
-the elementary tree for ``sleeps''. This tree satisfies the need to convey
-the semantic information, but introduces a need to generate a noun phrase
-(NP) for the subject; this NP must refer uniquely to the target referent
-$r_1$. In a second step, we substitute the tree for ``the rabbit'' into the
-open NP leaf, which makes the derivation grammatically complete. Since
-there are two different individuals that could be described as ``the
-rabbit''---technically, $r_2$ is still a \emph{distractor} (i.e., based on
-the description ``the rabbit'', the hearer might erroneously think that
-we're talking about $r_2$ and not $r_1$)---we are still not finished. To
-complete the derivation, the tree for ``white'' is added to the existing
-structure by an \emph{adjunction} operation, making the derivation
-both syntactically and semantically complete.
+We can now consider various different algorithms for solving the SPUD
+problem. The naive algorithm builds the TAG derivation top-down. In
+the example, it starts with the elementary tree for ``sleeps''. This
+tree satisfies the need to convey the semantic information, but
+introduces a need to generate a noun phrase (NP) for the subject,
+which must refer uniquely to the target referent $r_1$. In a second
+step, we substitute the tree for ``the rabbit'' into the open NP leaf,
+which makes the derivation grammatically complete. Since there are two
+different individuals that could be described as ``the
+rabbit''---technically, $r_2$ is still a \emph{distractor} (i.e.,
+based on the description ``the rabbit'', the hearer might erroneously
+think that we're talking about $r_2$ and not $r_1$)---we are still not
+finished. To complete the derivation, the tree for ``white'' is added
+to the existing structure by an \emph{adjunction} operation, making
+the derivation both syntactically and semantically complete. This
+search algorithm takes worst-case exponential time because it may have
+to consider many combinations of elementary trees; and in fact, the
+SPUD problem generalizes the surface realization problem considered by
+\citet{KolStr02} and is therefore NP-complete.
 
@@
 
-The process described above has clear parallels to planning: we manipulate
-states by applying sequences of actions to achieve a goal. We can make this
-connection even more precise by translating the SPUD problem into a
-planning problem.\footnote{See
-  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
-  which this conversion is implemented.}  For instance,
+The SPUD system itself worked around this complexity by using a
+greedy, incomplete search algorithm. On the other hand,
+\citet{KolSto07} present an alternative to controlling the necessary
+search, by encoding the generation problem into a planning
+problem.\footnote{See \url{http://code.google.com/p/crisp-nlg/} for
+  the CRISP system, in which this conversion is implemented.} Each
+operator of this planning problem models the addition of some
+elementary tree to the TAG derivation. For instance,
 Figure~\ref{fig:white-rabbit-as-planning} shows the corresponding PDDL
-actions for the above generation task, where each action corresponds
-to an operation that adds a single elementary tree to the derivation.
-In each case, the first parameter of the action is a node name in the
-derivation tree, and the remaining parameters stand for the
-individuals to which the semantic roles will be instantiated.  The
-syntactic preconditions and effects are encoded using $\mathsf{open}$
-and $\mathsf{canadjoin}$ literals; the status of each referring expression
-is tracked using $\mathsf{distractor}$ literals. The resulting action
-effects contain terms of the form $\mathsf{subj}(u)$, which construct new
-node names, but aren't strictly permitted by PDDL. In order to meet the
-syntactic requirements of PDDL, these terms can be eliminated by estimating
-an upper bound $n$ for the plan length, making $n$ copies of each action,
-ensuring that copy $i$ can only be applied in step $i$, and replacing the
-term $\mathsf{subj}(u)$ in an action copy by the constant
-$\mathsf{subj}_i$.
-
-Once the appropriate actions are defined, we can solve the generation
-problem as an ordinary planning problem. For instance, the following plan
-solves our previous example:
+actions for the above generation task. The \texttt{add-sleeps}
+operator represents the addition of the elementary tree for ``sleeps''
+to the derivation. $\verb?add-sleep?(\mathsf{root},e,r_1)$ removes the
+atom $\mathsf{subst}(S,\mathsf{root})$, which expresses that there is
+currently an open substitution node $\mathsf{root}$ with label $S$,
+from the planning state and replaces it with an atom
+$\mathsf{subst}(NP,\mathsf{subj}(\mathsf{root}))$; here
+$\mathsf{subj}(\mathsf{root})$ is meant as shorthand for a fresh
+individual name.\footnote{These terms, which are not valid in ordinary
+  PDDL, can be eliminated by estimating an upper bound $n$ for the
+  plan length, making $n$ copies of each action, ensuring that copy
+  $i$ can only be applied in step $i$, and replacing the term
+  $\mathsf{subj}(u)$ in an action copy by the constant
+  $\mathsf{subj}_i$. Notice that $S$, $NP$, and $N$ are constants.}
+At the same time, the operator records that the semantic information
+$\mathsf{sleep}(e,r_1)$ has now been expressed, and introduces all
+individuals except for $r_1$ as distractors for the new RE at
+$\mathsf{subj}(\mathsf{root})$. These distractors can then be removed
+by subsequent applications of the other two operators. By assuming an
+initial state that encodes the knowledge base and contains an atom
+$\mathsf{subst}(S,\emph{root})$, and a goal that contains, among
+others, $\forall x \forall y. \neg \mathsf{subst}(x,y)$ and
+$\mathsf{expressed}(\emph{sleep},e,r_1)$, a planner will find plans
+like the following:
 %
@@
 
-The object of the GIVE Challenge (``Generating Instructions in Virtual
-Environments''; \citealt{alexander07:_shared_task_propos}) is to build an
-NLG system which is able to produce natural-language instructions which
-will guide a human user in performing some task in a virtual environment.
-From an NLG perspective, GIVE makes for an interesting challenge because it
-is a theory-neutral task that exercises all components of an NLG system,
-and emphasizes the study of communication in a (simulated) physical
-environment. It also has the advantage that the user and the NLG system can
-be in physically different places, as long as the 3D client and the NLG
-system are connected over a network. This makes it possible to evaluate
-GIVE NLG systems on a large scale over the Internet.
-%
-The first GIVE evaluation is currently underway. Five NLG systems from
-three countries are in the process of being evaluated. At the time of
-writing, 1028 different evaluation runs---each consisting of one user
-interacting with one instance of the virtual environment and being guided
-by an NLG system---have been collected, making GIVE the largest-scale
-NLG evaluation effort ever.\footnote{See the GIVE website at
- \url{http://www.give-challenge.org/} for more details about this
- challenge.}
+Let's now turn to a second recent application of planning in NLG, the
+GIVE Challenge (``Generating Instructions in Virtual Environments'';
+\citealt{ByrKolStrCasDalMooObe09}). The object of this shared task is
+to build an NLG system which produces natural-language instructions
+which will guide a human user in performing some task in a virtual
+environment.  From an NLG perspective, GIVE makes for an interesting
+challenge because it is a theory-neutral task that exercises all
+components of an NLG system, and emphasizes the study of communication
+in a (simulated) physical environment. Another advantage is that GIVE
+makes it possible to connect NLG systems to users over the Internet,
+and thus offer cheap access to human experimental subjects. The first
+installment of GIVE (GIVE-1) evaluated five NLG systems on the
+performance of 1143 users, making it the largest NLG evaluation effort
+in terms of human users ever.
 
@@
 \includegraphics[width=0.75\columnwidth]{give_world_2}
-\caption{Map of the GIVE development world.}
+\caption{Map of a GIVE world.}
   \label{fig:give-development-world}
@@
 
-An example GIVE world map is shown in
-Figure~\ref{fig:give-development-world}. To simplify both the planning and
-the NLG task, the world is discretised into a set of tiles of equal size.
-The user can turn by 90 degree steps in either direction, and can move from
-the centre of one tile to the centre of the next tile, provided the path
-between two tiles is not blocked. The world also consists of a set of
-objects which can be manipulated by the user in various ways. For
-instance, in the example world the user's task is to pick up a trophy in
-the top left room. The trophy is hidden in a safe behind a picture, so the
-user must first move the picture out of the way and open the safe by
-pushing a sequence of buttons (the small square boxes on the walls) in the
-correct order. In order to get to all these buttons, the user must also
-open the door in the centre of the map and deactivate the alarm tile by
-pushing further buttons. Figure~\ref{fig:give-planning} shows the encoding
-of some of the available GIVE domain actions, in PDDL syntax.
+Planning plays a central role in the GIVE task. To see this, consider
+the example GIVE world map in
+Figure~\ref{fig:give-development-world}. To simplify both the planning
+and the NLG task, this world is discretised into a set of tiles of
+equal size.  The user can turn by 90 degree steps in either direction,
+and can move from the centre of one tile to the centre of the next
+tile, provided the path between two tiles is not blocked. The world
+also consists of a set of objects which can be manipulated by the user
+in various ways. For instance, in the example world the user's task is
+to pick up a trophy in the top left room. The trophy is hidden in a
+safe behind a picture; to access it, the user must push certain
+buttons in order to move the picture out of the way, open the safe,
+and open doors. Figure~\ref{fig:give-planning} shows the encoding of
+some of the available GIVE domain actions, in PDDL syntax. In the
+example, the shortest plan to solve the task consists of 108 action
+steps, with the first few steps as follows:
+%
+\begin{enumerate}
+\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north},
+\mathsf{west})$,
+\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
+\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$,
+\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
+\end{enumerate}
 
@@
 \end{minipage}
-\caption{PDDL actions for the GIVE domain.}
+\caption{Some PDDL actions for the GIVE domain.}
 \label{fig:give-planning}
@@
 
-The plans generated in the GIVE domain are often nontrivial. For instance,
-in the above example the shortest plan consists of 108 action steps, with
-the first few steps as follows:
+A GIVE NLG system must be able to compute such plans. At a minimum,
+the discourse planner will call a planner in order to determine the
+content of the instructions that should be presented to the
+user. Under this view, the planning problem is very similar to the
+Gridworld problem (see, e.g., \citep{Tovey-Koenig:2000}), which also
+involves finding a route through a two-dimensional world map with
+discrete positions. GIVE tasks are usually much more complex, however,
+as illustrated by the above example map: a successful plan must
+include steps for pressing buttons in the right order, reasoning about
+large numbers of world objects, and navigating through complicated
+room shapes. This relatively loose integration of NLG system and
+planner is the state of the art of the systems that participated in
+GIVE-1.
+
+However, in order to solve the GIVE task in a more satisfying way, we
+expect that planning and generation will have to be interleaved more
+closely, making the planning a genuine part of the NLG task. For
+instance, the NLG system could generate an instruction sequence
+starting with ``turn left; walk forward; press the button; turn right;
+walk forward; walk forward; turn right; walk forward; walk forward;
+turn left; walk forward''. But such instructions are clumsy and
+uninteresting; it would be much better to say ``turn around and walk
+through the door''. That is, the system should \emph{summarise} plans
+by merging multiple action instances into single instructions before
+presenting them to the user.\footnote{This task is similar to the
+  construction of ``macro'' actions, but is primarily intended for
+  plan presentation purposes. See, e.g., \citep{Botea-etal:05} which
+  contains a survey of recent approaches for generating macro actions
+  in planning.}  Conversely, it may also be necessary to
+\emph{elaborate} on a single planning step by expressing it with
+several instructions. Having just entered the top left room, it may be
+easier for the user to understand the instruction ``walk to the centre
+of the room; turn right; now press the green button in front of you'',
+rather than the instruction ``press the green button on the wall to
+your right''. Furthermore, in order to plan the referring expression
+``the green button in front of you'' at a time when the button is not
+yet in front of the user, the NLG system must keep track of the
+hypothetical changes in what objects are visible to the user. Thus,
+the NLG acts of referring and instructing must be tightly integrated
+with the modules for plan generation. This makes the GIVE domain
+planning problem an integral part of the NLG system, which makes this
+a second planning problem of immediate relevance to NLG that is
+completely different from the first one.
+
+% It seems to me that this paragraph misses the point. We're not
+% trying to sell GIVE as a cool problem here; we take the perspective
+% that GIVE exists and has already been shown to be cool, and we only
+% talk about solving it here. - AK
 %
-\begin{enumerate}
-\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north},
-\mathsf{west})$,
-\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
-\end{enumerate}
-
-Most of the actions in the above plan are ``move'' actions. This makes the
-bulk of the GIVE problem very similar to that of a Gridworld problem (see,
-e.g., \citep{Tovey-Koenig:2000}), which also involves finding a route
-through a two-dimensional world map with discrete positions. GIVE tasks are
-usually much more complex, however, as illustrated by the above example
-map: a successful plan must include steps for pressing buttons in the right
-order, reasoning about large numbers of world objects, and navigating
-through complicated room shapes.
-
-The GIVE task comes with a number of interesting research challenges for
-both the NLG and planning communities. On the NLG side, one major problem
-is that verbalising each action instance in a plan makes for very
-poor natural-language instructions. For instance, the NLG system could
-generate an instruction sequence starting with ``turn left; walk forward;
-press the button; turn right; walk forward; walk forward; turn right; walk
-forward; walk forward; turn left; walk forward''. However, directions of
-this form are linguistically clumsy and uninteresting, and it is to be
-expected that frustrated users will quickly cancel an interaction with such
-a system, resulting in a bad evaluation score. Having just pressed the
-button, it would be much better to say ``turn around and walk through the
-door''. That is, the system should \emph{summarise} plans by merging
-multiple action instances into single instructions before presenting them
-to the user.\footnote{This task is similar to the construction of
-  ``macro'' actions, but is primarily intended for plan presentation
-  purposes. See, e.g., \citep{Botea-etal:05} which contains a survey of
-  recent approaches for generating macro actions in planning.}
-Conversely, it may also be necessary to \emph{elaborate} on a single
-planning step by expressing it with several instructions. Having just
-entered the top left room, it may be easier for the user to understand the
-instruction ``walk to the centre of the room; turn right; now press the
-green button in front of you'', rather than the instruction ``press the
-green button on the wall to your right''. Furthermore, in order to plan the
-referring expression ``the green button in front of you'' at a time when
-the button is not yet in front of the user, the NLG system must keep track
-of the hypothetical changes in what objects are visible to the user. Thus,
-the NLG acts of referring and instructing must be tightly integrated with
-the modules for plan generation.
-
-From a planning perspective, GIVE imposes very strict runtime
-requirements on the plan generation task: planning must
-happen in real time and the system must respond to a user in a timely
-fashion.  If the system takes too much time deliberating over an
-instruction to give, rather than actually giving this instruction, the
-user may have walked or turned away, thus making the instruction
-invalid.  Furthermore, plan execution monitoring also plays an
-important role in the GIVE problem.  At a high level, the system needs
-to monitor a user's actions and compare them against the generated
-instruction set to determine if the user has correctly followed
-directions or not. In the case of the latter, new instructions may
-have to be generated. In practice, the situation can be quite
-complicated since the mental state of the user is not known and so the
-system must be inferred from observing the user's actions in real
-time. For example, a user directed to ``turn around and walk through
-the door'' may not necessarily perform these actions to the letter,
-i.e., immediately turning 180 degrees and proceeding directly to the
-door. Instead, the user might take a roundabout route through the
-room, eventually exiting out the door.  Although the user's actions do
-not match the generated instructions exactly, they meet the intended
-goal. The system must be able to identify such ``equivalent'' plans
-and not immediately generate new instructions as soon as the user's
-actions have gone off course. Furthermore, a user can communicate
-certain intentions to the system, both through action and
-inaction. For instance, the system should infer that a user has failed
-to follow instructions if the user exits a room when given a directive
-to ``walk to the centre of the room''. The system should also make a
-similar conclusion if a user simply does nothing when given the
-instruction.
+% From a planning perspective, GIVE imposes very strict runtime
+% requirements on the plan generation task: planning must
+% happen in real time and the system must respond to a user in a timely
+% fashion.  If the system takes too much time deliberating over an
+% instruction to give, rather than actually giving this instruction, the
+% user may have walked or turned away, thus making the instruction
+% invalid.  Furthermore, plan execution monitoring also plays an
+% important role in the GIVE problem.  At a high level, the system needs
+% to monitor a user's actions and compare them against the generated
+% instruction set to determine if the user has correctly followed
+% directions or not. In the case of the latter, new instructions may
+% have to be generated. In practice, the situation can be quite
+% complicated since the mental state of the user is not known and so the
+% system must be inferred from observing the user's actions in real
+% time. For example, a user directed to ``turn around and walk through
+% the door'' may not necessarily perform these actions to the letter,
+% i.e., immediately turning 180 degrees and proceeding directly to the
+% door. Instead, the user might take a roundabout route through the
+% room, eventually exiting out the door.  Although the user's actions do
+% not match the generated instructions exactly, they meet the intended
+% goal. The system must be able to identify such ``equivalent'' plans
+% and not immediately generate new instructions as soon as the user's
+% actions have gone off course. Furthermore, a user can communicate
+% certain intentions to the system, both through action and
+% inaction. For instance, the system should infer that a user has failed
+% to follow instructions if the user exits a room when given a directive
+% to ``walk to the centre of the room''. The system should also make a
+% similar conclusion if a user simply does nothing when given the
+% instruction.
 
@@
 
+\todo{planning claims general applicability; it's clear that IPC
+  benchmarks focus areas of improvement; but it's not so clear to what
+  extent this limits usefulness for non-IPC application areas; this
+  paper evaluates this (for the first time??)}
+
 \section*{Acknowledgements}

### 2009-05-07 trunk/Papers/ci-spark/abstract.tex
split paper up into individual files; changes to abstract and intro

%%%
%%% Abstract
%%%
\begin{abstract}
  Natural language generation (NLG) is a field of computational
  linguistics that has a long tradition as an application area of
  planning systems. While this connection had not received much
  attention for a while, interest in it has been renewed recently in
  several publications. In this paper, we investigate the extent to
  which these new NLG systems profit from the recent advances in
  planner efficiency. Our findings are mixed: While modern planners do
  very well on solving the search problems that come up in our NLG
  experiments, their overall runtime is dominated by the grounding
  step they perform as preprocessing. Because this makes current
  off-the-shelf planners unusably slow for nontrivial NLG problem
  instances, we offer our domains as challenges for the planning
  community.
\end{abstract}

\bigskip\noindent
\textbf{Keywords:} natural language generation, planning

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "manuscript"
%%% End:

### 2009-05-07 trunk/Papers/ci-spark/conclusion.tex
split paper up into individual files; changes to abstract and intro

--- 2009-04-27 trunk/Papers/ci-spark/manuscript.tex
@@
-%%%
-%%% Computational Intelligence manuscript based on SPARK'08 paper
-%%%
-%%% Alexander Koller and Ronald Petrick
-%%% 
-%%% Updated: 2008/12/22
-%%%
-\documentclass[letterpaper]{article}
-\usepackage{times}
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage[round]{natbib}
-\usepackage{setspace}
-\usepackage[paper=letterpaper,margin=1.0in]{geometry}
-
-\title{Experiences with Planning for Natural Language Generation}
-\author{\textbf{Alexander Koller} \\
-Exzellenzcluster / FR 4.7 Computerlinguistik \\
-Saarland University \\
-Postfach 151150 \\
-66041 Saarbrücken \\
-Germany \\
-Phone: +49 681 302 70040 \\
-Fax: +49 681 302 4351 \\
-E-mail: \url{koller@mmci.uni-saarland.de}
-\and
-\textbf{Ronald P. A. Petrick} \\
-School of Informatics \\
-University of Edinburgh \\
-Informatics Forum, 10 Crichton Street \\
-Edinburgh \ EH8 9AB \\
-Scotland, United Kingdom \\
-Phone: +44 131 650 4426 \\
-Fax: +44 131 650 6899 \\
-E-mail: \url{rpetrick@inf.ed.ac.uk}}
-
-%\date{22 December 2008}
-\date{}
-
-\pagestyle{plain}
-\pagenumbering{arabic}
-\doublespacing
-
-\bibpunct{(}{)}{;}{a}{}{,}
-
-\begin{document}
-
-\maketitle
-
-%%%
-%%% Abstract
-%%%
-\pagebreak
-\begin{abstract}
-We investigate the application of modern planning techniques to domains
-arising from problems in natural language generation (NLG). In particular,
-we consider two novel NLG-inspired planning problems, the sentence
-generation domain and the GIVE (``Generating Instructions in Virtual
-Environment'') domain, and investigate the efficiency of the FF and SGPLAN
-planners in these domains. We also compare our results against an ad-hoc
-implementation of GraphPlan in Java. Our results are mixed. While modern
-planners are able to quickly solve many moderately-sized instances of our
-problems, the overall planning time is dominated by the grounding step that
-these planners perform, rather than search. This has a pronounced effect on
-our domains which require relatively short plans but have large universes.
-We share our experiences and offer these domains as challenges for the
-planning community.
-\end{abstract}
-
-\bigskip\noindent
-\textbf{Keywords:} natural language generation, planning
-
-%%%
-%%% Main text
-%%%
-\pagebreak
-\section{Introduction}
-\label{sec:introduction}
-
-Natural language generation (NLG; \citealp{reiter00building}) is one of the
-major subfields of natural language processing, concerned with computing
-natural language sentences or texts that convey a given piece of
-information to the user. Intuitively, this task can be viewed as a problem
-involving actions, beliefs, and goals: an agent communicating with another
-agent tries to change the mental state of the hearer by applying actions
-which correspond to the utterance of words or sentences.
-This characterization of NLG suggests obvious parallels to automated
-planning---a view which has a long tradition in NLG.
-
-Early approaches to generation as planning
-\citep{perrault80,appelt:planning,hovy88,young94dpocl} focused primarily on
-planning high-level structures, such as speech acts and discourse
-relations.  While these approaches suffered from the inefficiency of the
-planners that were available at the time, there has recently been a
-resurgence of interest in applying modern planning techniques to NLG.
-These newer papers revisit speech act planning based on the more expressive
-planning formalisms available today \citep{Steedman-Petrick:07,benotti08b}
-and exploit the improved efficiency of modern planners for an integrated
-model of sentence planning and surface realization \citep{KolSto07}.
-Problems in NLG are often computationally challenging (see, e.g.,
-\citep{KolStr02}), and these new approaches combine an interest in planning
-as a modelling tool for natural language, with a hope of improved
-efficiency by exploiting modern approaches and algorithms.
-
-The focus of this paper is twofold. First, we present two novel planning
-domains that arise in the context of natural language generation: the
-sentence generation domain and the GIVE domain. In the sentence generation
-task, the goal is to generate a single sentence that expresses a given
-meaning. \citet{KolSto07} cast this domain as a planning problem, where the
-plan encodes the necessary sentence and the actions correspond to uttering
-individual words.  In the GIVE domain (``Generating Instructions in Virtual
-Environments''), we describe a new shared task that was recently posed as a
-challenge for the NLG community \citep{alexander07:_shared_task_propos}.
-GIVE uses planning as part of an NLG system that generates natural-language
-instructions to guide a user in performing a given task in a virtual
-environment.
-
-Second, we discuss some of our experiences using off-the-shelf
-planners in our two NLG-inspired domains. In particular, we explore the
-efficiency of FF \citep{HoffmannNebel01} and SGPLAN
-\citep{hsu06:_new_featur_in_sgplan_for}: two planners that are readily
-available, support an expressive subset of the Planning Domain Definition
-Language (PDDL; McDermott et al.~\citeyear{PDDL}) for encoding domains, and
-have been successful on traditional benchmarks and problems from the
-International Planning Competition (IPC)\footnote{See
- \url{http://ipc.icaps-conference.org/} for information about past
- editions of the IPC.}.
-We test these planners on a range of problem instances in our planning
-domains, and compare our results against an ad-hoc Java implementation of
-GraphPlan \citep{Blum1997}. Overall, our findings are mixed. On the one
-hand, we demonstrate that modern planners can easily handle some of the
-moderately-sized search problems that arise in NLG. In certain cases, the
-planners we tested are more efficient than specialized NLG algorithms
-applied to the same problem. On the other hand, these same planners spend
-tremendous amounts of time on preprocessing. For instance, in the sentence
-generation domain, FF spends 90\% of its runtime grounding out literals and
-actions, and ends up being slower than our version of GraphPlan that avoids
-grounding. For domains like ours, which are dominated by the number of
-actions and the universe size, rather than the combinatorics of the search
-problem, this observation suggests that the overall runtime of a planner
-could be improved by grounding more selectively. This is particularly true
-for large problem instances, which still remain a challenge for the
-planners we tested, and where the effects of the grounding approach become
-quite pronounced. We therefore offer our NLG domains as challenges for the
-planning community.
-
-\section{Planning in NLG}
-\label{sec:domains}
-
-We begin by presenting our two planning domains: the sentence generation
-domain and the instruction giving domain.
-
-\subsection{Sentence generation as planning}
-
-Sentence generation is the problem of computing from a grammar and a
-semantic representation, a single sentence that expresses this piece of
-meaning. Traditionally, this problem is split into several steps
-\citep{reiter00building}. In a first step, called \emph{sentence planning},
-the semantic representation is enriched with additional information. For
-instance, \emph{referring expressions}, which refer to individuals (e.g.,
-objects, people, etc.) in the domain of discourse, are determined at this
-point. In a second step, called \emph{surface realization}, this enriched
-representation is translated into a natural-language sentence, using the
-grammar.
-
-In practice, the process of determining referring expressions typically
-interacts with the realization step and so it often turns out to be
-beneficial to perform these two steps together. For instance, the SPUD
-system (``Sentence Planning Using Description''; \citealt{Stone2003a})
-performed this task using a top-down generation algorithm based on
-tree-adjoining grammars \citep{joshi;etal1997} whose lexical entries were
-equipped with semantic and pragmatic information. Unfortunately, SPUD had
-to resort to a non-optimal greedy search strategy to cope with its large
-search space and retain reasonable efficiency. To improve the efficiency
-of SPUD, \citet{KolSto07} translate the sentence generation problem into a
-planning problem and use a planning algorithm to perform the generation
-task.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{pic-grammar}
-  \caption{An example grammar in the sentence generation domain.}
-  \label{fig:white-rabbit-sleeps-grammar}
-\end{figure}
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{pic-derivation-2}
-  \caption{Derivation of ``The white rabbit sleeps.''}
-  \label{fig:white-rabbit-sleeps-deriv}
-\end{figure}
-
-We illustrate this process on a simplified example.  Consider a
-knowledge base containing the individuals $e$, $r_1$ and $r_2$, and a
-set of attributes encoding the fact that $r_1$ and $r_2$ are rabbits,
-$r_1$ is white and $r_2$ is brown, and $e$ is an event in which $r_1$
-sleeps.  Say that we want to express the information
-$\{\mathsf{sleep}(e,r_1)\}$ using the tree-adjoining grammar shown in
-Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
-\emph{elementary trees} (i.e., the disjoint trees in the figure), each
-of which contributes certain \emph{semantic content}. We can
-instantiate these trees by substituting individuals for \emph{semantic
-  roles}, such as $\mathsf{self}$ and $\mathsf{subj}$, and then
-combine the tree instances as shown in
-Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
-``The white rabbit sleeps''.
-
-We compute this grammatical derivation in a top-down manner, starting with
-the elementary tree for ``sleeps''. This tree satisfies the need to convey
-the semantic information, but introduces a need to generate a noun phrase
-(NP) for the subject; this NP must refer uniquely to the target referent
-$r_1$. In a second step, we substitute the tree for ``the rabbit'' into the
-open NP leaf, which makes the derivation grammatically complete. Since
-there are two different individuals that could be described as ``the
-rabbit''---technically, $r_2$ is still a \emph{distractor} (i.e., based on
-the description ``the rabbit'', the hearer might erroneously think that
-we're talking about $r_2$ and not $r_1$)---we are still not finished. To
-complete the derivation, the tree for ``white'' is added to the existing
-structure by an \emph{adjunction} operation, making the derivation
-both syntactically and semantically complete.
-
-\begin{figure}[p]
-\centering
-\begin{minipage}{0.5\textwidth}
-{\small%
-\begin{verbatim}
-(:action add-sleeps
-   :parameters (?u - node
-                ?xself - individual
-                ?xsubj - individual)
-   :precondition
-       (and (subst S ?u)
-            (referent ?u ?xself)
-            (sleep ?xself ?xsubj))
-   :effect 
-       (and (not (subst S ?u))
-            (expressed sleep ?xself ?xsubj)
-            (subst NP (subj ?u))
-            (referent (subj ?u) ?xsubj)
-            (forall (?y - individual)
-                (when (not (= ?y ?xself))
-                    (distractor (subj ?u) ?y)))))
-
-(:action add-rabbit
-   :parameters (?u - node
-                ?xself - individual)
-   :precondition 
-       (and (subst NP ?u)
-            (referent ?u ?xself)
-            (rabbit ?xself))
-   :effect 
-       (and (not (subst NP ?u))
-            (canadjoin N ?u)
-            (forall (?y - individual)
-                (when (not (rabbit ?y))
-                    (not (distractor ?u ?y))))))
-
-(:action add-white
-   :parameters (?u - node
-                ?xself - individual)
-   :precondition 
-       (and (canadjoin N ?u)
-            (referent ?u ?xself)
-            (rabbit ?xself))
-   :effect 
-       (forall (?y - individual)
-           (when (not (white ?y))
-               (not (distractor ?u ?y)))))
-\end{verbatim}}%
-\end{minipage}
-\caption{PDDL actions for generating the sentence ``The white rabbit
-sleeps.''}
-\label{fig:white-rabbit-as-planning}
-\end{figure}
-
-The process described above has clear parallels to planning: we manipulate
-states by applying sequences of actions to achieve a goal. We can make this
-connection even more precise by translating the SPUD problem into a
-planning problem.\footnote{See
-  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
-  which this conversion is implemented.}  For instance,
-Figure~\ref{fig:white-rabbit-as-planning} shows the corresponding PDDL
-actions for the above generation task, where each action corresponds
-to an operation that adds a single elementary tree to the derivation.
-In each case, the first parameter of the action is a node name in the
-derivation tree, and the remaining parameters stand for the
-individuals to which the semantic roles will be instantiated.  The
-syntactic preconditions and effects are encoded using $\mathsf{open}$
-and $\mathsf{canadjoin}$ literals; the status of each referring expression
-is tracked using $\mathsf{distractor}$ literals. The resulting action
-effects contain terms of the form $\mathsf{subj}(u)$, which construct new
-node names, but aren't strictly permitted by PDDL. In order to meet the
-syntactic requirements of PDDL, these terms can be eliminated by estimating
-an upper bound $n$ for the plan length, making $n$ copies of each action,
-ensuring that copy $i$ can only be applied in step $i$, and replacing the
-term $\mathsf{subj}(u)$ in an action copy by the constant
-$\mathsf{subj}_i$.
-
-Once the appropriate actions are defined, we can solve the generation
-problem as an ordinary planning problem. For instance, the following plan
-solves our previous example:
-%
-\begin{enumerate}
-\item $\mathsf{add}\textsf{-}\mathsf{sleeps}(\mathsf{root}, r_1)$,
-\item $\mathsf{add}\textsf{-}\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}),r_1)$,
-\item $\mathsf{add}\textsf{-}\mathsf{white}(\mathsf{subj}(\mathsf{root}),r_1)$.
-\end{enumerate}
-%
-Using this plan, the grammatical derivation in
-Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
-generated sentence ``the white rabbit sleeps'', can be systematically
-reconstructed. Thus, we can solve the sentence generation problem via
-the detour through planning and bring current search heuristics for
-planning to bear on generation.
-
-\subsection{Planning in instruction giving}
-
-The object of the GIVE Challenge (``Generating Instructions in Virtual
-Environments''; \citealt{alexander07:_shared_task_propos}) is to build an
-NLG system which is able to produce natural-language instructions which
-will guide a human user in performing some task in a virtual environment.
-From an NLG perspective, GIVE makes for an interesting challenge because it
-is a theory-neutral task that exercises all components of an NLG system,
-and emphasizes the study of communication in a (simulated) physical
-environment. It also has the advantage that the user and the NLG system can
-be in physically different places, as long as the 3D client and the NLG
-system are connected over a network. This makes it possible to evaluate
-GIVE NLG systems on a large scale over the Internet.
-%
-The first GIVE evaluation is currently underway. Five NLG systems from
-three countries are in the process of being evaluated. At the time of
-writing, 1028 different evaluation runs---each consisting of one user
-interacting with one instance of the virtual environment and being guided
-by an NLG system---have been collected, making GIVE the largest-scale
-NLG evaluation effort ever.\footnote{See the GIVE website at
- \url{http://www.give-challenge.org/} for more details about this
- challenge.}
-
-\begin{figure}[t]
-\centering
-%\includegraphics[width=1 \columnwidth]{give_world_no_expl}
-\includegraphics[width=0.75\columnwidth]{give_world_2}
-\caption{Map of the GIVE development world.}
-  \label{fig:give-development-world}
-\end{figure}
-
-An example GIVE world map is shown in
-Figure~\ref{fig:give-development-world}. To simplify both the planning and
-the NLG task, the world is discretised into a set of tiles of equal size.
-The user can turn by 90 degree steps in either direction, and can move from
-the centre of one tile to the centre of the next tile, provided the path
-between two tiles is not blocked. The world also consists of a set of
-objects which can be manipulated by the user in various ways. For
-instance, in the example world the user's task is to pick up a trophy in
-the top left room. The trophy is hidden in a safe behind a picture, so the
-user must first move the picture out of the way and open the safe by
-pushing a sequence of buttons (the small square boxes on the walls) in the
-correct order. In order to get to all these buttons, the user must also
-open the door in the centre of the map and deactivate the alarm tile by
-pushing further buttons. Figure~\ref{fig:give-planning} shows the encoding
-of some of the available GIVE domain actions, in PDDL syntax.
-
-\begin{figure}[p]
-\centering
-\begin{minipage}{0.5\textwidth}
-{\small%
-\begin{verbatim}
-(:action move
-   :parameters (?from - position
-                ?to - position
-                ?ori - orientation)
-   :precondition 
-       (and (player-pos ?from) 
-            (adjacent ?from ?to ?ori) 
-            (player-orient ?ori)
-            (not-blocked ?from ?to)
-            (not-alarmed ?to))
-   :effect 
-       (and (not (player-pos ?from))
-            (player-pos ?to)))
-
-(:action turn-left
-   :parameters (?ori - orientation
-                ?newOri - orientation)
-   :precondition 
-       (and (player-orient ?ori)
-            (next-orient-left ?ori ?newOri))
-   :effect 
-       (and (not (player-orient ?ori))
-            (player-orient ?newOri)))
-
-(:action turn-right
-   :parameters (?ori - orientation
-                ?newOri - orientation)
-   :precondition 
-       (and (player-orient ?ori)
-            (next-orient-right ?ori ?newOri))
-   :effect 
-       (and (not (player-orient ?ori))
-            (player-orient ?newOri)))
-
-(:action manipulate-b1-off-on
-   :parameters (?pos - position)
-   :precondition 
-       (and (state b1 off)
-            (player-pos ?pos)
-            (position b1 ?pos))
-   :effect 
-       (and (not (state b1 off))
-            (state b1 on)
-            (not (state d1 closed))
-            (state d1 open) 
-            (not (blocked pos_6_5 pos_6_4))
-            (not (blocked pos_6_4 pos_6_5))))
-\end{verbatim}}%
-\end{minipage}
-\caption{PDDL actions for the GIVE domain.}
-\label{fig:give-planning}
-\end{figure}
-
-The plans generated in the GIVE domain are often nontrivial. For instance,
-in the above example the shortest plan consists of 108 action steps, with
-the first few steps as follows:
-%
-\begin{enumerate}
-\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north},
-\mathsf{west})$,
-\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
-\end{enumerate}
-
-Most of the actions in the above plan are ``move'' actions. This makes the
-bulk of the GIVE problem very similar to that of a Gridworld problem (see,
-e.g., \citep{Tovey-Koenig:2000}), which also involves finding a route
-through a two-dimensional world map with discrete positions. GIVE tasks are
-usually much more complex, however, as illustrated by the above example
-map: a successful plan must include steps for pressing buttons in the right
-order, reasoning about large numbers of world objects, and navigating
-through complicated room shapes.
-
-The GIVE task comes with a number of interesting research challenges for
-both the NLG and planning communities. On the NLG side, one major problem
-is that verbalising each action instance in a plan makes for very
-poor natural-language instructions. For instance, the NLG system could
-generate an instruction sequence starting with ``turn left; walk forward;
-press the button; turn right; walk forward; walk forward; turn right; walk
-forward; walk forward; turn left; walk forward''. However, directions of
-this form are linguistically clumsy and uninteresting, and it is to be
-expected that frustrated users will quickly cancel an interaction with such
-a system, resulting in a bad evaluation score. Having just pressed the
-button, it would be much better to say ``turn around and walk through the
-door''. That is, the system should \emph{summarise} plans by merging
-multiple action instances into single instructions before presenting them
-to the user.\footnote{This task is similar to the construction of
-  ``macro'' actions, but is primarily intended for plan presentation
-  purposes. See, e.g., \citep{Botea-etal:05} which contains a survey of
-  recent approaches for generating macro actions in planning.}
-Conversely, it may also be necessary to \emph{elaborate} on a single
-planning step by expressing it with several instructions. Having just
-entered the top left room, it may be easier for the user to understand the
-instruction ``walk to the centre of the room; turn right; now press the
-green button in front of you'', rather than the instruction ``press the
-green button on the wall to your right''. Furthermore, in order to plan the
-referring expression ``the green button in front of you'' at a time when
-the button is not yet in front of the user, the NLG system must keep track
-of the hypothetical changes in what objects are visible to the user. Thus,
-the NLG acts of referring and instructing must be tightly integrated with
-the modules for plan generation.
-
-From a planning perspective, GIVE imposes very strict runtime
-requirements on the plan generation task: planning must
-happen in real time and the system must respond to a user in a timely
-fashion.  If the system takes too much time deliberating over an
-instruction to give, rather than actually giving this instruction, the
-user may have walked or turned away, thus making the instruction
-invalid.  Furthermore, plan execution monitoring also plays an
-important role in the GIVE problem.  At a high level, the system needs
-to monitor a user's actions and compare them against the generated
-instruction set to determine if the user has correctly followed
-directions or not. In the case of the latter, new instructions may
-have to be generated. In practice, the situation can be quite
-complicated since the mental state of the user is not known and so the
-system must be inferred from observing the user's actions in real
-time. For example, a user directed to ``turn around and walk through
-the door'' may not necessarily perform these actions to the letter,
-i.e., immediately turning 180 degrees and proceeding directly to the
-door. Instead, the user might take a roundabout route through the
-room, eventually exiting out the door.  Although the user's actions do
-not match the generated instructions exactly, they meet the intended
-goal. The system must be able to identify such ``equivalent'' plans
-and not immediately generate new instructions as soon as the user's
-actions have gone off course. Furthermore, a user can communicate
-certain intentions to the system, both through action and
-inaction. For instance, the system should infer that a user has failed
-to follow instructions if the user exits a room when given a directive
-to ``walk to the centre of the room''. The system should also make a
-similar conclusion if a user simply does nothing when given the
-instruction.
-
-\section{Experiments}
-\label{sec:experiments}
-
-We now describe the results of four simple experiments designed to evaluate
-the performance of three planners---FF \citep{HoffmannNebel01}, SGPLAN
-5.2.2 \citep{hsu06:_new_featur_in_sgplan_for}, and an ad-hoc implementation
-of GraphPlan \citep{Blum1997} written in Java---in our NLG domains. The FF
-and SGPLAN planners were chosen due to their success in the International
-Planning Competitions. Our implementation of GraphPlan is included as a
-counterpoint to the other two, more complex, planners. Most notably, our
-version of GraphPlan does not perform a preprocessing ``grounding'' step
-like FF and SGPLAN but is much more selective in its predicate and action
-instantiations.
-
-On first impressions, our results indicate that planning is a promising
-tool for both domains.  In the sentence generation domain, FF dramatically
-outperforms the best previously known algorithm for the same problem (a
-reimplementation of \citep{Stone2003a}), although the latter approach uses
-a greedy search algorithm with a heuristic that is hand-tailored to the
-domain. In the GIVE domain, SGPLAN computes a domain plan from the initial
-state to the goal in 0.3 seconds, which is fast enough for moderately-sized
-problems instances in the application.\footnote{All
-  runtimes were measured on a Pentium 4 CPU running at 3 GHz. Java programs
-  were allowed to ``warm up'', i.e., the planner was run five times and the
-  first four measurements discarded to ensure that the JVM had just-in-time
-  compiled all relevant bytecode.}
-
-We also observe that although FF and SGPLAN manage the search for plans
-very efficiently, they both spend comparatively large amounts of time
-computing predicate and action instantiations in their preprocessing
-stages, many of which are then never used during the search. FF's
-``grounding time'' in particular dominates its overall planning time,
-leading to some conflicting results. In some cases, our Java implementation
-of GraphPlan significantly outperforms both state-of-the-art planners. The
-experiments described below seek to improve our understanding of this
-situation.
-
-\subsection{Experiment 1: Sentence generation}
-\label{sec:exper-1:-sent}
-
-In the first experiment, we construct a series of sentence generation
-problems which require the planner to compute a plan representing the
-sentence ``Mary likes the Adj$_1$ \ldots Adj$_n$ rabbit.''  Each problem
-instance assumes a certain number $m$ of rabbits that were distinguished by
-$n \leq m$ different properties, such that all $n$ properties are required
-to distinguish the target referent from all other rabbits.  The $n$
-properties are realized as $n$ different adjectives, in any order.  This
-setup allows us to control the plan length (a plan with $n$ properties will
-have length $n+4$) and the universe size (the universe will contain $m+1$
-individuals in addition to the differently-typed individuals used to encode
-the grammar).
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp1}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-modifiers-with-sgplan}
-  \caption{Results for the sentence generation domain. The
-    horizontal axis represents parameters $(m,n)$ from $(1,1)$ to
-    $(10,10)$ in lexicographical order. The vertical axis is the
-    runtime in milliseconds.}
-  \label{fig:runtimes-crisp}
-\end{figure}
-
-The results of this experiment are shown in the graph in
-Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
-in lexicographic order on the horizontal axis and the runtime is shown in
-milliseconds on the vertical axis. These results reveal a number of
-interesting insights. First, FF significantly outperforms SGPLAN in this
-domain, sometimes by a factor of 10 or more.\footnote{Experiments with
- SGPLAN use a pre-release version of SGPLAN, kindly provided to us by
- Chih-Wei Hsu.The release version of SGPLAN 5.2.2 had a bug which caused it
- to crash on some of our problem instances.}
-Second, FF's runtime is dominated by its initial grounding step, in which
-it computes the ground instances of all predicates and actions used in the
-planning problem in order to avoid unnecessary instantiation during search.
-In particular, the ratio of grounding time to total runtime is generally
-above 85\%, and rises to above 99\% at $m=11$, which is still a relatively
-small universe for this application.\footnote{The
-  ``grounding'' time reported here is what FF reports as ``time spent:
-  instantiating action templates''.}
-
-In our testing examples, the time spent by FF on grounding is such that it
-is consistently outperformed by our Java implementation of GraphPlan, which
-only computes instances of predicates and actions as they are discovered
-during the construction of the planning graph. Although FF is consistently
-much faster as far as pure search time is concerned, our results indicate
-that FF's performance is much more sensitive to the domain size: if we fix
-$n=1$, FF takes 60 ms to compute a plan at $m=1$, but 2.4 seconds to
-compute the same plan at $m=10$. By comparison, our GraphPlan
-implementation takes 30 ms at $m=1$ and still only requires 50 ms at
-$m=10$. Conversely, GraphPlan's runtime grows much faster with the plan
-size (i.e., with growing values of $n$ for a fixed $m$). Larger, but still
-realistically-sized, instances of the sentence generation problem are still
-problematic for all the planners we tested.
-
-\subsection{Experiment 2: Minimal GIVE worlds}
-\label{sec:exper-2:-minim}
-
-In the second experiment, we evaluate the performance of the planners on
-problems arising in the GIVE domain. We construct a series of test worlds,
-similar to the one illustrated in Figure~\ref{fig:give-minimal}. These
-worlds consist of a $2n$ by $h$ grid of positions, such that there are
-buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The
-player starts in position $(1,1)$ and must press all the buttons
-to successfully complete the game. The world is generated as a GIVE world
-description, and then automatically converted into a planning problem by
-the GIVE software. For instance, Figure~\ref{fig:give-planning} shows some
-of the actions available in the GIVE domain in PDDL syntax.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.5\columnwidth]{pic-buttons}
-  \caption{Minimal GIVE world.}
-  \label{fig:give-minimal}
-\end{figure}
-
-Results for the $h=20$ case, with $n$ ranging from $1$ to $40$, are shown
-in Figure~\ref{fig:give-runtime-minimal}. The most obvious result is that
-FF is unable to solve any problems beyond $n=13$ on our experimentation
-machine within the memory limit of 1 GB. SGPLAN, on the other hand, solves
-instances beyond $n=40$ without major problems. In this case, our
-implementation of GraphPlan is unable to find any of these plans. The time
-spent on grounding is not a major factor in either planner, probably
-because the planners need more time to actually compute the plan---for
-instance, the optimal plan for the problem $n=40$ has a length of about
-1600 steps. As a concrete example, the following is a minimal plan for the
-case of a $2$ by $2$ grid with $2$ buttons for the player to press (i.e.,
-$n=1$, $h=2$):
-
-\begin{enumerate}
-\item $\mathsf{move}(\mathsf{pos\_1\_1},\mathsf{pos\_1\_2}, \mathsf{north})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{u1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_1\_2})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{north}, \mathsf{east})$,
-\item $\mathsf{move}(\mathsf{pos\_1\_2}, \mathsf{pos\_2\_2},
-  \mathsf{east})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{east}, \mathsf{south})$,
-\item $\mathsf{move}(\mathsf{pos\_2\_2}, \mathsf{pos\_2\_1}, \mathsf{south})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{l1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_2\_1})$.
-\end{enumerate}
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp2}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-buttons}
-  \caption{Results on the minimal GIVE
-    domain for $h=20$. The horizontal axis is $n$. The vertical axis
-    is the runtime in seconds.}
-  \label{fig:give-runtime-minimal}
-\end{figure}
-
-\subsection{Experiment 3: GIVE worlds with extra positions}
-\label{sec:experiment-3:-give}
-
-In the third experiment, we vary the structure of the GIVE world in order
-to judge the effect that universe size has on the planning problem in this
-domain. Starting with the ordinary GIVE world described in Experiment~2, we
-extend the world map by adding another $w$ by $h$ empty ``junk'' positions
-to the right of the minimal world, as shown in Figure~\ref{fig:give-junk}.
-These new positions are not actually required in any plan, but extend the
-size of the state space and approximate the situation in the actual GIVE
-domain where most grid positions are never used. We leave the initial state
-and goal untouched. As before, we generate a GIVE world description and
-then convert it into a planning problem in PDDL.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.80\columnwidth]{pic-empty-buttons}
-  \caption{GIVE world with extra ``junk'' positions.}
-  \label{fig:give-junk}
-\end{figure}
-
-Results for the $h=20$, $n=5$ case with $w$ ranging from $1$ to $70$ are
-shown in Figure~\ref{fig:give-runtime-junk}. As in Experiment~2, FF again
-runs out of memory, this time at $w=17$, while SGPLAN easily solves inputs
-beyond $w=70$. As in Experiment~2, our implementation of GraphPlan is
-unable to generate any of these plans. Unlike Experiment 2, however, both
-FF and SGPLAN now spend a substantial proportion of their time on
-grounding. In SGPLAN, this translates to a ``parsing time'' (which we
-assume includes grounding) which grows from 180 ms to 21.7 seconds as $w$
-grows from $1$ to $75$. The rest of the runtime, which also includes the
-search time, only grows from 400 ms to 2.3 seconds. This difference is
-particularly dramatic given that the actual optimal plan in each case is an
-identical plan of about 100 steps (i.e., the one we would have found in
-Experiment 2 for the $h=20$, $n=5$ case). The planning times for these
-instances are also concerning since times over a couple seconds will
-negatively affect the overall response time of the system, which must react
-in real time to user actions.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp3}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
-  \caption{Results on the GIVE domain with junk
-    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
-    The vertical axis is the runtime in seconds.}
-  \label{fig:give-runtime-junk}
-\end{figure}
-
-\subsection{Experiment 4: GIVE worlds with inaccessible positions}
-\label{sec:experiment-4:-give}
-
-In the final experiment, we consider a variation on the GIVE worlds from
-Experiment~3. As before, we begin with a base grid of $2n$ by $h$
-positions, with buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i
-\leq n$. We then generate a second grid of $w$ by $h$ positions and place
-an additional button in this grid. Unlike Experiment~3, the second grid
-does not extend the first grid but is ``disconnected'' so its positions are
-inaccessible from the first grid, as shown in
-Figure~\ref{fig:give-junk-nosoln}. Since the geometry of the grid makes it
-impossible to construct a plan for pressing all the buttons in the world,
-we instead investigate the time it takes a planner to arrive at the
-conclusion that the problem cannot be solved.
-
-Results for the $h=20$, $n=5$ case with $w$ ranging from 1 to 50 are shown
-in Figure~\ref{fig:give-runtime-nosoln}.\footnote{FF and SGPLAN do
- not display runtimes when they fail to construct a plan so an external timing
- program was used to generate these results.}
-In this case, the runtimes for FF rise sharply around $w=28$, producing
-times that are unacceptable for an NLG system in practice. FF is unable
-to complete problem instances above $w=35$. SGPLAN, by comparison, performs
-exceptionally well and is able to complete problems instances up to $w=50$
-in about 0.3 seconds and $w=80$ in about 0.6 seconds. These results
-demonstrate that SGPLAN has a significant advantage over FF in this
-experiment. 
-%Our implementation of GraphPlan is unable to complete any of
-%these problem instances.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.85\columnwidth]{pic-empty-inaccessible}
-  \caption{GIVE world with inaccessible positions.}
-  \label{fig:give-junk-nosoln}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.85\columnwidth]{graph-exp4}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
-  \caption{Results on the GIVE domain with inaccessible
-    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
-    The vertical axis is the runtime in seconds.}
-  \label{fig:give-runtime-nosoln}
-\end{figure}
-
-\section{Discussion}
-\label{sec:discussion}
-
-The positive conclusion we draw from our simple set of experiments is that
-modern planners are fairly good at controlling the search for many of the
-moderately-sized NLG problems we looked at. This is particularly true for
-FF in the sentence generation domain, which generates nontrivial 14-word
-sentences in about two seconds. Our results also indicate there is still
-room for improvement, however: special-purpose algorithms are able to
-generate much larger referring expressions in milliseconds
-\citep{AreKolStr08}. Although search efficiency is not the main concern in
-these domains, it nevertheless becomes a factor in scenarios like
-Experiment~2, where the world is almost completely unconstrained and
-(unsurprisingly) the search becomes much slower.
-
-The most restrictive bottleneck in the two NLG domains we investigated is
-the initial grounding step that many modern planners perform. While this
-may be a good strategy for traditional planning benchmarks and IPC
-domains---where plan length often dominates universe size and an initial
-investment into grounding pays off in saved instantiations during
-search---this approach has a very pronounced effect in domains like ours.
-As our experiments show, there are natural planning domains in which
-relatively short plans must be computed in large universes. For instance,
-in the sentence generation domain, it is not unrealistic to look for plans
-of length 20 over a universe consisting of thousands of domain individuals
-and tens of thousands of possible action instantiations, some of which
-require three domain individuals as parameters. Planners like FF and SGPLAN
-are less effective in such domains, to the point that grounding
-often dominates the total running time of these planners. (In the case of
-Experiment~1, our ad-hoc implementation of GraphPlan outperforms both
-planners in the sentence generation task.)
-
-Since our domains are not that unusual in their structure and composition,
-we hope that the lessons learned from our experiences can help improve the
-performance of current planning systems. For instance, the common strategy
-of splitting the universe into several types of individuals will not be a
-very effective solution in our NLG domains given the large number of
-individuals that could exist for a given type. On the other hand,
-sophisticated techniques based on reachability analysis may have the
-potential to improve this situation, provided such approaches can avoid the
-drawbacks of unnecessary grounding.\footnote{See
-  \citep{Bryce:07} for a good survey on reachability heuristics for planning
-  graph-based approaches to plan generation.}
-While recent trends in planning research have (rightfully) focused on the
-development of algorithms that control search in sophisticated ways,
-resulting in state-of-the-art planners that are more powerful and
-successful than their predecessors, we nevertheless believe that
-improvements are necessary if planning is to become a more mature
-technology that can offer tools to a wider community of users and
-researchers. At the end of the day, real-world users will care about the
-\emph{total} runtime of a planner, which is more than just the search time.
-
-By and large, our experiences with the planning community from a user's
-point of view have been favourable. Thanks to the planning competitions, it
-is easy to identify and download a fast state-of-the-art planner, at least
-for Linux. However, in the course of our experiments we found and reported
-bugs in both FF and SGPLAN. We also discovered that deciding as to which
-planner is appropriate for a particular task is not always straightforward,
-as the best performing planner in the IPC is not necessarily suitable for
-every planning task. As our experiments have shown, even planners as
-closely related as FF and SGPLAN can differ significantly in their
-performance on different domains (for instance, our results in
-Experiment~4).
-
 \section{Conclusion}
@@
 
-\bibliography{references}
-\bibliographystyle{ci}
-
-\pagebreak
-\listoffigures
-
-\end{document}
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "manuscript"
+%%% End:

### 2009-05-07 trunk/Papers/ci-spark/discussion.tex
split paper up into individual files; changes to abstract and intro

--- 2009-04-27 trunk/Papers/ci-spark/manuscript.tex
@@
-%%%
-%%% Computational Intelligence manuscript based on SPARK'08 paper
-%%%
-%%% Alexander Koller and Ronald Petrick
-%%% 
-%%% Updated: 2008/12/22
-%%%
-\documentclass[letterpaper]{article}
-\usepackage{times}
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage[round]{natbib}
-\usepackage{setspace}
-\usepackage[paper=letterpaper,margin=1.0in]{geometry}
-
-\title{Experiences with Planning for Natural Language Generation}
-\author{\textbf{Alexander Koller} \\
-Exzellenzcluster / FR 4.7 Computerlinguistik \\
-Saarland University \\
-Postfach 151150 \\
-66041 Saarbrücken \\
-Germany \\
-Phone: +49 681 302 70040 \\
-Fax: +49 681 302 4351 \\
-E-mail: \url{koller@mmci.uni-saarland.de}
-\and
-\textbf{Ronald P. A. Petrick} \\
-School of Informatics \\
-University of Edinburgh \\
-Informatics Forum, 10 Crichton Street \\
-Edinburgh \ EH8 9AB \\
-Scotland, United Kingdom \\
-Phone: +44 131 650 4426 \\
-Fax: +44 131 650 6899 \\
-E-mail: \url{rpetrick@inf.ed.ac.uk}}
-
-%\date{22 December 2008}
-\date{}
-
-\pagestyle{plain}
-\pagenumbering{arabic}
-\doublespacing
-
-\bibpunct{(}{)}{;}{a}{}{,}
-
-\begin{document}
-
-\maketitle
-
-%%%
-%%% Abstract
-%%%
-\pagebreak
-\begin{abstract}
-We investigate the application of modern planning techniques to domains
-arising from problems in natural language generation (NLG). In particular,
-we consider two novel NLG-inspired planning problems, the sentence
-generation domain and the GIVE (``Generating Instructions in Virtual
-Environment'') domain, and investigate the efficiency of the FF and SGPLAN
-planners in these domains. We also compare our results against an ad-hoc
-implementation of GraphPlan in Java. Our results are mixed. While modern
-planners are able to quickly solve many moderately-sized instances of our
-problems, the overall planning time is dominated by the grounding step that
-these planners perform, rather than search. This has a pronounced effect on
-our domains which require relatively short plans but have large universes.
-We share our experiences and offer these domains as challenges for the
-planning community.
-\end{abstract}
-
-\bigskip\noindent
-\textbf{Keywords:} natural language generation, planning
-
-%%%
-%%% Main text
-%%%
-\pagebreak
-\section{Introduction}
-\label{sec:introduction}
-
-Natural language generation (NLG; \citealp{reiter00building}) is one of the
-major subfields of natural language processing, concerned with computing
-natural language sentences or texts that convey a given piece of
-information to the user. Intuitively, this task can be viewed as a problem
-involving actions, beliefs, and goals: an agent communicating with another
-agent tries to change the mental state of the hearer by applying actions
-which correspond to the utterance of words or sentences.
-This characterization of NLG suggests obvious parallels to automated
-planning---a view which has a long tradition in NLG.
-
-Early approaches to generation as planning
-\citep{perrault80,appelt:planning,hovy88,young94dpocl} focused primarily on
-planning high-level structures, such as speech acts and discourse
-relations.  While these approaches suffered from the inefficiency of the
-planners that were available at the time, there has recently been a
-resurgence of interest in applying modern planning techniques to NLG.
-These newer papers revisit speech act planning based on the more expressive
-planning formalisms available today \citep{Steedman-Petrick:07,benotti08b}
-and exploit the improved efficiency of modern planners for an integrated
-model of sentence planning and surface realization \citep{KolSto07}.
-Problems in NLG are often computationally challenging (see, e.g.,
-\citep{KolStr02}), and these new approaches combine an interest in planning
-as a modelling tool for natural language, with a hope of improved
-efficiency by exploiting modern approaches and algorithms.
-
-The focus of this paper is twofold. First, we present two novel planning
-domains that arise in the context of natural language generation: the
-sentence generation domain and the GIVE domain. In the sentence generation
-task, the goal is to generate a single sentence that expresses a given
-meaning. \citet{KolSto07} cast this domain as a planning problem, where the
-plan encodes the necessary sentence and the actions correspond to uttering
-individual words.  In the GIVE domain (``Generating Instructions in Virtual
-Environments''), we describe a new shared task that was recently posed as a
-challenge for the NLG community \citep{alexander07:_shared_task_propos}.
-GIVE uses planning as part of an NLG system that generates natural-language
-instructions to guide a user in performing a given task in a virtual
-environment.
-
-Second, we discuss some of our experiences using off-the-shelf
-planners in our two NLG-inspired domains. In particular, we explore the
-efficiency of FF \citep{HoffmannNebel01} and SGPLAN
-\citep{hsu06:_new_featur_in_sgplan_for}: two planners that are readily
-available, support an expressive subset of the Planning Domain Definition
-Language (PDDL; McDermott et al.~\citeyear{PDDL}) for encoding domains, and
-have been successful on traditional benchmarks and problems from the
-International Planning Competition (IPC)\footnote{See
- \url{http://ipc.icaps-conference.org/} for information about past
- editions of the IPC.}.
-We test these planners on a range of problem instances in our planning
-domains, and compare our results against an ad-hoc Java implementation of
-GraphPlan \citep{Blum1997}. Overall, our findings are mixed. On the one
-hand, we demonstrate that modern planners can easily handle some of the
-moderately-sized search problems that arise in NLG. In certain cases, the
-planners we tested are more efficient than specialized NLG algorithms
-applied to the same problem. On the other hand, these same planners spend
-tremendous amounts of time on preprocessing. For instance, in the sentence
-generation domain, FF spends 90\% of its runtime grounding out literals and
-actions, and ends up being slower than our version of GraphPlan that avoids
-grounding. For domains like ours, which are dominated by the number of
-actions and the universe size, rather than the combinatorics of the search
-problem, this observation suggests that the overall runtime of a planner
-could be improved by grounding more selectively. This is particularly true
-for large problem instances, which still remain a challenge for the
-planners we tested, and where the effects of the grounding approach become
-quite pronounced. We therefore offer our NLG domains as challenges for the
-planning community.
-
-\section{Planning in NLG}
-\label{sec:domains}
-
-We begin by presenting our two planning domains: the sentence generation
-domain and the instruction giving domain.
-
-\subsection{Sentence generation as planning}
-
-Sentence generation is the problem of computing from a grammar and a
-semantic representation, a single sentence that expresses this piece of
-meaning. Traditionally, this problem is split into several steps
-\citep{reiter00building}. In a first step, called \emph{sentence planning},
-the semantic representation is enriched with additional information. For
-instance, \emph{referring expressions}, which refer to individuals (e.g.,
-objects, people, etc.) in the domain of discourse, are determined at this
-point. In a second step, called \emph{surface realization}, this enriched
-representation is translated into a natural-language sentence, using the
-grammar.
-
-In practice, the process of determining referring expressions typically
-interacts with the realization step and so it often turns out to be
-beneficial to perform these two steps together. For instance, the SPUD
-system (``Sentence Planning Using Description''; \citealt{Stone2003a})
-performed this task using a top-down generation algorithm based on
-tree-adjoining grammars \citep{joshi;etal1997} whose lexical entries were
-equipped with semantic and pragmatic information. Unfortunately, SPUD had
-to resort to a non-optimal greedy search strategy to cope with its large
-search space and retain reasonable efficiency. To improve the efficiency
-of SPUD, \citet{KolSto07} translate the sentence generation problem into a
-planning problem and use a planning algorithm to perform the generation
-task.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{pic-grammar}
-  \caption{An example grammar in the sentence generation domain.}
-  \label{fig:white-rabbit-sleeps-grammar}
-\end{figure}
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{pic-derivation-2}
-  \caption{Derivation of ``The white rabbit sleeps.''}
-  \label{fig:white-rabbit-sleeps-deriv}
-\end{figure}
-
-We illustrate this process on a simplified example.  Consider a
-knowledge base containing the individuals $e$, $r_1$ and $r_2$, and a
-set of attributes encoding the fact that $r_1$ and $r_2$ are rabbits,
-$r_1$ is white and $r_2$ is brown, and $e$ is an event in which $r_1$
-sleeps.  Say that we want to express the information
-$\{\mathsf{sleep}(e,r_1)\}$ using the tree-adjoining grammar shown in
-Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
-\emph{elementary trees} (i.e., the disjoint trees in the figure), each
-of which contributes certain \emph{semantic content}. We can
-instantiate these trees by substituting individuals for \emph{semantic
-  roles}, such as $\mathsf{self}$ and $\mathsf{subj}$, and then
-combine the tree instances as shown in
-Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
-``The white rabbit sleeps''.
-
-We compute this grammatical derivation in a top-down manner, starting with
-the elementary tree for ``sleeps''. This tree satisfies the need to convey
-the semantic information, but introduces a need to generate a noun phrase
-(NP) for the subject; this NP must refer uniquely to the target referent
-$r_1$. In a second step, we substitute the tree for ``the rabbit'' into the
-open NP leaf, which makes the derivation grammatically complete. Since
-there are two different individuals that could be described as ``the
-rabbit''---technically, $r_2$ is still a \emph{distractor} (i.e., based on
-the description ``the rabbit'', the hearer might erroneously think that
-we're talking about $r_2$ and not $r_1$)---we are still not finished. To
-complete the derivation, the tree for ``white'' is added to the existing
-structure by an \emph{adjunction} operation, making the derivation
-both syntactically and semantically complete.
-
-\begin{figure}[p]
-\centering
-\begin{minipage}{0.5\textwidth}
-{\small%
-\begin{verbatim}
-(:action add-sleeps
-   :parameters (?u - node
-                ?xself - individual
-                ?xsubj - individual)
-   :precondition
-       (and (subst S ?u)
-            (referent ?u ?xself)
-            (sleep ?xself ?xsubj))
-   :effect 
-       (and (not (subst S ?u))
-            (expressed sleep ?xself ?xsubj)
-            (subst NP (subj ?u))
-            (referent (subj ?u) ?xsubj)
-            (forall (?y - individual)
-                (when (not (= ?y ?xself))
-                    (distractor (subj ?u) ?y)))))
-
-(:action add-rabbit
-   :parameters (?u - node
-                ?xself - individual)
-   :precondition 
-       (and (subst NP ?u)
-            (referent ?u ?xself)
-            (rabbit ?xself))
-   :effect 
-       (and (not (subst NP ?u))
-            (canadjoin N ?u)
-            (forall (?y - individual)
-                (when (not (rabbit ?y))
-                    (not (distractor ?u ?y))))))
-
-(:action add-white
-   :parameters (?u - node
-                ?xself - individual)
-   :precondition 
-       (and (canadjoin N ?u)
-            (referent ?u ?xself)
-            (rabbit ?xself))
-   :effect 
-       (forall (?y - individual)
-           (when (not (white ?y))
-               (not (distractor ?u ?y)))))
-\end{verbatim}}%
-\end{minipage}
-\caption{PDDL actions for generating the sentence ``The white rabbit
-sleeps.''}
-\label{fig:white-rabbit-as-planning}
-\end{figure}
-
-The process described above has clear parallels to planning: we manipulate
-states by applying sequences of actions to achieve a goal. We can make this
-connection even more precise by translating the SPUD problem into a
-planning problem.\footnote{See
-  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
-  which this conversion is implemented.}  For instance,
-Figure~\ref{fig:white-rabbit-as-planning} shows the corresponding PDDL
-actions for the above generation task, where each action corresponds
-to an operation that adds a single elementary tree to the derivation.
-In each case, the first parameter of the action is a node name in the
-derivation tree, and the remaining parameters stand for the
-individuals to which the semantic roles will be instantiated.  The
-syntactic preconditions and effects are encoded using $\mathsf{open}$
-and $\mathsf{canadjoin}$ literals; the status of each referring expression
-is tracked using $\mathsf{distractor}$ literals. The resulting action
-effects contain terms of the form $\mathsf{subj}(u)$, which construct new
-node names, but aren't strictly permitted by PDDL. In order to meet the
-syntactic requirements of PDDL, these terms can be eliminated by estimating
-an upper bound $n$ for the plan length, making $n$ copies of each action,
-ensuring that copy $i$ can only be applied in step $i$, and replacing the
-term $\mathsf{subj}(u)$ in an action copy by the constant
-$\mathsf{subj}_i$.
-
-Once the appropriate actions are defined, we can solve the generation
-problem as an ordinary planning problem. For instance, the following plan
-solves our previous example:
-%
-\begin{enumerate}
-\item $\mathsf{add}\textsf{-}\mathsf{sleeps}(\mathsf{root}, r_1)$,
-\item $\mathsf{add}\textsf{-}\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}),r_1)$,
-\item $\mathsf{add}\textsf{-}\mathsf{white}(\mathsf{subj}(\mathsf{root}),r_1)$.
-\end{enumerate}
-%
-Using this plan, the grammatical derivation in
-Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
-generated sentence ``the white rabbit sleeps'', can be systematically
-reconstructed. Thus, we can solve the sentence generation problem via
-the detour through planning and bring current search heuristics for
-planning to bear on generation.
-
-\subsection{Planning in instruction giving}
-
-The object of the GIVE Challenge (``Generating Instructions in Virtual
-Environments''; \citealt{alexander07:_shared_task_propos}) is to build an
-NLG system which is able to produce natural-language instructions which
-will guide a human user in performing some task in a virtual environment.
-From an NLG perspective, GIVE makes for an interesting challenge because it
-is a theory-neutral task that exercises all components of an NLG system,
-and emphasizes the study of communication in a (simulated) physical
-environment. It also has the advantage that the user and the NLG system can
-be in physically different places, as long as the 3D client and the NLG
-system are connected over a network. This makes it possible to evaluate
-GIVE NLG systems on a large scale over the Internet.
-%
-The first GIVE evaluation is currently underway. Five NLG systems from
-three countries are in the process of being evaluated. At the time of
-writing, 1028 different evaluation runs---each consisting of one user
-interacting with one instance of the virtual environment and being guided
-by an NLG system---have been collected, making GIVE the largest-scale
-NLG evaluation effort ever.\footnote{See the GIVE website at
- \url{http://www.give-challenge.org/} for more details about this
- challenge.}
-
-\begin{figure}[t]
-\centering
-%\includegraphics[width=1 \columnwidth]{give_world_no_expl}
-\includegraphics[width=0.75\columnwidth]{give_world_2}
-\caption{Map of the GIVE development world.}
-  \label{fig:give-development-world}
-\end{figure}
-
-An example GIVE world map is shown in
-Figure~\ref{fig:give-development-world}. To simplify both the planning and
-the NLG task, the world is discretised into a set of tiles of equal size.
-The user can turn by 90 degree steps in either direction, and can move from
-the centre of one tile to the centre of the next tile, provided the path
-between two tiles is not blocked. The world also consists of a set of
-objects which can be manipulated by the user in various ways. For
-instance, in the example world the user's task is to pick up a trophy in
-the top left room. The trophy is hidden in a safe behind a picture, so the
-user must first move the picture out of the way and open the safe by
-pushing a sequence of buttons (the small square boxes on the walls) in the
-correct order. In order to get to all these buttons, the user must also
-open the door in the centre of the map and deactivate the alarm tile by
-pushing further buttons. Figure~\ref{fig:give-planning} shows the encoding
-of some of the available GIVE domain actions, in PDDL syntax.
-
-\begin{figure}[p]
-\centering
-\begin{minipage}{0.5\textwidth}
-{\small%
-\begin{verbatim}
-(:action move
-   :parameters (?from - position
-                ?to - position
-                ?ori - orientation)
-   :precondition 
-       (and (player-pos ?from) 
-            (adjacent ?from ?to ?ori) 
-            (player-orient ?ori)
-            (not-blocked ?from ?to)
-            (not-alarmed ?to))
-   :effect 
-       (and (not (player-pos ?from))
-            (player-pos ?to)))
-
-(:action turn-left
-   :parameters (?ori - orientation
-                ?newOri - orientation)
-   :precondition 
-       (and (player-orient ?ori)
-            (next-orient-left ?ori ?newOri))
-   :effect 
-       (and (not (player-orient ?ori))
-            (player-orient ?newOri)))
-
-(:action turn-right
-   :parameters (?ori - orientation
-                ?newOri - orientation)
-   :precondition 
-       (and (player-orient ?ori)
-            (next-orient-right ?ori ?newOri))
-   :effect 
-       (and (not (player-orient ?ori))
-            (player-orient ?newOri)))
-
-(:action manipulate-b1-off-on
-   :parameters (?pos - position)
-   :precondition 
-       (and (state b1 off)
-            (player-pos ?pos)
-            (position b1 ?pos))
-   :effect 
-       (and (not (state b1 off))
-            (state b1 on)
-            (not (state d1 closed))
-            (state d1 open) 
-            (not (blocked pos_6_5 pos_6_4))
-            (not (blocked pos_6_4 pos_6_5))))
-\end{verbatim}}%
-\end{minipage}
-\caption{PDDL actions for the GIVE domain.}
-\label{fig:give-planning}
-\end{figure}
-
-The plans generated in the GIVE domain are often nontrivial. For instance,
-in the above example the shortest plan consists of 108 action steps, with
-the first few steps as follows:
-%
-\begin{enumerate}
-\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north},
-\mathsf{west})$,
-\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
-\end{enumerate}
-
-Most of the actions in the above plan are ``move'' actions. This makes the
-bulk of the GIVE problem very similar to that of a Gridworld problem (see,
-e.g., \citep{Tovey-Koenig:2000}), which also involves finding a route
-through a two-dimensional world map with discrete positions. GIVE tasks are
-usually much more complex, however, as illustrated by the above example
-map: a successful plan must include steps for pressing buttons in the right
-order, reasoning about large numbers of world objects, and navigating
-through complicated room shapes.
-
-The GIVE task comes with a number of interesting research challenges for
-both the NLG and planning communities. On the NLG side, one major problem
-is that verbalising each action instance in a plan makes for very
-poor natural-language instructions. For instance, the NLG system could
-generate an instruction sequence starting with ``turn left; walk forward;
-press the button; turn right; walk forward; walk forward; turn right; walk
-forward; walk forward; turn left; walk forward''. However, directions of
-this form are linguistically clumsy and uninteresting, and it is to be
-expected that frustrated users will quickly cancel an interaction with such
-a system, resulting in a bad evaluation score. Having just pressed the
-button, it would be much better to say ``turn around and walk through the
-door''. That is, the system should \emph{summarise} plans by merging
-multiple action instances into single instructions before presenting them
-to the user.\footnote{This task is similar to the construction of
-  ``macro'' actions, but is primarily intended for plan presentation
-  purposes. See, e.g., \citep{Botea-etal:05} which contains a survey of
-  recent approaches for generating macro actions in planning.}
-Conversely, it may also be necessary to \emph{elaborate} on a single
-planning step by expressing it with several instructions. Having just
-entered the top left room, it may be easier for the user to understand the
-instruction ``walk to the centre of the room; turn right; now press the
-green button in front of you'', rather than the instruction ``press the
-green button on the wall to your right''. Furthermore, in order to plan the
-referring expression ``the green button in front of you'' at a time when
-the button is not yet in front of the user, the NLG system must keep track
-of the hypothetical changes in what objects are visible to the user. Thus,
-the NLG acts of referring and instructing must be tightly integrated with
-the modules for plan generation.
-
-From a planning perspective, GIVE imposes very strict runtime
-requirements on the plan generation task: planning must
-happen in real time and the system must respond to a user in a timely
-fashion.  If the system takes too much time deliberating over an
-instruction to give, rather than actually giving this instruction, the
-user may have walked or turned away, thus making the instruction
-invalid.  Furthermore, plan execution monitoring also plays an
-important role in the GIVE problem.  At a high level, the system needs
-to monitor a user's actions and compare them against the generated
-instruction set to determine if the user has correctly followed
-directions or not. In the case of the latter, new instructions may
-have to be generated. In practice, the situation can be quite
-complicated since the mental state of the user is not known and so the
-system must be inferred from observing the user's actions in real
-time. For example, a user directed to ``turn around and walk through
-the door'' may not necessarily perform these actions to the letter,
-i.e., immediately turning 180 degrees and proceeding directly to the
-door. Instead, the user might take a roundabout route through the
-room, eventually exiting out the door.  Although the user's actions do
-not match the generated instructions exactly, they meet the intended
-goal. The system must be able to identify such ``equivalent'' plans
-and not immediately generate new instructions as soon as the user's
-actions have gone off course. Furthermore, a user can communicate
-certain intentions to the system, both through action and
-inaction. For instance, the system should infer that a user has failed
-to follow instructions if the user exits a room when given a directive
-to ``walk to the centre of the room''. The system should also make a
-similar conclusion if a user simply does nothing when given the
-instruction.
-
-\section{Experiments}
-\label{sec:experiments}
-
-We now describe the results of four simple experiments designed to evaluate
-the performance of three planners---FF \citep{HoffmannNebel01}, SGPLAN
-5.2.2 \citep{hsu06:_new_featur_in_sgplan_for}, and an ad-hoc implementation
-of GraphPlan \citep{Blum1997} written in Java---in our NLG domains. The FF
-and SGPLAN planners were chosen due to their success in the International
-Planning Competitions. Our implementation of GraphPlan is included as a
-counterpoint to the other two, more complex, planners. Most notably, our
-version of GraphPlan does not perform a preprocessing ``grounding'' step
-like FF and SGPLAN but is much more selective in its predicate and action
-instantiations.
-
-On first impressions, our results indicate that planning is a promising
-tool for both domains.  In the sentence generation domain, FF dramatically
-outperforms the best previously known algorithm for the same problem (a
-reimplementation of \citep{Stone2003a}), although the latter approach uses
-a greedy search algorithm with a heuristic that is hand-tailored to the
-domain. In the GIVE domain, SGPLAN computes a domain plan from the initial
-state to the goal in 0.3 seconds, which is fast enough for moderately-sized
-problems instances in the application.\footnote{All
-  runtimes were measured on a Pentium 4 CPU running at 3 GHz. Java programs
-  were allowed to ``warm up'', i.e., the planner was run five times and the
-  first four measurements discarded to ensure that the JVM had just-in-time
-  compiled all relevant bytecode.}
-
-We also observe that although FF and SGPLAN manage the search for plans
-very efficiently, they both spend comparatively large amounts of time
-computing predicate and action instantiations in their preprocessing
-stages, many of which are then never used during the search. FF's
-``grounding time'' in particular dominates its overall planning time,
-leading to some conflicting results. In some cases, our Java implementation
-of GraphPlan significantly outperforms both state-of-the-art planners. The
-experiments described below seek to improve our understanding of this
-situation.
-
-\subsection{Experiment 1: Sentence generation}
-\label{sec:exper-1:-sent}
-
-In the first experiment, we construct a series of sentence generation
-problems which require the planner to compute a plan representing the
-sentence ``Mary likes the Adj$_1$ \ldots Adj$_n$ rabbit.''  Each problem
-instance assumes a certain number $m$ of rabbits that were distinguished by
-$n \leq m$ different properties, such that all $n$ properties are required
-to distinguish the target referent from all other rabbits.  The $n$
-properties are realized as $n$ different adjectives, in any order.  This
-setup allows us to control the plan length (a plan with $n$ properties will
-have length $n+4$) and the universe size (the universe will contain $m+1$
-individuals in addition to the differently-typed individuals used to encode
-the grammar).
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp1}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-modifiers-with-sgplan}
-  \caption{Results for the sentence generation domain. The
-    horizontal axis represents parameters $(m,n)$ from $(1,1)$ to
-    $(10,10)$ in lexicographical order. The vertical axis is the
-    runtime in milliseconds.}
-  \label{fig:runtimes-crisp}
-\end{figure}
-
-The results of this experiment are shown in the graph in
-Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
-in lexicographic order on the horizontal axis and the runtime is shown in
-milliseconds on the vertical axis. These results reveal a number of
-interesting insights. First, FF significantly outperforms SGPLAN in this
-domain, sometimes by a factor of 10 or more.\footnote{Experiments with
- SGPLAN use a pre-release version of SGPLAN, kindly provided to us by
- Chih-Wei Hsu.The release version of SGPLAN 5.2.2 had a bug which caused it
- to crash on some of our problem instances.}
-Second, FF's runtime is dominated by its initial grounding step, in which
-it computes the ground instances of all predicates and actions used in the
-planning problem in order to avoid unnecessary instantiation during search.
-In particular, the ratio of grounding time to total runtime is generally
-above 85\%, and rises to above 99\% at $m=11$, which is still a relatively
-small universe for this application.\footnote{The
-  ``grounding'' time reported here is what FF reports as ``time spent:
-  instantiating action templates''.}
-
-In our testing examples, the time spent by FF on grounding is such that it
-is consistently outperformed by our Java implementation of GraphPlan, which
-only computes instances of predicates and actions as they are discovered
-during the construction of the planning graph. Although FF is consistently
-much faster as far as pure search time is concerned, our results indicate
-that FF's performance is much more sensitive to the domain size: if we fix
-$n=1$, FF takes 60 ms to compute a plan at $m=1$, but 2.4 seconds to
-compute the same plan at $m=10$. By comparison, our GraphPlan
-implementation takes 30 ms at $m=1$ and still only requires 50 ms at
-$m=10$. Conversely, GraphPlan's runtime grows much faster with the plan
-size (i.e., with growing values of $n$ for a fixed $m$). Larger, but still
-realistically-sized, instances of the sentence generation problem are still
-problematic for all the planners we tested.
-
-\subsection{Experiment 2: Minimal GIVE worlds}
-\label{sec:exper-2:-minim}
-
-In the second experiment, we evaluate the performance of the planners on
-problems arising in the GIVE domain. We construct a series of test worlds,
-similar to the one illustrated in Figure~\ref{fig:give-minimal}. These
-worlds consist of a $2n$ by $h$ grid of positions, such that there are
-buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The
-player starts in position $(1,1)$ and must press all the buttons
-to successfully complete the game. The world is generated as a GIVE world
-description, and then automatically converted into a planning problem by
-the GIVE software. For instance, Figure~\ref{fig:give-planning} shows some
-of the actions available in the GIVE domain in PDDL syntax.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.5\columnwidth]{pic-buttons}
-  \caption{Minimal GIVE world.}
-  \label{fig:give-minimal}
-\end{figure}
-
-Results for the $h=20$ case, with $n$ ranging from $1$ to $40$, are shown
-in Figure~\ref{fig:give-runtime-minimal}. The most obvious result is that
-FF is unable to solve any problems beyond $n=13$ on our experimentation
-machine within the memory limit of 1 GB. SGPLAN, on the other hand, solves
-instances beyond $n=40$ without major problems. In this case, our
-implementation of GraphPlan is unable to find any of these plans. The time
-spent on grounding is not a major factor in either planner, probably
-because the planners need more time to actually compute the plan---for
-instance, the optimal plan for the problem $n=40$ has a length of about
-1600 steps. As a concrete example, the following is a minimal plan for the
-case of a $2$ by $2$ grid with $2$ buttons for the player to press (i.e.,
-$n=1$, $h=2$):
-
-\begin{enumerate}
-\item $\mathsf{move}(\mathsf{pos\_1\_1},\mathsf{pos\_1\_2}, \mathsf{north})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{u1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_1\_2})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{north}, \mathsf{east})$,
-\item $\mathsf{move}(\mathsf{pos\_1\_2}, \mathsf{pos\_2\_2},
-  \mathsf{east})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{east}, \mathsf{south})$,
-\item $\mathsf{move}(\mathsf{pos\_2\_2}, \mathsf{pos\_2\_1}, \mathsf{south})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{l1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_2\_1})$.
-\end{enumerate}
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp2}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-buttons}
-  \caption{Results on the minimal GIVE
-    domain for $h=20$. The horizontal axis is $n$. The vertical axis
-    is the runtime in seconds.}
-  \label{fig:give-runtime-minimal}
-\end{figure}
-
-\subsection{Experiment 3: GIVE worlds with extra positions}
-\label{sec:experiment-3:-give}
-
-In the third experiment, we vary the structure of the GIVE world in order
-to judge the effect that universe size has on the planning problem in this
-domain. Starting with the ordinary GIVE world described in Experiment~2, we
-extend the world map by adding another $w$ by $h$ empty ``junk'' positions
-to the right of the minimal world, as shown in Figure~\ref{fig:give-junk}.
-These new positions are not actually required in any plan, but extend the
-size of the state space and approximate the situation in the actual GIVE
-domain where most grid positions are never used. We leave the initial state
-and goal untouched. As before, we generate a GIVE world description and
-then convert it into a planning problem in PDDL.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.80\columnwidth]{pic-empty-buttons}
-  \caption{GIVE world with extra ``junk'' positions.}
-  \label{fig:give-junk}
-\end{figure}
-
-Results for the $h=20$, $n=5$ case with $w$ ranging from $1$ to $70$ are
-shown in Figure~\ref{fig:give-runtime-junk}. As in Experiment~2, FF again
-runs out of memory, this time at $w=17$, while SGPLAN easily solves inputs
-beyond $w=70$. As in Experiment~2, our implementation of GraphPlan is
-unable to generate any of these plans. Unlike Experiment 2, however, both
-FF and SGPLAN now spend a substantial proportion of their time on
-grounding. In SGPLAN, this translates to a ``parsing time'' (which we
-assume includes grounding) which grows from 180 ms to 21.7 seconds as $w$
-grows from $1$ to $75$. The rest of the runtime, which also includes the
-search time, only grows from 400 ms to 2.3 seconds. This difference is
-particularly dramatic given that the actual optimal plan in each case is an
-identical plan of about 100 steps (i.e., the one we would have found in
-Experiment 2 for the $h=20$, $n=5$ case). The planning times for these
-instances are also concerning since times over a couple seconds will
-negatively affect the overall response time of the system, which must react
-in real time to user actions.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp3}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
-  \caption{Results on the GIVE domain with junk
-    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
-    The vertical axis is the runtime in seconds.}
-  \label{fig:give-runtime-junk}
-\end{figure}
-
-\subsection{Experiment 4: GIVE worlds with inaccessible positions}
-\label{sec:experiment-4:-give}
-
-In the final experiment, we consider a variation on the GIVE worlds from
-Experiment~3. As before, we begin with a base grid of $2n$ by $h$
-positions, with buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i
-\leq n$. We then generate a second grid of $w$ by $h$ positions and place
-an additional button in this grid. Unlike Experiment~3, the second grid
-does not extend the first grid but is ``disconnected'' so its positions are
-inaccessible from the first grid, as shown in
-Figure~\ref{fig:give-junk-nosoln}. Since the geometry of the grid makes it
-impossible to construct a plan for pressing all the buttons in the world,
-we instead investigate the time it takes a planner to arrive at the
-conclusion that the problem cannot be solved.
-
-Results for the $h=20$, $n=5$ case with $w$ ranging from 1 to 50 are shown
-in Figure~\ref{fig:give-runtime-nosoln}.\footnote{FF and SGPLAN do
- not display runtimes when they fail to construct a plan so an external timing
- program was used to generate these results.}
-In this case, the runtimes for FF rise sharply around $w=28$, producing
-times that are unacceptable for an NLG system in practice. FF is unable
-to complete problem instances above $w=35$. SGPLAN, by comparison, performs
-exceptionally well and is able to complete problems instances up to $w=50$
-in about 0.3 seconds and $w=80$ in about 0.6 seconds. These results
-demonstrate that SGPLAN has a significant advantage over FF in this
-experiment. 
-%Our implementation of GraphPlan is unable to complete any of
-%these problem instances.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.85\columnwidth]{pic-empty-inaccessible}
-  \caption{GIVE world with inaccessible positions.}
-  \label{fig:give-junk-nosoln}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.85\columnwidth]{graph-exp4}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
-  \caption{Results on the GIVE domain with inaccessible
-    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
-    The vertical axis is the runtime in seconds.}
-  \label{fig:give-runtime-nosoln}
-\end{figure}
-
 \section{Discussion}
@@
 
-\section{Conclusion}
-\label{sec:conclusion}
-
-In this paper, we introduced two novel planning domains arising from
-problems in natural language generation: the sentence generation domain and
-the Generating Instructions in Virtual Environments (GIVE) challenge
-domain. We also reported the results of a set of experiments designed to
-evaluate the suitability of off-the-shelf planners (particularly FF and
-SGPLAN) in these domains.
-
-Our results were mixed. While the planners we tested do an impressive job
-of controlling the complexity of search in many domains, they also suffer
-from practical problems that limit their performance in unexpected ways: in
-both of our NLG domains the grounding step performed by FF and SGPLAN
-dominates the time it takes to perform the search itself. As future work,
-we hope to extend our analysis to other planners that do not perform the
-initial grounding step, planners such as SHOP2
-\citep{DBLP:journals/jair/NauAIKMWY03} that have traditionally performed
-well on a wide range of domains, and new planners introduced in the latest
-edition of the IPC.\footnote{See
- \url{http://ipc.informatik.uni-freiburg.de/} for details of the 2008
- International Planning Competition (deterministic track).}
-
-The NLG community's recent interest in planning presents a valuable
-opportunity for planning researchers. Domains like GIVE highlight certain
-challenges, such as plan execution monitoring and plan presentation (i.e.,
-summarisation and elaboration), but also offer a platform on which such
-technologies can be evaluated in experiments with human users. Although we
-have focused on classical planning problems in this work, research related
-to reasoning under uncertainty, resource management, and planning with
-knowledge and sensing, can also be investigated in these settings. As such,
-we believe our domains would provide suitable challenges for planners
-entered in future editions of the IPC. Provided some of the challenges we
-have identified can be addressed, NLG-inspired problem domains offer a
-constructive platform for the planning community to showcase their
-techniques to a wider audience---and to improve the quality of their tools
-for real-world planning tasks.
-
-\section*{Acknowledgements}
-
-This work arose in the context of the Planning and Language Interest Group
-at the University of Edinburgh. The authors would like to thank all members
-of this group, especially Hector Geffner and Mark Steedman, for interesting
-discussions. This work was partially supported by the DFG Research
-Fellowship ``CRISP: Efficient integrated realization and microplanning''
-and by the European Commission through the PACO-PLUS project
-(FP6-2004-IST-4-27657).
-
-\bibliography{references}
-\bibliographystyle{ci}
-
-\pagebreak
-\listoffigures
-
-\end{document}
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "manuscript"
+%%% End:

### 2009-05-07 trunk/Papers/ci-spark/domains.tex
split paper up into individual files; changes to abstract and intro

--- 2009-05-04 trunk/Papers/ci-spark/manuscript.tex
@@
-%%%
-%%% Computational Intelligence manuscript based on SPARK'08 paper
-%%%
-%%% Alexander Koller and Ronald Petrick
-%%% 
-%%% Updated: 2008/12/22
-%%%
-\documentclass[letterpaper]{article}
-\usepackage{times}
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage[round]{natbib}
-\usepackage{setspace}
-\usepackage[paper=letterpaper,margin=1.0in]{geometry}
-
-\title{Experiences with Planning for Natural Language Generation}
-\author{\textbf{Alexander Koller} \\
-Exzellenzcluster / FR 4.7 Computerlinguistik \\
-Saarland University \\
-Postfach 151150 \\
-66041 Saarbr\"ucken \\
-Germany \\
-Phone: +49 681 302 70040 \\
-Fax: +49 681 302 4351 \\
-E-mail: \url{koller@mmci.uni-saarland.de}
-\and
-\textbf{Ronald P. A. Petrick} \\
-School of Informatics \\
-University of Edinburgh \\
-Informatics Forum, 10 Crichton Street \\
-Edinburgh \ EH8 9AB \\
-Scotland, United Kingdom \\
-Phone: +44 131 650 4426 \\
-Fax: +44 131 650 6899 \\
-E-mail: \url{rpetrick@inf.ed.ac.uk}}
-
-%\date{22 December 2008}
-\date{}
-
-\pagestyle{plain}
-\pagenumbering{arabic}
-\doublespacing
-
-\bibpunct{(}{)}{;}{a}{}{,}
-
-\newcommand{\todo}[1]{\textbf{(#1)}}
-
-\begin{document}
-
-\maketitle
-
-%%%
-%%% Abstract
-%%%
-\pagebreak
-\begin{abstract}
-We investigate the application of modern planning techniques to domains
-arising from problems in natural language generation (NLG). In particular,
-we consider two novel NLG-inspired planning problems, the sentence
-generation domain and the GIVE (``Generating Instructions in Virtual
-Environment'') domain, and investigate the efficiency of the FF and SGPLAN
-planners in these domains. We also compare our results against an ad-hoc
-implementation of GraphPlan in Java. Our results are mixed. While modern
-planners are able to quickly solve many moderately-sized instances of our
-problems, the overall planning time is dominated by the grounding step that
-these planners perform, rather than search. This has a pronounced effect on
-our domains which require relatively short plans but have large universes.
-We share our experiences and offer these domains as challenges for the
-planning community.
-\end{abstract}
-
-\bigskip\noindent
-\textbf{Keywords:} natural language generation, planning
-
-%%%
-%%% Main text
-%%%
-\pagebreak
-\section{Introduction}
-\label{sec:introduction}
-
-Natural language generation (NLG; \citealp{reiter00building}) is one
-of the major subfields of natural language processing, concerned with
-computing natural language sentences or texts that convey a given
-piece of information to the user. This task can be viewed as a problem
-of achieving a (communicative) goal by successively applying
-(communicative) actions, and thus has clear intuitive parallels to
-automated planning.
-
-This view of generation as planning has a long tradition in NLG
-\citep{perrault80,appelt:planning,hovy88,young94dpocl}. Despite their
-theoretical attractiveness, these early approaches never gained
-mainstream use in applied systems, in part because of the inefficiency
-of the planners that were available at the time. However, the recent
-improvements in planner efficiency and expressiveness has sparked a
-renewed interest in applying modern planning techniques to NLG---both
-on the level of speech act planning
-\citep{Steedman-Petrick:07,benotti08b} and and on the level of
-sentence generation \citep{KolSto07}. The goal of this paper is to
-determine, by looking at some representative generation problems,
-whether planning has advanced to the point that this enterprise can be
-successful this time---and thus, to determine the usefulness of
-current planning technology to an application area that has a long
-tradition, but is not currently in the focus of the planning
-competitions.
-
-The focus of this paper is twofold. First, we present two generation
-problems that have recently been cast as planning problems: the
-sentence generation task and the GIVE task. In the sentence generation
-task, the goal is to generate a single sentence that expresses a given
-meaning. \citet{KolSto07} cast this task as a planning problem, where
-the plan encodes the necessary sentence and the actions correspond to
-uttering individual words.  In the GIVE domain (``Generating
-Instructions in Virtual Environments''), we describe a new shared task
-that was recently posed as a challenge for the NLG community
-\citep{ByrKolStrCasDalMooObe09}.  GIVE uses planning as part of an NLG
-system that generates natural-language instructions to guide a user in
-performing a given task in a virtual environment.
-
-Second, we discuss some of our experiences using off-the-shelf
-planners in our two NLG-inspired domains. In particular, we explore
-the efficiency of FF \citep{HoffmannNebel01} and SGPLAN
-\citep{hsu06:_new_featur_in_sgplan_for}: two planners that are readily
-available, support an expressive subset of the Planning Domain
-Definition Language (PDDL; McDermott et al.~\citeyear{PDDL}) for
-encoding domains, and have been successful on traditional benchmarks
-and problems from the International Planning Competition
-(IPC)\footnote{See \url{http://ipc.icaps-conference.org/} for
-  information about past editions of the IPC.}.  We test these
-planners on a range of problem instances in our planning domains, and
-compare our results against an ad-hoc Java implementation of GraphPlan
-\citep{Blum1997}. Overall, our findings are mixed. On the one hand, we
-demonstrate that modern planners can easily handle the \emph{search}
-problems that arise in NLG on realistic inputs, which is promising
-given that the sentence generation task, at least, is NP-complete
-\citep{KolStr02}. On the other hand, these same planners spend
-tremendous amounts of time on preprocessing. For instance, in the
-sentence generation domain, FF spends 90\% of its runtime grounding
-out literals and actions, and ends up being slower than our version of
-GraphPlan that avoids grounding. For domains like ours, which are
-dominated by the number of actions and the universe size, rather than
-the combinatorics of the search problem, this observation suggests
-that the overall runtime of a planner could be improved by grounding
-more selectively. This is particularly true for large problem
-instances, which still remain a challenge for the planners we tested,
-and where the effects of the grounding approach become quite
-pronounced. We conclude that planning \emph{algorithms} have advanced
-to the point that generation as planning is now a realistic endeavour,
-but the concrete \emph{implementations} of these algorithms must be
-improved to be useful in this application, and offer our NLG domains
-as challenges for the planning community.
-
-This paper is structured as follows. We begin by briefly reviewing
-some of the literature on NLG as planning
-(Section~\ref{sec:nlg-as-planning}. Then we define introduce the
-planning problems arising in two particular NLG tasks in more detail
-in Section~\ref{sec:domains}. We report on some experiments with these
-planning problems in Section~\ref{sec:experiments}, and discuss the
-results in Section~\ref{sec:discussion}. Section~\ref{sec:conclusion}
-concludes.
-
-\section{NLG as Planning} \label{sec:nlg-as-planning}
-
-The task of generating natural language from semantic representations
-is typically split up into two parts: the \emph{domain planning} task,
-which selects the information that is to be conveyed and structures it
-into sentence-sized chunks, and the \emph{sentence generation} tasks,
-which then translates each of these chunks into natural language
-sentences. The sentence generation task is often split up into two
-parts of its own---the \emph{sentence planning} task, which enriches
-the input by e.g.\ determining how to refer to objects and selects
-some lexical material, and the \emph{surface realization} task, which
-maps the enriched meaning representation into a sentence using a
-grammar. The chain of domain planning, sentence planning, and surface
-realization is sometimes called the ``NLG pipeline''
-\cite{reiter00building}.
-
-Viewing generation as a planning problem has a long tradition in the
-NLG literature, especially in discourse planning. \todo{Sketch
-  Perrault et al here. Say something very brief about Appelt, Hovy,
-  Moore. Say why all this didn't really work.}
-
-Most recently, interest in generation as planning has been revived for
-NLG tasks on various levels of the NLG pipeline. Below, we will look
-in some detail at \cite{KolSto07}'s recasting of sentence generation
-as planning; furthermore, \cite{Steedman-Petrick:07} and
-\cite{benotti08b} have applied planning to \todo{something}. Notice
-that although all of these approaches encode ``communicative actions''
-as planning operators, they still solve fundamentally different
-problems. For instance, Perrault et al.\ tackled discourse planning
-with planning operators that encode speech acts, each of which
-represents the meaning of a whole sentence and models the contribution
-of this sentence to the intentional structure of the entire
-discourse. By contrast, Koller and Stone solve a sentence generation
-problem using planning operators that encode utterances of single
-words and model the contribution of each word to the grammatical
-structure of the sentence. Whether these different levels of ``NLG as
-planning'' could be combined into a single planning problem, and
-whether this would make computational sense, is an open question.
-
-Problems in NLG at all levels of description can become
-computationally complex very easily. For instance, \citet{KolStr02}
-showed that even surface realization is NP-complete under reasonable
-assumptions. This makes it necessary to investigate methods for
-guiding the search, which makes modern planning techniques a promising
-approach to NLG. The point of this paper is to investigate whether
-existing planners fulfill this promise of efficiency for the
-application to NLG.
-
 \section{Two NLG Tasks}
@@
 
-\section{Experiments}
-\label{sec:experiments}
-
-We now describe the results of four simple experiments designed to evaluate
-the performance of three planners---FF \citep{HoffmannNebel01}, SGPLAN
-5.2.2 \citep{hsu06:_new_featur_in_sgplan_for}, and an ad-hoc implementation
-of GraphPlan \citep{Blum1997} written in Java---in our NLG domains. The FF
-and SGPLAN planners were chosen due to their success in the International
-Planning Competitions. Our implementation of GraphPlan is included as a
-counterpoint to the other two, more complex, planners. Most notably, our
-version of GraphPlan does not perform a preprocessing ``grounding'' step
-like FF and SGPLAN but is much more selective in its predicate and action
-instantiations.
-
-On first impressions, our results indicate that planning is a promising
-tool for both domains.  In the sentence generation domain, FF dramatically
-outperforms the best previously known algorithm for the same problem (a
-reimplementation of \citep{Stone2003a}), although the latter approach uses
-a greedy search algorithm with a heuristic that is hand-tailored to the
-domain. In the GIVE domain, SGPLAN computes a domain plan from the initial
-state to the goal in 0.3 seconds, which is fast enough for moderately-sized
-problems instances in the application.\footnote{All
-  runtimes were measured on a Pentium 4 CPU running at 3 GHz. Java programs
-  were allowed to ``warm up'', i.e., the planner was run five times and the
-  first four measurements discarded to ensure that the JVM had just-in-time
-  compiled all relevant bytecode.}
-
-We also observe that although FF and SGPLAN manage the search for plans
-very efficiently, they both spend comparatively large amounts of time
-computing predicate and action instantiations in their preprocessing
-stages, many of which are then never used during the search. FF's
-``grounding time'' in particular dominates its overall planning time,
-leading to some conflicting results. In some cases, our Java implementation
-of GraphPlan significantly outperforms both state-of-the-art planners. The
-experiments described below seek to improve our understanding of this
-situation.
-
-\subsection{Experiment 1: Sentence generation}
-\label{sec:exper-1:-sent}
-
-In the first experiment, we construct a series of sentence generation
-problems which require the planner to compute a plan representing the
-sentence ``Mary likes the Adj$_1$ \ldots Adj$_n$ rabbit.''  Each problem
-instance assumes a certain number $m$ of rabbits that were distinguished by
-$n \leq m$ different properties, such that all $n$ properties are required
-to distinguish the target referent from all other rabbits.  The $n$
-properties are realized as $n$ different adjectives, in any order.  This
-setup allows us to control the plan length (a plan with $n$ properties will
-have length $n+4$) and the universe size (the universe will contain $m+1$
-individuals in addition to the differently-typed individuals used to encode
-the grammar).
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp1}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-modifiers-with-sgplan}
-  \caption{Results for the sentence generation domain. The
-    horizontal axis represents parameters $(m,n)$ from $(1,1)$ to
-    $(10,10)$ in lexicographical order. The vertical axis is the
-    runtime in milliseconds.}
-  \label{fig:runtimes-crisp}
-\end{figure}
-
-The results of this experiment are shown in the graph in
-Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
-in lexicographic order on the horizontal axis and the runtime is shown in
-milliseconds on the vertical axis. These results reveal a number of
-interesting insights. First, FF significantly outperforms SGPLAN in this
-domain, sometimes by a factor of 10 or more.\footnote{Experiments with
- SGPLAN use a pre-release version of SGPLAN, kindly provided to us by
- Chih-Wei Hsu.The release version of SGPLAN 5.2.2 had a bug which caused it
- to crash on some of our problem instances.}
-Second, FF's runtime is dominated by its initial grounding step, in which
-it computes the ground instances of all predicates and actions used in the
-planning problem in order to avoid unnecessary instantiation during search.
-In particular, the ratio of grounding time to total runtime is generally
-above 85\%, and rises to above 99\% at $m=11$, which is still a relatively
-small universe for this application.\footnote{The
-  ``grounding'' time reported here is what FF reports as ``time spent:
-  instantiating action templates''.}
-
-In our testing examples, the time spent by FF on grounding is such that it
-is consistently outperformed by our Java implementation of GraphPlan, which
-only computes instances of predicates and actions as they are discovered
-during the construction of the planning graph. Although FF is consistently
-much faster as far as pure search time is concerned, our results indicate
-that FF's performance is much more sensitive to the domain size: if we fix
-$n=1$, FF takes 60 ms to compute a plan at $m=1$, but 2.4 seconds to
-compute the same plan at $m=10$. By comparison, our GraphPlan
-implementation takes 30 ms at $m=1$ and still only requires 50 ms at
-$m=10$. Conversely, GraphPlan's runtime grows much faster with the plan
-size (i.e., with growing values of $n$ for a fixed $m$). Larger, but still
-realistically-sized, instances of the sentence generation problem are still
-problematic for all the planners we tested.
-
-\subsection{Experiment 2: Minimal GIVE worlds}
-\label{sec:exper-2:-minim}
-
-In the second experiment, we evaluate the performance of the planners on
-problems arising in the GIVE domain. We construct a series of test worlds,
-similar to the one illustrated in Figure~\ref{fig:give-minimal}. These
-worlds consist of a $2n$ by $h$ grid of positions, such that there are
-buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The
-player starts in position $(1,1)$ and must press all the buttons
-to successfully complete the game. The world is generated as a GIVE world
-description, and then automatically converted into a planning problem by
-the GIVE software. For instance, Figure~\ref{fig:give-planning} shows some
-of the actions available in the GIVE domain in PDDL syntax.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.5\columnwidth]{pic-buttons}
-  \caption{Minimal GIVE world.}
-  \label{fig:give-minimal}
-\end{figure}
-
-Results for the $h=20$ case, with $n$ ranging from $1$ to $40$, are shown
-in Figure~\ref{fig:give-runtime-minimal}. The most obvious result is that
-FF is unable to solve any problems beyond $n=13$ on our experimentation
-machine within the memory limit of 1 GB. SGPLAN, on the other hand, solves
-instances beyond $n=40$ without major problems. In this case, our
-implementation of GraphPlan is unable to find any of these plans. The time
-spent on grounding is not a major factor in either planner, probably
-because the planners need more time to actually compute the plan---for
-instance, the optimal plan for the problem $n=40$ has a length of about
-1600 steps. As a concrete example, the following is a minimal plan for the
-case of a $2$ by $2$ grid with $2$ buttons for the player to press (i.e.,
-$n=1$, $h=2$):
-
-\begin{enumerate}
-\item $\mathsf{move}(\mathsf{pos\_1\_1},\mathsf{pos\_1\_2}, \mathsf{north})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{u1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_1\_2})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{north}, \mathsf{east})$,
-\item $\mathsf{move}(\mathsf{pos\_1\_2}, \mathsf{pos\_2\_2},
-  \mathsf{east})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{east}, \mathsf{south})$,
-\item $\mathsf{move}(\mathsf{pos\_2\_2}, \mathsf{pos\_2\_1}, \mathsf{south})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{l1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_2\_1})$.
-\end{enumerate}
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp2}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-buttons}
-  \caption{Results on the minimal GIVE
-    domain for $h=20$. The horizontal axis is $n$. The vertical axis
-    is the runtime in seconds.}
-  \label{fig:give-runtime-minimal}
-\end{figure}
-
-\subsection{Experiment 3: GIVE worlds with extra positions}
-\label{sec:experiment-3:-give}
-
-In the third experiment, we vary the structure of the GIVE world in order
-to judge the effect that universe size has on the planning problem in this
-domain. Starting with the ordinary GIVE world described in Experiment~2, we
-extend the world map by adding another $w$ by $h$ empty ``junk'' positions
-to the right of the minimal world, as shown in Figure~\ref{fig:give-junk}.
-These new positions are not actually required in any plan, but extend the
-size of the state space and approximate the situation in the actual GIVE
-domain where most grid positions are never used. We leave the initial state
-and goal untouched. As before, we generate a GIVE world description and
-then convert it into a planning problem in PDDL.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.80\columnwidth]{pic-empty-buttons}
-  \caption{GIVE world with extra ``junk'' positions.}
-  \label{fig:give-junk}
-\end{figure}
-
-Results for the $h=20$, $n=5$ case with $w$ ranging from $1$ to $70$ are
-shown in Figure~\ref{fig:give-runtime-junk}. As in Experiment~2, FF again
-runs out of memory, this time at $w=17$, while SGPLAN easily solves inputs
-beyond $w=70$. As in Experiment~2, our implementation of GraphPlan is
-unable to generate any of these plans. Unlike Experiment 2, however, both
-FF and SGPLAN now spend a substantial proportion of their time on
-grounding. In SGPLAN, this translates to a ``parsing time'' (which we
-assume includes grounding) which grows from 180 ms to 21.7 seconds as $w$
-grows from $1$ to $75$. The rest of the runtime, which also includes the
-search time, only grows from 400 ms to 2.3 seconds. This difference is
-particularly dramatic given that the actual optimal plan in each case is an
-identical plan of about 100 steps (i.e., the one we would have found in
-Experiment 2 for the $h=20$, $n=5$ case). The planning times for these
-instances are also concerning since times over a couple seconds will
-negatively affect the overall response time of the system, which must react
-in real time to user actions.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp3}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
-  \caption{Results on the GIVE domain with junk
-    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
-    The vertical axis is the runtime in seconds.}
-  \label{fig:give-runtime-junk}
-\end{figure}
-
-\subsection{Experiment 4: GIVE worlds with inaccessible positions}
-\label{sec:experiment-4:-give}
-
-In the final experiment, we consider a variation on the GIVE worlds from
-Experiment~3. As before, we begin with a base grid of $2n$ by $h$
-positions, with buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i
-\leq n$. We then generate a second grid of $w$ by $h$ positions and place
-an additional button in this grid. Unlike Experiment~3, the second grid
-does not extend the first grid but is ``disconnected'' so its positions are
-inaccessible from the first grid, as shown in
-Figure~\ref{fig:give-junk-nosoln}. Since the geometry of the grid makes it
-impossible to construct a plan for pressing all the buttons in the world,
-we instead investigate the time it takes a planner to arrive at the
-conclusion that the problem cannot be solved.
-
-Results for the $h=20$, $n=5$ case with $w$ ranging from 1 to 50 are shown
-in Figure~\ref{fig:give-runtime-nosoln}.\footnote{FF and SGPLAN do
- not display runtimes when they fail to construct a plan so an external timing
- program was used to generate these results.}
-In this case, the runtimes for FF rise sharply around $w=28$, producing
-times that are unacceptable for an NLG system in practice. FF is unable
-to complete problem instances above $w=35$. SGPLAN, by comparison, performs
-exceptionally well and is able to complete problems instances up to $w=50$
-in about 0.3 seconds and $w=80$ in about 0.6 seconds. These results
-demonstrate that SGPLAN has a significant advantage over FF in this
-experiment. 
-%Our implementation of GraphPlan is unable to complete any of
-%these problem instances.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.85\columnwidth]{pic-empty-inaccessible}
-  \caption{GIVE world with inaccessible positions.}
-  \label{fig:give-junk-nosoln}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.85\columnwidth]{graph-exp4}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
-  \caption{Results on the GIVE domain with inaccessible
-    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
-    The vertical axis is the runtime in seconds.}
-  \label{fig:give-runtime-nosoln}
-\end{figure}
-
-\section{Discussion}
-\label{sec:discussion}
-
-The positive conclusion we draw from our simple set of experiments is that
-modern planners are fairly good at controlling the search for many of the
-moderately-sized NLG problems we looked at. This is particularly true for
-FF in the sentence generation domain, which generates nontrivial 14-word
-sentences in about two seconds. Our results also indicate there is still
-room for improvement, however: special-purpose algorithms are able to
-generate much larger referring expressions in milliseconds
-\citep{AreKolStr08}. Although search efficiency is not the main concern in
-these domains, it nevertheless becomes a factor in scenarios like
-Experiment~2, where the world is almost completely unconstrained and
-(unsurprisingly) the search becomes much slower.
-
-The most restrictive bottleneck in the two NLG domains we investigated is
-the initial grounding step that many modern planners perform. While this
-may be a good strategy for traditional planning benchmarks and IPC
-domains---where plan length often dominates universe size and an initial
-investment into grounding pays off in saved instantiations during
-search---this approach has a very pronounced effect in domains like ours.
-As our experiments show, there are natural planning domains in which
-relatively short plans must be computed in large universes. For instance,
-in the sentence generation domain, it is not unrealistic to look for plans
-of length 20 over a universe consisting of thousands of domain individuals
-and tens of thousands of possible action instantiations, some of which
-require three domain individuals as parameters. Planners like FF and SGPLAN
-are less effective in such domains, to the point that grounding
-often dominates the total running time of these planners. (In the case of
-Experiment~1, our ad-hoc implementation of GraphPlan outperforms both
-planners in the sentence generation task.)
-
-Since our domains are not that unusual in their structure and composition,
-we hope that the lessons learned from our experiences can help improve the
-performance of current planning systems. For instance, the common strategy
-of splitting the universe into several types of individuals will not be a
-very effective solution in our NLG domains given the large number of
-individuals that could exist for a given type. On the other hand,
-sophisticated techniques based on reachability analysis may have the
-potential to improve this situation, provided such approaches can avoid the
-drawbacks of unnecessary grounding.\footnote{See
-  \citep{Bryce:07} for a good survey on reachability heuristics for planning
-  graph-based approaches to plan generation.}
-While recent trends in planning research have (rightfully) focused on the
-development of algorithms that control search in sophisticated ways,
-resulting in state-of-the-art planners that are more powerful and
-successful than their predecessors, we nevertheless believe that
-improvements are necessary if planning is to become a more mature
-technology that can offer tools to a wider community of users and
-researchers. At the end of the day, real-world users will care about the
-\emph{total} runtime of a planner, which is more than just the search time.
-
-By and large, our experiences with the planning community from a user's
-point of view have been favourable. Thanks to the planning competitions, it
-is easy to identify and download a fast state-of-the-art planner, at least
-for Linux. However, in the course of our experiments we found and reported
-bugs in both FF and SGPLAN. We also discovered that deciding as to which
-planner is appropriate for a particular task is not always straightforward,
-as the best performing planner in the IPC is not necessarily suitable for
-every planning task. As our experiments have shown, even planners as
-closely related as FF and SGPLAN can differ significantly in their
-performance on different domains (for instance, our results in
-Experiment~4).
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-In this paper, we introduced two novel planning domains arising from
-problems in natural language generation: the sentence generation domain and
-the Generating Instructions in Virtual Environments (GIVE) challenge
-domain. We also reported the results of a set of experiments designed to
-evaluate the suitability of off-the-shelf planners (particularly FF and
-SGPLAN) in these domains.
-
-Our results were mixed. While the planners we tested do an impressive job
-of controlling the complexity of search in many domains, they also suffer
-from practical problems that limit their performance in unexpected ways: in
-both of our NLG domains the grounding step performed by FF and SGPLAN
-dominates the time it takes to perform the search itself. As future work,
-we hope to extend our analysis to other planners that do not perform the
-initial grounding step, planners such as SHOP2
-\citep{DBLP:journals/jair/NauAIKMWY03} that have traditionally performed
-well on a wide range of domains, and new planners introduced in the latest
-edition of the IPC.\footnote{See
- \url{http://ipc.informatik.uni-freiburg.de/} for details of the 2008
- International Planning Competition (deterministic track).}
-
-The NLG community's recent interest in planning presents a valuable
-opportunity for planning researchers. Domains like GIVE highlight certain
-challenges, such as plan execution monitoring and plan presentation (i.e.,
-summarisation and elaboration), but also offer a platform on which such
-technologies can be evaluated in experiments with human users. Although we
-have focused on classical planning problems in this work, research related
-to reasoning under uncertainty, resource management, and planning with
-knowledge and sensing, can also be investigated in these settings. As such,
-we believe our domains would provide suitable challenges for planners
-entered in future editions of the IPC. Provided some of the challenges we
-have identified can be addressed, NLG-inspired problem domains offer a
-constructive platform for the planning community to showcase their
-techniques to a wider audience---and to improve the quality of their tools
-for real-world planning tasks.
-
-\todo{planning claims general applicability; it's clear that IPC
-  benchmarks focus areas of improvement; but it's not so clear to what
-  extent this limits usefulness for non-IPC application areas; this
-  paper evaluates this (for the first time??)}
-
-\section*{Acknowledgements}
-
-This work arose in the context of the Planning and Language Interest Group
-at the University of Edinburgh. The authors would like to thank all members
-of this group, especially Hector Geffner and Mark Steedman, for interesting
-discussions. This work was partially supported by the DFG Research
-Fellowship ``CRISP: Efficient integrated realization and microplanning''
-and by the European Commission through the PACO-PLUS project
-(FP6-2004-IST-4-27657).
-
-\bibliography{references}
-\bibliographystyle{ci}
-
-\pagebreak
-\listoffigures
-
-\end{document}
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "manuscript"
+%%% End:

### 2009-05-07 trunk/Papers/ci-spark/experiments.tex
split paper up into individual files; changes to abstract and intro

--- 2009-04-27 trunk/Papers/ci-spark/manuscript.tex
@@
-%%%
-%%% Computational Intelligence manuscript based on SPARK'08 paper
-%%%
-%%% Alexander Koller and Ronald Petrick
-%%% 
-%%% Updated: 2008/12/22
-%%%
-\documentclass[letterpaper]{article}
-\usepackage{times}
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage[round]{natbib}
-\usepackage{setspace}
-\usepackage[paper=letterpaper,margin=1.0in]{geometry}
-
-\title{Experiences with Planning for Natural Language Generation}
-\author{\textbf{Alexander Koller} \\
-Exzellenzcluster / FR 4.7 Computerlinguistik \\
-Saarland University \\
-Postfach 151150 \\
-66041 Saarbrücken \\
-Germany \\
-Phone: +49 681 302 70040 \\
-Fax: +49 681 302 4351 \\
-E-mail: \url{koller@mmci.uni-saarland.de}
-\and
-\textbf{Ronald P. A. Petrick} \\
-School of Informatics \\
-University of Edinburgh \\
-Informatics Forum, 10 Crichton Street \\
-Edinburgh \ EH8 9AB \\
-Scotland, United Kingdom \\
-Phone: +44 131 650 4426 \\
-Fax: +44 131 650 6899 \\
-E-mail: \url{rpetrick@inf.ed.ac.uk}}
-
-%\date{22 December 2008}
-\date{}
-
-\pagestyle{plain}
-\pagenumbering{arabic}
-\doublespacing
-
-\bibpunct{(}{)}{;}{a}{}{,}
-
-\begin{document}
-
-\maketitle
-
-%%%
-%%% Abstract
-%%%
-\pagebreak
-\begin{abstract}
-We investigate the application of modern planning techniques to domains
-arising from problems in natural language generation (NLG). In particular,
-we consider two novel NLG-inspired planning problems, the sentence
-generation domain and the GIVE (``Generating Instructions in Virtual
-Environment'') domain, and investigate the efficiency of the FF and SGPLAN
-planners in these domains. We also compare our results against an ad-hoc
-implementation of GraphPlan in Java. Our results are mixed. While modern
-planners are able to quickly solve many moderately-sized instances of our
-problems, the overall planning time is dominated by the grounding step that
-these planners perform, rather than search. This has a pronounced effect on
-our domains which require relatively short plans but have large universes.
-We share our experiences and offer these domains as challenges for the
-planning community.
-\end{abstract}
-
-\bigskip\noindent
-\textbf{Keywords:} natural language generation, planning
-
-%%%
-%%% Main text
-%%%
-\pagebreak
-\section{Introduction}
-\label{sec:introduction}
-
-Natural language generation (NLG; \citealp{reiter00building}) is one of the
-major subfields of natural language processing, concerned with computing
-natural language sentences or texts that convey a given piece of
-information to the user. Intuitively, this task can be viewed as a problem
-involving actions, beliefs, and goals: an agent communicating with another
-agent tries to change the mental state of the hearer by applying actions
-which correspond to the utterance of words or sentences.
-This characterization of NLG suggests obvious parallels to automated
-planning---a view which has a long tradition in NLG.
-
-Early approaches to generation as planning
-\citep{perrault80,appelt:planning,hovy88,young94dpocl} focused primarily on
-planning high-level structures, such as speech acts and discourse
-relations.  While these approaches suffered from the inefficiency of the
-planners that were available at the time, there has recently been a
-resurgence of interest in applying modern planning techniques to NLG.
-These newer papers revisit speech act planning based on the more expressive
-planning formalisms available today \citep{Steedman-Petrick:07,benotti08b}
-and exploit the improved efficiency of modern planners for an integrated
-model of sentence planning and surface realization \citep{KolSto07}.
-Problems in NLG are often computationally challenging (see, e.g.,
-\citep{KolStr02}), and these new approaches combine an interest in planning
-as a modelling tool for natural language, with a hope of improved
-efficiency by exploiting modern approaches and algorithms.
-
-The focus of this paper is twofold. First, we present two novel planning
-domains that arise in the context of natural language generation: the
-sentence generation domain and the GIVE domain. In the sentence generation
-task, the goal is to generate a single sentence that expresses a given
-meaning. \citet{KolSto07} cast this domain as a planning problem, where the
-plan encodes the necessary sentence and the actions correspond to uttering
-individual words.  In the GIVE domain (``Generating Instructions in Virtual
-Environments''), we describe a new shared task that was recently posed as a
-challenge for the NLG community \citep{alexander07:_shared_task_propos}.
-GIVE uses planning as part of an NLG system that generates natural-language
-instructions to guide a user in performing a given task in a virtual
-environment.
-
-Second, we discuss some of our experiences using off-the-shelf
-planners in our two NLG-inspired domains. In particular, we explore the
-efficiency of FF \citep{HoffmannNebel01} and SGPLAN
-\citep{hsu06:_new_featur_in_sgplan_for}: two planners that are readily
-available, support an expressive subset of the Planning Domain Definition
-Language (PDDL; McDermott et al.~\citeyear{PDDL}) for encoding domains, and
-have been successful on traditional benchmarks and problems from the
-International Planning Competition (IPC)\footnote{See
- \url{http://ipc.icaps-conference.org/} for information about past
- editions of the IPC.}.
-We test these planners on a range of problem instances in our planning
-domains, and compare our results against an ad-hoc Java implementation of
-GraphPlan \citep{Blum1997}. Overall, our findings are mixed. On the one
-hand, we demonstrate that modern planners can easily handle some of the
-moderately-sized search problems that arise in NLG. In certain cases, the
-planners we tested are more efficient than specialized NLG algorithms
-applied to the same problem. On the other hand, these same planners spend
-tremendous amounts of time on preprocessing. For instance, in the sentence
-generation domain, FF spends 90\% of its runtime grounding out literals and
-actions, and ends up being slower than our version of GraphPlan that avoids
-grounding. For domains like ours, which are dominated by the number of
-actions and the universe size, rather than the combinatorics of the search
-problem, this observation suggests that the overall runtime of a planner
-could be improved by grounding more selectively. This is particularly true
-for large problem instances, which still remain a challenge for the
-planners we tested, and where the effects of the grounding approach become
-quite pronounced. We therefore offer our NLG domains as challenges for the
-planning community.
-
-\section{Planning in NLG}
-\label{sec:domains}
-
-We begin by presenting our two planning domains: the sentence generation
-domain and the instruction giving domain.
-
-\subsection{Sentence generation as planning}
-
-Sentence generation is the problem of computing from a grammar and a
-semantic representation, a single sentence that expresses this piece of
-meaning. Traditionally, this problem is split into several steps
-\citep{reiter00building}. In a first step, called \emph{sentence planning},
-the semantic representation is enriched with additional information. For
-instance, \emph{referring expressions}, which refer to individuals (e.g.,
-objects, people, etc.) in the domain of discourse, are determined at this
-point. In a second step, called \emph{surface realization}, this enriched
-representation is translated into a natural-language sentence, using the
-grammar.
-
-In practice, the process of determining referring expressions typically
-interacts with the realization step and so it often turns out to be
-beneficial to perform these two steps together. For instance, the SPUD
-system (``Sentence Planning Using Description''; \citealt{Stone2003a})
-performed this task using a top-down generation algorithm based on
-tree-adjoining grammars \citep{joshi;etal1997} whose lexical entries were
-equipped with semantic and pragmatic information. Unfortunately, SPUD had
-to resort to a non-optimal greedy search strategy to cope with its large
-search space and retain reasonable efficiency. To improve the efficiency
-of SPUD, \citet{KolSto07} translate the sentence generation problem into a
-planning problem and use a planning algorithm to perform the generation
-task.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{pic-grammar}
-  \caption{An example grammar in the sentence generation domain.}
-  \label{fig:white-rabbit-sleeps-grammar}
-\end{figure}
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{pic-derivation-2}
-  \caption{Derivation of ``The white rabbit sleeps.''}
-  \label{fig:white-rabbit-sleeps-deriv}
-\end{figure}
-
-We illustrate this process on a simplified example.  Consider a
-knowledge base containing the individuals $e$, $r_1$ and $r_2$, and a
-set of attributes encoding the fact that $r_1$ and $r_2$ are rabbits,
-$r_1$ is white and $r_2$ is brown, and $e$ is an event in which $r_1$
-sleeps.  Say that we want to express the information
-$\{\mathsf{sleep}(e,r_1)\}$ using the tree-adjoining grammar shown in
-Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
-\emph{elementary trees} (i.e., the disjoint trees in the figure), each
-of which contributes certain \emph{semantic content}. We can
-instantiate these trees by substituting individuals for \emph{semantic
-  roles}, such as $\mathsf{self}$ and $\mathsf{subj}$, and then
-combine the tree instances as shown in
-Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
-``The white rabbit sleeps''.
-
-We compute this grammatical derivation in a top-down manner, starting with
-the elementary tree for ``sleeps''. This tree satisfies the need to convey
-the semantic information, but introduces a need to generate a noun phrase
-(NP) for the subject; this NP must refer uniquely to the target referent
-$r_1$. In a second step, we substitute the tree for ``the rabbit'' into the
-open NP leaf, which makes the derivation grammatically complete. Since
-there are two different individuals that could be described as ``the
-rabbit''---technically, $r_2$ is still a \emph{distractor} (i.e., based on
-the description ``the rabbit'', the hearer might erroneously think that
-we're talking about $r_2$ and not $r_1$)---we are still not finished. To
-complete the derivation, the tree for ``white'' is added to the existing
-structure by an \emph{adjunction} operation, making the derivation
-both syntactically and semantically complete.
-
-\begin{figure}[p]
-\centering
-\begin{minipage}{0.5\textwidth}
-{\small%
-\begin{verbatim}
-(:action add-sleeps
-   :parameters (?u - node
-                ?xself - individual
-                ?xsubj - individual)
-   :precondition
-       (and (subst S ?u)
-            (referent ?u ?xself)
-            (sleep ?xself ?xsubj))
-   :effect 
-       (and (not (subst S ?u))
-            (expressed sleep ?xself ?xsubj)
-            (subst NP (subj ?u))
-            (referent (subj ?u) ?xsubj)
-            (forall (?y - individual)
-                (when (not (= ?y ?xself))
-                    (distractor (subj ?u) ?y)))))
-
-(:action add-rabbit
-   :parameters (?u - node
-                ?xself - individual)
-   :precondition 
-       (and (subst NP ?u)
-            (referent ?u ?xself)
-            (rabbit ?xself))
-   :effect 
-       (and (not (subst NP ?u))
-            (canadjoin N ?u)
-            (forall (?y - individual)
-                (when (not (rabbit ?y))
-                    (not (distractor ?u ?y))))))
-
-(:action add-white
-   :parameters (?u - node
-                ?xself - individual)
-   :precondition 
-       (and (canadjoin N ?u)
-            (referent ?u ?xself)
-            (rabbit ?xself))
-   :effect 
-       (forall (?y - individual)
-           (when (not (white ?y))
-               (not (distractor ?u ?y)))))
-\end{verbatim}}%
-\end{minipage}
-\caption{PDDL actions for generating the sentence ``The white rabbit
-sleeps.''}
-\label{fig:white-rabbit-as-planning}
-\end{figure}
-
-The process described above has clear parallels to planning: we manipulate
-states by applying sequences of actions to achieve a goal. We can make this
-connection even more precise by translating the SPUD problem into a
-planning problem.\footnote{See
-  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
-  which this conversion is implemented.}  For instance,
-Figure~\ref{fig:white-rabbit-as-planning} shows the corresponding PDDL
-actions for the above generation task, where each action corresponds
-to an operation that adds a single elementary tree to the derivation.
-In each case, the first parameter of the action is a node name in the
-derivation tree, and the remaining parameters stand for the
-individuals to which the semantic roles will be instantiated.  The
-syntactic preconditions and effects are encoded using $\mathsf{open}$
-and $\mathsf{canadjoin}$ literals; the status of each referring expression
-is tracked using $\mathsf{distractor}$ literals. The resulting action
-effects contain terms of the form $\mathsf{subj}(u)$, which construct new
-node names, but aren't strictly permitted by PDDL. In order to meet the
-syntactic requirements of PDDL, these terms can be eliminated by estimating
-an upper bound $n$ for the plan length, making $n$ copies of each action,
-ensuring that copy $i$ can only be applied in step $i$, and replacing the
-term $\mathsf{subj}(u)$ in an action copy by the constant
-$\mathsf{subj}_i$.
-
-Once the appropriate actions are defined, we can solve the generation
-problem as an ordinary planning problem. For instance, the following plan
-solves our previous example:
-%
-\begin{enumerate}
-\item $\mathsf{add}\textsf{-}\mathsf{sleeps}(\mathsf{root}, r_1)$,
-\item $\mathsf{add}\textsf{-}\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}),r_1)$,
-\item $\mathsf{add}\textsf{-}\mathsf{white}(\mathsf{subj}(\mathsf{root}),r_1)$.
-\end{enumerate}
-%
-Using this plan, the grammatical derivation in
-Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
-generated sentence ``the white rabbit sleeps'', can be systematically
-reconstructed. Thus, we can solve the sentence generation problem via
-the detour through planning and bring current search heuristics for
-planning to bear on generation.
-
-\subsection{Planning in instruction giving}
-
-The object of the GIVE Challenge (``Generating Instructions in Virtual
-Environments''; \citealt{alexander07:_shared_task_propos}) is to build an
-NLG system which is able to produce natural-language instructions which
-will guide a human user in performing some task in a virtual environment.
-From an NLG perspective, GIVE makes for an interesting challenge because it
-is a theory-neutral task that exercises all components of an NLG system,
-and emphasizes the study of communication in a (simulated) physical
-environment. It also has the advantage that the user and the NLG system can
-be in physically different places, as long as the 3D client and the NLG
-system are connected over a network. This makes it possible to evaluate
-GIVE NLG systems on a large scale over the Internet.
-%
-The first GIVE evaluation is currently underway. Five NLG systems from
-three countries are in the process of being evaluated. At the time of
-writing, 1028 different evaluation runs---each consisting of one user
-interacting with one instance of the virtual environment and being guided
-by an NLG system---have been collected, making GIVE the largest-scale
-NLG evaluation effort ever.\footnote{See the GIVE website at
- \url{http://www.give-challenge.org/} for more details about this
- challenge.}
-
-\begin{figure}[t]
-\centering
-%\includegraphics[width=1 \columnwidth]{give_world_no_expl}
-\includegraphics[width=0.75\columnwidth]{give_world_2}
-\caption{Map of the GIVE development world.}
-  \label{fig:give-development-world}
-\end{figure}
-
-An example GIVE world map is shown in
-Figure~\ref{fig:give-development-world}. To simplify both the planning and
-the NLG task, the world is discretised into a set of tiles of equal size.
-The user can turn by 90 degree steps in either direction, and can move from
-the centre of one tile to the centre of the next tile, provided the path
-between two tiles is not blocked. The world also consists of a set of
-objects which can be manipulated by the user in various ways. For
-instance, in the example world the user's task is to pick up a trophy in
-the top left room. The trophy is hidden in a safe behind a picture, so the
-user must first move the picture out of the way and open the safe by
-pushing a sequence of buttons (the small square boxes on the walls) in the
-correct order. In order to get to all these buttons, the user must also
-open the door in the centre of the map and deactivate the alarm tile by
-pushing further buttons. Figure~\ref{fig:give-planning} shows the encoding
-of some of the available GIVE domain actions, in PDDL syntax.
-
-\begin{figure}[p]
-\centering
-\begin{minipage}{0.5\textwidth}
-{\small%
-\begin{verbatim}
-(:action move
-   :parameters (?from - position
-                ?to - position
-                ?ori - orientation)
-   :precondition 
-       (and (player-pos ?from) 
-            (adjacent ?from ?to ?ori) 
-            (player-orient ?ori)
-            (not-blocked ?from ?to)
-            (not-alarmed ?to))
-   :effect 
-       (and (not (player-pos ?from))
-            (player-pos ?to)))
-
-(:action turn-left
-   :parameters (?ori - orientation
-                ?newOri - orientation)
-   :precondition 
-       (and (player-orient ?ori)
-            (next-orient-left ?ori ?newOri))
-   :effect 
-       (and (not (player-orient ?ori))
-            (player-orient ?newOri)))
-
-(:action turn-right
-   :parameters (?ori - orientation
-                ?newOri - orientation)
-   :precondition 
-       (and (player-orient ?ori)
-            (next-orient-right ?ori ?newOri))
-   :effect 
-       (and (not (player-orient ?ori))
-            (player-orient ?newOri)))
-
-(:action manipulate-b1-off-on
-   :parameters (?pos - position)
-   :precondition 
-       (and (state b1 off)
-            (player-pos ?pos)
-            (position b1 ?pos))
-   :effect 
-       (and (not (state b1 off))
-            (state b1 on)
-            (not (state d1 closed))
-            (state d1 open) 
-            (not (blocked pos_6_5 pos_6_4))
-            (not (blocked pos_6_4 pos_6_5))))
-\end{verbatim}}%
-\end{minipage}
-\caption{PDDL actions for the GIVE domain.}
-\label{fig:give-planning}
-\end{figure}
-
-The plans generated in the GIVE domain are often nontrivial. For instance,
-in the above example the shortest plan consists of 108 action steps, with
-the first few steps as follows:
-%
-\begin{enumerate}
-\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north},
-\mathsf{west})$,
-\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
-\end{enumerate}
-
-Most of the actions in the above plan are ``move'' actions. This makes the
-bulk of the GIVE problem very similar to that of a Gridworld problem (see,
-e.g., \citep{Tovey-Koenig:2000}), which also involves finding a route
-through a two-dimensional world map with discrete positions. GIVE tasks are
-usually much more complex, however, as illustrated by the above example
-map: a successful plan must include steps for pressing buttons in the right
-order, reasoning about large numbers of world objects, and navigating
-through complicated room shapes.
-
-The GIVE task comes with a number of interesting research challenges for
-both the NLG and planning communities. On the NLG side, one major problem
-is that verbalising each action instance in a plan makes for very
-poor natural-language instructions. For instance, the NLG system could
-generate an instruction sequence starting with ``turn left; walk forward;
-press the button; turn right; walk forward; walk forward; turn right; walk
-forward; walk forward; turn left; walk forward''. However, directions of
-this form are linguistically clumsy and uninteresting, and it is to be
-expected that frustrated users will quickly cancel an interaction with such
-a system, resulting in a bad evaluation score. Having just pressed the
-button, it would be much better to say ``turn around and walk through the
-door''. That is, the system should \emph{summarise} plans by merging
-multiple action instances into single instructions before presenting them
-to the user.\footnote{This task is similar to the construction of
-  ``macro'' actions, but is primarily intended for plan presentation
-  purposes. See, e.g., \citep{Botea-etal:05} which contains a survey of
-  recent approaches for generating macro actions in planning.}
-Conversely, it may also be necessary to \emph{elaborate} on a single
-planning step by expressing it with several instructions. Having just
-entered the top left room, it may be easier for the user to understand the
-instruction ``walk to the centre of the room; turn right; now press the
-green button in front of you'', rather than the instruction ``press the
-green button on the wall to your right''. Furthermore, in order to plan the
-referring expression ``the green button in front of you'' at a time when
-the button is not yet in front of the user, the NLG system must keep track
-of the hypothetical changes in what objects are visible to the user. Thus,
-the NLG acts of referring and instructing must be tightly integrated with
-the modules for plan generation.
-
-From a planning perspective, GIVE imposes very strict runtime
-requirements on the plan generation task: planning must
-happen in real time and the system must respond to a user in a timely
-fashion.  If the system takes too much time deliberating over an
-instruction to give, rather than actually giving this instruction, the
-user may have walked or turned away, thus making the instruction
-invalid.  Furthermore, plan execution monitoring also plays an
-important role in the GIVE problem.  At a high level, the system needs
-to monitor a user's actions and compare them against the generated
-instruction set to determine if the user has correctly followed
-directions or not. In the case of the latter, new instructions may
-have to be generated. In practice, the situation can be quite
-complicated since the mental state of the user is not known and so the
-system must be inferred from observing the user's actions in real
-time. For example, a user directed to ``turn around and walk through
-the door'' may not necessarily perform these actions to the letter,
-i.e., immediately turning 180 degrees and proceeding directly to the
-door. Instead, the user might take a roundabout route through the
-room, eventually exiting out the door.  Although the user's actions do
-not match the generated instructions exactly, they meet the intended
-goal. The system must be able to identify such ``equivalent'' plans
-and not immediately generate new instructions as soon as the user's
-actions have gone off course. Furthermore, a user can communicate
-certain intentions to the system, both through action and
-inaction. For instance, the system should infer that a user has failed
-to follow instructions if the user exits a room when given a directive
-to ``walk to the centre of the room''. The system should also make a
-similar conclusion if a user simply does nothing when given the
-instruction.
-
 \section{Experiments}
@@
 
-\section{Discussion}
-\label{sec:discussion}
-
-The positive conclusion we draw from our simple set of experiments is that
-modern planners are fairly good at controlling the search for many of the
-moderately-sized NLG problems we looked at. This is particularly true for
-FF in the sentence generation domain, which generates nontrivial 14-word
-sentences in about two seconds. Our results also indicate there is still
-room for improvement, however: special-purpose algorithms are able to
-generate much larger referring expressions in milliseconds
-\citep{AreKolStr08}. Although search efficiency is not the main concern in
-these domains, it nevertheless becomes a factor in scenarios like
-Experiment~2, where the world is almost completely unconstrained and
-(unsurprisingly) the search becomes much slower.
-
-The most restrictive bottleneck in the two NLG domains we investigated is
-the initial grounding step that many modern planners perform. While this
-may be a good strategy for traditional planning benchmarks and IPC
-domains---where plan length often dominates universe size and an initial
-investment into grounding pays off in saved instantiations during
-search---this approach has a very pronounced effect in domains like ours.
-As our experiments show, there are natural planning domains in which
-relatively short plans must be computed in large universes. For instance,
-in the sentence generation domain, it is not unrealistic to look for plans
-of length 20 over a universe consisting of thousands of domain individuals
-and tens of thousands of possible action instantiations, some of which
-require three domain individuals as parameters. Planners like FF and SGPLAN
-are less effective in such domains, to the point that grounding
-often dominates the total running time of these planners. (In the case of
-Experiment~1, our ad-hoc implementation of GraphPlan outperforms both
-planners in the sentence generation task.)
-
-Since our domains are not that unusual in their structure and composition,
-we hope that the lessons learned from our experiences can help improve the
-performance of current planning systems. For instance, the common strategy
-of splitting the universe into several types of individuals will not be a
-very effective solution in our NLG domains given the large number of
-individuals that could exist for a given type. On the other hand,
-sophisticated techniques based on reachability analysis may have the
-potential to improve this situation, provided such approaches can avoid the
-drawbacks of unnecessary grounding.\footnote{See
-  \citep{Bryce:07} for a good survey on reachability heuristics for planning
-  graph-based approaches to plan generation.}
-While recent trends in planning research have (rightfully) focused on the
-development of algorithms that control search in sophisticated ways,
-resulting in state-of-the-art planners that are more powerful and
-successful than their predecessors, we nevertheless believe that
-improvements are necessary if planning is to become a more mature
-technology that can offer tools to a wider community of users and
-researchers. At the end of the day, real-world users will care about the
-\emph{total} runtime of a planner, which is more than just the search time.
-
-By and large, our experiences with the planning community from a user's
-point of view have been favourable. Thanks to the planning competitions, it
-is easy to identify and download a fast state-of-the-art planner, at least
-for Linux. However, in the course of our experiments we found and reported
-bugs in both FF and SGPLAN. We also discovered that deciding as to which
-planner is appropriate for a particular task is not always straightforward,
-as the best performing planner in the IPC is not necessarily suitable for
-every planning task. As our experiments have shown, even planners as
-closely related as FF and SGPLAN can differ significantly in their
-performance on different domains (for instance, our results in
-Experiment~4).
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-In this paper, we introduced two novel planning domains arising from
-problems in natural language generation: the sentence generation domain and
-the Generating Instructions in Virtual Environments (GIVE) challenge
-domain. We also reported the results of a set of experiments designed to
-evaluate the suitability of off-the-shelf planners (particularly FF and
-SGPLAN) in these domains.
-
-Our results were mixed. While the planners we tested do an impressive job
-of controlling the complexity of search in many domains, they also suffer
-from practical problems that limit their performance in unexpected ways: in
-both of our NLG domains the grounding step performed by FF and SGPLAN
-dominates the time it takes to perform the search itself. As future work,
-we hope to extend our analysis to other planners that do not perform the
-initial grounding step, planners such as SHOP2
-\citep{DBLP:journals/jair/NauAIKMWY03} that have traditionally performed
-well on a wide range of domains, and new planners introduced in the latest
-edition of the IPC.\footnote{See
- \url{http://ipc.informatik.uni-freiburg.de/} for details of the 2008
- International Planning Competition (deterministic track).}
-
-The NLG community's recent interest in planning presents a valuable
-opportunity for planning researchers. Domains like GIVE highlight certain
-challenges, such as plan execution monitoring and plan presentation (i.e.,
-summarisation and elaboration), but also offer a platform on which such
-technologies can be evaluated in experiments with human users. Although we
-have focused on classical planning problems in this work, research related
-to reasoning under uncertainty, resource management, and planning with
-knowledge and sensing, can also be investigated in these settings. As such,
-we believe our domains would provide suitable challenges for planners
-entered in future editions of the IPC. Provided some of the challenges we
-have identified can be addressed, NLG-inspired problem domains offer a
-constructive platform for the planning community to showcase their
-techniques to a wider audience---and to improve the quality of their tools
-for real-world planning tasks.
-
-\section*{Acknowledgements}
-
-This work arose in the context of the Planning and Language Interest Group
-at the University of Edinburgh. The authors would like to thank all members
-of this group, especially Hector Geffner and Mark Steedman, for interesting
-discussions. This work was partially supported by the DFG Research
-Fellowship ``CRISP: Efficient integrated realization and microplanning''
-and by the European Commission through the PACO-PLUS project
-(FP6-2004-IST-4-27657).
-
-\bibliography{references}
-\bibliographystyle{ci}
-
-\pagebreak
-\listoffigures
-
-\end{document}
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "manuscript"
+%%% End:

### 2009-05-07 trunk/Papers/ci-spark/introduction.tex
split paper up into individual files; changes to abstract and intro

--- 2009-04-27 trunk/Papers/ci-spark/manuscript.tex
@@
-%%%
-%%% Computational Intelligence manuscript based on SPARK'08 paper
-%%%
-%%% Alexander Koller and Ronald Petrick
-%%% 
-%%% Updated: 2008/12/22
-%%%
-\documentclass[letterpaper]{article}
-\usepackage{times}
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage[round]{natbib}
-\usepackage{setspace}
-\usepackage[paper=letterpaper,margin=1.0in]{geometry}
-
-\title{Experiences with Planning for Natural Language Generation}
-\author{\textbf{Alexander Koller} \\
-Exzellenzcluster / FR 4.7 Computerlinguistik \\
-Saarland University \\
-Postfach 151150 \\
-66041 Saarbrücken \\
-Germany \\
-Phone: +49 681 302 70040 \\
-Fax: +49 681 302 4351 \\
-E-mail: \url{koller@mmci.uni-saarland.de}
-\and
-\textbf{Ronald P. A. Petrick} \\
-School of Informatics \\
-University of Edinburgh \\
-Informatics Forum, 10 Crichton Street \\
-Edinburgh \ EH8 9AB \\
-Scotland, United Kingdom \\
-Phone: +44 131 650 4426 \\
-Fax: +44 131 650 6899 \\
-E-mail: \url{rpetrick@inf.ed.ac.uk}}
-
-%\date{22 December 2008}
-\date{}
-
-\pagestyle{plain}
-\pagenumbering{arabic}
-\doublespacing
-
-\bibpunct{(}{)}{;}{a}{}{,}
-
-\begin{document}
-
-\maketitle
-
-%%%
-%%% Abstract
-%%%
-\pagebreak
-\begin{abstract}
-We investigate the application of modern planning techniques to domains
-arising from problems in natural language generation (NLG). In particular,
-we consider two novel NLG-inspired planning problems, the sentence
-generation domain and the GIVE (``Generating Instructions in Virtual
-Environment'') domain, and investigate the efficiency of the FF and SGPLAN
-planners in these domains. We also compare our results against an ad-hoc
-implementation of GraphPlan in Java. Our results are mixed. While modern
-planners are able to quickly solve many moderately-sized instances of our
-problems, the overall planning time is dominated by the grounding step that
-these planners perform, rather than search. This has a pronounced effect on
-our domains which require relatively short plans but have large universes.
-We share our experiences and offer these domains as challenges for the
-planning community.
-\end{abstract}
-
-\bigskip\noindent
-\textbf{Keywords:} natural language generation, planning
-
-%%%
-%%% Main text
-%%%
-\pagebreak
 \section{Introduction}
@@
 
-Natural language generation (NLG; \citealp{reiter00building}) is one of the
-major subfields of natural language processing, concerned with computing
-natural language sentences or texts that convey a given piece of
-information to the user. Intuitively, this task can be viewed as a problem
-involving actions, beliefs, and goals: an agent communicating with another
-agent tries to change the mental state of the hearer by applying actions
-which correspond to the utterance of words or sentences.
-This characterization of NLG suggests obvious parallels to automated
-planning---a view which has a long tradition in NLG.
+Natural language generation (NLG; \citealp{reiter00building}) is one
+of the major subfields of natural language processing, concerned with
+computing natural language sentences or texts that convey a given
+piece of information to the user. This task can be viewed as a problem
+of achieving a (communicative) goal by successively applying
+(communicative) actions, and thus has clear intuitive parallels to
+automated planning.
 
-Early approaches to generation as planning
-\citep{perrault80,appelt:planning,hovy88,young94dpocl} focused primarily on
-planning high-level structures, such as speech acts and discourse
-relations.  While these approaches suffered from the inefficiency of the
-planners that were available at the time, there has recently been a
-resurgence of interest in applying modern planning techniques to NLG.
-These newer papers revisit speech act planning based on the more expressive
-planning formalisms available today \citep{Steedman-Petrick:07,benotti08b}
-and exploit the improved efficiency of modern planners for an integrated
-model of sentence planning and surface realization \citep{KolSto07}.
-Problems in NLG are often computationally challenging (see, e.g.,
-\citep{KolStr02}), and these new approaches combine an interest in planning
-as a modelling tool for natural language, with a hope of improved
-efficiency by exploiting modern approaches and algorithms.
+This view of generation as planning has a long tradition in NLG
+\citep{perrault80,appelt:planning,hovy88,young94dpocl}. Despite their
+theoretical attractiveness, these early approaches never gained
+mainstream use in applied systems, in part because of the inefficiency
+of the planners that were available at the time. However, the recent
+improvements in planner efficiency and expressiveness have sparked a
+renewed interest in applying modern planning techniques to NLG---both
+on the level of speech act planning
+\citep{Steedman-Petrick:07,benotti08b} and and on the level of
+sentence generation \citep{KolSto07}. The goal of this paper is to
+determine, by looking at some representative generation problems,
+whether planning has advanced to the point that this enterprise can be
+successful this time---and thus, to determine the usefulness of
+current planning technology to an application area that has a long
+tradition, but is not currently in the focus of the planning
+competitions.
 
-The focus of this paper is twofold. First, we present two novel planning
-domains that arise in the context of natural language generation: the
-sentence generation domain and the GIVE domain. In the sentence generation
+The focus of this paper is twofold. First, we present two generation
+problems that have recently been cast as planning problems: the
+sentence generation task and the GIVE task. In the sentence generation
 task, the goal is to generate a single sentence that expresses a given
-meaning. \citet{KolSto07} cast this domain as a planning problem, where the
-plan encodes the necessary sentence and the actions correspond to uttering
-individual words.  In the GIVE domain (``Generating Instructions in Virtual
-Environments''), we describe a new shared task that was recently posed as a
-challenge for the NLG community \citep{alexander07:_shared_task_propos}.
-GIVE uses planning as part of an NLG system that generates natural-language
-instructions to guide a user in performing a given task in a virtual
-environment.
+meaning. \citet{KolSto07} cast this task as a planning problem, where
+the plan encodes the necessary sentence and the actions correspond to
+uttering individual words.  In the GIVE domain (``Generating
+Instructions in Virtual Environments''), we describe a new shared task
+that was recently posed as a challenge for the NLG community
+\citep{ByrKolStrCasDalMooObe09}.  GIVE uses planning as part of an NLG
+system that generates natural-language instructions to guide a user in
+performing a given task in a virtual environment.
 
 Second, we discuss some of our experiences using off-the-shelf
-planners in our two NLG-inspired domains. In particular, we explore the
+planners in these two domains. In particular, we explore the
 efficiency of FF \citep{HoffmannNebel01} and SGPLAN
 \citep{hsu06:_new_featur_in_sgplan_for}: two planners that are readily
-available, support an expressive subset of the Planning Domain Definition
-Language (PDDL; McDermott et al.~\citeyear{PDDL}) for encoding domains, and
-have been successful on traditional benchmarks and problems from the
-International Planning Competition (IPC)\footnote{See
+available, support an expressive subset of the Planning Domain
+Definition Language (PDDL; McDermott et al.~\citeyear{PDDL}) for
+encoding domains, and have been successful on the benchmarks and
+problems in the International Planning Competition (IPC)\footnote{See
   \url{http://ipc.icaps-conference.org/} for information about past
- editions of the IPC.}.
-We test these planners on a range of problem instances in our planning
-domains, and compare our results against an ad-hoc Java implementation of
-GraphPlan \citep{Blum1997}. Overall, our findings are mixed. On the one
-hand, we demonstrate that modern planners can easily handle some of the
-moderately-sized search problems that arise in NLG. In certain cases, the
-planners we tested are more efficient than specialized NLG algorithms
-applied to the same problem. On the other hand, these same planners spend
-tremendous amounts of time on preprocessing. For instance, in the sentence
-generation domain, FF spends 90\% of its runtime grounding out literals and
-actions, and ends up being slower than our version of GraphPlan that avoids
-grounding. For domains like ours, which are dominated by the number of
-actions and the universe size, rather than the combinatorics of the search
-problem, this observation suggests that the overall runtime of a planner
-could be improved by grounding more selectively. This is particularly true
-for large problem instances, which still remain a challenge for the
-planners we tested, and where the effects of the grounding approach become
-quite pronounced. We therefore offer our NLG domains as challenges for the
-planning community.
+  editions of the IPC.}. We test these planners on a range of problem
+instances in our planning domains, and compare our results against an
+ad-hoc Java implementation of GraphPlan \citep{Blum1997}.
 
-\section{Planning in NLG}
-\label{sec:domains}
+Overall, our findings are mixed. On the one hand, we demonstrate that
+modern planners can easily handle the \emph{search} problems that
+arise in NLG on realistic inputs, which is promising given that the
+sentence generation task, at least, is NP-complete
+\citep{KolStr02}. On the other hand, the same planners spend
+tremendous amounts of time on preprocessing: Because our domains use a
+large number of actions and constants, FF spends 90\% of its runtime
+in the sentence generation domain on grounding out all the actions and
+literals. As a consequence, the off-the-shelf planners we looked at
+are too slow to be useful in the NLG application -- but this is due to
+implementation issues rather than the quality of the search
+algorithms. We offer our NLG domains as challenges for the planning
+community in the hope that these implementation issues might gain more
+attention in the future.
 
-We begin by presenting our two planning domains: the sentence generation
-domain and the instruction giving domain.
+This paper is structured as follows. We begin by briefly reviewing
+some of the literature on NLG as planning
+(Section~\ref{sec:nlg-as-planning}). Then we define introduce the
+planning problems arising in two particular NLG tasks in more detail
+in Section~\ref{sec:domains}. We report on some experiments with these
+planning problems in Section~\ref{sec:experiments}, and discuss the
+results in Section~\ref{sec:discussion}. Section~\ref{sec:conclusion}
+concludes.
 
-\subsection{Sentence generation as planning}
-
-Sentence generation is the problem of computing from a grammar and a
-semantic representation, a single sentence that expresses this piece of
-meaning. Traditionally, this problem is split into several steps
-\citep{reiter00building}. In a first step, called \emph{sentence planning},
-the semantic representation is enriched with additional information. For
-instance, \emph{referring expressions}, which refer to individuals (e.g.,
-objects, people, etc.) in the domain of discourse, are determined at this
-point. In a second step, called \emph{surface realization}, this enriched
-representation is translated into a natural-language sentence, using the
-grammar.
-
-In practice, the process of determining referring expressions typically
-interacts with the realization step and so it often turns out to be
-beneficial to perform these two steps together. For instance, the SPUD
-system (``Sentence Planning Using Description''; \citealt{Stone2003a})
-performed this task using a top-down generation algorithm based on
-tree-adjoining grammars \citep{joshi;etal1997} whose lexical entries were
-equipped with semantic and pragmatic information. Unfortunately, SPUD had
-to resort to a non-optimal greedy search strategy to cope with its large
-search space and retain reasonable efficiency. To improve the efficiency
-of SPUD, \citet{KolSto07} translate the sentence generation problem into a
-planning problem and use a planning algorithm to perform the generation
-task.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{pic-grammar}
-  \caption{An example grammar in the sentence generation domain.}
-  \label{fig:white-rabbit-sleeps-grammar}
-\end{figure}
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{pic-derivation-2}
-  \caption{Derivation of ``The white rabbit sleeps.''}
-  \label{fig:white-rabbit-sleeps-deriv}
-\end{figure}
-
-We illustrate this process on a simplified example.  Consider a
-knowledge base containing the individuals $e$, $r_1$ and $r_2$, and a
-set of attributes encoding the fact that $r_1$ and $r_2$ are rabbits,
-$r_1$ is white and $r_2$ is brown, and $e$ is an event in which $r_1$
-sleeps.  Say that we want to express the information
-$\{\mathsf{sleep}(e,r_1)\}$ using the tree-adjoining grammar shown in
-Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
-\emph{elementary trees} (i.e., the disjoint trees in the figure), each
-of which contributes certain \emph{semantic content}. We can
-instantiate these trees by substituting individuals for \emph{semantic
-  roles}, such as $\mathsf{self}$ and $\mathsf{subj}$, and then
-combine the tree instances as shown in
-Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
-``The white rabbit sleeps''.
-
-We compute this grammatical derivation in a top-down manner, starting with
-the elementary tree for ``sleeps''. This tree satisfies the need to convey
-the semantic information, but introduces a need to generate a noun phrase
-(NP) for the subject; this NP must refer uniquely to the target referent
-$r_1$. In a second step, we substitute the tree for ``the rabbit'' into the
-open NP leaf, which makes the derivation grammatically complete. Since
-there are two different individuals that could be described as ``the
-rabbit''---technically, $r_2$ is still a \emph{distractor} (i.e., based on
-the description ``the rabbit'', the hearer might erroneously think that
-we're talking about $r_2$ and not $r_1$)---we are still not finished. To
-complete the derivation, the tree for ``white'' is added to the existing
-structure by an \emph{adjunction} operation, making the derivation
-both syntactically and semantically complete.
-
-\begin{figure}[p]
-\centering
-\begin{minipage}{0.5\textwidth}
-{\small%
-\begin{verbatim}
-(:action add-sleeps
-   :parameters (?u - node
-                ?xself - individual
-                ?xsubj - individual)
-   :precondition
-       (and (subst S ?u)
-            (referent ?u ?xself)
-            (sleep ?xself ?xsubj))
-   :effect 
-       (and (not (subst S ?u))
-            (expressed sleep ?xself ?xsubj)
-            (subst NP (subj ?u))
-            (referent (subj ?u) ?xsubj)
-            (forall (?y - individual)
-                (when (not (= ?y ?xself))
-                    (distractor (subj ?u) ?y)))))
-
-(:action add-rabbit
-   :parameters (?u - node
-                ?xself - individual)
-   :precondition 
-       (and (subst NP ?u)
-            (referent ?u ?xself)
-            (rabbit ?xself))
-   :effect 
-       (and (not (subst NP ?u))
-            (canadjoin N ?u)
-            (forall (?y - individual)
-                (when (not (rabbit ?y))
-                    (not (distractor ?u ?y))))))
-
-(:action add-white
-   :parameters (?u - node
-                ?xself - individual)
-   :precondition 
-       (and (canadjoin N ?u)
-            (referent ?u ?xself)
-            (rabbit ?xself))
-   :effect 
-       (forall (?y - individual)
-           (when (not (white ?y))
-               (not (distractor ?u ?y)))))
-\end{verbatim}}%
-\end{minipage}
-\caption{PDDL actions for generating the sentence ``The white rabbit
-sleeps.''}
-\label{fig:white-rabbit-as-planning}
-\end{figure}
-
-The process described above has clear parallels to planning: we manipulate
-states by applying sequences of actions to achieve a goal. We can make this
-connection even more precise by translating the SPUD problem into a
-planning problem.\footnote{See
-  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
-  which this conversion is implemented.}  For instance,
-Figure~\ref{fig:white-rabbit-as-planning} shows the corresponding PDDL
-actions for the above generation task, where each action corresponds
-to an operation that adds a single elementary tree to the derivation.
-In each case, the first parameter of the action is a node name in the
-derivation tree, and the remaining parameters stand for the
-individuals to which the semantic roles will be instantiated.  The
-syntactic preconditions and effects are encoded using $\mathsf{open}$
-and $\mathsf{canadjoin}$ literals; the status of each referring expression
-is tracked using $\mathsf{distractor}$ literals. The resulting action
-effects contain terms of the form $\mathsf{subj}(u)$, which construct new
-node names, but aren't strictly permitted by PDDL. In order to meet the
-syntactic requirements of PDDL, these terms can be eliminated by estimating
-an upper bound $n$ for the plan length, making $n$ copies of each action,
-ensuring that copy $i$ can only be applied in step $i$, and replacing the
-term $\mathsf{subj}(u)$ in an action copy by the constant
-$\mathsf{subj}_i$.
-
-Once the appropriate actions are defined, we can solve the generation
-problem as an ordinary planning problem. For instance, the following plan
-solves our previous example:
-%
-\begin{enumerate}
-\item $\mathsf{add}\textsf{-}\mathsf{sleeps}(\mathsf{root}, r_1)$,
-\item $\mathsf{add}\textsf{-}\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}),r_1)$,
-\item $\mathsf{add}\textsf{-}\mathsf{white}(\mathsf{subj}(\mathsf{root}),r_1)$.
-\end{enumerate}
-%
-Using this plan, the grammatical derivation in
-Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
-generated sentence ``the white rabbit sleeps'', can be systematically
-reconstructed. Thus, we can solve the sentence generation problem via
-the detour through planning and bring current search heuristics for
-planning to bear on generation.
-
-\subsection{Planning in instruction giving}
-
-The object of the GIVE Challenge (``Generating Instructions in Virtual
-Environments''; \citealt{alexander07:_shared_task_propos}) is to build an
-NLG system which is able to produce natural-language instructions which
-will guide a human user in performing some task in a virtual environment.
-From an NLG perspective, GIVE makes for an interesting challenge because it
-is a theory-neutral task that exercises all components of an NLG system,
-and emphasizes the study of communication in a (simulated) physical
-environment. It also has the advantage that the user and the NLG system can
-be in physically different places, as long as the 3D client and the NLG
-system are connected over a network. This makes it possible to evaluate
-GIVE NLG systems on a large scale over the Internet.
-%
-The first GIVE evaluation is currently underway. Five NLG systems from
-three countries are in the process of being evaluated. At the time of
-writing, 1028 different evaluation runs---each consisting of one user
-interacting with one instance of the virtual environment and being guided
-by an NLG system---have been collected, making GIVE the largest-scale
-NLG evaluation effort ever.\footnote{See the GIVE website at
- \url{http://www.give-challenge.org/} for more details about this
- challenge.}
-
-\begin{figure}[t]
-\centering
-%\includegraphics[width=1 \columnwidth]{give_world_no_expl}
-\includegraphics[width=0.75\columnwidth]{give_world_2}
-\caption{Map of the GIVE development world.}
-  \label{fig:give-development-world}
-\end{figure}
-
-An example GIVE world map is shown in
-Figure~\ref{fig:give-development-world}. To simplify both the planning and
-the NLG task, the world is discretised into a set of tiles of equal size.
-The user can turn by 90 degree steps in either direction, and can move from
-the centre of one tile to the centre of the next tile, provided the path
-between two tiles is not blocked. The world also consists of a set of
-objects which can be manipulated by the user in various ways. For
-instance, in the example world the user's task is to pick up a trophy in
-the top left room. The trophy is hidden in a safe behind a picture, so the
-user must first move the picture out of the way and open the safe by
-pushing a sequence of buttons (the small square boxes on the walls) in the
-correct order. In order to get to all these buttons, the user must also
-open the door in the centre of the map and deactivate the alarm tile by
-pushing further buttons. Figure~\ref{fig:give-planning} shows the encoding
-of some of the available GIVE domain actions, in PDDL syntax.
-
-\begin{figure}[p]
-\centering
-\begin{minipage}{0.5\textwidth}
-{\small%
-\begin{verbatim}
-(:action move
-   :parameters (?from - position
-                ?to - position
-                ?ori - orientation)
-   :precondition 
-       (and (player-pos ?from) 
-            (adjacent ?from ?to ?ori) 
-            (player-orient ?ori)
-            (not-blocked ?from ?to)
-            (not-alarmed ?to))
-   :effect 
-       (and (not (player-pos ?from))
-            (player-pos ?to)))
-
-(:action turn-left
-   :parameters (?ori - orientation
-                ?newOri - orientation)
-   :precondition 
-       (and (player-orient ?ori)
-            (next-orient-left ?ori ?newOri))
-   :effect 
-       (and (not (player-orient ?ori))
-            (player-orient ?newOri)))
-
-(:action turn-right
-   :parameters (?ori - orientation
-                ?newOri - orientation)
-   :precondition 
-       (and (player-orient ?ori)
-            (next-orient-right ?ori ?newOri))
-   :effect 
-       (and (not (player-orient ?ori))
-            (player-orient ?newOri)))
-
-(:action manipulate-b1-off-on
-   :parameters (?pos - position)
-   :precondition 
-       (and (state b1 off)
-            (player-pos ?pos)
-            (position b1 ?pos))
-   :effect 
-       (and (not (state b1 off))
-            (state b1 on)
-            (not (state d1 closed))
-            (state d1 open) 
-            (not (blocked pos_6_5 pos_6_4))
-            (not (blocked pos_6_4 pos_6_5))))
-\end{verbatim}}%
-\end{minipage}
-\caption{PDDL actions for the GIVE domain.}
-\label{fig:give-planning}
-\end{figure}
-
-The plans generated in the GIVE domain are often nontrivial. For instance,
-in the above example the shortest plan consists of 108 action steps, with
-the first few steps as follows:
-%
-\begin{enumerate}
-\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north},
-\mathsf{west})$,
-\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
-\end{enumerate}
-
-Most of the actions in the above plan are ``move'' actions. This makes the
-bulk of the GIVE problem very similar to that of a Gridworld problem (see,
-e.g., \citep{Tovey-Koenig:2000}), which also involves finding a route
-through a two-dimensional world map with discrete positions. GIVE tasks are
-usually much more complex, however, as illustrated by the above example
-map: a successful plan must include steps for pressing buttons in the right
-order, reasoning about large numbers of world objects, and navigating
-through complicated room shapes.
-
-The GIVE task comes with a number of interesting research challenges for
-both the NLG and planning communities. On the NLG side, one major problem
-is that verbalising each action instance in a plan makes for very
-poor natural-language instructions. For instance, the NLG system could
-generate an instruction sequence starting with ``turn left; walk forward;
-press the button; turn right; walk forward; walk forward; turn right; walk
-forward; walk forward; turn left; walk forward''. However, directions of
-this form are linguistically clumsy and uninteresting, and it is to be
-expected that frustrated users will quickly cancel an interaction with such
-a system, resulting in a bad evaluation score. Having just pressed the
-button, it would be much better to say ``turn around and walk through the
-door''. That is, the system should \emph{summarise} plans by merging
-multiple action instances into single instructions before presenting them
-to the user.\footnote{This task is similar to the construction of
-  ``macro'' actions, but is primarily intended for plan presentation
-  purposes. See, e.g., \citep{Botea-etal:05} which contains a survey of
-  recent approaches for generating macro actions in planning.}
-Conversely, it may also be necessary to \emph{elaborate} on a single
-planning step by expressing it with several instructions. Having just
-entered the top left room, it may be easier for the user to understand the
-instruction ``walk to the centre of the room; turn right; now press the
-green button in front of you'', rather than the instruction ``press the
-green button on the wall to your right''. Furthermore, in order to plan the
-referring expression ``the green button in front of you'' at a time when
-the button is not yet in front of the user, the NLG system must keep track
-of the hypothetical changes in what objects are visible to the user. Thus,
-the NLG acts of referring and instructing must be tightly integrated with
-the modules for plan generation.
-
-From a planning perspective, GIVE imposes very strict runtime
-requirements on the plan generation task: planning must
-happen in real time and the system must respond to a user in a timely
-fashion.  If the system takes too much time deliberating over an
-instruction to give, rather than actually giving this instruction, the
-user may have walked or turned away, thus making the instruction
-invalid.  Furthermore, plan execution monitoring also plays an
-important role in the GIVE problem.  At a high level, the system needs
-to monitor a user's actions and compare them against the generated
-instruction set to determine if the user has correctly followed
-directions or not. In the case of the latter, new instructions may
-have to be generated. In practice, the situation can be quite
-complicated since the mental state of the user is not known and so the
-system must be inferred from observing the user's actions in real
-time. For example, a user directed to ``turn around and walk through
-the door'' may not necessarily perform these actions to the letter,
-i.e., immediately turning 180 degrees and proceeding directly to the
-door. Instead, the user might take a roundabout route through the
-room, eventually exiting out the door.  Although the user's actions do
-not match the generated instructions exactly, they meet the intended
-goal. The system must be able to identify such ``equivalent'' plans
-and not immediately generate new instructions as soon as the user's
-actions have gone off course. Furthermore, a user can communicate
-certain intentions to the system, both through action and
-inaction. For instance, the system should infer that a user has failed
-to follow instructions if the user exits a room when given a directive
-to ``walk to the centre of the room''. The system should also make a
-similar conclusion if a user simply does nothing when given the
-instruction.
-
-\section{Experiments}
-\label{sec:experiments}
-
-We now describe the results of four simple experiments designed to evaluate
-the performance of three planners---FF \citep{HoffmannNebel01}, SGPLAN
-5.2.2 \citep{hsu06:_new_featur_in_sgplan_for}, and an ad-hoc implementation
-of GraphPlan \citep{Blum1997} written in Java---in our NLG domains. The FF
-and SGPLAN planners were chosen due to their success in the International
-Planning Competitions. Our implementation of GraphPlan is included as a
-counterpoint to the other two, more complex, planners. Most notably, our
-version of GraphPlan does not perform a preprocessing ``grounding'' step
-like FF and SGPLAN but is much more selective in its predicate and action
-instantiations.
-
-On first impressions, our results indicate that planning is a promising
-tool for both domains.  In the sentence generation domain, FF dramatically
-outperforms the best previously known algorithm for the same problem (a
-reimplementation of \citep{Stone2003a}), although the latter approach uses
-a greedy search algorithm with a heuristic that is hand-tailored to the
-domain. In the GIVE domain, SGPLAN computes a domain plan from the initial
-state to the goal in 0.3 seconds, which is fast enough for moderately-sized
-problems instances in the application.\footnote{All
-  runtimes were measured on a Pentium 4 CPU running at 3 GHz. Java programs
-  were allowed to ``warm up'', i.e., the planner was run five times and the
-  first four measurements discarded to ensure that the JVM had just-in-time
-  compiled all relevant bytecode.}
-
-We also observe that although FF and SGPLAN manage the search for plans
-very efficiently, they both spend comparatively large amounts of time
-computing predicate and action instantiations in their preprocessing
-stages, many of which are then never used during the search. FF's
-``grounding time'' in particular dominates its overall planning time,
-leading to some conflicting results. In some cases, our Java implementation
-of GraphPlan significantly outperforms both state-of-the-art planners. The
-experiments described below seek to improve our understanding of this
-situation.
-
-\subsection{Experiment 1: Sentence generation}
-\label{sec:exper-1:-sent}
-
-In the first experiment, we construct a series of sentence generation
-problems which require the planner to compute a plan representing the
-sentence ``Mary likes the Adj$_1$ \ldots Adj$_n$ rabbit.''  Each problem
-instance assumes a certain number $m$ of rabbits that were distinguished by
-$n \leq m$ different properties, such that all $n$ properties are required
-to distinguish the target referent from all other rabbits.  The $n$
-properties are realized as $n$ different adjectives, in any order.  This
-setup allows us to control the plan length (a plan with $n$ properties will
-have length $n+4$) and the universe size (the universe will contain $m+1$
-individuals in addition to the differently-typed individuals used to encode
-the grammar).
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp1}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-modifiers-with-sgplan}
-  \caption{Results for the sentence generation domain. The
-    horizontal axis represents parameters $(m,n)$ from $(1,1)$ to
-    $(10,10)$ in lexicographical order. The vertical axis is the
-    runtime in milliseconds.}
-  \label{fig:runtimes-crisp}
-\end{figure}
-
-The results of this experiment are shown in the graph in
-Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
-in lexicographic order on the horizontal axis and the runtime is shown in
-milliseconds on the vertical axis. These results reveal a number of
-interesting insights. First, FF significantly outperforms SGPLAN in this
-domain, sometimes by a factor of 10 or more.\footnote{Experiments with
- SGPLAN use a pre-release version of SGPLAN, kindly provided to us by
- Chih-Wei Hsu.The release version of SGPLAN 5.2.2 had a bug which caused it
- to crash on some of our problem instances.}
-Second, FF's runtime is dominated by its initial grounding step, in which
-it computes the ground instances of all predicates and actions used in the
-planning problem in order to avoid unnecessary instantiation during search.
-In particular, the ratio of grounding time to total runtime is generally
-above 85\%, and rises to above 99\% at $m=11$, which is still a relatively
-small universe for this application.\footnote{The
-  ``grounding'' time reported here is what FF reports as ``time spent:
-  instantiating action templates''.}
-
-In our testing examples, the time spent by FF on grounding is such that it
-is consistently outperformed by our Java implementation of GraphPlan, which
-only computes instances of predicates and actions as they are discovered
-during the construction of the planning graph. Although FF is consistently
-much faster as far as pure search time is concerned, our results indicate
-that FF's performance is much more sensitive to the domain size: if we fix
-$n=1$, FF takes 60 ms to compute a plan at $m=1$, but 2.4 seconds to
-compute the same plan at $m=10$. By comparison, our GraphPlan
-implementation takes 30 ms at $m=1$ and still only requires 50 ms at
-$m=10$. Conversely, GraphPlan's runtime grows much faster with the plan
-size (i.e., with growing values of $n$ for a fixed $m$). Larger, but still
-realistically-sized, instances of the sentence generation problem are still
-problematic for all the planners we tested.
-
-\subsection{Experiment 2: Minimal GIVE worlds}
-\label{sec:exper-2:-minim}
-
-In the second experiment, we evaluate the performance of the planners on
-problems arising in the GIVE domain. We construct a series of test worlds,
-similar to the one illustrated in Figure~\ref{fig:give-minimal}. These
-worlds consist of a $2n$ by $h$ grid of positions, such that there are
-buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The
-player starts in position $(1,1)$ and must press all the buttons
-to successfully complete the game. The world is generated as a GIVE world
-description, and then automatically converted into a planning problem by
-the GIVE software. For instance, Figure~\ref{fig:give-planning} shows some
-of the actions available in the GIVE domain in PDDL syntax.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.5\columnwidth]{pic-buttons}
-  \caption{Minimal GIVE world.}
-  \label{fig:give-minimal}
-\end{figure}
-
-Results for the $h=20$ case, with $n$ ranging from $1$ to $40$, are shown
-in Figure~\ref{fig:give-runtime-minimal}. The most obvious result is that
-FF is unable to solve any problems beyond $n=13$ on our experimentation
-machine within the memory limit of 1 GB. SGPLAN, on the other hand, solves
-instances beyond $n=40$ without major problems. In this case, our
-implementation of GraphPlan is unable to find any of these plans. The time
-spent on grounding is not a major factor in either planner, probably
-because the planners need more time to actually compute the plan---for
-instance, the optimal plan for the problem $n=40$ has a length of about
-1600 steps. As a concrete example, the following is a minimal plan for the
-case of a $2$ by $2$ grid with $2$ buttons for the player to press (i.e.,
-$n=1$, $h=2$):
-
-\begin{enumerate}
-\item $\mathsf{move}(\mathsf{pos\_1\_1},\mathsf{pos\_1\_2}, \mathsf{north})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{u1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_1\_2})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{north}, \mathsf{east})$,
-\item $\mathsf{move}(\mathsf{pos\_1\_2}, \mathsf{pos\_2\_2},
-  \mathsf{east})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{east}, \mathsf{south})$,
-\item $\mathsf{move}(\mathsf{pos\_2\_2}, \mathsf{pos\_2\_1}, \mathsf{south})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{l1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_2\_1})$.
-\end{enumerate}
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp2}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-buttons}
-  \caption{Results on the minimal GIVE
-    domain for $h=20$. The horizontal axis is $n$. The vertical axis
-    is the runtime in seconds.}
-  \label{fig:give-runtime-minimal}
-\end{figure}
-
-\subsection{Experiment 3: GIVE worlds with extra positions}
-\label{sec:experiment-3:-give}
-
-In the third experiment, we vary the structure of the GIVE world in order
-to judge the effect that universe size has on the planning problem in this
-domain. Starting with the ordinary GIVE world described in Experiment~2, we
-extend the world map by adding another $w$ by $h$ empty ``junk'' positions
-to the right of the minimal world, as shown in Figure~\ref{fig:give-junk}.
-These new positions are not actually required in any plan, but extend the
-size of the state space and approximate the situation in the actual GIVE
-domain where most grid positions are never used. We leave the initial state
-and goal untouched. As before, we generate a GIVE world description and
-then convert it into a planning problem in PDDL.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.80\columnwidth]{pic-empty-buttons}
-  \caption{GIVE world with extra ``junk'' positions.}
-  \label{fig:give-junk}
-\end{figure}
-
-Results for the $h=20$, $n=5$ case with $w$ ranging from $1$ to $70$ are
-shown in Figure~\ref{fig:give-runtime-junk}. As in Experiment~2, FF again
-runs out of memory, this time at $w=17$, while SGPLAN easily solves inputs
-beyond $w=70$. As in Experiment~2, our implementation of GraphPlan is
-unable to generate any of these plans. Unlike Experiment 2, however, both
-FF and SGPLAN now spend a substantial proportion of their time on
-grounding. In SGPLAN, this translates to a ``parsing time'' (which we
-assume includes grounding) which grows from 180 ms to 21.7 seconds as $w$
-grows from $1$ to $75$. The rest of the runtime, which also includes the
-search time, only grows from 400 ms to 2.3 seconds. This difference is
-particularly dramatic given that the actual optimal plan in each case is an
-identical plan of about 100 steps (i.e., the one we would have found in
-Experiment 2 for the $h=20$, $n=5$ case). The planning times for these
-instances are also concerning since times over a couple seconds will
-negatively affect the overall response time of the system, which must react
-in real time to user actions.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp3}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
-  \caption{Results on the GIVE domain with junk
-    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
-    The vertical axis is the runtime in seconds.}
-  \label{fig:give-runtime-junk}
-\end{figure}
-
-\subsection{Experiment 4: GIVE worlds with inaccessible positions}
-\label{sec:experiment-4:-give}
-
-In the final experiment, we consider a variation on the GIVE worlds from
-Experiment~3. As before, we begin with a base grid of $2n$ by $h$
-positions, with buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i
-\leq n$. We then generate a second grid of $w$ by $h$ positions and place
-an additional button in this grid. Unlike Experiment~3, the second grid
-does not extend the first grid but is ``disconnected'' so its positions are
-inaccessible from the first grid, as shown in
-Figure~\ref{fig:give-junk-nosoln}. Since the geometry of the grid makes it
-impossible to construct a plan for pressing all the buttons in the world,
-we instead investigate the time it takes a planner to arrive at the
-conclusion that the problem cannot be solved.
-
-Results for the $h=20$, $n=5$ case with $w$ ranging from 1 to 50 are shown
-in Figure~\ref{fig:give-runtime-nosoln}.\footnote{FF and SGPLAN do
- not display runtimes when they fail to construct a plan so an external timing
- program was used to generate these results.}
-In this case, the runtimes for FF rise sharply around $w=28$, producing
-times that are unacceptable for an NLG system in practice. FF is unable
-to complete problem instances above $w=35$. SGPLAN, by comparison, performs
-exceptionally well and is able to complete problems instances up to $w=50$
-in about 0.3 seconds and $w=80$ in about 0.6 seconds. These results
-demonstrate that SGPLAN has a significant advantage over FF in this
-experiment. 
-%Our implementation of GraphPlan is unable to complete any of
-%these problem instances.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.85\columnwidth]{pic-empty-inaccessible}
-  \caption{GIVE world with inaccessible positions.}
-  \label{fig:give-junk-nosoln}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.85\columnwidth]{graph-exp4}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
-  \caption{Results on the GIVE domain with inaccessible
-    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
-    The vertical axis is the runtime in seconds.}
-  \label{fig:give-runtime-nosoln}
-\end{figure}
-
-\section{Discussion}
-\label{sec:discussion}
-
-The positive conclusion we draw from our simple set of experiments is that
-modern planners are fairly good at controlling the search for many of the
-moderately-sized NLG problems we looked at. This is particularly true for
-FF in the sentence generation domain, which generates nontrivial 14-word
-sentences in about two seconds. Our results also indicate there is still
-room for improvement, however: special-purpose algorithms are able to
-generate much larger referring expressions in milliseconds
-\citep{AreKolStr08}. Although search efficiency is not the main concern in
-these domains, it nevertheless becomes a factor in scenarios like
-Experiment~2, where the world is almost completely unconstrained and
-(unsurprisingly) the search becomes much slower.
-
-The most restrictive bottleneck in the two NLG domains we investigated is
-the initial grounding step that many modern planners perform. While this
-may be a good strategy for traditional planning benchmarks and IPC
-domains---where plan length often dominates universe size and an initial
-investment into grounding pays off in saved instantiations during
-search---this approach has a very pronounced effect in domains like ours.
-As our experiments show, there are natural planning domains in which
-relatively short plans must be computed in large universes. For instance,
-in the sentence generation domain, it is not unrealistic to look for plans
-of length 20 over a universe consisting of thousands of domain individuals
-and tens of thousands of possible action instantiations, some of which
-require three domain individuals as parameters. Planners like FF and SGPLAN
-are less effective in such domains, to the point that grounding
-often dominates the total running time of these planners. (In the case of
-Experiment~1, our ad-hoc implementation of GraphPlan outperforms both
-planners in the sentence generation task.)
-
-Since our domains are not that unusual in their structure and composition,
-we hope that the lessons learned from our experiences can help improve the
-performance of current planning systems. For instance, the common strategy
-of splitting the universe into several types of individuals will not be a
-very effective solution in our NLG domains given the large number of
-individuals that could exist for a given type. On the other hand,
-sophisticated techniques based on reachability analysis may have the
-potential to improve this situation, provided such approaches can avoid the
-drawbacks of unnecessary grounding.\footnote{See
-  \citep{Bryce:07} for a good survey on reachability heuristics for planning
-  graph-based approaches to plan generation.}
-While recent trends in planning research have (rightfully) focused on the
-development of algorithms that control search in sophisticated ways,
-resulting in state-of-the-art planners that are more powerful and
-successful than their predecessors, we nevertheless believe that
-improvements are necessary if planning is to become a more mature
-technology that can offer tools to a wider community of users and
-researchers. At the end of the day, real-world users will care about the
-\emph{total} runtime of a planner, which is more than just the search time.
-
-By and large, our experiences with the planning community from a user's
-point of view have been favourable. Thanks to the planning competitions, it
-is easy to identify and download a fast state-of-the-art planner, at least
-for Linux. However, in the course of our experiments we found and reported
-bugs in both FF and SGPLAN. We also discovered that deciding as to which
-planner is appropriate for a particular task is not always straightforward,
-as the best performing planner in the IPC is not necessarily suitable for
-every planning task. As our experiments have shown, even planners as
-closely related as FF and SGPLAN can differ significantly in their
-performance on different domains (for instance, our results in
-Experiment~4).
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-In this paper, we introduced two novel planning domains arising from
-problems in natural language generation: the sentence generation domain and
-the Generating Instructions in Virtual Environments (GIVE) challenge
-domain. We also reported the results of a set of experiments designed to
-evaluate the suitability of off-the-shelf planners (particularly FF and
-SGPLAN) in these domains.
-
-Our results were mixed. While the planners we tested do an impressive job
-of controlling the complexity of search in many domains, they also suffer
-from practical problems that limit their performance in unexpected ways: in
-both of our NLG domains the grounding step performed by FF and SGPLAN
-dominates the time it takes to perform the search itself. As future work,
-we hope to extend our analysis to other planners that do not perform the
-initial grounding step, planners such as SHOP2
-\citep{DBLP:journals/jair/NauAIKMWY03} that have traditionally performed
-well on a wide range of domains, and new planners introduced in the latest
-edition of the IPC.\footnote{See
- \url{http://ipc.informatik.uni-freiburg.de/} for details of the 2008
- International Planning Competition (deterministic track).}
-
-The NLG community's recent interest in planning presents a valuable
-opportunity for planning researchers. Domains like GIVE highlight certain
-challenges, such as plan execution monitoring and plan presentation (i.e.,
-summarisation and elaboration), but also offer a platform on which such
-technologies can be evaluated in experiments with human users. Although we
-have focused on classical planning problems in this work, research related
-to reasoning under uncertainty, resource management, and planning with
-knowledge and sensing, can also be investigated in these settings. As such,
-we believe our domains would provide suitable challenges for planners
-entered in future editions of the IPC. Provided some of the challenges we
-have identified can be addressed, NLG-inspired problem domains offer a
-constructive platform for the planning community to showcase their
-techniques to a wider audience---and to improve the quality of their tools
-for real-world planning tasks.
-
-\section*{Acknowledgements}
-
-This work arose in the context of the Planning and Language Interest Group
-at the University of Edinburgh. The authors would like to thank all members
-of this group, especially Hector Geffner and Mark Steedman, for interesting
-discussions. This work was partially supported by the DFG Research
-Fellowship ``CRISP: Efficient integrated realization and microplanning''
-and by the European Commission through the PACO-PLUS project
-(FP6-2004-IST-4-27657).
-
-\bibliography{references}
-\bibliographystyle{ci}
-
-\pagebreak
-\listoffigures
-
-\end{document}
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "manuscript"
+%%% End:

### 2009-05-07 trunk/Papers/ci-spark/manuscript.tex
split paper up into individual files; changes to abstract and intro

--- 2009-04-27 trunk/Papers/ci-spark/manuscript.tex
@@
 
+\newcommand{\todo}[1]{\textbf{(#1)}}
+
 \begin{document}
@@
 
-%%%
-%%% Abstract
-%%%
 \pagebreak
-\begin{abstract}
-We investigate the application of modern planning techniques to domains
-arising from problems in natural language generation (NLG). In particular,
-we consider two novel NLG-inspired planning problems, the sentence
-generation domain and the GIVE (``Generating Instructions in Virtual
-Environment'') domain, and investigate the efficiency of the FF and SGPLAN
-planners in these domains. We also compare our results against an ad-hoc
-implementation of GraphPlan in Java. Our results are mixed. While modern
-planners are able to quickly solve many moderately-sized instances of our
-problems, the overall planning time is dominated by the grounding step that
-these planners perform, rather than search. This has a pronounced effect on
-our domains which require relatively short plans but have large universes.
-We share our experiences and offer these domains as challenges for the
-planning community.
-\end{abstract}
+\input abstract
 
-\bigskip\noindent
-\textbf{Keywords:} natural language generation, planning
+\pagebreak
 
-%%%
-%%% Main text
-%%%
-\pagebreak
-\section{Introduction}
-\label{sec:introduction}
-
-Natural language generation (NLG; \citealp{reiter00building}) is one of the
-major subfields of natural language processing, concerned with computing
-natural language sentences or texts that convey a given piece of
-information to the user. Intuitively, this task can be viewed as a problem
-involving actions, beliefs, and goals: an agent communicating with another
-agent tries to change the mental state of the hearer by applying actions
-which correspond to the utterance of words or sentences.
-This characterization of NLG suggests obvious parallels to automated
-planning---a view which has a long tradition in NLG.
-
-Early approaches to generation as planning
-\citep{perrault80,appelt:planning,hovy88,young94dpocl} focused primarily on
-planning high-level structures, such as speech acts and discourse
-relations.  While these approaches suffered from the inefficiency of the
-planners that were available at the time, there has recently been a
-resurgence of interest in applying modern planning techniques to NLG.
-These newer papers revisit speech act planning based on the more expressive
-planning formalisms available today \citep{Steedman-Petrick:07,benotti08b}
-and exploit the improved efficiency of modern planners for an integrated
-model of sentence planning and surface realization \citep{KolSto07}.
-Problems in NLG are often computationally challenging (see, e.g.,
-\citep{KolStr02}), and these new approaches combine an interest in planning
-as a modelling tool for natural language, with a hope of improved
-efficiency by exploiting modern approaches and algorithms.
-
-The focus of this paper is twofold. First, we present two novel planning
-domains that arise in the context of natural language generation: the
-sentence generation domain and the GIVE domain. In the sentence generation
-task, the goal is to generate a single sentence that expresses a given
-meaning. \citet{KolSto07} cast this domain as a planning problem, where the
-plan encodes the necessary sentence and the actions correspond to uttering
-individual words.  In the GIVE domain (``Generating Instructions in Virtual
-Environments''), we describe a new shared task that was recently posed as a
-challenge for the NLG community \citep{alexander07:_shared_task_propos}.
-GIVE uses planning as part of an NLG system that generates natural-language
-instructions to guide a user in performing a given task in a virtual
-environment.
-
-Second, we discuss some of our experiences using off-the-shelf
-planners in our two NLG-inspired domains. In particular, we explore the
-efficiency of FF \citep{HoffmannNebel01} and SGPLAN
-\citep{hsu06:_new_featur_in_sgplan_for}: two planners that are readily
-available, support an expressive subset of the Planning Domain Definition
-Language (PDDL; McDermott et al.~\citeyear{PDDL}) for encoding domains, and
-have been successful on traditional benchmarks and problems from the
-International Planning Competition (IPC)\footnote{See
- \url{http://ipc.icaps-conference.org/} for information about past
- editions of the IPC.}.
-We test these planners on a range of problem instances in our planning
-domains, and compare our results against an ad-hoc Java implementation of
-GraphPlan \citep{Blum1997}. Overall, our findings are mixed. On the one
-hand, we demonstrate that modern planners can easily handle some of the
-moderately-sized search problems that arise in NLG. In certain cases, the
-planners we tested are more efficient than specialized NLG algorithms
-applied to the same problem. On the other hand, these same planners spend
-tremendous amounts of time on preprocessing. For instance, in the sentence
-generation domain, FF spends 90\% of its runtime grounding out literals and
-actions, and ends up being slower than our version of GraphPlan that avoids
-grounding. For domains like ours, which are dominated by the number of
-actions and the universe size, rather than the combinatorics of the search
-problem, this observation suggests that the overall runtime of a planner
-could be improved by grounding more selectively. This is particularly true
-for large problem instances, which still remain a challenge for the
-planners we tested, and where the effects of the grounding approach become
-quite pronounced. We therefore offer our NLG domains as challenges for the
-planning community.
-
-\section{Planning in NLG}
-\label{sec:domains}
-
-We begin by presenting our two planning domains: the sentence generation
-domain and the instruction giving domain.
-
-\subsection{Sentence generation as planning}
-
-Sentence generation is the problem of computing from a grammar and a
-semantic representation, a single sentence that expresses this piece of
-meaning. Traditionally, this problem is split into several steps
-\citep{reiter00building}. In a first step, called \emph{sentence planning},
-the semantic representation is enriched with additional information. For
-instance, \emph{referring expressions}, which refer to individuals (e.g.,
-objects, people, etc.) in the domain of discourse, are determined at this
-point. In a second step, called \emph{surface realization}, this enriched
-representation is translated into a natural-language sentence, using the
-grammar.
-
-In practice, the process of determining referring expressions typically
-interacts with the realization step and so it often turns out to be
-beneficial to perform these two steps together. For instance, the SPUD
-system (``Sentence Planning Using Description''; \citealt{Stone2003a})
-performed this task using a top-down generation algorithm based on
-tree-adjoining grammars \citep{joshi;etal1997} whose lexical entries were
-equipped with semantic and pragmatic information. Unfortunately, SPUD had
-to resort to a non-optimal greedy search strategy to cope with its large
-search space and retain reasonable efficiency. To improve the efficiency
-of SPUD, \citet{KolSto07} translate the sentence generation problem into a
-planning problem and use a planning algorithm to perform the generation
-task.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{pic-grammar}
-  \caption{An example grammar in the sentence generation domain.}
-  \label{fig:white-rabbit-sleeps-grammar}
-\end{figure}
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{pic-derivation-2}
-  \caption{Derivation of ``The white rabbit sleeps.''}
-  \label{fig:white-rabbit-sleeps-deriv}
-\end{figure}
-
-We illustrate this process on a simplified example.  Consider a
-knowledge base containing the individuals $e$, $r_1$ and $r_2$, and a
-set of attributes encoding the fact that $r_1$ and $r_2$ are rabbits,
-$r_1$ is white and $r_2$ is brown, and $e$ is an event in which $r_1$
-sleeps.  Say that we want to express the information
-$\{\mathsf{sleep}(e,r_1)\}$ using the tree-adjoining grammar shown in
-Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
-\emph{elementary trees} (i.e., the disjoint trees in the figure), each
-of which contributes certain \emph{semantic content}. We can
-instantiate these trees by substituting individuals for \emph{semantic
-  roles}, such as $\mathsf{self}$ and $\mathsf{subj}$, and then
-combine the tree instances as shown in
-Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
-``The white rabbit sleeps''.
-
-We compute this grammatical derivation in a top-down manner, starting with
-the elementary tree for ``sleeps''. This tree satisfies the need to convey
-the semantic information, but introduces a need to generate a noun phrase
-(NP) for the subject; this NP must refer uniquely to the target referent
-$r_1$. In a second step, we substitute the tree for ``the rabbit'' into the
-open NP leaf, which makes the derivation grammatically complete. Since
-there are two different individuals that could be described as ``the
-rabbit''---technically, $r_2$ is still a \emph{distractor} (i.e., based on
-the description ``the rabbit'', the hearer might erroneously think that
-we're talking about $r_2$ and not $r_1$)---we are still not finished. To
-complete the derivation, the tree for ``white'' is added to the existing
-structure by an \emph{adjunction} operation, making the derivation
-both syntactically and semantically complete.
-
-\begin{figure}[p]
-\centering
-\begin{minipage}{0.5\textwidth}
-{\small%
-\begin{verbatim}
-(:action add-sleeps
-   :parameters (?u - node
-                ?xself - individual
-                ?xsubj - individual)
-   :precondition
-       (and (subst S ?u)
-            (referent ?u ?xself)
-            (sleep ?xself ?xsubj))
-   :effect 
-       (and (not (subst S ?u))
-            (expressed sleep ?xself ?xsubj)
-            (subst NP (subj ?u))
-            (referent (subj ?u) ?xsubj)
-            (forall (?y - individual)
-                (when (not (= ?y ?xself))
-                    (distractor (subj ?u) ?y)))))
-
-(:action add-rabbit
-   :parameters (?u - node
-                ?xself - individual)
-   :precondition 
-       (and (subst NP ?u)
-            (referent ?u ?xself)
-            (rabbit ?xself))
-   :effect 
-       (and (not (subst NP ?u))
-            (canadjoin N ?u)
-            (forall (?y - individual)
-                (when (not (rabbit ?y))
-                    (not (distractor ?u ?y))))))
-
-(:action add-white
-   :parameters (?u - node
-                ?xself - individual)
-   :precondition 
-       (and (canadjoin N ?u)
-            (referent ?u ?xself)
-            (rabbit ?xself))
-   :effect 
-       (forall (?y - individual)
-           (when (not (white ?y))
-               (not (distractor ?u ?y)))))
-\end{verbatim}}%
-\end{minipage}
-\caption{PDDL actions for generating the sentence ``The white rabbit
-sleeps.''}
-\label{fig:white-rabbit-as-planning}
-\end{figure}
-
-The process described above has clear parallels to planning: we manipulate
-states by applying sequences of actions to achieve a goal. We can make this
-connection even more precise by translating the SPUD problem into a
-planning problem.\footnote{See
-  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
-  which this conversion is implemented.}  For instance,
-Figure~\ref{fig:white-rabbit-as-planning} shows the corresponding PDDL
-actions for the above generation task, where each action corresponds
-to an operation that adds a single elementary tree to the derivation.
-In each case, the first parameter of the action is a node name in the
-derivation tree, and the remaining parameters stand for the
-individuals to which the semantic roles will be instantiated.  The
-syntactic preconditions and effects are encoded using $\mathsf{open}$
-and $\mathsf{canadjoin}$ literals; the status of each referring expression
-is tracked using $\mathsf{distractor}$ literals. The resulting action
-effects contain terms of the form $\mathsf{subj}(u)$, which construct new
-node names, but aren't strictly permitted by PDDL. In order to meet the
-syntactic requirements of PDDL, these terms can be eliminated by estimating
-an upper bound $n$ for the plan length, making $n$ copies of each action,
-ensuring that copy $i$ can only be applied in step $i$, and replacing the
-term $\mathsf{subj}(u)$ in an action copy by the constant
-$\mathsf{subj}_i$.
-
-Once the appropriate actions are defined, we can solve the generation
-problem as an ordinary planning problem. For instance, the following plan
-solves our previous example:
-%
-\begin{enumerate}
-\item $\mathsf{add}\textsf{-}\mathsf{sleeps}(\mathsf{root}, r_1)$,
-\item $\mathsf{add}\textsf{-}\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}),r_1)$,
-\item $\mathsf{add}\textsf{-}\mathsf{white}(\mathsf{subj}(\mathsf{root}),r_1)$.
-\end{enumerate}
-%
-Using this plan, the grammatical derivation in
-Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
-generated sentence ``the white rabbit sleeps'', can be systematically
-reconstructed. Thus, we can solve the sentence generation problem via
-the detour through planning and bring current search heuristics for
-planning to bear on generation.
-
-\subsection{Planning in instruction giving}
-
-The object of the GIVE Challenge (``Generating Instructions in Virtual
-Environments''; \citealt{alexander07:_shared_task_propos}) is to build an
-NLG system which is able to produce natural-language instructions which
-will guide a human user in performing some task in a virtual environment.
-From an NLG perspective, GIVE makes for an interesting challenge because it
-is a theory-neutral task that exercises all components of an NLG system,
-and emphasizes the study of communication in a (simulated) physical
-environment. It also has the advantage that the user and the NLG system can
-be in physically different places, as long as the 3D client and the NLG
-system are connected over a network. This makes it possible to evaluate
-GIVE NLG systems on a large scale over the Internet.
-%
-The first GIVE evaluation is currently underway. Five NLG systems from
-three countries are in the process of being evaluated. At the time of
-writing, 1028 different evaluation runs---each consisting of one user
-interacting with one instance of the virtual environment and being guided
-by an NLG system---have been collected, making GIVE the largest-scale
-NLG evaluation effort ever.\footnote{See the GIVE website at
- \url{http://www.give-challenge.org/} for more details about this
- challenge.}
-
-\begin{figure}[t]
-\centering
-%\includegraphics[width=1 \columnwidth]{give_world_no_expl}
-\includegraphics[width=0.75\columnwidth]{give_world_2}
-\caption{Map of the GIVE development world.}
-  \label{fig:give-development-world}
-\end{figure}
-
-An example GIVE world map is shown in
-Figure~\ref{fig:give-development-world}. To simplify both the planning and
-the NLG task, the world is discretised into a set of tiles of equal size.
-The user can turn by 90 degree steps in either direction, and can move from
-the centre of one tile to the centre of the next tile, provided the path
-between two tiles is not blocked. The world also consists of a set of
-objects which can be manipulated by the user in various ways. For
-instance, in the example world the user's task is to pick up a trophy in
-the top left room. The trophy is hidden in a safe behind a picture, so the
-user must first move the picture out of the way and open the safe by
-pushing a sequence of buttons (the small square boxes on the walls) in the
-correct order. In order to get to all these buttons, the user must also
-open the door in the centre of the map and deactivate the alarm tile by
-pushing further buttons. Figure~\ref{fig:give-planning} shows the encoding
-of some of the available GIVE domain actions, in PDDL syntax.
-
-\begin{figure}[p]
-\centering
-\begin{minipage}{0.5\textwidth}
-{\small%
-\begin{verbatim}
-(:action move
-   :parameters (?from - position
-                ?to - position
-                ?ori - orientation)
-   :precondition 
-       (and (player-pos ?from) 
-            (adjacent ?from ?to ?ori) 
-            (player-orient ?ori)
-            (not-blocked ?from ?to)
-            (not-alarmed ?to))
-   :effect 
-       (and (not (player-pos ?from))
-            (player-pos ?to)))
-
-(:action turn-left
-   :parameters (?ori - orientation
-                ?newOri - orientation)
-   :precondition 
-       (and (player-orient ?ori)
-            (next-orient-left ?ori ?newOri))
-   :effect 
-       (and (not (player-orient ?ori))
-            (player-orient ?newOri)))
-
-(:action turn-right
-   :parameters (?ori - orientation
-                ?newOri - orientation)
-   :precondition 
-       (and (player-orient ?ori)
-            (next-orient-right ?ori ?newOri))
-   :effect 
-       (and (not (player-orient ?ori))
-            (player-orient ?newOri)))
-
-(:action manipulate-b1-off-on
-   :parameters (?pos - position)
-   :precondition 
-       (and (state b1 off)
-            (player-pos ?pos)
-            (position b1 ?pos))
-   :effect 
-       (and (not (state b1 off))
-            (state b1 on)
-            (not (state d1 closed))
-            (state d1 open) 
-            (not (blocked pos_6_5 pos_6_4))
-            (not (blocked pos_6_4 pos_6_5))))
-\end{verbatim}}%
-\end{minipage}
-\caption{PDDL actions for the GIVE domain.}
-\label{fig:give-planning}
-\end{figure}
-
-The plans generated in the GIVE domain are often nontrivial. For instance,
-in the above example the shortest plan consists of 108 action steps, with
-the first few steps as follows:
-%
-\begin{enumerate}
-\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north},
-\mathsf{west})$,
-\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
-\end{enumerate}
-
-Most of the actions in the above plan are ``move'' actions. This makes the
-bulk of the GIVE problem very similar to that of a Gridworld problem (see,
-e.g., \citep{Tovey-Koenig:2000}), which also involves finding a route
-through a two-dimensional world map with discrete positions. GIVE tasks are
-usually much more complex, however, as illustrated by the above example
-map: a successful plan must include steps for pressing buttons in the right
-order, reasoning about large numbers of world objects, and navigating
-through complicated room shapes.
-
-The GIVE task comes with a number of interesting research challenges for
-both the NLG and planning communities. On the NLG side, one major problem
-is that verbalising each action instance in a plan makes for very
-poor natural-language instructions. For instance, the NLG system could
-generate an instruction sequence starting with ``turn left; walk forward;
-press the button; turn right; walk forward; walk forward; turn right; walk
-forward; walk forward; turn left; walk forward''. However, directions of
-this form are linguistically clumsy and uninteresting, and it is to be
-expected that frustrated users will quickly cancel an interaction with such
-a system, resulting in a bad evaluation score. Having just pressed the
-button, it would be much better to say ``turn around and walk through the
-door''. That is, the system should \emph{summarise} plans by merging
-multiple action instances into single instructions before presenting them
-to the user.\footnote{This task is similar to the construction of
-  ``macro'' actions, but is primarily intended for plan presentation
-  purposes. See, e.g., \citep{Botea-etal:05} which contains a survey of
-  recent approaches for generating macro actions in planning.}
-Conversely, it may also be necessary to \emph{elaborate} on a single
-planning step by expressing it with several instructions. Having just
-entered the top left room, it may be easier for the user to understand the
-instruction ``walk to the centre of the room; turn right; now press the
-green button in front of you'', rather than the instruction ``press the
-green button on the wall to your right''. Furthermore, in order to plan the
-referring expression ``the green button in front of you'' at a time when
-the button is not yet in front of the user, the NLG system must keep track
-of the hypothetical changes in what objects are visible to the user. Thus,
-the NLG acts of referring and instructing must be tightly integrated with
-the modules for plan generation.
-
-From a planning perspective, GIVE imposes very strict runtime
-requirements on the plan generation task: planning must
-happen in real time and the system must respond to a user in a timely
-fashion.  If the system takes too much time deliberating over an
-instruction to give, rather than actually giving this instruction, the
-user may have walked or turned away, thus making the instruction
-invalid.  Furthermore, plan execution monitoring also plays an
-important role in the GIVE problem.  At a high level, the system needs
-to monitor a user's actions and compare them against the generated
-instruction set to determine if the user has correctly followed
-directions or not. In the case of the latter, new instructions may
-have to be generated. In practice, the situation can be quite
-complicated since the mental state of the user is not known and so the
-system must be inferred from observing the user's actions in real
-time. For example, a user directed to ``turn around and walk through
-the door'' may not necessarily perform these actions to the letter,
-i.e., immediately turning 180 degrees and proceeding directly to the
-door. Instead, the user might take a roundabout route through the
-room, eventually exiting out the door.  Although the user's actions do
-not match the generated instructions exactly, they meet the intended
-goal. The system must be able to identify such ``equivalent'' plans
-and not immediately generate new instructions as soon as the user's
-actions have gone off course. Furthermore, a user can communicate
-certain intentions to the system, both through action and
-inaction. For instance, the system should infer that a user has failed
-to follow instructions if the user exits a room when given a directive
-to ``walk to the centre of the room''. The system should also make a
-similar conclusion if a user simply does nothing when given the
-instruction.
-
-\section{Experiments}
-\label{sec:experiments}
-
-We now describe the results of four simple experiments designed to evaluate
-the performance of three planners---FF \citep{HoffmannNebel01}, SGPLAN
-5.2.2 \citep{hsu06:_new_featur_in_sgplan_for}, and an ad-hoc implementation
-of GraphPlan \citep{Blum1997} written in Java---in our NLG domains. The FF
-and SGPLAN planners were chosen due to their success in the International
-Planning Competitions. Our implementation of GraphPlan is included as a
-counterpoint to the other two, more complex, planners. Most notably, our
-version of GraphPlan does not perform a preprocessing ``grounding'' step
-like FF and SGPLAN but is much more selective in its predicate and action
-instantiations.
-
-On first impressions, our results indicate that planning is a promising
-tool for both domains.  In the sentence generation domain, FF dramatically
-outperforms the best previously known algorithm for the same problem (a
-reimplementation of \citep{Stone2003a}), although the latter approach uses
-a greedy search algorithm with a heuristic that is hand-tailored to the
-domain. In the GIVE domain, SGPLAN computes a domain plan from the initial
-state to the goal in 0.3 seconds, which is fast enough for moderately-sized
-problems instances in the application.\footnote{All
-  runtimes were measured on a Pentium 4 CPU running at 3 GHz. Java programs
-  were allowed to ``warm up'', i.e., the planner was run five times and the
-  first four measurements discarded to ensure that the JVM had just-in-time
-  compiled all relevant bytecode.}
-
-We also observe that although FF and SGPLAN manage the search for plans
-very efficiently, they both spend comparatively large amounts of time
-computing predicate and action instantiations in their preprocessing
-stages, many of which are then never used during the search. FF's
-``grounding time'' in particular dominates its overall planning time,
-leading to some conflicting results. In some cases, our Java implementation
-of GraphPlan significantly outperforms both state-of-the-art planners. The
-experiments described below seek to improve our understanding of this
-situation.
-
-\subsection{Experiment 1: Sentence generation}
-\label{sec:exper-1:-sent}
-
-In the first experiment, we construct a series of sentence generation
-problems which require the planner to compute a plan representing the
-sentence ``Mary likes the Adj$_1$ \ldots Adj$_n$ rabbit.''  Each problem
-instance assumes a certain number $m$ of rabbits that were distinguished by
-$n \leq m$ different properties, such that all $n$ properties are required
-to distinguish the target referent from all other rabbits.  The $n$
-properties are realized as $n$ different adjectives, in any order.  This
-setup allows us to control the plan length (a plan with $n$ properties will
-have length $n+4$) and the universe size (the universe will contain $m+1$
-individuals in addition to the differently-typed individuals used to encode
-the grammar).
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp1}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-modifiers-with-sgplan}
-  \caption{Results for the sentence generation domain. The
-    horizontal axis represents parameters $(m,n)$ from $(1,1)$ to
-    $(10,10)$ in lexicographical order. The vertical axis is the
-    runtime in milliseconds.}
-  \label{fig:runtimes-crisp}
-\end{figure}
-
-The results of this experiment are shown in the graph in
-Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
-in lexicographic order on the horizontal axis and the runtime is shown in
-milliseconds on the vertical axis. These results reveal a number of
-interesting insights. First, FF significantly outperforms SGPLAN in this
-domain, sometimes by a factor of 10 or more.\footnote{Experiments with
- SGPLAN use a pre-release version of SGPLAN, kindly provided to us by
- Chih-Wei Hsu.The release version of SGPLAN 5.2.2 had a bug which caused it
- to crash on some of our problem instances.}
-Second, FF's runtime is dominated by its initial grounding step, in which
-it computes the ground instances of all predicates and actions used in the
-planning problem in order to avoid unnecessary instantiation during search.
-In particular, the ratio of grounding time to total runtime is generally
-above 85\%, and rises to above 99\% at $m=11$, which is still a relatively
-small universe for this application.\footnote{The
-  ``grounding'' time reported here is what FF reports as ``time spent:
-  instantiating action templates''.}
-
-In our testing examples, the time spent by FF on grounding is such that it
-is consistently outperformed by our Java implementation of GraphPlan, which
-only computes instances of predicates and actions as they are discovered
-during the construction of the planning graph. Although FF is consistently
-much faster as far as pure search time is concerned, our results indicate
-that FF's performance is much more sensitive to the domain size: if we fix
-$n=1$, FF takes 60 ms to compute a plan at $m=1$, but 2.4 seconds to
-compute the same plan at $m=10$. By comparison, our GraphPlan
-implementation takes 30 ms at $m=1$ and still only requires 50 ms at
-$m=10$. Conversely, GraphPlan's runtime grows much faster with the plan
-size (i.e., with growing values of $n$ for a fixed $m$). Larger, but still
-realistically-sized, instances of the sentence generation problem are still
-problematic for all the planners we tested.
-
-\subsection{Experiment 2: Minimal GIVE worlds}
-\label{sec:exper-2:-minim}
-
-In the second experiment, we evaluate the performance of the planners on
-problems arising in the GIVE domain. We construct a series of test worlds,
-similar to the one illustrated in Figure~\ref{fig:give-minimal}. These
-worlds consist of a $2n$ by $h$ grid of positions, such that there are
-buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The
-player starts in position $(1,1)$ and must press all the buttons
-to successfully complete the game. The world is generated as a GIVE world
-description, and then automatically converted into a planning problem by
-the GIVE software. For instance, Figure~\ref{fig:give-planning} shows some
-of the actions available in the GIVE domain in PDDL syntax.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.5\columnwidth]{pic-buttons}
-  \caption{Minimal GIVE world.}
-  \label{fig:give-minimal}
-\end{figure}
-
-Results for the $h=20$ case, with $n$ ranging from $1$ to $40$, are shown
-in Figure~\ref{fig:give-runtime-minimal}. The most obvious result is that
-FF is unable to solve any problems beyond $n=13$ on our experimentation
-machine within the memory limit of 1 GB. SGPLAN, on the other hand, solves
-instances beyond $n=40$ without major problems. In this case, our
-implementation of GraphPlan is unable to find any of these plans. The time
-spent on grounding is not a major factor in either planner, probably
-because the planners need more time to actually compute the plan---for
-instance, the optimal plan for the problem $n=40$ has a length of about
-1600 steps. As a concrete example, the following is a minimal plan for the
-case of a $2$ by $2$ grid with $2$ buttons for the player to press (i.e.,
-$n=1$, $h=2$):
-
-\begin{enumerate}
-\item $\mathsf{move}(\mathsf{pos\_1\_1},\mathsf{pos\_1\_2}, \mathsf{north})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{u1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_1\_2})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{north}, \mathsf{east})$,
-\item $\mathsf{move}(\mathsf{pos\_1\_2}, \mathsf{pos\_2\_2},
-  \mathsf{east})$,
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{east}, \mathsf{south})$,
-\item $\mathsf{move}(\mathsf{pos\_2\_2}, \mathsf{pos\_2\_1}, \mathsf{south})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{l1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_2\_1})$.
-\end{enumerate}
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp2}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-buttons}
-  \caption{Results on the minimal GIVE
-    domain for $h=20$. The horizontal axis is $n$. The vertical axis
-    is the runtime in seconds.}
-  \label{fig:give-runtime-minimal}
-\end{figure}
-
-\subsection{Experiment 3: GIVE worlds with extra positions}
-\label{sec:experiment-3:-give}
-
-In the third experiment, we vary the structure of the GIVE world in order
-to judge the effect that universe size has on the planning problem in this
-domain. Starting with the ordinary GIVE world described in Experiment~2, we
-extend the world map by adding another $w$ by $h$ empty ``junk'' positions
-to the right of the minimal world, as shown in Figure~\ref{fig:give-junk}.
-These new positions are not actually required in any plan, but extend the
-size of the state space and approximate the situation in the actual GIVE
-domain where most grid positions are never used. We leave the initial state
-and goal untouched. As before, we generate a GIVE world description and
-then convert it into a planning problem in PDDL.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.80\columnwidth]{pic-empty-buttons}
-  \caption{GIVE world with extra ``junk'' positions.}
-  \label{fig:give-junk}
-\end{figure}
-
-Results for the $h=20$, $n=5$ case with $w$ ranging from $1$ to $70$ are
-shown in Figure~\ref{fig:give-runtime-junk}. As in Experiment~2, FF again
-runs out of memory, this time at $w=17$, while SGPLAN easily solves inputs
-beyond $w=70$. As in Experiment~2, our implementation of GraphPlan is
-unable to generate any of these plans. Unlike Experiment 2, however, both
-FF and SGPLAN now spend a substantial proportion of their time on
-grounding. In SGPLAN, this translates to a ``parsing time'' (which we
-assume includes grounding) which grows from 180 ms to 21.7 seconds as $w$
-grows from $1$ to $75$. The rest of the runtime, which also includes the
-search time, only grows from 400 ms to 2.3 seconds. This difference is
-particularly dramatic given that the actual optimal plan in each case is an
-identical plan of about 100 steps (i.e., the one we would have found in
-Experiment 2 for the $h=20$, $n=5$ case). The planning times for these
-instances are also concerning since times over a couple seconds will
-negatively affect the overall response time of the system, which must react
-in real time to user actions.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp3}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
-  \caption{Results on the GIVE domain with junk
-    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
-    The vertical axis is the runtime in seconds.}
-  \label{fig:give-runtime-junk}
-\end{figure}
-
-\subsection{Experiment 4: GIVE worlds with inaccessible positions}
-\label{sec:experiment-4:-give}
-
-In the final experiment, we consider a variation on the GIVE worlds from
-Experiment~3. As before, we begin with a base grid of $2n$ by $h$
-positions, with buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i
-\leq n$. We then generate a second grid of $w$ by $h$ positions and place
-an additional button in this grid. Unlike Experiment~3, the second grid
-does not extend the first grid but is ``disconnected'' so its positions are
-inaccessible from the first grid, as shown in
-Figure~\ref{fig:give-junk-nosoln}. Since the geometry of the grid makes it
-impossible to construct a plan for pressing all the buttons in the world,
-we instead investigate the time it takes a planner to arrive at the
-conclusion that the problem cannot be solved.
-
-Results for the $h=20$, $n=5$ case with $w$ ranging from 1 to 50 are shown
-in Figure~\ref{fig:give-runtime-nosoln}.\footnote{FF and SGPLAN do
- not display runtimes when they fail to construct a plan so an external timing
- program was used to generate these results.}
-In this case, the runtimes for FF rise sharply around $w=28$, producing
-times that are unacceptable for an NLG system in practice. FF is unable
-to complete problem instances above $w=35$. SGPLAN, by comparison, performs
-exceptionally well and is able to complete problems instances up to $w=50$
-in about 0.3 seconds and $w=80$ in about 0.6 seconds. These results
-demonstrate that SGPLAN has a significant advantage over FF in this
-experiment. 
-%Our implementation of GraphPlan is unable to complete any of
-%these problem instances.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.85\columnwidth]{pic-empty-inaccessible}
-  \caption{GIVE world with inaccessible positions.}
-  \label{fig:give-junk-nosoln}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.85\columnwidth]{graph-exp4}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
-  \caption{Results on the GIVE domain with inaccessible
-    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
-    The vertical axis is the runtime in seconds.}
-  \label{fig:give-runtime-nosoln}
-\end{figure}
-
-\section{Discussion}
-\label{sec:discussion}
-
-The positive conclusion we draw from our simple set of experiments is that
-modern planners are fairly good at controlling the search for many of the
-moderately-sized NLG problems we looked at. This is particularly true for
-FF in the sentence generation domain, which generates nontrivial 14-word
-sentences in about two seconds. Our results also indicate there is still
-room for improvement, however: special-purpose algorithms are able to
-generate much larger referring expressions in milliseconds
-\citep{AreKolStr08}. Although search efficiency is not the main concern in
-these domains, it nevertheless becomes a factor in scenarios like
-Experiment~2, where the world is almost completely unconstrained and
-(unsurprisingly) the search becomes much slower.
-
-The most restrictive bottleneck in the two NLG domains we investigated is
-the initial grounding step that many modern planners perform. While this
-may be a good strategy for traditional planning benchmarks and IPC
-domains---where plan length often dominates universe size and an initial
-investment into grounding pays off in saved instantiations during
-search---this approach has a very pronounced effect in domains like ours.
-As our experiments show, there are natural planning domains in which
-relatively short plans must be computed in large universes. For instance,
-in the sentence generation domain, it is not unrealistic to look for plans
-of length 20 over a universe consisting of thousands of domain individuals
-and tens of thousands of possible action instantiations, some of which
-require three domain individuals as parameters. Planners like FF and SGPLAN
-are less effective in such domains, to the point that grounding
-often dominates the total running time of these planners. (In the case of
-Experiment~1, our ad-hoc implementation of GraphPlan outperforms both
-planners in the sentence generation task.)
-
-Since our domains are not that unusual in their structure and composition,
-we hope that the lessons learned from our experiences can help improve the
-performance of current planning systems. For instance, the common strategy
-of splitting the universe into several types of individuals will not be a
-very effective solution in our NLG domains given the large number of
-individuals that could exist for a given type. On the other hand,
-sophisticated techniques based on reachability analysis may have the
-potential to improve this situation, provided such approaches can avoid the
-drawbacks of unnecessary grounding.\footnote{See
-  \citep{Bryce:07} for a good survey on reachability heuristics for planning
-  graph-based approaches to plan generation.}
-While recent trends in planning research have (rightfully) focused on the
-development of algorithms that control search in sophisticated ways,
-resulting in state-of-the-art planners that are more powerful and
-successful than their predecessors, we nevertheless believe that
-improvements are necessary if planning is to become a more mature
-technology that can offer tools to a wider community of users and
-researchers. At the end of the day, real-world users will care about the
-\emph{total} runtime of a planner, which is more than just the search time.
-
-By and large, our experiences with the planning community from a user's
-point of view have been favourable. Thanks to the planning competitions, it
-is easy to identify and download a fast state-of-the-art planner, at least
-for Linux. However, in the course of our experiments we found and reported
-bugs in both FF and SGPLAN. We also discovered that deciding as to which
-planner is appropriate for a particular task is not always straightforward,
-as the best performing planner in the IPC is not necessarily suitable for
-every planning task. As our experiments have shown, even planners as
-closely related as FF and SGPLAN can differ significantly in their
-performance on different domains (for instance, our results in
-Experiment~4).
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-In this paper, we introduced two novel planning domains arising from
-problems in natural language generation: the sentence generation domain and
-the Generating Instructions in Virtual Environments (GIVE) challenge
-domain. We also reported the results of a set of experiments designed to
-evaluate the suitability of off-the-shelf planners (particularly FF and
-SGPLAN) in these domains.
-
-Our results were mixed. While the planners we tested do an impressive job
-of controlling the complexity of search in many domains, they also suffer
-from practical problems that limit their performance in unexpected ways: in
-both of our NLG domains the grounding step performed by FF and SGPLAN
-dominates the time it takes to perform the search itself. As future work,
-we hope to extend our analysis to other planners that do not perform the
-initial grounding step, planners such as SHOP2
-\citep{DBLP:journals/jair/NauAIKMWY03} that have traditionally performed
-well on a wide range of domains, and new planners introduced in the latest
-edition of the IPC.\footnote{See
- \url{http://ipc.informatik.uni-freiburg.de/} for details of the 2008
- International Planning Competition (deterministic track).}
-
-The NLG community's recent interest in planning presents a valuable
-opportunity for planning researchers. Domains like GIVE highlight certain
-challenges, such as plan execution monitoring and plan presentation (i.e.,
-summarisation and elaboration), but also offer a platform on which such
-technologies can be evaluated in experiments with human users. Although we
-have focused on classical planning problems in this work, research related
-to reasoning under uncertainty, resource management, and planning with
-knowledge and sensing, can also be investigated in these settings. As such,
-we believe our domains would provide suitable challenges for planners
-entered in future editions of the IPC. Provided some of the challenges we
-have identified can be addressed, NLG-inspired problem domains offer a
-constructive platform for the planning community to showcase their
-techniques to a wider audience---and to improve the quality of their tools
-for real-world planning tasks.
-
-\section*{Acknowledgements}
-
-This work arose in the context of the Planning and Language Interest Group
-at the University of Edinburgh. The authors would like to thank all members
-of this group, especially Hector Geffner and Mark Steedman, for interesting
-discussions. This work was partially supported by the DFG Research
-Fellowship ``CRISP: Efficient integrated realization and microplanning''
-and by the European Commission through the PACO-PLUS project
-(FP6-2004-IST-4-27657).
+\input introduction
+\input nlg-as-planning
+\input domains
+\input experiments
+\input discussion
+\input conclusion
 

### 2009-05-07 trunk/Papers/ci-spark/nlg-as-planning.tex
split paper up into individual files; changes to abstract and intro

\section{NLG as Planning} \label{sec:nlg-as-planning}

The task of generating natural language from semantic representations
is typically split up into two parts: the \emph{domain planning} task,
which selects the information that is to be conveyed and structures it
into sentence-sized chunks, and the \emph{sentence generation} tasks,
which then translates each of these chunks into natural language
sentences. The sentence generation task is often split up into two
parts of its own---the \emph{sentence planning} task, which enriches
the input by e.g.\ determining how to refer to objects and selects
some lexical material, and the \emph{surface realization} task, which
maps the enriched meaning representation into a sentence using a
grammar. The chain of domain planning, sentence planning, and surface
realization is sometimes called the ``NLG pipeline''
\cite{reiter00building}.

Viewing generation as a planning problem has a long tradition in the
NLG literature, especially in discourse planning. \todo{Sketch
  Perrault et al here. Say something very brief about Appelt, Hovy,
  Moore. Say why all this didn't really work.}

Most recently, interest in generation as planning has been revived for
NLG tasks on various levels of the NLG pipeline. Below, we will look
in some detail at \cite{KolSto07}'s recasting of sentence generation
as planning; furthermore, \cite{Steedman-Petrick:07} and
\cite{benotti08b} have applied planning to \todo{something}. Notice
that although all of these approaches encode ``communicative actions''
as planning operators, they still solve fundamentally different
problems. For instance, Perrault et al.\ tackled discourse planning
with planning operators that encode speech acts, each of which
represents the meaning of a whole sentence and models the contribution
of this sentence to the intentional structure of the entire
discourse. By contrast, Koller and Stone solve a sentence generation
problem using planning operators that encode utterances of single
words and model the contribution of each word to the grammatical
structure of the sentence. Whether these different levels of ``NLG as
planning'' could be combined into a single planning problem, and
whether this would make computational sense, is an open question.

Problems in NLG at all levels of description can become
computationally complex very easily. For instance, \citet{KolStr02}
showed that even surface realization is NP-complete under reasonable
assumptions. This makes it necessary to investigate methods for
guiding the search, which makes modern planning techniques a promising
approach to NLG. The point of this paper is to investigate whether
existing planners fulfill this promise of efficiency for the
application to NLG.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "manuscript"
%%% End:

### 2009-05-07 trunk/Papers/ci-spark/domains.tex
started revising the crisp presentation

@@
 
-Although the sentence generation task is often split up into two
-separate steps for sentence planning and surface realization, the
-decisions that an NLG system must make in these two modules often
-interact. For instance, in order to generate \emph{referring
-  expressions (REs)}, i.e.\ unique descriptions of individuals in the
-universe of discourse, it is necessary to know what properties of
-individuals can be expressed as nouns and adjectives. Grammatical
-information like this should belong exclusively to the surface
-realization module; but on the other hand, RE generation enriches the
-input meaning representation and is therefore usually seen as part of
-sentence planning.
-
-One seminal approach to solving the sentence planning and surface
-realization problems together is offered by the SPUD generation system
-(``Sentence Planning Using Description''; \citealt{Stone2003a}). This
-approach assumes a tree-adjoining grammar (TAG) \citep{joshi;etal1997}
-whose lexical entries are equipped with semantic and pragmatic
-information, as well as two sets of ground atoms: the
-\emph{communicative goal} that specifies what semantic information we
-must express, and a \emph{knowledge base}. Let's consider an example
-to illustrate the problem that SPUD solves. Consider a knowledge base
-containing the individuals $e$, $r_1$ and $r_2$, and a set of
-attributes encoding the fact that $r_1$ and $r_2$ are rabbits, $r_1$
-is white and $r_2$ is brown, and $e$ is an event in which $r_1$
-sleeps.  Say that we want to express the information
-$\{\mathsf{sleep}(e,r_1)\}$ using the tree-adjoining grammar shown in
+One way of modelling the sentence generation problem is to assume a
+lexicalized grammar in which each lexicon entry specifies how it can
+be combined grammatically with the other lexicon entries, what piece
+of meaning it expresses, and what the pragmatic conditions on using it
+are. Sentence generation can then be seen as constructing a
+grammatical derivation that is syntactically complete, respects the
+semantic and pragmatic conditions, and achieves all the
+\emph{communicative goals}. An example is the tree-adjoining grammar
+(TAG; \citep{joshi;etal1997}) shown in
 Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
 \emph{elementary trees} (i.e., the disjoint trees in the figure), each
-of which contributes certain \emph{semantic content}. We can
-instantiate these trees by substituting individuals for \emph{semantic
-  roles}, such as $\mathsf{self}$ and $\mathsf{subj}$, and then
-combine the tree instances as shown in
-Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
-``The white rabbit sleeps''.
+of which contributes certain \emph{semantic content}. Let's say that
+we assume a knowledge base containing the individuals $e$, $r_1$ and
+$r_2$ and the facts that $r_1$ and $r_2$ are rabbits, $r_1$ is white
+and $r_2$ is brown, and $e$ is an event in which $r_1$ sleeps. Then we
+would want to construct a sentence expressing the information
+$\{\mathsf{sleep}(e,r_1)\}$ by combining instances of the elementary
+trees (in which the \emph{semantic roles}, such as $\mathsf{self}$ and
+$\mathsf{subj}$, have been substituted by constants from the knowledge
+base) into a TAG derivation as as shown in
+Figure~\ref{fig:white-rabbit-sleeps-deriv}. We can then read off the
+sentence ``The white rabbit sleeps'' from the derivation. \todo{need
+  to mention distractors and unique REs here}
 
@@
 
-We can now consider various different algorithms for solving the SPUD
-problem. The naive algorithm builds the TAG derivation top-down. In
-the example, it starts with the elementary tree for ``sleeps''. This
-tree satisfies the need to convey the semantic information, but
-introduces a need to generate a noun phrase (NP) for the subject,
-which must refer uniquely to the target referent $r_1$. In a second
-step, we substitute the tree for ``the rabbit'' into the open NP leaf,
-which makes the derivation grammatically complete. Since there are two
-different individuals that could be described as ``the
-rabbit''---technically, $r_2$ is still a \emph{distractor} (i.e.,
-based on the description ``the rabbit'', the hearer might erroneously
-think that we're talking about $r_2$ and not $r_1$)---we are still not
-finished. To complete the derivation, the tree for ``white'' is added
-to the existing structure by an \emph{adjunction} operation, making
-the derivation both syntactically and semantically complete. This
-search algorithm takes worst-case exponential time because it may have
-to consider many combinations of elementary trees; and in fact, the
-SPUD problem generalizes the surface realization problem considered by
-\citet{KolStr02} and is therefore NP-complete.
+This perspective on sentence generation has the advantage of solving
+the sentence planning and surface realization problems simultaneously:
+in the example, we can require that referring expressions ``the white
+rabbit'' can be resolved uniquely to $r_1$ by the hearer, in addition
+to the requirement that the derivation is grammatically correct. This
+is useful in cases where sentence planning and surface realization
+interact, because syntactic information about the individual words is
+available when the REs are generated (see
+e.g. \cite{stone98textual}). However, the problem of whether a given
+communicative goal can be achieved with a given grammar is NP-complete
+\cite{KolStr02}. A naive search algorithm that computes the derivation
+top-down takes exponential time and is clearly infeasible to use in
+practice. The SPUD system \cite{Stone2003a}, which worked this idea
+out for the first time, uses a greedy algorithm, which circumvents the
+combinatorial explosion but is incomplete.
 
@@
 
-The SPUD system itself worked around this complexity by using a
-greedy, incomplete search algorithm. On the other hand,
-\citet{KolSto07} present an alternative to controlling the necessary
-search, by encoding the generation problem into a planning
-problem.\footnote{See \url{http://code.google.com/p/crisp-nlg/} for
-  the CRISP system, in which this conversion is implemented.} Each
-operator of this planning problem models the addition of some
-elementary tree to the TAG derivation. For instance,
+Here we will look at one recent approach which aims at controlling the
+search by converting the sentence generation problem into a planning
+problem and then running a planner \citet{KolSto07}.\footnote{See
+  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
+  which this conversion is implemented.} \todo{XXX} Each operator of this
+planning problem models the addition of some elementary tree to the
+TAG derivation. For instance,
 Figure~\ref{fig:white-rabbit-as-planning} shows the corresponding PDDL

### 2009-05-07 trunk/Papers/ci-spark/domains.tex

@@
   \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
-  which this conversion is implemented.} \todo{XXX} Each operator of this
+  which this conversion is implemented.} Each operator of this
 planning problem models the addition of some elementary tree to the
@@
 Figure~\ref{fig:white-rabbit-as-planning} shows the corresponding PDDL
-actions for the above generation task. The \texttt{add-sleeps}
-operator represents the addition of the elementary tree for ``sleeps''
-to the derivation. $\verb?add-sleep?(\mathsf{root},e,r_1)$ removes the
-atom $\mathsf{subst}(S,\mathsf{root})$, which expresses that there is
+actions for the above generation task. The $\addact{sleeps}$ operator
+represents the addition of the elementary tree for ``sleeps'' to the
+derivation. $\addact{sleeps}(\mathsf{root},e,r_1)$ removes the atom
+$\mathsf{subst}(S,\mathsf{root})$, which expresses that there is
 currently an open substitution node $\mathsf{root}$ with label $S$,
@@
 initial state that encodes the knowledge base and contains an atom
-$\mathsf{subst}(S,\emph{root})$, and a goal that contains, among
+$\mathsf{subst}(S,\mathsf{root})$, and a goal that contains, among
 others, $\forall x \forall y. \neg \mathsf{subst}(x,y)$ and
-$\mathsf{expressed}(\emph{sleep},e,r_1)$, a planner will find plans
+$\mathsf{expressed}(\mathsf{sleep},e,r_1)$, a planner will find plans
 like the following:
@@
 
+However, it is desirable to integrate the planner and the generation
+system more closely than this. For instance, let's say that the NLG
+system wants to generate the instruction sequence ``walk to the centre
+of the room; turn right; now press the green button in front of you''.
+Experiments \todo{cite} show that this is a pattern that human
+instruction givers use frequently: The instruction follower is made to
+walk to a certain point in the world at which the instruction giver
+can then use a a referring expression (``the green button'') that is
+very easy to interpret. \todo{XXX}
+
 However, in order to solve the GIVE task in a more satisfying way, we

### 2009-05-07 trunk/Papers/ci-spark/manuscript.tex
changed to official CI style

@@
 %%%
-\documentclass[letterpaper]{article}
-\usepackage{times}
+\documentclass[useAMS,usenatbib]{coin}
+
+%\usepackage{times}
 \usepackage{graphicx}
 \usepackage{url}
-\usepackage[round]{natbib}
-\usepackage{setspace}
-\usepackage[paper=letterpaper,margin=1.0in]{geometry}
+\usepackage{natbib}
+%\usepackage{setspace}
+%\usepackage[paper=letterpaper,margin=1.0in]{geometry}
 
 \title{Experiences with Planning for Natural Language Generation}
-\author{\textbf{Alexander Koller} \\
-Exzellenzcluster / FR 4.7 Computerlinguistik \\
-Saarland University \\
-Postfach 151150 \\
-66041 Saarbr\"ucken \\
-Germany \\
-Phone: +49 681 302 70040 \\
-Fax: +49 681 302 4351 \\
-E-mail: \url{koller@mmci.uni-saarland.de}
+
+\author{{\sc Alexander Koller}\\
+        {\it Cluster of Excellence, Saarland University, Saarbr\"ucken, Germany}\\ 
 		\and
-\textbf{Ronald P. A. Petrick} \\
-School of Informatics \\
-University of Edinburgh \\
-Informatics Forum, 10 Crichton Street \\
-Edinburgh \ EH8 9AB \\
-Scotland, United Kingdom \\
-Phone: +44 131 650 4426 \\
-Fax: +44 131 650 6899 \\
-E-mail: \url{rpetrick@inf.ed.ac.uk}}
+        {\sc Ronald P. A. Petrick}\\
+        {\it School of Informatics, University of Edinburgh,
+          Edinburgh, UK\thanks{Address correspondence to XXXX.}}
+}
+
+% \author{\textbf{Alexander Koller} \\
+% Exzellenzcluster / FR 4.7 Computerlinguistik \\
+% Saarland University \\
+% Postfach 151150 \\
+% 66041 Saarbr\"ucken \\
+% Germany \\
+% Phone: +49 681 302 70040 \\
+% Fax: +49 681 302 4351 \\
+% E-mail: \url{koller@mmci.uni-saarland.de}
+% \and
+% \textbf{Ronald P. A. Petrick} \\
+% School of Informatics \\
+% University of Edinburgh \\
+% Informatics Forum, 10 Crichton Street \\
+% Edinburgh \ EH8 9AB \\
+% Scotland, United Kingdom \\
+% Phone: +44 131 650 4426 \\
+% Fax: +44 131 650 6899 \\
+% E-mail: \url{rpetrick@inf.ed.ac.uk}}
 
@@
 
-\pagestyle{plain}
-\pagenumbering{arabic}
-\doublespacing
+%\pagestyle{plain}
+%\pagenumbering{arabic}
+%\doublespacing
 
@@
 \newcommand{\todo}[1]{\textbf{(#1)}}
+\newcommand{\addact}[1]{\mathsf{add}\textsf{-}\mathsf{#1}}
 
@@
 
+%\pagebreak
+\input abstract
+
 \maketitle
 
-\pagebreak
-\input abstract
-
-\pagebreak
+%\pagebreak
 

### 2009-05-07 trunk/Papers/ci-spark/domains.tex
finished going over 1-3

@@
 can then use a a referring expression (``the green button'') that is
-very easy to interpret. \todo{XXX}
-
-However, in order to solve the GIVE task in a more satisfying way, we
-expect that planning and generation will have to be interleaved more
-closely, making the planning a genuine part of the NLG task. For
-instance, the NLG system could generate an instruction sequence
-starting with ``turn left; walk forward; press the button; turn right;
-walk forward; walk forward; turn right; walk forward; walk forward;
-turn left; walk forward''. But such instructions are clumsy and
-uninteresting; it would be much better to say ``turn around and walk
-through the door''. That is, the system should \emph{summarise} plans
-by merging multiple action instances into single instructions before
-presenting them to the user.\footnote{This task is similar to the
-  construction of ``macro'' actions, but is primarily intended for
-  plan presentation purposes. See, e.g., \citep{Botea-etal:05} which
-  contains a survey of recent approaches for generating macro actions
-  in planning.}  Conversely, it may also be necessary to
-\emph{elaborate} on a single planning step by expressing it with
-several instructions. Having just entered the top left room, it may be
-easier for the user to understand the instruction ``walk to the centre
-of the room; turn right; now press the green button in front of you'',
-rather than the instruction ``press the green button on the wall to
-your right''. Furthermore, in order to plan the referring expression
-``the green button in front of you'' at a time when the button is not
-yet in front of the user, the NLG system must keep track of the
-hypothetical changes in what objects are visible to the user. Thus,
-the NLG acts of referring and instructing must be tightly integrated
-with the modules for plan generation. This makes the GIVE domain
-planning problem an integral part of the NLG system, which makes this
-a second planning problem of immediate relevance to NLG that is
-completely different from the first one.
-
-% It seems to me that this paragraph misses the point. We're not
-% trying to sell GIVE as a cool problem here; we take the perspective
-% that GIVE exists and has already been shown to be cool, and we only
-% talk about solving it here. - AK
-%
-% From a planning perspective, GIVE imposes very strict runtime
-% requirements on the plan generation task: planning must
-% happen in real time and the system must respond to a user in a timely
-% fashion.  If the system takes too much time deliberating over an
-% instruction to give, rather than actually giving this instruction, the
-% user may have walked or turned away, thus making the instruction
-% invalid.  Furthermore, plan execution monitoring also plays an
-% important role in the GIVE problem.  At a high level, the system needs
-% to monitor a user's actions and compare them against the generated
-% instruction set to determine if the user has correctly followed
-% directions or not. In the case of the latter, new instructions may
-% have to be generated. In practice, the situation can be quite
-% complicated since the mental state of the user is not known and so the
-% system must be inferred from observing the user's actions in real
-% time. For example, a user directed to ``turn around and walk through
-% the door'' may not necessarily perform these actions to the letter,
-% i.e., immediately turning 180 degrees and proceeding directly to the
-% door. Instead, the user might take a roundabout route through the
-% room, eventually exiting out the door.  Although the user's actions do
-% not match the generated instructions exactly, they meet the intended
-% goal. The system must be able to identify such ``equivalent'' plans
-% and not immediately generate new instructions as soon as the user's
-% actions have gone off course. Furthermore, a user can communicate
-% certain intentions to the system, both through action and
-% inaction. For instance, the system should infer that a user has failed
-% to follow instructions if the user exits a room when given a directive
-% to ``walk to the centre of the room''. The system should also make a
-% similar conclusion if a user simply does nothing when given the
-% instruction.
+very easy to interpret. Such an NLG system must integrate discourse
+planning and planning in the domain of the world map closely: On the
+one hand, the structure of the discourse is determined by the needs of
+the NLG system rather than the domain plan; on the other hand, the
+discourse planner must be aware of the way in which the instruction
+``turn right'' is likely to change the visibility of objects. Even if
+the NLG system doesn't implement the generation of this discourse as
+planning, it still solves a problem that subsumes the domain planning
+problem. For these reasons, we consider the GIVE domain planning
+problem as a natural part of a GIVE NLG system.
+\todo{Make sure that this paragraph still talks about the things that
+  the reviewers found interesting about it.}
 

### 2009-05-19 trunk/Papers/crisp-rtg-09/crisp-rtg-09.tex
started with definitions

\documentclass[11pt,a4]{article}

\usepackage{url}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsfonts,amsthm}

\def\N{\mathbb{N}}
\newcommand{\sem}{\mathsf{sem}}
\newcommand{\self}{\mathsf{self}}
\newcommand{\produ}{\mathsf{prod}}
\newcommand{\roles}{\mathsf{roles}}
\newcommand{\Neq}{{:}}
\newcommand{\refr}{\mathsf{ref}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\idsem}{\mathsf{idsem}}
\newcommand{\Vars}{\mathsf{Vars}}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{prop}[theorem]{Proposition}
\newtheorem{kor}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{verm}{Conjecture}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{bem}{Remark}

\title{Sentence generation with RTGs}
\author{Alexander Koller \\ Saarland University \\
  \url{koller@mmci.uni-saarland.de}}
\date{\today\ (v3.0)}

\begin{document}

\maketitle

\section{Introduction}

\section{The generation problem of regular tree grammars}

I write $\N_n$ for the set $\{1,\ldots,n\}$.  I write $T_A(B)$ for the
set of terms $f(a_1,\ldots,a_n)$ with $f \in A$ and $a_1,\ldots,a_n
\in B$.  I also assume that the set of nodes in a tree is a set of
strings in $\N^*$ that is closed under left sister (i.e., if $ui$ is a
node of some tree and $i>1$, then $u(i-1)$ is also a node) and prefix
(i.e., if $ui$ is a node of the tree, then $u$ is too).

\begin{definition}
  A \emph{regular generation grammar (RGG)} is a tuple $G =
  (N,\Sigma,S,Pred,R,L)$ consisting of a terminal alphabet $\Sigma$, a
  nonterminal alphabet $N$, a start symbol $S \in N$, a set $Pred$ of
  \emph{predicate symbols}, a set $R$ of \emph{roles}, and a finite
  set $L$ of \emph{lexicon entries} $l = (P,\sem,r)$, where $P = A
  \rightarrow f(B_1,\ldots,B_n)$ is an RTG production rule over $N$
  and $\Sigma$, $r:\N_n \rightarrow R$ an assignment of roles to
  right-hand nonterminal occurrences that assigns the role $\self$ to
  exactly one number, and $\sem \subseteq
  T_{Pred}(\{r(1),\ldots,r(n)\})$ is the semantic representation.
\end{definition}

We write $A \rightarrow f(B_1\Neq r_1,\ldots,B_n \Neq r_n)$ to
abbreviate a production rule with a role assignment that maps $i$ to
$r_i$ for each $1 \leq i \leq n$. We write $\produ(l)$, $\roles(l)$,
and $\sem(l)$ for the production rule, the set of roles
$\{r(1),\ldots,r(n)\}$, and the $\sem$ value of the lexicon entry
respectively.

\begin{definition}
  Assume a universe $U$ and an infinite set $\Vars$ of variables. A
  \emph{derivation} of a regular generation grammar $G$ is a sequence
  $d = l_1,\ldots,l_n$ of lexicon entries from $G$ such that there are
  functions $\id:\N_n \times R \rightsquigarrow \Vars$ and
  $\refr:\Vars \rightarrow U$ such that
  \begin{enumerate}
  \item grammaticality: the sequence $\produ(l_1),\ldots,\produ(l_n)$
    is a complete RTG derivation that maps $S$ into a tree of terminal
    symbols;
  \item definedness: $\id(i,r)$ is defined iff $r \in
    \roles(l_i)$;
  \item consistent reference: if $\produ(l_k)$ expands the right-hand
    nonterminal $A:r$ in $\produ(l_i)$, then $\id(k,\self) =
    \id(j,r)$.
  \end{enumerate}

  We write $\idsem(l_i) = \{P(x_1,\ldots,x_m) \;|\;
  \mbox{$P(r_1,\ldots,r_m) \in \sem(l_i)$ and $x_k = \id(i,r_k)$ for
    all $k$}\}$, and take $||S||_M$ to be the set of all variable
  assignments $\Vars \rightarrow U$ that satisfy $S$ in all models
  over $U$ that satisfy all formulas in the set $M$.

  Then we can define a \emph{successful derivation} of $G$ for the
  communicative goal $C \subseteq T_{Pred}(U)$, the target referent $e
  \in U$, and the speaker and hearer knowledge bases $SKB, HKB
  \subseteq T_{Pred}(U)$ to be a derivation of $G$ that also satisfies
  the following conditions:

  \begin{enumerate}
  \item truthfulness: for all $i$, $\refr(\idsem(l_i)) \subseteq SKB$;
  \item communicative goal achieved: $C \subseteq \cup_{k=1}^n \refr(\idsem(l_k))$;
  \item unique reference: for all $1 \leq i \leq n$ and $r \in
    \roles(l_i)$, $|| \cup_{k=1}^n \idsem(l_k) ||_{HKB}(\id(i,r)) =
    \{\refr(\id(i,r))\}$.
  \end{enumerate}

  The \emph{sentence generation problem} of RGGs is to decide for a
  grammar $G$ and the communicative goal, target referent, and
  knowledge bases, whether there is a successful derivation.
\end{definition}

\section{RTG generation as planning}

\section{NP-completeness of the RTG generation problem}

\section{Conclusion}

\section*{Version History}

\begin{tabular}{lll}
  v 3.0 & \today & Completely revised \\
  v 2.0 & 29/10/07 & Version for Hector Geffner
\end{tabular}

\bibliographystyle{plainnat}
\bibliography{gen}

\end{document}

### 2009-05-20 trunk/Papers/crisp-rtg-09/crisp-rtg-09.tex
finished def; put in example

@@
 
+We start by defining the generation problem of regular tree
+grammars. We will first show how RTGs can be equipped with semantic
+information; we call the result a \emph{regular generation grammar
+  (RGG)} (see Def.~\ref{def:rgg}). Then we will define the
+grammatically correct derivations of an RGG, and a \emph{successful}
+derivation as one that also achieves a given communicative goal in
+Def.~\ref{def:rgg-deriv}. We will then discuss an example in
+Section~\ref{sec:rgg-example}.
+
+\subsection{Definitions}
+
 I write $\N_n$ for the set $\{1,\ldots,n\}$.  I write $T_A(B)$ for the
@@
 
-\begin{definition}
+\begin{definition} \label{def:rgg}
   A \emph{regular generation grammar (RGG)} is a tuple $G =
@@
 
-\begin{definition}
+We will now define the derivations of RGGs. Derivations are sequences
+of lexicon entries. It is intuitively clear that the sequence of RTG
+production rules in each of these lexicon entries should constitute an
+ordinary RTG derivation. Furthermore, we require that there is a
+function $\id$ that maps (the position of) a lexicon entry in the
+derivation and a role to a variable, and a variable assignment $\refr$
+that maps variables to individuals in the universe. For a derivation
+to be correct, we insist that the variables assigned by $\id$ are
+chosen consistently, i.e.\ if we use some rule $\produ(l_k)$ to expand
+a right-hand nonterminal $A \Neq r$ introduced by the rule
+$\produ(l_i)$, then the $\self$ variable of $k$ must be the same as
+the $r$ variable of $i$.
+
+\begin{definition} \label{def:rgg-deriv}
   Assume a universe $U$ and an infinite set $\Vars$ of variables. A
@@
     nonterminal $A:r$ in $\produ(l_i)$, then $\id(k,\self) =
-    \id(j,r)$.
+    \id(j,r)$;
+  \item naming apart: if $r_1,r_2 \in \roles(l_i)$ and $r_1\neq r_2$,
+    then $\id(i,r_1) \neq \id(i,r_2)$.
   \end{enumerate}
+\end{definition}
 
 We write $\idsem(l_i) = \{P(x_1,\ldots,x_m) \;|\;
-  \mbox{$P(r_1,\ldots,r_m) \in \sem(l_i)$ and $x_k = \id(i,r_k)$ for
-    all $k$}\}$, and take $||S||_M$ to be the set of all variable
-  assignments $\Vars \rightarrow U$ that satisfy $S$ in all models
-  over $U$ that satisfy all formulas in the set $M$.
+\mbox{$P(r_1,\ldots,r_m) \in \sem(l_i)$ and $x_k = \id(i,r_k)$ for all
+  $k$}\}$, and take $||S||_M$ to be the set of all variable
+assignments $\Vars \rightarrow U$ that satisfy $S$ in all models over
+$U$ that satisfy all formulas in the set $M$.
 
-  Then we can define a \emph{successful derivation} of $G$ for the
-  communicative goal $C \subseteq T_{Pred}(U)$, the target referent $e
-  \in U$, and the speaker and hearer knowledge bases $SKB, HKB
-  \subseteq T_{Pred}(U)$ to be a derivation of $G$ that also satisfies
-  the following conditions:
+Then we can define a \emph{successful derivation} as a derivation that
+also achieves a given communicative goal. We assume two knowledge
+bases $SKB$ and $HKB$, which are sets of ground atoms representing the
+speaker's and hearer's knowledge, respectively. Then we insist that
+the generated statement is truthful (i.e., the semantic
+representations of all used lexicon entries are supported by the SKB),
+the communicative goal was achieved, and all referring expressions can
+be resolved uniquely by the hearer.
+
+\begin{definition} \label{def:rgg-succ-deriv} A \emph{successful
+    derivation} of $G$ for the communicative goal $C \subseteq
+  T_{Pred}(U)$, the target referent $e \in U$, and the speaker and
+  hearer knowledge bases $SKB, HKB \subseteq T_{Pred}(U)$ is a
+  derivation $l_1,\ldots,l_n$ of $G$ such that:
 
@@
 \end{definition}
+
+Compared to a system like SPUD \citep{Stone2003a}, the mapping from
+the semantic roles in the lexicon entries to individuals in the
+universe is relatively indirect: The semantic roles are first mapped
+to variables by the $\id$ function, and then the variables are
+assigned individuals by $\refr$. The point about this two-step process
+is that it makes it possible to specify the ``unique reference''
+condition. By mapping the roles to variables first, we can ask for the
+set of all variable assignments that satisfy the combined semantic
+representations of all lexicon entries (given the hearer's
+knowledge). We can then convince ourselves that the information in the
+semantic representations was precise enough that the hearer can
+identify the referent uniquely by checking that all these variable
+assignments assign the same individual to the variable for the
+referring expression, and that this individual is indeed the target
+referent. If we had mapped roles to individuals directly, rather than
+distinguishing $\refr$ and $\i$, we could have verified that the
+assignment satisfies all the semantic representations, but we couldn't
+have stated that it is the \emph{only} such assignment.
+
+\subsection{An example} \label{sec:rgg-example}
+
+Let's go through an example to illustrate these definitions.
+
+Consider the following grammar, which is defined over the nonterminals
+$\{S, NP, Det, Adj\}$, start symbol $S$, roles $\{\self,
+\mathsf{subj}, \mathsf{obj}\}$, and the following lexicon
+entries:\footnote{I assume a predicate ``$\cdot = \mathsf{peter}$'' of
+  which  $\self = \mathsf{peter}$ is an instance.}
+
+$$\begin{array}{lp{1cm}l}
+S \rightarrow \mathrm{likes}(NP \Neq \mathsf{subj}, NP \Neq
+\mathsf{obj}) 
+&&
+Det \rightarrow \mathrm{the} \\
+\mathsf{sem} = \{ \mathsf{likes}(\self, \mathsf{subj}, \mathsf{obj})
+\} 
+&&
+\mathsf{sem} = \emptyset \\
+&&\\
+%
+%
+NP \rightarrow \mathrm{Peter} 
+&&
+Adj \rightarrow \mathrm{white} \\
+\mathsf{sem} = \{ \self = \mathsf{peter} \}
+&&
+\mathsf{sem} = \{ \mathsf{white}(\self) \} \\
+&&\\
+%
+%
+NP \rightarrow \mathrm{rabbit}(Det \Neq \self, Adj \Neq \mathsf{self})
+&&
+Adj \rightarrow \epsilon \\
+\mathsf{sem} = \{ \mathsf{rabbit}(\self) \}
+&&
+\mathsf{sem} = \emptyset
+\end{array}
+$$
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.6]{pic-rdg-crisp}
+  \caption{Verbalization of the example. Each node displays its
+    terminal, nonterminal, and $\id$ function, in this order. $\refr$
+    is $\{x \mapsto e, y \mapsto p, z \mapsto r_1\}$.}
+  \label{fig:rdg-crisp}
+\end{figure}
+
+There are several different derivations for this grammar. Identifying
+words with lexicon entries, we can write two of these as:
+$$
+\begin{array}{lll}
+  d_1 &= &(\mathrm{likes}, \mathrm{Peter}, \mathrm{rabbit},
+  \mathrm{the}, \mathrm{white}) \\
+  d_2 &= &(\mathrm{likes}, \mathrm{Peter}, \mathrm{rabbit},
+  \mathrm{the}, \epsilon) 
+\end{array}
+$$
+
+In each of these cases, we can assume an $\id$ function that looks as
+follows:
+$$\begin{array}{ll|l}
+i&r&\id(i,r) \\\hline
+1&\self&x_1 \\
+1&\mathsf{subj}&x_2 \\
+1&\mathsf{obj}&x_3 \\
+2&\self&x_2 \\
+3&\self&x_3 \\
+4&\self&x_3 \\
+5&\self&x_3
+\end{array}
+$$
+
+The tree represented by the derivation $d_1$ is shown in
+Fig.~\ref{fig:rdg-crisp}.  The values of $\refr$ are irrelevant at
+this point.
+
+Now let's determine which of these derivations is
+\emph{successful}. We assume the following knowledge bases, the
+communicative goal $C=\{\mathsf{likes}(e,p,r_1)\}$, and the target
+referent $e$.
+
+$$\begin{array}{lll}
+  SKB &=
+  &\{\mathsf{likes}(e,p,r_1), p = \mathsf{peter}, \mathsf{rabbit}(r_1),
+  \mathsf{rabbit}(r_2), \mathsf{white}(r_1)\} \\
+  HKB &= &\{p = \mathsf{peter}, \mathsf{rabbit}(r_1),
+  \mathsf{rabbit}(r_2), \mathsf{white}(r_1)\} 
+\end{array}
+$$
+
+Then both $d_1$ and $d_2$ are truthful and convey the communicative
+goal for $\refr$ being $\{x_1 \mapsto e, x_2 \mapsto p, x_3 \mapsto
+r_1\}$. However, the reference in $d_2$ is not unique: $||\cup
+\idsem(l_k)||(x_3)$ is $\{r_1,r_2\}$, i.e.\ the hearer might mistake
+the object of the liking as the rabbit $r_2$. Therefore $d_2$ is not
+successful. On the other hand, the unique reference condition is
+satisfied in $d_1$, and so it is a successful derivation. Thus we can
+read off the sentence ``Peter likes the white rabbit'' as a correct
+verbalization of $C$ from Fig.~\ref{fig:rdg-crisp}.
 

### 2009-05-25 trunk/Papers/ci-spark/discussion.tex
started working on experiments

@@
 \label{sec:discussion}
+
+this was in Sect 4 before:
+
+On first impressions, our results indicate that planning is a promising
+tool for both domains.  In the sentence generation domain, FF dramatically
+outperforms the best previously known algorithm for the same problem (a
+reimplementation of \citep{Stone2003a}), although the latter approach uses
+a greedy search algorithm with a heuristic that is hand-tailored to the
+domain. In the GIVE domain, SGPLAN computes a domain plan from the initial
+state to the goal in 0.3 seconds, which is fast enough for moderately-sized
+problems instances in the application.\footnote{All
+  runtimes were measured on a Pentium 4 CPU running at 3 GHz. Java programs
+  were allowed to ``warm up'', i.e., the planner was run five times and the
+  first four measurements discarded to ensure that the JVM had just-in-time
+  compiled all relevant bytecode.}
+
+We also observe that although FF and SGPLAN manage the search for plans
+very efficiently, they both spend comparatively large amounts of time
+computing predicate and action instantiations in their preprocessing
+stages, many of which are then never used during the search. FF's
+``grounding time'' in particular dominates its overall planning time,
+leading to some conflicting results. In some cases, our Java implementation
+of GraphPlan significantly outperforms both state-of-the-art planners. The
+experiments described below seek to improve our understanding of this
+situation.
+
+---
 

### 2009-05-25 trunk/Papers/ci-spark/experiments.tex
started working on experiments

@@
 
-We now describe the results of four simple experiments designed to evaluate
-the performance of three planners---FF \citep{HoffmannNebel01}, SGPLAN
-5.2.2 \citep{hsu06:_new_featur_in_sgplan_for}, and an ad-hoc implementation
-of GraphPlan \citep{Blum1997} written in Java---in our NLG domains. The FF
-and SGPLAN planners were chosen due to their success in the International
-Planning Competitions. Our implementation of GraphPlan is included as a
-counterpoint to the other two, more complex, planners. Most notably, our
-version of GraphPlan does not perform a preprocessing ``grounding'' step
-like FF and SGPLAN but is much more selective in its predicate and action
-instantiations.
+We will now return to the original question of the paper: Is planning
+technology ready for applications in natural language generation? We
+will tackle this question by evaluating the performance of several
+planners on the NLG planning domains from the previous section.
 
-On first impressions, our results indicate that planning is a promising
-tool for both domains.  In the sentence generation domain, FF dramatically
-outperforms the best previously known algorithm for the same problem (a
-reimplementation of \citep{Stone2003a}), although the latter approach uses
-a greedy search algorithm with a heuristic that is hand-tailored to the
-domain. In the GIVE domain, SGPLAN computes a domain plan from the initial
-state to the goal in 0.3 seconds, which is fast enough for moderately-sized
-problems instances in the application.\footnote{All
-  runtimes were measured on a Pentium 4 CPU running at 3 GHz. Java programs
-  were allowed to ``warm up'', i.e., the planner was run five times and the
-  first four measurements discarded to ensure that the JVM had just-in-time
-  compiled all relevant bytecode.}
+We start with two experiments in the CRISP domain. First, we present a
+scenario which focuses on the generation of referring expressions with
+a tiny grammar (Section~\ref{sec:exper-1:-sent}); then we look at a
+setting in which CRISP is used for surface realization with the XTAG
+Grammar \citep{xtag01:_tr}, a large-scale TAG grammar of English
+(Section~\ref{sec:exper-2:-sent-xtag}).
 
-We also observe that although FF and SGPLAN manage the search for plans
-very efficiently, they both spend comparatively large amounts of time
-computing predicate and action instantiations in their preprocessing
-stages, many of which are then never used during the search. FF's
-``grounding time'' in particular dominates its overall planning time,
-leading to some conflicting results. In some cases, our Java implementation
-of GraphPlan significantly outperforms both state-of-the-art planners. The
-experiments described below seek to improve our understanding of this
-situation.
+We then go through a series of experiments in the GIVE domain. We
+start with a domain that is essentially Gridworld
+(Section~\ref{sec:exper-2:-minim}), and then add positions that are
+either not needed (Section~\ref{sec:experiment-3:-give}) or even
+inaccessible (Section~\ref{sec:experiment-4:-give}).
 
-\subsection{Experiment 1: Sentence generation}
+\subsection{Experiment 1: Sentence generation (referring expressions)}
 \label{sec:exper-1:-sent}
@@
 \end{figure}
+
+ The
+planners we will use include FF \citep{HoffmannNebel01} and SGPLAN
+5.2.2 \citep{hsu06:_new_featur_in_sgplan_for}: These were highly
+successful in recent International Planning Competitions and support a
+fragment of
+
+and an ad-hoc implementation
+of GraphPlan \citep{Blum1997} written in Java---in our NLG domains. The FF
+and SGPLAN planners were chosen due to their success in the International
+Planning Competitions. Our implementation of GraphPlan is included as a
+counterpoint to the other two, more complex, planners. Most notably, our
+version of GraphPlan does not perform a preprocessing ``grounding'' step
+like FF and SGPLAN but is much more selective in its predicate and action
+instantiations.
 
@@
 problematic for all the planners we tested.
+
+\subsection{Experiment 2: Sentence generation (XTAG)}
+\label{sec:exper-2:-sent-xtag}
 

### 2009-05-26 trunk/Papers/ci-spark/experiments.tex
iterated over rabbit experiment

@@
 either not needed (Section~\ref{sec:experiment-3:-give}) or even
-inaccessible (Section~\ref{sec:experiment-4:-give}).
+inaccessible (Section~\ref{sec:experiment-4:-give}). This series of
+experiments is set up to illustrate what we perceive as the main
+limitation of current off-the-shelf planners for our applications:
+They spend a long time on computing all ground instances, even if most
+of these instances are never used in the plan search.
 
@@
 
-In the first experiment, we construct a series of sentence generation
-problems which require the planner to compute a plan representing the
-sentence ``Mary likes the Adj$_1$ \ldots Adj$_n$ rabbit.''  Each problem
-instance assumes a certain number $m$ of rabbits that were distinguished by
-$n \leq m$ different properties, such that all $n$ properties are required
-to distinguish the target referent from all other rabbits.  The $n$
-properties are realized as $n$ different adjectives, in any order.  This
-setup allows us to control the plan length (a plan with $n$ properties will
-have length $n+4$) and the universe size (the universe will contain $m+1$
-individuals in addition to the differently-typed individuals used to encode
-the grammar).
+For the first experiment on sentence generation, we focus on the
+generation of referring expressions, a problem that is usually handled
+by the sentence planner if sentence planning and surface realization
+are separated. We consider a series of sentence generation problems
+which require the planner to compute a plan representing the sentence
+``Mary likes the Adj$_1$ \ldots Adj$_n$ rabbit.''  Each problem
+instance assumes a target referent $r$, which is a rabbit, and a
+certain number $m$ of further rabbits $r_1,\ldots,r_m$ that are
+distinguished by properties $P_1,\ldots,P_n$ with $n \leq m$. The
+problem instance is set up such that $r$ has all properties except for
+$P_i$ in common with each $r_i$. That is, all $n$ properties are
+required to describe $r$ uniquely. The $n$ properties are realized as
+$n$ different adjectives, in any order.  This setup allows us to vary
+the plan length (a plan with $n$ properties will have length $n+4$)
+and the universe size (the universe will contain $m+1$ rabbit
+individuals in addition to the differently-typed individuals used to
+encode the grammar).
 
@@
 
- The
-planners we will use include FF \citep{HoffmannNebel01} and SGPLAN
-5.2.2 \citep{hsu06:_new_featur_in_sgplan_for}: These were highly
-successful in recent International Planning Competitions and support a
-fragment of
-
-and an ad-hoc implementation
-of GraphPlan \citep{Blum1997} written in Java---in our NLG domains. The FF
-and SGPLAN planners were chosen due to their success in the International
-Planning Competitions. Our implementation of GraphPlan is included as a
-counterpoint to the other two, more complex, planners. Most notably, our
-version of GraphPlan does not perform a preprocessing ``grounding'' step
-like FF and SGPLAN but is much more selective in its predicate and action
-instantiations.
+We converted these generation problem instances into planning problem
+instances as described in Section~\ref{sec:domains}, and then ran
+three different planners on them. We used the off-the-shelf planners
+FF 2.3 \citep{HoffmannNebel01} and SGPLAN 5.2.2
+\citep{hsu06:_new_featur_in_sgplan_for}, because they were highly
+successful in the recent International Planning Competitions, and
+support a fragment of PDDL with quantified and conditional effects,
+which is necessary in our domain (see
+Fig.~\ref{fig:white-rabbit-as-planning}). Common to both of these
+planners is that they perform a preprocessing step in which all ground
+instances of all predicates and actions are computed.  Furthermore, we
+used an ad-hoc implementation of GraphPlan \citep{Blum1997} written in
+Java which avoids the grounding preprocessing step and only computes
+instances of actions and predicates by need, as they are actually
+discovered when the planning graph is built. Our expectation was that
+this planner would not be able to compete with FF or SGPLAN in terms
+of search efficiency, but we were interested in comparing the
+contribution of the prior grounding step on the planner performance.
 
 The results of this experiment are shown in the graph in
-Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
-in lexicographic order on the horizontal axis and the runtime is shown in
-milliseconds on the vertical axis. These results reveal a number of
-interesting insights. First, FF significantly outperforms SGPLAN in this
-domain, sometimes by a factor of 10 or more.\footnote{Experiments with
- SGPLAN use a pre-release version of SGPLAN, kindly provided to us by
- Chih-Wei Hsu.The release version of SGPLAN 5.2.2 had a bug which caused it
- to crash on some of our problem instances.}
-Second, FF's runtime is dominated by its initial grounding step, in which
-it computes the ground instances of all predicates and actions used in the
-planning problem in order to avoid unnecessary instantiation during search.
-In particular, the ratio of grounding time to total runtime is generally
-above 85\%, and rises to above 99\% at $m=11$, which is still a relatively
-small universe for this application.\footnote{The
-  ``grounding'' time reported here is what FF reports as ``time spent:
-  instantiating action templates''.}
+Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are
+plotted in lexicographic order on the horizontal axis, and the runtime
+is shown in milliseconds on the vertical axis.\footnote{These times
+  were measured on XXX} These results reveal a number of interesting
+insights. First, FF significantly outperforms SGPLAN in this domain,
+sometimes by a factor of 10 or more.\footnote{Experiments with SGPLAN
+  use a pre-release version of SGPLAN, kindly provided to us by
+  Chih-Wei Hsu.The release version of SGPLAN 5.2.2 had a bug which
+  caused it to crash on some of our problem instances.}  Second, FF's
+runtime is dominated by its initial grounding step, in which it
+computes the ground instances of all predicates and actions used in
+the planning problem in order to avoid unnecessary instantiation
+during search.  In particular, the ratio of grounding time to total
+runtime is generally above 85\%, and rises to above 99\% at $m=11$,
+which is still a relatively small universe for this
+application.\footnote{The ``grounding'' time reported here is what FF
+  reports as ``time spent: instantiating action templates''.}
 
-In our testing examples, the time spent by FF on grounding is such that it
-is consistently outperformed by our Java implementation of GraphPlan, which
-only computes instances of predicates and actions as they are discovered
-during the construction of the planning graph. Although FF is consistently
-much faster as far as pure search time is concerned, our results indicate
-that FF's performance is much more sensitive to the domain size: if we fix
-$n=1$, FF takes 60 ms to compute a plan at $m=1$, but 2.4 seconds to
-compute the same plan at $m=10$. By comparison, our GraphPlan
-implementation takes 30 ms at $m=1$ and still only requires 50 ms at
-$m=10$. Conversely, GraphPlan's runtime grows much faster with the plan
-size (i.e., with growing values of $n$ for a fixed $m$). Larger, but still
-realistically-sized, instances of the sentence generation problem are still
-problematic for all the planners we tested.
+In our testing examples, the time spent by FF on grounding is so high
+that it is consistently outperformed by our Java implementation of
+GraphPlan, which only computes instances by need. Although FF is
+consistently much faster as far as pure search time is concerned, our
+results indicate that FF's performance is much more sensitive to the
+domain size: if we fix $n=1$, FF takes 60 ms to compute a plan at
+$m=1$, but 2.4 seconds to compute the same plan at $m=10$. By
+comparison, our GraphPlan implementation takes 30 ms at $m=1$ and
+still only requires 50 ms at $m=10$. Conversely, GraphPlan's runtime
+grows much faster with the plan size (i.e., with growing values of $n$
+for a fixed $m$). Larger, but still realistically-sized, instances of
+the sentence generation problem are still problematic for all the
+planners we tested.
 

### 2009-05-26 trunk/Papers/ci-spark/experiments.tex
wrote something about xtag experiment

@@
 \label{sec:exper-2:-sent-xtag}
+
+The first experiment already gives us some initial insights into the
+appropriateness of planning for the sentence generation domain: On the
+examples we looked at, search times were low, but FF and SGPLAN spend
+a lot of time on the initial grounding step.  However, one weakness of
+this experiment is that it uses a tiny grammar, consisting of just
+those 12 lexicon entries that are needed for the experiment. While the
+grounding problem can only get worse with larger grammars, the
+experiment by itself does not allow us to make clear statements about
+the search efficiency.
+
+To address this problem, we ran a second sentence generation
+experiment. This time, we used the XTAG Grammar \citep{xtag01:_tr}, a
+large-scale TAG grammar of English. XTAG contains lexicon entries for
+about 17,000 uninflected words using about 1100 different elementary
+trees. Although XTAG does not contain semantic information, it is
+possible to automatically equip the lexicon entries with inferred
+semantic representations based on the words in the lexicalized
+elementary tree. The result is a highly ambiguous grammar: The most
+ambiguous word, ``ask'', is the anchor of 314 lexicon entries.
+
+In our experiment, we were especially interested in two
+questions. First, how would the planners handle the search problem
+involved in generating with such a large and ambiguous grammar?
+Second, would it be harder to generate sentences containing verbs with
+multiple arguments, given that verbs with more arguments lead to
+actions with more parameters and therefore more instances? To answer
+these questions, we generated sentences of the form ``S and S and
+\ldots and S'', where each S was a sentence; we called the number of
+sentences in the conjunction $n$. Each S was a sentence of the form
+``the businessman sneezes'', ``the businessman admires the girl'', or
+``the businessman gives the girl the book''---that is, they varied in
+the number $k$ of syntactic arguments the verb expects (1 for the
+intransitive verb ``sneeze'', 2 for the transitive verb ``admire'',
+and 3 for the ditransitive verb ``give''). This means that the output
+sentence for parameters $n$ and $k$ contained $n(2k+2)-1$ words. The
+instances were set up in such a way that the generation of referring
+expressions was trivial, so this experiment was purely a surface
+realization task.
+
+\begin{figure}
+  \centering
+
+  \caption{Results for the XTAG experiment. The horizontal axis
+    represents parameters $(k,n)$ from $(1,1)$ to $(3,10)$ in
+    lexicographic order. The vertical axis is runtime in milliseconds.}
+  \label{fig:xtag-graph}
+\end{figure}
+
+Again we ran the experiment with FF 2.3 and SGPLAN 5.2.2; this time,
+we also ran the experiment with the Metric-FF implementation of FF
+because FF 2.3 had a bug that caused it to switch to a slower search
+heuristic for larger problem instances, and which was fixed in
+Metric-FF. The results are shown in
+Fig.~\ref{fig:xtag-graph}. \todo{blah blah talk about results a bit}
+Unlike in the referring expressions experiment above, our own
+Graphplan implementation was unusably slow here for serious problem
+instances---for $k=1$ and $n=3$, this planner took over two minutes,
+and it exceeded its memory limit of 16 GB for $n>3$. \todo{blah blah
+  talk about what we learn about FF etc}
 

### 2009-05-27 trunk/Papers/ci-spark/experiments.tex

@@
 
-Again we ran the experiment with FF 2.3 and SGPLAN 5.2.2; this time,
-we also ran the experiment with the Metric-FF implementation of FF
-because FF 2.3 had a bug that caused it to switch to a slower search
-heuristic for larger problem instances, and which was fixed in
-Metric-FF. The results are shown in
-Fig.~\ref{fig:xtag-graph}. \todo{blah blah talk about results a bit}
-Unlike in the referring expressions experiment above, our own
-Graphplan implementation was unusably slow here for serious problem
-instances---for $k=1$ and $n=3$, this planner took over two minutes,
-and it exceeded its memory limit of 16 GB for $n>3$. \todo{blah blah
-  talk about what we learn about FF etc}
+Fig.~\ref{fig:xtag-graph} reports the runtimes we measured in this
+experiment for FF. Because the computations in this experiment were
+very memory-intensive, we recompiled FF 2.3 as a 64-bit binary. We
+also ran the experiments with a 64-bit version of Metric-FF; although
+our planning problem is purely sequential and not metric, Metric-FF
+avoids a bug which caused FF 2.3 to switch to a best-first search
+strategy beyond a certain point; the jump in runtime is clearly
+visible in the figure. We did not run this experiment with SGPLAN
+because we didn not have access to a 64-bit version of either SGPLAN
+5.2.2 or SGPLAN 6; some initial experiences with the 32-bit version
+indicate that SGPLAN switches to FF's best-first strategy even for
+small problem instances, and thus would be slower than the versions of
+FF we reported here. We also do not report runtimes for our Java
+implementation of Graphplan, because it was unusuably slow for serious
+problem instances: For $k=1$ and $n=3$, it already took over two
+minutes, and it exceeded its memory limit of 16 GB for
+$n>3$.
+
+\todo{blah blah talk about what we learn about FF etc}
+search strategy matters; EHC search works pretty well; but again,
+preprocessing takes tons of time.
 

### 2009-05-28 trunk/Papers/ci-spark/experiments.tex
mostly done with new experiment

@@
 expressions was trivial, so this experiment was purely a surface
-realization task.
+realization task. To achieve reasonable performance, we only generated
+planning operators for those elementary trees for which all predicate
+symbols in the semantic representation also appeared in the knowledge
+base.
 
@@
 
-\todo{blah blah talk about what we learn about FF etc}
-search strategy matters; EHC search works pretty well; but again,
-preprocessing takes tons of time.
+There are a number of observations we can make in this
+experiment. First of all, the experiment confirms that FF's Enforced
+Hill-Climbing search strategy works very well for the sentence
+generation task: Although we are now generating with a large grammar,
+FF generates a 39-word sentence ($k=1,n=6$) in under a second of
+search time. This efficiency is a direct result of this particular
+search strategy: Whenever FF falls back to the best-first strategy,
+runtimes become much slower (compare the graphs for FF 2.3 and
+Metric-FF). Second, FF's runtime is still dominated by the
+preprocessing stage. For instance, Metric-FF spends about 10 seconds
+on search for $k=1,n=10$, compared to its total runtime of about 65
+seconds. This effect becomes more pronounced as we increase $k$: For
+$k=2$, we reach 65 seconds of total runtime at $n=4$, but here
+Metric-FF only spends about a second on search. For $k=3$, \todo{say
+  something true}. This is to be expected, given that the planning
+operators for the verbs have $k+2$ parameters (see
+Fig.~\ref{fig:white-rabbit-as-planning}), and thus the number of
+action instances grows by a factor of the universe size every time we
+increase $k$ by one. A planner which computes all ground instances of
+the operators thus takes exponential time in $k$ for preprocessing.
 

### 2009-05-28 trunk/Papers/ci-spark/nlg-as-planning.tex
gen as planning

@@
 The task of generating natural language from semantic representations
-is typically split up into two parts: the \emph{domain planning} task,
-which selects the information that is to be conveyed and structures it
-into sentence-sized chunks, and the \emph{sentence generation} tasks,
-which then translates each of these chunks into natural language
-sentences. The sentence generation task is often split up into two
-parts of its own---the \emph{sentence planning} task, which enriches
-the input by e.g.\ determining how to refer to objects and selects
-some lexical material, and the \emph{surface realization} task, which
-maps the enriched meaning representation into a sentence using a
-grammar. The chain of domain planning, sentence planning, and surface
-realization is sometimes called the ``NLG pipeline''
-\cite{reiter00building}.
+(NLG) is typically split up into two parts: the \emph{discourse
+  planning} task, which selects the information that is to be conveyed
+and structures it into sentence-sized chunks, and the \emph{sentence
+  generation} tasks, which then translates each of these chunks into
+natural language sentences. The sentence generation task is often
+split up into two parts of its own---the \emph{sentence planning}
+task, which enriches the input by e.g.\ determining how to refer to
+objects and selects some lexical material, and the \emph{surface
+  realization} task, which maps the enriched meaning representation
+into a sentence using a grammar. The chain of domain planning,
+sentence planning, and surface realization is sometimes called the
+``NLG pipeline'' \cite{reiter00building}.
 
 Viewing generation as a planning problem has a long tradition in the
-NLG literature, especially in discourse planning. \todo{Sketch
-  Perrault et al here. Say something very brief about Appelt, Hovy,
-  Moore. Say why all this didn't really work.}
+NLG literature.  \citet{perrault80} presented an approach to discourse
+planning in which the planning operators represent individual speech
+acts such as ``request'' and ``inform''. This idea was later worked
+out further, e.g.\ by \citet{young94dpocl}. On the other hand,
+researchers such as \citet{appelt:planning} and \citet{hovy88} used
+techniques from hierarchical planning to expand a high-level plan
+consisting of speech acts into more detailed specifications of
+individual sentences. Their systems also covered some aspects of
+sentence planning. However, these systems used very expressive logics,
+which were designed to reason about beliefs and intentions, to
+represent the planning state and the planning operators. Furthermore,
+most of them used ad-hoc planning algorithms with rather naive search
+strategies, which did not scale to realistic inputs. As a consequence,
+the NLG-as-planning approach was mostly marginalized throughout the
+1990s.
 
-Most recently, interest in generation as planning has been revived for
-NLG tasks on various levels of the NLG pipeline. Below, we will look
-in some detail at \cite{KolSto07}'s recasting of sentence generation
-as planning; furthermore, \cite{Steedman-Petrick:07} and
-\cite{benotti08b} have applied planning to \todo{something}. Notice
-that although all of these approaches encode ``communicative actions''
-as planning operators, they still solve fundamentally different
-problems. For instance, Perrault et al.\ tackled discourse planning
-with planning operators that encode speech acts, each of which
-represents the meaning of a whole sentence and models the contribution
-of this sentence to the intentional structure of the entire
-discourse. By contrast, Koller and Stone solve a sentence generation
-problem using planning operators that encode utterances of single
-words and model the contribution of each word to the grammatical
-structure of the sentence. Whether these different levels of ``NLG as
-planning'' could be combined into a single planning problem, and
-whether this would make computational sense, is an open question.
-
-Problems in NLG at all levels of description can become
-computationally complex very easily. For instance, \citet{KolStr02}
-showed that even surface realization is NP-complete under reasonable
-assumptions. This makes it necessary to investigate methods for
-guiding the search, which makes modern planning techniques a promising
-approach to NLG. The point of this paper is to investigate whether
-existing planners fulfill this promise of efficiency for the
-application to NLG.
+However, there has recently been a string of publications by various
+authors with renewed interest in the generation-as-planning approach,
+motivated by the recent development of efficient and expressive
+planners. Below (Section~\ref{sec:domain-crisp}) we will look in some
+detail at \cite{KolSto07}'s approach to sentence generation (i.e., the
+sentence planning and surface realization modules of the pipeline) as
+planning. In addition, \citet{Steedman-Petrick:07} revisit the
+analysis of indirect speech acts with modern planning technology;
+\citet{benotti08b} uses planning to explain the accommodation of
+presuppositions; and
+\citet{brenner08:_contin_multiag_plann_approac_to_situat_dialog} use
+multi-agent planning to model the joint problem-solving behavior of
+agents in a situated dialogue. These approaches focus on different
+issues than the 1980's NLG-as-planning literature. But the main
+difference is that they apply \emph{existing}, well-understood
+planning approaches to linguistic problems because these offer new
+modeling tools, and because they hope that modern planners can solve
+the hard search problems that arise in NLG \citep{KolStr02}
+efficiently. The point of this paper is to investigate whether
+existing planners fulfill this latter promise.
 

### 2009-05-30 trunk/Papers/ci-spark/discussion.tex
rewrote discussion

--- 2009-05-07 trunk/Papers/ci-spark/discussion.tex
@@
 
-The positive conclusion we draw from our simple set of experiments is that
-modern planners are fairly good at controlling the search for many of the
-moderately-sized NLG problems we looked at. This is particularly true for
-FF in the sentence generation domain, which generates nontrivial 14-word
-sentences in about two seconds. Our results also indicate there is still
-room for improvement, however: special-purpose algorithms are able to
-generate much larger referring expressions in milliseconds
-\citep{AreKolStr08}. Although search efficiency is not the main concern in
-these domains, it nevertheless becomes a factor in scenarios like
-Experiment~2, where the world is almost completely unconstrained and
-(unsurprisingly) the search becomes much slower.
+There are both positive and negative conclusions about the state of
+planning for modern NLG applications that we can draw from these
+experiments.
 
-The most restrictive bottleneck in the two NLG domains we investigated is
-the initial grounding step that many modern planners perform. While this
-may be a good strategy for traditional planning benchmarks and IPC
-domains---where plan length often dominates universe size and an initial
-investment into grounding pays off in saved instantiations during
-search---this approach has a very pronounced effect in domains like ours.
-As our experiments show, there are natural planning domains in which
-relatively short plans must be computed in large universes. For instance,
-in the sentence generation domain, it is not unrealistic to look for plans
-of length 20 over a universe consisting of thousands of domain individuals
-and tens of thousands of possible action instantiations, some of which
-require three domain individuals as parameters. Planners like FF and SGPLAN
-are less effective in such domains, to the point that grounding
-often dominates the total running time of these planners. (In the case of
-Experiment~1, our ad-hoc implementation of GraphPlan outperforms both
-planners in the sentence generation task.)
+On the one hand, we found that modern planners are very good at
+dealing with the \emph{search} problems that are involved in the
+NLG-based planning problems we looked at. In the sentence generation
+domain, FF's Enforced Hill-Climbing strategy finds plans corresponding
+to 25-word sentences in about a second. It is hard to compare this
+number to a baseline because there are no shared benchmark problems;
+but FF's search outperforms a greedy search algorithm, and the
+performance is competitive with other sentence generators, which
+similarly employ special-purpose algorithms. Thus research on search
+strategies for planning has paid off: In particular, the Enforced
+Hill-Climbing heuristic outperforms the best-first strategy to which
+FF 2.3 switches for some problem instances. Similarly, SGPLAN's
+performance on the GIVE domain is very convincing and fast enough for
+the application.
 
-Since our domains are not that unusual in their structure and composition,
-we hope that the lessons learned from our experiences can help improve the
-performance of current planning systems. For instance, the common strategy
-of splitting the universe into several types of individuals will not be a
-very effective solution in our NLG domains given the large number of
-individuals that could exist for a given type. On the other hand,
-sophisticated techniques based on reachability analysis may have the
-potential to improve this situation, provided such approaches can avoid the
-drawbacks of unnecessary grounding.\footnote{See
-  \citep{Bryce:07} for a good survey on reachability heuristics for planning
-  graph-based approaches to plan generation.}
-While recent trends in planning research have (rightfully) focused on the
-development of algorithms that control search in sophisticated ways,
-resulting in state-of-the-art planners that are more powerful and
-successful than their predecessors, we nevertheless believe that
-improvements are necessary if planning is to become a more mature
-technology that can offer tools to a wider community of users and
-researchers. At the end of the day, real-world users will care about the
-\emph{total} runtime of a planner, which is more than just the search time.
+However, each of the off-the-shelf planners we tested spent
+substantial amounts of time on preprocessing. This is most apparent in
+the sentence generation domain, where the planners spent almost their
+entire runtime on grounding the predicates and operators for some
+instances. This effect is much weaker in the GIVE domain, which has a
+much smaller number of operators; but as we added dummy predicates,
+the ratio of time that the planners spent on preprocessing rose in a
+similar way. While the GIVE domain can be carefully defined in such a
+way that the number of operators is minimized, this is not possible
+for a domain in which the operators encode the different communicative
+actions that the NLG system can use. For instance, in the sentence
+generation domain, the XTAG planning problem for $k=2$ and $n=5$
+consists of about 1000 operators for the different lexicon entries for
+all the words in the sentence, some of which take four parameters. It
+is not unrealistic to assume a knowledge base with a few hundred
+individuals. All this adds up to trillions of ground instances, i.e.\
+it is completely infeasible to compute ground instances naively.
 
-By and large, our experiences with the planning community from a user's
-point of view have been favourable. Thanks to the planning competitions, it
-is easy to identify and download a fast state-of-the-art planner, at least
-for Linux. However, in the course of our experiments we found and reported
-bugs in both FF and SGPLAN. We also discovered that deciding as to which
-planner is appropriate for a particular task is not always straightforward,
-as the best performing planner in the IPC is not necessarily suitable for
-every planning task. As our experiments have shown, even planners as
-closely related as FF and SGPLAN can differ significantly in their
-performance on different domains (for instance, our results in
-Experiment~4).
+This means that while current planners deal well with the search
+problems we have looked at, they are still unusable for practical NLG
+applications because of the time they spend on preprocessing. In other
+words, generation-as-planning research is still not in a much better
+position than it was in the 1980s. While we are aware that the time
+that a planner invests in preprocessing can pay off during search, we
+still suggest that the inability of current planners to scale to
+larger domains still limits their usefulness for applications beyond
+NLG as well. Furthermore, we feel that preprocessing gets less
+research attention than it deserves: If it is scientifically trivial,
+we challenge the planning community to come up with implementations
+that only ground operators by need; otherwise, we look forward to
+publications on this topic. To support this, we offer our planning
+domains as benchmarks for future competitions. They are available at
+\todo{upload them somewhere}
+
+Finally, we found it very convenient that the recent IPC planning
+competitions provide an entry point for selecting and obtaining
+current planners. Nevertheless, our experiments exposed several bugs
+in the planners we tested, and required us to make changes to the
+source code to make them scale to our inputs. We also found that
+different planners differ in the variants of PDDL that they can
+process. These differences range from fragments of ADL that can be
+parsed to sensitivity to the order of declarations and the use of
+``objects'' rather than ``individuals'' as the keyword for declaring
+the universe. We propose that the case for planning as a mature
+technology with professional-quality implementations could be made
+more strongly if such discrepancies were harmonized.
 

### 2009-05-30 trunk/Papers/ci-spark/experiments.tex
rewrote discussion

@@
 The results of this experiment are shown in the graph in
-Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are
+Figure~\ref{fig:runtimes-crisp}.\footnote{All
+  runtimes were measured on a Pentium 4 CPU running at 3 GHz. Java programs
+  were allowed to ``warm up'', i.e., the planner was run five times and the
+  first four measurements discarded to ensure that the JVM had just-in-time
+  compiled all relevant bytecode.} The input parameters $(m,n)$ are
 plotted in lexicographic order on the horizontal axis, and the runtime
@@
 minutes, and it exceeded its memory limit of 16 GB for
-$n>3$.
+$n>3$. \todo{compare to SPUD reconstruction here}
 
@@
 $k=2$, we reach 65 seconds of total runtime at $n=4$, but here
-Metric-FF only spends about a second on search. For $k=3$, \todo{say
-  something true}. This is to be expected, given that the planning
+Metric-FF only spends about a second on search. For $k=3$, neither FF
+nor Metric-FF were able to solve any of the input
+instances. \todo{bla} This is to be expected, given that the planning
 operators for the verbs have $k+2$ parameters (see

### 2009-05-31 trunk/Papers/ci-spark/introduction.tex
changes to Sections 1 and 2

@@
 
-Natural language generation (NLG; \citealp{reiter00building}) is one
-of the major subfields of natural language processing, concerned with
-computing natural language sentences or texts that convey a given
-piece of information to the user. This task can be viewed as a problem
-of achieving a (communicative) goal by successively applying
-(communicative) actions, and thus has clear intuitive parallels to
-automated planning.
+\emph{Natural language generation} (NLG; \citealp{reiter00building}) is one of
+the major subfields of natural language processing, concerned with computing
+natural language sentences or texts that convey a given piece of information to
+an audience. While the output of a generation task can take many forms,
+including written text, synthesised speech, or embodied multimodal
+presentations, the underlying NLG problem in each case can be modelled as a
+problem of achieving a (communicative) goal by successively applying a set of
+(communicative) actions. This view of NLG as goal-directed action has clear
+parallels to \emph{automated planning}, which seeks to find general techniques
+for efficiently solving the action sequencing problem.
 
-This view of generation as planning has a long tradition in NLG
-\citep{perrault80,appelt:planning,hovy88,young94dpocl}. Despite their
-theoretical attractiveness, these early approaches never gained
-mainstream use in applied systems, in part because of the inefficiency
-of the planners that were available at the time. However, the recent
-improvements in planner efficiency and expressiveness have sparked a
-renewed interest in applying modern planning techniques to NLG---both
-on the level of speech act planning
-\citep{Steedman-Petrick:07,benotti08b} and and on the level of
-sentence generation \citep{KolSto07}. The goal of this paper is to
-determine, by looking at some representative generation problems,
-whether planning has advanced to the point that this enterprise can be
-successful this time---and thus, to determine the usefulness of
-current planning technology to an application area that has a long
-tradition, but is not currently in the focus of the planning
-competitions.
+Treating generation as planning has a long history in NLG, ranging from the
+initial attempts of the field to utilise early planning approaches
+\citep{perrault80,appelt:planning,hovy88,young94dpocl}, to the recent surge of
+research \citep{Steedman-Petrick:07,KolSto07,benotti08b} seeking to capitalise
+on the improvements modern planners offer in terms of efficiency and
+expressiveness. This paper attempts to assess the usefulness of current planning
+techniques to NLG by investigating some representative generation problems, and
+evaluating whether automated planning has advanced to the point that it can
+provide solutions to such NLG applications---applications that are not currently
+being investigated by mainstream planning research.
 
-The focus of this paper is twofold. First, we present two generation
-problems that have recently been cast as planning problems: the
-sentence generation task and the GIVE task. In the sentence generation
-task, the goal is to generate a single sentence that expresses a given
-meaning. \citet{KolSto07} cast this task as a planning problem, where
-the plan encodes the necessary sentence and the actions correspond to
-uttering individual words.  In the GIVE domain (``Generating
-Instructions in Virtual Environments''), we describe a new shared task
-that was recently posed as a challenge for the NLG community
-\citep{ByrKolStrCasDalMooObe09}.  GIVE uses planning as part of an NLG
-system that generates natural-language instructions to guide a user in
+The focus of this paper is twofold. First, we present two generation problems
+that have recently been cast as planning problems: the sentence generation task
+and the GIVE task. In the sentence generation task, we concentrate on generating
+a single sentence that expresses a given meaning. In this case, a plan encodes
+the necessary sentence with the actions in the plan corresponding to the
+utterance of individual words \citep{KolSto07}. In the GIVE domain (``Generating
+Instructions in Virtual Environments''), we describe a new shared task that was
+recently posed as a challenge for the NLG community
+\citep{ByrKolStrCasDalMooObe09}. GIVE uses planning as part of a larger NLG
+system for generating natural-language instructions that guide a human user in
 performing a given task in a virtual environment.
 
-Second, we discuss some of our experiences using off-the-shelf
-planners in these two domains. In particular, we explore the
-efficiency of FF \citep{HoffmannNebel01} and SGPLAN
-\citep{hsu06:_new_featur_in_sgplan_for}: two planners that are readily
-available, support an expressive subset of the Planning Domain
-Definition Language (PDDL; McDermott et al.~\citeyear{PDDL}) for
-encoding domains, and have been successful on the benchmarks and
-problems in the International Planning Competition (IPC)\footnote{See
-  \url{http://ipc.icaps-conference.org/} for information about past
-  editions of the IPC.}. We test these planners on a range of problem
-instances in our planning domains, and compare our results against an
-ad-hoc Java implementation of GraphPlan \citep{Blum1997}.
+Second, we discuss our experiences using a set of off-the-shelf planners in
+these two generation domains. Among the planners we test, we explore the
+efficiency of FF \citep{HoffmannNebel01}---a planner that has arguably had the
+greatest impact on recent approaches to deterministic planning---and its many
+descendants, such as SGPLAN \citep{hsu06:_new_featur_in_sgplan_for}. All of the
+planners we test are freely available, support an expressive subset of the
+Planning Domain Definition Language (PDDL; McDermott et al.~\citeyear{PDDL}),
+and have been successful on both standard planning benchmarks and the problems
+of the International Planning Competition (IPC).\footnote{See
+  \url{http://ipc.icaps-conference.org/} for information about recent and past
+  editions of the IPC. A good overview of the deterministic track of the 2004
+  instance of the IPC can be found in \citep{Hoffmann-Edelkamp:05}.}
+Using these planners---and an ad-hoc Java implementation of GraphPlan
+\citep{Blum1997} which serves as a baseline for certain experiments---we perform
+a series of tests on a range of problem instances in our NLG domains.
 
-Overall, our findings are mixed. On the one hand, we demonstrate that
-modern planners can easily handle the \emph{search} problems that
-arise in NLG on realistic inputs, which is promising given that the
-sentence generation task, at least, is NP-complete
-\citep{KolStr02}. On the other hand, the same planners spend
-tremendous amounts of time on preprocessing: Because our domains use a
-large number of actions and constants, FF spends 90\% of its runtime
-in the sentence generation domain on grounding out all the actions and
-literals. As a consequence, the off-the-shelf planners we looked at
-are too slow to be useful in the NLG application -- but this is due to
-implementation issues rather than the quality of the search
-algorithms. We offer our NLG domains as challenges for the planning
-community in the hope that these implementation issues might gain more
-attention in the future.
+Overall, our findings are mixed. On the one hand, we demonstrate that certain
+planners can readily handle the \emph{search} problems that arise in our testing
+domains on realistic inputs, which is promising given the challenging nature of
+these tasks (e.g., the sentence generation task is NP-complete \citep{KolStr02})
+On the other hand, these same planners spend tremendous amounts of time on
+\emph{preprocessing} techniques that analyse the problem domain in support of
+the search. (For instance, FF spends 90\% of its runtime in the sentence
+generation domain on grounding out actions and literals.) As a consequence, the
+off-the-shelf planners we investigated are often too slow to be useful in real
+NLG applications.  Nevertheless, we offer our NLG domains and experiences as
+both challenges and lessons for the planning community, in the hope that the
+issues we raise might gain more attention in the future.
 
-This paper is structured as follows. We begin by briefly reviewing
-some of the literature on NLG as planning
-(Section~\ref{sec:nlg-as-planning}). Then we define introduce the
-planning problems arising in two particular NLG tasks in more detail
-in Section~\ref{sec:domains}. We report on some experiments with these
-planning problems in Section~\ref{sec:experiments}, and discuss the
-results in Section~\ref{sec:discussion}. Section~\ref{sec:conclusion}
-concludes.
+The remainder of this paper is structured as follows. In
+Section~\ref{sec:nlg-as-planning}, we introduce the idea of NLG as planning and
+briefly review the relevant literature. In Section~\ref{sec:domains}, we
+describe a set of planning problems associated with two NLG tasks, namely
+sentence planning and instruction generation. In Section~\ref{sec:experiments},
+we report on the results of our experiments with these planning problems.
+In Section~\ref{sec:discussion} we discuss our results and overall
+experiences, and conclude in Section~\ref{sec:conclusion}.
 

### 2009-05-31 trunk/Papers/ci-spark/nlg-as-planning.tex
changes to Sections 1 and 2

@@
 
-The task of generating natural language from semantic representations
-(NLG) is typically split up into two parts: the \emph{discourse
-  planning} task, which selects the information that is to be conveyed
-and structures it into sentence-sized chunks, and the \emph{sentence
-  generation} tasks, which then translates each of these chunks into
-natural language sentences. The sentence generation task is often
-split up into two parts of its own---the \emph{sentence planning}
-task, which enriches the input by e.g.\ determining how to refer to
-objects and selects some lexical material, and the \emph{surface
-  realization} task, which maps the enriched meaning representation
-into a sentence using a grammar. The chain of domain planning,
-sentence planning, and surface realization is sometimes called the
-``NLG pipeline'' \cite{reiter00building}.
+The task of generating natural language from semantic representations (NLG) is
+typically split into two parts: the \emph{discourse planning} task, which
+selects the information to be conveyed and structures it into sentence-sized
+chunks, and the \emph{sentence generation} task, which then translates each of
+these chunks into natural language sentences. The sentence generation task is
+often divided into two parts of its own---the \emph{sentence planning} task,
+which enriches the input by, e.g., determining object references and selecting
+some lexical material, and the \emph{surface realization} task, which maps the
+enriched meaning representation into a sentence using a grammar. The chain of
+domain planning, sentence planning, and surface realization is sometimes called
+the ``NLG pipeline'' \citep{reiter00building}.
 
-Viewing generation as a planning problem has a long tradition in the
-NLG literature.  \citet{perrault80} presented an approach to discourse
-planning in which the planning operators represent individual speech
-acts such as ``request'' and ``inform''. This idea was later worked
-out further, e.g.\ by \citet{young94dpocl}. On the other hand,
-researchers such as \citet{appelt:planning} and \citet{hovy88} used
-techniques from hierarchical planning to expand a high-level plan
-consisting of speech acts into more detailed specifications of
-individual sentences. Their systems also covered some aspects of
-sentence planning. However, these systems used very expressive logics,
-which were designed to reason about beliefs and intentions, to
-represent the planning state and the planning operators. Furthermore,
-most of them used ad-hoc planning algorithms with rather naive search
-strategies, which did not scale to realistic inputs. As a consequence,
-the NLG-as-planning approach was mostly marginalized throughout the
-1990s.
+Viewing generation as a planning problem has a long tradition in the NLG
+literature. \citet{perrault80} presented an approach to discourse planning in
+which the planning operators represented individual speech acts such as
+``request'' and ``inform''. This idea was later expanded, e.g., by
+\citet{young94dpocl}. On the other hand, researchers such as
+\citet{appelt:planning} and \citet{hovy88} used techniques from hierarchical
+planning to expand a high-level plan consisting of speech acts into more
+detailed specifications of individual sentences. Their systems also covered some
+aspects of sentence planning. However, these systems used very expressive
+logics, which were designed to reason about beliefs and intentions, to represent
+the planning state and the planning operators. Furthermore, most of them used
+ad-hoc planning algorithms with rather naive search strategies, which did not
+scale to realistic inputs. As a consequence, the NLG-as-planning approach was
+mostly marginalized throughout the 1990s.
 
-However, there has recently been a string of publications by various
-authors with renewed interest in the generation-as-planning approach,
-motivated by the recent development of efficient and expressive
-planners. Below (Section~\ref{sec:domain-crisp}) we will look in some
-detail at \cite{KolSto07}'s approach to sentence generation (i.e., the
+More recently, there has been a string of publications by various authors with a
+renewed interest in the generation-as-planning approach, motivated by the
+ongoing development of increasingly more efficient and expressive planners. For
+instance, \cite{KolSto07} propose an approach to sentence generation (i.e., the
 sentence planning and surface realization modules of the pipeline) as
-planning. In addition, \citet{Steedman-Petrick:07} revisit the
-analysis of indirect speech acts with modern planning technology;
-\citet{benotti08b} uses planning to explain the accommodation of
-presuppositions; and
-\citet{brenner08:_contin_multiag_plann_approac_to_situat_dialog} use
-multi-agent planning to model the joint problem-solving behavior of
-agents in a situated dialogue. These approaches focus on different
-issues than the 1980's NLG-as-planning literature. But the main
-difference is that they apply \emph{existing}, well-understood
-planning approaches to linguistic problems because these offer new
-modeling tools, and because they hope that modern planners can solve
-the hard search problems that arise in NLG \citep{KolStr02}
-efficiently. The point of this paper is to investigate whether
-existing planners fulfill this latter promise.
+planning---an approach we will explore in more detail below
+(Section~\ref{sec:domain-crisp}). \citet{Steedman-Petrick:07} revisit the
+analysis of indirect speech acts with modern planning technology, viewing the
+problem as an instance of the general problem of planning with incomplete
+information and sensing actions. In addition, \citet{benotti08b} uses planning
+to explain the accommodation of presuppositions, and
+\citet{brenner08:_contin_multiag_plann_approac_to_situat_dialog} use multi-agent
+planning to model the joint problem-solving behaviour of agents in a situated
+dialogue. While these approaches focus on different issues compared to the
+1980's NLG-as-planning literature, they all apply \emph{existing},
+well-understood planning approaches to linguistic problems, in order to utilise
+the rich set of modelling tools provided by modern planners, and in the hope
+that such planners can efficiently solve the hard search problems that arise in
+NLG \citep{KolStr02}. This paper aims to investigate whether existing planners
+achieve this latter goal.
 

### 2009-06-01 trunk/Papers/ci-spark/domains.tex
gen as planning
figures moved to the right places
sect 3

@@
 
-We will now focus the discussion towards two specific NLG problems:
-sentence generation in the sense of \cite{KolSto07}, and the
+We will now focus the discussion on two specific NLG problems:
+sentence generation in the sense of \citet{KolSto07}, and the
 generation of instructions in virtual environments
-\cite{ByrKolStrCasDalMooObe09}.
+\citep{ByrKolStrCasDalMooObe09}. For each, we will introduce the task
+and show by example how the task can be seen as a planning problem.
 
 \subsection{Sentence generation as planning}
+\label{sec:domain-crisp}
 
@@
 semantic and pragmatic conditions, and achieves all the
-\emph{communicative goals}. An example is the tree-adjoining grammar
-(TAG; \citep{joshi;etal1997}) shown in
+\emph{communicative goals}.
+
+An example is the tree-adjoining grammar (TAG;
+\citealt{joshi;etal1997}) shown in
 Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
@@
 base) into a TAG derivation as as shown in
-Figure~\ref{fig:white-rabbit-sleeps-deriv}. We can then read off the
-sentence ``The white rabbit sleeps'' from the derivation. \todo{need
-  to mention distractors and unique REs here}
+Figure~\ref{fig:white-rabbit-sleeps-deriv}. In the picture, the dashed
+arrow indicates TAG's \emph{substitution} operation, which ``plugs''
+an elementary into the leaf of another tree, and the dotted arrows
+stand for \emph{adjunction}, which splices an elementary tree into an
+internal node. We can then read off the sentence ``The white rabbit
+sleeps'' from the derivation. Note that the sentence ``The rabbit
+sleeps'' would not have been an appropriate result, because ``the
+rabbit'' could refer to either $r_1$ or $r_2$, and thus $r_2$ remains
+as a \emph{distractor}, i.e.\ an incorrect possible interpretation of
+the phrase.
 
-\begin{figure}[t]
+\begin{figure}
   \centering
@@
 
-\begin{figure}[t]
+\begin{figure}
   \centering
@@
 This perspective on sentence generation has the advantage of solving
-the sentence planning and surface realization problems simultaneously:
-in the example, we can require that referring expressions ``the white
-rabbit'' can be resolved uniquely to $r_1$ by the hearer, in addition
-to the requirement that the derivation is grammatically correct. This
-is useful in cases where sentence planning and surface realization
-interact, because syntactic information about the individual words is
-available when the REs are generated (see
-e.g. \cite{stone98textual}). However, the problem of whether a given
-communicative goal can be achieved with a given grammar is NP-complete
-\cite{KolStr02}. A naive search algorithm that computes the derivation
-top-down takes exponential time and is clearly infeasible to use in
-practice. The SPUD system \cite{Stone2003a}, which worked this idea
-out for the first time, uses a greedy algorithm, which circumvents the
-combinatorial explosion but is incomplete.
+the sentence planning and surface realization problems
+simultaneously. The generation of referring expressions (REs) is
+usually seen as a sentence planning task; in the example, we required
+that the referring expression ``the white rabbit'' can be resolved
+uniquely to $r_1$ by the hearer, in addition to the requirement that
+the derivation is grammatically correct. This is useful in cases where
+sentence planning and surface realization interact, because syntactic
+information about the individual words is available when the REs are
+generated (see e.g. \citealt{stone98textual}). However, the problem of
+whether a given communicative goal can be achieved with a given
+grammar is NP-complete \citep{KolStr02}. A naive search algorithm that
+computes the derivation top-down takes exponential time and is clearly
+infeasible to use in practice. The seminal SPUD system
+\citep{Stone2003a}, which first worked out the idea of integrated
+TAG-based sentence generation, used a greedy algorithm to cope with
+the search, which circumvents the combinatorial explosion but is
+incomplete.
 
-\begin{figure}[p]
+\begin{figure}
 \centering
-\begin{minipage}{0.5\textwidth}
+\begin{minipage}{0.8\textwidth}
 {\small%
@@
 (:action add-sleeps
-   :parameters (?u - node
-                ?xself - individual
-                ?xsubj - individual)
+   :parameters (?u - node  ?xself - individual  ?xsubj - individual)
    :precondition
-       (and (subst S ?u)
-            (referent ?u ?xself)
-            (sleep ?xself ?xsubj))
+       (and (subst S ?u)  (referent ?u ?xself)  (sleep ?xself ?xsubj))
    :effect 
-       (and (not (subst S ?u))
-            (expressed sleep ?xself ?xsubj)
-            (subst NP (subj ?u))
-            (referent (subj ?u) ?xsubj)
+       (and (not (subst S ?u))  (expressed sleep ?xself ?xsubj)
+            (subst NP (subj ?u))  (referent (subj ?u) ?xsubj)
             (forall (?y - individual)
-                (when (not (= ?y ?xself))
-                    (distractor (subj ?u) ?y)))))
+                (when (not (= ?y ?xself)) (distractor (subj ?u) ?y)))))
 
 (:action add-rabbit
-   :parameters (?u - node
-                ?xself - individual)
+   :parameters (?u - node  ?xself - individual)
    :precondition 
-       (and (subst NP ?u)
-            (referent ?u ?xself)
-            (rabbit ?xself))
+       (and (subst NP ?u)  (referent ?u ?xself)  (rabbit ?xself))
    :effect 
-       (and (not (subst NP ?u))
-            (canadjoin N ?u)
+       (and (not (subst NP ?u))  (canadjoin N ?u)
             (forall (?y - individual)
-                (when (not (rabbit ?y))
-                    (not (distractor ?u ?y))))))
+                (when (not (rabbit ?y)) (not (distractor ?u ?y))))))
 
 (:action add-white
-   :parameters (?u - node
-                ?xself - individual)
+   :parameters (?u - node  ?xself - individual)
    :precondition 
-       (and (canadjoin N ?u)
-            (referent ?u ?xself)
-            (rabbit ?xself))
+       (and (canadjoin N ?u)  (referent ?u ?xself)  (rabbit ?xself))
    :effect 
        (forall (?y - individual)
-           (when (not (white ?y))
-               (not (distractor ?u ?y)))))
+           (when (not (white ?y)) (not (distractor ?u ?y)))))
 \end{verbatim}}%
@@
 
-Here we will look at one recent approach which aims at controlling the
+As an alternative, \citet{KolSto07} recently proposed to control the
 search by converting the sentence generation problem into a planning
@@
   \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
-  which this conversion is implemented.} Each operator of this
-planning problem models the addition of some elementary tree to the
-TAG derivation. For instance,
-Figure~\ref{fig:white-rabbit-as-planning} shows the corresponding PDDL
-actions for the above generation task. The $\addact{sleeps}$ operator
-represents the addition of the elementary tree for ``sleeps'' to the
-derivation. $\addact{sleeps}(\mathsf{root},e,r_1)$ removes the atom
-$\mathsf{subst}(S,\mathsf{root})$, which expresses that there is
-currently an open substitution node $\mathsf{root}$ with label $S$,
-from the planning state and replaces it with an atom
-$\mathsf{subst}(NP,\mathsf{subj}(\mathsf{root}))$; here
-$\mathsf{subj}(\mathsf{root})$ is meant as shorthand for a fresh
-individual name.\footnote{These terms, which are not valid in ordinary
-  PDDL, can be eliminated by estimating an upper bound $n$ for the
-  plan length, making $n$ copies of each action, ensuring that copy
-  $i$ can only be applied in step $i$, and replacing the term
+  which this conversion is implemented.} This planning problem assumes
+an initial state containing an atom $\mathsf{subst}(S,\mathsf{root})$,
+encoding that we must generate a sentence ($S$) starting at the node
+named $\mathsf{root}$ in the TAG derivation tree, and an atom
+$\mathsf{referent}(\mathsf{root},e)$ encoding that the (event)
+individual which the entire sentence describes is $e$. Then it encodes
+the addition of elementary trees to the TAG derivation in the
+individual planning operators; the ones that are needed for our
+example above are shown in
+Fig.~\ref{fig:white-rabbit-as-planning}.
+
+Here the action instance $\addact{sleeps}(\mathsf{root}, e, r_1)$
+replaces the atom $\mathsf{subst}(S,\mathsf{root})$ with the atom
+$\mathsf{subst}(NP,\mathsf{subj}(\mathsf{root}))$. In an abuse of PDDL
+syntax, we write $\mathsf{subj}(\mathsf{root})$ as shorthand for a
+fresh individual name.\footnote{These terms, which are not valid in
+  ordinary PDDL, can be eliminated by estimating an upper bound $n$
+  for the plan length, making $n$ copies of each action, ensuring that
+  copy $i$ can only be applied in step $i$, and replacing the term
   $\mathsf{subj}(u)$ in an action copy by the constant
-  $\mathsf{subj}_i$. Notice that $S$, $NP$, and $N$ are constants.}
-At the same time, the operator records that the semantic information
-$\mathsf{sleep}(e,r_1)$ has now been expressed, and introduces all
-individuals except for $r_1$ as distractors for the new RE at
-$\mathsf{subj}(\mathsf{root})$. These distractors can then be removed
-by subsequent applications of the other two operators. By assuming an
-initial state that encodes the knowledge base and contains an atom
-$\mathsf{subst}(S,\mathsf{root})$, and a goal that contains, among
-others, $\forall x \forall y. \neg \mathsf{subst}(x,y)$ and
-$\mathsf{expressed}(\mathsf{sleep},e,r_1)$, a planner will find plans
-like the following:
+  $\mathsf{subj}_i$. The terms $S$, $NP$, and $N$ in the planning
+  problem are constants.}  At the same time, the operator records that
+the semantic information $\mathsf{sleep}(e,r_1)$ has now been
+expressed, and introduces all individuals except for $r_1$ as
+distractors for the new RE at $\mathsf{subj}(\mathsf{root})$. These
+distractors can then be removed by subsequent applications of the
+other two operators. Eventually we reach a goal state, which is
+characterized by goals including $\forall x \forall y. \neg
+\mathsf{subst}(x,y)$, $\forall x \forall y. \neg
+\mathsf{distractor}(x,y)$, and
+$\mathsf{expressed}(\mathsf{sleep},e,r_1)$. In summary, the following
+is a correct plan:
 %
@@
 %
-Using this plan, the grammatical derivation in
+The grammatical derivation in
 Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
 generated sentence ``the white rabbit sleeps'', can be systematically
-reconstructed. Thus, we can solve the sentence generation problem via
-the detour through planning and bring current search heuristics for
-planning to bear on generation.
+reconstructed from this plan. Thus, we can solve the sentence
+generation problem via the detour through planning and bring current
+search heuristics for planning to bear on generation.
 
 \subsection{Planning in instruction giving}
+\label{sec:domain-give}
 
-Let's now turn to a second recent application of planning in NLG, the
+Let us now turn to a second recent application of planning in NLG, the
 GIVE Challenge (``Generating Instructions in Virtual Environments'';
@@
 components of an NLG system, and emphasizes the study of communication
-in a (simulated) physical environment. Another advantage is that GIVE
-makes it possible to connect NLG systems to users over the Internet,
-and thus offer cheap access to human experimental subjects. The first
-installment of GIVE (GIVE-1) evaluated five NLG systems on the
-performance of 1143 users, making it the largest NLG evaluation effort
-in terms of human users ever.
+in a (simulated) physical environment. Furthermore, because the client
+which displays the 3D environment to the user and the NLG system can
+be physically separated, as long as they are connected over a network,
+such systems can be cheaply evaluated over the Internet. This provides
+a potential solution to the long-standing problem of evaluating NLG
+systems. The first installment of GIVE (GIVE-1) evaluated five NLG
+systems on the performance of 1143 users, making it the largest NLG
+evaluation effort in terms of human users ever.
 
@@
 the example GIVE world map in
-Figure~\ref{fig:give-development-world}. To simplify both the planning
-and the NLG task, this world is discretised into a set of tiles of
+Figure~\ref{fig:give-development-world}. In this example world, the
+user's task is to pick up a trophy in the top left room. The trophy is
+hidden in a safe behind a picture; to access it, the user must push
+certain buttons in order to move the picture out of the way, open the
+safe, and open doors. The user must to navigate the world and perform
+these actions in the 3D client in a certain order; the NLG system must
+instruct the user on how to do this.  To simplify both the planning
+and the NLG task, the world is discretised into a set of tiles of
 equal size. The user can turn by 90 degree steps in either direction,
 and can move from the centre of one tile to the centre of the next
-tile, provided the path between two tiles is not blocked. The world
-also consists of a set of objects which can be manipulated by the user
-in various ways. For instance, in the example world the user's task is
-to pick up a trophy in the top left room. The trophy is hidden in a
-safe behind a picture; to access it, the user must push certain
-buttons in order to move the picture out of the way, open the safe,
-and open doors. Figure~\ref{fig:give-planning} shows the encoding of
-some of the available GIVE domain actions, in PDDL syntax. In the
-example, the shortest plan to solve the task consists of 108 action
-steps, with the first few steps as follows:
+tile, provided the path between two tiles is not blocked.
+Figure~\ref{fig:give-planning} shows the encoding of some of the
+available GIVE domain actions, in PDDL syntax. In the example, the
+shortest plan to solve the task consists of 108 action steps, with the
+first few steps as follows:
 %
@@
 
-\begin{figure}[p]
+\begin{figure}
 \centering
@@
 user. Under this view, the planning problem is very similar to the
-Gridworld problem (see, e.g., \citep{Tovey-Koenig:2000}), which also
+Gridworld problem (see, e.g., \citealt{Tovey-Koenig:2000}), which also
 involves finding a route through a two-dimensional world map with
-discrete positions. GIVE tasks are usually much more complex, however,
-as illustrated by the above example map: a successful plan must
-include steps for pressing buttons in the right order, reasoning about
-large numbers of world objects, and navigating through complicated
-room shapes. This relatively loose integration of NLG system and
-planner is the state of the art of the systems that participated in
-GIVE-1.
+discrete positions. As in Gridworld, the domain requires not only
+route finding in the discretised world, but also the execution of
+certain actions (such as finding keys and unlocking doors in
+Gridworld, or pushing the correct buttons to open doors and the safe
+in GIVE). However, the GIVE worlds tend to be bigger than the
+Gridworld instances that were used in the 1990s planning competitions.
 
-However, it is desirable to integrate the planner and the generation
-system more closely than this. For instance, let's say that the NLG
-system wants to generate the instruction sequence ``walk to the centre
-of the room; turn right; now press the green button in front of you''.
-Experiments \todo{cite} show that this is a pattern that human
-instruction givers use frequently: The instruction follower is made to
-walk to a certain point in the world at which the instruction giver
-can then use a a referring expression (``the green button'') that is
-very easy to interpret. Such an NLG system must integrate discourse
-planning and planning in the domain of the world map closely: On the
-one hand, the structure of the discourse is determined by the needs of
-the NLG system rather than the domain plan; on the other hand, the
-discourse planner must be aware of the way in which the instruction
-``turn right'' is likely to change the visibility of objects. Even if
-the NLG system doesn't implement the generation of this discourse as
-planning, it still solves a problem that subsumes the domain planning
-problem. For these reasons, we consider the GIVE domain planning
-problem as a natural part of a GIVE NLG system.
-\todo{Make sure that this paragraph still talks about the things that
-  the reviewers found interesting about it.}
+This relatively loose integration of NLG system and planner is the
+state of the art of the systems that participated in GIVE-1. However,
+it is desirable to integrate the planner and the generation system
+more closely than this. For instance, let's say that the NLG system
+wants to generate the instruction sequence ``walk to the centre of the
+room; turn right; now press the green button in front of
+you''. Experiments with human instruction givers \cite{stoia:08} show
+that this is a pattern that they use frequently: The instruction
+follower is made to walk to a certain point in the world at which the
+instruction giver can then use a a referring expression (``the green
+button'') that is very easy to interpret. Such an NLG system must
+integrate discourse planning and planning in the domain of the world
+map closely. On the one hand, the structure of the discourse is
+determined by the needs of the NLG system rather than the domain plan;
+on the other hand, the discourse planner must be aware of the way in
+which the instruction ``turn right'' is likely to change the
+visibility of objects. Even if the NLG system doesn't implement the
+generation of this discourse as planning, it still solves a problem
+that subsumes the domain planning problem. For these reasons, we
+consider the GIVE domain planning problem as a natural part of a GIVE
+NLG system.  \todo{Make sure that this paragraph still talks about the
+  things that the reviewers found interesting about it.}
 

### 2009-06-01 trunk/Papers/ci-spark/introduction.tex
sect 3

@@
 The remainder of this paper is structured as follows. In
-Section~\ref{sec:nlg-as-planning}, we introduce the idea of NLG as planning and
-briefly review the relevant literature. In Section~\ref{sec:domains}, we
-describe a set of planning problems associated with two NLG tasks, namely
-sentence planning and instruction generation. In Section~\ref{sec:experiments},
-we report on the results of our experiments with these planning problems.
-In Section~\ref{sec:discussion} we discuss our results and overall
+Section~\ref{sec:nlg-as-planning}, we introduce the idea of NLG as
+planning and briefly review the relevant literature. In
+Section~\ref{sec:domains}, we describe a set of planning problems
+associated with two NLG tasks, namely sentence planning and situated
+instruction generation. In Section~\ref{sec:experiments}, we report on
+the results of our experiments with these planning problems.  In
+Section~\ref{sec:discussion} we discuss our results and overall
 experiences, and conclude in Section~\ref{sec:conclusion}.

### 2009-06-01 trunk/Papers/ci-spark/conclusion.tex
fixed conclusion

@@
 
-In this paper, we introduced two novel planning domains arising from
-problems in natural language generation: the sentence generation domain and
-the Generating Instructions in Virtual Environments (GIVE) challenge
-domain. We also reported the results of a set of experiments designed to
-evaluate the suitability of off-the-shelf planners (particularly FF and
-SGPLAN) in these domains.
+In this paper, we have investigated the usefulness of current planning
+technology to natural language generation, an application area with a
+long tradition that has recently experienced renewed interest from NLG
+researchers. In particular, we have evaluated the performance of
+several off-the-shelf planners on a series of planning domains that
+arise in the context of sentence generation and situated instruction
+generation.
 
-Our results were mixed. While the planners we tested do an impressive job
-of controlling the complexity of search in many domains, they also suffer
-from practical problems that limit their performance in unexpected ways: in
-both of our NLG domains the grounding step performed by FF and SGPLAN
-dominates the time it takes to perform the search itself. As future work,
-we hope to extend our analysis to other planners that do not perform the
-initial grounding step, planners such as SHOP2
-\citep{DBLP:journals/jair/NauAIKMWY03} that have traditionally performed
-well on a wide range of domains, and new planners introduced in the latest
-edition of the IPC.\footnote{See
- \url{http://ipc.informatik.uni-freiburg.de/} for details of the 2008
- International Planning Competition (deterministic track).}
+Our results were mixed. While some of the planners we tested---in
+particular, FF and SGPLAN, in different domains---did an impressive
+job of controlling the complexity of the search, we also found that
+all planners we tested spent too much time on preprocessing to be
+useful. In the sentence generation domain, FF spent 90\% of its
+runtime on computing the ground instances of the planning operators;
+in the instruction-giving domain, which is very similar to Gridworld,
+a similar effect happened once we added more predicates. As things
+stand, we found that this overly long preprocessing time makes current
+planners inapplicable to NLG application for any but the smallest
+instances. Users who come to planning from the outside, such as NLG
+researchers, take planners as a black box. This means that search
+efficiency alone is not helpful when other modules of the planner are
+slow. From this perspective, we propose that the planning community
+might spend some attention on optimizing the preprocessing module with
+similar vigor as the search itself. In particular, we propose to
+investigate planning algorithms that do not have to rely on grounding
+out all operators before starting the search.
 
-The NLG community's recent interest in planning presents a valuable
-opportunity for planning researchers. Domains like GIVE highlight certain
-challenges, such as plan execution monitoring and plan presentation (i.e.,
-summarisation and elaboration), but also offer a platform on which such
-technologies can be evaluated in experiments with human users. Although we
-have focused on classical planning problems in this work, research related
-to reasoning under uncertainty, resource management, and planning with
-knowledge and sensing, can also be investigated in these settings. As such,
-we believe our domains would provide suitable challenges for planners
-entered in future editions of the IPC. Provided some of the challenges we
-have identified can be addressed, NLG-inspired problem domains offer a
-constructive platform for the planning community to showcase their
-techniques to a wider audience---and to improve the quality of their tools
-for real-world planning tasks.
-
-\todo{planning claims general applicability; it's clear that IPC
-  benchmarks focus areas of improvement; but it's not so clear to what
-  extent this limits usefulness for non-IPC application areas; this
-  paper evaluates this (for the first time??)}
+NLG and planning have a long history in common. The recent surge in
+NLG-as-planning research presents valuable opportunities for both
+sides. Clearly, NLG researchers who apply planning technology will
+benefit directly from any improvements in planner effiency. But
+conversely, NLG may be a worthwhile application area for planning
+researchers to keep in mind. Domains like GIVE highlight certain
+challenges, such as plan execution monitoring and plan presentation
+(i.e., summarisation and elaboration), but also offer a platform on
+which such technologies can be evaluated in experiments with human
+users. Further, although we have focused on classical planning
+problems in this work, research related to reasoning under
+uncertainty, resource management, and planning with knowledge and
+sensing, can also be investigated in these settings. As such, we
+believe our domains would provide interesting challenges for planners
+entered in future editions of the IPC.
 
@@
 
-This work arose in the context of the Planning and Language Interest Group
-at the University of Edinburgh. The authors would like to thank all members
-of this group, especially Hector Geffner and Mark Steedman, for interesting
-discussions. This work was partially supported by the DFG Research
-Fellowship ``CRISP: Efficient integrated realization and microplanning''
-and by the European Commission through the PACO-PLUS project
-(FP6-2004-IST-4-27657).
+This work arose in the context of the Planning and Language Interest
+Group at the University of Edinburgh. The authors would like to thank
+all members of this group, especially Hector Geffner and Mark
+Steedman, for interesting discussions. We also thank our reviewers for
+their insightful and challenging comments.  This work was supported by
+the DFG Research Fellowship ``CRISP: Efficient integrated realization
+and microplanning'', the DFG Cluster of Excellence ``Multimodal
+Computing and Interaction'', and by the European Commission through
+the PACO-PLUS project (FP6-2004-IST-4-27657).
 

### 2009-06-01 trunk/Papers/ci-spark/discussion.tex
fixed conclusion

@@
 
-This means that while current planners deal well with the search
-problems we have looked at, they are still unusable for practical NLG
-applications because of the time they spend on preprocessing. In other
-words, generation-as-planning research is still not in a much better
-position than it was in the 1980s. While we are aware that the time
-that a planner invests in preprocessing can pay off during search, we
-still suggest that the inability of current planners to scale to
-larger domains still limits their usefulness for applications beyond
-NLG as well. Furthermore, we feel that preprocessing gets less
-research attention than it deserves: If it is scientifically trivial,
-we challenge the planning community to come up with implementations
-that only ground operators by need; otherwise, we look forward to
-publications on this topic. To support this, we offer our planning
-domains as benchmarks for future competitions. They are available at
-\todo{upload them somewhere}
+Of course, it would be premature to judge the usefulness of current
+planners for NLG as a whole based on just two NLG
+domains. Nevertheless, we believe that the structure of our planning
+problems, which are dominated by a large number of operators and
+individuals, is typical for NLG-related planning problems as a
+whole. This suggests strongly that while current planners deal well
+with the search problems we have looked at, they are still unusable
+for practical NLG applications because of the time they spend on
+preprocessing. In other words, generation-as-planning research is
+still not in a much better position than it was in the 1980s. While we
+are aware that the time that a planner invests in preprocessing can
+pay off during search, we still suggest that the inability of current
+planners to scale to larger domains still limits their usefulness for
+applications beyond NLG as well. Furthermore, we feel that
+preprocessing gets less research attention than it deserves: If it is
+scientifically trivial, we challenge the planning community to come up
+with implementations that only ground operators by need; otherwise, we
+look forward to publications on this topic. To support this, we offer
+our planning domains as benchmarks for future competitions. They are
+available at \todo{upload them somewhere}
 

### 2009-06-01 trunk/Papers/ci-spark/introduction.tex
fixed conclusion

@@
 
-Treating generation as planning has a long history in NLG, ranging from the
-initial attempts of the field to utilise early planning approaches
-\citep{perrault80,appelt:planning,hovy88,young94dpocl}, to the recent surge of
-research \citep{Steedman-Petrick:07,KolSto07,benotti08b} seeking to capitalise
-on the improvements modern planners offer in terms of efficiency and
-expressiveness. This paper attempts to assess the usefulness of current planning
-techniques to NLG by investigating some representative generation problems, and
-evaluating whether automated planning has advanced to the point that it can
-provide solutions to such NLG applications---applications that are not currently
-being investigated by mainstream planning research.
+Treating generation as planning has a long history in NLG, ranging
+from the initial attempts of the field to utilise early planning
+approaches \citep{perrault80,appelt:planning,hovy88,young94dpocl}, to
+a recent surge of research
+\citep{Steedman-Petrick:07,KolSto07,brenner08:_contin_multiag_plann_approac_to_situat_dialog,benotti08b}
+seeking to capitalise on the improvements modern planners offer in
+terms of efficiency and expressiveness. This paper attempts to assess
+the usefulness of current planning techniques to NLG by investigating
+some representative generation problems, and evaluating whether
+automated planning has advanced to the point that it can provide
+solutions to such NLG applications---applications that are not
+currently being investigated by mainstream planning research.
 
-The focus of this paper is twofold. First, we present two generation problems
-that have recently been cast as planning problems: the sentence generation task
-and the GIVE task. In the sentence generation task, we concentrate on generating
-a single sentence that expresses a given meaning. In this case, a plan encodes
-the necessary sentence with the actions in the plan corresponding to the
-utterance of individual words \citep{KolSto07}. In the GIVE domain (``Generating
-Instructions in Virtual Environments''), we describe a new shared task that was
-recently posed as a challenge for the NLG community
-\citep{ByrKolStrCasDalMooObe09}. GIVE uses planning as part of a larger NLG
-system for generating natural-language instructions that guide a human user in
-performing a given task in a virtual environment.
+To answer this question, we proceed in two steps. First, we present
+two generation problems that have recently been cast as planning
+problems: the sentence generation task and the GIVE task. In the
+sentence generation task, we concentrate on generating a single
+sentence that expresses a given meaning. In this case, a plan encodes
+the necessary sentence with the actions in the plan corresponding to
+the utterance of individual words \citep{KolSto07}. In the GIVE domain
+(``Generating Instructions in Virtual Environments''), we describe a
+new shared task that was recently posed as a challenge for the NLG
+community \citep{ByrKolStrCasDalMooObe09}. GIVE uses planning as part
+of a larger NLG system for generating natural-language instructions
+that guide a human user in performing a given task in a virtual
+environment.
 
-Second, we discuss our experiences using a set of off-the-shelf planners in
-these two generation domains. Among the planners we test, we explore the
-efficiency of FF \citep{HoffmannNebel01}---a planner that has arguably had the
-greatest impact on recent approaches to deterministic planning---and its many
-descendants, such as SGPLAN \citep{hsu06:_new_featur_in_sgplan_for}. All of the
-planners we test are freely available, support an expressive subset of the
-Planning Domain Definition Language (PDDL; McDermott et al.~\citeyear{PDDL}),
-and have been successful on both standard planning benchmarks and the problems
-of the International Planning Competition (IPC).\footnote{See
-  \url{http://ipc.icaps-conference.org/} for information about recent and past
-  editions of the IPC. A good overview of the deterministic track of the 2004
-  instance of the IPC can be found in \citep{Hoffmann-Edelkamp:05}.}
-Using these planners---and an ad-hoc Java implementation of GraphPlan
-\citep{Blum1997} which serves as a baseline for certain experiments---we perform
-a series of tests on a range of problem instances in our NLG domains.
+Second, we evaluate the performance of current off-the-shelf planners
+in the planning domains into which these two generation problems
+translate. Among the planners we test, we explore the efficiency of FF
+\citep{HoffmannNebel01}---a planner that has arguably had the greatest
+impact on recent approaches to deterministic planning---and some of its
+descendants, such as SGPLAN
+\citep{hsu06:_new_featur_in_sgplan_for}. All of the planners we test
+are freely available, support an expressive subset of the Planning
+Domain Definition Language (PDDL; McDermott et al.~\citeyear{PDDL}),
+and have been successful on both standard planning benchmarks and the
+problems of the International Planning Competition (IPC).\footnote{See
+  \url{http://ipc.icaps-conference.org/} for information about recent
+  and past editions of the IPC. A good overview of the deterministic
+  track of the 2004 instance of the IPC can be found in
+  \citep{Hoffmann-Edelkamp:05}.}  Using these planners---and an ad-hoc
+Java implementation of GraphPlan \citep{Blum1997} which serves as a
+baseline for certain experiments---we perform a series of tests on a
+range of problem instances in our NLG domains.
 
-Overall, our findings are mixed. On the one hand, we demonstrate that certain
-planners can readily handle the \emph{search} problems that arise in our testing
-domains on realistic inputs, which is promising given the challenging nature of
-these tasks (e.g., the sentence generation task is NP-complete \citep{KolStr02})
-On the other hand, these same planners spend tremendous amounts of time on
-\emph{preprocessing} techniques that analyse the problem domain in support of
-the search. (For instance, FF spends 90\% of its runtime in the sentence
-generation domain on grounding out actions and literals.) As a consequence, the
-off-the-shelf planners we investigated are often too slow to be useful in real
-NLG applications.  Nevertheless, we offer our NLG domains and experiences as
-both challenges and lessons for the planning community, in the hope that the
-issues we raise might gain more attention in the future.
+Overall, our findings are mixed. On the one hand, we demonstrate that
+some planners can readily handle the \emph{search} problems that arise
+in our testing domains on realistic inputs, which is promising given
+the challenging nature of these tasks (e.g., the sentence generation
+task is NP-complete; see \citealt{KolStr02}). On the other hand, these
+same planners spend tremendous amounts of time on \emph{preprocessing}
+to analyse the problem domain in support of the search. On many of our
+problem instances, the preprocessing time overshadows the search time.
+For instance, FF spends 90\% of its runtime in the sentence generation
+domain on grounding out actions and literals. As a consequence, we
+conclude that the off-the-shelf planners we investigated are too slow
+to be useful in real NLG applications.  We hope that our results will
+spark some interest for more efficient implementations of planning
+preprocessors, and offer our NLG domains and experiences as challenges
+for the planning community.
 

### 2009-06-01 trunk/Papers/ci-spark/experiments.tex
xtag graphs
figures moved to the right places
using subfig
added modifier runtime picture

@@
   \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp1}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-modifiers-with-sgplan}
+  \includegraphics[width=\columnwidth]{data/modifiers-all}
   \caption{Results for the sentence generation domain. The
@@
 The results of this experiment are shown in the graph in
-Figure~\ref{fig:runtimes-crisp}.\footnote{All
-  runtimes were measured on a Pentium 4 CPU running at 3 GHz. Java programs
-  were allowed to ``warm up'', i.e., the planner was run five times and the
-  first four measurements discarded to ensure that the JVM had just-in-time
-  compiled all relevant bytecode.} The input parameters $(m,n)$ are
-plotted in lexicographic order on the horizontal axis, and the runtime
-is shown in milliseconds on the vertical axis.\footnote{These times
-  were measured on XXX} These results reveal a number of interesting
-insights. First, FF significantly outperforms SGPLAN in this domain,
-sometimes by a factor of 10 or more.\footnote{Experiments with SGPLAN
-  use a pre-release version of SGPLAN, kindly provided to us by
-  Chih-Wei Hsu.The release version of SGPLAN 5.2.2 had a bug which
-  caused it to crash on some of our problem instances.}  Second, FF's
-runtime is dominated by its initial grounding step, in which it
-computes the ground instances of all predicates and actions used in
-the planning problem in order to avoid unnecessary instantiation
-during search.  In particular, the ratio of grounding time to total
-runtime is generally above 85\%, and rises to above 99\% at $m=11$,
-which is still a relatively small universe for this
-application.\footnote{The ``grounding'' time reported here is what FF
-  reports as ``time spent: instantiating action templates''.}
+Figure~\ref{fig:runtimes-crisp}.\footnote{All runtimes in
+  Sections~\ref{sec:exper-1:-sent} and \ref{sec:exper-2:-sent-xtag}
+  were measured on an AMD Opteron 8220 CPU running at 2.8 GHz, under
+  Linux. FF 2.3 and Metric-FF were recompiled as 64-bit binaries. Java
+  programs were executed under Java 1.6.0\_13 in 64-bit mode. They
+  were allowed to ``warm up'', i.e. the JVM was given the opportunity
+  to just-in-time compile the relevant bytecode by running the planner
+  three times and discarding the runtimes before taking the actual
+  measurements. All runtimes are averaged over three runs of
+  the planners.} The input parameters $(m,n)$ are plotted in
+lexicographic order on the horizontal axis, and the runtime is shown
+in seconds on the vertical axis. These results reveal a number of
+interesting insights. First, FF significantly outperforms SGPLAN in
+this domain, sometimes by a factor of 10 or more. Second, FF's runtime
+is dominated by its initial grounding step, in which it computes the
+ground instances of all predicates and actions used in the planning
+problem in order to avoid unnecessary instantiation during search.  In
+particular, the ratio of grounding time to total runtime is generally
+above 85\%, and rises to above 99\% at $m=11$, which is still a
+relatively small universe for this application.\footnote{The
+  ``grounding'' time reported here is what FF reports as ``time spent:
+  instantiating action templates''.}
 
@@
 
-\begin{figure}
-  \centering
-
-  \caption{Results for the XTAG experiment. The horizontal axis
-    represents parameters $(k,n)$ from $(1,1)$ to $(3,10)$ in
-    lexicographic order. The vertical axis is runtime in milliseconds.}
+\begin{figure}[t]
+\subfloat[$k=1$]{\includegraphics[width=8cm]{data/xtag-k1.pdf}}
+\subfloat[$k=2$]{\includegraphics[width=8cm]{data/xtag-k2.pdf}}
+  \caption{Results for the XTAG experiment, at $k=1$ and $k=2$.}
   \label{fig:xtag-graph}

### 2009-06-02 trunk/Papers/ci-spark/abstract.tex
changed to official CI style
changes to abstract, section1, and section 2

@@
-%%%
-%%% Abstract
-%%%
 \begin{abstract}
-  Natural language generation (NLG) is a field of computational
-  linguistics that has a long tradition as an application area of
-  planning systems. While this connection had not received much
-  attention for a while, interest in it has been renewed recently in
-  several publications. In this paper, we investigate the extent to
-  which these new NLG systems profit from the recent advances in
-  planner efficiency. Our findings are mixed: While modern planners do
-  very well on solving the search problems that come up in our NLG
-  experiments, their overall runtime is dominated by the grounding
-  step they perform as preprocessing. Because this makes current
-  off-the-shelf planners unusably slow for nontrivial NLG problem
-  instances, we offer our domains as challenges for the planning
-  community.
+  Natural language generation (NLG) is a major subfield of
+  computational linguistics with a long tradition as an application
+  area of automated planning systems. While the connection between
+  NLG and planning has not received much attention in current
+  mainstream NLG research, several recent publications have sparked
+  a renewed interest in this area. In this paper, we investigate the
+  extent to which these new NLG approaches profit from the advances
+  in planner expressiveness and efficiency. Our findings are mixed.
+  While modern planners can readily handle the search problems that
+  arise in our NLG experiments, their overall runtime is often
+  dominated by the grounding step they perform as preprocessing.
+  Furthermore, small changes in the structure of a domain can
+  significantly shift the balance between search and preprocessing.
+  Overall, our experiments show that the off-the-shelf planners we
+  tested are unusably slow for nontrivial NLG problem instances. As
+  a result, we offer our domains and experiences as challenges for
+  the planning community.
 \end{abstract}
 
-\bigskip\noindent
-\textbf{Keywords:} natural language generation, planning
+\begin{keywords}
+  natural language generation, planning
+\end{keywords}
 

### 2009-06-02 trunk/Papers/ci-spark/introduction.tex
changes to abstract, section1, and section 2

@@
 
-\emph{Natural language generation} (NLG; \citealp{reiter00building}) is one of
-the major subfields of natural language processing, concerned with computing
-natural language sentences or texts that convey a given piece of information to
-an audience. While the output of a generation task can take many forms,
-including written text, synthesised speech, or embodied multimodal
-presentations, the underlying NLG problem in each case can be modelled as a
-problem of achieving a (communicative) goal by successively applying a set of
-(communicative) actions. This view of NLG as goal-directed action has clear
-parallels to \emph{automated planning}, which seeks to find general techniques
-for efficiently solving the action sequencing problem.
+\emph{Natural language generation} (NLG; \citealp{reiter00building})
+is one of the major subfields of natural language processing,
+concerned with computing natural language sentences or texts that
+convey a given piece of information to an audience. While the output
+of a generation task can take many forms, including written text,
+synthesised speech, or embodied multimodal presentations, the
+underlying NLG problem in each case can be modelled as a problem of
+achieving a (communicative) goal by successively applying a set of
+(communicative) actions. This view of NLG as goal-directed action has
+clear parallels to \emph{automated planning}, which seeks to find
+general techniques for efficiently solving the action sequencing
+problem.
 
@@
 the usefulness of current planning techniques to NLG by investigating
-some representative generation problems, and evaluating whether
+some representative generation problems, and by evaluating whether
 automated planning has advanced to the point that it can provide
@@
 
-Second, we evaluate the performance of current off-the-shelf planners
-in the planning domains into which these two generation problems
+Second, we evaluate the performance of several off-the-shelf planners
+on the planning domains into which these two generation problems
 translate. Among the planners we test, we explore the efficiency of FF
@@
 problems of the International Planning Competition (IPC).\footnote{See
-  \url{http://ipc.icaps-conference.org/} for information about recent
-  and past editions of the IPC. A good overview of the deterministic
-  track of the 2004 instance of the IPC can be found in
-  \citep{Hoffmann-Edelkamp:05}.}  Using these planners---and an ad-hoc
-Java implementation of GraphPlan \citep{Blum1997} which serves as a
-baseline for certain experiments---we perform a series of tests on a
-range of problem instances in our NLG domains.
+  \url{http://ipc.icaps-conference.org/} for information about past
+  editions of the IPC. Also see \citep{Hoffmann-Edelkamp:05} for a good
+  overview of the deterministic track of the 2004 competition.}
+Using these planners---together with an ad-hoc Java implementation of
+GraphPlan \citep{Blum1997} serving as a baseline for certain
+experiments---we perform a series of tests on a range of problem
+instances in our NLG domains.
 
@@
 task is NP-complete; see \citealt{KolStr02}). On the other hand, these
-same planners spend tremendous amounts of time on \emph{preprocessing}
-to analyse the problem domain in support of the search. On many of our
-problem instances, the preprocessing time overshadows the search time.
-For instance, FF spends 90\% of its runtime in the sentence generation
-domain on grounding out actions and literals. As a consequence, we
-conclude that the off-the-shelf planners we investigated are too slow
-to be useful in real NLG applications.  We hope that our results will
-spark some interest for more efficient implementations of planning
-preprocessors, and offer our NLG domains and experiences as challenges
-for the planning community.
+same planners often spend tremendous amounts of time on
+\emph{preprocessing} to analyse the problem domain in support of the
+search. On many of our problem instances, the preprocessing time
+overshadows the search time. (For instance, FF spends 90\% of its
+runtime in the sentence generation domain on preprocessing.)
+Furthermore, small changes in the structure of a planning domain can
+dramatically shift the balance between preprocessing and search. As a
+consequence, we are forced to conclude that the off-the-shelf planners
+we investigated are generally too slow to be useful in real NLG
+applications. It is also our hope, however, that these results will
+spark an interest to improve the quality of planner
+implementations---especially in the area of preprocessing
+techniques---and to this end we offer our domains and experiences as
+challenges for the planning community.
 
@@
 Section~\ref{sec:domains}, we describe a set of planning problems
-associated with two NLG tasks, namely sentence planning and situated
+associated with two NLG tasks: sentence planning and situated
 instruction generation. In Section~\ref{sec:experiments}, we report on
-the results of our experiments with these planning problems.  In
+our experiments with these planning problems.  In
 Section~\ref{sec:discussion} we discuss our results and overall

### 2009-06-02 trunk/Papers/ci-spark/domains.tex
changes to section 3

@@
 
-We will now focus the discussion on two specific NLG problems:
-sentence generation in the sense of \citet{KolSto07}, and the
-generation of instructions in virtual environments
-\citep{ByrKolStrCasDalMooObe09}. For each, we will introduce the task
-and show by example how the task can be seen as a planning problem.
+We begin by considering two specific NLG problems: sentence generation in
+the sense of \citet{KolSto07}, and the generation of instructions in
+virtual environments \citep{ByrKolStrCasDalMooObe09}. For each problem, we
+introduce the task and show by example how the task can be viewed as a
+planning problem.
 
@@
 
-An example is the tree-adjoining grammar (TAG;
-\citealt{joshi;etal1997}) shown in
+An example of such a lexicalized grammar is the tree-adjoining grammar
+(TAG; \citealt{joshi;etal1997}) shown in
 Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
-\emph{elementary trees} (i.e., the disjoint trees in the figure), each
-of which contributes certain \emph{semantic content}. Let's say that
-we assume a knowledge base containing the individuals $e$, $r_1$ and
-$r_2$ and the facts that $r_1$ and $r_2$ are rabbits, $r_1$ is white
-and $r_2$ is brown, and $e$ is an event in which $r_1$ sleeps. Then we
-would want to construct a sentence expressing the information
-$\{\mathsf{sleep}(e,r_1)\}$ by combining instances of the elementary
-trees (in which the \emph{semantic roles}, such as $\mathsf{self}$ and
-$\mathsf{subj}$, have been substituted by constants from the knowledge
-base) into a TAG derivation as as shown in
-Figure~\ref{fig:white-rabbit-sleeps-deriv}. In the picture, the dashed
-arrow indicates TAG's \emph{substitution} operation, which ``plugs''
-an elementary into the leaf of another tree, and the dotted arrows
-stand for \emph{adjunction}, which splices an elementary tree into an
-internal node. We can then read off the sentence ``The white rabbit
-sleeps'' from the derivation. Note that the sentence ``The rabbit
-sleeps'' would not have been an appropriate result, because ``the
-rabbit'' could refer to either $r_1$ or $r_2$, and thus $r_2$ remains
-as a \emph{distractor}, i.e.\ an incorrect possible interpretation of
-the phrase.
+\emph{elementary trees} (i.e., the disjoint trees in the figure), each of
+which contributes certain \emph{semantic content}. For instance, say that a
+knowledge base contains the individuals $e$, $r_1$ and $r_2$, and the facts
+that $r_1$ and $r_2$ are rabbits, $r_1$ is white and $r_2$ is brown, and
+$e$ is an event in which $r_1$ sleeps. We could then construct a sentence
+expressing the information $\{\mathsf{sleep}(e,r_1)\}$ by combining
+instances of the elementary trees (in which the \emph{semantic roles}, such
+as $\mathsf{self}$ and $\mathsf{subj}$, have been substituted by constants
+from the knowledge base) into a TAG derivation as shown in
+Figure~\ref{fig:white-rabbit-sleeps-deriv}. In the figure, the dashed arrow
+indicates TAG's \emph{substitution} operation, which ``plugs'' an
+elementary into the leaf of another tree; the dotted arrows stands for
+\emph{adjunction}, which splices an elementary tree into an internal node.
+We can then read the sentence ``The white rabbit sleeps'' from the
+derivation. Note that the sentence ``The rabbit sleeps'' would not have
+been an appropriate result, because ``the rabbit'' could refer to either
+$r_1$ or $r_2$. Thus, $r_2$ remains as a \emph{distractor}, i.e., an
+incorrect possible interpretation of the phrase.
 
@@
 
-This perspective on sentence generation has the advantage of solving
-the sentence planning and surface realization problems
-simultaneously. The generation of referring expressions (REs) is
-usually seen as a sentence planning task; in the example, we required
-that the referring expression ``the white rabbit'' can be resolved
-uniquely to $r_1$ by the hearer, in addition to the requirement that
-the derivation is grammatically correct. This is useful in cases where
-sentence planning and surface realization interact, because syntactic
-information about the individual words is available when the REs are
-generated (see e.g. \citealt{stone98textual}). However, the problem of
-whether a given communicative goal can be achieved with a given
-grammar is NP-complete \citep{KolStr02}. A naive search algorithm that
-computes the derivation top-down takes exponential time and is clearly
-infeasible to use in practice. The seminal SPUD system
-\citep{Stone2003a}, which first worked out the idea of integrated
-TAG-based sentence generation, used a greedy algorithm to cope with
-the search, which circumvents the combinatorial explosion but is
-incomplete.
+This perspective on sentence generation also has the advantage of solving
+the sentence planning and surface realization problems simultaneously,
+which is particularly useful in cases where these two problems interact.
+For instance, the generation of referring expressions (REs) is usually seen
+as a sentence planning task, however, syntactic information about
+individual words in available when the REs are generated (see, e.g.,
+\citealt{stone98textual}). (In the example, we require the referring
+expression ``the white rabbit'' to be resolved uniquely to $r_1$ by the
+hearer, in addition to the requirement that the derivation be grammatically
+correct.)
 
@@
 
-As an alternative, \citet{KolSto07} recently proposed to control the
-search by converting the sentence generation problem into a planning
-problem and then running a planner \citet{KolSto07}.\footnote{See
+However, the problem of deciding whether a given communicative goal can be
+achieved with a given grammar is NP-complete \citep{KolStr02}: a naive
+search algorithm that computes a derivation top-down takes exponential time
+and is clearly infeasible to use in practice. In order to circumvent this
+combinatorial explosion, the seminal SPUD system \citep{Stone2003a}, which
+first established the idea of integrated TAG-based sentence generation,
+used a greedy, but incomplete, search algorithm.
+%
+To better control the search, \citet{KolSto07} recently proposed an
+alternative approach which converts the sentence generation problem into a
+planning problem, and solves the transformed search problem using a planner
+\citep{KolSto07}.\footnote{See
   \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
-  which this conversion is implemented.} This planning problem assumes
-an initial state containing an atom $\mathsf{subst}(S,\mathsf{root})$,
-encoding that we must generate a sentence ($S$) starting at the node
-named $\mathsf{root}$ in the TAG derivation tree, and an atom
-$\mathsf{referent}(\mathsf{root},e)$ encoding that the (event)
-individual which the entire sentence describes is $e$. Then it encodes
-the addition of elementary trees to the TAG derivation in the
-individual planning operators; the ones that are needed for our
-example above are shown in
-Fig.~\ref{fig:white-rabbit-as-planning}.
+  which this conversion is implemented.} The resulting planning problem in
+this case assumes an initial state containing an atom
+$\mathsf{subst}(S,\mathsf{root})$, encoding the fact that a sentence ($S$)
+must be generated starting at the node named $\mathsf{root}$ in the TAG
+derivation tree, and a second atom $\mathsf{referent}(\mathsf{root},e)$
+which encodes the fact that the entire sentence describes the (event)
+individual $e$. The elementary trees in the TAG derivation are encoded as
+individual planning operators.
 
-Here the action instance $\addact{sleeps}(\mathsf{root}, e, r_1)$
-replaces the atom $\mathsf{subst}(S,\mathsf{root})$ with the atom
+Figure~\ref{fig:white-rabbit-as-planning} shows the
+transformed planning operators needed to generate the above example
+sentence, ``The white rabbit sleeps.'' Here the action instance
+$\addact{sleeps}(\mathsf{root}, e, r_1)$ replaces the atom
+$\mathsf{subst}(S,\mathsf{root})$ with the atom
 $\mathsf{subst}(NP,\mathsf{subj}(\mathsf{root}))$. In an abuse of PDDL
-syntax, we write $\mathsf{subj}(\mathsf{root})$ as shorthand for a
-fresh individual name.\footnote{These terms, which are not valid in
-  ordinary PDDL, can be eliminated by estimating an upper bound $n$
+syntax, we write $\mathsf{subj}(\mathsf{root})$ as a shorthand for a fresh
+individual name.\footnote{These terms are not valid in
+  ordinary PDDL but can be eliminated by estimating an upper bound $n$
   for the plan length, making $n$ copies of each action, ensuring that
@@
 the semantic information $\mathsf{sleep}(e,r_1)$ has now been
-expressed, and introduces all individuals except for $r_1$ as
+expressed, and introduces all individuals except $r_1$ as
 distractors for the new RE at $\mathsf{subj}(\mathsf{root})$. These
@@
 \mathsf{distractor}(x,y)$, and
-$\mathsf{expressed}(\mathsf{sleep},e,r_1)$. In summary, the following
-is a correct plan:
+$\mathsf{expressed}(\mathsf{sleep},e,r_1)$. For instance, the following
+plan correctly performs the necessary derivation:
 %
@@
 \item $\mathsf{add}\textsf{-}\mathsf{sleeps}(\mathsf{root}, r_1)$,
+\todo{FIX: add-sleeps takes 3 arguments in the PDDL spec!}
 \item $\mathsf{add}\textsf{-}\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}),r_1)$,
@@
 Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
-generated sentence ``the white rabbit sleeps'', can be systematically
+generated sentence ``the white rabbit sleeps,'' can be systematically
 reconstructed from this plan. Thus, we can solve the sentence

### 2009-06-02 trunk/Papers/ci-spark/domains.tex
more changes to section 3

@@
 
-Let us now turn to a second recent application of planning in NLG, the
-GIVE Challenge (``Generating Instructions in Virtual Environments'';
-\citealt{ByrKolStrCasDalMooObe09}). The object of this shared task is
-to build an NLG system which produces natural-language instructions
-which will guide a human user in performing some task in a virtual
-environment.  From an NLG perspective, GIVE makes for an interesting
-challenge because it is a theory-neutral task that exercises all
-components of an NLG system, and emphasizes the study of communication
-in a (simulated) physical environment. Furthermore, because the client
-which displays the 3D environment to the user and the NLG system can
-be physically separated, as long as they are connected over a network,
-such systems can be cheaply evaluated over the Internet. This provides
-a potential solution to the long-standing problem of evaluating NLG
-systems. The first installment of GIVE (GIVE-1) evaluated five NLG
-systems on the performance of 1143 users, making it the largest NLG
-evaluation effort in terms of human users ever.
+In the second application of planning in NLG, we consider the recent GIVE
+Challenge (``Generating Instructions in Virtual Environments'';
+\citealt{ByrKolStrCasDalMooObe09}). The object of this shared task is to
+build an NLG system which produces natural language instructions which
+guide a human user in performing a task in a virtual environment. From an
+NLG perspective, GIVE makes for an interesting challenge since it is a
+theory-neutral task that exercises all components of an NLG system, and
+emphasizes the study of communication in a (simulated) physical
+environment. Furthermore, because the client displaying the 3D environment
+to the user can be physically separated from the NLG system (provided they
+are connected over a network), such systems can be cheaply evaluated over
+the Internet. This provides a potential solution to the long-standing
+problem of evaluating NLG systems. The first instalment of GIVE (GIVE-1)
+evaluated five NLG systems on the performance of 1143 users, making it the
+largest ever NLG evaluation effort to date in terms of human users.
 
@@
 \includegraphics[width=0.75\columnwidth]{give_world_2}
-\caption{Map of a GIVE world.}
+\caption{Map of an example GIVE world.}
   \label{fig:give-development-world}
@@
 
-Planning plays a central role in the GIVE task. To see this, consider
-the example GIVE world map in
-Figure~\ref{fig:give-development-world}. In this example world, the
-user's task is to pick up a trophy in the top left room. The trophy is
-hidden in a safe behind a picture; to access it, the user must push
-certain buttons in order to move the picture out of the way, open the
-safe, and open doors. The user must to navigate the world and perform
-these actions in the 3D client in a certain order; the NLG system must
-instruct the user on how to do this.  To simplify both the planning
-and the NLG task, the world is discretised into a set of tiles of
-equal size. The user can turn by 90 degree steps in either direction,
-and can move from the centre of one tile to the centre of the next
-tile, provided the path between two tiles is not blocked.
-Figure~\ref{fig:give-planning} shows the encoding of some of the
-available GIVE domain actions, in PDDL syntax. In the example, the
-shortest plan to solve the task consists of 108 action steps, with the
-first few steps as follows:
+Planning plays a central role in the GIVE task. For instance, consider the
+example GIVE world shown map in Figure~\ref{fig:give-development-world}. In
+this world, the user's task is to pick up a trophy in the top left room.
+The trophy is hidden in a safe behind a picture; to access it, the user
+must push certain buttons in order to move the picture out of the way, open
+the safe, and open doors. The user must navigate the world and perform
+these actions in the 3D client; the NLG system must instruct the user on
+how to do this. To simplify both the planning and the NLG task, the world
+is discretised into a set of tiles of equal size. The user can turn by 90
+degree steps in either direction, and can move from the centre of one tile
+to the centre of the next tile, provided the path between two tiles is not
+blocked. Figure~\ref{fig:give-planning} shows the encoding of some of the
+available GIVE domain actions in PDDL syntax. In the example, the shortest
+plan to solve the task consists of 108 action steps, with the first few
+steps as follows:
 %
@@
 \item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$,
+\item $\mathsf{manipulate}\textsf{-}\mathsf{alarm}\textsf{-}\mathsf{button}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{b1, pos\_5\_2})$,
 \item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
@@
 \centering
-\begin{minipage}{0.5\textwidth}
+\begin{minipage}{0.9\textwidth}
 {\small%
@@
 (:action move
-   :parameters (?from - position
-                ?to - position
-                ?ori - orientation)
+   :parameters (?from - position  ?to - position  ?ori - orientation)
    :precondition 
-       (and (player-pos ?from) 
-            (adjacent ?from ?to ?ori) 
-            (player-orient ?ori)
-            (not-blocked ?from ?to)
-            (not-alarmed ?to))
+       (and (player-position ?from)  (player-orientation ?ori)
+            (adjacent ?from ?to ?ori)  (not (alarmed ?to)))
    :effect 
-       (and (not (player-pos ?from))
-            (player-pos ?to)))
+       (and (not (player-position ?from))  (player-position ?to)))
 
 (:action turn-left
-   :parameters (?ori - orientation
-                ?newOri - orientation)
+   :parameters (?ori - orientation  ?newOri - orientation)
    :precondition 
-       (and (player-orient ?ori)
-            (next-orient-left ?ori ?newOri))
+       (and (player-orientation ?ori)  (next-orientation ?ori ?newOri))
    :effect 
-       (and (not (player-orient ?ori))
-            (player-orient ?newOri)))
+       (and (not (player-orientation ?ori))  (player-orientation ?newOri)))
 
 (:action turn-right
-   :parameters (?ori - orientation
-                ?newOri - orientation)
+   :parameters (?ori - orientation  ?newOri - orientation)
    :precondition 
-       (and (player-orient ?ori)
-            (next-orient-right ?ori ?newOri))
+       (and (player-orientation ?ori)  (next-orientation ?newOri ?ori))
    :effect 
-       (and (not (player-orient ?ori))
-            (player-orient ?newOri)))
+       (and (not (player-orientation ?ori))  (player-orientation ?newOri)))
 
-(:action manipulate-b1-off-on
-   :parameters (?pos - position)
+(:action manipulate-alarm-button-off-on
+   :parameters (?b - button  ?pos - position  ?alarm - position)
    :precondition 
-       (and (state b1 off)
-            (player-pos ?pos)
-            (position b1 ?pos))
+       (and (state ?b off)  (player-position ?pos)  (position ?b ?pos)
+            (controls-alarm ?b ?alarm))
    :effect
-       (and (not (state b1 off))
-            (state b1 on)
-            (not (state d1 closed))
-            (state d1 open) 
-            (not (blocked pos_6_5 pos_6_4))
-            (not (blocked pos_6_4 pos_6_5))))
+       (and (not (state ?b off))  (not (alarmed ?alarm))  (state ?b on)))
 \end{verbatim}}%
 \end{minipage}
-\caption{Some PDDL actions for the GIVE domain.}
+\caption{Simplified PDDL actions for the GIVE domain.}
 \label{fig:give-planning}
@@
 
-A GIVE NLG system must be able to compute such plans. At a minimum,
-the discourse planner will call a planner in order to determine the
-content of the instructions that should be presented to the
-user. Under this view, the planning problem is very similar to the
-Gridworld problem (see, e.g., \citealt{Tovey-Koenig:2000}), which also
-involves finding a route through a two-dimensional world map with
-discrete positions. As in Gridworld, the domain requires not only
-route finding in the discretised world, but also the execution of
-certain actions (such as finding keys and unlocking doors in
-Gridworld, or pushing the correct buttons to open doors and the safe
-in GIVE). However, the GIVE worlds tend to be bigger than the
-Gridworld instances that were used in the 1990s planning competitions.
+Our description of GIVE as a planning problem makes it very similar to the
+classic Gridworld problem (see, e.g., \citealt{Tovey-Koenig:2000}
+or the 1998 edition of the IPC\footnote{See
+\texttt{ftp://ftp.cs.yale.edu/pub/mcdermott/aipscomp-results.html}.}),
+which also involves route finding through a two-dimensional world map with
+discrete positions. As in Gridworld, the domain also requires the
+execution of certain object-manipulation actions (e.g., finding keys and
+opening locks in Gridworld, or pushing the correct buttons to open doors
+and the safe in GIVE). However, the worlds we consider in GIVE tend to be
+much bigger than the Gridworld instances used in the 1998 planning
+competition, with more complex room shapes and more object types in the
+world.
 
-This relatively loose integration of NLG system and planner is the
-state of the art of the systems that participated in GIVE-1. However,
-it is desirable to integrate the planner and the generation system
-more closely than this. For instance, let's say that the NLG system
-wants to generate the instruction sequence ``walk to the centre of the
-room; turn right; now press the green button in front of
-you''. Experiments with human instruction givers \cite{stoia:08} show
-that this is a pattern that they use frequently: The instruction
-follower is made to walk to a certain point in the world at which the
-instruction giver can then use a a referring expression (``the green
-button'') that is very easy to interpret. Such an NLG system must
-integrate discourse planning and planning in the domain of the world
-map closely. On the one hand, the structure of the discourse is
-determined by the needs of the NLG system rather than the domain plan;
-on the other hand, the discourse planner must be aware of the way in
-which the instruction ``turn right'' is likely to change the
-visibility of objects. Even if the NLG system doesn't implement the
-generation of this discourse as planning, it still solves a problem
-that subsumes the domain planning problem. For these reasons, we
-consider the GIVE domain planning problem as a natural part of a GIVE
-NLG system.  \todo{Make sure that this paragraph still talks about the
-  things that the reviewers found interesting about it.}
+To be successful in GIVE, an NLG system must be able to compute plans of
+the form described above. At a minimum, a discourse planner will call a
+domain planner in order to determine the content of the instructions that
+should be presented to the user. This relatively loose integration of
+NLG system and planner is the state of the art of the systems that
+participated in GIVE-1. However, it is generally desirable to integrate the
+planner and the generation system more closely than this. For instance,
+consider an NLG system that wants to generate the instruction sequence
+``walk to the centre of the room; turn right; now press the green button in
+front of you''. Experiments with human instruction givers \citep{stoia:08}
+show that this is a pattern that they use frequently: the instruction
+follower is made to walk to a certain point in the world where the
+instruction giver can then use a referring expression (``the green
+button'') that is easy for the follower to interpret. An NLG system must
+therefore integrate discourse planning and planning in the domain of the
+world map closely. On the one hand, the structure of the discourse is
+determined by the needs of the NLG system rather than the domain plan; on
+the other hand, the discourse planner must be aware of the way in which the
+instruction ``turn right'' is likely to change the visibility of objects.
+Even if an NLG system doesn't implement the generation of such discourse as
+planning, it must still solve a problem that subsumes the domain planning
+problem. For these reasons, we consider the GIVE domain planning problem as
+a natural part of a GIVE NLG system.
 

### 2009-06-03 trunk/Papers/ci-spark/discussion.tex
changes to section 5

--- 2009-05-30 trunk/Papers/ci-spark/discussion.tex
@@
 
-There are both positive and negative conclusions about the state of
-planning for modern NLG applications that we can draw from these
-experiments.
+We can draw both positive and negative conclusions from our experiments about
+the state of planning for modern NLG applications. On the one hand, we found
+that modern planners are very good at dealing with the \emph{search} problems
+that arise in the NLG-based planning problems we investigated; the search
+time reported by most planners was often a very small fraction of the total
+running time of the planner. In the sentence generation domain, FF's Enforced
+Hill-Climbing strategy finds plans corresponding to 25-word sentences in
+about a second. It is hard to compare this number to a baseline because there
+are no shared benchmark problems, but FF's search outperforms a greedy search
+algorithm, and the performance is competitive with other sentence generators,
+which similarly employ special-purpose algorithms. Thus research on search
+strategies for planning has paid off; in particular, the Enforced
+Hill-Climbing heuristic outperforms the best-first strategy to which FF 2.3
+switches for some problem instances. Similarly, SGPLAN's performance on the
+GIVE domain is very convincing and fast enough for this application.
 
-On the one hand, we found that modern planners are very good at
-dealing with the \emph{search} problems that are involved in the
-NLG-based planning problems we looked at. In the sentence generation
-domain, FF's Enforced Hill-Climbing strategy finds plans corresponding
-to 25-word sentences in about a second. It is hard to compare this
-number to a baseline because there are no shared benchmark problems;
-but FF's search outperforms a greedy search algorithm, and the
-performance is competitive with other sentence generators, which
-similarly employ special-purpose algorithms. Thus research on search
-strategies for planning has paid off: In particular, the Enforced
-Hill-Climbing heuristic outperforms the best-first strategy to which
-FF 2.3 switches for some problem instances. Similarly, SGPLAN's
-performance on the GIVE domain is very convincing and fast enough for
-the application.
+On the other hand, each of the off-the-shelf planners we tested spent
+substantial amounts of time on preprocessing. This is most apparent in the
+sentence generation domain, where the planners spent almost their entire
+runtime on grounding the predicates and operators for some problem instances.
+This effect is much weaker in the GIVE domain, which has a much smaller
+number of operators and less interactions between the predicates in the
+domain. However, our GIVE experiments also illustrate that altering the
+structure of a domain, even minimally, can significantly change a planner's
+performance on a problem. For instance, in some of our GIVE experiments with
+extra grid positions, increasing the number of buttons in the world, while
+keeping the dimensions of the grid fixed, resulted in a significantly larger
+search time while the preprocessing time remained essentially unchanged.
 
-However, each of the off-the-shelf planners we tested spent
-substantial amounts of time on preprocessing. This is most apparent in
-the sentence generation domain, where the planners spent almost their
-entire runtime on grounding the predicates and operators for some
-instances. This effect is much weaker in the GIVE domain, which has a
-much smaller number of operators; but as we added dummy predicates,
-the ratio of time that the planners spent on preprocessing rose in a
-similar way. While the GIVE domain can be carefully defined in such a
-way that the number of operators is minimized, this is not possible
-for a domain in which the operators encode the different communicative
-actions that the NLG system can use. For instance, in the sentence
-generation domain, the XTAG planning problem for $k=2$ and $n=5$
-consists of about 1000 operators for the different lexicon entries for
-all the words in the sentence, some of which take four parameters. It
-is not unrealistic to assume a knowledge base with a few hundred
-individuals. All this adds up to trillions of ground instances, i.e.\
-it is completely infeasible to compute ground instances naively.
+While the GIVE domain can be defined in such a way that the number of
+operators is minimized, this is not possible for an encoding of a domain in
+which the operators model the different communicative actions that the NLG
+system can use. For instance, in the sentence generation domain, the XTAG
+planning problem for $k=2$ and $n=5$ consists of about 1000 operators for the
+different lexicon entries for all the words in the sentence, some of which
+take four parameters. It is not unrealistic to assume a knowledge base with a
+few hundred individuals. All this adds up to trillions of ground
+instances: a set which is completely infeasible to compute naively.
 
-This means that while current planners deal well with the search
-problems we have looked at, they are still unusable for practical NLG
-applications because of the time they spend on preprocessing. In other
-words, generation-as-planning research is still not in a much better
-position than it was in the 1980s. While we are aware that the time
-that a planner invests in preprocessing can pay off during search, we
-still suggest that the inability of current planners to scale to
-larger domains still limits their usefulness for applications beyond
-NLG as well. Furthermore, we feel that preprocessing gets less
-research attention than it deserves: If it is scientifically trivial,
-we challenge the planning community to come up with implementations
-that only ground operators by need; otherwise, we look forward to
-publications on this topic. To support this, we offer our planning
-domains as benchmarks for future competitions. They are available at
-\todo{upload them somewhere}
+Of course, it would be premature to judge the usefulness of current planners
+as a whole, based on just two NLG domains. Nevertheless, we believe that the
+structure of our planning problems, which are dominated by large numbers of
+operators and individuals, is typical of NLG-related planning problems as a
+whole. This strongly suggests that while current planners are able to manage
+many of the search problems in the domains we looked at, they are still
+unusable for practical NLG applications because of the time they spend on
+preprocessing. In other words, the state of generation-as-planning research
+is still not in a much better position than it was in the 1980s.
 
-Finally, we found it very convenient that the recent IPC planning
-competitions provide an entry point for selecting and obtaining
-current planners. Nevertheless, our experiments exposed several bugs
-in the planners we tested, and required us to make changes to the
-source code to make them scale to our inputs. We also found that
-different planners differ in the variants of PDDL that they can
-process. These differences range from fragments of ADL that can be
-parsed to sensitivity to the order of declarations and the use of
-``objects'' rather than ``individuals'' as the keyword for declaring
-the universe. We propose that the case for planning as a mature
-technology with professional-quality implementations could be made
-more strongly if such discrepancies were harmonized.
+We are also aware that the time a planner invests in preprocessing can pay
+off during search, and that such techniques have been invaluable in improving
+the overall running time of modern planners. However, we still suggest that
+the inability of current planners to scale to larger domains limits their
+usefulness for applications beyond NLG as well. Furthermore, we feel that the
+problem of preprocessing receives less research attention than it deserves:
+if the problem is scientifically trivial then we challenge the planning
+community to develop more efficient implementations that only ground
+operators by need; otherwise, we look forward to future publications on this
+topic. To support this effort, we offer our planning domains as benchmarks
+for future research and competitions.\footnote{The PDDL problem generators
+for our NLG domains are available at \texttt{http://\todo{TODO}}.}
+
+Finally, we found it very convenient that the recent International Planning
+Competitions provide a useful entry point for selecting and obtaining current
+planners. Nevertheless, our experiments exposed several bugs in the planners
+we tested, which required us to change their source code to make them scale
+to our inputs. We also found that different planners that solve the same
+class of planning problems (e.g., STRIPS, ADL, etc.) sometimes differ in the
+variants of PDDL that they support. These differences range from fragments of
+ADL that can be parsed to sensitivity, to the order of declarations and the
+use of ``objects'' rather than ``individuals'' as the keyword for declaring
+the universe. We propose that the case for planning as a mature technology
+with professional-quality implementations could be made more strongly if such
+discrepancies were harmonized.
 

### 2009-06-03 trunk/Papers/ci-spark/domains.tex
changes to section 5

@@
 
-We begin by considering two specific NLG problems: sentence generation in
-the sense of \citet{KolSto07}, and the generation of instructions in
-virtual environments \citep{ByrKolStrCasDalMooObe09}. For each problem, we
-introduce the task and show by example how the task can be viewed as a
-planning problem.
+We begin by considering two specific NLG problems: sentence generation in the
+sense of \citet{KolSto07}, and the generation of instructions in virtual
+environments \citep{ByrKolStrCasDalMooObe09}. In each case, we introduce the
+task and show by example how it can be viewed as a planning problem.
 
@@
 \citep{KolSto07}.\footnote{See
-  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
-  which this conversion is implemented.} The resulting planning problem in
+  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, which
+  implements this conversion.} The resulting planning problem in
 this case assumes an initial state containing an atom
@@
 \item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{alarm}\textsf{-}\mathsf{button}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{b1, pos\_5\_2})$,
+\item $\mathsf{manipulate}\textsf{-}\mathsf{button}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{b1, pos\_5\_2})$,
 \item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
@@
 
-(:action manipulate-alarm-button-off-on
+(:action manipulate-button-off-on
    :parameters (?b - button  ?pos - position  ?alarm - position)

### 2009-06-03 trunk/Papers/ci-spark/experiments.tex
revising desc of my experiments

@@
 They spend a long time on computing all ground instances, even if most
-of these instances are never used in the plan search.
+of these instances are never used in the plan search. \todo{Ron, this
+  is something you need to adapt to the new story -- AK}
 
@@
 
-For the first experiment on sentence generation, we focus on the
-generation of referring expressions, a problem that is usually handled
-by the sentence planner if sentence planning and surface realization
-are separated. We consider a series of sentence generation problems
-which require the planner to compute a plan representing the sentence
-``Mary likes the Adj$_1$ \ldots Adj$_n$ rabbit.''  Each problem
-instance assumes a target referent $r$, which is a rabbit, and a
-certain number $m$ of further rabbits $r_1,\ldots,r_m$ that are
-distinguished by properties $P_1,\ldots,P_n$ with $n \leq m$. The
-problem instance is set up such that $r$ has all properties except for
-$P_i$ in common with each $r_i$. That is, all $n$ properties are
-required to describe $r$ uniquely. The $n$ properties are realized as
-$n$ different adjectives, in any order.  This setup allows us to vary
-the plan length (a plan with $n$ properties will have length $n+4$)
-and the universe size (the universe will contain $m+1$ rabbit
-individuals in addition to the differently-typed individuals used to
-encode the grammar).
+For the first experiment on sentence generation, we exercise the
+ability of the CRISP system described in
+Section~\ref{sec:domain-crisp} to generate referring expressions. This
+problem is usually handled by the sentence planner if sentence
+planning and surface realization are separated; here it happens as
+part of the overall generation process.
+
+We consider a series of sentence generation problems which require the
+planner to compute a plan representing the sentence ``Mary likes the
+Adj$_1$ \ldots Adj$_n$ rabbit.''  Each problem instance assumes a
+target referent $r$, which is a rabbit, and a certain number $m$ of
+further rabbits $r_1,\ldots,r_m$ that are distinguished by properties
+$P_1,\ldots,P_n$ with $n \leq m$. The problem instance is set up such
+that $r$ has all properties except for $P_i$ in common with each $r_i$
+for $1 \leq i \leq n$, and $r_{n+1},\ldots,r_m$ have none of the
+properties $P_i$. That is, all $n$ properties are required to describe
+$r$ uniquely. The $n$ properties are realized as $n$ different
+adjectives, in any order.  This setup allows us to vary the plan
+length (a plan with $n$ properties will have length $n+4$) and the
+universe size (the universe will contain $m+1$ rabbit individuals in
+addition to the individuals used to encode the grammar, which have
+different types).
 
@@
 instances as described in Section~\ref{sec:domains}, and then ran
-three different planners on them. We used the off-the-shelf planners
-FF 2.3 \citep{HoffmannNebel01} and SGPLAN 5.2.2
-\citep{hsu06:_new_featur_in_sgplan_for}, because they were highly
-successful in the recent International Planning Competitions, and
-support a fragment of PDDL with quantified and conditional effects,
-which is necessary in our domain (see
-Fig.~\ref{fig:white-rabbit-as-planning}). Common to both of these
-planners is that they perform a preprocessing step in which all ground
-instances of all predicates and actions are computed.  Furthermore, we
+several different planners on them. We used three off-the-shelf
+planners: FF 2.3 \citep{HoffmannNebel01}, Metric-FF
+\citep{hoffmann:ecai-02}, and SGPLAN 6
+\citep{hsu06:_new_featur_in_sgplan_for}; all of these were highly
+successful at the recent IPC competitions, and unlike many other IPC
+participants, support a fragment of PDDL with quantified and
+conditional effects, which is necessary in our domain. In addition, we
 used an ad-hoc implementation of GraphPlan \citep{Blum1997} written in
-Java which avoids the grounding preprocessing step and only computes
-instances of actions and predicates by need, as they are actually
-discovered when the planning graph is built. Our expectation was that
-this planner would not be able to compete with FF or SGPLAN in terms
-of search efficiency, but we were interested in comparing the
-contribution of the prior grounding step on the planner performance.
+Java; unlike the three off-the-shelf planners, it only computes
+instances of literals and operators as they are needed in the course
+of the plan search, instead of computing all ground instances in a
+separate preprocessing step. Finally, we reimplemented the incomplete
+greedy search algorithm used in the SPUD system \citep{Stone2003a}.
 
@@
   Sections~\ref{sec:exper-1:-sent} and \ref{sec:exper-2:-sent-xtag}
-  were measured on an AMD Opteron 8220 CPU running at 2.8 GHz, under
-  Linux. FF 2.3 and Metric-FF were recompiled as 64-bit binaries. Java
-  programs were executed under Java 1.6.0\_13 in 64-bit mode. They
-  were allowed to ``warm up'', i.e. the JVM was given the opportunity
-  to just-in-time compile the relevant bytecode by running the planner
-  three times and discarding the runtimes before taking the actual
-  measurements. All runtimes are averaged over three runs of
-  the planners.} The input parameters $(m,n)$ are plotted in
+  were measured on a single core of an AMD Opteron 8220 CPU running at
+  2.8 GHz, under Linux. FF 2.3 and Metric-FF were recompiled as 64-bit
+  binaries. Java programs were executed under Java 1.6.0\_13 in 64-bit
+  mode. They were allowed to ``warm up'', i.e. the JVM was given the
+  opportunity to just-in-time compile the relevant bytecode by running
+  the planner three times and discarding the runtimes before taking
+  the actual measurements. All runtimes are averaged over three runs
+  of the planners.} The input parameters $(m,n)$ are plotted in
 lexicographic order on the horizontal axis, and the runtime is shown
-in seconds on the vertical axis. These results reveal a number of
-interesting insights. First, FF significantly outperforms SGPLAN in
-this domain, sometimes by a factor of 10 or more. Second, FF's runtime
-is dominated by its initial grounding step, in which it computes the
-ground instances of all predicates and actions used in the planning
-problem in order to avoid unnecessary instantiation during search.  In
-particular, the ratio of grounding time to total runtime is generally
-above 85\%, and rises to above 99\% at $m=11$, which is still a
-relatively small universe for this application.\footnote{The
-  ``grounding'' time reported here is what FF reports as ``time spent:
-  instantiating action templates''.}
+in seconds on the vertical axis, on a logarithmic scale. These results
+reveal a number of interesting insights. First, the search times of FF
+and Metric-FF (shown as thinner lines) significantly outperform
+SGPLAN's search in this domain---on the largest instances, by a factor
+of over 100.\footnote{For FF and Metric-FF, we report the
+  ``searching'' and ``total'' times reported by the planners. For
+  SGPLAN, we report the ``total'' time and the difference between the
+  ``total'' and ``parsing'' times.} Second, FF and Metric-FF perform
+very similarly to each other, and their search times are almost the
+same as those of the SPUD algorithm, which is impressive because they
+are complete search algorithms, whereas SPUD's greedy algorithm is
+not.
 
-In our testing examples, the time spent by FF on grounding is so high
-that it is consistently outperformed by our Java implementation of
-GraphPlan, which only computes instances by need. Although FF is
-consistently much faster as far as pure search time is concerned, our
-results indicate that FF's performance is much more sensitive to the
-domain size: if we fix $n=1$, FF takes 60 ms to compute a plan at
-$m=1$, but 2.4 seconds to compute the same plan at $m=10$. By
-comparison, our GraphPlan implementation takes 30 ms at $m=1$ and
-still only requires 50 ms at $m=10$. Conversely, GraphPlan's runtime
-grows much faster with the plan size (i.e., with growing values of $n$
-for a fixed $m$). Larger, but still realistically-sized, instances of
-the sentence generation problem are still problematic for all the
-planners we tested.
+Finally, it is striking that for all three off-the-shelf planners, the
+search only accounts for a tiny fraction of the total runtime; in each
+case, the preprocessing times are higher than the search times by one
+or two orders of magnitude. As a consequence, even our relatively
+naive Java implementation of GraphPlan outperforms them all in terms
+of total runtime, because it only computes instances by need. Although
+FF is consistently much faster as far as pure search time is
+concerned, our results indicate that FF's performance is much more
+sensitive to the domain size: if we fix $n=1$, FF takes 27
+milliseconds to compute a plan at $m=1$, but 4.4 seconds to compute
+the same plan at $m=10$. By comparison, our GraphPlan implementation
+takes 20 ms at $m=1$ and still only requires 22 ms at
+$m=10$.
 

### 2009-06-03 trunk/Papers/ci-spark/conclusion.tex
more changes to section 3
changes to the conclusion

@@
 
-In this paper, we have investigated the usefulness of current planning
-technology to natural language generation, an application area with a
-long tradition that has recently experienced renewed interest from NLG
-researchers. In particular, we have evaluated the performance of
-several off-the-shelf planners on a series of planning domains that
-arise in the context of sentence generation and situated instruction
-generation.
+In this paper, we investigated the usefulness of current planning technology
+to natural language generation, an application area with a long tradition of
+using automated planning that has recently experienced renewed interest from
+NLG researchers. In particular, we evaluated the performance of several
+off-the-shelf planners on a series of planning domains that arose in the
+context of sentence generation and situated instruction generation.
 
 Our results were mixed. While some of the planners we tested---in
-particular, FF and SGPLAN, in different domains---did an impressive
-job of controlling the complexity of the search, we also found that
-all planners we tested spent too much time on preprocessing to be
-useful. In the sentence generation domain, FF spent 90\% of its
-runtime on computing the ground instances of the planning operators;
-in the instruction-giving domain, which is very similar to Gridworld,
-a similar effect happened once we added more predicates. As things
-stand, we found that this overly long preprocessing time makes current
-planners inapplicable to NLG application for any but the smallest
-instances. Users who come to planning from the outside, such as NLG
-researchers, take planners as a black box. This means that search
-efficiency alone is not helpful when other modules of the planner are
-slow. From this perspective, we propose that the planning community
-might spend some attention on optimizing the preprocessing module with
-similar vigor as the search itself. In particular, we propose to
-investigate planning algorithms that do not have to rely on grounding
-out all operators before starting the search.
+particular, FF and SGPLAN---did an impressive job of controlling the
+complexity of the search, we also found that all the planners we tested
+spent too much time on preprocessing to be useful. For instance, in the
+sentence generation domain, FF spent 90\% of its runtime on computing the
+ground instances of the planning operators; in the instruction-giving
+domain, which is very similar to Gridworld, a similar effect happened for
+certain combinations of grid sizes and buttons. As things stand, we found
+that this overly long preprocessing time makes current planners an
+inappropriate choice for NLG applications, in any but the smallest problem
+instances. Users who come to planning from outside the field, such as NLG
+researchers, treat planners as black boxes. This means that search
+efficiency alone is not helpful when other modules of the planner are slow.
+From this perspective, we propose that the planning community should spend
+some attention on optimising the preprocessing component of the problem with
+similar vigour as the search itself. In particular, we propose that one line
+of research might be to investigate planning algorithms that do not rely on
+grounding out all operators prior to the search, but instead selectively
+perform this operation when needed.
 
@@
 NLG-as-planning research presents valuable opportunities for both
-sides. Clearly, NLG researchers who apply planning technology will
-benefit directly from any improvements in planner effiency. But
-conversely, NLG may be a worthwhile application area for planning
-researchers to keep in mind. Domains like GIVE highlight certain
-challenges, such as plan execution monitoring and plan presentation
-(i.e., summarisation and elaboration), but also offer a platform on
-which such technologies can be evaluated in experiments with human
-users. Further, although we have focused on classical planning
-problems in this work, research related to reasoning under
-uncertainty, resource management, and planning with knowledge and
-sensing, can also be investigated in these settings. As such, we
-believe our domains would provide interesting challenges for planners
+disciplines. Clearly, NLG researchers who apply planning technology will
+benefit directly from any improvements in planner effiency. Conversely, NLG
+may also be a worthwhile application area for planning researchers to keep
+in mind. Domains like GIVE highlight certain challenges, such as plan
+execution monitoring and plan presentation (i.e., summarisation and
+elaboration), but also offer a platform on which such technologies can be
+evaluated in experiments with human users. Furthermore, although we have
+focused on classical planning problems in this work, research related to
+reasoning under uncertainty, resource management, and planning with
+knowledge and sensing, can also be investigated in these settings. As such,
+we believe our domains would provide interesting challenges for planners
 entered in future editions of the IPC.
@@
 Group at the University of Edinburgh. The authors would like to thank
-all members of this group, especially Hector Geffner and Mark
+all members of this group, especially H{\'{e}}ctor Geffner and Mark
 Steedman, for interesting discussions. We also thank our reviewers for

### 2009-06-03 trunk/Papers/ci-spark/experiments.tex
done with my experiments

@@
 separate preprocessing step. Finally, we reimplemented the incomplete
-greedy search algorithm used in the SPUD system \citep{Stone2003a}.
+greedy search algorithm used in the SPUD system \citep{Stone2003a} in
+Java.
 
@@
   2.8 GHz, under Linux. FF 2.3 and Metric-FF were recompiled as 64-bit
-  binaries. Java programs were executed under Java 1.6.0\_13 in 64-bit
-  mode. They were allowed to ``warm up'', i.e. the JVM was given the
-  opportunity to just-in-time compile the relevant bytecode by running
-  the planner three times and discarding the runtimes before taking
-  the actual measurements. All runtimes are averaged over three runs
-  of the planners.} The input parameters $(m,n)$ are plotted in
-lexicographic order on the horizontal axis, and the runtime is shown
-in seconds on the vertical axis, on a logarithmic scale. These results
-reveal a number of interesting insights. First, the search times of FF
-and Metric-FF (shown as thinner lines) significantly outperform
-SGPLAN's search in this domain---on the largest instances, by a factor
-of over 100.\footnote{For FF and Metric-FF, we report the
-  ``searching'' and ``total'' times reported by the planners. For
-  SGPLAN, we report the ``total'' time and the difference between the
-  ``total'' and ``parsing'' times.} Second, FF and Metric-FF perform
-very similarly to each other, and their search times are almost the
-same as those of the SPUD algorithm, which is impressive because they
-are complete search algorithms, whereas SPUD's greedy algorithm is
-not.
+  binaries and run with a memory limit of 32 GB. Java programs were
+  executed under Java 1.6.0\_13 in 64-bit mode. They were allowed to
+  ``warm up'', i.e. the JVM was given the opportunity to just-in-time
+  compile the relevant bytecode by running the planner three times and
+  discarding the runtimes before taking the actual measurements. All
+  runtimes are averaged over three runs of the planners.} The input
+parameters $(m,n)$ are plotted in lexicographic order on the
+horizontal axis, and the runtime is shown in seconds on the vertical
+axis, on a logarithmic scale. These results reveal a number of
+interesting insights. First, the search times of FF and Metric-FF
+(shown as thinner lines) significantly outperform SGPLAN's search in
+this domain---on the largest instances, by a factor of over
+100.\footnote{For FF and Metric-FF, we report the ``searching'' and
+  ``total'' times reported by the planners. For SGPLAN, we report the
+  ``total'' time and the difference between the ``total'' and
+  ``parsing'' times.} Second, FF and Metric-FF perform very similarly
+to each other, and their search times are almost the same as those of
+the SPUD algorithm, which is impressive because they are complete
+search algorithms, whereas SPUD's greedy algorithm is not.
 
@@
 appropriateness of planning for the sentence generation domain: On the
-examples we looked at, search times were low, but FF and SGPLAN spend
-a lot of time on the initial grounding step.  However, one weakness of
-this experiment is that it uses a tiny grammar, consisting of just
-those 12 lexicon entries that are needed for the experiment. While the
-grounding problem can only get worse with larger grammars, the
-experiment by itself does not allow us to make clear statements about
-the search efficiency.
-
+examples we looked at, the search times were quite acceptable, but FF
+and SGPLAN spent a lot of time on the initial grounding step.
+However, one weakness of this experiment is that it uses a tiny
+grammar, consisting of just those 12 lexicon entries that are needed
+for the experiment. While the grounding problem can only get worse
+with larger grammars, the experiment by itself does not allow us to
+make clear statements about the search efficiency.
 To address this problem, we ran a second sentence generation
@@
 semantic representations based on the words in the lexicalized
-elementary tree. The result is a highly ambiguous grammar: The most
+elementary trees. The result is a highly ambiguous grammar: The most
 ambiguous word, ``ask'', is the anchor of 314 lexicon entries.
@@
 Fig.~\ref{fig:xtag-graph} reports the runtimes we measured in this
-experiment for FF. Because the computations in this experiment were
-very memory-intensive, we recompiled FF 2.3 as a 64-bit binary. We
-also ran the experiments with a 64-bit version of Metric-FF; although
-our planning problem is purely sequential and not metric, Metric-FF
-avoids a bug which caused FF 2.3 to switch to a best-first search
-strategy beyond a certain point; the jump in runtime is clearly
-visible in the figure. We did not run this experiment with SGPLAN
-because we didn not have access to a 64-bit version of either SGPLAN
-5.2.2 or SGPLAN 6; some initial experiences with the 32-bit version
-indicate that SGPLAN switches to FF's best-first strategy even for
-small problem instances, and thus would be slower than the versions of
-FF we reported here. We also do not report runtimes for our Java
-implementation of Graphplan, because it was unusuably slow for serious
-problem instances: For $k=1$ and $n=3$, it already took over two
-minutes, and it exceeded its memory limit of 16 GB for
-$n>3$. \todo{compare to SPUD reconstruction here}
+experiment for FF, Metric FF, and the SPUD reimplementation. We do not
+report runtimes for SGPLAN, because we could not recompile SGPLAN as a
+64-bit binary, and the 32-bit version ran out of memory very
+quickly. We also do not report runtimes for our Java implementation of
+Graphplan, because it was unusuably slow for serious problem
+instances: For $k=1$ and $n=3$, it already took over two minutes, and
+it exceeded its memory limit of 16 GB for $n>3$. This may be a
+limitation of our naive implementation rather than the GraphPlan
+algorithm itself.
 
-There are a number of observations we can make in this
+Nonetheless, there are a number of observations we can make in this
 experiment. First of all, the experiment confirms that FF's Enforced
@@
 FF generates a 39-word sentence ($k=1,n=6$) in under a second of
-search time. This efficiency is a direct result of this particular
-search strategy: Whenever FF falls back to the best-first strategy,
-runtimes become much slower (compare the graphs for FF 2.3 and
-Metric-FF). Second, FF's runtime is still dominated by the
-preprocessing stage. For instance, Metric-FF spends about 10 seconds
-on search for $k=1,n=10$, compared to its total runtime of about 65
-seconds. This effect becomes more pronounced as we increase $k$: For
-$k=2$, we reach 65 seconds of total runtime at $n=4$, but here
-Metric-FF only spends about a second on search. For $k=3$, neither FF
-nor Metric-FF were able to solve any of the input
-instances. \todo{bla} This is to be expected, given that the planning
+search time. This level of efficiency is a direct result of using this
+particular search strategy: For $k=1$ and $n>6$, FF 2.3 (but not
+Metric-FF) fell back to the best-first search strategy, which causes a
+dramatic loss of search effiency. It is also encouraging that
+Metric-FF still performs comparably to SPUD in terms of pure search
+time. We believe that by FF's technique of evaluating actions by
+estimating the distance to a goal state for the relaxed problem
+essentially picks out the same evaluation function as SPUD's domain-specific
+heuristic, and the enforced hill-climbing strategy needs to backtrack
+very little in this domain and thus performs similarly to SPUD's
+greedy search. However, SPUD's incompleteness manifests itself in this
+experiment by its inability to find any plan for $k>1$ and $n>1$,
+whereas FF and its variants still (correctly) find these plans.
+
+Second, FF's runtime is still dominated by the preprocessing
+stage. For instance, Metric-FF spends about 10 seconds on search for
+$k=1,n=10$, compared to its total runtime of about 65 seconds. This
+effect becomes more pronounced as we increase $k$: For $k=2$, we reach
+65 seconds of total runtime at $n=4$, but here Metric-FF only spends
+about a second on search. For $k=3$, neither FF nor Metric-FF were
+able to solve any of the input instances within their memory
+limit. This is consistent with the observation that the planning
 operators for the verbs have $k+2$ parameters (see

### 2009-06-03 trunk/Papers/ci-spark/domains.tex
changes to section 4

@@
   $\mathsf{subj}(u)$ in an action copy by the constant
-  $\mathsf{subj}_i$. The terms $S$, $NP$, and $N$ in the planning
-  problem are constants.}  At the same time, the operator records that
+  $\mathsf{subj}_i$. The terms $S$, $NP$, and $N$ are constants.}
+At the same time, the operator records that
 the semantic information $\mathsf{sleep}(e,r_1)$ has now been

### 2009-06-03 trunk/Papers/ci-spark/experiments.tex
changes to section 4

@@
 
-We will now return to the original question of the paper: Is planning
-technology ready for applications in natural language generation? We
-will tackle this question by evaluating the performance of several
-planners on the NLG planning domains from the previous section.
-
-We start with two experiments in the CRISP domain. First, we present a
-scenario which focuses on the generation of referring expressions with
-a tiny grammar (Section~\ref{sec:exper-1:-sent}); then we look at a
-setting in which CRISP is used for surface realization with the XTAG
-Grammar \citep{xtag01:_tr}, a large-scale TAG grammar of English
-(Section~\ref{sec:exper-2:-sent-xtag}).
-
-We then go through a series of experiments in the GIVE domain. We
-start with a domain that is essentially Gridworld
-(Section~\ref{sec:exper-2:-minim}), and then add positions that are
-either not needed (Section~\ref{sec:experiment-3:-give}) or even
-inaccessible (Section~\ref{sec:experiment-4:-give}). This series of
-experiments is set up to illustrate what we perceive as the main
-limitation of current off-the-shelf planners for our applications:
-They spend a long time on computing all ground instances, even if most
-of these instances are never used in the plan search. \todo{Ron, this
-  is something you need to adapt to the new story -- AK}
+We now return to the original question of the paper: is planning technology
+ready for realistic applications in natural language generation? To investigate
+this question we consider two sets of experiments, designed to evaluate the
+performance of several planners on the NLG planning domains from the previous
+section. Starting with the CRISP domain, we first present a scenario which
+focuses on the generation of referring expressions with a tiny grammar
+(Section~\ref{sec:exper-1:-sent}). We then look at a setting in which CRISP is
+used for surface realization with the XTAG Grammar \citep{xtag01:_tr}, a
+large-scale TAG grammar for English (Section~\ref{sec:exper-2:-sent-xtag}). In
+the second set of experiments we investigate the GIVE domain. We begin with a
+domain that is similar to the classic Gridworld
+(Section~\ref{sec:exper-2:-minim}), and then add extra grid cells to the world
+that are neither needed (Section~\ref{sec:experiment-3:-give}) nor even
+accessible (Section~\ref{sec:experiment-4:-give}). We also investigate the role
+that goal ordering plays in some of these problems. These experiments are
+configured in a way that lets us explore the scalability of a planner's search
+and preprocessing capabilities, and illustrates what we perceive to be one of
+the main limitations of current off-the-shelf planners for our applications:
+they often spend a long time computing ground instances, even when most of
+these instances are not required during plan search.
 
@@
   binaries and run with a memory limit of 32 GB. Java programs were
-  executed under Java 1.6.0\_13 in 64-bit mode. They were allowed to
-  ``warm up'', i.e. the JVM was given the opportunity to just-in-time
+  executed under Java 1.6.0\_13 in 64-bit mode and were allowed to
+  ``warm up'', i.e., the JVM was given the opportunity to just-in-time
   compile the relevant bytecode by running the planner three times and
@@
 \begin{figure}[t]
-\subfloat[$k=1$]{\includegraphics[width=8cm]{data/xtag-k1.pdf}}
-\subfloat[$k=2$]{\includegraphics[width=8cm]{data/xtag-k2.pdf}}
+\subfloat[$k=1$]{\includegraphics[width=7cm]{data/xtag-k1.pdf}}
+\subfloat[$k=2$]{\includegraphics[width=7cm]{data/xtag-k2.pdf}}
   \caption{Results for the XTAG experiment, at $k=1$ and $k=2$.}

### 2009-06-03 trunk/Papers/ci-spark/experiments.tex
changes to section 4

@@
 problems arising in the GIVE domain. We construct a series of test worlds,
-similar to the one illustrated in Figure~\ref{fig:give-minimal}. These
+similar to the one illustrated in Figure~\ref{fig:give-maps}(a). These
 worlds consist of a $2n$ by $h$ grid of positions, such that there are
@@
   \centering
-  \includegraphics[width=0.5\columnwidth]{pic-buttons}
-  \caption{Minimal GIVE world.}
-  \label{fig:give-minimal}
+  \subfloat[position=bottom][(a) Minimal world]{%
+      \includegraphics[height=2.5cm,clip=true,trim=0 -10px 0 -3px]{pic-buttons}}
+  \qquad
+  \subfloat[position=bottom][(b) Extra grid cells with accessible positions]{%
+      \includegraphics[height=2.5cm]{pic-empty-buttons}} \\
+  \subfloat[(c) Extra grid cells with inaccessible positions]{%
+      \includegraphics[height=2.5cm]{pic-empty-inaccessible}}
+  \caption{Experimental GIVE world configuration.}
+  \label{fig:give-maps}
 \end{figure}
@@
 extend the world map by adding another $w$ by $h$ empty ``junk'' positions
-to the right of the minimal world, as shown in Figure~\ref{fig:give-junk}.
+to the right of the minimal world, as shown in Figure~\ref{fig:give-maps}(b).
 These new positions are not actually required in any plan, but extend the
@@
 then convert it into a planning problem in PDDL.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.80\columnwidth]{pic-empty-buttons}
-  \caption{GIVE world with extra ``junk'' positions.}
-  \label{fig:give-junk}
-\end{figure}
 
@@
 inaccessible from the first grid, as shown in
-Figure~\ref{fig:give-junk-nosoln}. Since the geometry of the grid makes it
+Figure~\ref{fig:give-maps}(c). Since the geometry of the grid makes it
 impossible to construct a plan for pressing all the buttons in the world,
@@
 
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.85\columnwidth]{pic-empty-inaccessible}
-  \caption{GIVE world with inaccessible positions.}
-  \label{fig:give-junk-nosoln}
-\end{figure}
-
 \begin{figure}

### 2009-06-03 trunk/Papers/ci-spark/discussion.tex
some details

@@
 
-We can draw both positive and negative conclusions from our experiments about
-the state of planning for modern NLG applications. On the one hand, we found
-that modern planners are very good at dealing with the \emph{search} problems
-that arise in the NLG-based planning problems we investigated; the search
-time reported by most planners was often a very small fraction of the total
-running time of the planner. In the sentence generation domain, FF's Enforced
-Hill-Climbing strategy finds plans corresponding to 25-word sentences in
-about a second. It is hard to compare this number to a baseline because there
-are no shared benchmark problems, but FF's search outperforms a greedy search
-algorithm, and the performance is competitive with other sentence generators,
-which similarly employ special-purpose algorithms. Thus research on search
-strategies for planning has paid off; in particular, the Enforced
-Hill-Climbing heuristic outperforms the best-first strategy to which FF 2.3
-switches for some problem instances. Similarly, SGPLAN's performance on the
-GIVE domain is very convincing and fast enough for this application.
+We can draw both positive and negative conclusions from our
+experiments about the state of planning for modern NLG
+applications. On the one hand, we found that modern planners are very
+good at dealing with the \emph{search} problems that arise in the
+NLG-based planning problems we investigated.
+% I don't think we want to talk about this yet - AK
+%; the search time reported
+%by most planners was often a very small fraction of the total running
+%time of the planner. 
+In the sentence generation domain, FF's Enforced
+Hill-Climbing strategy finds plans corresponding to 25-word sentences
+in about a second. It is hard to compare this number to a baseline
+because there are no shared benchmark problems, but FF's search
+performance is similar to that of a greedy, incomplete special-purpose
+algorithm, and competitive to other sentence generators as well. Thus
+research on search strategies for planning has paid off; in
+particular, the Enforced Hill-Climbing heuristic outperforms the
+best-first strategy to which FF 2.3 switches for some problem
+instances. Similarly, SGPLAN's performance on the GIVE domain is very
+convincing and fast enough for this application.
 
@@
 for future research and competitions.\footnote{The PDDL problem generators
-for our NLG domains are available at \texttt{http://\todo{TODO}}.}
+for our NLG domains are available at
+\url{http://www.coli.uni-saarland.de/~koller/projects/crisp}.}
 
@@
 variants of PDDL that they support. These differences range from fragments of
-ADL that can be parsed to sensitivity, to the order of declarations and the
+ADL that can be parsed, to sensitivity to the order of declarations and the
 use of ``objects'' rather than ``individuals'' as the keyword for declaring

### 2009-06-03 trunk/Papers/ci-spark/experiments.tex
changes to data and section 4

@@
 
+\begin{figure}[t]
+  \centering
+  \subfloat[position=bottom][(a) Minimal world]{%
+      \includegraphics[height=2.5cm,clip=true,trim=0 -10px 0 -3px]{pic-buttons}}
+  \qquad
+  \subfloat[position=bottom][(b) World with extra grid cells]{%
+      \includegraphics[height=2.5cm]{pic-empty-buttons}} \\
+  %\subfloat[(c) Extra grid cells with inaccessible positions]{%
+  %    \includegraphics[height=2.5cm]{pic-empty-inaccessible}}
+  \caption{Experimental GIVE world configurations.}
+  \label{fig:give-maps}
+\end{figure}
+
 \subsection{Experiment 2: Minimal GIVE worlds}
@@
 of the actions available in the GIVE domain in PDDL syntax.
-
-\begin{figure}[t]
-  \centering
-  \subfloat[position=bottom][(a) Minimal world]{%
-      \includegraphics[height=2.5cm,clip=true,trim=0 -10px 0 -3px]{pic-buttons}}
-  \qquad
-  \subfloat[position=bottom][(b) Extra grid cells with accessible positions]{%
-      \includegraphics[height=2.5cm]{pic-empty-buttons}} \\
-  \subfloat[(c) Extra grid cells with inaccessible positions]{%
-      \includegraphics[height=2.5cm]{pic-empty-inaccessible}}
-  \caption{Experimental GIVE world configuration.}
-  \label{fig:give-maps}
-\end{figure}
 
@@
 \begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp2}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-buttons}
-  \caption{Results on the minimal GIVE
-    domain for $h=20$. The horizontal axis is $n$. The vertical axis
-    is the runtime in seconds.}
+\subfloat[(a) Unordered]{\includegraphics[width=7cm]{data/give-minimal.pdf}}
+\subfloat[(b) Ordered]{\includegraphics[width=7cm]{data/give2-minimal.pdf}}
+  \caption{Results for the unordered and ordered minimal GIVE domains with grid
+  height $h=20$. The horizontal axis is the grid width, $b = 2n$. The
+  vertical axis is the total runtime in seconds.}
   \label{fig:give-runtime-minimal}
@@
 Results for the $h=20$, $n=5$ case with $w$ ranging from $1$ to $70$ are
-shown in Figure~\ref{fig:give-runtime-junk}. As in Experiment~2, FF again
+shown in Figure~\ref{fig:give-extracells}. As in Experiment~2, FF again
 runs out of memory, this time at $w=17$, while SGPLAN easily solves inputs
@@
 
+%\begin{figure}[t]
+%  \centering
+%  \includegraphics[width=0.75\columnwidth]{graph-exp3}
+%  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
+%  \caption{Results on the GIVE domain with junk
+%    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
+%    The vertical axis is the runtime in seconds.}
+%\end{figure}
+
 \begin{figure}[t]
-  \centering
-  \includegraphics[width=0.75\columnwidth]{graph-exp3}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
-  \caption{Results on the GIVE domain with junk
-    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
-    The vertical axis is the runtime in seconds.}
-  \label{fig:give-runtime-junk}
+\subfloat[(a) Unordered]{\includegraphics[width=7cm]{data/give-extracells.pdf}}
+\subfloat[(b) Ordered]{\includegraphics[width=7cm]{data/give2-extracells.pdf}}
+  \caption{Results for the unordered and ordered GIVE domains with $h = 20$ and
+  $n = 10$. The horizontal axis is the extra grid width $w$.
+  The vertical axis is the total runtime in seonds.}
+  \label{fig:give-extracells}
+\end{figure}
+
+\begin{figure}[t]
+\subfloat[(a) Unordered]{\includegraphics[width=7cm]{data/give-button-unord.pdf}}
+\subfloat[(b) Ordered]{\includegraphics[width=7cm]{data/give-button-ord.pdf}}
+  \caption{Results for the unordered and ordered GIVE domains with a grid size
+  of height 20 and width 40. The horizontal axis is the number of buttons $b$.
+  The vertical axis is the runtime in seonds.}
+  \label{fig:give-buttons}
 \end{figure}

### 2009-06-03 trunk/Papers/ci-spark/experiments.tex
changes to section 4

@@
 Results for the $h=20$ case, with $n$ ranging from $1$ to $40$, are shown
-in Figure~\ref{fig:give-runtime-minimal}. The most obvious result is that
+in Figure~\ref{fig:give-minimal}. The most obvious result is that
 FF is unable to solve any problems beyond $n=13$ on our experimentation
@@
   vertical axis is the total runtime in seconds.}
-  \label{fig:give-runtime-minimal}
+  \label{fig:give-minimal}
+\end{figure}
+
+\begin{figure}[t]
+\subfloat[(a) Unordered]{\includegraphics[width=7cm]{data/give-minimal-grounding.pdf}}
+\subfloat[(b) Ordered]{\includegraphics[width=7cm]{data/give2-minimal-grounding.pdf}}
+  \caption{Comparison of the total running time and grounding time for selected
+  planners in the $h=20$ minimal GIVE domain. The horizontal axis is the grid
+  width, $b = 2n$. The vertical axis is the total runtime in seconds.}
+  \label{fig:give-minimal-grounding}
 \end{figure}
@@
   $n = 10$. The horizontal axis is the extra grid width $w$.
-  The vertical axis is the total runtime in seonds.}
+  The vertical axis is the total runtime in seconds.}
   \label{fig:give-extracells}
@@
 \subfloat[(b) Ordered]{\includegraphics[width=7cm]{data/give-button-ord.pdf}}
-  \caption{Results for the unordered and ordered GIVE domains with a grid size
-  of height 20 and width 40. The horizontal axis is the number of buttons $b$.
-  The vertical axis is the runtime in seonds.}
+  \caption{Results for the GIVE domains with a fixed grid size of
+  height 20 and width 40. The horizontal axis is the number of buttons $b$.
+  The vertical axis is the runtime in seconds (log scale).}
   \label{fig:give-buttons}

### 2009-06-03 trunk/Papers/ci-spark/experiments.tex
changes

@@
 
-In the second experiment, we evaluate the performance of the planners on
-problems arising in the GIVE domain. We construct a series of test worlds,
-similar to the one illustrated in Figure~\ref{fig:give-maps}(a). These
-worlds consist of a $2n$ by $h$ grid of positions, such that there are
-buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The
-player starts in position $(1,1)$ and must press all the buttons
-to successfully complete the game. The world is generated as a GIVE world
-description, and then automatically converted into a planning problem by
-the GIVE software. For instance, Figure~\ref{fig:give-planning} shows some
-of the actions available in the GIVE domain in PDDL syntax.
+In the second set of experiments, we evaluate the performance of several
+planners on problems arising in the GIVE domain. Besides the FF, Metric-FF, and
+SGPLAN planners we used in the previous set of experiments, we expand our
+testing to include FF(h$_a$), LAMA, and C$^3$ \todo{citations}. Each of these
+additional planners competed in the deterministic ``sequential, satifying''
+track of the 2008 International Planning Competition; all planners generally
+performed well on the competition domains, with LAMA the overall winner of
+the track.\footnote{See \texttt{http://ipc.informatik.uni-freiburg.de/} for
+details of the 2008 IPC.}
 
-Results for the $h=20$ case, with $n$ ranging from $1$ to $40$, are shown
-in Figure~\ref{fig:give-minimal}. The most obvious result is that
-FF is unable to solve any problems beyond $n=13$ on our experimentation
-machine within the memory limit of 1 GB. SGPLAN, on the other hand, solves
-instances beyond $n=40$ without major problems. In this case, our
-implementation of GraphPlan is unable to find any of these plans. The time
-spent on grounding is not a major factor in either planner, probably
-because the planners need more time to actually compute the plan---for
-instance, the optimal plan for the problem $n=40$ has a length of about
-1600 steps. As a concrete example, the following is a minimal plan for the
-case of a $2$ by $2$ grid with $2$ buttons for the player to press (i.e.,
-$n=1$, $h=2$):
+In the first GIVE test, we construct a series of grid worlds, similar to the one
+illustrated in Figure~\ref{fig:give-maps}(a). These worlds consist of a $2n$ by
+$h$ grid of positions, such that there are buttons at positions $(2i-1,1)$ and
+$(2i,h)$ for $1 \leq i \leq n$. The player starts in position $(1,1)$ and must
+press all the buttons to successfully complete the game.  (The actions in this
+domain are similar to the PDDL actions in Figure~\ref{fig:give-planning}.)
+We consider two variants of this problem in our tests. In the \emph{unordered}
+problem, the player is permitted to press the buttons in any order to
+successfully achieve the goal. In the \emph{ordered} version of the problem, the
+player is unable to initially move to any grid cell containing a button, except
+for the cell containing the first button, $u_1$. Pressing $u_1$ releases the
+position of the next button, $l_1$, allowing the player to move into this cell.
+Similarly, pressing button $l_1$ frees button $u_2$, and so on. The end result
+is a set of constraints that forces the buttons to be pressed in a particular
+order to achieve the goal. As a concrete example, the following is a minimal
+plan (in either variant of the problem) for the case of a $2$ by $2$ grid with
+$2$ buttons (i.e., $n=1$, $h=2$):
 
@@
 
+Results for the $h=20$ case, with the grid width $N$ ranging from $0$ to $40$,
+are shown in Figure~\ref{fig:give-minimal}. In the unordered case
+(Figure~\ref{fig-give-minimal}(a)), the most obvious result is that some of the
+planners tested---Metric-FF, FF(h$_a$), and C$^3$---are unable to solve any
+problems beyond $N=24$ on our experimentation machine within the memory limit of
+2 GB.\footnote{\todo{describe testing environment}} While FF, Lama, and SGPLAN
+are all able to solve problem instances up to $N=40$, the total running time
+varies greatly between this planners. For instance, FF takes almost 35
+seconds to solve the $N=40$ problem, LAMA takes around 6.5 seconds, and SGPLAN
+shows impressive performance, taking well less than a second.
+
+In a real NLG system, where response time is
+
+machine within the memory limit of 1 GB. SGPLAN, on the other hand, solves
+instances beyond $n=40$ without major problems. In this case, our
+implementation of GraphPlan is unable to find any of these plans. The time
+spent on grounding is not a major factor in either planner, probably
+because the planners need more time to actually compute the plan---for
+instance, the optimal plan for the problem $n=40$ has a length of about
+1600 steps.
+
 \begin{figure}[t]
@@
   \caption{Results for the unordered and ordered minimal GIVE domains with grid
-  height $h=20$. The horizontal axis is the grid width, $b = 2n$. The
+  height $h=20$. The horizontal axis is the grid width, $N = 2n$. The
   vertical axis is the total runtime in seconds.}
@@
   planners in the $h=20$ minimal GIVE domain. The horizontal axis is the grid
-  width, $b = 2n$. The vertical axis is the total runtime in seconds.}
+  width, $N = 2n$. The vertical axis is the total runtime in seconds.}
   \label{fig:give-minimal-grounding}
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=0.85\columnwidth]{graph-exp4}
-  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
-  \caption{Results on the GIVE domain with inaccessible
-    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
-    The vertical axis is the runtime in seconds.}
-  \label{fig:give-runtime-nosoln}
-\end{figure}
+%\begin{figure}
+%  \centering
+%  \includegraphics[width=0.85\columnwidth]{graph-exp4}
+%  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
+%  \caption{Results on the GIVE domain with inaccessible
+%    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
+%    The vertical axis is the runtime in seconds.}
+%  \label{fig:give-runtime-nosoln}
+%\end{figure}
 

### 2009-06-03 trunk/Papers/ci-spark/experiments.tex
adjusted colors; added references to planenrs

@@
 
-In the second set of experiments, we evaluate the performance of several
-planners on problems arising in the GIVE domain. Besides the FF, Metric-FF, and
-SGPLAN planners we used in the previous set of experiments, we expand our
-testing to include FF(h$_a$), LAMA, and C$^3$ \todo{citations}. Each of these
-additional planners competed in the deterministic ``sequential, satifying''
-track of the 2008 International Planning Competition; all planners generally
-performed well on the competition domains, with LAMA the overall winner of
-the track.\footnote{See \texttt{http://ipc.informatik.uni-freiburg.de/} for
-details of the 2008 IPC.}
+In the second set of experiments, we evaluate the performance of
+several planners on problems arising in the GIVE domain. Besides the
+FF, Metric-FF, and SGPLAN planners we used in the previous set of
+experiments, we expand our testing to include FF($h_a$)
+\citep{keyder08:_ff_h_plann_for_plann}, LAMA
+\citep{richter08:_lama_plann}, and C$^3$
+\citep{lipovetzky08:_c3}. Each of these additional planners competed
+in the deterministic ``sequential, satifying'' track of the 2008
+International Planning Competition; all planners generally performed
+well on the competition domains, with LAMA the overall winner of the
+track.\footnote{See \texttt{http://ipc.informatik.uni-freiburg.de/}
+  for details of the 2008 IPC.}
 

### 2009-06-03 trunk/Papers/ci-spark/experiments.tex
changes to section 4

--- 2009-06-03 trunk/Papers/ci-spark/experiments.tex
@@
 
-\subsection{Experiment 2: Minimal GIVE worlds}
+\subsection{Experiment 3: Minimal GIVE worlds}
 \label{sec:exper-2:-minim}
 
-In the second set of experiments, we evaluate the performance of several
-planners on problems arising in the GIVE domain. Besides the FF, Metric-FF, and
-SGPLAN planners we used in the previous set of experiments, we expand our
-testing to include FF(h$_a$), LAMA, and C$^3$ \todo{citations}. Each of these
-additional planners competed in the deterministic ``sequential, satifying''
-track of the 2008 International Planning Competition; all planners generally
-performed well on the competition domains, with LAMA the overall winner of
-the track.\footnote{See \texttt{http://ipc.informatik.uni-freiburg.de/} for
-details of the 2008 IPC.}
+We now turn our attention to a set of experiments arising from the GIVE domain.
+Besides using some of the planners from the previous set of experiments (FF,
+Metric-FF, and SGPLAN), we expand our testing to include the FF(h$_a$), LAMA, and
+C$^3$ planners \todo{citations}. Each of these additional planners competed in the
+deterministic ``sequential, satifying'' track of the 2008 International Planning
+Competition; all planners generally performed well on the competition domains,
+with LAMA the overall winner of the track.\footnote{See
+\texttt{http://ipc.informatik.uni-freiburg.de/} for details of the 2008 IPC.}
 
-In the first GIVE test, we construct a series of grid worlds, similar to the one
-illustrated in Figure~\ref{fig:give-maps}(a). These worlds consist of a $2n$ by
-$h$ grid of positions, such that there are buttons at positions $(2i-1,1)$ and
-$(2i,h)$ for $1 \leq i \leq n$. The player starts in position $(1,1)$ and must
-press all the buttons to successfully complete the game.  (The actions in this
-domain are similar to the PDDL actions in Figure~\ref{fig:give-planning}.)
-We consider two variants of this problem in our tests. In the \emph{unordered}
-problem, the player is permitted to press the buttons in any order to
-successfully achieve the goal. In the \emph{ordered} version of the problem, the
-player is unable to initially move to any grid cell containing a button, except
-for the cell containing the first button, $u_1$. Pressing $u_1$ releases the
-position of the next button, $l_1$, allowing the player to move into this cell.
-Similarly, pressing button $l_1$ frees button $u_2$, and so on. The end result
-is a set of constraints that forces the buttons to be pressed in a particular
-order to achieve the goal. As a concrete example, the following is a minimal
-plan (in either variant of the problem) for the case of a $2$ by $2$ grid with
-$2$ buttons (i.e., $n=1$, $h=2$):
+In the first GIVE experiment, we construct a series of grid worlds, similar to
+the one illustrated in Figure~\ref{fig:give-maps}(a). These worlds consist of a
+$N = 2n$ by $h$ grid of positions, such that there are buttons at positions
+$(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The player starts in position
+$(1,1)$ and must press all the buttons to successfully complete the game.  (The
+actions in this domain are similar to the PDDL actions in
+Figure~\ref{fig:give-planning}.) We consider two variants of this problem in our
+tests. In the \emph{unordered} problem, the player is permitted to press the
+buttons in any order to successfully achieve the goal. In the \emph{ordered}
+version of the problem, the player is unable to initially move to any grid cell
+containing a button, except for the cell containing the first button, $u_1$.
+Pressing $u_1$ releases the position of the next button, $l_1$, allowing the
+player to move into this cell.  Similarly, pressing button $l_1$ frees button
+$u_2$, and so on. The end result is a set of constraints that forces the buttons
+to be pressed in a particular order to achieve the goal. As a concrete example,
+the following is a minimal plan (in either variant of the problem) for the case
+of a $2$ by $2$ grid with $2$ buttons (i.e., $n=1$, $h=2$):
 
@@
 \item $\mathsf{move}(\mathsf{pos\_1\_1},\mathsf{pos\_1\_2}, \mathsf{north})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{u1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_1\_2})$,
+\item $\mathsf{manipulate}\textsf{-}\mathsf{button}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{u1, pos\_1\_2})$,
 \item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{north}, \mathsf{east})$,
@@
 \item $\mathsf{move}(\mathsf{pos\_2\_2}, \mathsf{pos\_2\_1}, \mathsf{south})$,
-\item $\mathsf{manipulate}\textsf{-}\mathsf{l1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_2\_1})$.
+\item $\mathsf{manipulate}\textsf{-}\mathsf{button}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{l1, pos\_2\_1})$.
 \end{enumerate}
 
-Results for the $h=20$ case, with the grid width $N$ ranging from $0$ to $40$,
+Results for the $h=20$ case, with the grid width $N$ ranging from $1$ to $40$,
 are shown in Figure~\ref{fig:give-minimal}. In the unordered case
-(Figure~\ref{fig-give-minimal}(a)), the most obvious result is that some of the
+(Figure~\ref{fig:give-minimal}(a)), the most obvious result is that some of the
 planners tested---Metric-FF, FF(h$_a$), and C$^3$---are unable to solve any
 problems beyond $N=24$ on our experimentation machine within the memory limit of
-2 GB.\footnote{\todo{describe testing environment}} While FF, Lama, and SGPLAN
-are all able to solve problem instances up to $N=40$, the total running time
-varies greatly between this planners. For instance, FF takes almost 35
-seconds to solve the $N=40$ problem, LAMA takes around 6.5 seconds, and SGPLAN
-shows impressive performance, taking well less than a second.
+2 GB.\footnote{\todo{describe testing environment}} While FF, LAMA, and SGPLAN
+are able to solve all problem instances up to $N=40$, the total running time
+varies greatly between these planners. For instance, FF takes almost 35 seconds
+to solve the $N=40$ problem, while LAMA takes around 6.5 seconds. SGPLAN shows
+impressive performance, generating a 240 step plan in the $N=40$ case in well
+less than a second. In the ordered case (Figure~\ref{fig:give-minimal}(b)), we
+again have the situation where Metric-FF, FF(h$_a$), and C$^3$ are unable to
+solve all problem instances. Furthermore, both SGPLAN and LAMA, which performed
+well on the unordered problem, now perform much worse than FF: FF takes 39
+seconds for the $N=40$ case, while SGPLAN takes 50 seconds and LAMA takes 90
+seconds. In real NLG systems, where response time is essential, running times
+over a few seconds are unacceptable.
 
-In a real NLG system, where response time is
-
-machine within the memory limit of 1 GB. SGPLAN, on the other hand, solves
-instances beyond $n=40$ without major problems. In this case, our
-implementation of GraphPlan is unable to find any of these plans. The time
-spent on grounding is not a major factor in either planner, probably
-because the planners need more time to actually compute the plan---for
-instance, the optimal plan for the problem $n=40$ has a length of about
-1600 steps.
+Preprocessing (parsing, grounding, etc.) time generally plays less of a role in
+GIVE, compared with the sentence generation domain, however, its effects still
+contribute significantly to the overall running time of a number of planners.
+Figure~\ref{fig:give-minimal-grounding} shows the grounding time for FF, LAMA,
+and SGPLAN on the minimal GIVE problems, compared with the total running time.
+In the unordered variant of the minimal GIVE domain
+(Figure~\ref{fig:give-minimal-grounding}(a)), the grounding time in LAMA and
+SGPLAN accounts for a significant fraction of the total runtime: SGPLAN spends
+around 40\% of its total runtime on preprocessing; for LAMA, this number
+rises to at least 80\% for our test problems. For FF, the preprocessing time is
+much less important than the search time, especially for large problem
+instances. In the ordered case (Figure~\ref{fig:give-minimal-grounding}(b)), the
+actual time spent on preprocessing is essentially unchanged from the unordered
+case, and search time dominates the total runtime for all three planners.
+However, FF is now much better at controlling the search,
+compared with the unordered variant of the problem.
 
@@
   \caption{Results for the unordered and ordered minimal GIVE domains with grid
-  height $h=20$. The horizontal axis is the grid width, $N = 2n$. The
+  height $h=20$. The horizontal axis is the grid width, $N$. The
   vertical axis is the total runtime in seconds.}
@@
   planners in the $h=20$ minimal GIVE domain. The horizontal axis is the grid
-  width, $N = 2n$. The vertical axis is the total runtime in seconds.}
+  width, $N$. The vertical axis is the total runtime in seconds.}
   \label{fig:give-minimal-grounding}
@@
 
-\subsection{Experiment 3: GIVE worlds with extra positions}
+\subsection{Experiment 4: GIVE worlds with extra positions}
 \label{sec:experiment-3:-give}
@@
 
-\subsection{Experiment 4: GIVE worlds with inaccessible positions}
+\subsection{Experiment 5: GIVE worlds with inaccessible positions}
 \label{sec:experiment-4:-give}

### 2009-06-03 trunk/Papers/ci-spark/experiments.tex
changes to section 4

@@
 We now turn our attention to a set of experiments arising from the GIVE domain.
-Besides using some of the planners from the previous set of experiments (FF,
-Metric-FF, and SGPLAN), we expand our testing to include the FF(h$_a$), LAMA, and
-C$^3$ planners \todo{citations}. Each of these additional planners competed in the
-deterministic ``sequential, satifying'' track of the 2008 International Planning
-Competition; all planners generally performed well on the competition domains,
-with LAMA the overall winner of the track.\footnote{See
+Besides using many of the planners from the previous set of experiments (FF,
+Metric-FF, and SGPLAN), we also expand our testing to include the FF(h$_a$)
+\citep{keyder08:_ff_h_plann_for_plann}, LAMA \citep{richter08:_lama_plann}, and
+C$^3$ \citep{lipovetzky08:_c3} planners. Each of these additional planners
+competed in the deterministic ``sequential, satifying'' track of the 2008
+International Planning Competition; all planners performed well on the
+competition domains, with LAMA the overall winner of the track.\footnote{See
 \texttt{http://ipc.informatik.uni-freiburg.de/} for details of the 2008 IPC.}
@@
 case, and search time dominates the total runtime for all three planners.
-However, FF is now much better at controlling the search,
-compared with the unordered variant of the problem.
+However, overall, FF is now much better at controlling the search, compared with
+the other planners and its performance on the unordered variant of the problem.
 

### 2009-06-04 trunk/Papers/ci-spark/experiments.tex
changes to section 4

@@
 domain that is similar to the classic Gridworld
-(Section~\ref{sec:exper-2:-minim}), and then add extra grid cells to the world
-that are neither needed (Section~\ref{sec:experiment-3:-give}) nor even
-accessible (Section~\ref{sec:experiment-4:-give}). We also investigate the role
-that goal ordering plays in some of these problems. These experiments are
-configured in a way that lets us explore the scalability of a planner's search
-and preprocessing capabilities, and illustrates what we perceive to be one of
-the main limitations of current off-the-shelf planners for our applications:
-they often spend a long time computing ground instances, even when most of
-these instances are not required during plan search.
+(Section~\ref{sec:exper-3:-minim}), and then add extra grid cells to the world
+that are not necessary to complete the task
+(Section~\ref{sec:experiment-4:-give}). We also investigate the role that goal
+ordering plays in these problems. These experiments are configured in a way that
+lets us explore the scalability of a planner's search and preprocessing
+capabilities, and illustrates what we perceive to be one of the main limitations
+of current off-the-shelf planners for our applications: they often spend a long
+time computing ground instances, even when most of these instances are not
+required during plan search.
 
@@
   \centering
-  \subfloat[position=bottom][(a) Minimal world]{%
+  \subfloat[position=bottom][(a) Minimal GIVE world]{%
       \includegraphics[height=2.5cm,clip=true,trim=0 -10px 0 -3px]{pic-buttons}}
   \qquad
-  \subfloat[position=bottom][(b) World with extra grid cells]{%
+  \subfloat[position=bottom][(b) GIVE world with extra grid cells]{%
       \includegraphics[height=2.5cm]{pic-empty-buttons}} \\
@@
 \subsection{Experiment 3: Minimal GIVE worlds}
-\label{sec:exper-2:-minim}
+\label{sec:exper-3:-minim}
 
@@
 
+\begin{figure}[t]
+\subfloat[(a) Unordered]{\includegraphics[width=7cm]{data/give-minimal.pdf}}
+\subfloat[(b) Ordered]{\includegraphics[width=7cm]{data/give2-minimal.pdf}}
+  \caption{Results for the unordered and ordered minimal GIVE domains with grid
+  height $h=20$. The horizontal axis is the grid width, $N$. 
+  The vertical axis is the total runtime in seconds.}
+  \label{fig:give-minimal}
+\end{figure}
+
 Results for the $h=20$ case, with the grid width $N$ ranging from $1$ to $40$,
@@
 problems beyond $N=24$ on our experimentation machine within the memory limit of
-2 GB.\footnote{\todo{describe testing environment}} While FF, LAMA, and SGPLAN
-are able to solve all problem instances up to $N=40$, the total running time
-varies greatly between these planners. For instance, FF takes almost 35 seconds
-to solve the $N=40$ problem, while LAMA takes around 6.5 seconds. SGPLAN shows
-impressive performance, generating a 240 step plan in the $N=40$ case in well
-less than a second. In the ordered case (Figure~\ref{fig:give-minimal}(b)), we
-again have the situation where Metric-FF, FF(h$_a$), and C$^3$ are unable to
-solve all problem instances. Furthermore, both SGPLAN and LAMA, which performed
-well on the unordered problem, now perform much worse than FF: FF takes 39
-seconds for the $N=40$ case, while SGPLAN takes 50 seconds and LAMA takes 90
-seconds. In real NLG systems, where response time is essential, running times
-over a few seconds are unacceptable.
+2 GB.\footnote{All runtimes in Sections~\ref{sec:exper-3:-minim} and
+ \ref{sec:experiment-4:-give} were measured on a single core of an Intel Xeon
+ CPU running at 3GHz, under Linux. All runtimes are averaged over three runs of
+ the planners. Only 32-bit versions of the planners were used for testing in
+ each case.}
+While FF, LAMA, and SGPLAN are able to solve all problem instances up to $N=40$,
+the total running time varies greatly between these planners. For instance, FF
+takes almost 35 seconds to solve the $N=40$ problem, while LAMA takes around 6.5
+seconds. SGPLAN shows impressive performance on $N=40$, generating a 240 step
+plan in well under a second. In the ordered case
+(Figure~\ref{fig:give-minimal}(b)), we again have the situation where Metric-FF,
+FF(h$_a$), and C$^3$ are unable to solve all problem instances. Furthermore,
+both SGPLAN and LAMA, which performed well on the unordered problem, now perform
+much worse than FF: FF takes 39 seconds for the $N=40$ case, while SGPLAN takes
+50 seconds and LAMA takes 90 seconds. In real NLG systems, where response time
+is essential, running times over a few seconds are unacceptable.
 
@@
 case, and search time dominates the total runtime for all three planners.
-However, overall, FF is now much better at controlling the search, compared with
+Overall, however, FF is now much better at controlling the search, compared with
 the other planners and its performance on the unordered variant of the problem.
-
-\begin{figure}[t]
-\subfloat[(a) Unordered]{\includegraphics[width=7cm]{data/give-minimal.pdf}}
-\subfloat[(b) Ordered]{\includegraphics[width=7cm]{data/give2-minimal.pdf}}
-  \caption{Results for the unordered and ordered minimal GIVE domains with grid
-  height $h=20$. The horizontal axis is the grid width, $N$. The
-  vertical axis is the total runtime in seconds.}
-  \label{fig:give-minimal}
-\end{figure}
 
@@
 
-\subsection{Experiment 4: GIVE worlds with extra positions}
-\label{sec:experiment-3:-give}
+\subsection{Experiment 4: GIVE worlds with extra grid cells}
+\label{sec:experiment-4:-give}
 
-In the third experiment, we vary the structure of the GIVE world in order
-to judge the effect that universe size has on the planning problem in this
-domain. Starting with the ordinary GIVE world described in Experiment~2, we
-extend the world map by adding another $w$ by $h$ empty ``junk'' positions
-to the right of the minimal world, as shown in Figure~\ref{fig:give-maps}(b).
-These new positions are not actually required in any plan, but extend the
-size of the state space and approximate the situation in the actual GIVE
-domain where most grid positions are never used. We leave the initial state
-and goal untouched. As before, we generate a GIVE world description and
-then convert it into a planning problem in PDDL.
+In our last set of experiments, we vary the structure of the GIVE world in order
+to judge the effect that universe size has on the resulting planning problem.
+Starting with the GIVE world described in Experiment~3, we extend the world map
+by adding another $w$ by $h$ empty cell positions to the right of the minimal
+world, as shown in Figure~\ref{fig:give-maps}(b). These new positions are not
+actually required in any plan, but extend the size of the state space and
+approximate the situation in the actual GIVE domain where most grid positions
+are never used. We leave the initial state and goal untouched and, again,
+consider both unordered and ordered variants of the problem.
 
-Results for the $h=20$, $n=5$ case with $w$ ranging from $1$ to $70$ are
-shown in Figure~\ref{fig:give-extracells}. As in Experiment~2, FF again
-runs out of memory, this time at $w=17$, while SGPLAN easily solves inputs
-beyond $w=70$. As in Experiment~2, our implementation of GraphPlan is
-unable to generate any of these plans. Unlike Experiment 2, however, both
-FF and SGPLAN now spend a substantial proportion of their time on
-grounding. In SGPLAN, this translates to a ``parsing time'' (which we
-assume includes grounding) which grows from 180 ms to 21.7 seconds as $w$
-grows from $1$ to $75$. The rest of the runtime, which also includes the
-search time, only grows from 400 ms to 2.3 seconds. This difference is
-particularly dramatic given that the actual optimal plan in each case is an
-identical plan of about 100 steps (i.e., the one we would have found in
-Experiment 2 for the $h=20$, $n=5$ case). The planning times for these
-instances are also concerning since times over a couple seconds will
-negatively affect the overall response time of the system, which must react
-in real time to user actions.
+Results for the $h=20$, $n=10$ case with $w$ ranging from $1$ to $40$ are shown
+in Figure~\ref{fig:give-extracells}. As in Experiment~3, a number of planners
+again fail to solve all the problems: Metric-FF, FF(h$_a$), and C$^3$ solve only
+a few instances, while FF only scales to $w=23$. In the unordered version of the
+domain, SGPLAN easily solves inputs beyond $w=40$ in well less than a second.
+LAMA is also reasonably successful on these problems, however, its runtimes grow
+more quickly than SGPLAN, with LAMA taking almost 5 seconds to solve the $w=40$
+problem instance. In the ordered case, we again see behaviour similar to that of
+Experiment~3: for the problem instances FF is able to solve, it performs
+significantly better than LAMA and SGPLAN. (SGPLAN's long term runtime
+appears to be growing at a slower rate than FF's, and so even if FF could be
+scaled to larger problem instances, it seems possible that SGPLAN might overtake
+FF as the better performer.) However, the overall planning times for most of
+these instances are concerning since times over a couple seconds will negatively
+affect the response time of an NLG system, which must react in real time to user
+actions.
 
-%\begin{figure}[t]
-%  \centering
-%  \includegraphics[width=0.75\columnwidth]{graph-exp3}
-%  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
-%  \caption{Results on the GIVE domain with junk
-%    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
-%    The vertical axis is the runtime in seconds.}
-%\end{figure}
+Finally, we also performed a set of experiments designed to investigate the
+tradeoff between grounding time and search time on certain grid configurations.
+For these experiments, we initially fixed the size of the grid and then varied
+the number of buttons $b$ in the world, thereby creating a series of
+``snapshots'' of particular extra-cell GIVE domains.
+Figure~\ref{fig:give-buttons} shows the results of these experiments for the FF
+and SGPLAN planners, for a fixed size grid of height 20 and width 40, and the
+number of buttons $b$ ranging from 1 to 40. In each case, the amount of time a
+planner spends on grounding is relatively unchanged as we vary the number of
+buttons in a grid, while the search time continues to rise (sometimes quite
+dramatically), as $b$ increases (we saw a similar effect for other grid
+configurations we tried). This observation has important consequences for the
+design of our GRID worlds: changing the underlying domain structure, even
+minimally, may result in significant---and often unexpected---performance
+differences for the planners that must operate in these domains.
 
@@
 
-\subsection{Experiment 5: GIVE worlds with inaccessible positions}
-\label{sec:experiment-4:-give}
+%In the final experiment, we consider a variation on the GIVE worlds from
+%Experiment~3. As before, we begin with a base grid of $2n$ by $h$
+%positions, with buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i
+%\leq n$. We then generate a second grid of $w$ by $h$ positions and place
+%an additional button in this grid. Unlike Experiment~3, the second grid
+%does not extend the first grid but is ``disconnected'' so its positions are
+%inaccessible from the first grid, as shown in
+%Figure~\ref{fig:give-maps}(c). Since the geometry of the grid makes it
+%impossible to construct a plan for pressing all the buttons in the world,
+%we instead investigate the time it takes a planner to arrive at the
+%conclusion that the problem cannot be solved.
 
-In the final experiment, we consider a variation on the GIVE worlds from
-Experiment~3. As before, we begin with a base grid of $2n$ by $h$
-positions, with buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i
-\leq n$. We then generate a second grid of $w$ by $h$ positions and place
-an additional button in this grid. Unlike Experiment~3, the second grid
-does not extend the first grid but is ``disconnected'' so its positions are
-inaccessible from the first grid, as shown in
-Figure~\ref{fig:give-maps}(c). Since the geometry of the grid makes it
-impossible to construct a plan for pressing all the buttons in the world,
-we instead investigate the time it takes a planner to arrive at the
-conclusion that the problem cannot be solved.
-
-Results for the $h=20$, $n=5$ case with $w$ ranging from 1 to 50 are shown
-in Figure~\ref{fig:give-runtime-nosoln}.\footnote{FF and SGPLAN do
- not display runtimes when they fail to construct a plan so an external timing
- program was used to generate these results.}
-In this case, the runtimes for FF rise sharply around $w=28$, producing
-times that are unacceptable for an NLG system in practice. FF is unable
-to complete problem instances above $w=35$. SGPLAN, by comparison, performs
-exceptionally well and is able to complete problems instances up to $w=50$
-in about 0.3 seconds and $w=80$ in about 0.6 seconds. These results
-demonstrate that SGPLAN has a significant advantage over FF in this
-experiment. 
-%Our implementation of GraphPlan is unable to complete any of
-%these problem instances.
-
-%\begin{figure}
-%  \centering
-%  \includegraphics[width=0.85\columnwidth]{graph-exp4}
-%  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
-%  \caption{Results on the GIVE domain with inaccessible
-%    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
-%    The vertical axis is the runtime in seconds.}
-%  \label{fig:give-runtime-nosoln}
-%\end{figure}
+%Results for the $h=20$, $n=5$ case with $w$ ranging from 1 to 50 are shown
+%in Figure.\footnote{FF and SGPLAN do
+% not display runtimes when they fail to construct a plan so an external timing
+% program was used to generate these results.}
+%In this case, the runtimes for FF rise sharply around $w=28$, producing
+%times that are unacceptable for an NLG system in practice. FF is unable
+%to complete problem instances above $w=35$. SGPLAN, by comparison, performs
+%exceptionally well and is able to complete problems instances up to $w=50$
+%in about 0.3 seconds and $w=80$ in about 0.6 seconds. These results
+%demonstrate that SGPLAN has a significant advantage over FF in this
+%experiment.
 

### 2009-06-04 trunk/Papers/ci-spark/experiments.tex
spell check

@@
 quickly. We also do not report runtimes for our Java implementation of
-Graphplan, because it was unusuably slow for serious problem
+GraphPlan, because it was unusably slow for serious problem
 instances: For $k=1$ and $n=3$, it already took over two minutes, and
@@
 Metric-FF) fell back to the best-first search strategy, which causes a
-dramatic loss of search effiency. It is also encouraging that
+dramatic loss of search efficiency. It is also encouraging that
 Metric-FF still performs comparably to SPUD in terms of pure search
@@
 C$^3$ \citep{lipovetzky08:_c3} planners. Each of these additional planners
-competed in the deterministic ``sequential, satifying'' track of the 2008
+competed in the deterministic ``sequential, satisfying'' track of the 2008
 International Planning Competition; all planners performed well on the

### 2009-06-17 trunk/Papers/crisp-rtg-09/crisp-rtg-09.tex
ported npc proof and crisp encoding

@@
 \usepackage{graphicx}
-\usepackage{natbib}
+\usepackage[round]{natbib}
 \usepackage{amsfonts,amsthm}
@@
 
+\newcommand{\actionname}[3]{\mathbf{\mathtt{#1}\mbox{-}\mathtt{#2}\mbox{-}\mathtt{#3}}}
+\newcommand{\action}[4]{\textbf{Action \texttt{#1}$(#2)$.}\\
+\strut\quad   Precond:$\;$ \parbox[t]{12cm}{\ensuremath{#3}}\\
+\strut\quad   Effect:$\;$ \parbox[t]{12cm}{\ensuremath{#4}}}
+\newcommand{\actionc}[4]{\textbf{Action \texttt{#1}$(#2)$.}
+Precond:$\;$ \parbox[t]{3cm}{\ensuremath{#3}}\\
+\strut\quad   Effect:$\;$ \parbox[t]{12cm}{\ensuremath{#4}}}
+
+\newcommand{\open}{\ensuremath{\mathsf{open}}}
+\newcommand{\gensym}{\ensuremath{\mathsf{gensym}()}}
+
 \title{Sentence generation with RTGs}
@@
 \section{The generation problem of regular tree grammars}
+\label{sec:gener-probl-regul}
 
@@
 \section{RTG generation as planning} 
+\label{sec:rtg-generation-as-planning}
+
+It is straightforward to extend the encoding of TAG sentence
+generation to planning \citep{KolSto07} to regular tree grammars.
+Notice that this makes the \citet{KolSto07} encoding a special case of
+the algorithm presented here, as the derivation trees of a given TAG
+grammar form regular tree languages
+\citep{schmitz08:_featur_in_tag_deriv_trees} and can therefore be
+described by RTGs.\footnote{This is not strictly speaking true,
+  because we have not shown how to represent the pragmatic conditions
+  and semantic requirements in an RGG, in order to keep the
+  presentation simple; but it is obvious how these could be added.}
+
+Let $l$ be a lexicon entry with production rule $A \rightarrow
+\sigma(B_1 \Neq r_1, \ldots, B_n \Neq r_n)$.  Then the grammaticality
+condition is encoded by the precondition and effects of a planning
+operator as follows:\footnote{I write $r(u)$ to indicate the
+  construction of a new symbol representing the $r$-child of $u$. I
+  take $\self(u) = u$ as a special case. This cannot be expressed in
+  PDDL, but can be encoded with the same planning-step-counting
+  construction as in the ACL paper.}
+
+\strut\\
+\action{$\sigma$/$L$}{u}
+{\open(u,A)}
+{\neg \open(u,A), \open(r_1(u),,A_1), \ldots, \open(r_n(u),A_n)}\\
+
+Notice that I write $\open$ to encode the unprocessed nodes in the
+tree we are constructing, rather than $\mathsf{subst}$ and
+$\mathsf{adj}$ as in the paper. Those were TAG-specific predicates,
+encoding TAG's substitution and adjunction operation, and are no
+longer necessary here. Effectively, what happens here is that the plan
+encodes a derivation of the underlying RTG grammar; an atom
+$\open(u,A)$ in the current planning state encodes the fact that the
+current partial RTG derivation contains a leaf $u$ that is labeled
+with the nonterminal symbol $A$ (and still needs to be expanded in the
+future).
+
+The translation of the semantic and reference conditions then proceeds
+essentially as in the paper.  However, one crucial difference is that
+we no longer distinguish between substitution and adjunction, and thus
+it is not completely obvious where distractors for new referring
+expressions get introduced. This used to happen exactly at the
+substitution nodes. Now we can get the same effect by introducing
+distractors only for those nonterminal occurrences on the right-hand
+side that are labeled with a semantic role that is not $\self$.
+
+Note that the encoding of RGG generation as a planning problem is not
+entirely correct.  The difference is that in the problem as defined in
+Section~\ref{sec:gener-probl-regul}, we took $a$ to be a possible
+distractor for a new referring expression if there was any tuple
+$(a_1,\ldots,a,\ldots,a_n)$ that satisfied the semantic condition of
+the lexicon entry that introduced the RE.  In the encoding into
+planning in the ACL paper, however, we only accept $a$ as a possible
+distractor if $a_1,\ldots,a_n$ are the intended referents for the
+other semantic roles in the lexicon entry. If we continue using this
+encoding here, we end up with a planning problem that generates too
+few distractors, and therefore can compute referring expressions that
+are not unique. This is something that needs to be fixed at some
+point, but for now the planning encoding seems to work pretty well in
+practice.
 
 \section{NP-completeness of the RTG generation problem}
+\label{sec:np-completeness}
+
+Finally, we can show that the sentence generation problem of RGGs is
+NP-complete. This means that it is combinatorially hard, and hence
+solving it with a worst-case exponential-time algorithm like a planner
+is appropriate.
+
+The NP-hardness proof below works by reducing the HAMILTONIAN-CYCLE
+problem \cite{garey79:_comput_and_intrac} to the sentence generation
+problem. HAMILTONIAN-CYCLE is the (NP-complete) problem of deciding
+whether a directed graph has a Hamiltonian cycle.  A Hamiltonian cycle
+is a path $(u_1,u_2,\ldots,u_n,u_1)$ in the graph such that the graph
+contains an edge $(u_i,u_{i+1})$ for every $1 \leq i <n$ and the edge
+$(u_n,u_1)$ -- i.e., it is a cycle --, and such that the cycle visits
+every node in the graph exactly once. In the example graph in
+Fig.~\ref{fig:npc}a, $(1,3,2,1)$ is a Hamiltonian cycle; $(1,2,1)$ is
+a cycle that is not Hamiltonian because it doesn't visit 3.
+
+\begin{prop} \label{prop:npc} The sentence generation problem of
+  regular generation grammars is NP-complete.
+\end{prop}
+\begin{proof}
+  It is obvious that the problem is in NP; once we have
+  nondeterministically guessed a candidate tree, we can test whether
+  it satisfies all conditions in polynomial time.
+
+  For the NP-hardness, let's say we have a graph $G =
+  (\{1,\ldots,n\},E)$ with $m$ edges of the form $(i,k)$ of which we
+  want to check whether it has a Hamiltonian cycle.  We construct a
+  regular generation grammar $G_G$ with nonterminals $N = \{S,
+  A_{11}\} \cup \{A_{ik} \;|\; 2 \leq i \leq n+1, 1 \leq k \leq n\}$,
+  start symbol $S$, terminal alphabet $\Sigma=\{f,a\}$, role alphabet
+  $R = \{\mathsf{r}, \self\}$, and the following lexicon entries:
+
+  $$\begin{array}{l}
+    S \rightarrow f(A_{11} \Neq \mathsf{r}) \\
+    \mathsf{sem} = \{ \mathsf{start}(\self,\mathsf{r})\} \\
+    \\
+    A_{ij} \rightarrow f(A_{(i+1)k} \Neq \self) \quad \mbox{for all
+      $(j,k) \in E$ and $1 \leq i \leq n$} \\ 
+    \mathsf{sem} = \{ \mathsf{neq}_j(\self) \} \\
+    \\
+    A_{(n+1)1} \rightarrow a \\
+    \mathsf{sem} = \emptyset
+  \end{array}
+  $$
+
+  Now the target referent $e$ and the communicative goal $\emptyset$
+  have a successful derivation given $G_G$ and the knowledge bases
+  $SKB = HKB = \{\mathsf{start}(e,1)\} \cup \{ \mathsf{neq}_i(k) \;|\;
+  1 \leq i,k \leq n \;\mbox{and}\; i \neq k \}$ iff $G$ has a
+  Hamiltonian cycle.  To see this, notice that any tree generated by
+  $G_G$ must be a linear tree with nonterminal symbols $S, A_{11},
+  A_{2u_2}, \ldots, A_{(n+1)1}$ from top to bottom, and for any two
+  subsequent nonterminals $A_{ij}$ and $A_{(i+1)k}$ in this sequence,
+  there must be an edge $(j,k)$ in the graph.  So the tree encodes a
+  cycle of length $n$.  Now, if any node in the graph, say $j$, was
+  not visited on this cycle, then the semantics in the subtree of the
+  referring expression with nonterminal $A_{11}$ does not contain the
+  atom $\mathsf{neq}_j(1)$.  In other words, the tree does not contain
+  sufficient information to distinguish 1 (the target referent of the
+  referring expression) from $j$, so the referring expression is not
+  unique, and the derivation is not successful.  Hence, the cycle
+  visits every node exactly once and is therefore Hamiltonian.
+\end{proof}
+
+\begin{figure}
+  \centering
+  \begin{tabular}{cp{2cm}c}
+  \includegraphics[scale=0.5]{pic-npc-graph}
+  &&
+  \includegraphics[scale=0.5]{pic-npc} \\
+  (a) && (b)
+  \end{tabular}
+  \caption{(a) A graph with a Hamiltonian cycle, and (b) a tree
+    encoding this cycle according to the construction in
+    Prop.~\ref{prop:npc}.} 
+  \label{fig:npc}
+\end{figure}
+
+An example of this construction (i.e., a graph together with the tree
+that encodes its Hamiltonian cycle and that is generated by the RGG)
+is shown in Fig.~\ref{fig:npc}. Notice that nothing in this
+construction particularly hinges in the generalization of the TAG
+generation problem to RGGs. The proof can thus be ported directly to a
+proof of the NP-completeness of the TAG sentence generation problem.
 

### 2009-07-30 tags/ant-crisp/Papers/ci-spark/experiments.tex
Tagged the last version of CRISP with Ant buildfile and directory structure before converting to Maven.

--- 2009-06-04 trunk/Papers/ci-spark/experiments.tex
@@
 lets us explore the scalability of a planner's search and preprocessing
-capabilities, and illustrates what we perceive to be one of the main limitations
+capabilities, and illustrate what we perceive to be one of the main limitations
 of current off-the-shelf planners for our applications: they often spend a long
@@
 C$^3$ \citep{lipovetzky08:_c3} planners. Each of these additional planners
-competed in the deterministic ``sequential, satisfying'' track of the 2008
+competed in the deterministic ``sequential, satisficing'' track of the 2008
 International Planning Competition; all planners performed well on the
@@
 
-Preprocessing (parsing, grounding, etc.) time generally plays less of a role in
-GIVE, compared with the sentence generation domain, however, its effects still
+Preprocessing time (parsing, grounding, etc.) generally plays less of a role in
+GIVE, compared with the sentence generation domain; however, its effects still
 contribute significantly to the overall running time of a number of planners.
@@
 domain, SGPLAN easily solves inputs beyond $w=40$ in well less than a second.
-LAMA is also reasonably successful on these problems, however, its runtimes grow
+LAMA is also reasonably successful on these problems; however, its runtimes grow
 more quickly than SGPLAN, with LAMA taking almost 5 seconds to solve the $w=40$

### 2009-07-30 tags/ant-crisp/Papers/ci-spark/nlg-as-planning.tex
Tagged the last version of CRISP with Ant buildfile and directory structure before converting to Maven.

--- 2009-05-31 trunk/Papers/ci-spark/nlg-as-planning.tex
@@
 Viewing generation as a planning problem has a long tradition in the NLG
-literature. \citet{perrault80} presented an approach to discourse planning in
-which the planning operators represented individual speech acts such as
+literature. \citet{perrault80} presented an approach to discourse planning
+in which the planning operators represented individual speech acts such as
 ``request'' and ``inform''. This idea was later expanded, e.g., by
 \citet{young94dpocl}. On the other hand, researchers such as
-\citet{appelt:planning} and \citet{hovy88} used techniques from hierarchical
-planning to expand a high-level plan consisting of speech acts into more
-detailed specifications of individual sentences. Their systems also covered some
-aspects of sentence planning. However, these systems used very expressive
-logics, which were designed to reason about beliefs and intentions, to represent
-the planning state and the planning operators. Furthermore, most of them used
-ad-hoc planning algorithms with rather naive search strategies, which did not
-scale to realistic inputs. As a consequence, the NLG-as-planning approach was
-mostly marginalized throughout the 1990s.
+\citet{appelt:planning} and \citet{hovy88} used techniques from
+hierarchical planning to expand a high-level plan consisting of speech acts
+into more detailed specifications of individual sentences. Although these
+systems covered some aspects of sentence planning, they also used very
+expressive logics designed to reason about beliefs and intentions, in order
+to represent the planning state and the planning operators. Most of these
+systems also used ad-hoc planning algorithms with rather naive search
+strategies, which did not scale well to realistic inputs. As a consequence,
+the NLG-as-planning approach was mostly marginalized throughout the 1990s.
 
-More recently, there has been a string of publications by various authors with a
-renewed interest in the generation-as-planning approach, motivated by the
-ongoing development of increasingly more efficient and expressive planners. For
-instance, \cite{KolSto07} propose an approach to sentence generation (i.e., the
-sentence planning and surface realization modules of the pipeline) as
-planning---an approach we will explore in more detail below
+More recently, there has been a string of publications by various authors
+with a renewed interest in the generation-as-planning approach, motivated
+by the ongoing development of increasingly more efficient and expressive
+planners. For instance, \cite{KolSto07} propose an approach to sentence
+generation (i.e., the sentence planning and surface realization modules of
+the pipeline) as planning---an approach we explore in more detail below
 (Section~\ref{sec:domain-crisp}). \citet{Steedman-Petrick:07} revisit the
-analysis of indirect speech acts with modern planning technology, viewing the
-problem as an instance of the general problem of planning with incomplete
-information and sensing actions. In addition, \citet{benotti08b} uses planning
-to explain the accommodation of presuppositions, and
-\citet{brenner08:_contin_multiag_plann_approac_to_situat_dialog} use multi-agent
-planning to model the joint problem-solving behaviour of agents in a situated
-dialogue. While these approaches focus on different issues compared to the
-1980's NLG-as-planning literature, they all apply \emph{existing},
-well-understood planning approaches to linguistic problems, in order to utilise
-the rich set of modelling tools provided by modern planners, and in the hope
-that such planners can efficiently solve the hard search problems that arise in
-NLG \citep{KolStr02}. This paper aims to investigate whether existing planners
-achieve this latter goal.
+analysis of indirect speech acts with modern planning technology, viewing
+the problem as an instance of planning with incomplete information and
+sensing actions. In addition, \citet{benotti08b} uses planning to explain
+the accommodation of presuppositions, and
+\citet{brenner08:_contin_multiag_plann_approac_to_situat_dialog} use
+multi-agent planning to model the joint problem solving behaviour of agents
+in a situated dialogue. While these approaches focus on different issues
+compared to the 1980's NLG-as-planning literature, they all apply
+\emph{existing}, well-understood planning approaches to linguistic
+problems, in order to utilise the rich set of modelling tools provided by
+modern planners, and in the hope that such planners can efficiently solve
+the hard search problems that arise in NLG \citep{KolStr02}. This paper
+aims to investigate whether existing planners achieve this latter goal.
 

### 2009-07-30 tags/ant-crisp/Papers/dl-gre-jair/conclusion.tex
Tagged the last version of CRISP with Ant buildfile and directory structure before converting to Maven.

--- 2008-05-16 trunk/Papers/dl-gre-08/conclusion.tex
@@
 non-relational and 67\% of the relational REs from
-\newcite{viethen06:_algor_for_gener_refer_expres}.  Both algorithms
+\citeA{viethen06:_algor_for_gener_refer_expres}.  Both algorithms
 are extremely efficient (350 ms and 140 ms respectively to generate
@@
 model changes. Similarly, it would be interesting to explore how
-linguistic constraints and different attribute orderings can be taken
-into account without slowing the algorithms down, how our algorithms
-could be integrated with more standard DL T-Box inferences, and how the
-algorithms can be adapted to use inverse relations or to compute REs
+different linguistic constraints and attribute orderings can be taken
+into account efficiently, how our algorithms
+could be integrated with more standard DL T-Box inferences, and how they
+ can be adapted to use inverse relations or to compute REs
 for sets. In exploring these extensions we will be able to draw on a
-rich body of literature that considers many variants of simulation
-algorithms.
+rich body of literature that has already considered many variants of simulation
+algorithms addressing similar issues.
 
@@
 
-\begin{small}
-\paragraph{Acknowledgments.} We are grateful to Hector Geffner (who
-independently suggested to view GRE as computation of DL formulas),
-Kees van Deemter, and Emiel Krahmer for interesting discussions.  We
-also thank Jette Viethen and Robert Dale for making their corpus
-available, and the reviewers for their comments.
-\end{small}
-
 %%% Local Variables: 

### 2009-07-30 tags/ant-crisp/Papers/dl-gre-jair/introduction.tex
Tagged the last version of CRISP with Ant buildfile and directory structure before converting to Maven.

--- 2008-05-16 trunk/Papers/dl-gre-08/introduction.tex
@@
 Referring expressions (REs) involving relations, in particular, have
-received increasing attention recently, specially in the context of
-spatial referring expressions in situated generation (e.g.\
-\cite{kelleher06:_increm_gener_of_spatial_refer}), where it seems
+received increasing attention recently; especially in the context of
+spatial referring expressions in situated generation
+\cite<e.g.,>{kelleher06:_increm_gener_of_spatial_refer}, where it seems
 particularly natural to use expressions such as ``the book on the
 table''.  However, the classical algorithm by
-\newcite{dale91:_gener_refer_expres_invol_relat} was recently shown to
+\citeA{dale91:_gener_refer_expres_invol_relat} was recently shown to
 be unable to generate satisfying REs in practice
 \cite{viethen06:_algor_for_gener_refer_expres}.  Furthermore, the Dale
-and Haddock algorithm and most of its successors (such as
-\cite{kelleher06:_increm_gener_of_spatial_refer}) are vulnerable to
-the problem of ``infinite regress'', where the algorithm jumps back
-and forth between generating descriptions for two related individuals
+and Haddock algorithm and most of its successors \cite<such
+as>{kelleher06:_increm_gener_of_spatial_refer} are vulnerable to the
+problem of ``infinite regress'', where the algorithm jumps back and
+forth between generating descriptions for two related individuals
 infinitely, as in ``the book on the table which supports a book on the
@@
 compute purely conjunctive formulas;
-\newcite{deemter02:_gener_refer_expres} extends this language by
+\citeA{deemter02:_gener_refer_expres} extends this language by
 adding the other propositional connectives, whereas
-\newcite{dale91:_gener_refer_expres_invol_relat} extends it by
+\citeA{dale91:_gener_refer_expres_invol_relat} extends it by
 allowing existential quantification.
@@
 \el\ algorithm is capable of generating 67\% of the relational REs in
-the \newcite{viethen06:_algor_for_gener_refer_expres} dataset, in about
+the \citeA{viethen06:_algor_for_gener_refer_expres} dataset, in about
 15 milliseconds.  The \alc\ algorithm is even faster; it computes

### 2009-07-30 tags/ant-crisp/Papers/experiences/conclusion.tex
Tagged the last version of CRISP with Ant buildfile and directory structure before converting to Maven.

--- 2008-07-25 trunk/Papers/experiences/conclusion.tex
@@
 opportunity for planning researchers.  While GIVE highlights
-planning-related challenges such as plan execution monitoring and plan
-presentation (i.e., summarisation and elaboration), it also offers a
-platform on which such technologies can be evaluated in experiments with
-human users.  Furthermore, more mainstream research on reasoning under
-uncertainty, planning with incomplete information and sensing, and resource
-management can also be applied in this domain setting. Provided some of the
-challenges we have highlighted can be addressed, projects like GIVE (and
-other NLG-inspired problem domains) offer a constructive platform for the
+certain challenges such as plan execution monitoring and plan presentation
+(i.e., summarisation and elaboration), it also offers a platform on which
+such technologies can be evaluated in experiments with human users.
+Furthermore, research related to reasoning under uncertainty, resource
+management, and planning with knowledge and sensing, can also be
+investigated in this domain setting. Provided some of the challenges we
+have highlighted can be addressed, projects like GIVE (and other
+NLG-inspired problem domains) offer a constructive platform for the
 planning community to showcase their techniques to a wider audience---and

### 2009-07-30 tags/ant-crisp/Papers/experiences/domains.tex
Tagged the last version of CRISP with Ant buildfile and directory structure before converting to Maven.

--- 2008-07-25 trunk/Papers/experiences/domains.tex
@@
 both for the NLG and the planning communities.  On the NLG side, one
-major problem is that the natural language instruction sequences can't
-simply be obtained by verbalising each action instance in a plan in turn.
-For instance, the NLG system could generate an instruction sequence
-starting with ``turn left; walk forward; press the button; turn right; walk
-forward; walk forward; turn right; walk forward; walk forward; turn left;
-walk forward''.  However, such directions would be clumsy and boring, and
+major problem is that verbalising each action instance in the plan in
+turn makes for very poor natural-language instructions.  For instance,
+the NLG system could generate an instruction sequence starting with
+``turn left; walk forward; press the button; turn right; walk forward;
+walk forward; turn right; walk forward; walk forward; turn left; walk
+forward''.  However, such directions would be clumsy and boring, and
 it is to be expected that frustrated users will quickly cancel an
@@
 \emph{summarised} by merging multiple action instances into a single
-instruction, in order to present them to the
-user.\footnote{This task is similar to the construction of ``macro''
-actions, but is primarily intended for plan presentation.}
-Conversely, it may also be necessary to \emph{elaborate} on a single
-planning step by expressing it with several instructions.  Having just
-entered the top left room, it may be easier for the user to understand the
-instruction ``walk to the centre of the room; turn right; now press the
-green button in front of you'', rather than the instruction ``press the
-green button on the wall to your right''.  To plan the referring expression
-``the green button in front of you'' at a time when the button is not yet
-in front of the user, the NLG system must keep track of the hypothetical
-changes in what objects are visible to the user.  Thus the acts of
-referring and instructing, and the modules for planning and NLG, must be
-tightly integrated.
+instruction, in order to present them to the user.\footnote{This task
+  is similar to the construction of ``macro'' actions, but is
+  primarily intended for plan presentation.}  Conversely, it may also
+be necessary to \emph{elaborate} on a single planning step by
+expressing it with several instructions.  Having just entered the top
+left room, it may be easier for the user to understand the instruction
+``walk to the centre of the room; turn right; now press the green
+button in front of you'', rather than the instruction ``press the
+green button on the wall to your right''.  To plan the referring
+expression ``the green button in front of you'' at a time when the
+button is not yet in front of the user, the NLG system must keep track
+of the hypothetical changes in what objects are visible to the user.
+Thus the acts of referring and instructing, and the modules for
+planning and NLG, must be tightly integrated.
 
-From a planning perspective, GIVE imposes very strict runtime requirements
-on the planner and its associated modules: planning must happen in real
-time and the system must respond to a user in a timely fashion.  If the
-system takes too much time deliberating over an instruction to give, rather
-than actually giving this instruction, the user may have walked or turned
-away, thus making the instruction invalid.  Furthermore, plan execution
-monitoring also plays an important role in the GIVE problem.  At a high
-level, the system needs to monitor a user's actions and compare them
-against the generated instruction set to determine if the user has
-correctly followed directions or not. In the case of the latter, new
-instructions may have to be generated. In practice, the situation can be
-quite complicated since the mental state of the user is not known and so
-the system must observe the user's actions in real time. For example, a
-user directed to ``turn around and walk through the door'' may not
-necessarily perform these actions to the letter, i.e., immediately turning
-180 degrees and proceeding directly to the door. Instead, the user might
-take a roundabout route through the room, eventually exiting out the door.
-Although the user's actions do not match the generated instructions
-exactly, they meet the intended goal. The system must be able to identify
-such ``equivalent'' plans and not immediately generate new instructions as
-soon as the user's actions have gone off course. Furthermore, a user can
-communicate certain intentions to the system, both through action and
-inaction. For instance, the system should infer that a user has failed to
-follow instructions if the user exits a room when given a directive to
-``walk to the centre of the room''. The system should also make a
+From a planning perspective, GIVE imposes very strict runtime
+requirements on the planner and its associated modules: planning must
+happen in real time and the system must respond to a user in a timely
+fashion.  If the system takes too much time deliberating over an
+instruction to give, rather than actually giving this instruction, the
+user may have walked or turned away, thus making the instruction
+invalid.  Furthermore, plan execution monitoring also plays an
+important role in the GIVE problem.  At a high level, the system needs
+to monitor a user's actions and compare them against the generated
+instruction set to determine if the user has correctly followed
+directions or not. In the case of the latter, new instructions may
+have to be generated. In practice, the situation can be quite
+complicated since the mental state of the user is not known and so the
+system must be inferred from observing the user's actions in real
+time. For example, a user directed to ``turn around and walk through
+the door'' may not necessarily perform these actions to the letter,
+i.e., immediately turning 180 degrees and proceeding directly to the
+door. Instead, the user might take a roundabout route through the
+room, eventually exiting out the door.  Although the user's actions do
+not match the generated instructions exactly, they meet the intended
+goal. The system must be able to identify such ``equivalent'' plans
+and not immediately generate new instructions as soon as the user's
+actions have gone off course. Furthermore, a user can communicate
+certain intentions to the system, both through action and
+inaction. For instance, the system should infer that a user has failed
+to follow instructions if the user exits a room when given a directive
+to ``walk to the centre of the room''. The system should also make a
 similar conclusion if a user simply does nothing when given the

### 2009-07-30 tags/ant-crisp/Papers/rmrs-08/old/motivation.tex
Tagged the last version of CRISP with Ant buildfile and directory structure before converting to Maven.

% Motivate what we're doing and provide background.  Example outputs of
% deep and shallow parsers.  Illustrate and state the problem.
%
% Try to find an eg where the comparison isn't easy to see (from Anna
% Ritchie's deliverable?). 
% NOTE: on Ritichie: Perhaps "Three dogs bark on Tuesday in
% June" from page 11?  It's hard to follow what's going on, from the
% sheer size of the RMRSs and the need to allow for variation on the
% variable names.  But I don't know how to extract this figure
% techinically; nor whether we have room.

We motivate and describe {\sc rmrs} with a toy
example.  Sentence 
(\ref{eg:cat}a) features a semantic scope ambiguity that, according to
most deep grammars, syntax
doesn't resolve.    The {\sc mrs} (\ref{eg:cat}b) derived by the
{\sc erg} \cite{copestake:flickinger:2000}
reflects this: (i) each predication is
labelled ($l_1$, $l_4$, etc); (ii) scopal arguments are indicated by
holes (e.g., $h_2$, $h_3$); (iii)
constraints on scope are expressed by the qeq conditions (roughly, $h
=_q l$ means that only zero or more quantifiers lie between the scopal
argument $h$ and the predication labelled by $l$ in any fully-resolved
logical form); and (iv) these scope
constraints allow for more than one way of `plugging' the holes with
labels.\footnote{The event variable in $\_\mbox{{\em fat}}\_j\_1$ is 
  for copulas. {\em The
    cat is fat} has logical form 
  $\mbox{\_the\_q\_1}(x,\mbox{\_cat\_n\_1}(x),\mbox{\_fat\_j\_1}(e,x))$.}
\begin{examples}
\item   \label{eg:cat}
\begin{subexamples}
\item
Every fat cat chased some dog.
% \item
% $\mbox{\_every\_q\_1}(x, \mbox{\_fat\_j\_1}(e',x) \wedge
% \mbox{\_cat\_n\_1}(x),$\\
% \hspace*{0.5in}$\mbox{\_some\_q\_1}(y, \mbox{\_dog\_n\_1}(y),
% \mbox{\_chase\_v\_1}(e_{spast},x) )$\\
% $\mbox{\_some\_q\_1}(y, \mbox{\_dog\_n\_1}(y),$\\
% \hspace*{0.5in}$\mbox{\_every\_q\_1}(x, \mbox{\_fat\_j\_1}(e',x)
% \wedge 
% \mbox{\_cat\_n\_1}(x), 
% \mbox{\_chase\_v\_1}(e_{spast},x)))$
\item   $l_1\handel\mbox{\_every\_q\_1}(x,h_2,h_3)$\\
$l_4\handel\mbox{\_fat\_j\_1}(e',x)$\\
$l_4\handel\mbox{\_cat\_n\_1}(x)$\\
$l_5\handel\mbox{\_chased\_n\_1}(e,x,y)$\\
$l_6\handel\mbox{\_some\_n\_1}(y,h_7,h_8)$\\
$l_9\handel\mbox{\_dog\_n\_1}(y)$\\
$h_2 =_q l_4, h_7 =_q l_9$
\end{subexamples}
\end{examples}
(\ref{eg:cat}b) also exhibits {\sc mrs}' compact description of
conjunction through label sharing (see $l_4$), and the {\sc
  erg}'s naming convention for predicate symbols: each lexical predicate is
constructed from the word lemma, its syntactic category and a sense
number, and the leading underscore distinguishes lexical predicates
from non-lexical ones.

Conceptually, this {\sc mrs} is a partial description of all its
possible (fully specific) interpretations in context.  
These are expressed in a so-called {\em base language}, where each
predicate symbol in the {\sc mrs} is a proxy for one or more
base-language constructors of the same arity.
Assuming a very close
correspondence between the {\sc mrs} predicate symbols and the
base-language constructors they 
stand for, the {\sc mrs}
(\ref{eg:cat}b) (partially) describes both the base-language formulae in
Figure~\ref{fig:1}, shown as trees to emphasise that an {\sc mrs} is a
partial description of the formulae's {\em form}; not its
(base-language) entailments. 
In essence, the labels in the {\sc mrs} reify the
nodes in the tree, and each of these trees encodes one way of equating
holes with labels that respects qeq constraints and binding conditions
on the variables.

The {\sc mrs} (\ref{eg:cat}b) is a notational variant of the {\sc
  rmrs} (\ref{eg:cat}c). {\sc rmrs}s offer a more factorised
representation: each predicate symbol is unary and the other
arguments to its corresonding base-language constructor
are represented by separate binary relations on the unique {\em
  anchor} of the predicate symbol ($a_1, a_{41},\ldots$) together with
variable equalities (e.g., $x=x_1$).
\begin{examples}
\item   [\ref{eg:cat}]
\begin{subexamples}
\item   [c]

\end{subexamples}
\end{examples}

This factored representation allows one to build semantic components
to shallow parsers where lexical or syntactic information is absent.
An extreme example would be a {\sc 
  pos} tagger: one can build its semantic component simply by deriving
lexical predicate symbols from the word lemmas and tags and assigning each
predication a unique label, anchor and
argument, as in (\ref{eg:cat-pos}):
\begin{examples}
\item   \label{eg:cat-pos}
\begin{subexamples}
\item   Every\_{\sc at1} fat\_{\sc jj} cat\_{\sc nn1} chased\_{\sc
    vvd} some\_{\sc dd} dog\_{\sc nn1}
\item   $l_1:a_1:\mbox{\_every\_q}(x)$, \\
$l_{41}:a_{41}:\mbox{\_fat\_a}(e')$,\\
$l_{42}:a_{42}:\mbox{\_cat\_n}(x_3)$\\
$l_5:a_5:\mbox{\_chased\_v}(e_{past})$, \\
$l_6:a_6:\mbox{\_some\_q}(x_4)$, \\
$l_9:a_9:\mbox{\_dog\_n}(x_5)$
\end{subexamples}
\end{examples}

Semantic relations, sense tags and the arity of the base-language
constructors are
missing from (\ref{eg:cat-pos}b) because the {\sc pos} tagger doesn't
reveal information about syntactic constituency, word sense or lexical
subcategorisation respectively.  
But the {\sc rmrs}s (\ref{eg:cat}c) and
(\ref{eg:cat-pos}b) are entirely compatible, the former being more
specific than the latter.  At least, this holds so long as the
base-language constructors that the {\sc rmrs} predicate symbol
{\_name\_t\_s} corresponds to is a 
subset of those corresponding to the predicate symbol {\em \_name\_t},
for any word lemma {\em name}, {\sc pos} tag $t$ and sense $s$ (written 
$\mbox{{\em \_name\_t\_s}} \sqsubset \mbox{{\em name\_t}}$).

The values of non-scopal arguments are underspecified through the
absence of ARG relations and/or through the absence of
equalities and inequalities among the variables $x_1, x_2,\ldots$.
The need to
underspecify such semantic dependencies is required when
the shallow parser doesn't record syntactic dependencies.
Indeed, there is no single {\sc mrs} which is satisfied by both the
trees in Figure~\ref{fig:2}, assuming of course that disjunction is not part of
the {\sc mrs} language.  In words, these trees collectively specify
the (partial) semantic information that $x$ is either the second
argument (i.e., the direct object) or the third argument (i.e., the
indirect object) to {\em give}, but we don't know which.  
A shallow
processor that misses long distance dependencies in
(\ref{eg:which-dog}) cannot
discern that the bone is the second argument to {\em
  give} in (\ref{eg:which-dog}a), while the dog is the third argument
in (\ref{eg:which-dog}b).
\begin{examples}
\item   \label{eg:which-dog}
\begin{subexamples}
\item   To which dog did Kim give the bone?
\item   Which bone did Kim give the dog?
\end{subexamples}
\end{examples}
Thus to maximise the semantic contribution
of the parser while not overdetermining it, the phrases
{\em give the bone} vs.\
{\em give the dog} should whether the semantic index of the noun
phrase is the second or third argument to {\em give},
as shown in Figure~\ref{fig:2}.   This will be expressed in
{\sc rmrs} via the binary relation  $\mbox{{\em ARG}}_{\{2,3\}}(a,x)$
between the anchor $a$ of the predicate symbol for {\em give} and the
individual $x$.   More generally, $\mbox{{\em ARG}}_n(a,x)$ means that
$x$ is an argument to the predicate anchored at $a$, but its argument
position is unspecified.

\begin{figure}

\leaf{$e$}
\leaf{$\ldots$}
\leaf{$x$}
\leaf{$\ldots$}
\branch{4}{\_give\_v\_b1}
\tree
\hfill
\leaf{$e$}
\leaf{$\ldots$}
\leaf{$\ldots$}
\leaf{$x$}
\branch{4}{\_give\_v\_b1}
\tree

\caption{$x$ is a participant in a
  giving event, but $x$'s role is unknown.}
\label{fig:2}
\end{figure}

The binary ARG relations also supports underspecifying the {\em arity}
of the base-language constructors, as is required when lexical
subcategorisation information is missing from the shallow parser.
While a deep parser assumes complete information about lexical
subcategorisation in any of its derivations, thereby fixing the arity
of the base-language constructors in the fully-specific logical form
to be the same as the {\sc mrs} predicate symbol $P$ that's introduced by
the lexeme, the predicate symbols in an {\sc rmrs} yielded from a
shallow parse should, in principle at least,
correspond to base language constructors of mixed arity.  For
instance, the English verb {\em bake} can be used causatively ({\em
  Kim baked the potatoes}), in which case semantically it corresponds
to a 3-place predicate symbol {\em \_bake\_v\_b1} in the base
language, or it may be used 
inchoatively ({\em the potatoes are baking}), that semantically
corresponds to a 2-place predicate {\em \_bake\_v\_b2}.  So the 
predicate symbol {\em 
  \_bake\_v} in the {\sc rmrs} that's introduced by a {\sc pos} tagger
must be able to denote either of these
base-language predicates in the {\sc rmrs}'s interpretation.

\hidden{
Finally, {\sc rmrs} allows one to underspecify the {\em sort} of an
argument by including in its language meta-variables $u_1,u_2,\ldots$
over individual, event, and handle variables.  For instance, the predicate
symbol {\em \_be\_v} produced by a {\sc pos} tagger is resolvable to several
3-place base-language constructors whose argument-types 
vary: the base-language constructor corresponding
to its use in {\em John is a man} takes an event and two individuals
as arguments; but in {\em It is John who talked} the constructor takes
an event, an individual and a proposition as arguments.
}

### 2009-07-30 tags/ant-crisp/Papers/rmrs-08/rmrs.tex
Tagged the last version of CRISP with Ant buildfile and directory structure before converting to Maven.

--- 2009-01-28 trunk/Papers/rmrs-08/rmrs.tex
@@
 finite constructor tree $\tau$ is defined in terms of several
-assignment functions.  First, a node variable assignment
-function $\alpha: \nvar \rightarrow D(\tau)$ maps the node
-variables in an \rmrs\ to the nodes of $\tau$.  Second, a
-base language assignment function $g:\ovar \rightarrow \Sigma_0$ 
-maps the base variables to nullary constructors
-representing variables in the base language. Finally, a
-function $\sigma$ from $\pred$ to the power set of $\Sigma_{\geq 1}$ maps
-each \rmrs\ predicate symbol to a set of constructors from $\Sigma$.
-As we'll see shortly, this relation allows an \rmrs\ to underspecify lexical
-ambiguities.
+assignment functions.  First, a node variable assignment function
+$\alpha: \nvar \rightarrow D(\tau)$ maps the node variables in an
+\rmrs\ to the nodes of $\tau$.  Second, a base language assignment
+function $g:\ovar \rightarrow \Sigma_0$ maps the base variables to
+nullary constructors representing variables in the base
+language. Finally, a function $\sigma$ from $\pred$ to the power set
+of $\Sigma_{\geq 1}$ maps each \rmrs\ predicate symbol to a set of
+constructors from $\Sigma$.  As we'll see shortly, this function
+allows an \rmrs\ to underspecify lexical ambiguities.
 

### 2009-12-09 trunk/Papers/crisp-ff-10/crisp-ff.tex
ICAPS submission for CRISP as challenge problem

\section{Experiences with FF}
\label{sec:crisp-ff}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2009-12-09 trunk/Papers/crisp-ff-10/main.tex
ICAPS submission for CRISP as challenge problem

\documentclass[letterpaper]{article}
\usepackage{aaai}

\usepackage{times}
\usepackage{helvet}
\usepackage{courier}

\pdfinfo{
/Title (Formatting Instructions for Authors)
/Subject (Proceedings of the AAAI Press Conference)
/Author (AAAI)}

\setcounter{secnumdepth}{0}

\begin{document}
\title{Challenge: Using Planning to Generate Natural-Language
  Sentences}
\author{J\"org Hoffmann \\ LORIA \\ joerg.hoffmann@loria.fr
\And
Alexander Koller \\ Saarland University \\ koller@mmci.uni-saarland.de
\And
Ronald Petrick \\ University of Edinburgh \\ rpetrick@inf.ed.ac.uk}
\maketitle

\input abstract
\input introduction
\input crisp
\input crisp-ff
\input conclusion

\bibliography{bibliography}
\bibliographystyle{aaai}

\end{document}

### 2009-12-09 trunk/Papers/reasoning-challenges/abstract.tex
ICAPS submission for Reasoning Challenges 2.0

\begin{abstract}
\begin{quote}
AAAI creates proceedings, working notes, and technical reports directly from electronic source furnished by the authors. To ensure that all papers in the publication have a uniform appearance, authors must adhere to the following 
instructions. 
\end{quote}
\end{abstract}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2009-12-09 trunk/Papers/reasoning-challenges/conclusion.tex
ICAPS submission for Reasoning Challenges 2.0

--- 2008-07-24 trunk/Papers/experiences/conclusion.tex
@@
 \section{Conclusion} \label{sec:conclusion}
-
-In this paper, we introduced two novel planning domains arising from
-problems in natural language generation: the sentence generation
-domain and the GIVE navigation domain. We also reported on the results
-of a number of experiments in which we applied off-the-shelf planners
-to these domains.
-
-Our results were mixed. While modern planners do a pretty good job of
-controlling the complexity of search, they also suffer from practical
-problems that limit their performance in unexpected ways. In
-particular, in both domains the grounding step performed by both
-planners dominates the time it takes to perform the search itself. In
-future work, it would be interesting to explore planners such as SHOP2
-\cite{DBLP:journals/jair/NauAIKMWY03}, which do not perform the
-initial grounding step.  We are also considering to propose some of
-our domains as IPC benchmarks.
-
-The NLG community's recent interest in planning presents a valuable
-opportunity for planning researchers: provided some of the challenges
-we have highlighted can be addressed, projects like GIVE (and other
-NLG-inspired problem domains) offer a constructive platform for the
-planning community to showcase their techniques to a wider
-audience---and to improve the quality of their tools for real-world
-planning tasks.  Furthermore, GIVE highlights planning-related
-challenges such as plan execution monitoring and plan presentation
-and summarisation and offers a platform on which such technologies can
-be evaluated in experiments with human users.
-
-\section{Acknowledgements}
-This work arose in the context of the Planning and Language Interest Group
-at the University of Edinburgh. We thank all members of this group,
-especially Hector Geffner and Mark Steedman, for interesting discussions.
-This work was partially supported by the DFG Research Fellowship ``CRISP:
-Efficient integrated realization and microplanning'' and by the European
-Commission through the PACO-PLUS project (FP6-2004-IST-4-27657).
-
-%\paragraph{Acknowledgments.} This work is a collaboration that arose
-%in the context of the ``planning and language'' group at the
-%University of Edinburgh.  We thank all members of this group,
-%especially Hector Geffner and Mark Steedman, for interesting
-%discussions.  This work was supported by the DFG Research Fellowship
-%``CRISP: Efficient integrated realization and microplanning'' and the
-%EU project XYZ PACO-PLUS.
 
@@
 %%% mode: latex
-%%% TeX-master: "experiences"
+%%% TeX-master: "main"
 %%% End:

### 2009-12-09 trunk/Papers/reasoning-challenges/introduction.tex
ICAPS submission for Reasoning Challenges 2.0

--- 2009-07-30 tags/ant-crisp/Papers/dl-gre-jair/introduction.tex
@@
 \section{Introduction} \label{sec:introduction}
-
-The generation of referring expressions (GRE) is one of the most
-active and successful research areas in natural language generation.
-Building upon Dale and Reiter's work \cite{dale89cooking,Dale1995},
-various researchers have added extensions such as reference to sets
-\cite{Stone2000}, more expressive logical connectives
-\cite{deemter02:_gener_refer_expres}, and relational expressions
-\cite{dale91:_gener_refer_expres_invol_relat}.
-
-Referring expressions (REs) involving relations, in particular, have
-received increasing attention recently; especially in the context of
-spatial referring expressions in situated generation
-\cite<e.g.,>{kelleher06:_increm_gener_of_spatial_refer}, where it seems
-particularly natural to use expressions such as ``the book on the
-table''.  However, the classical algorithm by
-\citeA{dale91:_gener_refer_expres_invol_relat} was recently shown to
-be unable to generate satisfying REs in practice
-\cite{viethen06:_algor_for_gener_refer_expres}.  Furthermore, the Dale
-and Haddock algorithm and most of its successors \cite<such
-as>{kelleher06:_increm_gener_of_spatial_refer} are vulnerable to the
-problem of ``infinite regress'', where the algorithm jumps back and
-forth between generating descriptions for two related individuals
-infinitely, as in ``the book on the table which supports a book on the
-table \ldots''.
-
-In this paper, we propose to view GRE as the problem of computing a
-formula of description logic (DL) that denotes exactly the set of
-individuals that we want to refer to.  This very natural idea has been
-mentioned in passing before
-\cite{Krahmer2003,gardent07:_gener_bridg_defin_descr}; however, we
-take it one step further by proposing DL as an interlingua for
-comparing the REs produced by different approaches to GRE.  In this
-way, we can organize existing GRE approaches in an expressiveness
-hierarchy.  For instance, the classical Dale and Reiter algorithms
-compute purely conjunctive formulas;
-\citeA{deemter02:_gener_refer_expres} extends this language by
-adding the other propositional connectives, whereas
-\citeA{dale91:_gener_refer_expres_invol_relat} extends it by
-allowing existential quantification.
-
-Furthermore, the view of GRE as a problem of computing DL formulas
-with a given extension allows us to apply existing algorithms for the
-latter problem to obtain efficient algorithms for GRE.  We present
-algorithms that compute such formulas for the description logics \el\
-(which allows only conjunction and existential quantification) and
-\alc\ (which also allows negation).  These algorithms effectively
-compute REs for all individuals in the domain at the same time, which
-allows them to systematically avoid the infinite regress problem.  The
-\el\ algorithm is capable of generating 67\% of the relational REs in
-the \citeA{viethen06:_algor_for_gener_refer_expres} dataset, in about
-15 milliseconds.  The \alc\ algorithm is even faster; it computes
-relational REs for all 100 individuals in a random model in 140
-milliseconds.
-
-The paper is structured as follows.  In Section~\ref{sec:bisim}, we
-will first define description logics.  We will then show how to
-generate REs by computing DL similarity sets for \alc\ and \el\ in
-Section~\ref{sec:gre}.  In Section~\ref{sec:discussion}, we evaluate
-our algorithms and discuss our results.  Section~\ref{sec:related}
-compares our approach to related research; in particular, it shows how
-various prominent GRE algorithms fit into the DL framework.
-Section~\ref{sec:conclusion} concludes and points to future work.
 
@@
 %%% mode: latex
-%%% TeX-master: "dl-gre-08"
+%%% TeX-master: "main"
 %%% End:

### 2009-12-10 trunk/Papers/crisp-ff-10/crisp-ff.tex
some initial stuff

@@
 \label{sec:crisp-ff}
+
+Things that went wrong with FF:
+
+\begin{enumerate}
+\item Enforced hillclimbing: Initial state of example problem has a
+  relaxed graph of length one because ``expressed'' and ``subst''
+  goals are achieved, and newly introduced violations of ``subst'' and
+  ``distractor'' are not noticed. As a consequence, EHC on the second
+  state spends all of its time in breadth-first search.
+\item Goal agenda: For some reason that I don't understand in detail,
+  the goal agenda heuristic can choose that the ``subst'' goal must be
+  reached before even the ``expressed'' goal. This leads the planner
+  to generating some arbitrary sentence. Because the example grammar
+  is so small, this leads to a state from which the ``expressed'' goal
+  can no longer be achieved. (A realistic big grammar should be able
+  to recover from this by introducing a conjunction or relative clause
+  or some such -- but of course, this would still lead to a stupid
+  sentence.)
+\item No helpful actions in best-first search: After the initial
+  failure of the goal agenda, FF fell back to best-first search, but
+  didn't use the helpful actions heuristic in this search.
+\item Tautology detection in reachability analysis: The old FF
+  performed a pairwise comparison of the ground literals in the
+  instantiated goal to remove tautologies and obvious contradictions.
+  In the sentence generation problems, the instantiated goal can
+  contain lots of literals, so this takes a long time.  Removing the
+  check helped a lot.  Another way to fix this problem would have been
+  to sort the list of ground literals appropriately and then only
+  compare neighboring literals.
+\item Unnecessary computation of hard action template data structure:
+  FF computed a big data structure that was only useful for hard
+  action templates, although the sentence generation planning problems
+  only contain easy action templates.  Because the universe can be
+  relatively big, this again wasted a lot of time.
+\end{enumerate}
+
+Other observations:
+
+\begin{enumerate}
+\item Subst and distractor atoms have a weird lifecycle: They start as
+  false, then get made true at some point, and then false again.
+  After they have been made false, they can never become true again.
+\end{enumerate}
 

### 2009-12-10 trunk/Papers/crisp-ff-10/crisp.tex
ICAPS submission for CRISP as challenge problem
some initial stuff

--- 2009-12-09 trunk/Papers/crisp-ff-10/crisp-ff.tex
@@
-\section{Experiences with FF}
-\label{sec:crisp-ff}
+\section{Sentence generation as planning}
+\label{sec:crisp}
+
+\subsection{Sentence generation with tree-adjoining grammars}
+
+We consider a planning domain that comes up in the generation of
+natural language sentences. This is a standard problem in
+computational linguistics and can be construed as follows.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.75\columnwidth]{pic-grammar}
+  \caption{An example grammar in the sentence generation domain.}
+  \label{fig:white-rabbit-sleeps-grammar}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.75\columnwidth]{pic-derivation}
+  \caption{Derivation of ``The white rabbit sleeps.''}
+  \label{fig:white-rabbit-sleeps-deriv}
+\end{figure}
+
+We assume a tree-adjoining grammar (TAG; \cite{joshi;etal1997}), which
+consists of a finite set of \emph{elementary trees} as shown in
+Fig.~\ref{fig:white-rabbit-sleeps-grammar}. Elementary trees can be
+combined using the operations of \emph{substitution} and
+\emph{adjunction}, which are illustrated in
+Fig.~\ref{fig:white-rabbit-sleeps-deriv}. Substitution replaces a
+substitution node (marked with a downarrow) by a subtree whose root
+has the same label, whereas adjunction splits a node in two and
+replaces it with a subtree whose root and foot node (marked with an
+asterisk) have the same label. A TAG derivation starts with a single
+substitution node labeled with a start symbol (in the example, S) and
+terminates when all leaves of the tree are labeled with words. In this
+way, a TAG grammar defines a language of trees, from which we can read
+off a language of strings.
+
+To model the sentence generation problem, we equip every elementary
+tree with a set of atoms that specify the \emph{semantic
+  representation} of the tree \cite{Stone2003a}.  These terms in these
+atoms may be \emph{semantic roles}, such as $\mathsf{subj}$ and
+$\mathsf{obj}$ indicating the subject and object positions of a verb;
+all substitution nodes of an elementary tree must and the other nodes
+may be annotated with semantic roles as well.  In an \emph{instance}
+of an elementary tree, all semantic roles have been substituted with
+constants from the knowledge base.  Now we can assume a TAG grammar
+and two finite sets of ground atoms for the knowledge base and the
+semantic representation we want to express in a sentence.  The
+sentence generation problem is then to compute a correct TAG
+derivation using elementary tree instances such that the input
+semantic representation is covered by the semantic representations of
+the tree instances; that is, the derivation expresses the intended
+meaning.  Furthermore, the assignment of constants to the semantic
+roles of the elementary trees must be \emph{unique}: If we replace
+each semantic role of every elementary tree in the derivation by a
+different variable, there must be exactly one variable assignment that
+makes the combined semantic representations of the elementary trees a
+subset of the knowledge base.  This ensures that the hearer can
+uniquely understand each referring expression.
+
+For example, assume the grammar in
+Fig.~\ref{fig:white-rabbit-sleeps-grammar} and a knowledge base that
+contains the atoms $\{\mathsf{sleep}(e,r_1), \mathsf{rabbit}(r_1),
+\mathsf{rabbit}(r_2), \mathsf{white}(r_1), \mathsf{rabbit}(r_2)\}$,
+and let's say we want to express the semantic representation
+$\{\mathsf{sleep}(e,r_1)\}$.  Then the derivation in
+Fig.~\ref{fig:white-rabbit-sleeps-deriv} would be a correct output of
+a sentence generator.  Notice that a derivation for the sentence ``the
+rabbit sleeps'' would not be a correct result because the expression
+``the rabbit'' is not a unique description of $r_1$: the set
+$\{\mathsf{sleep}(x,y), \mathsf{rabbit}(y)\}$ can also be mapped into
+the knowledge base by substituting $r_2$ for $y$, so the hearer might
+misunderstand the subject as $r_2$. We call $r_2$ a \emph{distractor}
+of the referring expression.  Note also that the input semantic
+representation did not specify that we should refer to $r_1$ as
+``white'' or ``rabbit''; this decision was made by the generation
+algorithm.
+
+\subsection{Sentence generation as a planning problem}
+
+The TAG sentence generation problem can be encoded as a planning
+problem \cite{KolSto07}.  The key idea is that each operator encodes
+the addition of an elementary tree to the derivation; the syntactic
+and semantic requirements and effects of doing this are captured in
+the operator's precondition and effect.
+
+\begin{figure}
+\centering
+\begin{minipage}{0.8\textwidth}
+{\small%
+\begin{verbatim}
+(:action sleeps
+   :parameters (?u - node  ?subjnode - node  ?nextnode - node
+                ?xself - individual  ?xsubj - individual)
+   :precondition
+       (and (subst S ?u)  (referent ?u ?xself)  (sleep ?xself ?xsubj)
+            (current ?subjnode) (next ?subjnode ?nextnode)
+   :effect 
+       (and (not (subst S ?u))  (expressed sleep ?xself ?xsubj)
+            (subst NP ?subjnode)  (referent ?subjnode ?xsubj)
+            (not (next ?subjnode)) (next ?nextnode)
+            (forall (?y - individual)
+                (when (not (= ?y ?xself)) (distractor ?subjnode ?y)))))
+
+(:action rabbit
+   :parameters (?u - node  ?xself - individual)
+   :precondition 
+       (and (subst NP ?u)  (referent ?u ?xself)  (rabbit ?xself))
+   :effect 
+       (and (not (subst NP ?u))  (canadjoin N ?u)
+            (forall (?y - individual)
+                (when (not (rabbit ?y)) (not (distractor ?u ?y))))))
+
+(:action white
+   :parameters (?u - node  ?xself - individual)
+   :precondition 
+       (and (canadjoin N ?u)  (referent ?u ?xself)  (rabbit ?xself))
+   :effect 
+       (forall (?y - individual)
+           (when (not (white ?y)) (not (distractor ?u ?y)))))
+\end{verbatim}}%
+\end{minipage}
+\caption{PDDL actions for generating the sentence ``The white rabbit
+sleeps.''}
+\label{fig:white-rabbit-as-planning}
+\end{figure}
+
+More precisely, each operator has a parameter \verb|?u| representing
+the syntax node into which the elementary tree is substituted or
+adjoined and a parameter \verb|?xR| for each semantic role \verb|R| of
+the elementary tree which will be instantiated with the constants with
+which the role gets instantiated in the derivation. Furthermore, if
+the tree has substitution nodes, there are parameters \verb|?Rnode|
+for each role representing these nodes, and an extra parameter
+\verb|?nextnode|. The state encodes a list of possible node names;
+$\mathsf{current}(n_i)$ expresses that $n_i$ is the next node name
+that should be picked when a new substitution node is created, and
+$\mathsf{next}(n_i,n_{i+1})$ says that the next node after $n_i$ is
+$n_{i+1}$. The atom $\mathsf{subst}(A,u)$ expresses that we need to
+still substitute something into the substitution node $u$ with label
+$A$; the initial state contains an atom
+$\mathsf{subst}(A,\mathsf{root})$ where $\mathsf{root}$ is an
+arbitrary name for the root of the derivation.  The values of the
+semantic roles are maintained using the $\mathsf{referent}$ atoms.
+Finally, we keep track of the uniqueness of referring expressions in
+the $\mathsf{distractor}$ atoms: $\mathsf{distractor}(u,a)$ expresses
+that the expression at syntax node $u$ could still be misunderstood as
+$a$.
+
+The example generation problem above translates into a planning
+problem whose domain is shown, in PDDL format, in
+Fig.~\ref{fig:white-rabbit-as-planning}. The initial state of the
+problem encodes the generation problem's knowledge base; it contains
+atoms $\mathsf{rabbit}(r_1)$, $\mathsf{sleep}(e,r_1)$, etc.
+Furthermore, the initial state expresses the beginning of the
+derivation in atoms $\mathsf{subst}(S,\mathsf{root})$ and
+$\mathsf{referent}(\mathsf{root},e)$.  The goal requires syntactic
+completeness as $\forall u \forall A \neg \mathsf{subst}(A,u)$ and
+unique reference as $\forall u \forall x \neg
+\mathsf{distractor}(u,x)$; it also specifies the semantic
+representation we want to express in an atom
+$\mathsf{expressed}(\mathsf{sleep},e,r_1)$.  A minimal plan for this
+problem looks as follows:
+
+\begin{enumerate}
+\item $\mathsf{sleeps}(\mathsf{root}, n_1, n_2, e, r_1)$
+\item $\mathsf{rabbit}(n_1, r_1)$
+\item $\mathsf{white}(n_1, r_1)$
+\end{enumerate}
+
+This plan can be automatically decoded into the derivation shown in
+Fig.~\ref{fig:white-rabbit-sleeps-deriv}.  The analogous plan with
+$r_2$ instead of $r_1$ is not correct because the $\mathsf{sleep}$
+precondition of the $\mathsf{sleeps}$ operator is not satisfied.  The
+first two steps of the plan alone would not be a correct plan because
+they leave an atom $\mathsf{distractor}(n_1, r_2)$ in the state, which
+contradicts the goal that all distractors have been eliminated.
 

### 2009-12-11 trunk/Papers/crisp-ff-10/crisp-ff.tex
title, structure, ff sec outline modifications

@@
-\section{Experiences with FF}
+\section{Why FF off-the-shelf doesn't work, and how to fix it (partially)}
 \label{sec:crisp-ff}
+
+\subsection{Pre-Processing}
+\label{sec:crisp-ff:preprocess}
 
@@
 \begin{enumerate}
-\item Enforced hillclimbing: Initial state of example problem has a
-  relaxed graph of length one because ``expressed'' and ``subst''
-  goals are achieved, and newly introduced violations of ``subst'' and
-  ``distractor'' are not noticed. As a consequence, EHC on the second
-  state spends all of its time in breadth-first search.
+\item Tautology detection in reachability analysis: The old FF
+  performed a pairwise comparison of the ground literals in the
+  instantiated goal to remove tautologies and obvious contradictions.
+  In the sentence generation problems, the instantiated goal can
+  contain lots of literals, so this takes a long time.  FIX: Removing
+  the check helped a lot.  Another way to fix this problem would have
+  been to sort the list of ground literals appropriately and then only
+  compare neighboring literals.
+\item Unnecessary computation of hard action template data structure:
+  FF computed a big data structure that was only useful for hard
+  action templates, although the sentence generation planning problems
+  only contain easy action templates.  Because the universe can be
+  relatively big, this again wasted a lot of time. FIX: just add this
+  tuiny bit of intelligence to FF to not do the data structure if it
+  is not needed. .. interesting that no-one had this problem before.
+\end{enumerate}
+
+\subsection{Heuristics and Search}
+\label{sec:crisp-ff:preprocess:heuristics}
+
+\begin{enumerate}
 \item Goal agenda: For some reason that I don't understand in detail,
-  the goal agenda heuristic can choose that the ``subst'' goal must be
+  the goal agenda heuristic chooses that the ``subst'' goal must be
   reached before even the ``expressed'' goal. This leads the planner
@@
   or some such -- but of course, this would still lead to a stupid
-  sentence.)
-\item No helpful actions in best-first search: After the initial
-  failure of the goal agenda, FF fell back to best-first search, but
-  didn't use the helpful actions heuristic in this search.
-\item Tautology detection in reachability analysis: The old FF
-  performed a pairwise comparison of the ground literals in the
-  instantiated goal to remove tautologies and obvious contradictions.
-  In the sentence generation problems, the instantiated goal can
-  contain lots of literals, so this takes a long time.  Removing the
-  check helped a lot.  Another way to fix this problem would have been
-  to sort the list of ground literals appropriately and then only
-  compare neighboring literals.
-\item Unnecessary computation of hard action template data structure:
-  FF computed a big data structure that was only useful for hard
-  action templates, although the sentence generation planning problems
-  only contain easy action templates.  Because the universe can be
-  relatively big, this again wasted a lot of time.
+  sentence.) FIX: turn goal agenda off. alternatively, develop
+  appropriate goal ordering techniques.
+\item Subst and distractor atoms have a weird lifecycle: They start as
+  false, then get made true at some point, and then false again.
+  After they have been made false, they can never become true
+  again. FIX: no fix just yet. highly non-trivial challenge to
+  generation of h fns. {\bf (QUESTION: does lama do better? do any of
+    the known admissible heuristics do better?)}
+\item As a consequence, Enforced hillclimbing: Initial state of
+  example problem has a relaxed graph of length one because
+  ``expressed'' and ``subst'' goals are achieved, and newly introduced
+  violations of ``subst'' and ``distractor'' are not noticed. As a
+  consequence, EHC on the second state spends all of its time in
+  breadth-first search. FIX: don't use EHC.
+\item The only thing that works well is helpful actions. However,
+  these didn't get around to be used due to no helpful actions in
+  best-first search: After failure of EHC, FF falls back to best-first
+  search, but doesn't use the helpful actions heuristic in this
+  search. FIX: simply go straight to BFS with helpful actions.
 \end{enumerate}

### 2009-12-11 trunk/Papers/crisp-ff-10/experiments.tex
title, structure, ff sec outline modifications

\section{Experiments} 
\label{sec:experiments}

.. describe before/after picture, claim that now the performance is
``for the moment ok for this application'', but not for foreseeable
future (bigger grammers etc)

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2009-12-11 trunk/Papers/crisp-ff-10/main.tex
some initial stuff
title, structure, ff sec outline modifications

@@
 \usepackage{courier}
+
+\usepackage{graphicx}
 
@@
 \begin{document}
-\title{Challenge: Using Planning to Generate Natural-Language
-  Sentences}
-\author{J\"org Hoffmann \\ LORIA \\ joerg.hoffmann@loria.fr
+\title{Using Planning to Generate Natural-Language Sentences:\\
+  Experiences and Challenges}
+
+\author{
+Alexander Koller \\ Saarland University, Germany \\ koller@mmci.uni-saarland.de
 \And
-Alexander Koller \\ Saarland University \\ koller@mmci.uni-saarland.de
+J\"org Hoffmann \\ INRIA, France \\ joerg.hoffmann@inria.fr
 \And
-Ronald Petrick \\ University of Edinburgh \\ rpetrick@inf.ed.ac.uk}
+Ronald Petrick \\ University of Edinburgh, UK \\ rpetrick@inf.ed.ac.uk
+}
+
 \maketitle
@@
 \input crisp-ff
+\input experiments
 \input conclusion

### 2009-12-11 trunk/Papers/crisp-ff-10/abstract.tex
abstract revised

\begin{abstract}
We present a planning domain that encodes the problem of generating
natural language sentences. This domain has a number of features that
make it significantly different from existing IPC benchmark
domains. In particular, we show how several techniques used in the FF
planner, which has been highly successful on IPC problems, are led
astray. We discuss possible fixes to FF's shortcomings, some of which
we have already implemented, but some of which pose interesting
challenges to the AI Planning community. The performance of our
modified FF is up to three orders of magnitude better than that of the
original FF. While this makes FF much more practical for this
application, we show that it is still very limited with respect to
problem parameters such as the complexity of the underlying
grammar. We outline how other planning domains might share the
features that make the generation domain difficult. {\bf do we? how?}
\end{abstract}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2009-12-11 trunk/Papers/reasoning-challenges/abstract.tex
new abstract and title

@@
 \begin{abstract}
-\begin{quote}
-AAAI creates proceedings, working notes, and technical reports directly from electronic source furnished by the authors.
-To ensure that all papers in the publication have a uniform appearance, authors must adhere to the following
-instructions. 
-\end{quote}
+The International Planning Competition (IPC) recently celebrated its
+10-year anniversary in 2008. During this time, the IPC has helped drive the
+development of new planning technology, by providing a regular forum for
+showcasing the latest trends in planner design, domain description
+languages, and challenging planning domains. Given the absence of a
+competition in 2010, we feel the time is right to revisit the structure of
+the existing IPC. Building on the strengths of previous competitions, we
+propose a new type of ``rolling'' planning competition, as a permanent
+fixture for the planning community. A central competition server would
+continually evaluate planners and generate performance data, while members
+of the community could directly upload new planners and problem domains,
+with the server automatically reevaluating existing planners as necessary.
+Real-time results would continuously be available, while more formal
+competitions could be ``run'' at regular intervals by producing snapshots
+of current planners and problems. Such a proposal offers significant
+benefits: researchers would have access to a central repository offering
+the latest domains and planner performance statistics, running on a common
+hardware platform, plus a facility for related activities such as posting
+challenges and demonstrating new technologies. Since much of the proposed
+infrastructure is not tied to a specific research area, we believe such an
+approach also offers the possibility of joint ventures with other
+interested research communities.
 \end{abstract}
-
-%%% Local Variables: 
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:

### 2009-12-11 trunk/Papers/crisp-ff-10/crisp-ff.tex
ff sec 1st drafty version

--- 2009-12-09 trunk/Papers/crisp-ff-10/crisp-ff.tex
@@
-\section{Experiences with FF}
+\section{FF Pitfalls and Fixes}
 \label{sec:crisp-ff}
+
+It turns out that, as far as the suitability of commonly used
+pre-processing and search techniques are concerned, sentence
+generation diverges significantly from most known benchmark
+domains. We examine this here, in detail, for the case of FF
+\cite{HoffmannNebel01}. We show how certain hitherto irrelevant
+details of the pre-processing suddenly become deadly obstacles to
+performance, and how certain widely used search techniques -- namely
+goal orderings and the relaxed plan heuristic -- are completely off
+the mark and partly even detrimental. For some of the observed
+pitfalls, we are able to provide simple fixes; others pose serious
+challenges to planning research. We focus first on the pre-processing,
+then on the search techniques.
+
+\subsection{Pre-Processing}
+\label{sec:crisp-ff:preprocess}
+
+FF's pre-process contains two details that, for a long time until they
+were detected, were so detrimental in our domain that we were not able
+to seriously propose FF as a practical solution. Simple fixes to these
+details are so effective that FF is now a practical (if not perfect)
+possibility.
+
+First, while pre-processing the ADL formulas, FF has a sub-routine
+that checks the instantiated formulas -- quantifiers already compiled
+into grounded conjunctions and disjunctions -- for tautologies. This
+involves a loop checking all pairs of sub-formulas within every
+conjunction and disjunction, testing whether they are
+identical/identical modulo their sign. For example, if we find $\phi$
+and $\neg \phi$ within a disjunction, then the disjunction is replaced
+by TRUE. Now, we haven't tested thoroughly whether this sub-routine
+has any notable effect on the formulas in other benchmarks. It
+certainly has no such effect in our domain. The effect it {\em does}
+have is take a huge amount of runtime on the long conjunctions that
+characterize {\bf \dots} For example {\bf \dots before/after}.
+
+Second, FF's pre-process contains a very small imprudence that
+apparently didn't surface in other domains, but that does in ours. FF
+distinguishes operators into ``easy'' (DNF precondition) and ``hard''
+(non-DNF precondition). The easy ones are instantiated in a
+spezialized very effectively implemented procedure, while the hard
+ones go through a much less effectively implemented more general
+procedure. In particular, that latter procedure uses a data structure
+encompassing an integer for every possible instantiation of all
+predicates. Since this does not take into account reachability, it is
+prohibitively huge in our domain. Now, there actually are no hard
+operators in our domain. But FF builds the data structure anyway, even
+though it is not needed. For example {\bf \dots before/after.}
+
+\subsection{Search Techniques}
+\label{sec:crisp-ff:preprocess:heuristics}
+
+FF's pre-processing pitfalls -- and the fact that they didn't occur to
+anyone before -- are certainly baffling, although easily fixed. We now
+turn our attention to some of FF's search techniques, variations of
+which are used in many planners and which are very much not easily
+fixed.
+
+%\item Goal agenda: For some reason that I don't understand in detail,
+%  the goal agenda heuristic chooses that the ``subst'' goal must be
+%  reached before even the ``expressed'' goal. This leads the planner
+%  to generating some arbitrary sentence. Because the example grammar
+%  is so small, this leads to a state from which the ``expressed'' goal
+%  can no longer be achieved. (A realistic big grammar should be able
+%  to recover from this by introducing a conjunction or relative clause
+%  or some such -- but of course, this would still lead to a stupid
+%  sentence.) FIX: turn goal agenda off. alternatively, develop
+%  appropriate goal ordering techniques.
+
+First, we consider Koehler and Hoffmann's
+\shortcite{koehler:hoffmann:jair-00} ``reasonable goal orderings'',
+variants and extensions of which are used in several planning systems
+today, e.g.\ \cite{hoffmann:etal:jair-04,richter:etal:aaai-08}. FF as
+used in the 2000 competition implements the ideas from
+\cite{koehler:hoffmann:jair-00}, approximating reasonable orders as a
+pre-process and then partitioning the goal set into a {\em goal
+  agenda}, a sequence of goal subsets. Each subset is posed to the
+planner in isolation, ignoring all other goals. The final plan is the
+concatenation of the plans for all entries.
+
+While the goal agenda and related techniques have proved quite useful
+in many domains, in the sentence generation domain they are completely
+off-target. Reasonable orders exist {\bf \dots explain using the
+  example dedicated to this point how we get to have only the
+  ``subst'' goal in the first entry \dots} When achieving this first
+goal agenda entry, the planner decides to generate {\em any}
+sentence. If it happens to be the wrong sentence, the planner is stuck
+in a dead-end. While that dead end is recognized by the relaxed plan
+heuristic -- i.e., it is easy to see that the actions needed for
+achieving the goal cannot be applied anymore -- certainly this is an
+entirely unintended (and a bit amusing) outcome of this
+technique. Switching it off, we {\bf \dots example before/after.}
+
+{\bf \dots pose as a challenge \dots if sensible based on example
+  observations, speculate a bit how/whether goal ordering techniques
+  could be modified to work}
+
+%\item Subst and distractor atoms have a weird lifecycle: They start as
+%  false, then get made true at some point, and then false again.
+%  After they have been made false, they can never become true
+%  again. FIX: no fix just yet. highly non-trivial challenge to
+%  generation of h fns. {\bf (QUESTION: does lama do better? do any of
+%    the known admissible heuristics do better?)}
+
+While the goal agenda meltdown may be more amusing than alarming, we
+next report that the relaxed plan heuristic is not far from such a
+meltdown either. They are fatally confused by the weird lifecycle of
+the ``subst'' and ``distractor'' atoms: in valid plans, these start
+off being false, then they get made true at some point, and then they
+are made false again in order to achieve the goal. In other words,
+from the perspective of the relaxed plan heuristic, from one state to
+the next whole classes of goals suddenly appear ``out of the
+blue''. These goals are not accounted for in the preceding state, and
+hence the relaxed plan is completely un-informative.
+
+To give just one concrete example, say we are in the initial state,
+and have not yet committed to a sentence. The relaxed plan has length
+$1$ because the we only need to choose a sentence communicating the
+desired meaning. However, in the state $s$ resulting from this action,
+the mechanisms of the planning task introduce all the sub-goals that
+are required in order to pin-point that meaning (in particular we get
+a new goal for all the distractors that must be removed). Hence the
+relaxed plan is much longer than before -- arbitrarily longer, in
+general. For FF's particular search algorithm, this means that {\em
+  enforced hill-climbing turns into a breadth-first search}: it starts
+at $h$ value $1$ and thereafter needs to solve the whole problem in
+order to find a (goal) state with lower $h$ value.
+
+%\item As a consequence, Enforced hillclimbing: Initial state of
+%  example problem has a relaxed graph of length one because
+%  ``expressed'' and ``subst'' goals are achieved, and newly introduced
+%  violations of ``subst'' and ``distractor'' are not noticed. As a
+%  consequence, EHC on the second state spends all of its time in
+%  breadth-first search. FIX: don't use EHC.
+
+More generally, from this example we immediately see that, in the
+terms of Hoffmann \shortcite{hoffmann:jair-05}, the exit distance from
+local minima under $h^+$ is unbounded. Also, we have unrecognized
+dead-ends, i.e., dead-end states that have a relaxed plan. An example
+is the initial state of a task that features distractors for which
+there is no means of removal -- ignoring the distractors, the initial
+$h$ value will still be $1$. So our domain is in the most difficult
+class of Hoffmann's \shortcite{hoffmann:jair-05} ``plannin domain
+taxonomy''. While it shares this property with several other
+challenging domains, the extremity of turning enforced hill-climbing
+into a breadth-first search is unheard of.
+
+{\bf \dots pose as a challenge \dots if sensible based on example
+  observations, speculate a bit how/whether heuristics could be
+  modified to work \dots a question that we should ideally at least
+  partially answer in here is whether other existing heuristics
+  actually work better.. we could try at least a little bit with PDBs,
+  merge\&shrink, etc}
+
+The only FF technique that appears to work reasonably well is its
+action pruning technique, the ``helpful actions''. While the relaxed
+plan is often much too short, it appears to always contain at least
+one of the actions that actually are useful, and it certainly prunes
+the set of avaibale actions quite effectively. {\bf \dots now
+  something on example and BFS with helpful actions vs. without \dots}
+
+%\item The only thing that works well is helpful actions. However,
+%  these didn't get around to be used due to no helpful actions in
+%  best-first search: After failure of EHC, FF falls back to best-first
+%  search, but doesn't use the helpful actions heuristic in this
+%  search. FIX: simply go straight to BFS with helpful actions.
 

### 2009-12-13 trunk/Papers/reasoning-challenges/abstract.tex
changes to abstract, introduction

--- 2009-12-09 trunk/Papers/reasoning-challenges/abstract.tex
@@
 \begin{abstract}
-\begin{quote}
-AAAI creates proceedings, working notes, and technical reports directly from electronic source furnished by the authors.
-To ensure that all papers in the publication have a uniform appearance, authors must adhere to the following
-instructions. 
-\end{quote}
+We revisit the structure of the International Planning Competition (IPC)
+and propose a new ``rolling'' competition, as a permanent fixture for the
+planning community. A central competition server would continually evaluate
+planners and generate performance data, while members of the community
+could directly upload new planners and problem domains, with the server
+automatically reevaluating existing planners as necessary. Real-time
+results would continuously be available, while more formal competitions
+could be ``run'' at regular intervals by producing snapshots of current
+planners and problems. Researchers would have access to a central
+repository with the latest domains and performance statistics, running on a
+common hardware platform, plus a facility for related activities such as
+posting challenges and demonstrating new technologies. Since much of the
+proposed infrastructure is not tied to a specific research area, this
+approach offers the possibility of joint ventures with other research
+communities.
 \end{abstract}
-
-%%% Local Variables: 
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:

### 2009-12-13 trunk/Papers/reasoning-challenges/infrastructure.tex
changes to abstract, introduction

\section{Infrastructure}
\label{sec:infrastructure}

Talk a little bit about the current planning competition and (online)
competitions from other research areas (GIVE, CP?, others?)... Outline the
infrastructure we're proposing. We need some references here...

### 2009-12-13 trunk/Papers/reasoning-challenges/introduction.tex
changes to abstract, introduction

@@
-\section{Introduction} \label{sec:introduction}
+\section{Introduction}
+\label{sec:introduction}
 
-%%% Local Variables: 
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:
+The International Planning Competition (IPC)\footnote{See
+\url{http://ipc.icaps-conference.org/}.} recently celebrated its 10-year
+anniversary in 2008. During this time, six instances of the IPC have been
+run, each of which has helped drive the development of new planning
+technology, by providing a regular forum for showcasing the latest trends
+in planner design, domain description languages, and challenging planning
+domains. Given the absence of a competition in 2010---breaking the
+competition's established two year cycle for the first time since its
+inception---we feel the time is right to revisit the structure of the
+existing IPC, and suggest a new direction for the competition in the
+future.
+
+This paper proposes a new direction for the IPC, building on the strength
+of previous competitions, but centred around a novel online infrastructure.
+Most notably, we envision a more interactive and potentially more
+wide-ranging competition featuring:
+%
+\begin{itemize}
+\item A ``rolling'' competition format with community-initiated track and
+problem specifications,
+
+\item Continuous evaluation of contributed planners on competition
+problem sets, with real-time access to performance data generated on a
+common hardware platform, and
+
+\item A centralised repository offering researchers access to the latest
+domains and planners, and a forum for community news, discussion, and
+interaction.
+\end{itemize}
+
+Thus, while previous instances of the IPC have traditionally been
+event-based competitions, scheduled at regular intervals, we propose an
+ongoing competition as a permanent fixture for the planning community. In
+particular, members of the community would be able to upload new planners
+and problem domains to a central competition server that continually
+re-evaluates planners and generates performance data. Up-to-the-minute
+results would be centrally available, while more formal competitions could
+be ``run'' at regular intervals by producing snapshots of current planners
+and problems.
+
+From a technical point of view, it is important to note that such an idea
+would not have been practical ten years ago during the time of the first
+IPC. It is only recently, due to advances in computing hardware and network
+bandwidth, that a scenario such as this is now feasible. Even so, our
+proposed infrastructure nevertheless presents some technical challenges
+that must be addressed (e.g., security, , accessibility, maintenance,
+etc.). However, we believe that for the most part, solutions to these
+problems can already be found by looking at similar (although smaller
+scale) competition efforts in the wider research community.
+
+Most importantly, we believe this proposal offers significant benefits to
+competition organisers, participants, and the wider planning community.
+\textbf{[RP: say something more here]} Furthermore, since much of the
+infrastructure we describe here is not tied to a specific research area,
+attempts to establish similar competitions could benefit from this
+approach, leading to joint ventures with other research fields---and the
+possibility of future inter-disciplinary challenges.
+
+In the remainder of the paper we outline our proposal for a new competition
+infrastructure, and the benefits we believe this approach could bring to
+the planning community and beyond.

### 2009-12-16 trunk/Papers/reasoning-challenges/infrastructure.tex
Some more details on infrastructure, some references (not cited yet)

@@
 
-Talk a little bit about the current planning competition and (online)
+\textbf{[RP: Talk a little bit about the current planning competition and (online)
 competitions from other research areas (GIVE, CP?, others?)... Outline the
-infrastructure we're proposing. We need some references here...
+infrastructure we're proposing. We need some references here...]}
+
+There are three groups of people involved in a competition: the organisers, the participants, and the audience (which usually includes organisers and participants).
+In the following, we outline an infrastructure that makes competitions easier to run and more valuable for each of these groups.
+
+\begin{itemize}
+  \item Organisers benefit from a mostly automated platform that is maintained by the community, which should greatly reduce the effort needed to run a competition.
+  \item Participants benefit from a well-defined environment for running their systems, short feedback times after submitting a system, and a standardised test and benchmark platform for experimenting with novel algorithms.
+  \item The audience, i.e., the scientific community, benefits from the improved discussion and report facilities.
+The scientific value of competitions is to find out \emph{why} a system shows a certain performance, not only \emph{that} it does.
+\end{itemize}
+
+We want to stress that the proposed platform is not limited to the domain of planning problems.
+Therefore, we will talk about \emph{problem classes} and \emph{instances} in the following, instead of planning domains and problems.
+
+\subsection{Submitting Systems}
+
+A system is uploaded as an archive containing
+\begin{itemize}
+  \item A specification of the input format type
+  \item An executable file
+  \item Optional format conversion rules
+  \item Optional other files necessary for running the system
+\end{itemize}
+
+The environment in which the system will be run is available as a virtual machine image, such that participants can compile and test their systems before uploading them.
+
+To upload a system, the participants have to register with the maintainer. We hope that this sets the bar high enough to avoid the submission of malicious systems.
+
+\subsection{Submitting Instances}
+
+There are three levels at which instances are submitted: the problem class, the instance, and the \emph{data}.
+The problem class groups similar instances (for example those from the same planning domain), and possibly data describing the class (such as a PDDL planning domain).
+The instance itself may contain a human-readable description and several sets of data.
+Each set of data has a \emph{type}, which determines what systems the data is suitable for (e.g.\ different subsets of PDDL, or EUROPA).
+All sets of data for one instance should of course model the same problem.
+
+Problem classes, instances, and data can be submitted by any registered user.
+In order to avoid abuse, the number of instances should be limited (e.g.\ per hour), and a simple validity check should be possible (such as running a parser if the format is PDDL).
+
+\subsection{Running Systems}
+
+Let us now come to the backend of the platform.
+Its main task is to prepare the data for the different systems, to distribute the work of running the systems among a number of (identical) machines, and to collect the results in a central database.
+
+Systems will be run in a sandbox environment with well-defined resource limits such as number of processors, available memory and processor time, and limited disk space and network connectivity.
+
+The platform provides automatic conversion between different formats.
+When new instance data is submitted, it is automatically converted into all formats for which conversions are available.
+Then it is scheduled to be run on all suitable systems.
+
+When a system is updated or a new system is submitted, all suitable instances are scheduled to be run on the system.
+
+A number of statistics is gathered during execution.
+The obvious statistics would be runtime, memory consumption, and whether the system could solve the problem.
+More detailed statistics will require system-specific support.
+
+\subsection{Reports and Discussion}
+
+The public interface of the competition platform serves two main purposes: \emph{interactivity}, which lets users describe and discuss systems and problems, and \emph{reporting}, which presents the results of running the systems.
+
+We propose to achieve interactivity by making use of three standard technologies:
+\begin{itemize}
+  \item a \emph{wiki} for all the descriptive content,
+  \item a \emph{discussion board} for each wiki page (similar to the discussion feature in Wikipedia), and
+  \item a \emph{tagging} system, which allows arbitrary grouping of systems and problems.
+\end{itemize}
+
+The reporting engine connects to the central database that is populated by the backend.
+Reports can be embedded into wiki pages and customised using tags.
+For example, a user can create a page with the report comparing the runtimes of three systems, \emph{A}, \emph{B}, and \emph{C}, on all instances labeled \emph{mediumHard}.

### 2009-12-16 trunk/Papers/crisp-ff-10/crisp-ff.tex
ff sec ueberarbeitung teil 1

@@
 
+%../ff -o xtag-3-2-domain.lisp -f xtag-3-2-problem.lisp -T -H: 0.10
+%../ff -o xtag-3-2-domain.lisp -f xtag-3-2-problem.lisp -T: 1.54
+%../ff -o xtag-3-2-domain.lisp -f xtag-3-2-problem.lisp -H: 0.30
+%../ff -o xtag-3-2-domain.lisp -f xtag-3-2-problem.lisp: 1.80
+
+%../ff -o xtag-5-2-domain.lisp -f xtag-5-2-problem.lisp -T -H: 0.29
+%../ff -o xtag-5-2-domain.lisp -f xtag-5-2-problem.lisp -T: seg fault ie too much memory
+%../ff -o xtag-5-2-domain.lisp -f xtag-5-2-problem.lisp -H: 11.08
+%../ff -o xtag-5-2-domain.lisp -f xtag-5-2-problem.lisp: seg fault ie too much memory
+
+%../ff -o xtag-3-3-domain.lisp -f xtag-3-3-problem.lisp -T -H: 
+%../ff -o xtag-3-3-domain.lisp -f xtag-3-3-problem.lisp -T: seg fault ie too much memory
+%../ff -o xtag-3-3-domain.lisp -f xtag-3-3-problem.lisp -H: 
+%../ff -o xtag-3-3-domain.lisp -f xtag-3-3-problem.lisp: seg fault ie too much memory
+
 FF's pre-process contains two details that, for a long time until they
@@
 certainly has no such effect in our domain. The effect it {\em does}
-have is take a huge amount of runtime on the long conjunctions that
-characterize {\bf \dots} For example {\bf \dots before/after}.
+have is take a huge amount of runtime on the long goal conjunctions
+stating that we do not wish to have any open substitution nodes nor
+any distractors. The simple fix is hence to switch this sub-routine
+off. In smaller examples, this gives advantages like $0.29s$
+vs.\ $11s$ pre-processing time; in a larger example we ran, original
+FF's pre-process doesn't finish within 15 minutes whereas our fix
+yields a pre-processing time of about $3s$.
 
@@
 operators in our domain. But FF builds the data structure anyway, even
-though it is not needed. For example {\bf \dots before/after.}
+though it is not needed. In small examples, this results in a couple
+of seconds runtime overhead. In larger examples, it exhausts memory
+and causes FF to terminate with a segmentation fault.
 
@@
 
+%EXAMPLE missing
+
 While the goal agenda and related techniques have proved quite useful
@@
 While the goal agenda meltdown may be more amusing than alarming, we
-next report that the relaxed plan heuristic is not far from such a
+next report that relaxed plan heuristics are not far from such a
 meltdown either. They are fatally confused by the weird lifecycle of
-the ``subst'' and ``distractor'' atoms: in valid plans, these start
-off being false, then they get made true at some point, and then they
-are made false again in order to achieve the goal. In other words,
-from the perspective of the relaxed plan heuristic, from one state to
-the next whole classes of goals suddenly appear ``out of the
-blue''. These goals are not accounted for in the preceding state, and
-hence the relaxed plan is completely un-informative.
+facts encoding substitution nodes and distractors. In instances where
+these constructs are important for the construction of a valid
+sentence, these facts are required to be false in the goal, and most
+of them are initially false. However, any valid plan must make many of
+them true at some intermediate time point (e.g.\ distractors arise
+depending on the decisions taken by the planner). Hence, from the
+perspective of the relaxed plan heuristic, from one state to the next
+whole classes of goals suddenly appear ``out of the blue''. These
+goals are not accounted for in the preceding state, and hence the
+relaxed plan is completely un-informative.
 
 To give just one concrete example, say we are in the initial state,
-and have not yet committed to a sentence. The relaxed plan has length
-$1$ because the we only need to choose a sentence communicating the
-desired meaning. However, in the state $s$ resulting from this action,
-the mechanisms of the planning task introduce all the sub-goals that
-are required in order to pin-point that meaning (in particular we get
-a new goal for all the distractors that must be removed). Hence the
-relaxed plan is much longer than before -- arbitrarily longer, in
-general. For FF's particular search algorithm, this means that {\em
-  enforced hill-climbing turns into a breadth-first search}: it starts
-at $h$ value $1$ and thereafter needs to solve the whole problem in
-order to find a (goal) state with lower $h$ value.
+and have not yet committed to a sentence. For certain kinds of
+sentences ({\bf more concrete here???}), the relaxed plan has length
+$1$ because the we only need to choose the desired meaning. However,
+in the state $s$ resulting from this action, the mechanisms of the
+planning task introduce all the sub-goals that are required in order
+to pin-point that meaning (in particular we get a new goal for all the
+distractors that must be removed). Hence the relaxed plan is much
+longer than before -- arbitrarily longer, in general. For FF's
+particular search algorithm, enforced hill-climbing (EHC), this means
+that {\em FF performs a blind breadth-first search}: EHC starts at $h$
+value $1$ and thereafter needs to solve the whole problem in order to
+find a (goal) state with lower $h$ value.
+
+%../ff -o modifiers-10-10-domain.lisp -f modifiers-10-10-problem.lisp -T -H
 
@@
 $h$ value will still be $1$. So our domain is in the most difficult
-class of Hoffmann's \shortcite{hoffmann:jair-05} ``plannin domain
+class of Hoffmann's \shortcite{hoffmann:jair-05} ``planning domain
 taxonomy''. While it shares this property with several other
-challenging domains, the extremity of turning enforced hill-climbing
-into a breadth-first search is unheard of.
+challenging domains, the extremity of turning EHC into a breadth-first
+search is unheard of.
 
@@
 one of the actions that actually are useful, and it certainly prunes
-the set of avaibale actions quite effectively. {\bf \dots now
-  something on example and BFS with helpful actions vs. without \dots}
+the set of avaibale actions quite effectively. For example, in one of
+the instances where EHC turns into a breadth-first search, $16875$
+nodes are explored without helpful actions; with helpful actions, this
+reduces to $3078$. For larger examples, the factor between the two
+numbers tends to become larger. {\bf use here larger version of
+  modifiers-10-10???}
+
+Since EHC is often completely inadequate, and helpful actions are
+quite effective, our FF configuration directly runs best-first search
+with helpful actions (as opposed to original FF which switches to
+best-first with{\em out} helpful actions if EHC fails).
 

### 2009-12-16 trunk/Papers/crisp-ff-10/crisp-ff.tex
ff sec revision, teil 2

@@
 any distractors. The simple fix is hence to switch this sub-routine
-off. In smaller examples, this gives advantages like $0.29s$
-vs.\ $11s$ pre-processing time; in a larger example we ran, original
-FF's pre-process doesn't finish within 15 minutes whereas our fix
-yields a pre-processing time of about $3s$.
+off. In small instances, this gives advantages like $0.29s$ vs.\ $11s$
+pre-processing time; in a larger instance we ran, original FF's
+pre-process doesn't finish within 15 minutes whereas our fix yields a
+pre-processing time of about $3s$.
 
@@
 operators in our domain. But FF builds the data structure anyway, even
-though it is not needed. In small examples, this results in a couple
-of seconds runtime overhead. In larger examples, it exhausts memory
+though it is not needed. In small instances, this results in a couple
+of seconds runtime overhead. In larger instances, it exhausts memory
 and causes FF to terminate with a segmentation fault.
@@
 While the goal agenda and related techniques have proved quite useful
-in many domains, in the sentence generation domain they are completely
-off-target. Reasonable orders exist {\bf \dots explain using the
-  example dedicated to this point how we get to have only the
-  ``subst'' goal in the first entry \dots} When achieving this first
-goal agenda entry, the planner decides to generate {\em any}
-sentence. If it happens to be the wrong sentence, the planner is stuck
-in a dead-end. While that dead end is recognized by the relaxed plan
-heuristic -- i.e., it is easy to see that the actions needed for
-achieving the goal cannot be applied anymore -- certainly this is an
-entirely unintended (and a bit amusing) outcome of this
-technique. Switching it off, we {\bf \dots example before/after.}
+in many domains, in the sentence generation domain they often are
+completely off-target. The most striking point concerns the
+interaction between the goal to create a sentence -- $\neg
+\mathsf{subst}(S,\mathsf{root})$ in the rabbit example -- and the goal
+to express a certain meaning -- $\mathsf{sleep}(e,r_1)$ in the
+rabbit example. The goal agenda orders the former before the latter
+because it detects that, once a meaning is expressed, the sentence
+creation has started and it is not possible to start anew
+anymore. Unless the inverse also holds,\footnote{The inverse holds if,
+  in the given grammar, the only way to create a meaning is to create
+  a full sentence -- i.e., there are no side sentences like ``the
+  sleeping rabbit''.} $\mathsf{subst}(S,\mathsf{root})$ is alone in
+the first goal agenda entry, i.e., there is no goal stating the actual
+meaning the sentence is supposed to have. Consequently, the planner
+decides to generate {\em any} sentence. If it happens to be the wrong
+sentence, the planner is stuck in a dead-end. While that dead end is
+recognized by the relaxed plan heuristic -- i.e., it is easy to see
+that the actions needed for achieving the goal cannot be applied
+anymore -- certainly this is an entirely unintended (and a bit
+amusing) outcome of this technique. 
+% Switching it off, we {\bf \dots
+%example before/after.}
 
-{\bf \dots pose as a challenge \dots if sensible based on example
-  observations, speculate a bit how/whether goal ordering techniques
-  could be modified to work}
+It is not clear to us, at this point, how goal ordering techniques
+could be modified so that they return sensible results in the sentence
+generation domain. The issue explained above regarding $\neg
+\mathsf{subst}(S,\mathsf{root})$ and $\mathsf{sleep}(e,r_1)$ could be
+addressed by recognizing that, once the latter is achieved, the former
+is already true anyway. More generally however, the life-cycle of
+substitution nodes and distractors is weird -- as we outline below, it
+follows a false-true-false pattern -- and seems difficult to detect
+based on domain analysis.
 
@@
 
-To give just one concrete example, say we are in the initial state,
-and have not yet committed to a sentence. For certain kinds of
-sentences ({\bf more concrete here???}), the relaxed plan has length
-$1$ because the we only need to choose the desired meaning. However,
-in the state $s$ resulting from this action, the mechanisms of the
-planning task introduce all the sub-goals that are required in order
-to pin-point that meaning (in particular we get a new goal for all the
-distractors that must be removed). Hence the relaxed plan is much
-longer than before -- arbitrarily longer, in general. For FF's
-particular search algorithm, enforced hill-climbing (EHC), this means
-that {\em FF performs a blind breadth-first search}: EHC starts at $h$
-value $1$ and thereafter needs to solve the whole problem in order to
-find a (goal) state with lower $h$ value.
+Consider again the rabbit example. In the initial state, the relaxed
+plan has length $1$ because we only need to achieve $\neg
+\mathsf{subst}(S,\mathsf{root})$ and $\mathsf{sleep}(e,r_1)$, which is
+done by the single action ``sleeps''. However, that action has the
+effect ``forall (?y - individual) (when (not (= ?y ?xself))
+(distractor ?subjnode ?y))'' which introduces new distractors. Those
+must be removed, so they form ``new'' goals. In the given case, the
+relaxed plan now contains two actions to achieve these goals. In
+general (if there are $n$ rabbits, say), the relaxed plan may become
+arbitrarily long.
+%To give just one concrete example, say we are in the initial state,
+%and have not yet committed to a sentence. For certain kinds of
+%sentences ({\bf more concrete here???}), the relaxed plan has length
+%$1$ because the we only need to choose the desired meaning. However,
+%in the state $s$ resulting from this action, the mechanisms of the
+%planning task introduce all the sub-goals that are required in order
+%to pin-point that meaning (in particular we get a new goal for all the
+%distractors that must be removed). Hence the relaxed plan is much
+%longer than before -- arbitrarily longer, in general.  
+For FF's particular search algorithm, enforced hill-climbing (EHC),
+this means that {\em FF performs a blind breadth-first search}: EHC
+starts at $h$ value $1$ and thereafter needs to solve the whole
+problem in order to find a (goal) state with lower $h$ value.
 
@@
 
-{\bf \dots pose as a challenge \dots if sensible based on example
-  observations, speculate a bit how/whether heuristics could be
-  modified to work \dots a question that we should ideally at least
-  partially answer in here is whether other existing heuristics
-  actually work better.. we could try at least a little bit with PDBs,
-  merge\&shrink, etc}
+Can we design heuristics that are better at deadling with the
+lifecycle of substitution nodes and distractors? The problem of not
+noticing them in the initial state is present for all approaches that
+are dominated by $h^+$,
+e.g.\ \cite{karpas:domshlak-ijcai-09,helmert:domshlak:icaps-09}, and
+is also present for the known inadmissible landmarks heuristics
+\cite{richter:etal:aaai-08}. As far as known heuristics are concerned,
+hope thus resides mainly in the $h^m$ family and in abstraction
+heuristics. We ran a few tests with an example like the rabbit one,
+except that there are 10 distractors. The Graphplan estimate for the
+initial state is $4$, i.e., $h^2$ is better than $h^+$ here. Still,
+$4$ is far away from the real goal distance $14$. With merge\&shrink
+\cite{helmert:etal:icaps07}, {\bf \dots (malte agreed to run a test on
+  modifiers-10-10-)}
 
@@
 reduces to $3078$. For larger examples, the factor between the two
-numbers tends to become larger. {\bf use here larger version of
-  modifiers-10-10???}
+numbers tends to become larger. 
+%{\bf use here larger version of
+%  modifiers-10-10???}
 

### 2009-12-16 trunk/Papers/crisp-ff-10/crisp-ff.tex
minor

@@
 initial state is $4$, i.e., $h^2$ is better than $h^+$ here. Still,
-$4$ is far away from the real goal distance $14$. With merge\&shrink
-\cite{helmert:etal:icaps07}, {\bf \dots (malte agreed to run a test on
-  modifiers-10-10-)}
+$4$ is far away from the real goal distance $14$ {\bf here I could use
+  differently scaled versions to see whether the distance estimate is
+  constant across the number of distractors; I would think that
+  yes}. With merge\&shrink \cite{helmert:etal:icaps07}, {\bf \dots
+  (malte agreed to run a test on modifiers-10-10-).. larger versions
+  would also be useful, to see how the search spaces develop}
 
@@
 one of the actions that actually are useful, and it certainly prunes
-the set of avaibale actions quite effectively. For example, in one of
-the instances where EHC turns into a breadth-first search, $16875$
-nodes are explored without helpful actions; with helpful actions, this
-reduces to $3078$. For larger examples, the factor between the two
-numbers tends to become larger. 
+the set of avaibale actions quite effectively. For example, in the
+10-distractor test instance, $16875$ nodes are explored without
+helpful actions pruning; with pruning, this reduces to $3078$. For
+larger examples, the factor between the two numbers tends to become
+larger. {\bf here, again, having actual scaled examples would help}
 %{\bf use here larger version of

### 2009-12-17 trunk/Papers/crisp-ff-10/crisp-ff.tex
fertig basierend auf *alten* beispielen

--- 2009-12-16 trunk/Papers/crisp-ff-10/crisp-ff.tex
@@
 initial state is $4$, i.e., $h^2$ is better than $h^+$ here. Still,
-$4$ is far away from the real goal distance $14$. With merge\&shrink
-\cite{helmert:etal:icaps07}, {\bf \dots (malte agreed to run a test on
-  modifiers-10-10-)}
+$4$ is far away from the real goal distance $14$. %
+%{\bf here I could use
+%  differently scaled versions to see whether the distance estimate is
+%  constant across the number of distractors; I would think that
+%  yes}. 
+With merge\&shrink \cite{helmert:etal:icaps07}, we experimented with
+various settings for the parameters (number of abstractions and size
+limit for each abstraction) but could not find a setting that reduced
+the search nodes to less than $5000$. So, all in all, finding better
+heuristic functions for this kind of problem remains a challenge.
+
+%{\bf \dots
+%  (malte agreed to run a test on modifiers-10-10-).. larger versions
+%  would also be useful, to see how the search spaces develop}
 
@@
 one of the actions that actually are useful, and it certainly prunes
-the set of avaibale actions quite effectively. For example, in one of
-the instances where EHC turns into a breadth-first search, $16875$
-nodes are explored without helpful actions; with helpful actions, this
-reduces to $3078$. For larger examples, the factor between the two
-numbers tends to become larger. 
+the set of avaibale actions quite effectively. For example, in the
+10-distractor test instance, $16875$ nodes are explored without
+helpful actions pruning; with pruning, this reduces to $3078$. For
+larger examples, the factor between the two numbers tends to become
+larger. %{\bf here, again, having actual scaled examples would help}
 %{\bf use here larger version of

### 2009-12-18 trunk/Papers/crisp-ff-10/experiments.tex
experiments, v1

@@
 
-.. describe before/after picture, claim that now the performance is
-``for the moment ok for this application'', but not for foreseeable
-future (bigger grammers etc)
+To evaluate the effect of our changes on the efficiency of the planner
+on practical inputs, we ran a series of experiments using the XTAG
+Grammar \cite{xtag01:_tr}, a large-scale tree-adjoining grammar of
+English, from which we selected all lexicon entries for seven words we
+used in the experiments. We set up inputs to generate sentences of the
+form
+
+$$S_1 \;\mathrm{and}\; S_2 \;\mathrm{and}\; \ldots \;\mathrm{and}\;
+S_n,$$
+
+i.e.\ a conjunction of $n$ sentences. Each sentence $S_i$ is of the
+form ``the X admires the Y'' or ``the X gives the Y the Z''; that is,
+we vary the arity $k$ of the verb to be 2 or 3. For X, Y, and Z are
+all of a form like ``the rich businessman'', i.e.\ like the rabbit in
+Fig.~\ref{fig:white-rabbit-as-planning}, the individual must be
+described with one noun and one adjective to distinguish it from all
+other individuals. We translate these generation problems into
+planning problems with the new encoding as shown in
+Fig.~\ref{fig:white-rabbit-as-planning}.
+
+\begin{figure}
+  \centering
+  la la la
+  \caption{Runtimes for $k=2$.}
+  \label{fig:runtimes-k2}
+\end{figure}
+
+The results for $k=2$ are shown in Fig.~\ref{fig:runtimes-k2}; the
+curves end where the planner exceeded its five-minute timeout.  As the
+graph shows, the improved preprocessor has a huge impact on the
+runtimes: \todo{refer to data}.  We can also see that best-first
+search with the helpful actions heuristic is the best-performing
+search algorithm.  While the EHC search is still faster than
+best-first search without helpful actions, this says less about the
+quality of the search algorithm itself than it does about the
+usefulness of the helpful actions heuristic.
+
+The improvements we have made in this paper bring the overall planner
+runtimes into a range where they can be useful in practical natural
+language generation applications.  For instance, the 24-word sentence
+for $k=2$ and $n=3$ is generated in under four seconds; before our
+changes, FF was not able to generate this sentence in less than
+\todo{XXX time}.  Nevertheless, our experiment also shows that the
+current technology does not yet scale to larger instances.  It should
+be noted that although we used all lexicon entries from XTAG for the
+words in our sentences, our experiment grammar still only contained
+lexicon entries for seven different words.  It would not be
+unreasonable to assume a grammar with a larger vocabulary, which would
+then be difficult for the current planner to handle.
+
+One final aspect of our experiments concerns the difference between
+the new encoding of generation problems into planning problems
+proposed above and the old encoding mentioned in Section \todo{XXX}.
+We ran a variant of the above experiment where X, Y, and Z could be
+realized with a single noun (``the businessman'') and did not require
+an adjective to be uniquely identifiable. On this data, search on the
+new encoding tended to be mildly faster for small to medium-sized
+instances than on the old encoding (1.6 sec vs.\ 2.7 sec for EHC
+search on $k=2,n=4$), whereas preprocessing was considerably faster
+for the new encoding even after our improvements (0.7 sec vs.\ 2.7 sec
+on $k=2,n=4$).  However, for larger instances the new encoding
+exceeded the five-minute timeout earlier than the old encoding did.
+Overall, the picture for the old and new encoding was roughly
+comparable, with one notable exception: Under the old encoding, EHC
+search was slightly faster than best-first search with helpful
+actions (7.2 sec vs.\ 10.9 sec at $k=2,n=5$).  This is an artefact of
+the fact that because the input specified the sentence word for word,
+the relaxed plan heuristic did not underestimate the plan length as
+grossly as in other cases.  As we have seen, adding even a single
+distractor for every noun phrase already negates this advantage.
 

### 2009-12-18 trunk/Papers/crisp-ff-10/main.tex
author order
pre-final version ff-sec
experiments, v1

@@
 
+\newcommand{\todo}[1]{\textbf{(#1)}}
+\usepackage{url}
+
 \pdfinfo{
-/Title (Formatting Instructions for Authors)
-/Subject (Proceedings of the AAAI Press Conference)
-/Author (AAAI)}
+/Title (Using Planning to Generate Natural-Language Sentences)
+/Subject (Proceedings of ICAPS, hopefully)
+/Author (Anonymous authors)}
 
@@
 Alexander Koller \\ Saarland University, Germany \\ koller@mmci.uni-saarland.de
+%\And
+%Ronald Petrick \\ University of Edinburgh, UK \\ rpetrick@inf.ed.ac.uk
 \And
 J\"org Hoffmann \\ INRIA, France \\ joerg.hoffmann@inria.fr
-\And
-Ronald Petrick \\ University of Edinburgh, UK \\ rpetrick@inf.ed.ac.uk
 }

### 2009-12-18 trunk/Papers/reasoning-challenges/infrastructure.tex
Structure, some more content

@@
-\section{Infrastructure}
-\label{sec:infrastructure}
+\section{Rationale}
+\label{sec:rationale}
 
-\textbf{[RP: Talk a little bit about the current planning competition and (online)
-competitions from other research areas (GIVE, CP?, others?)... Outline the
-infrastructure we're proposing. We need some references here...]}
+Let us introduce the proposed platform by comparing it to how competitions are organised currently, and pointing out the benefits of the new approach.
 
-There are three groups of people involved in a competition: the organisers, the participants, and the audience (which usually includes organisers and participants).
-In the following, we outline an infrastructure that makes competitions easier to run and more valuable for each of these groups.
+We want to stress that what we propose is not limited to the domain of planning problems.
+Therefore, we will talk about \emph{systems}, \emph{problem classes} and \emph{instances} in the following, instead of planners, planning domains and problems.
+
+Any competition has four main tasks: collecting systems, collecting problem specifications, running the systems, and reporting the results.
+
+\begin{description}
+  \item[Collecting systems.] Currently, systems are uploaded manually by the participants, and possibly even manually compiled, installed, and tested by the competition organiser.
+New approach: participants submit their systems without any organiser interaction.
+  \item[Collecting problems.] This is currently done by a special committee, possibly with manual submission from outside users.
+New approach: a community-built library of problems, where competitions can be run by picking from the library.
+This approach has been proposed before \cite{ToughNuts}, and it would avoid problems like the ones discussed by \cite{pg2008002}.
+  \item[Running the systems.] There is usually some degree of automation, distributing work over a cluster of computers, but the competitions are started manually for a particular event.
+New approach: continually run the competition as soon as new systems or new problems are submitted.
+  \item[Reporting the results.] Currently, the organisers choose the metrics for comparing systems as well as the type of presentation.
+New approach: reports are generated dynamically from the gathered statistics, users can retrieve data they find interesting.
+\end{description}
+
+This new approach has advantages for each of the three groups of people involved in a competition, the organisers, the participants, and the audience (which usually includes organisers and participants).
 
@@
 
-We want to stress that the proposed platform is not limited to the domain of planning problems.
-Therefore, we will talk about \emph{problem classes} and \emph{instances} in the following, instead of planning domains and problems.
+\section{Infrastructure}
+\label{sec:infrastructure}
 
-\subsection{Submitting Systems}
+The previous section identified the four main tasks in a competition: collecting systems, collecting problems, running the systems, and reporting the results.
+In this section, we propose an infrastructure consisting of four modules, one for each task, that together yield a complete platform for community-organised competitions.
 
-A system is uploaded as an archive containing
+\subsection{Collecting Systems}
+
+The proposed platform allows users to upload new systems or new versions of their systems at any time through a web interface. A system is uploaded as an archive containing
 \begin{itemize}
@@
   \item An executable file
-  \item Optional format conversion rules
-  \item Optional other files necessary for running the system
+  \item Optionally other files necessary for running the system
 \end{itemize}
@@
 
-To upload a system, the participants have to register with the maintainer. We hope that this sets the bar high enough to avoid the submission of malicious systems.
+Letting users upload arbitrary code that is run automatically obviously presents security risks.
+We believe that two simple measures are sufficient to prevent abuse: personal registration of participants, and a sandboxed execution environment.
 
-\subsection{Submitting Instances}
+\subsection{Collecting Problems}
 
-There are three levels at which instances are submitted: the problem class, the instance, and the \emph{data}.
+There are three levels at which problems are submitted: the \emph{problem class}, the \emph{problem instance}, and the \emph{data}.
 The problem class groups similar instances (for example those from the same planning domain), and possibly data describing the class (such as a PDDL planning domain).
@@
 
-Let us now come to the backend of the platform.
+We now come to the backend of the platform.
 Its main task is to prepare the data for the different systems, to distribute the work of running the systems among a number of (identical) machines, and to collect the results in a central database.
-
-Systems will be run in a sandbox environment with well-defined resource limits such as number of processors, available memory and processor time, and limited disk space and network connectivity.
 
@@
 
+Systems will be run in a sandbox environment with well-defined resource limits such as number of processors, available memory and processor time, and limited disk space and network connectivity.
+
 When a system is updated or a new system is submitted, all suitable instances are scheduled to be run on the system.
@@
 The obvious statistics would be runtime, memory consumption, and whether the system could solve the problem.
-More detailed statistics will require system-specific support.
+We discuss more detailed statistics later.
 
-\subsection{Reports and Discussion}
+\subsection{Reporting Results}
 
-The public interface of the competition platform serves two main purposes: \emph{interactivity}, which lets users describe and discuss systems and problems, and \emph{reporting}, which presents the results of running the systems.
+The public interface of competition is currently mostly static: a web page or document that contains the various results.
+In a community-organised competition, the interface must be dynamic and created by the community.
 
-We propose to achieve interactivity by making use of three standard technologies:
+The most successful example of community-generated content is a \emph{wiki}. We therefore propose to build the user interface of the competition platform around a wiki engine.
+
+Users can add and edit pages that describe systems as well as problems.
+The platform provides a reporting engine that connects to the central result database, and reports can be embedded into wiki pages.
+Any user can therefore present competition results by creating a wiki page with a set of reports.
+
+A competition, in the classic sense, can be run on the new platform as follows:
 \begin{itemize}
-  \item a \emph{wiki} for all the descriptive content,
-  \item a \emph{discussion board} for each wiki page (similar to the discussion feature in Wikipedia), and
-  \item a \emph{tagging} system, which allows arbitrary grouping of systems and problems.
+  \item Users submit systems and problems.
+  \item A committee picks a set of problems (and possibly submits new problems that the participants have not seen before).
+  \item The platform automatically runs the systems, producing results in a database.
+  \item The organisers decide on a set of metrics, categories etc., and set up a web page with the results, using the reporting features of the platform.
 \end{itemize}
 
-The reporting engine connects to the central database that is populated by the backend.
-Reports can be embedded into wiki pages and customised using tags.
-For example, a user can create a page with the report comparing the runtimes of three systems, \emph{A}, \emph{B}, and \emph{C}, on all instances labeled \emph{mediumHard}.
+\section{Additional Features}
+\label{sec:additional_features}
+
+This section discusses additional features that can be added to the basic system.
+
+\subsection{Community Features}
+
+The user interface module from the previous section can be augmented with \emph{social networking} features.
+This section presents features and how they may prove useful for a community-organised competition.
+
+\begin{description}
+  \item[Discussion.]
+  \item[Voting.]
+  \item[Tagging.]
+\end{description}
+
+The community features enable a new form of community-organised ``rolling competition''.
+For instance, one can generate a dynamic top-ten list of systems on the problems with the highest votes, or a list of the hardest problems with the highest votes.
+These features can help the community to focus on the problems that most people find interesting.
+
+\subsection{Statistics}
+
+Additional statistics yield more interesting competition results, a deeper understanding of system behaviour, or greater comparability between different systems. Possibly interesting numbers include
+\begin{itemize}
+  \item Search tree size in search-based systems
+  \item Percentage of problems solved per system and category
+  \item Solution quality
+  \item Overall system quality
+\end{itemize}
+
+The exact metrics (in particular for features like system quality) are beyond the scope of this paper.
+The important point is that the platform we propose is extensible enough to support these features.
+
+\section{Applicability and Limits}
+\label{sec:limits}
+
+The competition platform as outlined in this paper certainly has its limits. It is mostly applicable to competitions with the following characteristics:
+\begin{itemize}
+  \item Highly dependent on running all systems on the exact same hardware
+  \item High ratio of instances to systems
+  \item ...?
+\end{itemize}

### 2009-12-18 trunk/Papers/crisp-ff-10/introduction.tex
intro, v1

\section{Introduction} \label{sec:introduction}

Natural language generation (NLG) -- the problem of computing a
sentence or text that communicates a given piece of information -- has
a long-standing connection to automated planning
\cite{perrault80,appelt:planning}. In the past few years, the idea has
been picked up again in the NLG community, both to use planning as an
approach for modeling language problems
\cite{Steedman-Petrick:07,benotti08b} and to tackle the search
problems in existing NLG problems by using techniques from planning
\cite{KolSto07}.

Existing planners are relatively good at solving modest instances of
the planning problems that arise in this context. However, Koller and
Petrick \shortcite{koller10:_exper_with_plann_for_natur_languag_gener}
recently showed that as the planning problem instances from Koller and
Stone \shortcite{KolSto07} scale up, off-the-shelf planners become too
slow to be practically useful. They noted that there are major sources
of inefficiency both in the search itself and in the preprocessor,
which is put under pressure by the sheer number of ground instances of
facts and actions that come up in this domain.

In this paper, we revisit this situation, analyze the sources of
inefficiency in more detail in the context of the FF planner
\cite{HoffmannNebel01}, and show how FF can be optimized to perform
dramatically better -- and in fact, well enough to be useful in
practical NLG applications.  We achieve this by tweaking the
implementation of the FF preprocessor and analyzing the behavior of
the enforced hillclimbing and best-first search strategies on the NLG
domain.  Neither of these changes degrades FF's performance on other
domains.

Nevertheless, we can still generate realistic instances of the NLG
planning domain on which our modified FF becomes too slow for
practical use.  As FF is still a competitive planner, this suggests
that NLG might be interesting as a future challenge domain.  In
particular, we show that by modifying the linguistic grammar from
which we generate, we can vary the planning problems quite widely and
emphasize different aspects that make our domain hard.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2009-12-18 trunk/Papers/crisp-ff-10/introduction.tex
revision of text reg ff and challenges

@@
 slow to be practically useful. They noted that there are major sources
-of inefficiency both in the search itself and in the preprocessor,
-which is put under pressure by the sheer number of ground instances of
-facts and actions that come up in this domain.
+of inefficiency both in the search itself and in the commonly used
+preprocesses, which are put under pressure by several features of the
+domain (e.g.\ comparatively large predicate and operator arities) that
+diverge from most IPC benchmarks.
+%the sheer number of
+%ground instances of facts and actions that come up in this domain.
 
@@
 inefficiency in more detail in the context of the FF planner
-\cite{HoffmannNebel01}, and show how FF can be optimized to perform
-dramatically better -- and in fact, well enough to be useful in
-practical NLG applications.  We achieve this by tweaking the
-implementation of the FF preprocessor and analyzing the behavior of
-the enforced hillclimbing and best-first search strategies on the NLG
-domain.  Neither of these changes degrades FF's performance on other
-domains.
+\cite{HoffmannNebel01}, and show how FF can be modified to perform
+dramatically better. We achieve this by tweaking the implementation of
+the FF preprocessor, and by acting on some simple but shocking
+observations regarding the behavior of some of FF's search techniques,
+in particular its goal ordering technique and the relaxed plan
+heuristic. The changes we make are all simple, but make FF perform
+well enough to be useful in practical NLG applications.
 
-Nevertheless, we can still generate realistic instances of the NLG
-planning domain on which our modified FF becomes too slow for
-practical use.  As FF is still a competitive planner, this suggests
-that NLG might be interesting as a future challenge domain.  In
-particular, we show that by modifying the linguistic grammar from
-which we generate, we can vary the planning problems quite widely and
-emphasize different aspects that make our domain hard.
+%the enforced hillclimbing and best-first search strategies on the NLG
+%domain.  Neither of these changes degrades FF's performance on other
+%domains.
+
+%Nevertheless
+
+Our performance gains with simple FF configuration changes certainly
+point out that automatic planner configuration may be quite
+useful. Apart from that, our domain remains an interesting challenge
+for the development of better heuristics. There still are realistic
+instances of the NLG planning domain on which our modified FF becomes
+too slow for practical use. Overcoming this weakness would entail
+overcoming the observed weaknesses of FF's search techniques --
+variants of which are used in many planners today.
+
+%As FF is still a competitive
+%planner, this suggests that NLG might be interesting as a future
+%challenge domain. 
+%
+% In particular, we show that by modifying the
+%linguistic grammar from which we generate, we can vary the planning
+%problems quite widely and emphasize different aspects that make our
+%domain hard.
 

### 2009-12-18 trunk/Papers/crisp-ff-10/crisp-ff.tex
pre-final version ff-sec
additional cites

@@
 
-Can we design heuristics that are better at deadling with the
-lifecycle of substitution nodes and distractors? The problem of not
-noticing them in the initial state is present for all approaches that
-are dominated by $h^+$,
-e.g.\ \cite{karpas:domshlak-ijcai-09,helmert:domshlak:icaps-09}, and
-is also present for the known inadmissible landmarks heuristics
-\cite{richter:etal:aaai-08}. As far as known heuristics are concerned,
+Can we design heuristics that are better at dealing with the lifecycle
+of substitution nodes and distractors? The problem of not noticing
+them in the initial state is present for all approaches that are
+dominated by $h^+$,
+e.g.\ \cite{karpas:domshlak-ijcai-09,helmert:domshlak:icaps-09}. The
+same problem is also present for the inadmissible landmarks heuristic
+\cite{richter:etal:aaai-08}, as well as for $h^{\mathsf{cea}}$
+\cite{helmert:geffner:icaps-08} and its variant $h^{\mathsf{pcc}}$
+\cite{cai:etal:icaps-09}, and hence also for $h^{\mathsf{CG}}$
+\cite{helmert:icaps-04}. As far as known heuristics are concerned,
 hope thus resides mainly in the $h^m$ family and in abstraction
@@
 %  yes}. 
-With merge\&shrink \cite{helmert:etal:icaps07}, we experimented with
-various settings for the parameters (number of abstractions and size
-limit for each abstraction) but could not find a setting that reduced
-the search nodes to less than $5000$. So, all in all, finding better
-heuristic functions for this kind of problem remains a challenge.
+% .. have tried IPP on NEUES-MODIFIER-SET but get seg faults and shit..
+% oh, whatever.
+With merge\&shrink \cite{helmert:etal:icaps07}, we obtained the
+estimate $11$ for the initial state, which is much better. Still, we
+could not find a parameter setting that reduced the search nodes to
+less than $5200$, which is not much better than blind search which
+took $13800$ nodes. So, all in all, finding better heuristic functions
+for this kind of problem remains a challenge.
 
@@
 %  would also be useful, to see how the search spaces develop}
+
+%NEUES-MODIFIER-SET$ ../ff -o modifiers-5-5-domain.lisp  -f modifiers-5-5-problem.lisp -T -H: 102
+%NEUES-MODIFIER-SET$ ../ff -o modifiers-5-5-domain.lisp  -f modifiers-5-5-problem.lisp -T -H -N: 282
+%NEUES-MODIFIER-SET$ ../ff -o modifiers-10-10-domain.lisp  -f modifiers-10-10-problem.lisp -T -H: 3078
+%NEUES-MODIFIER-SET$ ../ff -o modifiers-10-10-domain.lisp  -f modifiers-10-10-problem.lisp -T -H -N: 16875
+%NEUES-MODIFIER-SET$ ../ff -o modifiers-15-15-domain.lisp  -f modifiers-15-15-problem.lisp -T -H: ?? sehr lange
+%NEUES-MODIFIER-SET$ ../ff -o modifiers-15-15-domain.lisp  -f modifiers-15-15-problem.lisp -T -H -N: ?? sehr lange
 
@@
 larger examples, the factor between the two numbers tends to become
-larger. %{\bf here, again, having actual scaled examples would help}
-%{\bf use here larger version of
-%  modifiers-10-10???}
+larger.
 

### 2009-12-18 trunk/Papers/crisp-ff-10/abstract.tex
skipped last sentence
shorter, new title

@@
 natural language sentences. This domain has a number of features that
-make it significantly different from existing IPC benchmark
-domains. In particular, we show how several techniques used in the FF
-planner, which has been highly successful on IPC problems, are led
-astray. We discuss possible fixes to FF's shortcomings, some of which
-we have already implemented, but some of which pose interesting
-challenges to the AI Planning community. The performance of our
-modified FF is up to three orders of magnitude better than that of the
-original FF. While this makes FF much more practical for this
-application, we show that it is still very limited with respect to
-problem parameters such as the complexity of the underlying
-grammar. We outline how other planning domains might share the
-features that make the generation domain difficult. {\bf do we? how?}
+provoke fairly unusual behavior in existing planners. In particular,
+we show how several techniques used in the FF planner, which has been
+highly successful on IPC problems, are led astray. We discuss possible
+fixes to FF's shortcomings, some of which we have already implemented,
+but some of which pose interesting challenges to the AI Planning
+community. The performance of our modified FF is up to several orders
+of magnitude better than that of the original FF. While this makes FF
+much more practical for this application, we show that it is still
+quite limited with respect to the complexity of the underlying
+grammar.
+%We outline how other planning domains might share the
+%features that make the generation domain difficult. {\bf do we? how?}
 \end{abstract}

### 2009-12-18 trunk/Papers/crisp-ff-10/introduction.tex
shorter, new title

@@
 \cite{Steedman-Petrick:07,benotti08b} and to tackle the search
-problems in existing NLG problems by using techniques from planning
-\cite{KolSto07}.
+problems in NLG by using techniques from planning \cite{KolSto07}.
 
@@
 of inefficiency both in the search itself and in the commonly used
-preprocesses, which are put under pressure by several features of the
-domain (e.g.\ comparatively large predicate and operator arities) that
+preprocesses, due to several features of the domain
+(e.g.\ comparatively large predicate and operator arities) that
 diverge from most IPC benchmarks.
@@
 inefficiency in more detail in the context of the FF planner
-\cite{HoffmannNebel01}, and show how FF can be modified to perform
-dramatically better. We achieve this by tweaking the implementation of
-the FF preprocessor, and by acting on some simple but shocking
-observations regarding the behavior of some of FF's search techniques,
-in particular its goal ordering technique and the relaxed plan
-heuristic. The changes we make are all simple, but make FF perform
-well enough to be useful in practical NLG applications.
+\cite{HoffmannNebel01}, and show how to ``wake up the rabbit'': we
+modify FF to obtain dramatically better performance. We achieve this
+by tweaking the implementation of FF's preprocessor, and by acting on
+some simple but shocking observations regarding the behavior of some
+of FF's search techniques, in particular its goal ordering technique
+and the relaxed plan heuristic. The changes we make are all simple,
+but make FF perform well enough to be useful in practical NLG
+applications.
 
@@
 for the development of better heuristics. There still are realistic
-instances of the NLG planning domain on which our modified FF becomes
-too slow for practical use. Overcoming this weakness would entail
-overcoming the observed weaknesses of FF's search techniques --
-variants of which are used in many planners today.
+instances of the NLG planning domain on which our modified FF is not
+practical. Overcoming this weakness would entail overcoming the
+observed weaknesses of FF's search techniques -- variants of which are
+used in many planners today.
 

### 2009-12-18 trunk/Papers/crisp-ff-10/crisp-ff.tex
shorter ff sec

@@
 
-It turns out that, as far as the suitability of commonly used
-pre-processing and search techniques are concerned, sentence
-generation diverges significantly from most known benchmark
-domains. We examine this here, in detail, for the case of FF
-\cite{HoffmannNebel01}. We show how certain hitherto irrelevant
-details of the pre-processing suddenly become deadly obstacles to
-performance, and how certain widely used search techniques -- namely
-goal orderings and the relaxed plan heuristic -- are completely off
-the mark and partly even detrimental. For some of the observed
-pitfalls, we are able to provide simple fixes; others pose serious
-challenges to planning research. We focus first on the pre-processing,
-then on the search techniques.
+%It turns out that, as far as the suitability of commonly used
+%pre-processing and search techniques are concerned, sentence
+%generation diverges significantly from most known benchmark
+%domains. We examine this here, in detail, for the case of FF
+%\cite{HoffmannNebel01}. We show how certain hitherto irrelevant
+%details of the pre-processing suddenly become deadly obstacles to
+%performance, and how certain widely used search techniques -- namely
+%goal orderings and the relaxed plan heuristic -- are completely off
+%the mark and partly even detrimental. For some of the observed
+%pitfalls, we are able to provide simple fixes; others pose serious
+%challenges to planning research. We focus first on the pre-processing,
+%then on the search techniques.
+
+It turns out that FF's exhibits, in this domain, a number of
+detrimental phenomena that were previously unheard of. We now examine
+these in some detail.  For some of the observed pitfalls, we are able
+to provide simple fixes. Others pose serious challenges to planning
+research.
+
+%We focus first on the pre-processing,
+%then on the search techniques.
 
@@
 
-FF's pre-process contains two details that, for a long time until they
-were detected, were so detrimental in our domain that we were not able
-to seriously propose FF as a practical solution. Simple fixes to these
-details are so effective that FF is now a practical (if not perfect)
-possibility.
+FF's pre-process contains two details that, for a long time until the
+writing of this paper, were so detrimental that we were not able to
+seriously propose FF as a practical solution.
+%Simple fixes
+%to these details are so effective that FF is now a practical (if not
+%perfect) possibility.
 
-First, while pre-processing the ADL formulas, FF has a sub-routine
-that checks the instantiated formulas -- quantifiers already compiled
-into grounded conjunctions and disjunctions -- for tautologies. This
-involves a loop checking all pairs of sub-formulas within every
-conjunction and disjunction, testing whether they are
-identical/identical modulo their sign. For example, if we find $\phi$
-and $\neg \phi$ within a disjunction, then the disjunction is replaced
-by TRUE. Now, we haven't tested thoroughly whether this sub-routine
-has any notable effect on the formulas in other benchmarks. It
-certainly has no such effect in our domain. The effect it {\em does}
-have is take a huge amount of runtime on the long goal conjunctions
-stating that we do not wish to have any open substitution nodes nor
-any distractors. The simple fix is hence to switch this sub-routine
-off. In small instances, this gives advantages like $0.29s$ vs.\ $11s$
-pre-processing time; in a larger instance we ran, original FF's
-pre-process doesn't finish within 15 minutes whereas our fix yields a
-pre-processing time of about $3s$.
+While pre-processing the ADL formulas, FF has a sub-routine that
+checks the grounded formulas
+% -- quantifiers already compiled into
+%grounded conjunctions and disjunctions -- 
+for tautologies. This involves a loop checking all pairs of
+sub-formulas within every conjunction and disjunction, testing whether
+they are identical/identical modulo their sign. For example, if we
+find $\phi$ and $\neg \phi$ within a disjunction, then the disjunction
+is replaced by TRUE. Now, we haven't tested whether this sub-routine
+has a notable effect on the formulas in other benchmarks. It certainly
+has no such effect in our domain, other than taking a huge amount of
+runtime on the long goal conjunctions stating that we do not wish to
+have any open substitution nodes nor any distractors. The fix is to
+switch this sub-routine off. This trivial operation alone can render
+previously infeasible instances (pre-processing time $>15$ minutes)
+harmless (instance solved in $6$ seconds {\em total} time).
 
-Second, FF's pre-process contains a very small imprudence that
-apparently didn't surface in other domains, but that does in ours. FF
+%In small instances, this gives advantages like
+%$0.29s$ vs.\ $11s$ pre-processing time. In a larger instance we ran,
+%original FF's pre-process doesn't finish within 15 minutes whereas our
+%fix yields a pre-processing time of about $3s$.
+
+FF's pre-process contains a very small imprudence that apparently
+didn't surface in other domains, but that does in ours. FF
 distinguishes operators into ``easy'' (DNF precondition) and ``hard''
-(non-DNF precondition). The easy ones are instantiated in a
-spezialized very effectively implemented procedure, while the hard
-ones go through a much less effectively implemented more general
-procedure. In particular, that latter procedure uses a data structure
+(non-DNF precondition), which go through different instantiation
+procedures. The procedure for hard operators uses a data structure
 encompassing an integer for every possible instantiation of all
 predicates. Since this does not take into account reachability, it is
-prohibitively huge in our domain. Now, there actually are no hard
-operators in our domain. But FF builds the data structure anyway, even
-though it is not needed. In small instances, this results in a couple
-of seconds runtime overhead. In larger instances, it exhausts memory
-and causes FF to terminate with a segmentation fault.
+prohibitively huge in our domain. Now, we actually have no hard
+operators. But FF builds the data structure anyway. This sometimes
+results in a couple of seconds runtime overhead; it also sometimes
+exhausts memory and causes a segmentation fault.
 
@@
 \shortcite{koehler:hoffmann:jair-00} ``reasonable goal orderings'',
-variants and extensions of which are used in several planning systems
-today, e.g.\ \cite{hoffmann:etal:jair-04,richter:etal:aaai-08}. FF as
-used in the 2000 competition implements the ideas from
-\cite{koehler:hoffmann:jair-00}, approximating reasonable orders as a
-pre-process and then partitioning the goal set into a {\em goal
-  agenda}, a sequence of goal subsets. Each subset is posed to the
-planner in isolation, ignoring all other goals. The final plan is the
-concatenation of the plans for all entries.
+variants and extensions of which are in prominent use today, e.g.\ in
+LAMA \cite{richter:etal:aaai-08}. FF as used in the 2000 competition
+approximates reasonable orders as a pre-process and then partitions
+the goal set into a {\em goal agenda}, a sequence of goal
+subsets. Each subset is posed to the planner in isolation. The final
+plan is the concatenation of the plans for all entries.
 
@@
 
-While the goal agenda and related techniques have proved quite useful
-in many domains, in the sentence generation domain they often are
-completely off-target. The most striking point concerns the
-interaction between the goal to create a sentence -- $\neg
-\mathsf{subst}(S,\mathsf{root})$ in the rabbit example -- and the goal
-to express a certain meaning -- $\mathsf{sleep}(e,r_1)$ in the
-rabbit example. The goal agenda orders the former before the latter
-because it detects that, once a meaning is expressed, the sentence
-creation has started and it is not possible to start anew
-anymore. Unless the inverse also holds,\footnote{The inverse holds if,
-  in the given grammar, the only way to create a meaning is to create
-  a full sentence -- i.e., there are no side sentences like ``the
-  sleeping rabbit''.} $\mathsf{subst}(S,\mathsf{root})$ is alone in
-the first goal agenda entry, i.e., there is no goal stating the actual
-meaning the sentence is supposed to have. Consequently, the planner
-decides to generate {\em any} sentence. If it happens to be the wrong
-sentence, the planner is stuck in a dead-end. While that dead end is
-recognized by the relaxed plan heuristic -- i.e., it is easy to see
-that the actions needed for achieving the goal cannot be applied
-anymore -- certainly this is an entirely unintended (and a bit
-amusing) outcome of this technique. 
+The goal agenda is completely off-target in our domain. The most
+striking point concerns the interaction between the goal to create a
+sentence ($\neg \mathsf{subst}(S,\mathsf{root})$ in the rabbit
+example), and the goal to express a certain meaning
+($\mathsf{sleep}(e,r_1)$ in the rabbit example). The goal agenda
+orders the former before the latter because it detects that, once a
+meaning is expressed, the sentence creation has started and it is not
+possible to start anew anymore. Unless the inverse also
+holds,\footnote{The inverse holds if, in the given grammar, the only
+  way to create a meaning is to create a full sentence -- i.e., there
+  are no side sentences like ``the sleeping rabbit''.}
+$\mathsf{subst}(S,\mathsf{root})$ is alone in the first goal agenda
+entry, i.e., there is no goal stating the actual meaning the sentence
+is supposed to have. Consequently, the planner decides to generate
+{\em any} sentence. If it happens to be the wrong sentence, the
+planner is stuck in a dead-end. 
+%While that dead end is recognized by
+%the relaxed plan heuristic -- i.e., it is easy to see that the actions
+%needed for achieving the goal cannot be applied anymore -- certainly
+%this is an entirely unintended (and a bit amusing) outcome of this
+%technique.
 % Switching it off, we {\bf \dots
@@
 
-While the goal agenda meltdown may be more amusing than alarming, we
-next report that relaxed plan heuristics are not far from such a
-meltdown either. They are fatally confused by the weird lifecycle of
-facts encoding substitution nodes and distractors. In instances where
-these constructs are important for the construction of a valid
-sentence, these facts are required to be false in the goal, and most
-of them are initially false. However, any valid plan must make many of
-them true at some intermediate time point (e.g.\ distractors arise
-depending on the decisions taken by the planner). Hence, from the
-perspective of the relaxed plan heuristic, from one state to the next
-whole classes of goals suddenly appear ``out of the blue''. These
-goals are not accounted for in the preceding state, and hence the
-relaxed plan is completely un-informative.
+In our improved FF, the goal agenda is simply switched off. Having
+disposed of goal orderings in that way, let us turn our attention to
+the relaxed plan heuristic. %
+%While the goal agenda meltdown may be more amusing than alarming, we
+%next report that relaxed plan heuristics are not far from such a
+%meltdown either. They are fatally 
+This meets its Waterloo as well, due to the weird lifecycle of facts
+encoding substitution nodes and distractors. In instances where these
+constructs are important for the construction of a valid sentence,
+these facts are required to be false in the goal, and most of them are
+initially false. However, any valid plan must make many of them true
+at some intermediate time point (e.g.\ distractors arise depending on
+the decisions taken by the planner). Hence, from the perspective of
+the relaxed plan heuristic, from one state to the next whole classes
+of goals suddenly appear ``out of the blue''. 
+%These goals are not
+%accounted for at all in the preceding state.
+%, and hence the relaxed plan is
+%completely un-informative.
 
@@
 effect ``forall (?y - individual) (when (not (= ?y ?xself))
-(distractor ?subjnode ?y))'' which introduces new distractors. Those
+(distractor ?subjnode ?y))'' that introduces new distractors. Those
 must be removed, so they form ``new'' goals. In the given case, the
@@
 %longer than before -- arbitrarily longer, in general.  
-For FF's particular search algorithm, enforced hill-climbing (EHC),
-this means that {\em FF performs a blind breadth-first search}: EHC
+Given FF's particular search algorithm, enforced hill-climbing (EHC),
+this means that {\em FF turns into a blind breadth-first search}: EHC
 starts at $h$ value $1$ and thereafter needs to solve the whole
@@
 
-Can we design heuristics that are better at dealing with the lifecycle
-of substitution nodes and distractors? The problem of not noticing
-them in the initial state is present for all approaches that are
-dominated by $h^+$,
-e.g.\ \cite{karpas:domshlak-ijcai-09,helmert:domshlak:icaps-09}. The
-same problem is also present for the inadmissible landmarks heuristic
-\cite{richter:etal:aaai-08}, as well as for $h^{\mathsf{cea}}$
-\cite{helmert:geffner:icaps-08} and its variant $h^{\mathsf{pcc}}$
-\cite{cai:etal:icaps-09}, and hence also for $h^{\mathsf{CG}}$
-\cite{helmert:icaps-04}. As far as known heuristics are concerned,
-hope thus resides mainly in the $h^m$ family and in abstraction
-heuristics. We ran a few tests with an example like the rabbit one,
-except that there are 10 distractors. The Graphplan estimate for the
-initial state is $4$, i.e., $h^2$ is better than $h^+$ here. Still,
-$4$ is far away from the real goal distance $14$. %
+%Can we design heuristics that are better at dealing with the lifecycle
+%of substitution nodes and distractors? The problem of not noticing
+%them in the initial state is present for all approaches that are
+%dominated by $h^+$,
+%e.g.\ \cite{karpas:domshlak-ijcai-09,helmert:domshlak:icaps-09}. The
+%same problem is also present for the inadmissible landmarks heuristic
+%\cite{richter:etal:aaai-08}, as well as for $h^{\mathsf{cea}}$
+%\cite{helmert:geffner:icaps-08} and its variant $h^{\mathsf{pcc}}$
+%\cite{cai:etal:icaps-09}, and hence also for $h^{\mathsf{CG}}$
+%\cite{helmert:icaps-04}. As far as known heuristics are concerned,
+%hope thus resides mainly in the $h^m$ family and in abstraction
+%heuristics. We ran a few tests with an example like the rabbit one,
+%except that there are 10 distractors. The Graphplan estimate for the
+%initial state is $4$, i.e., $h^2$ is better than $h^+$ here. Still,
+%$4$ is far away from the real goal distance $14$. %
 %{\bf here I could use
@@
 % oh, whatever.
-With merge\&shrink \cite{helmert:etal:icaps07}, we obtained the
-estimate $11$ for the initial state, which is much better. Still, we
-could not find a parameter setting that reduced the search nodes to
-less than $5200$, which is not much better than blind search which
-took $13800$ nodes. So, all in all, finding better heuristic functions
-for this kind of problem remains a challenge.
+%With merge\&shrink \cite{helmert:etal:icaps07}, we obtained the
+%estimate $11$ for the initial state, which is much better. Still, we
+%could not find a parameter setting that reduced the search nodes to
+%less than $5200$, which is not much better than blind search ($13800$
+%nodes). All in all, finding better heuristic functions for this kind
+%of problem remains a challenge.
 
-%{\bf \dots
-%  (malte agreed to run a test on modifiers-10-10-).. larger versions
-%  would also be useful, to see how the search spaces develop}
+Can we design heuristics that are better at dealing with the lifecycle
+of substitution nodes and distractors? It is easy to see that the
+problem of not noticing them in the initial state is present for most
+known heuristics
+\cite{helmert:icaps-04,richter:etal:aaai-08,karpas:domshlak-ijcai-09,helmert:domshlak:icaps-09,cai:etal:icaps-09}. The
+two notable eceptions are the $h^m$ family, and abstraction
+heuristics. We ran a few tests with an example like the rabbit one,
+except that there are 10 distractors. The Graphplan estimate for the
+initial state is $4$, i.e., $h^2$ is better than $h^+$ here. Still,
+$4$ is far away from the real goal distance $14$. With merge\&shrink
+\cite{helmert:etal:icaps07}, we obtained the estimate $11$ for the
+initial state, which is much better. Still the search space with that
+latter heuristic remained fairly large ($5200$ nodes vs.\ $13800$
+nodes for blind search) and it is not clear how well the technique
+will scale to larger instances and instances with a more complex
+grammar. Certainly, interesting challenges remain.
 
@@
 plan is often much too short, it appears to always contain at least
-one of the actions that actually are useful, and it certainly prunes
-the set of avaibale actions quite effectively. For example, in the
-10-distractor test instance, $16875$ nodes are explored without
-helpful actions pruning; with pruning, this reduces to $3078$. For
-larger examples, the factor between the two numbers tends to become
-larger.
+one of the actions that actually are useful. As our experiments data
+will show, it certainly prunes the set of available actions quite
+effectively.
 
-Since EHC is often completely inadequate, and helpful actions are
-quite effective, our FF configuration directly runs best-first search
-with helpful actions (as opposed to original FF which switches to
-best-first with{\em out} helpful actions if EHC fails).
+%For example, in the
+%10-distractor test instance, $16875$ nodes are explored without
+%helpful actions pruning; with pruning, this reduces to $3078$. For
+%larger examples, the factor between the two numbers tends to become
+%larger.
+
+%Since EHC is often completely inadequate, and helpful actions are
+%quite effective,
+
+Given the above observations regarding EHC and helpful actions, our
+modified FF directly runs best-first search with helpful actions (as
+opposed to original FF which switches to best-first with{\em out}
+helpful actions if EHC fails).
 

### 2009-12-18 trunk/Papers/crisp-ff-10/experiments.tex
experiments v2, added pictures

--- 2009-12-11 trunk/Papers/crisp-ff-10/experiments.tex
@@
 
-.. describe before/after picture, claim that now the performance is
-``for the moment ok for this application'', but not for foreseeable
-future (bigger grammers etc)
+To evaluate the effect of our changes on the efficiency of the planner
+on practical inputs, we ran a series of experiments using the XTAG
+Grammar \cite{xtag01:_tr}, a large-scale tree-adjoining grammar of
+English, from which we selected all lexicon entries for seven words we
+used in the experiments. We set up inputs to generate sentences of the
+form ``$S_1$ and $S_2$ and \ldots and $S_n$'', i.e.\ a conjunction of
+$n$ sentences. Each sentence $S_i$ is of the form ``X admires Y''. X
+and Y are unique referring expressions to individuals, such as ``the
+rich businessman''. For each experiment, we have a parameter $d$ that
+specifies how many adjectives are needed to distinguish the individual
+uniquely -- in particular, $d=0$ means we can say ``the businessman'',
+and $d=2$ means we must say ``the rich sad businessman''. We translate
+these generation problems into planning problems with the new encoding
+as shown in Fig.~\ref{fig:white-rabbit-as-planning}.
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\columnwidth]{pics/xtag-k2-dist0}
+  \caption{Runtimes for $d=0$.}
+  \label{fig:runtimes-k2-dist0}
+\end{figure}
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\columnwidth]{pics/xtag-k2-dist2}
+  \caption{Runtimes for $d=2$.}
+  \label{fig:runtimes-k2-dist2}
+\end{figure}
+
+Consider first the results for $d=0$, which are shown in
+Fig.~\ref{fig:runtimes-k2-dist0}; the lines end where the planner
+exceeded its five-minute timeout.  The top line in the figure
+represents the off-the-shelf version of FF. The other two lines
+represent FF with the improved preprocessor, using the standard
+enforced hillclimbing strategy and the best-first search with helpful
+actions (``New FF'') respectively.  By improving the preprocessor, we
+have improved its runtime on small instances by three orders of
+magnitude; this factor only increases for larger inputs.  After the
+preprocessor has been improved, both search strategies can solve these
+problem instances in practically useful time -- for instance, the
+24-word sentence at $n=4$ is generated in under two seconds.
+
+To get a clearer view of the relative efficiency of the two search
+strategies, we also ran the experiment with $d=2$; the results are
+shown in Fig.~\ref{fig:runtimes-k2-dist2}.  Here the old FF
+preprocessor crashed with a segmentation fault even at
+$n=1$. Best-first search with helpful actions is clearly the
+best-performing search strategy here.  It takes about 14 seconds for
+$n=3$, but this sentence is already quite long at 29 words, and EHC
+takes 3.5 minutes on the same input.  Best-first search without
+helpful actions timed out at $n=3$, which illustrates the advantage we
+gain from using the helpful actions heuristic.
+
+While this brings the overall planner runtimes into a range where FF
+can now be useful in practical NLG applications, it is important to
+note that it still has serious limitations.  The best-first search
+timed out for any $n>4$ in either experiment.  While many NLG
+applications can get by with shorter sentences, improving the search
+on our domain is still an open question for the future.  Even within
+our domain, the relative quality of EHC and BFS+H depended on the
+exact generation problem instance: While BFS+H outperformed EHC at
+$d=2$, EHC was actually slightly faster at $d=0$ because the absence
+of distractors meant there was not much room for the characteristic
+underestimation of the relaxed plan evaluation discussed above.  This
+means that sentence generation remains as a varied and challenging
+domain for planning.
 

### 2009-12-18 trunk/Papers/crisp-ff-10/conclusion.tex
exps revise, conclusion

\section{Conclusion} \label{sec:conclusion}

Generation of natural language sentences is interesting, as a planning
problem, both in its application value and in the challenges it
poses. Our observations herein have above all the message that, as of
now, FF is a practical option for this application. It is baffling
with which ease this milestone came about, and certainly this suggests
to look more closely at automatic planner configuration techniques.

The domain is also far from ``solved'' in all its aspects. As we
pointed out, several features of the domain are very hard to capture
with existing search heuristics. In Fig.~\ref{fig:runtimes}, this
shows in the fact that no planner version scales beyond $n>4$ -- even
in case (a) where no adjectives at all are required. It remains an
interesting challenge to address this.

%While this brings the overall planner runtimes into a range where FF
%can now be useful in practical NLG applications, it is important to
%note that it still has serious limitations.  The best-first search
%timed out for any $n>4$ in either experiment.  While many NLG
%applications can get by with shorter sentences, improving the search
%on our domain is still an open question for the future.  Even within
%our domain, the relative quality of EHC and BFS+H depended on the
%exact generation problem instance: While BFS+H outperformed EHC at
%$d=2$, EHC was actually slightly faster at $d=0$ because the absence
%of distractors meant there was not much room for the characteristic
%underestimation of the relaxed plan evaluation discussed above.  This
%means that sentence generation remains as a varied and challenging
%domain for planning.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2009-12-18 trunk/Papers/crisp-ff-10/experiments.tex
exps revise, conclusion

--- 2009-12-11 trunk/Papers/crisp-ff-10/experiments.tex
@@
+\begin{figure*}[t]
+  \centering
+\begin{tabular}{cc}
+  \includegraphics[width=0.9\columnwidth]{pics/xtag-k2-dist0} & 
+\includegraphics[width=0.9\columnwidth]{pics/xtag-k2-dist2}\\[-0.2cm]
+(a) & (b)
+\end{tabular}
+\vspace{-0.3cm}
+\caption{Runtimes for $d=0$ (a) and $d=2$ (b).}
+  \label{fig:runtimes}
+\vspace{-0.5cm}
+\end{figure*}
+
 \section{Experiments} 
@@
 
-.. describe before/after picture, claim that now the performance is
-``for the moment ok for this application'', but not for foreseeable
-future (bigger grammers etc)
+To evaluate the effect of our changes on the efficiency of the planner
+on practical inputs, we ran a series of experiments using the XTAG
+Grammar \cite{xtag01:_tr}, a large-scale tree-adjoining grammar of
+English, from which we selected all lexicon entries for seven words we
+used in the experiments. We set up inputs to generate sentences of the
+form ``$S_1$ and $S_2$ and \ldots and $S_n$'', i.e.\ a conjunction of
+$n$ sentences. Each sentence $S_i$ is of the form ``X admires Y''. X
+and Y are unique referring expressions to individuals, such as ``the
+rich businessman''. For each experiment, we have a parameter $d$ that
+specifies how many adjectives are needed to distinguish the individual
+uniquely -- in particular, $d=0$ means we can say ``the businessman'',
+and $d=2$ means we must say ``the rich sad businessman''. 
+% JOERG: ``new encoding'' gibt's ja gar nicht mehr als unterscheidung
+%
+%We translate
+%these generation problems into planning problems with the new encoding
+%as shown in Fig.~\ref{fig:white-rabbit-as-planning}.
+
+%\begin{figure}[t]
+%  \centering
+%  \includegraphics[width=\columnwidth]{pics/xtag-k2-dist2}
+%  \caption{Runtimes for $d=2$.}
+%  \label{fig:runtimes-k2-dist2}
+%\end{figure}
+
+Consider first the results for $d=0$, shown in Fig.~\ref{fig:runtimes}
+(a); the lines end where the planner exceeded its five-minute timeout.
+``Old FF'' in the figure is original FF, ``New preprocessor'' applies
+only our fixes to FF's preprocess, ``New FF'' applies all fixes.
+% The other two lines represent FF with the improved preprocessor,
+%using the standard enforced hillclimbing strategy and the best-first
+%search with helpful actions (``New FF'') respectively.  
+Just the preprocessor fixes lead to a speed-up of at least $3$ orders
+of magnitude, leading to practically useful runtime -- for instance,
+the 24-word sentence at $n=4$ is generated in under two seconds. Note
+that, by contrast, the search configuration does not make much of a
+difference here. This changes in our next experiment.
+
+Results for $d=2$ are shown in Fig.~\ref{fig:runtimes} (b).  Here the
+old FF preprocessor crashed with a segmentation fault even at
+$n=1$. Best-first search with helpful actions is clearly the
+best-performing search strategy here.  It takes about 14 seconds for
+$n=3$, but this sentence is already quite long at 29 words, and EHC
+takes 3.5 minutes on the same input.  Best-first search without
+helpful actions timed out at $n=3$, illustrating the importance of
+that heuristic.
+
+We remark that the relative behavior of the different search methods
+is not uniform in our domain, which exhibits a huge amount of
+structural variance due to the many different forms that the input
+grammer may take. We are far from an exhaustive exploration of all the
+possible parameter settings. Suffice it to say that, in particular
+encoding variants and with particular grammars (e.g.\ grammars not
+featuring any distractors), in stark contrast to
+Fig.~\ref{fig:runtimes} (b) EHC is much more effective than best-first
+search.
+
+%While this brings the overall planner runtimes into a range where FF
+%can now be useful in practical NLG applications, it is important to
+%note that it still has serious limitations.  The best-first search
+%timed out for any $n>4$ in either experiment.  While many NLG
+%applications can get by with shorter sentences, improving the search
+%on our domain is still an open question for the future.  Even within
+%our domain, the relative quality of EHC and BFS+H depended on the
+%exact generation problem instance: While BFS+H outperformed EHC at
+%$d=2$, EHC was actually slightly faster at $d=0$ because the absence
+%of distractors meant there was not much room for the characteristic
+%underestimation of the relaxed plan evaluation discussed above.  This
+%means that sentence generation remains as a varied and challenging
+%domain for planning.
 

### 2009-12-18 trunk/Papers/crisp-ff-10/crisp-ff.tex
..

@@
 {\em any} sentence. If it happens to be the wrong sentence, the
-planner is stuck in a dead-end. 
+planner is stuck in a dead-end. %
 %While that dead end is recognized by
@@
 %example before/after.}
+It is not clear to us, at this point, how goal ordering techniques
+could be modified so that they return more sensible results.
 
-It is not clear to us, at this point, how goal ordering techniques
-could be modified so that they return sensible results in the sentence
-generation domain. The issue explained above regarding $\neg
-\mathsf{subst}(S,\mathsf{root})$ and $\mathsf{sleep}(e,r_1)$ could be
-addressed by recognizing that, once the latter is achieved, the former
-is already true anyway. More generally however, the life-cycle of
-substitution nodes and distractors is weird -- as we outline below, it
-follows a false-true-false pattern -- and seems difficult to detect
-based on domain analysis.
+%The issue regarding $\neg
+%\mathsf{subst}(S,\mathsf{root})$ and $\mathsf{sleep}(e,r_1)$ could
+%potentially be addressed by recognizing that, once the latter is
+%achieved, the former is already true anyway. Whether such techniques
+%suffice to obtain useful goal orderings overall remains to be seen.
+
+%More generally however,
+%as outlined below the life-cycle of substitution nodes and distractors
+%is weird; %
+% -- as we
+%outline below, it follows a false-true-false pattern --
+%it seems difficult to detect this based on domain analysis.
 

### 2009-12-18 trunk/Papers/reasoning-challenges/conclusion.tex
Some more details on infrastructure, some references (not cited yet)
new levels

@@
-\section{Conclusion} \label{sec:conclusion}
+\section{Conclusion}
+\label{sec:conclusion}
 
-%%% Local Variables: 
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:
+This new approach has advantages for each of the three groups of people involved in a competition, the organisers, the participants, and the audience (which usually includes organisers and participants).
+
+\begin{itemize}
+  \item Organisers benefit from a mostly automated platform that is maintained by the community, which should greatly reduce the effort needed to run a competition.
+  \item Participants benefit from a well-defined environment for running their systems, short feedback times after submitting a system, and a standardised test and benchmark platform for experimenting with novel algorithms.
+  \item The audience, i.e., the scientific community, benefits from the improved discussion and report facilities.
+The scientific value of competitions is to find out \emph{why} a system shows a certain performance, not only \emph{that} it does.
+\end{itemize}
+
+\subsubsection*{Acknowledgements}
+The original idea to organize a ``rolling'' competition in the CP community is due to Christian Schulte and Peter Stuckey.

### 2009-12-18 trunk/Papers/reasoning-challenges/levels.tex
new levels

\section{Levels of community involvement} \label{sec:levels}

We present our proposal for a new planning competition infrastructure
in terms of an ascending scale of levels, each of which makes it
easier or more rewarding for the community to participate in the
competition process.  As we will discuss below, we believe that the
competition format we propose is applicable in fields beyond planning
as well.  Therefore we will talk about \emph{systems}, \emph{problem
  classes}, and \emph{problem instances} instead of planners, planning
domains, and problems.

\subsection{Level 0: The current IPC}

As background and starting point of our proposal, consider the process
that underlies the current IPC competition.  Every two years, the IPC
committee appoints one or two organizers for each track of that year's
IPC competition.  These organizers assemble a number of problem
classes and instances, by developing them themselves or soliciting
them from the IPC committee or the community at large; sometimes this
requires extending the Planning Domain Description Language (PDDL,
\cite{PDDL}) to more advanced classes of problems.  After the
participants have developed their planners, they submit the source
code to the organizers, who then compile and run the planners on the
test machines.  Finally, the organizers prepare and present a report,
and the best-performing systems win awards.

\todo{Ron, can you fix this so it becomes true? - ak}

\subsection{Level 1: Collecting problems}

While this format for the IPC has been highly successful in the past
ten years, there is also a number of ways in which it can be improved
and streamlined.  One such direction concerns the way in which the
problem classes on which an IPC runs are collected.  There is no doubt
that the selection of IPC domains is a fundamental driving force for
the development of new planning algorithms.  Unfortunately, the
current process makes it difficult for users of planning technology,
who may come from a different community, to get a new
application-motivated domain accepted as an IPC benchmark.  This is
unsurprising, given that an IPC organizer must be able to justify
their selection of domains, and picking untested domains from other
communities is not the best way to do this.  However, this leads to a
selection process that is less dynamic and open to suggestions from
users than it could be.  A converse movement is that to keep the
number of domains that must be tested in each round of the competition
manageable, older domains are sometimes ``retired'', i.e.\ removed
from subsequent competitions.  This comes with a risk of a false sense
of progress, in which retired domains are remembered as ``solved''
although they were removed for completely independent reasons.
Finally, it has been argued \cite{ToughNuts} that some IPC organizers
tend to favor the development of new types of planning problems and
new dialects of PDDL over trying to find challenging benchmark
problems for existing versions of PDDL.

We propose that one way in which these problems can be alleviated is a
web repository in which any user can upload problem classes.  This
takes up Hoffmann's \shortcite{ToughNuts} idea to create a
community-driven repository for ``tough nuts'' -- i.e., instances that
were shown to be hard to solve in the past --, and takes it one step
further: In our proposal, \emph{all} IPC problem classes are suggested
by the community, much in line with the proposal by Roberts and Howe
\shortcite{roberts07:_call_for_partic}. This approach offers a
principled, publicly documented and accessible avenue for proposing
new problem classes, which benefits both application-oriented users of
planning technology and planning researchers who come up with tough
challenges for existing PDDL dialects. It also makes it easy to record
retired problem classes alongside more current ones, and makes it easy
to keep them accessible.

Submitting a problem class should involve three pieces of data. First,
the user must submit a collection of instances of the problem class or
a generator for such instances.  Second, the user must specify some
metadata including the class name and a human-readable description,
maybe with links to the application literature, complexity results,
and so on.  Third, the user must specify the \emph{type} of the
problem class, which determines what systems the data is suitable for
-- i.e., the language in which the problem is specified (such as a
particular dialect of PDDL or EUROPA \todo{cite}) and the type of
planning problem this is (propositional, temporal, uncertainty,
etc.).  If it is possible to convert between one specification
language and another automatically, this conversion could be performed
each time a problem is submitted.

Problem classes and instances can be submitted by any registered
user. In order to avoid abuse, the number of instances should be
limited (e.g.\ per hour), and a simple validity check should be
possible (such as running a parser if the format is PDDL).

\subsection{Level 2: Continuous evaluation}

But given that all the domains are available on a server, why stop
there?  We propose to make it possible to upload the \emph{planners}
to the same server as well, and then continuously and automatically
run the planners on the domains.  While the ability to upload domains
in Level 1 represents a conservative extension of the existing IPC
infrastructure, it is here that we depart more radically from the
existing format: Instead of a regular competition with organizers who
run the planner evaluation every other year, we propose a rolling
evaluation scheme, in which the current evaluation results would be
available online at any given time.

Of course, several issues need to be addressed to make this possible.
The first is how we can allow users to upload their systems and still
expect to run and maybe compile them automatically, without requiring
the manual intervention of an organizer if things break.  This problem
can be solved by imposing strict requirements on the execution
environment of the planner.  One way in which we could envision this
is by defining a single image for a virtual machine, such as
VirtualBox.\footnote{\url{http://www.virtualbox.org/}} When a planner
is evaluated, this happens in a fresh instance of this virtual
machine.  We can offer the virtual machine image for download; a
participant can then run the image on their own machine and ensure
that their planner works, in the exact environment in which it will be
evaluated.  This would even make it technically possible for
participants to submit not the source code, but a binary executable of
the planner (perhaps along with support files).  However, even if we
require the submission of source code, the move to a clean virtual
machine environment that contains only standard versions of tools and
libraries enforces at least a certain degree of portability from which
users will benefit when they download the planners.  Even when the
virtual machine image changes to reflect updates in the standard
development environments (new version of the C compiler, etc.), old
images could remain available, and so old planners would continue
running until the virtual machine emulator itself becomes unavailable
or incompatible.

We propose that every newly submitted planner is run automatically on
all available problem instances, and every newly submitted problem
instance is sent to all available planners.  The scheduling of this
process can be under the control of an automated system for continuous
integration, such as Hudson.\footnote{\url{http://hudson-ci.org/}}
Each run of a planner on a problem instance or problem class can take
place in a fresh instance of the virtual machine image.  The
evaluation infrastructure then inserts the problem data and the
planner into the virtual machine, runs the planner, measures
performance statistics (including which problems they could solve and
how much time this took them; see below), and stores the results in a
database.  Users will be able to view the results in dynamically
generated reports on the main website; there may also be functionality
to customize these reports to focus on certain types of problems.
Because all planners are run on all problem instances, problems in
which only a small and somewhat arbitrary selection of instances of a
certain problem are actually used in the evaluation \cite{pg2008002}
can be avoided.

In order to minimize security risks -- which arise when user-submitted
programs are being run automatically --, the individual virtual
machines need to be sandboxed quite tightly.  There is no need for an
individual virtual machine to be connected to any network, beyond
being able to enter performance measurements into a database.  By
further imposing resource limits on the number of available
processors, memory and processor time, and disk space, we believe the
potential for abuse is limited.  It can be limited further by only
allowing planner uploads to registered users.

While we see our proposed system's ability to generate up-to-date
reports automatically as a major strength, there may still be a
benefit in maintaining the current style of regular competitions at
discrete timepoints, e.g.\ to encourage the development of faster
planners through awards.  Such a competition could easily be based on
our system, by announcing a deadline and a specified subset of problem
classes. The selection of this subset could be informed, among other
things, by the report on problems that many planners can't
solve. Participants could then upload their planners at their leisure;
``running'' the competition amounts to taking a snapshot of the
performance reports on the selected problems when the deadline comes
and announcing the results.  By recording timestamps whenever a
performance measurements is made, this snapshot could remain available
on the website simply as one special report for a certain point in
time.  In such a case, the authors of planners should have the option
of marking them as ``non-competing'' to exclude them from the
competition snapshot, and maybe also from the up-to-date performance
reports.

\subsection{Level 3: Social networking}

As a final step, we justify the ``2.0'' in the title of our paper by
proposing the addition of mechanisms for social networking to the
continuous evaluation website.

The simplest such mechanism is to add Wiki pages to the evaluation
website.  This would allow the users who upload problem classes to
document these problems and link to existing literature; it would
allow the developers of planners to document these planners; and it
would make it easy for the organizers of regular competitions to add
more information to the snapshot reports.  The exchange of ideas and
opinions can be further facilitated by providing a discussion board
which links into these Wiki pages.

A second social networking tool which we believe would be useful in
the context of a planning evaluation website is to let users tag
problem classes and vote on them.  Any registered user can be allowed
to comment on any problem class and vote for how relevant, important,
or exciting they find it -- say, using a one to five star system as on
Amazon.  This has a number of crucial advantages.  First, an incoming
user or a competition organizer can take the average score as an
indication for the problems that are considered important in the
community.  Next to the difficulty of problems, this can be another
factor in the selection of problem classes for a competition snapshot.
Second, a relevance score for problem classes can also serve as a
guideline for the scheduling of planner runs: Assuming that there is
not enough computing power to run all incoming planners on all domains
in real time (e.g., just before a competition deadline), the runs on
popular domains can be prioritized.

\section{Implementation}

The obvious criticism that can be brought up against this proposal is
that of cost -- both in terms of implementing the necessary software
and in terms of running the hardware on which the continuous
evaluation is performed.

We believe that the actual hardware cost will be bearable.  While the
idea of running all planners on all domains sounds intimidating at
first, we don't anticipate that there will ever be more than a couple
dozen current versions of planners entered in the system.  This means
that any new domain that is entered will not be attempted to be solved
too many times.  Conversely, while there may (hopefully!) be a
significant number of domains in the system, and every new planner
must try to solve all domains whose types it claims to be able to
process, this cost can also be controlled by prioritizing popular or
hard domains.  A single computer should be sufficient for these needs;
if it turns out not to be, the virtual machines can easily be run on
machines anywhere.

implementing software

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2009-12-18 trunk/Papers/reasoning-challenges/conclusion.tex
a first complete draft

@@
 
-This new approach has advantages for each of the three groups of people involved in a competition, the organisers, the participants, and the audience (which usually includes organisers and participants).
+\todo{this is cut and paste from Guido's text, which I like but which
+  was in a different place. The conclusion should be cleaned up and
+  made real text. - ak}
+
+This new approach has advantages for each of the three groups of
+people involved in a competition, the organisers, the participants,
+and the audience (which usually includes organisers and
+participants). \todo{do we want to use the term ``stakeholders'' for
+  all of these people together? - ak}
 
@@
 
+\todo{I don't think we should have acknowledgments in the anonymous
+  version. - ak}
+
 \subsubsection*{Acknowledgements}
 The original idea to organize a ``rolling'' competition in the CP community is due to Christian Schulte and Peter Stuckey.
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2009-12-18 trunk/Papers/reasoning-challenges/levels.tex
a first complete draft

@@
 indication for the problems that are considered important in the
-community.  Next to the difficulty of problems, this can be another
-factor in the selection of problem classes for a competition snapshot.
-Second, a relevance score for problem classes can also serve as a
-guideline for the scheduling of planner runs: Assuming that there is
-not enough computing power to run all incoming planners on all domains
-in real time (e.g., just before a competition deadline), the runs on
-popular domains can be prioritized.
+community; imagine a ``top ten'' list of domains and planners on the
+website's front page.  Next to the difficulty of problems, this can be
+another factor in the selection of problem classes for a competition
+snapshot.  Second, a relevance score for problem classes can also
+serve as a guideline for the scheduling of planner runs: Assuming that
+there is not enough computing power to run all incoming planners on
+all domains in real time (e.g., just before a competition deadline),
+the runs on popular domains can be prioritized.
+
+\section{Evaluation measures}
+
+The continuous evaluation infrastructure we just presented is
+fundamentally agnostic to the type of concrete evaluation measures
+that it collects.  In particular, traditional measures such as the
+percentage of problems solved per system, average solution quality,
+average runtime and memory use, and search tree sizes can be
+collected -- with the caveat that the virtual machine must be
+configured so as to allow comparable runtime measurements.
+Furthermore, popularity scores from the social networking aspect can
+be incorporated, or different measures could be combined into an
+overall score.
 
@@
 
-implementing software
+On the other hand, we expect that implementing the software for
+running the continuous evaluation and presenting the results on a
+website would not be a prohibitive effort either.  Most of the
+components of the system (continuous integration server, virtual
+machines, Wiki) already exist as open-source off-the-shelf
+components.  We estimate that a prototype implementation of a complete
+system could be built in a person-month of work; surely this is the
+amount of work for which funding or volunteers could be found.
+
+\section{Applicability and limitations}
+
+One advantage of the evaluation infrastructure proposed above is that
+its applicability is not limited to the evaluation of planners.  Many
+reasoning tasks, including constraint solving and theorem proving, are
+being evaluated by running solvers over problem instances and
+reporting performance measures.  At the level of abstraction at which
+our continuous evaluation system could be implemented, there is no
+fundamental difference between the problem type ``propositional
+planning problem in PDDL 1.0'' and the problem type ``\todo{something
+  from CP or similar}'' -- in both cases, the system can collect
+problem classes and instances of this type and run them through
+solvers that declare themselves to accept such inputs.  This means
+that there is a potential for sharing development and maintenance
+costs for the evaluation platform among different communities.
+
+However, there are of course limitations to the type of problem for
+which it makes sense to run a community-driven continuous evaluation.
+For one thing, it must be acceptable in the community that every
+system runs on the same hardware and software configuration, as
+implemented in the virtual machine images.  On the other hand, our
+system makes most sense where the ratio of problem classes to systems
+is relatively high.  Neither of the two is true, for instance, for
+most natural language processing competitions: Different research
+groups tend to use a variety of programming languages and development
+environments to build their systems, and there are typically many more
+systems in any competition than there are problem classes (corpora).
+However, our approach is completely consistent with the requirements
+of a community where solving a single hard problem instance for the
+first time counts as an achievement, such as in theorem proving.
 

### 2009-12-19 trunk/Papers/reasoning-challenges/conclusion.tex
changes to abstract, intro, conclusions

--- 2009-12-09 trunk/Papers/reasoning-challenges/conclusion.tex
@@
-\section{Conclusion} \label{sec:conclusion}
+\section{Conclusion}
+\label{sec:conclusion}
+
+%\todo{this is cut and paste from Guido's text, which I like but which
+%  was in a different place. The conclusion should be cleaned up and
+%  made real text. - ak}
+
+The new direction we propose for the IPC has significant advantages for
+each of the competition's three main groups of stakeholders: the
+organisers, the participants, and the audience.  Organisers benefit from a
+mostly automated platform that is maintained by the community, which should
+greatly reduce the effort needed to run a competition. Participants benefit
+from a well-defined environment for running their systems, short feedback
+times for system evaluation, and a standardised test and benchmark platform
+for experimenting with novel algorithms. The audience -- in this case the
+wider planning and scientific communities -- benefits from improved
+discussion and reporting facilities. Since the scientific value of
+competitions is to find out \emph{why} a system performs the way it does,
+not only \emph{that} it performs a certain way, enabling users to
+contribute to the competition process reduces the chance that essential
+avenues of investigation and evaluation are missed. Instead, a robust and
+active competition community only serves to enhance the wider planning
+community as a whole.
+
+%This new approach has advantages for each of the three groups of
+%people involved in a competition, the organisers, the participants,
+%and the audience (which usually includes organisers and
+%participants). \todo{do we want to use the term ``stakeholders'' for
+%  all of these people together? - ak} %RP: I like stakeholders.
+
+%\todo{I don't think we should have acknowledgments in the anonymous
+%  version. - ak}
+%RP: I agree; the acknowledgements should be left out to anonymise the
+%paper
+
+%\subsubsection*{Acknowledgements}
+%The original idea to organize a ``rolling'' competition in the CP
+%community is due to Christian Schulte and Peter Stuckey.
 

### 2009-12-19 trunk/Papers/reasoning-challenges/introduction.tex
changes to abstract, intro, conclusions

@@
 This paper proposes a new direction for the IPC, building on the strength
-of previous competitions, but centred around a novel online infrastructure.
-Most notably, we envision a more interactive and potentially more
-wide-ranging competition featuring:
+of previous competitions and inspired by competitions in other research
+communities. Most notably, we envision an interactive and user-centric
+%potentially more wide-ranging
+competition, focused around a novel online infrastructure featuring:
+%Most notably, we envision a more interactive and potentially more
+%wide-ranging competition featuring:
 %
@@
 \item A centralised repository offering researchers access to the latest
-domains and planners, and a forum for community news, discussion, and
-interaction.
+domains and planners, and a social networking forum for community news,
+discussion, and interaction.
 \end{itemize}
@@
 proposed infrastructure nevertheless presents some technical challenges
-that must be addressed (e.g., security, , accessibility, maintenance,
+that must be addressed (e.g., security, accessibility, maintenance,
 etc.). However, we believe that for the most part, solutions to these
@@
 Most importantly, we believe this proposal offers significant benefits to
-competition organisers, participants, and the wider planning community.
-\textbf{[RP: say something more here]} Furthermore, since much of the
-infrastructure we describe here is not tied to a specific research area,
-attempts to establish similar competitions could benefit from this
-approach, leading to joint ventures with other research fields---and the
+the competition stakeholders: organisers, participants, and the wider
+planning community. Furthermore, since much of the infrastructure we
+describe is not tied to a specific research area, attempts to establish
+similar competitions could benefit from this approach, leading to joint
+ventures with research fields beyond planning---and the
 possibility of future inter-disciplinary challenges.
 
-In the remainder of the paper we outline our proposal for a new competition
-infrastructure, and the benefits we believe this approach could bring to
-the planning community and beyond.
+%RP: Alexander's first couple sentences in the next section are a good lead
+%in to the content of the paper. I don't think we need much or anything
+%as a transition here, especially if we're short on space.
+
+%In the remainder of the paper we outline our proposal for a new competition
+%infrastructure, and the benefits we believe this approach could bring to
+%the planning community and beyond.

### 2009-12-19 trunk/Papers/reasoning-challenges/levels.tex
changes to abstract, intro, conclusions

@@
 competition process. As we will discuss below, we believe that the
-competition format we propose is applicable in fields beyond planning
-as well.  Therefore we will talk about \emph{systems}, \emph{problem
-  classes}, and \emph{problem instances} instead of planners, planning
-domains, and problems.
+competition format we propose is applicable to fields beyond planning.  We
+will therefore refer to \emph{systems}, \emph{problem classes}, and
+\emph{problem instances}, instead of planners, planning domains, and
+planning problems.
 
@@
 
-As background and starting point of our proposal, consider the process
-that underlies the current IPC competition.  Every two years, the IPC
-committee appoints one or two organizers for each track of that year's
-IPC competition.  These organizers assemble a number of problem
-classes and instances, by developing them themselves or soliciting
-them from the IPC committee or the community at large; sometimes this
-requires extending the Planning Domain Description Language (PDDL,
-\cite{PDDL}) to more advanced classes of problems.  After the
-participants have developed their planners, they submit the source
-code to the organizers, who then compile and run the planners on the
-test machines.  Finally, the organizers prepare and present a report,
-and the best-performing systems win awards.
+As background, and a starting point for our proposal, consider the process
+that underlies the current IPC. Every two years, the IPC committee looks
+for volunteers to organize and run the various tracks of that year's
+competition. These organizers are responsible for assembling a number of
+problem classes and instances, by developing them themselves or soliciting
+them from the wider community. Sometimes this process results in, or
+coincides with, a new dialect of the Planning Domain Description Language
+(PDDL, \cite{PDDL}), which describes for participants the supported
+language for that year's competition. After the participants have developed
+their systems, they submit the source code to the organizers, who then
+compile and run the planners on the test machines.  Performance data is
+collected and analysed.  Finally, the organizers prepare and present a
+report, and the best-performing systems win awards.
 
-\todo{Ron, can you fix this so it becomes true? - ak}
-
-\subsection{Level 1: Collecting problems}
+\subsection{Level 1: Problem collection}
 
 While this format for the IPC has been highly successful in the past
-ten years, there is also a number of ways in which it can be improved
-and streamlined.  One such direction concerns the way in which the
-problem classes on which an IPC runs are collected.  There is no doubt
-that the selection of IPC domains is a fundamental driving force for
-the development of new planning algorithms.  Unfortunately, the
-current process makes it difficult for users of planning technology,
-who may come from a different community, to get a new
-application-motivated domain accepted as an IPC benchmark.  This is
-unsurprising, given that an IPC organizer must be able to justify
-their selection of domains, and picking untested domains from other
+ten years, there is also a number of ways in which it can be improved and
+streamlined.  One such direction concerns the way in which problem classes
+are collected.  There is no doubt that the existing collection of IPC
+domains is a fundamental driving force for the development of new planning
+algorithms.  Unfortunately, the current selection process makes it
+difficult for users of these systems, who may come from different
+communities, to get a new application-motivated domain accepted as an IPC
+benchmark.  This is unsurprising, given that IPC organizers must be able to
+justify their selection of domains, and picking untested domains from other
 communities is not the best way to do this.  However, this leads to a
-selection process that is less dynamic and open to suggestions from
-users than it could be.  A converse movement is that to keep the
-number of domains that must be tested in each round of the competition
-manageable, older domains are sometimes ``retired'', i.e.\ removed
-from subsequent competitions.  This comes with a risk of a false sense
-of progress, in which retired domains are remembered as ``solved''
-although they were removed for completely independent reasons.
-Finally, it has been argued \cite{ToughNuts} that some IPC organizers
-tend to favor the development of new types of planning problems and
-new dialects of PDDL over trying to find challenging benchmark
-problems for existing versions of PDDL.
+selection process that is less dynamic than it could be, and more
+restrictive for users wanting to contribute suggestions.  Conversely, to
+keep the number of domains that must be tested in each round of the
+competition manageable, older domains are sometimes ``retired'', i.e.\
+removed from subsequent competitions.  This comes with a risk of a false
+sense of progress, in which retired domains are remembered as ``solved''
+although they were removed for completely independent reasons.  Finally, it
+has been argued \cite{ToughNuts} that some IPC organizers tend to favor the
+development of new types of planning problems and new dialects of PDDL over
+trying to find challenging benchmark problems for existing versions of
+PDDL. Such approaches run the risk of ``reinventing'' past problem classes
+(albeit in different guises), or redirecting investigation away from
+current challenges.
 
-We propose that one way in which these problems can be alleviated is a
-web repository in which any user can upload problem classes.  This
-takes up Hoffmann's \shortcite{ToughNuts} idea to create a
-community-driven repository for ``tough nuts'' -- i.e., instances that
-were shown to be hard to solve in the past --, and takes it one step
-further: In our proposal, \emph{all} IPC problem classes are suggested
-by the community, much in line with the proposal by Roberts and Howe
-\shortcite{roberts07:_call_for_partic}. This approach offers a
-principled, publicly documented and accessible avenue for proposing
-new problem classes, which benefits both application-oriented users of
-planning technology and planning researchers who come up with tough
-challenges for existing PDDL dialects. It also makes it easy to record
-retired problem classes alongside more current ones, and makes it easy
-to keep them accessible.
+One way to alleviate these problems is by introducing a web repository
+where users can upload problem classes. This revisits Hoffmann's
+\shortcite{ToughNuts} idea to create a community-driven repository for
+``tough nuts'' -- i.e., instances that were shown to be hard to solve in
+the past -- and takes it one step further: in our proposal, \emph{all} IPC
+problem classes are provided by the community, much in line with the
+proposal by Roberts and Howe \shortcite{roberts07:_call_for_partic}. Such a
+repository offers a principled, publicly documented and accessible avenue
+for proposing new problem classes, which benefits both application-oriented
+users of planning technology and researchers who invent tough challenges
+for existing PDDL dialects. It also makes it easy to record retired problem
+classes alongside active ones, and keep them all accessible.
 
+Problem classes and instances can be submitted by any registered user. 
 Submitting a problem class should involve three pieces of data. First,
-the user must submit a collection of instances of the problem class or
-a generator for such instances.  Second, the user must specify some
+the user should submit a collection of instances of the problem class or
+a generator for such instances.  Second, the user should specify some
 metadata including the class name and a human-readable description,
-maybe with links to the application literature, complexity results,
-and so on.  Third, the user must specify the \emph{type} of the
-problem class, which determines what systems the data is suitable for
--- i.e., the language in which the problem is specified (such as a
-particular dialect of PDDL or EUROPA \todo{cite}) and the type of
-planning problem this is (propositional, temporal, uncertainty,
-etc.).  If it is possible to convert between one specification
-language and another automatically, this conversion could be performed
-each time a problem is submitted.
-
-Problem classes and instances can be submitted by any registered
-user. In order to avoid abuse, the number of instances should be
-limited (e.g.\ per hour), and a simple validity check should be
-possible (such as running a parser if the format is PDDL).
+possibly with links to the applicable literature, complexity results, and
+so on.  Third, the user should specify the \emph{type} of problem class,
+which determines what systems the data is suitable for, i.e., the language
+in which the problem is specified (such as a particular dialect of PDDL or
+EUROPA \todo{cite}) and the type of planning problem (e.g., propositional,
+temporal, uncertainty, etc.).  If it is possible to convert between one
+specification language and another automatically, this conversion could be
+performed each time a problem is submitted.
+In order to avoid abuse, the number of instances should be limited (e.g.,
+per hour), and a simple validity check should be possible (e.g., running a
+parser to verify that the format is PDDL).
 
@@
 
-But given that all the domains are available on a server, why stop
+Given that all the domains are available on a server, why stop
 there?  We propose to make it possible to upload the \emph{planners}
@@
 infrastructure, it is here that we depart more radically from the
-existing format: Instead of a regular competition with organizers who
+existing format: instead of a regular competition with organizers who
 run the planner evaluation every other year, we propose a rolling
@@
 database.  Users will be able to view the results in dynamically
-generated reports on the main website; there may also be functionality
-to customize these reports to focus on certain types of problems.
-Because all planners are run on all problem instances, problems in
-which only a small and somewhat arbitrary selection of instances of a
-certain problem are actually used in the evaluation \cite{pg2008002}
-can be avoided.
+generated reports on the main website; there may also be functionality to
+customize these reports to focus on certain types of problems.  Because all
+planners are run on all problem instances, and users can ensure that
+challenging problem instances are included in the competition, the chance
+of systems being evaluated on only a small and somewhat arbitrary selection
+of certain problems \cite{pg2008002} can be avoided.
 
-In order to minimize security risks -- which arise when user-submitted
-programs are being run automatically --, the individual virtual
+In order to minimize security risks, which can arise when user-submitted
+programs are being run automatically, the individual virtual
 machines need to be sandboxed quite tightly.  There is no need for an
@@
 benefit in maintaining the current style of regular competitions at
-discrete timepoints, e.g.\ to encourage the development of faster
+discrete timepoints, e.g., to encourage the development of faster
 planners through awards.  Such a competition could easily be based on
@@
 
-The obvious criticism that can be brought up against this proposal is
+The obvious criticism that can be brought against this proposal is
 that of cost -- both in terms of implementing the necessary software
@@
 One advantage of the evaluation infrastructure proposed above is that
-its applicability is not limited to the evaluation of planners.  Many
-reasoning tasks, including constraint solving and theorem proving, are
-being evaluated by running solvers over problem instances and
-reporting performance measures.  At the level of abstraction at which
-our continuous evaluation system could be implemented, there is no
-fundamental difference between the problem type ``propositional
-planning problem in PDDL 1.0'' and the problem type ``\todo{something
-  from CP or similar}'' -- in both cases, the system can collect
-problem classes and instances of this type and run them through
-solvers that declare themselves to accept such inputs.  This means
-that there is a potential for sharing development and maintenance
+its applicability is not limited to the evaluation of planners.
+%RP: I don't see why this needs to be restricted to "reasoning" tasks
+Many reasoning tasks, including constraint solving and theorem proving, are
+being evaluated by running solvers over problem instances and reporting
+performance measures.  At the level of abstraction at which our continuous
+evaluation system could be implemented, there is no fundamental difference
+between the problem type ``deterministic planning problem in PDDL 1.0'' and
+the problem type ``\todo{something from CP or similar}'' -- in both cases,
+the system can collect problem classes and instances of this type and run
+them through solvers that declare themselves to accept such inputs.  This
+means that there is a potential for sharing development and maintenance
 costs for the evaluation platform among different communities.
@@
 is relatively high.  Neither of the two is true, for instance, for
-most natural language processing competitions: Different research
+most natural language processing competitions: different research
 groups tend to use a variety of programming languages and development

### 2009-12-19 trunk/Papers/crisp-ff-10/conclusion.tex
4 pages. whew.

@@
 
-Generation of natural language sentences is interesting, as a planning
+Generation of natural language sentences is interesting as a planning
 problem, both in its application value and in the challenges it
-poses. Our observations herein have above all the message that, as of
-now, FF is a practical option for this application. It is baffling
-with which ease this milestone came about, and certainly this suggests
-to look more closely at automatic planner configuration techniques.
+poses. After the changes described in this paper, FF is a practical
+option for this application. It is striking how simple these changes
+were, and certainly this suggests to look more closely at automatic
+planner configuration techniques.
 

### 2009-12-19 trunk/Papers/crisp-ff-10/crisp-ff.tex
4 pages. whew.

@@
 
-%It turns out that, as far as the suitability of commonly used
-%pre-processing and search techniques are concerned, sentence
-%generation diverges significantly from most known benchmark
-%domains. We examine this here, in detail, for the case of FF
-%\cite{HoffmannNebel01}. We show how certain hitherto irrelevant
-%details of the pre-processing suddenly become deadly obstacles to
-%performance, and how certain widely used search techniques -- namely
-%goal orderings and the relaxed plan heuristic -- are completely off
-%the mark and partly even detrimental. For some of the observed
-%pitfalls, we are able to provide simple fixes; others pose serious
-%challenges to planning research. We focus first on the pre-processing,
-%then on the search techniques.
+FF exhibits a number of problems in this domain that were previously
+unknown.  The basic observation of FF's inefficiency on these inputs
+was made before
+\cite{koller10:_exper_with_plann_for_natur_languag_gener}; here we
+examine its causes in detail. For some of the observed pitfalls, we
+are able to provide simple fixes. Others pose serious challenges to
+planning research.
 
-It turns out that FF's exhibits, in this domain, a number of
-detrimental phenomena that were previously unheard of. We now examine
-these in some detail.  For some of the observed pitfalls, we are able
-to provide simple fixes. Others pose serious challenges to planning
-research.
-
-%We focus first on the pre-processing,
-%then on the search techniques.
-
-\subsection{Pre-Processing}
+\subsection{Preprocessing}
 \label{sec:crisp-ff:preprocess}
@@
 
-FF's pre-process contains two details that, for a long time until the
-writing of this paper, were so detrimental that we were not able to
-seriously propose FF as a practical solution.
+FF's preprocessor contains two details that, before this paper, were
+so detrimental that we were not able to seriously propose FF as a
+practical solution.
 %Simple fixes
@@
 
-While pre-processing the ADL formulas, FF has a sub-routine that
+While preprocessing the ADL formulas, FF has a subroutine that
 checks the grounded formulas
@@
 find $\phi$ and $\neg \phi$ within a disjunction, then the disjunction
-is replaced by TRUE. Now, we haven't tested whether this sub-routine
-has a notable effect on the formulas in other benchmarks. It certainly
-has no such effect in our domain, other than taking a huge amount of
-runtime on the long goal conjunctions stating that we do not wish to
-have any open substitution nodes nor any distractors. The fix is to
-switch this sub-routine off. This trivial operation alone can render
-previously infeasible instances (pre-processing time $>15$ minutes)
-harmless (instance solved in $6$ seconds {\em total} time).
+is replaced by TRUE. We haven't tested whether this subroutine has a
+notable effect in other benchmarks. In our domain, the only effect it
+has is to take a huge amount of runtime on the long goal conjunctions
+stating that we do not want any open substitution nodes or
+distractors. The fix is to switch this subroutine off. This trivial
+operation alone can render previously infeasible instances
+(preprocessing time $>15$ minutes) harmless (instance solved in $6$
+seconds {\em total} time).
 
@@
 
-FF's pre-process contains a very small imprudence that apparently
+FF's preprocessor contains a very small imprudence that apparently
 didn't surface in other domains, but that does in ours. FF
@@
 prohibitively huge in our domain. Now, we actually have no hard
-operators. But FF builds the data structure anyway. This sometimes
-results in a couple of seconds runtime overhead; it also sometimes
-exhausts memory and causes a segmentation fault.
+operators. But FF builds the data structure anyway, which exhausts
+memory and causes a segmentation fault on serious instances.
 
@@
 
-FF's pre-processing pitfalls -- and the fact that they didn't occur to
+FF's preprocessing pitfalls -- and the fact that they didn't occur to
 anyone before -- are certainly baffling, although easily fixed. We now
@@
 example), and the goal to express a certain meaning
-($\mathsf{sleep}(e,r_1)$ in the rabbit example). The goal agenda
-orders the former before the latter because it detects that, once a
-meaning is expressed, the sentence creation has started and it is not
-possible to start anew anymore. Unless the inverse also
-holds,\footnote{The inverse holds if, in the given grammar, the only
-  way to create a meaning is to create a full sentence -- i.e., there
-  are no side sentences like ``the sleeping rabbit''.}
+($\f{expressed}(\f{sleep},e,r_1)$). The goal agenda orders the former
+before the latter because it detects that, once a meaning is
+expressed, the sentence creation has started and it is not possible to
+start anew anymore. Unless the inverse also holds,\footnote{The
+  inverse holds if, in the given grammar, the only way to communicate
+  a meaning is in the main clause -- i.e., there are no subordinate
+  clauses like ``the sleeping rabbit''.}
 $\mathsf{subst}(S,\mathsf{root})$ is alone in the first goal agenda
-entry, i.e., there is no goal stating the actual meaning the sentence
-is supposed to have. Consequently, the planner decides to generate
-{\em any} sentence. If it happens to be the wrong sentence, the
-planner is stuck in a dead-end. %
+entry. Consequently, the planner decides to generate {\em any}
+sentence, with some arbitrary meaning. If it happens to be the wrong
+sentence, the planner is stuck in a dead-end. %
 %While that dead end is recognized by
@@
 %meltdown either. They are fatally 
-This meets its Waterloo as well, due to the weird lifecycle of facts
-encoding substitution nodes and distractors. In instances where these
-constructs are important for the construction of a valid sentence,
-these facts are required to be false in the goal, and most of them are
+This meets its Waterloo as well, due to the unusual lifecycle of facts
+encoding substitution nodes and distractors. 
+%In instances where these
+%constructs are important for the construction of a valid sentence,
+These facts are required to be false in the goal, and most of them are
 initially false. However, any valid plan must make many of them true
@@
 plan has length $1$ because we only need to achieve $\neg
-\mathsf{subst}(S,\mathsf{root})$ and $\mathsf{sleep}(e,r_1)$, which is
-done by the single action ``sleeps''. However, that action has the
-effect ``forall (?y - individual) (when (not (= ?y ?xself))
-(distractor ?subjnode ?y))'' that introduces new distractors. Those
-must be removed, so they form ``new'' goals. In the given case, the
-relaxed plan now contains two actions to achieve these goals. In
+\mathsf{subst}(S,\mathsf{root})$ and $\f{expressed}(\f{sleep},e,r_1)$,
+which is done by the single action ``sleeps''. However, that action
+has the effect $\forall y. y \neq x_1 \rightarrow
+\f{distractor}(u_1,y)$, which introduces new distractors. Thus the
+relaxed plan for the second state contains \emph{two} actions. In
 general (if there are $n$ rabbits, say), the relaxed plan may become
@@
 %longer than before -- arbitrarily longer, in general.  
-Given FF's particular search algorithm, enforced hill-climbing (EHC),
+For FF's particular search algorithm, enforced hill-climbing (EHC),
 this means that {\em FF turns into a blind breadth-first search}: EHC
@@
 dead-ends, i.e., dead-end states that have a relaxed plan. An example
-is the initial state of a task that features distractors for which
-there is no means of removal -- ignoring the distractors, the initial
-$h$ value will still be $1$. So our domain is in the most difficult
-class of Hoffmann's \shortcite{hoffmann:jair-05} ``planning domain
-taxonomy''. While it shares this property with several other
-challenging domains, the extremity of turning EHC into a breadth-first
-search is unheard of.
+is the initial state of a generation problem that it unsolvable
+because not all distractors can be removed; its $h$ value will still
+be $1$. So our domain is in the most difficult class of Hoffmann's
+\shortcite{hoffmann:jair-05} ``planning domain taxonomy''. While it
+shares this property with several other challenging domains, the
+extremity of turning EHC into a breadth-first search is unheard of.
 
@@
 two notable eceptions are the $h^m$ family, and abstraction
-heuristics. We ran a few tests with an example like the rabbit one,
-except that there are 10 distractors. The Graphplan estimate for the
-initial state is $4$, i.e., $h^2$ is better than $h^+$ here. Still,
-$4$ is far away from the real goal distance $14$. With merge\&shrink
+heuristics. We ran a few tests with a variant of the example problem
+involving 10 rabbits.  The Graphplan estimate for the initial state is
+$4$, i.e., $h^2$ is better than $h^+$ here (although still far from
+the real goal distance of $14$). With merge\&shrink
 \cite{helmert:etal:icaps07}, we obtained the estimate $11$ for the
@@
 latter heuristic remained fairly large ($5200$ nodes vs.\ $13800$
-nodes for blind search) and it is not clear how well the technique
+nodes for blind search), and it is not clear how well the technique
 will scale to larger instances and instances with a more complex
@@
 
-The only FF technique that appears to work reasonably well is its
+The only FF technique that appears to be clearly useful here is its
 action pruning technique, the ``helpful actions''. While the relaxed
 plan is often much too short, it appears to always contain at least
-one of the actions that actually are useful. As our experiments data
-will show, it certainly prunes the set of available actions quite
-effectively.
-
+one of the actions that actually are useful. 
+%
+%As our experiments data
+%will show, it certainly prunes the set of available actions quite
+%effectively.
+%
 %For example, in the
@@
 %larger. 
-
+%
+%
 %Since EHC is often completely inadequate, and helpful actions are
 %quite effective, 
-
-Given the above observations regarding EHC and helpful actions, our
-modified FF directly runs best-first search with helpful actions (as
-opposed to original FF which switches to best-first with{\em out}
+%
+Given the above observations, we modified FF to directly run
+best-first search with helpful actions (as
+% regarding EHC and helpful actions, our
+%modified FF directly runs best-first search with helpful actions (as
+opposed to original FF, which switches to best-first with{\em out}
 helpful actions if EHC fails).

### 2009-12-19 trunk/Papers/crisp-ff-10/crisp.tex
4 pages. whew.

--- 2009-12-09 trunk/Papers/crisp-ff-10/crisp-ff.tex
@@
-\section{Experiences with FF}
-\label{sec:crisp-ff}
+\section{Sentence generation as planning}
+\label{sec:crisp}
+
+We consider a planning domain that comes up in the generation of
+natural language sentences based on tree-adjoining grammar (TAG;
+\cite{joshi;etal1997}).  This is a standard problem in computational
+linguistics, which we sketch by example; see
+\cite{Stone2003a,KolSto07} for details.
+
+% \begin{figure}
+%   \centering
+%   \includegraphics[width=0.75\columnwidth]{pic-grammar}
+%   \caption{An example grammar in the sentence generation domain.}
+%   \label{fig:white-rabbit-sleeps-grammar}
+% \end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1\columnwidth]{pic-derivation}
+  \caption{Derivation of ``The white rabbit sleeps.''}
+  \label{fig:white-rabbit-sleeps-deriv}
+\end{figure}
+
+A TAG grammar consists of a finite set of \emph{elementary trees},
+each of which contains one or more words; the left of
+Fig.~\ref{fig:white-rabbit-sleeps-deriv} shows three trees for ``the
+rabbit'', ``sleeps'', and ``white''.  Trees can be combined using the
+operations of \emph{substitution} (purple dashed arrow in the figure)
+and \emph{adjunction} (blue dotted arrows) to form larger trees.
+The end result of a grammatically correct derivation is a tree all of
+whose leaves are labeled with words, as on the right of
+Fig.~\ref{fig:white-rabbit-sleeps-deriv}. We can read off a sentence
+from such a tree from left to right.
+
+To use TAG grammars for generation \cite{Stone2003a}, we assume a set
+of ground atoms expressing the information we want the sentence to
+convey, such as $\{\mathsf{sleep}(e,r_1)\}$, and a knowledge base that
+contains all ground atoms we know to be true; say,
+$\{\mathsf{sleep}(e,r_1), \mathsf{rabbit}(r_1), \mathsf{rabbit}(r_2),
+\mathsf{white}(r_1), \mathsf{brown}(r_2)\}$.  We then add variables
+ranging over constants from the knowledge base to the nodes of each
+elementary tree, and equip every elementary tree with a set of atoms
+over these variables to encode the meaning this tree expresses.
+Fig.~\ref{fig:white-rabbit-sleeps-deriv} already shows specific
+instances of the elementary trees from the grammar, in which constants
+have been substituted for these variables.  The derivation in the
+figure conveys the intended meaning -- in particular, that $r_1$
+sleeps.  Crucially, it also describes uniquely who does the sleeping:
+The sentence ``the rabbit sleeps'' would not have done this, as ``the
+rabbit'' could be understood either as $r_1$ or $r_2$.  We say that
+$r_2$ is a \emph{distractor} for the subject of the sentence, which is
+removed by adjoining the tree for ``white''.  The sentence generation
+problem consists in computing a grammatically correct TAG derivation
+from instances of the elementary trees in the grammar that conveys the
+intended meaning and describes all referents uniquely.
+
+\subsection{Sentence generation as a planning problem}
+
+The TAG sentence generation problem can be encoded as a planning
+problem \cite{KolSto07}.  The key idea is that each operator encodes
+the addition of an elementary tree to the derivation; the syntactic
+and semantic requirements and effects of doing this are captured in
+the operator's precondition and effect.
+
+\newcommand{\action}[4]{\textbf{#1$(#2)$:}\\
+\strut\quad   Precond:$\;$ \parbox[t]{12cm}{\ensuremath{#3}}\\
+\strut\quad   Effect:$\;$ \parbox[t]{12cm}{\ensuremath{#4}}}
+\newcommand{\f}[1]{\mathsf{#1}}
+
+\begin{figure}
+%\centering
+%\begin{minipage}{0.8\textwidth}
+{\small%
+\action{sleeps}{u, u_1, u_n, x_0, x_1}{
+  \f{subst}(\f{S},u) \wedge \f{ref}(u,x_0) \wedge
+  \f{sleep}(x_0,x_1) \\ \wedge \f{current}(u_1) \wedge
+  \f{next}(u_1,u_n)
+}{
+  \neg \f{subst}(\f{S},u) \wedge \f{expressed}(\f{sleep}, x_0, x_1) \\
+  \wedge \f{subst}(\f{NP},u) \wedge \f{ref}(u_1,x_1) \\ \wedge
+  \neg \f{current}(u_1) \wedge \f{current}(u_n) \\ \wedge
+  \forall y. y \neq x_1 \rightarrow \f{distractor}(u_1,y)
+}\\
+
+\action{rabbit}{u, x_0}{
+  \f{subst}(\f{NP},u) \wedge \f{ref}(u,x_0) \wedge \f{rabbit}(x_0)
+}{
+  \neg \f{subst}(\f{NP},u) \wedge \f{canadjoin}(\f{N},u) \\
+  \wedge \forall y. \neg \f{rabbit}(y) \rightarrow \neg \f{distractor}(u,y)
+}\\
+
+\action{white}{u,x_0}{
+  \f{canadjoin}(\f{N},u) \wedge \f{ref}(u,x_0) \wedge \f{rabbit}(x_0)
+}{
+  \forall y. \neg \f{white}(y) \rightarrow \neg \f{distractor}(u,y)
+}
+}\strut\\[-5ex]
+%\end{minipage}
+\caption{Actions for generating ``The white rabbit
+sleeps.''}
+\label{fig:white-rabbit-as-planning}
+\end{figure}
+
+More precisely, each operator has a parameter $u$ representing the
+syntax node into which the elementary tree is substituted or adjoined,
+and a parameter $u_i$ for each substitution node.  There are also
+parameters $x_0,\ldots,x_k$ for the variables in the semantic
+representation of the elementary tree; $x_0$ is the variable that
+occurs at the root of the tree.  The planning state encodes a list of
+possible node names; $\f{current}(u)$ expresses that $u$ is the next
+node name that should be picked when a new substitution node is
+created, and $\f{next}(u,v)$ says that the next node after $u$ is
+$v$. These atoms are used to select names for the substitution nodes
+that are introduced by adding an elementary tree; the parameter $u_n$
+represents the node that will be current after the addition.
+
+The atom $\f{subst}(A,u)$ expresses that we need to still substitute
+something into the substitution node $u$ with label $A$; the initial
+state contains an atom $\f{subst}(\f{S},\f{root})$ where $\f{S}$
+stands for ``sentence'' and $\f{root}$ is an arbitrary name for the
+root of the derivation.  The mapping from nodes to semantic constants
+is maintained using the $\f{ref}$ atoms; the initial state for
+generating a sentence about the event $e$ contains an atom
+$\f{ref}(\f{root},e)$.  Finally, we keep track of the uniqueness of
+referring expressions in the $\f{distractor}$ atoms:
+$\f{distractor}(u,a)$ expresses that the expression at syntax node $u$
+could still be misunderstood as $a$.
+
+The example generation problem above translates into a planning
+problem whose domain is shown, in PDDL format, in
+Fig.~\ref{fig:white-rabbit-as-planning}. The initial state of the
+problem encodes the generation problem's knowledge base; it contains
+atoms $\f{rabbit}(r_1)$, $\f{sleep}(e,r_1)$, etc.  The goal requires
+syntactic completeness as $\forall u \forall A \neg \f{subst}(A,u)$
+and unique reference as $\forall u \forall x \neg
+\f{distractor}(u,x)$; it also specifies the semantic representation we
+want to express in an atom $\f{expressed}(\mathsf{sleep},e,r_1)$.
+
+A minimal plan for the example problem is
+$\mathsf{sleeps}(\mathsf{root}, n_1, n_2, e, r_1)$;
+$\mathsf{rabbit}(n_1, r_1)$; $\mathsf{white}(n_1, r_1)$. This plan can
+be automatically decoded into the derivation shown in
+Fig.~\ref{fig:white-rabbit-sleeps-deriv}.  The analogous plan with
+$r_2$ instead of $r_1$ is not correct because the $\mathsf{sleep}$
+precondition of the $\mathsf{sleeps}$ operator is not satisfied.  The
+first two steps of the plan alone would not be a correct plan because
+they leave an atom $\mathsf{distractor}(n_1, r_2)$ in the state, which
+contradicts the goal that all distractors have been eliminated.
 

### 2009-12-19 trunk/Papers/crisp-ff-10/experiments.tex
4 pages. whew.

@@
   \centering
-\begin{tabular}{cc}
-  \includegraphics[width=0.9\columnwidth]{pics/xtag-k2-dist0} & 
-\includegraphics[width=0.9\columnwidth]{pics/xtag-k2-dist2}\\[-0.2cm]
-(a) & (b)
-\end{tabular}
-\vspace{-0.3cm}
+(a)
+\vtop{
+\vskip-2ex
+\hbox{\includegraphics[width=0.8\columnwidth]{pics/xtag-k2-dist0}}
+}
+$\qquad\qquad$
+(b)
+\vtop{
+\vskip-2ex
+\hbox{\includegraphics[width=0.8\columnwidth]{pics/xtag-k2-dist2}}
+}
+\vspace{-0.5cm}
 \caption{Runtimes for $d=0$ (a) and $d=2$ (b).}
@@
 $n$ sentences. Each sentence $S_i$ is of the form ``X admires Y''. X
-and Y are unique referring expressions to individuals, such as ``the
-rich businessman''. For each experiment, we have a parameter $d$ that
-specifies how many adjectives are needed to distinguish the individual
+and Y are unique referring expressions, such as ``the rich
+businessman''. For each experiment, we have a parameter $d$ that
+specifies how many adjectives are needed to distinguish the referent
 uniquely -- in particular, $d=0$ means we can say ``the businessman'',
@@
 
-We remark that the relative behavior of the different search methods
-is not uniform in our domain, which exhibits a huge amount of
-structural variance due to the many different forms that the input
-grammer may take. We are far from an exhaustive exploration of all the
-possible parameter settings. Suffice it to say that, in particular
-encoding variants and with particular grammars (e.g.\ grammars not
-featuring any distractors), in stark contrast to
-Fig.~\ref{fig:runtimes} (b) EHC is much more effective than best-first
-search.
+The relative behavior of the different search methods is not uniform
+in our domain, which exhibits a huge amount of structural variance due
+to the many different forms that the input grammar may take. We are
+far from an exhaustive exploration of all the possible parameter
+settings. For instance, with particular inputs (no distractors), EHC
+can be more effective than best-first search.
+
+%Suffice it to say that, in particular encoding variants and
+%with particular grammars (e.g.\ grammars not featuring any
+%distractors), in stark contrast to Fig.~\ref{fig:runtimes} (b) EHC is
+%much more effective than best-first search.
 

### 2009-12-19 trunk/Papers/crisp-ff-10/introduction.tex
minor detail
minor
4 pages. whew.

@@
 sentence or text that communicates a given piece of information -- has
-a long-standing connection to automated planning
-\cite{perrault80,appelt:planning}. In the past few years, the idea has
-been picked up again in the NLG community, both to use planning as an
-approach for modeling language problems
-\cite{Steedman-Petrick:07,benotti08b} and to tackle the search
+a long-standing connection to automated planning \cite{perrault80}. In
+the past few years, the idea has been picked up again in the NLG
+community, both to use planning as an approach for modeling language
+problems (e.g., \cite{Steedman-Petrick:07}) and to tackle the search
 problems in NLG by using techniques from planning \cite{KolSto07}.
@@
 inefficiency in more detail in the context of the FF planner
-\cite{HoffmannNebel01}, and show how to ``wake up the rabbit'': we
-modify FF to obtain dramatically better performance. We achieve this
-by tweaking the implementation of FF's preprocessor, and by acting on
-some simple but shocking observations regarding the behavior of some
-of FF's search techniques, in particular its goal ordering technique
-and the relaxed plan heuristic. The changes we make are all simple,
-but make FF perform well enough to be useful in practical NLG
-applications.
+\cite{HoffmannNebel01}, and show how to modify FF to obtain
+dramatically better performance. We achieve this by tweaking the
+implementation of FF's preprocessor, and by acting on some simple but
+shocking observations regarding the behavior of some of FF's search
+techniques, in particular its goal ordering technique and the relaxed
+plan heuristic. The changes we make are all simple, but make FF
+perform well enough to be useful in practical NLG applications.
 

### 2009-12-19 trunk/Papers/crisp-ff-10/main.tex
experiments v2, added pictures
minor
4 pages. whew.

@@
 \begin{document}
-\title{Using Planning to Generate Natural-Language Sentences:\\
-  Experiences and Challenges}
+\title{Waking up a sleeping rabbit:\\ On Natural-Language Sentence Generation with FF}
+
+%\title{Using Planning to Generate Natural-Language Sentences:\\
+%  Experiences and Challenges}
 
 \author{
-Alexander Koller \\ Saarland University, Germany \\ koller@mmci.uni-saarland.de
+%Alexander Koller \\ Saarland University, Germany \\ koller@mmci.uni-saarland.de
 %\And
-%Ronald Petrick \\ University of Edinburgh, UK \\ rpetrick@inf.ed.ac.uk
-\And
-J\"org Hoffmann \\ INRIA, France \\ joerg.hoffmann@inria.fr
+%J\"org Hoffmann \\ INRIA, France \\ joerg.hoffmann@inria.fr
 }
+
+\nocopyright
 

### 2009-12-19 trunk/Papers/reasoning-challenges/levels.tex
done

@@
 While this format for the IPC has been highly successful in the past
-ten years, there is also a number of ways in which it can be improved and
-streamlined.  One such direction concerns the way in which problem classes
-are collected.  There is no doubt that the existing collection of IPC
-domains is a fundamental driving force for the development of new planning
-algorithms.  Unfortunately, the current selection process makes it
-difficult for users of these systems, who may come from different
-communities, to get a new application-motivated domain accepted as an IPC
-benchmark.  This is unsurprising, given that IPC organizers must be able to
-justify their selection of domains, and picking untested domains from other
-communities is not the best way to do this.  However, this leads to a
-selection process that is less dynamic than it could be, and more
-restrictive for users wanting to contribute suggestions.  Conversely, to
-keep the number of domains that must be tested in each round of the
-competition manageable, older domains are sometimes ``retired'', i.e.\
-removed from subsequent competitions.  This comes with a risk of a false
-sense of progress, in which retired domains are remembered as ``solved''
-although they were removed for completely independent reasons.  Finally, it
-has been argued \cite{ToughNuts} that some IPC organizers tend to favor the
-development of new types of planning problems and new dialects of PDDL over
-trying to find challenging benchmark problems for existing versions of
-PDDL. Such approaches run the risk of ``reinventing'' past problem classes
-(albeit in different guises), or redirecting investigation away from
-current challenges.
+ten years, there is also a number of ways in which it can be improved
+and streamlined.  One such direction concerns the way in which problem
+classes are collected.  There is no doubt that the existing collection
+of IPC domains is a fundamental driving force for the development of
+new planning algorithms.  Unfortunately, the current selection process
+makes it difficult for users of these systems, who may come from
+different communities, to get a new application-motivated domain
+accepted as an IPC benchmark.  This is unsurprising, given that IPC
+organizers must be able to justify their selection of domains, and
+picking untested domains from other communities is not the best way to
+do this.  However, this leads to a selection process that is less
+dynamic than it could be, and more restrictive for users wanting to
+contribute suggestions.  Conversely, to keep the number of domains
+that must be tested in each round of the competition manageable, older
+domains are sometimes ``retired'', i.e.\ removed from subsequent
+competitions.  This comes with a risk of a false sense of progress, in
+which retired domains are remembered as ``solved'' although they were
+removed for completely independent reasons.  Finally, it has been
+argued \cite{ToughNuts} that IPC organizers may be tempted to favor
+the development of new types of planning problems and new dialects of
+PDDL over trying to find challenging benchmark problems for existing
+versions of PDDL. Such approaches run the risk of ``reinventing'' past
+problem classes (albeit in different guises), or redirecting
+investigation away from current challenges.
 
@@
 environment of the planner.  One way in which we could envision this
-is by defining a single image for a virtual machine, such as
-VirtualBox.\footnote{\url{http://www.virtualbox.org/}} When a planner
-is evaluated, this happens in a fresh instance of this virtual
-machine.  We can offer the virtual machine image for download; a
-participant can then run the image on their own machine and ensure
-that their planner works, in the exact environment in which it will be
+is by defining a single hardware and software environment in which
+systems will be evaluated, and making it available for download as a
+virtual machine image.  Developers can then run this image on their
+own machine, using a virtual machine such as
+VirtualBox,\footnote{\url{http://www.virtualbox.org/}} and ensure
+that their system works, in the exact environment in which it will be
 evaluated.  This would even make it technically possible for
-participants to submit not the source code, but a binary executable of
-the planner (perhaps along with support files).  However, even if we
-require the submission of source code, the move to a clean virtual
-machine environment that contains only standard versions of tools and
-libraries enforces at least a certain degree of portability from which
-users will benefit when they download the planners.  Even when the
-virtual machine image changes to reflect updates in the standard
-development environments (new version of the C compiler, etc.), old
-images could remain available, and so old planners would continue
-running until the virtual machine emulator itself becomes unavailable
-or incompatible.
+participants to submit not the source code, but a binary executable
+(perhaps along with support files).  However, even if we require the
+submission of source code, the move to a precisely defined environment
+that contains only standard versions of tools and libraries and on
+which the code must compile without human intervention enforces at
+least a certain degree of portability from which users will benefit
+when they download the systems.  Even when the environment changes to
+reflect updates in the standard development environments (new version
+of the C compiler, etc.), old images could remain available, and so
+old systems would continue running until the virtual machine emulator
+itself becomes unavailable or incompatible.
 
-We propose that every newly submitted planner is run automatically on
+We propose that every newly submitted system is run automatically on
 all available problem instances, and every newly submitted problem
-instance is sent to all available planners.  The scheduling of this
+instance is sent to all available systems.  The scheduling of this
 process can be under the control of an automated system for continuous
-integration, such as Hudson.\footnote{\url{http://hudson-ci.org/}}
-Each run of a planner on a problem instance or problem class can take
-place in a fresh instance of the virtual machine image.  The
-evaluation infrastructure then inserts the problem data and the
-planner into the virtual machine, runs the planner, measures
-performance statistics (including which problems they could solve and
-how much time this took them; see below), and stores the results in a
-database.  Users will be able to view the results in dynamically
-generated reports on the main website; there may also be functionality to
-customize these reports to focus on certain types of problems.  Because all
-planners are run on all problem instances, and users can ensure that
-challenging problem instances are included in the competition, the chance
-of systems being evaluated on only a small and somewhat arbitrary selection
-of certain problems \cite{pg2008002} can be avoided.
+integration, such as Hudson.\footnote{\url{http://hudson-ci.org/}} The
+evaluation infrastructure sends the problem data and the system to the
+test machine, runs the system, measures performance statistics
+(including which problems they could solve and how much time this took
+them; see below), and stores the results in a database. Users will be
+able to view the results in dynamically generated reports on the main
+website; there may also be functionality to customize these reports to
+focus on certain types of problems.  Because all systems are run on
+all problem instances, and users can ensure that challenging problem
+instances are included in the competition, the chance of systems being
+evaluated on only a small and somewhat arbitrary selection of problem
+instances \cite{pg2008002} can be avoided.
 
-In order to minimize security risks, which can arise when user-submitted
-programs are being run automatically, the individual virtual
-machines need to be sandboxed quite tightly.  There is no need for an
-individual virtual machine to be connected to any network, beyond
-being able to enter performance measurements into a database.  By
+In order to minimize security risks, which can arise when
+user-submitted programs are being run automatically, the test machines
+need to be sandboxed quite tightly.  In particular, the machines do
+not need to be able to initiate any network connections, beyond being
+able to enter performance measurements into a database, and can work
+on virtual file systems that are deleted after each system.  By
 further imposing resource limits on the number of available
@@
 potential for abuse is limited.  It can be limited further by only
-allowing planner uploads to registered users.
+allowing systems uploads to registered users.
 
@@
 discrete timepoints, e.g., to encourage the development of faster
-planners through awards.  Such a competition could easily be based on
+systems through awards.  Such a competition could easily be based on
 our system, by announcing a deadline and a specified subset of problem
 classes. The selection of this subset could be informed, among other
-things, by the report on problems that many planners can't
-solve. Participants could then upload their planners at their leisure;
+things, by the report on problems that many systems can't
+solve. Participants could then upload their systems at their leisure;
 ``running'' the competition amounts to taking a snapshot of the
@@
 on the website simply as one special report for a certain point in
-time.  In such a case, the authors of planners should have the option
+time.  In such a case, the authors of systems should have the option
 of marking them as ``non-competing'' to exclude them from the
@@
 document these problems and link to existing literature; it would
-allow the developers of planners to document these planners; and it
+allow the developers of systems to document these systems; and it
 would make it easy for the organizers of regular competitions to add
@@
 indication for the problems that are considered important in the
-community; imagine a ``top ten'' list of domains and planners on the
-website's front page.  Next to the difficulty of problems, this can be
-another factor in the selection of problem classes for a competition
-snapshot.  Second, a relevance score for problem classes can also
-serve as a guideline for the scheduling of planner runs: Assuming that
-there is not enough computing power to run all incoming planners on
-all domains in real time (e.g., just before a competition deadline),
-the runs on popular domains can be prioritized.
+community; imagine a ``top ten'' list of problem classes and systems
+on the website's front page.  Next to the difficulty of problems, this
+can be another factor in the selection of problem classes for a
+competition snapshot.  Second, a relevance score for problem classes
+can also serve as a guideline for the scheduling of system runs:
+Assuming that there is not enough computing power to run all incoming
+planners on all problem instances in real time (e.g., just before a
+competition deadline), the runs on popular domains can be prioritized.
 
@@
 percentage of problems solved per system, average solution quality,
-average runtime and memory use, and search tree sizes can be
-collected -- with the caveat that the virtual machine must be
-configured so as to allow comparable runtime measurements.
-Furthermore, popularity scores from the social networking aspect can
-be incorporated, or different measures could be combined into an
-overall score.
+average runtime and memory use, and search tree sizes can be collected
+-- with the caveat that the test machines must be configured so as to
+allow comparable runtime measurements.  Furthermore, popularity scores
+from the social networking aspect can be incorporated, or different
+measures could be combined into an overall score.
 
@@
 We believe that the actual hardware cost will be bearable.  While the
-idea of running all planners on all domains sounds intimidating at
-first, we don't anticipate that there will ever be more than a couple
-dozen current versions of planners entered in the system.  This means
-that any new domain that is entered will not be attempted to be solved
-too many times.  Conversely, while there may (hopefully!) be a
-significant number of domains in the system, and every new planner
-must try to solve all domains whose types it claims to be able to
-process, this cost can also be controlled by prioritizing popular or
-hard domains.  A single computer should be sufficient for these needs;
-if it turns out not to be, the virtual machines can easily be run on
-machines anywhere.
+idea of running all systems on all problem instances sounds
+intimidating at first, we don't anticipate that there will ever be
+more than a couple dozen current versions of systems entered in the
+evaluation process.  This means that any new domain that is entered
+will not be attempted to be solved too many times.  Conversely, while
+there may (hopefully!) be a significant number of domains in the
+system, and every new system must try to solve all problem instances
+whose types it claims to be able to process, this cost can also be
+controlled by prioritizing popular or hard domains.  A single computer
+should be sufficient for these needs for the most part; in times of
+high demand (e.g., just before a competition deadline), the work can
+be automatically distributed over ``slave'' test machines by an
+off-the-shelf continuous integration engine.
 
@@
 %RP: I don't see why this needs to be restricted to "reasoning" tasks
-Many reasoning tasks, including constraint solving and theorem proving, are
-being evaluated by running solvers over problem instances and reporting
-performance measures.  At the level of abstraction at which our continuous
-evaluation system could be implemented, there is no fundamental difference
-between the problem type ``deterministic planning problem in PDDL 1.0'' and
-the problem type ``\todo{something from CP or similar}'' -- in both cases,
-the system can collect problem classes and instances of this type and run
-them through solvers that declare themselves to accept such inputs.  This
-means that there is a potential for sharing development and maintenance
-costs for the evaluation platform among different communities.
+Many reasoning tasks, including constraint solving and theorem
+proving, are being evaluated by running solvers over problem instances
+and reporting performance measures.  At the level of abstraction at
+which our continuous evaluation system could be implemented, there is
+no fundamental difference between the problem type ``deterministic
+planning problem in PDDL 1.0'' and the problem type ``theorem of
+first-order logic in TPTP format'' -- in both cases, the system can
+collect problem classes and instances of this type and run them
+through solvers that declare themselves to accept such inputs.  In
+fact, other fields are already ahead of the planning community in
+maintaining problem repositories as on Level 1 above
+\cite{SS98,csplib,Asuncion+Newman:2007}. This means that there is a
+potential for sharing development and maintenance costs for the
+evaluation platform among different communities.
 
@@
 environments to build their systems, and there are typically many more
-systems in any competition than there are problem classes (corpora).
+systems in any competition than there are problem classes (text corpora).
 However, our approach is completely consistent with the requirements

### 2009-12-19 trunk/Papers/crisp-ff-10/crisp.tex
submitted version

@@
 elementary tree, and equip every elementary tree with a set of atoms
-over these variables to encode the meaning this tree expresses.
+over these variables to encode the meaning this tree expresses
+(printed in italics in the figure).
 Fig.~\ref{fig:white-rabbit-sleeps-deriv} already shows specific
@@
 $r_2$ is a \emph{distractor} for the subject of the sentence, which is
-removed by adjoining the tree for ``white''.  The sentence generation
-problem consists in computing a grammatically correct TAG derivation
-from instances of the elementary trees in the grammar that conveys the
-intended meaning and describes all referents uniquely.
+removed by adjoining the tree for ``white''.  In short, the sentence
+generation problem consists in computing a grammatically correct TAG
+derivation from instances of the elementary trees in the grammar that
+conveys the intended meaning and describes all referents uniquely.
 
@@
 occurs at the root of the tree.  The planning state encodes a list of
-possible node names; $\f{current}(u)$ expresses that $u$ is the next
+possible node names: $\f{current}(u)$ expresses that $u$ is the next
 node name that should be picked when a new substitution node is
@@
 The example generation problem above translates into a planning
-problem whose domain is shown, in PDDL format, in
+problem whose domain is shown in
 Fig.~\ref{fig:white-rabbit-as-planning}. The initial state of the
@@
 be automatically decoded into the derivation shown in
-Fig.~\ref{fig:white-rabbit-sleeps-deriv}.  The analogous plan with
-$r_2$ instead of $r_1$ is not correct because the $\mathsf{sleep}$
-precondition of the $\mathsf{sleeps}$ operator is not satisfied.  The
-first two steps of the plan alone would not be a correct plan because
-they leave an atom $\mathsf{distractor}(n_1, r_2)$ in the state, which
-contradicts the goal that all distractors have been eliminated.
+Fig.~\ref{fig:white-rabbit-sleeps-deriv}.  The first two steps of the
+plan alone would not be a correct plan because they leave an atom
+$\mathsf{distractor}(n_1, r_2)$ in the state, which contradicts the
+goal that all distractors have been eliminated.
 

### 2010-02-26 trunk/Papers/crisp-ff-10/conclusion.tex
intro umgeschrieben, platz gemacht.

@@
 problem, both in its application value and in the challenges it
-poses. After the changes described in this paper, FF is a practical
-option for this application. It is striking how simple these changes
-were, and certainly this suggests to look more closely at automatic
-planner configuration techniques.
+poses. After the changes to FF described in this paper, for the first
+time an automated planner is a practical option for this
+application. The simplicity of the changes required is striking, and
+strongly suggests to look more closely at automatic planner
+configuration techniques.
 
-The domain is also far from ``solved'' in all its aspects. As we
-pointed out, several features of the domain are very hard to capture
-with existing search heuristics. In Fig.~\ref{fig:runtimes}, this
-shows in the fact that no planner version scales beyond $n>4$ -- even
-in case (a) where no adjectives at all are required. It remains an
-interesting challenge to address this.
+%It is striking how simple these changes were, and certainly this
+%suggests to look more closely at automatic planner configuration
+%techniques.
+
+The domain is far from ``solved'' in all its aspects. As we pointed
+out, several features of the domain are very hard to capture with
+existing search heuristics. In Fig.~\ref{fig:runtimes}, this shows in
+the fact that no planner version scales beyond $n>4$ -- even in case
+(a) where no adjectives at all are required. It remains an interesting
+challenge to address this.
 

### 2010-02-26 trunk/Papers/crisp-ff-10/crisp-ff.tex
intro umgeschrieben, platz gemacht.

@@
 
-FF exhibits a number of problems in this domain that were previously
-unknown.  The basic observation of FF's inefficiency on these inputs
-was made before
-\cite{koller10:_exper_with_plann_for_natur_languag_gener}; here we
-examine its causes in detail. For some of the observed pitfalls, we
+%FF exhibits a number of problems in this domain that were previously
+%unknown.
+
+The basic observation of FF's inefficiency in the NLG domain was made
+before \cite{koller10:_exper_with_plann_for_natur_languag_gener}; here
+we examine its causes in detail. For some of the observed pitfalls, we
 are able to provide simple fixes. Others pose serious challenges to
@@
 
-FF's preprocessor contains two details that, before this paper, were
-so detrimental that we were not able to seriously propose FF as a
-practical solution.
+%FF's preprocessor contains two details that, before this paper, were
+%so detrimental that we were not able to seriously propose FF as a
+%practical solution.
 %Simple fixes
@@
 predicates. Since this does not take into account reachability, it is
-prohibitively huge in our domain. Now, we actually have no hard
-operators. But FF builds the data structure anyway, which exhausts
+prohibitively huge in our domain. Now, {\em we actually have no hard
+  operators}. But FF builds the data structure anyway, which exhausts
 memory and causes a segmentation fault on serious instances.
@@
 
-FF's preprocessing pitfalls -- and the fact that they didn't occur to
-anyone before -- are certainly baffling, although easily fixed. We now
-turn our attention to some of FF's search techniques, variations of
-which are used in many planners and which are very much not easily
-fixed.
+FF's preprocessing pitfalls are certainly baffling, but easily
+fixed. We now turn our attention to some of FF's search techniques,
+variations of which are used in many planners, and which are very much
+not easily fixed. Our solution for the moment are simple configuration
+changes. %; these give much better performance but do .
 
@@
 LAMA \cite{richter:etal:aaai-08}. FF as used in the 2000 competition
-approximates reasonable orders as a pre-process and then partitions
-the goal set into a {\em goal agenda}, a sequence of goal
-subsets. Each subset is posed to the planner in isolation. The final
-plan is the concatenation of the plans for all entries.
+approximates reasonable orders as a preprocess and then partitions the
+goal set into a {\em goal agenda}, a sequence of goal subsets. Each
+subset is posed to the planner in isolation. The final plan is the
+concatenation of the plans for all entries.
 
@@
 In our improved FF, the goal agenda is simply switched off. Having
-disposed of goal orderings in that way, let us turn our attention to
-the relaxed plan heuristic. %
+thus disposed of goal orderings, let us turn our attention to the
+relaxed plan heuristic. %
 %While the goal agenda meltdown may be more amusing than alarming, we
@@
 %meltdown either. They are fatally 
-This meets its Waterloo as well, due to the unusual lifecycle of facts
-encoding substitution nodes and distractors. 
+This meets its Waterloo in the unusual life-cycle of facts encoding
+substitution nodes and distractors.
 %In instances where these
@@
 dead-ends, i.e., dead-end states that have a relaxed plan. An example
-is the initial state of a generation problem that it unsolvable
-because not all distractors can be removed; its $h$ value will still
-be $1$. So our domain is in the most difficult class of Hoffmann's
-\shortcite{hoffmann:jair-05} ``planning domain taxonomy''. While it
-shares this property with several other challenging domains, the
-extremity of turning EHC into a breadth-first search is unheard of.
+is the initial state of a task that it unsolvable because not all
+distractors can be removed; its $h$ value will still be $1$. So the
+domain is in the most difficult class of Hoffmann's
+\shortcite{hoffmann:jair-05} planning domain taxonomy. While it shares
+this property with several other challenging domains, turning EHC into
+a breadth-first search is unheard of.
 
@@
 
-Can we design heuristics that are better at dealing with the lifecycle
+Can we design heuristics that are better at dealing with the life-cycle
 of substitution nodes and distractors? It is easy to see that the
@@
 \cite{helmert:icaps-04,richter:etal:aaai-08,karpas:domshlak-ijcai-09,helmert:domshlak:icaps-09,cai:etal:icaps-09}. The
-two notable eceptions are the $h^m$ family, and abstraction
+two notable exceptions are the $h^m$ family, and abstraction
 heuristics. We ran a few tests with a variant of the example problem
@@
 The only FF technique that appears to be clearly useful here is its
-action pruning technique, the ``helpful actions''. While the relaxed
-plan is often much too short, it appears to always contain at least
-one of the actions that actually are useful. 
+action pruning technique. While the relaxed plan is often much too
+short, it appears to always contain at least one of the actions that
+actually are useful.
 %

### 2010-02-26 trunk/Papers/crisp-ff-10/introduction.tex
intro umgeschrieben, platz gemacht.

@@
 recently showed that as the planning problem instances from Koller and
-Stone \shortcite{KolSto07} scale up, off-the-shelf planners become too
-slow to be practically useful. They noted that there are major sources
-of inefficiency both in the search itself and in the commonly used
-preprocesses, due to several features of the domain
-(e.g.\ comparatively large predicate and operator arities) that
+Stone \shortcite{KolSto07} scale up, off-the-shelf planners become
+slow -- indeed, {\em too slow to be practically useful in NLG
+  applications.} They showed this for SGplan and several variants of
+FF \cite{HoffmannNebel01}; we show it herein also for LAMA
+\cite{richter:etal:aaai-08}. Koller and Petrick noted that there are
+major sources of inefficiency both in the search itself and in the
+commonly used preprocesses; they identified several features of the
+domain (e.g.\ comparatively large predicate and operator arities) that
 diverge from most IPC benchmarks.
@@
 
-In this paper, we revisit this situation, analyze the sources of
-inefficiency in more detail in the context of the FF planner
-\cite{HoffmannNebel01}, and show how to modify FF to obtain
-dramatically better performance. We achieve this by tweaking the
-implementation of FF's preprocessor, and by acting on some simple but
-shocking observations regarding the behavior of some of FF's search
-techniques, in particular its goal ordering technique and the relaxed
-plan heuristic. The changes we make are all simple, but make FF
-perform well enough to be useful in practical NLG applications.
+In this paper, we revisit this situation, and analyze the sources of
+inefficiency in detail for FF. Based on the analysis, we modify FF and
+obtain {\em dramatically better performance that, for the first time,
+  makes automated planners a realistic possibility for NLG
+  applications.} To achieve this quantum leap, all that is needed are
+a few minor implementation fixes and basic reconfiguration of search
+options. While these fixes certainly are no contribution to planning
+in their own right, we believe it a valuable -- and slightly
+unsettling -- lesson learned that such banalities can make the
+difference between success and failure. They can hold up progress in
+planning applications for years (until a planning expert comes by and
+does some code profiling). We interpret this as a strong call for
+automated configuration techniques, e.g.\ along the lines of Hutter et
+al.\ \shortcite{HutHooStu07}.
+
+Apart from these insights, our domain remains an interesting challenge
+for the development of better heuristics. There still are realistic
+instances of the NLG planning domain on which our modified FF is not
+practical. Overcoming this weakness would entail overcoming
+fundamental weaknesses of FF's goal ordering techniques and the
+relaxed plan heuristic -- variants of which are used in many planners
+today.
+
+% show how to modify FF to obtain dramatically better performance. We
+% achieve this by tweaking the implementation of FF's preprocessor, and
+% by acting on some simple but shocking observations regarding the
+% behavior of some of FF's search techniques, in particular its goal
+% ordering technique and the relaxed plan heuristic. The changes we
+% make are all simple, but make FF perform well enough to be useful in
+% practical NLG applications.
 
@@
 
-Our performance gains with simple FF configuration changes certainly
-point out that automatic planner configuration may be quite
-useful. Apart from that, our domain remains an interesting challenge
-for the development of better heuristics. There still are realistic
-instances of the NLG planning domain on which our modified FF is not
-practical. Overcoming this weakness would entail overcoming the
-observed weaknesses of FF's search techniques -- variants of which are
-used in many planners today.
+%Our performance gains with simple FF configuration changes certainly
+%point out that automatic planner configuration may be quite
+%useful. Apart from that, our domain remains an interesting challenge
+%for the development of better heuristics. There still are realistic
+%instances of the NLG planning domain on which our modified FF is not
+%practical. Overcoming this weakness would entail overcoming the
+%observed weaknesses of FF's search techniques -- variants of which are
+%used in many planners today.
 

### 2010-02-27 trunk/Papers/crisp-ff-10/experiments.tex
bin zufrieden so

@@
 Consider first the results for $d=0$, shown in Fig.~\ref{fig:runtimes}
-(a); the lines end where the planner exceeded its five-minute timeout.
-``Old FF'' in the figure is original FF, ``New preprocessor'' applies
+(a); the lines end where the planner exceeded its five-minute
+timeout. Notice that the vertical axis is logarithmic. Clearly, ``Old
+FF'' (the original FF) is much too slow for practical use. The same is
+true -- although to a less dramatic extent -- for LAMA, which already
+takes three seconds to generate the relatively short sentence at $n=3$
+and times out for larger instances. ``New preprocessor'' is FF with
 only our fixes to FF's preprocess, ``New FF'' applies all fixes.
@@
 %search with helpful actions (``New FF'') respectively.  
-Just the preprocessor fixes lead to a speed-up of at least $3$ orders
-of magnitude, leading to practically useful runtime -- for instance,
-the 24-word sentence at $n=4$ is generated in under two seconds. Note
-that, by contrast, the search configuration does not make much of a
-difference here. This changes in our next experiment.
+Even the preprocessor fixes alone lead to a speed-up of at least $3$
+orders of magnitude, leading to practically useful runtime -- for
+instance, the 24-word sentence at $n=4$ is generated in under two
+seconds. Note that, by contrast, the search configuration does not
+make much of a difference here. This changes in our next experiment.
 
@@
 old FF preprocessor crashed with a segmentation fault even at
-$n=1$. Best-first search with helpful actions is clearly the
-best-performing search strategy here.  It takes about 14 seconds for
-$n=3$, but this sentence is already quite long at 29 words, and EHC
-takes 3.5 minutes on the same input.  Best-first search without
-helpful actions timed out at $n=3$, illustrating the importance of
-that heuristic.
+$n=1$. LAMA performs better, but times out beyond $n>2$.  Best-first
+search with helpful actions is clearly the best-performing search
+strategy here.  It takes about 14 seconds for $n=3$, but this sentence
+is already quite long at 29 words, and EHC takes 3.5 minutes on the
+same input.  Best-first search without helpful actions timed out at
+$n=3$, illustrating the importance of that heuristic.
 

### 2010-02-27 trunk/Papers/crisp-ff-10/introduction.tex
bin zufrieden so

@@
 Stone \shortcite{KolSto07} scale up, off-the-shelf planners become
-slow -- indeed, {\em too slow to be practically useful in NLG
-  applications.} They showed this for SGplan and several variants of
-FF \cite{HoffmannNebel01}; we show it herein also for LAMA
+slow -- indeed, too slow to be practically useful in NLG
+applications. They showed this for SGPlan and several variants of FF
+\cite{HoffmannNebel01}; we show it herein also for LAMA
 \cite{richter:etal:aaai-08}. Koller and Petrick noted that there are
@@
 inefficiency in detail for FF. Based on the analysis, we modify FF and
-obtain {\em dramatically better performance that, for the first time,
-  makes automated planners a realistic possibility for NLG
-  applications.} To achieve this quantum leap, all that is needed are
-a few minor implementation fixes and basic reconfiguration of search
+obtain dramatically better performance that, for the first time, makes
+automated planners a realistic possibility for NLG applications. To
+achieve this quantum leap, all that is needed are a few minor
+implementation fixes and basic reconfiguration of search
 options. While these fixes certainly are no contribution to planning
-in their own right, we believe it a valuable -- and slightly
+in their own right, we consider it a valuable -- and slightly
 unsettling -- lesson learned that such banalities can make the
 difference between success and failure. They can hold up progress in
-planning applications for years (until a planning expert comes by and
-does some code profiling). We interpret this as a strong call for
-automated configuration techniques, e.g.\ along the lines of Hutter et
-al.\ \shortcite{HutHooStu07}.
+planning applications for years, unless users of a planner have direct
+access to its developer. We interpret this as illustrative of the
+importance of implementation details such as the preprocessor for the
+overall runtime, and as a strong call for automated configuration
+techniques, e.g.\ along the lines of Hutter et al.\
+\shortcite{HutHooStu07}.
 

### 2010-03-01 trunk/Papers/crisp-ff-10/experiments.tex
Joerg final

@@
 same input.  Best-first search without helpful actions timed out at
-$n=3$, illustrating the importance of that heuristic.
+$n=3$, illustrating the importance of that pruning technique.
 
-The relative behavior of the different search methods is not uniform
-in our domain, which exhibits a huge amount of structural variance due
-to the many different forms that the input grammar may take. We are
-far from an exhaustive exploration of all the possible parameter
-settings. For instance, with particular inputs (no distractors), EHC
-can be more effective than best-first search.
+We remark that the relative behavior of the different search methods
+is not uniform in our domain, which exhibits a huge amount of
+structural variance due to the many different forms that the input
+grammar may take. We are far from an exhaustive exploration of all the
+possible parameter settings. For instance, with particular inputs (no
+distractors), EHC can be more effective than best-first search.
 

### 2010-03-01 trunk/Papers/crisp-ff-10/main.tex
intro umgeschrieben, platz gemacht.
final PDF-Ok-according-to-AAAI-standards

--- 2009-12-18 trunk/Papers/crisp-ff-10/main.tex
@@
 
-\pdfinfo{
-/Title (Using Planning to Generate Natural-Language Sentences)
-/Subject (Proceedings of ICAPS, hopefully)
-/Author (Anonymous authors)}
+\special{! /pdfmark where
+{pop} {userdict /pdfmark /cleartomark load put} ifelse
+[ /Author (Alexander Koller, Joerg Hoffmann)
+/Title (Waking up a sleeping rabbit: On Natural-Language Sentence Generation with FF)
+/Keywords (AAAI, artificial intelligence)
+/DOCINFO pdfmark}
+
+%\pdfinfo{
+%/Title (Waking up a sleeping rabbit:\\ On Natural-Language Sentence Generation with FF)
+%/Subject (Proceedings of ICAPS, hopefully)
+%/Author (Anonymous authors)}
 
@@
 \begin{document}
-\title{Using Planning to Generate Natural-Language Sentences:\\
-  Experiences and Challenges}
+\title{Waking up a sleeping rabbit:\\ On Natural-Language Sentence Generation with FF}
 
@@
 Alexander Koller \\ Saarland University, Germany \\ koller@mmci.uni-saarland.de
-%\And
-%Ronald Petrick \\ University of Edinburgh, UK \\ rpetrick@inf.ed.ac.uk
 \And
@@
 }
+
+\nocopyright
 
@@
 
+{\small
 \bibliography{bibliography}
 \bibliographystyle{aaai}
+}
 

### 2010-03-01 trunk/Papers/crisp-ff-10/main.tex
minor..

@@
 [ /Author (Alexander Koller, Joerg Hoffmann)
-/Title (Waking up a sleeping rabbit: On Natural-Language Sentence Generation with FF)
+/Title (Waking Up a Sleeping Rabbit: On Natural-Language Sentence Generation with FF)
 /Keywords (AAAI, artificial intelligence)
@@
 \begin{document}
-\title{Waking up a sleeping rabbit:\\ On Natural-Language Sentence Generation with FF}
+\title{Waking Up a Sleeping Rabbit:\\ On Natural-Language Sentence Generation with FF}
 

### 2010-03-01 trunk/Papers/crisp-ff-10/abstract.tex
new abstract, copyright notice, formatting back into 4 pages

@@
 natural language sentences. This domain has a number of features that
-provoke fairly unusual behavior in existing planners. In particular,
-we show how several techniques used in the FF planner, which has been
-highly successful on IPC problems, are led astray. We discuss possible
-fixes to FF's shortcomings, some of which we have already implemented,
-but some of which pose interesting challenges to the AI Planning
-community. The performance of our modified FF is up to several orders
-of magnitude better than that of the original FF. While this makes FF
-much more practical for this application, we show that it is still
-quite limited with respect to the complexity of the underlying
-grammar.
+provoke fairly unusual behavior in planners. In particular, hitherto
+no existing automated planner was sufficiently effective to be of
+practical value in this application. We analyze in detail the reasons
+for ineffectiveness in FF, resulting in a few minor implementation
+fixes in FF's preprocessor, and in a basic reconfiguration of its
+search options. The performance of the modified FF is up to several
+orders of magnitude better than that of the original FF, and for the
+first time makes automated planners a practical possibility for this
+application. Beside thus highlighting the importance of preprocessing
+and automated configuration techniques, we show that the domain still
+poses several interesting challenges to the development of search
+heuristics.
+%In particular, we show how several techniques used in the FF planner,
+%which has been highly successful on IPC problems, are led astray. We
+%discuss possible fixes to FF's shortcomings, some of which we have
+%already implemented, but some of which pose interesting challenges to
+%the AI Planning community. The performance of our modified FF is up to
+%several orders of magnitude better than that of the original FF, and
+%for the first time makes automated planners a pratical possibility for
+%this application.
+%
+%. While this makes FF
+%much more practical for this application, we show that it is still
+%quite limited with respect to the complexity of the underlying
+%grammar.
 %We outline how other planning domains might share the

### 2010-03-01 trunk/Papers/crisp-ff-10/experiments.tex
new abstract, copyright notice, formatting back into 4 pages

@@
 (a); the lines end where the planner exceeded its five-minute
-timeout. Notice that the vertical axis is logarithmic. Clearly, ``Old
+timeout. Note that the vertical axis is logarithmic. Clearly, ``Old
 FF'' (the original FF) is much too slow for practical use. The same is
 true -- although to a less dramatic extent -- for LAMA, which already
-takes three seconds to generate the relatively short sentence at $n=3$
-and times out for larger instances. ``New preprocessor'' is FF with
-only our fixes to FF's preprocess, ``New FF'' applies all fixes.
+takes $3$ seconds to generate the relatively short sentence at $n=3$ and
+times out for larger instances. ``New preprocessor'' is FF with only
+our fixes to FF's preprocess, ``New FF'' applies all fixes.
 % The other two lines represent FF with the improved preprocessor,
@@
 %search with helpful actions (``New FF'') respectively.  
-Even the preprocessor fixes alone lead to a speed-up of at least $3$
-orders of magnitude, leading to practically useful runtime -- for
-instance, the 24-word sentence at $n=4$ is generated in under two
-seconds. Note that, by contrast, the search configuration does not
-make much of a difference here. This changes in our next experiment.
+The preprocessor fixes alone lead to a speed-up of at least $3$ orders
+of magnitude, leading to practically useful runtime -- e.g., the
+24-word sentence at $n=4$ is generated in under two seconds. By
+contrast, the search configuration does not make much of a
+difference. This changes in our next experiment.
 

### 2010-05-19 trunk/Papers/pcrisp-10/conclusion.tex
Added draft for TAG+ paper about pcrisp

\section{Conclusion}
\label{sec:conclusion}

We describe an approach to sentence generation based on probabilistic tree adjoining
grammar and AI planning. Following previous work by \newcite{stonedoran1997,kollerstone2007}, we treat sentence 
generation as an integrated task comprising sentence planning and surface realization using declarative descriptions of the syntactic
and semantic contribution of each word and contextual constraints on its use. Our system uses AI planning technology to find the best {\sc tag} 
derivation that fulfills a given communicative goal, while satisfying these constraints. 
Our experiments show that our generation system based on {\sc pltag} and planning makes it possible to use an integrated approach to sentence generation with treebank induced grammars for the first time. The results also highlight the trade-off between output quality, efficiency and robustness. Future work work should optimize the smoothing model to boost output quality. Improved and/or tailored heuristics could help to improve runtime behavior. Finally it would be interesting to evaluate the system in an extrinsic way that also addresses sentence planning.

### 2010-05-19 trunk/Papers/pcrisp-10/crisp.tex
Added draft for TAG+ paper about pcrisp

\section{Sentence Generation as Planning}
\label{sec:crisp}
In this section we review the original non-statistical {\sc crisp} system \cite{kollerstone2007}. 
Following {\sc spud} \cite{stonedoran1997}, {\sc crisp} is based on a declarative description of the sentence generation problem using {\sc tag}. Given a knowledge base, a communicative goal and a grammar, we require to find a grammatical {\sc tag} derivation that is consistent with this knowledge base and satisfies a communicative goal. A number of semantic and pragmatic constraints that must be satisfied by the solution can be added, for instance to enforce generation of unambiguous refering expressions. \newcite{kollerstone2007} describe how to encode this problem into an AI planning problem, that can be solved efficiently by off-the-shelf planners. We describe the general mechanism in the following section and then review the encoding into planning in section \ref{ssec:crispdomain}. 
\begin{figure*}[th]
\begin{center}
\includegraphics[width=.8\textwidth]{figures/grammar.pdf}
\caption{\label{fig:grammar}(a) An example grammar with semantic content. (b) A derivation for ``the cat eats raw tuna''. }
\end{center}
\end{figure*}

\subsection{Sentence Generation in CRISP}
Like {\sc spud}, {\sc crisp} uses an {\sc ltag} in which elementary trees are assigned semantic content. Each node in a {\sc crisp} elementary tree node is associated with a semantic role. Semantic content is expressed as a set of literals, encoding relations between these roles. All nodes that dominate the lexical anchor are assigned the role `self', which intuitively corresponds to the event or individual described by this tree. Figure \ref{fig:grammar}(a) shows an example grammar of this type.

In a derivation we may only include elementary trees whose semantic content has an instantiation in the knowledge base. For each substitution and adjunction, the semantic role associated with the role of the target node is unified with the `self' role of the child tree. For example, given the knowledge base 
{\it \{cat($i_2$), tuna($i_3$), raw($i_3$), eats($i_1$,$i_2$,$i_3$)\}} and the grammar in \ref{fig:grammar}(a), {\sc crisp} could produce the derivation in \ref{fig:grammar}(b).

\subsection{CRISP Planning Domains} 
\label{ssec:crispdomain}
Before we describe {\sc crisp}'s encoding of sentence generation as planning, we briefly review AI planning in general. 
The crucial concept in AI planning is a state, a conjunction of first order literals describing relations between some individuals. A planning problem consist of an initial state, a set of goal states and a set of planning operators that describe possible state transitions.  A solution to the problem is any sequence of actions (instantiated operators) that leads from the initial state to one of the goal states. Planning problems can be written in a standard notation ({\sc pddl}\footnote{Planning Domain Description Language}, \newcite{gerevinietal2009}) and can be solved efficiently by planning systems such as {\sc ff}\footnote{Fast Forward} \cite{hoffmannnebel2001}.

In {\sc crisp} planning states correspond to partial {\sc tag} derivations and record open substitution and adjunction sites, semantic individuals associated with them, and parts of the communicative goal that have not yet been expressed.  The initial state also encodes the knowledge base and the communicative goal. Each planning operator contributes a new elementary tree to the derivation and at the same time satisfies part of the communicative goal, as described in the previous section. In a goal state there are no open substitution sites left and all literals in the communicative goal have been expressed. 
\begin{figure}[t]
\begin{center}
\planaction{\bf S-eats-1(u,~x1,~x2,~x3)}{step(step1),referent(u,~x1),\\ subst(S,~u), eats(x1,~x2,~x3)}{
$\lnot$needtoexpr(pred-eats,~x1,~x2,~x3),\\ $\lnot$subst(S,~u),\\
subst(NP,~subj-1), subst(NP,~obj-1),\\ refer(subj-1,~x2), refer(obj-1,~x3)\\
canadj(VP, u), canadj(V, u), canadj(S, u)\\
$\lnot$ step(step1), step(step2)}\\\smallskip

\planaction{\bf NP-cat-2(u,~x1)}{step(step1),referent(u,~x1),\\ subst(NP,~u), cat(x1)}{
$\lnot$needtoexpress(pred-cat,~x1),\\ $\lnot$subst(NP,~u),\\
canadj(N, u), canadj(NP, u)\\
$\lnot$ step(step1), step(step2)}\\\smallskip

\planaction{\bf N-raw-4(u,~x1)}{step(step1),referent(u,~x1),\\ canadj(N,~u), raw(x1)}{
$\lnot$needtoexpress(pred-raw,~x1),\\ $\lnot$canadj(N,~u),\\
$\lnot$ step(step4), step(step5)}\\\smallskip
\end{center}
\caption{\label{fig:crisp-operators} {\sc crisp} operators for some of the elementary trees in figure \ref{fig:grammar}.}
\end{figure}

Figure \ref{fig:crisp-operators} shows planning operators for part of the grammar in figure \ref{fig:grammar}(a).
The preconditions of the operators require that a suitable open substitution node (i.e. of the correct category) or internal node for adjunction exists in the partial derivation. In the operator effect, open substitution nodes are closed and new identifiers are created for each substitution node and internal node in the new tree. Given the knowlede base from above, a plan corresponding to the derivation in figure \ref{fig:grammar} would be $\langle${\it S-eats-1(root, i1, i2, i3), NP-cat-2(subj-1, i2), NP-tuna-3(obj-1, i3), N-raw-4(obj-1, i3)}$\rangle$

\subsection{CRISP and Large Grammars}
Trees in large, possibly treebank induced grammars are often applicable in a number of different syntactic contexts, leading to overgeneration. Such trees contain many internal nodes that are open for adjunction but will not necessarily be filled in a derivation for a given input communicative goal. Attaching to the wrong node can produce invalid or disprefered word order. 
In addition, in such grammars the ambiguity of possible, structurally different trees (supertags) for a given word is high. These problems are illustrated in figure \ref{fig:overgen}. Assuming a grammar that includes trees for both right attaching ($t13$) and left attaching PPs ($t252$), both derivations (a) and (b) are {\it grammatical} derivations that satisfy the same communicative goal. However, most readers disprefer the reading in (b). 
Clearly, to use {\sc crisp} with such a grammar we need a method of distinguishing good derivations from bad ones. 
\begin{figure*}
\begin{center}
\includegraphics[width=.6\textwidth]{figures/overgen.pdf}
\caption{\label{fig:overgen} Two derivations with a large grammar, that satisfy the same communicative goal. Sentence (b) is disprefered by most readers.}
\end{center}
\end{figure*}

### 2010-05-19 trunk/Papers/pcrisp-10/experiments.tex
Added draft for TAG+ paper about pcrisp

\section{Experiments}
\label{sec:experiments}
In this section we perform experiments with the statistical generation system described in the previous section.  We are interested in the output quality and runtime behavior of different combinations of planning systems and probability models. We can assume that the sentence planning abilities are inherited from {\sc crisp} and only evaluate the realization abilities of our system by comparing generated sentences against a gold standard testset. The next section describes our evaluation data and section \ref{ssec:eval-measures} describes the evaluation measures used.

\subsection{Evaluation Data}

For our experiments we use a grammar and treebank, that was automatically extracted from the Wall Street Journal portion of the Penn Treebank (41576 sentences)
using the algorithm described by \newcite{chenschanker2004}.  We preprocess this grammar by linearizing multiple adjunctions and by re-attaching prepositions to trees they appear with.

%{\bf Adding semantics to elementary trees:}
We then automatically create semantic content for all lexicalized elementary trees by assigning a single semantic predicate to each tree in the treebank, using the lexical anchor as a name and setting the arity to the number of substitution nodes in the tree plus one for the `self' role.  We calculate the role associated with each node in each tree by assigning role-names to each substitution node (`self' is assigned to the lexical anchor) and then percolating the roles up the tree, giving preference to the self role.

We estimate our probability models on a training portion of the dataset (96\%) using maximum likelihood estimation.
For the testing set we selected from the remaining 4\% all sentences that were shorter than 16 words (416 sentences). For longer sentences the search time often exceeded 5 minutes.

%{\bf Creating realization problems:} 
For this testing set we compute semantic representation for each sentence, by traversing its derivation tree top-down and collects a conjunction of terms, by instantiating the automatically created semantic representations. Our algorithm introduces a new individual for each substitution node marked with role $r$ and equates it to the `self' role of any tree that adjoins or substitutes into a node marked with $r$.

\subsection{Evaluation Measures}
\label{ssec:eval-measures}
To evaluate the ouput quality of our statistical generation system we compare the system output against the reference sentences in the treebank, from which the input representations were generated. We adopt the {\bf generation tree accuracy} measure discussed by \cite{bangalorerambowwhittaker2001}.
This measure is computed by first creating a list of all 'treelets' from the reference derivation tree. A 'treelet' is a subtree, consisting only of a node and its direct children. For each treelet we calculate edit distance, sum the distances over all treelets and then divide by the total length of the reference string.
$$1 - \frac{\sum\limits_{d \in treelets(D)} editDist(O|_d, R|_d)}{\textit{length}(R)},$$
where $D$ is the reference derivation tree and $S|_d$ are the tokens associated with the nodes of treelet $d$ in the order they appear in $S$ (if at all). 
Edit distance is modified to treat insertion-deletion pairs as single movement errors. 
This metric account for the observation that errors that change word order on a local level (i.e. within a treelet) are less severe than errors that swap words between different treelets. Intuitively errors of the latter type violate the continuity of sub-phrases. Because of this property we prefer generation tree accuracy over {\sc blue}, which is overly sensitive to movement errors on an individual sentence level.

%\newcite{belzreiter2006} empirically investigate the usefulness of {\sc bleu} and {\sc nist} for evaluating NLG systems. They find that both metrics correlate with human judgements of output quality on a corpus level, but require a large number of different references sentences to be reliable. Multiple references are necessary because {\sc blue} favors sentences with the same word order as the reference. In NLG, more than in Machine Translation, word order of output sentences can vary.

%{\it String Accuracy} is defined as 
%$$1 - \frac{\textit{editDist(O,R)}}{\textit{length}(R)},$$ where $O$ is the surface system output, $R$ is the reference sentence and $editDist(A,B)$ is the minimal edit distance between the token strings $A$ and $B$ defined as the minimal number of substitutions, insertions and deletions needed to transform $A$ into $B$.

%Like {\sc bleu}, string accuracy often underestimates generation performance if the system output differs from the reference only in word order. 
%For example if a word occurs in the beginning of the string instead of the end, blocked by a number of correct words, edit distance will count two errors, one for removing the word in the beginning and one for adding it in the end. In this case it would be more reasonable to count a single movement error.
%The revised {\it generation string accuracy} does this in a post-processing step: Each insertion-deletion pair sharing the same token, but in different places is counted as a single movement error. 
%\newcite{bangalorerambowwhittaker2001}  refine these string based measures by referring to structural information from the reference derivation tree.

%Tree accuracy measure only the quality of the surface string. To compare the syntactic structure of the output with the reference we propose {\it derivation tree accuracy}, the ratio of derivation-treelets (nodes with their direct children in their specific order) that occurs in both the system output derivation and the reference derivation. 
%$$ \frac{|treelets(D_O) \cap treelets(D_R)|}{\textit{length}(D_R)}$$
% Notice that measures used in statistical parsing, e.g. bracketing and labelling accuracy, cannot be used to evaluate generation output because they can only measure similarity between two different syntactic structures for the same surface string.

\subsection{Results}
Table \ref{results} presents our results.
The planning systems we used was a reimplementation of {\sc FF} \cite{hoffmannnebel2001}, a heuristic search planner. We extended this planner to take action costs into account when it calculates heuristic values for successor states. We also experimented with the greedy search heuristic used by {\sc spud} \cite{stonedoran1997}.

The first two columns confirm our hypothesis that {\sc crisp} does not scale up to the automatically induced treebank. The greedy {\sc spud} heuristic does not generate a single correct sentence, as the combinatorics of the planning domain becomes too large with the treebank induced grammar. 
With the standard {\sc pltag} definition our system manages to generate only very few sentences due to data sparseness. However these sentences are almost identical to the ones found in the test set. Both generalized models ease the data sparseness problem, but output quality drops rapidly. For the unlexicalized model the search problem becomes very hard due to the large planning domain, so that 22\% of the generation problems require more than 5 minutes runtime. As expected the smoothed model offers a compromise between efficiency, output quality and coverage 
\footnote{In our experiments we set the weight between level 1 and 2 to 0.9 and did not consider level 3, after empirically testing different settings. We set a probability threshold to include operators into the planning domain to $10^{-4}$}.

\begin{table}
    \caption{\label{results} Evaluation results.}
    \begin{center}
    \begin{tabular}{|l|l|r|r|r|}
    \multicolumn{5}{c}{\sc crisp}\\ 
    \hline
     {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     greedy         & ??? & 0\% & 100\% & 0\%  \\ \hline 
     FF         &  0.66 & 88\% & 0\% & 12\% \\ \hline 
    \multicolumn{5}{c}{\bf planning with {\sc pltag}}\\\hline
    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     FF + costs &  0.99 &  97\% & 0\% & 3\% \\ \hline
     FF         &  0.99 & 71\% & 0\% &  29\% \\ \hline
    \multicolumn{5}{c}{\bf planning with {\sc pltag} (unlexicalized)}\\\hline
    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     FF + costs & 0.74 & 62\% & 16\% & 22\%\\ \hline
     FF & 0.76 & 62\% & 16\% &  22\%\\ \hline
    \multicolumn{5}{c}{\bf planning with {\sc pltag} (smoothed)}\\\hline
    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     FF + costs & 0.78 & 19\% & 74\% & 7\% \\ \hline
     FF & 0.77 & 19\% & 74\% & 7\% \\ \hline
    \end{tabular}
    \end{center}
\end{table}

### 2010-05-19 trunk/Papers/pcrisp-10/introduction.tex
Added draft for TAG+ paper about pcrisp

\section{Introduction}
Traditional sentence generation systems adopt a pipeline architecture that includes sentence planning and surface realization as two independent modules \cite{reiterdale2000}.  A different view is advocated by \newcite{stonedoran1997}. Their sentence generation system {\sc spud} allows a combined treatment of sentence planning and surface realization.
{\sc spud} exploits {\sc ltag}'s localization of argument structure and uses a top-down construction algorithm to incrementally builds both syntactic and flat semantic structure that satisfies the communicative goal and is consistent with semantic and pragmatic constraints. 
This unified approach to sentence generation is attractive because it tackles different phenomena in generation in an elegant way using a unified, linguistically well motivated framework.

While {\sc spud} uses an incomplete heuristic greedy search, to solve the {\sc NP-}complete problem of finding such a derivation \cite{kollerstriegnitz2002}, \newcite{kollerstone2007} present an encoding of this problem into an AI planning problem.

We expect that using larger grammars with this approach leads to problems in both performance and output quality, due to over-generation. 
Our work approaches this problem by combining integrated sentence generation as planning (following \newcite{kollerstone2007}), numeric AI planning and a probability model ({\sc pltag}, \newcite{resnik1992}).

{\bf Structure of the paper:}
In section \ref{sec:related} we discuss related work. Section \ref{sec:crisp} reviews the original generation as planning approach ({\sc crisp}) suggested by \newcite{kollerstone2007}. We describe our statistical extension of this approach ({\sc pcrisp}) in section \ref{sec:pcrisp} and evaluate our new system in section \ref{sec:experiments}. Finally we conclude this paper in section \ref{sec:conclusion}.

### 2010-05-19 trunk/Papers/pcrisp-10/pcrisp-10.tex
Added draft for TAG+ paper about pcrisp

%
% File naaclhlt2010.tex
%
% Contact: nasmith@cs.cmu.edu

\documentclass[11pt,letterpaper]{article}
\usepackage{naaclhlt2010}
\usepackage{times}
\usepackage{latexsym}
\usepackage{graphicx}
\setlength\titlebox{6.5cm}    % Expanding the titlebox

\newlength{\rhwidth}
\newcommand{\planaction}[3]{
\begin{minipage}{.5\textwidth}
\footnotesize
\setlength{\rhwidth}{.5\textwidth}
\addtolength{\rhwidth}{2.0cm}
{#1:}

\noindent \hspace{.2cm} \begin{minipage}{1.5cm}{Precond:}\end{minipage}\begin{minipage}[t]{\rhwidth} #2 \end{minipage}

\noindent \hspace{.2cm} \begin{minipage}{1.5cm}{Effect:}\end{minipage} \begin{minipage}[t]{\rhwidth}\ #3 \end{minipage}
\end{minipage}
}

\newcommand{\cplanaction}[4]{
\begin{minipage}{.5\textwidth}
\setlength{\rhwidth}{.5\textwidth}
\addtolength{\rhwidth}{2.0cm}
{#1:}

\noindent \hspace{.2cm} \begin{minipage}{1.5cm}{Precond:}\end{minipage}\begin{minipage}[t]{\rhwidth} #2 \end{minipage}

\noindent \hspace{.2cm} \begin{minipage}{1.5cm}{Effect:}\end{minipage} \begin{minipage}[t]{\rhwidth} #3 \end{minipage}

\noindent \hspace{.2cm} \begin{minipage}{1.5cm}{Cost:}\end{minipage} \begin{minipage}[t]{\rhwidth} \ensuremath{#4} \end{minipage}
\end{minipage}
}

\title{Sentence Generation as Planning with Probabilistic LTAG\\
%\Thanks{}
}

\author{Daniel Bauer\\
  Columbia University\\
  New York, NY, USA\\
  {\tt bauer@cs.columbia.edu} 
  \And
  Alexander Koller\\
  Saarland University\\
  Saarbr\"{u}cken, Germany\\ 
  {\tt koller@mmci.uni-saarland.de}}

\date{}

\begin{document}
\maketitle
\begin{abstract}
Our work merges combined sentence planning and surface realization using 
{\sc ltag} and statistical generation. This makes it possible to address
integrated sentence generation with large, treebank induced grammars for
the first time.
\end{abstract}

\input{introduction}
\input{related} % Related work
\input{crisp} % Sentence Generation as Planning
\input{pcrisp} % Statistical Generation as Planning
\input{experiments} % Experiments
\input{conclusion}

\bibliographystyle{naaclhlt2010}
\bibliography{literature}

\end{document}

### 2010-05-19 trunk/Papers/pcrisp-10/pcrisp.tex
Added draft for TAG+ paper about pcrisp

\section{Statistical Generation as Planning}
\label{sec:pcrisp}

We now extend {\sc crisp} to statistical generation ({\sc pcrisp}). The basic idea is to add a statistical grammar model while leaving the sentence generation mechanism untouched. This way we can select the highest scoring derivation, that satisfies all constraints (grammaticality, expresses the communicative goal, use unambiguous refering expressions etc.). We first propose three probability models in the next section and adapt {\sc crisp} planning problems to acomodate probabilities in section \ref{ssec:pcrisp-domains}.

\subsection{Probability Models}
\label{ssec:probmodels}

{\bf Probabilistic TAG:}
As a basic probability model over {\sc ltag} derivations we use probabilistic {\sc tag}\footnote{we refer to this formalism as \sc{pltag}} \cite{resnik1992}, 
which provides a probability for each substitution and adjunction event in a derivation. In addition, it assigns a probability to the event of starting a derivation with a specific initial tree and to the event of not adjoining anything to an open adjunction node at all. Assuming that the probabilities of these events within a derivation are independent of each other, the probability of a derivation is then simply defined as the product of all elementary event probabilities.
Formally,

\newcommand{\adj}[0]{\textit{ adj}}
\newcommand{\init}[0]{\textit{ init}}
\newcommand{\subst}[0]{\textit{ subst}}
$$ \sum\limits_{\alpha \in I}\sum\limits_{lex \textit{ of } \alpha} P_{init}(\init(\alpha, lex)) = 1 $$

$$\sum\limits_{\alpha \in I}\sum\limits_{lex \textit{ of }\alpha} P_{subst}(\subst(\tau,lex_\tau, \alpha, lex_\alpha, n)) = 1,$$
for each $\tau \in I \cup A$, each lexicalization $lex_\tau$  of $\tau$ and for each substitution node $n$ in $\tau$.

$$ P_{adj}(\adj(\tau, lex_\tau, \textit{None},n)) +$$
$$ \sum\limits_{\beta \in A}\sum\limits_{lex \textit{ of }\beta} P_{adj}(\adj(\tau,lex_\tau, \alpha, lex_\alpha, n)) = 1,$$ for each $\tau \in I \cup A$, each lexicalization $lex_\tau$  of $\tau$ and for each node $n$ that is open for adjunction in $\tau$.

Our choice of {\sc pltag} for sentence generation is motivated by a number of attractive properties.
 First, as {\sc pltag} is lexicalized, it does not only assign probabilities to operations in the grammar (as for example plain {\sc pcfg}), but also accounts for binary dependencies between words. Unlike n-gram models however, these co-occurences are structured according to local syntactic context as a result of {\sc tag}'s extended domain of locality. The probability model describes how the syntactic arguments of a word are typically filled. 
Furthermore, as {\sc tag} factors recursion from the domain of dependencies, the probability for core constructions remains the same independent of additional adjunctions. 
Second, {\sc pltag} is a generative model. This is crucial because it allows us to assign probabilities to partial derivations. In addition we use the probability of partial derivations as a cost function, to help guide search through the space of possible PLTAG derivations. In contrast, other statistical generation systems use discriminative models that select the best possible generation output from a set of candidate sentences.
Finally, the independence assumption makes it easy to estimate {\sc pltag}s from a treebank.

To alleviate data sparseness, caused by the specific event definition of {\sc pltag}, we propose two alternative probability models. The first model is an unlexicalized version of {\sc pltag}, the second model uses a simple smoothing approach.\\ 
{\bf Unlexicalized Probabilities:}
  An easy way to deal with data sparseness is to drop all lexicalization from event definitions, as illustrated in figure \ref{fig:modelillustration}, A. Unfortunately the model does no longer account for bilexical dependencies between words. Since our system has to add a lexicalized tree in each step, lexicalization for this child tree should always be taken into account by the probability model, if available. Despite these drawbacks, we perform experiments with the unlexicalized model as a baseline. This allows us to investigate if purely syntactic information is sufficient to achieve high quality generation output. \\  
{\bf Linear Interpolation:}
 The third model computes linear interpolation between three back-off levels. The first level is just standard {\sc pltag} (figure \ref{fig:modelillustration}, B.1), for the second level the lexicalization of the parent tree is dropped (figure \ref{fig:modelillustration}, B.2), for the third level the model describes only the distribution of lexicalized child trees over each category (figure \ref{fig:modelillustration}, B.3). 
The third level is similar in spirit to a probabilistic version of the original generation as planning formulation in \cite{kollerstone2007}.
\begin{figure}[t]
\begin{center}
\includegraphics[width=.6\textwidth]{figures/modelillustration}
\caption{\label{fig:modelillustration} Illustration of the unlexicalized probability model (A) and the three back-off levels of the linear interpolation model (B). B.1 corresponds to the original {\sc pltag} definition.}

\end{center}
\end{figure}

%It is obvious that the number of total planning operators grows quadratically with the number of possible input words in the grammar. In practice this causes a problem for most heuristic search planners, which will instantiate planning operators to all possible actions before attempting to solve the problem. Our generation system therefore only selects operators that are compatible with the input semantics before running the planner. 
%On the other hand,

\subsection{PCRISP Planning Domains}
\label{ssec:pcrisp-domains}
In this section we reformulate the {\sc crisp} planning operators described in section \ref{ssec:crispdomain}.
The independence assumption in {\sc pltag} allows us to maintain that planning operator add a single tree, but are now assigned a probability score. However, while {\sc crisp} planning operators can add an elementary tree to any site of the right category, {\sc pltag} events are binary events between lexicalized trees at a specific node. We therefore adapt the literals that record open substitution and adjunction sites in partial derivations accordingly and create one operator for each node in each possible combination of lexicalized trees.

Different approaches to planning with numeric values exist. We can encode probabilities either as timespans required by each operator and use a temporal planner, or as as generic numeric variable that is increased by each operator. In both cases this value (which can be seen as a cost) is summed over actions and a resulting plan should seek to minimize it. We therefore set the cost of an operator to be its negative log probability. 
Figure \ref{example-action} shows an example planning operator. We adopt this scheme for all planning operators regardless of the probability model.
\begin{figure}[t]
\begin{center}
\cplanaction{\bf subst-t3-cat-t28-eats-n2(u,~x1)}{step(step1),referent(u,~x1),\\ subst(t28-eats,n2,~u), cat(x1)}{
$\lnot$needtoexpress(pred-cat,~x1),\\ $\lnot$subst(t-28-eats, n2,~u),\\
adj(t3-cat, n0, u), adj(t3-cat, n1 u)\\
$\lnot$ step(step1), step(step2)}{4.3012}\\\smallskip
\caption{\label{example-action} {\sc pcrisp} Operator to substitute $t3$ lexicalized with 'cat' for node 2 of $t28$ lexicalized with `eats'.}
\end{center}
\end{figure}
%\newcommand{\init}[0]{\textit{ init}}
%\newcommand{\subst}[0]{\textit{ subst}}

%\begin{figure}[p]
%\caption{\label{modelillustration} The three back-off levels.}
%\begin{center}
%\includegraphics[width=.5\textwidth]{modelillustration}
%\end{center}
%\end{figure}

### 2010-05-19 trunk/Papers/pcrisp-10/related.tex
Added draft for TAG+ paper about pcrisp

\section{Related Work}
\label{sec:related}

Previous statistical NLG systems (e.g. \newcite{knight1995}, {\sc nitrogen} \cite{langkildeknight1998}, {\sc openCCG} \cite{whitebaldridge2003}
and the {\sc ltag} based realizer {\sc fergus} \cite{bangalorerambow2000}), follow a generate-and-select strategy using n-gram models over output structures
for discrimination. But since this approach focuses on discriminating realization output for the same sentence plan only, it is incompatible with the
integrated treatment of sentence planning and realization.

One of the few systems uses generative models for sentence generation is {\sc pCru} \newcite{belz2008}. However, her model is defined on possible states of the generation system itself, not on the underlying grammar formalism.

$\cdots$

### 2010-05-19 trunk/Papers/pcrisp-10/crisp.tex
Typos.

@@
 \includegraphics[width=.8\textwidth]{figures/grammar.pdf}
-\caption{\label{fig:grammar}(a) An example grammar with semantic content. (b) A derivation for ``the cat eats raw tuna''. }
+\caption{\label{fig:grammar}(a) An example grammar with semantic content. (b) A derivation for ``the cat eats the raw fish''. }
 \end{center}
@@
 For example, given the knowledge base
-{\it \{cat($i_2$), tuna($i_3$), raw($i_3$), eats($i_1$,$i_2$,$i_3$)\}} and the grammar in \ref{fig:grammar}(a), {\sc crisp} could produce the derivation in \ref{fig:grammar}(b).
+{\it \{cat($i_2$), fish($i_3$), raw($i_3$), eats($i_1$,$i_2$,$i_3$)\}} and the grammar in \ref{fig:grammar}(a), {\sc crisp} could produce the derivation in \ref{fig:grammar}(b).
 
@@
 In the operator effect, open substitution nodes are closed and new identifiers are created for each substitution node and internal node in the new tree.
-Given the knowlede base from above, a plan corresponding to the derivation in figure \ref{fig:grammar} would be $\langle${\it S-eats-1(root, i1, i2, i3), NP-cat-2(subj-1, i2), NP-tuna-3(obj-1, i3), N-raw-4(obj-1, i3)}$\rangle$
+Given the knowlede base from above, a plan corresponding to the derivation in figure \ref{fig:grammar} would be $\langle${\it S-eats-1(root, i1, i2, i3), NP-cat-2(subj-1, i2), NP-fish-3(obj-1, i3), N-raw-4(obj-1, i3)}$\rangle$
 

### 2010-05-19 trunk/Papers/pcrisp-10/crisp.tex
Many local changes, restructured section 4

@@
 Like {\sc spud}, {\sc crisp} uses an {\sc ltag} in which elementary trees are assigned semantic content.
-Each node in a {\sc crisp} elementary tree node is associated with a semantic role.
+Each node in a {\sc crisp} elementary tree is associated with a semantic role.
 Semantic content is expressed as a set of literals, encoding relations between these roles.
@@
 For example, given the knowledge base
-{\it \{cat($i_2$), fish($i_3$), raw($i_3$), eats($i_1$,$i_2$,$i_3$)\}} and the grammar in \ref{fig:grammar}(a), {\sc crisp} could produce the derivation in \ref{fig:grammar}(b).
+{\it \{cat($i_2$), fish($i_3$), raw($i_3$), fish($i_4$), eats($i_1$,$i_2$,$i_3$)\}} and the grammar in \ref{fig:grammar}(a), {\sc crisp} could produce the derivation in \ref{fig:grammar}(b).
+Notice that {\sc crisp} can generate the unambiguos refering expression `the raw fish' for $i_3$, to distinguish it from $i_4$.
+For details on  refering expression generation in {\sc crisp} see \newcite{garoufikoller2010}.
 
@@
 Before we describe {\sc crisp}'s encoding of sentence generation as planning, we briefly review AI planning in general. 
-The crucial concept in AI planning is a state, a conjunction of first order literals describing relations between some individuals.
+A planning state is a conjunction of first order literals describing relations between some individuals.
 A planning problem consist of an initial state, a set of goal states and a set of planning operators that describe possible state transitions.
 A solution to the problem is any sequence of actions (instantiated operators) that leads from the initial state to one of the goal states.
-Planning problems can be written in a standard notation ({\sc pddl}\footnote{Planning Domain Description Language}, \newcite{gerevinietal2009}) and can be solved efficiently by planning systems such as {\sc ff}\footnote{Fast Forward} \cite{hoffmannnebel2001}.
+Planning problems can be solved efficiently by general purpose planning systems such as
+{\sc ff} \cite{hoffmannnebel2001}.  
+% Planning problems can be written in a standard notation ({\sc pddl}\footnote{Planning Domain Description Language}, \newcite{gerevinietal2009}).
 
@@
 \begin{center}
-\planaction{\bf S-eats-1(u,~x1,~x2,~x3)}{step(step1),referent(u,~x1),\\ subst(S,~u), eats(x1,~x2,~x3)}{
+\planaction{\bf subst-t28-eats-S(u,~x1,~x2,~x3)}{referent(u,~x1),\\ subst(S,~u), eats(x1,~x2,~x3)}{
 $\lnot$needtoexpr(pred-eats,~x1,~x2,~x3),\\ $\lnot$subst(S,~u),\\
-subst(NP,~subj-1), subst(NP,~obj-1),\\ refer(subj-1,~x2), refer(obj-1,~x3)\\
-canadj(VP, u), canadj(V, u), canadj(S, u)\\
-$\lnot$ step(step1), step(step2)}\\\smallskip
+subst(NP,~subj-1), subst(NP,~obj-1),\\ referent(subj-1,~x2), referent(obj-1,~x3)\\
+canadj(VP, u), canadj(V, u), canadj(S, u)}\\\smallskip
 
-\planaction{\bf NP-cat-2(u,~x1)}{step(step1),referent(u,~x1),\\ subst(NP,~u), cat(x1)}{
-$\lnot$needtoexpress(pred-cat,~x1),\\ $\lnot$subst(NP,~u),\\
-canadj(N, u), canadj(NP, u)\\
-$\lnot$ step(step1), step(step2)}\\\smallskip
+\planaction{\bf subst-t3-cat-NP(u,~x1)}{referent(u,~x1),\\ subst(NP,~u), cat(x1)}{
+$\lnot$needtoexpr(pred-cat,~x1),\\ $\lnot$subst(NP,~u),\\
+canadj(N, u), canadj(NP, u)}\\ \smallskip
 
-\planaction{\bf N-raw-4(u,~x1)}{step(step1),referent(u,~x1),\\ canadj(N,~u), raw(x1)}{
-$\lnot$needtoexpress(pred-raw,~x1),\\ $\lnot$canadj(N,~u),\\
-$\lnot$ step(step4), step(step5)}\\\smallskip
+\planaction{\bf subst-t5-raw-N(u,~x1)}{referent(u,~x1),\\ canadj(N,~u), raw(x1)}{
+$\lnot$needtoexpr(pred-raw,~x1),\\ $\lnot$canadj(N,~u)}\\\smallskip
 \end{center}
@@
 \subsection{CRISP and Large Grammars}
-Trees in large, possibly treebank induced grammars are often applicable in a number of different syntactic contexts, leading to overgeneration.
-Such trees contain many internal nodes that are open for adjunction but will not necessarily be filled in a derivation for a given input communicative goal.
+Using {\sc crisp} with large, treebank induced grammars can lead to overgeneration, 
+%Such trees contain many internal nodes that are open for adjunction but will not necessarily be filled in a derivation for a given input communicative goal.
 Attaching to the wrong node can produce invalid or disprefered word order.
-In addition, in such grammars the ambiguity of possible, structurally different trees (supertags) for a given word is high.
-These problems are illustrated in figure \ref{fig:overgen}.
-Assuming a grammar that includes trees for both right attaching ($t13$) and left attaching PPs ($t252$), both derivations (a) and (b) are {\it grammatical} derivations that satisfy the same communicative goal.
+due to the high lexical ambiguity of possible, structurally different trees (supertags) for a given word.
+We illustrate this problem in figure \ref{fig:overgen}.
+Assuming a grammar that includes trees for both right adjoining ($t13$) and left adjoining PPs ($t252$), both derivations (a) and (b) are {\it grammatical} derivations that satisfy the same communicative goal.
 However, most readers disprefer the reading in (b).
 Clearly, to use {\sc crisp} with such a grammar we need a method of distinguishing good derivations from bad ones.
-\begin{figure*}
+In section \ref{sec:experiments} we provide empirical evidence for this problem and show how it is solved by the approach we describe in the next section.
+\begin{figure}
 \begin{center}
-\includegraphics[width=.6\textwidth]{figures/overgen.pdf}
+\includegraphics[width=.3\textwidth]{figures/overgen.pdf}
 \caption{\label{fig:overgen} Two derivations with a large grammar, that satisfy the same communicative goal. Sentence (b) is disprefered by most readers.}
 \end{center}
-\end{figure*}
+\end{figure}

### 2010-05-19 trunk/Papers/pcrisp-10/experiments.tex
Many local changes, restructured section 4

@@
 We are interested in the output quality and runtime behavior of different combinations of planning systems and probability models.
-We can assume that the sentence planning abilities are inherited from {\sc crisp} and only evaluate the realization abilities of our system by comparing generated sentences against a gold standard testset.
+Like {\sc crisp}, {\sc pcrisp} is a sentence generation system and can for instance generate refering expressions.
+Unfortunately, since we do not have an evaluation dataset for full sentence generation, we only evaluate the realization abilities of our system by comparing generated sentences against a gold standard testset.
 The next section describes our evaluation data and section \ref{ssec:eval-measures} describes the evaluation measures used.
@@
 
-For our experiments we use a grammar and treebank, that was automatically extracted from the Wall Street Journal portion of the Penn Treebank (41576 sentences)
-using the algorithm described by \newcite{chenschanker2004}.  We preprocess this grammar by linearizing multiple adjunctions and by re-attaching prepositions to trees they appear with.
+For our experiments we use a grammar and treebank that was automatically extracted from the Wall Street Journal using the algorithm described by \newcite{chenschanker2004}.
+
+This algorithm outputs a grammar that allows multiple adjunctions at the same node.
+For such a grammar {\sc ptag} is not a suitable probability model.
+Models that can deal with multiple adjunction are discussed by \newcite{nasrrambow2006}, but employing them would require non-trivial modifications to our encoding as a metric planning problem.
+We therefore preprocess the treebank by `linearizing' multiple adjunctions.
+Furthermore, we attach prepositions to the trees they substitute in, to increase the expressiveness of the bilexical probability distribution.
+For details on these preprocessing steps see \cite{bauer2009}.
 

### 2010-05-19 trunk/Papers/pcrisp-10/pcrisp.tex
Many local changes, restructured section 4

@@
 This way we can select the highest scoring derivation, that satisfies all constraints (grammaticality, expresses the communicative goal, use unambiguous refering expressions etc.).
-We first propose three probability models in the next section and adapt {\sc crisp} planning problems to acomodate probabilities in section \ref{ssec:pcrisp-domains}.
 
-\subsection{Probability Models}
+ As a straightforward probability model over {\sc ltag} derivations we choose probabilistic {\sc tag} ({\sc ptag}) \cite{resnik1992}.
+Our choice of {\sc ptag} for sentence generation is motivated by a number of attractive properties.
+ {\sc Ptag} is lexicalized and therefore does not only assign probabilities to operations in the grammar (as for example plain {\sc pcfg}), but also accounts for binary dependencies between words.
+Unlike n-gram models however, these co-occurences are structured according to local syntactic context as a result of {\sc tag}'s extended domain of locality.
+The probability model describes how the syntactic arguments of a word are typically filled.
+Furthermore, as {\sc tag} factors recursion from the domain of dependencies, the probability for core constructions remains the same independent of additional adjunctions. 
+%Second, {\sc ptag} is a generative model.
+This is crucial because it allows us to assign probabilities to partial derivations.
+In addition we use the probability of partial derivations as a cost function, to help guide search through the space of possible PLTAG derivations.
+In contrast, other statistical generation systems use discriminative models that select the best possible generation output from a set of candidate sentences.
+%Finally, the independence assumption makes it easy to estimate {\sc ptag}s from a treebank.
+ We review {\sc ptag} in section \ref{ssec:probmodels}.
+
+While we leave the basic sentence generation mechanism intact, we need to modify the concrete formulation of {\sc crisp} planning operators, as each of them now corresponds to some probability event.
+Likewise, we need to take the step from classical planning to metric planning systems, that can use the probabilities.
+In metric planning \cite{fox2002}, instead of changing only boolean state variables, planning actions can decrease or increase a numeric variable.
+The goal state specifies constraints on this variable.
+In the simplest case the variable can only be increased by a static cost value in each action, and the goal state contains the objective to minimize the total cost.
+Systems such as metric-{\sc FF} \cite{hoffmann2003} do not guarantee optimality, but generally offer good results.
+We address our encoding of sentence generation with {\sc ptag} as metric planning in section \ref{ssec:pcrisp-domains}.
+
+\subsection{Probabilistic TAG}
 \label{ssec:probmodels}
-
-{\bf Probabilistic TAG:}
-As a basic probability model over {\sc ltag} derivations we use probabilistic {\sc tag}\footnote{we refer to this formalism as \sc{pltag}} \cite{resnik1992}, 
-which provides a probability for each substitution and adjunction event in a derivation.
+{\sc ptag} \cite{resnik1992} views {\sc tag} derivations as sequences of probability events, each corresponding to a substitution or adjunction.
 In addition, it assigns a probability to the event of starting a derivation with a specific initial tree and to the event of not adjoining anything to an open adjunction node at all.
@@
 
-Our choice of {\sc pltag} for sentence generation is motivated by a number of attractive properties.
- First, as {\sc pltag} is lexicalized, it does not only assign probabilities to operations in the grammar (as for example plain {\sc pcfg}), but also accounts for binary dependencies between words.
-Unlike n-gram models however, these co-occurences are structured according to local syntactic context as a result of {\sc tag}'s extended domain of locality.
-The probability model describes how the syntactic arguments of a word are typically filled.
-Furthermore, as {\sc tag} factors recursion from the domain of dependencies, the probability for core constructions remains the same independent of additional adjunctions. 
-Second, {\sc pltag} is a generative model.
-This is crucial because it allows us to assign probabilities to partial derivations.
-In addition we use the probability of partial derivations as a cost function, to help guide search through the space of possible PLTAG derivations.
-In contrast, other statistical generation systems use discriminative models that select the best possible generation output from a set of candidate sentences.
-Finally, the independence assumption makes it easy to estimate {\sc pltag}s from a treebank.
+\subsection{PCRISP Planning Domains}
+\label{ssec:pcrisp-domains}
+Here, we reformulate the {\sc crisp} planning operators described in section \ref{ssec:crispdomain}.
+The independence assumption in {\sc ptag} allows us to maintain that planning operator add a single tree, but are now assigned a probability score.
+However, while {\sc crisp} planning operators can add an elementary tree to any site of the right category, {\sc ptag} events are binary events between lexicalized trees at a specific node.
+We therefore adapt the literals that record open substitution and adjunction sites in partial derivations accordingly and create one operator for each node in each possible combination of lexicalized trees.
+Figure \ref{example-action} shows an example planning operator. We adopt this scheme for all planning operators regardless of the probability model.
+\begin{figure}[t]
+\begin{center}
+\cplanaction{\bf subst-t3-cat-t28-eats-n2(u,~x1)}{referent(u,~x1),\\ subst(t28-eats,n2,~u), cat(x1)}{
+$\lnot$needtoexpress(pred-cat,~x1),\\ $\lnot$subst(t-28-eats, n2,~u),\\
+adj(t3-cat, n0, u), adj(t3-cat, n1 u)}{4.3012}\\\smallskip
+\caption{\label{example-action} {\sc pcrisp} Operator to substitute $t3$ lexicalized with 'cat' for node 2 of $t28$ lexicalized with `eats'.}
+\end{center}
+\end{figure}
 
-To alleviate data sparseness, caused by the specific event definition of {\sc pltag}, we propose two alternative probability models.
-The first model is an unlexicalized version of {\sc pltag}, the second model uses a simple smoothing approach.\\
-{\bf Unlexicalized Probabilities:}
-  An easy way to deal with data sparseness is to drop all lexicalization from event definitions, as illustrated in figure \ref{fig:modelillustration}, A.
-Unfortunately the model does no longer account for bilexical dependencies between words.
+Finally, we set the cost of an operator to be its negative log probability.
+For example for the operator in figure \ref{example-action} $$Cost( \subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2) ) =$$
+$$ -\log( P(\subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2)) ). $$ This way the plan with the smallest cost sum corresponds to the {\sc tag} derivation with the highest probability.
+
+\subsection{Dealing with Data Sparseness}
+\label{ssec:sparseness}
+%To alleviate data sparseness, caused by the specific event definition of {\sc ptag}, we propose two alternative probability models.
+The first model is an unlexicalized version of {\sc ptag}, the second model uses a simple smoothing approach.\\
+The probability event definition of {\sc ptag} is very fine grained.
+Substitution and adjunction events depend on specific parent and child trees with specific lexicalizations and on a node in the parent tree, as illustrated in figure \ref{fig:modelillustration}, B.1.
+When we estimate a probability model from training data, we cannot expect to observe evidence for all combinations of trees.
+Derivations that include such unseen events, have zero probability and are therefore impossible.
+As we show in section \ref{sec:experiments}, this gives rise to a massive data sparseness problem.
+
+  A straightforward way to deal with data sparseness is to drop all lexicalization from event definitions, as illustrated in figure \ref{fig:modelillustration}, A.
+Unfortunately this model does no longer account for bilexical dependencies between words.
 Since our system has to add a lexicalized tree in each step, lexicalization for this child tree should always be taken into account by the probability model, if available.
@@
 \\
-{\bf Linear Interpolation:}
- The third model computes linear interpolation between three back-off levels.
-The first level is just standard {\sc pltag} (figure \ref{fig:modelillustration}, B.1), for the second level the lexicalization of the parent tree is dropped (figure \ref{fig:modelillustration}, B.2), for the third level the model describes only the distribution of lexicalized child trees over each category (figure \ref{fig:modelillustration}, B.3).
-The third level is similar in spirit to a probabilistic version of the original generation as planning formulation in \cite{kollerstone2007}.
+An alternative model computes linear interpolation between three back-off levels.
+The first level is just standard {\sc ptag} (figure \ref{fig:modelillustration}, B.1), for the second level the lexicalization of the parent tree is dropped (figure \ref{fig:modelillustration}, B.2), for the third level the model describes only the distribution of lexicalized child trees over each category (figure \ref{fig:modelillustration}, B.3).
+Notice also, that the third level is similar in spirit to a probabilistic version of original {\sc crisp} operators (compare figure \ref{fig:modelillustration}, B.3 and figure \ref{fig:crisp-operators}.
 \begin{figure}[t]
 \begin{center}
-\includegraphics[width=.6\textwidth]{figures/modelillustration}
+\includegraphics[width=.5\textwidth]{figures/modelillustration}
 \caption{\label{fig:modelillustration} Illustration of the unlexicalized probability model (A) and the three back-off levels of the linear interpolation model (B).
-B.1 corresponds to the original {\sc pltag} definition.}
+B.1 is the original {\sc ptag} definition and corresponds directly to {\sc pcrisp} operators as in figure \ref{example-action}, while B.3 resembles {\sc crisp} operators (c.f.
+figure \ref{fig:crisp-operators}).}
 
@@
 
-\subsection{PCRISP Planning Domains}
-\label{ssec:pcrisp-domains}
-In this section we reformulate the {\sc crisp} planning operators described in section \ref{ssec:crispdomain}.
-The independence assumption in {\sc pltag} allows us to maintain that planning operator add a single tree, but are now assigned a probability score.
-However, while {\sc crisp} planning operators can add an elementary tree to any site of the right category, {\sc pltag} events are binary events between lexicalized trees at a specific node.
-We therefore adapt the literals that record open substitution and adjunction sites in partial derivations accordingly and create one operator for each node in each possible combination of lexicalized trees.
-
-Different approaches to planning with numeric values exist.
-We can encode probabilities either as timespans required by each operator and use a temporal planner, or as as generic numeric variable that is increased by each operator.
-In both cases this value (which can be seen as a cost) is summed over actions and a resulting plan should seek to minimize it.
-We therefore set the cost of an operator to be its negative log probability.
-Figure \ref{example-action} shows an example planning operator. We adopt this scheme for all planning operators regardless of the probability model.
-\begin{figure}[t]
-\begin{center}
-\cplanaction{\bf subst-t3-cat-t28-eats-n2(u,~x1)}{step(step1),referent(u,~x1),\\ subst(t28-eats,n2,~u), cat(x1)}{
-$\lnot$needtoexpress(pred-cat,~x1),\\ $\lnot$subst(t-28-eats, n2,~u),\\
-adj(t3-cat, n0, u), adj(t3-cat, n1 u)\\
-$\lnot$ step(step1), step(step2)}{4.3012}\\\smallskip
-\caption{\label{example-action} {\sc pcrisp} Operator to substitute $t3$ lexicalized with 'cat' for node 2 of $t28$ lexicalized with `eats'.}
-\end{center}
-\end{figure}
 %\newcommand{\init}[0]{\textit{ init}}

### 2010-05-20 trunk/Papers/pcrisp-10/crisp.tex
more changes. example for all pcrisp operator types

@@
 $\lnot$needtoexpr(pred-eats,~x1,~x2,~x3),\\ $\lnot$subst(S,~u),\\
-subst(NP,~subj-1), subst(NP,~obj-1),\\ referent(subj-1,~x2), referent(obj-1,~x3)\\
-canadj(VP, u), canadj(V, u), canadj(S, u)}\\\smallskip
+subst(NP,~subj), subst(NP,~obj),\\ referent(subj,~x2), referent(obj,~x3)\\
+adj(VP, u), adj(V, u), adj(S, u)}\\\smallskip
 
@@
 $\lnot$needtoexpr(pred-cat,~x1),\\ $\lnot$subst(NP,~u),\\
-canadj(N, u), canadj(NP, u)}\\ \smallskip
+adj(N, u), adj(NP, u)}\\ \smallskip
 
-\planaction{\bf subst-t5-raw-N(u,~x1)}{referent(u,~x1),\\ canadj(N,~u), raw(x1)}{
-$\lnot$needtoexpr(pred-raw,~x1),\\ $\lnot$canadj(N,~u)}\\\smallskip
+\planaction{\bf subst-t5-raw-N(u,~x1)}{referent(u,~x1),\\ adj(N,~u), raw(x1)}{
+$\lnot$needtoexpr(pred-raw,~x1),\\ $\lnot$adj(N,~u)}\\\smallskip
 \end{center}

### 2010-05-20 trunk/Papers/pcrisp-10/experiments.tex
more changes. example for all pcrisp operator types

@@
 \subsection{Evaluation Data}
-
 For our experiments we use a grammar and treebank that was automatically extracted from the Wall Street Journal using the algorithm described by \newcite{chenschanker2004}.
@@
 \label{ssec:eval-measures}
-To evaluate the ouput quality of our statistical generation system we compare the system output against the reference sentences in the treebank, from which the input representations were generated.
+To evaluate the ouput quality of our statistical generation system we compare the system output $O$ against the reference sentence $R$in the treebank, from which the input representation was generated.
 We adopt the {\bf generation tree accuracy} measure discussed by \cite{bangalorerambowwhittaker2001}.
-This measure is computed by first creating a list of all 'treelets' from the reference derivation tree.
+This measure is computed by first creating a list of all 'treelets' from the reference derivation tree $D$.
 A 'treelet' is a subtree, consisting only of a node and its direct children.
@@
 
+\subsection{Planning Systems}
+The planning system we used was a reimplementation of {\sc FF} \cite{hoffmannnebel2001}, a heuristic search planner.
+We extended this planner to take action costs into account when it calculates heuristic values for successor states.
+We also experimented with the greedy search heuristic used by {\sc spud} \cite{stonedoran1997}.
+
 \subsection{Results}
 Table \ref{results} presents our results.
-The planning systems we used was a reimplementation of {\sc FF} \cite{hoffmannnebel2001}, a heuristic search planner.
-We extended this planner to take action costs into account when it calculates heuristic values for successor states.
-We also experimented with the greedy search heuristic used by {\sc spud} \cite{stonedoran1997}.
 

### 2010-05-20 trunk/Papers/pcrisp-10/pcrisp.tex
more changes. example for all pcrisp operator types

@@
 \label{ssec:probmodels}
-{\sc ptag} \cite{resnik1992} views {\sc tag} derivations as sequences of probability events, each corresponding to a substitution or adjunction.
-In addition, it assigns a probability to the event of starting a derivation with a specific initial tree and to the event of not adjoining anything to an open adjunction node at all.
-Assuming that the probabilities of these events within a derivation are independent of each other, the probability of a derivation is then simply defined as the product of all elementary event probabilities.
-Formally,
+{\sc ptag} \cite{resnik1992} views {\sc tag} derivations as sequences of probability events of three types, initial events, substitution events and adjunction events.
 
@@
 \newcommand{\subst}[0]{\textit{ subst}}
-$$ \sum\limits_{\alpha \in I}\sum\limits_{lex \textit{ of } \alpha} P_{init}(\init(\alpha, lex)) = 1 $$
+The probability distribution for initial events describes how likely it is to start any derivation with a given initial tree. 
+It is defined over all initial trees $\alpha$ with their possible lexicalizations $w \in W_\alpha$: 
+$$ \sum\limits_{\alpha \in I}\sum\limits_{w \in W_\alpha} P_{init}(\init(\alpha, w)) = 1 $$
 
-$$\sum\limits_{\alpha \in I}\sum\limits_{lex \textit{ of }\alpha} P_{subst}(\subst(\tau,lex_\tau, \alpha, lex_\alpha, n)) = 1,$$
-for each $\tau \in I \cup A$, each lexicalization $lex_\tau$  of $\tau$ and for each substitution node $n$ in $\tau$.
+For substitution events, there is a probability distribution for each substitution node $n$ of each initial tree $\tau$ lexicalized with $v$, which describes how likely it is to substitute it with an initial tree $\alpha$ lexicalized with $w$.
+$$\sum\limits_{\alpha \in I}\sum\limits_{w \in W_\alpha} P_{subst}(\subst(\tau,v , \alpha, w, n)) = 1$$
+
+Similarly for each internal node there is a distribution that describes the probability to adjoin each auxiliary tree $\beta$ lexicalized with $w$.
+In addition some probability mass is reserved for the event of not adjoining anything to such a node at all.
 
 $$ P_{adj}(\adj(\tau, lex_\tau, \textit{None},n)) +$$
-$$ \sum\limits_{\beta \in A}\sum\limits_{lex \textit{ of }\beta} P_{adj}(\adj(\tau,lex_\tau, \alpha, lex_\alpha, n)) = 1,$$ for each $\tau \in I \cup A$, each lexicalization $lex_\tau$  of $\tau$ and for each node $n$ that is open for adjunction in $\tau$.
+$$ \sum\limits_{\beta \in A}\sum\limits_{w_\beta} P_{adj}(\adj(\tau, w, \tau, v, n)) = 1.$$
+
+{\sc Ptag} assumes that all events occur independent of each other. Therefore it defines the total probability for a derivation as the product of the probability of its individual events.
 
@@
 \label{ssec:pcrisp-domains}
-Here, we reformulate the {\sc crisp} planning operators described in section \ref{ssec:crispdomain}.
+Using the definition of {\sc ptag}, we now reformulate the {\sc crisp} planning operators described in section \ref{ssec:crispdomain}.
 The independence assumption in {\sc ptag} allows us to maintain that planning operator add a single tree, but are now assigned a probability score.
-However, while {\sc crisp} planning operators can add an elementary tree to any site of the right category, {\sc ptag} events are binary events between lexicalized trees at a specific node.
+However, while {\sc crisp} planning operators can add an elementary tree to any site of the right category, {\sc ptag} substitution and adjunction events are binary events between lexicalized trees at a specific node.
 We therefore adapt the literals that record open substitution and adjunction sites in partial derivations accordingly and create one operator for each node in each possible combination of lexicalized trees.
-Figure \ref{example-action} shows an example planning operator. We adopt this scheme for all planning operators regardless of the probability model.
+Figure \ref{example-action} shows an example planning operator for each type.
 \begin{figure}[t]
 \begin{center}
-\cplanaction{\bf subst-t3-cat-t28-eats-n2(u,~x1)}{referent(u,~x1),\\ subst(t28-eats,n2,~u), cat(x1)}{
-$\lnot$needtoexpress(pred-cat,~x1),\\ $\lnot$subst(t-28-eats, n2,~u),\\
-adj(t3-cat, n0, u), adj(t3-cat, n1 u)}{4.3012}\\\smallskip
+\cplanaction{\bf subst-t3-cat-t28-eats-n1(u,~x1)}{referent(u,~x1),\\ subst(t28-eats,~n1,~u), cat(x1)}{
+$\lnot$needtoexpress(pred-cat,~x1),\\ $\lnot$subst(t-28-eats, n1,~u),\\
+adj(t3-cat, n2 u)}{4.3012}\\\smallskip
+
+\cplanaction{\bf adj-t5-raw-t3-fish-n2(u,~x1)}{referent(u,~x1),\\ adj(t28-eats,n2,~u), raw(x1)}{
+$\lnot$needtoexpress(pred-raw,~x1),\\ $\lnot$adj(t-28-eats, n2,~u)}{6.9076}\\\smallskip
+
+\cplanaction{\bf init-t28-eats(u,~x1,~x2,~x3)}{referent(u,~x1),\\ eats(x1,~x2,~x3)}{
+$\lnot$needtoexpress(pred-eats,~x1,~x2,~x3), \\
+ subst(t-28-eats,~n1,subj), subst(t28-eats,~n4,obj),\\
+adj(t-28-eats, n2, u), adj(t-28-eats, n3, u)}{8.5172}\\\smallskip
+
+\cplanaction{\bf noadj-t28-eats-n3(u)}{
+ adj(t-28-eats,~n3,u)}
+{$\lnot$ adj(t-28-eats, n3, u)}{0.1054}\\\smallskip
+
 \caption{\label{example-action} {\sc pcrisp} Operator to substitute $t3$ lexicalized with 'cat' for node 2 of $t28$ lexicalized with `eats'.}

### 2010-05-20 trunk/Papers/pcrisp-10/introduction.tex
intro

@@
 \section{Introduction}
-Traditional sentence generation systems adopt a pipeline architecture that includes sentence planning and surface realization as two independent modules \cite{reiterdale2000}.
-A different view is advocated by \newcite{stonedoran1997}.
-Their sentence generation system {\sc spud} allows a combined treatment of sentence planning and surface realization.
-{\sc spud} exploits {\sc ltag}'s localization of argument structure and uses a top-down construction algorithm to incrementally builds both syntactic and flat semantic structure that satisfies the communicative goal and is consistent with semantic and pragmatic constraints.
-This unified approach to sentence generation is attractive because it tackles different phenomena in generation in an elegant way using a unified, linguistically well motivated framework.
 
-While {\sc spud} uses an incomplete heuristic greedy search, to solve the {\sc NP-}complete problem of finding such a derivation \cite{kollerstriegnitz2002}, \newcite{kollerstone2007} present an encoding of this problem into an AI planning problem.
+Sentence generation systems are typically organized in a pipeline
+architecture, in which the semantic input representation is first
+enriched by such things as referring expressions by a \emph{sentence
+  planner} and only then transformed into natural language strings by
+a \emph{surface realizer} \cite{reiterdale2000}.  An alternative
+approach is \emph{integrated} sentence generation, in which both steps
+are performed by the same algorithm, as in the SPUD system
+\cite{Stone2003a}.  An integrated algorithm can sometimes generate
+better and more succinct sentences \cite{stone98textual}.  SPUD itself
+gives up some of this advantage by using a greedy search heuristic
+for efficiency reasons.  The CRISP system, a recent reimplementation
+of SPUD using search techniques from AI planning, achieves high
+efficiency without sacrificing complete search
+\cite{kollerstone2007,KolHof10}.
 
-We expect that using larger grammars with this approach leads to problems in both performance and output quality, due to over-generation. 
-Our work approaches this problem by combining integrated sentence generation as planning (following \newcite{kollerstone2007}), numeric AI planning and a probability model ({\sc pltag}, \newcite{resnik1992}).
+While CRISP is efficient enough to perform well on large-scale
+grammars \cite{kollerpetrick2009}, such grammars tend to offer many
+different ways to express them same semantic representation.  This
+makes it necessary for the generation system to be able to compute not
+just grammatical sentences, but to identify which of these sentences
+are the \emph{best}. This problem is exacerbated when using
+treebank-derived grammars, which tend to underspecify the actual
+constraints on grammaticality and instead rely on statistical
+information learned from the treebank.  Indeed, there have been a
+number of systems for statistical generation, which can exploit such
+information to rank sentences appropriately
+\cite{langkildeknight1998,whitebaldridge2003,belz2008}.  However,
+these systems are restricted to performing surface realization.
 
-{\bf Structure of the paper:}
-In section \ref{sec:related} we discuss related work.
-Section \ref{sec:crisp} reviews the original generation as planning approach ({\sc crisp}) suggested by \newcite{kollerstone2007}.
-We describe our statistical extension of this approach ({\sc pcrisp}) in section \ref{sec:pcrisp} and evaluate our new system in section \ref{sec:experiments}.
-Finally we conclude this paper in section \ref{sec:conclusion}.
+In this paper, we bring these two strands of research together for the
+first time.  We present the PCRISP system, which redefines the SPUD
+generation problem in terms of probabilistic TAG grammars (PTAG,
+\cite{resnik1992}) and then extends CRISP to solving the probabilistic
+SPUD generation problem using metric planning \cite{hoffmann2003}.  We
+evaluate PCRISP on a PTAG treebank extracted from the Wall Street
+Journal Corpus \cite{chenschanker2004}.  While the evaluation results
+are not competitive in terms of efficiency and suffer from data
+sparseness problems, we still believe our system is a first step
+towards probabilistic integrated generation.
+
+\paragraph{Plan of the paper.} We start by putting our research in the
+context of related work in Section~\ref{sec:related} and reviewing
+CRISP in Section~\ref{sec:crisp}.  We then describe PCRISP, our
+probabilistic extension of CRISP, in Section~\ref{sec:pcrisp} and
+evaluate it in Section~\ref{sec:experiments}. We conclude in
+Section~\ref{sec:conclusion}.
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "pcrisp-10"
+%%% End:

### 2010-05-20 trunk/Papers/pcrisp-10/abstract.tex

\begin{abstract}
We present PCRISP, a sentence generation system for probabilistic TAG
grammars which performs sentence planning and surface realization in
an integrated fashion, in the style of the SPUD system.  PCRISP
operates by converting the generation problem into a metric planning
problem and solving it using an off-the-shelf planner.  We evaluate
PCRISP on the WSJ corpus and identify trade-offs between coverage,
efficiency, and accuracy.
\end{abstract}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "pcrisp-10"
%%% End:

### 2010-05-20 trunk/Papers/pcrisp-10/conclusion.tex

@@
 
-We describe an approach to sentence generation based on probabilistic tree adjoining
-grammar and AI planning. Following previous work by \newcite{stonedoran1997,kollerstone2007}, we treat sentence 
-generation as an integrated task comprising sentence planning and surface realization using declarative descriptions of the syntactic
-and semantic contribution of each word and contextual constraints on its use. Our system uses AI planning technology to find the best {\sc tag} 
-derivation that fulfills a given communicative goal, while satisfying these constraints. 
-Our experiments show that our generation system based on {\sc pltag} and planning makes it possible to use an integrated approach to sentence generation with treebank induced grammars for the first time.
-The results also highlight the trade-off between output quality, efficiency and robustness.
-Future work work should optimize the smoothing model to boost output quality.
-Improved and/or tailored heuristics could help to improve runtime behavior.
-Finally it would be interesting to evaluate the system in an extrinsic way that also addresses sentence planning.
+We have described PCRISP, an approach to integrated sentence generation which can compute the best derivation according to a probabilistic TAG grammar.
+This brings two strands of research -- statistical generation and integrated sentence planning and realization -- together for the first time.
+Our generation algorithm operates by converting the generation problem into a metric planning problem and solving it with an off-the-shelf planner.
+An evaluation on the WSJ corpus revealed that while PCRISP is excellent at computing the preferred derivation on seen data, it is still seriously hampered by data sparseness issues on unseen data.
+Because the size of the planning problem is quadratic in the number of lexicalized trees in the grammar, current planning algorithms are also too slow to be used for longer sentences.
+
+An obvious issue for future research would be to apply improved smoothing techniques to deal with the data sparseness.
+Planning runtimes should be improved by further tweaking the exact planning problems we generate, and will benefit from any future improvements in metric planning.
+It is interesting to note that the extensions we made to CRISP to accommodate statistical generation here are compatible with recent work in which CRISP is applied to situated generation \cite{garoufikoller2010}; we expect that this will be true for other future extensions to CRISP as well.
+Finally, we have only evaluated PCRISP on a surface realization problem in this paper.
+It would be interesting to carry out an extrinsic, task-based evaluation of PCRISP that also addresses sentence planning.
+
+\paragraph{Acknowledgments.} We are grateful to Owen Rambow for providing us with the Chen WSJ-TAG corpus and to Malte Helmert and Silvia Richter for their help with running LAMA, another metric planner with which we experimented.
+We thank Owen and Konstantina Garoufi for helpful discussions, and our reviewers for their insightful comments.
+
+%%% Local Variables:  
+%%% mode: latex 
+%%% TeX-master: "pcrisp-10" 
+%%% End:

### 2010-05-20 trunk/Papers/pcrisp-10/introduction.tex
Related Work

@@
 grammars \cite{kollerpetrick2009}, such grammars tend to offer many
-different ways to express them same semantic representation.  This
+different ways to express the same semantic representation.  This
 makes it necessary for the generation system to be able to compute not
@@
 \cite{resnik1992}) and then extends CRISP to solving the probabilistic
-SPUD generation problem using metric planning \cite{hoffmann2003}.  We
+SPUD generation problem using metric planning \cite{fox2002,hoffmann2003}.  We
 evaluate PCRISP on a PTAG treebank extracted from the Wall Street

### 2010-05-20 trunk/Papers/pcrisp-10/related.tex
Typos.
Related Work

@@
 \label{sec:related}
+Statistical methods are extremely popular for surface realization, but have not been used in systems that integrate sentence planning.
+Most statistical generation approaches follow a generate-and-select strategy, first proposed by \newcite{knight1995} in their {\sc nitrogen} system.
+Such systems generate a set of candidate sentences using a conventional (non-statistical) grammar and then select the best output sentence by applying a statistical language model.
+This family includes systems such as {\sc HAL}ogen \cite{langkildeknight1998,langkilde2000} and open{\sc CCG} \cite{whitebaldridge2003}.
+The {\sc fergus} system by \newcite{bangalorerambow2000} is a variant of this approach that, like {\sc pcrisp}, employs {\sc ltag}.
+It first assigns elementary trees to each entity in the input sentence plan using a statistical tree model and then computes the most likely derivation using using only these trees with an $n$-gram model on the output sentence.
+An alternative to the generate and select approach is to use a probabilistic grammar model, like {\sc ptag}, trained on automatic parses \cite{zhongstent2005}.
+A related approach uses a model over local decisions of the generation system itself \cite{belz2008}.
+Both model can either be used to discriminate a set of output candidates, or more directly to choose the next best decision locally.
+Our approach is similar, because it uses {\sc ptag} to find the most likely output structure.
+However, the previous work discussed so far addresses surface realization only.
+The generate and select approach is even incompatible with the integrated treatment of sentence planning and realization, because it discriminates between possible linearization of the same lexical material.
+In contrast, {\sc pcrisp} addresses sentence planning as well.
 
-Previous statistical NLG systems (e.g. \newcite{knight1995}, {\sc nitrogen} \cite{langkildeknight1998}, {\sc openCCG} \cite{whitebaldridge2003}
-and the {\sc ltag} based realizer {\sc fergus} \cite{bangalorerambow2000}), follow a generate-and-select strategy using n-gram models over output structures
-for discrimination. But since this approach focuses on discriminating realization output for the same sentence plan only, it is incompatible with the
-integrated treatment of sentence planning and realization.
-
-One of the few systems uses generative models for sentence generation is {\sc pCru} \newcite{belz2008}.
-However, her model is defined on possible states of the generation system itself, not on the underlying grammar formalism.
-
-$\cdots$
+Our treatment of integrated sentence planning and surface realization as planning, is inherited directly from {\sc crisp} \cite{kollerstone2007}.
+Planning has long played a role in generation, but has focused on discourse planning instead of specifically addressing sentence generation \cite{hovy1988,appelt1992}.
+The applicability of these ideas was limited at that time because efficient planning technology was not available.
+Recently the development of more efficient planning algorithms (e.g.
+\newcite{hoffmannnebel2001,richterwestphal2008}) spawned a renewed interest in planning for NLG.
+{\sc Crisp} uses such efficient algorithms to efficiently solve the sentence generation problem defined by {\sc spud} \cite{stonedoran1997}.
+{\sc Spud}, which instead used an incomplete greedy algorithm, is based on an {\sc ltag} whose trees are augmented with semantic and pragmatic constraints.
+Given a communicative goal, a solution to the {\sc spud} problem realizes this goal and simultaneosly selects refering expressions.
+The next section explores this idea in more detail.

### 2010-05-20 trunk/Papers/pcrisp-10/crisp.tex
Spell checked

@@
 Given a knowledge base, a communicative goal and a grammar, we require to find a grammatical {\sc tag} derivation that is consistent with this knowledge base and satisfies a communicative goal.
-A number of semantic and pragmatic constraints that must be satisfied by the solution can be added, for instance to enforce generation of unambiguous refering expressions.
+A number of semantic and pragmatic constraints that must be satisfied by the solution can be added, for instance to enforce generation of unambiguous referring expressions.
 \newcite{kollerstone2007} describe how to encode this problem into an AI planning problem, that can be solved efficiently by off-the-shelf planners.
@@
 {\it \{cat($i_2$), fish($i_3$), raw($i_3$), fish($i_4$), eats($i_1$,$i_2$,$i_3$)\}} and the grammar in \ref{fig:grammar}(a), {\sc crisp} could produce the derivation in \ref{fig:grammar}(b).
-Notice that {\sc crisp} can generate the unambiguos refering expression `the raw fish' for $i_3$, to distinguish it from $i_4$.
-For details on  refering expression generation in {\sc crisp} see \newcite{garoufikoller2010}.
+Notice that {\sc crisp} can generate the unambiguous referring expression `the raw fish' for $i_3$, to distinguish it from $i_4$.
+For details on  referring expression generation in {\sc crisp} see \newcite{garoufikoller2010}.
 
@@
 In the operator effect, open substitution nodes are closed and new identifiers are created for each substitution node and internal node in the new tree.
-Given the knowlede base from above, a plan corresponding to the derivation in figure \ref{fig:grammar} would be $\langle${\it S-eats-1(root, i1, i2, i3), NP-cat-2(subj-1, i2), NP-fish-3(obj-1, i3), N-raw-4(obj-1, i3)}$\rangle$
+Given the knowledge base from above, a plan corresponding to the derivation in figure \ref{fig:grammar} would be $\langle${\it S-eats-1(root, i1, i2, i3), NP-cat-2(subj-1, i2), NP-fish-3(obj-1, i3), N-raw-4(obj-1, i3)}$\rangle$
 
@@
 %Such trees contain many internal nodes that are open for adjunction but will not necessarily be filled in a derivation for a given input communicative goal.
-Attaching to the wrong node can produce invalid or disprefered word order.
+Attaching to the wrong node can produce invalid or dispreferred word order.
 due to the high lexical ambiguity of possible, structurally different trees (supertags) for a given word.
@@
 \includegraphics[width=.3\textwidth]{figures/overgen.pdf}
-\caption{\label{fig:overgen} Two derivations with a large grammar, that satisfy the same communicative goal. Sentence (b) is disprefered by most readers.}
+\caption{\label{fig:overgen} Two derivations with a large grammar, that satisfy the same communicative goal. Sentence (b) is dispreferred by most readers.}
 \end{center}

### 2010-05-20 trunk/Papers/pcrisp-10/experiments.tex
Spell checked

@@
 We are interested in the output quality and runtime behavior of different combinations of planning systems and probability models.
-Like {\sc crisp}, {\sc pcrisp} is a sentence generation system and can for instance generate refering expressions.
+Like {\sc crisp}, {\sc pcrisp} is a sentence generation system and can for instance generate referring expressions.
 Unfortunately, since we do not have an evaluation dataset for full sentence generation, we only evaluate the realization abilities of our system by comparing generated sentences against a gold standard testset.
@@
 \label{ssec:eval-measures}
-To evaluate the ouput quality of our statistical generation system we compare the system output $O$ against the reference sentence $R$in the treebank, from which the input representation was generated.
+To evaluate the output quality of our statistical generation system we compare the system output $O$ against the reference sentence $R$in the treebank, from which the input representation was generated.
 We adopt the {\bf generation tree accuracy} measure discussed by \cite{bangalorerambowwhittaker2001}.
@@
 %\newcite{belzreiter2006} empirically investigate the usefulness of {\sc bleu} and {\sc nist} for evaluating NLG systems.
-They find that both metrics correlate with human judgements of output quality on a corpus level, but require a large number of different references sentences to be reliable.
+They find that both metrics correlate with human judgments of output quality on a corpus level, but require a large number of different references sentences to be reliable.
 Multiple references are necessary because {\sc blue} favors sentences with the same word order as the reference.
@@
 % Notice that measures used in statistical parsing, e.g.
-bracketing and labelling accuracy, cannot be used to evaluate generation output because they can only measure similarity between two different syntactic structures for the same surface string.
+bracketing and labeling accuracy, cannot be used to evaluate generation output because they can only measure similarity between two different syntactic structures for the same surface string.
 
@@
 The greedy {\sc spud} heuristic does not generate a single correct sentence, as the combinatorics of the planning domain becomes too large with the treebank induced grammar.
-With the standard {\sc pltag} definition our system manages to generate only very few sentences due to data sparseness.
+With the standard {\sc ptag} definition our system manages to generate only very few sentences due to data sparseness.
 However these sentences are almost identical to the ones found in the test set.
@@
      FF         &  0.66 & 88\% & 0\% & 12\% \\ \hline 
-    \multicolumn{5}{c}{\bf planning with {\sc pltag}}\\\hline
+    \multicolumn{5}{c}{\bf planning with {\sc ptag}}\\\hline
     {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
@@
      FF         &  0.99 & 71\% & 0\% &  29\% \\ \hline
-    \multicolumn{5}{c}{\bf planning with {\sc pltag} (unlexicalized)}\\\hline
+    \multicolumn{5}{c}{\bf planning with {\sc ptag} (unlexicalized)}\\\hline
     {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
@@
      FF & 0.76 & 62\% & 16\% &  22\%\\ \hline
-    \multicolumn{5}{c}{\bf planning with {\sc pltag} (smoothed)}\\\hline
+    \multicolumn{5}{c}{\bf planning with {\sc ptag} (smoothed)}\\\hline
     {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 

### 2010-05-20 trunk/Papers/pcrisp-10/pcrisp.tex
Related Work
Spell checked

@@
 The basic idea is to add a statistical grammar model while leaving the sentence generation mechanism untouched.
-This way we can select the highest scoring derivation, that satisfies all constraints (grammaticality, expresses the communicative goal, use unambiguous refering expressions etc.).
+This way we can select the highest scoring derivation, that satisfies all constraints (grammaticality, expresses the communicative goal, use unambiguous referring expressions etc.).
 
@@
  {\sc Ptag} is lexicalized and therefore does not only assign probabilities to operations in the grammar (as for example plain {\sc pcfg}), but also accounts for binary dependencies between words.
-Unlike n-gram models however, these co-occurences are structured according to local syntactic context as a result of {\sc tag}'s extended domain of locality.
+Unlike n-gram models however, these co-occurrences are structured according to local syntactic context as a result of {\sc tag}'s extended domain of locality.
 The probability model describes how the syntactic arguments of a word are typically filled.
@@
 
-$$ P_{adj}(\adj(\tau, lex_\tau, \textit{None},n)) +$$
-$$ \sum\limits_{\beta \in A}\sum\limits_{w_\beta} P_{adj}(\adj(\tau, w, \tau, v, n)) = 1.$$
+$$ P_{adj}(\adj(\tau, v, \textit{None}, n)) +$$
+$$ \sum\limits_{\beta \in A}\sum\limits_{w_\beta} P_{adj}(\adj(\tau, v, \beta, w, n)) = 1.$$
 
@@
 \cplanaction{\bf subst-t3-cat-t28-eats-n1(u,~x1)}{referent(u,~x1),\\ subst(t28-eats,~n1,~u), cat(x1)}{
-$\lnot$needtoexpress(pred-cat,~x1),\\ $\lnot$subst(t-28-eats, n1,~u),\\
+$\lnot$needtoexpr(pred-cat,~x1),\\ $\lnot$subst(t-28-eats, n1,~u),\\
 adj(t3-cat, n2 u)}{4.3012}\\\smallskip
@@
 \cplanaction{\bf adj-t5-raw-t3-fish-n2(u,~x1)}{referent(u,~x1),\\ adj(t28-eats,n2,~u), raw(x1)}{
-$\lnot$needtoexpress(pred-raw,~x1),\\ $\lnot$adj(t-28-eats, n2,~u)}{6.9076}\\\smallskip
+$\lnot$needtoexpr(pred-raw,~x1),\\ $\lnot$adj(t-28-eats, n2,~u)}{6.9076}\\\smallskip
 
 \cplanaction{\bf init-t28-eats(u,~x1,~x2,~x3)}{referent(u,~x1),\\ eats(x1,~x2,~x3)}{
-$\lnot$needtoexpress(pred-eats,~x1,~x2,~x3), \\
+$\lnot$needtoexpr(pred-eats,~x1,~x2,~x3), \\
  subst(t-28-eats,~n1,subj), subst(t28-eats,~n4,obj),\\
@@
 
-  A straightforward way to deal with data sparseness is to drop all lexicalization from event definitions, as illustrated in figure \ref{fig:modelillustration}, A.
+  A straightforward way to deal with data sparseness is to drop all lexicalizations from event definitions, as illustrated in figure \ref{fig:modelillustration}, A.
 Unfortunately this model does no longer account for bilexical dependencies between words.
-Since our system has to add a lexicalized tree in each step, lexicalization for this child tree should always be taken into account by the probability model, if available.
+Since our system has to add a lexicalized tree in each step, lexicalizations for this child tree should always be taken into account by the probability model, if available.
 Despite these drawbacks, we perform experiments with the unlexicalized model as a baseline.
@@
 An alternative model computes linear interpolation between three back-off levels.
-The first level is just standard {\sc ptag} (figure \ref{fig:modelillustration}, B.1), for the second level the lexicalization of the parent tree is dropped (figure \ref{fig:modelillustration}, B.2), for the third level the model describes only the distribution of lexicalized child trees over each category (figure \ref{fig:modelillustration}, B.3).
+The first level is just standard {\sc ptag} (figure \ref{fig:modelillustration}, B.1), for the second level the lexicalizations of the parent tree is dropped (figure \ref{fig:modelillustration}, B.2), for the third level the model describes only the distribution of lexicalized child trees over each category (figure \ref{fig:modelillustration}, B.3).
 Notice also, that the third level is similar in spirit to a probabilistic version of original {\sc crisp} operators (compare figure \ref{fig:modelillustration}, B.3 and figure \ref{fig:crisp-operators}.

### 2010-05-21 trunk/Papers/pcrisp-10/experiments.tex
experiment section

@@
 
-We estimate our probability models on a training portion of the dataset (96\%) using maximum likelihood estimation.
-For the testing set we selected from the remaining 4\% all sentences that were shorter than 16 words (416 sentences). For longer sentences the search time often exceeded 5 minutes.
+We estimate our probability models on section 1-23 of the converted WSJ using maximum likelihood estimation.
+We use section 0 as a testing set.
+However, since the number of {\sc pcrisp} operators grows quadratically, generating long sentences requires too much time to run batch experiments.
+We therefore restricted our evaluation to the 416 sentences shorter than 16 words.
 
 %{\bf Creating realization problems:} 
-For this testing set we compute semantic representation for each sentence, by traversing its derivation tree top-down and collects a conjunction of terms, by instantiating the automatically created semantic representations.
-Our algorithm introduces a new individual for each substitution node marked with role $r$ and equates it to the `self' role of any tree that adjoins or substitutes into a node marked with $r$.
+For this testing set we automatically create semantic representation for each sentence, by instantiating the semantic content of each elementary used in its derivation.
+We use these representations as input for our system and compare the system output against the original sentence.
 
@@
 \label{ssec:eval-measures}
-To evaluate the output quality of our statistical generation system we compare the system output $O$ against the reference sentence $R$in the treebank, from which the input representation was generated.
+To evaluate the output quality of our statistical generation system we compare the system output $O$ against the reference sentence $R$ in the treebank, from which the input representation was generated.
 We adopt the {\bf generation tree accuracy} measure discussed by \cite{bangalorerambowwhittaker2001}.
@@
 A 'treelet' is a subtree, consisting only of a node and its direct children.
-For each treelet we calculate edit distance, sum the distances over all treelets and then divide by the total length of the reference string.
+For each treelet we calculate edit distance, sum the distances over all treelets and then divide by the total length of the reference string:
 $$1 - \frac{\sum\limits_{d \in treelets(D)} editDist(O|_d, R|_d)}{\textit{length}(R)},$$
@@
 Intuitively errors of the latter type violate the continuity of sub-phrases.
-Because of this property we prefer generation tree accuracy over {\sc blue}, which is overly sensitive to movement errors on an individual sentence level.
+Because of this property we prefer generation tree accuracy over {\sc bleu}, which is overly sensitive to movement errors on an individual sentence level.
 
@@
 \subsection{Planning Systems}
-The planning system we used was a reimplementation of {\sc FF} \cite{hoffmannnebel2001}, a heuristic search planner.
-We extended this planner to take action costs into account when it calculates heuristic values for successor states.
-We also experimented with the greedy search heuristic used by {\sc spud} \cite{stonedoran1997}.
+For our experiments we use a reimplementation of {\sc FF} \cite{hoffmannnebel2001}, a heuristic search planner.
+We extend this planner to take action costs into account when it calculates heuristic values for successor states.
+We also experiment with the greedy heuristic search strategy used by {\sc spud}, which we adapted to problem specific search strategy for the planner.
 
@@
 Table \ref{results} presents our results.
+In each experiment we count how many sentences in the test set produce an output at all (column `generated') and for how many the planner requires more than 5 minutes (column `timeout'.
+The search is terminated in this case).
+Some sentences cannot be realized at all (column `fail'), due to data sparseness problems.
+In such cases either no elementary tree with the required lexical content was seen in the training data, or the model assigns a zero probability to a required adjunction or substitution operation.
+We compute the average generation tree accuracy (column `acc') over all generated sentences.
 
-The first two columns confirm our hypothesis that {\sc crisp} does not scale up to the automatically induced treebank.
-The greedy {\sc spud} heuristic does not generate a single correct sentence, as the combinatorics of the planning domain becomes too large with the treebank induced grammar.
+The first two rows show results for non-statistical {\sc crisp} and confirm our hypothesis that this problem formulation does not scale up to the automatically induced treebank.
+The greedy search with {\sc spud} heuristics does not generate a single correct sentence, as the combinatorics of the planning domain becomes too large with our grammar.
 With the standard {\sc ptag} definition our system manages to generate only very few sentences due to data sparseness.
-However these sentences are almost identical to the ones found in the test set.
-Both generalized models ease the data sparseness problem, but output quality drops rapidly.
-For the unlexicalized model the search problem becomes very hard due to the large planning domain, so that 22\% of the generation problems require more than 5 minutes runtime.
-As expected the smoothed model offers a compromise between efficiency, output quality and coverage
+However these sentences are very similar to the ones found in the test set.
+Both generalized models ease the data sparseness problem, but output quality drops.
+For the unlexicalized model the search problem becomes very hard due to the large planning domain, so that 22\% of the generation problems time out.
+As expected the smoothed model provides better coverage than {\sc ptag}, while producing higher output quality than the unlexicalized model.
 \footnote{In our experiments we set the weight between level 1 and 2 to 0.9 and did not consider level 3, after empirically testing different settings.
@@
 \begin{table}
-    \caption{\label{results} Evaluation results.}
+    \caption{\label{results} Results for the realization experiment. Planners: `greedy'= adapted {\sc spud} strategy, `FF' = reimplementation of {\sc ff}, `FF + costs' = cost-sensitive heuristics } 
     \begin{center}
-    \begin{tabular}{|l|l|r|r|r|}
+    \begin{tabular}{|l||l||r|r|r|}
     \multicolumn{5}{c}{\sc crisp}\\ 
@@
      greedy         & ??? & 0\% & 100\% & 0\%  \\ \hline 
-     FF         &  0.66 & 88\% & 0\% & 12\% \\ \hline 
+     FF         &  0.66 & 45\% & 42\% & 13\% \\ \hline 
     \multicolumn{5}{c}{\bf planning with {\sc ptag}}\\\hline
     {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
-     FF + costs &  0.99 &  97\% & 0\% & 3\% \\ \hline
-     FF         &  0.99 & 71\% & 0\% &  29\% \\ \hline
-    \multicolumn{5}{c}{\bf planning with {\sc ptag} (unlexicalized)}\\\hline
+     FF + costs &  0.90 &  10\% &88\% & 2\% \\ \hline
+     FF         &  0.87 & 7\% & 67\% &  26\% \\ \hline
+    \multicolumn{5}{c}{\bf planning with the unlexicalized model}\\\hline
     {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
@@
      FF & 0.76 & 62\% & 16\% &  22\%\\ \hline
-    \multicolumn{5}{c}{\bf planning with {\sc ptag} (smoothed)}\\\hline
+    \multicolumn{5}{c}{\bf planning with linear interpolation}\\\hline
     {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
-     FF + costs & 0.78 & 19\% & 74\% & 7\% \\ \hline
-     FF & 0.77 & 19\% & 74\% & 7\% \\ \hline
+     FF + costs & 0.88 & 19\% & 74\% & 7\% \\ \hline
+     FF & 0.78 & 19\% & 74\% & 7\% \\ \hline
     \end{tabular}

### 2010-05-21 trunk/Papers/pcrisp-10/pcrisp.tex
experiment section

@@
 
-Finally, we set the cost of an operator to be its negative log probability.
-For example for the operator in figure \ref{example-action} $$Cost( \subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2) ) =$$
-$$ -\log( P(\subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2)) ). $$ This way the plan with the smallest cost sum corresponds to the {\sc tag} derivation with the highest probability.
+Finally, we set the cost of an operator to be its negative log probability. For example 
+\[
+Cost( \subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2) ) =
+\]
+\[ 
+ -\log( P(\subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2)) ). 
+\] 
+This way the plan with the smallest cost sum corresponds to the {\sc tag} derivation with the highest probability.
 

### 2010-05-21 trunk/Papers/pcrisp-10/related.tex
Spell checked
experiment section

@@
 This family includes systems such as {\sc HAL}ogen \cite{langkildeknight1998,langkilde2000} and open{\sc CCG} \cite{whitebaldridge2003}.
-The {\sc fergus} system by \newcite{bangalorerambow2000} is a variant of this approach that, like {\sc pcrisp}, employs {\sc ltag}.
+The {\sc Fergus} system \cite{bangalorerambow2000} is a variant of this approach that, like {\sc pcrisp}, employs {\sc ltag}.
 It first assigns elementary trees to each entity in the input sentence plan using a statistical tree model and then computes the most likely derivation using using only these trees with an $n$-gram model on the output sentence.
@@
 {\sc Spud}, which instead used an incomplete greedy algorithm, is based on an {\sc ltag} whose trees are augmented with semantic and pragmatic constraints.
-Given a communicative goal, a solution to the {\sc spud} problem realizes this goal and simultaneosly selects refering expressions.
+Given a communicative goal, a solution to the {\sc spud} problem realizes this goal and simultaneously selects referring expressions.
 The next section explores this idea in more detail.

### 2010-05-21 trunk/Papers/pcrisp-10/experiments.tex

@@
 Table \ref{results} presents our results.
-In each experiment we count how many sentences in the test set produce an output at all (column `generated') and for how many the planner requires more than 5 minutes (column `timeout'.
+In each experiment we count how many sentences in the test set produce an output at all (column `success') and for how many the planner requires more than 5 minutes (column `timeout'.
 The search is terminated in this case).

### 2010-05-21 trunk/Papers/pcrisp-10/related.tex

@@
 The applicability of these ideas was limited at that time because efficient planning technology was not available.
-Recently the development of more efficient planning algorithms (e.g.
-\newcite{hoffmannnebel2001,richterwestphal2008}) spawned a renewed interest in planning for NLG.
-{\sc Crisp} uses such efficient algorithms to efficiently solve the sentence generation problem defined by {\sc spud} \cite{stonedoran1997}.
+Recently the development of more efficient planning algorithms \cite{hoffmannnebel2001,richterwestphal2008} spawned a renewed interest in planning for NLG.
+{\sc Crisp} uses such algorithms to efficiently solve the sentence generation problem defined by {\sc spud} \cite{stonedoran1997}.
 {\sc Spud}, which instead used an incomplete greedy algorithm, is based on an {\sc ltag} whose trees are augmented with semantic and pragmatic constraints.
 Given a communicative goal, a solution to the {\sc spud} problem realizes this goal and simultaneously selects referring expressions.
-The next section explores this idea in more detail.
+The next section explores {\sc crisp} in more detail.

### 2010-05-21 trunk/Papers/pcrisp-10/conclusion.tex
einmal ganz drueber

@@
 Our generation algorithm operates by converting the generation problem into a metric planning problem and solving it with an off-the-shelf planner.
-An evaluation on the WSJ corpus revealed that while PCRISP is excellent at computing the preferred derivation on seen data, it is still seriously hampered by data sparseness issues on unseen data.
+An evaluation on the WSJ corpus reveals that \pcrisp, like PTAG in general, is susceptible to data sparseness problems.
 Because the size of the planning problem is quadratic in the number of lexicalized trees in the grammar, current planning algorithms are also too slow to be used for longer sentences.
 
-An obvious issue for future research would be to apply improved smoothing techniques to deal with the data sparseness.
+An obvious issue for future research is to apply improved smoothing techniques to deal with the data sparseness.
 Planning runtimes should be improved by further tweaking the exact planning problems we generate, and will benefit from any future improvements in metric planning.
@@
 \paragraph{Acknowledgments.} We are grateful to Owen Rambow for providing us with the Chen WSJ-TAG corpus and to Malte Helmert and Silvia Richter for their help with running LAMA, another metric planner with which we experimented.
-We thank Owen and Konstantina Garoufi for helpful discussions, and our reviewers for their insightful comments.
+We thank Konstantina Garoufi and Owen Rambow for helpful discussions, and our reviewers for their insightful comments.
 

### 2010-05-21 trunk/Papers/pcrisp-10/crisp.tex
einmal ganz drueber

@@
 In this section we review the original non-statistical {\sc crisp} system \cite{kollerstone2007}. 
-Following {\sc spud} \cite{stonedoran1997}, {\sc crisp} is based on a declarative description of the sentence generation problem using {\sc tag}.
+Following {\sc spud} \cite{Stone2003a}, {\sc crisp} is based on a declarative description of the sentence generation problem using {\sc tag}.
 Given a knowledge base, a communicative goal and a grammar, we require to find a grammatical {\sc tag} derivation that is consistent with this knowledge base and satisfies a communicative goal.
 A number of semantic and pragmatic constraints that must be satisfied by the solution can be added, for instance to enforce generation of unambiguous referring expressions.
-\newcite{kollerstone2007} describe how to encode this problem into an AI planning problem, that can be solved efficiently by off-the-shelf planners.
+\newcite{kollerstone2007} describe how to encode this problem into an AI planning problem which can be solved efficiently by off-the-shelf planners.
 We describe the general mechanism in the following section and then review the encoding into planning in section \ref{ssec:crispdomain}.
@@
 All nodes that dominate the lexical anchor are assigned the role `self', which intuitively corresponds to the event or individual described by this tree.
-Figure \ref{fig:grammar}(a) shows an example grammar of this type.
+Fig.~\ref{fig:grammar}(a) shows an example grammar of this type.
 
@@
 For each substitution and adjunction, the semantic role associated with the role of the target node is unified with the `self' role of the child tree.
-For example, given the knowledge base
-{\it \{cat($i_2$), fish($i_3$), raw($i_3$), fish($i_4$), eats($i_1$,$i_2$,$i_3$)\}} and the grammar in \ref{fig:grammar}(a), {\sc crisp} could produce the derivation in \ref{fig:grammar}(b).
-Notice that {\sc crisp} can generate the unambiguous referring expression `the raw fish' for $i_3$, to distinguish it from $i_4$.
-For details on  referring expression generation in {\sc crisp} see \newcite{garoufikoller2010}.
+For example, given the knowledge base $\{\f{cat}(c), \f{fish}(f_1), \f{raw}(f_1), \f{fish}(f_2), \f{eats}(e,c,f_1)\}$ and the grammar in \ref{fig:grammar}(a), {\sc crisp} could produce the derivation in \ref{fig:grammar}(b).
+Notice that {\sc crisp} can generate the unambiguous referring expression `the raw fish' for $f_1$, to distinguish it from $f_2$.
 
@@
 A planning problem consist of an initial state, a set of goal states and a set of planning operators that describe possible state transitions.
-A solution to the problem is any sequence of actions (instantiated operators) that leads from the initial state to one of the goal states.
-Planning problems can be solved efficiently by general purpose planning systems such as
-{\sc ff} \cite{hoffmannnebel2001}.  
-% Planning problems can be written in a standard notation ({\sc pddl}\footnote{Planning Domain Description Language}, \newcite{gerevinietal2009}).
+A \emph{plan} is any sequence of actions (instantiated operators) that leads from the initial state to one of the goal states.
+Planning problems can be solved efficiently by general purpose planning systems such as {\sc ff} \cite{hoffmannnebel2001}.
 
-In {\sc crisp} planning states correspond to partial {\sc tag} derivations and record open substitution and adjunction sites, semantic individuals associated with them, and parts of the communicative goal that have not yet been expressed.
+In {\sc crisp}, planning states correspond to partial {\sc tag} derivations and record open substitution and adjunction sites, semantic individuals associated with them, and parts of the communicative goal that have not yet been expressed.
 The initial state also encodes the knowledge base and the communicative goal.
-Each planning operator contributes a new elementary tree to the derivation and at the same time satisfies part of the communicative goal, as described in the previous section.
+Each planning operator contributes a new elementary tree to the derivation and at the same time can satisfy part of the communicative goal, as described in the previous section.
 In a goal state there are no open substitution sites left and all literals in the communicative goal have been expressed.
@@
 
-\planaction{\bf subst-t5-raw-N(u,~x1)}{referent(u,~x1),\\ adj(N,~u), raw(x1)}{
+\planaction{\bf adj-t5-raw-N(u,~x1)}{referent(u,~x1),\\ adj(N,~u), raw(x1)}{
 $\lnot$needtoexpr(pred-raw,~x1),\\ $\lnot$adj(N,~u)}\\\smallskip
 \end{center}
-\caption{\label{fig:crisp-operators} {\sc crisp} operators for some of the elementary trees in figure \ref{fig:grammar}.}
+\caption{\label{fig:crisp-operators} {\sc crisp} operators for some of the elementary trees in Fig.~\ref{fig:grammar}.}
 \end{figure}
 
-Figure \ref{fig:crisp-operators} shows planning operators for part of the grammar in figure \ref{fig:grammar}(a).
+Fig.~\ref{fig:crisp-operators} shows planning operators for part of the grammar in Fig.~\ref{fig:grammar}(a).
+The operators are simplified for lack of space, and in particular we do not show the preconditions and effects that enforce uniqueness of referring expressions; see \newcite{kollerstone2007} for details.
 The preconditions of the operators require that a suitable open substitution node (i.e.
@@
 In the operator effect, open substitution nodes are closed and new identifiers are created for each substitution node and internal node in the new tree.
-Given the knowledge base from above, a plan corresponding to the derivation in figure \ref{fig:grammar} would be $\langle${\it S-eats-1(root, i1, i2, i3), NP-cat-2(subj-1, i2), NP-fish-3(obj-1, i3), N-raw-4(obj-1, i3)}$\rangle$
+Given the knowledge base from above, a plan corresponding to the derivation in Fig.~\ref{fig:grammar} would be $\f{subst-t28-eats-S-eats}(\f{root}, e, c, f_1)$; $\f{subst-t3-cat-NP}(\f{subj},c)$; $\f{subst-t3-fish-NP}(\f{obj},f_1)$; $\f{adj-t5-raw-N}(\f{obj}, f_1)$.
+This plan can be automatically decoded into a derivation tree for Fig.~\ref{fig:grammar}b.
 
 \subsection{CRISP and Large Grammars}
-Using {\sc crisp} with large, treebank induced grammars can lead to overgeneration, 
-%Such trees contain many internal nodes that are open for adjunction but will not necessarily be filled in a derivation for a given input communicative goal.
-Attaching to the wrong node can produce invalid or dispreferred word order.
-due to the high lexical ambiguity of possible, structurally different trees (supertags) for a given word.
-We illustrate this problem in figure \ref{fig:overgen}.
-Assuming a grammar that includes trees for both right adjoining ($t13$) and left adjoining PPs ($t252$), both derivations (a) and (b) are {\it grammatical} derivations that satisfy the same communicative goal.
+
+\newcite{KolHof10} report on experiments that show CRISP can generate sentences with the large-scale XTAG grammar \cite{xtag2001} quite efficiently.
+However, because CRISP has no notion of how ``good'' a generated sentence is compared to other grammatical alternatives, it will sometimes compute dispreferred sentences with large grammars.
+This is especially true for treebank-induced grammars, which tend to overgenerate and rely on statistical methods to rank good sentences highly.
+Fig.~\ref{fig:overgen} illustrates this problem.
+Assuming a (treebank) grammar that includes trees for both right adjoining (t13) and left adjoining PPs (t252), both derivations (a) and (b) are {\it grammatical} derivations that satisfy the same communicative goal.
 However, most readers disprefer the reading in (b).
 Clearly, to use {\sc crisp} with such a grammar we need a method of distinguishing good derivations from bad ones.
-In section \ref{sec:experiments} we provide empirical evidence for this problem and show how it is solved by the approach we describe in the next section.
+
 \begin{figure}
@@
 \end{figure}
+
+%%% Local Variables:  %%% mode: latex %%% TeX-master: "pcrisp-10" %%% End:

### 2010-05-21 trunk/Papers/pcrisp-10/experiments.tex
einmal ganz drueber

--- 2010-05-20 trunk/Papers/pcrisp-10/experiments.tex
@@
-\section{Experiments}
+\section{Evaluation}
 \label{sec:experiments}
-In this section we perform experiments with the statistical generation system described in the previous section.
+
+We now report on some experiments with \pcrisp.
 We are interested in the output quality and runtime behavior of different combinations of planning systems and probability models.
-Like {\sc crisp}, {\sc pcrisp} is a sentence generation system and can for instance generate referring expressions.
-Unfortunately, since we do not have an evaluation dataset for full sentence generation, we only evaluate the realization abilities of our system by comparing generated sentences against a gold standard testset.
-The next section describes our evaluation data and section \ref{ssec:eval-measures} describes the evaluation measures used.
+Like \crisp, \pcrisp\ is an integrated sentence generation system capable of generating referring expressions during surface realization.
+However, for lack of an appropriate evaluation dataset for full sentence generation, we only evaluate \pcrisp\ on a realization task here.
+Further experiments and more details can be found in \newcite{bauer2009}.
 
 \subsection{Evaluation Data}
-For our experiments we use a grammar and treebank that was automatically extracted from the Wall Street Journal using the algorithm described by \newcite{chenschanker2004}.
+For our experiments we use an \ltag\ grammar and treebank that was automatically extracted from the Wall Street Journal using the algorithm described by \newcite{chenschanker2004}.
 
@@
 Models that can deal with multiple adjunction are discussed by \newcite{nasrrambow2006}, but employing them would require non-trivial modifications to our encoding as a metric planning problem.
-We therefore preprocess the treebank by `linearizing' multiple adjunctions.
-Furthermore, we attach prepositions to the trees they substitute in, to increase the expressiveness of the bilexical probability distribution.
-For details on these preprocessing steps see \cite{bauer2009}.
+We therefore preprocess the treebank by linearizing multiple adjunctions.
+Furthermore, we reattach prepositions to the trees they substitute in, to increase the expressiveness of the bilexical probability distribution.
 
-%{\bf Adding semantics to elementary trees:}
-We then automatically create semantic content for all lexicalized elementary trees by assigning a single semantic predicate to each tree in the treebank, using the lexical anchor as a name and setting the arity to the number of substitution nodes in the tree plus one for the `self' role.
-We calculate the role associated with each node in each tree by assigning role-names to each substitution node (`self' is assigned to the lexical anchor) and then percolating the roles up the tree, giving preference to the self role.
+We then automatically create semantic content for all lexicalized elementary trees by assigning a single semantic literal to each tree in the treebank, using the lexical anchor as the predicate symbol and variables for each substitution node and the `self' role as arguments.
+We calculate the role associated with each node in each tree by assigning role names to each substitution node (`self' is assigned to the lexical anchor) and then percolating the roles up the tree, giving preference to the `self' role.
 
-We estimate our probability models on a training portion of the dataset (96\%) using maximum likelihood estimation.
-For the testing set we selected from the remaining 4\% all sentences that were shorter than 16 words (416 sentences). For longer sentences the search time often exceeded 5 minutes.
+We estimate our probability models on section 1 to 23 of the converted WSJ using maximum likelihood estimation.
+We use section 0 as a testing set.
+However, since the number of \pcrisp\ operators grows quadratically with the grammar size, generating long sentences requires too much time to run batch experiments.
+We therefore restricted our evaluation to the 416 sentences in Section 0 that are shorter than 16 words.
 
-%{\bf Creating realization problems:} 
-For this testing set we compute semantic representation for each sentence, by traversing its derivation tree top-down and collects a conjunction of terms, by instantiating the automatically created semantic representations.
-Our algorithm introduces a new individual for each substitution node marked with role $r$ and equates it to the `self' role of any tree that adjoins or substitutes into a node marked with $r$.
+For this testing set we automatically create semantic representation for each sentence, by instantiating the semantic content of each elementary used in its derivation.
+We use these representations as input for our system and compare the system output against the original sentence.
 
-\subsection{Evaluation Measures}
+\subsection{Generation tree accuracy}
 \label{ssec:eval-measures}
-To evaluate the output quality of our statistical generation system we compare the system output $O$ against the reference sentence $R$in the treebank, from which the input representation was generated.
-We adopt the {\bf generation tree accuracy} measure discussed by \cite{bangalorerambowwhittaker2001}.
+To evaluate the output quality of our statistical generation system we compare the system output $O$ against the reference sentence $R$ in the treebank from which the input representation was generated.
+We adopt the {\bf generation tree accuracy} (GTA) measure discussed by \cite{bangalorerambowwhittaker2001}.
 This measure is computed by first creating a list of all 'treelets' from the reference derivation tree $D$.
-A 'treelet' is a subtree, consisting only of a node and its direct children.
-For each treelet we calculate edit distance, sum the distances over all treelets and then divide by the total length of the reference string.
+A `treelet' is a subtree consisting only of a node and its direct children.
+For each treelet we calculate the edit distance, sum the distances over all treelets and then divide by the total length of the reference string:
 $$1 - \frac{\sum\limits_{d \in treelets(D)} editDist(O|_d, R|_d)}{\textit{length}(R)},$$
@@
 Edit distance is modified to treat insertion-deletion pairs as single movement errors.
-This metric account for the observation that errors that change word order on a local level (i.e.
-within a treelet) are less severe than errors that swap words between different treelets.
-Intuitively errors of the latter type violate the continuity of sub-phrases.
-Because of this property we prefer generation tree accuracy over {\sc blue}, which is overly sensitive to movement errors on an individual sentence level.
-
-%\newcite{belzreiter2006} empirically investigate the usefulness of {\sc bleu} and {\sc nist} for evaluating NLG systems.
-They find that both metrics correlate with human judgments of output quality on a corpus level, but require a large number of different references sentences to be reliable.
-Multiple references are necessary because {\sc blue} favors sentences with the same word order as the reference.
-In NLG, more than in Machine Translation, word order of output sentences can vary.
-
-%{\it String Accuracy} is defined as 
-%$$1 - \frac{\textit{editDist(O,R)}}{\textit{length}(R)},$$ where $O$ is the surface system output, $R$ is the reference sentence and $editDist(A,B)$ is the minimal edit distance between the token strings $A$ and $B$ defined as the minimal number of substitutions, insertions and deletions needed to transform $A$ into $B$.
-
-%Like {\sc bleu}, string accuracy often underestimates generation performance if the system output differs from the reference only in word order. 
-%For example if a word occurs in the beginning of the string instead of the end, blocked by a number of correct words, edit distance will count two errors, one for removing the word in the beginning and one for adding it in the end.
-In this case it would be more reasonable to count a single movement error.
-%The revised {\it generation string accuracy} does this in a post-processing step: Each insertion-deletion pair sharing the same token, but in different places is counted as a single movement error. 
-%\newcite{bangalorerambowwhittaker2001}  refine these string based measures by referring to structural information from the reference derivation tree.
-
-%Tree accuracy measure only the quality of the surface string.
-To compare the syntactic structure of the output with the reference we propose {\it derivation tree accuracy}, the ratio of derivation-treelets (nodes with their direct children in their specific order) that occurs in both the system output derivation and the reference derivation.
-%$$ \frac{|treelets(D_O) \cap treelets(D_R)|}{\textit{length}(D_R)}$$
-% Notice that measures used in statistical parsing, e.g.
-bracketing and labeling accuracy, cannot be used to evaluate generation output because they can only measure similarity between two different syntactic structures for the same surface string.
-
-\subsection{Planning Systems}
-The planning system we used was a reimplementation of {\sc FF} \cite{hoffmannnebel2001}, a heuristic search planner.
-We extended this planner to take action costs into account when it calculates heuristic values for successor states.
-We also experimented with the greedy search heuristic used by {\sc spud} \cite{stonedoran1997}.
+Compared to a purely string-based metric like {\sc bleu}, GTA penalizes swapped words less harshly if they can be explained by local tree movements.
 
 \subsection{Results}
-Table \ref{results} presents our results.
 
-The first two columns confirm our hypothesis that {\sc crisp} does not scale up to the automatically induced treebank.
-The greedy {\sc spud} heuristic does not generate a single correct sentence, as the combinatorics of the planning domain becomes too large with the treebank induced grammar.
-With the standard {\sc ptag} definition our system manages to generate only very few sentences due to data sparseness.
-However these sentences are almost identical to the ones found in the test set.
-Both generalized models ease the data sparseness problem, but output quality drops rapidly.
-For the unlexicalized model the search problem becomes very hard due to the large planning domain, so that 22\% of the generation problems require more than 5 minutes runtime.
-As expected the smoothed model offers a compromise between efficiency, output quality and coverage
-\footnote{In our experiments we set the weight between level 1 and 2 to 0.9 and did not consider level 3, after empirically testing different settings.
-We set a probability threshold to include operators into the planning domain to $10^{-4}$}.
+Table \ref{results} presents the results of the experiment for five different generation systems.
+We compare three variants of \pcrisp: the fully lexicalized PTAG model (``PTAG''), the fully unlexicalized model (``unlexicalized''), and a linear interpolation model in which we (manually) set the weight of level 2 to 0.9 \todo{stimmt das?} and the weight of level 3 to 0 (``interpolation'').
+We also list results for the non-statistical \crisp\ system of  \newcite{kollerstone2007} (``\crisp'') and the greedy search heuristic used by \spud.
+All these systems are based on a reimplementation of the FF planner \cite{hoffmannnebel2001}, to which we added a search heuristic that takes action costs into account for the \pcrisp\ systems.
+
+For each system, we determine the proportion of sentences in the test set for which the system produced an output (``success''), did not find a plan (``fail''), and exceeded the five-minute timeout (``timeout'').
+The column ``gta'' records the mean generation tree accuracy for those sentences where the system produced an output.
+
+The table confirms that the non-statistical \crisp\ system has considerable trouble reconstructing the original sentences from the treebank, with a mean GTA of 0.66.
+This is still better than our reimplementation of \spud, which fails to recover from early mistakes of its greedy search heuristic for every single sentence in the test set.
+
+By contrast, the fully lexicalized \pcrisp\ model achieves a much better mean GTA.
+However, this comes at the cost of a very low success rate of only 10\%, reflecting a serious data sparseness problem on unseen inputs.
+The data sparseness problem is reduced in the unlexicalized version of \pcrisp, but this comes at the cost of decreased accuracy and much increased runtimes.
+The linear interpolation model strikes a balance between these two, by improving the success rate over the lexicalized model while sacrificing only a small amount of accuracy.
+This suggests that smoothing is a promising approach to balancing coverage, efficiency, and accuracy, but clearly further experimentation is needed to substantiate this.
 
 \begin{table}
-    \caption{\label{results} Evaluation results.}
     \begin{center}
-    \begin{tabular}{|l|l|r|r|r|}
-    \multicolumn{5}{c}{\sc crisp}\\ 
+    \begin{tabular}{|l||l||r|r|r|}
     \hline
-     {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
-     greedy         & ??? & 0\% & 100\% & 0\%  \\ \hline 
-     FF         &  0.66 & 88\% & 0\% & 12\% \\ \hline 
-    \multicolumn{5}{c}{\bf planning with {\sc ptag}}\\\hline
-    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
-     FF + costs &  0.99 &  97\% & 0\% & 3\% \\ \hline
-     FF         &  0.99 & 71\% & 0\% &  29\% \\ \hline
-    \multicolumn{5}{c}{\bf planning with {\sc ptag} (unlexicalized)}\\\hline
-    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
-     FF + costs & 0.74 & 62\% & 16\% & 22\%\\ \hline
-     FF & 0.76 & 62\% & 16\% &  22\%\\ \hline
-    \multicolumn{5}{c}{\bf planning with {\sc ptag} (smoothed)}\\\hline
-    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
-     FF + costs & 0.78 & 19\% & 74\% & 7\% \\ \hline
-     FF & 0.77 & 19\% & 74\% & 7\% \\ \hline
+     {\bf System }  & {\bf gta}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
+     \spud         & n/a & 0\% & 100\% & 0\%  \\ \hline 
+     \crisp         &  0.66 & 45\% & 42\% & 13\% \\ \hline
+     PTAG &  0.90 &  10\% &88\% & 2\% \\ \hline
+     unlexicalized & 0.74 & 62\% & 16\% & 22\%\\ \hline
+     interpolation & 0.88 & 19\% & 74\% & 7\% \\ \hline
     \end{tabular}
     \end{center}
+    \caption{\label{results} Results for the realization experiment.} 
 \end{table}
+
+%%% Local Variables:  %%% mode: latex %%% TeX-master: "pcrisp-10" %%% End:

### 2010-05-21 trunk/Papers/pcrisp-10/introduction.tex
einmal ganz drueber

@@
 
-Sentence generation systems are typically organized in a pipeline
-architecture, in which the semantic input representation is first
-enriched by such things as referring expressions by a \emph{sentence
+Many sentence generation systems are organized in a pipeline
+architecture, in which the input semantic representation is first
+enriched, e.g.\ by referring expressions, by a \emph{sentence
   planner} and only then transformed into natural language strings by
@@
 While CRISP is efficient enough to perform well on large-scale
-grammars \cite{kollerpetrick2009}, such grammars tend to offer many
+grammars \cite{KolHof10}, such grammars tend to offer many
 different ways to express the same semantic representation.  This
@@
 just grammatical sentences, but to identify which of these sentences
-are the \emph{best}. This problem is exacerbated when using
+are \emph{good}. This problem is exacerbated when using
 treebank-derived grammars, which tend to underspecify the actual
@@
 information to rank sentences appropriately
-\cite{langkildeknight1998,whitebaldridge2003,belz2008}.  However,
-these systems are restricted to performing surface realization.
+\cite{langkildeknight1998,whitebaldridge2003,belz2008}.  However, to our 
+knowledge, all such systems are currently restricted to performing surface realization,
+and must rely on separate modules to perform sentence planning.
 
@@
 evaluate PCRISP on a PTAG treebank extracted from the Wall Street
-Journal Corpus \cite{chenschanker2004}.  While the evaluation results
-are not competitive in terms of efficiency and suffer from data
-sparseness problems, we still believe our system is a first step
-towards probabilistic integrated generation.
+Journal Corpus \cite{chenschanker2004}.  The evaluation reveals
+a tradeoff between coverage, efficiency, and accuracy which we think
+are worth exploring further in future work.
 

### 2010-05-21 trunk/Papers/pcrisp-10/pcrisp.tex
einmal ganz drueber

@@
 
-We now extend {\sc crisp} to statistical generation ({\sc pcrisp}).
+We now extend \crisp\ to statistical generation (\pcrisp).
 The basic idea is to add a statistical grammar model while leaving the sentence generation mechanism untouched.
-This way we can select the highest scoring derivation, that satisfies all constraints (grammaticality, expresses the communicative goal, use unambiguous referring expressions etc.).
+This way we can select the highest scoring derivation which satisfies all constraints (grammaticality, expresses the communicative goal, uses unambiguous referring expressions, etc.).
 
@@
 
-While we leave the basic sentence generation mechanism intact, we need to modify the concrete formulation of {\sc crisp} planning operators, as each of them now corresponds to some probability event.
-Likewise, we need to take the step from classical planning to metric planning systems, that can use the probabilities.
-In metric planning \cite{fox2002}, instead of changing only boolean state variables, planning actions can decrease or increase a numeric variable.
+While we leave the basic sentence generation mechanism intact, we need to modify the concrete formulation of \crisp\ planning operators to accommodate bilexical dependencies.
+Likewise, we need to take the step from classical planning to \emph{metric} planning systems which can use the probabilities.
+In metric planning \cite{fox2002}, planning actions can modify the value of numeric variables in addition to adding and deleting logical literals to the state.
 The goal state specifies constraints on this variable.
 In the simplest case the variable can only be increased by a static cost value in each action, and the goal state contains the objective to minimize the total cost.
-Systems such as metric-{\sc FF} \cite{hoffmann2003} do not guarantee optimality, but generally offer good results.
+While systems such as Metric-{\sc FF} \cite{hoffmann2003} do not guarantee optimality, they do generally offer good results.
 We address our encoding of sentence generation with {\sc ptag} as metric planning in section \ref{ssec:pcrisp-domains}.
@@
 \label{ssec:probmodels}
-{\sc ptag} \cite{resnik1992} views {\sc tag} derivations as sequences of probability events of three types, initial events, substitution events and adjunction events.
+{\sc ptag} \cite{resnik1992} views {\sc tag} derivations as sequences of events of three types: initial events, substitution events, and adjunction events.
 
-\newcommand{\adj}[0]{\textit{ adj}}
-\newcommand{\init}[0]{\textit{ init}}
-\newcommand{\subst}[0]{\textit{ subst}}
 The probability distribution for initial events describes how likely it is to start any derivation with a given initial tree. 
-It is defined over all initial trees $\alpha$ with their possible lexicalizations $w \in W_\alpha$: 
-$$ \sum\limits_{\alpha \in I}\sum\limits_{w \in W_\alpha} P_{init}(\init(\alpha, w)) = 1 $$
+It is defined over all initial trees $\alpha \in I$ with their possible lexicalizations $w \in W_\alpha$: 
+$$ \sum\limits_{\alpha \in I}\sum\limits_{w \in W_\alpha} P_i(\init(\alpha, w)) = 1 $$
 
-For substitution events, there is a probability distribution for each substitution node $n$ of each initial tree $\tau$ lexicalized with $v$, which describes how likely it is to substitute it with an initial tree $\alpha$ lexicalized with $w$.
-$$\sum\limits_{\alpha \in I}\sum\limits_{w \in W_\alpha} P_{subst}(\subst(\tau,v , \alpha, w, n)) = 1$$
+For substitution events, there is a probability distribution for each substitution node $n$ of each elementary tree $\tau$ lexicalized with $v$, which describes how likely it is to substitute it with an initial tree $\alpha$ lexicalized with $w$.
+$$\sum\limits_{\alpha \in I}\sum\limits_{w \in W_\alpha} P_s(\subst(\tau,v , \alpha, w, n)) = 1$$
 
-Similarly for each internal node there is a distribution that describes the probability to adjoin each auxiliary tree $\beta$ lexicalized with $w$.
+Similarly for each internal node there is a distribution that describes the probability to adjoin an auxiliary tree $\beta \in A$ lexicalized with $w$.
 In addition some probability mass is reserved for the event of not adjoining anything to such a node at all.
 
-$$ P_{adj}(\adj(\tau, v, \textit{None}, n)) +$$
-$$ \sum\limits_{\beta \in A}\sum\limits_{w_\beta} P_{adj}(\adj(\tau, v, \beta, w, n)) = 1.$$
+$$ P_a(\noadj(\tau, v, n)) + \strut$$
+$$ \sum\limits_{\beta \in A}\sum\limits_{w \in W_\beta} P_a(\adj(\tau, v, \beta, w, n)) = 1.$$
 
-{\sc Ptag} assumes that all events occur independent of each other. Therefore it defines the total probability for a derivation as the product of the probability of its individual events.
+{\sc Ptag} assumes that all events occur independently of each other. Therefore it defines the total probability for a derivation as the product of the probability of its individual events.
 
@@
 \label{ssec:pcrisp-domains}
-Using the definition of {\sc ptag}, we now reformulate the {\sc crisp} planning operators described in section \ref{ssec:crispdomain}.
-The independence assumption in {\sc ptag} allows us to maintain that planning operator add a single tree, but are now assigned a probability score.
-However, while {\sc crisp} planning operators can add an elementary tree to any site of the right category, {\sc ptag} substitution and adjunction events are binary events between lexicalized trees at a specific node.
+Using the definition of {\sc ptag}, we now reformulate the \crisp\ planning operators described in section \ref{ssec:crispdomain}.
+The independence assumption in {\sc ptag} allows us to continue to model each addition of a single elementary tree to the derivation (with a certain probability score).
+However, while \crisp\ planning operators can add an elementary tree to any site of the correct category, {\sc ptag} substitution and adjunction events are binary events between lexicalized trees at a specific node.
 We therefore adapt the literals that record open substitution and adjunction sites in partial derivations accordingly and create one operator for each node in each possible combination of lexicalized trees.
-Figure \ref{example-action} shows an example planning operator for each type.
+Fig.~\ref{example-action} shows an example planning operator for each type.
 \begin{figure}[t]
@@
 
-\caption{\label{example-action} {\sc pcrisp} Operator to substitute $t3$ lexicalized with 'cat' for node 2 of $t28$ lexicalized with `eats'.}
+\caption{\label{example-action} Some \pcrisp\ operators for the grammar from Fig.~\ref{fig:grammar}.}
 \end{center}
@@
 Finally, we set the cost of an operator to be its negative log probability. For example 
-\[
-Cost( \subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2) ) =
-\]
-\[ 
- -\log( P(\subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2)) ). 
-\] 
-This way the plan with the smallest cost sum corresponds to the {\sc tag} derivation with the highest probability.
+\begin{align*}
+&Cost(\mbox{subst-t3-cat-t28-eats-n1}) = \\
+ & \qquad -\log P_s(\subst(\mathrm{t3},\textit{`cat'},\mathrm{t28},\textit{`eats'},\mathrm{n1})) ).
+\end{align*}
+This way the plan which minimizes the sum of the costs of its actions  corresponds to the {\sc tag} derivation with the highest probability.
 
@@
 \label{ssec:sparseness}
-%To alleviate data sparseness, caused by the specific event definition of {\sc ptag}, we propose two alternative probability models.
-The first model is an unlexicalized version of {\sc ptag}, the second model uses a simple smoothing approach.\\
-The probability event definition of {\sc ptag} is very fine grained.
-Substitution and adjunction events depend on specific parent and child trees with specific lexicalizations and on a node in the parent tree, as illustrated in figure \ref{fig:modelillustration}, B.1.
+The event definition of {\sc ptag} is very-fine grained.
+Substitution and adjunction events depend on specific parent and child trees with specific lexicalizations and on a node in the parent tree, as illustrated in Fig.~\ref{fig:modelillustration}, B.1.
 When we estimate a probability model from training data, we cannot expect to observe evidence for all combinations of trees.
-Derivations that include such unseen events, have zero probability and are therefore impossible.
+Derivations that include such unseen events have zero probability and are therefore impossible.
 As we show in section \ref{sec:experiments}, this gives rise to a massive data sparseness problem.
 
-  A straightforward way to deal with data sparseness is to drop all lexicalizations from event definitions, as illustrated in figure \ref{fig:modelillustration}, A.
-Unfortunately this model does no longer account for bilexical dependencies between words.
+  A straightforward way to deal with data sparseness is to drop all lexicalizations from event definitions, as illustrated in Fig.~\ref{fig:modelillustration}, A.
+Unfortunately this model  no longer accounts for bilexical dependencies between words.
 Since our system has to add a lexicalized tree in each step, lexicalizations for this child tree should always be taken into account by the probability model, if available.
@@
 This allows us to investigate if purely syntactic information is sufficient to achieve high quality generation output.
-\\
+
 An alternative model computes linear interpolation between three back-off levels.
-The first level is just standard {\sc ptag} (figure \ref{fig:modelillustration}, B.1), for the second level the lexicalizations of the parent tree is dropped (figure \ref{fig:modelillustration}, B.2), for the third level the model describes only the distribution of lexicalized child trees over each category (figure \ref{fig:modelillustration}, B.3).
-Notice also, that the third level is similar in spirit to a probabilistic version of original {\sc crisp} operators (compare figure \ref{fig:modelillustration}, B.3 and figure \ref{fig:crisp-operators}.
+The first level is just standard {\sc ptag} (Fig.~\ref{fig:modelillustration}, B.1), for the second level the lexicalizations of the parent tree is dropped (Fig.~\ref{fig:modelillustration}, B.2), for the third level the model describes only the distribution of lexicalized child trees over each category (Fig.~\ref{fig:modelillustration}, B.3).
+Notice also that the third level is similar in spirit to a probabilistic version of original \crisp\ operators (compare Fig.~\ref{fig:modelillustration}, B.3 and Fig.~\ref{fig:crisp-operators}). 
 \begin{figure}[t]
@@
 \includegraphics[width=.5\textwidth]{figures/modelillustration}
-\caption{\label{fig:modelillustration} Illustration of the unlexicalized probability model (A) and the three back-off levels of the linear interpolation model (B).
-B.1 is the original {\sc ptag} definition and corresponds directly to {\sc pcrisp} operators as in figure \ref{example-action}, while B.3 resembles {\sc crisp} operators (c.f.
-figure \ref{fig:crisp-operators}).}
+\caption{\label{fig:modelillustration} Illustration of the unlexicalized probability model (A) and the three back-off levels of the linear interpolation model (B).}
 
@@
 %\end{figure}
+
+%%% Local Variables:  %%% mode: latex %%% TeX-master: "pcrisp-10" %%% End:

### 2010-05-21 trunk/Papers/pcrisp-10/related.tex
einmal ganz drueber

@@
 \label{sec:related}
-Statistical methods are extremely popular for surface realization, but have not been used in systems that integrate sentence planning.
+
+Statistical methods are popular for surface realization, but have not been used in systems that integrate sentence planning.
 Most statistical generation approaches follow a generate-and-select strategy, first proposed by \newcite{knight1995} in their {\sc nitrogen} system.
-Such systems generate a set of candidate sentences using a conventional (non-statistical) grammar and then select the best output sentence by applying a statistical language model.
-This family includes systems such as {\sc HAL}ogen \cite{langkildeknight1998,langkilde2000} and open{\sc CCG} \cite{whitebaldridge2003}.
-The {\sc Fergus} system \cite{bangalorerambow2000} is a variant of this approach that, like {\sc pcrisp}, employs {\sc ltag}.
+Such systems generate a set of candidate sentences using a (possibly overgenerating) grammar and then select the best output sentence by applying a statistical language model.
+This family includes systems such as {\sc HAL}ogen \cite{langkildeknight1998,langkilde2000} and OpenCCG \cite{whitebaldridge2003}.
+The {\sc Fergus} system \cite{bangalorerambow2000} is a variant of this approach which, like {\sc pcrisp}, employs {\sc ltag}.
 It first assigns elementary trees to each entity in the input sentence plan using a statistical tree model and then computes the most likely derivation using using only these trees with an $n$-gram model on the output sentence.
@@
 A related approach uses a model over local decisions of the generation system itself \cite{belz2008}.
-Both model can either be used to discriminate a set of output candidates, or more directly to choose the next best decision locally.
-Our approach is similar, because it uses {\sc ptag} to find the most likely output structure.
+Both models can either be used to discriminate a set of output candidates, or more directly to choose the next best decision locally.
+Our approach is similar in that it uses {\sc ptag} to find the most likely output structure.
 However, the previous work discussed so far addresses surface realization only.
-The generate and select approach is even incompatible with the integrated treatment of sentence planning and realization, because it discriminates between possible linearization of the same lexical material.
-In contrast, {\sc pcrisp} addresses sentence planning as well.
+We extend this to a statistical NLG algorithm which does surface realization and sentence planning at the same time.
 
@@
 The applicability of these ideas was limited at that time because efficient planning technology was not available.
-Recently the development of more efficient planning algorithms \cite{hoffmannnebel2001,richterwestphal2008} spawned a renewed interest in planning for NLG.
-{\sc Crisp} uses such algorithms to efficiently solve the sentence generation problem defined by {\sc spud} \cite{stonedoran1997}.
+Recently the development of more efficient planning algorithms \cite{hoffmannnebel2001} spawned a renewed interest in planning for NLG.
+{\sc Crisp} uses such algorithms to efficiently solve the sentence generation problem defined by {\sc spud} \cite{Stone2003a}.
 {\sc Spud}, which instead used an incomplete greedy algorithm, is based on an {\sc ltag} whose trees are augmented with semantic and pragmatic constraints.
@@
 The next section explores {\sc crisp} in more detail.
+
+%%% Local Variables:  
+%%% mode: latex 
+%%% TeX-master: "pcrisp-10" 
+%%% End:

### 2010-05-21 trunk/Papers/pcrisp-10/experiments.tex
Minor changes. Camera ready. :)

@@
 However, since the number of \pcrisp\ operators grows quadratically with the grammar size, generating long sentences requires too much time to run batch experiments.
-We therefore restricted our evaluation to the 416 sentences in Section 0 that are shorter than 16 words.
+We therefore restrict our evaluation to the 416 sentences in Section 0 that are shorter than 16 words.
 
-For this testing set we automatically create semantic representation for each sentence, by instantiating the semantic content of each elementary used in its derivation.
+For this testing set we automatically create semantic representation for each sentence, by instantiating the semantic content of each elementary tree used in its derivation.
 We use these representations as input for our system and compare the system output against the original sentence.
@@
 Table \ref{results} presents the results of the experiment for five different generation systems.
-We compare three variants of \pcrisp: the fully lexicalized PTAG model (``PTAG''), the fully unlexicalized model (``unlexicalized''), and a linear interpolation model in which we (manually) set the weight of level 2 to 0.9 \todo{stimmt das?} and the weight of level 3 to 0 (``interpolation'').
+We compare three variants of \pcrisp: the fully lexicalized PTAG model (``PTAG''), the fully unlexicalized model (``unlexicalized''), and a linear interpolation model (``interpolation'') in which we (manually) set the weight of level 2 to 0.9 and the weight of level 3 to 0.
 We also list results for the non-statistical \crisp\ system of  \newcite{kollerstone2007} (``\crisp'') and the greedy search heuristic used by \spud.

### 2010-05-21 trunk/Papers/pcrisp-10/introduction.tex
Minor changes. Camera ready. :)

@@
 architecture, in which the input semantic representation is first
-enriched, e.g.\ by referring expressions, by a \emph{sentence
+enriched, e.g.\ with referring expressions, by a \emph{sentence
   planner} and only then transformed into natural language strings by

### 2010-05-21 trunk/Papers/pcrisp-10/pcrisp.tex
Minor changes. Camera ready. :)

@@
 Likewise, we need to take the step from classical planning to \emph{metric} planning systems which can use the probabilities.
-In metric planning \cite{fox2002}, planning actions can modify the value of numeric variables in addition to adding and deleting logical literals to the state.
+In metric planning \cite{fox2002}, planning actions can modify the value of numeric variables in addition to adding and deleting logical literals from the state.
 The goal state specifies constraints on this variable.

### 2010-05-21 trunk/Papers/pcrisp-10/related.tex
Minor changes. Camera ready. :)

@@
 The {\sc Fergus} system \cite{bangalorerambow2000} is a variant of this approach which, like {\sc pcrisp}, employs {\sc ltag}.
-It first assigns elementary trees to each entity in the input sentence plan using a statistical tree model and then computes the most likely derivation using using only these trees with an $n$-gram model on the output sentence.
-An alternative to the generate and select approach is to use a probabilistic grammar model, like {\sc   ptag}, trained on automatic parses \cite{zhongstent2005}.
+It first assigns elementary trees to each entity in the input sentence plan using a statistical tree model and then computes the most likely derivation using only these trees with an $n$-gram model on the output sentence.
+An alternative to the $n$-gram based generate and select approach is to use a probabilistic grammar model, like {\sc   ptag}, trained on automatic parses \cite{zhongstent2005}.
 A related approach uses a model over local decisions of the generation system itself \cite{belz2008}.
@@
 
-Our treatment of integrated sentence planning and surface realization as planning, is inherited directly from {\sc crisp} \cite{kollerstone2007}.
+Our treatment of integrated sentence planning and surface realization as planning is inherited directly from {\sc crisp} \cite{kollerstone2007}.
 Planning has long played a role in generation, but has focused on discourse planning instead of specifically addressing sentence generation \cite{hovy1988,appelt1992}.
@@
 {\sc Crisp} uses such algorithms to efficiently solve the sentence generation problem defined by {\sc spud} \cite{Stone2003a}.
-{\sc Spud}, which instead used an incomplete greedy algorithm, is based on an {\sc ltag} whose trees are augmented with semantic and pragmatic constraints.
+{\sc Spud}, which instead uses an incomplete greedy algorithm, is based on an {\sc ltag} whose trees are augmented with semantic and pragmatic constraints.
 Given a communicative goal, a solution to the {\sc spud} problem realizes this goal and simultaneously selects referring expressions.

### 2010-05-21 trunk/Papers/pcrisp-10/pcrisp-10.tex
Many local changes, restructured section 4
Fixed bibtex file.
einmal ganz drueber
addresses and minor changes

@@
 \usepackage{graphicx}
+\usepackage{amsmath}
+\usepackage{color}
+
+\usepackage{url}
+
 \setlength\titlebox{6.5cm}    % Expanding the titlebox
@@
 \newlength{\rhwidth}
+
+\newcommand{\f}[1]{\mbox{#1}}
+\newcommand{\adj}[0]{\textit{adj}}
+\newcommand{\noadj}[0]{\textit{noadj}}
+\newcommand{\init}[0]{\textit{init}}
+\newcommand{\subst}[0]{\textit{subst}}
+
+\newcommand{\pcrisp}{PCRISP}
+\newcommand{\crisp}{CRISP}
+\newcommand{\ltag}{LTAG}
+\newcommand{\ptag}{PTAG}
+\renewcommand{\tag}{TAG}
+\newcommand{\spud}{SPUD}
+
+\newcommand{\todo}[1]{\textcolor{red}{(#1)}}
+
 \newcommand{\planaction}[3]{
@@
 \begin{minipage}{.5\textwidth}
+\footnotesize
 \setlength{\rhwidth}{.5\textwidth}
@@
 \author{Daniel Bauer\\
+  Department of Computer Science\\
   Columbia University\\
-  New York, NY, USA\\
+  1214 Amsterdam Ave.\\
+  New York, NY 10027, USA\\
   {\tt bauer@cs.columbia.edu} 
@@
   Alexander Koller\\
+  Cluster of Excellence\\ ``Multimodal Computing and Interaction''\\
   Saarland University\\
-  Saarbr\"{u}cken, Germany\\ 
+  66041 Saarbr\"{u}cken, Germany\\ 
   {\tt koller@mmci.uni-saarland.de}}
@@
 \maketitle
-\begin{abstract}
-Our work merges combined sentence planning and surface realization using 
-{\sc ltag} and statistical generation. This makes it possible to address
-integrated sentence generation with large, treebank induced grammars for
-the first time.
-\end{abstract}
+
+\input abstract
 

### 2010-05-21 trunk/Papers/pcrisp-10/pcrisp.tex
addresses and minor changes

@@
 
-We now extend \crisp\ to statistical generation (\pcrisp).
+We now extend CRISP to statistical generation (PCRISP).
 The basic idea is to add a statistical grammar model while leaving the sentence generation mechanism untouched.
@@
 
- As a straightforward probability model over {\sc ltag} derivations we choose probabilistic {\sc tag} ({\sc ptag}) \cite{resnik1992}.
-Our choice of {\sc ptag} for sentence generation is motivated by a number of attractive properties.
- {\sc Ptag} is lexicalized and therefore does not only assign probabilities to operations in the grammar (as for example plain {\sc pcfg}), but also accounts for binary dependencies between words.
-Unlike n-gram models however, these co-occurrences are structured according to local syntactic context as a result of {\sc tag}'s extended domain of locality.
+ As a straightforward probability model over LTAG derivations we choose probabilistic TAG (PTAG) \cite{resnik1992}.
+Our choice of PTAG for sentence generation is motivated by a number of attractive properties.
+ PTAG is lexicalized and therefore does not only assign probabilities to operations in the grammar (as for example plain PCFG), but also accounts for binary dependencies between words.
+Unlike n-gram models however, these co-occurrences are structured according to local syntactic context as a result of TAG's extended domain of locality.
 The probability model describes how the syntactic arguments of a word are typically filled.
-Furthermore, as {\sc tag} factors recursion from the domain of dependencies, the probability for core constructions remains the same independent of additional adjunctions. 
-%Second, {\sc ptag} is a generative model.
+Furthermore, as TAG factors recursion from the domain of dependencies, the probability for core constructions remains the same independent of additional adjunctions. 
+%Second, PTAG is a generative model.
 This is crucial because it allows us to assign probabilities to partial derivations.
@@
 In contrast, other statistical generation systems use discriminative models that select the best possible generation output from a set of candidate sentences.
-%Finally, the independence assumption makes it easy to estimate {\sc ptag}s from a treebank.
- We review {\sc ptag} in section \ref{ssec:probmodels}.
+%Finally, the independence assumption makes it easy to estimate PTAGs from a treebank.
+ We review PTAG in section \ref{ssec:probmodels}.
 
-While we leave the basic sentence generation mechanism intact, we need to modify the concrete formulation of \crisp\ planning operators to accommodate bilexical dependencies.
+While we leave the basic sentence generation mechanism intact, we need to modify the concrete formulation of CRISP planning operators to accommodate bilexical dependencies.
 Likewise, we need to take the step from classical planning to \emph{metric} planning systems which can use the probabilities.
@@
 While systems such as Metric-{\sc FF} \cite{hoffmann2003} do not guarantee optimality, they do generally offer good results.
-We address our encoding of sentence generation with {\sc ptag} as metric planning in section \ref{ssec:pcrisp-domains}.
+We address our encoding of sentence generation with PTAG as metric planning in section \ref{ssec:pcrisp-domains}.
 
@@
 \label{ssec:probmodels}
-{\sc ptag} \cite{resnik1992} views {\sc tag} derivations as sequences of events of three types: initial events, substitution events, and adjunction events.
+PTAG \cite{resnik1992} views TAG derivations as sequences of events of three types: initial events, substitution events, and adjunction events.
 
@@
 
-{\sc Ptag} assumes that all events occur independently of each other. Therefore it defines the total probability for a derivation as the product of the probability of its individual events.
+PTAG assumes that all events occur independently of each other. Therefore it defines the total probability for a derivation as the product of the probability of its individual events.
 
@@
 \label{ssec:pcrisp-domains}
-Using the definition of {\sc ptag}, we now reformulate the \crisp\ planning operators described in section \ref{ssec:crispdomain}.
-The independence assumption in {\sc ptag} allows us to continue to model each addition of a single elementary tree to the derivation (with a certain probability score).
-However, while \crisp\ planning operators can add an elementary tree to any site of the correct category, {\sc ptag} substitution and adjunction events are binary events between lexicalized trees at a specific node.
+Using the definition of PTAG, we now reformulate the CRISP planning operators described in section \ref{ssec:crispdomain}.
+The independence assumption in PTAG allows us to continue to model each addition of a single elementary tree to the derivation (with a certain probability score).
+However, while CRISP planning operators can add an elementary tree to any site of the correct category, PTAG substitution and adjunction events are binary events between lexicalized trees at a specific node.
 We therefore adapt the literals that record open substitution and adjunction sites in partial derivations accordingly and create one operator for each node in each possible combination of lexicalized trees.
@@
 
-\caption{\label{example-action} Some \pcrisp\ operators for the grammar from Fig.~\ref{fig:grammar}.}
+\caption{\label{example-action} Some PCRISP operators for the grammar from Fig.~\ref{fig:grammar}.}
 \end{center}
@@
 \end{align*}
-This way the plan which minimizes the sum of the costs of its actions  corresponds to the {\sc tag} derivation with the highest probability.
+This way the plan which minimizes the sum of the costs of its actions  corresponds to the TAG derivation with the highest probability.
 
@@
 \label{ssec:sparseness}
-The event definition of {\sc ptag} is very-fine grained.
+The event definition of PTAG is very-fine grained.
 Substitution and adjunction events depend on specific parent and child trees with specific lexicalizations and on a node in the parent tree, as illustrated in Fig.~\ref{fig:modelillustration}, B.1.
@@
 An alternative model computes linear interpolation between three back-off levels.
-The first level is just standard {\sc ptag} (Fig.~\ref{fig:modelillustration}, B.1), for the second level the lexicalizations of the parent tree is dropped (Fig.~\ref{fig:modelillustration}, B.2), for the third level the model describes only the distribution of lexicalized child trees over each category (Fig.~\ref{fig:modelillustration}, B.3).
-Notice also that the third level is similar in spirit to a probabilistic version of original \crisp\ operators (compare Fig.~\ref{fig:modelillustration}, B.3 and Fig.~\ref{fig:crisp-operators}). 
+The first level is just standard PTAG (Fig.~\ref{fig:modelillustration}, B.1), for the second level the lexicalizations of the parent tree is dropped (Fig.~\ref{fig:modelillustration}, B.2), for the third level the model describes only the distribution of lexicalized child trees over each category (Fig.~\ref{fig:modelillustration}, B.3).
+Notice also that the third level is similar in spirit to a probabilistic version of original CRISP operators (compare Fig.~\ref{fig:modelillustration}, B.3 and Fig.~\ref{fig:crisp-operators}). 
 \begin{figure}[t]

### 2010-05-21 trunk/Papers/pcrisp-10/pcrisp-10.tex
done.

@@
   Alexander Koller\\
-  Cluster of Excellence\\ ``Multimodal Computing and Interaction''\\
+  Cluster of Excellence\\
   Saarland University\\
+  Campus C7 4\\
   66041 Saarbr\"{u}cken, Germany\\ 

### 2010-10-04 trunk/Papers/operator-splitting/definition-v1.tex
joergs version v2 from email

\section{Definition Version 1}
\label{definition-v1}

\joerg{this entire section is outdated and kept only as a point of
  reference.}

Given an operator $o$ with parameters $P(o)$, define as the
\emph{parameter-groups} the set $G(o) = \{G_1, \dots, G_n\}$ of
parameter subsets $G_i \subseteq P(o)$ that arises from starting with
$G(o) = \emptyset$, then looping over all predicate $p$ in $o$'s
description and inserting the set of $p$'s parameters as a new group
$G_i$.

A \emph{$k$-split} of $o$ is a set $S(o) = \{S_1, \dots, S_k\}$ of
parameter subsets so that: (1) $\bigcup_{j=1}^k S_j = P(o)$; (2) for
all $1 \leq i \leq n$ there exists $1 \leq j \leq k$ so that $G_i
\subseteq S_j$. The \emph{arity} of a split $S(o)$ is $max_{1\leq j
  \leq k} |S_j|$. A split is \emph{arity-minimizing} if there exists
no other split (for any $k$) that has smaller arity. The
\emph{overlap} of a split $S(o)$ is $|\bigcap_{j=1}^k S_j|$. A split
is \emph{overlap-minimizing} if there exists no other split that has
smaller overlap.
%$S'(o)$ so that $max_{1\leq j \leq k} |S_j| > max_{1\leq j \leq k'}
%  |S'_j|$. A $k$-split $S(o)$ is \emph{overlap-minimizing} if there
%  exists no other split (for any $k$) $S'(o)$ so that
%  $|\bigcap_{j=1}^k S_j| > |\bigcap_{j=1}^{k'} S'_j|$.
\joerg{I'm not sure if overlap-minimzation is meaningful, but
  I thought I can just as well state it. I haven't yet thought at all
  about any possible implications between the two, neither whether
  finding minimal splits is hard. I would guess it's not.}

We observe that, given a split, we can re-write $o$ into $k$ operators
$o_1, \dots, o_k$ so that executing $o$ is equivalent to executing
$o_1, \dots, o_k$ in sequence -- note that this is basically inverting
all the known work on macro-operators... :-)

We introduce new $0$-ary predicates $OK$, $do-o_2, \dots,
do-o_k$. $OK$ becomes a precondition of every operator other than $o$,
as well as of $o_1$; it becomes a delete of $o_1$. $do-o_i$ becomes an
add effect of $o_{i-1}$, a precondition of $o_i$, and a delete effect
of $o_i$. The precondition/add list/delete list of $o_i$ simply
consists of those predicates $p$ of $o$ whose parameters are a subset
of $S_i$ -- more precisely, we put each predicates into exactly one
(arbitrary) $o_i$ for which this is the case. \joerg{Unless I'm
  mistaken, this simple construction does the job -- the modified
  planning task has ``the same''set of plans, modulo replacing $o_1,
  \dots, o_k$ with an according instantiation of $o$ (of course we
  need to prove this formally). Generalizing this to splitting up
  several operators should be straightforward. ... I think there will
  also be an inter-action with predicate-splitting as known in logics:
  smaller predicate arity will yield smaller parameter groups and thus
  splits with smaller arity!}

For example, consider the blocksworld operator ``move(A,B,C)'', whose
precondition is ``on(A,B),clear(A),clear(C)'', whose add is ``on(A,C),
clear(B)'' and whose del is ``on(A,B), clear(C)''. A 2-split is given
by ``A,B'' and ``A,C''. We get the actions ``move1(A,B)'' with pre
``OK, on(A,B), clear(A)'', add ``do-move2, clear(B)'', del ``on(A B),
OK''; and ``move2(A,C)''with pre ``do-move2, clear(C)'', add ``on(A
C), OK'', del ``clear(C), do-move2''.

\joerg{I note that this construction may run planners into trouble
  because, somewhere along the way $o_1, \dots, o_k$, a precondition
  may not be satisfied, ie there may not exists any continuation of
  the operator $o$ -- an instantiation of the remaining parameters --
  so that the op execution can be completed. The planner then finds
  itself in a dead-end. One may think of giving the option to ``roll
  back'' an execution. .. well, maybe this is nonsense because
  branching factor inside $o_1, \dots, o_k$ is zero anyway so
  back-tracking is ``easy''. Well, something to keep in mind.}

### 2010-10-04 trunk/Papers/operator-splitting/definition-v2.tex
joergs version v2 from email

\section{Initial Definition}
\label{initial-definition}

\joerg{Note: our email text is contained at the end of the tex file as
  a comment}

Given an operator $o$ with parameters $P(o)$, define as the
\emph{parameter-groups} the set $G(o) = \{G_1, \dots, G_n\}$ of
parameter subsets $G_i \subseteq P(o)$ that arises from starting with
$G(o) = \emptyset$, then looping over all predicate $p$ in $o$'s
description and inserting the set of $p$'s parameters as a new group
$G_i$.

A \emph{$k$-split} of $o$ is a set $S(o) = \{S_1, \dots, S_k\}$ of
parameter subsets so that: (1) $\bigcup_{j=1}^k S_j = P(o)$; (2) for
all $1 \leq i \leq n$ there exists $1 \leq j \leq k$ so that $G_i
\subseteq S_j$. The \emph{arity} of a split $S(o)$ is $max_{1\leq j
  \leq k} |S_j|$. A split is \emph{arity-minimizing} if there exists
no other split (for any $k$) that has smaller arity. The
\emph{overlap} of a split $S(o)$ is $|\bigcap_{j=1}^k S_j|$. A split
is \emph{overlap-minimizing} if there exists no other split that has
smaller overlap.
%$S'(o)$ so that $max_{1\leq j \leq k} |S_j| > max_{1\leq j \leq k'}
%  |S'_j|$. A $k$-split $S(o)$ is \emph{overlap-minimizing} if there
%  exists no other split (for any $k$) $S'(o)$ so that
%  $|\bigcap_{j=1}^k S_j| > |\bigcap_{j=1}^{k'} S'_j|$.
\joerg{I'm not sure if overlap-minimzation is meaningful, but
  I thought I can just as well state it. I haven't yet thought at all
  about any possible implications between the two, neither whether
  finding minimal splits is hard. I would guess it's not.}

We observe that, given a split, we can re-write $o$ into $k$ operators
$o_1, \dots, o_k$ so that executing $o$ is equivalent to executing
$o_1, \dots, o_k$ in sequence -- note that this is basically inverting
all the known work on macro-operators... :-)

We introduce new $0$-ary predicates $processing\mbox{-}none$,
$do\mbox{-}o_2, \dots, do\mbox{-}o_k$. We introduce new $1$-are
predicates $arg_1(.), \dots, arg_l(.)$ where $l = |P(o)|$ is the
number of parameters of $o$, and $x_1, \dots, x_l$ is an arbitrary
ordering of $P(o)$. $processing\mbox{-}none$ becomes a precondition of
every operator other than $o$, as well as of $o_1$; it becomes a
delete of $o_1$, and an add effect of $o_k$. $do\mbox{-}o_i$ becomes
an add effect of $o_{i-1}$, a precondition of $o_i$, and a delete
effect of $o_i$. The precondition/add list/delete list of $o_i$
consist: (a) of those predicates $p$ of $o$ whose parameters are a
subset of $S_i$ -- more precisely, we put each predicates into exactly
one (arbitrary) $o_i$ for which this is the case; (b) for any operator
$op_i$ in which parameter $x_j$ appears for the first time in the
sequence $o_1, \dots, o_k$, $arg_j(x_j)$ is inserted into the add
effect of $o_1$, and into the precondition of any $o_g$, $g > i$,
where $o_g$ also uses parameter $x_j$. \joerg{Unless I'm mistaken,
  this simple construction does the job -- the modified planning task
  has ``the same''set of plans, modulo replacing $o_1, \dots, o_k$
  with an according instantiation of $o$ (of course we need to prove
  this formally). Generalizing this to splitting up several operators
  should be straightforward. ... I think there will also be an
  inter-action with predicate-splitting as known in logics: smaller
  predicate arity will yield smaller parameter groups and thus splits
  with smaller arity!}

For example, consider the blocksworld operator ``move(A,B,C)'', whose
precondition is ``on(A,B),clear(A),clear(C)'', whose add is ``on(A,C),
clear(B)'' and whose del is ``on(A,B), clear(C)''. A 2-split is given
by ``A,B'' and ``A,C''. We get the actions ``move1(A,B)'' with pre
``processing-none, on(A,B), clear(A)'', add ``do-move2, clear(B),
arg1(A), arg2(B)'', del ``on(A B), processing-none''; and
``move2(A,C)''with pre ``do-move2, clear(C), arg1(A)'', add ``on(A C),
processing-none, arg3(C)'', del ``clear(C), do-move2''.

\joerg{I note that this construction may run planners into trouble
  because, somewhere along the way $o_1, \dots, o_k$, a precondition
  may not be satisfied, ie there may not exists any continuation of
  the operator $o$ -- an instantiation of the remaining parameters --
  so that the op execution can be completed. The planner then finds
  itself in a dead-end. One may think of giving the option to ``roll
  back'' an execution. .. well dunno if this would help or be
  detrimental. something to keep in mind.}

%OLD stuff / meta info

\commentout{

%email feedback by alex and carlos, replies by joerg (1 october 2010)

> The encoding of move(A,B,C) in your draft does not work, for two
>reasons. First, nothing in your encoding enforces that between move1
>and move2, only component actions of "move" can be
>executed. Presumably, putdown2 (created by splitting the putdown
>operator) can also be executed whenever OK is false, even if do-move2
>is true.

Nope -- putdown2 has the precondition "do-putdown2" which will not be fulfilled here.

> The second problem is that nothing in the encoding enforces that A
> is bound to the same value in move1 and move2. One could easily
> "split" move(a1,b,c) into the two actions move1(a1,b) and
> move2(a2,c), and the planner would see nothing wrong with it. This
> means that we need a mechanism for keeping track of the variable
> bindings.

>
> We propose the following variant of your mechanism:
>
> move1(A,B)
>  - pre: processing(none), on(A,B), clear(A)
>  - effects: ~processing(none), processing(move), clear(B), ~on(A,B), arg1(A), arg2(B)
>
> move2(A,C)
>  - pre: processing(move), arg1(A), clear(C)
>  - effects: ~processing(move), processing(none), arg3(C), on(A,C)

Yes, good point. I incorporated this into the draft.

> - At each point in the plan, exactly one instance of processing(X)
>is true. While a split of the action "a" is processed, this is the
>atom processing(a); otherwise it is processing(none).

Yep; the same effect is achieved via my construction, where the "do-*"
predicate is true only for one operator (and one part i of that
operator). I did adopt, in the draft, your name "processing-none"
rather than my nondescript "OK".

Note that I say "operator" not "action". This is old-school talk for
distinguishing between the high-level "operators", using variables, in
the usual PDDL file, and the grounded propositional-logic "actions". I
know, it's not very consistent because in PDDL the denotation actually
is "action"... ;-) ... anyway, let's please stick to this.

> If the order in which the component actions of the split are
>  executed is important, one could keep track of the action number in
>  a separate predicate.

The order is important, at least to the extent that we need to know
what the *last* action will be -- the one to re-establish
"processing-none". From a more general perspective, I think that a
total ordering could be beneficial for planners because it gives less
options to the search -- it is more explicit about the fact that the
only choice we have is how to instantiate the remaining parameters.

> However, one thing that may be inconvenient about this encoding is
> that only one action can be executed at a time. This would make it
> impossible e.g. for Graphplan to parallelize the execution of
> different actions, because the components of splits for different
> operators have mutex preconditions. Is this going to be a problem
> for FF or other modern planners?

The only potential problem I see is that the heuristic functions may
get "confused" by this construction, ie yield worse search guidance
than on the original planning task. ... we'll just have to see about
this empirically.
}

### 2010-10-04 trunk/Papers/operator-splitting/total.tex
joergs version v2 from email

\documentclass{llncs}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage{epsfig,graphics,latexsym}
\usepackage{amsmath,amssymb,theorem,enumerate}

%% Farbige Kommentar-Felder
\usepackage[usenames]{color} % Only used in comment commands

\newcommand{\commentout}[1]{}

\definecolor{Blue}{rgb}{0,0.16,0.90}
\definecolor{Red}{rgb}{0.90,0.16,0}
\definecolor{DarkBlue}{rgb}{0,0.08,0.45}
\definecolor{ChangedColor}{rgb}{0.9,0.08,0}
\definecolor{CommentColor}{rgb}{0.2,0.8,0.2}
\definecolor{ToDoColor}{rgb}{0.1,0.2,1}

\definecolor{verylightgray}{rgb}{0.91,0.91,0.91}

% *** Use this definition of the command to show the comments ***
\newcommand{\todo}[1]{\textbf{\color{ToDoColor} TODO: #1}}
\newcommand{\joerg}[1]{\textbf{\color{CommentColor} /* #1  (joerg)*/}}
\newcommand{\alex}[1]{\textbf{\color{CommentColor} /* #1  (alex)*/}}
\newcommand{\carlos}[1]{\textbf{\color{CommentColor} /* #1  (carlos)*/}}

%\newenvironment{proof}{\noindent {\bf Proof:}}%
%{\hfill \rule[0.3ex]{1ex}{1ex} \par \addvspace{\bigskipamount}}

\newenvironment{proofsketch}{\noindent {\bf Proof Sketch:}}%
{\hfill \rule[0.3ex]{1ex}{1ex} \par \addvspace{\bigskipamount}}

\title{When Operators Learned How to Split Up}

\author{Koller, Areces, Hoffmann, \& Friends}

\institute{I'm sure they're working somewhere \dots}

\begin{document}

\maketitle

\begin{abstract}
PDDL is arguably a programming language for planners. While that
language, in the case of ``fully automatic'' planners, is limited to a
description of the mechanics of the domain of interest, it is
well-known that different models of the same domain can yield immense
differences in planner performance. Indeed, to a considerable extent
the domain-designing activities of hitherto IPC organizers consisted
of ``spoon-feeding'' the targeted domains to the planners, in that
they used their inside knowledge of planning algorithms to design the
domains in a way so that existing planners could attack them in a
sensible way. However, what if actual ``end-users'', people that want
to use planners but that haven't spent the last 10 years implementing
them, wish to ``spoon-feed'' their domain to the planners?

To some extent, this can be supported by documentation and best
practices explanations, as well as graphical environments for domain
modeling. However, the holy grail would be to provide
\emph{domain-rewriting} techniques, in a manner similar to what is
known as program re-writing: Given an input model designed by the
end-user, can we automatically modify that model to yield better
performance, while still encoding essentially the same domain?

We herein begin to address this question, focussing on one particular
technical detail that affects the pre-processing (and in certain ways
also the search techniques) of many existing planners: the number of
operator parameters. For example, in the development of the Pipesworld
domain as used at IPC 2004, one of the essential steps was to split up
some of the operators, so that each part would have less parameters
and thus be more feasible for common pre-processing techniques. We
herein provide an automatic method that does just that. We evaluate
the method on IPC benchmarks, as well as on the domain of sentence
generation, where pre-processing, in particular related to the number
of operator parameters, is known to be one of the major obstacles.
\end{abstract}

\input{definition-v1}

\input{definition-v2}

\bibliographystyle{plain}
\bibliography{biblio}

\end{document}

### 2010-10-27 trunk/Papers/operator-splitting/equiv-definition-carlos.tex
structure for "homework" defs

\subsection{Definition: Carlos}
\label{definition-carlos}

### 2010-10-27 trunk/Papers/operator-splitting/equiv-definition-gabi.tex
structure for "homework" defs

\subsection{Definition: Gabi}
\label{definition-gabi}

### 2010-11-15 trunk/Papers/operator-splitting/equiv-definition-gabi.tex
first draft of my definition of equivalence

@@
 \label{definition-gabi}
+
+Altough I started from the compilation schemes, I ended up with a definition
+that is much closer to reductions (an idea that Joerg also mentioned in the
+Skype call).
+
+For my definition, I use the following terminology:
+A (PDDL) planning task is a tuple $\left<D, P\right>$
+where $D$ is the domain description and $P=(O,I,\gamma)$ 
+is the instance description, consisting of the set of objects $O$, the
+initial state $I$ and the goal specification $\gamma$.
+
+\begin{theorem}[Reduction]
+A reduction of planning domain $D$ to planning domain $D'$ is a tuple
+$\mathbf f=\left<f_P, f_{\pi'}\right>$ of functions
+such that for all planning tasks $\Pi = \left<D, P\right>$:
+\begin{enumerate}
+    \item $\left<D', f_P(P)\right>$ is a planning task,
+    \item for each plan $\pi'$ of $\left<D', f_P(P)\right>$, $f_{\pi'}(\pi')$ is
+          a plan for $\Pi$,
+    \item $f_P$ and $f_{\pi'}$ are polynomial-time computable,
+    \item for each two plans $\pi'_1$ and $\pi'_2$ of $\left<D', f_P(P)\right>$
+          it holds that $\textit{cost}(f_{\pi'}(\pi'_1)) >
+          \textit{cost}(f_{\pi'}(\pi'_2))$ iff $\textit{cost}(\pi'_1) >
+          \textit{cost}(\pi'_2)$, and
+    \item for at least one optimal plan $\pi^*$ of $\Pi$ there is a
+          plan $\pi'$ of $\left<D', f_P(P)\right>$
+          whose size is polynomial in the size of $\pi^*$ and
+          $\textit{cost}(\pi^*) = \textit{cost}(f_{\pi'}(\pi'))$.
+\end{enumerate}
+\end{theorem}
+
+\begin{theorem}[Reducibility]
+A planning domain $D$ is reducible to planning domain $D'$ (written $D\leq D'$)
+if there is a reduction $\mathbf f$ from $D$ to $D'$.
+\end{theorem}
+
+\begin{theorem}[Equivalence of planning domains]
+Two planning domains $D$ and $D'$ are equivalent (written $D\equiv D'$) iff
+$D\leq D'$ and $D'\leq D$.
+\end{theorem}
+
+\noindent Some random remarks:
+\begin{itemize}
+    \item From a practical perspective we are surely interested into generating
+        a more expressive domain (a domain which our original domain is 
+        reducible to) plus the reduction. However, from a theoretical point of
+        view, if we want to show that two domains are \emph{not} equivalent, we
+        want to prove that there is no reduction.
+    \item Reductions are composable, i.e.\ if $\left<f_P^1, f_{\pi'}^1\right>$
+          is a reduction of $D$ to $D_1$ and $\left<f_P^2, f_{\pi'}^2\right>$
+          is a reduction of $D_1$ to $D_2$, then $\left<f_P^2\circ f_P^1, 
+          f_{\pi'}^2\circ f_{\pi'}^1\right>$ is a reduction of $D$ to $D_2$.
+    \item Condition $2$ ensures (together with the existence requirement in
+          condition $5$) that there is a plan for $\Pi$ iff there is a
+          plan for $\left<D', f_P(P)\right>$.
+    \item Condition $4$ ensures that the domains caputure the same idea
+          of (relative) plan quality.
+    \item Condition $5$ makes the reduction optimality preserving. In practice
+          it is probably most of the time the case (and the easiest to show) that
+          this condition holds for all plans $\pi$ of $\Pi$ but the current
+          definition is sufficient to ensure that 
+          \begin{itemize}
+            \item for optimal planning, we can use the reduction to find an
+                optimal plan if there is a plan.
+            \item for satisficing planning, we can use the reduction to
+                find a plan if there is a plan.
+          \end{itemize}
+          However, if we do not extend condition $5$ to all plans $\pi$ of
+          $\Pi$, we are not guaranteed that we can find a plan for each
+          cost for which a plan in the original task exists.
+    \item I still have to think more about for what the time and space
+          restrictions are necessary and whether there must be a 
+          restriction on the size of $D'$ (currently I do not see, why).
+\end{itemize}

### 2010-11-18 trunk/Papers/operator-splitting/equiv-definition-carlos.tex

@@
+\newcommand{\tup}[1]{\langle #1 \rangle}
+
 \subsection{Definition: Carlos}
 \label{definition-carlos}
+
+Let start by introducing some basic notation
+
+\begin{definition}
+An \emph{action} is a pair $a = \tup{P,E}$ where both $P$ and $E$ are finite sets of 
+propositional literals. $P$ is the set of \emph{preconditions} of $a$. $E$ is the set of \emph{effects} of $a$.
+
+A \emph{planning domain} $D$ is a finite set of actions.
+\fixme{If we want, we can define something like a `\emph{restricted planning domain}' as a pair $D=\tup{A,R}$ where $A$ is a finite set of actions, and $R$ is a finite set of
+propositional formulas$.  We would use $R$ to restrict planning tasks by requesting consistency with $R$.}
+
+A \emph{planning task} is a tuple $\tup{D, P}$
+where $D$ is a planning domain and  $P=\tup{I,F)$ 
+is an \emph{instance description}, consisting of a finite set of literals $I$
+(the \emph{initial state}) and the a finite set of literals $O$ (the \emph{final state}).
+
+\end{definition}
+
+\begin{definition}[Reduction]
+A reduction of planning domain $D$ to planning domain $D'$ is a tuple
+$\mathbf f=\left<f_P, f_{\pi'}\right>$ of functions
+such that for all planning tasks $\Pi = \left<D, P\right>$:
+\begin{enumerate}
+    \item $\left<D', f_P(P)\right>$ is a planning task,
+    \item for each plan $\pi'$ of $\left<D', f_P(P)\right>$, $f_{\pi'}(\pi')$ is
+          a plan for $\Pi$,
+    \item $f_P$ and $f_{\pi'}$ are polynomial-time computable,
+    \item for each two plans $\pi'_1$ and $\pi'_2$ of $\left<D', f_P(P)\right>$
+          it holds that $\textit{cost}(f_{\pi'}(\pi'_1)) >
+          \textit{cost}(f_{\pi'}(\pi'_2))$ iff $\textit{cost}(\pi'_1) >
+          \textit{cost}(\pi'_2)$, and
+    \item for at least one optimal plan $\pi^*$ of $\Pi$ there is a
+          plan $\pi'$ of $\left<D', f_P(P)\right>$
+          whose size is polynomial in the size of $\pi^*$ and
+          $\textit{cost}(\pi^*) = \textit{cost}(f_{\pi'}(\pi'))$.
+\end{enumerate}
+\end{definition}
+
+\begin{definition}[Reducibility]
+A planning domain $D$ is reducible to planning domain $D'$ (written $D\leq D'$)
+if there is a reduction $\mathbf f$ from $D$ to $D'$.
+\end{definition}
+
+\begin{definition}[Equivalence of planning domains]
+Two planning domains $D$ and $D'$ are equivalent (written $D\equiv D'$) iff
+$D\leq D'$ and $D'\leq D$.
+\end{definition}

### 2010-11-18 trunk/Papers/operator-splitting/equiv-definition-gabi.tex

@@
 
-\begin{theorem}[Reduction]
+\begin{definition}[Reduction]
 A reduction of planning domain $D$ to planning domain $D'$ is a tuple
@@
 \end{enumerate}
-\end{theorem}
+\end{definition}
 
-\begin{theorem}[Reducibility]
+\begin{definition}[Reducibility]
 A planning domain $D$ is reducible to planning domain $D'$ (written $D\leq D'$)
 if there is a reduction $\mathbf f$ from $D$ to $D'$.
-\end{theorem}
+\end{definition}
 
-\begin{theorem}[Equivalence of planning domains]
+\begin{definition}[Equivalence of planning domains]
 Two planning domains $D$ and $D'$ are equivalent (written $D\equiv D'$) iff
 $D\leq D'$ and $D'\leq D$.
-\end{theorem}
+\end{definition}
 

### 2010-11-18 trunk/Papers/operator-splitting/equiv-definition-carlos.tex
fixed some bugs in carlos part

@@
 \fixme{If we want, we can define something like a `\emph{restricted planning domain}' as a pair $D=\tup{A,R}$ where $A$ is a finite set of actions, and $R$ is a finite set of
-propositional formulas$.  We would use $R$ to restrict planning tasks by requesting consistency with $R$.}
+propositional formulas.  We would use $R$ to restrict planning tasks by requesting consistency with $R$.}
 
 A \emph{planning task} is a tuple $\tup{D, P}$
-where $D$ is a planning domain and  $P=\tup{I,F)$ 
+where $D$ is a planning domain and  $P=\tup{I,F}$ 
 is an \emph{instance description}, consisting of a finite set of literals $I$

### 2010-11-18 trunk/Papers/operator-splitting/equiv-definition-carlos.tex

--- 2010-10-27 trunk/Papers/operator-splitting/equiv-definition-carlos.tex
@@
+\newcommand{\tup}[1]{\langle #1 \rangle}
+
 \subsection{Definition: Carlos}
 \label{definition-carlos}
+
+Let start by introducing some basic notation
+
+\begin{definition}
+An \emph{action} is a pair $a = \tup{P,E}$ where both $P$ and $E$ are finite sets of 
+propositional literals. $P$ is the set of \emph{preconditions} of $a$. $E$ is the set of \emph{effects} of $a$.
+
+A \emph{planning domain} $D$ is a finite set of actions. 
+\fxnote{\tiny If we want, we can define something like a `\emph{restricted planning domain}' as a pair $D=\tup{A,R}$ where $A$ is a finite set of actions, and $R$ is a finite set of 
+propositional formulas.  We would use $R$ to restrict planning tasks by requesting consistency with $R$.}
+
+A \emph{planning task} is a tuple $\Pi=\tup{D, P}$
+where $D$ is a planning domain and  $P=\tup{I,F}$ 
+is an \emph{instance description}, consisting of a finite set of literals $I$
+(the \emph{initial state}) and the a finite set of literals $O$ (the \emph{final state}).
+
+\fxnote{\tiny I don't see that we are using objects anywere, so I abtracted from them and I'm only considering propositional literals.}
+\end{definition}
+
+\begin{definition}[Reduction]
+A reduction of planning domain $D$ to planning domain $D'$ is a tuple
+$\mathbf r=\left<f_1, f_2\right>$ of functions
+such that for all planning tasks $\Pi = \left<D, P\right>$:
+\begin{enumerate}
+    \item $f_1$ and $f_2$ are computable functions,
+    \item $\left<D', f_1(P)\right>$ is a planning task,
+    \item for each plan $\pi'$ of $\left<D', f_1(P)\right>$, $f_2(\pi')$ is
+          a plan for $\Pi$,
+\end{enumerate}
+
+We say that a reduction $\mathbf r = \tup{f_1,f_2}$ 
+\begin{enumerate}
+\item is \emph{polynomial} if both $f_1$ and $f_2$ are 
+polynomial-time computable;
+
+\fxnote{\tiny I suppose that \textit{cost} is a function from plans to natural numbers.}
+\item \emph{preserves costs} if for each two plans $\pi'_1$ and $\pi'_2$ of $\left<D', f_1(P)\right>$
+          it holds that $\textit{cost}(f_2(\pi'_1)) >
+          \textit{cost}(f_2(\pi'_2))$ iff $\textit{cost}(\pi'_1) >
+          \textit{cost}(\pi'_2)$;
+
+\item \emph{preserves optimal plans} if for at least one optimal plan $\pi^*$ of $\Pi$ there is a
+          plan $\pi'$ of $\left<D', f_1(P)\right>$
+          whose size is polynomial in the size of $\pi^*$ and
+          $\textit{cost}(\pi^*) = \textit{cost}(f_2(\pi'))$.
+\end{enumerate}
+
+\end{definition}
+
+\begin{definition}[Reducibility]
+A planning domain $D$ is reducible to planning domain $D'$ (written $D\leq D'$)
+if there is a reduction $\mathbf r$ from $D$ to $D'$.
+\end{definition}
+
+\begin{definition}[Equivalence of planning domains]
+Two planning domains $D$ and $D'$ are equivalent (written $D\equiv D'$) iff
+$D\leq D'$ and $D'\leq D$.
+\end{definition}
+
+Some comments:
+
+\begin{itemize}
+
+\item If we are interesting just in expressivity (when a planning domain can be reduced 
+to another) I don't see why we need to impose time and space constrains.
+
+\item On the other hand, a polynomial reduction preserves complexity bounds.
+
+\item In the same way, reductions with stronger properties preserve more properties of the planning domain.
+
+\end{itemize}

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/converter/CRISPtoPDDL.java
oops

@@ -1,63 +1,0 @@
-package crisp.converter;
-
-import crisp.planningproblem.Domain;
-import crisp.planningproblem.Problem;
-import crisp.planningproblem.codec.PddlOutputCodec;
-
-
-
-import de.saar.penguin.tag.grammar.CrispGrammar;
-import de.saar.penguin.tag.grammar.SituatedCrispXmlInputCodec;
-import java.io.File;
-import java.io.FileReader;
-
-public class CRISPtoPDDL {
-
-    /**
-     * Print a usage message
-     */
-    public static void usage() {
-        System.out.println("Usage: java crisp.converter.CRISPtoPPDL [CRISP grammar] [CRISP problem]");
-    }
-
-    /**
-     * Main program.  When running the converter from the command line, pass
-     * the name of the CRISP problem file as the first argument.
-     *
-     * @param args
-     * @throws Exception
-     */
-    public static void main(String[] args) throws Exception {
-
-        if (args.length < 1) {
-            System.err.println("No crisp problem specified");
-            usage();
-            System.exit(1);
-        }
-
-        // TODO some exception handling
-
-        Domain domain = new Domain();
-        Problem problem = new Problem();
-
-        long start = System.currentTimeMillis();
-
-        SituatedCrispXmlInputCodec codec = new SituatedCrispXmlInputCodec();
-        CrispGrammar grammar = new CrispGrammar();
-        codec.parse(new FileReader(new File(args[0])), grammar);
-
-        File problemfile = new File(args[1]);
-
-        CurrentNextCrispConverter converter = new CurrentNextCrispConverter();
-        converter.convert(grammar, problemfile, domain, problem);
-        long end = System.currentTimeMillis();
-
-        System.out.println("Total runtime: " + (end - start) + "ms");
-
-        System.out.println("Domain: " + domain);
-        System.out.println("Problem: " + problem);
-
-        new PddlOutputCodec().writeToDisk(domain, problem, "");
-        
-    }
-}

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/converter/CRISPtoPKS.java

@@ -1,46 +1,0 @@
-package crisp.converter;
-
-import crisp.planningproblem.Domain;
-import crisp.planningproblem.Problem;
-import crisp.planningproblem.codec.PksOutputCodec;
-
-import de.saar.penguin.tag.codec.CrispXmlInputCodec;
-import de.saar.penguin.tag.grammar.Grammar;
-import de.saar.chorus.term.Term;
-
-import java.io.FileReader;
-import java.io.File;
-
-public class CRISPtoPKS {
-
-	/**
-	 * Main program.  When running the converter from the command line, pass
-	 * the name of the CRISP problem file as the first argument.
-	 *
-	 * @param args
-	 * @throws Exception
-	 */
-	public static void main(String[] args) throws Exception {
-		Domain domain = new Domain();
-		Problem problem = new Problem();
-
-		long start = System.currentTimeMillis();
-
-                CrispXmlInputCodec codec = new CrispXmlInputCodec();
-		Grammar<Term> grammar = new Grammar<Term>();
-		codec.parse(new FileReader(new File(args[0])), grammar);
-
-                File problemFile =  new File(args[1]);
-
-		new FastCRISPConverter().convert(grammar, problemFile, domain, problem);
-		long end = System.currentTimeMillis();
-
-		System.err.println("Total runtime: " + (end-start) + "ms");
-
-		System.out.println("Domain: " + domain);
-		System.out.println("Problem: " + problem);
-
-		new PksOutputCodec().writeToDisk(domain, problem, "");
-	}
-
-}

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/converter/CurrentNextCrispConverter.java

@@ -1,1183 +1,0 @@
-package crisp.converter;
-
-import java.io.File;
-import java.io.FileReader;
-import java.io.IOException;
-import java.io.Reader;
-import java.io.StringWriter;
-import java.util.ArrayList;
-import java.util.Collection;
-import java.util.HashMap;
-import java.util.HashSet;
-import java.util.List;
-import java.util.Map;
-import java.util.Set;
-import java.util.Stack;
-
-import javax.xml.parsers.ParserConfigurationException;
-import javax.xml.parsers.SAXParser;
-import javax.xml.parsers.SAXParserFactory;
-
-import org.xml.sax.Attributes;
-import org.xml.sax.InputSource;
-import org.xml.sax.SAXException;
-import org.xml.sax.SAXParseException;
-import org.xml.sax.helpers.DefaultHandler;
-
-import crisp.planningproblem.Action;
-import crisp.planningproblem.Domain;
-import crisp.planningproblem.Problem;
-import crisp.planningproblem.formula.Conjunction;
-import crisp.planningproblem.formula.Formula;
-import crisp.planningproblem.formula.Literal;
-import crisp.planningproblem.formula.Negation;
-import crisp.planningproblem.formula.Universal;
-import crisp.planningproblem.formula.Conditional;
-import de.saar.chorus.term.Compound;
-import de.saar.chorus.term.Constant;
-import de.saar.chorus.term.Substitution;
-import de.saar.chorus.term.Term;
-import de.saar.chorus.term.Variable;
-import de.saar.chorus.term.parser.TermParser;
-import de.saar.penguin.tag.grammar.Constraint;
-import de.saar.penguin.tag.grammar.CrispGrammar;
-import de.saar.penguin.tag.grammar.CrispLexiconEntry;
-import de.saar.penguin.tag.grammar.ElementaryTree;
-import de.saar.penguin.tag.grammar.ElementaryTreeType;
-import de.saar.penguin.tag.grammar.Grammar;
-import de.saar.penguin.tag.grammar.LexiconEntry;
-import de.saar.penguin.tag.grammar.NodeType;
-import de.saar.penguin.tag.grammar.filter.SemanticsPredicateListFilter;
-
-/**
- * This class provides a fast converter from XML CRISP problem descriptions to
- * planning domains and problems. This class only processes non-probabilistic grammar descriptions.
- * The parser is implemented as a plain SAX parser, thereby improving
- * processing speed and decreasing memory requirements over the old xpath based
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/converter/PCrispActionCreator.java

@@ -1,775 +1,0 @@
-package crisp.converter;
-
-import java.io.StringWriter;
-import java.util.ArrayList;
-import java.util.HashMap;
-import java.util.HashSet;
-import java.util.List;
-import java.util.Map;
-import java.util.Collection;
-import java.util.Collections;
-
-import crisp.planningproblem.Action;
-
-import crisp.planningproblem.formula.Formula;
-import crisp.planningproblem.formula.Conjunction;
-import crisp.planningproblem.formula.Literal;
-
-import de.saar.chorus.term.Compound;
-import de.saar.chorus.term.Constant;
-import de.saar.chorus.term.Term;
-import de.saar.chorus.term.Variable;
-
-import de.saar.penguin.tag.grammar.ProbabilisticGrammar;
-import de.saar.penguin.tag.grammar.LexiconEntry;
-import de.saar.penguin.tag.grammar.ElementaryTree;
-import de.saar.penguin.tag.grammar.NodeType;
-import de.saar.penguin.tag.grammar.Constraint;
-
-
-
-
-
-/**
- * Create CRISP actions whose preconditions specify a specific lexicalized
- * parent tree and node as target for substitution and adjunction.
- * @author dbauer
- */
-public class PCrispActionCreator {
-    
-    /******************** Methods to create actions ***************/    
-    
-    private static int maximumArity = 0; // the maximum arity of any predicate in the problem file
-    
-    private static String getTreeName(LexiconEntry entry){
-        String actionName = entry.tree + "-" + entry.word;
-        for (String aux : entry.auxLexicalItems.keySet()) {
-            actionName += "-" + aux;
-        }
-        return actionName;
-    }
-    
-    public static Action createInitAction(ProbabilisticGrammar grammar, LexiconEntry entry, double prob, Map<String,HashSet<String>> roles){               
-        
-        // Get the tree for this lexical entry from the hash map        
-        ElementaryTree<Term> tree = grammar.getTree(entry.tree);
-        Collection<String> allNodes = tree.getAllNodes();
-        
-        // Get lists of nodes that are open for substitution and adjunction
-        List<String> substNodes = new ArrayList<String>();
-        List<String> adjNodes = new ArrayList<String>();                
-        
-        Map<String,String> constants = new HashMap<String,String>();
-        
-        for (String node : allNodes) {
-            String cat = tree.getNodeLabel(node).toLowerCase();
-            if (cat==null || cat.equals("")) {
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/converter/ProbCRISPConverter.java

@@ -1,739 +1,0 @@
-package crisp.converter;
-
-import java.io.File;
-import java.io.FileReader;
-import java.io.IOException;
-import java.io.Reader;
-import java.io.StringWriter;
-import java.util.ArrayList;
-import java.util.Collection;
-import java.util.HashMap;
-import java.util.HashSet;
-import java.util.List;
-import java.util.Map;
-import java.util.Set;
-import java.util.Stack;
-
-import javax.xml.parsers.ParserConfigurationException;
-import javax.xml.parsers.SAXParser;
-import javax.xml.parsers.SAXParserFactory;
-
-import org.xml.sax.Attributes;
-import org.xml.sax.InputSource;
-import org.xml.sax.SAXException;
-import org.xml.sax.SAXParseException;
-import org.xml.sax.helpers.DefaultHandler;
-
-import crisp.planningproblem.Action;
-import crisp.planningproblem.Domain;
-import crisp.planningproblem.Problem;
-import crisp.planningproblem.formula.Formula;
-import crisp.planningproblem.formula.Conjunction;
-import crisp.planningproblem.formula.Universal;
-import crisp.planningproblem.formula.Literal;
-import de.saar.chorus.term.Compound;
-import de.saar.chorus.term.Constant;
-import de.saar.chorus.term.Term;                         
-import de.saar.chorus.term.Variable;
-import de.saar.chorus.term.parser.TermParser;
-import de.saar.penguin.tag.grammar.ElementaryTree;
-import de.saar.penguin.tag.grammar.Grammar;
-import de.saar.penguin.tag.grammar.ProbabilisticGrammar;
-import de.saar.penguin.tag.grammar.LexiconEntry;
-import de.saar.penguin.tag.grammar.filter.GrammarFilterer;
-import de.saar.penguin.tag.grammar.filter.SemanticsPredicateListFilter;
-import de.saar.penguin.tag.grammar.filter.LexiconEntryFilter;
-
-
-/**
-* This class provides a fast converter from XML CRISP problem descriptions to
-* planning domains with costs and problems using a probabilistic TAG grammar. 
-*/
-
-public class ProbCRISPConverter implements ProblemConverter {
-                                                          // Default Handler already does a lot of work like
-                                                          // parse error handling and registering the handler
-                                                          // with a parser.                
-    
-    protected int plansize;         // the maximum plan size as specified in the problem file
-    protected int maximumArity = 0; // the maximum arity of any predicate in the problem file
-    
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/evaluation/MySQLInterface.java

@@ -1,195 +1,0 @@
-package crisp.evaluation;
-
-import java.sql.Connection;
-import java.sql.DriverManager;
-import java.sql.PreparedStatement;
-import java.sql.ResultSet;
-import java.sql.SQLException;
-import java.sql.Statement;
-import java.sql.Timestamp;
-
-import java.util.Set;
-import java.util.HashSet;
-import java.util.Properties;
-
-import de.saar.chorus.term.Term;
-import de.saar.chorus.term.parser.TermParser;
-
-import de.saar.penguin.tag.derivation.DerivationTree;
-import de.saar.penguin.tag.derivation.DerivedTree;
-
-/**
- * An interface for connecting to a database that stores planning problems and result. 
- *
- * @author Daniel Bauer
- *
- */
-public class MySQLInterface extends DatabaseInterface {
-
-    private Connection connection;
-    private String url;
-    private String username;
-    private String password;
-    private Properties connectionProperties;        
-
-    /**
-     * Open a new database interface.
-     */
-    public MySQLInterface (String url, String username, String password) throws Exception{
-        Class.forName ("com.mysql.jdbc.Driver").newInstance();
-                               
-        connectionProperties = new Properties();
-        if( username != null ) {
-            connectionProperties.put("user", username);
-        }
-        if( password != null ) {
-            connectionProperties.put("password", password);
-        }
-                        
-        connection = DriverManager.getConnection(url, connectionProperties);
-    }
-    
-     /**
-      * Execute a given SQL query on the database and return a result set.
-      */
-     private ResultSet executeQuery(String sql) throws SQLException{
-        Statement stmt = connection.createStatement(ResultSet.TYPE_SCROLL_SENSITIVE,  ResultSet.CONCUR_READ_ONLY);
-        ResultSet srs = stmt.executeQuery(sql);
-        return srs;
-     }
-
-
-     private int executeUpdate(String sql) throws SQLException{
-         Statement stmt = connection.createStatement();
-         int updatecount = stmt.executeUpdate(sql);
-         return updatecount;
-     }
-     
-    /**
-     * Retrieve a semantic representation for a sentence in the database.
-     * @return Semantic representation for the sentence as a set of ground positive literals.
-     * @param sentenceID the id of the sentence for which to return semantics.
-     */
-    public Set<Term> getSentenceSemantics(String semanticsTableName, int sentenceID) throws Exception{
-        
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/evaluation/ffplanparser/FfPlanParser.java

@@ -1,383 +1,0 @@
-/* Generated By:JavaCC: Do not edit this line. FfPlanParser.java */
-package crisp.evaluation.ffplanparser;
-
-import de.saar.chorus.term.Term;
-import de.saar.chorus.term.Compound;
-import de.saar.chorus.term.Constant;
-
-import java.io.Reader;
-import java.io.StringReader;
-import java.util.List;
-import java.util.ArrayList;
-import java.util.Map;
-
-/**
- * A parser for reading a plan that is output by the FF planner and returning
- * it as a list of {@link give.formula.Atom}s representing action instances.
- *
- * @author Alexander Koller
- *
- */
- public class FfPlanParser implements FfPlanParserConstants {
-
-    public String normalize(String s) {
-        return s.toLowerCase();
-    }
-
-
-    public static List<Term> parse(String string) {
-        try {
-            FfPlanParser p = new FfPlanParser(new StringReader(string));
-            return p.plan();
-        } catch(ParseException e) {
-            System.err.println("Exception while parsing " + string + ": " + e);
-            return null;
-        }
-    }
-
-  final public List<Term> plan() throws ParseException {
-        List<Term> ret = new ArrayList<Term>();
-    Term atom;
-    jj_consume_token(STEP);
-    label_1:
-    while (true) {
-      if (jj_2_1(2)) {
-        ;
-      } else {
-        break label_1;
-      }
-      jj_consume_token(STEPNO);
-      atom = planstep();
-                                           System.out.println(atom); ret.add(atom);
-    }
-                {if (true) return ret;}
-    throw new Error("Missing return statement in function");
-  }
-
-  final public Compound planstep() throws ParseException {
-        List<Term> arguments = new ArrayList<Term>();
-        Token predicate, arg;
-    predicate = jj_consume_token(IDENTIFIER);
-    label_2:
-    while (true) {
-      if (jj_2_2(2)) {
-        ;
-      } else {
-        break label_2;
-      }
-      arg = jj_consume_token(IDENTIFIER);
-                                                   arguments.add(new Constant(normalize(arg.image)));
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/evaluation/ffplanparser/FfPlanParserTokenManager.java

@@ -1,396 +1,0 @@
-/* Generated By:JavaCC: Do not edit this line. FfPlanParserTokenManager.java */
-package crisp.evaluation.ffplanparser;
-import de.saar.chorus.term.Term;
-import de.saar.chorus.term.Compound;
-import de.saar.chorus.term.Constant;
-import java.io.Reader;
-import java.io.StringReader;
-import java.util.List;
-import java.util.ArrayList;
-import java.util.Map;
-
-public class FfPlanParserTokenManager implements FfPlanParserConstants
-{
-  public  java.io.PrintStream debugStream = System.out;
-  public  void setDebugStream(java.io.PrintStream ds) { debugStream = ds; }
-private final int jjStopStringLiteralDfa_0(int pos, long active0)
-{
-   switch (pos)
-   {
-      case 0:
-         if ((active0 & 0x20L) != 0L)
-         {
-            jjmatchedKind = 7;
-            return 3;
-         }
-         return -1;
-      case 1:
-         if ((active0 & 0x20L) != 0L)
-         {
-            jjmatchedKind = 7;
-            jjmatchedPos = 1;
-            return 3;
-         }
-         return -1;
-      case 2:
-         if ((active0 & 0x20L) != 0L)
-         {
-            jjmatchedKind = 7;
-            jjmatchedPos = 2;
-            return 3;
-         }
-         return -1;
-      default :
-         return -1;
-   }
-}
-private final int jjStartNfa_0(int pos, long active0)
-{
-   return jjMoveNfa_0(jjStopStringLiteralDfa_0(pos, active0), pos + 1);
-}
-private final int jjStopAtPos(int pos, int kind)
-{
-   jjmatchedKind = kind;
-   jjmatchedPos = pos;
-   return pos + 1;
-}
-private final int jjStartNfaWithStates_0(int pos, int kind, int state)
-{
-   jjmatchedKind = kind;
-   jjmatchedPos = pos;
-   try { curChar = input_stream.readChar(); }
-   catch(java.io.IOException e) { return pos + 1; }
-   return jjMoveNfa_0(state, pos + 1);
-}
-private final int jjMoveStringLiteralDfa0_0()
-{
-   switch(curChar)
-   {
-      case 115:
-         return jjMoveStringLiteralDfa1_0(0x20L);
-      default :
-         return jjMoveNfa_0(2, 0);
-   }
-}
-private final int jjMoveStringLiteralDfa1_0(long active0)
-{
-   try { curChar = input_stream.readChar(); }
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/evaluation/ffplanparser/Token.java

@@ -1,81 +1,0 @@
-/* Generated By:JavaCC: Do not edit this line. Token.java Version 3.0 */
-package crisp.evaluation.ffplanparser;
-
-/**
- * Describes the input token stream.
- */
-
-public class Token {
-
-  /**
-   * An integer that describes the kind of this token.  This numbering
-   * system is determined by JavaCCParser, and a table of these numbers is
-   * stored in the file ...Constants.java.
-   */
-  public int kind;
-
-  /**
-   * beginLine and beginColumn describe the position of the first character
-   * of this token; endLine and endColumn describe the position of the
-   * last character of this token.
-   */
-  public int beginLine, beginColumn, endLine, endColumn;
-
-  /**
-   * The string image of the token.
-   */
-  public String image;
-
-  /**
-   * A reference to the next regular (non-special) token from the input
-   * stream.  If this is the last token from the input stream, or if the
-   * token manager has not read tokens beyond this one, this field is
-   * set to null.  This is true only if this token is also a regular
-   * token.  Otherwise, see below for a description of the contents of
-   * this field.
-   */
-  public Token next;
-
-  /**
-   * This field is used to access special tokens that occur prior to this
-   * token, but after the immediately preceding regular (non-special) token.
-   * If there are no such special tokens, this field is set to null.
-   * When there are more than one such special token, this field refers
-   * to the last of these special tokens, which in turn refers to the next
-   * previous special token through its specialToken field, and so on
-   * until the first special token (whose specialToken field is null).
-   * The next fields of special tokens refer to other special tokens that
-   * immediately follow it (without an intervening regular token).  If there
-   * is no such token, this field is null.
-   */
-  public Token specialToken;
-
-  /**
-   * Returns the image.
-   */
-  public String toString()
-  {
-     return image;
-  }
-
-  /**
-   * Returns a new Token object, by default. However, if you want, you
-   * can create and return subclass objects based on the value of ofKind.
-   * Simply add the cases to the switch for all those special cases.
-   * For example, if you have a subclass of Token called IDToken that
-   * you want to create if ofKind is ID, simlpy add something like :
-   *
-   *    case MyParserConstants.ID : return new IDToken();
-   *
-   * to the following switch statement. Then you can cast matchedToken
-   * variable to the appropriate type and use it in your lexical actions.
-   */
-  public static final Token newToken(int ofKind)
-  {
-     switch(ofKind)
-     {
-       default : return new Token();
-     }
-  }
-
-}

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/evaluation/lamaplanparser/LamaPlanParser.java

@@ -1,371 +1,0 @@
-/* Generated By:JavaCC: Do not edit this line. LamaPlanParser.java */
-package crisp.evaluation.lamaplanparser;
-
-import de.saar.chorus.term.Term;
-import de.saar.chorus.term.Compound;
-import de.saar.chorus.term.Constant;
-
-import java.io.Reader;
-import java.io.StringReader;
-import java.util.List;
-import java.util.ArrayList;
-import java.util.Map;
-
-/**
- * A parser for reading a plan that is output by the LAMA lanner and returning
- * it as a list of Atoms representing action instances.  
- * 
- * @author Daniel Bauer 
- *
- */
- public class LamaPlanParser implements LamaPlanParserConstants {
-
-        private String normalize(String s) {
-                        return s.toLowerCase();
-        }
-
-    public static List<Term> parse(String string, Map<String,String> normalizer) {
-        try {
-            LamaPlanParser p = new LamaPlanParser(new StringReader(string));
-            return p.plan();
-        } catch(ParseException e) {
-            System.err.println("Exception while parsing " + string + ": " + e);
-            return null;
-        }
-    }
-
-  final public List<Term> plan() throws ParseException {
-        List<Term> ret = new ArrayList<Term>();
-        Term atom;
-    label_1:
-    while (true) {
-      if (jj_2_1(2)) {
-        ;
-      } else {
-        break label_1;
-      }
-      jj_consume_token(7);
-      atom = planstep();
-                               ret.add(atom);
-    }
-                {if (true) return ret;}
-    throw new Error("Missing return statement in function");
-  }
-
-  final public Compound planstep() throws ParseException {
-        List<Term> arguments = new ArrayList<Term>();
-    Token predicate, arg;
-    predicate = jj_consume_token(IDENTIFIER);
-    label_2:
-    while (true) {
-      if (jj_2_2(2)) {
-        ;
-      } else {
-        break label_2;
-      }
-      arg = jj_consume_token(IDENTIFIER);
-                                                   arguments.add(new Constant(normalize(arg.image)));
-    }
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/evaluation/sgplanparser/ParseException.java

@@ -1,192 +1,0 @@
-/* Generated By:JavaCC: Do not edit this line. ParseException.java Version 3.0 */
-package crisp.evaluation.sgplanparser;
-
-/**
- * This exception is thrown when parse errors are encountered.
- * You can explicitly create objects of this exception type by
- * calling the method generateParseException in the generated
- * parser.
- *
- * You can modify this class to customize your error reporting
- * mechanisms so long as you retain the public fields.
- */
-public class ParseException extends Exception {
-
-  /**
-   * This constructor is used by the method "generateParseException"
-   * in the generated parser.  Calling this constructor generates
-   * a new object of this type with the fields "currentToken",
-   * "expectedTokenSequences", and "tokenImage" set.  The boolean
-   * flag "specialConstructor" is also set to true to indicate that
-   * this constructor was used to create this object.
-   * This constructor calls its super class with the empty string
-   * to force the "toString" method of parent class "Throwable" to
-   * print the error message in the form:
-   *     ParseException: <result of getMessage>
-   */
-  public ParseException(Token currentTokenVal,
-                        int[][] expectedTokenSequencesVal,
-                        String[] tokenImageVal
-                       )
-  {
-    super("");
-    specialConstructor = true;
-    currentToken = currentTokenVal;
-    expectedTokenSequences = expectedTokenSequencesVal;
-    tokenImage = tokenImageVal;
-  }
-
-  /**
-   * The following constructors are for use by you for whatever
-   * purpose you can think of.  Constructing the exception in this
-   * manner makes the exception behave in the normal way - i.e., as
-   * documented in the class "Throwable".  The fields "errorToken",
-   * "expectedTokenSequences", and "tokenImage" do not contain
-   * relevant information.  The JavaCC generated code does not use
-   * these constructors.
-   */
-
-  public ParseException() {
-    super();
-    specialConstructor = false;
-  }
-
-  public ParseException(String message) {
-    super(message);
-    specialConstructor = false;
-  }
-
-  /**
-   * This variable determines which constructor was used to create
-   * this object and thereby affects the semantics of the
-   * "getMessage" method (see below).
-   */
-  protected boolean specialConstructor;
-
-  /**
-   * This is the last token that has been consumed successfully.  If
-   * this object has been created due to a parse error, the token
-   * followng this token will (therefore) be the first error token.
-   */
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/evaluation/sgplanparser/SgPlanParser.java

@@ -1,380 +1,0 @@
-/* Generated By:JavaCC: Do not edit this line. SgPlanParser.java */
-package crisp.evaluation.sgplanparser;
-
-import de.saar.chorus.term.Term;
-import de.saar.chorus.term.Compound;
-import de.saar.chorus.term.Constant;
-
-import java.io.Reader;
-import java.io.StringReader;
-import java.util.List;
-import java.util.ArrayList;
-import java.util.Map;
-
-/**
- * A parser for reading a plan that is output by the SGPLAN planner and returning
- * it as a list of {@link give.formula.Atom}s representing action instances.
- *
- * @author Alexander Koller
- *
- */
- public class SgPlanParser implements SgPlanParserConstants {
-        private Map<String,String> normalizeUppercaseNames;
-
-        public SgPlanParser(Reader reader, Map<String,String> normalizeUppercaseNames) {
-                this(reader);
-
-                this.normalizeUppercaseNames = normalizeUppercaseNames;
-        }
-
-        private String normalize(String s) {
-                if( normalizeUppercaseNames.containsKey(s) ) {
-                        return normalizeUppercaseNames.get(s);
-                } else {
-                        return s;
-                }
-        }
-
-    public static List<Term> parse(String string, Map<String,String> normalizer) {
-        try {
-            SgPlanParser p = new SgPlanParser(new StringReader(string), normalizer);
-            return p.plan();
-        } catch(ParseException e) {
-            System.err.println("Exception while parsing " + string + ": " + e);
-            return null;
-        }
-    }
-
-  final public List<Term> plan() throws ParseException {
-        List<Term> ret = new ArrayList<Term>();
-        Term atom;
-    label_1:
-    while (true) {
-      if (jj_2_1(2)) {
-        ;
-      } else {
-        break label_1;
-      }
-      jj_consume_token(STEPNO);
-      jj_consume_token(10);
-      atom = planstep();
-      jj_consume_token(11);
-      jj_consume_token(COST);
-                                                   ret.add(atom);
-    }
-                {if (true) return ret;}
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/evaluation/sgplanparser/SgPlanParserConstants.java

@@ -1,29 +1,0 @@
-/* Generated By:JavaCC: Do not edit this line. SgPlanParserConstants.java */
-package crisp.evaluation.sgplanparser;
-
-public interface SgPlanParserConstants {
-
-  int EOF = 0;
-  int STEP = 6;
-  int STEPNO = 7;
-  int COST = 8;
-  int IDENTIFIER = 9;
-
-  int DEFAULT = 0;
-
-  String[] tokenImage = {
-    "<EOF>",
-    "\" \"",
-    "\"\\r\"",
-    "\"\\t\"",
-    "\"\\n\"",
-    "\";.*\\n\"",
-    "\"step\"",
-    "<STEPNO>",
-    "<COST>",
-    "<IDENTIFIER>",
-    "\"(\"",
-    "\")\"",
-  };
-
-}

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/evaluation/sgplanparser/SgPlanParserTokenManager.java

@@ -1,437 +1,0 @@
-/* Generated By:JavaCC: Do not edit this line. SgPlanParserTokenManager.java */
-package crisp.evaluation.sgplanparser;
-import de.saar.chorus.term.Term;
-import de.saar.chorus.term.Compound;
-import de.saar.chorus.term.Constant;
-import java.io.Reader;
-import java.io.StringReader;
-import java.util.List;
-import java.util.ArrayList;
-import java.util.Map;
-
-public class SgPlanParserTokenManager implements SgPlanParserConstants
-{
-  public  java.io.PrintStream debugStream = System.out;
-  public  void setDebugStream(java.io.PrintStream ds) { debugStream = ds; }
-private final int jjStopStringLiteralDfa_0(int pos, long active0)
-{
-   switch (pos)
-   {
-      case 0:
-         if ((active0 & 0x40L) != 0L)
-         {
-            jjmatchedKind = 9;
-            return 6;
-         }
-         return -1;
-      case 1:
-         if ((active0 & 0x40L) != 0L)
-         {
-            jjmatchedKind = 9;
-            jjmatchedPos = 1;
-            return 6;
-         }
-         return -1;
-      case 2:
-         if ((active0 & 0x40L) != 0L)
-         {
-            jjmatchedKind = 9;
-            jjmatchedPos = 2;
-            return 6;
-         }
-         return -1;
-      default :
-         return -1;
-   }
-}
-private final int jjStartNfa_0(int pos, long active0)
-{
-   return jjMoveNfa_0(jjStopStringLiteralDfa_0(pos, active0), pos + 1);
-}
-private final int jjStopAtPos(int pos, int kind)
-{
-   jjmatchedKind = kind;
-   jjmatchedPos = pos;
-   return pos + 1;
-}
-private final int jjStartNfaWithStates_0(int pos, int kind, int state)
-{
-   jjmatchedKind = kind;
-   jjmatchedPos = pos;
-   try { curChar = input_stream.readChar(); }
-   catch(java.io.IOException e) { return pos + 1; }
-   return jjMoveNfa_0(state, pos + 1);
-}
-private final int jjMoveStringLiteralDfa0_0()
-{
-   switch(curChar)
-   {
-      case 40:
-         return jjStopAtPos(0, 10);
-      case 41:
-         return jjStopAtPos(0, 11);
-      case 59:
-         return jjMoveStringLiteralDfa1_0(0x20L);
-      case 115:
-         return jjMoveStringLiteralDfa1_0(0x40L);
-      default :
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/evaluation/sgplanparser/SimpleCharStream.java

@@ -1,439 +1,0 @@
-/* Generated By:JavaCC: Do not edit this line. SimpleCharStream.java Version 4.0 */
-package crisp.evaluation.sgplanparser;
-
-/**
- * An implementation of interface CharStream, where the stream is assumed to
- * contain only ASCII characters (without unicode processing).
- */
-
-public class SimpleCharStream
-{
-  public static final boolean staticFlag = false;
-  int bufsize;
-  int available;
-  int tokenBegin;
-  public int bufpos = -1;
-  protected int bufline[];
-  protected int bufcolumn[];
-
-  protected int column = 0;
-  protected int line = 1;
-
-  protected boolean prevCharIsCR = false;
-  protected boolean prevCharIsLF = false;
-
-  protected java.io.Reader inputStream;
-
-  protected char[] buffer;
-  protected int maxNextCharInd = 0;
-  protected int inBuf = 0;
-  protected int tabSize = 8;
-
-  protected void setTabSize(int i) { tabSize = i; }
-  protected int getTabSize(int i) { return tabSize; }
-
-
-  protected void ExpandBuff(boolean wrapAround)
-  {
-     char[] newbuffer = new char[bufsize + 2048];
-     int newbufline[] = new int[bufsize + 2048];
-     int newbufcolumn[] = new int[bufsize + 2048];
-
-     try
-     {
-        if (wrapAround)
-        {
-           System.arraycopy(buffer, tokenBegin, newbuffer, 0, bufsize - tokenBegin);
-           System.arraycopy(buffer, 0, newbuffer,
-                                             bufsize - tokenBegin, bufpos);
-           buffer = newbuffer;
-
-           System.arraycopy(bufline, tokenBegin, newbufline, 0, bufsize - tokenBegin);
-           System.arraycopy(bufline, 0, newbufline, bufsize - tokenBegin, bufpos);
-           bufline = newbufline;
-
-           System.arraycopy(bufcolumn, tokenBegin, newbufcolumn, 0, bufsize - tokenBegin);
-           System.arraycopy(bufcolumn, 0, newbufcolumn, bufsize - tokenBegin, bufpos);
-           bufcolumn = newbufcolumn;
-
-           maxNextCharInd = (bufpos += (bufsize - tokenBegin));
-        }
-        else
-        {
-           System.arraycopy(buffer, tokenBegin, newbuffer, 0, bufsize - tokenBegin);
-           buffer = newbuffer;
-
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/evaluation/sgplanparser/Token.java

@@ -1,81 +1,0 @@
-/* Generated By:JavaCC: Do not edit this line. Token.java Version 3.0 */
-package crisp.evaluation.sgplanparser;
-
-/**
- * Describes the input token stream.
- */
-
-public class Token {
-
-  /**
-   * An integer that describes the kind of this token.  This numbering
-   * system is determined by JavaCCParser, and a table of these numbers is
-   * stored in the file ...Constants.java.
-   */
-  public int kind;
-
-  /**
-   * beginLine and beginColumn describe the position of the first character
-   * of this token; endLine and endColumn describe the position of the
-   * last character of this token.
-   */
-  public int beginLine, beginColumn, endLine, endColumn;
-
-  /**
-   * The string image of the token.
-   */
-  public String image;
-
-  /**
-   * A reference to the next regular (non-special) token from the input
-   * stream.  If this is the last token from the input stream, or if the
-   * token manager has not read tokens beyond this one, this field is
-   * set to null.  This is true only if this token is also a regular
-   * token.  Otherwise, see below for a description of the contents of
-   * this field.
-   */
-  public Token next;
-
-  /**
-   * This field is used to access special tokens that occur prior to this
-   * token, but after the immediately preceding regular (non-special) token.
-   * If there are no such special tokens, this field is set to null.
-   * When there are more than one such special token, this field refers
-   * to the last of these special tokens, which in turn refers to the next
-   * previous special token through its specialToken field, and so on
-   * until the first special token (whose specialToken field is null).
-   * The next fields of special tokens refer to other special tokens that
-   * immediately follow it (without an intervening regular token).  If there
-   * is no such token, this field is null.
-   */
-  public Token specialToken;
-
-  /**
-   * Returns the image.
-   */
-  public String toString()
-  {
-     return image;
-  }
-
-  /**
-   * Returns a new Token object, by default. However, if you want, you
-   * can create and return subclass objects based on the value of ofKind.
-   * Simply add the cases to the switch for all those special cases.
-   * For example, if you have a subclass of Token called IDToken that
-   * you want to create if ofKind is ID, simlpy add something like :
-   *
-   *    case MyParserConstants.ID : return new IDToken();
-   *
-   * to the following switch statement. Then you can cast matchedToken
-   * variable to the appropriate type and use it in your lexical actions.
-   */
-  public static final Token newToken(int ofKind)
-  {
-     switch(ofKind)
-     {
-       default : return new Token();
-     }
-  }
-
-}

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/evaluation/sgplanparser/TokenMgrError.java

@@ -1,133 +1,0 @@
-/* Generated By:JavaCC: Do not edit this line. TokenMgrError.java Version 3.0 */
-package crisp.evaluation.sgplanparser;
-
-public class TokenMgrError extends Error
-{
-   /*
-    * Ordinals for various reasons why an Error of this type can be thrown.
-    */
-
-   /**
-    * Lexical error occured.
-    */
-   static final int LEXICAL_ERROR = 0;
-
-   /**
-    * An attempt wass made to create a second instance of a static token manager.
-    */
-   static final int STATIC_LEXER_ERROR = 1;
-
-   /**
-    * Tried to change to an invalid lexical state.
-    */
-   static final int INVALID_LEXICAL_STATE = 2;
-
-   /**
-    * Detected (and bailed out of) an infinite loop in the token manager.
-    */
-   static final int LOOP_DETECTED = 3;
-
-   /**
-    * Indicates the reason why the exception is thrown. It will have
-    * one of the above 4 values.
-    */
-   int errorCode;
-
-   /**
-    * Replaces unprintable characters by their espaced (or unicode escaped)
-    * equivalents in the given string
-    */
-   protected static final String addEscapes(String str) {
-      StringBuffer retval = new StringBuffer();
-      char ch;
-      for (int i = 0; i < str.length(); i++) {
-        switch (str.charAt(i))
-        {
-           case 0 :
-              continue;
-           case '\b':
-              retval.append("\\b");
-              continue;
-           case '\t':
-              retval.append("\\t");
-              continue;
-           case '\n':
-              retval.append("\\n");
-              continue;
-           case '\f':
-              retval.append("\\f");
-              continue;
-           case '\r':
-              retval.append("\\r");
-              continue;
-           case '\"':
-              retval.append("\\\"");
-              continue;
-           case '\'':
-              retval.append("\\\'");
-              continue;
-           case '\\':
-              retval.append("\\\\");
-              continue;
-           default:
-              if ((ch = str.charAt(i)) < 0x20 || ch > 0x7e) {
-                 String s = "0000" + Integer.toString(ch, 16);
-                 retval.append("\\u" + s.substring(s.length() - 4, s.length()));
-              } else {
-                 retval.append(ch);
-              }
-              continue;
-        }
-      }
-      return retval.toString();
-   }
-
-   /**
-    * Returns a detailed message for the Error when it is thrown by the
-    * token manager to indicate a lexical error.
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/main/BasicCrispGenerator.java

@@ -1,222 +1,0 @@
-package crisp.main;
-
-import crisp.converter.CurrentNextCrispConverter;
-import crisp.converter.FastCRISPConverter;
-import crisp.planner.external.FfPlannerInterface;
-import crisp.planner.external.PlannerInterface;
-import crisp.planningproblem.Action;
-import crisp.planningproblem.Domain;
-import crisp.planningproblem.Problem;
-import crisp.result.CrispDerivationTreeBuilder;
-import crisp.result.DerivationTreeBuilder;
-import de.saar.chorus.term.Term;
-import de.saar.penguin.tag.codec.CrispXmlInputCodec;
-import de.saar.penguin.tag.codec.ParserException;
-import de.saar.penguin.tag.derivation.DerivationTree;
-import de.saar.penguin.tag.derivation.DerivedTree;
-import de.saar.penguin.tag.grammar.CrispGrammar;
-import de.saar.penguin.tag.grammar.Grammar;
-import de.saar.penguin.tag.grammar.SituatedCrispXmlInputCodec;
-
-import java.io.File;
-import java.io.FileNotFoundException;
-import java.io.FileReader;
-import java.io.Reader;
-import java.io.IOException;
-import java.io.StringReader;
-import java.util.ArrayList;
-import java.util.List;
-
-/**
- * A simple implementation of the @{CRISPGeneratorInterface} using the current state
- * of the standard CRISP generation system (with "steps" andwithout pragmatics or
- * context or probabilistic grammars). It also creates a new PDDL domain for every
- * problem instance and thus does not benefit from Konstantinas optimizations.
- * <p/>
- * Use this class only through the @{CRISPGeneratorInterface} as it is likely to be modified
- * very soon.
- *
- * @author dbauer
- */
-public class BasicCrispGenerator implements CrispGeneratorInterface {
-
-    CrispGrammar grammar = null;
-    CurrentNextCrispConverter converter;
-    PlannerInterface planner;
-    DerivationTreeBuilder planDecoder = null;
-    private List<Term> plan = new ArrayList<Term>();
-    private Domain domain;
-    private Problem problem;
-
-    public BasicCrispGenerator(String ffFlags) {
-        converter = new CurrentNextCrispConverter();
-        planner = new FfPlannerInterface(ffFlags);
-    }
-
-    public BasicCrispGenerator() {
-	this("");
-    }
-
-    // *** Methods to load the grammar ***
-
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/main/CrispGeneratorInterface.java

@@ -1,37 +1,0 @@
-package crisp.main;
-
-import de.saar.chorus.term.Term;
-import de.saar.penguin.tag.codec.ParserException;
-import de.saar.penguin.tag.derivation.DerivationTree;
-import de.saar.penguin.tag.grammar.Grammar;
-import java.io.IOException;
-import java.io.Reader;
-import java.util.List;
-
-/**
- * An interface that makes it easy to use a CRISP generation system
- * as part of other java projects (e.g. a GIVE NLG Server).
- * @author dbauer
- */
-public interface CrispGeneratorInterface {
-
-    /**
-     * Preset a CRISP domain which is used to solve all generation problems.
-     * @param xmlDomainReader A reader that accesses a CRISP grammar specification in XML format.
-     */
-    public void setGrammar(Grammar<Term> grammar);
-    public void setGrammar(Reader xmlGrammarReader) throws ParserException, IOException;
-    public void setGrammar(String xmlGrammar) throws ParserException;
-
-    /**
-     * Solve a generation problem using a preset domain. 
-     * @param xmlProblemReader
-     * @return The output sentence as a String
-     */
-    public String generateSentence(Reader xmlProblemReader) throws CrispGeneratorException;
-    public String generateSentence(String xmlProblem) throws CrispGeneratorException;
-    public List<String> generateWordSequence(Reader xmlProblemReader) throws CrispGeneratorException;
-    public List<String> generateWordSequence(String xmlProblem) throws CrispGeneratorException;
-    public DerivationTree generate(Reader xmlProblemReader) throws CrispGeneratorException;
-    public DerivationTree generate(String xmlProblem) throws CrispGeneratorException;
-}

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/main/GenerateSentence.java

@@ -1,93 +1,0 @@
-package crisp.main;
-
-import crisp.converter.CurrentNextCrispConverter;
-import crisp.planner.external.FfPlannerInterface;
-import crisp.planningproblem.Domain;
-import crisp.planningproblem.Problem;
-import de.saar.chorus.term.Term;
-import de.saar.penguin.tag.codec.ParserException;
-import de.saar.penguin.tag.grammar.CrispGrammar;
-import de.saar.penguin.tag.grammar.SituatedCrispXmlInputCodec;
-import java.io.File;
-import java.io.FileReader;
-import java.io.IOException;
-import java.util.List;
-import java.util.Properties;
-import javax.xml.parsers.ParserConfigurationException;
-
-
-public class GenerateSentence {       
-
-        public static void usage(){
-                    System.out.println("crisp.main.GenerateSentence [grammar file] [problem file]");
-        }
-
-	public static void main(String[] args) throws Exception {
-
-        Properties crispProps = new Properties();
-        crispProps.load(new FileReader(new File(Generate.PROPERTIES_FILE)));
-
-        if (args.length != 2) {
-            System.err.println("Wrong number of arguments.");
-            usage();
-            System.exit(1);
-        }
-
-         Domain domain = new Domain();
-        Problem problem = new Problem();
-
-        /// read CRISP problem specification and convert it to PDDL domain/problem
-        long start = System.currentTimeMillis();
-
-        SituatedCrispXmlInputCodec codec = new SituatedCrispXmlInputCodec();
-        CrispGrammar grammar = new CrispGrammar();
-        try {
-            codec.parse(new FileReader(new File(args[0])), grammar);
-        } catch (ParserException ex) {
-            System.err.println("Could not parse grammar File. Exiting.");
-            System.exit(1);
-        } catch (IOException ex) {
-            System.err.println("Could not read grammar File " + args[0] + ". Exiting.");
-            System.exit(1);
-        }
-        System.out.println("Grammar parsed in " + (System.currentTimeMillis() - start) + "ms .");
-
-        File problemfile = new File(args[1]);
-        try {
-            new CurrentNextCrispConverter().convert(grammar, problemfile, domain, problem);
-        } catch (ParserConfigurationException ex) {
-            System.err.println("Error configuring the XML parser. Exiting.");
-            System.exit(1);        
-        } catch (IOException ex) {
-            System.err.println("Could not read problem File " + args[1] + ". Exiting.");
-        }
-
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/planner/external/FfPlannerInterface.java

@@ -1,221 +1,0 @@
-package crisp.planner.external;
-
-
-import crisp.planner.external.PlannerInterface;
-import crisp.evaluation.ffplanparser.ParseException;
-
-import java.io.FileInputStream;
-import java.io.FileWriter;
-import java.io.File;
-import java.io.IOException;
-import java.io.Reader;
-import java.io.BufferedReader;
-import java.io.InputStreamReader;
-
-
-import crisp.converter.ProbCRISPConverter;
-import crisp.planningproblem.Domain;
-import crisp.planningproblem.Problem;
-import crisp.planningproblem.codec.PddlOutputCodec;
-
-import crisp.evaluation.ffplanparser.FfPlanParser;
-
-import crisp.result.CrispDerivationTreeBuilder;
-import crisp.result.DerivationTreeBuilder;
-
-import de.saar.penguin.tag.grammar.ProbabilisticGrammar;
-import de.saar.penguin.tag.codec.PCrispXmlInputCodec;
-import de.saar.penguin.tag.derivation.DerivationTree;
-import de.saar.penguin.tag.derivation.DerivedTree;
-
-import de.saar.chorus.term.Term;
-
-import java.io.FileReader;
-import java.io.PrintWriter;
-import java.io.StringReader;
-import java.io.StringWriter;
-import java.util.List;
-import java.util.Properties;
-import java.util.regex.Matcher;
-import java.util.regex.Pattern;
-
-//import javax.swing.JFrame;
-//import org.jgraph.JGraph;
-
-
-public class FfPlannerInterface implements PlannerInterface {
-
-    public final String FF_BIN = "/proj/penguin/planners/FF-v2.3/new-ff-mac";
-
-    public static final String TEMPDOMAIN_FILE = "/tmp/tmpdomain.lisp";
-    public static final String TEMPPROBLEM_FILE = "/tmp/tmpproblem.lisp";
-    public static final String TEMPRESULT_FILE = "/tmp/tmpresult";
-
-    private String ffFlags;
-
-    private long preprocessingTime;
-    private long searchTime;
-    private String binaryLocation;
-
-    public FfPlannerInterface(String ffFlags) {
-        this.ffFlags = ffFlags;
-        preprocessingTime = 0;
-        searchTime = 0;
-
-        try {
-            Properties crispProps = new Properties();
-            crispProps.load(new FileReader(new File(crisp.main.Generate.PROPERTIES_FILE)));
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/planner/external/InterruptScheduler.java

@@ -1,17 +1,0 @@
-package crisp.planner.external;
-
-import java.util.TimerTask;
-
-class InterruptScheduler extends TimerTask { 
- 
-    Thread target = null;
-    public InterruptScheduler(Thread target)
-    {
-        this.target = target;
-    }
-    @Override
-    public void run()
-    {
-        target.interrupt();
-    }
-}

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/planner/external/LazyFfInterface.java

@@ -1,282 +1,0 @@
-package crisp.planner.external;
-
-
-import crisp.converter.FastCRISPConverter;
-import crisp.planner.external.PlannerInterface;
-import java.io.File;
-import java.io.FileReader;
-
-
-import crisp.planningproblem.Domain;
-import crisp.planningproblem.Problem;
-import crisp.planningproblem.codec.CostPddlOutputCodec;
-import crisp.planningproblem.codec.OutputCodec;
-
-
-import crisp.pddl.PddlParser;
-import crisp.planner.lazyff.FindAllBFS;
-import crisp.planner.lazyff.GoalStateCondition;
-import crisp.planner.lazyff.HelpfulActionFinder;
-import crisp.planner.lazyff.RelaxedGraphplanEvaluator;
-import crisp.planner.lazyff.BestFirstSearch;
-import crisp.planner.lazyff.Search;
-import crisp.planner.lazyff.State;
-import crisp.planningproblem.Action;
-import crisp.planner.reachability.ReachabilityAnalyzer;
-import crisp.planner.reachability.GroundPlanningProblem;
-import crisp.planningproblem.codec.PddlOutputCodec;
-import crisp.planningproblem.codec.TempPddlOutputCodec;
-import crisp.result.CrispDerivationTreeBuilder;
-import crisp.result.PCrispDerivationTreeBuilder;
-import crisp.result.DerivationTreeBuilder;
-
-import de.saar.penguin.tag.codec.PCrispXmlInputCodec;
-import de.saar.penguin.tag.derivation.DerivationTree;
-import de.saar.penguin.tag.derivation.DerivedTree;
-
-import de.saar.chorus.term.Term;
-
-import de.saar.penguin.tag.grammar.Grammar;
-import java.io.PrintWriter;
-import java.io.StringReader;
-import java.io.StringWriter;
-import java.util.ArrayList;
-
-import java.util.List;
-import java.util.Timer;
-
-//import javax.swing.JFrame;
-//import org.jgraph.JGraph;
-
-
-public class LazyFfInterface implements PlannerInterface {
-
-
-    private class SearchRunnable implements Runnable {
-
-        State initState;
-        State result;
-        GoalStateCondition condition;
-        Search search;
-
-        public SearchRunnable(Search search, State initState, GoalStateCondition condition) {
-            this.initState = initState;
-            this.condition = condition;
-            this.search = search;
-        }
-      
-        public void run() {
-            result = search.search(initState, condition);
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/planningproblem/codec/CostPddlOutputCodec.java

@@ -1,169 +1,0 @@
-package crisp.planningproblem.codec;
-
-import java.io.PrintWriter;
-
-
-import crisp.planningproblem.Action;
-import crisp.planningproblem.Domain;
-import crisp.planningproblem.Problem;
-import crisp.planningproblem.TypeHierarchy;
-import crisp.planningproblem.formula.Conditional;
-import crisp.planningproblem.formula.Conjunction;
-import crisp.planningproblem.formula.Formula;
-import crisp.planningproblem.formula.Literal;
-import crisp.planningproblem.formula.Negation;
-import crisp.planningproblem.formula.Universal;
-import de.saar.basic.StringTools;
-import de.saar.chorus.term.Term;
-import java.util.List;
-
-public class CostPddlOutputCodec extends PddlOutputCodec {
-    
-
-    @Override
-    public void writeDomain(Domain domain, PrintWriter dw) {
-        dw.println("(define (domain " + domain.getName() + ")");
-        dw.println("        (:requirements " + StringTools.join(domain.getRequirements(), " ") + " :action-costs )");        
-        dw.print("        (:types");
-        for( String type : domain.getTypeHierarchy().getTypes() ) {
-            if( !type.equals(TypeHierarchy.TOP) ) {
-                dw.print("  " + type);
-                if( ! domain.getTypeHierarchy().getDirectSupertype(type).equals(TypeHierarchy.TOP) ) {
-                    dw.print(" - " + domain.getTypeHierarchy().getDirectSupertype(type));
-                }
-            }
-        }
-        dw.println(")");
-        
-        dw.println("       (:constants");
-        for( String constant : domain.getUniverse().keySet()) {
-            dw.println("         " + constant + " - " + domain.getUniverse().get(constant));
-        }
-        dw.println("       )");
-        
-        dw.println("       (:predicates");
-
-        for (String pred : domain.getPredicates()) {
-            if (!pred.equals("**equals**")) {
-                dw.println("         (" + pred + " " + makeLispString(domain.getSignature(pred)) + ")");
-            }
-        }
-        dw.println("        )");
-        
-        dw.println("        (:functions (total-cost) - number)\n");
-
-        for( Action a : domain.getActions() ) {
-            dw.println("\n" + toPddlString(a));
-        }
-        
-        dw.println(")");
-        dw.flush(); //otherwise output is incomplete 
-    }
-    
-
-    protected String toPddlString(Action action) {
-        StringBuffer buf = new StringBuffer();
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/planningproblem/codec/TempPddlOutputCodec.java

@@ -1,143 +1,0 @@
-/*
- * To change this template, choose Tools | Templates
- * and open the template in the editor.
- */
-
-package crisp.planningproblem.codec;
-
-import crisp.planningproblem.Action;
-import crisp.planningproblem.Domain;
-import crisp.planningproblem.TypeHierarchy;
-import crisp.planningproblem.formula.Conditional;
-import crisp.planningproblem.formula.Conjunction;
-import crisp.planningproblem.formula.Formula;
-import crisp.planningproblem.formula.Literal;
-import crisp.planningproblem.formula.Negation;
-import crisp.planningproblem.formula.Universal;
-import de.saar.basic.StringTools;
-import de.saar.chorus.term.Term;
-import java.io.PrintWriter;
-import java.util.List;
-
-/**
- *
- * @author dbauer
- */
-public class TempPddlOutputCodec extends PddlOutputCodec{
-
-    public void writeDomain(Domain domain, PrintWriter dw) {
-        dw.println("(define (domain " + domain.getName() + ")");
-        dw.println("        (:requirements " + StringTools.join(domain.getRequirements(), " ") + " :durative-actions )");
-        dw.print("        (:types");
-        for( String type : domain.getTypeHierarchy().getTypes() ) {
-            if( !type.equals(TypeHierarchy.TOP) ) {
-                dw.print("  " + type);
-                if( ! domain.getTypeHierarchy().getDirectSupertype(type).equals(TypeHierarchy.TOP) ) {
-                    dw.print(" - " + domain.getTypeHierarchy().getDirectSupertype(type));
-                }
-            }
-        }
-        dw.println(")");
-
-        dw.println("       (:constants");
-        for( String constant : domain.getUniverse().keySet()) {
-            dw.println("         " + constant + " - " + domain.getUniverse().get(constant));
-        }
-        dw.println("       )");
-
-        dw.println("       (:predicates");
-        for (String pred : domain.getPredicates()) {
-            if (!pred.equals("**equals**")) {
-                dw.println("         (" + pred + " " + makeLispString(domain.getSignature(pred)) + ")");
-            }
-        }
-        dw.println("        )");
-        
-        for( Action a : domain.getActions() ) {
-            dw.println("\n" + toPddlString(a));
-        }
-
-        dw.println(")");
-        dw.flush(); //otherwise output is incomplete
-    }
-
-    protected String toPddlString(Action action) {
-        StringBuffer buf = new StringBuffer();
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/result/CrispDerivationTreeBuilder.java

@@ -1,229 +1,0 @@
-package crisp.result;
-
-import de.saar.penguin.tag.derivation.DerivationTree;
-import de.saar.penguin.tag.grammar.Grammar;
-import de.saar.penguin.tag.grammar.LexiconEntry;
-import de.saar.penguin.tag.grammar.ElementaryTree;
-
-import crisp.planningproblem.Domain;
-import crisp.planningproblem.Action;
-
-
-import de.saar.chorus.term.Term;
-import de.saar.chorus.term.Constant;
-import de.saar.chorus.term.Compound;
-
-import java.util.Collections;
-import java.util.List;
-import java.util.Map;
-import java.util.HashMap;
-
-
-public class CrispDerivationTreeBuilder extends DerivationTreeBuilder {
-
-    public static final String DEFAULT_ROOT_CATEGORY = "s";
-
-    private class Site {
-
-        public String derivationNode;
-        //public String tree;
-        //public String word;
-        public String treeNode;
-        public String cat;
-
-        public Site(String derivationNode, String treeNode, String cat) {
-            this.derivationNode = derivationNode;
-            //this.tree = tree;
-            //this.word = word;
-            this.treeNode = treeNode;
-            this.cat = cat;
-        }
-    }
-
-    private Map<String, Site> substitutionSites;
-    private Map<String, Site> adjunctionSites;
-
-    public CrispDerivationTreeBuilder(Grammar<Term> grammar) {
-        super(grammar);
-
-    }
-
-
-    // TODO - this looks like a terrible hack. What if there are two substitution nodes with the same category?? - ak, jan 10
-    private String findNodeWithCat(String cat, ElementaryTree t) {
-        for (String node : t.getAllNodes()) {
-            if (t.getNodeLabel(node).equals(cat)) {
-                return node;
-            }
-        }
-        return null;
-
-    }
-
-    private String findLeafWithCat(String cat, ElementaryTree t) {
-        for (String node : t.getAllNodes()) {
-            if (t.getNodeLabel(node).equals(cat) && t.getChildren(node).isEmpty()) {
-                return node;
-            }
-        }
-        return null;
-
-    }
-
-
-    private void addNewSubstAndAdjSites(Action action, ElementaryTree tree, String derivationNode) {
-
-        //System.out.println("processing action " + action );
-        List<Compound> substEffects = getSubstEffects(action);
-        List<Compound> adjEffects = getAdjEffects(action);
-
-
-        for (Compound c : substEffects) {
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/result/DerivationTreeBuilder.java

@@ -1,145 +1,0 @@
-package crisp.result;
-
-import de.saar.penguin.tag.derivation.DerivationTree;
-import de.saar.penguin.tag.grammar.Grammar;
-
-import crisp.planningproblem.Domain;
-import crisp.planningproblem.Action;
-import crisp.planningproblem.formula.Formula;
-import crisp.planningproblem.formula.Literal;
-import crisp.planningproblem.formula.Conjunction;
-
-
-import crisp.converter.TagActionType;
-
-import de.saar.chorus.term.Term;
-import de.saar.chorus.term.Compound;
-import de.saar.chorus.term.Variable;
-import de.saar.chorus.term.Substitution;
-
-import java.util.List;
-import java.util.ArrayList;
-import java.util.Map;
-
-
-/**
- * Build @{de.saar.penguin.tag.derivation.DerivationTree}s from
- * a plan. 
- *
- * @author Daniel Bauer
- */
-public abstract class DerivationTreeBuilder{
-       
-    
-    protected Grammar<Term> grammar;
-    
-    protected Map<String,String> syntaxnodeToTreeId; // Store a mapping from tree#word#syntaxnode -> derivation 
-                                                       // tree tree ID.   
-    protected DerivationTree currentDerivation;
-    
-    public DerivationTreeBuilder(Grammar<Term> grammar) {
-        this.grammar = grammar;
-    }   
-            
-    /** 
-     * Retrieve the list of positive preconditions with a certain label from an instantiated action.
-     */
-    protected List<Compound> getPreconditionByLabel(Action action, String label) {
-        Conjunction preconds = (Conjunction) action.getPrecondition();        
-        List<Compound> result = new ArrayList<Compound>();
-        for (Formula conjunct : preconds.getConjuncts()) {
-            if (conjunct instanceof Literal){
-                Literal literal = (Literal) conjunct;
-                if (literal.getAtom() instanceof Compound){                
-                    // Take posisitive subst terms
-                    Compound comp = (Compound) literal.getAtom();
-                    if ((label.equals(comp.getLabel())) && literal.getPolarity()) {
-                        result.add(comp);
-                    }
-                }
-            }
-        }
-        return result;
-    }
-
-    protected List<Compound> getSubstPreconditions(Action action) {
-        return this.getPreconditionByLabel(action, "subst");
-    }
-    
-    protected List<Compound> getAdjPreconditions(Action action) {
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/crisp/result/PCrispDerivationTreeBuilder.java

@@ -1,111 +1,0 @@
-package crisp.result;
-
-import de.saar.penguin.tag.derivation.DerivationTree;
-import de.saar.penguin.tag.grammar.Grammar;
-import de.saar.penguin.tag.grammar.LexiconEntry;
-
-import crisp.planningproblem.Domain;
-import crisp.planningproblem.Action;
-
-
-import crisp.converter.TagActionType;
-
-import de.saar.chorus.term.Term;
-import de.saar.chorus.term.Compound;
-
-import java.util.List;
-import java.util.HashMap;
-
-public class PCrispDerivationTreeBuilder extends DerivationTreeBuilder {
-
-    public PCrispDerivationTreeBuilder(Grammar<Term> grammar) {
-        super(grammar);
-    }   
-    
-    @Override
-    protected List<Compound> getAdjEffects(Action action) {
-        return this.getEffectsByLabel(action, "mustadjoin");
-    }
-  
-    @Override 
-    public void processPlanStep(Action action) {
-                  
-            Compound pred = (Compound) action.getPredicate();
-            List<Term> variables = pred.getSubterms();
-            List<String> variableTypes = action.getParameterTypes();
-
-            String[] predicateParts = pred.getLabel().split("-");                        
-            TagActionType operation = decodeActionType(predicateParts[0]);
-            String childTree = predicateParts[1];
-            String word = predicateParts[2];
-                
-            String parentTree = null;
-            String parentWord = null;
-            String parentNode = null;            
-            int step;                                  
-            
-            String syntaxnode = pred.getSubterms().get(0).toString();
-            
-            
-            if (operation == TagActionType.NO_ADJUNCTION) { // this is a null adjunction
-               // Ignore no adjunction operations for now
-            } else {
-                                                
-            
-                String treeId;
-                String snode;
-                String node;                
-                
-                if (operation == TagActionType.INIT ) { // this is an init operation                                
-                    step = Integer.valueOf(predicateParts[3]).intValue();                    
-                    LexiconEntry entry = grammar.getLexiconEntry(word,childTree);
-                    // Really add the derivation
-                    treeId = currentDerivation.addNode(null, null, childTree, entry);                                                            
-                                                                                
-                } else {// This is an adjunction or substitution
-                    parentTree = predicateParts[3];
-                    parentWord = predicateParts[4];
-                    parentNode = predicateParts[5];
[…]

### 2011-05-23 tags/crisp-1.1.5/src/main/java/de/saar/penguin/tag/grammar/CrispGrammar.java

@@ -1,52 +1,0 @@
-/*
- * To change this template, choose Tools | Templates
- * and open the template in the editor.
- */
-
-package de.saar.penguin.tag.grammar;
-
-import de.saar.chorus.term.Term;
-import java.util.ArrayList;
-import java.util.Collection;
-import java.util.HashSet;
-import java.util.List;
-import java.util.Map;
-
-/**
- * A TAG Grammar that contains @{CrispLexiconEntry}s.
- * @author dbauer
- */
-public class CrispGrammar extends Grammar<Term>{
-
-    @Override
-    public void addLexiconEntry(String word, String elementaryTreeName, Map<String, String> auxLexicalItems, List<Term> semantics) {
-        List<LexiconEntry> hereTrees = lexicon.get(word);
-
-        if (hereTrees == null) {
-            hereTrees = new ArrayList<LexiconEntry>();
-            lexicon.put(word, hereTrees);
-        }
-
-        hereTrees.add(new CrispLexiconEntry(word, elementaryTreeName, auxLexicalItems, semantics));
-    }
-
-    public void addLexiconEntry(CrispLexiconEntry entry) {
-        List<LexiconEntry> hereTrees = lexicon.get(entry.word);
-
-        if (hereTrees == null) {
-            hereTrees = new ArrayList<LexiconEntry>();
-            lexicon.put(entry.word, hereTrees);
-        }
-
-        hereTrees.add(entry);
-    }
-    
-    public List<CrispLexiconEntry> getCrispLexiconEntries(String word) {
-        List<CrispLexiconEntry> ret = new ArrayList<CrispLexiconEntry>();
-        for (LexiconEntry entry : lexicon.get(word)) {
-            ret.add((CrispLexiconEntry) entry);
-        }
-        return ret;
-    }
-
-}

### 2011-05-23 tags/crisp-1.1.5/src/test/java/crisp/converter/CurrentNextCrispConverterTest.java

@@ -1,144 +1,0 @@
-package crisp.converter;
-
-import static org.junit.Assert.*;
-
-import java.io.*;
-import java.util.ArrayList;
-import java.util.List;
-
-import javax.xml.parsers.ParserConfigurationException;
-
-import crisp.planner.external.FfPlannerInterface;
-import crisp.planner.external.PlannerInterface;
-import crisp.result.CrispDerivationTreeBuilder;
-import crisp.result.DerivationTreeBuilder;
-import de.saar.chorus.term.Term;
-import de.saar.penguin.tag.derivation.DerivationTree;
-import de.saar.penguin.tag.derivation.DerivedTree;
-import org.junit.Before;
-import org.junit.Test;
-import org.xml.sax.SAXException;
-
-import crisp.planningproblem.Action;
-import crisp.planningproblem.Domain;
-import crisp.planningproblem.Problem;
-import de.saar.penguin.tag.codec.ParserException;
-import de.saar.penguin.tag.grammar.CrispGrammar;
-import de.saar.penguin.tag.grammar.SituatedCrispXmlInputCodec;
-
-public class CurrentNextCrispConverterTest {
-    private Domain domain;
-    private Problem problem;
-
-    private DerivationTreeBuilder planDecoder;
-
-    private CrispGrammar grammar;
-
-    @Before
-    public void setUp() throws FileNotFoundException, ParserException, IOException, ParserConfigurationException, SAXException {
-        CurrentNextCrispConverter converter;
-        domain = new Domain();
-        problem = new Problem();
-
-        SituatedCrispXmlInputCodec codec = new SituatedCrispXmlInputCodec();
-        grammar = new CrispGrammar();
-        codec.parse(new InputStreamReader(getClass().getResourceAsStream("/grammar-scrisp.xml")), grammar);
-
-        Reader problemfile = new InputStreamReader(getClass().getResourceAsStream("/problem-scrisp-give1-generated.xml"));
-
-        converter = new CurrentNextCrispConverter();
-        converter.convert(grammar, problemfile, domain, problem);
-
-        planDecoder = new CrispDerivationTreeBuilder(grammar);
-    }
-
-    @Test
-    public void testConvertOnce() {
-        assert domain != null;
-        assert problem != null;
-    }
-
-    @Test
-    public void testConvertTwiceFreshGrammar() throws IOException, SAXException, ParserConfigurationException, ParserException {
-        Domain domain2 = new Domain();
[…]

### 2011-05-23 tags/crisp-1.1.5/src/test/java/crisp/result/CrispDerivationTreeBuilderTest.java

@@ -1,61 +1,0 @@
-package crisp.result;
-
-import crisp.converter.CurrentNextCrispConverter;
-import crisp.planner.external.FfPlannerInterface;
-import crisp.planner.external.PlannerInterface;
-import crisp.planningproblem.Domain;
-import crisp.planningproblem.Problem;
-import de.saar.chorus.term.Term;
-import de.saar.penguin.tag.codec.ParserException;
-import de.saar.penguin.tag.derivation.DerivationTree;
-import de.saar.penguin.tag.derivation.DerivedTree;
-import de.saar.penguin.tag.grammar.CrispGrammar;
-import de.saar.penguin.tag.grammar.SituatedCrispXmlInputCodec;
-import org.junit.Test;
-import org.xml.sax.SAXException;
-
-import javax.xml.parsers.ParserConfigurationException;
-import java.io.IOException;
-import java.io.InputStreamReader;
-import java.io.Reader;
-import java.util.List;
-
-import static org.junit.Assert.assertEquals;
-
-/**
- * Created by IntelliJ IDEA.
- * User: koller
- * Date: Jan 18, 2010
- * Time: 7:58:41 PM
- * To change this template use File | Settings | File Templates.
- */
-public class CrispDerivationTreeBuilderTest {
-    @Test
-    public void testModifierOrder() throws Exception, ParserException, SAXException, ParserConfigurationException {
-        CurrentNextCrispConverter converter;
-        Domain domain = new Domain();
-        Problem problem = new Problem();
-
-        SituatedCrispXmlInputCodec codec = new SituatedCrispXmlInputCodec();
-        CrispGrammar grammar = new CrispGrammar();
-        codec.parse(new InputStreamReader(getClass().getResourceAsStream("/modifiers-grammar.xml")), grammar);
-
-        Reader problemfile = new InputStreamReader(getClass().getResourceAsStream("/modifiers-problem.xml"));
-
-        converter = new CurrentNextCrispConverter();
-        converter.convert(grammar, problemfile, domain, problem);
-
-        CrispDerivationTreeBuilder planDecoder = new CrispDerivationTreeBuilder(grammar);
-        PlannerInterface planner = new FfPlannerInterface();
-        List<Term> plan = planner.runPlanner(domain, problem);
-        DerivationTree derivationTree = planDecoder.buildDerivationTreeFromPlan(plan, domain, "s");
-        DerivedTree derivedTree = derivationTree.computeDerivedTree(grammar);
-
[…]

### 2011-05-23 tags/crisp-1.1.6/tags/ant-crisp/Papers/crisp-rtg-09/crisp-rtg-09.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2009-06-17 trunk/Papers/crisp-rtg-09/crisp-rtg-09.tex
@@
 
-\section{Introduction}
+\section{Introduction} \label{sec:introduction}
+
+In \citet{KolSto07}, Matthew Stone and I showed how the sentence
+generation problem of TAG grammars can be encoded into a planning
+problem. A planner could then be used to compute a plan, and this plan
+could be decoded into a TAG derivation, and further into a sentence.
+
+In this manuscript, I generalize this result to sentence generation
+with regular tree grammars \citep{tata2007}.  Our older TAG encoding
+is indeed a special case of this because it is known that for each TAG
+grammar, the language of derivation trees forms a regular tree
+language \citep{schmitz08:_featur_in_tag_deriv_trees}. On the other
+hand, there are many other grammar formalisms whose derivational
+structures can be seen as regular tree languages -- including regular
+dependency grammars \citep{kuhlmann2007mildly}, linear context-free
+rewrite systems
+\citep{weir88:_charac_mildl_contex_sensit_gramm_formal}, and
+combinatory categorial grammars
+\citep{steedman01:_syntac_proces,KolKuh09}. What distinguishes these
+grammar formalisms from each other is how they assign word order to
+the trees in the regular tree language. But as long as we can read the
+trees we generate with our RTG as having each node decorated with
+lexicon entries of the underlying grammar formalism, we can leave it
+up to the grammar formalism to compute the word order (e.g., by
+computing a derived tree from a TAG derivation tree and reading the
+leaves left to right). As a consequence, the construction in this
+paper generalizes our sentence-generation-as-planning approach to a
+whole range of mildly context-sensitive grammar formalisms.
+
+The paper is structured as follows. In
+Section~\ref{sec:gener-probl-regul}, I will define the sentence
+generation problem of regular tree grammars and illustrate it with an
+example. In Section~\ref{sec:rtg-generation-as-planning}, I will
+sketch how to adapt the planning construction by \citet{KolSto07} to
+regular tree grammars. In Section~\ref{sec:np-completeness}, I prove
+that the sentence generation problem of RTGs is NP-complete, which
+justifies the use of a worst-case exponential algorithm like a planner
+for solving the problem. Section~\ref{sec:conclusion} concludes and
+points to future work.
 
@@
 \section{Conclusion}
+\label{sec:conclusion}
 

### 2011-05-23 tags/crisp-1.1.6/tags/ant-crisp/Papers/dl-gre-jair/related.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2008-05-16 trunk/Papers/dl-gre-08/related.tex
@@
 Furthermore, the non-relational but negative and disjunctive
-descriptions generated by \newcite{deemter02:_gener_refer_expres} are
+descriptions generated by \citeA{deemter02:_gener_refer_expres} are
 simply formulas of $\mathcal{PL}$; and
-\newcite{gardent02:_gener_minim_defin_descr} generalizes this into
+\citeA{gardent02:_gener_minim_defin_descr} generalizes this into
 generating formulas of $\mathcal{ELU}_{(\neg)}$, i.e., \el\ plus
@@
 subsets.  In this way, it is reminiscent of the Incremental Algorithm,
-which in fact can be seen as the special case of the \el\ algorithm.
-However, unlike
-\newcite{dale91:_gener_refer_expres_invol_relat} and its successors,
-such as \newcite{kelleher06:_increm_gener_of_spatial_refer},
-we do not have to take special precautions to avoid infinite
-regress. While Dale and Haddock's algorithm attempts to generate a RE
-for a single individual, for successive individuals in the model, our
-algorithms consider all individuals in parallel.  It monotonically
-refines a partition of the model and never needs to backtrack, and
-therefore is always guaranteed to terminate.
+which in fact can be seen as a special case of the \el\ algorithm.
+However, unlike \citeA{dale91:_gener_refer_expres_invol_relat} and
+its successors, such as
+\citeA{kelleher06:_increm_gener_of_spatial_refer}, we do not have to
+take special precautions to avoid infinite regress. While Dale and
+Haddock's algorithm attempts to generate a RE for a single individual,
+for successive individuals in the model, our algorithms consider all
+individuals in parallel.  It monotonically refines a partition of the
+model and never needs to backtrack, and therefore is always guaranteed
+to terminate.
 
@@
     GRE algorithm & DL variant \\ \hline
-    \newcite{Dale1995} & $\mathcal{CL}$ \\
-    \newcite{deemter02:_gener_refer_expres} & $\mathcal{PL}$ \\
-    \newcite{dale91:_gener_refer_expres_invol_relat} & \el \\
-    \newcite{kelleher06:_increm_gener_of_spatial_refer} & \el \\
-    \newcite{gardent02:_gener_minim_defin_descr} & $\mathcal{ELU}_{(\neg)}$\\
-%    \newcite{Krahmer2003} & \el\ + equality between individuals
+    \citeA{Dale1995} & $\mathcal{CL}$ \\
+    \citeA{deemter02:_gener_refer_expres} & $\mathcal{PL}$ \\
+    \citeA{dale91:_gener_refer_expres_invol_relat} & \el \\
+    \citeA{kelleher06:_increm_gener_of_spatial_refer} & \el \\
+    \citeA{gardent02:_gener_minim_defin_descr} & $\mathcal{ELU}_{(\neg)}$\\
+%    \citeA{Krahmer2003} & \el\ + equality between individuals
   \end{tabular}

### 2011-05-23 tags/crisp-1.1.6/tags/inlg-08-final/Papers/rabbit-from-hat/introduction.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

\section{Introduction}
\label{sec:introduction}

In the standard architecture of a natural language generation (NLG)
system, the problem of sentence generation is split into two separate
steps -- sentence planning and surface realization
\cite{reiter00building}.  However, it has been shown that by coupling
sentence planning and realization more closely, an NLG system such as
SPUD \cite{Stone2003a} can sometimes generate more succinct
expressions than a pipelined system could \cite{Stone1998a}.  In order
to express the instruction to remove the indicated rabbit from the
indicated hat in Fig.~\ref{fig:rabbits}, a pipelined system would have
to generate the sentence

\enumsentence{Take the rabbit in the hat from the hat
that contains a rabbit.\label{ex:rabbit1}}

  A system that can exploit the fact that the
two noun phrases are connected via the verb (i.e., which makes
realization information available while referring expressions are
generated) could instead generate the more succinct

\enumsentence{Take the rabbit from the hat. \label{ex:rabbit2}}

\newcite{KolSto07} presented an implementation of the core of SPUD as
a planning problem, and demonstrated that SPUD-style integrated
sentence planning and realization can be performed efficiently by
applying modern planning technology.  However, their approach is
incapable of generating (\ref{ex:rabbit2}) because it requires the
first NP to distinguish the rabbit from all other things that are
inside other things, and likewise for the second NP.

In this paper, we will show how to extend their approach to sentence
generation as planning in such a way that it is capable of generating
referring expressions that are as succinct as SPUD's.  The relevance
of this change goes beyond isolated examples such as
(\ref{ex:rabbit1}): It requires us to model in what way adding
information to one referring expression constraints the distractors of
another.  In a first step, we will show how the Koller and Stone model
can be directly extended to take different divisions of labor between
the two referring expressions into account.  However, we will then
show that this leads to unacceptable runtimes.  The reason for this
slowdown is that the initial state and the universe considered by the
planner grow, and that the planner must make decisions about how to
distribute distractors over the different REs top-down, rather than
letting the different REs constrain each other.

In a second step, we will then present a novel approach to generation
as planning in which the planning system (in our case, PKS
\cite{Petrick-Bacchus:02}) is allowed to access external reasoning
modules.  We now keep track of a constraint network representing the
distractors of all REs \cite{dale91:_gener_refer_expres_invol_relat}
in a separate module; the planner can add new information to this
module and query it.  This allows us to keep the planning state
manageable and delay the decision about the distractors for each RE
until these REs are actually being built.  We evaluate the runtime of
our system \todo{blah}.  Attaching external reasoning modules to a
planner is a novel technique in planning, and far more general than
just keeping track of accessing a constraint network; this increase in
expressive power will be very useful for other applications of
planning-based generation systems in the future.

\todo{need to explain how distractor sets for the REs are created}

The paper is structured as follows. We will first briefly review
planning (Section~\ref{sec:planning}) and Koller and Stone's
generation as planning approach (Section~\ref{sec:gen-as-planning}).
Then we will discuss the examples (\ref{ex:rabbit1}) and
(\ref{ex:rabbit2}) in more detail and present the first, failed
attempt to computing (\ref{ex:rabbit1}) by planning, in
Section~\ref{sec:problem}.  Section~\ref{sec:howto} then describes how
we get a more efficient encoding of the generation problem by
attaching external modules to the planner.  Finally,
Section~\ref{sec:evaluation} evaluates the performance of our system,
and Section~\ref{sec:conclusion} concludes.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "rabbit-from-hat"
%%% End:

### 2011-05-23 tags/crisp-1.1.6/tags/spark-08-submit/discussion.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2008-06-23 trunk/Papers/experiences/discussion.tex
@@
 development of algorithms that control search in sophisticated ways,
-resulting in a host of planners that are more powerful and more successful
-than their predecessors. Depending on the particular algorithm, however,
-these planners often ground out sets of predicates and actions before
-starting the search itself, which has a very pronounced effect on domains
-such as those described above. Since our domains are not that unusual in
-their structure and composition, we hope that the lessons learnt from our
-experiences can help improve the performance of current systems. We believe
-that such improvements are necessary if planning is to become a more mature
-technology that can offer tools to a wider community of users. At the end
-of the day, real-world users will care about the \emph{total} runtime of a
-planner, and this is more than just the search time.
+resulting in a host of planners that are more powerful and more
+successful than their predecessors. However, the common strategy of
+grounding out sets of predicates and actions in implementations of
+these algorithms has a very pronounced effect on domains such as those
+described above. Since our domains are not that unusual in their
+structure and composition, we hope that the lessons learnt from our
+experiences can help improve the performance of current systems. We
+believe that such improvements are necessary if planning is to become
+a more mature technology that can offer tools to a wider community of
+users. At the end of the day, real-world users will care about the
+\emph{total} runtime of a planner, and this is more than just the
+search time.
 %We offer our larger problem instances as future challenges for the

### 2011-05-23 tags/crisp-1.1.6/tags/spark-08-submit/domains.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2008-07-20 trunk/Papers/experiences/domains.tex
@@
 
-The process described above has clear parallels to planning: we
-manipulate a state by applying actions in order to achieve a goal. We
-can make this connection even more precise by translating the SPUD
-problem into a planning problem.\footnote{See
-  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
-  which this conversion is implemented.}  For instance,
-Figure~\ref{fig:white-rabbit-as-planning} shows the corresponding PDDL
-actions for the above generation task, where each action corresponds
-to an operation that adds a single elementary tree to the derivation.
-In each case, the first parameter of the action is a node name in the
-derivation tree, and the remaining parameters stand for the
-individuals to which the semantic roles will be instantiated.  The
-syntactic preconditions and effects are encoded using $\mathsf{open}$
-and $\mathsf{canadjoin}$ literals; the status of each referring
-expression is tracked using $\mathsf{distractor}$ literals. Notice
-that the action effects contain terms of the form $\mathsf{subj}(u)$,
-which construct new node names. In order to meet the syntactic
-requirements of PDDL, these terms can be eliminated by estimating an
-upper bound $n$ for the plan length, making $n$ copies of each action,
-ensuring that copy $i$ can only be applied in step $i$, and replacing
-the term $\mathsf{subj}(u)$ in an action copy by the constant
-$\mathsf{subj}_i$.
+The process described above has clear parallels to planning: we manipulate
+a state by applying actions in order to achieve a goal. We can make this
+connection even more precise by translating the SPUD problem into a
+planning problem. For instance, Figure~\ref{fig:white-rabbit-as-planning}
+shows the corresponding PDDL actions for the above generation task, where
+each action corresponds to an operation that adds a single elementary tree
+to the derivation.  In each case, the first parameter of the action is a
+node name in the derivation tree, and the remaining parameters stand for
+the individuals to which the semantic roles will be instantiated.  The
+syntactic preconditions and effects are encoded using $\mathsf{open}$ and
+$\mathsf{canadjoin}$ literals; the status of each referring expression is
+tracked using $\mathsf{distractor}$ literals. Notice that the action
+effects contain terms of the form $\mathsf{subj}(u)$, which construct new
+node names. In order to meet the syntactic requirements of PDDL, these
+terms can be eliminated by estimating an upper bound $n$ for the plan
+length, making $n$ copies of each action, ensuring that copy $i$ can only
+be applied in step $i$, and replacing the term $\mathsf{subj}(u)$ in an
+action copy by the constant $\mathsf{subj}_i$.
 
@@
 Using this plan, the grammatical derivation in
-Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
-generated sentence ``the white rabbit sleeps'', can be systematically
-reconstructed. Thus, we can solve the sentence generation problem via
-the detour through planning and bring current search heuristics for
-planning to bear on generation.
+Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the generated
+sentence, can be systematically reconstructed. Thus, we can solve the
+sentence generation problem via the detour through planning and bring
+current search heuristics for planning to bear on generation.
 
@@
 The object of the GIVE Challenge (``Generating Instructions in Virtual
-Environments''; Koller et al.\
-\citeyear{alexander07:_shared_task_propos}) is to build an NLG system
-which is able to produce natural-language instructions which will
-guide a human user in performing some task in a virtual
-environment. From an NLG perspective, GIVE makes for an interesting
-challenge because it is a theory-neutral task that exercises all
-components of an NLG system, and emphasizes the study of communication
-in a (simulated) physical environment. It also has the advantage that
-the user and the NLG system can be in physically different places, as
-long as the 3D client and the NLG system are connected over a
-network. This makes it possible to evaluate GIVE NLG systems on a
-large scale over the Internet. The first GIVE evaluation will take
-place in late 2008; currently eight research teams from five countries
-are working on developing systems to participate in the
-challenge.\footnote{See the GIVE website for more details about this
-  project: \url{http://www.give-challenge.org/}.}
+Environments''; Koller et al.\ \citeyear{alexander07:_shared_task_propos})
+is to build an NLG system which is able to produce natural-language
+instructions which will guide a human user in performing some task in a
+virtual environment. From an NLG perspective, GIVE makes for an
+interesting challenge because it is a theory-neutral task that exercises
+all components of an NLG system, and emphasizes the study of communication
+in a (simulated) physical environment. It also has the advantage that the
+user and the NLG system can be in physically different places, as long as
+the 3D client and the NLG system are connected over a network. This makes
+it possible to evaluate GIVE NLG systems on a large scale over the
+Internet. The first GIVE evaluation will take place in late 2008;
+currently eight research teams from five countries are working on
+developing systems to participate in the challenge.\footnote{See the GIVE
+ website for more details about this project:
+ \url{homepages.inf.ed.ac.uk/v1akolle/proj/give/}.}
 
@@
 evaluation score.  Having just pressed the button, it would be much
-better to say ``turn around and walk through the door''. That is,
-multiple action instances should be merged into a single instruction;
-plans must be \emph{summarised} in order to present them to the
-user. Conversely, it may also be necessary to express a single
+better to say ``turn around and walk through the door''; i.e.,
+multiple action instances should be merged into a single
+instruction. Conversely, it may also be necessary to express a single
 planning step with several instructions.  Having just entered the top

### 2011-05-23 tags/crisp-1.1.6/tags/spark-08-submit/introduction.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2008-07-20 trunk/Papers/experiences/introduction.tex
@@
 \cite{HoffmannNebel01} and SGPLAN
-\cite{hsu06:_new_featur_in_sgplan_for}---planners that are easily
-available, support expressive preconditions and effects out of the
-box, and have been successful on traditional benchmarks, and in the
-International Planning Competition (IPC)---on a range of problem
-instances from our planning domains.  We also compare these results
-against an ad-hoc Java implementation of GraphPlan \cite{Blum1997}.
-Our findings are mixed. On the one hand, it turns out that modern
-planners handle some of the search problems that arise in NLG quite
-easily (although large problem instances still remain a challenge).
-On the other hand, these same planners spend tremendous amounts of
-time on preprocessing. For instance, FF spends 90\% of its runtime in
-the sentence generation domain on grounding out literals and actions,
-and ends up slower than our version of GraphPlan that avoids
-grounding. For domains like ours, which are dominated by the number of
-actions and the universe size, rather than the combinatorics of the
-search problem, this observation suggests that the overall runtime of
-a planner could be improved by grounding more selectively. We
-therefore offer these domains as challenges for the planning
-community.
+\cite{hsu06:_new_featur_in_sgplan_for}---planners that have been successful
+on traditional benchmarks, and in the International Planning Competition
+(IPC)---on a range of problem instances from our planning domains.  We also
+compare these results against an ad-hoc Java implementation of GraphPlan
+\cite{Blum1997}.  Our findings are mixed. On the one hand, it turns out
+that modern planners handle some of the search problems that arise in NLG
+quite easily (although large problem instances still remain a challenge).
+On the other hand, these same planners spend tremendous amounts of time on
+preprocessing. For instance, FF spends 90\% of its runtime in the sentence
+generation domain on grounding out literals and actions, and ends up slower
+than our version of GraphPlan that avoids grounding. For domains like ours,
+which are dominated by the number of actions and the universe size, rather
+than the combinatorics of the search problem, this observation suggests
+that the overall runtime of a planner could be improved by grounding more
+selectively. We therefore offer these domains as challenges for the
+planning community.
 

### 2011-05-23 tags/crisp-1.1.6/trunk/Papers/ci-spark/abstract.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2009-06-02 trunk/Papers/ci-spark/abstract.tex
@@
   computational linguistics with a long tradition as an application
-  area of automated planning systems. While the connection between
-  NLG and planning has not received much attention in current
-  mainstream NLG research, several recent publications have sparked
-  a renewed interest in this area. In this paper, we investigate the
-  extent to which these new NLG approaches profit from the advances
-  in planner expressiveness and efficiency. Our findings are mixed.
-  While modern planners can readily handle the search problems that
-  arise in our NLG experiments, their overall runtime is often
-  dominated by the grounding step they perform as preprocessing.
-  Furthermore, small changes in the structure of a domain can
-  significantly shift the balance between search and preprocessing.
-  Overall, our experiments show that the off-the-shelf planners we
-  tested are unusably slow for nontrivial NLG problem instances. As
-  a result, we offer our domains and experiences as challenges for
-  the planning community.
+  area of automated planning systems. While things were relatively
+  quiet with the planning approach to NLG for a while, several recent
+  publications have sparked a renewed interest in this area. In this
+  paper, we investigate the extent to which these new NLG approaches
+  profit from the advances in planner expressiveness and
+  efficiency. Our findings are mixed.  While modern planners can
+  readily handle the search problems that arise in our NLG
+  experiments, their overall runtime is often dominated by the
+  grounding step they perform as preprocessing.  Furthermore, small
+  changes in the structure of a domain can significantly shift the
+  balance between search and preprocessing.  Overall, our experiments
+  show that the off-the-shelf planners we tested are unusably slow for
+  nontrivial NLG problem instances. As a result, we offer our domains
+  and experiences as challenges for the planning community.
 \end{abstract}

### 2011-05-23 tags/crisp-1.1.6/trunk/Papers/crisp-ff-10/introduction.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2010-02-27 trunk/Papers/crisp-ff-10/introduction.tex
@@
 achieve this quantum leap, all that is needed are a few minor
-implementation fixes and basic reconfiguration of search
-options. While these fixes certainly are no contribution to planning
-in their own right, we consider it a valuable -- and slightly
-unsettling -- lesson learned that such banalities can make the
-difference between success and failure. They can hold up progress in
-planning applications for years, unless users of a planner have direct
-access to its developer. We interpret this as illustrative of the
-importance of implementation details such as the preprocessor for the
-overall runtime, and as a strong call for automated configuration
-techniques, e.g.\ along the lines of Hutter et al.\
-\shortcite{HutHooStu07}.
+implementation fixes in FF's preprocessor, and basic reconfiguration
+of its search options. While these changes certainly are no
+contribution to planning in their own right, we consider it a valuable
+-- and slightly unsettling -- lesson learned that such banalities can
+make the difference between success and failure. They can hold up
+progress in planning applications for years, unless users of a planner
+have direct access to its developer. We interpret this as a strong
+call for more research on automated configuration techniques,
+e.g.\ along the lines of Hutter et al.\ \shortcite{HutHooStu07}. The
+detrimental effect of preprocessing details also sheds a light on the
+importance of this often neglected part of the planning process.
+
+%illustrative of the
+%importance of implementation details such as the preprocessor for the
+%overall runtime, and as a strong call for automated configuration
+%techniques, e.g.\ along the lines of Hutter et
+%al.\ \shortcite{HutHooStu07}.
 

### 2011-05-23 tags/crisp-1.1.6/trunk/Papers/experiences/experiments.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2008-07-20 trunk/Papers/experiences/experiments.tex
@@
 
-In the first experiment, we generate a series of sentence generation
+In the first experiment, we construct a series of sentence generation
 problems which require the planner to compute a plan representing the
@@
   \centering
-  \includegraphics[width=1\columnwidth]{pic-runtime-modifiers-with-sgplan}
+  \includegraphics[width=1\columnwidth]{graph-exp1}
+  %\includegraphics[width=1\columnwidth]{pic-runtime-modifiers-with-sgplan}
   \caption{Runtimes (in ms) in the sentence generation domain.  The
@@
 
-In fact, the time spent by FF on grounding is such that it is consistently
-outperformed by an ad-hoc Java implementation of GraphPlan which only
-computes instances of predicates and actions as they are discovered (i.e.,
-while the planning graph is being built)---despite the fact that FF is
-consistently much faster as far as pure search time is concerned.  Looking
-at the difference from another angle, FF's performance is much more
-sensitive to the domain size: if we fix $n=1$, FF takes 60 ms to compute a
-plan at $m=1$, but 2.4 seconds (for the same plan) at $m=10$; our GraphPlan
-implementation takes 30 ms at $m=1$ and still only requires 50 ms at
-$m=10$.  Conversely, GraphPlan's runtime grows much faster with the plan
-size (i.e., with growing values of $n$ for a fixed $m$).  Larger (but still
-realistically-sized) instances of the sentence generation problem are still
-problematic for the planners we tested.
+In our examples, the time spent by FF on grounding is such that it is
+consistently outperformed by an ad-hoc Java implementation of GraphPlan
+which only computes instances of predicates and actions as they are
+discovered (i.e., while the planning graph is being built)---despite the
+fact that FF is consistently much faster as far as pure search time is
+concerned.  Looking at this difference from another angle, FF's performance
+is much more sensitive to the domain size: if we fix $n=1$, FF takes 60 ms
+to compute a plan at $m=1$, but 2.4 seconds to compute the same plan at
+$m=10$; our GraphPlan implementation takes 30 ms at $m=1$ and still only
+requires 50 ms at $m=10$.  Conversely, GraphPlan's runtime grows much
+faster with the plan size (i.e., with growing values of $n$ for a fixed
+$m$).  Larger, but still realistically-sized, instances of the sentence
+generation problem are still problematic for the planners we tested.
 
@@
 more time to actually compute the plan---for instance, the optimal
-plan for the problem $n=40$ has a length of about 1600 steps.  As an
-example, the plan for $n=h=1$ looks as follows:
+plan for the problem $n=40$ has a length of about 1600 steps.  As a
+concrete example, the following is a plan for $n=h=1$:
 
 \begin{enumerate}
-\item $\mathsf{move}(\mathsf{pos\_1\_1},\mathsf{pos\_1\_2}, \mathsf{north})$
-\item $\mathsf{manipulate}\textsf{-}\mathsf{u1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_1\_2})$
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{north}, \mathsf{east})$
+\item $\mathsf{move}(\mathsf{pos\_1\_1},\mathsf{pos\_1\_2}, \mathsf{north})$,
+\item $\mathsf{manipulate}\textsf{-}\mathsf{u1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_1\_2})$,
+\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{north}, \mathsf{east})$,
 \item $\mathsf{move}(\mathsf{pos\_1\_2}, \mathsf{pos\_2\_2},
-  \mathsf{east})$
-\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{east}, \mathsf{south})$
-\item $\mathsf{move}(\mathsf{pos\_2\_2}, \mathsf{pos\_2\_1}, \mathsf{south})$
-\item $\mathsf{manipulate}\textsf{-}\mathsf{l1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_2\_1})$
+  \mathsf{east})$,
+\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{east}, \mathsf{south})$,
+\item $\mathsf{move}(\mathsf{pos\_2\_2}, \mathsf{pos\_2\_1}, \mathsf{south})$,
+\item $\mathsf{manipulate}\textsf{-}\mathsf{l1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_2\_1})$.
 \end{enumerate}
@@
   \centering
-  \includegraphics[width=1\columnwidth]{pic-runtime-buttons}
+  \includegraphics[width=1\columnwidth]{graph-exp2}
+  %\includegraphics[width=1\columnwidth]{pic-runtime-buttons}
   \caption{Runtimes (in seconds) of FF and SGPLAN on the minimal GIVE
@@
   \centering
-  \includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
+  \includegraphics[width=1\columnwidth]{graph-exp3}
+  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
   \caption{Runtimes (in seconds) of FF and SGPLAN on the GIVE worlds with junk

### 2011-05-23 tags/crisp-1.1.6/trunk/Papers/operator-splitting/equiv-definition-alex.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

\subsection{Definition: Alex}
\label{definition-alex}

My definition is inspired by existing work on comparing the generative
capacity of grammar formalisms \cite{miller99:_stron}.  The basic idea
is that in order to compare two parse trees (which may be in terms of
two different grammar formalisms, and therefore look quite different),
we interpret them both into the same algebra $A$.  We take the parse
trees to be equivalent if they map to the same object in $A$.  Two
grammars are equivalent if their parse tree languages map to the same
set of objects in $A$.  One example is classical weak equivalence, in
which $A$ is the string algebra and the interpretation function just
takes the yield of the parse trees; but clearly, more powerful
interpretation algebras can be used for a more fine-grained view on
equivalence.

\newcommand{\A}{{\mathcal A}}

The fundamental parallel I see is that plans are essentially like
parse trees, and planning problem instances are essentially like
grammars.  Let's say we have a planning domain $D$, and let $A =
\{a_1,\ldots,a_n\}$ be the set of all operator instances of $D$.  Then
a plan is a string in $A^*$. Note that $(A^*,\cdot,a_1,\ldots,a_n)$ is
an algebra where the $a_i$ are constants and $\cdot$ is binary
concatenation.  I write $\A_D$ for the string algebra I get from the
planning domain $D$.  I can now define plan equivalence as follows.

\begin{definition}
  Let $D$ and $D'$ be planning domains.  Let $I$ be some algebra, and
  let $h:\A_D \rightarrow I$ and $h':\A_{D'} \rightarrow I$ be algebra
  homomorphisms. Now if $\pi$ is a sequence of action instances of $D$
  and $\pi'$ is a sequence of action instances of $D'$, $\pi$ is
  equivalent to $\pi'$ with respect to $I,h,h'$ ($\pi \equiv \pi'$)
  iff $h(\pi) = h'(\pi')$.
\end{definition}

Note that equivalence of plans (i.e., sequences of action instances
that go from a specific initial state to a specific goal) is a special
case of this. It is straightforward to lift this definition to
planning tasks and domains:

\begin{definition}
  Let $t = \langle D,P \rangle$ and $t' = \langle D',P'\rangle$ be
  planning tasks, and let $plans(t)$ be the set of all plans for the task
  $t$. Then $t$ and $t'$ are equivalent if $h(plans(t)) =
  h'(plans(t'))$.

  Let $plans(D)$ be the union over all $plans(\langle D,P \rangle)$
  for all tasks $\langle D,P \rangle$ that can be formed with the
  signature defined by $D$.  Then we take two domains $D$ and $D'$ to
  be equivalent if $h(plans(D)) = h'(plans(D'))$.
\end{definition}

Let's look at some special cases to illustrate the definitions.

\newcommand{\states}{{\mathcal S}}
\newcommand{\partialfunctions}{{\mathcal PF}}

\begin{itemize}
\item The most trivial special case is to require plan equivalence to
  mean plan identity.  For this, we take $I = \A_D$ and $h$ and $h'$
  the identity function.
\item The other extreme case is a notion of plan equivalence in which
  all plans are equivalent.  Clearly, this can be achieved if we take
  $I$ to be an algebra that contains a single element $e$, and $h$ and
  $h'$ map all plans to $e$.
\item A more interesting case is Gabi's question about reordering of
  actions in a plan: We want to see two plans over the same domain as
  equivalent if they take us from the same initial states to the same
  goals, but we don't care how the actions are ordered inside the
  plans, as long as the ``meaning'' of the plan stays the same.  One
  way to approach this is to let $I$  be the algebra
  $\partialfunctions_D$ of partial functions $f:\states_D
  \rightsquigarrow \states_D$, where $\states_D$ is the set of all
  states specified by $D$.  If $a$ is an action instance, we say that
  $h(a)$ is defined on a given state $s$ if the preconditions of $a$
  are satisfied in $s$.  $h(a)(s)$ is then the state we get by
  executing $a$ in $s$.  Concatenation of action sequences is
  interpreted as composition of partial functions.  In this way, any
  action sequence is interpreted as the partial function that is
  defined if the action sequence can be executed in a state, and that
  modifies the state as the actions specify.
\item Finally, operator splitting.  This is basically like the
  previous point, except that one domain has a richer signature,
  containing predicates like ``processing-none'' and such.  Let's say
  that $D$ was the original domain, and $D'$ is the enriched domain
  for the split operators.  We can then interpret both the original
  plan $\pi$ and the operator-split plan $\pi'$ into
  $\partialfunctions_{D}$.  The homomorphism $h:\A_D
  \rightarrow \partialfunctions_D$ is as above.  The homomorphism
  $h':\A_{D'} \rightarrow \partialfunctions_D$ maps actions of $D'$
  into partial functions over $\states_D$ by ignoring all
  preconditions and effects that use predicates that exist in $D'$ but
  not in $D$.  In this way, the fact that plans over $D$ are shorter
  than the split-operator plans over $D'$ is irrelevant: Because $\pi$
  can be executed in the same states in which $\pi'$ can be executed,
  and the effects of the two plans are the same (as far as predicates
  of $D$ are concerned), the plans count as equivalent.
\end{itemize}

My definition proposal is quite different in character from those of
Carlos and Gabi. It is a model-theoretic rather than a computational
notion: Plan equivalence is defined in terms of an interpretation of
the plans into a third algebra, rather than in terms of a reduction.
However, I feel that this is actually an advantage because the
definition is more flexible and, as far as I can tell, simpler, and
relies on very standard concepts such as algebras and homomorphisms.

It is clear that we will typically be interested in algorithms that
transform one class of plans into another, such as the
operator-splitting algorithm in Section~\ref{initial-definition}.  But
this does not mean that equivalence itself must be defined
procedurally.  I would bet that any procedural definition of
equivalence that we can come up with can be cast in terms of my
definition, by selecting the appropriate interpretation algebra and
homomorphisms for which the reduction is sound and complete.  One
interesting twist is that Gabi's definition suggests that we may care
about runtime and space bounds for the reduction.  This is one point
for which it is not entirely clear to me how to capture it in my
world.  Perhaps we can talk about why we care about them.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "total"
%%% End:

### 2011-05-23 tags/crisp-1.1.6/trunk/Papers/operator-splitting/opsplit-definition-v2.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2010-10-04 trunk/Papers/operator-splitting/definition-v2.tex
@@
-\section{Initial Definition}
+\subsection{Initial Definition}
 \label{initial-definition}
@@
 arg1(A), arg2(B)'', del ``on(A B), processing-none''; and
-``move2(A,C)''with pre ``do-move2, clear(C), arg1(A)'', add ``on(A C),
+``move2(A,C)'' with pre ``do-move2, clear(C), arg1(A)'', add ``on(A C),
 processing-none, arg3(C)'', del ``clear(C), do-move2''.
@@
   detrimental. something to keep in mind.}
+
+One important remark concerning planner performance: splitting up
+operators in the above way may produce dead-ends. After the first
+operator is applied, we do not have $processing\mbox{-}none$ anymore
+and thus there is no chance of reaching the goal other than by
+completing the split, i.e., reaching an operator
+$do\mbox{-}o_k$. That, however, may not be possible. As of the first
+operator, we commit to a partial variable instantiation $i$ for the
+operator in question, in terms of the $arg_1(.), \dots, arg_l(.)$
+predicates. Any completion of the split -- at any point during the
+execution of the split -- must stick to $i$. It may be the case that
+some ``future'' precondition, i.e., a precondition included in a
+later-on part of the split, cannot be satisfied in the current state
+in conformance with $i$. As a simple example, if we enrich the
+blocksworld with an additional static predicate ``can-stack(x y)'' and
+allow ``move(A,B,C)'' only if ``can-stack(A C)'' holds, then in the
+above split ``move2(A,C)'' has the additional precondition
+``can-stack(A C)''. However, there may not exist ``C'' for which both
+``can-stack(A C)'' and ``clear(C)'' hold. In this case, upon
+application of ``move1(A,B)'' we're in a dead-end. However, if the
+planner is using a relaxed-plan based heuristic, then these dead-ends
+``do not do a lot of damage'', in the sense that for any such dead-end
+state $s$, no relaxed plan will exist and thus the heuristic value
+will be $\infty$ and the state will be immediately pruned. The simple
+reason for this is that $processing\mbox{-}none$ must be re-achieved
+also in the relaxed plan, and doing so is not possible because at some
+point along the split sequence no more actions will be applicable.
 

### 2011-05-23 tags/crisp-1.1.6/trunk/Papers/operator-splitting/total.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2010-10-04 trunk/Papers/operator-splitting/total.tex
@@
 \documentclass{llncs}
-\usepackage{times}
-\usepackage{helvet}
-\usepackage{courier}
 \usepackage{epsfig,graphics,latexsym}
 \usepackage{amsmath,amssymb,theorem,enumerate}
+\usepackage[draft]{fixme}
 
@@
 of operator parameters, is known to be one of the major obstacles.
+
+\joerg{NOTE: instead of selling this as a re-writing technique, we
+  could just as well sell it is a way to ``get complex operators
+  through common pre-processors'': after all, once the pre-processing
+  is through, inside the planner we could easily ``unsplit'' the
+  operators again, yielding exactly the same grounded representation
+  that we would have got from the original operators. ... the
+  advantage is that we will not suffer from any potential search
+  degradation, and can focus exclusively on reducing instantiation
+  time. the disadvantage of course is in cases where the grounded
+  representation is prohibitively big -- the number of ``original''
+  ground ops can be exponentially (in the number of op params) larger
+  than the number of ``splitted'' ground ops.}
 \end{abstract}
 
-\input{definition-v1}
+\setcounter{tocdepth}{4}
 
-\input{definition-v2}
+\tableofcontents
+
+\section{Operator Splitting}
+
+\input{opsplit-definition-v1}
+
+\input{opsplit-definition-v2}
+
+\section{Task/Domain Equivalence}
+
+\input{equiv-definition-carlos}
+
+\input{equiv-definition-alex}
+
+\input{equiv-definition-gabi}
 

### 2011-05-23 tags/crisp-1.1.6/trunk/Papers/pcrisp-10/crisp.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2010-05-21 trunk/Papers/pcrisp-10/crisp.tex
@@
 \label{sec:crisp}
-In this section we review the original non-statistical {\sc crisp} system \cite{kollerstone2007}. 
-Following {\sc spud} \cite{Stone2003a}, {\sc crisp} is based on a declarative description of the sentence generation problem using {\sc tag}.
-Given a knowledge base, a communicative goal and a grammar, we require to find a grammatical {\sc tag} derivation that is consistent with this knowledge base and satisfies a communicative goal.
+In this section we review the original non-statistical CRISP system \cite{kollerstone2007}. 
+Following SPUD \cite{Stone2003a}, CRISP is based on a declarative description of the sentence generation problem using TAG.
+Given a knowledge base, a communicative goal and a grammar, we require to find a grammatical TAG derivation that is consistent with this knowledge base and satisfies a communicative goal.
 A number of semantic and pragmatic constraints that must be satisfied by the solution can be added, for instance to enforce generation of unambiguous referring expressions.
@@
 \subsection{Sentence Generation in CRISP}
-Like {\sc spud}, {\sc crisp} uses an {\sc ltag} in which elementary trees are assigned semantic content.
-Each node in a {\sc crisp} elementary tree is associated with a semantic role.
+Like SPUD, CRISP uses an LTAG in which elementary trees are assigned semantic content.
+Each node in a CRISP elementary tree is associated with a semantic role.
 Semantic content is expressed as a set of literals, encoding relations between these roles.
@@
 For each substitution and adjunction, the semantic role associated with the role of the target node is unified with the `self' role of the child tree.
-For example, given the knowledge base $\{\f{cat}(c), \f{fish}(f_1), \f{raw}(f_1), \f{fish}(f_2), \f{eats}(e,c,f_1)\}$ and the grammar in \ref{fig:grammar}(a), {\sc crisp} could produce the derivation in \ref{fig:grammar}(b).
-Notice that {\sc crisp} can generate the unambiguous referring expression `the raw fish' for $f_1$, to distinguish it from $f_2$.
+For example, given the knowledge base $\{\f{cat}(c), \f{fish}(f_1), \f{raw}(f_1), \f{fish}(f_2), \f{eats}(e,c,f_1)\}$ and the grammar in \ref{fig:grammar}(a), CRISP could produce the derivation in \ref{fig:grammar}(b).
+Notice that CRISP can generate the unambiguous referring expression `the raw fish' for $f_1$, to distinguish it from $f_2$.
 
@@
 \label{ssec:crispdomain}
-Before we describe {\sc crisp}'s encoding of sentence generation as planning, we briefly review AI planning in general.
+Before we describe CRISP's encoding of sentence generation as planning, we briefly review AI planning in general.
 A planning state is a conjunction of first order literals describing relations between some individuals.
@@
 
-In {\sc crisp}, planning states correspond to partial {\sc tag} derivations and record open substitution and adjunction sites, semantic individuals associated with them, and parts of the communicative goal that have not yet been expressed.
+In CRISP, planning states correspond to partial TAG derivations and record open substitution and adjunction sites, semantic individuals associated with them, and parts of the communicative goal that have not yet been expressed.
 The initial state also encodes the knowledge base and the communicative goal.
@@
 \end{center}
-\caption{\label{fig:crisp-operators} {\sc crisp} operators for some of the elementary trees in Fig.~\ref{fig:grammar}.}
+\caption{\label{fig:crisp-operators} CRISP operators for some of the elementary trees in Fig.~\ref{fig:grammar}.}
 \end{figure}
@@
 However, most readers disprefer the reading in (b).
-Clearly, to use {\sc crisp} with such a grammar we need a method of distinguishing good derivations from bad ones.
+Clearly, to use CRISP with such a grammar we need a method of distinguishing good derivations from bad ones.
 

### 2011-05-23 tags/crisp-1.1.6/trunk/Papers/pcrisp-10/experiments.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2010-05-21 trunk/Papers/pcrisp-10/experiments.tex
@@
 
-We now report on some experiments with \pcrisp.
+We now report on some experiments with PCRISP.
 We are interested in the output quality and runtime behavior of different combinations of planning systems and probability models.
-Like \crisp, \pcrisp\ is an integrated sentence generation system capable of generating referring expressions during surface realization.
-However, for lack of an appropriate evaluation dataset for full sentence generation, we only evaluate \pcrisp\ on a realization task here.
+Like CRISP, PCRISP is an integrated sentence generation system capable of generating referring expressions during surface realization.
+However, for lack of an appropriate evaluation dataset for full sentence generation, we only evaluate PCRISP on a realization task here.
 Further experiments and more details can be found in \newcite{bauer2009}.
@@
 \subsection{Evaluation Data}
-For our experiments we use an \ltag\ grammar and treebank that was automatically extracted from the Wall Street Journal using the algorithm described by \newcite{chenschanker2004}.
+For our experiments we use an LTAG grammar and treebank that was automatically extracted from the Wall Street Journal using the algorithm described by \newcite{chenschanker2004}.
 
 This algorithm outputs a grammar that allows multiple adjunctions at the same node.
-For such a grammar {\sc ptag} is not a suitable probability model.
+For such a grammar PTAG is not a suitable probability model.
 Models that can deal with multiple adjunction are discussed by \newcite{nasrrambow2006}, but employing them would require non-trivial modifications to our encoding as a metric planning problem.
@@
 We use section 0 as a testing set.
-However, since the number of \pcrisp\ operators grows quadratically with the grammar size, generating long sentences requires too much time to run batch experiments.
+However, since the number of PCRISP operators grows quadratically with the grammar size, generating long sentences requires too much time to run batch experiments.
 We therefore restrict our evaluation to the 416 sentences in Section 0 that are shorter than 16 words.
@@
 Edit distance is modified to treat insertion-deletion pairs as single movement errors.
-Compared to a purely string-based metric like {\sc bleu}, GTA penalizes swapped words less harshly if they can be explained by local tree movements.
+Compared to a purely string-based metric like BLEU, GTA penalizes swapped words less harshly if they can be explained by local tree movements.
 
@@
 Table \ref{results} presents the results of the experiment for five different generation systems.
-We compare three variants of \pcrisp: the fully lexicalized PTAG model (``PTAG''), the fully unlexicalized model (``unlexicalized''), and a linear interpolation model (``interpolation'') in which we (manually) set the weight of level 2 to 0.9 and the weight of level 3 to 0.
-We also list results for the non-statistical \crisp\ system of  \newcite{kollerstone2007} (``\crisp'') and the greedy search heuristic used by \spud.
-All these systems are based on a reimplementation of the FF planner \cite{hoffmannnebel2001}, to which we added a search heuristic that takes action costs into account for the \pcrisp\ systems.
+We compare three variants of PCRISP: the fully lexicalized PTAG model (``PTAG''), the fully unlexicalized model (``unlexicalized''), and a linear interpolation model (``interpolation'') in which we (manually) set the weight of level 2 to 0.9 and the weight of level 3 to 0.
+We also list results for the non-statistical CRISP system of  \newcite{kollerstone2007} (``CRISP'') and the greedy search heuristic used by SPUD.
+All these systems are based on a reimplementation of the FF planner \cite{hoffmannnebel2001}, to which we added a search heuristic that takes action costs into account for the PCRISP systems.
 
@@
 
-The table confirms that the non-statistical \crisp\ system has considerable trouble reconstructing the original sentences from the treebank, with a mean GTA of 0.66.
-This is still better than our reimplementation of \spud, which fails to recover from early mistakes of its greedy search heuristic for every single sentence in the test set.
+The table confirms that the non-statistical CRISP system has considerable trouble reconstructing the original sentences from the treebank, with a mean GTA of 0.66.
+This is still better than our reimplementation of SPUD, which fails to recover from early mistakes of its greedy search heuristic for every single sentence in the test set.
 
-By contrast, the fully lexicalized \pcrisp\ model achieves a much better mean GTA.
+By contrast, the fully lexicalized PCRISP model achieves a much better mean GTA.
 However, this comes at the cost of a very low success rate of only 10\%, reflecting a serious data sparseness problem on unseen inputs.
-The data sparseness problem is reduced in the unlexicalized version of \pcrisp, but this comes at the cost of decreased accuracy and much increased runtimes.
+The data sparseness problem is reduced in the unlexicalized version of PCRISP but this comes at the cost of decreased accuracy and much increased runtimes.
 The linear interpolation model strikes a balance between these two, by improving the success rate over the lexicalized model while sacrificing only a small amount of accuracy.
@@
      {\bf System }  & {\bf gta}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
-     \spud         & n/a & 0\% & 100\% & 0\%  \\ \hline 
-     \crisp         &  0.66 & 45\% & 42\% & 13\% \\ \hline
+     SPUD         & n/a & 0\% & 100\% & 0\%  \\ \hline 
+     CRISP         &  0.66 & 45\% & 42\% & 13\% \\ \hline
      PTAG &  0.90 &  10\% &88\% & 2\% \\ \hline

### 2011-05-23 tags/crisp-1.1.6/trunk/Papers/pcrisp-10/related.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2010-05-21 trunk/Papers/pcrisp-10/related.tex
@@
 Statistical methods are popular for surface realization, but have not been used in systems that integrate sentence planning.
-Most statistical generation approaches follow a generate-and-select strategy, first proposed by \newcite{knight1995} in their {\sc nitrogen} system.
+Most statistical generation approaches follow a generate-and-select strategy, first proposed by \newcite{knight1995} in their NITROGEN system.
 Such systems generate a set of candidate sentences using a (possibly overgenerating) grammar and then select the best output sentence by applying a statistical language model.
-This family includes systems such as {\sc HAL}ogen \cite{langkildeknight1998,langkilde2000} and OpenCCG \cite{whitebaldridge2003}.
-The {\sc Fergus} system \cite{bangalorerambow2000} is a variant of this approach which, like {\sc pcrisp}, employs {\sc ltag}.
-It first assigns elementary trees to each entity in the input sentence plan using a statistical tree model and then computes the most likely derivation using using only these trees with an $n$-gram model on the output sentence.
-An alternative to the generate and select approach is to use a probabilistic grammar model, like {\sc   ptag}, trained on automatic parses \cite{zhongstent2005}.
+This family includes systems such as HALogen \cite{langkildeknight1998,langkilde2000} and OpenCCG \cite{whitebaldridge2003}.
+The FERGUS system \cite{bangalorerambow2000} is a variant of this approach which, like PCRISP, employs TAG.
+It first assigns elementary trees to each entity in the input sentence plan using a statistical tree model and then computes the most likely derivation using only these trees with an n-gram model on the output sentence.
+An alternative to the n-gram based generate and select approach is to use a probabilistic grammar model, like PTAG, trained on automatic parses \cite{zhongstent2005}.
 A related approach uses a model over local decisions of the generation system itself \cite{belz2008}.
 Both models can either be used to discriminate a set of output candidates, or more directly to choose the next best decision locally.
-Our approach is similar in that it uses {\sc ptag} to find the most likely output structure.
+Our approach is similar in that it uses PTAG to find the most likely output structure.
 However, the previous work discussed so far addresses surface realization only.
@@
 
-Our treatment of integrated sentence planning and surface realization as planning, is inherited directly from {\sc crisp} \cite{kollerstone2007}.
+Our treatment of integrated sentence planning and surface realization as planning is inherited directly from CRISP \cite{kollerstone2007}.
 Planning has long played a role in generation, but has focused on discourse planning instead of specifically addressing sentence generation \cite{hovy1988,appelt1992}.
@@
 Recently the development of more efficient planning algorithms \cite{hoffmannnebel2001} spawned a renewed interest in planning for NLG.
-{\sc Crisp} uses such algorithms to efficiently solve the sentence generation problem defined by {\sc spud} \cite{Stone2003a}.
-{\sc Spud}, which instead used an incomplete greedy algorithm, is based on an {\sc ltag} whose trees are augmented with semantic and pragmatic constraints.
-Given a communicative goal, a solution to the {\sc spud} problem realizes this goal and simultaneously selects referring expressions.
-The next section explores {\sc crisp} in more detail.
+CRISP uses such algorithms to efficiently solve the sentence generation problem defined by SPUD \cite{Stone2003a}.
+SPUD, which instead uses an incomplete greedy algorithm, is based on a TAG  whose trees are augmented with semantic and pragmatic constraints.
+Given a communicative goal, a solution to the SPUD problem realizes this goal and simultaneously selects referring expressions.
+The next section explores CRISP in more detail.
 

### 2011-05-23 tags/crisp-1.1.6/trunk/Papers/rabbit-from-hat/problem.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

\section{Referring succinctly}
\label{sec:problem}

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{pic-rabbits}
  \caption{``Take the rabbit from the hat'' (from \cite{Stone1998a}).}
  \label{fig:rabbits}
\end{figure}

\subsection{Coordinated referring expressions}

Consider the scene in Fig.~\ref{fig:rabbits}, and let's say we want to
instruct the hearer to remove the indicated rabbit, $r_1$, from the
indicated hat, $h_1$.  In the absence of further attributes that
distinguish $r_1$ from the other rabbits and $h_1$ from the other
hats, the best unique referring expression for $r_1$ is ``the rabbit
in the hat'' and the best unique RE for $h_1$ is ``the hat that
contains a rabbit''.  An NLG system which first computes the referring
expressions and then performs realization in a separate, subsequent
step would thus generate the sentence ``remove the rabbit in the hat
from the hat that contains a rabbit''.

\begin{figure}
  \centering

  \caption{A small SPUD-style lexicon.}
  \label{fig:spud-lexicon}
\end{figure}

\newcite{Stone1998a} showed that by coupling sentence planning and
realization more closely, the SPUD system is able to generate an
instruction that uses more succinct referring expressions: It is
sufficient to say ``remove the rabbit from the hat''.  There are two
aspects of SPUD that conspire to make this expression possible.
First, the lexicon entry for ``take X from Y'' in SPUD (shown in a
simplified version modelled after \newcite{KolSto07} in
Fig.~\ref{fig:spud-lexicon}) includes the ``semantic requirement''
that X must be in Y for the utterance to be felicitous.  Second, this
information is then used to coordinate the meanings of the two
referring expressions by means of a constraint network
\cite{dale91:_gener_refer_expres_invol_relat}.  In the example, the
semantic requirement of the verb would contribute the information
$in(x,y)$; the first NP would contribute the information $rabbit(x)$;
and the second NP would contribute $hat(y)$.  Taken together,
$\{in(x,y), rabbit(x), hat(y)\}$ is only satisfied by a single tuple
$(r_1,h_1)$, and therefore the reference is unique -- the two NPs can
restrict each other's extensions, mediated by the verb.

\subsection{A first planning approach}

Now let's talk about how this interaction between the two REs can be
modelled in a planning framework.  From the perspective of the
planning problem presented in Section~\ref{sec:gen-as-planning}, the
action instance for the verb's elementary tree instantiates the two
new REs with sets of distractors.  The purpose of each RE is then to
eliminate all distractors so a goal state can be reached.

\newcite{KolSto07} already go one step beyond what a pipelined model
could do by choosing all objects except for $r_1$ that are inside
something else (in the example, $\{r_2,f_2\}$) and all objects except
for $h_1$ that contains something ($\{h_3,b_1\}$) as the initial
distractor sets for the two NPs.  This exploits the semantic
requirement of the lexicon entry partially.  In this particular
example, it is not sufficient to generate ``the rabbit'' and ``the
hat'', because $r_2$ is also a rabbit and $h_3$ is also a hat.

However, this particular choice of distractor sets is not the only one
which, together with the information that X is in Y, would make the
joint reference unique.  For instance, if we can establish that the
first NP does not refer to $f_2$ and the second NP does not refer to
$b_1$, then we have again referred to $(r_1,h_1)$ uniquely.  More
generally, if we can find sets $A$ and $B$ such that 
$$\{(x,y) \mid x \notin A \;\mbox{and}\; y \notin B \;\mbox{and}\; in(x,y) \}
= \{(r_1,h_1)\},$$ 
then we will obtain a unique joint referring expression if we can
eliminate the distractors $A$ in the first NP and the distractors $B$
in the second NP.  The workload of establishing unique reference can
be shifted between $A$ and $B$: The more elements $A$ contains, the
more informative the first RE will have to be; but this will also
allow $B$ to be smaller, and so the second RE can be shorter.

Unfortunately, the computation of suitable sets $A$ and $B$ is beyond
the expressive power of a typical planning problem.  We can get around
this limitation by computing all possible (minimal) values for $A$ and
$B$ beforehand and encoding them in the planning problem.  To encode
the four minimal distractor sets for $r_1$ and $h_1$ in the example,
we can add the following conjunction to the initial state of the
planning problem:

$$
\begin{array}{l}
  \mathsf{distractorset}(s_1,\mathsf{in},r_1,h_1) \wedge
  \mathsf{distractor}_L(s_1,f_1) \wedge \\ \mathsf{distractor}_L(s_1,f_2)
  \wedge \mathsf{distractor}_L(s_1,r_3)  \wedge \\
%
  \mathsf{distractorset}(s_2,\mathsf{in},r_1,h_1) \wedge
  \mathsf{distractor}_L(s_2,f_2) \wedge \\ \mathsf{distractor}_R(s_2,h_3)
  \wedge \mathsf{distractor}_R(s_2,b_1) \wedge \\
%
  \mathsf{distractorset}(s_3,\mathsf{in},r_1,h_1) \wedge
  \mathsf{distractor}_L(s_3,f_1) \wedge \\ \mathsf{distractor}_L(s_3,f_2)
  \wedge \mathsf{distractor}_R(s_3,b_1) \wedge \\
%
  \mathsf{distractorset}(s_4,\mathsf{in},r_1,h_1) \wedge
  \mathsf{distractor}_L(s_4,r_3) \wedge \\ \mathsf{distractor}_L(s_4,f_2)
  \wedge \mathsf{distractor}_R(s_4,h_3) 
\end{array}
$$

These sets can then be picked up by the planning operators.  Operators
now receive an additional parameter for the choice of distractor sets,
and additional preconditions and effects as follows:

\action{S-takefrom-1}{u,x_1,x_2,x_3,s}
{\mathsf{distractorset}(s,\mathsf{in},x_2,x_3) \wedge \ldots}
{\forall y. \mathsf{distractor}_L(s,y) \rightarrow
  \mathsf{distractor}(\mathsf{pat}(u),y)
\wedge \\
\forall y. \mathsf{distractor}_R(s,y) \rightarrow
  \mathsf{distractor}(\mathsf{src}(u),y)
\wedge \ldots
}

We have implemented this conversion an evaluated the performance of
the FF planner \cite{HoffmannNebel01} on it.  FF is widely accepted as
an extremely efficient planner.  Unfortunately, it performs very
poorly on these examples.  For the above example, FF takes 22 seconds
to compute the sentence ``take the rabbit from the
hat''.\footnote{\todo{Measure runtimes on the same machine and say
    here what it is.}}  As the domain size increases, so does FF's
runtime: On a domain with four rabbits, four hats, three flowers, and
three bathtubs, plus random ``in'' relations between these, FF took
about 150 seconds on average.  For three rabbits and hats and four
flowers and bathtubs, FF ran out of memory before it was able to find
a plan.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "rabbit-from-hat"
%%% End:

### 2011-05-23 tags/crisp-1.1.6/trunk/Papers/reasoning-challenges/abstract.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2009-12-13 trunk/Papers/reasoning-challenges/abstract.tex
@@
 planners and problems. Researchers would have access to a central
-repository with the latest domains and performance statistics, running on a
-common hardware platform, plus a facility for related activities such as
-posting challenges and demonstrating new technologies. Since much of the
-proposed infrastructure is not tied to a specific research area, this
+repository with the latest domains and performance data, running on a
+common hardware platform, plus a social networking forum for posting
+challenges, announcing new technologies, and discussion. Since much of
+the proposed infrastructure is not tied to a specific research area, this
 approach offers the possibility of joint ventures with other research

### 2011-05-23 tags/crisp-1.1.6/trunk/Papers/reasoning-challenges/introduction.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2009-12-19 trunk/Papers/reasoning-challenges/introduction.tex
@@
 run, each of which has helped drive the development of new planning
-technology, by providing a regular forum for showcasing the latest trends
+technology by providing a regular forum for showcasing the latest trends
 in planner design, domain description languages, and challenging planning
@@
 
-From a technical point of view, it is important to note that such an idea
-would not have been practical ten years ago during the time of the first
-IPC. It is only recently, due to advances in computing hardware and network
-bandwidth, that a scenario such as this is now feasible. Even so, our
-proposed infrastructure nevertheless presents some technical challenges
-that must be addressed (e.g., security, accessibility, maintenance,
+From a technical point of view, it is important to note that such an
+idea would not have been practical ten years ago during the time of
+the first IPC. It is only recently, due to advances in computing
+hardware, network bandwidth, and available software infrastructure,
+that a scenario such as this is now feasible. Even so, our proposed
+infrastructure nevertheless presents some technical challenges that
+must be addressed (e.g., security, accessibility, maintenance,
 etc.). However, we believe that for the most part, solutions to these
@@
 %the planning community and beyond.
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2011-05-23 tags/crisp-1.1.6/trunk/Papers/reasoning-challenges/levels.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2009-12-19 trunk/Papers/reasoning-challenges/levels.tex
@@
 While this format for the IPC has been highly successful in the past
-ten years, there is also a number of ways in which it can be improved
+ten years, there are also a number of ways in which it can be improved
 and streamlined.  One such direction concerns the way in which problem
@@
 that must be tested in each round of the competition manageable, older
-domains are sometimes ``retired'', i.e.\ removed from subsequent
+domains are sometimes ``retired'', i.e., removed from subsequent
 competitions.  This comes with a risk of a false sense of progress, in
@@
 which determines what systems the data is suitable for, i.e., the language
-in which the problem is specified (such as a particular dialect of PDDL or
-EUROPA \todo{cite}) and the type of planning problem (e.g., propositional,
-temporal, uncertainty, etc.).  If it is possible to convert between one
+in which the problem is specified (such as a particular dialect of PDDL)
+and the type of planning problem (e.g., propositional,
+temporal, uncertainty).  If it is possible to convert between one
 specification language and another automatically, this conversion could be
@@
 VirtualBox,\footnote{\url{http://www.virtualbox.org/}} and ensure
-that their system works, in the exact environment in which it will be
+that their system works in the exact environment in which it will be
 evaluated.  This would even make it technically possible for
@@
 when they download the systems.  Even when the environment changes to
-reflect updates in the standard development environments (new version
-of the C compiler, etc.), old images could remain available, and so
+reflect updates in the standard development environments (e.g., a new version
+of the C compiler), old images could remain available, and so
 old systems would continue running until the virtual machine emulator
@@
 
-We propose that every newly submitted system is run automatically on
+We propose that every newly submitted system be run automatically on
 all available problem instances, and every newly submitted problem
-instance is sent to all available systems.  The scheduling of this
+instance be sent to all available systems.  The scheduling of this
 process can be under the control of an automated system for continuous
 integration, such as Hudson.\footnote{\url{http://hudson-ci.org/}} The
-evaluation infrastructure sends the problem data and the system to the
-test machine, runs the system, measures performance statistics
+evaluation infrastructure would send the problem data and the system to the
+test machine, run the system, measure performance statistics
 (including which problems they could solve and how much time this took
-them; see below), and stores the results in a database. Users will be
+them; see below), and store the results in a database. Users would be
 able to view the results in dynamically generated reports on the main
@@
 In order to minimize security risks, which can arise when
-user-submitted programs are being run automatically, the test machines
+user-submitted programs are run automatically, the test machines
 need to be sandboxed quite tightly.  In particular, the machines do
@@
 able to enter performance measurements into a database, and can work
-on virtual file systems that are deleted after each system.  By
+on virtual file systems that are deleted after each execution.  By
 further imposing resource limits on the number of available
@@
 potential for abuse is limited.  It can be limited further by only
-allowing systems uploads to registered users.
+allowing system uploads to registered users.
 
@@
 classes. The selection of this subset could be informed, among other
-things, by the report on problems that many systems can't
-solve. Participants could then upload their systems at their leisure;
-``running'' the competition amounts to taking a snapshot of the
+things, by the report on problems that many systems had difficulty
+solving. Participants could then upload their systems at their leisure;
+``running'' the competition would amount to taking a snapshot of the
 performance reports on the selected problems when the deadline comes
 and announcing the results.  By recording timestamps whenever a
-performance measurements is made, this snapshot could remain available
+performance measurement is made, this snapshot could remain available
 on the website simply as one special report for a certain point in
@@
 
-The continuous evaluation infrastructure we just presented is
+The continuous evaluation infrastructure we have presented is
 fundamentally agnostic to the type of concrete evaluation measures
@@
 Many reasoning tasks, including constraint solving and theorem
-proving, are being evaluated by running solvers over problem instances
+proving, are evaluated by running solvers over problem instances
 and reporting performance measures.  At the level of abstraction at

### 2011-05-23 tags/crisp-1.1.6/trunk/Papers/rmrs-08/introduction.tex
[maven-release-plugin]  copy for tag crisp-1.1.6

--- 2008-10-10 trunk/Papers/rmrs-08/introduction.tex
@@
 English Resource Grammar ({\sc erg}, \cite{copestake:flickinger:2000})
-are special cases of {\sc rmrs}.
+are special cases of {\sc rmrs}.  But {\sc rmrs}, unlike {\sc mrs} and
+related formalisms like dominance constraints \cite{egg:etal:2001}, is
+able to express semantic information in the absence of full predicate
+argument structure and lexical subcategorisation.
 
@@
 We finish in Section~\ref{sec:entailment} by pointing out some avenues
-for future research, and conclude in Section~\ref{sec:conclusion}.
+for future research. %, and conclude in Section~\ref{sec:conclusion}.
 

### 2011-05-23 tags/crisp-1.1.6/trunk/Code/crisp/src/main/java/crisp/converter/CRISPtoPDDL.java

@@ -1,0 +1,63 @@
+package crisp.converter;
+
+import crisp.planningproblem.Domain;
+import crisp.planningproblem.Problem;
+import crisp.planningproblem.codec.PddlOutputCodec;
+
+
+
+import de.saar.penguin.tag.grammar.CrispGrammar;
+import de.saar.penguin.tag.grammar.SituatedCrispXmlInputCodec;
+import java.io.File;
+import java.io.FileReader;
+
+public class CRISPtoPDDL {
+
+    /**
+     * Print a usage message
+     */
+    public static void usage() {
+        System.out.println("Usage: java crisp.converter.CRISPtoPPDL [CRISP grammar] [CRISP problem]");
+    }
+
+    /**
+     * Main program.  When running the converter from the command line, pass
+     * the name of the CRISP problem file as the first argument.
+     *
+     * @param args
+     * @throws Exception
+     */
+    public static void main(String[] args) throws Exception {
+
+        if (args.length < 1) {
+            System.err.println("No crisp problem specified");
+            usage();
+            System.exit(1);
+        }
+
+        // TODO some exception handling
+
+        Domain domain = new Domain();
+        Problem problem = new Problem();
+
+        long start = System.currentTimeMillis();
+
+        SituatedCrispXmlInputCodec codec = new SituatedCrispXmlInputCodec();
+        CrispGrammar grammar = new CrispGrammar();
+        codec.parse(new FileReader(new File(args[0])), grammar);
+
+        File problemfile = new File(args[1]);
+
+        CurrentNextCrispConverter converter = new CurrentNextCrispConverter();
+        converter.convert(grammar, problemfile, domain, problem);
+        long end = System.currentTimeMillis();
+
+        System.out.println("Total runtime: " + (end - start) + "ms");
+
+        System.out.println("Domain: " + domain);
+        System.out.println("Problem: " + problem);
+
+        new PddlOutputCodec().writeToDisk(domain, problem, "");
+
+    }
+}

### 2011-05-23 tags/crisp-1.1.6/trunk/Code/crisp/src/main/java/crisp/converter/CurrentNextCrispConverter.java

@@ -1,0 +1,1172 @@
+package crisp.converter;
+
+import java.io.File;
+import java.io.FileReader;
+import java.io.IOException;
+import java.io.Reader;
+import java.io.StringWriter;
+import java.util.ArrayList;
+import java.util.Collection;
+import java.util.HashMap;
+import java.util.HashSet;
+import java.util.List;
+import java.util.Map;
+import java.util.Set;
+import java.util.Stack;
+
+import javax.xml.parsers.ParserConfigurationException;
+import javax.xml.parsers.SAXParser;
+import javax.xml.parsers.SAXParserFactory;
+
+import org.xml.sax.Attributes;
+import org.xml.sax.InputSource;
+import org.xml.sax.SAXException;
+import org.xml.sax.SAXParseException;
+import org.xml.sax.helpers.DefaultHandler;
+
+import crisp.planningproblem.Action;
+import crisp.planningproblem.Domain;
+import crisp.planningproblem.Problem;
+import crisp.planningproblem.formula.Conjunction;
+import crisp.planningproblem.formula.Formula;
+import crisp.planningproblem.formula.Literal;
+import crisp.planningproblem.formula.Negation;
+import crisp.planningproblem.formula.Universal;
+import crisp.planningproblem.formula.Conditional;
+import de.saar.chorus.term.Compound;
+import de.saar.chorus.term.Constant;
+import de.saar.chorus.term.Substitution;
+import de.saar.chorus.term.Term;
+import de.saar.chorus.term.Variable;
+import de.saar.chorus.term.parser.TermParser;
+import de.saar.penguin.tag.grammar.Constraint;
+import de.saar.penguin.tag.grammar.CrispGrammar;
+import de.saar.penguin.tag.grammar.CrispLexiconEntry;
+import de.saar.penguin.tag.grammar.ElementaryTree;
+import de.saar.penguin.tag.grammar.ElementaryTreeType;
+import de.saar.penguin.tag.grammar.NodeType;
+
+/**
+ * This class provides a fast converter from XML CRISP problem descriptions to
+ * planning domains and problems. This class only processes non-probabilistic grammar descriptions.
+ * The parser is implemented as a plain SAX parser, thereby improving
+ * processing speed and decreasing memory requirements over the old xpath based
+ * parser in CRISPConverter, which is particularly important for parsing large grammar
+ * description files.
+ */
+public class CurrentNextCrispConverter {
+
[…]

### 2011-05-23 tags/crisp-1.1.6/trunk/Code/crisp/src/main/java/crisp/main/BasicCrispGenerator.java

@@ -1,0 +1,216 @@
+package crisp.main;
+
+import crisp.converter.CurrentNextCrispConverter;
+import crisp.planner.external.FfPlannerInterface;
+import crisp.planner.external.PlannerInterface;
+import crisp.planningproblem.Domain;
+import crisp.planningproblem.Problem;
+import crisp.result.CrispDerivationTreeBuilder;
+import crisp.result.DerivationTreeBuilder;
+import de.saar.chorus.term.Term;
+import de.saar.penguin.tag.codec.ParserException;
+import de.saar.penguin.tag.derivation.DerivationTree;
+import de.saar.penguin.tag.derivation.DerivedTree;
+import de.saar.penguin.tag.grammar.CrispGrammar;
+import de.saar.penguin.tag.grammar.Grammar;
+import de.saar.penguin.tag.grammar.SituatedCrispXmlInputCodec;
+
+import java.io.*;
+import java.util.ArrayList;
+import java.util.List;
+
+/**
+ * A simple implementation of the @{CRISPGeneratorInterface} using the current state
+ * of the standard CRISP generation system (with "steps" andwithout pragmatics or
+ * context or probabilistic grammars). It also creates a new PDDL domain for every
+ * problem instance and thus does not benefit from Konstantina's optimizations.
+ * <p/>
+ * Please don't use this class any more. It relies on older versions of CRISP,
+ * and is no longer being maintained. Use GenerateSentence instead.
+ *
+ * @author dbauer
+ */
+@Deprecated
+public class BasicCrispGenerator implements CrispGeneratorInterface {
+    CrispGrammar grammar = null;
+    CurrentNextCrispConverter converter;
+    PlannerInterface planner;
+    DerivationTreeBuilder planDecoder = null;
+    private List<Term> plan = new ArrayList<Term>();
+    private Domain domain;
+    private Problem problem;
+
+    public BasicCrispGenerator(String ffFlags) {
+        converter = new CurrentNextCrispConverter();
+        planner = new FfPlannerInterface(ffFlags);
+    }
+
+    public BasicCrispGenerator() {
+        this("");
+    }
+
+    // *** Methods to load the grammar ***
+
+    public void setGrammar(Grammar grammar) {
+        this.grammar = (CrispGrammar) grammar;
+        planDecoder = new CrispDerivationTreeBuilder(grammar);
+    }
+
+    private void setGrammar(CrispGrammar grammar) {
+        this.grammar = grammar;
[…]

### 2011-05-23 tags/crisp-1.1.6/trunk/Code/crisp/src/test/java/crisp/converter/CurrentNextCrispConverterTest.java

@@ -1,0 +1,185 @@
+package crisp.converter;
+
+import static org.junit.Assert.*;
+
+import java.io.*;
+import java.util.ArrayList;
+import java.util.List;
+
+import javax.xml.parsers.ParserConfigurationException;
+
+import crisp.planner.external.FfPlannerInterface;
+import crisp.planner.external.PlannerInterface;
+import crisp.result.CrispDerivationTreeBuilder;
+import crisp.result.DerivationTreeBuilder;
+import de.saar.chorus.term.Term;
+import de.saar.penguin.tag.derivation.DerivationTree;
+import de.saar.penguin.tag.derivation.DerivedTree;
+import org.junit.Test;
+import org.xml.sax.SAXException;
+
+import crisp.planningproblem.Action;
+import crisp.planningproblem.Domain;
+import crisp.planningproblem.Problem;
+import de.saar.penguin.tag.codec.ParserException;
+import de.saar.penguin.tag.grammar.CrispGrammar;
+import de.saar.penguin.tag.grammar.SituatedCrispXmlInputCodec;
+
+public class CurrentNextCrispConverterTest {
+
+    private Domain domain;
+    private Problem problem;
+    private SituatedCrispXmlInputCodec codec = new SituatedCrispXmlInputCodec();
+    private DerivationTreeBuilder planDecoder;
+    private CrispGrammar grammar;
+
+    /*
+    @Before
+    public void setUp() throws FileNotFoundException, ParserException, IOException, ParserConfigurationException, SAXException {
+        CurrentNextCrispConverter converter;
+        domain = new Domain();
+        problem = new Problem();
+
+        grammar = new CrispGrammar();
+        codec.parse(new InputStreamReader(getClass().getResourceAsStream("/grammar-scrisp.xml")), grammar);
+
+        Reader problemfile = new InputStreamReader(getClass().getResourceAsStream("/problem-scrisp-give1-generated.xml"));
+
+        converter = new CurrentNextCrispConverter();
+        converter.convert(grammar, problemfile, domain, problem);
+
+        planDecoder = new CrispDerivationTreeBuilder(grammar);
+    }
+     * 
+     */
+
+    private void loadAndConvert(String grammarResourceName, String problemResourceName) throws ParserException, IOException, ParserConfigurationException, SAXException {
+        CurrentNextCrispConverter converter;
+        domain = new Domain();
+        problem = new Problem();
+
+        grammar = new CrispGrammar();
[…]
